### 2018-10-25 inst/table_tennis_serve.Rmd
add table tennis serve analysis

---
output:
  pdf_document: default
  html_document: default
---
# The importance of serve in table tennis.

Suppose We have three players, "a","b", and "c" who play table tennis.
We seek to quantify the effect of the serve.  The scoreline is as
follows:

* A vs B, A serves, 5-1
* A vs B, B serves, 1-3
* A vs C, A serves, 5-1
* A vs C, C serves, 1-3

This is easily analysed with the ```hyper2``` package.

```{r}
library("hyper2",quietly=TRUE)
```

The first step is to define a ```hyper2``` object:

```{r}
H <- hyper2(pnames=c("a","b","c","S"))
```

Now we have to input the data:

### A vs B, A serves, score 5-1:
```{r}
H[c("a","S")]     %<>% "+"(5  ) # a+S win 5 games
H["b"]            %<>% "+"(  1) # b wins 1 game
H[c("a","b","S")] %<>% "-"(5+1) # a+b+S play 1+5=6 games
```

### A vs B, B serves, score 1-3:
```{r}
H[c("b","S")]     %<>% "+"(3  ) # b+S win 3 games
H["a"]            %<>% "+"(  1) # a wins 1 game 
H[c("a","b","S")] %<>% "-"(3+1) # a+b+S play 1+3=4 games
```

### A vs C, A serves, score 4-1:
```{r}
H[c("a","S")]     %<>% "+"(4  ) # a+S win 4 games
H["c"]            %<>% "+"(  1) # c wins 1 game 
H[c("a","c","S")] %<>% "-"(4+1) # a+c+S play 1+4=5 games
```

### A vs C, C serves, score 1-2:
```{r}
H["a"]            %<>% "+"(1  ) # a wins 1 game
H[c("c","S")]     %<>% "+"(  2) # c+S win 2 games
H[c("a","c","S")] %<>% "-"(1+2) # a+c+S play 1+2=3 games
```

So H is the likelihood function for three players and a serve.
We now find the unconstrained MLE:

```{r}
p1 <- maxp(H)
p1
```

And we can find the support at this point:

```{r}
maxlike_free <- loglik(H,indep(p1))
maxlike_free
```

Now find the MLE conditional on the serve having zero strength:

```{r}
small <- 1e-6
maxlike_constrained <-
    constrOptim(
        f = function(p){loglik(H,p)},
        theta=rep(1/3,3)-small,
        grad=NULL,
        ui=rbind(diag(3),-diag(3),1,-1),
        ci=c(rep(small,3),rep(small,3)-1, 1-4*small, small*2-1),
        control=list(fnscale = -1)  # finds max, not min
    )
```

The difference in support is:

```{r}    	    
delta_support <- maxlike_free - maxlike_constrained$value
delta_support
```

and the likelihood ratio
```{r}
LR <- exp(delta_support)
LR
```

On the null, Wilks's theorem says that $\Lambda\sim\chi^2_1$, where
$\Lambda=2\log(L_1/L_2)$, so the
p-value is

```{r}
Lambda <- 2*delta_support
pval <- pchisq(Lambda,df=1,lower.tail=FALSE)
pval 
```

The difference in support, ```delta_support```, is about 2.62 which
exceeds Edwards's two-units-of-support criterion.  The likelihood
ratio LR is about 13.8, and Wilks's theorem gives us that
$\Lambda=2\log(LR)$ is distributed as chi-squared with one degree of
freedom; the tail area, pval, is about 0.022 which gives us a
significant result.

### 2018-10-25 inst/table_tennis_serve.Rmd
change docs to match actual scores used

@@
 * A vs B, B serves, 1-3
-* A vs C, A serves, 5-1
-* A vs C, C serves, 1-3
+* A vs C, A serves, 4-1
+* A vs C, C serves, 1-2
 
@@
 
-and the likelihood ratio
+and the likelihood ratio is:
 ```{r}

### 2018-10-25 inst/table_tennis_serve.Rmd
Change the order of the players from a,b,c,S to S,a,b,c.
The previous ordering, a,b,c,S, was unnecessarily difficult because
the additional constraint (additional in the sense of additional over
and above "0 <= p_A, p_B, p_C, p_S<1" and "p_A + p_B + p_C <= 1") was
on the fillup value p_S, so "p_S<=small" was forced to appear as
"p_A+p_B+p_C>=1-small", which was confusing as the constraint on p_S
was written in terms of the other strengths.

Now we have S,a,b,c this is easier.  We have the additional constraint
appearing as the much simpler form "p_S<small".

Actually, now that I write it like

@@
 ```{r}
-H <- hyper2(pnames=c("a","b","c","S"))
+H <- hyper2(pnames=c("S","a","b","c"))
 ```
@@
         f = function(p){loglik(H,p)},
-        theta=rep(1/3,3)-small,
+        theta=c(small, 1/3-small/2, 1/3-small/2),
         grad=NULL,
-        ui=rbind(diag(3),-diag(3),1,-1),
-        ci=c(rep(small,3),rep(small,3)-1, 1-4*small, small*2-1),
+        ui=rbind(diag(3),-diag(3),-1),
+        ci=c(rep(small/2,3),c(-small*2,-1,-1),-1),
         control=list(fnscale = -1)  # finds max, not min

### 2018-11-07 inst/table_tennis_serve.Rmd
more work on interpreting the MLE

@@
 
-This is easily analysed with the ```hyper2``` package.
+(note that A does not play C).  This dataset is easily analysed with
+the ```hyper2``` package.
 
@@
 So H is the likelihood function for three players and a serve.
+
+# Results 
 We now find the unconstrained MLE:
@@
 significant result.
+
+Further, note that our estimated strengths as obtained by MLE allow us
+to assess what might happen when A plays C (which was not observed).
+If A serves, we would have $A+S:C\simeq 0.24+0.45:0.16\equiv 69:16$,
+or A wins with probability of about $0.81$.  If C serves we have
+$A:C+S\simeq 0.24:0.16+0.45\equiv 24:61$, so A wins with a probability
+of about $0.28$.

### 2018-11-07 inst/table_tennis_serve.Rmd
slight rewording for clarity

@@
 exceeds Edwards's two-units-of-support criterion.  The likelihood
-ratio LR is about 13.8, and Wilks's theorem gives us that
+ratio LR is about 13.7, and Wilks's theorem gives us that
 $\Lambda=2\log(LR)$ is distributed as chi-squared with one degree of
@@
 
-Further, note that our estimated strengths as obtained by MLE allow us
-to assess what might happen when A plays C (which was not observed).
+Further, note that our maximum likelihood estimate for strengths
+allows us to assess what might happen when A plays C (which was not
+observed).  For convenience, here is the evaluate:
+
+```{r}
+maxp(H)
+```
+
 If A serves, we would have $A+S:C\simeq 0.24+0.45:0.16\equiv 69:16$,

### 2018-11-07 inst/table_tennis_serve.Rmd
rewording

@@
 
-On the null, Wilks's theorem says that $\Lambda\sim\chi^2_1$, where
-$\Lambda=2\log(L_1/L_2)$, so the
-p-value is
+The difference in support, ```delta_support```, is about 2.62 which
+exceeds Edwards's two-units-of-support criterion.  We can reject the
+null.  Alternatively, Wilks's theorem gives us that
+$\Lambda=2\log(LR)$ is distributed as chi-squared with one degree of
+freedom:
 
@@
 
-The difference in support, ```delta_support```, is about 2.62 which
-exceeds Edwards's two-units-of-support criterion.  The likelihood
-ratio LR is about 13.7, and Wilks's theorem gives us that
-$\Lambda=2\log(LR)$ is distributed as chi-squared with one degree of
-freedom; the tail area, pval, is about 0.022 which gives us a
-significant result.
+Thus the tail area, pval, is about 0.022 which gives us a significant
+result.
 

### 2018-11-15 inst/skating_analysis.R
various changes to change the skating analysis into a vignette

@@ -1,58 +1,0 @@
-## This script creates hyper2 object 'skating' which is a likelihood
-## function for the strengths of the competitors in the Ladies' Figure
-## Skating at the 2002 Winter Olympics.  File 'skating.txt' is copied
-## from Lock and Lock (see skating.Rd for more details).
-
-##  Note that file is structured so that each competitor is a row, and
-##  each judge is a column.  Function \code{order_likelihood()}
-##  considers each row to be a race [or a judge]
-
-library("hyper2")
-
-
-## The following R idiom is plausible but incorrect:
-
-
-a <- as.matrix(read.table("skating.txt"))
-skating_incorrect <- order_likelihood(t(a))
-
-## the above is incorrect because we want the ranks, and we have the
-## order.   Correct analysis follows.
-
-b <- apply(a,2,order)
-
-
-## Column 1 of 'a' and column 1 of 'b' are confusingly simular. So
-## consider column 2, corresponding to judge number 2. If aa=a[,2] and
-## bb=b[,2] we have aa=c(4,1,3,2, ...) and bb=c(2,4,3,1, ...).  Thus
-## aa means that hughes was judged to be fourth best, slutskaya was
-## judged to be the best, kwan third best, cohen second best, and so
-## on.  Conversely, bb means that the best competitor was number two
-## [slutskaya], the second best number was number four [cohen], the
-## third best was number three [kwan], the fourth best number 1
-## [hughes], and so on.  Note that this means the rows of b cannot be
-## named.
-
-## The R idiom requires the transpose of the matrix because function
-## order_likelihood() treats the rows as independent observations
-## [here order statistics].
-
-skating <- order_likelihood(t(b))
-pnames(skating) <- rownames(a)
-
-
-
-## Now some data visualization.  First the MLE for the strengths:
-dotchart(maxp(skating))
-
-
-## Now compare my rating with the official point-tallying method:
-dev.new()
-par(pty='s')
-lik <- order(maxp(skating),decreasing=TRUE)
-plot(1:23,lik,asp=1,pch=16,xlab='official order',ylab='likelihood order')
-text(1:23,lik,asp=1,pch=16,pos=c(rep(4,19),rep(2,4)),rownames(a))
-abline(0,1)
-
-
-

### 2018-11-17 inst/oneill.R
explicitly calculate the icons likelihood function in the vignette, and remove the (now) redundant oneill.R

@@ -1,38 +1,0 @@
-## This file creates 'icons' as appearing on page 10 of RKS Hankin
-## 2010, 'A Generalization of the Dirichlet distribution', Journal of
-## Statistical Software, volume 33, issue 11.  Note that this file
-## corrects a typo appearing in the paper.
-
-
-## First specify the matrix as appearing in Hankin 2010
-M <- matrix(c(
-    5 , 3 , NA,   4, NA,   3,
-    3 , NA,  5,   8, NA,   2,
-    NA,  4,  9,   2, NA,   1,
-    10,  3, NA,   3,  4,  NA,
-    4 , NA,  5,   6,  3,  NA,
-    NA,  4,  3,   1,  3,  NA,
-    5 ,  1, NA,  NA,  1,   2,
-    5 , NA,  1,  NA,  1,   1,
-    NA,  9,  7,  NA,  2,   0)
-  , byrow=TRUE,ncol=6)
-colnames(M) <- c("NB","L","PB","THC","OA","WAIS")
-
-## set up the hyper2 object with correct names:
-icons <- hyper2(pnames=colnames(M))
-
-## iterate through each row to work out the denominators:
-for(i in seq_len(nrow(M))){
-  onerow <- M[i,]
-  ## variable 'choices' = icons presented to that row of respondents:
-  choices <- names(onerow[!is.na(onerow)])
-  icons[choices] <-   icons[choices] - sum(onerow,na.rm=TRUE)
-}
-
-## Now iterate through columns to find numerators:
-for(j in seq_len(ncol(M))){
-  icons[colnames(M)[j]] <- icons[colnames(M)[j]] + sum(M[,j],na.rm=TRUE)
-}
-
-
-

### 2019-02-09 R/hyper2.R
more arguments to maxp() allowing easy optimization with further constraints

@@ -364,7 +364,7 @@ setGeneric("pnames<-",function(x,value){standardGeneric("pnames<-")})
   }
 }
 
-`maxp` <- function(H, startp=NULL, give=FALSE, ...){
+`maxp` <- function(H, startp=NULL, give=FALSE, fcm=NULL, fcv=NULL, ...){
     SMALL <- 1e-6
     
     n <- size(H)
@@ -375,8 +375,14 @@ setGeneric("pnames<-",function(x,value){standardGeneric("pnames<-")})
     objective <- function(p){ -loglik(H,p) }
     gradfun   <- function(p){ -(gradient(H,p))} #NB minus signs
     
-    UI <- rbind(diag(nrow=n-1),-1)
-    CI <- c(rep(SMALL,n-1),-1+SMALL)
+    UI <- rbind(
+        diag(nrow=n-1),  # regular: p1 >=0, p2 >= 0, ..., p_{n-1} >= 0
+        -1,              # fillup: p_n = 1-(p1+p2+...+p_{n-1}) >= 0
+        fcm)             # further problem-specific constraints
+    CI <- c(
+        rep(SMALL,n-1),  # regular
+        -1+SMALL,        # fillup
+        fcv)             # further contraint vector
     
     out <- constrOptim(
         theta = startp,

### 2019-04-24 inst/home_advantage.Rmd
brief analysis of home advantage

---
title: "Quantifying the home ground advantage"
author: "Robin K. S. Hankin"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{home_ground_advantage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Here I analyse a small synthetic dataset for home ground advantage.

First load the package:
```{r}
library(hyper2)
```

Now generate the dataset:

```{r}
home_games_won <- matrix(c(
    NA, 16, 12, 11,
    19, NA, 19, 16,
    17, 12, NA, 11,
    11, 12, 12, NA),
    nrow=4,ncol=4,byrow=TRUE)

away_games_won <- matrix(c(
    NA, 05, 02, 02,
    02, NA, 10, 02,
    00, 04, NA, 07,
    02, 06, 04, NA),
    nrow=4,ncol=4,byrow=TRUE)

teams <- LETTERS[1:4]
rownames(home_games_won) <- teams
colnames(home_games_won) <- teams
rownames(away_games_won) <- teams
colnames(away_games_won) <- teams

home_games_won
away_games_won
```

Thus ```home_games_won[1,2] == 16``` means A played at home against B
and won 16 times; ```home_games_won[2,1] == 19``` means B played at
home against A and won 19 times.  Also, ```away_games_won[1,2] == 5```
means A played away against B and won 5 times, and
```away_games_won[2,1] == 2``` means B played away against A and won 2
times.  Alternatively, A played B 16+19+5+2=45 times; A played at home
16+2=18 times and won 16 times and lost 2 times; and B played at home
19+5=24 times and won 19 times and lost 5 times.

We will create a hyper2 object with the teams and home ground advantage term:

```{r}
H <- hyper2(pnames=c(teams,"home"))
```

First include the home games won:

```{r}
for(i in seq_len(nrow(home_games_won))){
  for(j in seq_len(ncol(home_games_won))){
    if(i != j){  # diagonal means a team plays itself, meaningless
      game_winner <- teams[i]
      game_loser  <- teams[j]
      no_of_matches <- home_games_won[i,j] # won with home strength helping
      H[c(game_winner,           "home")] %<>% inc(no_of_matches)  # won with home strength helping
      H[c(game_winner,game_loser,"home")] %<>% dec(no_of_matches)
    }
  }
}
```

Now away games won:

```{r}
for(i in seq_len(nrow(away_games_won))){
  for(j in seq_len(ncol(away_games_won))){
    if(i != j){
      game_winner <- teams[i]
      game_loser  <- teams[j]
      no_of_matches <- away_games_won[i,j] # won with home strength helping
      H[c(game_winner                  )] %<>% inc(no_of_matches)  # won with home ground helping
      H[c(game_winner,game_loser,"home")] %<>% dec(no_of_matches)
    }
  }
}
```

Thus

```{r}
H
```

Visually:

```{r}
dotchart(maxp(H))
```

Now a likelihood ratio for the hypothesis that home strength (or home
advantage) is zero.  First define a function ```f()``` that maximises
the likelihood subject to the home strength being *less than* its
argument:

```{r}
small <- 1e-4
f <- function(home_strength){
  maxp(H,startp=rep((1-small)/4,4),give=TRUE,fcm=t(rep(1,4)),fcv=1-home_strength)$value
}
```

See how the ```fcm``` and ```fcv``` constraints make the sum of
a,b,c,d, greater than 1-home_strength.  Thus 'home' [which is the
fillup value] is constrained to be less than home_strength.  Using
this function, the likelihood ratio test is straightforward:

```{r}
Lambda <- -2*(f(2e-4)-maxp(H,give=TRUE)$value)
Lambda
```

Under the null, Wilks's theorem gives us that Lambda is distributed as
a chi-square with 1 degree of freedom (clearly rejected).  However, it
is always interesting to plot likelihood functions.To plot a profile
likelihood we can constrain the home to be greater than a particular
value:

```{r}
g <- function(home_strength){
  maxp(H,startp=rep(small/4,4),give=TRUE,fcm=-t(rep(1,4)),fcv=home_strength-1)$value
}
```

```{r}
home_strength <- seq(from=0.2,to=0.6,length=30)
support <- pmin(sapply(home_strength,f),sapply(home_strength,g))
plot(home_strength,support-max(support),
     xlab="home strength",ylab="support",main="profile likelihood for home advantage")
abline(h=c(0,-2))
```

The figure give a support interval between about 0.3 and 0.5 for home
ground advantage.

### 2019-04-24 inst/home_advantage.Rmd
typos and clarification

@@
       game_loser  <- teams[j]
-      no_of_matches <- away_games_won[i,j] # won with home strength helping
-      H[c(game_winner                  )] %<>% inc(no_of_matches)  # won with home ground helping
-      H[c(game_winner,game_loser,"home")] %<>% dec(no_of_matches)
+      no_of_matches <- away_games_won[i,j] # won without the benefit of home strength
+      H[c(game_winner                  )] %<>% inc(no_of_matches)  # won without home ground helping...
+      H[c(game_winner,game_loser,"home")] %<>% dec(no_of_matches)  # home strength nevertheless on denominator
     }

### 2019-04-30 inst/collusion_chess_draws.Rmd
start to write this up more formally

---
title: "Collusion at the 1962 chess world championships"
author: "Robin K. S. Hankin"
date: "`r Sys.Date()`"
vignette: |
  %\VignetteIndexEntry{collusion at the 1962 world chess championships}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

First load the package:
```{r}
library("hyper2")
options("width" = 90)
```

Now generate the dataset:

```{r}
jj <- read.table("stockholm1962.txt",header=FALSE)
results <- as.matrix(jj[,-(1:2)])
players <- as.character(jj$V1)
nationality <- as.character(jj$V2)
rownames(results) <- players
colnames(results) <- substr(players,1,1)
results
```

(this taken from Wikipedia
\url{https://en.wikipedia.org/wiki/World_Chess_Championship_1963}).
In this dataset, "0" means a loss, "1" a draw and "2" a win.  Thus
Fisher (row 1) beat Korchnoi (column 4).  As a consistency check we
verify that the results table is antisymmetric:

```{r}
all(t(results) == 2-results)
which(results != 2-t(results),arr.ind=TRUE)
```

Thus Barcza vs Portisch is problematic.  Glossing over this

```{r}
H <- hyper2(pnames=c("draw","collusion",players))

for(i in seq_len(nrow(results)-1)){  # i=row
    for(j in seq(from=i+1,to=ncol(results))){ # j = col
        o <- results[i,j]
        if((nationality[i]=="USSR") & (nationality[j]=="USSR")){
            monster <- "collusion"
        } else {
            monster <- "draw" 
        }
        if(o==2){ # row player wins
            H[players[i]] %<>% inc(1)
        } else if(o==0){  # col player wins
            H[players[j]] %<>% inc(1)
        } else if(o==1){ # game drawn
            H[monster] %<>% inc(1)
        } else {
            stop("this cannot happen")
        }
        H[c(players[i],players[j],monster)] %<>% dec(1)
    }  # j loop closes
} # i loop closes
```
and we have

```{r}
head(H)
dotchart(maxp(H),pch=16)
dotchart(log(maxp(H)),pch=16)
```

which would indicate strongly that collusion is indeed a strong
feature of the tournament.  Now some optimization.  First optimize freely:

```{r}
max_support_free <- maxp(H,give=TRUE)$value
ml_p   <- maxp(H)

s <- indep(equalp(H))
small <- 1e-3
s[1] <- s[1] + small
s[2] <- s[2] - small
```
Perform the constrained optimization:
```{r}
ml_p_constrained <- maxp(H,fcm=c(1,-1,rep(0,22)),fcv=0,startp=s)
max_support_constrained <- loglik(H,indep(ml_p_constrained))
max_support_constrained
```

Now compare the supports:

```{r}
extra_support <- max_support_free - max_support_constrained
extra_support
```

well in excess of Edwards's two-unit criterion.  Alternatively, we may
apply Wilks's theorem and observe that the test statistic is in the
tail region of its asymptotic distribution:

```{r}
pchisq(2*extra_support,df=1,lower.tail=FALSE)
```

### 2019-04-30 inst/collusion_chess_draws.Rmd
more analysis

@@
 ```
+
+# Now Curacao
+
+This is the 5th candidates Curacao 1962 results table:
+
+```{r}
+jj <- read.table("curacao1962.txt",header=FALSE)
+curacao <- as.matrix(jj[,-(1:2)])
+players <- as.character(jj$V1)
+nationality <- as.character(jj$V2)
+rownames(curacao) <- players
+colnames(curacao) <- players
+curacao[is.na(curacao)] <- ""
+curacao <- noquote(curacao)
+curacao
+```
+
+Now convert to a likelihood function:
+```{r}
+Hcur <- hyper2(pnames=c("draw","collusion",players))
+for(i in seq_len(nrow(curacao)-1)){  # i=row
+  for(j in seq(from=i+1,to=ncol(curacao))){ # j = col
+    r <- strsplit(curacao[i,j],'')[[1]]
+    if((nationality[i]=="USSR") & (nationality[j]=="USSR")){
+      monster <- "collusion"
+    } else {
+      monster <- "draw"
+    }
+    if(any(r=="W")){  # row player wins
+      wins <- sum(r=="W")
+      Hcur[  players[i]                    ] %<>% inc(wins)
+      Hcur[c(players[i],players[j],monster)] %<>% dec(wins)
+    }
+    if(any(r=="L")){  # row player loses
+      losses <- sum(r=="L")
+      Hcur[             players[j]         ] %<>% inc(losses)
+      Hcur[c(players[i],players[j],monster)] %<>% dec(losses)
+    }
+    if(any(r=="W")){  # row player draws
+      draws <- sum(r=="D")
+      Hcur[                        monster ] %<>% inc(draws)
+      Hcur[c(players[i],players[j],monster)] %<>% dec(draws)
+    }
+  }  # j loop closes
+} # i loop closes
+Hcur
+```
+
+```{r}
+dotchart(maxp(Hcur))
+```

### 2019-05-02 inst/collusion_chess_draws.Rmd
more work

@@
 
-Now generate the dataset:
+In inference problems where the dataset is outcomes of paired
+comparisons, such as chess, The Bradley-Terry model can be augmented
+to accommodate draws in a number of ways.  Here I present a new
+augmentation of Bradley-Terry in which a drawn chess match is regarded
+as a three-way competition between the players and an entity that wins
+if the match is drawn.  The resulting likelihood functions are easily
+optimised numerically and I present a number of results.
+
+Consider first the standard Bradley-Terry model for $n$ players.
+First we ignore the effect of playing white.  We wish to find numbers
+("strengths") $p_i,1\leqslant i\leqslant n$ with $p_i\geqslant 0, \sum
+p_i=1$ such that the probability of $i$ beating $j\neq i$ is
+$\frac{p_i}{p_i+p_j}$.  A suitable likelihood function, on the
+assumption of independence of trials, would be
+
+$$
+\prod_{i<j}
+\left(\frac{p_i}{p_i+p_j}\right)^{w_{ij}}
+\left(\frac{p_j}{p_i+p_j}\right)^{l_{ij}}
+$$
+
+where $w_{ij}$ is the number of times $i$ beats $j$ and $l_{ij}$ is
+the number of times $i$ loses to $j$.  Incorporation of draws is
+straightforward: we consider a chess match to be a three-way trial
+between the two players and an entity that wins if the match is drawn.
+We thus seek a likelihood function for non-negative
+$p_1,\ldots,p_n,p_D$ $p_D+\sum p_i=1$ and this would be
+
+\[
+\prod_{i<j}
+\left(\frac{p_i}{p_i+p_j+p_D}\right)^{W_{ij}}
+\left(\frac{p_j}{p_i+p_j+p_D}\right)^{L_{ij}}
+\left(\frac{p_D}{p_i+p_j+p_D}\right)^{D_{ij}}
+\]
+
+where $W_{ij}$ is the number of games won by $i$ when playing $j$,
+$L_{ij}$ is games lost and $D_{ij}$ is games drawn.  However, this
+approach ignores the extra strength conferred by playing white.  We
+thus consider a chess match between two players to be a three-way
+entity but with a friendly ghost with strength $p_W$ whose strength is
+added to the white player.
+
+\[
+\prod_{i<j}
+\left(\frac{p_i+p_W}{p_i+p_j+p_D+p_W}\right)^{W_{ij}^\mathrm{white}}
+\left(\frac{p_j}{p_i+p_j+p_D+p_W    }\right)^{L_{ij}^\mathrm{white}}
+\left(\frac{p_i}{p_i+p_j+p_D+p_W}\right)^{W_{ij}^\mathrm{black}}
+\left(\frac{p_j+p_W}{p_i+p_j+p_D+p_W    }\right)^{L_{ij}^\mathrm{black}}
+\left(\frac{p_D}{p_i+p_j+p_D}\right)^{D_{ij}}
+\]
+
+where $W_{ij}^\mathrm{white}$ is the number of games won by $i$ when
+playing white against $j$, and so on.  The final term is algebraically
+equal to
+
+\[
+\prod_{i<j}
+\left(\frac{p_D}{p_i+p_j+p_D}\right)^{D_{ij}^\mathrm{white}}
+\left(\frac{p_D}{p_i+p_j+p_D}\right)^{D_{ij}^\mathrm{black}}
+\]
+
+which might be preferable on grounds of clarity.  We now apply this
+likelihood model to a real dataset.
 

### 2019-05-03 inst/collusion_chess_draws.Rnw
use Rnw format, as I need sophisticated tables not available in Rmd format

--- 2019-05-02 inst/collusion_chess_draws.Rmd
@@
----
-title: "Collusion at the 1962 chess world championships"
-author: "Robin K. S. Hankin"
-date: "`r Sys.Date()`"
-vignette: |
-  %\VignetteIndexEntry{collusion at the 1962 world chess championships}
-  %\VignetteEngine{knitr::rmarkdown}
-  %\VignetteEncoding{UTF-8}
----
+% -*- mode: noweb; noweb-default-code-mode: R-mode; -*-
+\documentclass[nojss]{jss}
+\usepackage{dsfont}
+\usepackage{bbm}
+\usepackage{amsfonts}
+\usepackage{amsmath}
+\usepackage{wasysym}
+\usepackage{amssymb}
+\usepackage{yfonts}
+\usepackage{booktabs}
 
-First load the package:
-```{r}
-library("hyper2")
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%% just as usual
+\author{Robin K. S. Hankin\\Auckland University of Technology}
+\title{Partial rank data with the \pkg{hyper2} package: likelihood
+  functions for generalized Bradley-Terry models}
+
+%\VignetteIndexEntry{The hyper2 package}
+
+%% for pretty printing and a nice hypersummary also set:
+\Plainauthor{Robin K. S. Hankin}
+\Plaintitle{Chess Draws}
+\Shorttitle{Chess Draws}
+
+%% an abstract and keywords
+\Abstract{
+
+Here I analyse chess games for draws using the hyper2 package.
+}
+
+\Keywords{Dirichlet distribution, hyperdirichlet distribution,
+  combinatorics, \proglang{R}, multinomial distribution, constrained
+optimization, Bradley-Terry}
+\Plainkeywords{Dirichlet distribution, hyperdirichlet distribution,
+  combinatorics, R, multinomial distribution, constrained
+optimization, Bradley-Terry}
+
+%% publication information
+%% NOTE: This needs to filled out ONLY IF THE PAPER WAS ACCEPTED.
+%% If it was not (yet) accepted, leave them commented.
+%% \Volume{13}
+%% \Issue{9}
+%% \Month{September}
+%% \Year{2004}
+%% \Submitdate{2004-09-29}
+%% \Acceptdate{2004-09-29}
+
+%% The address of (at least) one author should be given
+%% in the following format:
+\Address{
+  Robin K. S. Hankin\\
+  Auckland University of Technology\\
+  E-mail: \email{hankin.robin@gmail.com}
+}
+%% It is also possible to add a telephone and fax number
+%% before the e-mail in the following format:
+%% Telephone: +43/1/31336-5053
+%% Fax: +43/1/31336-734
+
+%% for those who use Sweave please include the following line (with % symbols):
+%% need no \usepackage{Sweave.sty}
+
+%% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\newcommand{\bolda}{\mathbf a}
+\newcommand{\boldp}{\mathbf p}
+\newcommand{\boldV}{\mathbf V}
+\newcommand{\boldalpha}{\mbox{\boldmath$\alpha$}}
+\newcommand{\boldzero}{\mathbf 0}
+
+\newcommand{\mcf}{\mathcal F}
+\newcommand{\mcm}{\mathcal M}
+\newcommand{\mcs}{\mathcal S}
+\newcommand{\mcw}{\mathcal W}
+
+\newcommand{\pG}{p_{\mbox{\frakfamily g}}}
+
+\SweaveOpts{}
+\begin{document}
+\section{Introduction}
+
+<<echo=FALSE,print=FALSE>>=
+ignore <- require(magrittr,quietly=TRUE)
+ignore <- require(hyper2,quietly=TRUE)
 options("width" = 90)
-```
+@
+
+\section{Introduction: the Bradley-Terry model}
 
@@
 
-\[
+\begin{equation}
 \prod_{i<j}
@@
 \left(\frac{p_D}{p_i+p_j+p_D}\right)^{D_{ij}}
-\]
+\end{equation}
 
@@
 \left(\frac{p_D}{p_i+p_j+p_D}\right)^{D_{ij}^\mathrm{black}}
-\]
+\end{equation}
 
@@
 
-```{r}
+\newcommand\headercell[1]{%
+   \smash[b]{\begin{tabular}[t]{@{}c@{}} #1 \end{tabular}}}
+
+\begin{table}
+\centering
+\begin{tabular}{@{} *{4}{c} @{}}
+\headercell{\small{Plays White,}\\ \small{wins}} & \multicolumn{3}{c@{}}{\small{Plays black, loses}}\\
+\cmidrule(l){2-4}
+& An &  Kr & Ks        \\ 
+\midrule
+  An & -  &  18 &   6  \\
+  Kr &  7 &  -  &  18  \\
+  Ks & 15 &  30 &  -   \\
+\end{tabular}\quad
+\begin{tabular}{@{} *{4}{c} @{}}
+\headercell{\small{Plays W,}\\ \small{draws}} & \multicolumn{3}{c@{}}{\small{Plays black, draws}}\\
+\cmidrule(l){2-4}
+& An &  Kr & Ks        \\ 
+\midrule
+  An & -  &  20 &  20  \\
+  Kr & 29 &  -  &  72  \\
+  Ks & 26 &  57 &  -   \\
+\end{tabular}\quad
+\begin{tabular}{@{} *{4}{c} @{}}
+\headercell{\small{Plays W,}\\ \small{loses}} & \multicolumn{3}{c@{}}{\small{Plays white, loses}}\\
+\cmidrule(l){2-4}
+& An &  Kr & Ks        \\ 
+\midrule
+  An & -  &  13 &   2  \\
+  Kr &  5 &  -  &   7  \\
+  Ks & 11 &   9 &  -   \\
+\end{tabular}
+\caption{Results of 365 chess matches \label{kka} between three
+  Grandmasters.  Thus, for example, Anand played white against Karpov
+  a total of $18+20+13=51$ times and won 18, drew 20, lost 13}
+\end{table}
+
+The likelihood function is
+
+<<>>=
 jj <- read.table("stockholm1962.txt",header=FALSE)
@@
 results
-```
+@
 
@@
 
-```{r}
+<<>>=
+
 all(t(results) == 2-results)
 which(results != 2-t(results),arr.ind=TRUE)
-```
+@
 
@@
 
-```{r}
-H <- hyper2(pnames=c("draw","collusion",players))
-
-for(i in seq_len(nrow(results)-1)){  # i=row
-    for(j in seq(from=i+1,to=ncol(results))){ # j = col
-        o <- results[i,j]
-        if((nationality[i]=="USSR") & (nationality[j]=="USSR")){
-            monster <- "collusion"
-        } else {
-            monster <- "draw" 
-        }
-        if(o==2){ # row player wins
-            H[players[i]] %<>% inc(1)
-        } else if(o==0){  # col player wins
-            H[players[j]] %<>% inc(1)
-        } else if(o==1){ # game drawn
-            H[monster] %<>% inc(1)
-        } else {
-            stop("this cannot happen")
-        }
-        H[c(players[i],players[j],monster)] %<>% dec(1)
-    }  # j loop closes
-} # i loop closes
-```
-and we have
-
-```{r}
-head(H)
-dotchart(maxp(H),pch=16)
-dotchart(log(maxp(H)),pch=16)
-```
-
-which would indicate strongly that collusion is indeed a strong
-feature of the tournament.  Now some optimization.  First optimize freely:
-
-```{r}
-max_support_free <- maxp(H,give=TRUE)$value
-ml_p   <- maxp(H)
-
-s <- indep(equalp(H))
-small <- 1e-3
-s[1] <- s[1] + small
-s[2] <- s[2] - small
-```
-Perform the constrained optimization:
-```{r}
-ml_p_constrained <- maxp(H,fcm=c(1,-1,rep(0,22)),fcv=0,startp=s)
-max_support_constrained <- loglik(H,indep(ml_p_constrained))
-max_support_constrained
-```
-
-Now compare the supports:
-
-```{r}
-extra_support <- max_support_free - max_support_constrained
-extra_support
-```
-
-well in excess of Edwards's two-unit criterion.  Alternatively, we may
-apply Wilks's theorem and observe that the test statistic is in the
-tail region of its asymptotic distribution:
-
-```{r}
-pchisq(2*extra_support,df=1,lower.tail=FALSE)
-```
-
-# Now Curacao
-
-This is the 5th candidates Curacao 1962 results table:
-
-```{r}
-jj <- read.table("curacao1962.txt",header=FALSE)
-curacao <- as.matrix(jj[,-(1:2)])
-players <- as.character(jj$V1)
-nationality <- as.character(jj$V2)
-rownames(curacao) <- players
-colnames(curacao) <- players
-curacao[is.na(curacao)] <- ""
-curacao <- noquote(curacao)
-curacao
-```
-
-Now convert to a likelihood function:
-```{r}
-Hcur <- hyper2(pnames=c("draw","collusion",players))
-for(i in seq_len(nrow(curacao)-1)){  # i=row
-  for(j in seq(from=i+1,to=ncol(curacao))){ # j = col
-    r <- strsplit(curacao[i,j],'')[[1]]
-    if((nationality[i]=="USSR") & (nationality[j]=="USSR")){
-      monster <- "collusion"
-    } else {
-      monster <- "draw"
-    }
-    if(any(r=="W")){  # row player wins
-      wins <- sum(r=="W")
-      Hcur[  players[i]                    ] %<>% inc(wins)
-      Hcur[c(players[i],players[j],monster)] %<>% dec(wins)
-    }
-    if(any(r=="L")){  # row player loses
-      losses <- sum(r=="L")
-      Hcur[             players[j]         ] %<>% inc(losses)
-      Hcur[c(players[i],players[j],monster)] %<>% dec(losses)
-    }
-    if(any(r=="W")){  # row player draws
-      draws <- sum(r=="D")
-      Hcur[                        monster ] %<>% inc(draws)
-      Hcur[c(players[i],players[j],monster)] %<>% dec(draws)
-    }
-  }  # j loop closes
-} # i loop closes
-Hcur
-```
-
-```{r}
-dotchart(maxp(Hcur))
-```
+\bibliography{hyper2}
+\end{document}

### 2019-05-03 inst/collusion_chess_draws.Rnw
more work

@@
 
-\newcommand{\bolda}{\mathbf a}
-\newcommand{\boldp}{\mathbf p}
-\newcommand{\boldV}{\mathbf V}
-\newcommand{\boldalpha}{\mbox{\boldmath$\alpha$}}
-\newcommand{\boldzero}{\mathbf 0}
-
-\newcommand{\mcf}{\mathcal F}
-\newcommand{\mcm}{\mathcal M}
-\newcommand{\mcs}{\mathcal S}
-\newcommand{\mcw}{\mathcal W}
-
-\newcommand{\pG}{p_{\mbox{\frakfamily g}}}
+\newcommand\headercell[1]{%
+   \smash[b]{\begin{tabular}[t]{@{}c@{}} #1 \end{tabular}}}
 
@@
 We thus seek a likelihood function for non-negative
-$p_1,\ldots,p_n,p_D$ $p_D+\sum p_i=1$ and this would be
+$p_1,\ldots,p_n,p_D$ with $p_D+\sum p_i=1$ and this would be
 
@@
 
-\begin{equation}
-\prod_{i<j}
+\begin{align}\nonumber
+\prod_{i<j}&{}
 \left(\frac{p_i+p_W}{p_i+p_j+p_D+p_W}\right)^{W_{ij}^\mathrm{white}}
-\left(\frac{p_j}{p_i+p_j+p_D+p_W    }\right)^{L_{ij}^\mathrm{white}}
-\left(\frac{p_i}{p_i+p_j+p_D+p_W}\right)^{W_{ij}^\mathrm{black}}
-\left(\frac{p_j+p_W}{p_i+p_j+p_D+p_W    }\right)^{L_{ij}^\mathrm{black}}
-\left(\frac{p_D}{p_i+p_j+p_D}\right)^{D_{ij}}
-\end{equation}
+\left(\frac{p_j}{p_i+p_j+p_D+p_W    }\right)^{L_{ij}^\mathrm{white}}\\
+&{} \left(\frac{p_i}{p_i+p_j+p_D+p_W}\right)^{W_{ij}^\mathrm{black}}\label{likelihood_kka}
+\left(\frac{p_j+p_W}{p_i+p_j+p_D+p_W    }\right)^{L_{ij}^\mathrm{black}}\\
+&{} \left(\frac{p_D}{p_i+p_j+p_D}\right)^{D_{ij}}
+\nonumber
+\end{align}
 
@@
 which might be preferable on grounds of clarity.  We now apply this
-likelihood model to a real dataset.
+likelihood model to a real dataset, shown in table~\ref{kka}.  The
+likelihood function is easily maximized,
+figure~\ref{maxp_karpov_kasparov_anand}.
 
-\newcommand\headercell[1]{%
-   \smash[b]{\begin{tabular}[t]{@{}c@{}} #1 \end{tabular}}}
+\begin{figure}[htbp]
+\begin{center}
+<<rowing_maxp,echo=TRUE,fig=TRUE>>=
+dotchart(maxp(karpov_kasparov_anand),pch=16,xlim=c(0,0.55))
+@
+\caption{Maximum likelihood estimate for the strengths of the 3 Grandmasters in Table~\ref{kka} 
+\label{maxp_karpov_kasparov_anand} under a likelihood defined as Equation~\ref{likelihood_kka}}
+  \end{center}
+\end{figure}
 
@@
 \begin{tabular}{@{} *{4}{c} @{}}
-\headercell{\small{Plays W,}\\ \small{loses}} & \multicolumn{3}{c@{}}{\small{Plays white, loses}}\\
+\headercell{\small{Plays W,}\\ \small{loses}} & \multicolumn{3}{c@{}}{\small{Plays black, wins}}\\
 \cmidrule(l){2-4}
@@
 \end{tabular}
-\caption{Results of 365 chess matches \label{kka} between three
-  Grandmasters.  Thus, for example, Anand played white against Karpov
-  a total of $18+20+13=51$ times and won 18, drew 20, lost 13}
+\caption{Results of 365 chess matches \label{kka} between Grandmaster
+  Anand, Grandmaster Karpov, and Grandmaster Kasparov.  Thus, for
+  example, Anand played white against Karpov a total of $18+20+13=51$
+  times and won 18, drew 20, lost 13}
 \end{table}
 
-The likelihood function is
+Figure~\ref{kka} suggests a number of interesting hypotheses.
+Firstly, we observe that $p_D=0$ cannot possibly be the case: the
+likelihood would be identically zero, owing to the nonzero term in the
+numerator of the last term of Equation~\ref{likelihood_kka}.  However,
+it is conceivable that $p_W=0$ and this hypothesis can readily be
+tested using the concept of profile likelihood.
 

### 2019-05-03 inst/collusion_chess_draws.Rnw
more work

@@
 likelihood would be identically zero, owing to the nonzero term in the
-numerator of the last term of Equation~\ref{likelihood_kka}.  However,
-it is conceivable that $p_W=0$ and this hypothesis can readily be
-tested using the concept of profile likelihood.
+numerator of the last term of Equation~\ref{likelihood_kka}.  Indeed,
+we can give a maximum likelihood estimate for the various strengths
+using the \code{maxp()}``` functionality of the software:
+
+<<>>=
+maxp(karpov_kasparov_anand)
+@
+
+We see that $\widehat{p_D}\simeq 0.53$, compare the naive estimate of
+$\frac{224}{365}\simeq 0.61$ given by observing 224 draws among 265
+matches.  Secondly, we may make inferences about $p_W$.  It is
+conceivable that $p_W=0$ as this term does not appear on its own in
+the numerator of any term.  We may test $H_0\colon p_W=0$ using the
+method of support.
+
+<<>>=
+small <- 1e-4
+
+e <- function(white){
+    maxp(
+        karpov_kasparov_anand,
+        startp=c(1/3-small,1/3-small,1/3-small,small),
+        fcm=-cbind(0, 0, 0, 1),fcv= -white,give=TRUE)$value
+}
+e(0.001)
+e(0.002)
+e(0.003)
+maxp(karpov_kasparov_anand,give=TRUE)$value
+328.2975-335.6025
+@
+
+that is, well over 7 units of support, exceeding Edwards's
+two-units-of-support per degree of freedom and we can reject $p_W=0$.
+
+One very natural hypothesis would be $H_{A=K=K}\colon
+p_\mathrm{Karpov}=p_\mathrm{Kasparov}=p_\mathrm{Anand}$.  This too is
+readily testable:
 

### 2019-05-06 inst/collusion_chess_draws.Rnw
work

@@
 \usepackage{booktabs}
+\usepackage{units}
 
@@
 
-We see that $\widehat{p_D}\simeq 0.53$, compare the naive estimate of
-$\frac{224}{365}\simeq 0.61$ given by observing 224 draws among 265
+We see that $\widehat{p_D}\simeq 0.53$, compare the na\"ive estimate
+of $\frac{224}{365}\simeq 0.61$ given by observing 224 draws among 265
 matches.  Secondly, we may make inferences about $p_W$.  It is
@@
 two-units-of-support per degree of freedom and we can reject $p_W=0$.
-
-One very natural hypothesis would be $H_{A=K=K}\colon
+One very natural hypothesis would be $H_{A=Kr=Ks}\colon
 p_\mathrm{Karpov}=p_\mathrm{Kasparov}=p_\mathrm{Anand}$.  This too is
 readily testable:
+
+<<>>=
+objective <- function(V){
+  x <- V[1]  # strength of all three players
+  w <- V[2]  # white player advantage
+  D <- 1-3*x-w  # Draw monster
+  return(  
+      -loglik(karpov_kasparov_anand,c(x,x,x,w))  # negative as we minimize
+  )
+}
+
+jj <-
+constrOptim(
+    theta=c(0.1,0.4),
+    f=objective,
+    grad=NULL,
+    ui = rbind(diag(2),-diag(2),c(-3,-1)),
+    ci = c(0,0,-1,-1,-1)
+)
+jj
+
+(unconstrained <- maxp(karpov_kasparov_anand,give=TRUE)$value)
+(constrained <- jj$value)
+
+(extra_support <-   unconstrained + constrained)  # negative used
+
+pchisq(2*extra_support,df=2,lower.tail=FALSE)
+@
+
+highly significant.  We can justifiably assert that the players do
+indeed have different strengths, even if we account for white
+advantage and the proclivity to draw.
+
+\subsection{Each player has a draw monster}
+
+<<>>=
+objective <- function(V){
+  karpov <- V[1]
+  kasparov <- V[2]
+  anand <- V[3]
+  white <- V[4]
+
+  draw <- (1-karpov-kasparov-anand-white)/3
+
+    return(  
+        -loglik(kka_3draws,
+                c(karpov,kasparov,anand,white,draw,draw)
+                )
+        )
+}
+
+jj <-
+constrOptim(
+    theta=c(0.1,0.1,0.1,0.1),
+    f=objective,
+    grad=NULL,
+    ui <- rbind(
+        diag(nrow=4),
+        c(-1,-1,-1,-1)
+        ),
+    ci = c(
+        rep(0,4),
+        -1
+        )
+)
+jj
+
+(unconstrained <- maxp(kka_3draws,give=TRUE)$value)
+(constrained <- jj$value)
+
+(extra_support <-   unconstrained + constrained)
+
+pchisq(2*extra_support,df=2,lower.tail=FALSE)
+
+@
+
+Significant according to Edwards but not Fisher!
+
+\section{Chess collusion}
 
@@
 
+\newcommand{\half}{$\nicefrac{1}{2}$}
+
+\begin{table}
+\centering
+\begin{tabular}{l|ccccc}
+& \rotatebox{-90}{Fisher} 
+&\rotatebox{-90}{Geller} 
+&\rotatebox{-90}{Petrosian} 
+&\rotatebox{-90}{Korchnoi} 
+&\rotatebox{-90}{Filip} 
+\\ \hline
+  Fisher       & - & \half&\half& 1&\half\\
+Geller &\half&-&\half&\half&\half\\
+Petrosian &\half&\half&-&\half&\half\\
+Korchnoi &0&\half&\half&-&1\\
+Filip &\half&\half&\half&0\\
+\end{tabular}
+\end{table}
+
 \bibliography{hyper2}

### 2019-05-06 inst/collusion_chess_draws.Rnw
work

@@
 
-<<>>=
-jj <- read.table("stockholm1962.txt",header=FALSE)
-results <- as.matrix(jj[,-(1:2)])
-players <- as.character(jj$V1)
-nationality <- as.character(jj$V2)
-rownames(results) <- players
-colnames(results) <- substr(players,1,1)
-results
-@
-
-(this taken from Wikipedia
-\url{https://en.wikipedia.org/wiki/World_Chess_Championship_1963}).
-In this dataset, "0" means a loss, "1" a draw and "2" a win.  Thus
-Fisher (row 1) beat Korchnoi (column 4).  As a consistency check we
-verify that the results table is antisymmetric:
-
-<<>>=
-
-all(t(results) == 2-results)
-which(results != 2-t(results),arr.ind=TRUE)
-@
-
-Thus Barcza vs Portisch is problematic.  Glossing over this
-
 \newcommand{\half}{$\nicefrac{1}{2}$}
@@
 \centering
-\begin{tabular}{l|ccccc}
+\begin{tabular}{l|cccccc}
 & \rotatebox{-90}{Fisher} 
 &\rotatebox{-90}{Geller} 
-&\rotatebox{-90}{Petrosian} 
+&\rotatebox{-90}{Petrosian\ }
 &\rotatebox{-90}{Korchnoi} 
 &\rotatebox{-90}{Filip}
+& $\cdots$
 \\ \hline
-  Fisher       & - & \half&\half& 1&\half\\
-Geller &\half&-&\half&\half&\half\\
-Petrosian &\half&\half&-&\half&\half\\
-Korchnoi &0&\half&\half&-&1\\
-Filip &\half&\half&\half&0\\
+  Fisher   (USA)    & - & \half&\half& 1&\half&\\
+Geller (USSR) &\half&-&\half&\half&\half&\\
+Petrosian (USSR) &\half&\half&-&\half&\half&\\
+Korchnoi (USSR) &0&\half&\half&-&1&\\
+Filip (TCH)&\half&\half&\half&0&\\
+$\vdots$&&&&&\\
+
 \end{tabular}

### 2019-05-07 inst/collusion_chess_draws.Rnw
work

--- 2019-05-03 inst/collusion_chess_draws.Rnw
@@
 \usepackage{booktabs}
+\usepackage{units}
+\usepackage{xcolor}
 
@@
 First we ignore the effect of playing white.  We wish to find numbers
-("strengths") $p_i,1\leqslant i\leqslant n$ with $p_i\geqslant 0, \sum
-p_i=1$ such that the probability of $i$ beating $j\neq i$ is
-$\frac{p_i}{p_i+p_j}$.  A suitable likelihood function, on the
-assumption of independence of trials, would be
+(``strengths") $p_i$ for $1\leqslant i\leqslant n$ with $p_i\geqslant 0, \sum p_i=1$ such that the probability of $i$ beating $j\neq i$ is $\frac{p_i}{p_i+p_j}$.
+A suitable likelihood function, on the assumption of independence of trials, would be
 
@@
 which might be preferable on grounds of clarity.  We now apply this
-likelihood model to a real dataset, shown in table~\ref{kka}.  The
-likelihood function is easily maximized,
-figure~\ref{maxp_karpov_kasparov_anand}.
+likelihood model to real datasets.
 
-\begin{figure}[htbp]
-\begin{center}
-<<rowing_maxp,echo=TRUE,fig=TRUE>>=
-dotchart(maxp(karpov_kasparov_anand),pch=16,xlim=c(0,0.55))
-@
-\caption{Maximum likelihood estimate for the strengths of the 3 Grandmasters in Table~\ref{kka} 
-\label{maxp_karpov_kasparov_anand} under a likelihood defined as Equation~\ref{likelihood_kka}}
-  \end{center}
-\end{figure}
+\section{Anand, Karpov, Kasparov}
+
+Grandmasters Anand, Karpov, Kasparov have long-standing rivalry and
+have played many matches amongst themselves.  A summary may be found
+in Table~\ref{kka}.
 
@@
 
-Figure~\ref{kka} suggests a number of interesting hypotheses.
-Firstly, we observe that $p_D=0$ cannot possibly be the case: the
-likelihood would be identically zero, owing to the nonzero term in the
-numerator of the last term of Equation~\ref{likelihood_kka}.  Indeed,
-we can give a maximum likelihood estimate for the various strengths
-using the \code{maxp()}``` functionality of the software:
+\begin{equation}\label{kka_likelihood}
+\mathcal{L}\left(p_\mathrm{An},p_\mathrm{Kr},p_\mathrm{Ks},W,D\right)=
+\frac{
+p_\mathrm{An}^{15}
+p_\mathrm{Kr}^{12} 
+p_\mathrm{Ks}^{20}
+D^{224}
+\left(p_\mathrm{Kr} + W\right)^{25}
+\left(p_\mathrm{Ks} + W\right)^{45}
+\left(p_\mathrm{An} + W\right)^{24}
+}{
+\left(p_\mathrm{Kr} + p_\mathrm{Ks} + W + D\right)^{193}
+\left(p_\mathrm{Kr} + p_\mathrm{An} + W + D\right)^{92}
+\left(p_\mathrm{Ks} + p_\mathrm{An} + W + D\right)^{80}
+}
+\end{equation}
+
+This likelihood function is easily maximized:
 
@@
 
-We see that $\widehat{p_D}\simeq 0.53$, compare the naive estimate of
-$\frac{224}{365}\simeq 0.61$ given by observing 224 draws among 265
-matches.  Secondly, we may make inferences about $p_W$.  It is
-conceivable that $p_W=0$ as this term does not appear on its own in
-the numerator of any term.  We may test $H_0\colon p_W=0$ using the
-method of support.
+\begin{figure}[htbp]
+\begin{center}
+<<ddc,echo=TRUE,fig=TRUE>>=
+dotchart(maxp(karpov_kasparov_anand),pch=16,xlim=c(0,0.55))
+@
+\caption{Maximum likelihood estimate for the strengths of the 3 Grandmasters in Table~\ref{kka} 
+\label{maxp_karpov_kasparov_anand} under a likelihood defined as Equation~\ref{likelihood_kka}}
+  \end{center}
+\end{figure}
 
-<<>>=
+and this is show visually in figure~\ref{maxp_karpov_kasparov_anand},
+which suggests a number of interesting and testable statistical
+hypotheses.  Firstly, we observe that $D=0$ cannot possibly be the
+case: the likelihood would be identically zero, owing to the nonzero
+term in the numerator of the last term of
+Equation~\ref{likelihood_kka}.  However, we see that the maximum
+likelihood estimate for Draw strength $\hat{D}\simeq 0.53$: compare
+the na\"ive estimate of $\frac{224}{365}\simeq 0.61$ given by
+observing 224 draws among 265 matches.  The MLE is lower because the
+maximization process automatically adjusts for the different players'
+strengths.  Secondly, we may make inferences about $W$, the extra
+strength conferred by playing White: it is conceivable that $W=0$ as
+this term does not appear on its own in the numerator of any term.  We
+may test $H_0\colon W=0$ using the method of support by maximizing the
+likelihood subject the constraint $W=0$ and comparing with the
+likelihood when the optimization is free.
+
+<<echo=FALSE>>=
 small <- 1e-4
@@
 
-that is, well over 7 units of support, exceeding Edwards's
-two-units-of-support per degree of freedom and we can reject $p_W=0$.
+The result of numerical optimization is 7.304, that is, well over the
+two-units-of-support per degree of freedom suggested by Edwards; we
+can convincingly reject the hypothesis that $W=0$ and infer that
+playing white confers a significant advantage.  Of course, this is
+nothing new CITATION NEEDED, but it is now possible to go
+significantly beyond mere existence of the effect and to quantify it.
+Taking the MLE above, and considering a match between say Karpov and
+Anand, the strengths are Karpov=0.064, Anand=0.151, white =0.104 and
+draw =0.53.  Thus we would estimate that Karpov playing white against
+Kasparov would win with probability
+$\frac{0.064+0.104}{0.064+0.104+0.148+0.532}\simeq 0.198$, draw with
+probability $\frac{532}{64+104+148+532}\simeq 0.625$, and lose with
+probability $\simeq 0.174$ (compare $+0.075\,=0.627-0.297$ playing
+black).
 
-One very natural hypothesis would be $H_{A=K=K}\colon
-p_\mathrm{Karpov}=p_\mathrm{Kasparov}=p_\mathrm{Anand}$.  This too is
-readily testable:
+One very natural hypothesis would be that the players are of equal
+strength and we migth write $H_=\colon
+p_\mathrm{Kr}=p_\mathrm{Ks}=p_\mathrm{An}$.  This too is readily
+testable by the device of calculating the maximum support over the
+region in which $p_\mathrm{Kr}=p_\mathrm{Ks}=p_\mathrm{An}$.
+
+<<print=FALSE,echo=FALSE>>=
+objective <- function(V){
+  x <- V[1]  # strength of all three players
+  w <- V[2]  # white player advantage
+  D <- 1-3*x-w  # Draw monster
+  return(  
+      -loglik(karpov_kasparov_anand,c(x,x,x,w))  # negative as we minimize
+  )
+}
+
+jj <-
+constrOptim(
+    theta=c(0.1,0.4),
+    f=objective,
+    grad=NULL,
+    ui = rbind(diag(2),-diag(2),c(-3,-1)),
+    ci = c(0,0,-1,-1,-1)
+)
+
+unconstrained <- maxp(karpov_kasparov_anand,give=TRUE)$value
+constrained <- jj$value
+
+extra_support <-   unconstrained + constrained  # negative used
+
+pchisq(2*extra_support,df=2,lower.tail=FALSE)
+@
+
+In this case the extra support in going from the constrained to the
+free optimization is about 4.93.  However, this time the constraint
+corresponds to a loss of {\em two} degrees of freedom so we would need
+4 units of support, which is attained in this case.  Alternatively we
+might follow Wilks, and observe that the value of
+$2\Lambda=2*4.92=9.84$ lies in the tail region of its asymptotic
+distribution with a p-value of about 0.0072.  We can justifiably
+assert that the players do indeed have different strengths, even if we
+account for white advantage and the proclivity to draw.
+
+\subsection{Each player has a draw monster}
+
+Players with a cautious style tend to draw CITATION NEEDED and this is
+an important and interesting component of the game.  To investigate
+this we need a likelihood function that can account for possible
+differences in draw proclivity among the players.  This may be
+achieved by considering a match between two players to be a contest
+between five entities: the two real players, the white advantage, and
+two personalised draw entities, one per real player; a drawn match is
+counted as a win for either of the draw ghosts.  Under these
+assumptions, table~\ref{kka} translates into the following likelihood
+function:
+
+\begin{figure}[htbp]
+\begin{center}
+<<ddcc,echo=TRUE,fig=TRUE>>=
+pie(maxp(kka_3draws))
+@
+\caption{Maximum likelihood estimate for the strengths of the 3 Grandmasters in Table~\ref{kka} 
+under a likelihood function that admits differential proclivities to draw}
+  \end{center}
+\end{figure}
+
+\begin{equation}
+\mathcal{L}\left(
+p_\mathrm{An},p_\mathrm{Kr},p_\mathrm{Ks}
+,W,
+d_\mathrm{An},d_\mathrm{Kr},d_\mathrm{Ks}
+\right)=
+\frac{
+\parbox{3in}{$
+p_\mathrm{An}^{15}
+p_\mathrm{Kr}^{12}
+p_\mathrm{Ks}^{20}
+\left(p_\mathrm{Kr} + W\right)^{25}
+\left(p_\mathrm{Ks} + W\right)^{45} 
+\left(p_\mathrm{An} + W\right)^{24}\\ \rule{10mm}{0mm}
+\left(d_\mathrm{Kr} + d_\mathrm{Ks}\right)^{129}
+\left(d_\mathrm{Kr} + d_\mathrm{An}\right)^{49}
+\left(d_\mathrm{Ks} + d_\mathrm{An}\right)^{46}
+$}}{\parbox{3in}{$
+\left(p_\mathrm{Kr} + p_\mathrm{Ks} + W + d_\mathrm{Kr} + d_\mathrm{Ks}\right)^{193}\\ \rule{10mm}{0mm}
+\left(p_\mathrm{Kr} + p_\mathrm{An} + W + d_\mathrm{Kr} + d_\mathrm{An}\right)^{92}\\ \rule{20mm}{0mm}
+\left(p_\mathrm{Ks} + p_\mathrm{An} + W + d_\mathrm{Ks} + d_\mathrm{An}\right)^{80}
+$}}
+\end{equation}
 
 <<>>=
-jj <- read.table("stockholm1962.txt",header=FALSE)
-results <- as.matrix(jj[,-(1:2)])
-players <- as.character(jj$V1)
-nationality <- as.character(jj$V2)
-rownames(results) <- players
-colnames(results) <- substr(players,1,1)
-results
+round(maxp(kka_3draws),3)
 @
 
-(this taken from Wikipedia
-\url{https://en.wikipedia.org/wiki/World_Chess_Championship_1963}).
-In this dataset, "0" means a loss, "1" a draw and "2" a win.  Thus
-Fisher (row 1) beat Korchnoi (column 4).  As a consistency check we
-verify that the results table is antisymmetric:
+We see considerable variability in proclivity to draw and again this
+may be tested with the Method of Support, which gives a $p$-value of
+about 0.039, borderline significant: an interesting result, given that
+in this case no
+player's draw monster appears alone anywhere in the likelihood function.
 
-<<>>=
+<<echo=FALSE,print=FALSE>>=
+objective <- function(V){
+  karpov <- V[1]
+  kasparov <- V[2]
+  anand <- V[3]
+  white <- V[4]
 
-all(t(results) == 2-results)
-which(results != 2-t(results),arr.ind=TRUE)
+  draw <- (1-karpov-kasparov-anand-white)/3
+
+    return(  
+        -loglik(kka_3draws,
+                c(karpov,kasparov,anand,white,draw,draw)
+                )
+        )
+}
+
+jj <-
+constrOptim(
+    theta=c(0.1,0.1,0.1,0.1),
+    f=objective,
+    grad=NULL,
+    ui <- rbind(
+        diag(nrow=4),
+        c(-1,-1,-1,-1)
+        ),
+    ci = c(
+        rep(0,4),
+        -1
+        )
+)
+
+unconstrained <- maxp(kka_3draws,give=TRUE)$value
+constrained <- jj$value
+extra_support <-   unconstrained + constrained
+pchisq(2*extra_support,df=2,lower.tail=FALSE)
 @
 
-Thus Barcza vs Portisch is problematic.  Glossing over this
+\section{Chess collusion}
+
+\newcommand{\halfw}{$\nicefrac{1}{2}$}
+\newcommand{\halfb}{\colorbox{black}{\textcolor{white}{$\nicefrac{{\bf 1}}{{\bf 2}}$}}}
+\newcommand{\zerow}{$0$}
+\newcommand{\zerob}{\colorbox{black}{\textcolor{white}{$0$}}}
+\newcommand{\oneww}{$1$}
+\newcommand{\onebb}{\colorbox{black}{\textcolor{white}{$1$}}}
+
+\begin{table}
+\centering
+\begin{tabular}{l|cccccc}
+& \rotatebox{-90}{Fischer} 
+&\rotatebox{-90}{Geller} 
+&\rotatebox{-90}{Petrosian\ }
+&\rotatebox{-90}{Korchnoi} 
+&\rotatebox{-90}{Filip}
+& $\cdots$
+\\ \hline %       Fischer   Geller    Petrosian Korchnoi    Filip
+Fischer     (USA)  &   -     & \halfb  & \halfw  & \oneww    & \halfb   &\\
+Geller    (USSR)  & \halfw  &    -    & \halfb  & \halfw    & \halfw   &\\
+Petrosian (USSR)  & \halfb  & \halfw  &    -    & \halfb    & \halfw   &\\
+Korchnoi  (USSR)  & \zerob  & \halfb  & \halfw  &   -       & \oneww   &\\
+Filip      (TCH)  & \halfw  & \halfb  & \halfb  & \zerob    &  -       &\\
+$\vdots$&&&&&\\
+\end{tabular}\label{interzonal}
+\caption{Subset of the Fifth Interzonal round-robin, Stockholm 1962.
+  Results (0,\halfw,1) and box colour refer to row player.  Thus, for
+  example, Fischer played black against Geller and drew}
+\end{table}
+
+For the results shown in table~\ref{interzonal}, we would have
+
+\begin{equation}
+\mathcal{L}\left(F,G,P,K,L,W,D\right)=
+%                D/(F+G+W+D)  (F+P+D)/(F+P+D+W)   (F+W)/(F+W+D+K)   D/(F+W+L+D)
+%                             D/(G+P+D+W)         D/(G+K+W+D)       D/(G+F+W+D)
+%                                                 D/(P+K+W+D)       D/(P+L+W+D)
+%                                                                   (K+W)/(K+L+W+D)
+\frac{
+  D^8(F+W)(K+W)
+}{\parbox{4in}{$
+    (F+G+W+D)(F+P+W+D)\\ \rule{10mm}{0mm}(F+K+W+D)  (F+L+W+D)\\ \rule{20mm}{0mm}
+         (G+P+W+D)  (G+K+W+D)\\ \rule{30mm}{0mm}(G+F+W+D)
+                    (P+K+W+D)\\ \rule{40mm}{0mm}(P+L+W+D)
+                               (K+L+W+D)
+ $}}
+\end{equation}
+
+\noindent (the numerator includes 8 draws and 2 wins, one for Fischer
+and one for Korchnoi, both playing White); but the full likelihood
+function for the interzonal has 297 terms, including $D^{110}$.  To
+investigate the possibility of Soviet collusion we introduce a new
+draw entity: a soviet Draw monster $D_s$, who wins when two Soviet
+players draw, and is inactive whenever a game with at least one
+non-Soviet player is drawn.  Then $H_N\colon D_s=D$ corresponds to no
+collusion and the likelihood function for table~\ref{interzonal} would
+be
+
+\begin{equation}\label{collusion}
+\mathcal{L}\left(F,G,P,K,L,W,D,D_s\right)=
+%                D/(F+G+W+D)  (F+P+D)/(F+P+D+W)   (F+W)/(F+W+D+K)   D/(F+W+L+D)
+%                             D/(G+P+D+W)         D/(G+K+W+D)       D/(G+F+W+D)
+%                                                 D/(P+K+W+D)       D/(P+L+W+D)
+%                                                                   (K+W)/(K+L+W+D)
+\frac{
+  D_s^3D^5(F+W)(K+W)
+}{\parbox{4in}{$
+    (F+G+W+D)(F+P+W+D)\\ \rule{10mm}{0mm}(F+K+W+D)  (F+L+W+D)\\ \rule{20mm}{0mm}
+         (G+P+W+D_s)  (G+K+W+D_s)\\ \rule{30mm}{0mm}(G+F+W+D)
+                    (P+K+W+D_s)\\ \rule{40mm}{0mm}(P+L+W+D)
+                               (K+L+W+D)
+ $}}
+\end{equation}
+
+\begin{figure}[htbp]
+\begin{center}
+<<collusion,echo=TRUE,fig=TRUE>>=
+dotchart(maxp(collusion),pch=16)
+@
+\caption{sds\label{collusion_fig}}
+  \end{center}
+\end{figure}
+
+In figure~\ref{collusion_fig}, note the high estimate for $D_s$ (also
+note
+Fischer's dominance).
+The maximum likelihood estimate is $D\simeq 0.043, D_s=0.354$; a constrained maximization gives a likelihood ratio of about 5096, or a $p$-value of about $3.6\times 10^{-5}$.
+Indeed we could go further and plot a profile likelihood curve for the log-odds ratio $\ln\left(D_s/D\right)$ which, if Dirichlet, would be approximately Gaussian CITEKENDALANDSTUARTHERE.
+This is illustrated in Figure~\ref{DOESNOTEXIST}, that shows a credible interval of about HAVENOTDONETHISET.
 

### 2019-05-07 inst/collusion_chess_draws.Rnw
work

@@
 \author{Robin K. S. Hankin\\Auckland University of Technology}
-\title{Partial rank data with the \pkg{hyper2} package: likelihood
-  functions for generalized Bradley-Terry models}
+\title{A generalization of the Bradley-Terry model for draws in chess}
 
-%\VignetteIndexEntry{The hyper2 package}
+%\VignetteIndexEntry{chess draws}
 
@@
 In inference problems where the dataset is outcomes of paired
-comparisons, such as chess, The Bradley-Terry model can be augmented
+comparisons, such as chess, the Bradley-Terry model can be augmented
 to accommodate draws in a number of ways.  Here I present a new
@@
 First we ignore the effect of playing white.  We wish to find numbers
-(``strengths") $p_i$ for $1\leqslant i\leqslant n$ with $p_i\geqslant 0, \sum p_i=1$ such that the probability of $i$ beating $j\neq i$ is $\frac{p_i}{p_i+p_j}$.
-A suitable likelihood function, on the assumption of independence of trials, would be
+(``strengths") $p_i$ for $1\leqslant i\leqslant n$ with $p_i\geqslant
+0, \sum p_i=1$ such that the probability of $i$ beating $j\neq i$ is
+$\frac{p_i}{p_i+p_j}$.  A suitable likelihood function, on the
+assumption of independence of trials, would be
 
@@
 \left(\frac{p_j}{p_i+p_j+p_D+p_W    }\right)^{L_{ij}^\mathrm{white}}\\
-&{} \left(\frac{p_i}{p_i+p_j+p_D+p_W}\right)^{W_{ij}^\mathrm{black}}\label{likelihood_kka}
+&{} \left(\frac{p_i}{p_i+p_j+p_D+p_W}\right)^{W_{ij}^\mathrm{black}}\label{likelihood_WD}
 \left(\frac{p_j+p_W}{p_i+p_j+p_D+p_W    }\right)^{L_{ij}^\mathrm{black}}\\
@@
 which might be preferable on grounds of clarity.  We now apply this
-likelihood model to real datasets.
+likelihood model to real chess datasets.  The general approach to
+testing any hypothesis $H$ on any likelihood function $\mathcal{L}$ is
+first to identify the constraint on hypothesis space corresponding to
+$H$.  One then finds the unconstrained maximum likelihood
+$\mathcal{L}_\mathrm{free}$ and compares it with the maximum
+likelihood $\mathcal{L}_\mathrm{cons}$ when the optimization is
+constrained by $H$.  The test statistic is then the likelihood ratio
+$\mathcal{L}_\mathrm{free}/\mathcal{L}_\mathrm{cons}$ or, equivalently
+the support $\mathcal{S}=\log\mathcal{R}$.  Wilks's theorem states
+that the asymptotic distribution of $2\mathcal{S}$ is $\chi^2_n$ where
+$n$ is the number of degrees of freedom implemented by $H$.
 
@@
 have played many matches amongst themselves.  A summary may be found
-in Table~\ref{kka}.
+in Table~\ref{kka}.  Equation~\ref{likelihood_WD} becomes
 
@@
 \caption{Maximum likelihood estimate for the strengths of the 3 Grandmasters in Table~\ref{kka} 
-\label{maxp_karpov_kasparov_anand} under a likelihood defined as Equation~\ref{likelihood_kka}}
+\label{maxp_karpov_kasparov_anand} under a likelihood defined as Equation~\ref{likelihood_WD}}
   \end{center}
@@
 
-and this is show visually in figure~\ref{maxp_karpov_kasparov_anand},
+and this is shown visually in figure~\ref{maxp_karpov_kasparov_anand},
 which suggests a number of interesting and testable statistical
 hypotheses.  Firstly, we observe that $D=0$ cannot possibly be the
-case: the likelihood would be identically zero, owing to the nonzero
-term in the numerator of the last term of
-Equation~\ref{likelihood_kka}.  However, we see that the maximum
+case: the likelihood for this hypothesis would be identically zero,
+owing to the nonzero term in the numerator of the last term of
+Equation~\ref{likelihood_WD}.  However, we see that the maximum
 likelihood estimate for Draw strength $\hat{D}\simeq 0.53$: compare
@@
 }
-e(0.001)
-e(0.002)
-e(0.003)
-maxp(karpov_kasparov_anand,give=TRUE)$value
-328.2975-335.6025
+ignore1 <- e(0.001)
+ignore2 <- e(0.002)
+ignore3 <- e(0.003)
+ignore4 <- maxp(karpov_kasparov_anand,give=TRUE)$value
+support <- 328.2975-335.6025
 @
@@
 
-One very natural hypothesis would be that the players are of equal
-strength and we migth write $H_=\colon
-p_\mathrm{Kr}=p_\mathrm{Ks}=p_\mathrm{An}$.  This too is readily
-testable by the device of calculating the maximum support over the
-region in which $p_\mathrm{Kr}=p_\mathrm{Ks}=p_\mathrm{An}$.
-
-<<print=FALSE,echo=FALSE>>=
+<<kka_equal_strength, print=FALSE,echo=FALSE>>=
 objective <- function(V){
@@
 extra_support <-   unconstrained + constrained  # negative used
-
-pchisq(2*extra_support,df=2,lower.tail=FALSE)
+pvalue <- pchisq(2*extra_support,df=2,lower.tail=FALSE)
 @
 
-In this case the extra support in going from the constrained to the
-free optimization is about 4.93.  However, this time the constraint
-corresponds to a loss of {\em two} degrees of freedom so we would need
-4 units of support, which is attained in this case.  Alternatively we
-might follow Wilks, and observe that the value of
-$2\Lambda=2*4.92=9.84$ lies in the tail region of its asymptotic
-distribution with a p-value of about 0.0072.  We can justifiably
-assert that the players do indeed have different strengths, even if we
-account for white advantage and the proclivity to draw.
+might write $H_=\colon\mathbf{Kr}=\colon\mathbf{Ks}=\colon\mathbf{An}$
+
+One very natural and interesting hypothesis would be that the three
+players are of equal strength
+%equal_strength_is_assessed_in_chunk_kka_equal_strength_above and we
+might write $H_=\colon p_\mathrm{Kr}=p_\mathrm{Ks}=p_\mathrm{An}$.
+This too is readily testable by the device of calculating the maximum
+support over the region in which
+$p_\mathrm{Kr}=p_\mathrm{Ks}=p_\mathrm{An}$.  In this case the extra
+support in going from the constrained to the free optimization is
+about 4.93.  However, this time the constraint corresponds to a loss
+of {\em two} degrees of freedom so we would need 4 units of support,
+which is attained in this case.  Alternatively we might follow Wilks,
+and observe that the value of $2\mathcal{S}=2*4.92=9.84$ lies in the
+tail region of its asymptotic distribution with a p-value of about
+0.0072.  We can justifiably assert that the players do indeed have
+different strengths, even if we account for white advantage and the
+proclivity to draw.
 
@@
 In figure~\ref{collusion_fig}, note the high estimate for $D_s$ (also
-note
-Fischer's dominance).
-The maximum likelihood estimate is $D\simeq 0.043, D_s=0.354$; a constrained maximization gives a likelihood ratio of about 5096, or a $p$-value of about $3.6\times 10^{-5}$.
-Indeed we could go further and plot a profile likelihood curve for the log-odds ratio $\ln\left(D_s/D\right)$ which, if Dirichlet, would be approximately Gaussian CITEKENDALANDSTUARTHERE.
-This is illustrated in Figure~\ref{DOESNOTEXIST}, that shows a credible interval of about HAVENOTDONETHISET.
+note Fischer's dominance).  The maximum likelihood estimate is
+$D\simeq 0.043, D_s=0.354$; a constrained maximization gives a
+likelihood ratio of about 5096, or a $p$-value of about $3.6\times
+10^{-5}$.  Indeed we could go further and plot a profile likelihood
+curve for the log-odds ratio $\ln\left(D_s/D\right)$ which, if
+Dirichlet, would be approximately Gaussian CITEKENDALANDSTUARTHERE.
+This is illustrated in Figure~\ref{DOESNOTEXIST}, that shows a
+credible interval of about HAVENOTDONETHISET.
+
+<<echo=FALSE,print=FALSE>>=
+do_from_scratch  <- FALSE  # set to TRUE to recalculate (slow, maybe 2 mins)
+hotstart <- indep(maxp(collusion))
+
+f <- function(LO,give=FALSE){
+    small <- 1e-5
+
+    eq <- hotstart
+    m <- sum(eq[2:3])
+    a <- exp(LO)
+    eq[2:3] <- m/(1+a)*c(1,a)  # exact
+
+    eq[2:3] <- eq[2:3] + c(-small,small)
+
+    out <-
+        maxp(collusion,
+             startp=eq,
+             fcm=c(0, -1, exp(LO),rep(0,22)),
+             fcv=0,
+             give=TRUE)
+
+    if(give){return(out)}else{return(out$value)}
+}
+
+x <- seq(from= 1.5,to = 2.7,len=32)
+if(do_from_scratch){
+    jj <- sapply(x,f)
+} else {
+    jj <-
+        c(-246.6939, -245.7192, -244.7981, -243.9319, -243.1223, -242.3704, 
+          -241.6776, -241.0452, -240.4744, -239.9661, -239.5216, -239.1417, 
+          -238.8274, -238.5795, -238.3987, -238.2858, -238.2414, -238.2660, 
+          -238.3601, -238.5240, -238.7582, -239.0629, -239.4382, -239.8843, 
+          -240.4013, -240.9891, -241.6476, -242.3767, -243.1762, -244.0458, 
+          -244.9853, -245.9941)
+}
+@
+
+\begin{figure}[htbp]
+\begin{center}
+<<proflike,echo=TRUE,fig=TRUE>>=
+plot(x,jj-max(jj),type='b',xlab="log-odds",ylab="support")
+abline(h=c(0,-2))
+@
+\caption{Profile likelihood \label{proflike} for odds ratio
+  $\log\left(D_s/D\right)$}
+  \end{center}
+\end{figure}
 

### 2019-05-09 inst/collusion_chess_draws.Rnw
analyse curacao candidates match by match rather than using the table (which does not say who played white and who played black)

@@
 
+\cite{moul2009}
+
+\cite{davidson1970}
+
+\cite{joe1990}
+
+\cite{moul2009}: for instance, a player may have the choice of more
+and less conservative strategies with different chances for a
+draw. Risk- averse players may pursue less aggressive strategies,
+trading expected payout for reduced variance
+
+\cite{moul2009}: the question of Soviet collusion during the period
+when the USSR dominated the struggle for the world championship has
+been one of the long-standing debates in the chess world. The
+best-known (but not unique) criticism was aired by Bobby Fischer some
+weeks after the 1962 Candidates Tournament in Curacao
+
 In inference problems where the dataset is outcomes of paired
@@
 \left(\frac{p_D}{p_i+p_j+p_D}\right)^{D_{ij}^\mathrm{black}}
+\end{equation}
+
+\begin{equation}
+\prod_{i<j}
+\left(\frac{p_D}{p_i+p_j+p_D}\right)^{D_{ij}^\mathrm{white}+D_{ij}^\mathrm{black}}
 \end{equation}
@@
 
-\section{Chess collusion}
+\section{Chess collusion: Cura\c{c}ao 1962}
+\begin{table}
+\centering
+\begin{tabular}{ll|cccccccc}
+ &
+&\rotatebox{-90}{Petrosian} 
+&\rotatebox{-90}{Keres} 
+&\rotatebox{-90}{Geller}
+&\rotatebox{-90}{Fischer} 
+&\rotatebox{-90}{Korchnoi}
+&\rotatebox{-90}{Benko}
+&\rotatebox{-90}{Tal}
+&\rotatebox{-90}{Filip}\\
+Petrosian &USSR    &             &DDDD    &DDDD   &DWDD   &DDWW   &DDWD   &WWD    &DWWD\\
+Keres     &USSR    &             &NA      &DDDD   &LDWD   &DDWD   &WWWL   &WDW    &DWWD\\
+Geller    &USSR    &             &NA      &NA     &WWDL   &DDWD   &DDDW   &DWW    &DWWD\\
+Fischer   &USA     &             &NA      &NA     &NA     &LWLD   &LWDW   &DWD    &WDWD\\
+Korchnoi  &USSR    &             &NA      &NA     &NA     &NA     &DDDL   &WLD    &WWWW\\
+Benko     &USA     &             &NA      &NA     &NA     &NA     &NA     &WLD    &LWWD\\
+Tal       &USSR    &             &NA      &NA     &NA     &NA     &NA     &NA     &WLD \\
+Filip     &TCH     &             &NA      &NA     &NA     &NA     &NA     &NA     &A \\
+\end{tabular}
+\caption{Curacao 1962 SPELLED WRONG}
+\end{table}
+
+\section{Chess collusion: interzonal Stockholm 1962}
 
@@
 \begin{tabular}{l|cccccc}
-& \rotatebox{-90}{Fischer} 
+&\rotatebox{-90}{Fischer} 
 &\rotatebox{-90}{Geller} 
@@
 \\ \hline %       Fischer   Geller    Petrosian Korchnoi    Filip
-Fischer     (USA)  &   -     & \halfb  & \halfw  & \oneww    & \halfb   &\\
+Fischer    (USA)  &   -     & \halfb  & \halfw  & \oneww    &\halfb    &\\
 Geller    (USSR)  & \halfw  &    -    & \halfb  & \halfw    & \halfw   &\\
@@
 In figure~\ref{collusion_fig}, note the high estimate for $D_s$ (also
-note Fischer's dominance).  The maximum likelihood estimate is
-$D\simeq 0.043, D_s=0.354$; a constrained maximization gives a
-likelihood ratio of about 5096, or a $p$-value of about $3.6\times
-10^{-5}$.  Indeed we could go further and plot a profile likelihood
-curve for the log-odds ratio $\ln\left(D_s/D\right)$ which, if
-Dirichlet, would be approximately Gaussian CITEKENDALANDSTUARTHERE.
-This is illustrated in Figure~\ref{DOESNOTEXIST}, that shows a
-credible interval of about HAVENOTDONETHISET.
+note
+Fischer's dominance).
+
+The maximum likelihood estimate is $D\simeq 0.043, D_s=0.354$; a constrained maximization gives a likelihood ratio of about 5096, or a $p$-value of about $3.6\times 10^{-5}$.
+Indeed we could go further and plot a profile likelihood curve for the log-odds ratio $\ln\left(D_s/D\right)$ which, if Dirichlet, would be accurately parabolic over a wide range (OHAGAN AND FORSTER, p343).
+This is illustrated in Figure~\ref{DOESNOTEXIST}, that shows a credible interval of about HAVENOTDONETHISET.
 
@@
 \caption{Profile likelihood \label{proflike} for odds ratio
-  $\log\left(D_s/D\right)$}
+  $\log\left(D_s/D\right)$: note the accurately parabolic figure in
+  the region plotted}
   \end{center}
@@
 
-\bibliography{hyper2}
+\bibliography{chess}
 \end{document}

### 2019-05-09 inst/curacao1962_individual_games.R

@@ -1,0 +1,57 @@
+library("hyper2")
+jj <- read.table("curacao1962_candidates.txt",header=FALSE)
+
+players <- 
+c("Petrosian", "Keres", "Geller", "Fischer", "Korchnoi", "Benko", "Tal", "Filip")
+names(players) <-   # nationality
+c("USSR", "USSR", "USSR", "USA", "USSR", "USA", "USSR", "TCH")
+
+d <- as.matrix(cbind(as.character(jj$V1),as.character(jj$V2)))
+results <- as.character(jj$V3)
+colnames(d) <- c("white","black")
+
+stopifnot(all(c(d) %in% players))
+
+white <- "white"
+draw <- "draw"
+sovdraw <- "sovdraw"
+
+H <- hyper2(pnames=c(white,draw,sovdraw,players))
+
+
+for(i in seq_len(nrow(d))){
+  white_player <- d[i,1]
+  black_player <- d[i,2]
+  result <- results[i]
+  nationality_white <- names(players)[which(players == white_player)]
+  nationality_black <- names(players)[which(players == black_player)]
+  
+  if((nationality_black=="USSR") &   (nationality_white=="USSR")){  # two Soviets
+    drawmonster <- "sovdraw"  # Soviet draw
+  } else {
+    drawmonster <- "draw"  # regular draw
+  }
+  
+  if(result == "1-0"){  # white victory
+    winner <- white_player
+    loser <- black_player
+    H[c(winner,white                  )] %<>% inc
+    H[c(winner,white,drawmonster,loser)] %<>% dec
+  } else if(result == "0-1"){ # black victory
+    winner <- black_player
+    loser <- white_player
+    H[c(winner                        )] %<>% inc
+    H[c(winner,loser,white,drawmonster)] %<>% dec
+  } else if(result == "1/2-1/2"){
+    H[c(drawmonster                               )] %<>% inc
+    H[c(drawmonster,winner,loser,white,drawmonster)] %<>% inc
+  } else {
+    stop(paste("result = ", result, ". Should be 1-0, 0-1, or 1/2-1/2",sep=""))
+  }
+}  # i loop closes
+
+
+
+    
+
+ 

### 2019-05-13 inst/collusion_chess_draws.Rnw
more work, including correcting a transpose error on the third subtable of the kka results

@@
 draw.  Risk- averse players may pursue less aggressive strategies,
-trading expected payout for reduced variance
+trading expected payout for reduced variance.
+
+moul: White's relative advantage over Black will be dependent upon the
+comparative mastery of the two players.
+
+The most obvious issue regarding the threshold against which we compare
+this relative advantage of White is the inherent advantage of White's
+position in moving first. We correspondingly do not expect these base
+thresholds to be symmetric and impose no such restriction.
 
@@
 In inference problems where the dataset is outcomes of paired
-comparisons, such as chess, the Bradley-Terry model can be augmented
-to accommodate draws in a number of ways.  Here I present a new
-augmentation of Bradley-Terry in which a drawn chess match is regarded
-as a three-way competition between the players and an entity that wins
-if the match is drawn.  The resulting likelihood functions are easily
-optimised numerically and I present a number of results.
+comparisons, such as chess, the Bradley-Terry
+model~\citep{bradley1952} can be augmented to accommodate draws in a
+number of ways.  Here I present a new augmentation of Bradley-Terry in
+which a drawn chess match is regarded as a three-way competition
+between the players and an entity that wins if the match is drawn.
+The resulting likelihood functions are easily optimised numerically
+and I present a number of results.
 
@@
 \prod_{i<j}
-\left(\frac{p_D}{p_i+p_j+p_D}\right)^{D_{ij}^\mathrm{white}}
-\left(\frac{p_D}{p_i+p_j+p_D}\right)^{D_{ij}^\mathrm{black}}
-\end{equation}
-
-\begin{equation}
+\left(\frac{p_D}{p_i+p_j+p_D+p_W}\right)^{D_{ij}^\mathrm{white}}
+\left(\frac{p_D}{p_i+p_j+p_D+p_W}\right)^{D_{ij}^\mathrm{black}}
+=
 \prod_{i<j}
@@
 which might be preferable on grounds of clarity.  We now apply this
-likelihood model to real chess datasets.  The general approach to
-testing any hypothesis $H$ on any likelihood function $\mathcal{L}$ is
-first to identify the constraint on hypothesis space corresponding to
-$H$.  One then finds the unconstrained maximum likelihood
-$\mathcal{L}_\mathrm{free}$ and compares it with the maximum
-likelihood $\mathcal{L}_\mathrm{cons}$ when the optimization is
-constrained by $H$.  The test statistic is then the likelihood ratio
-$\mathcal{L}_\mathrm{free}/\mathcal{L}_\mathrm{cons}$ or, equivalently
-the support $\mathcal{S}=\log\mathcal{R}$.  Wilks's theorem states
-that the asymptotic distribution of $2\mathcal{S}$ is $\chi^2_n$ where
-$n$ is the number of degrees of freedom implemented by $H$.
+and related likelihood models to real chess datasets.  The general
+approach to testing any hypothesis $H$ on any likelihood function
+$\mathcal{L}$ is first to identify the constraint on hypothesis space
+corresponding to $H$.  One then finds the unconstrained maximum
+likelihood $\mathcal{L}_\mathrm{free}$ and compares it with the
+maximum likelihood $\mathcal{L}_\mathrm{cons}$ when the optimization
+is constrained by $H$.  The test statistic is then the likelihood
+ratio $\mathcal{L}_\mathrm{free}/\mathcal{L}_\mathrm{cons}$ or,
+equivalently the support $\mathcal{S}=\log\mathcal{R}$.  Wilks's
+theorem states that the asymptotic distribution of $2\mathcal{S}$ is
+$\chi^2_n$ where $n$ is the number of degrees of freedom implemented
+by $H$.
+
+\subsection{Discussion}
+
+The above likelihood functions are objective and reasonable
+representations of chess match results.  However, they differ markedly
+from previous likelihood functions of draws in that they include
+entities representing white player advantage ($p_W$), and the
+proclivity to draw $p_D$).   The $p_W$
 
@@
 \midrule
-  An & -  &  13 &   2  \\
-  Kr &  5 &  -  &   7  \\
-  Ks & 11 &   9 &  -   \\
+  An & -  &   5 &   11  \\
+  Kr & 13 &  -  &    9  \\
+  Ks &  2 &   7 &  -   \\
 \end{tabular}
@@
 
-\begin{equation}
+\begin{equation}\label{KKA_personalised_drawmonster}
 \mathcal{L}\left(
@@
 about 0.039, borderline significant: an interesting result, given that
-in this case no
-player's draw monster appears alone anywhere in the likelihood function.
+in this case no draw monster appears alone anywhere in the likelihood
+function.
 
@@
 
+\subsection{Each player has a distinct White monster}
+
+We now consider the possibility that the strength conferred by playing
+White, is specific to each player and we denote this $w_\mathrm{An}$,
+$w_\mathrm{Kr}$ and $w_\mathrm{Ks}$.  The corresponding likelihood
+function would be
+
+\begin{equation}\label{KKA_personalised_whitemonster}
+%\mathcal{L}\left(
+%p_\mathrm{An},p_\mathrm{Kr},p_\mathrm{Ks},
+%w_\mathrm{An},w_\mathrm{Kr},w_\mathrm{Ks},D
+%\right)=
+\frac{
+%\parbox{3in}
+%       {$
+p_\mathrm{An}^{15}
+p_\mathrm{Kr}^{12}
+p_\mathrm{Ks}^{20}
+D^{224}
+\left(p_\mathrm{An} + w_\mathrm{An}\right)^{24}%\\ \rule{10mm}{1mm}
+\left(p_\mathrm{Kr} + w_\mathrm{Kr}\right)^{25}
+\left(p_\mathrm{Ks} + w_\mathrm{Ks}\right)^{45}
+%$}
+}{\parbox{4in}{$
+\left(p_\mathrm{An} + p_\mathrm{Kr} + w_\mathrm{An} +  D\right)^{43}
+\left(p_\mathrm{An} + p_\mathrm{Kr} + w_\mathrm{Kr} +  D\right)^{49}\\ \rule{10mm}{0mm}
+\left(p_\mathrm{An} + p_\mathrm{Ks} + w_\mathrm{An} +  D\right)^{37}
+\left(p_\mathrm{An} + p_\mathrm{Ks} + w_\mathrm{Ks} +  D\right)^{43}\\ \rule{20mm}{0mm}
+\left(p_\mathrm{Kr} + p_\mathrm{Ks} + w_\mathrm{Kr} +  D\right)^{99}
+\left(p_\mathrm{Kr} + p_\mathrm{Ks} + w_\mathrm{Ks} +  D\right)^{94}
+$}}
+\end{equation}
+
+In the likelihood function above we see, for example, that Anand won a
+total of $18+6=24$ matches playing White, $13+2=15$ games playing
+Black, and that Anand played white against Karpov a total of
+$18+20+5=43$ times.  We fail to reject the null of
+$w_\mathrm{An}=w_\mathrm{Kr}=w_\mathrm{Ks}$, having a likelihood ratio
+of $4.5$, or an asymptotic $p$-value of $0.076$.  There is no strong
+evidence that the players differ in regard to their white advantage.
+
 \section{Chess collusion: Cura\c{c}ao 1962}
@@
 \end{tabular}
-\caption{Curacao 1962 SPELLED WRONG}
+\caption{Cura\c{c}ao 1962 candidates matches, world championship 1963}
 \end{table}
@@
 
+\section{Conclusions}
+
+In this context we note, following \cite{langer2013}, that a player
+may well adopt a riskier strategy after a sequence of defeats,
+adopting a ``nothing-to-lose'' mentality.  Such a phenomenon would be
+detectable, at least in principle, by generalizing the personalised
+draw likelihood function \ref{KKA_personalised_drawmonster} and
+seeking evidence for the draw monster's strength changing as a
+tournament progresses: it would become stronger for a player on a
+winning streak, and weaker for a losing player.
+
 \bibliography{chess}

### 2019-05-14 inst/collusion_chess_draws.Rnw
work

@@
 
-Here I analyse chess games for draws using the hyper2 package.
-}
+In inference problems where the dataset is outcomes of paired
+comparisons, such as chess, the Bradley-Terry model can be augmented
+to chess-specific considerations in a number of ways.  Here I present
+a new generalization of Bradley-Terry in which a chess match is
+regarded as a three-way competition between the two players and an
+entity that wins if the match is drawn.  Bradley-Terry can be further
+modified to account for the White player advantage in a number of
+ways.  This methodology affords insight into players' strengths,
+response to playing Black or White, and risk-aversion as manifested by
+probability of drawing.  The likelihood functions arising are easily
+optimised numerically.  I analyse a number of datasets of chess
+results, including the infamous World Chess Championships 1963 in
+which the Soviets were accused of collusion.  I conclude that the
+scoreline of the Candidates Tournament (Cura\c{c}ao 1962) exhibits no
+evidence of collusion, in agreement with previous work.  However, I
+find strong evidence of increased Soviet proclivity to draw in the
+Interzonal Tournament, Stockholm 1962.}
 
@@
 
-\cite{moul2009}: for instance, a player may have the choice of more
-and less conservative strategies with different chances for a
-draw.  Risk- averse players may pursue less aggressive strategies,
-trading expected payout for reduced variance.
-
 moul: White's relative advantage over Black will be dependent upon the
@@
 
-In inference problems where the dataset is outcomes of paired
-comparisons, such as chess, the Bradley-Terry
-model~\citep{bradley1952} can be augmented to accommodate draws in a
-number of ways.  Here I present a new augmentation of Bradley-Terry in
-which a drawn chess match is regarded as a three-way competition
-between the players and an entity that wins if the match is drawn.
-The resulting likelihood functions are easily optimised numerically
-and I present a number of results.
+Many situations involve pairwise comparisons that amount to a
+Bernoulli trial.  The canonical application would be to two-way
+competitive situations such as tennis in which one or other player is
+idenfied as the winner, but the system has been applied to many
+non-competitive situations such mating preference~\cite{hankin_aylmer}
+and consumer behaviour.  The Bradley-Terry model~\citep{bradley1952}
+considers $n$ players and assigns non-negative numbers
+$p_1,\ldots,p_n$ with unit sum to each player.  Then player $i$ beats
+player $j$ with probability $\frac{p_i}{p_i+p_j}$.  This simple system
+has been modified and extended many times, for example to account for
+multi-way choices such as icon selection~\cite{hankinhyperdirichlet}.
 
-Consider first the standard Bradley-Terry model for $n$ players.
-First we ignore the effect of playing white.  We wish to find numbers
-(``strengths") $p_i$ for $1\leqslant i\leqslant n$ with $p_i\geqslant
-0, \sum p_i=1$ such that the probability of $i$ beating $j\neq i$ is
-$\frac{p_i}{p_i+p_j}$.  A suitable likelihood function, on the
-assumption of independence of trials, would be
+Bradley-Terry is not directly applicable to Chess for two reasons:
+firstly, the White player is at a considerable
+advantage~\cite{asdfasdffdsfds}; and secondly, many chess matches end
+in a draw.  Considering White advantage first, we find an overall mean
+of about $(+0.375=0.394-0.276)$.  However, interpretations of this
+statistic vary and the issue is hotly debated with opinions ranging
+from Rowson who compares first-move advantage to that of serving in
+tennis, to that of commentators such as \citeauthor{adorjan} who
+claims that White's advantage is a ``delusion''; and that Black can
+mount an intelligent and dynamic response to white: ``chess is a game
+of complete information, and Black's information is always
+greater---by one move!''.
 
-$$
+Chess games may be won by either player, or may be drawn. The
+implication of a draw is that neither player is better than the other.
+Before 1867, tournament games that were drawn were replayed and in
+this context we note that this procedure reduces a chess match to a
+Bernoulli trial, with probability one.  This system became unworkable
+due to the large number of draws, resulting in the modern practice of
+awarding each player of a drawn match half a point, rendering chess a
+zero-sum game.  Drawn games are generally considered to be undesirable
+as they are generally poorer in spectacular or striking elements such
+as sharp tactical combinations or sacrifice.  Excessive drawing
+dissuades sponsors, leaves spectators feeling cheated, and leaves
+commentators with material to discuss with the public.
+\cite{moul2009} concurs, conceptualising a player as being able to
+adopt a strategy on a spectrum ranging from conservative to
+aggressive, thus being able to control the probability of a draw.  He
+suggests that risk-averse players may pursue less vigorous strategies,
+trading expected payout for reduced variance.
+
+such as chess, the Bradley-Terry model can
+be augmented to chess-specific considerations in a number of ways.
+Here I present a new generalization of Bradley-Terry in which a chess
+match is regarded as a three-way competition between the two players
+and an entity that wins if the match is drawn.  Bradley-Terry can be
+further modified to account for the White player advantage in a number
+of ways.  This methodology affords insight into players' strengths,
+response to playing Black or White, and risk-aversion as manifested by
+probability of drawing.  The likelihood functions arising are easily
+optimised numerically.  I analyse a number of datasets of chess
+results, including the infamous World Chess Championships 1963 in
+which the Soviets were accused of collusion.  I conclude that the
+scoreline of the Candidates Tournament (Cura\c{c}ao 1962) exhibits no
+evidence of collusion, in agreement with previous work.  However, I
+find strong evidence of increased Soviet proclivity to draw in the
+Interzonal Tournament, Stockholm 1962.
+
+Consider first the standard Bradley-Terry model for a finite number of
+chess players.  First we ignore the effect of playing white.  We wish
+to find numbers (``strengths") $p_i$ for $1\leqslant i\leqslant n$
+with $p_i\geqslant 0, \sum p_i=1$ such that the probability of $i$
+beating $j\neq i$ is $\frac{p_i}{p_i+p_j}$.  A suitable likelihood
+function, on the assumption of independence of trials, would be
+
+\begin{equation}
 \prod_{i<j}
@@
 \left(\frac{p_j}{p_i+p_j}\right)^{l_{ij}}
-$$
+\end{equation}
 
@@
 thus consider a chess match between two players to be a three-way
-entity but with a friendly ghost with strength $p_W$ whose strength is
-added to the white player.
+competition between the two ``real'' players and the draw monster, but
+this time we interpret the white player as the combined strength of
+the player himself and another entity, the White monster, whose
+strength $p_W$ is added to the white player.
 

### 2019-05-14 inst/collusion_chess_draws.Rnw
work

@@
 find strong evidence of increased Soviet proclivity to draw in the
-Interzonal Tournament, Stockholm 1962.}
+Interzonal Tournament, Stockholm 1962.
+
+The likelihood functions arising are easily optimised numerically.  I
+analyse a number of datasets of chess results, including the infamous
+World Chess Championships 1963 in which the Soviets were accused of
+collusion.
+
+}
 
@@
 
-\cite{moul2009}
-
 \cite{davidson1970}
@@
 \cite{joe1990}
-
-moul: White's relative advantage over Black will be dependent upon the
-comparative mastery of the two players.
-
-The most obvious issue regarding the threshold against which we compare
-this relative advantage of White is the inherent advantage of White's
-position in moving first. We correspondingly do not expect these base
-thresholds to be symmetric and impose no such restriction.
-
-\cite{moul2009}: the question of Soviet collusion during the period
-when the USSR dominated the struggle for the world championship has
-been one of the long-standing debates in the chess world. The
-best-known (but not unique) criticism was aired by Bobby Fischer some
-weeks after the 1962 Candidates Tournament in Curacao
 
@@
 idenfied as the winner, but the system has been applied to many
-non-competitive situations such mating preference~\cite{hankin_aylmer}
-and consumer behaviour.  The Bradley-Terry model~\citep{bradley1952}
+non-competitive situations such mating preference~\citep{west2008} and
+consumer behaviour.  The Bradley-Terry model~\citep{bradley1952}
 considers $n$ players and assigns non-negative numbers
@@
 has been modified and extended many times, for example to account for
-multi-way choices such as icon selection~\cite{hankinhyperdirichlet}.
+multi-way choices such as mating preferences~\citep{hankin2010}.
 
@@
 zero-sum game.  Drawn games are generally considered to be undesirable
-as they are generally poorer in spectacular or striking elements such
-as sharp tactical combinations or sacrifice.  Excessive drawing
-dissuades sponsors, leaves spectators feeling cheated, and leaves
-commentators with material to discuss with the public.
-\cite{moul2009} concurs, conceptualising a player as being able to
-adopt a strategy on a spectrum ranging from conservative to
-aggressive, thus being able to control the probability of a draw.  He
-suggests that risk-averse players may pursue less vigorous strategies,
-trading expected payout for reduced variance.
+as they are usually poorer in spectacular or striking elements such as
+sharp tactical combinations or sacrifice.  Excessive drawing dissuades
+sponsors, leaves spectators feeling cheated, and gives commentators
+less material to discuss.  Players themselves do not seem to value
+draws in the wider context of the game: \citeauthor{fischer1969}, for
+example, includes only 8 draws in his compilation of 60 memorable
+games, surely not representative of tournament play in which about
+34\% of games are drawn ($\chi^2$, $p<3\times 10^{-6}$).
 
-such as chess, the Bradley-Terry model can
-be augmented to chess-specific considerations in a number of ways.
-Here I present a new generalization of Bradley-Terry in which a chess
-match is regarded as a three-way competition between the two players
-and an entity that wins if the match is drawn.  Bradley-Terry can be
-further modified to account for the White player advantage in a number
-of ways.  This methodology affords insight into players' strengths,
-response to playing Black or White, and risk-aversion as manifested by
-probability of drawing.  The likelihood functions arising are easily
-optimised numerically.  I analyse a number of datasets of chess
-results, including the infamous World Chess Championships 1963 in
-which the Soviets were accused of collusion.  I conclude that the
-scoreline of the Candidates Tournament (Cura\c{c}ao 1962) exhibits no
-evidence of collusion, in agreement with previous work.  However, I
-find strong evidence of increased Soviet proclivity to draw in the
-Interzonal Tournament, Stockholm 1962.
+\cite{moul2009} conceptualise a player as being able to adopt a
+strategy on a spectrum ranging from conservative to aggressive, thus
+being able to exert some control over the probability of a draw.  They
+suggest that risk-averse players may pursue less vigorous strategies,
+trading expected payout for reduced variance. \cite{langer2013} {\bf
+  cite something better here} point out that rational players in an
+ongoing competitive situation modify their strategies depending on
+whether they are leading or lagging: the lagging player will adopt
+higher risk strategy (`nothing to lose'), while the leading player is
+incentivised to protect the lead by lowering the risk of losing.  In a
+chess context, this means the leading player will aim for a draw.
+
+Such arguments suggest that considering a drawn chess match as a
+qualitatively different outcome from a win for either player: the play
+itself tends to be different, the drawing players have a different
+mindset, and drawn games have less value in the wider context of the
+game.
+
+\section{Bradley-Terry}
 
@@
 where $w_{ij}$ is the number of times $i$ beats $j$ and $l_{ij}$ is
-the number of times $i$ loses to $j$.  Incorporation of draws is
-straightforward: we consider a chess match to be a three-way trial
-between the two players and an entity that wins if the match is drawn.
-We thus seek a likelihood function for non-negative
-$p_1,\ldots,p_n,p_D$ with $p_D+\sum p_i=1$ and this would be
+the number of times $i$ loses to $j$.  Here I present a new
+generalization of Bradley-Terry in which a chess match is regarded as
+a three-way competition between the two players and an entity that
+wins if the match is drawn.  The system is further modified to account
+for first-move advantage by positing an entity whose strength is added
+to that of the White player.  This methodology affords insight into
+players' strengths, response to playing Black or White, and
+risk-aversion as manifested by probability of drawing.
+
+If we consider a chess match to be a three-way trial, we define a
+likelihood function for non-negative $p_1,\ldots,p_n,p_D$ with
+$p_D+\sum p_i=1$ and this would be
 
@@
 $L_{ij}$ is games lost and $D_{ij}$ is games drawn.  However, this
-approach ignores the extra strength conferred by playing white.  We
-thus consider a chess match between two players to be a three-way
-competition between the two ``real'' players and the draw monster, but
-this time we interpret the white player as the combined strength of
-the player himself and another entity, the White monster, whose
-strength $p_W$ is added to the white player.
+approach ignores the extra strength conferred by the first-move
+advantage.  We thus consider another entity whose strength is added to
+the white player.  The strength of the white player is thus the sum of
+the player's strength and another entity, the White monster.
+Recalling that a game might be a draw, we obtain the probabilities as
+given in Table~\ref{probabilities}.  The corresponding likelihood
+function is given below in equation \ref{likelihood_WD}:
+
+\newcommand{\oo}{\vphantom{W^{W^W}}}
+
+\begin{table}
+\centering
+\begin{tabular}{l|cc}\\
+  &$i$ plays white & $j$ plays white\\ \hline
+$\Prob{(\mbox{$i$ wins})}$  & $\frac{p_i+p_W\oo}{p_i+p_j+p_W+D}$ & $\frac{p_i}{p_i+p_j+p_W+D}$  \\
+$\Prob{(\mbox{draw})}$      & $\frac{D\oo}{p_i+p_j+p_W+D}$       & $\frac{D}{p_i+p_j+p_W+D}$    \\
+$\Prob{(\mbox{$i$ loses})}$ & $\frac{p_j\oo}{p_i+p_j+p_W+D}$     & $\frac{p_j+p_W}{p_i+p_j+p_W+D}$\\
+\end{tabular}
+\caption{Probabilities of a win, draw, loss for the two players under a generalized Bradley-Terry model\label{probabilities}}
+\end{table}
 
@@
 \section{Chess collusion: Cura\c{c}ao 1962}
+
+\cite{moul2009}: the question of Soviet collusion during the period
+when the USSR dominated the struggle for the world championship has
+been one of the long-standing debates in the chess world. The
+best-known (but not unique) criticism was aired by Bobby Fischer some
+weeks after the 1962 Candidates Tournament in Curacao
+
 \begin{table}

### 2019-05-15 inst/collusion_chess_draws.Rnw
work

@@
 entity that wins if the match is drawn.  Bradley-Terry can be further
-modified to account for the White player advantage in a number of
-ways.  This methodology affords insight into players' strengths,
-response to playing Black or White, and risk-aversion as manifested by
+modified to account for the white player advantage in a number of
+ways.  Such reification affords insight into players' strengths,
+response to playing black or white, and risk-aversion as manifested by
 probability of drawing.  The likelihood functions arising are easily
@@
 Interzonal Tournament, Stockholm 1962.
-
-The likelihood functions arising are easily optimised numerically.  I
-analyse a number of datasets of chess results, including the infamous
-World Chess Championships 1963 in which the Soviets were accused of
-collusion.
-
 }
@@
 Bradley-Terry is not directly applicable to Chess for two reasons:
-firstly, the White player is at a considerable
+firstly, the white player is at a considerable
 advantage~\cite{asdfasdffdsfds}; and secondly, many chess matches end
-in a draw.  Considering White advantage first, we find an overall mean
+in a draw.  Considering white advantage first, we find an overall mean
 of about $(+0.375=0.394-0.276)$.  However, interpretations of this
@@
 tennis, to that of commentators such as \citeauthor{adorjan} who
-claims that White's advantage is a ``delusion''; and that Black can
+claims that white's advantage is a ``delusion''; and that Black can
 mount an intelligent and dynamic response to white: ``chess is a game
@@
 for first-move advantage by positing an entity whose strength is added
-to that of the White player.  This methodology affords insight into
-players' strengths, response to playing Black or White, and
+to that of the white player.  This methodology affords insight into
+players' strengths, response to playing black or white, and
 risk-aversion as manifested by probability of drawing.
@@
 If we consider a chess match to be a three-way trial, we define a
-likelihood function for non-negative $p_1,\ldots,p_n,p_D$ with
-$p_D+\sum p_i=1$ and this would be
+likelihood function for non-negative $p_1,\ldots,p_n,D$ with
+$D+\sum p_i=1$ and this would be
 
@@
 \prod_{i<j}
-\left(\frac{p_i}{p_i+p_j+p_D}\right)^{W_{ij}}
-\left(\frac{p_j}{p_i+p_j+p_D}\right)^{L_{ij}}
-\left(\frac{p_D}{p_i+p_j+p_D}\right)^{D_{ij}}
+\left(\frac{p_i}{p_i+p_j+D}\right)^{W_{ij}}
+\left(\frac{p_j}{p_i+p_j+D}\right)^{L_{ij}}
+\left(\frac{D  }{p_i+p_j+D}\right)^{D_{ij}}
 \end{equation}
@@
 the white player.  The strength of the white player is thus the sum of
-the player's strength and another entity, the White monster.
+the player's strength and another entity, the White Monster.
 Recalling that a game might be a draw, we obtain the probabilities as
@@
   &$i$ plays white & $j$ plays white\\ \hline
-$\Prob{(\mbox{$i$ wins})}$  & $\frac{p_i+p_W\oo}{p_i+p_j+p_W+D}$ & $\frac{p_i}{p_i+p_j+p_W+D}$  \\
-$\Prob{(\mbox{draw})}$      & $\frac{D\oo}{p_i+p_j+p_W+D}$       & $\frac{D}{p_i+p_j+p_W+D}$    \\
-$\Prob{(\mbox{$i$ loses})}$ & $\frac{p_j\oo}{p_i+p_j+p_W+D}$     & $\frac{p_j+p_W}{p_i+p_j+p_W+D}$\\
+$\Prob{(\mbox{$i$ wins})}$  & $\frac{p_i+W\oo}{p_i+p_j+W+D}$ & $\frac{p_i}{p_i+p_j+W+D}$  \\
+$\Prob{(\mbox{draw})}$      & $\frac{D\oo}{p_i+p_j+W+D}$     & $\frac{D}{p_i+p_j+W+D}$    \\
+$\Prob{(\mbox{$i$ loses})}$ & $\frac{p_j\oo}{p_i+p_j+W+D}$   & $\frac{p_j+W}{p_i+p_j+W+D}$\\
 \end{tabular}
@@
 \prod_{i<j}&{}
-\left(\frac{p_i+p_W}{p_i+p_j+p_D+p_W}\right)^{W_{ij}^\mathrm{white}}
-\left(\frac{p_j}{p_i+p_j+p_D+p_W    }\right)^{L_{ij}^\mathrm{white}}\\
-&{} \left(\frac{p_i}{p_i+p_j+p_D+p_W}\right)^{W_{ij}^\mathrm{black}}\label{likelihood_WD}
-\left(\frac{p_j+p_W}{p_i+p_j+p_D+p_W    }\right)^{L_{ij}^\mathrm{black}}\\
-&{} \left(\frac{p_D}{p_i+p_j+p_D}\right)^{D_{ij}}
+\left(\frac{p_i+W}{p_i+p_j+D+W}  \right)^{W_{ij}^\mathrm{white}}
+\left(\frac{p_j}{p_i+p_j+D+W    }\right)^{L_{ij}^\mathrm{white}}\\
+&{} \left(\frac{p_i}{p_i+p_j+D+W}\right)^{W_{ij}^\mathrm{black}}\label{likelihood_WD}
+\left(\frac{p_j+W}{p_i+p_j+D+W  }\right)^{L_{ij}^\mathrm{black}}\\
+&{} \left(\frac{D}{p_i+p_j+D    }\right)^{D_{ij}}
 \nonumber
@@
 \prod_{i<j}
-\left(\frac{p_D}{p_i+p_j+p_D+p_W}\right)^{D_{ij}^\mathrm{white}}
-\left(\frac{p_D}{p_i+p_j+p_D+p_W}\right)^{D_{ij}^\mathrm{black}}
+\left(\frac{D}{p_i+p_j+D+W}\right)^{D_{ij}^\mathrm{white}}
+\left(\frac{D}{p_i+p_j+D+W}\right)^{D_{ij}^\mathrm{black}}
 =
 \prod_{i<j}
-\left(\frac{p_D}{p_i+p_j+p_D}\right)^{D_{ij}^\mathrm{white}+D_{ij}^\mathrm{black}}
+\left(\frac{p_D}{p_i+p_j+D}\right)^{D_{ij}^\mathrm{white}+D_{ij}^\mathrm{black}}
 \end{equation}
@@
 and related likelihood models to real chess datasets.  The general
-approach to testing any hypothesis $H$ on any likelihood function
-$\mathcal{L}$ is first to identify the constraint on hypothesis space
-corresponding to $H$.  One then finds the unconstrained maximum
-likelihood $\mathcal{L}_\mathrm{free}$ and compares it with the
-maximum likelihood $\mathcal{L}_\mathrm{cons}$ when the optimization
-is constrained by $H$.  The test statistic is then the likelihood
-ratio $\mathcal{L}_\mathrm{free}/\mathcal{L}_\mathrm{cons}$ or,
-equivalently the support $\mathcal{S}=\log\mathcal{R}$.  Wilks's
-theorem states that the asymptotic distribution of $2\mathcal{S}$ is
-$\chi^2_n$ where $n$ is the number of degrees of freedom implemented
-by $H$.
+approach follows~\citet{edwards1972}.  To test any hypothesis $H$ on a
+likelihood function $\mathcal{L}$ one first identifies the constraint
+on hypothesis space corresponding to $H$.  One then finds the
+unconstrained maximum likelihood $\mathcal{L}_\mathrm{free}$ and
+compares it with the maximum likelihood $\mathcal{L}_\mathrm{cons}$
+when the optimization is constrained by $H$.  The test statistic is
+then the likelihood ratio
+$\mathcal{L}_\mathrm{free}/\mathcal{L}_\mathrm{cons}$ or, equivalently
+the support $\mathcal{S}=\log\mathcal{R}$.  Wilks's theorem states
+that the asymptotic distribution of $2\mathcal{S}$ is $\chi^2_n$ where
+$n$ is the number of degrees of freedom implemented by $H$.
 
@@
 from previous likelihood functions of draws in that they include
-entities representing white player advantage ($p_W$), and the
-proclivity to draw $p_D$).   The $p_W$
+entities representing proclivity to draw $D$, and first-mover
+advantage ($W$).
+
+Considering white player a
+
+The method of \citet{Edwards1972} involves only
+numerical optimization (not integration which is computationally
+expensive), and because gradients are available convergence is
+generally rapid.
 
@@
 The result of numerical optimization is 7.304, that is, well over the
-two-units-of-support per degree of freedom suggested by Edwards; we
-can convincingly reject the hypothesis that $W=0$ and infer that
-playing white confers a significant advantage.  Of course, this is
-nothing new CITATION NEEDED, but it is now possible to go
-significantly beyond mere existence of the effect and to quantify it.
-Taking the MLE above, and considering a match between say Karpov and
-Anand, the strengths are Karpov=0.064, Anand=0.151, white =0.104 and
-draw =0.53.  Thus we would estimate that Karpov playing white against
-Kasparov would win with probability
+two-units-of-support per degree of freedom suggested by
+\citeauthor{edwards1972}; we can convincingly reject the hypothesis that
+$W=0$ and infer that playing white confers a significant advantage.
+This is consistent with expectation, but it is now possible to go
+significantly beyond mere existence of the effect and to quantify it,
+accounting for the differential strengths of the players.  Taking the
+MLE above, and considering a match between say Karpov and Anand, the
+strengths are Karpov=0.064, Anand=0.151, white =0.104 and draw =0.53.
+Thus we would estimate that Karpov playing white against Kasparov
+would win with probability
 $\frac{0.064+0.104}{0.064+0.104+0.148+0.532}\simeq 0.198$, draw with
 probability $\frac{532}{64+104+148+532}\simeq 0.625$, and lose with
-probability $\simeq 0.174$ (compare $+0.075\,=0.627-0.297$ playing
-black).
+probability $\simeq 0.174$ (compare $+0.075\,=0.627-0.297$ if Karpov
+plays black).
 
@@
 
-might write $H_=\colon\mathbf{Kr}=\colon\mathbf{Ks}=\colon\mathbf{An}$
-
 One very natural and interesting hypothesis would be that the three
 players are of equal strength
-%equal_strength_is_assessed_in_chunk_kka_equal_strength_above and we
-might write $H_=\colon p_\mathrm{Kr}=p_\mathrm{Ks}=p_\mathrm{An}$.
-This too is readily testable by the device of calculating the maximum
-support over the region in which
-$p_\mathrm{Kr}=p_\mathrm{Ks}=p_\mathrm{An}$.  In this case the extra
-support in going from the constrained to the free optimization is
-about 4.93.  However, this time the constraint corresponds to a loss
-of {\em two} degrees of freedom so we would need 4 units of support,
-which is attained in this case.  Alternatively we might follow Wilks,
-and observe that the value of $2\mathcal{S}=2*4.92=9.84$ lies in the
-tail region of its asymptotic distribution with a p-value of about
-0.0072.  We can justifiably assert that the players do indeed have
-different strengths, even if we account for white advantage and the
-proclivity to draw.
+%equal_strength_is_assessed_in_chunk_kka_equal_strength_above____________
+and we might hypothesise that
+$p_\mathrm{Kr}=p_\mathrm{Ks}=p_\mathrm{An}$.  This too is readily
+testable by the device of maximizing the support, subject to the
+constraint that $p_\mathrm{Kr}=p_\mathrm{Ks}=p_\mathrm{An}$.  In this
+case the extra support in going from the constrained to the free
+optimization is about 4.93, or a likelihood ratio of about $138$.
+However, this time the constraint corresponds to a loss of {\em two}
+degrees of freedom so we would need 4 units of support, which is
+attained in this case.  Alternatively we might follow Wilks, and
+observe that the value of $2\mathcal{S}=2*4.92=9.84$ lies in the tail
+region of its asymptotic distribution with a p-value of about 0.0072.
+We can justifiably assert that the players do indeed have different
+strengths, even if we account for white advantage and the proclivity
+to draw.
 
@@
 
-\subsection{Each player has a distinct White monster}
+\subsection{Each player has a distinct White Monster}
 
 We now consider the possibility that the strength conferred by playing
-White, is specific to each player and we denote this $w_\mathrm{An}$,
+white, is specific to each player and we denote this $w_\mathrm{An}$,
 $w_\mathrm{Kr}$ and $w_\mathrm{Ks}$.  The corresponding likelihood
@@
 In the likelihood function above we see, for example, that Anand won a
-total of $18+6=24$ matches playing White, $13+2=15$ games playing
+total of $18+6=24$ matches playing white, $13+2=15$ games playing
 Black, and that Anand played white against Karpov a total of
@@
 \noindent (the numerator includes 8 draws and 2 wins, one for Fischer
-and one for Korchnoi, both playing White); but the full likelihood
+and one for Korchnoi, both playing white); but the full likelihood
 function for the interzonal has 297 terms, including $D^{110}$.  To

### 2019-05-15 inst/collusion_chess_draws.Rnw
work

@@
 to chess-specific considerations in a number of ways.  Here I present
-a new generalization of Bradley-Terry in which a chess match is
+a new generalization of Bradley-Terry in which a chess game is
 regarded as a three-way competition between the two players and an
-entity that wins if the match is drawn.  Bradley-Terry can be further
+entity that wins if the game is drawn.  Bradley-Terry can be further
 modified to account for the white player advantage in a number of
@@
 \begin{document}
-\section{Introduction}
 
@@
 firstly, the white player is at a considerable
-advantage~\cite{asdfasdffdsfds}; and secondly, many chess matches end
-in a draw.  Considering white advantage first, we find an overall mean
-of about $(+0.375=0.394-0.276)$.  However, interpretations of this
-statistic vary and the issue is hotly debated with opinions ranging
-from Rowson who compares first-move advantage to that of serving in
-tennis, to that of commentators such as \citeauthor{adorjan} who
-claims that white's advantage is a ``delusion''; and that Black can
-mount an intelligent and dynamic response to white: ``chess is a game
-of complete information, and Black's information is always
-greater---by one move!''.
+advantage~\cite{asdfasdffdsfds}; and secondly, many chess games end in
+a draw: white wins about $37.5\%$ of all games, black wins $27.6\%$,
+and about $39.4\%$ are drawn.  We consider the first-mover advantage
+first, then consider the effect of drawn games.
 
-Chess games may be won by either player, or may be drawn. The
-implication of a draw is that neither player is better than the other.
-Before 1867, tournament games that were drawn were replayed and in
-this context we note that this procedure reduces a chess match to a
-Bernoulli trial, with probability one.  This system became unworkable
-due to the large number of draws, resulting in the modern practice of
-awarding each player of a drawn match half a point, rendering chess a
-zero-sum game.  Drawn games are generally considered to be undesirable
-as they are usually poorer in spectacular or striking elements such as
-sharp tactical combinations or sacrifice.  Excessive drawing dissuades
-sponsors, leaves spectators feeling cheated, and gives commentators
-less material to discuss.  Players themselves do not seem to value
-draws in the wider context of the game: \citeauthor{fischer1969}, for
-example, includes only 8 draws in his compilation of 60 memorable
-games, surely not representative of tournament play in which about
-34\% of games are drawn ($\chi^2$, $p<3\times 10^{-6}$).
+\paragraph{First-mover advantage} Traditionally White moves first in
+any game, and this feature introduces an asymmetry into play.
+Historically, we see about 1.35 white wins for every black win.
+However, interpretations of this statistic vary and the issue is hotly
+debated with opinions ranging from Rowson who compares first-move
+advantage to that of serving in tennis, to that of commentators such
+as \citeauthor{adorjan} who claims that White's advantage is a
+``delusion''.  \cite{asdfasdasfd} goes further, asserting that Black
+can exploit the situation by mounting a directed response to White, of
+the type seen in analysis of sequential zero-sum games, in which the
+first-mover advantage can be positive or negative.
+
+\paragraph{Draws} Chess games may be won by either player, or may be
+drawn; the implication of a draw is that neither player is better than
+the other.  Before 1867, tournament games that were drawn were
+replayed and in this context we note that this procedure reduces a
+chess game to a Bernoulli trial, with probability one.  This system
+became unworkable due to the large number of draws, resulting in the
+modern practice of awarding each player of a drawn game half a point,
+rendering chess a zero-sum game.  Drawn games are generally considered
+to be undesirable as they are usually poorer in spectacular or
+striking elements such as sharp tactical combinations or sacrifice.
+Excessive drawing dissuades sponsors, leaves spectators feeling
+cheated, and gives commentators less material to discuss.  Players
+themselves do not seem to value draws in the wider context of the
+game: \citeauthor{fischer1969}, for example, includes only 8 draws in
+his compilation of 60 memorable games, surely not representative of
+tournament play in which about 39.4\% of games are drawn (binomial
+test, $p<10^{-5}$).
 
@@
 strategy on a spectrum ranging from conservative to aggressive, thus
-being able to exert some control over the probability of a draw.  They
+being able to exert some control over the probability of a draw and
 suggest that risk-averse players may pursue less vigorous strategies,
-trading expected payout for reduced variance. \cite{langer2013} {\bf
-  cite something better here} point out that rational players in an
-ongoing competitive situation modify their strategies depending on
-whether they are leading or lagging: the lagging player will adopt
-higher risk strategy (`nothing to lose'), while the leading player is
+trading expected payout for reduced variance.  Further,
+\cite{langer2013} point out that rational players in an ongoing
+competitive situation modify their strategies depending on whether
+they are leading or lagging: the lagging player will adopt a higher
+risk strategy (`nothing to lose'), while the leading player is
 incentivised to protect the lead by lowering the risk of losing.  In a
@@
 
-Such arguments suggest that considering a drawn chess match as a
-qualitatively different outcome from a win for either player: the play
-itself tends to be different, the drawing players have a different
-mindset, and drawn games have less value in the wider context of the
-game.
+Such arguments suggest considering a drawn chess game as a
+qualitatively different outcome from a statement of equality of skill:
+the play itself tends to be different, the drawing players have a
+different mindset, and drawn games have less value in the wider
+context of the game.
 
-\section{Bradley-Terry}
+\section{Generalized Bradley-Terry}
 
@@
 \prod_{i<j}
-\left(\frac{p_i}{p_i+p_j}\right)^{w_{ij}}
-\left(\frac{p_j}{p_i+p_j}\right)^{l_{ij}}
+\left(\frac{p_i}{p_i+p_j}\right)^{W_{ij}}
+\left(\frac{p_j}{p_i+p_j}\right)^{L_{ij}}
 \end{equation}
 
-where $w_{ij}$ is the number of times $i$ beats $j$ and $l_{ij}$ is
+where $W_{ij}$ is the number of times $i$ beats $j$ and $L_{ij}$ is
 the number of times $i$ loses to $j$.  Here I present a new
-generalization of Bradley-Terry in which a chess match is regarded as
+generalization of Bradley-Terry in which a chess game is regarded as
 a three-way competition between the two players and an entity that
-wins if the match is drawn.  The system is further modified to account
+wins if the game is drawn.  The system is further modified to account
 for first-move advantage by positing an entity whose strength is added
@@
 
-If we consider a chess match to be a three-way trial, we define a
+If we consider a chess game to be a three-way trial, we define a
 likelihood function for non-negative $p_1,\ldots,p_n,D$ with
@@
 
-\subsection{Discussion}
+The above likelihood functions are objective and reasonable
+representations of chess results.  However, they differ markedly from
+previous likelihood functions of draws in that they include entities
+representing proclivity to draw $D$, and first-mover advantage ($W$).
 
-The above likelihood functions are objective and reasonable
-representations of chess match results.  However, they differ markedly
-from previous likelihood functions of draws in that they include
-entities representing proclivity to draw $D$, and first-mover
-advantage ($W$).
+\subsection{Numerical considerations}
 
-Considering white player a
-
-The method of \citet{Edwards1972} involves only
-numerical optimization (not integration which is computationally
-expensive), and because gradients are available convergence is
-generally rapid.
+The Method of Support \citep{edwards1972} involves only numerical
+optimization (not integration which is computationally expensive).
+The analysis presented here uses the framework of the \pkg{hyper2}
+package~\citep{hankin2017} as a convenient and unified interface.
+Because gradients are available, convergence is generally rapid and
+hypothesis-specific constraints on parameter space are, in general,
+straightforward to implement.
 
@@
 Grandmasters Anand, Karpov, Kasparov have long-standing rivalry and
-have played many matches amongst themselves.  A summary may be found
+have played many games amongst themselves.  A summary may be found
 in Table~\ref{kka}.  Equation~\ref{likelihood_WD} becomes
@@
 \end{tabular}
-\caption{Results of 365 chess matches \label{kka} between Grandmaster
+\caption{Results of 365 chess games \label{kka} between Grandmaster
   Anand, Grandmaster Karpov, and Grandmaster Kasparov.  Thus, for
@@
 Equation~\ref{likelihood_WD}.  However, we see that the maximum
-likelihood estimate for Draw strength $\hat{D}\simeq 0.53$: compare
+likelihood estimate for draw strength $\hat{D}\simeq 0.53$: compare
 the na\"ive estimate of $\frac{224}{365}\simeq 0.61$ given by
-observing 224 draws among 265 matches.  The MLE is lower because the
+observing 224 draws among 265 games.  The MLE is lower because the
 maximization process automatically adjusts for the different players'
 strengths.  Secondly, we may make inferences about $W$, the extra
-strength conferred by playing White: it is conceivable that $W=0$ as
+strength conferred by playing white: it is conceivable that $W=0$ as
 this term does not appear on its own in the numerator of any term.  We
@@
 
-The result of numerical optimization is 7.304, that is, well over the
-two-units-of-support per degree of freedom suggested by
-\citeauthor{edwards1972}; we can convincingly reject the hypothesis that
-$W=0$ and infer that playing white confers a significant advantage.
-This is consistent with expectation, but it is now possible to go
-significantly beyond mere existence of the effect and to quantify it,
-accounting for the differential strengths of the players.  Taking the
-MLE above, and considering a match between say Karpov and Anand, the
-strengths are Karpov=0.064, Anand=0.151, white =0.104 and draw =0.53.
-Thus we would estimate that Karpov playing white against Kasparov
-would win with probability
+<<>>=
+ignore3
+ignore4
+ignore3-ignore4
+@
+
+The result of numerical optimization is about 7.01 units of support,
+or a likelihood ratio of $e^{7.01}\simeq 1107.7$: that is, well over
+the two-units-of-support per degree of freedom suggested by
+\citeauthor{edwards1972}; we can convincingly reject the hypothesis
+that $W=0$ and infer that playing white confers a significant
+advantage.  This is consistent with expectation, but it is now
+possible to go significantly beyond mere existence of the effect and
+to quantify it, accounting for the differential strengths of the
+players.  Taking the MLE above, and considering a game between say
+Karpov and Anand, the strengths are Karpov=0.064, Anand=0.151, white
+=0.104 and draw =0.53.  Thus we would estimate that Karpov playing
+white against Kasparov would win with probability
 $\frac{0.064+0.104}{0.064+0.104+0.148+0.532}\simeq 0.198$, draw with
 probability $\frac{532}{64+104+148+532}\simeq 0.625$, and lose with
-probability $\simeq 0.174$ (compare $+0.075\,=0.627-0.297$ if Karpov
-plays black).
+probability $\simeq 0.174$ (compare $+0.075\ {=}0.627\ {-}0.297$ if
+Karpov plays black).
 
@@
 observe that the value of $2\mathcal{S}=2*4.92=9.84$ lies in the tail
-region of its asymptotic distribution with a p-value of about 0.0072.
-We can justifiably assert that the players do indeed have different
-strengths, even if we account for white advantage and the proclivity
-to draw.
+region of its asymptotic $\chi^2_2$ distribution with a $p$-value of
+about 0.0072.  We can justifiably assert that the players do indeed
+have different strengths, even if we account for white advantage and
+the proclivity to draw.
 
@@
 
-Players with a cautious style tend to draw CITATION NEEDED and this is
-an important and interesting component of the game.  To investigate
+As argued above, players with a cautious style tend to draw and this
+is an important and interesting component of the game.  To investigate
 this we need a likelihood function that can account for possible
 differences in draw proclivity among the players.  This may be
-achieved by considering a match between two players to be a contest
+achieved by considering a game between two players to be a contest
 between five entities: the two real players, the white advantage, and
-two personalised draw entities, one per real player; a drawn match is
-counted as a win for either of the draw ghosts.  Under these
-assumptions, table~\ref{kka} translates into the following likelihood
-function:
+two {\em personalised} draw entities, one for each real player; a
+drawn game is counted as a win for either of the draw ``ghosts''.  Under
+these assumptions, table~\ref{kka} translates into the following
+likelihood function:
 
@@
 In the likelihood function above we see, for example, that Anand won a
-total of $18+6=24$ matches playing white, $13+2=15$ games playing
+total of $18+6=24$ games playing white, $13+2=15$ games playing
 Black, and that Anand played white against Karpov a total of
@@
 
-\cite{moul2009}: the question of Soviet collusion during the period
-when the USSR dominated the struggle for the world championship has
-been one of the long-standing debates in the chess world. The
-best-known (but not unique) criticism was aired by Bobby Fischer some
-weeks after the 1962 Candidates Tournament in Curacao
+The method of support allows one to account for proclivity to draw
+while accounting for players' strengths and first-move advantage.  We
+now consider one famous incident, the World Chess Championships 1963,
+in which Fischer famously and publicly claimed that the Soviet players
+colluded.  The accusation was that Soviet players agreed ahead of time
+to draw when playing one another, in order to conserve energy for
+subsequent matches~\citep{moul2009}.
 
-\begin{table}
-\centering
-\begin{tabular}{ll|cccccccc}
- &
-&\rotatebox{-90}{Petrosian} 
-&\rotatebox{-90}{Keres} 
-&\rotatebox{-90}{Geller}
-&\rotatebox{-90}{Fischer} 
-&\rotatebox{-90}{Korchnoi}
-&\rotatebox{-90}{Benko}
-&\rotatebox{-90}{Tal}
-&\rotatebox{-90}{Filip}\\
-Petrosian &USSR    &             &DDDD    &DDDD   &DWDD   &DDWW   &DDWD   &WWD    &DWWD\\
-Keres     &USSR    &             &NA      &DDDD   &LDWD   &DDWD   &WWWL   &WDW    &DWWD\\
-Geller    &USSR    &             &NA      &NA     &WWDL   &DDWD   &DDDW   &DWW    &DWWD\\
-Fischer   &USA     &             &NA      &NA     &NA     &LWLD   &LWDW   &DWD    &WDWD\\
-Korchnoi  &USSR    &             &NA      &NA     &NA     &NA     &DDDL   &WLD    &WWWW\\
-Benko     &USA     &             &NA      &NA     &NA     &NA     &NA     &WLD    &LWWD\\
-Tal       &USSR    &             &NA      &NA     &NA     &NA     &NA     &NA     &WLD \\
-Filip     &TCH     &             &NA      &NA     &NA     &NA     &NA     &NA     &A \\
-\end{tabular}
-\caption{Cura\c{c}ao 1962 candidates matches, world championship 1963}
-\end{table}
+%the question of Soviet collusion during the period when the USSR
+%dominated the struggle for the world championship has been one of the
+%long-standing debates in the chess world.  The best-known (but not
+%unique) criticism was aired by Bobby Fischer some weeks after the 1962
+%Candidates Tournament in Curacao
+
+% \begin{table}
+% \centering
+% \begin{tabular}{ll|cccccccc}
+% &
+% &%\rotatebox{-90}{Petrosian} 
+% &\rotatebox{-90}{Keres} 
+% &\rotatebox{-90}{Geller}
+% &\rotatebox{-90}{Fischer} 
+% &\rotatebox{-90}{Korchnoi}
+% &\rotatebox{-90}{Benko}
+% &\rotatebox{-90}{Tal}
+% &\rotatebox{-90}{Filip}\\
+% Petrosian &USSR    &             &DDDD    &DDDD   &DWDD   &DDWW   &DDWD   &WWD    &DWWD\\
+% Keres     &USSR    &             &NA      &DDDD   &LDWD   &DDWD   &WWWL   &WDW    &DWWD\\
+% Geller    &USSR    &             &NA      &NA     &WWDL   &DDWD   &DDDW   &DWW    &DWWD\\
+% Fischer   &USA     &             &NA      &NA     &NA     &LWLD   &LWDW   &DWD    &WDWD\\
+% Korchnoi  &USSR    &             &NA      &NA     &NA     &NA     &DDDL   &WLD    &WWWW\\
+% Benko     &USA     &             &NA      &NA     &NA     &NA     &NA     &WLD    &LWWD\\
+% Tal       &USSR    &             &NA      &NA     &NA     &NA     &NA     &NA     &WLD \\
+% Filip     &TCH     &             &NA      &NA     &NA     &NA     &NA     &NA     &NA \\
+% \end{tabular}
+% \caption{Cura\c{c}ao 1962 candidates, world championship 1963}
+% \end{table}
 
@@
 \caption{Subset of the Fifth Interzonal round-robin, Stockholm 1962.
-  Results (0,\halfw,1) and box colour refer to row player.  Thus, for
-  example, Fischer played black against Geller and drew}
+  Results $\in\{0,\halfw,1\}$ and box colour refer to row
+  player.  Thus, for example, Fischer played black against Geller and drew}
 \end{table}

### 2019-05-16 inst/collusion_chess_draws.Rnw
work

@@
 
-\begin{figure}[htbp]
-\begin{center}
-<<ddcc,echo=TRUE,fig=TRUE>>=
-pie(maxp(kka_3draws))
-@
-\caption{Maximum likelihood estimate for the strengths of the 3 Grandmasters in Table~\ref{kka} 
-under a likelihood function that admits differential proclivities to draw}
-  \end{center}
-\end{figure}
+%\begin{figure}[htbp]
+%\begin{center}
+%<<ddcc,echo=TRUE,fig=TRUE>>=
+%pie(maxp(kka_3draws))
+%@
+%\caption{Maximum likelihood estimate for the strengths of the 3 Grandmasters in Table~\ref{kka} 
+%under a likelihood function that admits differential proclivities to draw}
+%  \end{center}
+%\end{figure}
 
@@
 
-\section{Chess collusion: Cura\c{c}ao 1962}
+\section{Collusion in the World Chess Championship 1963}
 
@@
 while accounting for players' strengths and first-move advantage.  We
-now consider one famous incident, the World Chess Championships 1963,
-in which Fischer famously and publicly claimed that the Soviet players
-colluded.  The accusation was that Soviet players agreed ahead of time
-to draw when playing one another, in order to conserve energy for
-subsequent matches~\citep{moul2009}.
+now consider one famous incident, following the World Chess
+Championships 1963, in which Fischer famously and publicly claimed
+that the Soviet players colluded.  The accusation was that Soviet
+players agreed ahead of time to draw when playing one another, in
+order to conserve energy for subsequent matches~\citep{moul2009}.
+
+Here I analyse the original subject of the complaint, the Cura\c{c}ao
+1962 Candidates, Table~\ref{candidates}.  Fischer alleged that three
+of the Soviet players (Keres, Petrosian, and Geller) had agreed to
+draw amongst themselves.
 
@@
 
-% \begin{table}
-% \centering
-% \begin{tabular}{ll|cccccccc}
-% &
-% &%\rotatebox{-90}{Petrosian} 
-% &\rotatebox{-90}{Keres} 
-% &\rotatebox{-90}{Geller}
-% &\rotatebox{-90}{Fischer} 
-% &\rotatebox{-90}{Korchnoi}
-% &\rotatebox{-90}{Benko}
-% &\rotatebox{-90}{Tal}
-% &\rotatebox{-90}{Filip}\\
-% Petrosian &USSR    &             &DDDD    &DDDD   &DWDD   &DDWW   &DDWD   &WWD    &DWWD\\
-% Keres     &USSR    &             &NA      &DDDD   &LDWD   &DDWD   &WWWL   &WDW    &DWWD\\
-% Geller    &USSR    &             &NA      &NA     &WWDL   &DDWD   &DDDW   &DWW    &DWWD\\
-% Fischer   &USA     &             &NA      &NA     &NA     &LWLD   &LWDW   &DWD    &WDWD\\
-% Korchnoi  &USSR    &             &NA      &NA     &NA     &NA     &DDDL   &WLD    &WWWW\\
-% Benko     &USA     &             &NA      &NA     &NA     &NA     &NA     &WLD    &LWWD\\
-% Tal       &USSR    &             &NA      &NA     &NA     &NA     &NA     &NA     &WLD \\
-% Filip     &TCH     &             &NA      &NA     &NA     &NA     &NA     &NA     &NA \\
-% \end{tabular}
-% \caption{Cura\c{c}ao 1962 candidates, world championship 1963}
-% \end{table}
+ \begin{table}
+ \centering
+ \begin{tabular}{ll|cccccccc}
+ &
+ &%\rotatebox{-90}{Petrosian} 
+ &\rotatebox{-90}{Keres} 
+ &\rotatebox{-90}{Geller}
+ &\rotatebox{-90}{Fischer} 
+ &\rotatebox{-90}{Korchnoi}
+ &\rotatebox{-90}{Benko}
+ &\rotatebox{-90}{Tal}
+ &\rotatebox{-90}{Filip}\\
+ Petrosian &USSR    &             &DDDD    &DDDD   &DWDD   &DDWW   &DDWD   &WWD    &DWWD\\
+ Keres     &USSR    &             & -      &DDDD   &LDWD   &DDWD   &WWWL   &WDW    &DWWD\\
+ Geller    &USSR    &             & -      & -     &WWDL   &DDWD   &DDDW   &DWW    &DWWD\\
+ Fischer   &USA     &             & -      & -     & -     &LWLD   &LWDW   &DWD    &WDWD\\
+ Korchnoi  &USSR    &             & -      & -     & -     & -     &DDDL   &WLD    &WWWW\\
+ Benko     &USA     &             & -      & -     & -     & -     & -     &WLD    &LWWD\\
+ Tal       &USSR    &             & -      & -     & -     & -     & -     & -     &WLD \\
+ Filip     &TCH     &             & -      & -     & -     & -     & -     & -     & - \\
+ \end{tabular}
+ \caption{Cura\c{c}ao 1962 \label{candidates} Candidates, World Chess Championship 1963}
+ \end{table}
 
-\section{Chess collusion: interzonal Stockholm 1962}
+\subsection{Chess collusion: interzonal Stockholm 1962}
 

### 2019-05-16 inst/collusion_chess_draws.Rnw
work

@@
 
-\newcommand{\oo}{\vphantom{W^{W^W}}}
+\newcommand{\oo}{\vphantom{W^{W^{W^W}}}}
 
@@
 
-Here I analyse the original subject of the complaint, the Cura\c{c}ao
-1962 Candidates, Table~\ref{candidates}.  Fischer alleged that three
-of the Soviet players (Keres, Petrosian, and Geller) had agreed to
-draw amongst themselves.
-
-%the question of Soviet collusion during the period when the USSR
-%dominated the struggle for the world championship has been one of the
-%long-standing debates in the chess world.  The best-known (but not
-%unique) criticism was aired by Bobby Fischer some weeks after the 1962
-%Candidates Tournament in Curacao
-
  \begin{table}
@@
  \end{table}
+
+Here I analyse the original subject of the complaint, the Cura\c{c}ao
+1962 Candidates, Table~\ref{candidates}.  Fischer alleged that three
+of the Soviet players (Keres, Petrosian, and Geller) had agreed to
+draw amongst themselves.  Observe that all twelve of the games between
+the three accused players ended in a draw, a significant result given
+51 winning games and 54 draws (Fisher exact test, one-sided,
+$p=0.0079$).  The likelihood function for this dataset would have
+
+\begin{equation}
+\parbox{4in}{$
+P^5\,
+{Ke}^5\,
+G^4
+\mathit{Fis}\,  % Fischer, cf Filip
+\mathit{Ko}^5
+B
+T^2  % Tal
+\mathit{Fil}\,
+D^{42}C^{12}\oo\\ \rule{10mm}{0mm}
+\left(P+W\right)^3
+\left(\mathit{Ke}+W\right)^4
+\left(G+W\right)^4\
+\left(\mathit{Fis}+W\right)^7\oo\\ \rule{20mm}{0mm}
+\left(\mathit{Ko}+W\right)^2
+\left(B+W\right)^5
+\left(T+W\right)
+\left(\mathit{Fil}+W\right)\oo
+$}
+\end{equation}
+
+on the numerator (where $C$ is the strength of a collusive draw), and
+a total of 36 terms such as $\left(P + \mathit{Fis}+W+D\right)^{4}$
+and $\left(P + \mathit{Ke}+W+C\right)^{4}$ on the denominator.
 
@@
 $\vdots$&&&&&\\
-\end{tabular}\label{interzonal}
-\caption{Subset of the Fifth Interzonal round-robin, Stockholm 1962.
+\end{tabular}
+\caption{Subset of the Fifth Interzonal round-robin, Stockholm 1962.\label{interzonal}
   Results $\in\{0,\halfw,1\}$ and box colour refer to row

### 2019-05-17 inst/collusion_chess_draws.Rnw
work

@@
 results, including the infamous World Chess Championships 1963 in
-which the Soviets were accused of collusion.  I conclude that the
-scoreline of the Candidates Tournament (Cura\c{c}ao 1962) exhibits no
-evidence of collusion, in agreement with previous work.  However, I
-find strong evidence of increased Soviet proclivity to draw in the
-Interzonal Tournament, Stockholm 1962.
-}
+which the Soviets were accused of collusion.  I conclude, on the basis
+of a dataset that includes only the scorelines at the event itself,
+that the scoreline of the Candidates Tournament (Cura\c{c}ao 1962)
+exhibits evidence of collusion ($p<10^{-5}$), in agreement with
+previous work that uses additional information such as historical
+results including number of moves.}
 
@@
 the other.  Before 1867, tournament games that were drawn were
-replayed and in this context we note that this procedure reduces a
-chess game to a Bernoulli trial, with probability one.  This system
-became unworkable due to the large number of draws, resulting in the
-modern practice of awarding each player of a drawn game half a point,
-rendering chess a zero-sum game.  Drawn games are generally considered
-to be undesirable as they are usually poorer in spectacular or
-striking elements such as sharp tactical combinations or sacrifice.
-Excessive drawing dissuades sponsors, leaves spectators feeling
-cheated, and gives commentators less material to discuss.  Players
-themselves do not seem to value draws in the wider context of the
-game: \citeauthor{fischer1969}, for example, includes only 8 draws in
-his compilation of 60 memorable games, surely not representative of
-tournament play in which about 39.4\% of games are drawn (binomial
-test, $p<10^{-5}$).
+replayed until one or other player won, thus reducing chess game to a
+Bernoulli trial, almost surely.  This system became unworkable due to
+the large number of draws, resulting in the modern practice of
+awarding each player of a drawn game half a point, rendering chess a
+zero-sum game.  Drawn games are generally considered to be undesirable
+as they are usually poorer in spectacular or striking elements such as
+sharp tactical combinations or sacrifice.  Excessive drawing dissuades
+sponsors, leaves spectators feeling cheated, and gives commentators
+less material to discuss.  Players themselves do not seem to value
+draws in the wider context of the game: \citeauthor{fischer1969}, for
+example, includes only 8 draws in his compilation of 60 memorable
+games, surely not representative of tournament play in which about
+39.4\% of games are drawn (binomial test, $p<10^{-5}$).
 
@@
 representations of chess results.  However, they differ markedly from
-previous likelihood functions of draws in that they include entities
-representing proclivity to draw $D$, and first-mover advantage ($W$).
+previous likelihood functions of draws in that they include reified
+entities representing proclivity to draw $D$, and white or first-mover
+advantage ($W$).
 
@@
 
-\begin{equation}
-\parbox{4in}{$
+\begin{equation}\parbox{4in}{$
 P^5\,
@@
 \left(\mathit{Fil}+W\right)\oo
-$}
-\end{equation}
+$}\end{equation}
 
-on the numerator (where $C$ is the strength of a collusive draw), and
-a total of 36 terms such as $\left(P + \mathit{Fis}+W+D\right)^{4}$
-and $\left(P + \mathit{Ke}+W+C\right)^{4}$ on the denominator.
+\noindent on the numerator (where $C$ is the strength of a collusive
+draw), and a total of 36 terms such as
+$\left(P+\mathit{Fis}+W+D\right)^{4}$ and
+$\left(P+\mathit{Ke}+W+C\right)^{4}$ on the denominator.  The
+likelihood ratio for the hypothesis that collusive draws have the same
+strength as regular draws isabout $e^{11.8}\simeq 10^5$: we conclude
+that the high number of draws among the three accused players was
+unlikely to be due to chance.
+
+One plausible explanation for the previ
 
@@
 
-\begin{figure}[htbp]
-\begin{center}
-<<collusion,echo=TRUE,fig=TRUE>>=
-dotchart(maxp(collusion),pch=16)
-@
-\caption{sds\label{collusion_fig}}
-  \end{center}
-\end{figure}
-
-In figure~\ref{collusion_fig}, note the high estimate for $D_s$ (also
-note
-Fischer's dominance).
-
-The maximum likelihood estimate is $D\simeq 0.043, D_s=0.354$; a constrained maximization gives a likelihood ratio of about 5096, or a $p$-value of about $3.6\times 10^{-5}$.
-Indeed we could go further and plot a profile likelihood curve for the log-odds ratio $\ln\left(D_s/D\right)$ which, if Dirichlet, would be accurately parabolic over a wide range (OHAGAN AND FORSTER, p343).
-This is illustrated in Figure~\ref{DOESNOTEXIST}, that shows a credible interval of about HAVENOTDONETHISET.
+The maximum likelihood estimate is $D\simeq 0.043, D_s=0.354$; a
+constrained maximization gives a likelihood ratio of about 5096, or a
+$p$-value of about $3.6\times 10^{-5}$.  Indeed we could go further
+and plot a profile likelihood curve for the log-odds ratio
+$\ln\left(D_s/D\right)$ which, if Dirichlet, would be accurately
+parabolic over a wide range (OHAGAN AND FORSTER, p343).  This is
+illustrated in Figure~\ref{DOESNOTEXIST}, that shows a credible
+interval of about HAVENOTDONETHISET.
 
@@
 do_from_scratch  <- FALSE  # set to TRUE to recalculate (slow, maybe 2 mins)
-hotstart <- indep(maxp(collusion))
+hotstart <- indep(maxp(interzonal_collusion))
 
@@
 
-\section{Conclusions}
+\section{Conclusions and further work}
+
+The present work details an approach to chess analysis based on
+reification of entities representing propensity to draw, and the
+first-mover (or white player) advantage.  Likelihood functions for the
+players' and reified entities' strengths are defined using only
+specific tournament outcomes; further, the analysis uses only game
+outcome, not length.
 
@@
 
+\cite{moul2009} present detailed Monte Carlo simulations to assess the likely impact of any Soviet cartel on tournament outcomes.
+In principle, the present system of likelihood and reified entities would admit such simulations and might be interesting to compare results.
+
 \bibliography{chess}

### 2019-05-20 inst/collusion_chess_draws.Rnw
work

@@
 Bradley-Terry is not directly applicable to Chess for two reasons:
-firstly, the white player is at a considerable
-advantage~\cite{asdfasdffdsfds}; and secondly, many chess games end in
-a draw: white wins about $37.5\%$ of all games, black wins $27.6\%$,
-and about $39.4\%$ are drawn.  We consider the first-mover advantage
-first, then consider the effect of drawn games.
+firstly, the white player appears to be at a considerable advantage;
+and secondly, many chess games end in a draw: white wins about
+$37.5\%$ of all games, black wins $27.6\%$, and about $39.4\%$ are
+drawn.  We consider the first-mover advantage first, then consider the
+effect of drawn games.
 
@@
 However, interpretations of this statistic vary and the issue is hotly
-debated with opinions ranging from Rowson who compares first-move
-advantage to that of serving in tennis, to that of commentators such
-as \citeauthor{adorjan} who claims that White's advantage is a
-``delusion''.  \cite{asdfasdasfd} goes further, asserting that Black
-can exploit the situation by mounting a directed response to White, of
-the type seen in analysis of sequential zero-sum games, in which the
-first-mover advantage can be positive or negative.
+debated with opinions ranging from that of~\cite{adams2007}---``white
+to play and win''---to~\citet{adorjan1988}---``black is OK!''.
 
@@
 
+\subsection{Previous work}
+
+\cite{davidson1970} uses a threshhold approach to interpret a draw as
+performance of the players the same to within a threshhold.  Using a
+logistic expression for the difference in players' performance, he
+effectively uses, in current notation, $D=\nu\sqrt{p_ip_j}$.  Here we
+note that such a model automatically give stronger players a higher
+draw potential, and also note that it is difficult to account for
+first-player advantage in this framework.  Such an approach is
+numerically challenging in the current framework as the unit-sum
+constraint becomes nonlinear and likelihood maximization becomes
+difficult.
+
 \subsection{Numerical considerations}
@@
 package~\citep{hankin2017} as a convenient and unified interface.
-Because gradients are available, convergence is generally rapid and
-hypothesis-specific constraints on parameter space are, in general,
-straightforward to implement.
+Because gradients are available, convergence to the evaluate is
+generally rapid and hypothesis-specific constraints on parameter space
+are, in general, straightforward to implement.
 
@@
 
-This likelihood function is easily maximized:
-
-<<>>=
+<<echo=FALSE,print=FALSE,results=hide>>=
 maxp(karpov_kasparov_anand)
@@
 
-\begin{figure}[htbp]
+\begin{figure}[t]
 \begin{center}
-<<ddc,echo=TRUE,fig=TRUE>>=
+<<ddc,echo=FALSE,fig=TRUE>>=
 dotchart(maxp(karpov_kasparov_anand),pch=16,xlim=c(0,0.55))
@@
 
+This likelihood function is easily maximized numerically
 and this is shown visually in figure~\ref{maxp_karpov_kasparov_anand},
@@
 
-<<echo=FALSE>>=
+<<maxlikekka,echo=FALSE,print=FALSE,results=hide>>=
 small <- 1e-4
@@
 
-<<>>=
+<<ignore34diff,echo=FALSE,print=FALSE,results=hide>>=
 ignore3
@@
 
-<<>>=
-round(maxp(kka_3draws),3)
+\begin{figure}[t]
+\begin{center}
+<<echo=FALSE,fig=TRUE>>=
+dotchart(maxp(kka_3draws),pch=16)
 @ 
+\caption{Maximum likelihood estimate for \label{kka_3draws} equation \ref{KKA_personalised_drawmonster} equation} 
+  \end{center}
+\end{figure}
 
-We see considerable variability in proclivity to draw and again this
-may be tested with the Method of Support, which gives a $p$-value of
-about 0.039, borderline significant: an interesting result, given that
-in this case no draw monster appears alone anywhere in the likelihood
-function.
+In figure~\ref{kka_3draws}, we see considerable variability in
+proclivity to draw and again this may be tested with the Method of
+Support, which gives a $p$-value of about 0.039, borderline
+significant: an interesting result, given that in this case no draw
+monster appears alone anywhere in the likelihood function.
 
-<<echo=FALSE,print=FALSE>>=
+<<echo=FALSE,print=FALSE,results=hide>>=
 objective <- function(V){
@@
 players agreed ahead of time to draw when playing one another, in
-order to conserve energy for subsequent matches~\citep{moul2009}.
+order to conserve energy for subsequent matches.
+
+\citet{moul2009} consider this accusation from a statistical
+perspective and present a detailed analysis of Cura\c{c}ao.  Unlike
+the present analysis, which uses a dataset restricted to game result
+(win/loss/draw), \citet{moul2009} included number of moves in each
+game\footnote{The reasoning was that hard-fought games that happened
+  to end in a draw are more likely to last longer than collusive or
+  pre-arranged draws.} in their analysis, and used a dataset of 7355
+games played in the period 1940-1978 to calibrate the ability of the
+players.  In contrast, the analysis presented uses a dataset
+restricted to game outcome (win-loss-draw) in the competition itself.
 
@@
 
-One plausible explanation for the previ
+One plausible and charitable explanation for this result is that
+Soviet players have a higher propensity to draw than western players.
+We can investigate this by considering the interzonal tournament which
+took place in Stockholm in 1962.
 
@@
 draw entity: a soviet Draw monster $D_s$, who wins when two Soviet
-players draw, and is inactive whenever a game with at least one
-non-Soviet player is drawn.  Then $H_N\colon D_s=D$ corresponds to no
-collusion and the likelihood function for table~\ref{interzonal} would
-be
+players draw, and is inactive whenever a game has at least one
+non-Soviet player.  Then $H_N\colon D_s=D$ corresponds to no collusion
+and the likelihood function for table~\ref{interzonal} would be
 
@@
 $\ln\left(D_s/D\right)$ which, if Dirichlet, would be accurately
-parabolic over a wide range (OHAGAN AND FORSTER, p343).  This is
-illustrated in Figure~\ref{DOESNOTEXIST}, that shows a credible
-interval of about HAVENOTDONETHISET.
+parabolic over a wide range~\cite[page 343]{ohagan2004}.  This is
+illustrated in Figure~\ref{proflike}, that shows a credible interval
+of about 1.8-2.4.
 
@@
 \begin{center}
-<<proflike,echo=TRUE,fig=TRUE>>=
+<<proflike,echo=FALSE,fig=TRUE>>=
 plot(x,jj-max(jj),type='b',xlab="log-odds",ylab="support")
@@
 draw likelihood function \ref{KKA_personalised_drawmonster} and
-seeking evidence for the draw monster's strength changing as a
-tournament progresses: it would become stronger for a player on a
-winning streak, and weaker for a losing player.
+seeking evidence for the draw strength changing as a tournament
+progresses: it would become stronger for a player on a winning streak,
+and weaker for a losing player.
 
-\cite{moul2009} present detailed Monte Carlo simulations to assess the likely impact of any Soviet cartel on tournament outcomes.
-In principle, the present system of likelihood and reified entities would admit such simulations and might be interesting to compare results.
+\cite{moul2009} present detailed Monte Carlo simulations
+to assess the likely impact of any Soviet cartel on tournament outcomes.
+In principle, the present system of likelihood and reified entities
+would admit such simulations and might be interesting to compare results.
 

### 2019-05-20 inst/collusion_chess_draws.Rnw
minor wording

@@
 \documentclass[nojss]{jss}
-\usepackage{dsfont}
-\usepackage{bbm}
-\usepackage{amsfonts}
 \usepackage{amsmath}
-\usepackage{wasysym}
 \usepackage{amssymb}
-\usepackage{yfonts}
 \usepackage{booktabs}
 \usepackage{units}
-\usepackage{xcolor}
 
@@
 
-\cite{davidson1970}
-
-\cite{joe1990}
-
 Many situations involve pairwise comparisons that amount to a
@@
 the other.  Before 1867, tournament games that were drawn were
-replayed until one or other player won, thus reducing chess game to a
-Bernoulli trial, almost surely.  This system became unworkable due to
-the large number of draws, resulting in the modern practice of
-awarding each player of a drawn game half a point, rendering chess a
+replayed until one or other player won---reducing a chess game to a
+Bernoulli trial (with probability one).  This system became unworkable
+due to the large number of draws, resulting in the modern practice of
+awarding each player of a drawn game half a point, keeping chess a
 zero-sum game.  Drawn games are generally considered to be undesirable
@@
 logistic expression for the difference in players' performance, he
-effectively uses, in current notation, $D=\nu\sqrt{p_ip_j}$.  Here we
-note that such a model automatically give stronger players a higher
-draw potential, and also note that it is difficult to account for
-first-player advantage in this framework.  Such an approach is
+effectively uses, in current notation, $D=\nu\sqrt{p_ip_j}$.  However,
+\cite{joe1990} uses likelihood methods to show that this model does
+not fit data well.  Here we note that such a model automatically give
+stronger players a higher draw potential, and also note that it is
+difficult to account for first-player advantage in this framework.
+
+Such an approach is
 numerically challenging in the current framework as the unit-sum
@@
 @ 
-\caption{Maximum likelihood estimate for \label{kka_3draws} equation \ref{KKA_personalised_drawmonster} equation} 
+\caption{Maximum likelihood estimate for \label{kka_3draws} the support function given in
+  equation  \ref{KKA_personalised_drawmonster}
+} 
   \end{center}
@@
 
-The method of support allows one to account for proclivity to draw
-while accounting for players' strengths and first-move advantage.  We
-now consider one famous incident, following the World Chess
-Championships 1963, in which Fischer famously and publicly claimed
-that the Soviet players colluded.  The accusation was that Soviet
-players agreed ahead of time to draw when playing one another, in
-order to conserve energy for subsequent matches.
+The method of support allows one to estimate proclivity to draw while
+accounting for players' strengths and first-move advantage.  We now
+consider one famous incident, following the World Chess Championships
+1963, in which Fischer famously and publicly claimed that the Soviet
+players colluded.  The accusation was that Soviet players agreed ahead
+of time to draw when playing one another, in order to conserve energy
+for subsequent matches.
 
 \citet{moul2009} consider this accusation from a statistical
-perspective and present a detailed analysis of Cura\c{c}ao.  Unlike
-the present analysis, which uses a dataset restricted to game result
-(win/loss/draw), \citet{moul2009} included number of moves in each
-game\footnote{The reasoning was that hard-fought games that happened
-  to end in a draw are more likely to last longer than collusive or
-  pre-arranged draws.} in their analysis, and used a dataset of 7355
-games played in the period 1940-1978 to calibrate the ability of the
-players.  In contrast, the analysis presented uses a dataset
-restricted to game outcome (win-loss-draw) in the competition itself.
+perspective and present a detailed analysis.  Unlike the present work,
+which uses a dataset restricted to game result (win/loss/draw),
+\citet{moul2009} included number of moves in each game\footnote{The
+  reasoning was that hard-fought games that happened to end in a draw
+  are more likely to last longer than collusive or pre-arranged
+  draws.} in their analysis, and used a dataset of 7355 games played
+in the period 1940-1978 to calibrate the ability of the players.  In
+contrast, the analysis presented uses a dataset restricted to game
+outcome (win-loss-draw) in the competition itself.
 
@@
 Here I analyse the original subject of the complaint, the Cura\c{c}ao
-1962 Candidates, Table~\ref{candidates}.  Fischer alleged that three
-of the Soviet players (Keres, Petrosian, and Geller) had agreed to
-draw amongst themselves.  Observe that all twelve of the games between
-the three accused players ended in a draw, a significant result given
-51 winning games and 54 draws (Fisher exact test, one-sided,
-$p=0.0079$).  The likelihood function for this dataset would have
+1962 Candidates (match results shown in Table~\ref{candidates}).
+Fischer alleged that three of the Soviet players (Keres, Petrosian,
+and Geller) had agreed to draw amongst themselves.  Observe that all
+twelve of the games between the three accused players ended in a draw,
+a significant result given 51 winning games and 54 draws (Fisher exact
+test, one-sided, $p=0.0079$).
+
+To investigate the possibility of Soviet collusion we introduce a new
+draw entity: a soviet draw monster $C$, who wins when two colluding
+players draw, and is inactive whenever a game has at least one
+non-colluding player.  Then $H_N\colon C=D$ corresponds to no
+collusion and the likelihood function would be
 
@@
 \noindent on the numerator (where $C$ is the strength of a collusive
-draw), and a total of 36 terms such as
+draw), and a total of 30 terms such as
 $\left(P+\mathit{Fis}+W+D\right)^{4}$ and
-$\left(P+\mathit{Ke}+W+C\right)^{4}$ on the denominator.  The
-likelihood ratio for the hypothesis that collusive draws have the same
-strength as regular draws isabout $e^{11.8}\simeq 10^5$: we conclude
-that the high number of draws among the three accused players was
-unlikely to be due to chance.
+$\left(P+\mathit{Ke}+W+C\right)^{4}$ [indicating that Petrosian played
+  Keres 4 times] on the denominator.  The likelihood ratio for the
+hypothesis that collusive draws have the same strength as regular
+draws is about $e^{11.8}\simeq 10^5$: we conclude that the high number
+of draws among the three accused players was unlikely to be due to
+chance.
 
-One plausible and charitable explanation for this result is that
-Soviet players have a higher propensity to draw than western players.
-We can investigate this by considering the interzonal tournament which
-took place in Stockholm in 1962.
+One plausible explanation for this result is that Soviet players have
+a higher propensity to draw than western players.  We can investigate
+this by considering the interzonal tournament which took place in
+Stockholm in 1962.
 
@@
 
-For the results shown in table~\ref{interzonal}, we would have
+The fith interzonal took place in Stockholm in 1962, in the form of a
+23-player round-robin tournament.  For the results shown in
+table~\ref{interzonal}, we would have a likelihood function
 
@@
 function for the interzonal has 297 terms, including $D^{110}$.  To
-investigate the possibility of Soviet collusion we introduce a new
-draw entity: a soviet Draw monster $D_s$, who wins when two Soviet
-players draw, and is inactive whenever a game has at least one
-non-Soviet player.  Then $H_N\colon D_s=D$ corresponds to no collusion
-and the likelihood function for table~\ref{interzonal} would be
+investigate the possibility of increased Soviet drawing proclivity, we
+introduce a Soviet draw monster $D_s$, who wins when two Soviet
+players draw, and is inactive if a game has at least one non-Soviet
+player.  Then $H_N\colon D_s=D$ corresponds to no collusion and the
+likelihood function for table~\ref{interzonal} would be
 

### 2019-05-20 inst/collusion_chess_draws.Rnw
further tweaking

@@
 
-In inference problems where the dataset is outcomes of paired
-comparisons, such as chess, the Bradley-Terry model can be augmented
-to chess-specific considerations in a number of ways.  Here I present
-a new generalization of Bradley-Terry in which a chess game is
-regarded as a three-way competition between the two players and an
-entity that wins if the game is drawn.  Bradley-Terry can be further
-modified to account for the white player advantage in a number of
-ways.  Such reification affords insight into players' strengths,
-response to playing black or white, and risk-aversion as manifested by
-probability of drawing.  The likelihood functions arising are easily
-optimised numerically.  I analyse a number of datasets of chess
-results, including the infamous World Chess Championships 1963 in
-which the Soviets were accused of collusion.  I conclude, on the basis
-of a dataset that includes only the scorelines at the event itself,
-that the scoreline of the Candidates Tournament (Cura\c{c}ao 1962)
-exhibits evidence of collusion ($p<10^{-5}$), in agreement with
-previous work that uses additional information such as historical
-results including number of moves.}
+In inference problems where the dataset comprises Bernoulli outcomes
+of paired comparisons, the Bradley-Terry model offers a simple and
+easily interpreted framework.  However, it does not deal easily with
+chess because of the existence of draws, and the white player
+advantage.  Here I present a new generalization of Bradley-Terry in
+which a chess game is regarded as a three-way competition between the
+two players and an entity that wins if the game is drawn.
+Bradley-Terry is then further generalized to account for the white
+player advantage by positing an entity whose strength is added to that
+of the white player.  These techniques afford insight into players'
+strengths, response to playing black or white, and risk-aversion as
+manifested by probability of drawing.  The likelihood functions
+arising are easily optimised numerically.  I analyse a number of
+datasets of chess results, including the infamous World Chess
+Championships 1963 in which Fischer accused three Soviet players of
+collusion.  I conclude, on the basis of a dataset that includes only
+the scorelines at the event itself, that the Candidates Tournament
+(Cura\c{c}ao 1962) exhibits evidence of collusion ($p<10^{-5}$), in
+agreement with previous work that uses additional information such as
+historical results including number of moves.}
 
@@
 competitive situations such as tennis in which one or other player is
-idenfied as the winner, but the system has been applied to many
-non-competitive situations such mating preference~\citep{west2008} and
-consumer behaviour.  The Bradley-Terry model~\citep{bradley1952}
-considers $n$ players and assigns non-negative numbers
+identified as the winner.  The Bradley-Terry model~\citep{bradley1952}
+considers~$n$ players and assigns non-negative numbers
 $p_1,\ldots,p_n$ with unit sum to each player.  Then player $i$ beats
@@
 has been modified and extended many times, for example to account for
-multi-way choices such as mating preferences~\citep{hankin2010}.
+multi-way competitions such as {\em MasterChef
+  Australia}~\citep{hankin2017}, and the system has been applied to
+many non-competitive situations such as consumer
+behaviour~\citep{hankin2010} and mating preference~\citep{west2008}.
 
@@
 
-Such arguments suggest considering a drawn chess game as a
-qualitatively different outcome from a statement of equality of skill:
-the play itself tends to be different, the drawing players have a
-different mindset, and drawn games have less value in the wider
-context of the game.
+Such arguments suggest that one should infer more from a drawn game
+than simple equality of skill: the play itself tends to be different
+and more cautious, the drawing players have a different mindset, and
+drawn games have less value in the wider context of the game.
 
@@
 the number of times $i$ loses to $j$.  Here I present a new
-generalization of Bradley-Terry in which a chess game is regarded as
-a three-way competition between the two players and an entity that
-wins if the game is drawn.  The system is further modified to account
-for first-move advantage by positing an entity whose strength is added
-to that of the white player.  This methodology affords insight into
+generalization of Bradley-Terry in which a chess game is regarded as a
+three-way competition between the two players and an entity that wins
+if the game is drawn.  The system is further modified to account for
+first-move advantage by positing an entity whose strength is added to
+that of the white player.  This methodology affords insight into
 players' strengths, response to playing black or white, and
-risk-aversion as manifested by probability of drawing.
-
-If we consider a chess game to be a three-way trial, we define a
-likelihood function for non-negative $p_1,\ldots,p_n,D$ with
-$D+\sum p_i=1$ and this would be
+risk-aversion as manifested by probability of drawing.  If we consider
+a chess game to be a three-way trial, an appropriate likelihood
+function for non-negative $p_1,\ldots,p_n,D$ with $D+\sum p_i=1$ would
+be
 
@@
 the white player.  The strength of the white player is thus the sum of
-the player's strength and another entity, the White Monster.
-Recalling that a game might be a draw, we obtain the probabilities as
-given in Table~\ref{probabilities}.  The corresponding likelihood
-function is given below in equation \ref{likelihood_WD}:
+the player's strength and another entity with strength $W$.  Recalling
+that a game might be a draw, we obtain the probabilities as given in
+Table~\ref{probabilities}.  The corresponding likelihood function is
+given below in equation \ref{likelihood_WD}:
 
@@
 \end{tabular}
-\caption{Probabilities of a win, draw, loss for the two players under a generalized Bradley-Terry model\label{probabilities}}
+\caption{Probabilities of a win, draw, loss for the two players under a generalized
+  Bradley-Terry model\label{probabilities}}
 \end{table}
@@
 which might be preferable on grounds of clarity.  We now apply this
-and related likelihood models to real chess datasets.  The general
-approach follows~\citet{edwards1972}.  To test any hypothesis $H$ on a
+and related likelihood models to real chess datasets using the Method
+of Support~\citet{edwards1972}.  To test any hypothesis $H$ on
 likelihood function $\mathcal{L}$ one first identifies the constraint
@@
 
-This likelihood function is easily maximized numerically
-and this is shown visually in figure~\ref{maxp_karpov_kasparov_anand},
-which suggests a number of interesting and testable statistical
-hypotheses.  Firstly, we observe that $D=0$ cannot possibly be the
-case: the likelihood for this hypothesis would be identically zero,
-owing to the nonzero term in the numerator of the last term of
-Equation~\ref{likelihood_WD}.  However, we see that the maximum
-likelihood estimate for draw strength $\hat{D}\simeq 0.53$: compare
-the na\"ive estimate of $\frac{224}{365}\simeq 0.61$ given by
-observing 224 draws among 265 games.  The MLE is lower because the
-maximization process automatically adjusts for the different players'
-strengths.  Secondly, we may make inferences about $W$, the extra
-strength conferred by playing white: it is conceivable that $W=0$ as
-this term does not appear on its own in the numerator of any term.  We
-may test $H_0\colon W=0$ using the method of support by maximizing the
+This likelihood function is easily maximized numerically and this is
+shown visually in figure~\ref{maxp_karpov_kasparov_anand}, which
+suggests a number of interesting and testable statistical hypotheses.
+Firstly, we observe that $D=0$ cannot possibly be the case: the
+likelihood for this hypothesis would be identically zero, owing to the
+nonzero term in the numerator of equation~\ref{kka_likelihood}.  From
+the figure, we see that the maximum likelihood estimate for draw
+strength $\hat{D}\simeq 0.53$: compare the na\"ive estimate of
+$\frac{224}{365}\simeq 0.61$ given by observing 224 draws among 265
+games.  The MLE is lower because the maximization process
+automatically adjusts for the different players' strengths.  Secondly,
+we may make inferences about $W$, the extra strength conferred by
+playing white: it is conceivable that $W=0$ as this term does not
+appear on its own in the numerator of any term.  We may test
+$H_0\colon W=0$ using the Method of Support by maximizing the
 likelihood subject the constraint $W=0$ and comparing with the
@@
 
-The method of support allows one to estimate proclivity to draw while
+The Method of Support allows one to estimate proclivity to draw while
 accounting for players' strengths and first-move advantage.  We now
 consider one famous incident, following the World Chess Championships
-1963, in which Fischer famously and publicly claimed that the Soviet
-players colluded.  The accusation was that Soviet players agreed ahead
-of time to draw when playing one another, in order to conserve energy
-for subsequent matches.
+1963, in which Fischer famously and publicly claimed that three of the
+Soviet players colluded.  The accusation was that Keres, Petrosian,
+and Geller agreed ahead of time to draw when playing one another, in
+order to conserve energy for subsequent matches.
 
@@
 
-Here I analyse the original subject of the complaint, the Cura\c{c}ao
-1962 Candidates (match results shown in Table~\ref{candidates}).
-Fischer alleged that three of the Soviet players (Keres, Petrosian,
-and Geller) had agreed to draw amongst themselves.  Observe that all
-twelve of the games between the three accused players ended in a draw,
-a significant result given 51 winning games and 54 draws (Fisher exact
-test, one-sided, $p=0.0079$).
-
-To investigate the possibility of Soviet collusion we introduce a new
-draw entity: a soviet draw monster $C$, who wins when two colluding
-players draw, and is inactive whenever a game has at least one
-non-colluding player.  Then $H_N\colon C=D$ corresponds to no
-collusion and the likelihood function would be
+Here I consider only the game results, reproduced in
+table~\ref{candidates}.  We see that all twelve of the games between
+the three accused players ended in a draw, a significant result given
+51 winning games and 54 draws (Fisher exact test, one-sided,
+$p=0.0079$).  To quantify such Soviet collusion, accounting for the
+player skills, we introduce a new draw entity: a collusive draw
+monster $C$, who wins when two colluding players draw, and is inactive
+whenever a game has at least one non-colluding player.  Then
+$H_N\colon C=D$ corresponds to no collusion and the likelihood
+function would have
 
@@
 draw), and a total of 30 terms such as
-$\left(P+\mathit{Fis}+W+D\right)^{4}$ and
 $\left(P+\mathit{Ke}+W+C\right)^{4}$ [indicating that Petrosian played
@@
 
-The fith interzonal took place in Stockholm in 1962, in the form of a
+The fifth Interzonal took place in Stockholm in 1962, in the form of a
 23-player round-robin tournament.  For the results shown in
@@
 investigate the possibility of increased Soviet drawing proclivity, we
-introduce a Soviet draw monster $D_s$, who wins when two Soviet
-players draw, and is inactive if a game has at least one non-Soviet
-player.  Then $H_N\colon D_s=D$ corresponds to no collusion and the
-likelihood function for table~\ref{interzonal} would be
+introduce a Soviet draw monster with strength~$D_s$, who wins when two
+Soviet players draw, and is inactive if a game has at least one
+non-Soviet player.  Then $H_N\colon D_s=D$ corresponds to no collusion
+and the likelihood function for table~\ref{interzonal} would be
 

### 2019-05-23 inst/collusion_chess_draws.Rnw
various changes suggested by Jeff Nijsse.  Thanks Jeff\!

@@
 Bradley-Terry is then further generalized to account for the white
-player advantage by positing an entity whose strength is added to that
-of the white player.  These techniques afford insight into players'
-strengths, response to playing black or white, and risk-aversion as
-manifested by probability of drawing.  The likelihood functions
-arising are easily optimised numerically.  I analyse a number of
-datasets of chess results, including the infamous World Chess
-Championships 1963 in which Fischer accused three Soviet players of
+player advantage by positing a second entity whose strength is added
+to that of the white player.  These techniques afford insight into
+players' strengths, response to playing black or white, and
+risk-aversion as manifested by probability of drawing.  The likelihood
+functions arising are easily optimised numerically.  I analyse a
+number of datasets of chess results, including the infamous 1963 World
+Chess Championships, in which Fischer accused three Soviet players of
 collusion.  I conclude, on the basis of a dataset that includes only
@@
 Many situations involve pairwise comparisons that amount to a
-Bernoulli trial.  The canonical application would be to two-way
-competitive situations such as tennis in which one or other player is
+Bernoulli trial.  The canonical application would be two-way
+competitive situations such as tennis in which one player is
 identified as the winner.  The Bradley-Terry model~\citep{bradley1952}
@@
 multi-way competitions such as {\em MasterChef
-  Australia}~\citep{hankin2017}, and the system has been applied to
-many non-competitive situations such as consumer
-behaviour~\citep{hankin2010} and mating preference~\citep{west2008}.
+  Australia}~\citep{hankin2017}, and non-competitive situations such
+as consumer behaviour~\citep{hankin2010} and mating
+preference~\citep{west2008}.
 
-Bradley-Terry is not directly applicable to Chess for two reasons:
-firstly, the white player appears to be at a considerable advantage;
-and secondly, many chess games end in a draw: white wins about
-$37.5\%$ of all games, black wins $27.6\%$, and about $39.4\%$ are
-drawn.  We consider the first-mover advantage first, then consider the
-effect of drawn games.
+Bradley-Terry is not directly applicable to chess for two reasons:
+firstly, the white player appears to be at a considerable advantage
+winning about 37.5\% of games (black wins 27.6\%); and secondly, many
+chess games, about 39.4\%, are drawn (figures from Chessbase).
 
@@
 less material to discuss.  Players themselves do not seem to value
-draws in the wider context of the game: \citeauthor{fischer1969}, for
+draws in the wider context of the game: \citet{fischer1969}, for
 example, includes only 8 draws in his compilation of 60 memorable
@@
 incentivised to protect the lead by lowering the risk of losing.  In a
-chess context, this means the leading player will aim for a draw.
+chess context, this means the leading player will be more likely to
+aim for a draw.
 
@@
 Consider first the standard Bradley-Terry model for a finite number of
-chess players.  First we ignore the effect of playing white.  We wish
-to find numbers (``strengths") $p_i$ for $1\leqslant i\leqslant n$
-with $p_i\geqslant 0, \sum p_i=1$ such that the probability of $i$
-beating $j\neq i$ is $\frac{p_i}{p_i+p_j}$.  A suitable likelihood
-function, on the assumption of independence of trials, would be
+chess players.  Ignoring the effect of playing white, we wish to find
+numbers (``strengths") $p_i$ for $1\leqslant i\leqslant n$ with
+$p_i\geqslant 0, \sum p_i=1$ such that the probability of $i$ beating
+$j\neq i$ is $\frac{p_i}{p_i+p_j}$.  A suitable likelihood function,
+on the assumption of independence of trials, would be
 
@@
 generalization of Bradley-Terry in which a chess game is regarded as a
-three-way competition between the two players and an entity that wins
-if the game is drawn.  The system is further modified to account for
-first-move advantage by positing an entity whose strength is added to
-that of the white player.  This methodology affords insight into
-players' strengths, response to playing black or white, and
-risk-aversion as manifested by probability of drawing.  If we consider
-a chess game to be a three-way trial, an appropriate likelihood
-function for non-negative $p_1,\ldots,p_n,D$ with $D+\sum p_i=1$ would
-be
+three-way competition between the two players and a third entity (a
+``draw monster'') that wins if the game is drawn.  The system is
+further modified to account for first-move advantage by positing an
+entity whose strength is added to that of the white player.  This
+methodology affords insight into players' strengths, response to
+playing black or white, and risk-aversion as manifested by probability
+of drawing.  If we consider a chess game to be a three-way trial, an
+appropriate likelihood function would be for non-negative
+$p_1,\ldots,p_n,D$ ($D$ being the draw monster's strength), with
+$D+\sum p_i=1$ and
 
@@
 approach ignores the extra strength conferred by the first-move
-advantage.  We thus consider another entity whose strength is added to
-the white player.  The strength of the white player is thus the sum of
-the player's strength and another entity with strength $W$.  Recalling
-that a game might be a draw, we obtain the probabilities as given in
-Table~\ref{probabilities}.  The corresponding likelihood function is
+advantage.  We thus consider another entity whose strength $W$ is
+added to the white player.  The strength of the white player is thus
+the sum of the player's strength and $W$.  Recalling that a game might
+be a draw, we obtain the probabilities as given in
+table~\ref{probabilities}.  The corresponding likelihood function is
 given below in equation \ref{likelihood_WD}:
@@
 and related likelihood models to real chess datasets using the Method
-of Support~\citet{edwards1972}.  To test any hypothesis $H$ on
+of Support~\citep{edwards1972}.  To test any hypothesis $H$ on
 likelihood function $\mathcal{L}$ one first identifies the constraint
@@
 The above likelihood functions are objective and reasonable
-representations of chess results.  However, they differ markedly from
-previous likelihood functions of draws in that they include reified
-entities representing proclivity to draw $D$, and white or first-mover
-advantage ($W$).
+representations of chess game outcomes.  However, they differ markedly
+from previous likelihood functions of draws in that they include
+reified entities representing proclivity to draw $D$, and white or
+first-mover advantage ($W$).
 
@@
 
-\cite{davidson1970} uses a threshhold approach to interpret a draw as
-performance of the players the same to within a threshhold.  Using a
-logistic expression for the difference in players' performance, he
-effectively uses, in current notation, $D=\nu\sqrt{p_ip_j}$.  However,
-\cite{joe1990} uses likelihood methods to show that this model does
-not fit data well.  Here we note that such a model automatically give
-stronger players a higher draw potential, and also note that it is
-difficult to account for first-player advantage in this framework.
-
-Such an approach is
+\cite{davidson1970} uses a threshhold approach to interpret draws.
+From a chess draw, he infers that the performance of the players is
+identical, within a threshhold.  Using a logistic expression for the
+difference in players' performance he uses, in current notation,
+$D=\nu\sqrt{p_ip_j}$.  However, \cite{joe1990} uses likelihood methods
+to show that this model does not fit data well.  Here we note that
+such a model automatically give stronger players a higher draw
+potential, and also note that it is difficult to account for
+first-player advantage in this framework.  Such an approach is
 numerically challenging in the current framework as the unit-sum
-constraint becomes nonlinear and likelihood maximization becomes
-difficult.
+constraint becomes nonlinear, making likelihood maximization
+difficult.  \cite{moul2009} consider chess draws and use sophisticated
+regression techniques to analyse games in terms of not only the
+result, but also the ply-count, data not used in the present analysis.
 
@@
 optimization (not integration which is computationally expensive).
-The analysis presented here uses the framework of the \pkg{hyper2}
-package~\citep{hankin2017} as a convenient and unified interface.
-Because gradients are available, convergence to the evaluate is
-generally rapid and hypothesis-specific constraints on parameter space
-are, in general, straightforward to implement.
+The analysis presented here uses the framework of the
+R~\citep{rcore2019} \pkg{hyper2} package~\citep{hankin2017} as a
+convenient and unified interface.  Because gradients are available,
+convergence to the evaluate is generally rapid and hypothesis-specific
+constraints on parameter space are, in general, linear and
+straightforward to implement.
 
@@
 
-Grandmasters Anand, Karpov, Kasparov have long-standing rivalry and
-have played many games amongst themselves.  A summary may be found
-in Table~\ref{kka}.  Equation~\ref{likelihood_WD} becomes
+Grandmasters Anand, Karpov, and Kasparov have long-standing rivalry
+and have played many games amongst themselves.  A summary may be found
+in table~\ref{kka}.  Equation~\ref{likelihood_WD} becomes
 
@@
 \caption{Results of 365 chess games \label{kka} between Grandmaster
-  Anand, Grandmaster Karpov, and Grandmaster Kasparov.  Thus, for
-  example, Anand played white against Karpov a total of $18+20+13=51$
-  times and won 18, drew 20, lost 13}
+  Anand, Grandmaster Karpov, and Grandmaster Kasparov.  For example,
+  Anand played white against Karpov a total of $18+20+13=51$ times and
+  won 18, drew 20, lost 13}
 \end{table}
@@
 @
-\caption{Maximum likelihood estimate for the strengths of the 3 Grandmasters in Table~\ref{kka} 
+\caption{Maximum likelihood estimate for the strengths of the 3 Grandmasters in table~\ref{kka} 
 \label{maxp_karpov_kasparov_anand} under a likelihood defined as Equation~\ref{likelihood_WD}}
@@
 
-This likelihood function is easily maximized numerically and this is
-shown visually in figure~\ref{maxp_karpov_kasparov_anand}, which
-suggests a number of interesting and testable statistical hypotheses.
-Firstly, we observe that $D=0$ cannot possibly be the case: the
-likelihood for this hypothesis would be identically zero, owing to the
-nonzero term in the numerator of equation~\ref{kka_likelihood}.  From
-the figure, we see that the maximum likelihood estimate for draw
-strength $\hat{D}\simeq 0.53$: compare the na\"ive estimate of
-$\frac{224}{365}\simeq 0.61$ given by observing 224 draws among 265
-games.  The MLE is lower because the maximization process
-automatically adjusts for the different players' strengths.  Secondly,
-we may make inferences about $W$, the extra strength conferred by
-playing white: it is conceivable that $W=0$ as this term does not
-appear on its own in the numerator of any term.  We may test
-$H_0\colon W=0$ using the Method of Support by maximizing the
-likelihood subject the constraint $W=0$ and comparing with the
-likelihood when the optimization is free.
+This likelihood function is easily maximized numerically and the
+evaluate is shown visually in figure~\ref{maxp_karpov_kasparov_anand},
+which suggests a number of interesting and testable statistical
+hypotheses.  Firstly, we observe that $D=0$ cannot possibly be the
+case: the likelihood for this hypothesis would be identically zero,
+owing to the nonzero term in the numerator of
+equation~\ref{kka_likelihood}.  From the figure, we see that the
+maximum likelihood estimate for draw strength $\hat{D}\simeq 0.53$:
+compare the na\"ive estimate of $\frac{224}{365}\simeq 0.61$ given by
+observing 224 draws among 265 games.  The na\"{i}ve estimate is lower
+because the maximization process automatically adjusts for the
+different players' strengths.  Secondly, we may make inferences about
+$W$, the extra strength conferred by playing white: it is conceivable
+that $W=0$ as this term does not appear on its own in the numerator of
+any term.  We may test $H_0\colon W=0$ using the Method of Support by
+maximizing the likelihood subject to the constraint $W=0$ and
+comparing with the likelihood when the optimization is free.
 
@@
 In figure~\ref{kka_3draws}, we see considerable variability in
-proclivity to draw and again this may be tested with the Method of
+proclivity to draw.  Again this may be tested with the Method of
 Support, which gives a $p$-value of about 0.039, borderline
@@
 accounting for players' strengths and first-move advantage.  We now
-consider one famous incident, following the World Chess Championships
-1963, in which Fischer famously and publicly claimed that three of the
-Soviet players colluded.  The accusation was that Keres, Petrosian,
-and Geller agreed ahead of time to draw when playing one another, in
-order to conserve energy for subsequent matches.
+consider one infamous incident, following the World Chess
+Championships 1963, in which Fischer famously and publicly claimed
+that three of the Soviet players colluded.  The accusation was that
+Keres, Petrosian, and Geller agreed ahead of time to draw when playing
+one another, in order to conserve energy for subsequent matches.
 
@@
 the three accused players ended in a draw, a significant result given
-51 winning games and 54 draws (Fisher exact test, one-sided,
+51 winning games and 54 draws (Fisher's exact test, one-sided,
 $p=0.0079$).  To quantify such Soviet collusion, accounting for the
@@
 
-\noindent on the numerator (where $C$ is the strength of a collusive
+\noindent for the numerator (where $C$ is the strength of a collusive
 draw), and a total of 30 terms such as
 $\left(P+\mathit{Ke}+W+C\right)^{4}$ [indicating that Petrosian played
-  Keres 4 times] on the denominator.  The likelihood ratio for the
+  Keres 4 times] in the denominator.  The likelihood ratio for the
 hypothesis that collusive draws have the same strength as regular
@@
 and plot a profile likelihood curve for the log-odds ratio
-$\ln\left(D_s/D\right)$ which, if Dirichlet, would be accurately
+$\log\left(D_s/D\right)$ which, if Dirichlet, would be accurately
 parabolic over a wide range~\cite[page 343]{ohagan2004}.  This is
-illustrated in Figure~\ref{proflike}, that shows a credible interval
+illustrated in figure~\ref{proflike}, that shows a credible interval
 of about 1.8-2.4.
@@
 
-In this context we note, following \cite{langer2013}, that a player
-may well adopt a riskier strategy after a sequence of defeats,
-adopting a ``nothing-to-lose'' mentality.  Such a phenomenon would be
-detectable, at least in principle, by generalizing the personalised
-draw likelihood function \ref{KKA_personalised_drawmonster} and
-seeking evidence for the draw strength changing as a tournament
-progresses: it would become stronger for a player on a winning streak,
-and weaker for a losing player.
+In this context we note, following the observation of~\citet{moul2009}
+that a player may well adopt a riskier strategy after a sequence of
+defeats, adopting a ``nothing-to-lose'' mentality.  Such a phenomenon
+would be detectable, at least in principle, by generalizing the
+personalised draw likelihood function
+\ref{KKA_personalised_drawmonster} and seeking evidence for the draw
+strength changing as a tournament progresses: it would become stronger
+for a player on a winning streak, and weaker for a losing player.
 

### 2019-05-26 inst/collusion_chess_draws.Rnw
include pointer to github
various corrections suggested by Alna.  Thanks Alna\!

@@
 %% just as usual
-\author{Robin K. S. Hankin\\Auckland University of Technology}
+\author{Robin K. S. Hankin\\Department of Mathematical Sciences\\Auckland University of Technology}
 \title{A generalization of the Bradley-Terry model for draws in chess}
@@
 
-\Keywords{Dirichlet distribution, hyperdirichlet distribution,
-  combinatorics, \proglang{R}, multinomial distribution, constrained
-optimization, Bradley-Terry}
-\Plainkeywords{Dirichlet distribution, hyperdirichlet distribution,
-  combinatorics, R, multinomial distribution, constrained
-optimization, Bradley-Terry}
+\Keywords{Likelihood functions, method of support, \proglang{R},
+  constrained optimization, Bradley-Terry, Chess, Collusion, Chess
+  Draws}
+\Plainkeywords{Likelihood functions, method of support, R,
+  constrained optimization, Bradley-Terry, Chess, Collusion, Chess
+  Draws}
 
@@
 identified as the winner.  The Bradley-Terry model~\citep{bradley1952}
-considers~$n$ players and assigns non-negative numbers
-$p_1,\ldots,p_n$ with unit sum to each player.  Then player $i$ beats
-player $j$ with probability $\frac{p_i}{p_i+p_j}$.  This simple system
-has been modified and extended many times, for example to account for
-multi-way competitions such as {\em MasterChef
-  Australia}~\citep{hankin2017}, and non-competitive situations such
-as consumer behaviour~\citep{hankin2010} and mating
-preference~\citep{west2008}.
+considers~$n$ players and assigns non-negative number $p_i$ to player
+$i$, with $\sum p_i=1$.  Then player $i$ beats player $j$ with
+probability $\frac{p_i}{p_i+p_j}$.  This simple system has been
+modified and extended many times, for example to account for multi-way
+competitions such as {\em MasterChef Australia}~\citep{hankin2017},
+and non-competitive situations such as consumer
+behaviour~\citep{hankin2010} and mating preference~\citep{west2008}.
 
@@
 firstly, the white player appears to be at a considerable advantage
-winning about 37.5\% of games (black wins 27.6\%); and secondly, many
-chess games, about 39.4\%, are drawn (figures from Chessbase).
+winning about 37.5\% of tournament games (black wins 27.6\%); and
+secondly, many chess games, about 39.4\%, are drawn (figures from
+Chessbase).
 
-\paragraph{First-mover advantage} Traditionally White moves first in
-any game, and this feature introduces an asymmetry into play.
-Historically, we see about 1.35 white wins for every black win.
-However, interpretations of this statistic vary and the issue is hotly
-debated with opinions ranging from that of~\cite{adams2007}---``white
-to play and win''---to~\citet{adorjan1988}---``black is OK!''.
-
-\paragraph{Draws} Chess games may be won by either player, or may be
+\subsection{Draws} Chess games may be won by either player, or may be
 drawn; the implication of a draw is that neither player is better than
@@
 
-\section{Generalized Bradley-Terry}
+\subsection{First-mover advantage} Traditionally White moves first in
+any game, and this feature introduces an asymmetry into play.
+Historically, we see about 1.35 white wins for every black win.
+However, interpretations of this statistic vary and the issue is hotly
+debated with opinions ranging from that of~\cite{adams2007}---``white
+to play and win''---to~\citet{adorjan1988}---``black is OK!''.
 
-Consider first the standard Bradley-Terry model for a finite number of
-chess players.  Ignoring the effect of playing white, we wish to find
-numbers (``strengths") $p_i$ for $1\leqslant i\leqslant n$ with
-$p_i\geqslant 0, \sum p_i=1$ such that the probability of $i$ beating
-$j\neq i$ is $\frac{p_i}{p_i+p_j}$.  A suitable likelihood function,
-on the assumption of independence of trials, would be
+\subsection{Generalized Bradley-Terry}
+
+Here I show how Bradley-Terry may be generalized ot account for the
+existence of draws and the first-mover advantage.  Consider first the
+standard Bradley-Terry model for a finite number of chess players.
+Ignoring the effect of playing white, we wish to find numbers
+(``strengths") $p_i$ for $1\leqslant i\leqslant n$ with $p_i\geqslant
+0, \sum p_i=1$ such that the probability of $i$ beating $j\neq i$ is
+$\frac{p_i}{p_i+p_j}$.  A suitable likelihood function, on the
+assumption of independence of trials, would be
 
@@
 \prod_{i<j}
-\left(\frac{p_D}{p_i+p_j+D}\right)^{D_{ij}^\mathrm{white}+D_{ij}^\mathrm{black}}
+\left(\frac{D}{p_i+p_j+D}\right)^{D_{ij}^\mathrm{white}+D_{ij}^\mathrm{black}}
 \end{equation}
@@
 \citet{moul2009} consider this accusation from a statistical
-perspective and present a detailed analysis.  Unlike the present work,
-which uses a dataset restricted to game result (win/loss/draw),
-\citet{moul2009} included number of moves in each game\footnote{The
-  reasoning was that hard-fought games that happened to end in a draw
-  are more likely to last longer than collusive or pre-arranged
-  draws.} in their analysis, and used a dataset of 7355 games played
-in the period 1940-1978 to calibrate the ability of the players.  In
-contrast, the analysis presented uses a dataset restricted to game
-outcome (win-loss-draw) in the competition itself.
+perspective and present a detailed analysis based on number of moves
+in each game\footnote{The reasoning was that hard-fought games that
+  happened to end in a draw are more likely to last longer than
+  collusive or pre-arranged draws.}.  Their analysis incorporated a
+dataset of 7355 games played in the period 1940-1978, used to
+calibrate the ability of the players.  In contrast, the analysis
+presented here uses only outcome of games (win-loss-draw) in the
+competition itself.
 
@@
 
-Here I consider only the game results, reproduced in
-table~\ref{candidates}.  We see that all twelve of the games between
+Here I consider game results from Cura\c{c}ao 1962, reproduced in
+Table~\ref{candidates}.  We see that all twelve of the games between
 the three accused players ended in a draw, a significant result given
@@
 
+\subsection*{Supplementary material}
+
+R code for all statistical analysis given here is available at
+\url{https://github.com/RobinHankin/hyper2}.
+
 \bibliography{chess}

### 2019-05-27 inst/collusion_chess_draws.Rnw
more work, give ghosts a fraktur font

@@
    \smash[b]{\begin{tabular}[t]{@{}c@{}} #1 \end{tabular}}}
+\newcommand{\draw}{\ensuremath{\mathfrak{D}}}
+\newcommand{\cdraw}{\ensuremath{\mathfrak{C}}}
+\newcommand{\white}{\ensuremath{\mathfrak{W}}}
 
@@
 However, interpretations of this statistic vary and the issue is hotly
-debated with opinions ranging from that of~\cite{adams2007}---``white
-to play and win''---to~\citet{adorjan1988}---``black is OK!''.
+debated with opinions ranging from that of~\cite{adams2007}---``White
+to play and win'' to~\citet{adorjan1988}---``Black is OK!''.
 
@@
 
-Here I show how Bradley-Terry may be generalized ot account for the
+Here I show how Bradley-Terry may be generalized to account for the
 existence of draws and the first-mover advantage.  Consider first the
@@
 appropriate likelihood function would be for non-negative
-$p_1,\ldots,p_n,D$ ($D$ being the draw monster's strength), with
-$D+\sum p_i=1$ and
+$p_1,\ldots,p_n,\draw$ (here \draw\ is the draw monster's strength,
+fraktur font being used for reified entities' strengths) with $\draw+\sum
+p_i=1$ and
 
@@
 \prod_{i<j}
-\left(\frac{p_i}{p_i+p_j+D}\right)^{W_{ij}}
-\left(\frac{p_j}{p_i+p_j+D}\right)^{L_{ij}}
-\left(\frac{D  }{p_i+p_j+D}\right)^{D_{ij}}
+\left(\frac{p_i}{p_i+p_j+\draw}\right)^{W_{ij}}
+\left(\frac{p_j}{p_i+p_j+\draw}\right)^{L_{ij}}
+\left(\frac{\draw  }{p_i+p_j+\draw}\right)^{D_{ij}}
 \end{equation}
@@
 approach ignores the extra strength conferred by the first-move
-advantage.  We thus consider another entity whose strength $W$ is
+advantage.  We thus consider another entity whose strength \white\ is
 added to the white player.  The strength of the white player is thus
-the sum of the player's strength and $W$.  Recalling that a game might
-be a draw, we obtain the probabilities as given in
-table~\ref{probabilities}.  The corresponding likelihood function is
+the sum of the player's strength and \white.  Recalling that a game
+might be a draw, we obtain the probabilities as given in
+Table~\ref{probabilities}.  The corresponding likelihood function is
 given below in equation \ref{likelihood_WD}:
@@
   &$i$ plays white & $j$ plays white\\ \hline
-$\Prob{(\mbox{$i$ wins})}$  & $\frac{p_i+W\oo}{p_i+p_j+W+D}$ & $\frac{p_i}{p_i+p_j+W+D}$  \\
-$\Prob{(\mbox{draw})}$      & $\frac{D\oo}{p_i+p_j+W+D}$     & $\frac{D}{p_i+p_j+W+D}$    \\
-$\Prob{(\mbox{$i$ loses})}$ & $\frac{p_j\oo}{p_i+p_j+W+D}$   & $\frac{p_j+W}{p_i+p_j+W+D}$\\
+$\Prob{(\mbox{$i$ wins})}$  & $\frac{p_i+\white\oo}{p_i+p_j+\white+\draw}$ & $\frac{p_i}{p_i+p_j+\white+\draw}$  \\
+$\Prob{(\mbox{draw})}$      & $\frac{\draw\oo}{p_i+p_j+\white+\draw}$     & $\frac{\draw}{p_i+p_j+\white+\draw}$    \\
+$\Prob{(\mbox{$i$ loses})}$ & $\frac{p_j\oo}{p_i+p_j+\white+\draw}$   & $\frac{p_j+\white}{p_i+p_j+\white+\draw}$\\
 \end{tabular}
@@
 \prod_{i<j}&{}
-\left(\frac{p_i+W}{p_i+p_j+D+W}  \right)^{W_{ij}^\mathrm{white}}
-\left(\frac{p_j}{p_i+p_j+D+W    }\right)^{L_{ij}^\mathrm{white}}\\
-&{} \left(\frac{p_i}{p_i+p_j+D+W}\right)^{W_{ij}^\mathrm{black}}\label{likelihood_WD}
-\left(\frac{p_j+W}{p_i+p_j+D+W  }\right)^{L_{ij}^\mathrm{black}}\\
-&{} \left(\frac{D}{p_i+p_j+D    }\right)^{D_{ij}}
+\left(\frac{p_i+\white}{p_i+p_j+\draw+\white}  \right)^{W_{ij}^\mathrm{white}}
+\left(\frac{p_j}{p_i+p_j+\draw+\white    }\right)^{L_{ij}^\mathrm{white}}\\
+&{} \left(\frac{p_i}{p_i+p_j+\draw+\white}\right)^{W_{ij}^\mathrm{black}}\label{likelihood_WD}
+\left(\frac{p_j+\white}{p_i+p_j+\draw+\white  }\right)^{L_{ij}^\mathrm{black}}\\
+&{} \left(\frac{\draw}{p_i+p_j+\draw    }\right)^{D_{ij}}
 \nonumber
@@
 \prod_{i<j}
-\left(\frac{D}{p_i+p_j+D+W}\right)^{D_{ij}^\mathrm{white}}
-\left(\frac{D}{p_i+p_j+D+W}\right)^{D_{ij}^\mathrm{black}}
+\left(\frac{\draw}{p_i+p_j+\draw+\white}\right)^{D_{ij}^\mathrm{white}}
+\left(\frac{\draw}{p_i+p_j+\draw+\white}\right)^{D_{ij}^\mathrm{black}}
 =
 \prod_{i<j}
-\left(\frac{D}{p_i+p_j+D}\right)^{D_{ij}^\mathrm{white}+D_{ij}^\mathrm{black}}
+\left(\frac{\draw}{p_i+p_j+\draw}\right)^{D_{ij}^\mathrm{white}+D_{ij}^\mathrm{black}}
 \end{equation}
@@
 from previous likelihood functions of draws in that they include
-reified entities representing proclivity to draw $D$, and white or
-first-mover advantage ($W$).
+reified entities representing proclivity to draw \draw, and white or
+first-mover advantage \white.
 
@@
 difference in players' performance he uses, in current notation,
-$D=\nu\sqrt{p_ip_j}$.  However, \cite{joe1990} uses likelihood methods
+$\draw=\nu\sqrt{p_ip_j}$.  However, \cite{joe1990} uses likelihood methods
 to show that this model does not fit data well.  Here we note that
@@
 \begin{equation}\label{kka_likelihood}
-\mathcal{L}\left(p_\mathrm{An},p_\mathrm{Kr},p_\mathrm{Ks},W,D\right)=
+%\mathcal{L}\left(p_\mathrm{An},p_\mathrm{Kr},p_\mathrm{Ks},\white,\draw\right)=
 \frac{
@@
 p_\mathrm{Ks}^{20}
-D^{224}
-\left(p_\mathrm{Kr} + W\right)^{25}
-\left(p_\mathrm{Ks} + W\right)^{45}
-\left(p_\mathrm{An} + W\right)^{24}
+\draw^{224}
+\left(p_\mathrm{Kr} + \white\right)^{25}
+\left(p_\mathrm{Ks} + \white\right)^{45}
+\left(p_\mathrm{An} + \white\right)^{24}
 }{
-\left(p_\mathrm{Kr} + p_\mathrm{Ks} + W + D\right)^{193}
-\left(p_\mathrm{Kr} + p_\mathrm{An} + W + D\right)^{92}
-\left(p_\mathrm{Ks} + p_\mathrm{An} + W + D\right)^{80}
-}
+\left(p_\mathrm{Kr} + p_\mathrm{Ks} + \white + \draw\right)^{193}
+\left(p_\mathrm{Kr} + p_\mathrm{An} + \white + \draw\right)^{92}
+\left(p_\mathrm{Ks} + p_\mathrm{An} + \white + \draw\right)^{80}
+}.
 \end{equation}
@@
 which suggests a number of interesting and testable statistical
-hypotheses.  Firstly, we observe that $D=0$ cannot possibly be the
+hypotheses.  Firstly, we observe that $\draw=0$ cannot possibly be the
 case: the likelihood for this hypothesis would be identically zero,
@@
 equation~\ref{kka_likelihood}.  From the figure, we see that the
-maximum likelihood estimate for draw strength $\hat{D}\simeq 0.53$:
+maximum likelihood estimate for draw strength $\hat{\draw}\simeq 0.53$:
 compare the na\"ive estimate of $\frac{224}{365}\simeq 0.61$ given by
@@
 different players' strengths.  Secondly, we may make inferences about
-$W$, the extra strength conferred by playing white: it is conceivable
-that $W=0$ as this term does not appear on its own in the numerator of
-any term.  We may test $H_0\colon W=0$ using the Method of Support by
-maximizing the likelihood subject to the constraint $W=0$ and
+$\white$, the extra strength conferred by playing white: it is conceivable
+that $\white=0$ as this term does not appear on its own in the numerator of
+any term.  We may test $H_0\colon \white=0$ using the Method of Support by
+maximizing the likelihood subject to the constraint $\white=0$ and
 comparing with the likelihood when the optimization is free.
@@
 \citeauthor{edwards1972}; we can convincingly reject the hypothesis
-that $W=0$ and infer that playing white confers a significant
+that $\white=0$ and infer that playing white confers a significant
 advantage.  This is consistent with expectation, but it is now
@@
 players.  Taking the MLE above, and considering a game between say
-Karpov and Anand, the strengths are Karpov=0.064, Anand=0.151, white
-=0.104 and draw =0.53.  Thus we would estimate that Karpov playing
+Karpov and Anand, the strengths are Karpov=0.064, Anand=0.151, \white
+=0.104 and \draw =0.532.  Thus we would estimate that Karpov playing
 white against Kasparov would win with probability
@@
 two {\em personalised} draw entities, one for each real player; a
-drawn game is counted as a win for either of the draw ``ghosts''.  Under
+drawn game is counted as a win for either of the draw monsters.  Under
 these assumptions, table~\ref{kka} translates into the following
@@
 \begin{equation}\label{KKA_personalised_drawmonster}
-\mathcal{L}\left(
-p_\mathrm{An},p_\mathrm{Kr},p_\mathrm{Ks}
-,W,
-d_\mathrm{An},d_\mathrm{Kr},d_\mathrm{Ks}
-\right)=
+%\mathcal{L}\left(
+%p_\mathrm{An},p_\mathrm{Kr},p_\mathrm{Ks}
+%,\white,
+%\draw_\mathrm{An},\draw_\mathrm{Kr},\draw_\mathrm{Ks}
+%\right)=
 \frac{
@@
 p_\mathrm{Ks}^{20}
-\left(p_\mathrm{Kr} + W\right)^{25}
-\left(p_\mathrm{Ks} + W\right)^{45} 
-\left(p_\mathrm{An} + W\right)^{24}\\ \rule{10mm}{0mm}
-\left(d_\mathrm{Kr} + d_\mathrm{Ks}\right)^{129}
-\left(d_\mathrm{Kr} + d_\mathrm{An}\right)^{49}
-\left(d_\mathrm{Ks} + d_\mathrm{An}\right)^{46}
+\left(p_\mathrm{Kr} + \white\right)^{25}
+\left(p_\mathrm{Ks} + \white\right)^{45} 
+\left(p_\mathrm{An} + \white\right)^{24}\\ \rule{10mm}{0mm}
+\left(\draw_\mathrm{Kr} + \draw_\mathrm{Ks}\right)^{129}
+\left(\draw_\mathrm{Kr} + \draw_\mathrm{An}\right)^{49}
+\left(\draw_\mathrm{Ks} + \draw_\mathrm{An}\right)^{46}
 $}}{\parbox{3in}{$
-\left(p_\mathrm{Kr} + p_\mathrm{Ks} + W + d_\mathrm{Kr} + d_\mathrm{Ks}\right)^{193}\\ \rule{10mm}{0mm}
-\left(p_\mathrm{Kr} + p_\mathrm{An} + W + d_\mathrm{Kr} + d_\mathrm{An}\right)^{92}\\ \rule{20mm}{0mm}
-\left(p_\mathrm{Ks} + p_\mathrm{An} + W + d_\mathrm{Ks} + d_\mathrm{An}\right)^{80}
+\left(p_\mathrm{Kr} + p_\mathrm{Ks} + \white + \draw_\mathrm{Kr} + \draw_\mathrm{Ks}\right)^{193}\\ \rule{10mm}{0mm}
+\left(p_\mathrm{Kr} + p_\mathrm{An} + \white + \draw_\mathrm{Kr} + \draw_\mathrm{An}\right)^{92}\\ \rule{20mm}{0mm}
+\left(p_\mathrm{Ks} + p_\mathrm{An} + \white + \draw_\mathrm{Ks} + \draw_\mathrm{An}\right)^{80}
 $}}
@@
 
-In figure~\ref{kka_3draws}, we see considerable variability in
+In Figure~\ref{kka_3draws}, we see considerable variability in
 proclivity to draw.  Again this may be tested with the Method of
@@
 
-\subsection{Each player has a distinct White Monster}
+\subsection{Each player has a distinct white monster}
 
 We now consider the possibility that the strength conferred by playing
-white, is specific to each player and we denote this $w_\mathrm{An}$,
-$w_\mathrm{Kr}$ and $w_\mathrm{Ks}$.  The corresponding likelihood
-function would be
+white is specific to each player and we denote the strengths
+$\white_\mathrm{An}$, $\white_\mathrm{Kr}$ and $\white_\mathrm{Ks}$.
+The corresponding likelihood function would be
 
@@
 %p_\mathrm{An},p_\mathrm{Kr},p_\mathrm{Ks},
-%w_\mathrm{An},w_\mathrm{Kr},w_\mathrm{Ks},D
+%\white_\mathrm{An},\white_\mathrm{Kr},\white_\mathrm{Ks},\draw
 %\right)=
@@
 p_\mathrm{Ks}^{20}
-D^{224}
-\left(p_\mathrm{An} + w_\mathrm{An}\right)^{24}%\\ \rule{10mm}{1mm}
-\left(p_\mathrm{Kr} + w_\mathrm{Kr}\right)^{25}
-\left(p_\mathrm{Ks} + w_\mathrm{Ks}\right)^{45}
+\draw^{224}
+\left(p_\mathrm{An} + \white_\mathrm{An}\right)^{24}%\\ \rule{10mm}{1mm}
+\left(p_\mathrm{Kr} + \white_\mathrm{Kr}\right)^{25}
+\left(p_\mathrm{Ks} + \white_\mathrm{Ks}\right)^{45}
 %$}
 }{\parbox{4in}{$
-\left(p_\mathrm{An} + p_\mathrm{Kr} + w_\mathrm{An} +  D\right)^{43}
-\left(p_\mathrm{An} + p_\mathrm{Kr} + w_\mathrm{Kr} +  D\right)^{49}\\ \rule{10mm}{0mm}
-\left(p_\mathrm{An} + p_\mathrm{Ks} + w_\mathrm{An} +  D\right)^{37}
-\left(p_\mathrm{An} + p_\mathrm{Ks} + w_\mathrm{Ks} +  D\right)^{43}\\ \rule{20mm}{0mm}
-\left(p_\mathrm{Kr} + p_\mathrm{Ks} + w_\mathrm{Kr} +  D\right)^{99}
-\left(p_\mathrm{Kr} + p_\mathrm{Ks} + w_\mathrm{Ks} +  D\right)^{94}
+\left(p_\mathrm{An} + p_\mathrm{Kr} + \white_\mathrm{An} +  \draw\right)^{43}
+\left(p_\mathrm{An} + p_\mathrm{Kr} + \white_\mathrm{Kr} +  \draw\right)^{49}\\ \rule{10mm}{0mm}
+\left(p_\mathrm{An} + p_\mathrm{Ks} + \white_\mathrm{An} +  \draw\right)^{37}
+\left(p_\mathrm{An} + p_\mathrm{Ks} + \white_\mathrm{Ks} +  \draw\right)^{43}\\ \rule{20mm}{0mm}
+\left(p_\mathrm{Kr} + p_\mathrm{Ks} + \white_\mathrm{Kr} +  \draw\right)^{99}
+\left(p_\mathrm{Kr} + p_\mathrm{Ks} + \white_\mathrm{Ks} +  \draw\right)^{94}
 $}}
@@
 In the likelihood function above we see, for example, that Anand won a
-total of $18+6=24$ games playing white, $13+2=15$ games playing
-Black, and that Anand played white against Karpov a total of
-$18+20+5=43$ times.  We fail to reject the null of
-$w_\mathrm{An}=w_\mathrm{Kr}=w_\mathrm{Ks}$, having a likelihood ratio
-of $4.5$, or an asymptotic $p$-value of $0.076$.  There is no strong
-evidence that the players differ in regard to their white advantage.
+total of $18+6=24$ games playing white, $13+2=15$ games playing Black,
+and that Anand played white against Karpov a total of $18+20+5=43$
+times.  We fail to reject the null of
+$\white_\mathrm{An}=\white_\mathrm{Kr}=\white_\mathrm{Ks}$, having a
+likelihood ratio of $4.5$, or an asymptotic $p$-value of $0.076$.
+There is no strong evidence that the players differ in regard to their
+white advantage.
 
@@
 player skills, we introduce a new draw entity: a collusive draw
-monster $C$, who wins when two colluding players draw, and is inactive
-whenever a game has at least one non-colluding player.  Then
-$H_N\colon C=D$ corresponds to no collusion and the likelihood
-function would have
+monster with strength \cdraw, who wins when two colluding players
+draw, and is inactive whenever a game has at least one non-colluding
+player.  Then $H_N\colon\cdraw=\draw$ corresponds to no collusion and
+the likelihood function would have
 
@@
 \mathit{Ko}^5
-B
+B\,
 T^2  % Tal
 \mathit{Fil}\,
-D^{42}C^{12}\oo\\ \rule{10mm}{0mm}
-\left(P+W\right)^3
-\left(\mathit{Ke}+W\right)^4
-\left(G+W\right)^4\
-\left(\mathit{Fis}+W\right)^7\oo\\ \rule{20mm}{0mm}
-\left(\mathit{Ko}+W\right)^2
-\left(B+W\right)^5
-\left(T+W\right)
-\left(\mathit{Fil}+W\right)\oo
+\draw^{42}\,\cdraw^{12}\oo\\ \rule{10mm}{0mm}
+\left(P+\white\right)^3
+\left(\mathit{Ke}+\white\right)^4
+\left(G+\white\right)^4\
+\left(\mathit{Fis}+\white\right)^7\oo\\ \rule{20mm}{0mm}
+\left(\mathit{Ko}+\white\right)^2
+\left(B+\white\right)^5
+\left(T+\white\right)
+\left(\mathit{Fil}+\white\right)\oo
 $}\end{equation}
 
-\noindent for the numerator (where $C$ is the strength of a collusive
+\noindent for the numerator (where \cdraw\ is the strength of a collusive
 draw), and a total of 30 terms such as
-$\left(P+\mathit{Ke}+W+C\right)^{4}$ [indicating that Petrosian played
+$\left(P+\mathit{Ke}+\white+\cdraw\right)^{4}$ [indicating that Petrosian played
   Keres 4 times] in the denominator.  The likelihood ratio for the
@@
 
-\subsection{Chess collusion: interzonal Stockholm 1962}
+\subsection{Soviet playing style in the interzonal Stockholm 1962}
 
@@
 23-player round-robin tournament.  For the results shown in
-table~\ref{interzonal}, we would have a likelihood function
+Table~\ref{interzonal}, we would have a likelihood function
 
 \begin{equation}
-\mathcal{L}\left(F,G,P,K,L,W,D\right)=
+%\mathcal{L}\left(F,G,P,K,L,\white,\draw\right)=
 %                D/(F+G+W+D)  (F+P+D)/(F+P+D+W)   (F+W)/(F+W+D+K)   D/(F+W+L+D)
@@
 \frac{
-  D^8(F+W)(K+W)
+  \draw^8(F+\white)(K+\white)
 }{\parbox{4in}{$
-    (F+G+W+D)(F+P+W+D)\\ \rule{10mm}{0mm}(F+K+W+D)  (F+L+W+D)\\ \rule{20mm}{0mm}
-         (G+P+W+D)  (G+K+W+D)\\ \rule{30mm}{0mm}(G+F+W+D)
-                    (P+K+W+D)\\ \rule{40mm}{0mm}(P+L+W+D)
-                               (K+L+W+D)
+    (F+G+\white+\draw)(F+P+\white+\draw)\\ \rule{10mm}{0mm}(F+K+\white+\draw)  (F+L+\white+\draw)\\ \rule{20mm}{0mm}
+         (G+P+\white+\draw)  (G+K+\white+\draw)\\ \rule{30mm}{0mm}(G+F+\white+\draw)
+                    (P+K+\white+\draw)\\ \rule{40mm}{0mm}(P+L+\white+\draw)
+                               (K+L+\white+\draw)
  $}}
@@
 and one for Korchnoi, both playing white); but the full likelihood
-function for the interzonal has 297 terms, including $D^{110}$.  To
+function for the interzonal has 297 terms, including $\draw^{110}$.  To
 investigate the possibility of increased Soviet drawing proclivity, we
-introduce a Soviet draw monster with strength~$D_s$, who wins when two
+introduce a Soviet draw monster with strength~$\draw_s$, who wins when two
 Soviet players draw, and is inactive if a game has at least one
-non-Soviet player.  Then $H_N\colon D_s=D$ corresponds to no collusion
-and the likelihood function for table~\ref{interzonal} would be
+non-Soviet player.  Then $H_N\colon \draw_s=\draw$ corresponds to no collusion
+and the likelihood function for Table~\ref{interzonal} would be
 
 \begin{equation}\label{collusion}
-\mathcal{L}\left(F,G,P,K,L,W,D,D_s\right)=
+%\mathcal{L}\left(F,G,P,K,L,\white,\draw,\draw_s\right)=
 %                D/(F+G+W+D)  (F+P+D)/(F+P+D+W)   (F+W)/(F+W+D+K)   D/(F+W+L+D)
@@
 \frac{
-  D_s^3D^5(F+W)(K+W)
-}{\parbox{4in}{$
-    (F+G+W+D)(F+P+W+D)\\ \rule{10mm}{0mm}(F+K+W+D)  (F+L+W+D)\\ \rule{20mm}{0mm}
-         (G+P+W+D_s)  (G+K+W+D_s)\\ \rule{30mm}{0mm}(G+F+W+D)
-                    (P+K+W+D_s)\\ \rule{40mm}{0mm}(P+L+W+D)
-                               (K+L+W+D)
+  \draw_s^3\draw^5(F+\white)(K+\white)
+}{\parbox{4in}{$    (F+G+\white+\draw)(F+P+\white+\draw)\\ \rule{10mm}{0mm}(F+K+\white+\draw)  (F+L+\white+\draw)\\ \rule{20mm}{0mm}
+         (G+P+\white+\draw_s)  (G+K+\white+\draw_s)\\ \rule{30mm}{0mm}(G+F+\white+\draw)
+                    (P+K+\white+\draw_s)\\ \rule{40mm}{0mm}(P+L+\white+\draw)
+                               (K+L+\white+\draw)
  $}}
@@
 
-The maximum likelihood estimate is $D\simeq 0.043, D_s=0.354$; a
+The maximum likelihood estimate is $\draw\simeq 0.043, \draw_s=0.354$; a
 constrained maximization gives a likelihood ratio of about 5096, or a
@@
 and plot a profile likelihood curve for the log-odds ratio
-$\log\left(D_s/D\right)$ which, if Dirichlet, would be accurately
+$\log\left(\draw_s/\draw\right)$ which, if Dirichlet, would be accurately
 parabolic over a wide range~\cite[page 343]{ohagan2004}.  This is
@@
 \caption{Profile likelihood \label{proflike} for odds ratio
-  $\log\left(D_s/D\right)$: note the accurately parabolic figure in
+  $\log\left(\draw_s/\draw\right)$: note the accurately parabolic figure in
   the region plotted}

### 2019-05-27 inst/collusion_chess_draws.Rnw
typos

@@
 specific tournament outcomes; further, the analysis uses only game
-outcome, not length.
+outcome, not length.  Many hypotheses of interest to the chess
+community may be analysed straightforwardly.
 
@@
 In principle, the present system of likelihood and reified entities
-would admit such simulations and might be interesting to compare results.
+would admit such simulations and it might be interesting to compare results.
 

### 2019-05-27 inst/collusion_chess_draws.Rnw
minor rewording

@@
 identified as the winner.  The Bradley-Terry model~\citep{bradley1952}
-considers~$n$ players and assigns non-negative number $p_i$ to player
+considers~$n$ players and assigns a non-negative number $p_i$ to player
 $i$, with $\sum p_i=1$.  Then player $i$ beats player $j$ with
@@
 Bradley-Terry is not directly applicable to chess for two reasons:
-firstly, the white player appears to be at a considerable advantage
+firstly, the white player appears to be at a considerable advantage,
 winning about 37.5\% of tournament games (black wins 27.6\%); and
@@
 \subsection{Draws} Chess games may be won by either player, or may be
-drawn; the implication of a draw is that neither player is better than
+drawn; one inference of a draw is that neither player is better than
 the other.  Before 1867, tournament games that were drawn were
@@
 and more cautious, the drawing players have a different mindset, and
-drawn games have less value in the wider context of the game.
+drawn games have less value in the wider context of chess.
 
 \subsection{First-mover advantage} Traditionally White moves first in
-any game, and this feature introduces an asymmetry into play.
+chess, and this feature introduces an asymmetry into play.
 Historically, we see about 1.35 white wins for every black win.
@@
 Ignoring the effect of playing white, we wish to find numbers
-(``strengths") $p_i$ for $1\leqslant i\leqslant n$ with $p_i\geqslant
-0, \sum p_i=1$ such that the probability of $i$ beating $j\neq i$ is
+(``strengths") $p_i\geqslant 0$ for $1\leqslant i\leqslant n$ with
+$\sum p_i=1$ such that the probability of $i$ beating $j\neq i$ is
 $\frac{p_i}{p_i+p_j}$.  A suitable likelihood function, on the
@@
 three-way competition between the two players and a third entity (a
-``draw monster'') that wins if the game is drawn.  The system is
+``draw monster'') that wins if the game is drawn.  The system may be
 further modified to account for first-move advantage by positing an
@@
 playing black or white, and risk-aversion as manifested by probability
-of drawing.  If we consider a chess game to be a three-way trial, an
-appropriate likelihood function would be for non-negative
-$p_1,\ldots,p_n,\draw$ (here \draw\ is the draw monster's strength,
-fraktur font being used for reified entities' strengths) with $\draw+\sum
-p_i=1$ and
+of drawing.  If we consider a chess game to be a three-way trial, we
+seek a likelihood function for non-negative $p_1,\ldots,p_n,\draw$
+(here \draw\ is the draw monster's strength, fraktur font being used
+for reified entities' strengths) with $\draw+\sum p_i=1$.  A suitable
+likelihood function would be
 
@@
 Table~\ref{probabilities}.  The corresponding likelihood function is
-given below in equation \ref{likelihood_WD}:
+defined on $\white+\draw+\sum p_i=1$, given below in equation
+\ref{likelihood_WD}:
 
@@
 \left(\frac{p_j+\white}{p_i+p_j+\draw+\white  }\right)^{L_{ij}^\mathrm{black}}\\
-&{} \left(\frac{\draw}{p_i+p_j+\draw    }\right)^{D_{ij}}
+&{} \left(\frac{\draw }{p_i+p_j+\draw+\white  }\right)^{D_{ij}^\mathrm{white}+D_{ij}^\mathrm{black}}
 \nonumber
@@
 where $W_{ij}^\mathrm{white}$ is the number of games won by $i$ when
-playing white against $j$, and so on.  The final term is algebraically
-equal to
-
-\begin{equation}
-\prod_{i<j}
-\left(\frac{\draw}{p_i+p_j+\draw+\white}\right)^{D_{ij}^\mathrm{white}}
-\left(\frac{\draw}{p_i+p_j+\draw+\white}\right)^{D_{ij}^\mathrm{black}}
-=
-\prod_{i<j}
-\left(\frac{\draw}{p_i+p_j+\draw}\right)^{D_{ij}^\mathrm{white}+D_{ij}^\mathrm{black}}
-\end{equation}
-
-which might be preferable on grounds of clarity.  We now apply this
-and related likelihood models to real chess datasets using the Method
-of Support~\citep{edwards1972}.  To test any hypothesis $H$ on
-likelihood function $\mathcal{L}$ one first identifies the constraint
-on hypothesis space corresponding to $H$.  One then finds the
-unconstrained maximum likelihood $\mathcal{L}_\mathrm{free}$ and
-compares it with the maximum likelihood $\mathcal{L}_\mathrm{cons}$
-when the optimization is constrained by $H$.  The test statistic is
-then the likelihood ratio
+playing white against $j$, and so on.  Note that the power of the
+final term, $D_{ij}^\mathrm{white}+D_{ij}^\mathrm{black}$ is the total
+number of games drawn between $i$ and $j$ irrespective of colour.  We
+now apply this and related likelihood models to real chess datasets
+using the Method of Support~\citep{edwards1972}.  To test any
+hypothesis $H$ on likelihood function $\mathcal{L}$ one first
+identifies the constraint on hypothesis space corresponding to $H$.
+One then finds the unconstrained maximum likelihood
+$\mathcal{L}_\mathrm{free}$ and compares it with the maximum
+likelihood $\mathcal{L}_\mathrm{cons}$ when the optimization is
+constrained by $H$.  The test statistic is then the likelihood ratio
 $\mathcal{L}_\mathrm{free}/\mathcal{L}_\mathrm{cons}$ or, equivalently
@@
 constraints on parameter space are, in general, linear and
-straightforward to implement.
+straightforward to implement.  The total runtime for all the tests
+presented here is less than a few seconds on a desktop PC.
 
@@
 and have played many games amongst themselves.  A summary may be found
-in table~\ref{kka}.  Equation~\ref{likelihood_WD} becomes
+in Table~\ref{kka}.  Equation~\ref{likelihood_WD} becomes
 
@@
 @
-\caption{Maximum likelihood estimate for the strengths of the 3 Grandmasters in table~\ref{kka} 
+\caption{Maximum likelihood estimate for the strengths of the 3 Grandmasters in Table~\ref{kka} 
 \label{maxp_karpov_kasparov_anand} under a likelihood defined as Equation~\ref{likelihood_WD}}
@@
 This likelihood function is easily maximized numerically and the
-evaluate is shown visually in figure~\ref{maxp_karpov_kasparov_anand},
+evaluate is shown visually in Figure~\ref{maxp_karpov_kasparov_anand},
 which suggests a number of interesting and testable statistical
@@
 case: the likelihood for this hypothesis would be identically zero,
-owing to the nonzero term in the numerator of
+owing to the nonzero power of \draw\ in the numerator of
 equation~\ref{kka_likelihood}.  From the figure, we see that the
-maximum likelihood estimate for draw strength $\hat{\draw}\simeq 0.53$:
-compare the na\"ive estimate of $\frac{224}{365}\simeq 0.61$ given by
-observing 224 draws among 265 games.  The na\"{i}ve estimate is lower
-because the maximization process automatically adjusts for the
+maximum likelihood estimate for draw strength $\hat{\draw}\simeq
+0.53$: compare the na\"ive estimate of $\frac{224}{365}\simeq 0.61$
+given by observing 224 draws among 265 games.  The ML estimate is
+lower because the maximization process automatically adjusts for the
 different players' strengths.  Secondly, we may make inferences about
-$\white$, the extra strength conferred by playing white: it is conceivable
-that $\white=0$ as this term does not appear on its own in the numerator of
-any term.  We may test $H_0\colon \white=0$ using the Method of Support by
-maximizing the likelihood subject to the constraint $\white=0$ and
-comparing with the likelihood when the optimization is free.
+$\white$, the extra strength conferred by playing white: it is
+conceivable that $\white=0$ as this term does not appear on its own in
+the numerator of any term.  We may test $H_0\colon \white=0$ using the
+Method of Support by maximizing the likelihood subject to the
+constraint $\white=0$ and comparing with the likelihood when the
+optimization is free.
 
@@
 As argued above, players with a cautious style tend to draw and this
-is an important and interesting component of the game.  To investigate
-this we need a likelihood function that can account for possible
-differences in draw proclivity among the players.  This may be
-achieved by considering a game between two players to be a contest
-between five entities: the two real players, the white advantage, and
-two {\em personalised} draw entities, one for each real player; a
-drawn game is counted as a win for either of the draw monsters.  Under
-these assumptions, table~\ref{kka} translates into the following
-likelihood function:
+is an important and interesting feature of the game (even if drawn
+games themselves are boring).  One line of investigation might be a
+likelihood function that can account for differences in draw
+proclivity among the players.  This may be achieved by considering a
+game between two players to be a contest between five entities: the
+two real players, the white advantage, and two {\em personalised} draw
+entities, one for each real player; a drawn game is counted as a win
+for either of the draw monsters.  Under these assumptions,
+Table~\ref{kka} translates into the following likelihood function:
 
@@
 51 winning games and 54 draws (Fisher's exact test, one-sided,
-$p=0.0079$).  To quantify such Soviet collusion, accounting for the
-player skills, we introduce a new draw entity: a collusive draw
-monster with strength \cdraw, who wins when two colluding players
-draw, and is inactive whenever a game has at least one non-colluding
-player.  Then $H_N\colon\cdraw=\draw$ corresponds to no collusion and
-the likelihood function would have
+$p=0.0079$).  To quantify the evidence for any such collusion,
+accounting for the player skills, we introduce a new draw entity: a
+collusive draw monster with strength \cdraw, who wins when two
+colluding players draw, and is inactive whenever a game has at least
+one non-colluding player.  Then $H_N\colon\cdraw=\draw$ corresponds to
+no collusion and the likelihood function would have
 
@@
 \end{tabular}
-\caption{Subset of the Fifth Interzonal round-robin, Stockholm 1962.\label{interzonal}
-  Results $\in\{0,\halfw,1\}$ and box colour refer to row
-  player.  Thus, for example, Fischer played black against Geller and drew}
+\caption{Subset of the Fifth Interzonal round-robin, Stockholm
+  1962.\label{interzonal} Results $\in\{0,\halfw,1\}$ and box colour
+  refer to row player.  Thus, for example, Fischer played black
+  against Geller and drew.  The full dataset, not shown here, has 23
+  players}
 \end{table}
@@
 The fifth Interzonal took place in Stockholm in 1962, in the form of a
-23-player round-robin tournament.  For the results shown in
+23-player round-robin tournament.  For the 20 game results shown in
 Table~\ref{interzonal}, we would have a likelihood function
@@
 and one for Korchnoi, both playing white); but the full likelihood
-function for the interzonal has 297 terms, including $\draw^{110}$.  To
-investigate the possibility of increased Soviet drawing proclivity, we
-introduce a Soviet draw monster with strength~$\draw_s$, who wins when two
-Soviet players draw, and is inactive if a game has at least one
-non-Soviet player.  Then $H_N\colon \draw_s=\draw$ corresponds to no collusion
-and the likelihood function for Table~\ref{interzonal} would be
+function for the interzonal has 297 terms, including $\draw^{110}$.
+To investigate the possibility of increased Soviet drawing proclivity,
+we introduce a Soviet draw monster with strength~$\draw_s$, who wins
+when two Soviet players draw, and is inactive if a game has at least
+one non-Soviet player.  Then $H_N\colon \draw_s=\draw$ corresponds to
+no collusion and the likelihood function for the data of
+Table~\ref{interzonal} would be
 
@@
 
-The maximum likelihood estimate is $\draw\simeq 0.043, \draw_s=0.354$; a
-constrained maximization gives a likelihood ratio of about 5096, or a
-$p$-value of about $3.6\times 10^{-5}$.  Indeed we could go further
-and plot a profile likelihood curve for the log-odds ratio
-$\log\left(\draw_s/\draw\right)$ which, if Dirichlet, would be accurately
-parabolic over a wide range~\cite[page 343]{ohagan2004}.  This is
-illustrated in figure~\ref{proflike}, that shows a credible interval
-of about 1.8-2.4.
+The maximum likelihood estimate for the full 23-player dataset is
+$\draw\simeq 0.043, \draw_s=0.354$; a constrained maximization gives a
+likelihood ratio of about 5096, or a $p$-value of about $3.6\times
+10^{-5}$: there is strong evidence that Soviet players do indeed draw
+more frequently than non-Soviets.  Indeed we could go further and plot
+a profile likelihood curve for the log-odds ratio
+$\log\left(\draw_s/\draw\right)$ which, if Dirichlet, would be
+accurately parabolic over a wide range~\cite[page 343]{ohagan2004}.
+This is illustrated in figure~\ref{proflike}, that shows a credible
+interval of about 1.8-2.4: we can be reasonably sure that
+$\draw_s/\draw\geqslant e^{1.8}\simeq 6.05$.
 

### 2019-07-12 data/T20.R
work on cricket

@@ -1,0 +1,336 @@
+`T20_table` <-
+structure(list(team1 = structure(c(13L, 8L, 4L, 11L, 12L, 4L, 
+7L, 12L, 3L, 13L, 5L, 12L, 11L, 7L, 3L, 4L, 11L, 3L, 13L, 12L, 
+13L, 5L, 7L, 8L, 13L, 5L, 7L, 11L, 11L, 12L, 3L, 13L, 11L, 4L, 
+3L, 13L, 12L, 4L, 13L, 7L, 4L, 5L, 11L, 8L, 12L, 5L, 8L, 5L, 
+4L, 5L, 3L, 4L, 8L, 5L, 12L, 11L, 13L, 7L, 8L, 7L, 1L, 10L, 8L, 
+2L, 5L, 2L, 1L, 2L, 5L, 12L, 7L, 8L, 3L, 1L, 7L, 3L, 2L, 10L, 
+1L, 2L, 5L, 8L, 1L, 12L, 1L, 10L, 3L, 7L, 2L, 12L, 1L, 7L, 3L, 
+12L, 7L, 1L, 5L, 3L, 7L, 3L, 10L, 8L, 7L, 12L, 7L, 5L, 12L, 2L, 
+8L, 10L, 12L, 5L, 8L, 2L, 10L, 5L, 1L, 8L, 12L, 5L, 7L, 1L, 5L, 
+2L, 3L, 10L, 12L, 2L, 12L, 5L, 1L, 8L, 3L, 7L, 5L, 2L, 1L, 8L, 
+12L, 2L, 1L, 7L, 8L, 1L, 10L, 7L, 2L, 12L, 1L, 8L, 2L, 10L, 8L, 
+7L, 2L, 7L, 5L, 3L, 1L, 10L, 3L, 8L, 7L, 5L, 3L, 1L, 3L, 10L, 
+1L, 8L, 12L, 3L, 1L, 2L, 7L, 8L, 5L, 12L, 2L, 10L, 5L, 1L, 8L, 
+10L, 3L, 2L, 10L, 8L, 2L, 5L, 7L, 12L, 10L, 1L, 2L, 7L, 3L, 10L, 
+8L, 3L, 5L, 12L, 3L, 7L, 5L, 1L, 8L, 7L, 3L, 10L, 1L, 5L, 7L, 
[…]

### 2019-07-14 inst/T20_analysis.R
asymptotic analysis

@@ -1,0 +1,69 @@
+## Calculates the distribution of Lambda under the null, and
+## calculates a p-value for the hypothesis that all teams in the T20
+## dataset are of equal strength.  It does some resampling to verify
+## that the asymptotic distribution of 2*Lambda is attained in the case
+## of T20;
+
+
+## This takes quite a long time to run, which is why it's not included
+## in the package.
+
+library("hyper2")
+
+# Observed value of Lambda:
+Lambda_observed <-
+    maxp(T20,give=TRUE)$value-loglik(T20,indep(equalp(T20)))
+
+## Function to choose a random team from two teams:
+randomteam <- function(team1,team2){
+    as.vector(
+        cbind(team1,team2)[
+            cbind(seq_along(team1),
+                  sample(1:2,length(team1),replace=TRUE)
+                  )
+    ])
+}
+
+n <- 1000   # number of in-silico simulations
+Lambda <- rep(0,n)
+evaluate <- rep(0,n)
+equ <- rep(0,n)
+
+for(i in seq_len(n)){
+
+    ## Shuffle the results (but not the games):
+    team1 <- as.vector(T20_table$team1)  # not shuffled
+    team2 <- as.vector(T20_table$team2)  # not shuffled
+    toss_winner <- randomteam(team1,team2)
+    toss_decision <- as.vector(sample(T20$toss_decision))
+    match_winner <- randomteam(team1,team2)
+
+    # Now create virtual likelihood functions T20star:
+    T20star <- hyper2(pnames=c(levels(T20_table$team1)))
+    for(j in seq_along(team1)){
+        T20star[match_winner[j]] %<>% inc
+        T20star[c(team1[j],team2[j])] %<>% dec
+    }
+    evaluate[i] <- maxp(T20star,give=TRUE)$value
+    equ[i] <- loglik(T20star,indep(equalp(T20star)))
+    Lambda[i] <- evaluate[i] - equ[i]
+    print(i)
+}		       
+
+
+hist(Lambda)
+abline(v=Lambda_observed, lwd=7)
+cat(paste("resampled pvalue = ", sum(Lambda > Lambda_observed)/length(Lambda)),"\n")
+cat(paste("parametric pvalue = ", pchisq(Lambda_observed,df=13,lower.tail=FALSE)),"\n")
+
+
+
+## Now a qqplot of 2*Lambda vs random samples from chisq(df=13).
[…]

### 2019-09-23 inst/order.Rmd
some reflections on order

---
title: "Order statistics and numerical optimization"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The `hyper2` package

We consider the likelihood function given in section 5 (Men's sculling
in the 2016 Olympics) of the package vignette.  Here we investigate
some numerical problems arising from the use of `optim()` and
`constrOptim()`.  We will consider $n=10$ competitors and without loss
of generality, the competitors finish in order $1,2,3,\ldots,10$.  In
the vignette we see a formula

\[
{\mathcal L}\left(p_1,\ldots,p_{10}\right)=
\frac{p_1}{p_1+\cdots+p_{10}}\cdot
\frac{p_2}{p_2+\cdots+p_{10}}\cdot
\frac{p_3}{p_3+\cdots+p_{10}}\cdot
\frac{p_4}{p_4+\cdots+p_{10}}\cdot
\frac{p_5}{p_5+\cdots+p_{10}}\cdot\\
\frac{p_6}{p_6+\cdots+p_{10}}\cdot
\frac{p_7}{p_7+\cdots+p_{10}}\cdot
\frac{p_8}{p_8+p_9+p_{10}}\cdot
\frac{p_9}{p_9+p_{10}}\cdot
\frac{p_{10}}{p_{10}}
\]

We can calculate the evaluate (that is, the maximum likelihood
estimate) from the formula by inspection.  It is clear from the
penultimate term that $p_{10}=0$.  Similarly from the preceding term
it is easy to see that $p_9=0$ and so on.  The evaluate is thus
$p=(1,0,\ldots,0)$.  Let's see if we can reproduce this using
numerical methods.

The `hyper2` idiom for creating a likelihood function is as follows:

```{r defineH}
library("hyper2")
H <- order_likelihood(rbind(1:10))
H
```
And then maximizing it is carried out using the `maxp()` function:

```{r}
maxp(H)
loglik(H,indep(maxp(H)))
```

```{r}
dotchart(maxp(H))
```

So the numerics have not done a particularly good job in identifying
the evaluate, presumably as a result of the difficulty of honouring
the constraints.  However, if we give the system a hand by starting at
a point closer to the evaluate we can do much better:

```{r}
f <- function(p,n){c(p,rep((1-p)/n,n-1))}
sss <- f(0.9999,9)  # sensible start point, close to the evaluate
maxp(H,startp=sss)
loglik(H,indep(maxp(H,startp=sss)))
```

```{r}

H <- hyper2(
    pnames=c("Rahul","Kim-Joy","Ruby","Briony","Manon","Jon",
             "Dan","Karen","Terry","Antony","Luke","Imelda")
)
H
```

```{r}
	ranks <- c(Antony=12, Briony=2, Dan=4, Imelda=11, Jon=5, Karen=6, KimJoy=10, Luke=8, Manon=3, Rahul=7, Ruby=1, Terry=9)
```

### 2019-10-01 inst/cook.Rmd
nice analysis

---
title: "cooking"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

Suppose there are five people in a family.  Each day for five days, one person cooks a meal.  After the fifth day, each person puts the meals in order of preference, but is not allowed to vote on their own meal.  This very short vignette shows how to apply generalized Bradley-Terry to this situation using the `hyper2` package.

```{r cars}
library(hyper2)
H <- hyper2(pnames=letters[1:5])
H
```

Now add some data:

```{r}
H <- H +
    order_likelihood(c(5,2,3,4)) + # 1 cannot vote 
    order_likelihood(c(1,5,3,4)) + # 2 cannot vote
    order_likelihood(c(1,4,2,5)) + # 3 cannot vote
    order_likelihood(c(1,2,3,5)) + # 4 cannot vote
    order_likelihood(c(2,1,3,4))   # 5 cannot vote
H
```

graphically:

```{r maxlike, echo=FALSE}
dotchart(maxp(H),main="probabilities")
dotchart(log(maxp(H)),main="log probabilities")
```

Now we can assess the hypothesis that person `a` does in fact have the
highest strength of the five.  We can follow the reasoning in the
_icons_ demonstration:

```{r}
o <- function(Ul,Cl,startp,give=FALSE){
    small <- 1e-5  #  ensure start at an interior point
    if(missing(startp)){startp <- small*(1:4)+rep(0.1,4)}
    out <- maxp(H, startp=small*(1:4)+rep(0.1,4), give=TRUE, fcm=Ul,fcv=Cl)
		if(give){
			return(out)
		} else {
			return(out$value)
		}
}

p2max <- o(c(-1, 1, 0, 0), 0)
p3max <- o(c(-1, 0, 1, 0), 0)
p4max <- o(c(-1, 0, 0, 1), 0)
p5max <- o(c(-2,-1,-1,-1),-1)
```

(the final line is different because `p5` is the fillup value).

```{r}
(likes <- c(p2max,p3max,p4max,p5max))
max(likes)
```

(observe that the hypothesis with the maximum likelihood is
$p_2\geqslant p_1$).  Compare with the unconstrained maximum:

```{r}
maxp(H,give=TRUE)$value
```

So the support loss by the constraint is

```{r}
(L <- maxp(H,give=TRUE)$value - max(likes))
```

short of Edwards's two-units-of-support criterion.

### 2019-10-01 inst/cook.Rmd
more detail and rewriting

@@
 
-Suppose there are five people in a family.
-Each day for five days, one person cooks a meal.
-After the fifth day, each person puts the meals in order of preference, but is not allowed to vote on their own meal.
-This very short vignette shows how to apply generalized Bradley-Terry to this situation using the `hyper2` package.
+Suppose there are five people in a family.  Each day for five days,
+one person cooks a meal.  After the fifth day, each person puts the
+meals in order of preference, but is not allowed to vote on their own
+meal.  This very short vignette shows how to apply generalized
+Bradley-Terry to this situation using the `hyper2` package.
 
@@
 library(hyper2)
-H <- hyper2(pnames=letters[1:5])
+H <- hyper2(pnames=paste("p",1:5,sep=""))
 H
@@
 H <- H +
-    order_likelihood(c(5,2,3,4)) + # 1 cannot vote 
-    order_likelihood(c(1,5,3,4)) + # 2 cannot vote
-    order_likelihood(c(1,4,2,5)) + # 3 cannot vote
-    order_likelihood(c(1,2,3,5)) + # 4 cannot vote
-    order_likelihood(c(2,1,3,4))   # 5 cannot vote
+    order_likelihood(c(5,2,3,4)) + # person 1 says 5 is the best, 4 the worst.
+    order_likelihood(c(1,5,3,4)) + # person 2 says 1 is the best, 4 the worst.
+    order_likelihood(c(1,4,2,5)) + # person 3 says 1 is the best, 5 the worst.
+    order_likelihood(c(1,2,3,5)) + # person 4 says 1 is the best, 5 the worst.
+    order_likelihood(c(2,1,3,4))   # person 5 says 2 is the best, 4 the worst.
 H
@@
 
-graphically:
+Note that noone is allowed to vote for themselves.  We may find
+maximum likelihood estimate for the strengths,
+$\operatorname{argmax}\mathcal{L}\left(p_1,\ldots,p_5\right)$
+Graphically:
 
@@
 dotchart(log(maxp(H)),main="log probabilities")
+```
+
+We can assess the hypothesis that all players have the same strength:
+
+```{r} 
+L1 <- loglik(H,indep(equalp(H)))  # support at equal strengths
+L2 <- loglik(H,indep(maxp(H)))    # maximum support
+L2-L1
+pchisq(2*(L2-L1),df=1,lower.tail=FALSE) # p-value of asymptotic distribution
 ```

### 2019-10-02 inst/cook.Rmd
more work: clearer implications

@@
 maximum likelihood estimate for the strengths,
-$\operatorname{argmax}\mathcal{L}\left(p_1,\ldots,p_5\right)$
+$\operatorname{argmax}\mathcal{L}\left(p_1,\ldots,p_5\right)$:
+
+```{r}
+maxp(H)
+```
+
 Graphically:
@@
 
-(observe that the hypothesis with the maximum likelihood is
-$p_2\geqslant p_1$).  Compare with the unconstrained maximum:
+Observe that the hypothesis with the maximum likelihood is
+$p_2\geqslant p_1$, as might be expected on the grounds that in the
+unconstrained case we have
+$\hat{p_2}>\max\left(\hat{p_3},\hat{p_4},\hat{p_5}\right)$.  Now
+compare the constrained maximum with the unconstrained maximum:
 
@@
 
-short of Edwards's two-units-of-support criterion.
+short of Edwards's two-units-of-support criterion.  There is no strong
+evidence to support the assertion that person 1 is actually stronger
+than person 2 in the sense that $p_1>p_2$.

### 2019-10-02 inst/cook.Rmd
permutation test

@@
 than person 2 in the sense that $p_1>p_2$.
+
+## Suspect observation
+
+Suppose that we subsequently observe order statistic `5,4,3,2,1`, that
+is, person 5 is the best, 4 is the second best, and so on to 1 being
+the worst.  Is this observation consistent with the previous dataset?
+
+It is possible to perform a permutation test on this as follows.
+First, calculate the probability of each of the $5!=120$ possible
+observations (up to a constant):
+
+```{r}
+library(partitions)
+imH <- indep(maxp(H))
+f <- function(o){loglik(order_likelihood(o),imH,log=FALSE)}
+LL <- apply(perms(5),2,f)
+```
+
+Then calculate the pvalue as the probability of obtaining the
+observation or an observation more extreme:
+
+```{r}
+obs <- c(5,4,3,2,1)
+(pval <- sum(LL[LL <= f(obs)])/sum(LL))
+```

### 2019-11-15 inst/order.Rmd
rewording and more careful discussion

@@
 
-We consider the likelihood function given in section 5 (Men's sculling
-in the 2016 Olympics) of the package vignette.  Here we investigate
-some numerical problems arising from the use of `optim()` and
-`constrOptim()`.  We will consider $n=10$ competitors and without loss
-of generality, the competitors finish in order $1,2,3,\ldots,10$.  In
-the vignette we see a formula
+We consider the likelihood function given by a single observation of
+an order statistic.  Here we investigate some numerical problems
+arising from the use of `optim()` and `constrOptim()`.  We will
+consider $n=10$ competitors and without loss of generality, the
+competitors finish in order $1,2,3,\ldots,10$.  The likelihood
+function for this observation is:
 
@@
 $p=(1,0,\ldots,0)$.  Let's see if we can reproduce this using
-numerical methods.
+numerical methods.  Note that this is a highly pathological case; the
+likelihood is not defined at the evaluate.  Even considering limiting
+cases is not straightforward; consider two processes, approaching the
+evaluate in two different ways:
+
+\[
+\lim_{p\longrightarrow 0}
+{\mathcal L}(1-9p,p,p,p,p,p,p,p,p,p,p)=\frac{1}{8!}
+\]
+
+but
+
+\[
+\lim_{p\longrightarrow 0}
+{\mathcal L}(\alpha,\alpha p,\alpha p^2,\ldots,\alpha p^9)=1
+\]
+
+where $\alpha=(1-p^{10})/(1-p)$ is a normalizing constant.  Thus the
+likelihood _at_ the evaluate is not defined, even as a limit (because
+it depends on the direction of approach).
 
@@
 ```{r}
-maxp(H)
-loglik(H,indep(maxp(H)))
+evaluate1 <- maxp(H)
+(maxlike1 <- loglik(H,indep(evaluate1)))
+dotchart(evaluate1,pch=16)
 ```
 
-```{r}
-dotchart(maxp(H))
-```
-
-So the numerics have not done a particularly good job in identifying
-the evaluate, presumably as a result of the difficulty of honouring
-the constraints.  However, if we give the system a hand by starting at
-a point closer to the evaluate we can do much better:
+Comparing the numerically determined maximum likelihood estimate with
+the exact MLE of $(1,0,\ldots,0)$ shows that the numerics have not
+done a particularly good job in identifying the evaluate, presumably
+as a result of the difficulty of honouring the constraints.  However,
+if we give the system a hand by starting at a point closer to the
+evaluate we can do much better:
 
@@
 f <- function(p,n){c(p,rep((1-p)/n,n-1))}
-sss <- f(0.9999,9)  # sensible start point, close to the evaluate
-maxp(H,startp=sss)
-loglik(H,indep(maxp(H,startp=sss)))
+sss <- f(0.9999,9)  # sensible start point, close (?) to the evaluate
+evaluate2 <- maxp(H,startp=sss)
+maxlike2 <- loglik(H,indep(evaluate2))
 ```
 
+Graphically:
+
 ```{r}
-
-H <- hyper2(
-    pnames=c("Rahul","Kim-Joy","Ruby","Briony","Manon","Jon",
-             "Dan","Karen","Terry","Antony","Luke","Imelda")
-)
-H
+dotchart(evaluate2,pch=16)
 ```
 
+So it is much better than the coldstart above in the sense that there
+is more probability on $p_1$ and less everywhere else.  We can compare
+the hotstart above to the coldstart:
+
 ```{r}
-	ranks <- c(Antony=12, Briony=2, Dan=4, Imelda=11, Jon=5, Karen=6, KimJoy=10, Luke=8, Manon=3, Rahul=7, Ruby=1, Terry=9)
+maxlike2-maxlike1
 ```
+
+This is negative: the so-called "sensible start point" has produced an
+inferior solution.  This is because `sss` has $p_2=p_3=\ldots =
+p_{10}$, so predicts a draw between players $2,3,\ldots 10$, contrary
+to observation.
+
+A better way is to mimic the more appropriate of the two limiting processes above:
+
+```{r}
+vsssp <- function(x,n){
+  jj <- 1/x^seq_len(n)
+  indep(jj/sum(jj))
+}
+```
+
+Then
+
+```{r}
+evaluate3 <- maxp(H,vsssp(4,10))
+maxlike3 <- loglik(H,indep(evaluate3))
+dotchart(evaluate3,pch=16)
+```
+
+Observe that the likelihood at this point is marginally better than
+the `equalp()` start point:
+
+```{r}
+maxlike3-maxlike2
+```
+
+but the dotchart again shows a quite a lot of probability on
+$p_2,\ldots p_{10}$.
+
+# Discussion
+
+I am not sure how important these considerations are in practice.
+Firstly, the order statistic used here is extremely pathological, and
+perhaps it is unreasonable to expect any inferential tool to work
+under these circumstances.
+
+Secondly, and perhaps more importantly, likelihood is not defined at a
+point; likelihood is an equivalence class of functions whose domain is
+possible hypotheses.  We can only compare likelihoods for two
+different hypotheses [because likelihood is only defined up to a
+multiplicative constant].  And in this case, the evaluate
+$(1,0,\ldots,0)$ is not a hypothesis anyone would want a likelihood
+for: the likelihood is not defined at that point, and does not even
+have a well-defined limit there.

### 2019-11-15 inst/order.Rmd
more discussion

@@
 different hypotheses [because likelihood is only defined up to a
-multiplicative constant].  And in this case, the evaluate
-$(1,0,\ldots,0)$ is not a hypothesis anyone would want a likelihood
-for: the likelihood is not defined at that point, and does not even
-have a well-defined limit there.
+multiplicative constant].  And for the observation considered here,
+the evaluate $(1,0,\ldots,0)$ is not a hypothesis anyone would want a
+likelihood for: the likelihood is not defined at that point, and does
+not even have a well-defined limit there.
+
+To some extent this is not surprising: the numerics force one to have
+a strictly positive minimum value for each $p_i$ and the fillup value
+$p_{10}=1-\sum p_i$.  The likelihood maximization routine tries to
+maximize $p_9/p_{10}, p_8/p_9,\ldots,p_1/p_2$ simultaneously, all
+subject to the unit sum constraint $\sum p_i=1$.  And of course, if
+the ratio is too high then the unit sum constraint would be violated
+(given that $p_{10}$ is at its minimum value).

### 2019-11-15 inst/cook.Rmd
more work on the loser

@@
 ```
+
+# Just the loser
+
+Now we are going to look at what happens when our observation is only
+the _loser_ (which is me).  We have two hypotheses
+
+* $H_1: \mbox{I might very well be the worst chef in the family}$
+* $H_2: \mbox{I am not the worst chef in the family}$
+
+Observe that $H_2$ implies that someone is worse than me and this
+corresponds to a restricted parameter space.  The first step is to
+define a suitable likelihood function for our observation:
+
+```{r}
+n <- 6 # number of people in the family
+H <- hyper2(pnames=letters[1:6])
+L <- ggol(H,letters[2:n],letters[1])
+L[[1]]
+L[[2]]
+```
+
+In the above, `L` is a list of hyper2 objects corresponding to a
+possible order statistic (we see the first two).  Any order is
+permissible, so long as player `a` (me) is bottom.  Now we maximize
+the likelihood under $H_1$, that is, perform an unconstrained
+maximization:
+
+```{r}
+(ans_unconstrained <- maxplist(L,give=TRUE))
+```
+
+The likelihood at the evaluate is 1 (technically, this is wrong
+because likelihood is defined only up to a multiplicative constant.  A
+correct statement is: the likelihood function evaluated by
+`like_single_list()` returns 1).  The next step is to perform a
+constrained maximization, the constraint being that I am better than
+at least one other person, here chosen to be person `b`:
+
+```{r}
+small <- 1e-2  # numerical necessity
+ans_constrained <-
+maxplist(L,startp=c(small*2,rep(small,n-2)),fcm=rbind(c(1,-1,rep(0,n-3))),fcv=0,give=TRUE)
+ans_constrained
+```
+
+We can see that the likelihood ratio ${\mathcal L}(H_2)/{\mathcal
+L}(H_1)$ is 0.5 for an observation that I come last.  This is exactly
+the same as the likelihood ratio for the following situation: we toss
+a coin which is either fair ($H_\mbox{fair}$) or two-headed
+($H_\mbox{2heads}$).  We observe a head.  Then ${\mathcal
+L}(H_\mbox{2heads})/{\mathcal L}(H_\mbox{fair}) = 0.5$ as well (this
+is Edwards's valet and black and white ball thought experiment).

### 2019-11-16 inst/cook.Rmd
more work, nice conclusion

@@
 Now we are going to look at what happens when our observation is only
-the _loser_ (which is me).  We have two hypotheses
+the _loser_ (which is always me).  We have six competitors
+`a,b,c,d,e,f` with nonnegative strengths $p_1,p_2,p_3,p_4,p_5,p_6,
+\sum p_i=1$.  I am competitor `a` with strength $p_1$, and always
+lose.  We have two hypotheses:
 
-* $H_1: \mbox{I might very well be the worst chef in the family}$
-* $H_2: \mbox{I am not the worst chef in the family}$
+* $H_1\colon p_1=0$
+* $H_2\colon\exists i\neq 1 \mbox{ with } p_1\geq p_i>0$
 
-Observe that $H_2$ implies that someone is worse than me and this
-corresponds to a restricted parameter space.  The first step is to
-define a suitable likelihood function for our observation:
+(observe that $H_2$ implies that someone is worse than me).  Both
+hypotheses correspond to a restriction on allowable parameter space.
+The first step is to define a suitable likelihood function for our
+observation:
 
@@
 L <- ggol(H,letters[2:n],letters[1])
+length(L)
 L[[1]]
@@
 
-In the above, `L` is a list of hyper2 objects corresponding to a
-possible order statistic (we see the first two).  Any order is
-permissible, so long as player `a` (me) is bottom.  Now we maximize
-the likelihood under $H_1$, that is, perform an unconstrained
-maximization:
+In the above, `L` is a list of hyper2 objects corresponding to all
+$5!=120$ possible observed order statistics (only the first two,
+`L[[1]]` and `L[[2]]`, are printed, which correspond to orders
+`bcdefa` and `bcdfea` respectively).  Any order is permissible, so
+long as player `a` (me) is bottom.  Now we maximize the likelihood
+under $H_1$, that is, perform an unconstrained maximization:
 
@@
 
-The likelihood at the evaluate is 1 (technically, this is wrong
-because likelihood is defined only up to a multiplicative constant.  A
-correct statement is: the likelihood function evaluated by
+The constraint embodied in $H_1$ does not need to be enforced: the
+evaluate $\hat{p}$ has $p_1=0$, consistent with competitor `a` coming
+last; actually we have
+$\hat{p}=\left(0,\frac{1}{5},\frac{1}{5},\frac{1}{5},\frac{1}{5},\frac{1}{5}\right)$
+analytically, and the numerics are not too far from this.  The
+likelihood at the evaluate is 1 (technically, this is wrong because
+likelihood is defined only up to a multiplicative constant.  A correct
+statement is: the likelihood function evaluated by
 `like_single_list()` returns 1).  The next step is to perform a
 constrained maximization, the constraint being that I am better than
-at least one other person, here chosen to be person `b`:
+at least one other person, here chosen to be, without loss of
+generality, competitor `b`:
 
@@
 
-We can see that the likelihood ratio ${\mathcal L}(H_2)/{\mathcal
-L}(H_1)$ is 0.5 for an observation that I come last.  This is exactly
-the same as the likelihood ratio for the following situation: we toss
-a coin which is either fair ($H_\mbox{fair}$) or two-headed
-($H_\mbox{2heads}$).  We observe a head.  Then ${\mathcal
+Observe that the constraint is active: $p_1=p_2$ in the numerical
+result.  Analytically, the evaluate is indeterminate in
+$p_3,p_4,p_5,p_6$, but has $p_1=p_2=0$ (numerically this means that
+$p_1=p_2=\epsilon$).  We can see that the likelihood ratio ${\mathcal
+L}(H_2)/{\mathcal L}(H_1)$ is 0.5 for an observation that I come last.
+This is exactly the same as the likelihood ratio for the following
+situation: we toss a coin which is either fair ($H_\mbox{fair}$) or
+two-headed ($H_\mbox{2heads}$).  We observe a head.  Then ${\mathcal
 L}(H_\mbox{2heads})/{\mathcal L}(H_\mbox{fair}) = 0.5$ as well (this

### 2019-11-17 inst/cook.Rmd
work

--- 2019-10-02 inst/cook.Rmd
@@
 ---
-title: "cooking"
+title: "Domestic cooking and likelihood"
 output: html_document
@@
 
-## R Markdown
+Suppose there are five people in a family.  Each day for five days,
+one person cooks a meal.  After the fifth day, each person considers
+the four meals cooked by other family members, and puts them in order
+of preference.  This very short vignette shows how to apply
+generalized Bradley-Terry to this situation using the `hyper2`
+package.
 
-Suppose there are five people in a family.  Each day for five days,
-one person cooks a meal.  After the fifth day, each person puts the
-meals in order of preference, but is not allowed to vote on their own
-meal.  This very short vignette shows how to apply generalized
-Bradley-Terry to this situation using the `hyper2` package.
-
-```{r cars}
+```{r loadlib}
 library(hyper2)
@@
 
-```{r}
+```{r data}
 H <- H +
@@
 ```{r maxlike, echo=FALSE}
-dotchart(maxp(H),main="probabilities")
-dotchart(log(maxp(H)),main="log probabilities")
+dotchart(maxp(H),pch=16,main="probabilities")
+dotchart(log(maxp(H)),pch=16,main="log probabilities")
 ```
@@
 
-Now we can assess the hypothesis that person `a` does in fact have the
-highest strength of the five.  We can follow the reasoning in the
-_icons_ demonstration:
+Thus we reject the hypothesis of equal strength.  Now we can assess
+the hypothesis that person `a` does in fact have the highest strength
+of the five.  We can follow the reasoning in the _icons_
+demonstration:
 
@@
 LL <- apply(perms(5),2,f)
+length(LL)
+sum(LL)
+head(LL)
 ```
 
-Then calculate the pvalue as the probability of obtaining the
-observation or an observation more extreme:
+So `LL` is the probability of observing each permutation of 5 objects.
+We then calculate the $p$-value as the probability of obtaining the
+observation or an observation more extreme; in this case we define a
+permutation to be "more extreme" if it has a smaller probability than
+the observed permutation of `5,4,3,2,1` (under the null).  This is
+easily calculated:
 
@@
 ```
+
+Thus it is reasonable to reject the hypothesis that this particular
+observation was drawn from the same distribution as the others, and
+infer that it was suspect.

### 2019-11-17 inst/cook.Rmd

--- 2019-11-16 inst/cook.Rmd
@@
 
+Thus it is reasonable to reject the hypothesis that this particular
+observation was drawn from the same distribution as the others, and
+infer that it was suspect.
+
 # Just the loser
+
+Now we are going to look at what happens when our observation is only
+the _loser_ (which is always me).  We have six competitors
+`a,b,c,d,e,f` with nonnegative strengths $p_1,p_2,p_3,p_4,p_5,p_6,
+\sum p_i=1$.  I am competitor `a` with strength $p_1$, and always
+lose.  We have two hypotheses:
+
+* $H_1\colon p_1=0$
+* $H_2\colon\exists i\neq 1 \mbox{ with } p_1\geq p_i>0$
+
+(observe that $H_2$ implies that someone is worse than me).  Both
+hypotheses correspond to a restriction on allowable parameter space.
+The first step is to define a suitable likelihood function for our
+observation:
+
+```{r}
+n <- 6 # number of people in the family
+H <- hyper2(pnames=letters[1:6])
+L <- ggol(H,letters[2:n],letters[1])
+length(L)
+L[[1]]
+L[[2]]
+```
+
+In the above, `L` is a list of hyper2 objects corresponding to all
+$5!=120$ possible observed order statistics (only the first two,
+`L[[1]]` and `L[[2]]`, are printed, which correspond to orders
+`bcdefa` and `bcdfea` respectively).  Any order is permissible, so
+long as player `a` (me) is bottom.  Now we maximize the likelihood
+under $H_1$, that is, perform an unconstrained maximization:
+
+```{r}
+(ans_unconstrained <- maxplist(L,give=TRUE))
+```
+
+The constraint embodied in $H_1$ does not need to be enforced: the
+evaluate $\hat{p}$ has $p_1=0$, consistent with competitor `a` coming
+last; actually we have
+$\hat{p}=\left(0,\frac{1}{5},\frac{1}{5},\frac{1}{5},\frac{1}{5},\frac{1}{5}\right)$
+analytically, and the numerics are not too far from this.  The
+likelihood at the evaluate is 1 (technically, this is wrong because
+likelihood is defined only up to a multiplicative constant.  A correct
+statement is: the likelihood function evaluated by
+`like_single_list()` returns 1).  The next step is to perform a
+constrained maximization, the constraint being that I am better than
+at least one other person, here chosen to be, without loss of
+generality, competitor `b`:
+
+```{r}
+small <- 1e-2  # numerical necessity
+ans_constrained <-
+maxplist(L,startp=c(small*2,rep(small,n-2)),fcm=rbind(c(1,-1,rep(0,n-3))),fcv=0,give=TRUE)
+ans_constrained
+```
 

### 2019-11-18 inst/loser.Rmd
specific illustration of loser statistics

---
title: "Weakly informative observations: identifying the weakest player"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(hyper2)
```

# Partial rank observations: who lost?

Now we are going to look at what happens when our observation is only
the _loser_ (which is here taken to be player `a`).  We have six
competitors `a,b,c,d,e,f` with nonnegative strengths
$p_1,p_2,p_3,p_4,p_5,p_6, \sum p_i=1$.  Competitor `a` has strength
$p_1$, and loses.  We are interested in how informative this
observation is with regard to the hypothesis that `a` does in fact
have zero strength.  We have three hypotheses:

* $H_1\colon p_1=0$
* $H_2\colon\exists i\neq 1 \mbox{ with } p_1\geq p_i>0$
* $H_3\colon p_1=p_2=\ldots=p_n=\frac{1}{n}$

(observe that $H_2$ implies that someone is worse than me).  All three
hypotheses correspond to a restriction on allowable parameter space.
The first step is to define a suitable likelihood function for our
observation:

```{r}
n <- 6 # number of competitors
H <- hyper2(pnames=letters[1:6])
L <- ggol(H,letters[2:n],letters[1]) 
length(L)
L[[1]]
L[[2]]
```

In the above, `L` is a likelihood function for the observation that
competitor `a` came last.  In R idiom, it is a list of hyper2 objects
corresponding to all $5!=120$ possible observed order statistics (only
the first two, `L[[1]]` and `L[[2]]`, are printed, which correspond to
orders `bcdefa` and `bcdfea` respectively).  Any order is permissible,
so long as player `a` is bottom.

## Hypothesis 1

Now we maximize the likelihood
under $H_1$, that is, perform an unconstrained maximization:

```{r}
(ans_unconstrained <- maxplist(L,give=TRUE))
```

The constraint embodied in $H_1$ does not need to be enforced: the
evaluate $\hat{p}$ has $p_1=0$, consistent with competitor `a` coming
last; actually we have
$\hat{p}=\left(0,\frac{1}{5},\frac{1}{5},\frac{1}{5},\frac{1}{5},\frac{1}{5}\right)$
analytically, and the numerics are not too far from this.  The
likelihood at the evaluate is 1 (technically, this is wrong because
likelihood is defined only up to a multiplicative constant.  A correct
statement is: the likelihood function evaluated by
`like_single_list()` returns 1 at $\hat{p}$ returns 1).

## Hypothesis 2

The next step is to perform a constrained maximization, the constraint
being $H_2$.  This says that I am better than at least one other
person, here chosen to be, without loss of generality, competitor `b`:

```{r}
small <- 1e-2  # numerical necessity
ans_constrained <-
maxplist(L,startp=c(small*2,rep(small,n-2)),fcm=rbind(c(1,-1,rep(0,n-3))),fcv=0,give=TRUE)
ans_constrained
```

Observe that the constraint is active: $p_1=p_2$ in the numerical
result.  Analytically, the evaluate is indeterminate in
$p_3,p_4,p_5,p_6$, but has $p_1=p_2=0$ (numerically this means that
$p_1=p_2=\epsilon$, where $\epsilon$ is some small number chosen for
smallness and numerical expediency).

## Hypothesis 3

Recall that $H_3\colon p_1=p_2=\ldots=p_n=\frac{1}{n}$; this is
equivalent to a random chance.  Because this is a point hypothesis, we
do not need to use any maximization techniques; it is sufficient to
evaluate the likelihood at this point:

```{r}
like_single_list(indep(equalp(L[[1]])),L)
```

The likelihood can be shown to be $\frac{1}{n}$ (which is clear on
intuitive grounds: $H_3$ says that all players are equivalent and if
this is correct, then one would identify players `a` as the worst with
a probability of $\frac{1}{n}$).  Note again that this is loose
language as likelihood is only defined up to a multiplicative
constant.

# Discussion

We are now in a position to define a likelihood function for
$H-1,H_2,H_3$:

$$
{\mathcal L}(H_1)=1\qquad
{\mathcal L}(H_2)=\frac{1}{2}\qquad
{\mathcal L}(H_3)=\frac{1}{n}
$$

(recall that $n$ is the number of competitors).  We can see that the
likelihood ratio ${\mathcal L}(H_2)/{\mathcal L}(H_1)$ is 0.5 for an
observation that I come last.  This is exactly the same as the
likelihood ratio for the following situation: we toss a coin which is
either fair ($H_\mbox{fair}$) or two-headed ($H_\mbox{2heads}$).  We
observe a head.  Then ${\mathcal L}(H_\mbox{2heads})/{\mathcal
L}(H_\mbox{fair}) = 0.5$ as well (this is Edwards's valet and black
and white ball thought experiment).

## Repeated observations

Suppose we make $r$ independent observations, all of which are that
competitor `a` comes last.  Then the likelihood function becomes

$$
{\mathcal L}(H_1)=1\qquad
{\mathcal L}(H_2)=\frac{1}{2^r}\qquad
{\mathcal L}(H_3)=\frac{1}{n^r}
$$

The support for $H_1$ against $H_2$ would be $\log(2^r)=r\log 2$.  To
exceed Edwards's criterion of 2 units of support per degree of freedom
would require $2/\log(2)\simeq 2.9$ observations, or 3 rounded up;
alternatively, we might require the log-likelihood to lie in the tail
area of its asymptotic $\chi^2_1$ distribution, that is

```{r}
qchisq(0.95,1)
```

that is $2r\log 2\geq 3.84$ or $r\geq 2.77$.

The support for $H_1$ against $H_3$ would be $r\log n$ but this time
we need $n-1$ degrees of freedom.  The likelihood criterion would be
$r\log n\geq 2(n-1)$, or $r\geq 2(n-1)/\log n$.  Alternatively we
might adopt a frequentist view and require
$2\log(n^r)\geq\chi^2_{n-1}(0.95)$, or $r\geq Q/(2\log n)$, where Q is
the 95th percentile of the chi-square distribution with $n$ degrees of
freedom, `qchisq(0.95,df=n)`.  It might be interesting to compare
these two approaches.

### 2019-11-18 inst/loser.Rmd
minor rewording

@@
 The likelihood can be shown to be $\frac{1}{n}$ (which is clear on
-intuitive grounds: $H_3$ says that all players are equivalent and if
-this is correct, then one would identify players `a` as the worst with
-a probability of $\frac{1}{n}$).  Note again that this is loose
-language as likelihood is only defined up to a multiplicative
-constant.
+intuitive grounds: $H_3$ says that all players are equivalent and the
+order statistic is totally random.  If this is correct, then one would
+identify player `a` as the worst with a probability $\frac{1}{n}$).
+Note again that this is loose language as likelihood is only defined
+up to a multiplicative constant.
 
@@
 
-We are now in a position to define a likelihood function for
-$H-1,H_2,H_3$:
+We are now in a position to define a likelihood function on the three
+hypotheses $H_1,H_2,H_3$:
 
@@
 likelihood ratio ${\mathcal L}(H_2)/{\mathcal L}(H_1)$ is 0.5 for an
-observation that I come last.  This is exactly the same as the
-likelihood ratio for the following situation: we toss a coin which is
-either fair ($H_\mbox{fair}$) or two-headed ($H_\mbox{2heads}$).  We
-observe a head.  Then ${\mathcal L}(H_\mbox{2heads})/{\mathcal
-L}(H_\mbox{fair}) = 0.5$ as well (this is Edwards's valet and black
-and white ball thought experiment).
+observation that competitor `a` comes last.  This is exactly the same
+as the likelihood ratio for the following situation: we toss a coin
+which is either fair ($H_\mbox{fair}$) or two-headed
+($H_\mbox{2heads}$).  We observe a head.  Then ${\mathcal
+L}(H_\mbox{2heads})/{\mathcal L}(H_\mbox{fair}) = 0.5$ as well (this
+is Edwards's valet and black and white ball thought experiment).
 

### 2019-11-19 inst/loser.Rmd
work on issue #22

@@
 ```{r}
-(ans_unconstrained <- maxplist(L,give=TRUE))
+(ans_unconstrained <- maxp(L,give=TRUE))
 ```
@@
 ans_constrained <-
-maxplist(L,startp=c(small*2,rep(small,n-2)),fcm=rbind(c(1,-1,rep(0,n-3))),fcv=0,give=TRUE)
+maxp(L,startp=c(small*2,rep(small,n-2)),fcm=rbind(c(1,-1,rep(0,n-3))),fcv=0,give=TRUE)
 ans_constrained

### 2019-11-19 R/hyper2.R

@@ -366,6 +366,7 @@ setGeneric("pnames<-",function(x,value){standardGeneric("pnames<-")})
 }
 
 `maxp` <- function(H, startp=NULL, give=FALSE, fcm=NULL, fcv=NULL, ...){
+    if(inherits(H,"suplist")){return(maxplist(Hlist=H,startp=startp,give=give,fcm=fcm,fcv=fcv,...))}
     SMALL <- 1e-6
     
     n <- size(H)

### 2019-11-20 inst/collusion_chess_draws.Rnw
fixes #23

@@
   return(  
-      -loglik(karpov_kasparov_anand,c(x,x,x,w))  # negative as we minimize
+      -loglik(c(x,x,x,w),karpov_kasparov_anand)  # negative as we minimize
   )
@@
     return(  
-        -loglik(kka_3draws,
-                c(karpov,kasparov,anand,white,draw,draw)
+        -loglik(
+                c(karpov,kasparov,anand,white,draw,draw),
+             kka_3draws
                 )

### 2019-11-20 inst/cook.Rmd
look at finding the loser
fixes #23

--- 2019-11-17 inst/cook.Rmd
@@
 ```{r} 
-L1 <- loglik(H,indep(equalp(H)))  # support at equal strengths
-L2 <- loglik(H,indep(maxp(H)))    # maximum support
+L1 <- loglik(indep(equalp(H)),H)  # support at equal strengths
+L2 <- loglik(indep(maxp(H)),H)    # maximum support
 L2-L1
@@
 imH <- indep(maxp(H))
-f <- function(o){loglik(order_likelihood(o),imH,log=FALSE)}
+f <- function(o){loglik(imH,order_likelihood(o),log=FALSE)}
 LL <- apply(perms(5),2,f)

### 2019-11-20 inst/order.Rmd
fixes #23

@@
 evaluate1 <- maxp(H)
-(maxlike1 <- loglik(H,indep(evaluate1)))
+(maxlike1 <- loglik(indep(evaluate1),H))
 dotchart(evaluate1,pch=16)
@@
 evaluate2 <- maxp(H,startp=sss)
-maxlike2 <- loglik(H,indep(evaluate2))
+maxlike2 <- loglik(indep(evaluate2),H)
 ```
@@
 evaluate3 <- maxp(H,vsssp(4,10))
-maxlike3 <- loglik(H,indep(evaluate3))
+maxlike3 <- loglik(indep(evaluate3),H)
 dotchart(evaluate3,pch=16)

### 2019-11-20 R/hyper2.R

@@ -174,15 +174,15 @@ setGeneric("pnames<-",function(x,value){standardGeneric("pnames<-")})
   return(hyper2(out[[1]],out[[2]],pnames=jj))
 }
 
-`loglik` <- function(H,p,log=TRUE){
+`loglik` <- function(p,H,log=TRUE){
   if(is.matrix(p)){
-    return(apply(p,1,function(o){.loglik_single(H,p=o, log=log)}))
+    return(apply(p,1,function(o){.loglik_single(p=o,H, log=log)}))
   } else {
-    return(.loglik_single(H,p,log=log))
+    return(.loglik_single(p,H,log=log))
   }
 }
 
-`.loglik_single` <- function(H,p,log=TRUE){
+`.loglik_single` <- function(p,H,log=TRUE){
   stopifnot(length(p) == size(H)-1)
   stopifnot(all(p>=0))
   stopifnot(sum(p)<=1)
@@ -344,7 +344,7 @@ setGeneric("pnames<-",function(x,value){standardGeneric("pnames<-")})
     return(hyper2(out[[1]],out[[2]],pnames=pnames(x)))
 }
 
-`gradient` <- function(H,probs){
+`gradient` <- function(probs,H){
   stopifnot(length(probs) == size(H)-1)
   differentiate(brackets(H), powers(H), fillup(probs), size(H))$grad_comp
 }
@@ -374,8 +374,8 @@ setGeneric("pnames<-",function(x,value){standardGeneric("pnames<-")})
         startp <- rep(1/n,n-1)
         }
 
-    objective <- function(p){ -loglik(H,p) }
-    gradfun   <- function(p){ -(gradient(H,p))} #NB minus signs
+    objective <- function(p){ -loglik(p,H) }
+    gradfun   <- function(p){ -(gradient(p,H))} #NB minus signs
     
     UI <- rbind(
         diag(nrow=n-1),  # regular: p1 >=0, p2 >= 0, ..., p_{n-1} >= 0

### 2019-11-20 R/integration.R

@@ -31,7 +31,7 @@ function(e){
 
 `dhyper2` <- function(P,H,...){
   P <- rbind(P)
-  loglik(H,P,log=FALSE)/B(H,...)
+  loglik(P,H,log=FALSE)/B(H,...)
 }
 
 `dhyper2_e` <-  # analogous to dhyperdirichlet_e()
@@ -40,7 +40,7 @@ function(e, H, include.Jacobian=TRUE){
   e <- c(1,e)
   p <- e_to_p(e)
 
-  out <- loglik(H,indep(p),log=FALSE)
+  out <- loglik(indep(p),H,log=FALSE)
   if(include.Jacobian){out <- Jacobian(e)*out}
   return(out)
 }

### 2019-11-20 inst/T20_analysis.R

@@ -12,7 +12,7 @@ library("hyper2")
 
 # Observed value of Lambda:
 Lambda_observed <-
-    maxp(T20,give=TRUE)$value-loglik(T20,indep(equalp(T20)))
+    maxp(T20,give=TRUE)$value-loglik(indep(equalp(T20)),T20)
 
 ## Function to choose a random team from two teams:
 randomteam <- function(team1,team2){
@@ -45,7 +45,7 @@ for(i in seq_len(n)){
         T20star[c(team1[j],team2[j])] %<>% dec
     }
     evaluate[i] <- maxp(T20star,give=TRUE)$value
-    equ[i] <- loglik(T20star,indep(equalp(T20star)))
+    equ[i] <- loglik(indep(equalp(T20star)),T20star)
     Lambda[i] <- evaluate[i] - equ[i]
     print(i)
 }		       

### 2019-11-20 inst/collusion_curacao1962.R

@@ -52,7 +52,7 @@ s[2] %<>% `-`(small)
 
 ## Perform the constrained optimization:
 ml_p_constrained <- maxp(H,fcm=c(1,-1,rep(0,size(H)-3)),fcv=0,startp=s)
-max_support_constrained <- loglik(H,indep(ml_p_constrained))
+max_support_constrained <- loglik(indep(ml_p_constrained),H)
 
 support <- max_support_free - max_support_constrained
 

### 2019-11-20 inst/counterstrike_random.R

@@ -7,7 +7,7 @@
 
 source("counterstrike.R")
 equalstrengths <- rep(0.1, 9)
-datapoint <- loglik(H,indep(maxp(H))) - loglik(H,equalstrengths)
+datapoint <- loglik(indep(maxp(H)),H) - loglik(equalstrengths,H)
 
 
 n <- 1000  # takes about 5 minutes for n=1000 trials
@@ -16,9 +16,9 @@ for(i in seq_len(n)){
   Hrand <- hyper2(pnames=c(team1,team2))
   for(jj in 1:6){
     pp <- rdeath(team1,team2)
-    Hrand <- Hrand + counterstrike(team1,team2, pp)
+    Hrand <- Hrand + counterstrike_maker(team1,team2, pp)
   }
-  distrib[i] <- loglik(Hrand,indep(maxp(Hrand))) - loglik(Hrand,equalstrengths)
+  distrib[i] <- loglik(indep(maxp(Hrand)),Hrand) - loglik(equalstrengths,Hrand)
   print(i)
 }
 

### 2019-11-20 inst/gbbo_series10.R

@@ -45,7 +45,7 @@ H <- H + f(w10)
 
 ## Now some observed statistics:
 mp_obs  <- maxp(H)
-X_obs <-  2*(loglik(H,indep(mp_obs)) - loglik(H,indep(equalp(H))))
+X_obs <-  2*(loglik(indep(mp_obs),H) - loglik(indep(equalp(H)),H))
 print(paste("asymptotic pvalue = ",pchisq(X_obs,df=12,lower.tail=FALSE)))
 m_obs <- names(which.max(mp_obs))
 name_maxlike_obs <- names(which.max(mp_obs))
@@ -79,7 +79,7 @@ for(i in seq_len(n)){
   
   
   mp <- maxp(Hstar)
-  X[i] <- 2*(loglik(Hstar,indep(mp)) - loglik(Hstar,indep(equalp(Hstar))))
+  X[i] <- 2*(loglik(indep(mp),Hstar) - loglik(indep(equalp(Hstar)),Hstar))
   m[i] <- max(mp)
   name_maxlike[i] <- names(which.max(mp))
 }

### 2019-11-20 inst/kka_3draws.R

@@ -64,7 +64,7 @@ ml_p_free    <- maxp(H)
 
 
 objective <- function(x){
-  loglik(H,c(x[1],x[1],x[1],x[2:4]))
+  loglik(c(x[1],x[1],x[1],x[2:4]),H)
 }
 
 constrained_optimization <-

### 2019-12-08 inst/loser.Rmd
slight copyediting

@@
 
-Now we are going to look at what happens when our observation is only
-the _loser_ (which is here taken to be player `a`).  We have six
-competitors `a,b,c,d,e,f` with nonnegative strengths
-$p_1,p_2,p_3,p_4,p_5,p_6, \sum p_i=1$.  Competitor `a` has strength
-$p_1$, and loses.  We are interested in how informative this
-observation is with regard to the hypothesis that `a` does in fact
-have zero strength.  We have three hypotheses:
+Here we consider a situation in which we make only a rather
+uninformative observation: namely, the identity of the _loser_ (which
+is here taken to be me, player `a`).  We have six competitors
+`a,b,c,d,e,f` with nonnegative strengths $p_1,p_2,p_3,p_4,p_5,p_6,
+\sum p_i=1$.  Competitor `a` has strength $p_1$, and loses.  We are
+interested in how informative this observation is with regard to
+various assertions about the strength of `a`.  We have three
+hypotheses:
 
@@
 
-In the above, `L` is a likelihood function for the observation that
-competitor `a` came last.  In R idiom, it is a list of hyper2 objects
-corresponding to all $5!=120$ possible observed order statistics (only
-the first two, `L[[1]]` and `L[[2]]`, are printed, which correspond to
-orders `bcdefa` and `bcdfea` respectively).  Any order is permissible,
-so long as player `a` is bottom.
+In the above, `L` is a log-likelihood function for the observation
+that competitor `a` came last.  In R idiom, it is a list of hyper2
+objects corresponding to all $5!=120$ possible observed order
+statistics (only the first two, `L[[1]]` and `L[[2]]`, are printed,
+which correspond to orders `bcdefa` and `bcdfea` respectively).  Any
+order is permissible, so long as player `a` is bottom.
 

### 2019-12-08 inst/counterstrike.Rmd
nice Rmd file

---
title: "A statistical analysis of a counterstrike game: Boston 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("partitions") ## needed for perms()
library("magrittr") 
```
# Abstract

In this short vignette I use the `hyper2` package to analyse an
influential Counter-Strike match between two professional teams (_FaZe
Clan_ and _Cloud9_) in the major championships at Boston 2018.  I use
likelihood methods to demonstrate that one player, Skadoodle, has a
legitimate claim to be the strongest player on the field; but the data
is indistinguishable from a null of equal player strength.
Counter-strike dataset kindly supplied by Zachary Hankin.

# Counterstrike

E-sports are a form of competition using video games.  E-sports are
becoming increasingly popular, with high-profile tournaments attracting
over 400 million viewers, and prize pools exceeding US$20m.

_Counter Strike: Global Offensive_ (CS-GO) is a multiplayer
first-person shooter game in which two teams of five compete in an
immersive virtual reality combat environment.  CS-GO is distinguished
by the ability to download detailed gamefiles in which every (sic)
aspect of an entire match is recorded, it being possible to replay the
match at will.

Statistical analysis of such gamefiles is extremely difficult,
primarily due to complex gameplay features such as cooperative
teamwork, within-team communication, and real-time strategic fluidity.

## Boston 2018

"ELEAGUE Major: Boston 2018", here denoted "Boston 2018", was a major
championship held in Boston Massachusetts from 26-28 January 2018
featuring 24 professional teams; the purse was US$1000000.

The final pitted _FaZe Clan_ against _Cloud9_, with _Cloud9_ beating
favourites _FaZe Clan_ 2-1.

# Dataset

The dataset used here is a list whose six elements correspond to six
rounds of play, which are assumed to be statistically independent.
The R idiom is:

```{r, zachslist}
zachslist <- list(
    round1 =
      c("Skadoodle","olofmeister","tarik",
        "GuardiaN", "RUSH", "rain","Stewie2k",
        "karrigan","autimatic","NiKo"
        ),
    round2 =
      c("karrigan", "NiKo", "Stewie2K",
        "RUSH", "rain", "GuardiaN",
        "autimatic", "olofmeister"
        ),
    round3 =
      c("rain","tarik","autimatic",
        "karrigan","RUSH","GuardiaN",
        "Stewie2K","NiKo","olofmeister"
        ),
    round4 =
      c("rain","GuardiaN","karrigan",
        "NiKo","olofmeister"
        ),
    round5 =
      c("olofmeister","rain","karrigan",
        "tarik","Stewie2K","autimatic"
        ),
    round6 =
      c("GuardiaN","karrigan")
)
```

(data kindly supplied by Zachary Hankin).  Thus in round 1, the death
order was Skadoodle first, followed by olofmeister, then tarik.

## Likelihood function for a simple deathorder statistic

Suppose `team1 = (a,b,c,d)` and `team2 = (x,y,z)`; suppose the
deathorder were `(a,b,y,c)`.  Then the likelihood function for `a`'s
death would be `(x+y+z)/(a+b+c+d+x+y+z)` [that is, the likelihood
function for a Bernoulli trial between team `(x,y,z)` and
`(a,b,c,d)`].  Note the appearance of (`a`) in the denominator: he
[surely!] was on the losing team.  The likelihood function for the
full deathorder is then

```
 (x+y+z)/(a+b+c+d+x+y+z) * (x+y+z)/(b+c+d+x+y+z) * (c+d)/(c+d+x+y+z) * (x+z)/(c+d+x+z)
```

where the four terms correspond to the deaths of `a,b,y,c`
respectively.  See how the denominator gets shorter as the teams die
one by one.

# Likelihood function for the observed dataset

```{r, teamsetup}
team1  <- c("autimatic","tarik", "Skadoodle","Stewie2k","RUSH")   #Cloud9
team2 <- c("NiKo","olofmeister","karrigan","GuardiaN","rain")  # FaZe Clan
```

Consider the following function:

```{r, deffun}
`counterstrike_maker` <- function(team1,team2,deathorder){

  if(identical(sort(c(team1,team2)),sort(deathorder))){
    deathorder <- deathorder[-length(deathorder)]  # last player not killed
  }

  H <- hyper2(pnames=c(team1,team2))

  for(killed_player in deathorder){
    if(killed_player %in% team1){
      H[team2] %<>% inc
      H[c(team1,team2)] %<>% dec
      team1 %<>% "["(team1 != killed_player)      
    } else {
      H[team1] %<>% inc
      H[c(team1,team2)] %<>% dec
      team2 %<>% "["(team2 != killed_player)      
    }
  }
  return(H)
}
```

Function `counterstrike_maker()` defined above returns a
log-likelihood function for the strengths of the players in `team1`
and `team2` above.  The function needs the identities of all players
on each team.  It assumes that players are always killed by the
opposing team; if a player is killed by his own team, function
`counterstrike_maker()` may return an error.  The function does not
have strong error-checking functionality.

In the function, the `deathorder` argument specifies the order in
which players were killed (the first element is the first player to be
killed and so on).  Note that the identity of the killer is not needed
as it is assumed that a death is due to the combined strength of the
team, rather than the individual shooter who actually fired the shot.
Note that it is not required for all the players to be killed;
short rounds are OK (but are not as statistically informative).

We can use it:

```{r, useit}
H <- hyper2(pnames=c(team1,team2))
for(i in zachslist){
  H <- H + counterstrike_maker(team1,team2, deathorder=i)
}
```

then plot it:

```{r, plotit}
dotchart(maxp(H),pch=16,main='observed data')
```

showing that Skadoodle has the highest estimated strength.

# Random data

it is possible to analyse synthetic data generated randomly by
function `rdeath()`:

```{r, randomdata}

`rdeath` <- function(team1,team2){
  out <- NULL
  while(length(team1)>0 & length(team2)>0){
    if(runif(1)>length(team1)/(length(c(team1,team2)))){ ## killed person is on team1
      killed_player <- sample(team1,1)
      team1 %<>% "["(team1 != killed_player)      
    } else { ## killed person is on team2
      killed_player <- sample(team2,1)
      team2 %<>% "["(team2 != killed_player) 
    }
    out <- c(out,killed_player)
  }
  return(out)
}

Hrand <- hyper2(pnames=c(team1,team2))
for(i in 1:6){
  pp <- rdeath(team1,team2)
  Hrand <- Hrand + counterstrike_maker(team1,team2, pp)
}

dotchart(maxp(Hrand),pch=16,main='synthetic data')
```

We now run $n=1000$ trials on the assumption of equal player strength
(that is, random deathorder) and for each trial calculate the
log-likelihood ratio [for the hypothesis p=equal_strengths, vs the
hypothesis p=maxp(Hrand)].  It then draws a histogram of the
loglikelihood ratio thus obtained, and superimposes a line showing the
observed loglikelihood ratio.

```{r, random2}
equalstrengths <- rep(0.1, 9)
datapoint <- loglik(indep(maxp(H)),H) - loglik(equalstrengths,H)

n <- 100  # takes about 5 minutes for n=1000 trials
distrib <- rep(0,n)
for(i in seq_len(n)){
  Hrand <- hyper2(pnames=c(team1,team2))
  for(jj in 1:6){
    pp <- rdeath(team1,team2)
    Hrand <- Hrand + counterstrike_maker(team1,team2, pp)
	}
  distrib[i] <- loglik(indep(maxp(Hrand)),Hrand) - loglik(equalstrengths,Hrand)
}

hist(distrib)
abline(v=datapoint,lwd=6)
```

Thus we can see that the observation is not significant: from this
perspective, the deathorder may as well be totally random.

### 2019-12-12 inst/counterstrike.Rmd
whitespace, more compact display

@@
 zachslist <- list(
-    round1 =
-      c("Skadoodle","olofmeister","tarik",
-        "GuardiaN", "RUSH", "rain","Stewie2k",
-        "karrigan","autimatic","NiKo"
-        ),
-    round2 =
-      c("karrigan", "NiKo", "Stewie2K",
-        "RUSH", "rain", "GuardiaN",
-        "autimatic", "olofmeister"
-        ),
-    round3 =
-      c("rain","tarik","autimatic",
-        "karrigan","RUSH","GuardiaN",
-        "Stewie2K","NiKo","olofmeister"
-        ),
-    round4 =
-      c("rain","GuardiaN","karrigan",
-        "NiKo","olofmeister"
-        ),
-    round5 =
-      c("olofmeister","rain","karrigan",
-        "tarik","Stewie2K","autimatic"
-        ),
-    round6 =
-      c("GuardiaN","karrigan")
+    round1 = c("Skadoodle","olofmeister","tarik","GuardiaN","RUSH",
+         "rain", "Stewie2k","karrigan","autimatic","NiKo"),
+    round2 = c("karrigan", "NiKo", "Stewie2K", "RUSH", "rain", "GuardiaN",
+         "autimatic", "olofmeister"),
+    round3 = c("rain","tarik","autimatic", "karrigan","RUSH","GuardiaN",
+         "Stewie2K","NiKo","olofmeister"),
+    round4 = c("rain","GuardiaN","karrigan", "NiKo","olofmeister"),
+    round5 = c("olofmeister","rain","karrigan", "tarik","Stewie2K","autimatic"),
+    round6 = c("GuardiaN","karrigan")
 )

### 2019-12-12 inst/counterstrike.Rmd
use equalp()

@@
 ```{r, random2}
-equalstrengths <- rep(0.1, 9)
-datapoint <- loglik(indep(maxp(H)),H) - loglik(equalstrengths,H)
+
+datapoint <- loglik(indep(maxp(H)),H) - loglik(indep(equalp(H)),H)
 
@@
 	}
-  distrib[i] <- loglik(indep(maxp(Hrand)),Hrand) - loglik(equalstrengths,Hrand)
+  distrib[i] <- loglik(indep(maxp(Hrand)),Hrand) - loglik(indep(equalp(H)),Hrand)
 }

### 2019-12-15 inst/formula1.Rmd

---
title: "formula1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("magrittr")
```

## Formula 1

Formula 1 racing is an important and prestigious motor sport.  Here I
analyse a series of racing results using the `hyper` package which
implements the Plackett-Luce likelihood function.

Consider the following dataset, taken from Wikipedia:

```{r cars}
f2017 <- read.table("formula1_2017.txt",header=TRUE)
head(f2017)
```

Each row is a driver and each column (after the first) a venue.  We
see that Hamilton, the first row, came second in Australia, first in
China, second in Bahrain, fourthin Russia, and so on.  In the first
columnm, we see the result from Australia in which Hamilton came
second, Vettel first, Bottas third, and so on.  Notation used also
includes six different classes of no-score such as "Ret" for retired,
"WD" for withdrawn, and so on.  I now define a bespoke R function that
translates data of this type into a log-likelihood function using the
`hyper2` package.  The function is similar to, but slightly different
from, the analysis in `eurovision.R`.

```{r,defF1}
F1_likelihood <- function(wiki_table){

  noscore <- c("Ret", "WD", "DNS", "DSQ", "DNP", "NC")
  first_and_last <- c(1,ncol(wiki_table))

  racers <- wiki_table[,1]
  venues <- colnames(wiki_table)[-first_and_last]

  ## Now create a numeric matrix, fmat.  Two steps: first, count
  ## any no-score as zero:

  f <- function(x){
    if(any(x %in% noscore)){x[x%in%noscore] <- 0}
    return(x)
  }

  jj <- apply(wiki_table,2,f)

  ## Second, convert to numeric and strip out names; transpose of
  ## wiki_table (because we want each row to be a venue):

  fmat <- matrix(as.numeric(jj[,-first_and_last]),byrow=TRUE,ncol=nrow(wiki_table)) 
  colnames(fmat) <- racers
  rownames(fmat) <- venues

  F1 <- hyper2(d=ncol(fmat))

  ## Now cycle through the rows; each row is a venue [voter]
  for(i in seq_len(nrow(fmat))){
    d <- fmat[i,,drop=TRUE]
    while(any(d>0)){
      eligible <- which(d>=0)

      ## The first choice among eligible players has +1 power on the
      ## numerator:
      F1[which(d==1)] %<>% "+"(1)

      ## denominator of all eligible players; power -1
      F1[eligible] %<>% "-"(1)

      ## once you've come first in the field, you are ineligible to be first again:
      d[d==1] <- -1

      ## everyone moves down the list, so who *was* in second place
      ## becomes first place, who *was* third place becomes second,
      ## and so on:
      d[d>0] %<>% "-"(1)

    } # while() loop closes
  } # i loop closes

  ## syntatic sugar:
  pnames(F1) <- racers

  return(F1)
}
```

We can apply this to the dataset:

```{r,applydata}
L2017 <- F1_likelihood(f2017)
head(L2017,3)
```

Above, we see a small part of the Placket-Luce likelihood function for
the 2017 results.  We can find the maximum likelihood estimator by
using the `maxp()` function:

```{r,maxp2017}
dotchart(maxp(L2017),pch=16,main='2017 Formula 1')
dotchart(log(maxp(L2017)[1:23]),pch=16,main='2017 Formula 1, log scale')
```

# Likelihood scoring vs points scoring

The championship winning order is determined by a points system: the
winner is awarded 25 points, second place 18 points, and so on.  We
can now check to see how the points total matches up with the
likelihood estimation.

```{r,likelihoodandscore}
points <- f2017$points
names(points) <- f2017$driver

m <- maxp(L2017)
ox <- order(points,decreasing=TRUE)
oy <- order(m,decreasing=TRUE)
par(pty='s') # square plot
plot(ox,oy,asp=1,pty='s',xlim=c(0,25),ylim=c(0,25),pch=16,
xlab="official order",ylab="my order",main='Formula 1, 2017 season')
par(xpd=TRUE) # allow drivers' names to appear outside plotting region
for(i in seq_along(ox)){text(ox[i],oy[i],names(m)[ox[i]],pos=4,col='gray') }
par(xpd=FALSE) # stop diagonal line from protruding beyond plotting region
abline(0,1)
```

# Analysis of nulls:

## Null of equal strengths:

```{r,equalstrength}
L1 <- loglik(indep(equalp(L2017)),L2017)
L2 <- loglik(indep(maxp(L2017)),L2017)
pchisq(2*(L2-L1),df=1,lower.tail=FALSE)
```

highly signficant!

## Null of Hamilton being average

This script takes about a minute to run.  It creates hyper2 objects
`F1_2014` through `F1_2017`, which are the likelihood functions for
the F1 results table for each of the years.

Function `F1_likelihood()` is defined here; this creates objects such as
`F1_2017` by processing `formula1_2017.txt`.

At the end of the script, hyper2 object `F1_total` is created, which
is a likelihood function for all four years' data.  This object is
called 'formula1' in the package.  File `F1_allyears.R` is functionally
identical but uses slightly different R idiom.

### 2019-12-15 inst/formula1.Rmd
more work

@@
 
+In the graph above, we see that both systems consider Hamilton to be
+the best.  However, F1 points place Vettel second and Bottas third,
+while `hyper2` places Bottas second and Vettel third.  Overall, there
+is a high degree of correlation between the two, but differences in
+detail.
+
 # Analysis of nulls:
@@
 L2 <- loglik(indep(maxp(L2017)),L2017)
-pchisq(2*(L2-L1),df=1,lower.tail=FALSE)
+pchisq(2*(L2-L1),df=25-1,lower.tail=FALSE)
 ```

### 2019-12-16 inst/formula1.Rmd
more hypotheses

@@
 
+(in the diagram above, we remove the bottom-ranked two players, Resta
+and Button, whose estimated strength is zero).
+
 # Likelihood scoring vs points scoring
@@
 
-This script takes about a minute to run.  It creates hyper2 objects
+We can consider Hamilton and ask various questions about his playing
+strength.  One natural null would be that Hamilton's strength is equal
+to the mean strength, or $\frac{1}{25}$.
+
+* $H_2\colon p_\mathrm{Hamilton}\geq\frac{1}{25}$
+* $H_3\colon p_\mathrm{Hamilton} < \frac{1}{25}$
+
+Noting that the evaluate for Hamilton's strength
+$\widehat{p_\mathrm{Hamilton}}$ is about 0.22, we may consider the
+likelihood for $H_2$ to be equal to the unconstrained maximum
+likelhood as in `L2` above.  The likelihood for $H_3$ requires a
+constrained optimization:
+
+```{r,constham}
+small <- 0.01
+L3 <- maxp(L2017, startp=c(small, rep((1-small)/24,23)),fcm=rbind(c(-1,rep(0,23))),fcv=-1/25,give=TRUE)
+L3
+```
+
+(note that the constrained maximum attains the constraint of
+$p_\mathrm{Hamilton}=\frac{1}{25}$).   The likelihood ratio test gives
+
+```{r,liktestivediff}
+L2 - L3$value 
+```
+
+or a p-value of
+
+```{r,liktestive}
+pchisq(2*(L2 - L3$value),df=1,lower.tail=FALSE)
+```
+
+## Null of Hamilton being better than other players
+
+We may now test
+
+* $H_4\colon p_\mathrm{Hamilton} > \max_{i\neq\mathrm{Hamilton}} p_i$
+
+This is quite difficult to evaluate formally but we can evaluate it with no loss in accuracy by observing that the second strongest competitor is Bottas and considering instead
+
+* $H_4'\colon p_\mathrm{Hamilton} > p_\mathrm{Bottas}$.
+
+```{r,L4like}
+small <- 0.01
+L4 <- maxp(L2017, startp=c(small, rep((1-small)/24,23)),fcm=rbind(c(-1,1,rep(0,22))),fcv=0,give=TRUE)
+L4
+```
+
+Calculating the support and asymptotic p-values:
+```{r,l4val}
+L2 - L4$value 
+pchisq(2*(L2 - L4$value),df=1,lower.tail=FALSE)
+```
+
+this script takes about a minute to run.  It creates hyper2 objects
 `F1_2014` through `F1_2017`, which are the likelihood functions for

### 2019-12-16 inst/formula1.Rmd
use inc and dec

@@
       ## numerator:
-      F1[which(d==1)] %<>% "+"(1)
+      F1[which(d==1)] %<>% inc
 
       ## denominator of all eligible players; power -1
-      F1[eligible] %<>% "-"(1)
+      F1[eligible] %<>% dec
 
@@
       ## and so on:
-      d[d>0] %<>% "-"(1)
+      d[d>0] %<>% dec
 

### 2019-12-16 inst/formula1.Rmd
all years 2014-2017

@@
 
-this script takes about a minute to run.  It creates hyper2 objects
-`F1_2014` through `F1_2017`, which are the likelihood functions for
-the F1 results table for each of the years.
+## All years data
 
-Function `F1_likelihood()` is defined here; this creates objects such as
-`F1_2017` by processing `formula1_2017.txt`.
+We will consider four years data:
 
-At the end of the script, hyper2 object `F1_total` is created, which
-is a likelihood function for all four years' data.  This object is
-called 'formula1' in the package.  File `F1_allyears.R` is functionally
-identical but uses slightly different R idiom.
+```{r,fouryearsdata}
+F1_2014 <- "formula1_2014.txt" %>% read.table(header=TRUE) %>% F1_likelihood
+F1_2015 <- "formula1_2015.txt" %>% read.table(header=TRUE) %>% F1_likelihood
+F1_2016 <- "formula1_2016.txt" %>% read.table(header=TRUE) %>% F1_likelihood
+F1_2017 <- "formula1_2017.txt" %>% read.table(header=TRUE) %>% F1_likelihood
+a <- list(F1_2014, F1_2015, F1_2016, F1_2017)
+alldrivers <- all_pnames(a)
+
+F1_total <- hyper2(pnames=alldrivers)
+
+F1_total %<>% add(change_pnames(F1_2014,alldrivers)) 
+F1_total %<>% add(change_pnames(F1_2015,alldrivers)) 
+F1_total %<>% add(change_pnames(F1_2016,alldrivers)) 
+F1_total %<>% add(change_pnames(F1_2017,alldrivers))
+
+(mallyears <- maxp(F1_total))
+```
+
+and the plot:
+
+```{r,allyearplot,fig.height=7}
+dotchart(mallyears,pch=16,main='Formula 1, 2014-7')
+```

### 2019-12-23 inst/eurovision.Rmd
fixes #24

---
title: "Eurovision Song Contest"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("magrittr")
abbreviated <- TRUE  # change to FALSE for full names
```

## The Eurovision Song contest

The dataset `wiki_matrix`, defined below, is copied from "Eurovision
Song Contest 2009," Wikipedia, accessed May 13, 2018.  It refers to
semi-final 1.  More documentation is given in `euro.Rd` [type
`help(euro2009)` at the `R` prompt].

First we specify the matrix as appearing in the Wikipedia page:

```{r, specifymatrix}
wiki_matrix <- matrix(c(
#   ME  CZ  BE  BY  SW  AM  AD  CH  TR  IL  BG  IS  MK  RO  FI  PT  MT  BA  DE  UK
    NA, 00, 00, 03, 00, 05, 01, 02, 05, 01, 00, 00, 08, 00, 00, 01, 06, 10, 02, 00, # ME
    00, NA, 00, 00, 00, 00, 00, 00, 00, 00, 00, 00, 00, 00, 00, 00, 00, 00, 00, 00, # CZ
    00, 00, NA, 00, 00, 01, 00, 00, 00, 00, 00, 00, 00, 00, 00, 00, 00, 00, 00, 00, # BE
    02, 01, 00, NA, 01, 04, 00, 00, 00, 04, 01, 01, 06, 00, 04, 00, 01, 00, 00, 00, # BY
    00, 06, 04, 07, NA, 08, 07, 04, 04, 07, 00, 10, 03, 04, 10, 08, 08, 04, 04, 07, # SW
    04, 12, 10, 10, 05, NA, 00, 01, 10, 10, 08, 02, 02, 08, 01, 00, 00, 01, 10, 05, # AM
    00, 00, 00, 00, 00, 00, NA, 00, 01, 00, 00, 00, 00, 00, 00, 04, 03, 00, 00, 00, # AD
    00, 00, 00, 02, 02, 00, 02, NA, 00, 00, 00, 00, 00, 00, 05, 02, 00, 02, 00, 00, # CH
    08, 05, 12, 06, 07, 10, 05, 12, NA, 06, 12, 07, 12, 12, 07, 05, 10, 12, 12, 12, # TR
    05, 04, 03, 04, 06, 07, 08, 05, 03, NA, 04, 06, 01, 03, 06, 00, 04, 00, 05, 01, # IL
    00, 00, 00, 00, 00, 00, 00, 00, 02, 00, NA, 00, 05, 00, 00, 00, 00, 00, 00, 00, # BG
    07, 10, 07, 12, 12, 12, 10, 07, 08, 12, 06, NA, 04, 10, 12, 12, 12, 07, 06, 08, # IS
    10, 03, 00, 00, 00, 00, 00, 06, 06, 00, 10, 00, NA, 02, 00, 00, 00, 08, 00, 00, # MK
    06, 00, 02, 01, 00, 02, 04, 00, 07, 08, 05, 04, 07, NA, 00, 10, 02, 06, 01, 02, # RO
    03, 00, 01, 00, 10, 00, 03, 00, 00, 00, 00, 12, 00, 01, NA, 03, 05, 00, 00, 04, # FI
    00, 02, 06, 00, 03, 00, 12, 10, 00, 02, 02, 08, 00, 07, 02, NA, 00, 03, 07, 06, # PT
    01, 07, 08, 08, 04, 03, 06, 03, 00, 05, 03, 05, 00, 06, 03, 06, NA, 05, 03, 10, # MT
    12, 08, 05, 05, 08, 06, 00, 08, 12, 03, 07, 03, 10, 05, 08, 07, 07, NA, 08, 03) # BA
, nrow=18,byrow=TRUE)
```

Each row corresponds to a competitor and each column to a judge.  Note
that the matrix is not square: the last two columns correspond to
Germany and the UK, who were judges but not competitors.  I have
removed the first column in the Wikipedia table (which gave the total
of the points), replaced self-voting entries with `NA`, and replaced
blanks with `00`; every entry is padded to two digits.

```{r,pointsforplaces}
points <- c(12,10,8,7,6,5,4,3,2,1)
```

Variable `points` gives the number of points awarded, under Eurovision
rules, to voters' first, second, third, etc choice.  Points for any
competitor are added and the winner is the competitor with the most
points.  However, in this model, the numerical values themselves do
not affect the likelihood function; only the order of the voters'
preferences matters.

```{r,prefsetter}
preference <- wiki_matrix*0  
for(i in seq_along(points)){ preference[wiki_matrix == points[i]] <- i }

countries <- data.frame(
    fullname =
        c("Montenegro", "Czech rep", "Belgium", "Belarus", "Sweden",
          "Armenia", "Andorra", "Switzerland", "Turkey", "Israel",
          "Bulgaria", "Iceland", "Macedonia", "Romania", "Finland",
          "Portugal", "Malta", "Bosnia Herz", "Germany", "UK"),
    abbreviation = c("ME","CZ","BE", "BY", "SW", "AM", "AD", "CH",
                     "TR", "IL", "BG", "IS", "MK", "RO", "FI", "PT",
                     "MT", "BA", "DE", "UK")
)

if(abbreviated){ 
    jj <- countries$abbreviation
} else { 
    jj <- countries$fullname
}

competitors <- jj[1:18]
colnames(preference) <- jj      # voters; 20 countries (18 + DE + UK)
rownames(preference) <- competitors

rownames(wiki_matrix) <- jj[1:18]
colnames(wiki_matrix) <- jj
```

In the above, matrix `preference` records voters' first, second, third,
etc choice.  A zero entry means no points (nul punkte!) and `NA` means
that voter was forbidden from voting for that player (countries cannot
vote for themselves).

The competitors were the first 18 countries (the last two countries,
Germany and the UK, voted but did not compete).  We need the voters'
choices to be the *rows* for consistency with the rest of the package
(and, for that matter, the aylmer package and indeed the emulator
package), so we take the transpose:

```{r,taketranspose}
preference <- t(preference)
preference
```

Now, take the first row of 'preference'.  This represents the votes
cast *by* (sic) Montenegro (`"ME"`).  Their favourite was [last column]
Bosnia & Herzegovina (`"BH"`), who they gave rank 1 to.  Their second
favourite was Macedonia, their third was Turkey, and so on.  They were
not allowed to vote for themselves, which is why the first column is
`NA`.  So the order was: `bh`, `mc`, `tu`, `ic`, `ro`, `is`, `ar`,
`fi`, `br`, `ma`.

```{r,morestuff}
euro2009 <- hyper2(d=18)

for(i in seq_len(nrow(preference))){   # cycle through the rows; each row is a voter
    d <- preference[i,,drop=TRUE]
    d[is.na(d)] <- -1  # kludge: make the voting country ineligible to vote.
    while(any(d>0)){
        eligible <- which(d>=0)   # This is why we set NA values to -1
        euro2009[which(d==1)] %<>% "+"(1)
                                        # The first choice among
                                        # eligible players has +1
                                        # power on the numerator

        euro2009[eligible] %<>% dec()   # denominator of all eligible players

        d[d==1] <- -1  # once you've won, you are ineligible to be chosen again
                       # NB, pipe idiom is ugly:  d %<>% `[<-`(. == 1, -1)

        d[d>0] %<>% dec()  # everyone moves down the list, so who
                            # *was* second choice becomes first
                            # choice, who *was* third choice becomes
                            # second, and so on.

    } # while() loop closes
} # i loop closes

## syntatic sugar:
pnames(euro2009) <- competitors
```

Now we can check null of equal strengths:

```{r,checknull}
print(loglik(indep(maxp(euro2009)),euro2009))
print(loglik(indep(equalp(euro2009)),euro2009))
```

Thus the difference is about 122 units of support, surely significant.

```{r,plotdiffstrength}
plot(rowSums(wiki_matrix,na.rm=TRUE),maxp(euro2009),
     pch=16,xlab="points scored",ylab="MLE strength")
```

### 2019-12-23 inst/eurovision.Rmd
work

@@
 ---
-title: "Eurovision Song Contest"
+title: "Eurovision Song Contest: semifinal 1, 2009"
 output: html_document
@@
 of the points), replaced self-voting entries with `NA`, and replaced
-blanks with `00`; every entry is padded to two digits.
+blanks with `00`; every entry is padded to two digits.  Thus the first
+row corresponds to votes cast for `ME`, Montenegro.  We see that `BE`
+(Belgium) awarded them three points, `am` (Armenia) five points, and
+so on.
 
@@
 Now, take the first row of 'preference'.  This represents the votes
-cast *by* (sic) Montenegro (`"ME"`).  Their favourite was [last column]
-Bosnia & Herzegovina (`"BH"`), who they gave rank 1 to.  Their second
-favourite was Macedonia, their third was Turkey, and so on.  They were
-not allowed to vote for themselves, which is why the first column is
-`NA`.  So the order was: `bh`, `mc`, `tu`, `ic`, `ro`, `is`, `ar`,
-`fi`, `br`, `ma`.
+cast *by* (sic) Montenegro (`"ME"`).  Their favourite was [last
+column] Bosnia & Herzegovina (`"BA"`), who they gave rank 1 to.  Their
+second favourite was Macedonia, their third was Turkey, and so on.
+They were not allowed to vote for themselves, which is why the first
+column is `NA`.  So the order was: `BA` (1), `MK` (2) , `TR` (3), `IS`
+(4), `RO` (5), `IL` (6), `AM` (7), `FI` (8), `BY` (9), `MT` (10).  The
+other countries (`CZ`, `BE`, `SW`, `CH`, `BG`, `PT`) did not receive
+any points from Montenegro.
+
+We need to convert matrix `preference` into a likelihood function,
+here `euro2009`:
 
@@
 
-```{r,plotdiffstrength}
-plot(rowSums(wiki_matrix,na.rm=TRUE),maxp(euro2009),
+```{r,orderplot}
+m <- maxp(euro2009)  # likelihoods
+points_scored <- rowSums(wiki_matrix,na.rm=TRUE)  # points
+ox <- rank(-points_scored)
+oy <- rank(-m)
+par(pty='s') # square plot
+plot(ox,oy,asp=1,pty='s',xlim=c(0,18),ylim=c(0,18),pch=16,
+xlab="official order",ylab="my order",main="Eurovision 2009")
+par(xpd=TRUE) # allow country names to appear outside plotting region
+for(i in seq_along(ox)){text(ox[i],oy[i],names(m)[i],pos=4,col='gray') }
+par(xpd=FALSE) # stop diagonal line from protruding beyond plotting region
+abline(0,1)
+```
+
+In the graph above, we see that both systems agree on the first four places.
+We can also plot the numerical values:
+
+```{r}
+plot(log(rowSums(wiki_matrix,na.rm=TRUE)),log(maxp(euro2009)),
      pch=16,xlab="points scored",ylab="MLE strength")

### 2019-12-23 inst/eurovision.Rmd
typo

@@
 par(xpd=TRUE) # allow country names to appear outside plotting region
-for(i in seq_along(ox)){text(ox[i],oy[i],names(m)[i],pos=4,col='gray') }
+for(i in seq_along(ox)){text(ox[i],oy[i],names(m)[ox[i]],pos=4,col='gray') }
 par(xpd=FALSE) # stop diagonal line from protruding beyond plotting region

### 2019-12-23 inst/formula1.Rmd
slight tweak
rank not order

@@
 m <- maxp(L2017)
-ox <- order(points,decreasing=TRUE)
-oy <- order(m,decreasing=TRUE)
+ox <- rank(-points)
+oy <- rank(-m)
 par(pty='s') # square plot
@@
 
-In the graph above, we see that both systems consider Hamilton to be
-the best.  However, F1 points place Vettel second and Bottas third,
-while `hyper2` places Bottas second and Vettel third.  Overall, there
-is a high degree of correlation between the two, but differences in
-detail.
+In the graph above, we see a certain amount of disagreement.  
+For example, using F1 points scored places Hamilton first and Vettel
+second; while `hyper2` places Vettel first and Hamilton second.  But
+overall there seems to be some degree of correlation between the two,
+and differences in detail.
 
@@
 
-(mallyears <- maxp(F1_total))
+mallyears <- maxp(F1_total)
 ```
@@
 ```
+
+We see that Hamilton has the highest estimated strength.

### 2019-12-23 inst/formula1.Rmd
typo

@@
 par(xpd=TRUE) # allow drivers' names to appear outside plotting region
-for(i in seq_along(ox)){text(ox[i],oy[i],names(m)[ox[i]],pos=4,col='gray') }
+for(i in seq_along(ox)){text(ox[i],oy[i],names(m)[oy[i]],pos=4,col='gray') }
 par(xpd=FALSE) # stop diagonal line from protruding beyond plotting region
@@
 
-In the graph above, we see a certain amount of disagreement.  
-For example, using F1 points scored places Hamilton first and Vettel
+In the graph above, we see a certain amount of disagreement.  For
+example, using F1 points scored places Hamilton first and Vettel
 second; while `hyper2` places Vettel first and Hamilton second.  But

### 2019-12-23 inst/formula1.Rmd
better inline docs

@@
 m <- maxp(L2017)
-ox <- rank(-points)
-oy <- rank(-m)
+ox <- rank(-points) # negative because number 1 = largest number of points
+oy <- rank(-m)      # negative because number 1 = largest strength
 par(pty='s')        # square plot

### 2019-12-25 inst/cook.Rmd
more work on new maxp()

@@
 o <- function(Ul,Cl,startp,give=FALSE){
-    small <- 1e-5  #  ensure start at an interior point
+    small <- 1e-4  #  ensure start at an interior point
     if(missing(startp)){startp <- small*(1:4)+rep(0.1,4)}

### 2019-12-25 inst/eurovision.Rmd
more work on new maxp()

--- 2019-12-23 inst/eurovision.Rmd
@@
 ```{r,checknull}
-print(loglik(indep(maxp(euro2009)),euro2009))
+m <- maxp(euro2009)  # MLE
+print(loglik(indep(m),euro2009))
 print(loglik(indep(equalp(euro2009)),euro2009))
@@
 ```{r,orderplot}
-m <- maxp(euro2009)  # likelihoods
 points_scored <- rowSums(wiki_matrix,na.rm=TRUE)  # points
 ox <- rank(-points_scored)
-oy <- rank(-m)
+oy <- rank(-m)   # m = MLE
 par(pty='s') # square plot
@@
 ```{r}
-plot(log(rowSums(wiki_matrix,na.rm=TRUE)),log(maxp(euro2009)),
+plot(log(rowSums(wiki_matrix,na.rm=TRUE)),log(m),
      pch=16,xlab="points scored",ylab="MLE strength")

### 2019-12-25 inst/formula1.Rmd
more work on new maxp()

--- 2019-12-23 inst/formula1.Rmd
@@
 par(xpd=TRUE)       # allow drivers' names to appear outside plotting region
-for(i in seq_along(ox)){text(ox[i],oy[i],names(m)[oy[i]],pos=4,col='gray') }
+for(i in seq_along(ox)){text(ox[i],oy[i],names(ox)[i],pos=4,col='gray') }
 par(xpd=FALSE)      # stop diagonal line from protruding beyond plotting region
@@
 In the graph above, we see a certain amount of disagreement.  For
-example, using F1 points scored places Hamilton first and Vettel
-second; while `hyper2` places Vettel first and Hamilton second.  But
-overall there seems to be some degree of correlation between the two,
-and differences in detail.
+example, using F1 points scored places Hamilton first (363), Vettel
+second (317), and Bottas third (305); while `hyper2` places Hamilton
+first (0.29), Bottas second (0.17), and Vettel third (0.14).  Note
+also the large number of players with zero points (who rank equal 23rd
+by the F1 points system) but are nevertheless separated by
+Bradley-Terry.  These players appear as a line of points at $x=23$:
+Resta to Ericcson.  Note also Palmer and Alonso who managed to score a
+F1 point but rank below Ericcson according to Bradley-Terry.
+
+But overall there
+seems to be some degree of correlation between the two, and
+differences in detail.
 

### 2019-12-25 inst/home_advantage.Rmd
slight tweaks

@@
 ```{r}
-small <- 1e-4
-f <- function(home_strength){
-  maxp(H,startp=rep((1-small)/4,4),give=TRUE,fcm=t(rep(1,4)),fcv=1-home_strength)$value
+f <- function(home){
+	small <- 1e-3
+	maxp(H,startp=rep((1-small)/4,4),give=TRUE,fcm=t(rep(1,4)),fcv=1-home)$value
 }
@@
 See how the ```fcm``` and ```fcv``` constraints make the sum of
-a,b,c,d, greater than 1-home_strength.  Thus 'home' [which is the
-fillup value] is constrained to be less than home_strength.  Using
-this function, the likelihood ratio test is straightforward:
+a,b,c,d, greater than ```1-home```.  Thus 'home' [which is the fillup
+value, $p_\mathrm<{home}=1-(p_A+p_B+p_C+p_D)$] is constrained to be
+less than argument ```home```.  Using this function, the likelihood
+ratio test is straightforward:
 
 ```{r}
-Lambda <- -2*(f(2e-4)-maxp(H,give=TRUE)$value)
+Lambda <- -2*(f(0.01)-maxp(H,give=TRUE)$value)
 Lambda
@@
 a chi-square with 1 degree of freedom (clearly rejected).  However, it
-is always interesting to plot likelihood functions.To plot a profile
+is always interesting to plot likelihood functions.  To plot a profile
 likelihood we can constrain the home to be greater than a particular
@@
 ```{r}
-g <- function(home_strength){
-  maxp(H,startp=rep(small/4,4),give=TRUE,fcm=-t(rep(1,4)),fcv=home_strength-1)$value
+g <- function(home){
+	small <- 1e-3
+	maxp(H,startp=rep(small/4,4),give=TRUE,fcm=-t(rep(1,4)),fcv=home-1)$value
 }

### 2019-12-25 inst/order.Rmd
more tweaks on randomizer branch

@@
 f <- function(p,n){c(p,rep((1-p)/n,n-1))}
-sss <- f(0.9999,9)  # sensible start point, close (?) to the evaluate
+sss <- f(0.99,9)  # sensible start point, close (?) to the evaluate
 evaluate2 <- maxp(H,startp=sss)
@@
 ```{r}
-evaluate3 <- maxp(H,vsssp(4,10))
+evaluate3 <- maxp(H,vsssp(2,10))
 maxlike3 <- loglik(indep(evaluate3),H)

### 2020-01-07 inst/cook.Rmd
changes in cooking strength

@@
 infer that it was suspect.
+
+## Changes in cooking strength.
+
+We now consider a sequence of observations where one person, say
+person 1, is suspected of an increase in strength (perhaps they
+attended a domestic cookery course).  We make some observations before
+and after the training and seek evidence that it was effective.
+Unlike the previous cases, here we have a full order statistic without
+the complication of not being allowed to vote for oneself (perhaps a
+professional chef puts the dishes in order).  First, before:
+
+```{r,beforetraining}
+H <- hyper2(pnames=c(paste("p",1:5,sep=""),"t"))  # 't' the strength of training
+# observations before training:
+Hbefore <- H +
+    order_likelihood(c(5,2,3,4,1)) + #  5 is the best, 4 the worst.
+    order_likelihood(c(5,3,2,4,1)) + #  5 is the best, 1 the worst.
+    order_likelihood(c(4,5,3,2,1)) + #  4 is the best, 1 the worst.
+    order_likelihood(c(5,3,4,1,2)) + #  5 is the best, 2 the worst.
+    order_likelihood(c(2,5,3,1,4))   #  2 is the best, 4 the worst.
+Hbefore
+```
+
+Now, after the training we have two order observations: `1,2,3,4,5`
+and `3,1,5,2,4`.  Thus player 1 comes first and second (after player
+3) respectively.  The likelihood for the first would be
+
+\[
+\frac{p_1+t}{p_1+t+p_2+p_3+p_4+p_5}\cdot
+\frac{p_2  }{      p_2+p_3+p_4+p_5}\cdot
+\frac{p_3  }{          p_3+p_4+p_5}\cdot
+\frac{p_4  }{              p_4+p_5}\cdot
+\frac{p_5  }{                  p_5}\cdot
+\]
+
+and the likelihood for the second would be
+
+\[
+\frac{p_3  }{p_1+t+p_2+p_3+p_4+p_5}\cdot
+\frac{p_1+t}{p_1+t+p_2+    p_4+p_5}\cdot
+\frac{p_5  }{      p_2+    p_4+p_5}\cdot
+\frac{p_2  }{      p_2+    p_4    }\cdot
+\frac{p_4  }{              p_4    }\cdot
+\]
+
+We can translate these into package idiom by defining likelihood
+functions `H1after` and `H2after` for the two observations:
+
+```{r,aftertraining1}
+H1after <- H
+H1after[c("p1","t")] %<>% inc()
+H1after[  "p2"     ] %<>% inc()
+H1after[  "p3"     ] %<>% inc()
+H1after[  "p4"     ] %<>% inc()
+H1after[  "p5"     ] %<>% inc()
+H1after[c("p1","t","p2","p3","p4","p5")] %<>% dec()
+H1after[c(         "p2","p3","p4","p5")] %<>% dec()
+H1after[c(              "p3","p4","p5")] %<>% dec()
+H1after[c(                   "p4","p5")] %<>% dec()
+H1after[c(                        "p5")] %<>% dec()
+H1after
+```
+
+```{r,aftertraining2}
+H2after <- H
+H2after[  "p3"     ] %<>% inc()
+H2after[c("p1","t")] %<>% inc()
+H2after[  "p5"     ] %<>% inc()
+H2after[  "p2"     ] %<>% inc()
+H2after[  "p4"     ] %<>% inc()
+H2after[c("p1","t","p2","p3","p4","p5")] %<>% dec()
+H2after[c("p1","t","p2",     "p4","p5")] %<>% dec()
+H2after[c(         "p2",     "p4","p5")] %<>% dec()
+H2after[c(         "p2",     "p4"     )] %<>% dec()
+H2after[c(                   "p4"     )] %<>% dec()
+H2after
+```
+
+Combine:
+
+```{r,combineboth}
+(Htot <- Hbefore + H1after + H2after)  # observations are independent!
+(ML_unconstrained <- maxp(Htot))
+```
+
+We see the training entity has an estimated strength of about 0.63,
+higher than all real players combined.  At the evaluate the
+loglikelihood is
+
+```{r,loglikateval}
+loglik(indep(ML_unconstrained),Htot)
+```
+
+We can test the hypothesis that the training was ineffective by
+constraining the optimization to $t=0$:
+
+```{r}
+small <- 1e-4
+(ML_constrained <- maxp(Htot,startp=rep(0.2*(1-small*2.1),5),fcm=c(1,1,1,1,1),fcv=1-2*small))
+loglik(indep(ML_constrained),Htot)
+```
+
+So the support for our alternative hypothesis would be
+
+```{r}
+loglik(indep(ML_unconstrained),Htot)-loglik(indep(ML_constrained),Htot)
+```
+
+thus exceeding Edwards's two units of support.  We have strong
+evidence to suggest that the training was effective.

### 2020-01-07 inst/cook.Rmd
consistency check

@@
 
-So the support for our alternative hypothesis would be
+[as a consistency check we can perform the same calculation with a
+slightly different value for `small`:
+
+```{r}
+small <- 2e-4  # cf 1e-4 previously
+ML_constrained <- maxp(Htot,startp=rep(0.2*(1-small*2.1),5),fcm=c(1,1,1,1,1),fcv=1-2*small)
+loglik(indep(ML_constrained),Htot) # cf -29.973 previously
+```
+
+and we can see that this is almost the same].  Thus, the support for
+our alternative hypothesis would be
 

### 2020-01-09 inst/cook.Rmd
clarification

@@
 Now, after the training we have two order observations: `1,2,3,4,5`
-and `3,1,5,2,4`.  Thus player 1 comes first and second (after player
-3) respectively.  The likelihood for the first would be
+and `3,1,5,2,4`.  Thus in the first observation, player 1 comes first
+and in the second observation, player 1 comes second second (after
+player 3).  The likelihood for the first observation would be
 
@@
 
-and the likelihood for the second would be
+and the likelihood for the second observation would be
 

### 2020-01-26 inst/order.Rmd
more work on the limiting process
more work

@@
 ---
-title: "Order statistics and numerical optimization"
+title: "Order statistics and numerical optimization: the case of a single observation"
 output: html_document
@@
 \[
-\lim_{p\longrightarrow 0}
-{\mathcal L}(1-9p,p,p,p,p,p,p,p,p,p,p)=\frac{1}{8!}
+\lim_{x\longrightarrow 0}
+{\mathcal L}(1-9x,x,x,x,x,x,x,x,x,x,x)=\frac{1}{8!};\qquad
+\lim_{x\longrightarrow 0}
+{\mathcal L}(\alpha,\alpha x,\alpha x^2,\ldots,\alpha x^9)=1
 \]
 
-but
+where $\alpha=(1-x^{10})/(1-x)$ is a normalizing constant.  Thus the
+likelihood _at_ the evaluate is not defined, even as a limit (because
+it depends on the direction of approach).  However, it is reasonable
+to demand consistency in the limiting process.  Thus, writing
+$p_i(x),1\leq i\leq 10$, we require
 
 \[
-\lim_{p\longrightarrow 0}
-{\mathcal L}(\alpha,\alpha p,\alpha p^2,\ldots,\alpha p^9)=1
+\lim_{x\longrightarrow 0}\frac{p_i(x)}{p_{i+1}(x)}=0,\qquad i=1,\ldots,9
 \]
 
-where $\alpha=(1-p^{10})/(1-p)$ is a normalizing constant.  Thus the
-likelihood _at_ the evaluate is not defined, even as a limit (because
-it depends on the direction of approach).
+and if this is respected, then it is reasonable to say that the
+likelihood at the evaluate is 1, and the support 0.
+
+## Package idiom for likelihood functions
 
@@
 dotchart(evaluate2,pch=16)
+par(pty='s')
+plot(evaluate1,evaluate2,asp=1,log='xy')
+abline(0,1)
 ```
 
-So it is much better than the coldstart above in the sense that there
-is more probability on $p_1$ and less everywhere else.  We can compare
-the hotstart above to the coldstart:
+So it is almost identical to the coldstart above in the sense that the
+two evaluates are very close.  We can compare the hotstart to the
+coldstart in terms of support:
 
@@
 
-This is negative: the so-called "sensible start point" has produced an
-inferior solution.  This is because `sss` has $p_2=p_3=\ldots =
-p_{10}$, so predicts a draw between players $2,3,\ldots 10$, contrary
-to observation.
+This is almost exactly zero, but negative: the so-called "sensible
+start point" has produced an inferior solution.  This is because `sss`
+has $p_2=p_3=\ldots = p_{10}$, so predicts a draw between players
+$2,3,\ldots 10$, contrary to observation.
 
-A better way is to mimic the more appropriate of the two limiting processes above:
+Possibly, a better way is to mimic the more appropriate of the two
+limiting processes above:
 
@@
 
-Observe that the likelihood at this point is marginally better than
-the `equalp()` start point:
+Observe that the likelihood at this point is only very slightly better
+than the `equalp()` start point:
 
@@
 ```
-
-but the dotchart again shows a quite a lot of probability on
-$p_2,\ldots p_{10}$.
 

### 2020-01-27 inst/cook.Rmd
more detail

@@
 
-[as a consistency check we can perform the same calculation with a
-slightly different value for `small`:
+See how the constraint is active: the strength of the training effect
+is almost exactly equal to its maximal value of `small` [as a
+consistency check we can perform the same calculation with a slightly
+different value for `small`:
 

### 2020-02-03 inst/notthewinner.Rmd
more work

---
title: "Weakly informative observations: not the winner"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(hyper2)
```

# Partial rank observations: someone didn't win

Here we consider a situation in which we make the observation that a
particular player *does not* win.  We are ignorant of the order
statistic except for the fact that (wlog) player 1 is not first
placed.

As in `loser.Rmd`, we have three hypotheses:

* $H_1\colon p_1=0$
* $H_2\colon\exists i\neq 1 \mbox{ with } p_1\geq p_i>0$
* $H_3\colon p_1=p_2=\ldots=p_n=\frac{1}{n}$

(observe that $H_2$ implies that someone is worse than me).  All three
hypotheses correspond to a restriction on allowable parameter space.
The first step is to define a suitable likelihood function for our
observation.  Given that $P(D|H_1)=1$, $P(D|H_2)\leq\frac{1}{2}$ and
$P(D|H_3) =\frac{n-1}{n}$, this gives us a likelihood function for our
three hypotheses.

What we will do is combine this observation with other, perhaps more
informative, observations which we will take to be (conditionally)
independent of one another.  Firstly we will consider some highly
informative order statistics:

```{r}
(H <- hyper2(pnames=paste("p",1:5,sep="")) +
    order_likelihood(c(5,2,1,4)) + 
    order_likelihood(c(2,4,1)) + 
    order_likelihood(c(4,3,1)) + 
    order_likelihood(c(3,5)))
(mH <- maxp(H))
loglik(indep(mH),H)
```

Note the low estimated strength of `p1`.

## Hypothesis 1

```{r}
small <- 1e-3
(mH1 <- maxp(H, startp=rep(small/2,4),fcm=cbind(-1,0,0,0),fcv = -small))
loglik(indep(mH1),H)
```

and for a consistency we will use slightly different values of `small`:

```{r}
small <- 5e-4
(mH2 <- maxp(H, startp=rep(small/2,4),fcm=cbind(-1,0,0,0),fcv = -small))
loglik(indep(mH2),H)
```

```{r}
small <- 3e-4
(mH2 <- maxp(H, startp=rep(small/2,4),fcm=cbind(-1,0,0,0),fcv = -small))
loglik(indep(mH2),H)
```

Observe that decreasing the value of variable `small` corresponds to a
more restrictive constraint on the likelihood.  And as the value of
`small` decreases the likelihood at the evaluate decreases quite
substantially (the constraint is active as we can see from the first
element of the evaluate).  Usually this does not matter but in this
case we can deduce using elementary arguments that the likelihood of
$H_2$ is zero: if $p_1=0$, then $p_4=0$ from the first line, and then
$p_1=0$ from line 2; then $p_3=0$ from line 3, $p_5=0$ from line 4 and
finally $p+2=0$ from line 1.

Thus we can say for sure that $H_1$ has zero likelihood and we may
discount it.

## Hypothesis 2

```{r}
n <- 5 # number of players
small <- 1e-3  # numerical necessity
(ans_constrained <-
maxp(H,startp=c(small*2,rep(small,n-2)),fcm=rbind(c(1,-1,rep(0,n-3))),fcv=0,give=TRUE)
)
small <- 5e-4
(ans_constrained <-
maxp(H,startp=c(small*2,rep(small,n-2)),fcm=rbind(c(1,-1,rep(0,n-3))),fcv=0,give=TRUE)
)
```

In the above, note that the value of `small` is immaterial and the
evaluate has $p+1=p_2\neq 0$.

## Hypothesis 3

```{r}
loglik(indep(equalp(H)),H)
```

The support for $H_2$ against $H_3$ is about $-6.846-(-7.455) = 1.391$.

# Additional observation

Suppose we now make another observation: person 1 did not win.  How
does this change our relative support for $H_2$ and $H_1$?

```{r}
library("partitions")
M <- perms(5)
M <- M[,M[1,] != 1]
SL <- list()
for(i in seq_len(ncol(M))){
	SL[[i]] <- H + order_likelihood(M[,i,drop=TRUE])
}
SL <- as.suplist(SL)
1e4*like_single_list(indep(equalp(SL)),SL)
1e4*like_single_list(indep(maxp(SL)),SL)
```

Something is very wrong here because the support should be very
similar to the support applied to the order statistic without the
additional observation.

### 2020-02-03 inst/notthewinner.Rmd
typo

@@
 In the above, note that the value of `small` is immaterial and the
-evaluate has $p+1=p_2\neq 0$.
+evaluate has $p_1=p_2\neq 0$.
 
@@
 M <- perms(5)
-M <- M[,M[1,] != 1]
 SL <- list()

### 2020-02-04 inst/notthewinner.Rmd
total reworking

@@
 
-Here we consider a situation in which we make the observation that a
-particular player *does not* win.  We are ignorant of the order
-statistic except for the fact that (wlog) player 1 is not first
-placed.
+Here we consider a situation in which we make an observation that is
+only very very weakly informative.  We consider the observation
+${\mathcal O}$ that a particular player *does not* win: the order
+statistic is unobserved and we know nothing about it except for the
+fact that (wlog) player 1 (aka "me") is not first placed.
 
@@
 
-* $H_1\colon p_1=0$
-* $H_2\colon\exists i\neq 1 \mbox{ with } p_1\geq p_i>0$
-* $H_3\colon p_1=p_2=\ldots=p_n=\frac{1}{n}$
+* $H_0\colon\sum p_i=1$  (that is, no restriction)
+* $H_1\colon\exists i\neq 1 \mbox{ with } p_1\geq p_i>0$
+* $H_2\colon p_1=p_2=\ldots=p_n=\frac{1}{n}$
 
-(observe that $H_2$ implies that someone is worse than me).  All three
-hypotheses correspond to a restriction on allowable parameter space.
-The first step is to define a suitable likelihood function for our
-observation.  Given that $P(D|H_1)=1$, $P(D|H_2)\leq\frac{1}{2}$ and
-$P(D|H_3) =\frac{n-1}{n}$, this gives us a likelihood function for our
-three hypotheses.
+(see how $H_1$ implies that someone is worse than me).  Observe that
+$H_1$ and $H_2$ correspond to a restriction on allowable parameter
+space.  The first step is to define a suitable likelihood function for
+our data.
 
-What we will do is combine this observation with other, perhaps more
+What we will do is combine ${\mathcal O}$ with other, perhaps more
 informative, observations which we will take to be (conditionally)
@@
     order_likelihood(c(3,5)))
-(mH <- maxp(H))
-loglik(indep(mH),H)
+(mH0 <- maxp(H))
+loglik(indep(mH0),H)
 ```
 
-Note the low estimated strength of `p1`.
+Above, `mH0` corresponds to $H_0$.  Note the low but nonzero estimated
+strength of `p1`.
+
+Observe in passing that we may reject $H_z\colon p_1=0$
+unconditionally: we can deduce using elementary arguments that the
+likelihood of $H_z$ is zero: if $p_1=0$, then $p_4=0$ from the first
+line, and then $p_1=0$ from line 2; then $p_3=0$ from line 3, $p_5=0$
+from line 4 and finally $p_2=0$ from line 1.  Thus we can say for sure
+that $H_z$ is incorrect and we may safely ignore it henceforth.
 
@@
 
+Hypothesis 1 is rather tricky so we will use the ersatz $H_{1'}\colon
+p_1\geqslant p_4$ (we choose $p_4$ on the grounds of its low estimated
+value).  Then
+
 ```{r}
-small <- 1e-3
-(mH1 <- maxp(H, startp=rep(small/2,4),fcm=cbind(-1,0,0,0),fcv = -small))
+n <- 5 # number of players
+small <- 1e-3  # numerical necessity
+(mH1 <- maxp(H,startp=c(small*2,rep(small,n-2)),fcm=rbind(c(1,0,0,-1)),fcv=0))
 loglik(indep(mH1),H)
@@
 
-and for a consistency we will use slightly different values of `small`:
-
-```{r}
-small <- 5e-4
-(mH2 <- maxp(H, startp=rep(small/2,4),fcm=cbind(-1,0,0,0),fcv = -small))
-loglik(indep(mH2),H)
-```
-
-```{r}
-small <- 3e-4
-(mH2 <- maxp(H, startp=rep(small/2,4),fcm=cbind(-1,0,0,0),fcv = -small))
-loglik(indep(mH2),H)
-```
-
-Observe that decreasing the value of variable `small` corresponds to a
-more restrictive constraint on the likelihood.  And as the value of
-`small` decreases the likelihood at the evaluate decreases quite
-substantially (the constraint is active as we can see from the first
-element of the evaluate).  Usually this does not matter but in this
-case we can deduce using elementary arguments that the likelihood of
-$H_2$ is zero: if $p_1=0$, then $p_4=0$ from the first line, and then
-$p_1=0$ from line 2; then $p_3=0$ from line 3, $p_5=0$ from line 4 and
-finally $p+2=0$ from line 1.
-
-Thus we can say for sure that $H_1$ has zero likelihood and we may
-discount it.
+(the numerical value of `small` is immaterial).
 
@@
 ```{r}
-n <- 5 # number of players
-small <- 1e-3  # numerical necessity
-(ans_constrained <-
-maxp(H,startp=c(small*2,rep(small,n-2)),fcm=rbind(c(1,-1,rep(0,n-3))),fcv=0,give=TRUE)
-)
-small <- 5e-4
-(ans_constrained <-
-maxp(H,startp=c(small*2,rep(small,n-2)),fcm=rbind(c(1,-1,rep(0,n-3))),fcv=0,give=TRUE)
-)
+mH2 <- equalp(H)
+loglik(indep(mH2),H)
 ```
 
-In the above, note that the value of `small` is immaterial and the
-evaluate has $p_1=p_2\neq 0$.
+## Support function
 
-## Hypothesis 3
+We can present a support function for $H_0,H_1,H_2$:
 
-```{r}
-loglik(indep(equalp(H)),H)
+```{r,suppcalculator}
+Supp <- c(H0=loglik(indep(mH0),H), H1=loglik(indep(mH1),H), H2=loglik(indep(mH2),H)
+)
+Supp-max(Supp)
 ```
 
-The support for $H_2$ against $H_3$ is about $-6.846-(-7.455) = 1.391$.
+# Additional uninformative observation
 
-# Additional observation
+Suppose we now make a perfectly uninformative observation: in an
+independent trial, the order statistic for the five competitors was
+one of the $5!=120$ possible orders.  How does this change our
+relative support for $H_0$ against $H_1$ and $H_2$?  It should make no
+difference.
 
-Suppose we now make another observation: person 1 did not win.  How
-does this change our relative support for $H_2$ and $H_1$?
-
-```{r}
+```{r,uninformativesupportfunction}
 library("partitions")
-M <- perms(5)
+H # as before, here for convenience
+M <- perms(5)   # uninformative!
 SL <- list()
@@
 SL <- as.suplist(SL)
-1e4*like_single_list(indep(equalp(SL)),SL)
-1e4*like_single_list(indep(maxp(SL)),SL)
 ```
 
-Something is very wrong here because the support should be very
-similar to the support applied to the order statistic without the
-additional observation.
+So `SL` is a compound object that gives a likelihood function for the
+joint observation of the order statistics and the uninformative data a
+subsequent (independent) trial was in any of the $5!=120$ possible
+orders.  We can find the MLE for different observations:
+
+```{r,mls0}
+mSL0 <- maxp(SL)
+mSL1 <- maxp(SL,startp=c(small*2,rep(small,n-2)),fcm=rbind(c(1,0,0,-1)),fcv=0)
+mSL2 <- equalp(SL)
+```
+
+We use `like_single_list()` to calculate the likelihoods at the
+different evaluates; note that this function returns likelihoods, not
+supports.
+
+```{r,supp2calc}
+Supp2 <-
+      log(c(
+      H0 = like_single_list(indep(mSL0),SL),
+      H1 = like_single_list(indep(mSL1),SL),
+      H2 = like_single_list(indep(mSL2),SL)))
+(Supp2 <- Supp2-max(Supp2))
+```
+
+which is identical to the previous support function.
+
+# Additional weakly informative observation
+
+Now we make another observation ${\mathcal O}$ that is weakly
+informative: that person `1` did not win:
+
+```{r,slacalc}
+M <- perms(5)
+M <- M[, M[1,]!=1]  # remove any column with "1" in first place
+SLa <- list()
+for(i in seq_len(ncol(M))){
+	SLa[[i]] <- H + order_likelihood(M[,i,drop=TRUE])
+}
+SLa <- as.suplist(SLa)
+```
+
+So `SLa` is a compound object that gives a likelihood function for the
+joint observation of the order statistics and ${\mathcal O}$.  We can
+find the MLE for different observations:
+
+```{r,mlsa}
+mSLa0 <- maxp(SLa)
+mSLa1 <- maxp(SLa,startp=c(small*2,rep(small,n-2)),fcm=rbind(c(1,0,0,-1)),fcv=0)
+mSLa2 <- equalp(SLa)
+```
+
+```{r,suppacalc}
+Suppa <-
+      log(c(
+      H0 = like_single_list(indep(mSLa0),SLa),
+      H1 = like_single_list(indep(mSLa1),SLa),
+      H2 = like_single_list(indep(mSLa2),SLa)))
+(Suppa <- Suppa-max(Suppa))
+```
+
+To compare:
+
+```{r,comparesups}
+Supp2-Suppa
+```
+
+Thus ${\mathcal O}$ furnishes a tiny bit of extra evidence against
+$H_1$ (about $0.016$ units of support) and a little bit of extra
+evidence against $H_2$ (about $0.193$ units of support).

### 2020-02-07 inst/notthewinner.Rmd
more rewriting

@@
 $H_1$ and $H_2$ correspond to a restriction on allowable parameter
-space.  The first step is to define a suitable likelihood function for
-our data.
+space.
 
-What we will do is combine ${\mathcal O}$ with other, perhaps more
-informative, observations which we will take to be (conditionally)
-independent of one another.  Firstly we will consider some highly
-informative order statistics:
+To understand the informativeness of ${\mathcal O}$, we will consider
+it in conjunction with some highly informative observations ${\mathcal
+I}$ which are order statistics:
 
@@
     order_likelihood(c(3,5)))
+	```
+
+We will firstly consider these observations in isolation, then later
+combine them with ${\mathcal O}$ which we will take to be
+(conditionally) independent.  The first step is to define a suitable
+likelihood function for ${\mathcal I}$:q
+
+```{r}	
 (mH0 <- maxp(H))
@@
 
-(the numerical value of `small` is immaterial).
+See how the constraint of $H_1$ is active, for at the evaluate
+$p_1=p_4$ (the numerical value of `small` is immaterial).
 
@@
 
-We can present a support function for $H_0,H_1,H_2$:
+We can present a support function for $H_0,H_1,H_2$ for ${\mathcal
+I}$:
 
@@
 Now we make another observation ${\mathcal O}$ that is weakly
-informative: that person `1` did not win:
+informative: person `1` did not win.  Package idiom is similar to the
+perfectly uninformative observation considered above, except that we
+remove every order that has person 1 winning:
 
@@
 
-So `SLa` is a compound object that gives a likelihood function for the
-joint observation of the order statistics and ${\mathcal O}$.  We can
-find the MLE for different observations:
+So `SLa` is a compound object that gives a support function for the
+joint observation ${\mathcal I}$ and ${\mathcal O}$.  We can find the
+MLE:
 
@@
 
-To compare:
+We can now compare the support function for ${\mathcal I}$ with the
+support function for ${\mathcal I}$ and ${\mathcal H}$:
 

### 2020-02-19 inst/formula1.Rmd
typos

@@
 ```{r setup, include=FALSE}
+set.seed(0)
 knitr::opts_chunk$set(echo = TRUE)
@@
 see that Hamilton, the first row, came second in Australia, first in
-China, second in Bahrain, fourthin Russia, and so on.  In the first
-columnm, we see the result from Australia in which Hamilton came
+China, second in Bahrain, fourth in Russia, and so on.  In the first
+column, we see the result from Australia (AUS) in which Hamilton came
 second, Vettel first, Bottas third, and so on.  Notation used also

### 2020-02-19 inst/formula1.Rmd
time-saving: evaluate maxp(L2017) only once
timesaving: evaluate maxp(L2017) only once

@@
 ```{r,maxp2017}
-dotchart(maxp(L2017),pch=16,main='2017 Formula 1')
-dotchart(log(maxp(L2017)[1:23]),pch=16,main='2017 Formula 1, log scale')
+mL2017 <- maxp(L2017)
+dotchart(mL2017,pch=16,main='2017 Formula 1')
+dotchart(log(mL2017)[1:23],pch=16,main='2017 Formula 1, log scale')
 ```
@@
 
-m <- maxp(L2017)
+m <- mL2017   # maxp(L2017)
 ox <- rank(-points) # negative because number 1 = largest number of points
@@
 L1 <- loglik(indep(equalp(L2017)),L2017)
-L2 <- loglik(indep(maxp(L2017)),L2017)
+L2 <- loglik(indep(mL2017),L2017)   # mL2017 == maxp(L2017)
 pchisq(2*(L2-L1),df=25-1,lower.tail=FALSE)

### 2020-02-20 inst/eurovision.Rmd
slight rewording and reworking

@@
 competitor are added and the winner is the competitor with the most
-points.  However, in this model, the numerical values themselves do
-not affect the likelihood function; only the order of the voters'
-preferences matters.
+points.  However, in the `hyper2` model, the numerical values
+themselves do not affect the likelihood function; only the order of
+the voters' preferences matters.  The following R idiom translates
+`wiki_matrix` into a form suitable for analysis with `hyper2`:
 
@@
 
-Now, take the first row of 'preference'.  This represents the votes
+Now, take the first row of `preference`.  This represents the votes
 cast *by* (sic) Montenegro (`"ME"`).  Their favourite was [last
@@
 other countries (`CZ`, `BE`, `SW`, `CH`, `BG`, `PT`) did not receive
-any points from Montenegro.
-
-We need to convert matrix `preference` into a likelihood function,
-here `euro2009`:
+any points from Montenegro.  We need to convert matrix `preference`
+into a likelihood function, here `euro2009`:
 

### 2020-02-20 inst/formula1.Rmd
actually show the evaluate

@@
 mL2017 <- maxp(L2017)
+mL2017
 dotchart(mL2017,pch=16,main='2017 Formula 1')

### 2020-02-20 inst/formula1.Rmd
fixes #27

@@
 includes six different classes of no-score such as "Ret" for retired,
-"WD" for withdrawn, and so on.  I now define a bespoke R function that
-translates data of this type into a log-likelihood function using the
-`hyper2` package.  The function is similar to, but slightly different
+"WD" for withdrawn, and so on.  I now use bespoke R function
+\code{ranktable2supp()}, that translates data of this type into a
+log-likelihood.  The function is similar to, but slightly different
 from, the analysis in `eurovision.R`.
 
-```{r,defF1}
-F1_likelihood <- function(wiki_table){
-
-  noscore <- c("Ret", "WD", "DNS", "DSQ", "DNP", "NC")
-  first_and_last <- c(1,ncol(wiki_table))
-
-  racers <- wiki_table[,1]
-  venues <- colnames(wiki_table)[-first_and_last]
-
-  ## Now create a numeric matrix, fmat.  Two steps: first, count
-  ## any no-score as zero:
-
-  f <- function(x){
-    if(any(x %in% noscore)){x[x%in%noscore] <- 0}
-    return(x)
-  }
-
-  jj <- apply(wiki_table,2,f)
-
-  ## Second, convert to numeric and strip out names; transpose of
-  ## wiki_table (because we want each row to be a venue):
-
-  fmat <- matrix(as.numeric(jj[,-first_and_last]),byrow=TRUE,ncol=nrow(wiki_table)) 
-  colnames(fmat) <- racers
-  rownames(fmat) <- venues
-
-  F1 <- hyper2(d=ncol(fmat))
-
-  ## Now cycle through the rows; each row is a venue [voter]
-  for(i in seq_len(nrow(fmat))){
-    d <- fmat[i,,drop=TRUE]
-    while(any(d>0)){
-      eligible <- which(d>=0)
-
-      ## The first choice among eligible players has +1 power on the
-      ## numerator:
-      F1[which(d==1)] %<>% inc
-
-      ## denominator of all eligible players; power -1
-      F1[eligible] %<>% dec
-
-      ## once you've come first in the field, you are ineligible to be first again:
-      d[d==1] <- -1
-
-      ## everyone moves down the list, so who *was* in second place
-      ## becomes first place, who *was* third place becomes second,
-      ## and so on:
-      d[d>0] %<>% dec
-
-    } # while() loop closes
-  } # i loop closes
-
-  ## syntatic sugar:
-  pnames(F1) <- racers
-
-  return(F1)
-}
-```
-
-We can apply this to the dataset:
-
 ```{r,applydata}
-L2017 <- F1_likelihood(f2017)
+L2017 <- ranktable2supp(f2017)
 head(L2017,3)
@@
 ```{r,fouryearsdata}
-F1_2014 <- "formula1_2014.txt" %>% read.table(header=TRUE) %>% F1_likelihood
-F1_2015 <- "formula1_2015.txt" %>% read.table(header=TRUE) %>% F1_likelihood
-F1_2016 <- "formula1_2016.txt" %>% read.table(header=TRUE) %>% F1_likelihood
-F1_2017 <- "formula1_2017.txt" %>% read.table(header=TRUE) %>% F1_likelihood
+F1_2014 <- "formula1_2014.txt" %>% read.table(header=TRUE) %>% ranktable2supp
+F1_2015 <- "formula1_2015.txt" %>% read.table(header=TRUE) %>% ranktable2supp
+F1_2016 <- "formula1_2016.txt" %>% read.table(header=TRUE) %>% ranktable2supp
+F1_2017 <- "formula1_2017.txt" %>% read.table(header=TRUE) %>% ranktable2supp
 a <- list(F1_2014, F1_2015, F1_2016, F1_2017)

### 2020-02-20 R/hyper2.R

@@ -725,3 +725,63 @@ setGeneric("pnames<-",function(x,value){standardGeneric("pnames<-")})
   if(missing(pnames)){pnames(H) <- letters[seq_len(size(H))]}
   return(H)
 }
+
+`ranktable2supp` <- function(x, noscore, misslast=TRUE){
+    if(missing(noscore)){
+        noscore <- c("Ret", "WD", "DNS", "DSQ", "DNP", "NC")
+    }
+
+    if(misslast){
+        first_and_last <- c(1,ncol(x))
+    } else {
+        first_and_last <- 1
+    }
+
+    racers <- x[,1]
+    venues <- colnames(x)[-first_and_last] 
+
+    ## Now create a numeric matrix, fmat.  Two steps: first, count
+    ## any no-score as zero:
+  
+  jj <- apply(x,2,function(y){
+      if(any(y %in% noscore)){y[y%in%noscore] <- 0}
+      return(y)
+  })
+    
+    ## Second, convert to numeric and strip out names; transpose of
+    ## x (because we want each row to be a venue):
+    
+    fmat <- matrix(as.numeric(jj[,-first_and_last]),byrow=TRUE,ncol=nrow(x)) 
+    colnames(fmat) <- racers
+    rownames(fmat) <- venues
+    
+    out <- hyper2(d=ncol(fmat))
+    
+    ## Now cycle through the rows; each row is a venue [voter]
+    for(i in seq_len(nrow(fmat))){
+        d <- fmat[i,,drop=TRUE]
+        while(any(d>0)){
+            eligible <- which(d>=0)  #NB inclusive inequality; zero is DNC etc who
+            
+            ## Increment numerator power of the first choice among eligible players:
+            out[which(d==1)] %<>% inc
+            
+            ## Power of set of all eligible players decrements:
+            out[eligible] %<>% dec
+            
+            ## once you've come first in the field, you are ineligible to be first again:
+            d[d==1] <- -1  # NB strictly <0
+            
+            ## Now, everyone moves down the list, so who *was* in
+            ## second place becomes first place, who *was* third place
+            ## becomes second, and so on:
+            
+            d[d>0] %<>% dec
+            
+        } # while() loop closes
+    } # i loop closes
+    
+    ## syntatic sugar:
+    pnames(out) <- racers
+    return(out)
+} 

### 2020-02-20 data/formula1.R

@@ -287,3 +287,102 @@ structure(list(brackets = list(1, c(1, 2, 3, 4, 5, 6, 7, 8, 9,
 "Ocon", "Palmer", "Perez", "Raikkanen", "Resta", "Ricciardo", 
 "Sainz", "Stroll", "Vandoorne", "Verstappen", "Vettel", "Wehrlein"
 ), class = "factor")), class = "hyper2")
+
+`F1_table_2017` <-
+structure(list(driver = structure(c(8L, 24L, 2L, 17L, 19L, 23L, 
+16L, 14L, 20L, 10L, 13L, 21L, 7L, 12L, 1L, 22L, 15L, 25L, 11L, 
+4L, 5L, 6L, 9L, 3L, 18L), .Label = c("Alonso", "Bottas", "Button", 
+"Ericsson", "Gasly", "Giovinazzi", "Grosjean", "Hamilton", "Hartley", 
+"Hulkenberg", "Kvyat", "Magnussen", "Massa", "Ocon", "Palmer", 
+"Perez", "Raikkanen", "Resta", "Ricciardo", "Sainz", "Stroll", 
+"Vandoorne", "Verstappen", "Vettel", "Wehrlein"), class = "factor"), 
+    AUS = structure(c(7L, 2L, 8L, 9L, 15L, 10L, 12L, 3L, 13L, 
+    4L, 11L, 15L, 15L, 15L, 15L, 6L, 15L, 16L, 14L, 15L, 1L, 
+    5L, 1L, 1L, 1L), .Label = c("0", "1", "10", "11", "12", "13", 
+    "2", "3", "4", "5", "6", "7", "8", "9", "Ret", "WD"), class = "factor"), 
+    CHN = structure(c(2L, 9L, 13L, 12L, 11L, 10L, 16L, 3L, 14L, 
+    5L, 7L, 17L, 4L, 15L, 17L, 17L, 6L, 1L, 17L, 8L, 1L, 17L, 
+    1L, 1L, 1L), .Label = c("0", "1", "10", "11", "12", "13", 
+    "14", "15", "2", "3", "4", "5", "6", "7", "8", "9", "Ret"
+    ), class = "factor"), BHR = structure(c(8L, 2L, 9L, 10L, 
+    11L, 17L, 13L, 3L, 17L, 15L, 12L, 17L, 14L, 17L, 7L, 16L, 
[…]

### 2020-02-21 inst/moto.Rmd

---
title: "Grand Prix motorcycle racing"
output: html_document
---

```{r setup, include=FALSE}
set.seed(0)
options(width = 120)
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("magrittr")
```

## The 2019 FIM motoGP world championship

Here I analyse a series of Grand Prix motorcycling race results using
the `hyper` package which implements the Plackett-Luce likelihood
function.  Consider the following dataset, taken from Wikipedia:

```{r cars}
data(motoGP)
motoGP_2019_table
```

Each row is a driver (rider?) and each column (after the first) a
venue.  Takin the first row we see that Marquez came second in Qatar,
first in Argentina, retired in Austin, and so on.  In the first
column, we see the result from Qatar (QAT) in which Marquez came
second, Dovizioso first, Vinales third, and so on.  Notation used also
includes six different classes of no-score such as "Ret" for retired,
"WD" for withdrawn, and so on.  I now use bespoke R function
\code{ranktable2supp()}, that translates data of this type into a
log-likelihood.  The function is similar to, but slightly different
from, the analysis in `eurovision.R`.

```{r,applydata}
head(motoGP_2019)
```

(object `motoGP_2019` is equivalent to
`ranktable2supp(motoGP_2019_table)`.  Above, we see a small part of
the Placket-Luce likelihood function for the 2019 results.  We can
find the maximum likelihood estimator by using the `maxp()` function:

```{r,maxp2017}
options(width = 100)
m <- maxp(motoGP_2019)
m
dotchart(m,pch=16)
pie(m)
```

In the above, note the dominance of Marquez.  It is sometimes
preferable to plot the estimated strengths on a log scale:

```{r,maxp2017log}
dotchart(log(m[1:27]),pch=16)
```

(in the diagram above, we remove the bottom-ranked player, Lecuona,
whose estimated strength is zero).

# Likelihood scoring vs points scoring

The championship winning order is determined by a points system: the
winner is awarded 25 points, second place 18 points, and so on.  We
can now check to see how the points total matches up with the
likelihood estimation.

```{r,pointsforwinning}
(points_awarded <- c(`1st`=25, `2nd`=20, `3rd`=16, `4th`=13, `5th`= 11,
`6th`=10, `7th`= 9, `8th`=8,`9th`=7, `10th`=6, `11th`=5, `12th`=4,
`13th`=3, `14th`=2, `15th`=1))
```

```{r,likelihoodandscore}
points <- motoGP_2019_table$points
names(points) <- motoGP_2019_table$driver

ox <- rank(-points) # negative because number 1 = largest number of points
oy <- rank(-m)      # negative because number 1 = largest strength
par(pty='s')        # square plot
plot(ox,oy,asp=1,pty='s',xlim=c(0,28),ylim=c(0,28),pch=16,
xlab="official order",ylab="my order",main='motoGP, 2019 season')
par(xpd=TRUE)       # allow drivers' names to appear outside plotting region
for(i in seq_along(ox)){text(ox[i],oy[i],names(ox)[i],pos=4,col='gray') }
par(xpd=FALSE)      # stop diagonal line from protruding beyond plotting region
abline(0,1)
```

In the graph above, we see a certain amount of disagreement.  Only the
top two riders, Marquez and Dovizioso, have rankings common to
likelihood system and points system.  Considering the next four
players (Vinales, Rins, Quartaro, Petrucci) we see that the
points-based order is Vinales, Rins, Petrucci, but the likelihood
order is Petrucci, Vinales, Rins (Rossi), Quartararo.

### 2020-02-23 inst/skating_analysis.Rmd
issue #31

title: "Ladies' figure skating"
author: "Robin K. S. Hankin"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{figure_skating}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Ladies' figure skating in the 2002 Winter Olympics

This script creates and analyses hyper2 object "skating" which is a
likelihood function for the strengths of the competitors in the
Ladies' Figure Skating at the 2002 Winter Olympics.  Dataframe
`skating_table` is copied from Lock and Lock (see skating.Rd for
more details).

```{r}
library("hyper2",quietly=TRUE)
data("skating")
skating_table
```

Object `skating_table` is structured so that each competitor is a
row, and each judge is a column.  Function `order_likelihood()`
considers each row to be a race [or a judge], so we need to take a
transpose.  The following R idiom is plausible but incorrect:

```{r}
skating_incorrect <- order_likelihood(t(skating_table))
```

The above is incorrect because we want the ranks, and we have the
order.  Correct analysis follows.

```{r}
skating_rank <- apply(skating_table,2,order)
head(skating_rank)
```

Column 1 of `skating_table` and column 1 of `skating_rank` are
confusingly simular.  So consider column 2 instead, corresponding to
judge number 2.  If `a=skating_table[,2]` and
`b=skating_rank[,2]` we have `a=c(4,1,3,2, ...)` and
`b=c(2,4,3,1, ...)`.  Thus `a` means that hughes was judged to
be fourth best, slutskaya was judged to be the best, kwan third best,
cohen second best, and so on.  Conversely, `b` means that the best
competitor was number two [slutskaya], the second best number was
number four [cohen], the third best was number three [kwan], the
fourth best number 1 [hughes], and so on.  _Note that this means the
rows of `skating_rank` cannot be named_.

```{r}
skating <- order_likelihood(t(skating_rank))
pnames(skating) <- rownames(skating_table)
head(skating)
```

Further, we might ask how many judges ranked each competitor first,
second and so on:

```{r,alljudgeorder}
oo <- sapply(seq_len(23),function(i){rowSums(skating_table==i)})
colnames(oo) <- rep(" ",23)  # nicer print
oo
```

In the above, each row is a competitor and each column corresponds to
a ranking.  Thus the first column corresponds to "first place" and
	shows that five judges ranked Hughes first and four judges ranked
Slutskaya first.  The second column corresponds to "second place" and
shows that one judge gave Hughes second place, two judges placed
Slutskaya second, five gave Kawn second, and one gave Cohen second.
The formal ordering is, I think, to rank the competitors by number of
first places, then second, then third , and so on.  Ties at level $n$
are broken by considering $n+1$.

# Data visualization

Now some data visualization.  First the MLE for the strengths:

```{r,maxshow}
(m <- maxp(skating))
```

```{r,dpi=72}
dotchart(m)
dotchart(log(m))
```

## Evidence for medal positions

Looking at the dotcharts, it seems that the medallists---Hughes (gold),
Slutskya (silver), Kwan (bronze)---were considerably higher in
strength than the rest of the field.  Here I will test the hypothesis
that the medallists were in fact the strongest three competitors.
Technically you need to optimize over the union of the possibilities
that one of the three medallists did not come in the top three; but
this is hard.  We will do something much easier but numerically
equivalent: optimize over the union of outcomes where either Cohen or
Suguri (who placed fourth and fifth respectively) had higher strength
than any of the medallists.

```{r}
jj <-
    matrix(c(
        -1, 0, 0,   1,0,  # Hughes   < Cohen
         0,-1, 0,   1,0,  # Slutskya < Cohen
         0, 0,-1,   1,0,  # Kwan     < Cohen

        -1, 0, 0,   0,1,  # Hughes   < Suguri
         0,-1, 0,   0,1,  # Slutskya < Suguri
         0, 0,-1,   0,1), # Kwan     < Suguri
        byrow=TRUE,ncol=5)

problem_constraints <-  # fill with zeros for other competitors
    cbind(jj,matrix(0,nrow(jj),size(skating)-ncol(jj)-1))

small <-1e-4  # need a sensible start value satisfying the constraints
start <- c(rep(2*small,3),rep(3*small,2),rep(small,17))

out <- rep(0,nrow(problem_constraints))
fullout <- list()

for(i in seq_len(nrow(problem_constraints))){
jj <- maxp(skating, startp=start, give=TRUE,fcm=problem_constraints[i,], fcv=0)
   fullout[[i]] <- jj
   out[i] <- jj$value
}
out
```

Now compare these values with the unconstrained maximum likelihoods:

```{r,unconstrainedlikes}
mgv <- maxp(skating,give=TRUE)$value
mgv  - out
mgv  - max(out)
```

that is, a little over 1 unit of support, falling short of the two
units suggested by Edwards.  Observe that the maximum likelihood among
the six alternative hypotheses is that of number 3, in which the
maximization was constrained to obey `Kwan < Cohen`.

If, instead, we ask whether there is evidence that Suguri should not
have been a medallist, we find

```{r}
mgv  - max(out[4:6])
```

We can be more specific with the likelihoods:

```{r}
plot(out,ylab="log likelihood",axes=FALSE)
axis(2)
axis(side=1,at=1:6,srt=45,labels=c(
"Hug<Coh", "Slu<Coh", "Kwa<Coh",
"Hug<Sug", "Slu<Sug", "Kwa<Sug"
))
```

In the plot above, the vertical axis shows the support.  The six
points on the x-axis correspond to the six rows of
`problem_constraints`; names have been abbreviated to the first three
letters.  Thus the first three points are maximum likelihoods for
`Hughes < Cohen`, `Slutskya < Cohen`, and `Kwan < Cohen` respectively.

## Ranking methodologies.

 maximum likelihood estimated strengths furnish an ordering for the competitors.
We can compare ranking by strengths with the official point-tallying method:

```{r,dpi=72}
par(pty='s')
lik <- rank(-m)
plot(1:23,lik,asp=1,pch=16,xlab='official order',ylab='likelihood order')
text(1:23,lik,asp=1,pch=16,pos=c(rep(4,19),rep(2,4)),rownames(skating_table))
abline(0,1)
```

In the above diagram, there is close agreement between the two methods
in the large, but differences in detail.  For example, the official
ordering is hughes first, then slutskaya second, then kwan third;
likelihood winner is slutskya, followed by hughes and then kwan.

We can gain some understanding of this result by looking at the raw
judging results for the three medallists:

```{r,threemedallists}
head(skating_table,2)
```

and then calculating a table of the results:

```{r, tableofresults}
hughes    <- unlist(skating_table[1,])
slutskaya <- unlist(skating_table[2,])
table(hughes)
table(slutskaya)
```

Looking at the above, we see that Hughes has five judges who gave her
"1"s, while Slutskaya has only 4, which is why she was considered to
be the top.  It seems that Hughes's two "4"s (compared with
Slutskaya's one) have cost her more likelihood-based strength than her
extra "1" gave her; note also that the two judges who placed Highes
fourth (viz `J2` and `J4`) placed Slutskaya first.  Also:

```{r,compareslutskayahughes}
table(hughes < slutskaya)
```

(note that equality is disallowed) showing that five judges preferred Hughes
to Slutskaya and four preferred Slutsakaya to Hughes.

### 2020-02-24 inst/skating_analysis.Rmd
slight change to small

@@
 library("hyper2",quietly=TRUE)
+set.seed(0)
 data("skating")
@@
 
-small <-1e-4  # need a sensible start value satisfying the constraints
+small <- 1e-6  # need a sensible start value satisfying the constraints
 start <- c(rep(2*small,3),rep(3*small,2),rep(small,17))

### 2020-03-01 inst/cook.Rmd
fixes #39

@@
 H <- H +
-    order_likelihood(c(5,2,3,4)) + # person 1 says 5 is the best, 4 the worst.
-    order_likelihood(c(1,5,3,4)) + # person 2 says 1 is the best, 4 the worst.
-    order_likelihood(c(1,4,2,5)) + # person 3 says 1 is the best, 5 the worst.
-    order_likelihood(c(1,2,3,5)) + # person 4 says 1 is the best, 5 the worst.
-    order_likelihood(c(2,1,3,4))   # person 5 says 2 is the best, 4 the worst.
+    rank_likelihood(c(5,2,3,4)) + # person 1 says 5 is the best, 4 the worst.
+    rank_likelihood(c(1,5,3,4)) + # person 2 says 1 is the best, 4 the worst.
+    rank_likelihood(c(1,4,2,5)) + # person 3 says 1 is the best, 5 the worst.
+    rank_likelihood(c(1,2,3,5)) + # person 4 says 1 is the best, 5 the worst.
+    rank_likelihood(c(2,1,3,4))   # person 5 says 2 is the best, 4 the worst.
 H
@@
 imH <- indep(maxp(H))
-f <- function(o){loglik(imH,order_likelihood(o),log=FALSE)}
+f <- function(o){loglik(imH,rank_likelihood(o),log=FALSE)}
 LL <- apply(perms(5),2,f)
@@
 Hbefore <- H +
-    order_likelihood(c(5,2,3,4,1)) + #  5 is the best, 4 the worst.
-    order_likelihood(c(5,3,2,4,1)) + #  5 is the best, 1 the worst.
-    order_likelihood(c(4,5,3,2,1)) + #  4 is the best, 1 the worst.
-    order_likelihood(c(5,3,4,1,2)) + #  5 is the best, 2 the worst.
-    order_likelihood(c(2,5,3,1,4))   #  2 is the best, 4 the worst.
+    rank_likelihood(c(5,2,3,4,1)) + #  5 is the best, 4 the worst.
+    rank_likelihood(c(5,3,2,4,1)) + #  5 is the best, 1 the worst.
+    rank_likelihood(c(4,5,3,2,1)) + #  4 is the best, 1 the worst.
+    rank_likelihood(c(5,3,4,1,2)) + #  5 is the best, 2 the worst.
+    rank_likelihood(c(2,5,3,1,4))   #  2 is the best, 4 the worst.
 Hbefore

### 2020-03-01 inst/formula1.Rmd
fixes #39

@@
 "WD" for withdrawn, and so on.  I now use bespoke R function
-\code{ranktable2supp()}, that translates data of this type into a
+\code{ordertable2supp()}, that translates data of this type into a
 log-likelihood.  The function is similar to, but slightly different
@@
 ```{r,applydata}
-L2017 <- ranktable2supp(f2017)
+L2017 <- ordertable2supp(f2017)
 head(L2017,3)
@@
 ```{r,fouryearsdata}
-F1_2014 <- "formula1_2014.txt" %>% read.table(header=TRUE) %>% ranktable2supp
-F1_2015 <- "formula1_2015.txt" %>% read.table(header=TRUE) %>% ranktable2supp
-F1_2016 <- "formula1_2016.txt" %>% read.table(header=TRUE) %>% ranktable2supp
-F1_2017 <- "formula1_2017.txt" %>% read.table(header=TRUE) %>% ranktable2supp
+F1_2014 <- "formula1_2014.txt" %>% read.table(header=TRUE) %>% ordertable2supp
+F1_2015 <- "formula1_2015.txt" %>% read.table(header=TRUE) %>% ordertable2supp
+F1_2016 <- "formula1_2016.txt" %>% read.table(header=TRUE) %>% ordertable2supp
+F1_2017 <- "formula1_2017.txt" %>% read.table(header=TRUE) %>% ordertable2supp
 a <- list(F1_2014, F1_2015, F1_2016, F1_2017)

### 2020-03-01 inst/loser.Rmd
more work
fixes #39

@@
 H <- hyper2(pnames=letters[1:6])
-L <- ggol(H,letters[2:n],letters[1]) 
+L <- ggrl(H,letters[2:n],letters[1]) 
 length(L)
@@
 
-# Discussion
+# Discussion: repeated observations
 
@@
 is Edwards's valet and black and white ball thought experiment).
-
-## Repeated observations
 
@@
 these two approaches.
+
+Note carefully that the observation considered here is "I came last".
+This is different from the observation that "I did not win" which is
+considered separately in `inst/notthewinner.Rmd`.

### 2020-03-01 inst/moto.Rmd
fixes #39

@@
 (object `motoGP_2019` is equivalent to
-`ranktable2supp(motoGP_2019_table)`.  Above, we see a small part of
+`ordertable2supp(motoGP_2019_table)`.  Above, we see a small part of
 the Placket-Luce likelihood function for the 2019 results.  We can

### 2020-03-01 inst/notthewinner.Rmd
slight rewording
fixes #39

@@
 (H <- hyper2(pnames=paste("p",1:5,sep="")) +
-    order_likelihood(c(5,2,1,4)) + 
-    order_likelihood(c(2,4,1)) + 
-    order_likelihood(c(4,3,1)) + 
-    order_likelihood(c(3,5)))
+    rank_likelihood(c(5,2,1,4)) + 
+    rank_likelihood(c(2,4,1)) + 
+    rank_likelihood(c(4,3,1)) + 
+    rank_likelihood(c(3,5)))
 	```
@@
 for(i in seq_len(ncol(M))){
-	SL[[i]] <- H + order_likelihood(M[,i,drop=TRUE])
+	SL[[i]] <- H + rank_likelihood(M[,i,drop=TRUE])
 }
@@
 
-Now we make another observation ${\mathcal O}$ that is weakly
-informative: person `1` did not win.  Package idiom is similar to the
-perfectly uninformative observation considered above, except that we
-remove every order that has person 1 winning:
+Now we make another observation ${\mathcal O}$ which is independent of
+${\mathcal I}$: person `1` did not win.  It is clear that observation
+${\mathcal O}$ is only very weakly informative.  Package idiom is
+similar to the perfectly uninformative observation considered above,
+except that we remove every order that has person 1 winning:
 
@@
 for(i in seq_len(ncol(M))){
-	SLa[[i]] <- H + order_likelihood(M[,i,drop=TRUE])
+	SLa[[i]] <- H + rank_likelihood(M[,i,drop=TRUE])
 }

### 2020-03-01 inst/skating_analysis.Rmd
fixes #39

@@
 Object `skating_table` is structured so that each competitor is a
-row, and each judge is a column.  Function `order_likelihood()`
+row, and each judge is a column.  Function `rank_likelihood()`
 considers each row to be a race [or a judge], so we need to take a
@@
 ```{r}
-skating_incorrect <- order_likelihood(t(skating_table))
+skating_incorrect <- rank_likelihood(t(skating_table))
 ```
@@
 ```{r}
-skating <- order_likelihood(t(skating_rank))
+skating <- rank_likelihood(t(skating_rank))
 pnames(skating) <- rownames(skating_table)

### 2020-03-01 R/hyper2.R

@@ -468,7 +468,7 @@ setGeneric("pnames<-",function(x,value){standardGeneric("pnames<-")})
 
 `head.hyper2` <- function(x,...){ x[head(brackets(x),...)] }
 
-`order_likelihood` <- function(M,times=1){
+`rank_likelihood` <- function(M,times=1){
   M <- rbind(M)  # deals with vectors
   times <- cbind(seq_len(nrow(M)),times)[,2]
   if(is.null(colnames(M))){
@@ -487,7 +487,7 @@ setGeneric("pnames<-",function(x,value){standardGeneric("pnames<-")})
 }
 
 #`addrank` <- function(H,ranks){
-#     .Defunct(new = 'order_likelihood',msg='use H <- H+order_likelihood(...)')
+#     .Defunct(new = 'rank_likelihood',msg='use H <- H+rank_likelihood(...)')
 #     ranks <- rbind(ranks)
 # 
 #     for(r in seq_len(nrow(ranks))){
@@ -554,8 +554,8 @@ setGeneric("pnames<-",function(x,value){standardGeneric("pnames<-")})
   }
 }
 
-`general_grouped_order_likelihood` <- function(H, ...){
-  ## ggol(1:3,4,5:9) means group(1:3) came first, 4 in the middle, 5:9
+`general_grouped_rank_likelihood` <- function(H, ...){
+  ## ggrl(1:3,4,5:9) means group(1:3) came first, 4 in the middle, 5:9
   ## last
     dotargs <- list(...)
     if(any(unlist(lapply(dotargs, is.character)))){
@@ -567,22 +567,22 @@ setGeneric("pnames<-",function(x,value){standardGeneric("pnames<-")})
     }
 
     out <- mult_grid(lapply(dotargs, .allorders))[[1]]
-    out <- apply(out,2,function(rank){H+order_likelihood(rank)})
+    out <- apply(out,2,function(rank){H+rank_likelihood(rank)})
   return(as.suplist(out))
 }
 
-`ggol` <- general_grouped_order_likelihood
+`ggrl` <- general_grouped_rank_likelihood
 
 `choose_losers` <- function(H,all_players,losers){
   stopifnot(losers %in% all_players)
   winners <- all_players[!all_players %in% losers]
-  ggol(H,winners,losers)
+  ggrl(H,winners,losers)
 }
 
 `choose_winners` <- function(H,all_players, winners){
   stopifnot(all(winners %in% all_players))
[…]

### 2020-03-01 inst/gbbo_series9.R

@@ -38,7 +38,7 @@ L <- list()  # overall list
 
 if(doo[["week1"]]){
     L$week1 <-
-        ggol(H,
+        ggrl(H,
              SB = c("Manon"),  
              FB = c("Briony"),
              GT = c("Rahul","Kim-Joy","Jon","Dan","Karen","Antony","Luke"),
@@ -49,7 +49,7 @@ if(doo[["week1"]]){
 
 if(doo[["week2"]]){
     L$week2 <-
-        ggol(H,
+        ggrl(H,
              SB = c("Rahul"),  
              FB = c("Jon","Dan"),
              GT = c("Kim-Joy","Ruby","Manon","Karen","Antony"),
@@ -60,7 +60,7 @@ if(doo[["week2"]]){
 
 if(doo[["week3"]]){
     L$week3 <-
-        ggol(H,
+        ggrl(H,
              SB = c("Rahul"),  
              FB = c("Dan"),
              GT = c("Kim-Joy","Ruby","Manon","Jon","Karen","Terry"),
@@ -71,7 +71,7 @@ if(doo[["week3"]]){
 
 if(doo[["week4"]]){
     L$week4 <-
-        ggol(H,
+        ggrl(H,
              SB = c("Dan"),  
              FB = c("Rahul","Jon"),
              GT = c("Kim-Joy","Ruby","Manon"),
@@ -82,7 +82,7 @@ if(doo[["week4"]]){
 
 if(doo[["week5"]]){
     L$week5 <-
-        ggol(H,
+        ggrl(H,
              SB = c("Kim-Joy"),  
              FB = c("Rahul"),
              GT = c("Ruby","Manon","Dan"),
@@ -93,7 +93,7 @@ if(doo[["week5"]]){
 
 if(doo[["week6"]]){
     L$week6 <-
-        ggol(H,
+        ggrl(H,
              SB = c("Briony"),  
              FB = c("Rahul","Ruby"),
              GT = c("Kim-Joy"),
@@ -104,7 +104,7 @@ if(doo[["week6"]]){
 
 if(doo[["week7"]]){
     L$week7 <-
-        ggol(H,
+        ggrl(H,
              SB = c("Kim-Joy"),  
              FB = c("Rahul"),
              GT = c("Briony","Manon"),
@@ -115,7 +115,7 @@ if(doo[["week7"]]){
 
 if(doo[["week8"]]){
     L$week8 <-
-        ggol(H,
+        ggrl(H,
              SB = c("Ruby"),  
[…]

### 2020-03-01 inst/masterchef_series6_allrounds.R

@@ -78,7 +78,7 @@ L <- list()  # overall list
 
 if(doo[["L1"]]){
 L$week1 <-  # 8 May 2014
-ggol(H,
+ggrl(H,
      top3       = c("Laura","Jamie","Sean"),
      btm21      = c("Emelia", "Tracy", "Sarah", "Colin", "Kira",
                     "Georgia", "Tash", "Byron", "Scott", "Emily",
@@ -91,7 +91,7 @@ ggol(H,
 
 if(doo[["L2a"]]){
 L$week2a <- # elimination 4 chosen 11 May; Deepali eliminated 12 May 2014
-ggol(H,
+ggrl(H,
             win = c("Sarah"),
              IN = c("Brent","Laura","Emelia","Tracy","Ben","Amy","Renae",
                     "Colin","Kira","Georgia","Tash","Byron","Sam","Rachael",
@@ -117,7 +117,7 @@ L$week2b[c( # winning team union losing team; 22 players
 }
 
 if(doo[["L2c"]]){  # 14 May
-L$week2c <- ggol(H,  # elimination, field of losing team from 13 May.
+L$week2c <- ggrl(H,  # elimination, field of losing team from 13 May.
             team_lose  = c("Laura","Amy","Tash","Rachael","Emily"),
             btm6       = c("Brent","Tracy","Byron","Steven","Scott"),
             eliminated = c("Nicole")
@@ -126,7 +126,7 @@ L$week2c <- ggol(H,  # elimination, field of losing team from 13 May.
 
 if(doo[["L3a"]]){
 L$week3a <-  # Nick eliminated 19 May, 
-    ggol(H,
+    ggrl(H,
          top3       = c("Laura","Tracy","Amy"),
          IN         = c("Brent","Emelia","Jamie","Ben","Renae","Colin",
                         "Kira","Georgia","Tash","Byron","Sam","Rachael",
@@ -150,7 +150,7 @@ L$week3b[c(  # winning team union losing team
 }
 
 if(doo[["L3c"]]){
-L$week3c <- ggol(H, # Emily eliminated 22 May
+L$week3c <- ggrl(H, # Emily eliminated 22 May
             team_lose  = c("Emelia","Tracy","Ben","Sam"),
             btm6       = c("Brent","Jamie","Georgia"),
             btm3       = c("Steven","Sean"),
@@ -159,7 +159,7 @@ L$week3c <- ggol(H, # Emily eliminated 22 May
 }
[…]

### 2020-03-01 inst/rowing_analysis.R

@@ -24,7 +24,7 @@ for(v in o){
 
 I <- hyper2(pnames=rowers)
 for(v in o){
-  I <- I+order_likelihood(character_to_number(v,rowers))
+  I <- I + rank_likelihood(character_to_number(v,rowers))
 }
 
 

### 2020-03-04 inst/skating_analysis.Rmd
more extensive changes reflecting ordertable_to_ranktable()

@@
 
-This script creates and analyses hyper2 object "skating" which is a
-likelihood function for the strengths of the competitors in the
-Ladies' Figure Skating at the 2002 Winter Olympics.  Dataframe
-`skating_table` is copied from Lock and Lock (see skating.Rd for
-more details).
+```{r, echo=T, results = 'hide', message= FALSE}
+library("hyper2",quietly=TRUE)
+```
 
-```{r}
-library("hyper2",quietly=TRUE)
+This document analyses strengths of the competitors in the Ladies'
+Figure Skating at the 2002 Winter Olympics.  In the package, dataframe
+`skating_table` is copied from Lock and Lock (see skating.Rd for more
+details).
+
+```{r,showskating}
 set.seed(0)
@@
 
-Object `skating_table` is structured so that each competitor is a
-row, and each judge is a column.  Function `rank_likelihood()`
-considers each row to be a race [or a judge], so we need to take a
-transpose.  The following R idiom is plausible but incorrect:
+Object `skating_table` is an **order table**.  It is structured so
+that each competitor is a row, and each judge is a column.  Function
+`rank_likelihood()` considers each row to be a race [or a judge], so
+we need to take a transpose.  The following R idiom is plausible but
+incorrect:
 
@@
 
-The above is incorrect because we want the ranks, and we have the
-order.  Correct analysis follows.
+The above is incorrect because we are applying a rank technique to an
+order table.  Correct analysis follows; firstly we coerce to a
+**rank** table:
 
 ```{r}
-skating_rank <- apply(skating_table,2,order)
-head(skating_rank)
+(skating_order <- ordertable_to_ranktable(skating_table))
 ```
 
-Column 1 of `skating_table` and column 1 of `skating_rank` are
-confusingly simular.  So consider column 2 instead, corresponding to
-judge number 2.  If `a=skating_table[,2]` and
-`b=skating_rank[,2]` we have `a=c(4,1,3,2, ...)` and
-`b=c(2,4,3,1, ...)`.  Thus `a` means that hughes was judged to
-be fourth best, slutskaya was judged to be the best, kwan third best,
-cohen second best, and so on.  Conversely, `b` means that the best
-competitor was number two [slutskaya], the second best number was
-number four [cohen], the third best was number three [kwan], the
-fourth best number 1 [hughes], and so on.  _Note that this means the
-rows of `skating_rank` cannot be named_.
+In the above, the column names should be read "came first" (`c1`),
+"came second" (`c2`) and so on to "came 23rd" (`c23`).  From the first
+column, we see that Hughes came first 5 times and Slutskya came first
+four times, and from the last column we see that all judges except
+`J3` awarded Luca last place.  File `man/rrank.Rd` has a detailed
+discussion of the differences between rank and order.
+
+We can coerce from an order table to a likelihood function:
 
 ```{r}
-skating <- rank_likelihood(t(skating_rank))
-pnames(skating) <- rownames(skating_table)
-head(skating)
+head(ordertable2supp(skating_order))
 ```
@@
 Slutskaya second, five gave Kawn second, and one gave Cohen second.
-The formal ordering is, I think, to rank the competitors by number of
-first places, then second, then third , and so on.  Ties at level $n$
-are broken by considering $n+1$.
+
+The formal ordering used in competition is, according to Lock and
+Lock, given by the median ordinal:
+
+```{r,medianordinal}
+apply(skating_table,1,median)
+```
+
+but with ties broken using a complicated hierarchy, the first of which
+is the "size of the majority":
+
+```{r,sizeofthemajority}
+rowSums(sweep(skating_table,1,seq_len(23))>=0)
+```
+
+which would suggest that Slutskaya beats Kwan (5-4), Butyrskaya draws
+with Robinson (7-7), and so on.  The rows of `skating_table` are in
+the order given by this system.
 
@@
 
-small <- 1e-6  # need a sensible start value satisfying the constraints
+small <- 1e-4  # need a sensible start value satisfying the constraints
 start <- c(rep(2*small,3),rep(3*small,2),rep(small,17))
@@
 for(i in seq_len(nrow(problem_constraints))){
-jj <- maxp(skating, startp=start, give=TRUE,fcm=problem_constraints[i,], fcv=0)
+  jj <- maxp(skating, startp=start, give=TRUE,fcm=problem_constraints[i,], fcv=0,n=1,SMALL=1e-5)
    fullout[[i]] <- jj
@@
 ```{r,unconstrainedlikes}
-mgv <- maxp(skating,give=TRUE)$value
+mgv <- maxp(skating, give=TRUE, n=2)$value
 mgv  - out

### 2020-03-04 inst/skating_analysis.Rmd
copyediting

@@
 
-```{r}
+```{r,skatingwrong}
 skating_incorrect <- rank_likelihood(t(skating_table))
@@
 
-```{r}
+```{r,coercetoranktable}
 (skating_order <- ordertable_to_ranktable(skating_table))
@@
 
-We can coerce from an order table to a likelihood function:
+We can convert an order table to a support function:
 
-```{r}
-head(ordertable2supp(skating_order))
+```{r,convertordertosupport}
+head(ordertable2supp(skating_table))
 ```

### 2020-03-05 inst/eurovision.Rmd
euro -> eurovision

@@
 Song Contest 2009," Wikipedia, accessed May 13, 2018.  It refers to
-semi-final 1.  More documentation is given in `euro.Rd` [type
+semi-final 1.  More documentation is given in `eurovision.Rd` [type
 `help(euro2009)` at the `R` prompt].
@@
 ```{r, specifymatrix}
-wiki_matrix <- matrix(c(
-#   ME  CZ  BE  BY  SW  AM  AD  CH  TR  IL  BG  IS  MK  RO  FI  PT  MT  BA  DE  UK
-    NA, 00, 00, 03, 00, 05, 01, 02, 05, 01, 00, 00, 08, 00, 00, 01, 06, 10, 02, 00, # ME
-    00, NA, 00, 00, 00, 00, 00, 00, 00, 00, 00, 00, 00, 00, 00, 00, 00, 00, 00, 00, # CZ
-    00, 00, NA, 00, 00, 01, 00, 00, 00, 00, 00, 00, 00, 00, 00, 00, 00, 00, 00, 00, # BE
-    02, 01, 00, NA, 01, 04, 00, 00, 00, 04, 01, 01, 06, 00, 04, 00, 01, 00, 00, 00, # BY
-    00, 06, 04, 07, NA, 08, 07, 04, 04, 07, 00, 10, 03, 04, 10, 08, 08, 04, 04, 07, # SW
-    04, 12, 10, 10, 05, NA, 00, 01, 10, 10, 08, 02, 02, 08, 01, 00, 00, 01, 10, 05, # AM
-    00, 00, 00, 00, 00, 00, NA, 00, 01, 00, 00, 00, 00, 00, 00, 04, 03, 00, 00, 00, # AD
-    00, 00, 00, 02, 02, 00, 02, NA, 00, 00, 00, 00, 00, 00, 05, 02, 00, 02, 00, 00, # CH
-    08, 05, 12, 06, 07, 10, 05, 12, NA, 06, 12, 07, 12, 12, 07, 05, 10, 12, 12, 12, # TR
-    05, 04, 03, 04, 06, 07, 08, 05, 03, NA, 04, 06, 01, 03, 06, 00, 04, 00, 05, 01, # IL
-    00, 00, 00, 00, 00, 00, 00, 00, 02, 00, NA, 00, 05, 00, 00, 00, 00, 00, 00, 00, # BG
-    07, 10, 07, 12, 12, 12, 10, 07, 08, 12, 06, NA, 04, 10, 12, 12, 12, 07, 06, 08, # IS
-    10, 03, 00, 00, 00, 00, 00, 06, 06, 00, 10, 00, NA, 02, 00, 00, 00, 08, 00, 00, # MK
-    06, 00, 02, 01, 00, 02, 04, 00, 07, 08, 05, 04, 07, NA, 00, 10, 02, 06, 01, 02, # RO
-    03, 00, 01, 00, 10, 00, 03, 00, 00, 00, 00, 12, 00, 01, NA, 03, 05, 00, 00, 04, # FI
-    00, 02, 06, 00, 03, 00, 12, 10, 00, 02, 02, 08, 00, 07, 02, NA, 00, 03, 07, 06, # PT
-    01, 07, 08, 08, 04, 03, 06, 03, 00, 05, 03, 05, 00, 06, 03, 06, NA, 05, 03, 10, # MT
-    12, 08, 05, 05, 08, 06, 00, 08, 12, 03, 07, 03, 10, 05, 08, 07, 07, NA, 08, 03) # BA
-, nrow=18,byrow=TRUE)
+data(eurovision)
+(wiki_matrix <- as.matrix(eurovision2009_votingtable))
 ```
@@
 of the points), replaced self-voting entries with `NA`, and replaced
-blanks with `00`; every entry is padded to two digits.  Thus the first
-row corresponds to votes cast for `ME`, Montenegro.  We see that `BE`
-(Belgium) awarded them three points, `am` (Armenia) five points, and
-so on.
+blanks with `0`.  Thus the first row corresponds to votes cast for
+`ME`, Montenegro.  We see that `BE` (Belgium) awarded them three
+points, `am` (Armenia) five points, and so on.
 
@@
 any points from Montenegro.  We need to convert matrix `preference`
-into a likelihood function, here `euro2009`:
+into a likelihood function, here `eurovision2009`:
 
 ```{r,morestuff}
-euro2009 <- hyper2(d=18)
+eurovision2009 <- hyper2(d=18)
 
@@
         eligible <- which(d>=0)   # This is why we set NA values to -1
-        euro2009[which(d==1)] %<>% "+"(1)
+        eurovision2009[which(d==1)] %<>% "+"(1)
                                             # The first choice among
@@
 
-        euro2009[eligible] %<>% dec()   # denominator of all eligible players
+        eurovision2009[eligible] %<>% dec()   # denominator of all eligible players
 
@@
 ## syntatic sugar:
-pnames(euro2009) <- competitors
+pnames(eurovision2009) <- competitors
 ```
@@
 ```{r,checknull}
-m <- maxp(euro2009)  # MLE
-print(loglik(indep(m),euro2009))
-print(loglik(indep(equalp(euro2009)),euro2009))
+m <- maxp(eurovision2009)  # MLE
+print(loglik(indep(m),eurovision2009))
+print(loglik(indep(equalp(eurovision2009)),eurovision2009))
 ```

### 2020-03-08 inst/formula1.Rmd
conditions, start work

@@
 We see that Hamilton has the highest estimated strength.
+
+# Conditions
+
+Taking 2017 as an example:
+
+`
+venue pole conditions\
+AUS Hamilton   dry \
+CHN Hamilton   wet \
+BHR Bottas     dry \
+RUS Vettel     dry \
+ESP Hamilton   dry \
+MON Raikkonen  dry \
+CAN Hamilton   dry \
+AZE Hamilton   dry \
+AUT Bottas     dry \
+GBR Hamilton   dry \
+HUN Vettel     dry \
+BEL Hamilton   dry \
+ITA Hamilon    dry \
+SIN Vettel     wet \
+MAL Hamilton   dry \
+JPN Hamilton   dry \
+USA Hamilton   dry \
+MEX Vettel     dry \
+BRA Bottas     dry \
+ABU Bottas     dry \
+`

### 2020-03-10 inst/formula1_extra.Rmd
initial work on pole position

---
title: "formula1: further tests"
output: html_document
---

```{r setup, include=FALSE}
set.seed(0)
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("magrittr")
```

## Formula 1

(this document follows on from `formula1.Rmd`).

Here I analyse results from the 2017 series of Formula 1 motor racing
for two interesting hypotheses: that Lewis Hamilton does indeed
perform better in the wet, and also whether pole position makes a
difference.

Consider the following dataset, taken from Wikipedia:

```{r readformula1}
f2017 <- read.table("formula1_2017.txt",header=TRUE)
head(f2017)
```

Thus, reading the top row, Hamilton placed second in Australia, first
in China, second in Bahrain, and so on.  Reading the first column, we
see that in Australia, Hamilton came second, Vettel first, and so on.

```{r,readcond}
(cond <- read.table("formula1_conditions.txt",header=TRUE))
```

Thus from the fourth row, for example, we see that in Russia the
conditions were dry and Vettel drove in pole position.  We might ask whether

```{r dotable}
TT <- table(wins=as.vector(f2017[1,1:20]==1),pole=cond$pole == "Hamilton")
fisher.test(TT)
```

We are going to use a generalization of function
\code{ordertable2supp()} that incorporates pole position and condition
information:

```{r,applydata}
```

See how the rows of `a2017` correspond to the columns of `f2017`.  In
the following, we will define a `hyper2` object `H` which has a
reified player `pole` whose strength represents the advantage
conferred by driving pole position.

```{r definegeneralizedordertable2supp}
f2017 <- read.table("formula1_2017.txt",header=TRUE)
cond <- read.table("formula1_conditions.txt",header=TRUE)

x <- read.table("formula1_2017.txt",header=TRUE)
noscore <- c("Ret", "WD", "DNS", "DSQ", "DNP", "NC")
venues <- as.vector(colnames(x)[-ncol(x)])
jj <- apply(x[,-ncol(x)], 2, function(y){`[<-`(y,y %in% noscore,0)})

fmat <- matrix(as.numeric(jj), byrow = TRUE, ncol = nrow(x))
colnames(fmat) <- rownames(x)
rownames(fmat) <- venues

out <- hyper2(pnames = c(colnames(fmat),"pole"))

for(i in seq_len(nrow(fmat))){
  jj1 <- ordervec2supp(fmat[i,, drop=TRUE])
  pwa(jj1,as.vector(cond$pole[i]),"pole")
  out  <- out + pwa(jj1,as.vector(cond$pole[i]),"pole")
}
(mo <- maxp(out))
pie(mo)
```

Now we can use it to create a profile likelihood curve along the same
lines as table tennis

### 2020-03-11 inst/formula1.Rmd
more work

@@
 
-Each row is a driver and each column (after the first) a venue.  We
-see that Hamilton, the first row, came second in Australia, first in
-China, second in Bahrain, fourth in Russia, and so on.  In the first
-column, we see the result from Australia (AUS) in which Hamilton came
-second, Vettel first, Bottas third, and so on.  Notation used also
-includes six different classes of no-score such as "Ret" for retired,
-"WD" for withdrawn, and so on.  I now use bespoke R function
-\code{ordertable2supp()}, that translates data of this type into a
-log-likelihood.  The function is similar to, but slightly different
-from, the analysis in `eurovision.R`.
+(in the package this is object `F1_table_2017`).  Each row is a driver
+and each column (after the first) a venue.  We see that Hamilton, the
+first row, came second in Australia, first in China, second in
+Bahrain, fourth in Russia, and so on.  In the first column, we see the
+result from Australia (AUS) in which Hamilton came second, Vettel
+first, Bottas third, and so on.  Notation used also includes six
+different classes of no-score such as "Ret" for retired, "WD" for
+withdrawn, and so on.
+
+We can represent this data in a different form using
+`wikitable_to_ranktable()`:
+
+```{r}
+wikitable_to_ranktable(f2017)
+```
+
+Looking at the first column ("c1" is read "came first") we see who
+came first at each venue: Vettel in Australia (AUS), Hamilton in China
+(CHN), Vettel in Bahrain (BHR), and so on.  I now use bespoke R
+function \code{ordertable2supp()}, that translates data of this type
+into a log-likelihood.  The function is similar to, but slightly
+different from, the analysis in `eurovision.R`.
 
@@
 
+head(F1_total)
+```
+
+So object `F1_total` is a likelihood function for all four years data,
+on the assumption of independent observations and in this case driver
+strength being independent of year.  We can find the evaluate in the same way:
+
+```{r,findmallyears}
 mallyears <- maxp(F1_total)

### 2020-03-12 inst/formula1_extra.Rmd
just the strongest 8

@@
 
+Now we can consider only the strongest say 8 competitors.  There are
+two ways to proceed.  Firstly, we can retain just the 8 strongest
+competitors as estimated from the full likelihood function:
+
+```{r,strongestnineanotherapproach,cache=TRUE}
+mo8a <- sort(mo,decreasing=TRUE)[seq_len(8)]
+pie(mo8a)
+```
+
+Or secondly, we can simply erase everyone but the first 8 from the
+likelihood function `out`  and use that:
+
+```{r,strongestnine,cache=TRUE}
+out8 <- keep(out,seq_len(8))
+head(out8)
+mo8 <- maxp(out8)
+dotchart(log(mo8))
+pie(mo8)
+```
+
+Note how the two approaches are not equivalent: the second approach
+has redistributed the estimated strengths and, for example, Ricciardo
+has an estimated strength of essentially zero (because he consistently
+loses to the other 7 players).
+
+# Pole position
+
+Driving pole position is an important thing:
+
 ```{r,readcond}
@@
 Thus from the fourth row, for example, we see that in Russia the
-conditions were dry and Vettel drove in pole position.  We might ask whether
+conditions were dry and Vettel drove in pole position.  We might ask
+whether Hamilton's winning is associated with his driving at pole
+position:
 
-```{r dotable}
-TT <- table(wins=as.vector(f2017[1,1:20]==1),pole=cond$pole == "Hamilton")
+```{r makehamiltontable}
+(TT <- table(wins=as.vector(f2017[1,1:20]==1),pole=cond$pole == "Hamilton"))
+```
+
+Thus from the above we see that, of Hamilton's 9 wins, 8 of them were
+from pole; conversely, of his 11 pole races, he lost only 3.
+
+```{r,fishertestpoleham}
 fisher.test(TT)
@@
 
-We are going to use a generalization of function
-\code{ordertable2supp()} that incorporates pole position and condition
-information:
-
-```{r,applydata}
-```
+We are going to use a generalization of function `ordertable2supp()`
+that incorporates pole position and condition information:
 
@@
 
-for(i in seq_len(nrow(fmat))){
-  jj1 <- ordervec2supp(fmat[i,, drop=TRUE])
-  pwa(jj1,as.vector(cond$pole[i]),"pole")
-  out  <- out + pwa(jj1,as.vector(cond$pole[i]),"pole")
+for(i in seq_len(nrow(fmat))){  # iterate through the races
+  jj1 <- ordervec2supp(fmat[i,, drop=TRUE])      # jj1 is a LF for a single race...
+  jj2 <- pwa(jj1,as.vector(cond$pole[i]),"pole") # jj2 includes a pole ghost racer
+  out  <- out + jj2                              # now increment 'out' with ghost LF 
 }
+head(out)
+```
+
+In the above, we see ghost driver `pole`.
+
+```{r,maximizeoutandpie,cache=TRUE}
 (mo <- maxp(out))
@@
 
+# Profile likelihood of pole strength
+
 Now we can use it to create a profile likelihood curve along the same
-lines as table tennis
+lines as table tennis, but for simplicity we will keep just the
+strongest 9 competitors:
+
+```{r,justnine}
+4+5
+```

### 2020-03-16 inst/universities.Rmd
university rankings analysis

---
title: "universities"
output: html_document
---

```{r setup, include=FALSE}
library("hyper2")
library("magrittr")
knitr::opts_chunk$set(echo = TRUE)
```

## University ranking data

First load the data:

```{r cars}
a <- read.table("universities.txt",header=TRUE)
head(a,12)
```

Each row corresponds to a ranking in the THE ranking system.  Thus the
first row refers to The University of Auckland 2012, and we see that
they scored 28.3 for their teaching, 25.0 for their research, and so
on; rows 3 and 4 show that AUT and Lincoln were not given rankings
that year.  Rows 1-8 give results for 2012, rows 9-16 give 2013, and
so on to 2020.

We can coerce each year to a `hyper2` likelihood function by
converting to an order table with `order()`, thence to a support function:

```{r, coerceem}
f <- function(y,supp=TRUE,...){
  jj <- as.data.frame(a)
  jj <- subset(jj,jj$year==y)
  o <- function(v){
  good <- v != "DNP"
     v[good] <- order(v[good],decreasing=TRUE)
     return(v)
     }	
  x <- apply(jj,2,o)
  rownames(x) <- jj[,1]
  x <- x[,3:6]
  if(supp){
    return(ordertable2supp(x,...))
  } else {
  return(as.data.frame(x))
  }		
}
f(2012)
```

Now:
```{r,playerwithadvantage}
H <- f(2013) + f(2014) + f(2015)                          # pre  2016
H1 <- H + f(2016) + f(2017) + f(2018) + f(2019) + f(2020) # post 2016
```

Object `H1` is simply the support function for the universities'
strenghts for observations (assumed to be independent) 2013-2020:

```{r}
maxp(H)
```

Note the dominance of Auckland (UoA).  We observe that AUT has
increased its strength since 2016 and we can test whether this
increase is significant:

```{r}
pnames(H) <- c(pnames(H),"S")
H2 <- H + pwa(f(2016),"AUT") + pwa(f(2017),"AUT") +
          pwa(f(2018),"AUT") + pwa(f(2019),"AUT") +
  	  pwa(f(2020),"AUT")
head(H2)
```

Object `H2` is a support function for the Bradley-Terry strengths of
the universities, but with AUT having extra strength `S` post-2016.
We can test the null that `S=0` using package idiom
\code{specificp.test()}:

```{r}
specificp.test(H2,"S",0)
```

See how we reject $H_0\colon S=0$ with a p-value of about 0.00676,
significant at 1\%.

### 2020-03-19 inst/home_advantage.Rmd
explicitly name rownames and colnames

@@
 teams <- LETTERS[1:4]
-rownames(home_games_won) <- teams
-colnames(home_games_won) <- teams
-rownames(away_games_won) <- teams
-colnames(away_games_won) <- teams
+dimnames(home_games_won) <- list("@home" = teams,"@away"=teams)
+dimnames(away_games_won) <- list("@home" = teams,"@away"=teams)
 

### 2020-03-19 inst/home_advantage.Rmd
copyedit

@@
 
-Thus ```home_games_won[1,2] == 16``` means A played at home against B
-and won 16 times; ```home_games_won[2,1] == 19``` means B played at
-home against A and won 19 times.  Also, ```away_games_won[1,2] == 5```
-means A played away against B and won 5 times, and
-```away_games_won[2,1] == 2``` means B played away against A and won 2
-times.  Alternatively, A played B 16+19+5+2=45 times; A played at home
-16+2=18 times and won 16 times and lost 2 times; and B played at home
-19+5=24 times and won 19 times and lost 5 times.
+Thus `home_games_won[1,2] == 16` means A played at home against B and
+won 16 times; `home_games_won[2,1] == 19` means B played at home
+against A and won 19 times.  Also, `away_games_won[1,2] == 5` means A
+played away against B and won 5 times, and `away_games_won[2,1] == 2`
+means B played away against A and won 2 times.  Alternatively, A
+played B 16+19+5+2=45 times; A played at home 16+2=18 times and won 16
+times and lost 2 times; and B played at home 19+5=24 times and won 19
+times and lost 5 times.
 
@@
 Now a likelihood ratio for the hypothesis that home strength (or home
-advantage) is zero.  First define a function ```f()``` that maximises
-the likelihood subject to the home strength being *less than* its
+advantage) is zero.  First define a function `f()` that maximises the
+likelihood subject to the home strength being *less than* its
 argument:
@@
 
-See how the ```fcm``` and ```fcv``` constraints make the sum of
-a,b,c,d, greater than ```1-home```.  Thus 'home' [which is the fillup
-value, $p_\mathrm<{home}=1-(p_A+p_B+p_C+p_D)$] is constrained to be
-less than argument ```home```.  Using this function, the likelihood
-ratio test is straightforward:
+See how the `fcm` and `fcv` constraints make the sum of a,b,c,d,
+greater than `1-home`.  Thus 'home' [which is the fillup value,
+$p_\mathrm{home}=1-(p_A+p_B+p_C+p_D)$] is constrained to be less than
+argument `home`.  Using this function, the likelihood ratio test is
+straightforward:
 

### 2020-03-23 inst/soling.Rmd
start work on issue #46

---
title: "Soling at the 2000 Summer Olympics"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The Soling is a type of boat raced in the 2000 Summer Olympics.  Here I analyse race data taken from

  \url{https://en.wikipedia.org/w/index.php?title=Sailing_at_the_2000_Summer_Olympics_\%E2\%80\%93_Soling&oldid=945362535}

using a Plackett-Luce extension of the Bradley-Terry model.

```{r loadlib}
library("hyper2")
data(soling)
soling_table_2000
ordertable_to_ranktable(soling_table_2000)
head(soling2000)
```

We can also consider observations from the round robins and quarter finals:

```{r,rrqf}
soling2000_rr1
soling2000_rr2
soling2000_qf
```

Thus, for example, taking `soling2000_rr1[1,2] == TRUE` we see that
Germany beat Sweden in round robin 1.  To incorporate this information
we need a little bespoke function:

```{r, definef}
rr <- function(H,M){
	for(i in seq_len(nrow(M)-1)){
		for(j in (i+1):ncol(M)){
			if(M[i,j]){
				winner <- i
				loser  <- j
			} else {
				winner <- j
				loser <- i
			}
			H[winner] %<>% inc
			H[c(winner,loser)] %<>% dec
		}
	}
	return(H)
}
```

now

```{r}
before <- maxp(soling2000)
soling2000 %<>% rr(soling2000_rr1)
soling2000 %<>% rr(soling2000_rr2)
soling2000 %<>% rr(soling2000_qf)
after <- maxp(soling2000)
par(pty="s")
plot(before,after,asp=1,pch=16,xlim=c(0,0.3),ylim=c(0,0.3))
abline(0,1)
```

### 2020-03-23 data/soling.R

@@ -1,16 +1,54 @@
-`soling2000` <- 
-structure(list(I = c(5L, 3L, 1L, 4L, 2L, 8L, 13L, 12L, 9L, 14L, 
-7L, 6L, 10L, 16L, 11L, 15L), II = c(2L, 3L, 5L, 14L, 8L, 11L, 
-7L, 1L, 13L, 10L, 15L, 16L, 9L, 6L, 4L, 12L), III = c(1L, 10L, 
-13L, 2L, 4L, 6L, 8L, 7L, 11L, 3L, 5L, 12L, 9L, 14L, 15L, 16L), 
-    IV = c(3L, 2L, 11L, 16L, 8L, 1L, 4L, 6L, 9L, 7L, 10L, 12L, 
-    13L, 5L, 14L, 15L), V = c(5L, 7L, 1L, 2L, 6L, 9L, 4L, 15L, 
-    3L, 10L, 16L, 11L, 12L, 14L, 13L, 8L), VI = c(16L, 3L, 1L, 
-    2L, 9L, 14L, 13L, 12L, 7L, 11L, 5L, 4L, 8L, 10L, 15L, 6L)),
-class = "data.frame", row.names = c("norway", 
-"nz", "netherlands", "USA", "russia", "australia", "UK", "ukraine", 
-"france", "germany", "sweden", "denmark", "canada", "italy", 
-"finland", "spain"))
+`soling2000` <-
+    structure(list(
+        brackets = list(1, c(1, 2, 3, 4, 5, 6, 7, 8, 9, 
+10, 11, 12, 13, 14, 15, 16), c(1, 2, 3, 4, 5, 6, 7, 9, 10, 11, 
+12, 13, 14, 15, 16), c(1, 2, 3, 4, 5, 7, 8, 9, 10, 11, 12, 13, 
+14, 15, 16), c(1, 2, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 
+16), c(1, 2, 4, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16), c(1, 
+2, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16), c(1, 2, 5, 6, 
+7, 8, 10, 11, 12, 13, 14, 15, 16), c(1, 2, 5, 6, 8, 10, 11, 12, 
+13, 14, 15, 16), c(1, 3, 4, 5, 7, 8, 9, 10, 11, 12, 13, 14, 15, 
[…]

### 2020-04-01 inst/formula1.Rmd
Hamilton in the wet

@@
 library("magrittr")
+options("digits" = 5)
 ```
@@
 Above, we see a small part of the Placket-Luce likelihood function for
-the 2017 results.  We can find the maximum likelihood estimator by
-using the `maxp()` function:
+the 2017 results.  But from now on we will use built-in dataset
+`formula1`, which contains extra information including non-finishing
+competitors.  We can find the maximum likelihood estimator by using
+the `maxp()` function:
 
-```{r,maxp2017}
-mL2017 <- maxp(L2017)
+```{r,maxp2017,cache=TRUE}
+rm(L2017)
+mL2017 <- maxp(formula1)
 mL2017
@@
 
-m <- mL2017   # maxp(L2017)
+m <- mL2017   # maxp(formula1)
 ox <- rank(-points) # negative because number 1 = largest number of points
@@
 
-## Null of equal strengths:
+## Hypothesis: all players have equal strengths
 
-```{r,equalstrength}
-L1 <- loglik(indep(equalp(L2017)),L2017)
-L2 <- loglik(indep(mL2017),L2017)   # mL2017 == maxp(L2017)
-pchisq(2*(L2-L1),df=25-1,lower.tail=FALSE)
+Here we write $H_0\colon p_1=p_2=\cdots =p_n=\frac{1}{n}$.
+
+```{r,equalstrength,cache=TRUE}
+equalp.test(formula1)
 ```
@@
 We can consider Hamilton and ask various questions about his playing
-strength.  One natural null would be that Hamilton's strength is equal
-to the mean strength, or $\frac{1}{25}$.
+strength.  One natural hypothesis would be that Hamilton's strength is
+equal to the mean strength, or $\frac{1}{25}$.
 
-* $H_2\colon p_\mathrm{Hamilton}\geq\frac{1}{25}$
-* $H_3\colon p_\mathrm{Hamilton} < \frac{1}{25}$
+* $H_A\colon\sum p_i=1$
+* $H_0\colon p_\mathrm{Hamilton} < \frac{1}{25}$
 
-Noting that the evaluate for Hamilton's strength
-$\widehat{p_\mathrm{Hamilton}}$ is about 0.22, we may consider the
-likelihood for $H_2$ to be equal to the unconstrained maximum
-likelhood as in `L2` above.  The likelihood for $H_3$ requires a
-constrained optimization:
+(we attempt to reject the null $H_0$).  Noting that the evaluate for
+Hamilton's strength $\widehat{p_\mathrm{Hamilton}}$ is about 0.29, we
+ask how much support is lost, relative to the support at the
+unconstrained evaluate.
 
-```{r,constham}
-small <- 0.01
-L3 <- maxp(L2017, startp=c(small, rep((1-small)/24,23)),fcm=rbind(c(-1,rep(0,23))),fcv=-1/25,give=TRUE)
-L3
+```{r,specifichamiltonstrength,cache=TRUE}
+specificp.test(formula1,"Hamilton")  # default value is 1/25=0.04
 ```
 
-(note that the constrained maximum attains the constraint of
-$p_\mathrm{Hamilton}=\frac{1}{25}$).   The likelihood ratio test gives
-
-```{r,liktestivediff}
-L2 - L3$value 
-```
-
-or a p-value of
-
-```{r,liktestive}
-pchisq(2*(L2 - L3$value),df=1,lower.tail=FALSE)
-```
+(note that the constrained maximum is on the boundary of the
+admissible region with $p_\mathrm{Hamilton}\approxequal\frac{1}{25}$).
 
@@
 
-This is quite difficult to evaluate formally but we can evaluate it with no loss in accuracy by observing that the second strongest competitor is Bottas and considering instead
+This is quite difficult to evaluate formally but we can evaluate it
+with no loss in accuracy by observing that the second strongest
+competitor is Bottas and considering instead
 
@@
 
-```{r,L4like}
+```{r,L4like,cache=TRUE}
 small <- 0.01
-L4 <- maxp(L2017, startp=c(small, rep((1-small)/24,23)),fcm=rbind(c(-1,1,rep(0,22))),fcv=0,give=TRUE)
+L4 <- maxp(formula1, startp=c(small, rep((1-small)/24,23)),fcm=rbind(c(-1,1,rep(0,22))),fcv=0,give=TRUE)
 L4
@@
 ```{r,l4val}
-L2 - L4$value 
-pchisq(2*(L2 - L4$value),df=1,lower.tail=FALSE)
+#L2 - L4$value 
+#pchisq(2*(L2 - L4$value),df=1,lower.tail=FALSE)
+2+3
 ```
@@
 
-```{r,fouryearsdata}
+```{r,fouryearsdata, cache=TRUE}
 F1_2014 <- "formula1_2014.txt" %>% read.table(header=TRUE) %>% ordertable2supp
@@
 
-```{r,findmallyears}
+```{r,findmallyears,cache=TRUE}
 mallyears <- maxp(F1_total)
@@
 
-Taking 2017 as an example:
+```{r,loadtable}
+(details <- read.table("formula1_conditions.txt",header=TRUE))
+drivers <- pnames(formula1)
+```
 
-`
-venue pole conditions\
-AUS Hamilton   dry \
-CHN Hamilton   wet \
-BHR Bottas     dry \
-RUS Vettel     dry \
-ESP Hamilton   dry \
-MON Raikkonen  dry \
-CAN Hamilton   dry \
-AZE Hamilton   dry \
-AUT Bottas     dry \
-GBR Hamilton   dry \
-HUN Vettel     dry \
-BEL Hamilton   dry \
-ITA Hamilon    dry \
-SIN Vettel     wet \
-MAL Hamilton   dry \
-JPN Hamilton   dry \
-USA Hamilton   dry \
-MEX Vettel     dry \
-BRA Bottas     dry \
-ABU Bottas     dry \
-`
+We cannot use `ordertable2supp()` here because we somehow need to
+incorporate pole position and conditions into our likelihood function.
+The first step is to modify `f2017` so that non-finishers have a zero
+position:
+
+At this point, we introduce a new player: `Ham_wet`, for
+"Hamilton in the wet", and we can test the hypothesis that the
+`Hamilton == Ham_wet`.  We will cycle through all the venues and,
+if it is wet (according to dataframe `conditions`) we will credit
+Hamilton's placing to `Ham_wet`.
+
+```{r,hamiltonwetanddry,cache=TRUE}
+f2017 <- read.table("formula1_2017.txt",header=TRUE)
+f2017 <- f2017[,-ncol(f2017)]
+details <- read.table("formula1_conditions.txt",header=TRUE)
+
+drivers <- rownames(f2017)
+alldrivers <- c("Ham_dry","Ham_wet",drivers[-1])
+
+L2017 <- hyper2(pnames=alldrivers) # uninformative
+for(i in seq_len(ncol(f2017))){
+	venue <- colnames(f2017)[i]
+	conditions <- details[which(details$venue == venue),4]
+	o <- f2017[,i]
+        suppressWarnings(o <- as.numeric(levels(o))[o])
+        o[is.na(o)] <- 0
+	names(o) <- rownames(f2017)
+	if(conditions == "wet"){
+	   names(o)[names(o) == "Hamilton"] <- "Ham_wet"
+	} else if(conditions == "dry"){
+	   names(o)[names(o) == "Hamilton"] <- "Ham_dry"
+	} else {
+	   stop("neither wet nor dry?")
+	}
+	L2017 <- order_obs(L2017,o)
+}  # for(i) loop closes
+head(L2017)
+```
+
+We can now test the hypothesis that Hamilton's performance in the wet
+is the same as his performance in the dry:
+
+```{r,hamwetdrytest,cache=TRUE}
+options(width = 80)
+samep.test(L2017,c("Ham_wet","Ham_dry"))
+```

### 2020-04-01 inst/formula1.Rmd
idiom

@@
 	}
-	L2017 <- order_obs(L2017,o)
+	L2017 %<>% order_obs(o)
 }  # for(i) loop closes

### 2020-04-01 inst/soling.Rmd
consistentify

@@
 data(soling)
-soling_table_2000
+H <- soling_table_2000
 ordertable_to_ranktable(soling_table_2000)
@@
 
-now
+We can see the effect on the evaluate of incorporating the additional
+observations into the Plackett-Luce likelihoods:
 
 ```{r}
-before <- maxp(soling2000)
-soling2000 %<>% rr(soling2000_rr1)
-soling2000 %<>% rr(soling2000_rr2)
-soling2000 %<>% rr(soling2000_qf)
-after <- maxp(soling2000)
+H <- soling2000
+before <- maxp(H)
+H %<>% rr(soling2000_rr1)
+H %<>% rr(soling2000_rr2)
+H %<>% rr(soling2000_qf)
+after <- maxp(H)
 par(pty="s")
@@
 ```
+
+The diagram above shows a modest change.  We might ask whether the
+additional observations are consistent with the original Placket-Luce
+likelihood:
+
+```{r}
+(H2 <- H + (-1)*soling2000)
+```

### 2020-04-01 inst/formula1.Rmd
use pwa()

@@
 ```
+(the p-value seems to be too low to be believable).
+
+## Try the same thing but with function pwa()
+
+```{r,trypwa,cache=TRUE}
+f2017 <- read.table("formula1_2017.txt",header=TRUE)
+f2017 <- f2017[,-ncol(f2017)]
+details <- read.table("formula1_conditions.txt",header=TRUE)
+
+H <- hyper2(pnames=c(rownames(f2017),"S"))
+noscore <- c("Ret", "WD", "DNS", "DSQ", "DNP", "NC")
+
+for(i in seq_len(ncol(f2017))){
+	y <- f2017[,i]
+	if(any(y %in% noscore)){y[y%in%noscore] <- 0}
+	y <- as.numeric(levels(y)[y])
+	names(y) <- rownames(f2017)
+	cond <- details$conditions[i]
+        jj <- ordervec2supp(y)
+	if(cond == "dry"){
+            pnames(jj) <- pnames(H)  # do nothing!
+	} else if (cond == "wet"){
+            jj <- pwa(jj,"Hamilton")
+	} else {
+            stop("neither wet nor dry?")
+	}  # if wet closes
+        H <- H + jj
+} # i loop closes
+head(H)
+maxp(H)
+#specificp.test(H,"S",0)
+```
+
+see this

### 2020-04-02 inst/formula1.Rmd
forgot to change this

@@
 maxp(H)
-#specificp.test(H,"S",0)
+```
+
+Now we can test the hypothesis that Hamilton's wet strength `S` is zero:
+
+```{r,testwetstrengthzero}
+pnames(H)
+H9 <- keep(H,c("S",pnames(H)[1:8]))
+specificp.gt.test(H9,"S",0)
 ```

### 2020-04-03 inst/curling.Rmd
new curling analysis

---
title: "curling"
output: html_document
---

```{r setup, include=FALSE}
set.seed(0)
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("magrittr")
options("digits" = 5)
```

## Curling

Curling is a sport with stones and brooms.  Following dataset taken
from [wikipedia](https://www.google.com "Google's Homepage"), women's
tournament:

```{r readcurlingtable}
a <- read.table("curling.txt",header=TRUE)
a
```

Thus we see that Canada came first in 1998, third in 2002, and so on;
in 1998, Canada came first, China did not compete [`NA` meaning did
not compete], Denmark came second, and so on.  Note that `NA` entries
are different from the `DNF` category used in formula 1: the inference
from a `DNF` entry is that the player came last, but here we can make
no such inference as the team did not compete.  This means that
function `wikitable_to_ranktable()` is not appropriate and we need to use a different method:

```{r,makehyper2curling}
H <- hyper2(pnames=rownames(a))

for(i in seq_len(ncol(a))){
    x <- a[,i]
    names(x) <- rownames(a)
    H %<>% order_obs(x[!is.na(x)])
}
head(H)
(mH <- maxp(H))
pie(mH)
```

### 2020-04-04 inst/curling.Rmd
more work on curling

@@
 ```{r readcurlingtable}
-a <- read.table("curling.txt",header=TRUE)
-a
+curling_table  <- read.table("curling.txt",header=TRUE)
+curling_table
 ```
@@
 in 1998, Canada came first, China did not compete [`NA` meaning did
-not compete], Denmark came second, and so on.  Note that `NA` entries
-are different from the `DNF` category used in formula 1: the inference
-from a `DNF` entry is that the player came last, but here we can make
-no such inference as the team did not compete.  This means that
-function `wikitable_to_ranktable()` is not appropriate and we need to use a different method:
+not compete], Denmark came second, and so on.
+
+First we will interpret `NA` in the same way as `DNS` in the context
+of formula 1 motor racing.  We need to define a new data frame, `b`:
+
+```{r,defineb}
+ct <- curling_table
+ct[is.na(ct)] <- "DNS"
+ct
+```
+
+In this form we can easily convert to a support function:
+
+```{r,definsupportb}
+H1 <- ordertable2supp(ct)
+head(H1)
+(mH1 <- maxp(H1))
+pie(mH1)
+```
+
+## Interpretation of `NA`
+
+There is a different interpretation of `NA` entries.  We might hold
+that no inference should be made from a team's entry being `NA` entry
+because that that team did not compete.  This means that function
+`ordertable2supp()` is not appropriate and we need to use a different
+method:
 
 ```{r,makehyper2curling}
-H <- hyper2(pnames=rownames(a))
+H2 <- hyper2(pnames=rownames(curling_table))
 
-for(i in seq_len(ncol(a))){
-    x <- a[,i]
-    names(x) <- rownames(a)
-    H %<>% order_obs(x[!is.na(x)])
+for(i in seq_len(ncol(curling_table))){
+    x <- curling_table[,i]
+    names(x) <- rownames(curling_table)
+    H2 %<>% order_obs(x[!is.na(x)])
 }
-head(H)
-(mH <- maxp(H))
-pie(mH)
+head(H2)
+(mH2 <- maxp(H2))
+pie(mH2)
 ```
+
+##  Comparison
+
+```{r,plotcomplog}
+par(pty="s")
+plot(log10(mH1),log10(mH2),asp=1,pch=16,xlim=c(-2.6,0),ylim=c(-2.6,0))
+posn <- c(Canada=1, China=2, Denmark=4, Germany=4, GB=4, Italy=1, Japan=4, Norway=4,
+          Russia=2, South_Korea=4, Sweden=2, Switzerland=2, US=4)
+for(i in seq_len(nrow(curling_table))){
+	text(log10(mH1)[i],log10(mH2)[i],rownames(curling_table)[i],col="gray",pos=posn[i])
+}
+abline(0,1)
+grid()
+```
+
+Note that the two strongest competitors, Canada and Sweden, switch
+between first and second in the two systems.

### 2020-04-05 inst/cook.Rmd
use pwa() instead of working by hand

@@
 ```{r,beforetraining}
-H <- hyper2(pnames=c(paste("p",1:5,sep=""),"t"))  # 't' the strength of training
+H <- hyper2(pnames=c(paste("p",1:5,sep="")))  # 'S' the strength of training
 # observations before training:
@@
 
-We can translate these into package idiom by defining likelihood
-functions `H1after` and `H2after` for the two observations:
+We can translate these into package idiom with
+function `pwa()` ("person with advantage"):
 
-```{r,aftertraining1}
-H1after <- H
-H1after[c("p1","t")] %<>% inc()
-H1after[  "p2"     ] %<>% inc()
-H1after[  "p3"     ] %<>% inc()
-H1after[  "p4"     ] %<>% inc()
-H1after[  "p5"     ] %<>% inc()
-H1after[c("p1","t","p2","p3","p4","p5")] %<>% dec()
-H1after[c(         "p2","p3","p4","p5")] %<>% dec()
-H1after[c(              "p3","p4","p5")] %<>% dec()
-H1after[c(                   "p4","p5")] %<>% dec()
-H1after[c(                        "p5")] %<>% dec()
-H1after
+```{r}
+Hafter <- rank_likelihood(1:5)+rank_likelihood(c(3,1,5,2,4))
+pnames(Hafter) <- pnames(Hbefore)
+Hafter %<>%  pwa("p1")
+Hafter
 ```
 
-```{r,aftertraining2}
-H2after <- H
-H2after[  "p3"     ] %<>% inc()
-H2after[c("p1","t")] %<>% inc()
-H2after[  "p5"     ] %<>% inc()
-H2after[  "p2"     ] %<>% inc()
-H2after[  "p4"     ] %<>% inc()
-H2after[c("p1","t","p2","p3","p4","p5")] %<>% dec()
-H2after[c("p1","t","p2",     "p4","p5")] %<>% dec()
-H2after[c(         "p2",     "p4","p5")] %<>% dec()
-H2after[c(         "p2",     "p4"     )] %<>% dec()
-H2after[c(                   "p4"     )] %<>% dec()
-H2after
+In the above, the additional strength conferred by the training is
+represented by `S`.  We can estimate the effect of training:
+
+```{r}
+pnames(Hbefore) <- c(pnames(Hbefore),"S")
+maxp(Hbefore + Hafter)
 ```
 
-Combine:
+The above suggesting that the training had a sizable effect; and we
+may test the null that the training had no effect:
 
-```{r,combineboth}
-(Htot <- Hbefore + H1after + H2after)  # observations are independent!
-(ML_unconstrained <- maxp(Htot))
+```{r}
+specificp.gt.test(Hbefore + Hafter,"S",0)
 ```
 
-We see the training entity has an estimated strength of about 0.63,
-higher than all real players combined.  At the evaluate the
-loglikelihood is
-
-```{r,loglikateval}
-loglik(indep(ML_unconstrained),Htot)
-```
-
-We can test the hypothesis that the training was ineffective by
-constraining the optimization to $t=0$:
-
-```{r}
-small <- 1e-4
-(ML_constrained <- maxp(Htot,startp=rep(0.2*(1-small*2.1),5),fcm=c(1,1,1,1,1),fcv=1-2*small))
-loglik(indep(ML_constrained),Htot)
-```
-
-See how the constraint is active: the strength of the training effect
-is almost exactly equal to its maximal value of `small` [as a
-consistency check we can perform the same calculation with a slightly
-different value for `small`:
-
-```{r}
-small <- 2e-4  # cf 1e-4 previously
-ML_constrained <- maxp(Htot,startp=rep(0.2*(1-small*2.1),5),fcm=c(1,1,1,1,1),fcv=1-2*small)
-loglik(indep(ML_constrained),Htot) # cf -29.973 previously
-```
-
-and we can see that this is almost the same].  Thus, the support for
-our alternative hypothesis would be
-
-```{r}
-loglik(indep(ML_unconstrained),Htot)-loglik(indep(ML_constrained),Htot)
-```
-
-thus exceeding Edwards's two units of support.  We have strong
-evidence to suggest that the training was effective.
+thus we have strong evidence for the training being effective.

### 2020-04-05 inst/cook.Rmd
more detail

@@
 \[
-\frac{p_1+t}{p_1+t+p_2+p_3+p_4+p_5}\cdot
+\frac{p_1+S}{p_1+S+p_2+p_3+p_4+p_5}\cdot
 \frac{p_2  }{      p_2+p_3+p_4+p_5}\cdot
@@
 \[
-\frac{p_3  }{p_1+t+p_2+p_3+p_4+p_5}\cdot
-\frac{p_1+t}{p_1+t+p_2+    p_4+p_5}\cdot
+\frac{p_3  }{p_1+S+p_2+p_3+p_4+p_5}\cdot
+\frac{p_1+S}{p_1+S+p_2+    p_4+p_5}\cdot
 \frac{p_5  }{      p_2+    p_4+p_5}\cdot
@@
 
-We can translate these into package idiom with
-function `pwa()` ("person with advantage"):
+We can translate these into package idiom with function `pwa()`
+("person with advantage"):
 
-```{r}
+```{r,usepwafortraining}
 Hafter <- rank_likelihood(1:5)+rank_likelihood(c(3,1,5,2,4))
@@
 
-```{r}
+```{r,addbeforeandafter}
 pnames(Hbefore) <- c(pnames(Hbefore),"S")
-maxp(Hbefore + Hafter)
+maxp(Hbefore+Hafter)
 ```
 
-The above suggesting that the training had a sizable effect; and we
-may test the null that the training had no effect:
+[In the above, note the use of the overloaded "`+`", meaning to add
+two likelihood functions for independent observations].  The evaluate
+suggests that the training had a sizable effect; we may use
+\code{specficp.gt.test()} to test a null of zero strength:
 
-```{r}
-specificp.gt.test(Hbefore + Hafter,"S",0)
+```{r,testnullzerotrainingeffect}
+specificp.gt.test(Hbefore+Hafter, "S", 0)
 ```

### 2020-04-11 inst/cook_repeated.Rmd
repeated cooking observations

---
title: "Repeated incomplete rank observations"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Suppose there are five people in a family.  Each day for five days,
one person cooks a meal.  After the fifth day, each person considers
the four meals cooked by other family members, and puts them in order
of preference.  This process is repeated and here I consider how to
interpret such data.

```{r loadlib}
library(hyper2)
H <- hyper2(pnames=paste("p",1:5,sep=""))
H
```

Now add some data:

```{r data}
H <- H +
	# 3 April 2020:
    rank_likelihood(c(5,4,3,2)) + # person 1 says 5 is the best, 2 the worst.
    rank_likelihood(c(2,5,1,4)) + # person 2 says 1 is the best, 4 the worst.
    rank_likelihood(c(5,1,2,3)) + # person 3 says 1 is the best, 5 the worst.
    rank_likelihood(c(3,5,4,1)) + # person 4 says 1 is the best, 5 the worst.
    rank_likelihood(c(3,2,1,4)) + # person 5 says 2 is the best, 4 the worst.

	# 10 April 2020:
    rank_likelihood(c(5,3,4,1)) + # person 1 says 5 is the best, 4 the worst.
    rank_likelihood(c(2,3,4,1)) + # person 2 says 1 is the best, 4 the worst.
    rank_likelihood(c(3,5,2,1)) + # person 3 says 1 is the best, 5 the worst.
    rank_likelihood(c(2,4,5,3)) + # person 4 says 1 is the best, 5 the worst.
    rank_likelihood(c(2,5,4,1)) + # person 5 says 2 is the best, 4 the worst.

H
```

Note that noone is allowed to vote for themselves.  There are
$5\cdot4!=120$ distinct observations.  We may find maximum likelihood
estimate for the strengths,
$\operatorname{argmax}\mathcal{L}\left(p_1,\ldots,p_5\right)$:

```{r}
(mH <- maxp(H))
pie(mH)
```

### 2020-04-12 inst/jester.Rmd
incorporate the jester dataset

---
title: "The Jester joke recommender system"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Goldberg et al present a dataset in which respondents rated a number
of jokes.  Here, I analyse a small portion of this dataset using the
`hyper2` package.  This document is intended to illustrate an
extremely challenging application of the `hyper2` package and (without
cache) takes a long time to process.  Goldberg's dataset has 24938
lines, one per respondent, and 101 columns, one per joke (the first
column shows the number of jokes rated by each respondent); here I use
150 lines and 99 jokes (the 100th joke was not funny).

```{r }
library("hyper2",quietly=TRUE)
a <- read.csv("jester-data-3.csv",head=FALSE) # File is 150 lines only
a <- as.matrix(a[,-c(1,100)])
a[a==99] <- NA
colnames(a) <- paste("joke",sprintf("%02d",seq_len(ncol(a))),sep="")
a[1,]
```

Row 1 of `a` is displayed and shows that the first respondent rated
joke 5 at -1.65, joke 7 at -0.78, and so on; `NA` means that that
respondent did not rank that joke.  We can perform some summary
statistics of `a`:

```{r}
plot(rowSums(!is.na(a)),xlab="respondent")
```

The above plot shows how many jokes each of the 100 respondents rated.

```{r}
plot(colSums(!is.na(a)),xlab="joke index")
```

The above shows how many respondents rated each joke.  It would make
sense to remove the jokes that were not rated:

```{r}
dim(a)
a <- a[,colSums(!is.na(a))>1]
dim(a)
```

showing that 91 jokes were rated by at least one respondent.  We need
to transform the dataset:

```{r}
f <- function(x){
    x <- x[!is.na(x)]
    x[order(x,decreasing=TRUE)] <- seq_along(x)
    return(x)
}
f(a[1,])
```

Thus we see that this respondent rated `joke08` to be the funniest, having rank 1.

```{r,calculate_jester,cache=TRUE}
jester <- hyper2(pnames=colnames(a))
system.time(for(i in seq_len(nrow(a))){jester %<>% order_obs(f(a[i,]))})
```

now

```{r,cache=TRUE}
system.time(mjest <- maxp(jester,n=1))
```

and

```{r}
mjest
plot(mjest)
```

```{r,cache=TRUE}
equalp.test(H,startp=indep(mjest))
```

## Reference

Eigentaste: A Constant Time Collaborative Filtering Algorithm. Ken
Goldberg, Theresa Roeder, Dhruv Gupta, and Chris Perkins. _Information
Retrieval_, 4(2), 133-151. July 2001.

### 2020-04-13 inst/notthewinner.Rmd
fixes #51

@@
 statistic is unobserved and we know nothing about it except for the
-fact that (wlog) player 1 (aka "me") is not first placed.
+fact that (wlog) player 1 ($p_1$, aka "me") is not first placed.  We
+consider whether there is evidence to support two different assertions:
 
-As in `loser.Rmd`, we have three hypotheses:
+* $H_1\colon p_1=p_2=\ldots=p_n=\frac{1}{n}$
+* $H_2\colon\exists i\neq 1 \mbox{ with } p_1\geq p_i>0$
+
+Our alternative would be that $(p_1,\ldots,p_n) is unconstrained
+except for the unit sum constraint:
 
 * $H_0\colon\sum p_i=1$  (that is, no restriction)
-* $H_1\colon\exists i\neq 1 \mbox{ with } p_1\geq p_i>0$
-* $H_2\colon p_1=p_2=\ldots=p_n=\frac{1}{n}$
 
-(see how $H_1$ implies that someone is worse than me).  Observe that
+(see how $H_2$ implies that someone is worse than me).  Observe that
 $H_1$ and $H_2$ correspond to a restriction on allowable parameter
-space.
-
-To understand the informativeness of ${\mathcal O}$, we will consider
-it in conjunction with some highly informative observations ${\mathcal
-I}$ which are order statistics:
+space.  To understand the informativeness of ${\mathcal O}$, we will
+consider it in conjunction with some highly informative observations
+${\mathcal I}$ which are order statistics:
 
@@
 combine them with ${\mathcal O}$ which we will take to be
-(conditionally) independent.  The first step is to define a suitable
-likelihood function for ${\mathcal I}$:q
+(conditionally) independent.  The first step is estimate the players'
+strengths:
 
@@
 
-Above, `mH0` corresponds to $H_0$.  Note the low but nonzero estimated
-strength of `p1`.
+Above, `mH0` corresponds to an unconstrained optimization.  Note the
+low but nonzero estimated strength of `p_1`.  Observe in passing that
+we may reject $H_z\colon p_1=0$ unconditionally: we can deduce using
+elementary arguments that the likelihood of $H_z$ is zero: if $p_1=0$,
+then $p_4=0$ from the first line [because $p_4$ beat $p_1$ on that
+occasion], and then $p_1=0$ from line 2; then $p_3=0$ from line 3,
+$p_5=0$ from line 4 and finally $p_2=0$ from line 1.  Thus we can say
+for sure that $H_z$ is incorrect and we may safely ignore it
+henceforth.
 
-Observe in passing that we may reject $H_z\colon p_1=0$
-unconditionally: we can deduce using elementary arguments that the
-likelihood of $H_z$ is zero: if $p_1=0$, then $p_4=0$ from the first
-line, and then $p_1=0$ from line 2; then $p_3=0$ from line 3, $p_5=0$
-from line 4 and finally $p_2=0$ from line 1.  Thus we can say for sure
-that $H_z$ is incorrect and we may safely ignore it henceforth.
+## Hypothesis 1: equality
 
-## Hypothesis 1
+```{r}
+equalp.test(H)
+```
 
-Hypothesis 1 is rather tricky so we will use the ersatz $H_{1'}\colon
-p_1\geqslant p_4$ (we choose $p_4$ on the grounds of its low estimated
-value).  Then
+Thus there is no evidence to suggest that the players have different
+strengths: we are unable to reject $H_1$, the hypothesis of equality.
+Hypothesis $H_2$ is rather tricky so we will use the ersatz
+alternative that $p_1\geqslant p_4$ (we choose $p_4$ on the grounds of
+its low estimated value).  Then
 
@@
 loglik(indep(mH1),H)
+loglik(indep(mH0),H)-loglik(indep(mH1),H)   # support
 ```
 
-See how the constraint of $H_1$ is active, for at the evaluate
-$p_1=p_4$ (the numerical value of `small` is immaterial).
-
-## Hypothesis 2
-
-```{r}
-mH2 <- equalp(H)
-loglik(indep(mH2),H)
-```
-
-## Support function
-
-We can present a support function for $H_0,H_1,H_2$ for ${\mathcal
-I}$:
-
-```{r,suppcalculator}
-Supp <- c(H0=loglik(indep(mH0),H), H1=loglik(indep(mH1),H), H2=loglik(indep(mH2),H)
-)
-Supp-max(Supp)
-```
+There is no reason to reject our null that $p_1\geqslant p_4$, for the
+support falls below the two units of support criterion.  Further, see
+how the constraint is active, for at the evaluate $p_1=p_4$ (the
+numerical value of `small` is immaterial).
 
@@
 
-```{r,mls0}
+```{r,mls0,cache=TRUE}
 mSL0 <- maxp(SL)
@@
 
-which is identical to the previous support function.
+which is identical to the previous support function for $H_1$ and
+$H_2$ (apart from a traditional sign) taken from `equalp.test()` and
+`mSL2` above.
 
@@
 
-```{r,mlsa}
+```{r,mlsa,cache=TRUE}
 mSLa0 <- maxp(SLa)

### 2020-04-16 inst/formula1_top9.Rmd
new file, just the top 9 drivers

---
title: "formula1"
output: html_document
---

```{r setup, include=FALSE}
set.seed(0)
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("magrittr")
options("digits" = 5)
options(width = 120)
```

## Formula 1

Formula 1 racing is an important and prestigious motor sport.  Here I
analyse a series of racing results using the `hyper` package which
implements the Plackett-Luce likelihood function.

Consider the following dataset, taken from Wikipedia:

```{r cars}
f2017 <- read.table("formula1_2017.txt",header=TRUE)
f2017a <- f2017[1:9,]
rm(f2017)
f2017a
```

Now process it to give the ranks:

```{r}
f <- function(x){
	noscore <- c("Ret", "WD", "DNS", "DSQ", "DNP", "NC")
	x <- as.character(x)
	x[x %in% noscore] <-  "0"
	x <- as.numeric(x)
	wanted <- x != 0
	xw <- x[wanted]
	xw[order(xw,decreasing=FALSE)] <- seq_along(xw)
	x[wanted] <- xw
	x
}
o  <- apply(f2017a[,seq_len(20)],2,f)
rownames(o) <- rownames(f2017a)
(f2017a <- o)
```

Each row is a driver and each column (after the first) a venue.  We
see that Hamilton, the first row, came second in Australia, first in
China, second in Bahrain, fourth in Russia, and so on.  In the first
column, we see the result from Australia (AUS) in which Hamilton came
second, Vettel first, Bottas third, and so on.  Notation used also
includes six different classes of no-score such as "Ret" for retired,
"WD" for withdrawn, and so on.

We can represent this data in a different form using
`wikitable_to_ranktable()`:

```{r}
wikitable_to_ranktable(f2017a)
```

Looking at the first column ("c1" is read "came first") we see who
came first at each venue: Vettel in Australia (AUS), Hamilton in China
(CHN), Vettel in Bahrain (BHR), and so on.  Now create a likelihood
function `H`:

```{r,calcH,cache=TRUE}
H <- hyper2(pnames=rownames(o))
 for(i in seq_len(ncol(o))){
	H <- H + ordervec2supp(o[,i])
}
head(H)
```
Above, we see a small part of the Placket-Luce likelihood function for
the 2017 results, just the top 9 drivers.  We can find the maximum
likelihood estimator by using the `maxp()` function:

```{r,maxp2017,cache=TRUE}
mL2017 <- maxp(H)
mL2017
dotchart(mL2017,pch=16,main='2017 Formula 1')
pie(mL2017,main='2017 Formula 1')
```

We see Hamilton being the strongest, followed by Vettel and Bottas who
are approximately equal.

# Analysis of nulls:

## Hypothesis: all players have equal strengths

Here we write $H_0\colon p_1=p_2=\cdots =p_n=\frac{1}{n}$.

```{r,equalstrength,cache=TRUE}
equalp.test(H)
```

highly signficant!

## Null of Hamilton being average

We can consider Hamilton and ask various questions about his playing
strength.  One natural null hypothesis would be that Hamilton's
strength is equal to the mean strength, or $\frac{1}{9}$.  (we attempt
to reject the null $H_0$ against the alternative of unconstrained
$p$).  Noting that the evaluate for Hamilton's strength
$\widehat{p_\mathrm{Hamilton}}$ is about 0.28, we ask how much support
is lost, relative to the support at the unconstrained evaluate.

```{r,specifichamiltonstrength,cache=TRUE}
specificp.test(H,"Hamilton")  # default value is 1/9=11%
```

(note that the constrained maximum is on the boundary of the
admissible region with $p_\mathrm{Hamilton}=\frac{1}{9}$).

# Conditions

```{r,loadtable}
(details <- read.table("formula1_conditions.txt",header=TRUE))
drivers <- pnames(formula1)
```

At this point, we introduce a new player: `Ham_wet`, for "Hamilton in
the wet", and we can test the hypothesis that the `Hamilton ==
Ham_wet`.  We will cycle through all the venues and, if it is wet
(according to dataframe `conditions`) we will credit Hamilton's
placing to `Ham_wet`.

```{r,hamiltonwetanddry,cache=TRUE}
details <- read.table("formula1_conditions.txt",header=TRUE)
alldrivers <- c("Ham_dry","Ham_wet",rownames(f2017a)[-1])
(L2017 <- hyper2(pnames=alldrivers)) # uninformative
for(i in seq_len(ncol(f2017a))){
	venue <- colnames(f2017a)[i]
	conditions <- details[which(details$venue == venue),4]
	o <- f2017a[,i]
	names(o) <- rownames(f2017a)
	if(conditions == "wet"){
	   names(o)[names(o) == "Hamilton"] <- "Ham_wet"
	} else if(conditions == "dry"){
	   names(o)[names(o) == "Hamilton"] <- "Ham_dry"
	} else {
	   stop("neither wet nor dry?")
	}
	print(o)
	L2017 %<>% order_obs(o)
}  # for(i) loop closes
L2017
```

We can now test the hypothesis that Hamilton's performance in the wet
is the same as his performance in the dry:

```{r,hamwetdrytest,cache=TRUE}
samep.test(L2017,c("Ham_wet","Ham_dry"))
```
## Try the same thing but with function pwa()

```{r,trypwa,cache=TRUE}

details <- read.table("formula1_conditions.txt",header=TRUE)

H <- hyper2(pnames=c(rownames(f2017a),"S"))
noscore <- c("Ret", "WD", "DNS", "DSQ", "DNP", "NC")

for(i in seq_len(ncol(f2017a))){
	y <- f2017a[,i]
	cond <- details$conditions[i]
        jj <- ordervec2supp(y)
	if(cond == "dry"){
            pnames(jj) <- pnames(H)  # do nothing!
	} else if (cond == "wet"){
            jj <- pwa(jj,"Hamilton")
	} else {
            stop("neither wet nor dry?")
	}  # if wet closes
        H <- H + jj
} # i loop closes
head(H)
maxp(H)
```
Now we can test the hypothesis that Hamilton's wet strength `S` is zero:

```{r,testwetstrengthzero}
pnames(H)
specificp.gt.test(H,"S",0)
```

see this

### 2020-04-17 inst/formula1_points.Rmd
initial stab at points system

---
title: "formula1: the points system"
output: html_document
---

```{r setup, include=FALSE}
set.seed(0)
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("magrittr")
options("digits" = 5)
options(width = 140)
```

## Formula 1: top 9

Formula 1 racing is an important and prestigious motor sport.  Here I
analyse a series of racing results using the `hyper` package which
implements the Plackett-Luce likelihood function.

Consider the following dataset, taken from Wikipedia:

```{r cars}
f2017 <- read.table("formula1_2017.txt",header=TRUE)
no_of_drivers <- 9  # top 9
f2017 <- f2017[seq_len(no_of_drivers),]
f2017
```

Now process it to give the ranks:

```{r}
f <- function(x){
	noscore <- c("Ret", "WD", "DNS", "DSQ", "DNP", "NC")
	x <- as.character(x)
	x[x %in% noscore] <-  "0"
	x <- as.numeric(x)
	wanted <- x != 0
	xw <- x[wanted]
	xw[order(xw,decreasing=FALSE)] <- seq_along(xw)
	x[wanted] <- xw
	x
}
o  <- apply(f2017[,seq_len(20)],2,f)
rownames(o) <- rownames(f2017)
(f2017 <- o)
```

Each row is a driver and each column (after the first) a venue.  We
see that Hamilton, the first row, came second in Australia, first in
China, second in Bahrain, fourth in Russia, and so on.  In the first
column, we see the result from Australia (AUS) in which Hamilton came
second, Vettel first, Bottas third, and so on.  Notation used also
includes six different classes of no-score such as "Ret" for retired,
"WD" for withdrawn, and so on.

Now calculate points:

```{r}
points <- c(25,18,15,12,10,8,6,4,2,1,0,0,0,0,0,0,0,0,0,0)
stopifnot(all(diff(points)<=0))
points <- points/points[1]
jj <- f2017
jj[jj==0] <- length(points)
jj[] <- points[jj]
jj
real_points <- rowSums(jj)
real_points
sort(real_points,decreasing=TRUE)
names(sort(real_points,decreasing=TRUE))
```

Now do some random sampling.  First create a likelihood function:

```{r,calcH,cache=TRUE}
H <- hyper2(pnames=rownames(o))
 for(i in seq_len(ncol(o))){
	H <- H + ordervec2supp(o[,i])
}
head(H)
mL2017 <- maxp(H)
mL2017
dotchart(mL2017,pch=16,main='2017 Formula 1')
```

Now use random sampling:

```{r}
random_table <- rrank(n=ncol(f2017), p=mL2017)
rownames(random_table) <- colnames(f2017)
random_table
```

then see how close it is to reality:

```{r}
maxp(rank_likelihood(random_table)) # should be close to mL2017
```

```{r}
a <- random_table
class(a) <- "matrix"
a[] <- points[a]
a
p <- colSums(a)
p
rando <- order(p,decreasing=TRUE)
rando
```

Thus we see that

```{r}
jj <- sort(p,decreasing=TRUE)
jj
names(jj)
```

```{r,dothework,cache=TRUE}
n <- 200
goodness <- rep(-Inf,n)
synthetic_winner_points     <- rep("",n)
synthetic_winner_likelihood <- rep("",n)
f <- function(x,y){sum(cumprod(x==y))}
for(i in seq_len(n)){
	a <- rrank(n=ncol(f2017), p=mL2017)
	jj <- a
	class(jj) <- "matrix"
	jj[] <- points[jj]
	p <- colSums(jj)  # synthetic points
	names(p) <- names(mL2017)
	synthetic_rank_points <- names(sort(p,decreasing=TRUE))
	synthetic_winner_points[i] <- synthetic_rank_points[1]

	mH <- maxp(rank_likelihood(a))
	synthetic_rank_likelihood <- names(sort(mH,decreasing=TRUE))
	synthetic_winner_likelihood[i] <- synthetic_rank_likelihood[1]

	goodness[i] <- f(synthetic_rank_likelihood,synthetic_rank_points)
}
plot(table(goodness))
table(synthetic_winner_likelihood,synthetic_winner_points)
```

So from the table we see about half the times the winner was correctly
predicted by the points system.

### 2020-04-17 inst/cook_repeated.Rmd
more data

@@
 
-H
+	# 17 April 2020:
+    rank_likelihood(c(2,4,3,5)) + # person 1 says 5 is the best, 4 the worst.
+    rank_likelihood(c(3,5,4,1)) + # person 2 says 1 is the best, 4 the worst.
+    rank_likelihood(c(3,2,4,1)) + # person 3 says 1 is the best, 5 the worst.
+    rank_likelihood(c(2,5,4  )) + # person 4 says 1 is the best, 5 the worst.
+    rank_likelihood(c(5,3,1  ))   # person 5 says 2 is the best, 4 the worst.
+
 ```

### 2020-04-18 inst/formula1.Rmd
more work

@@
 
-see this
+now profile likelihood:
+
+```{r,doproflikn,cache=TRUE}
+H9
+maxp(H9)
+p <- seq(from=0.11,to=0.13,len=50)
+u <- profsupp(H9,"Hamilton",p,n=14)
+plot(p,u)
+```
+
+# Pole position
+
+We now introduce another reified player, `pole`, whose strength helps
+the driver in pole position.
+
+```{r,polepositionlikelihood,cache=TRUE}
+HP <- hyper2(pnames=c(rownames(f2017),"pole"))
+noscore <- c("Ret", "WD", "DNS", "DSQ", "DNP", "NC")
+
+for(i in seq_len(ncol(f2017))){
+	y <- f2017[,i]
+	if(any(y %in% noscore)){y[y%in%noscore] <- 0}
+	y <- as.numeric(levels(y)[y])
+	names(y) <- rownames(f2017)
+	HP <- HP + pwa(ordervec2supp(y),details$pole[i],"pole")
+} # i loop closes
+head(HP)
+(mHP <- maxp(HP))
+pie(mHP)
+```
+
+test Hamilton == bottas:
+
+```{r,dosometests,cache=TRUE}
+samep.test(HP,c("Hamilton","Vettel"))
+samep.test(HP,c("Hamilton","Bottas"))
+```
+
+Now profile likelihood for pole strength:
+
+```{r,profsupcalc,cache=TRUE}
+p <- seq(from=0.5,to=0.96,len=20)
+pnames(HP)
+u <- profsupp(HP,"Ricciardo",p,n=1)
+u
+plot(p,u-max(u))
+```

### 2020-04-18 inst/formula1_points.Rmd
more work

@@
 dotchart(mL2017,pch=16,main='2017 Formula 1')
+pie(mL2017)
 ```

### 2020-04-18 inst/formula1_top9.Rmd
change title
more work

@@
 ---
-title: "formula1"
+title: "formula1: top 9 drivers"
 output: html_document
@@
 
-## Formula 1
+## Formula 1: top 9
+
+(this document is largely the same as `formula1.Rmd` but considers
+only the top 9 drivers).
 
@@
 f2017 <- read.table("formula1_2017.txt",header=TRUE)
-f2017a <- f2017[1:9,]
-rm(f2017)
-f2017a
+n <- 9  # top 9
+f2017 <- f2017[seq_len(n),]
+f2017
 ```
@@
 }
-o  <- apply(f2017a[,seq_len(20)],2,f)
-rownames(o) <- rownames(f2017a)
-(f2017a <- o)
+o  <- apply(f2017[,seq_len(20)],2,f)
+rownames(o) <- rownames(f2017)
+(f2017 <- o)
 ```
@@
 ```{r}
-wikitable_to_ranktable(f2017a)
+wikitable_to_ranktable(f2017)
 ```
@@
 details <- read.table("formula1_conditions.txt",header=TRUE)
-alldrivers <- c("Ham_dry","Ham_wet",rownames(f2017a)[-1])
+alldrivers <- c("Ham_dry","Ham_wet",rownames(f2017)[-1])
 (L2017 <- hyper2(pnames=alldrivers)) # uninformative
-for(i in seq_len(ncol(f2017a))){
-	venue <- colnames(f2017a)[i]
+for(i in seq_len(ncol(f2017))){
+	venue <- colnames(f2017)[i]
 	conditions <- details[which(details$venue == venue),4]
-	o <- f2017a[,i]
-	names(o) <- rownames(f2017a)
+	o <- f2017[,i]
+	names(o) <- rownames(f2017)
 	if(conditions == "wet"){
@@
 
-H <- hyper2(pnames=c(rownames(f2017a),"S"))
+H <- hyper2(pnames=c(rownames(f2017),"S"))
 noscore <- c("Ret", "WD", "DNS", "DSQ", "DNP", "NC")
 
-for(i in seq_len(ncol(f2017a))){
-	y <- f2017a[,i]
+for(i in seq_len(ncol(f2017))){
+	y <- f2017[,i]
 	cond <- details$conditions[i]

### 2020-04-19 data/eurovision.R
make eurovision hyper2 pnames be character, not factor

@@ -1,162 +1,160 @@
 `eurovision2009` <-
-structure(list(brackets = list(1, c(1, 2, 3, 4, 5, 6, 7, 8, 9, 
-10, 11, 12, 13, 14, 15, 16, 17), c(1, 2, 3, 4, 5, 6, 7, 8, 9, 
-10, 11, 12, 13, 14, 15, 16, 17, 18), c(1, 2, 3, 4, 5, 6, 7, 8, 
-9, 10, 11, 12, 13, 14, 15, 16, 18), c(1, 2, 3, 4, 5, 6, 7, 8, 
-9, 10, 11, 12, 13, 14, 15, 17, 18), c(1, 2, 3, 4, 5, 6, 7, 8, 
-9, 10, 11, 12, 13, 14, 16, 17, 18), c(1, 2, 3, 4, 5, 6, 7, 8, 
-9, 10, 11, 12, 13, 15, 16, 17, 18), c(1, 2, 3, 4, 5, 6, 7, 8, 
-9, 10, 11, 12, 14, 15, 16, 17, 18), c(1, 2, 3, 4, 5, 6, 7, 8, 
-9, 10, 11, 13, 14, 15, 16, 17, 18), c(1, 2, 3, 4, 5, 6, 7, 8, 
-9, 10, 11, 13, 14, 15, 16, 18), c(1, 2, 3, 4, 5, 6, 7, 8, 9, 
-10, 11, 13, 14, 15, 17, 18), c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 
-11, 13, 14, 16, 17, 18), c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 
-13, 15, 17, 18), c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 12, 13, 14, 
-15, 16, 17, 18), c(1, 2, 3, 4, 5, 6, 7, 8, 9, 11, 12, 13, 14, 
-15, 16, 17, 18), c(1, 2, 3, 4, 5, 6, 7, 8, 9, 11, 13, 14, 15, 
-16, 17, 18), c(1, 2, 3, 4, 5, 6, 7, 8, 10, 11, 12, 13, 14, 15, 
-16, 17), c(1, 2, 3, 4, 5, 6, 7, 8, 10, 11, 12, 13, 14, 15, 16, 
[…]

### 2020-04-19 inst/table_tennis_serve.Rmd
move to inst/ (too big)

@@
 ---
-output:
-  pdf_document: default
-  html_document: default
+title: "The importance of serve in table tennis"
+author: "Robin K. S. Hankin"
+date: "`r Sys.Date()`"
+vignette: >
+  %\VignetteIndexEntry{table_tennis_serve}
+  %\VignetteEngine{knitr::rmarkdown}
+  %\VignetteEncoding{UTF-8}
 ---
+
 # The importance of serve in table tennis.
 
-Suppose We have three players, "a","b", and "c" who play table tennis.
+(In this vignette, object `H` is created.  It is identical to
+object `table_tennis` in the package; to see it type
+`data(table_tennis)` at the command line, and to see the help
+page, type `help(table_tennis)`).
+
+Suppose we have three players, "A","B", and "C" who play table tennis.
 We seek to quantify the effect of the serve.  The scoreline is as
@@
 
-(note that A does not play C).  This dataset is easily analysed with
-the ```hyper2``` package.
+(note that B does not play C).  This dataset is easily analysed with
+the `hyper2` package although the model implicitly assumes that the
+serve strength does not depend on who is serving.  The first step is
+to define a `hyper2` object:
 
-```{r}
+```{r,message=FALSE}
 library("hyper2",quietly=TRUE)
+(H <- hyper2(pnames=c("S","a","b","c")))
 ```
 
-The first step is to define a ```hyper2``` object:
-
-```{r}
-H <- hyper2(pnames=c("S","a","b","c"))
-```
-
-Now we have to input the data:
+so the likelihood function is uniform, for the only term is raised to
+the zeroth power (players' names are not capitalised in R idiom for
+clarity).  Now we have to input the data, and we will go through the
+four scorelines above, one by one.  Note that this approach implicitly
+assumes independence of scores.
 
@@
 H[c("a","b","S")] %<>% "-"(5+1) # a+b+S play 1+5=6 games
+H
 ```
+
+The likelihood function is of the form
+$\frac{\left(a+S\right)^5b}{\left(a+b+S\right)^6}$.  We can add the
+other observations to $H$:
 
@@
 
-So H is the likelihood function for three players and a serve.
+## Likelihood function
 
-# Results 
+For the entire dataset we have
+
+```{r}
+H
+```
+
+Here $H$ is a likelihood function on $a,b,c,S$, algebraically
+proportional to
+
+\[
+\frac{(S+a)^9(S+b)^3(S+c)^2a^2bc}{(S+a+b)^{10}(S+a+c)^8}
+\]
+
+# Maximum likelihood estimate for the strengths
+
 We now find the unconstrained MLE:
@@
 ```{r}
-maxlike_free <- loglik(H,indep(p1))
+maxlike_free <- loglik(indep(p1),H)
 maxlike_free
@@
 
-Now find the MLE conditional on the serve having zero strength:
+# Test hypothesis that serve strength S is zero
+
+To do this we follow the protocol in the `icons` vignette.  First find
+the MLE conditional on the serve having zero strength:
 
 ```{r}
-small <- 1e-6
+small <- 1e-3
 maxlike_constrained <-
-    constrOptim(
-        f = function(p){loglik(H,p)},
-        theta=c(small, 1/3-small/2, 1/3-small/2),
-        grad=NULL,
-        ui=rbind(diag(3),-diag(3),-1),
-        ci=c(rep(small/2,3),c(-small*2,-1,-1),-1),
-        control=list(fnscale = -1)  # finds max, not min
+    maxp(
+        H,
+        startp=c(small/2, 1/3-small/2, 1/3-small/2),
+        fcm=c(-1,0,0),  fcv=-small,give=TRUE
 )
+
+maxlike_constrained
 ```
@@
 
-The difference in support, ```delta_support```, is about 2.62 which
+The difference in support, `delta_support`, is about 2.62 which
 exceeds Edwards's two-units-of-support criterion.  We can reject the
@@
 Thus the tail area, pval, is about 0.022 which gives us a significant
-result.
+result: the serve has nonzero strength.
 
-Further, note that our maximum likelihood estimate for strengths
-allows us to assess what might happen when A plays C (which was not
-observed).  For convenience, here is the evaluate:
+## Assessment of equal player strength.
+
+We now assess the hypothesis that the three (real) players are of zero
+strength:  $H_E\colon p_A=p_B=p_C$.
+
+## Assessment of B vs C
+
+The maximum likelihood estimate for strengths allows us to assess what
+might happen when B plays C (which was not observed).  For
+convenience, here is the evaluate:
 
@@
 
-If A serves, we would have $A+S:C\simeq 0.24+0.45:0.16\equiv 69:16$,
-or A wins with probability of about $0.81$.  If C serves we have
-$A:C+S\simeq 0.24:0.16+0.45\equiv 24:61$, so A wins with a probability
-of about $0.28$.
+If B serves, we would have $B+S:C\simeq 0.14+0.45:0.15\equiv 59:15$,
+or B wins with probability of about $0.78$.  If C serves we have
+$B:C+S\simeq 0.14:0.16+0.45\equiv 14:61$, so B wins with a probability
+of about $0.19$.
+
+# Profile likelihood for serve strength
+
+We can give a profile likelihood for $S$ by maximizing the likelihood
+of $(S,p_a,p_b,p_c)$ subject to a particular value of $S$ (in addition
+to the unit sum constraint).  The following function defines a profile
+likelihood function for $S$:
+
+```{r}
+profile_likelihood <- function(S,give=FALSE){
+    small <- 1e-2  # cannot make this too small, optimization routine needs wiggle room
+    out <- maxp(
+       H, startp=c(S+small/2 , small,small),
+       give=TRUE,
+       fcm=rbind(c(1,0,0),c(-1,0,0)),fcv=c(S,-S-small)  # S <= p[1] <= S+small
+       )
+    if(give){
+       return(out)
+      } else {
+       return(out$value)
+      }
+}
+```
+
+We can then evaluate the profile likelihood for a range of possible
+values of $S$:
+
+```{r}
+S_vals <- seq(from=0.02,to=0.85,len=30)  # possible values for S
+prof_like <- sapply(S_vals,profile_likelihood)  
+plot(S_vals, prof_like-max(prof_like),xlab="serve strength",ylab="profile likelihood")
+abline(h= -2)
+```
+
+So we can see that the credible range is from about 0.07-0.8 (and in
+particular excludes zero).
+
+# Profile likelihood for B vs C
+
+The profile likelihood technique can be applied to the question of the
+strengths of players B and C, even though no competition between the
+players was observed:
+
+```{r,dpi=72}
+proflike <- function(bc,give=FALSE){
+  B <- bc[1]
+  C <- bc[2]
+  if(B+C>=1){return(NA)}
+
+  objective <- function(S){     #
+
+    jj <- c(S,A=1-(S+B+C),B,C)  # B,C fixed
+    loglik(indep(jj),H)         # no minus: we use maximum=T in optimize()
+  }
+
+  maxlike_constrained <-  # single DoF is S [B,C given, A is fillup]
+    optimize(             # single DoF -> use optimize() not optim()
+        f = objective,         
+        interval = c(0,1-(B+C)),
+        maximum = TRUE
+    )
+
+  if(give){
+    return(maxlike_constrained)
+  } else{
+    return(maxlike_constrained$objective)
+  }
+}
+
+small <- 1e-5
+n <- 50
+p <- seq(from=small,to=1-small,length=n)
+jj <- expand.grid(p,p)
+support <- apply(jj,1,proflike)
+support <- support-max(support,na.rm=TRUE)
+supportx <- support
+dim(supportx) <- c(n,n)
+
+x <- jj[,1]+jj[,2]/2
+y <- jj[,2]*sqrt(3)/2
+plot(x,y,cex=(support+6)/12, 
+   pch=16,asp=1,xlim=c(-0.1,1.1),ylim=c(-0.2,0.9),
+   axes=FALSE,xlab="",ylab="",main="profile likelihood for B,C")
+polygon(x=c(0,1/2,1),y=c(0,sqrt(3)/2,0))
+text(0.07,-0.04,'A+S=1, B=0, C=0')
+text(0.50,+0.90,'A+S=0, B=1, C=0')
+text(0.93,-0.04,'A+S=0, B=0, C=1')
+```
+
+But what one is really interested in is the relative strength of
+player B and player C.  We are indifferent between $p_B=0.1,p_C=0.2$
+and $p_B=0.3,p_C=0.6$.  To this end we need the profile likelihood
+curve of the log contrast, $X=\log\left(p_B/p_C\right)$.
+
+```{r}
+logcontrast <- apply(jj,1,function(x){log(x[1]/x[2])})
+plot(logcontrast,support,xlim=c(-5,5),ylim=c(-4,0))
+abline(h=-2)
+```
+
+From the graph, we can see that the two-units-of-support interval is
+from about $-3$ to $+3$ which would translate into a probability of B
+winning against C being in the interval
+$\left(\frac{1}{1+e^{3}},\frac{1}{1+e^{-3}}\right)$, or about
+$(0.047,0.953)$.  However, it is not clear if $p_B$ and $p_C$ are
+meaningful in isolation, as a real table tennis point has to have a
+server and the value of $S$ is itself uncertain.

### 2020-04-20 R/hyper2.R
consistentify gradient() and hessian()

@@ -347,7 +347,7 @@ setGeneric("pnames<-",function(x,value){standardGeneric("pnames<-")})
     return(hyper2(out[[1]],out[[2]],pnames=pnames(x)))
 }
 
-`gradient` <- function(probs,H){
+`gradient` <- function(H,probs=indep(maxp(H))){
   stopifnot(length(probs) == size(H)-1)
   differentiate(brackets(H), powers(H), fillup(probs), size(H))$grad_comp
 }
@@ -425,7 +425,7 @@ setGeneric("pnames<-",function(x,value){standardGeneric("pnames<-")})
         }
 
     objective <- function(p){ -loglik(p,H) }
-    gradfun   <- function(p){ -(gradient(p,H))} #NB minus signs
+    gradfun   <- function(p){ -(gradient(H,p))} #NB minus signs
     
     UI <- rbind(
         diag(nrow=n-1),  # regular: p1 >=0, p2 >= 0, ..., p_{n-1} >= 0

### 2020-04-25 inst/cook_repeated.Rmd
new data

@@
     rank_likelihood(c(2,5,4  )) + # person 4 says 1 is the best, 5 the worst.
-    rank_likelihood(c(5,3,1  ))   # person 5 says 2 is the best, 4 the worst.
+    rank_likelihood(c(5,3,1  )) + # person 5 says 2 is the best, 4 the worst.
+
+	# 25 April 2020
+    rank_likelihood(c(1,5,2,3)) + # person 1 says 5 is the best, 4 the worst.
+    rank_likelihood(c(3,2,1,4)) + # person 2 says 1 is the best, 4 the worst.
+    rank_likelihood(c(1,5,3,4)) + # person 3 says 1 is the best, 5 the worst.
+    rank_likelihood(c(5,2,1,4)) + # person 4 says 1 is the best, 5 the worst.
+    rank_likelihood(c(2,3,4,5)) + # person 5 says 2 is the best, 4 the worst.
 

### 2020-05-03 inst/cook_repeated.Rmd
typo
consistentify and improve cook_repeated.Rmd

--- 2020-04-11 inst/cook_repeated.Rmd
@@
 ---
-title: "Repeated incomplete rank observations"
+title: "Judging domestic cooking ability: repeated incomplete rank observations"
 output: html_document
@@
 
-```{r loadlib}
-library(hyper2)
-H <- hyper2(pnames=paste("p",1:5,sep=""))
+```{r loadlib,echo=FALSE}
+library(hyper2,quietly=TRUE)
+```
+
+```{r data}
+cooking_table <- as.matrix(read.table("cooking_weekly.txt", header=TRUE))
+head(cooking_table)
+```
+Thus from the first row we find that one judge considered Lesley to be
+fourth best, Alice third, Zac second and Annabel best (this must have
+been Robin: no-one is allowed to vote for themselves and a zero means
+no vote).  This is the transpose of an order table such as
+`skating.txt`.  In this context a zero contains no information (unlike
+the formula 1 tables in which zeros signify did not finish).  Document
+`curling.Rmd` contains a discussion and here we will use the same
+method.  First we will consider the support function for the first row only:
+
+### First row of the dataset
+
+```{r}
+jj <- cooking_table[1,]
+H_row1 <- order_obs(hyper2(pnames=colnames(cooking_table)),jj[jj>0])
+H_row1
+```
+
+In the above, we calculate the Plackett-Luce likelihood function;
+observe that it is uninformative about competitor number 1, which I
+show in two ways.  Firstly:
+
+```{r}
+p1 <- rdirichlet(1,dirichlet(rep(10,5)))
+# Now create p2 which is the same but competitor 1 is stronger:
+p2 <- p1
+p2[1] <- p2[1] + 0.1     # person 1 becomes stronger...
+p2[-1] <- p2[-1] - 0.1/4 # ...and everyone else becomes weaker
+
+loglik(indep(p1),H_row1)
+loglik(indep(p2),H_row1)
+```
+
+but more generally:
+
+```{r}
+(p <- rdirichlet(1,dirichlet(rep(1,5))))
+evaluate(brackets(H_row1), powers(H_row1), probs=p)
+evaluate(brackets(H_row1), powers(H_row1), probs=p*4)
+```
+
+## Complete dataset
+
+It is easy to iterate over the rows of the dataset:
+
+```{r}
+H <- hyper2(pnames=colnames(cooking_table))
+
+for(i in seq_len(nrow(cooking_table))){  # iterate over rows
+    x <- cooking_table[i,] 
+    names(x) <- colnames(cooking_table)
+    H %<>% order_obs(x[x>0])
+}
 H
@@
 
-Now add some data:
-
-```{r data}
-H <- H +
-	# 3 April 2020:
-    rank_likelihood(c(5,4,3,2)) + # person 1 says 5 is the best, 2 the worst.
-    rank_likelihood(c(2,5,1,4)) + # person 2 says 1 is the best, 4 the worst.
-    rank_likelihood(c(5,1,2,3)) + # person 3 says 1 is the best, 5 the worst.
-    rank_likelihood(c(3,5,4,1)) + # person 4 says 1 is the best, 5 the worst.
-    rank_likelihood(c(3,2,1,4)) + # person 5 says 2 is the best, 4 the worst.
-
-	# 10 April 2020:
-    rank_likelihood(c(5,3,4,1)) + # person 1 says 5 is the best, 4 the worst.
-    rank_likelihood(c(2,3,4,1)) + # person 2 says 1 is the best, 4 the worst.
-    rank_likelihood(c(3,5,2,1)) + # person 3 says 1 is the best, 5 the worst.
-    rank_likelihood(c(2,4,5,3)) + # person 4 says 1 is the best, 5 the worst.
-    rank_likelihood(c(2,5,4,1)) + # person 5 says 2 is the best, 4 the worst.
-
-H
-```
-
-Note that noone is allowed to vote for themselves.  There are
-$5\cdot4!=120$ distinct observations.  We may find maximum likelihood
-estimate for the strengths,
-$\operatorname{argmax}\mathcal{L}\left(p_1,\ldots,p_5\right)$:
+(compare file `curling.Rmd` in which we iterate over columns not rows).
+Standard techniques may be used:
 
 ```{r}
-(mH <- maxp(H))
-pie(mH)
+pie(maxp(H))
+equalp.test(H)
 ```

### 2020-05-04 inst/gradient.Rmd
gradientn vignette

---
title: "Numerical verification of function `gradientn()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("hyper2",quietly=TRUE)
```

We will numerically verify the `gradientn()` function for two `hyper2`
objects.

## Chess dataset

The chess dataset is as follows:

```{r}
chess
```

We will define two gradient functions: `chess_gradient_theoretical()`,
which calculates the partial derivatives using straightforward algebra
(specifically elementary calculus, $\partial(x+y)^n/\partial
x=n(x+y)^{n-1}$), and `chess_gradient_theoretical()`, which uses `C` code:

```{r}
chess_gradient_theoretical <- function(x){
    Topalov <- x[1]
    Anand   <- x[2]
    Karpov  <- x[3]

    out <- c(
        Topalov =  30/Topalov -35/(Topalov+Anand) -18/(Topalov+Karpov),
        Anand   = -35/(Topalov+Anand) +36/Anand -35/(Anand+Karpov),
        Karpov  = -18/(Topalov+Karpov) -35/(Anand+Karpov) +22/Karpov
    )

    names(out) <- pnames(chess)
    out

}
chess_gradient_numerical  <- function(x){gradientn(chess,x)}
```

So in the above we have two ways of calculating the gardient: one
theoretical and one numerical.  Do they agree?

```{r}
p_chess <- 
	c(
		Topalov = 0.5,
		Anand   = 0.3,
		Karpov  = 0.2
	)
chess_gradient_theoretical(p_chess) - chess_gradient_numerical(p_chess)
```

Yes, they agree

# Icons

The `icons` dataset is:

```{r}
icons
```

We can do the same thing but the algebra is more involved:

```{r}
icons_gradient_theoretical <- function(x){
    NB <- x[1]
    L <- x[2]
    PB <- x[3]
    THC <- x[4]
    OA <- x[5]
    WAIS <- x[6]

	out <- c(
		NB = (
			+32/NB
            -20/(NB + L + THC + OA)
            -15/(NB + L + THC + WAIS)
            -09/(NB + L + OA + WAIS)
            -18/(NB + PB + THC + OA)
            -18/(NB + PB + THC + WAIS)
            -08/(NB + PB + OA + WAIS)
        ),
        L = (
            -20/(NB + L + THC + OA)
            -15/(NB + L + THC + WAIS)
            -09/(NB + L + OA + WAIS)
            +24/L
            -11/(L + PB + THC + OA)
            -16/(L + PB + THC + WAIS)
            -18/(L + PB + OA + WAIS)
        ),
        PB = (
            -18/(NB + PB + THC + OA)
            -18/(NB + PB + THC + WAIS)
            -08/(NB + PB + OA + WAIS)
            -11/(L + PB + THC + OA)
            -16/(L + PB + THC + WAIS)
            -18/(L + PB + OA + WAIS)
            +30/PB
        ),
        THC = (
            -20/(NB + L + THC + OA)
            -15/(NB + L + THC + WAIS)
            -18/(NB + PB + THC + OA)
            -18/(NB + PB + THC + WAIS)
            -11/(L + PB + THC + OA)
            -16/(L + PB + THC + WAIS)
            +24/THC
        ),
        OA = (
            -20/(NB + L + THC + OA)
            -09/(NB + L + OA + WAIS)
            -18/(NB + PB + THC + OA)
            -08/(NB + PB + OA + WAIS)
            -11/(L + PB + THC + OA)
            -18/(L + PB + OA + WAIS)
            +14/OA
        ),
        WAIS = (
            -15/(NB + L + THC + WAIS)
            -09/(NB + L + OA + WAIS)
            -18/(NB + PB + THC + WAIS)
            -08/(NB + PB + OA + WAIS)
            -16/(L + PB + THC + WAIS)
            -18/(L + PB + OA + WAIS)
            +9/WAIS
        )
    )
	names(out) <- pnames(icons)
	return(out)
}

icons_gradient_numerical  <- function(x){gradientn(icons,x)}
```

Then

```{r}
p_icons <- c(NB=0.3, L=0.1, PB=0.2, THC=0.15, OA=0.15, WAIS=0.1)
```

```{r}
icons_gradient_theoretical(p_icons)
icons_gradient_theoretical(p_icons) - icons_gradient_numerical(p_icons)
```

also agreeing.

## Difference between `gradientn()` and `gradient()`

Taking the chess dataset we have

```{r}
rbind(indep(p_chess), gradient(chess,indep(p_chess)))
rbind(p_chess,gradientn(chess,p_chess))
```

Indeed, because the powers of `chess` sum to zero, we can test the
fact that the derivative in the direction parallel to `p_chess` is
zero:

```{r}
sum(powers(chess))
sum(p_chess*gradientn(chess,p_chess))
```

which is correct to numerical precision.  Further,

```{r}
sum(powers(icons))
sum(p_icons*gradientn(icons,p_icons))
```

### 2020-05-18 inst/formula1.Rmd
load vars here

@@
 ```{r,polepositionlikelihood,cache=TRUE}
+f2017 <- read.table("formula1_2017.txt",header=TRUE)
+details <- read.table("formula1_conditions.txt",header=TRUE)
 HP <- hyper2(pnames=c(rownames(f2017),"pole"))
@@
 
-for(i in seq_len(ncol(f2017))){
+for(i in seq_len(ncol(f2017)-1)){
 	y <- f2017[,i]

### 2020-05-18 inst/formula1.Rmd
slight changes, now works

@@
 	if(any(y %in% noscore)){y[y%in%noscore] <- 0}
-	y <- as.numeric(levels(y)[y])
+	y <- as.numeric(y)
 	names(y) <- rownames(f2017)
@@
 pie(mHP)
+loglik(indep(mHP),HP)
 ```
@@
 ```{r,profsupcalc,cache=TRUE}
-p <- seq(from=0.5,to=0.96,len=20)
-pnames(HP)
-u <- profsupp(HP,"Ricciardo",p,n=1)
+p <- seq(from=0.15,to=0.25,len=10)
+u <- profsupp(HP,"pole",p,n=27)
 u

### 2020-05-20 inst/formula1.Rmd
bugfix

@@
 ```
-(the p-value seems to be too low to be believable).
 
@@
 f2017 <- read.table("formula1_2017.txt",header=TRUE)
-f2017 <- f2017[,-ncol(f2017)]
 details <- read.table("formula1_conditions.txt",header=TRUE)
@@
 
-for(i in seq_len(ncol(f2017))){
+for(i in seq_len(ncol(f2017)-1)){
 	y <- f2017[,i]
 	if(any(y %in% noscore)){y[y%in%noscore] <- 0}
-	y <- as.numeric(levels(y)[y])
+	y <- as.numeric(y)
 	names(y) <- rownames(f2017)

### 2020-05-20 inst/formula1.Rmd
slight tweaks and use of better idiom

@@
 (note that the constrained maximum is on the boundary of the
-admissible region with $p_\mathrm{Hamilton}\approxequal\frac{1}{25}$).
+admissible region with $p_\mathrm{Hamilton}\approx\frac{1}{25}$).
 
-## Null of Hamilton being better than other players
+## Null of Hamilton being better than all other players
 
@@
 ```{r,L4like,cache=TRUE}
-small <- 0.01
-L4 <- maxp(formula1, startp=c(small, rep((1-small)/24,23)),fcm=rbind(c(-1,1,rep(0,22))),fcv=0,give=TRUE)
-L4
-```
-
-Calculating the support and asymptotic p-values:
-```{r,l4val}
-#L2 - L4$value 
-#pchisq(2*(L2 - L4$value),df=1,lower.tail=FALSE)
-2+3
+samep.test(formula1,c("Hamilton","Bottas"))
 ```
@@
 pie(mHP)
-loglik(indep(mHP),HP)
+mHP <- mHP[mHP > 1e-2]
+mHP
+pie(mHP)
 ```
@@
 p <- seq(from=0.15,to=0.25,len=10)
-u <- profsupp(HP,"pole",p,n=27)
-u
-plot(p,u-max(u))
+u07 <- profsupp(HP,"pole",p,n=7)
+u07
+plot(p,u07-max(u07))
+u27 <- profsupp(HP,"pole",p,n=27)
+u27
+plot(p,u27-max(u27))
 ```

### 2020-05-21 inst/formula1.Rmd
give proflike() a 'relative' argument

@@
 p <- seq(from=0.11,to=0.13,len=50)
-u <- profsupp(H9,"Hamilton",p,n=14)
-plot(p,u)
+u01 <- profsupp(H9,"Hamilton",p,n=01)
+plot(p,u01)
+u20 <- profsupp(H9,"Hamilton",p,n=20)
+plot(p,u20)
+u50 <- profsupp(H9,"Hamilton",p,n=50)
+plot(p,u50)
 ```

### 2020-05-21 R/profile.R

@@ -1,5 +1,7 @@
-`profsupp` <- function(H, i, p, ...){
-    sapply(p,function(x){profile_support_single(H=H, i=i, p=x, evaluate=FALSE, ...)})
+`profsupp` <- function(H, i, p, relative=TRUE, ...){
+    out <- sapply(p,function(x){profile_support_single(H=H, i=i, p=x, evaluate=FALSE, ...)})
+    if(relative){out <- out-max(out)}
+    return(out)
 }
     
 `profile_support_single` <- function(H, i, p, evaluate=FALSE, ...){

### 2020-05-21 inst/formula1.Rmd
plot all together

--- 2020-05-20 inst/formula1.Rmd
@@
 p <- seq(from=0.11,to=0.13,len=50)
-u <- profsupp(H9,"Hamilton",p,n=14)
-plot(p,u)
+u01 <- profsupp(H9,"Hamilton",p,n=01)
+plot(p,u01)
+u20 <- profsupp(H9,"Hamilton",p,n=20)
+plot(p,u20)
+u50 <- profsupp(H9,"Hamilton",p,n=50)
+plot(p,u50)
+uu <- pmax(u01,u20,u50)
+plot(p,uu)
 ```

### 2020-05-22 inst/cook_repeated.Rmd
more work

@@
 cooking_table <- as.matrix(read.table("cooking_weekly.txt", header=TRUE))
-head(cooking_table)
+cooking_table
 ```
@@
 ```{r}
-pie(maxp(H))
 equalp.test(H)
+mH <- maxp(H)
+mH
+pie(mH)
 ```
+
+# Preferential voting
+
+It seems that the boys tend to prefer the boys' cooking and the girls
+prefer the girls' cooking.  But simple things first:
+
+```{r}
+a <- cooking_table # saves typing
+a <- a[a[,4]>0,]  # just lines where zac does not vote
+jj <- table(robin_votes=a[,1]==0,zac_last=a[,4]==4)
+jj
+```
+
+The table above shows that Robin cast a total of `r sum(jj[2,])` votes
+for Zac, and everyone else combined cast a total of `r sum(jj[1,])`
+votes for Zac.  Everyone else voted Zac last a total of `r jj[1,2]`
+times and higher than last `r jj[1,1]` times:
+
+```{r}
+fisher.test(jj,alternative="less")
+```
+
+So it looks like Robin is less likely to vote Zac last than everyone
+else.  Does he do likewise for me?
+
+```{r}
+a <- cooking_table # saves typing
+a <- a[a[,1]>0,]  # just lines where robin does not vote
+jj <- table(zac_votes=a[,4]==0,robin_last=a[,1]==4)
+jj
+fisher.test(jj,alternative="less")
+```
+
+No evidence for that.
+
+We can do slightly better, and test a more general hypothesis,
+specifically that the girls systematically vote other girls higher,
+and boys systematically vote boys higher, than warranted.  To test
+this hypothesis, we posit two reified entities, `girl` and `boy`.
+Entity `girl` helps a girl when another girl is voting for her, and
+entity `boy` helps a boy when another boy is voting for him.
+
+```{r}
+cooking_table <-
+	as.matrix(read.table("cooking_weekly.txt", header=TRUE))
+cooking_table <- cooking_table[rowSums(cooking_table==0)==1,]
+boys <-c("robin","zac")
+girls <- c("lesley","alice","annabel")
+cooks <- colnames(cooking_table)
+P <- hyper2(pnames=c(cooks,"boy","girl"))
+P0 <- hyper2(pnames=cooks)
+for(i in seq_len(nrow(cooking_table))){  # iterate over rows
+    x <- cooking_table[i,] 
+	print(x)
+	if(sum(x==0)>1){break}
+    names(x) <- cooks
+	if(names(x[x==0]) %in% boys){  # if the voter is a boy
+		x <- x[x>0]  # remove the voter
+		Pjj <- order_obs(P0,x) # order statistic
+		Pjj %<>% pwa(names(x)[names(x) %in% boys], chameleon="boy")
+		pnames(Pjj) <- c(cooks,"boy","girl")
+	} else if (names(x[x==0]) %in% girls){ # voter a girl
+		x <- x[x>0]
+		Pjj <- order_obs(P0,x)
+		Pjj %<>% pwa(names(x)[names(x) %in% girls], chameleon="girl")
+		pnames(Pjj) <- c(cooks,"boy","girl")
+	} else {
+		stop()
+	}
+    P <- P + Pjj
+}
+x
+Pjj
+P
+maxp(P)
+```

### 2020-05-25 inst/formula1.Rmd
work on Zipf's law

@@
 and Button, whose estimated strength is zero).
+
+# Zipf's law
+
+Do the players follow Zipf's law?
+
+```{r,defzip}
+zipf <- function(n){
+	out <- 1/seq_len(n)
+	out/sum(out)
+}
+zipf(9)
+```
+
+Then there are two natural ways to proceed:
+
+```{r,cache=TRUE}
+mf <- maxp(formula1)
+l0 <- loglik(indep(mf),formula1)
+l1 <- loglik(indep(zipf(size(formula1))),formula1)
+c(l0,l1)
+pchisq(2*(l0-l1),df=size(formula1)-1,lower.tail=FALSE)
+```
+
+However, this uses Zipf's law according to the players' points rating,
+and it might be better to rank the players according to their
+likelihood rating.
+
+```{r}
+mf
+mf[rank(-mf)] <- zipf(size(formula1))
+mf
+l2 <- loglik(indep(mf),formula1)
+l2
+pchisq(2*(l0-l1),df=size(formula1)-1,lower.tail=FALSE)
+```
+
+which is even worse.
 

### 2020-05-25 inst/skating_analysis.Rmd
cache expensive results
keep just the top competitors, alternative to zapweak()

@@
 
-```{r,maxshow}
+```{r,maxshow, cache = TRUE}
 (m <- maxp(skating))
@@
 
-```{r,unconstrainedlikes}
-mgv <- maxp(skating, give=TRUE, n=2)$value
+```{r,unconstrainedlikes, cache=TRUE}
+mgv <- maxp(skating, give=TRUE, n=10)$value
 mgv  - out
@@
 to Slutskaya and four preferred Slutsakaya to Hughes.
+
+# Just the top seven competitors
+
+```{r}
+skating_table
+o <- ordertable_to_ranktable(skating_table)
+o
+class(o) <- "matrix"
+o
+o <- o[,1:7]
+o
+o <- t(apply(o,1,rank))
+colnames(o) <- rownames(skating_table)[1:7]
+o
+class(o) <- "ranktable"
+o
+o <- ranktable_to_ordertable(o)
+o
+o <- ordertable2supp(o)
+o
+o <- maxp(o)
+pie(o)
+```

### 2020-05-26 inst/skating_analysis.Rmd
fixes #56

@@
 ```{r,convertordertosupport}
-head(ordertable2supp(skating_table))
+head(ordertable2supp(skating_table,misslast=FALSE))
 ```
@@
 
-```{r}
+```{r,longthing,cache=TRUE}
 jj <-
@@
 o
+```
+
+```{r}
 class(o) <- "matrix"
 o
+```
+
+Select just the top seven:
+```{r}
 o <- o[,1:7]
 o
+```
+
+and now replace ranks with internal ranks (that is, rank among the top seven):
+
+```{r}
 o <- t(apply(o,1,rank))
@@
 o
+```
+
+Now coece to a `ranktable`:
+
+```{r}
 class(o) <- "ranktable"
 o
+```
+
+Thence to an order table:
+
+```{r}
 o <- ranktable_to_ordertable(o)
 o
-o <- ordertable2supp(o)
+```
+
+Now create a likelihood function:
+
+```{r}
+o <- ordertable2supp(o,misslast=FALSE)
 o
+```
+and have some fun with it:
+
+```{r}
 o <- maxp(o)

### 2020-05-27 inst/ordertable_to_ranktable.Rmd
more documentation

---
title: "The print method for class ranktable"
output: html_document
---

```{r setup, include=FALSE}
set.seed(0)
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("magrittr")
options("digits" = 5)
```

## Rank tables

Rank tables can be confusing, here is a tiny example that should
clarify things.

```{r}
OT <- cbind(1:5,c(1,5,2:4))
rownames(OT) <- letters[seq_len(nrow(OT))]
colnames(OT) <- c("J1","J2")
OT
```

Object `a` is a straightforward order table, just like
`skating_table`; we see, for example, that both judges put competitor
`a` first, but put `b` second and fifth respectively.  The judges
second choice was `b` and `c` respectively.  We can coerce order table
`OT` to a rank table `RT`:

```{r}
RT <- ordertable_to_ranktable(OT)
RT
```

Object `RT` gives the same information but in a different form.  Note
that above we see the print method in action, which uses
`ranktable_to_printable_object()`.  However, it is possible to view
`RT` directly by removing class `ranktable`:

```{r}
class(RT)
class(RT) <- "matrix"  # removes "ranktable" from class list
RT
```

Note carefully that the second row of `RT` (which corresponds to the
second judge) does *not* match the second column of `OT`.  The way to
read the second row of `RT` is as follows.  We see the entries `1 3 4
5 2` indicating that the first choice was competitor 1 [viz `a`, the
second choice competitor 3 [viz `c`], the third choice competitor 4
[viz `d`], fourth choice competitor 5 [`e`] and fifth choice
competitor 2 [`b`].  So the rows are *ranks* (see the discussion in
`rrank.Rd`).  But it is very tempting to read the second row of `RT`
as "judge 2 placed competitor `b` in third place", which is wrong
because this is interpreting the row as an *order*.  This is why the
print method is there.

### 2020-05-27 inst/retain.Rmd
a whole world of pain for keep()

---
title: "Illustrative use of retain()"
output: html_document
---

```{r setup, include=FALSE}
set.seed(0)
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("magrittr")
options("digits" = 5)
```

## Function keep() vs function retain()

Functions `keep()` and `retain()` can be confusing, here is a tiny
example that should clarify things.

We will use a subset of
`skating_table`:

```{r}
jj <- skating_table[1:3,1:6]
rownames(jj) <- letters[1:3]
jj
```

Object `jj` does not function as a rank table because the entries are
not sequential.  We need to convert it:

```{r}
a <- apply(jj,2,rank)
a
```

So each column is a permutation of `123`.  So, for example judge 1
ranked `a` first, `b` third, and `c` second.  We can coerce to a
`hyper2` object with `ordertable2supp()`:

```{r}
ordertable2supp(a,misslast=FALSE)
```

[note carefully how `misslast=FALSE` is needed here].  Now suppose we
are interested only in the first two competitors, `a` and `b`.  There
are two ways to discard competitor `c`, namely `keep()` and
`retain()`, which do slightly different things.  First, `keep()`:

```{r}
keep(ordertable2supp(a,misslast=FALSE),1:2)
```

Above, function `keep()` has effectively set `c=0` and then discarded
the `c^5` term.  This process, while natural, has the effect of
replacing `(a+c)^-4` with `a^-4`, which has no natural probabilistic
interpretation: we are ignoring `c`'s victories but nevertheless
interpreting joint strength term on the denominator [viz `a+c`] as a
*loss* for `a`.

```{r}
RA <- retain(a,1:2)
RA
```

Looking at object `ra` we see six comparisons, of which `a` wins 5 and
`b` wins 1.

```{r}
OR <- ordertable2supp(RA,misslast=FALSE)
OR
```

Above, object `or` is a perfectly sensible likelihood function.  What
happens if we `keep()` only `a`?

```{r}
keep(OR,1)
```

the above support function is meaningless (?) as, having set `b=0` we
then treat `a+b` as simply `a` which sort of makes sense until we
realise that `(a+b)^-6` reduces to `a^-6`.  It is as though `a` has
wins only 3 trials out of 8, against a competitor of known zero
strength whose wins we ignore.  Perhaps this has a natural
interpretation in probability, but if so I don't see it.

### 2020-05-27 R/hyper2.R

@@ -981,3 +981,11 @@ setGeneric("pnames<-",function(x,value){standardGeneric("pnames<-")})
     return(out)
 }
 
+`retain` <- function(x, wanted){
+    o <- ordertable_to_ranktable(x)
+    class(o) <- "matrix"
+    o <- t(apply(o[,wanted],1,rank))
+    colnames(o) <- rownames(x)[wanted]
+    class(o) <- "ranktable"
+    return(ranktable_to_ordertable(o))
+}

### 2020-05-27 inst/retain.Rmd
keep() and keep_flawed() implementation

@@
 ---
-title: "Illustrative use of retain()"
+title: "Keeping and discarding players"
 output: html_document
@@
 
-## Function keep() vs function retain()
+## Functions keep() and keep_flawed()
 
-Functions `keep()` and `retain()` can be confusing, here is a tiny
-example that should clarify things.
-
-We will use a subset of
+Functions `keep()` and `keep_flawed()` can be confusing, here is a
+tiny example that should clarify things.  We will use a subset of
 `skating_table`:
@@
 ```{r}
-ordertable2supp(a,misslast=FALSE)
+OT <- ordertable2supp(a,misslast=FALSE)
+OT
 ```
@@
 are two ways to discard competitor `c`, namely `keep()` and
-`retain()`, which do slightly different things.  First, `keep()`:
+`keep_flawed()`, which do slightly different things.  First, `keep()`:
 
 ```{r}
-keep(ordertable2supp(a,misslast=FALSE),1:2)
+keep_flawed(OT,1:2)
 ```
 
-Above, function `keep()` has effectively set `c=0` and then discarded
-the `c^5` term.  This process, while natural, has the effect of
-replacing `(a+c)^-4` with `a^-4`, which has no natural probabilistic
-interpretation: we are ignoring `c`'s victories but nevertheless
-interpreting joint strength term on the denominator [viz `a+c`] as a
-*loss* for `a`.
+Above, function `keep_flawed()` has effectively taken `OT`, set `c=0`,
+and then discarded the `c^5` term.  This process, while natural, has
+the effect of replacing `(a+c)^-4` with `a^-4`, which has no natural
+probabilistic interpretation: we are ignoring `c`'s victories but
+nevertheless interpreting joint strength term on the denominator [viz
+`a+c`] as a *loss* for `a`.  We will now use function `keep()`, which
+operates on order tables:
 
 ```{r}
-RA <- retain(a,1:2)
+RA <- keep(a,1:2)
 RA
@@
 Above, object `or` is a perfectly sensible likelihood function.  What
-happens if we `keep()` only `a`?
+happens if we keep only `a`?
 
 ```{r}
-keep(OR,1)
+keep_flawed(OR,1)
 ```
@@
 wins only 3 trials out of 8, against a competitor of known zero
-strength whose wins we ignore.  Perhaps this has a natural
-interpretation in probability, but if so I don't see it.
+strength whose (impossible) wins we ignore.  Perhaps this has a
+natural interpretation in probability, but if so I don't see it.

### 2020-05-27 R/hyper2.R

@@ -906,7 +906,7 @@ setGeneric("pnames<-",function(x,value){standardGeneric("pnames<-")})
     m <- maxp(H,n=1,...)
     too_weak <- m < minstrength
     if(any(too_weak)){
-      H %<>% discard(names(m[which.min(m)]))
+      H %<>% discard_flawed(names(m[which.min(m)]))
     } else {
      break
     }
@@ -981,11 +981,27 @@ setGeneric("pnames<-",function(x,value){standardGeneric("pnames<-")})
     return(out)
 }
 
-`retain` <- function(x, wanted){
-    o <- ordertable_to_ranktable(x)
+`keep` <- function(x, wanted){
+    if(is.character(wanted)){
+        wanted <- which(rownames(x)==wanted)
+    }
+    o <- wikitable_to_ranktable(x) # was o <- ordertable_to_ranktable(x)
     class(o) <- "matrix"
     o <- t(apply(o[,wanted],1,rank))
     colnames(o) <- rownames(x)[wanted]
     class(o) <- "ranktable"
     return(ranktable_to_ordertable(o))
 }
+
+`discard` <- function(x, unwanted){
+    if(is.character(unwanted)){
+        wanted <- !which(rownames(x)==unwanted)
+    }
+    o <- wikitable_to_ranktable(x) # was   o <- ordertable_to_ranktable(x)
+    class(o) <- "matrix"
+    o <- t(apply(o[,unwanted],1,rank))
+    colnames(o) <- rownames(x)[unwanted]
+    class(o) <- "ranktable"
+    return(ranktable_to_ordertable(o))
+}
+

### 2020-05-27 R/keep.R

@@ -21,22 +21,22 @@
   return(hyper2(bracketout,powerout,pout))
 }
 
-`keep` <- function(H, keep, tidy=TRUE){
+`keep_flawed` <- function(H, wanted, tidy=TRUE){
   p <- pnames(H)    # might be NA
-  if(is.character(keep)){
-    stopifnot(all(keep %in% p))
-    keep <- which(p %in% keep) # 'keep' now numeric
+  if(is.character(wanted)){
+    stopifnot(all(wanted %in% p))
+    wanted <- which(p %in% wanted) # 'wanted' now numeric
   } else {
     jj <- seq_len(size(H))
-    stopifnot(all(keep %in% jj))
-    keep <- which(jj %in% keep) # 'keep' now numeric
+    stopifnot(all(wanted %in% jj))
+    wanted <- which(jj %in% wanted) # 'wanted' now numeric
   }
 
   bracketout <- list()
   powerout <- NULL
   for(i in seq_along(H)){
     b <- brackets(H)[[i]]
-    jj <- b[b %in% keep]   # the meat
+    jj <- b[b %in% wanted]   # the meat
     if(length(jj)>0){
       bracketout <- c(bracketout, list(jj))
       powerout <- c(powerout, powers(H)[i])
@@ -47,15 +47,15 @@
   return(out)
 }
 
-`discard` <- function(H, discard, tidy=TRUE){
+`discard_flawed` <- function(H, unwanted, tidy=TRUE){
   p <- pnames(H)
-  if(is.character(discard)){
-    stopifnot(all(discard %in% p))
-    keep <- which(!(p %in% discard))
+  if(is.character(unwanted)){
+    stopifnot(all(unwanted %in% p))
+    wanted <- which(!(p %in% unwanted))
   } else {
-    jj <- seq_len(size(H))
-    stopifnot(all(discard %in% jj))
-    keep <- which(!(jj %in% discard))
+      jj <- seq_len(size(H))
+      stopifnot(all(unwanted %in% jj))
+      wanted <- which(!(jj %in% unwanted))
   }
-  return(keep(H,keep,tidy=tidy))  # the meat
+  return(keep_flawed(H,wanted=wanted, tidy=tidy))  # the meat
 }

### 2020-05-27 R/hyper2.R
points -> misslast

@@ -936,8 +936,8 @@ setGeneric("pnames<-",function(x,value){standardGeneric("pnames<-")})
   return(hyper2(B,powers(H),pnames=pnames(H)))
 }
 
-`wikitable_to_ranktable`  <- function(wikitable, points=TRUE, strict=FALSE){
-  if(points){wikitable <- wikitable[,-ncol(wikitable)]}
+`wikitable_to_ranktable`  <- function(wikitable, misslast=TRUE, strict=FALSE){
+  if(misslast){wikitable <- wikitable[,-ncol(wikitable)]}
 
   f <- function(x){  # deal with DNF etc and zero
     suppressWarnings(out <- as.numeric(as.vector(x)))

### 2020-05-29 inst/formula1.Rmd
now called keep_flawed()
get rid of that stupid 'misslast' argument

@@
 f2017 <- read.table("formula1_2017.txt",header=TRUE)
+f2017 <- f2017[,1:20]  # final column is points
 head(f2017)
@@
 ```{r,likelihoodandscore}
-points <- f2017$points
-names(points) <- f2017$driver
+jj <- read.table("formula1_2017.txt",header=TRUE)
+points <- jj[,21]
+names(points) <- rownames(jj)
 
@@
 pnames(H)
-H9 <- keep(H,c("S",pnames(H)[1:8]))
+H9 <- keep_flawed(H,c("S",pnames(H)[1:8]))
 specificp.gt.test(H9,"S",0)

### 2020-05-29 R/hyper2.R

@@ -864,16 +864,11 @@ setGeneric("pnames<-",function(x,value){standardGeneric("pnames<-")})
     return(H)
 }
 
-`ordertable2supp` <- function(x, noscore, misslast=TRUE){
+`ordertable2supp` <- function(x, noscore){
     if(missing(noscore)){
         noscore <- c("Ret", "WD", "DNS", "DSQ", "DNP", "NC")
     }
-
-    if(misslast){
-        venues <- colnames(x)[-ncol(x)]
-    } else {
-        venues <- colnames(x)
-    }
+    venues <- colnames(x)
 
     ## Now create a numeric matrix, fmat.  Two steps: first, count
     ## any no-score as zero:
@@ -885,7 +880,6 @@ setGeneric("pnames<-",function(x,value){standardGeneric("pnames<-")})
     
     ## Second, convert to numeric and strip out names; transpose of
     ## x (because we want each row to be a venue):
-    if(misslast){jj <- jj[,-ncol(x)]}
     fmat <- matrix(as.numeric(jj),byrow=TRUE,ncol=nrow(x)) 
     colnames(fmat) <- rownames(x)
     rownames(fmat) <- venues
@@ -936,9 +930,7 @@ setGeneric("pnames<-",function(x,value){standardGeneric("pnames<-")})
   return(hyper2(B,powers(H),pnames=pnames(H)))
 }
 
-`wikitable_to_ranktable`  <- function(wikitable, misslast=TRUE, strict=FALSE){
-  if(misslast){wikitable <- wikitable[,-ncol(wikitable)]}
-
+`wikitable_to_ranktable`  <- function(wikitable, strict=FALSE){
   f <- function(x){  # deal with DNF etc and zero
     suppressWarnings(out <- as.numeric(as.vector(x)))
     DNF <- is.na(out) | (out==0)

### 2020-05-29 data/formula1.R

@@ -286,96 +286,62 @@ structure(list(brackets = list(1, c(1, 2, 3, 4, 5, 6, 7, 8, 9,
 "Ericsson", "Gasly", "Giovinazzi", "Hartley", "Button", "Resta"
 )), class = "hyper2")
 
+`F1_points_2017` <- c(Hamilton = 363L, Vettel = 317L, Bottas = 305L, Raikkonen = 205L, 
+Ricciardo = 200L, Verstappen = 168L, Perez = 100L, Ocon = 87L, 
+Sainz = 54L, Hulkenberg = 43L, Massa = 43L, Stroll = 40L, Grosjean = 28L, 
+Magnussen = 19L, Alonso = 17L, Vandoorne = 13L, Palmer = 8L, 
+Wehrlein = 5L, Kvyat = 0L, Ericsson = 0L, Gasly = 0L, Giovinazzi = 0L, 
+Hartley = 0L, Button = 0L, Resta = 0L)
+
 `F1_table_2017` <-
-structure(list(AUS = structure(c(7L, 2L, 8L, 9L, 15L, 10L, 12L, 
-3L, 13L, 4L, 11L, 15L, 15L, 15L, 15L, 6L, 15L, 16L, 14L, 15L, 
-1L, 5L, 1L, 1L, 1L), .Label = c("0", "1", "10", "11", "12", "13", 
-"2", "3", "4", "5", "6", "7", "8", "9", "Ret", "WD"), class = "factor"), 
-    CHN = structure(c(2L, 9L, 13L, 12L, 11L, 10L, 16L, 3L, 14L, 
-    5L, 7L, 17L, 4L, 15L, 17L, 17L, 6L, 1L, 17L, 8L, 1L, 17L, 
-    1L, 1L, 1L), .Label = c("0", "1", "10", "11", "12", "13", 
-    "14", "15", "2", "3", "4", "5", "6", "7", "8", "9", "Ret"
-    ), class = "factor"), BHR = structure(c(8L, 2L, 9L, 10L, 
-    11L, 17L, 13L, 3L, 17L, 15L, 12L, 17L, 14L, 17L, 7L, 16L, 
-    6L, 4L, 5L, 17L, 1L, 1L, 1L, 1L, 1L), .Label = c("0", "1", 
-    "10", "11", "12", "13", "14", "2", "3", "4", "5", "6", "7", 
-    "8", "9", "DNS", "Ret"), class = "factor"), RUS = structure(c(12L, 
[…]

### 2020-05-31 inst/f1points.Rmd
draft file for submission

---
documentclass: jss
author:
  - name: Robin K. S. Hankin
    affiliation: Auckland University of Technology
    address: >
      2-14 Wakefield Street
      Auckland, New Zealand
    email: \email{hankin.robin@gmail.com}
    url: http://aut.ac.nz/
title:
  formatted: "An Objective Assessment of Points Scorings in Formula 1 Motor Racing"
  plain:     "An Objective Assessment of Points Scorings in Formula 1 Motor Racing"
  # For running headers, if needed
  short:     "Formula 1 Points Scoring"
abstract: >
  The points scoring system of Formula 1 motor racing is a long-standing
  contention among fans and competitors, with changes to the points
  scored occurring frequently.  Here, I apply generalized Plackett-Luce
  likelihood functions to Formula 1 racing, and assess historical points
  scoring systems objectively.  In a well-defined sense, the points
  scoring system of 1985 is the best of the historical ones.  I make
  similar observations about related competitive situations including
  the Eurovision Song contest and XXX.
keywords:
  # at least one keyword must be supplied
  formatted: [keywords, not capitalized, "\\proglang{Java}"]
  plain:     [keywords, not capitalized, Java]
preamble: >
  \usepackage{amsmath}
output: rticles::jss_article
---

# Introduction

Formula 1 racing is an important and prestigious motor sport
\citep{moul2009}.  Season ranking is based on a points system; points
are awarded to drivers and teams based on race order, and accumulate
additively.  The current points system, for example, is
(25-18-15-12-10-8-6-4-2-1) [that is, the winner is awarded 25 points,
second place 18 points, and so on], although since 2019 the fastest
lap attracts one point.

Given that racing is a zero-sum game---and that points are
monotonically decreasing---each player will try to get as high a rank
as possible regardless of the actual points system used.  However,
there are other consistent interpretations.  \cite{bakhrankova2011},
for example, considers the possibility of inter-driver collusion, a
phenomenon not pursued here; and authors such as
\citeauthor{mastromarco2009} suggest that the frequency of rule
changes is driven by factors such as driver safety and revenue
optimization.

Points systems similar to that of Formula 1 are common in other sports
and examples would include modern pentathlon, curling, chess, rowing,
and figure skating; all have the common feature of interpretation of
an order statistic to a points.  Further, we see points systems
frequently in the wider context of competitive situations such as the
Eurovision Song Contest, Australian MasterChef, University rankings,
educational league tables, and car safety ratings.

## Bradley-Terry and generalizations for rank statistics

The Bradley-Terry model \citep{bradley1952} assigns non-negative
strengths $p_1,\ldots, p_n$ to each of $n$ competitors in such a way
that the probability of $i$ beating $j\neq i$ in pairwise competition
is $\frac{p_i}{p_i+p_j}$; it is conventional to normalize so that
$\sum p_i=1$.  Further, we use a generalization due to
\citet{luce1959}, in which the probability of competitor $i$ winning
in a field of $j=1,\ldots, n$ is $\frac{p_i}{p_1+\cdots +p_n}$.
Noting that there is information in the whole of the finishing order,
not just the first across the line, we can follow \cite{plackett1975}
and consider the runner-up to be the winner among the remaining
competitors; and so on down the finishing order.  Without loss of
generality, if the order of finishing were $1,2,3,4$, then a suitable
\citeauthor{plackett1975} likelihood function would be

\begin{equation}\label{competitors_1_to_4_likelihood}
\frac{p_1}{p_1+p_2+p_3+p_4}\cdot
\frac{p_2}{p_2+p_3+p_4}\cdot
\frac{p_3}{p_3+p_4}\cdot
\frac{p_4}{p_4}\cdot
\end{equation}

We now use a technique due to \cite{hankin2010,hankin2017} and
introduce fictional (reified) entities whose nonzero Bradley-Terry
strength helps certain competitors or sets of competitors.  The
canonical example would be the home-ground advantage in association
football.  If players (teams) $1,2$ with strengths $p_1,p_2$ compete,
and if our observation were $a$ home wins and $b$ away wins for team
$1$, and $c$ home wins and $d$ away wins for team $2$, then a suitable
likelihood function would be

\[
\left(\frac{p_1+p_H}{p_1+p_2+p_H}\right)^a
\left(\frac{p_1}{p_1+p_2+p_H}\right)^b
\left(\frac{p_2+p_H}{p_1+p_2+p_H}\right)^c
\left(\frac{p_2+p_H}{p_1+p_2+p_H}\right)^d
\]

where $p_H$ is a quantification of the beneficial home ground effect.
Similar techniques have been used to account for the first-move
advantage in chess, and effective coordination between members of
doubles tennis teams; we may use a similar device to account for pole
position in Formula 1.  Here I analyse racing results from the 2017
season using the `hyper2` package \citep{hankin2017} which implements
the Plackett-Luce likelihood function with additional reified
entities.

# Formula 1 dataset

Here we use the results from 2017:

\small{
\begin{CodeChunk}
\begin{CodeOutput}
           AUS CHN BHR RUS ESP MON CAN AZE AUT GBR HUN BEL ITA SIN MAL JPN USA MEX BRA ABU
Hamilton     2   1   2   4   1   7   1   5   4   1   4   1   1   1   2   1   1   9   4   2
Vettel       1   2   1   2   2   1   4   4   2   7   1   2   3 Ret   4 Ret   2   4   1   3
Bottas       3   6   3   1 Ret   4   2   2   1   2   3   5   2   3   5   4   5   2   2   1
Raikkonen    4   5   4   3 Ret   2   7  14   5   3   2   4   5 Ret DNS   5   3   3   3   4
Ricciardo  Ret   4   5 Ret   3   3   3   1   3   5 Ret   3   4   2   3   3 Ret Ret   6 Ret
...
Hartley      0   0   0   0   0   0   0   0   0   0   0   0   0   0   0   0  13 Ret Ret  15
Button       0   0   0   0   0 Ret   0   0   0   0   0   0   0   0   0   0   0   0   0   0
Resta        0   0   0   0   0   0   0   0   0   0 Ret   0   0   0   0   0   0   0   0   0
\end{CodeOutput}
\end{CodeChunk}
}

Each row is a driver and each column (after the first) a venue.  We
see that Hamilton, the first row, came second in Australia, first in
China, second in Bahrain, fourth in Russia, and so on (Hartley,
Button, and Resta placed last).  In the first column, we see the
result from Australia (AUS) in which Hamilton came second, Vettel
first, Bottas third, and so on.  Notation used also includes six
different classes of no-score such as `Ret` for retired, `WD` for
withdrawn, and so on; a zero entry means "did not finish".  It is
straightforward to translate this table into a Plackett-Luce
log-likelihood function using the `hyper2` package; for simplicity we
will consider only the 10 top-ranked drivers\footnote{In the
Plackett-Luce likelihood function, the performance of lower-ranked
players can be informative about higher-ranked players' strengths.
For example, we see that Vettel retired twice (in Singapore and
Japan), so any player who placed in those venues will effectively
``steal'' strength from Vettel, and generally ``give'' it to Hamilton
or Bottas.}.  Although it has many terms, the first few are

```{r,echo=FALSE,print=FALSE}
library(hyper2,quietly=TRUE)
H <- ordertable2supp(keep(F1_table_2017,1:10))
head(H,5)
```

Finding the maximum likelihood estimate for the players' strengths is straightforward:

```{r,echo=FALSE}
mH <- maxp(H)
print(mH)
pie(maxp(H)[c(1,6,2,7,3,8,4,9,5,10)])
```

We see the that the top three players, Hamilton, Vettel, Bottas, are
of roughly equal strength and indeed a likelihood ratio test
[`samep.test()`] does not reject the null of equality ($p=0.72$).

## Likelihood scoring vs points scoring

The points scored by each driver give a ranking shown in the table;
but Plackett-Luce likelihoods also give a ranking that us unencumbered
by the arbitrary points given.  We can plot one ranking against the other:

```{r,label=likevspoints,cache=TRUE}
jj <- read.table("formula1_2017.txt",header=TRUE)
points <- jj[,21]
names(points) <- rownames(jj)

mL2017 <- maxp(H)

m <- mL2017   # maxp(H)
ox <- round(rank(-points)[1:10]) # negative because number 1 = largest number of points
oy <- rank(-m)      # negative because number 1 = largest strength
par(pty='s')        # square plot
plot(ox,oy,asp=1,pty='s',xlim=c(0,10.2),ylim=c(0,10.2),pch=16,
xlab="official order",ylab="my order",main='Formula 1, 2017 season')
par(xpd=TRUE)       # allow drivers' names to appear outside plotting region
for(i in seq_along(ox)){text(ox[i],oy[i],names(ox)[i],pos=4,col='gray') }
par(xpd=FALSE)      # stop diagonal line from protruding beyond plotting region
abline(0,1)
grid()
print(ox)
print(oy)
```

\bibliography{chess}

### 2020-06-01 inst/f1points.Rmd
incomplete argument to ordertable2supp()

@@
 
-\small{
+\small
 \begin{CodeChunk}
@@
 \end{CodeChunk}
-}
+
+```{r,echo=FALSE,print=FALSE}
+top <- 11
+```
 
@@
 log-likelihood function using the `hyper2` package; for simplicity we
-will consider only the 10 top-ranked drivers\footnote{In the
+will consider only the `r top` top-ranked drivers\footnote{In the
 Plackett-Luce likelihood function, the performance of lower-ranked
-players can be informative about higher-ranked players' strengths.
-For example, we see that Vettel retired twice (in Singapore and
-Japan), so any player who placed in those venues will effectively
-``steal'' strength from Vettel, and generally ``give'' it to Hamilton
-or Bottas.}.  Although it has many terms, the first few are
+players can be weakly informative about higher-ranked players'
+strengths.  For example, we see that Vettel retired twice (in
+Singapore and Japan), so any player who placed in those venues will
+effectively ``steal'' strength from Vettel, and generally ``give'' it
+to Hamilton or Bottas.}.  Although it has many terms, the first few
+are
 
@@
 library(hyper2,quietly=TRUE)
-H <- ordertable2supp(keep(F1_table_2017,1:10))
+top <- 11
+H <- ordertable2supp(F1_table_2017[seq_len(top),])
 head(H,5)
@@
 
-Finding the maximum likelihood estimate for the players' strengths is straightforward:
+Finding the maximum likelihood estimate for the players' strengths is
+straightforward:
 
@@
 print(mH)
-pie(maxp(H)[c(1,6,2,7,3,8,4,9,5,10)])
+pie(maxp(H)[c(1,6,11,2,7,3,8,4,9,5,10)])
 ```
 
-We see the that the top three players, Hamilton, Vettel, Bottas, are
+We see the that the top two players, Hamilton and Vettel
 of roughly equal strength and indeed a likelihood ratio test
-[`samep.test()`] does not reject the null of equality ($p=0.72$).
+[`samep.test()`] does not reject the null of equality ($p=0.09$).
+
+```{r,cache=TRUE}
+samep.test(H,1:2)
+```
 
@@
 The points scored by each driver give a ranking shown in the table;
-but Plackett-Luce likelihoods also give a ranking that us unencumbered
-by the arbitrary points given.  We can plot one ranking against the other:
+but Plackett-Luce likelihoods also give a ranking that does not depend
+on any arbitrary points system.  We can plot one ranking against the
+other:
 
-```{r,label=likevspoints,cache=TRUE}
+```{r,label=likevspoints,cache=TRUE,echo=FALSE}
 jj <- read.table("formula1_2017.txt",header=TRUE)
@@
 mL2017 <- maxp(H)
-
 m <- mL2017   # maxp(H)
-ox <- round(rank(-points)[1:10]) # negative because number 1 = largest number of points
+ox <- round(rank(-points[seq_len(top)]))  # negative because number 1 = largest number of points
 oy <- rank(-m)                            # negative because number 1 = largest strength
 par(pty='s')        # square plot
-plot(ox,oy,asp=1,pty='s',xlim=c(0,10.2),ylim=c(0,10.2),pch=16,
+plot(ox,oy,asp=1,pty='s',xlim=c(0,11.2),ylim=c(0,11.2),pch=16,
     xlab="official order",ylab="my order",main='Formula 1, 2017 season')
@@
 abline(0,1)
-grid()
-print(ox)
-print(oy)
 ```
 
+From the above, we see that
+
 \bibliography{chess}

### 2020-06-01 R/hyper2.R

@@ -864,7 +864,7 @@ setGeneric("pnames<-",function(x,value){standardGeneric("pnames<-")})
     return(H)
 }
 
-`ordertable2supp` <- function(x, noscore){
+`ordertable2supp` <- function(x, noscore, incomplete=TRUE){
     if(missing(noscore)){
         noscore <- c("Ret", "WD", "DNS", "DSQ", "DNP", "NC")
     }
@@ -888,7 +888,9 @@ setGeneric("pnames<-",function(x,value){standardGeneric("pnames<-")})
     
     ## Now cycle through the rows; each row is a venue [voter]
     for(i in seq_len(nrow(fmat))){
-        out %<>% `+`(ordervec2supp(fmat[i,,drop=TRUE]))
+        o <- fmat[i,,drop=TRUE]
+        if(incomplete){ o[o>0] <- order(o[o>0]) }
+        out %<>% `+`(ordervec2supp(o))
     } # i loop closes
     return(out)
 } 

### 2020-06-01 inst/f1points.Rmd
use ordertable2points() instead

@@
 
+```{r,label=calcpoints}
+p <- c(25, 18, 15, 12, 10, 8, 6, 4, 2, 1, 0, 0)
+dp <- ordertable2points(as.ordertable(F1_table_2017[seq_len(top),]),p)
+dp <- dp[order(-dp)]
+dp[] <- seq_along(dp)
+dp
+```
+nw:
+
 ```{r,label=likevspoints,cache=TRUE,echo=FALSE}
-jj <- read.table("formula1_2017.txt",header=TRUE)
-points <- jj[,21]
-names(points) <- rownames(jj)
 
-mL2017 <- maxp(H)
-m <- mL2017   # maxp(H)
-ox <- round(rank(-points[seq_len(top)]))  # negative because number 1 = largest number of points
-oy <- rank(-m)                            # negative because number 1 = largest strength
+m <- maxp(H)  # just 'top' players
+ox <- dp
+oy <- rank(-m)                           # minus as number 1 = largest strength
 par(pty='s')        # square plot
@@
 
-From the above, we see that
+From the above, we see that the first three drivers (Hamilton, Vettel,
+Bottas) are respectively first, second and third according to both
+ranking procedures.
 

### 2020-06-02 inst/f1points.Rmd
more work, proper comparison between points and likelihood ranking

--- 2020-06-01 inst/f1points.Rmd
@@
 competitors; and so on down the finishing order.  Without loss of
-generality, if the order of finishing were $1,2,3,4$, then a suitable
+generality, if the order of finishing were $1,2,3,4,5$, then a suitable
 \citeauthor{plackett1975} likelihood function would be
 
-\begin{equation}\label{competitors_1_to_4_likelihood}
-\frac{p_1}{p_1+p_2+p_3+p_4}\cdot
-\frac{p_2}{p_2+p_3+p_4}\cdot
-\frac{p_3}{p_3+p_4}\cdot
-\frac{p_4}{p_4}\cdot
+\begin{equation}\label{competitors_1_to_5_likelihood}
+\frac{p_1}{p_1+p_2+p_3+p_4+p_5}\cdot
+\frac{p_2}{p_2+p_3+p_4+p_5}\cdot
+\frac{p_3}{p_3+p_4+p_5}\cdot
+\frac{p_4}{p_4+p_5}\cdot
+\frac{p_5}{p_5}
 \end{equation}
 
-We now use a technique due to \cite{hankin2010,hankin2017} and
-introduce fictional (reified) entities whose nonzero Bradley-Terry
-strength helps certain competitors or sets of competitors.  The
-canonical example would be the home-ground advantage in association
-football.  If players (teams) $1,2$ with strengths $p_1,p_2$ compete,
-and if our observation were $a$ home wins and $b$ away wins for team
-$1$, and $c$ home wins and $d$ away wins for team $2$, then a suitable
-likelihood function would be
+and a slight generalization of this would allow the incorporation of
+non-finishers (DNF etc).  If, say, competitors 4 and 5 did not finish,
+we would have
+
+\begin{equation}\label{competitors_1_to_3_only_finished}
+\frac{p_1}{p_1+p_2+p_3+p_4+p_5}\cdot
+\frac{p_2}{p_2+p_3+p_4+p_5}\cdot
+\frac{p_3}{p_3+p_4+p_5}
+\end{equation}
+
+(observe how this likelihood function, while informative about
+$p_4+p_5$, is uninformative about $p_4\left|p_4+p_5\right.$).  We now
+use a technique due to \cite{hankin2010,hankin2017} and introduce
+fictional (reified) entities whose nonzero Bradley-Terry strength
+helps certain competitors or sets of competitors.  The canonical
+example would be the home-ground advantage in association football.
+If players (teams) $1,2$ with strengths $p_1,p_2$ compete, and if our
+observation were $a$ home wins and $b$ away wins for team $1$, and $c$
+home wins and $d$ away wins for team $2$, then a suitable likelihood
+function would be
 
@@
 
-We see the that the top two players, Hamilton and Vettel
-of roughly equal strength and indeed a likelihood ratio test
-[`samep.test()`] does not reject the null of equality ($p=0.09$).
+We see the Hamilton has the largest estimated strength, although it
+should be noted that a likelihood ratio test [`samep.test()`] does not
+reject the null of $p_\mathrm{Hamilton}=p_\mathrm{Vettel}$ ($p=0.09$).
 
@@
 
+```{r,label=calcpoints}
+p <- c(25, 18, 15, 12, 10, 8, 6, 4, 2, 1, 0, 0)
+dp <- ordertable2points(as.ordertable(F1_table_2017[seq_len(top),]),p)
+dp
+dp <- dp[order(-dp)]
+dp[] <- seq_along(dp)
+```
+now likelihood
+
+```{r,label=calclike}
+m <- maxp(H)
+m <- sort(m,decreasing=TRUE)
+m[] <- seq_along(m)
+```
+
+now ranks:
+
+```{r}
+jj <- outer(names(m),names(dp),"==")
+ox <- dp
+oy <- m[which(jj,arr.ind=TRUE)[,1]]
+```
+
+and plot:
+
 ```{r,label=likevspoints,cache=TRUE,echo=FALSE}
-jj <- read.table("formula1_2017.txt",header=TRUE)
-points <- jj[,21]
-names(points) <- rownames(jj)
-
-mL2017 <- maxp(H)
-m <- mL2017   # maxp(H)
-ox <- round(rank(-points[seq_len(top)]))  # negative because number 1 = largest number of points
-oy <- rank(-m)                            # negative because number 1 = largest strength
 par(pty='s')        # square plot
@@
 
-From the above, we see that
+From the above, we see that the first four drivers (Hamilton, Vettel,
+Raikkonen, Bottas) are respectively first, second, third and fourth
+according to both ranking procedures.  The lower-ranked players,
+having estimated strengths of less than 0.03, are more scattered as
+one might expect.
+
+We can use this observation to assess a given points system: using
+Monte-Carlo techniques we can generate an order table \emph{in
+silico}, using estimated driver strengths from the 2017 table, and
+then compare rankings generated by estimated \citeauthor{plackett1975}
+strengths with rankings generated by the points system.
+
+# Monte Carlo assumptions
+
+blah blah
 

### 2020-06-06 inst/f1points.Rmd
more commentary

@@
 
-From the above, we see that the first four drivers (Hamilton, Vettel,
-Raikkonen, Bottas) are respectively first, second, third and fourth
-according to both ranking procedures.  The lower-ranked players,
-having estimated strengths of less than 0.03, are more scattered as
-one might expect.
+From the above, we see that the first two drivers (Hamilton, Vettel)
+are are respectively first and second according to both ranking
+ranking procedures; but Raikkonen and Bottas are third and fourth, and
+fourth and third, according to points and likelihood respectively.
+The degree of agreement between the two systems might be quantified as
+"two", for the positions of the top two drivers agree.
 

### 2020-06-06 inst/f1points.Rmd
clarification and better idiom

@@
 dp
-dp <- dp[order(-dp)]
+dp <- sort(dp,decreasing=TRUE)
 dp[] <- seq_along(dp)
+dp
 ```

### 2020-06-07 inst/f1points.Rmd
more work, now have good comparison of points and likelihood ranking

--- 2020-06-06 inst/f1points.Rmd
@@
 
-```{r,label=calcpoints}
+```{r,label=calcpoints,echo=TRUE}
 p <- c(25, 18, 15, 12, 10, 8, 6, 4, 2, 1, 0, 0)
 dp <- ordertable2points(as.ordertable(F1_table_2017[seq_len(top),]),p)
-dp
-dp <- dp[order(-dp)]
+dp <- sort(dp,decreasing=TRUE)
 dp[] <- seq_along(dp)
-```
-now likelihood
-
-```{r,label=calclike}
 m <- maxp(H)
@@
 
-now ranks:
-
-```{r}
-jj <- outer(names(m),names(dp),"==")
-ox <- dp
-oy <- m[which(jj,arr.ind=TRUE)[,1]]
-```
-
-and plot:
-
 ```{r,label=likevspoints,cache=TRUE,echo=FALSE}
 par(pty='s')        # square plot
+ox <- dp
+oy <- ordertrans(m,dp)
 plot(ox,oy,asp=1,pty='s',xlim=c(0,11.2),ylim=c(0,11.2),pch=16,
@@
 
-blah blah
+```{r,cache=TRUE}
+set.seed(1)
+f2017 <- read.table("formula1_2017.txt",header=TRUE)[,1:20]
+random_table <- rrank(n=ncol(f2017), p=mH)
+rownames(random_table) <- colnames(f2017)
+Hran <- ordertable2supp(ranktable_to_ordertable(random_table))
+m <- maxp(Hran)
+sort(m,decreasing=TRUE)
+m[] <- seq_along(m)
+m
+p <- c(25, 18, 15, 12, 10, 8, 6, 4, 2, 1, 0, 0)
+dp <- ordertable2points(as.ordertable(ranktable_to_ordertable(random_table)),p)
+dp <- sort(dp,decreasing=TRUE)
+dp[] <- seq_along(dp)
+dp
+ox <- dp
+oy <- ordertrans(m,dp)
+par(pty="s")
+plot(1:10,1:10,asp=1,type="p",pch=NA,
+    xlab="official order",ylab="my order",main='Formula 1, 2017 virtual season',
+	xlim=c(0,11.2),ylim=c(0,11.2))
+points(ox,oy,pch=16)
+par(xpd=TRUE)       # allow drivers' names to appear outside plotting region
+for(i in seq_along(ox)){text(ox[i],oy[i],names(ox)[i],pos=4,col='gray') }
+par(xpd=FALSE)      # stop diagonal line from protruding beyond plotting region
+abline(0,1)
+```
 

### 2020-06-07 inst/f1points.Rmd
more work

@@
 
-```{r,cache=TRUE}
-set.seed(1)
-f2017 <- read.table("formula1_2017.txt",header=TRUE)[,1:20]
-random_table <- rrank(n=ncol(f2017), p=mH)
+```{r,montecarlofirst,cache=TRUE}
+top <- 11
+f2017 <- read.table("formula1_2017.txt",header=TRUE)[seq_len(top),1:20]
+m <- maxp(ordertable2supp(as.ordertable(f2017)))
+
+## Now work on synthetic dataset generation:
+points <- c(25, 18, 15, 12, 10, 8, 6, 4, 2, 1, 0, 0)
+resampling <- function(m,f2017,points){
+    random_table <- rrank(n=ncol(f2017), p=m)
     rownames(random_table) <- colnames(f2017)
-Hran <- ordertable2supp(ranktable_to_ordertable(random_table))
-m <- maxp(Hran)
-sort(m,decreasing=TRUE)
-m[] <- seq_along(m)
-m
-p <- c(25, 18, 15, 12, 10, 8, 6, 4, 2, 1, 0, 0)
-dp <- ordertable2points(as.ordertable(ranktable_to_ordertable(random_table)),p)
-dp <- sort(dp,decreasing=TRUE)
-dp[] <- seq_along(dp)
-dp
-ox <- dp
-oy <- ordertrans(m,dp)
-par(pty="s")
-plot(1:10,1:10,asp=1,type="p",pch=NA,
-    xlab="official order",ylab="my order",main='Formula 1, 2017 virtual season',
-	xlim=c(0,11.2),ylim=c(0,11.2))
-points(ox,oy,pch=16)
-par(xpd=TRUE)       # allow drivers' names to appear outside plotting region
-for(i in seq_along(ox)){text(ox[i],oy[i],names(ox)[i],pos=4,col='gray') }
-par(xpd=FALSE)      # stop diagonal line from protruding beyond plotting region
-abline(0,1)
+    ## "random_table" is a random table; now calculate lstar and pstar for
+    ## likelihood ranks and points ranks respectively:
+
+    l_star <- maxp(ordertable2supp(ranktable_to_ordertable(random_table)))
+    l_star[] <- seq_along(l_star)
+    ## l_star is the likelihood order statistic.
+
+    p_star <- ordertable2points(as.ordertable(ranktable_to_ordertable(random_table)),p)
+    p_star <- sort(p_star,decreasing=TRUE)
+    p_star[] <- seq_along(p_star)
+
+    goodness <- sum(cumprod(names(p_star)==names(l_star)))
+    return(goodness)
+}
+
+o <- replicate(1000,resampling(m,f2017,points))
+plot(table(o))
 ```

### 2020-06-07 inst/f1points.Rmd
another point system

@@
 ## Now work on synthetic dataset generation:
-points <- c(25, 18, 15, 12, 10, 8, 6, 4, 2, 1, 0, 0)
+points1 <- c(25, 18, 15, 12, 10, 8, 6, 4, 2, 1, 0, 0)
+points2 <- c(1, 0,0,0,0)
 resampling <- function(m,f2017,points){
@@
 
-o <- replicate(1000,resampling(m,f2017,points))
-plot(table(o))
+o1 <- replicate(100,resampling(m,f2017,points=points1))
+plot(table(o1))
+
+o2 <- replicate(100,resampling(m,f2017,points=points2))
+plot(table(o2))
+
+table(o1)
+table(o2)
+
 ```

### 2020-06-07 inst/f1points.Rmd
typos

@@
 
-```{r,montecarlofirst,cache=TRUE}
+```{r,montecarlosetup, echo=FALSE, cache=FALSE}
 top <- 11
 f2017 <- read.table("formula1_2017.txt",header=TRUE)[seq_len(top),1:20]
-m <- maxp(ordertable2supp(as.ordertable(f2017)))
+#m <- maxp(ordertable2supp(as.ordertable(f2017)))
 
 ## Now work on synthetic dataset generation:
-points1 <- c(25, 18, 15, 12, 10, 8, 6, 4, 2, 1, 0, 0)
-points2 <- c(1, 0,0,0,0)
 resampling <- function(m,f2017,points){
@@
 
-    p_star <- ordertable2points(as.ordertable(ranktable_to_ordertable(random_table)),p)
+    p_star <- ordertable2points(as.ordertable(ranktable_to_ordertable(random_table)),points)
     p_star <- sort(p_star,decreasing=TRUE)
@@
 }
+```
+
+Now
+
+```{r,montecarlosetup2, cache=TRUE}
+f2017 <- read.table("formula1_2017.txt",header=TRUE)[seq_len(top),1:20]
+m <- maxp(ordertable2supp(as.ordertable(f2017)))
+
+points1 <- c(25, 18, 15, 12, 10, 8, 6, 4, 2, 1, 0, 0)
+points2 <- c(1, 0,0,0,0)
 
@@
 table(o2)
-
 ```

### 2020-06-07 inst/f1points.Rmd
another points system

@@
 points2 <- c(1, 0,0,0,0)
+points3 <- c(1, 1,1,1,1,1,0,0,0,0,0)
 
-o1 <- replicate(100,resampling(m,f2017,points=points1))
+o1 <- replicate(400,resampling(m,f2017,points=points1))
+o2 <- replicate(400,resampling(m,f2017,points=points2))
+o3 <- replicate(400,resampling(m,f2017,points=points3))
+
 plot(table(o1))
-
-o2 <- replicate(100,resampling(m,f2017,points=points2))
 plot(table(o2))
+plot(table(o3))
 
@@
 table(o2)
+table(o3)
+```
+
+and, just for comparison, a stupid one:
+
+```{r,cache=TRUE}
+points4 <- c(1, 0,1,2,3,1,0,6,9,0,3)
+
+o4 <- replicate(400,resampling(m,f2017,points=points4))
+table(o4)
 ```

### 2020-06-09 inst/f1points.Rmd
label all chunks and more samples

--- 2020-06-07 inst/f1points.Rmd
@@
 
-```{r,echo=FALSE,print=FALSE}
+```{r,label=showanordertableandhead,echo=FALSE,print=FALSE}
 library(hyper2,quietly=TRUE)
@@
 
-```{r,echo=FALSE}
+```{r,label=printapiechart,echo=FALSE}
 mH <- maxp(H)
@@
 
-```{r,cache=TRUE}
+```{r,label=dosameptest,cache=TRUE}
 samep.test(H,1:2)
@@
 
-```{r,label=calcpoints,echo=TRUE}
+```{r,label=calcpoints,echo=TRUE,cache=TRUE}
 p <- c(25, 18, 15, 12, 10, 8, 6, 4, 2, 1, 0, 0)
@@
 
-```{r,montecarlosetup, echo=FALSE, cache=FALSE}
+```{r,label=montecarlosetup, echo=FALSE, cache=FALSE}
 top <- 11
@@
 
-```{r,montecarlosetup2, cache=TRUE}
+```{r,label=montecarlosetup2, cache=TRUE}
 f2017 <- read.table("formula1_2017.txt",header=TRUE)[seq_len(top),1:20]
 m <- maxp(ordertable2supp(as.ordertable(f2017)))
-
 points1 <- c(25, 18, 15, 12, 10, 8, 6, 4, 2, 1, 0, 0)
 points2 <- c(1, 0,0,0,0)
+points3 <- c(1, 1,1,1,1,1,0,0,0,0,0)
+```
 
-o1 <- replicate(100,resampling(m,f2017,points=points1))
+```{r,label=montecarlo_doer1, cache=TRUE}
+o1 <- replicate(1000,resampling(m,f2017,points=points1))
+```
+
+```{r,label=montecarlo_doer2, cache=TRUE}
+o2 <- replicate(1000,resampling(m,f2017,points=points2))
+
+```{r,label=montecarlo_doer3, cache=TRUE}
+o3 <- replicate(1000,resampling(m,f2017,points=points3))
+```
+
+```{r,label=montecarlo_shower}
 plot(table(o1))
-
-o2 <- replicate(100,resampling(m,f2017,points=points2))
 plot(table(o2))
+plot(table(o3))
 
@@
 table(o2)
+table(o3)
+
+mean(o1)
+mean(o2)
+mean(o3)
+
+M <- rbind(tabulate(o1+1),tabulate(o2+1),tabulate(o3+1))
+dimnames(M) <- list(system=LETTERS[1:3],correct=0:11)
+M
+```
+
+```{r}
+out <- data.frame(rbind(cbind("A",o1),cbind("B",o2),cbind("C",o3)))
+colnames(out) <- c("system","correct")
+head(out)
+tail(out)
+summary(lm(correct~system,data=out))
 ```

### 2020-06-09 inst/f1points.Rmd
complete caching and all results

@@
 p <- c(25, 18, 15, 12, 10, 8, 6, 4, 2, 1, 0, 0)
+points_real <- p
 dp <- ordertable2points(as.ordertable(F1_table_2017[seq_len(top),]),p)
@@
 m <- maxp(ordertable2supp(as.ordertable(f2017)))
-points1 <- c(25, 18, 15, 12, 10, 8, 6, 4, 2, 1, 0, 0)
-points2 <- c(1, 0,0,0,0)
-points3 <- c(1, 1,1,1,1,1,0,0,0,0,0)
 ```
 
-```{r,label=montecarlo_doer1, cache=TRUE}
-o1 <- replicate(1000,resampling(m,f2017,points=points1))
+```{r,label=montecarlosetup3}
+points_real
+o_real <- function(n){replicate(n,resampling(m,f2017,points=points_real))}
 ```
 
-```{r,label=montecarlo_doer2, cache=TRUE}
-o2 <- replicate(1000,resampling(m,f2017,points=points2))
+now do it:
 
-```{r,label=montecarlo_doer3, cache=TRUE}
-o3 <- replicate(1000,resampling(m,f2017,points=points3))
+```{r,label=montecarlorealpointsdoer,cache=TRUE}
+or <- o_real(1000)
 ```
 
-```{r,label=montecarlo_shower}
-plot(table(o1))
-plot(table(o2))
-plot(table(o3))
+Now some more, top n:
 
-table(o1)
-table(o2)
-table(o3)
+```{r,label=defineotop, cache = TRUE}
+o_top <- function(top,n=100){replicate(n,resampling(m,f2017,points=rep(1,top)))}
+```
 
-mean(o1)
-mean(o2)
-mean(o3)
+```{r,label=executeotop1, cache = TRUE}
+ot1 <- o_top(1,n=1000)
+```
 
-M <- rbind(tabulate(o1+1),tabulate(o2+1),tabulate(o3+1))
-dimnames(M) <- list(system=LETTERS[1:3],correct=0:11)
+```{r,label=executeotop2, cache = TRUE}
+ot2 <- o_top(2,n=1000)
+```
+
+```{r,label=executeotop3, cache = TRUE}
+ot3 <- o_top(3,n=1000)
+```
+
+```{r,label=executeotop4, cache = TRUE}
+ot4 <- o_top(4,n=1000)
+```
+
+```{r,label=executeotop5, cache = TRUE}
+ot5 <- o_top(5,n=1000)
+```
+
+now some other schemes:
+
+```{r,label=defineotherschemes}
+o_zip <- function(n=100){replicate(n,resampling(m,f2017,points=1/seq_len(top)))}
+o_exp <- function(p,n=100){replicate(n,resampling(m,f2017,points=p^seq_len(top)))}
+```
+
+and
+
+```{r,label=ozcalculator,cache=TRUE}
+oz <- o_zip(1000)
+```
+
+```{r,label=oecalculator,cache=TRUE}
+oe <- o_exp(0.5,1000)
+```
+
+look:
+
+```{r}
+M <- rbind(
+	tabulate(or +1,nbins=12),
+	tabulate(ot1+1,nbins=12),
+	tabulate(ot2+1,nbins=12),
+	tabulate(ot3+1,nbins=12),
+	tabulate(ot4+1,nbins=12),
+	tabulate(ot5+1,nbins=12),
+	tabulate(oz +1,nbins=12),
+	tabulate(oe +1,nbins=12)
+	)
+dimnames(M) <- list(system=c("real",paste("top",seq_len(5)),"oz","oe"),correct=0:11)
 M
@@
 
-```{r}
-out <- data.frame(rbind(cbind("A",o1),cbind("B",o2),cbind("C",o3)))
-colnames(out) <- c("system","correct")
-head(out)
-tail(out)
-summary(lm(correct~system,data=out))
-```
-
 \bibliography{chess}

### 2020-06-10 inst/f1points.Rmd
more rewriting

@@
 We can use this observation to assess a given points system: using
-Monte-Carlo techniques we can generate an order table \emph{in
-silico}, using estimated driver strengths from the 2017 table, and
-then compare rankings generated by estimated \citeauthor{plackett1975}
-strengths with rankings generated by the points system.
-
-# Monte Carlo assumptions
+random sampling techniques we can generate an order table \emph{in
+silico}, using estimated driver strengths from the 2017 table.  Given
+this table we can then then compare rankings generated by maximum
+likelihood \citeauthor{plackett1975} strengths with rankings generated
+by any particular points system; this furnishes an objective assessment of a
 
@@
 
-Now
-
-```{r,label=montecarlosetup2, cache=TRUE}
+```{r,label=montecarlosetup2, cache=TRUE,echo=FALSE}
 f2017 <- read.table("formula1_2017.txt",header=TRUE)[seq_len(top),1:20]
@@
 
-```{r,label=montecarlosetup3}
-points_real
+```{r,label=montecarlosetup3,echo=FALSE}
 o_real <- function(n){replicate(n,resampling(m,f2017,points=points_real))}
@@
 
-now do it:
-
-```{r,label=montecarlorealpointsdoer,cache=TRUE}
+```{r,label=montecarlorealpointsdoer,cache=TRUE,echo=FALSE}
 or <- o_real(1000)
@@
 
-Now some more, top n:
-
-```{r,label=defineotop, cache = TRUE}
+```{r,label=defineotop, cache = TRUE, echo=FALSE}
 o_top <- function(top,n=100){replicate(n,resampling(m,f2017,points=rep(1,top)))}
@@
 
-```{r,label=executeotop1, cache = TRUE}
+```{r,label=executeotop1, cache = TRUE, echo=FALSE}
 ot1 <- o_top(1,n=1000)
@@
 
-```{r,label=executeotop2, cache = TRUE}
+```{r,label=executeotop2, cache = TRUE,echo=FALSE}
 ot2 <- o_top(2,n=1000)
@@
 
-```{r,label=executeotop3, cache = TRUE}
+```{r,label=executeotop3, cache = TRUE,echo=FALSE}
 ot3 <- o_top(3,n=1000)
@@
 
-```{r,label=executeotop4, cache = TRUE}
+```{r,label=executeotop4, cache = TRUE,echo=FALSE}
 ot4 <- o_top(4,n=1000)
@@
 
-```{r,label=executeotop5, cache = TRUE}
+```{r,label=executeotop5, cache = TRUE,echo=FALSE}
 ot5 <- o_top(5,n=1000)
@@
 
-now some other schemes:
-
-```{r,label=defineotherschemes}
+```{r,label=defineotherschemes, echo=FALSE}
 o_zip <- function(n=100){replicate(n,resampling(m,f2017,points=1/seq_len(top)))}
@@
 
-and
-
-```{r,label=ozcalculator,cache=TRUE}
+```{r,label=ozcalculator,cache=TRUE, echo=FALSE}
 oz <- o_zip(1000)
@@
 
-```{r,label=oecalculator,cache=TRUE}
+```{r, label=oecalculator, cache=TRUE, echo=FALSE}
 oe <- o_exp(0.5,1000)
@@
 
-look:
-
-```{r}
+```{r, label=makeManddimlabel, echo=FALSE}
 M <- rbind(
@@
 	)
-dimnames(M) <- list(system=c("real",paste("top",seq_len(5)),"oz","oe"),correct=0:11)
+dimnames(M) <- list(system=c("real",paste("top",seq_len(5),sep=""),"zipf","exp"),correct=0:11)
+```
+
+```{r,showM}
 M
@@
 
+```{r,showcumsumM}
+M2 <- t(apply(M[,12:1],1,cumsum))[,12:1]
+dimnames(M2) <- list(system=rownames(M),correct=paste(">=",colnames(M2)))
+M2
+matplot(t(M2[,12:1]),type='b')
+```
+
 \bibliography{chess}

### 2020-06-10 inst/f1points.Rmd
more work (this one processes correctly)

@@
 likelihood \citeauthor{plackett1975} strengths with rankings generated
-by any particular points system; this furnishes an objective assessment of a
+by any particular points system; this furnishes an objective
+assessment of a
 
@@
 
+# more efficient ways
+
+```{r,label=multiplemontecarlosetup, echo=FALSE, cache=FALSE}
+top <- 11
+f2017 <- read.table("formula1_2017.txt",header=TRUE)[seq_len(top),1:20]
+#m <- maxp(ordertable2supp(as.ordertable(f2017)))
+
+## Now work on synthetic dataset generation:
+resampling_multiple <- function(m,f2017,pointslist){
+    random_table <- rrank(n=ncol(f2017), p=m)
+    rownames(random_table) <- colnames(f2017)
+    ## "random_table" is a random table; now calculate lstar and pstar for
+    ## likelihood ranks and points ranks respectively:
+
+    l_star <- maxp(ordertable2supp(ranktable_to_ordertable(random_table)))
+    l_star[] <- seq_along(l_star)
+    ## l_star is the likelihood order statistic.
+
+	goodnesses <- seq_along(pointslist)
+	for(i in seq_along(pointslist)){
+		p_star <- ordertable2points(as.ordertable(ranktable_to_ordertable(random_table)),pointslist[[i]])
+		p_star <- sort(p_star,decreasing=TRUE)
+		p_star[] <- seq_along(p_star)
+		goodness <- sum(cumprod(names(p_star)==names(l_star)))
+		goodnesses[i] <- goodness
+		}
+	return(goodnesses)
+}
+pointslist <- list(
+	real = points_real,
+	line = top:1,
+	top1 = rep(1,1),
+	top2 = rep(1,2),
+	top3 = rep(1,3),
+	top4 = rep(1,4),
+	zipf = 1/seq_len(top),
+	exp2 = 1/2^seq_along(top),	
+	exp3 = 1/3^seq_along(top),	
+	exp4 = 1/4^seq_along(top)
+)
+```
+
+```{r,label=multiplemontecarloexecute, cache=TRUE,echo=TRUE}
+f2017 <- read.table("formula1_2017.txt",header=TRUE)[seq_len(top),1:20]
+m <- maxp(ordertable2supp(as.ordertable(f2017)))
+OO <- replicate(500+500,resampling_multiple(m,f2017,pointslist=pointslist))
+```
+
+```{r,label=multiplemontecarlotabulation}
+M3 <- t(apply(OO,1,function(o){tabulate(o+1,nbins=12)}))
+dimnames(M3) <- list(system=names(pointslist),correct=0:11)
+M3
+Mc <- t(apply(M3,1,cumsum))
+Mc
+```
+
+and visuial:
+
+```{r,label=plotcumulativeM,fig=TRUE}
+matplot(t(Mc),type="b",lty=1,pch=16,col=rainbow(nrow(Mc)))
+legend("bottomright",lty=1,pch=16,col=rainbow(nrow(Mc)),legend=letters[seq_len(nrow(Mc))])
+plot(points_real)
+grid()
+```
+
 \bibliography{chess}

### 2020-06-11 inst/f1points.Rmd
more work

@@
 additively.  The current points system, for example, is
-(25-18-15-12-10-8-6-4-2-1) [that is, the winner is awarded 25 points,
+(25,18,15,12,10,8,6,4,2,1) [that is, the winner is awarded 25 points,
 second place 18 points, and so on], although since 2019 the fastest
@@
 and examples would include modern pentathlon, curling, chess, rowing,
-and figure skating; all have the common feature of interpretation of
-an order statistic to a points.  Further, we see points systems
-frequently in the wider context of competitive situations such as the
-Eurovision Song Contest, Australian MasterChef, University rankings,
-educational league tables, and car safety ratings.
+and figure skating; all have the common feature of translating ranks
+into scores which combine additively to generate an overall ranking.
+Further, we see points systems frequently in the wider context of
+competitive situations such as the Eurovision Song Contest, Australian
+MasterChef, University rankings, educational league tables, car safety
+ratings, and many others.
 
@@
 
+## Formula 1 racing: points systems
+
+The points system of Formula 1 Motor racing is a contentious topic;
+there have been many changes since systems has changed many times,
+although some changes involved only bonus points (typically, one point
+for fastest lap), which are not considered here.  The inaugural
+system, used from 1950 to 1959 was $(8,6,4,3,2)$, thus crediting only
+the first five finishers, and the current system of
+$(25,18,15,12,10,8,6,4,2,1)$ credits the first 10.  Arguably these two
+systems could introduce different rational behaviour under zero-sum
+assumptions: if, in a race, a driver knows he will place seventh under
+a low-risk strategy but may place sixth by dint of driving more
+aggressively, the low-risk strategy might be rational under the first
+points system (which does not reward the extra ranking), but not under
+the second, which does.  Still, it is reasonable to assume that each
+driver strives to maximise his rank.
+
 # Formula 1 dataset
@@
 We see the Hamilton has the largest estimated strength, although it
-should be noted that a likelihood ratio test [`samep.test()`] does not
-reject the null of $p_\mathrm{Hamilton}=p_\mathrm{Vettel}$ ($p=0.09$).
+should be noted that a likelihood ratio test [`samep.test()`, supplied
+with the package] does not reject the null of
+$p_\mathrm{Hamilton}=p_\mathrm{Vettel}$ , with a likelihood ratio of
+$e^{1.5}\simeq 4.48$, corresponding to an asymptotic $p$-value of
+about $0.08$.
 
-```{r,label=dosameptest,cache=TRUE}
-samep.test(H,1:2)
+```{r,label=dosameptest, echo=FALSE, print=FALSE, cache=TRUE}
+ignore <- samep.test(H,1:2)
 ```
@@
 
-```{r,label=calcpoints,echo=TRUE,cache=TRUE}
+```{r, label=calcpoints, echo=FALSE, cache=TRUE}
 p <- c(25, 18, 15, 12, 10, 8, 6, 4, 2, 1, 0, 0)
@@
 are are respectively first and second according to both ranking
-ranking procedures; but Raikkonen and Bottas are third and fourth, and
-fourth and third, according to points and likelihood respectively.
-The degree of agreement between the two systems might be quantified as
-"two", for the positions of the top two drivers agree.
+ranking procedures; but RAikkOnen 
+*crap problem: latex ignores  diaeresis directive* and Bottas are third and fourth, and fourth and
+third, according to points and likelihood respectively.  The degree of
+agreement between the two systems might be quantified as "two", for
+the positions of the top two drivers agree.
 
-We can use this observation to assess a given points system: using
-random sampling techniques we can generate an order table \emph{in
+We can now assess objectively a given points system: using random
+sampling techniques we can repeatedly generate an order table \emph{in
 silico}, using estimated driver strengths from the 2017 table.  Given
-this table we can then then compare rankings generated by maximum
-likelihood \citeauthor{plackett1975} strengths with rankings generated
-by any particular points system; this furnishes an objective
-assessment of a
+a synthetic table, calculate drivers' maximum likelihood
+\citeauthor{plackett1975} strengths, and also their points awarded
+according to any given points system.  We then compare rankings
+generated by the \citeauthor{plackett1975} strengths and the points
+awarded and note the degree of agreement between the two.  This
+furnishes an objective assessment of the points system used.
 
-```{r,label=montecarlosetup, echo=FALSE, cache=FALSE}
+# Numerical results
+
+Here we follow the above protocol to assess different plausible points
+systems.  There are a number of plausible points systems that might be
+used:
+
+\begin{itemize}
+\item winner takes all $(1,0,0,\ldots)$
+\item Zipfs law $(1,1/2,1/3,1/4,\ldots)$
+\item exponential decay $(1,r,r^2,r^3,\ldots)$
+\item the actual real one $(25,18,15,12,10,8,6,4,2,1)$
+\item equal points to the top $n$ competitors $\underbrace{1,1,\ldots
+,1}_{n},0,0,\ldots$ (noting that $n=1$ corresponds to winner takes all)
+\item linear decrease $(n,n-1,n-2,\ldots,3,2,1,0)$
+\end{itemize}
+
+```{r,label=definemultiplemontecarlo, echo=FALSE}
 top <- 11
 f2017 <- read.table("formula1_2017.txt",header=TRUE)[seq_len(top),1:20]
-#m <- maxp(ordertable2supp(as.ordertable(f2017)))
-
-## Now work on synthetic dataset generation:
-resampling <- function(m,f2017,points){
-    random_table <- rrank(n=ncol(f2017), p=m)
-    rownames(random_table) <- colnames(f2017)
-    ## "random_table" is a random table; now calculate lstar and pstar for
-    ## likelihood ranks and points ranks respectively:
-
-    l_star <- maxp(ordertable2supp(ranktable_to_ordertable(random_table)))
-    l_star[] <- seq_along(l_star)
-    ## l_star is the likelihood order statistic.
-
-    p_star <- ordertable2points(as.ordertable(ranktable_to_ordertable(random_table)),points)
-    p_star <- sort(p_star,decreasing=TRUE)
-    p_star[] <- seq_along(p_star)
-
-    goodness <- sum(cumprod(names(p_star)==names(l_star)))
-    return(goodness)
-}
-```
-
-```{r,label=montecarlosetup2, cache=TRUE,echo=FALSE}
-f2017 <- read.table("formula1_2017.txt",header=TRUE)[seq_len(top),1:20]
 m <- maxp(ordertable2supp(as.ordertable(f2017)))
-```
-
-```{r,label=montecarlosetup3,echo=FALSE}
-o_real <- function(n){replicate(n,resampling(m,f2017,points=points_real))}
-```
-
-```{r,label=montecarlorealpointsdoer,cache=TRUE,echo=FALSE}
-or <- o_real(1000)
-```
-
-```{r,label=defineotop, cache = TRUE, echo=FALSE}
-o_top <- function(top,n=100){replicate(n,resampling(m,f2017,points=rep(1,top)))}
-```
-
-```{r,label=executeotop1, cache = TRUE, echo=FALSE}
-ot1 <- o_top(1,n=1000)
-```
-
-```{r,label=executeotop2, cache = TRUE,echo=FALSE}
-ot2 <- o_top(2,n=1000)
-```
-
-```{r,label=executeotop3, cache = TRUE,echo=FALSE}
-ot3 <- o_top(3,n=1000)
-```
-
-```{r,label=executeotop4, cache = TRUE,echo=FALSE}
-ot4 <- o_top(4,n=1000)
-```
-
-```{r,label=executeotop5, cache = TRUE,echo=FALSE}
-ot5 <- o_top(5,n=1000)
-```
-
-```{r,label=defineotherschemes, echo=FALSE}
-o_zip <- function(n=100){replicate(n,resampling(m,f2017,points=1/seq_len(top)))}
-o_exp <- function(p,n=100){replicate(n,resampling(m,f2017,points=p^seq_len(top)))}
-```
-
-```{r,label=ozcalculator,cache=TRUE, echo=FALSE}
-oz <- o_zip(1000)
-```
-
-```{r, label=oecalculator, cache=TRUE, echo=FALSE}
-oe <- o_exp(0.5,1000)
-```
-
-```{r, label=makeManddimlabel, echo=FALSE}
-M <- rbind(
-	tabulate(or +1,nbins=12),
-	tabulate(ot1+1,nbins=12),
-	tabulate(ot2+1,nbins=12),
-	tabulate(ot3+1,nbins=12),
-	tabulate(ot4+1,nbins=12),
-	tabulate(ot5+1,nbins=12),
-	tabulate(oz +1,nbins=12),
-	tabulate(oe +1,nbins=12)
-	)
-dimnames(M) <- list(system=c("real",paste("top",seq_len(5),sep=""),"zipf","exp"),correct=0:11)
-```
-
-```{r,showM}
-M
-```
-
-```{r,showcumsumM}
-M2 <- t(apply(M[,12:1],1,cumsum))[,12:1]
-dimnames(M2) <- list(system=rownames(M),correct=paste(">=",colnames(M2)))
-M2
-matplot(t(M2[,12:1]),type='b')
-```
-
-# more efficient ways
-
-```{r,label=multiplemontecarlosetup, echo=FALSE, cache=FALSE}
-top <- 11
-f2017 <- read.table("formula1_2017.txt",header=TRUE)[seq_len(top),1:20]
-#m <- maxp(ordertable2supp(as.ordertable(f2017)))
 
@@
 
-```{r,label=multiplemontecarloexecute, cache=TRUE,echo=TRUE}
+```{r,label=multiplemontecarloexecute, cache=TRUE,echo=FALSE}
 f2017 <- read.table("formula1_2017.txt",header=TRUE)[seq_len(top),1:20]
@@
 
-```{r,label=multiplemontecarlotabulation}
+```{r,label=multiplemontecarlotabulation,echo=FALSE}
 M3 <- t(apply(OO,1,function(o){tabulate(o+1,nbins=12)}))
@@
 
-and visuial:
+and visual:
 
@@
 
+```{r,makeatable}
+table(real_winner_ok=OO[1,]>0,wta_winner_ok=OO[3,]>0)
+table(real_all=OO[1,]==11,wta_all=OO[3,]==11)
+write.csv(OO,file="~/f.txt")
+```
+
+# Discussion and conclusions
+
 \bibliography{chess}

### 2020-06-12 inst/f1points.Rmd
more work

@@
 are are respectively first and second according to both ranking
-ranking procedures; but RAikkOnen 
-*crap problem: latex ignores  diaeresis directive* and Bottas are third and fourth, and fourth and
-third, according to points and likelihood respectively.  The degree of
-agreement between the two systems might be quantified as "two", for
-the positions of the top two drivers agree.
+ranking procedures; but R\"{a}ikk\"{o}nen *yes I know it is wrong:
+rmarkdown is defective* and Bottas are third and fourth, and fourth
+and third, according to points and likelihood respectively.  The
+degree of agreement between the two systems might be quantified as
+"two", for the positions of the top two drivers agree.
 
-We can now assess objectively a given points system: using random
-sampling techniques we can repeatedly generate an order table \emph{in
-silico}, using estimated driver strengths from the 2017 table.  Given
-a synthetic table, calculate drivers' maximum likelihood
+Observing that both the likelihood ranking and the points ranking are
+random variables suggests a method whereby we can objectively assess a
+given points system.  Using sampling techniques we can repeatedly
+generate an order table \emph{in silico}, using estimated driver
+strengths from the observed 2017 table.  For each of, say, 1000 such
+synthetic tables, calculate drivers' maximum likelihood
 \citeauthor{plackett1975} strengths, and also their points awarded
@@
 	real = points_real,
-	line = top:1,
 	top1 = rep(1,1),
@@
 	top4 = rep(1,4),
+	top5 = rep(1,5),
+	top6 = rep(1,6),
+	top7 = rep(1,7),
+	top8 = rep(1,8),
+	top9 = rep(1,9),
+	topt = rep(1,10),  # effectively punishing the last finisher
+   	lin1 = 1,   # same as top1
+	lin2 = 2:1,
+	lin3 = 3:1,
+	lin4 = 4:1,
+	lin5 = 5:1,
+	lin6 = 6:1,
+	lin7 = 7:1,
+	lin8 = 8:1,
+	lin9 = 9:1,
+	lint = 10:1,
+	line = 11:1,
 	zipf = 1/seq_len(top),
-	exp2 = 1/2^seq_along(top),	
-	exp3 = 1/3^seq_along(top),	
-	exp4 = 1/4^seq_along(top)
+	expa = 1/1.01^seq_len(top),
+	expb = 1/1.10^seq_len(top),
+	expc = 1/1.20^seq_len(top),
+	expd = 1/1.50^seq_len(top),
+	exp2 = 1/2^seq_len(top),
+	exp3 = 1/3^seq_len(top),
+	exp4 = 1/4^seq_len(top)
 )
@@
 m <- maxp(ordertable2supp(as.ordertable(f2017)))
-OO <- replicate(500+500,resampling_multiple(m,f2017,pointslist=pointslist))
+OO <- replicate(1000,resampling_multiple(m,f2017,pointslist=pointslist))
 ```
@@
 
+`r sin(1)`
+
 \bibliography{chess}

### 2020-06-18 inst/f1points.Rmd
more work

@@
 
+Here we follow the above protocol to assess different points systems.
+There are a number of plausible points systems that might be used:
+
+\begin{itemize}
+\item Exponential decay $(1,r,r^2,r^3,\ldots)$ for some fixed $r<1$
+\item Equal points to the top $n$ competitors
+$(\underbrace{1,1,\ldots ,1}_{n},0,0,\ldots)$, noting that
+$n=1$ corresponds to a ``winner takes all''
+\item Linear decrease $(r,r-1,r-2,\ldots,3,2,1,0)$ for integer $0<r<n$
+\item Zipf's law $(1,1/2,1/3,1/4,\ldots)$
+\item The current Formula 1 system $(25,18,15,12,10,8,6,4,2,1)$
+\end{itemize}
+
+Applying Zipf's law, for example, to the 2017 table we would rank the
+drivers as follows:
+
+```{r,label=zipfpointsshower,echo=FALSE}
+zipf <- ordertable2points(ranktable_to_ordertable(wikitable_to_ranktable(F1_table_2017)),1/seq_len(top))
+zipf <- round(zipf[seq_len(top)],2)
+zipf <- sort(zipf,decreasing=TRUE)
+zipf[] <- seq_along(zipf)
+zipf
+```
+
+compare the real:
+
+```{r,label=realpointsshower,echo=FALSE}
+real <- ordertable2points(ranktable_to_ordertable(wikitable_to_ranktable(F1_table_2017)),p)
+real <- round(real[seq_len(top)],2)
+real <- sort(real,decreasing=TRUE)
+real[] <- seq_along(real)
+real
+```
+
+Thus these two systems agree on the first three places but fourth is
+awarded to Ricciardo and Raikkonen respectively.
+
 # Numerical results
 
-Here we follow the above protocol to assess different plausible points
-systems.  There are a number of plausible points systems that might be
-used:
-
-\begin{itemize}
-\item winner takes all $(1,0,0,\ldots)$
-\item Zipfs law $(1,1/2,1/3,1/4,\ldots)$
-\item exponential decay $(1,r,r^2,r^3,\ldots)$
-\item the actual real one $(25,18,15,12,10,8,6,4,2,1)$
-\item equal points to the top $n$ competitors $\underbrace{1,1,\ldots
-,1}_{n},0,0,\ldots$ (noting that $n=1$ corresponds to winner takes all)
-\item linear decrease $(n,n-1,n-2,\ldots,3,2,1,0)$
-\end{itemize}
+We now assess the various points systems discussed above, using 1000
+_in silico_ trials:
 
@@
 Mc
+Md <- t(apply(M3[,-1],1,cumsum))
+colnames(Md) <- paste("<=",colnames(Md),sep="")
+Md
 ```

### 2020-06-19 inst/f1points.Rmd
more work

@@
 
+Have a quick look at the result:
+
+```{r,label=quicklookatOO}
+dim(OO)
+rownames(OO) <- names(pointslist)
+plot(apply(OO,1,mean),ylab='mean number of ranks correct')
+```
+
+And more work:
+
 ```{r,label=multiplemontecarlotabulation,echo=FALSE}
@@
 dimnames(M3) <- list(system=names(pointslist),correct=0:11)
-M3
-Mc <- t(apply(M3,1,cumsum))
-Mc
-Md <- t(apply(M3[,-1],1,cumsum))
+head(M3)
+```
+
+now cumulative:
+
+```{r,label=cummulttab,echo=FALSE}
+Md <- t(apply(M3,1,cumsum))
 colnames(Md) <- paste("<=",colnames(Md),sep="")
@@
 ```{r,label=plotcumulativeM,fig=TRUE}
-matplot(t(Mc),type="b",lty=1,pch=16,col=rainbow(nrow(Mc)))
-legend("bottomright",lty=1,pch=16,col=rainbow(nrow(Mc)),legend=letters[seq_len(nrow(Mc))])
+plot(Md[,1],ylab='number wrong')
+plot(1000-Md[,11],ylab='number perfect')
+matplot(t(Md),type="b",lty=1,pch=16,col=rainbow(nrow(Md)))
+legend("bottomright",lty=1,pch=16,col=rainbow(nrow(Md)),legend=letters[seq_len(nrow(Md))])
 plot(points_real)
@@
 
-# Discussion and conclusions
+Consider the first column of the table above which shows the number of
+virtual races in which each scoring system gets the winner incorrect;
+a good points system has a low value here.  Note the peculiar finding
+that `top3` has by far the lowest number, a finding which I have no
+explanation for.
+
+however,
 

### 2020-06-20 inst/f1points.Rmd
rewording

@@
 fictional (reified) entities whose nonzero Bradley-Terry strength
-helps certain competitors or sets of competitors.  The canonical
-example would be the home-ground advantage in association football.
-If players (teams) $1,2$ with strengths $p_1,p_2$ compete, and if our
-observation were $a$ home wins and $b$ away wins for team $1$, and $c$
-home wins and $d$ away wins for team $2$, then a suitable likelihood
-function would be
+helps certain competitors or sets of competitors under certain
+conditions.  The canonical example would be the home-ground advantage
+in association football.  If players (teams) $1,2$ with strengths
+$p_1,p_2$ compete, and if our observation were $a$ home wins and $b$
+away wins for team $1$, and $c$ home wins and $d$ away wins for team
+$2$, then a suitable likelihood function would be
 
@@
 points_real <- p
+points_inaugural <- c(8,6,4,3,2)
 dp <- ordertable2points(as.ordertable(F1_table_2017[seq_len(top),]),p)
@@
 	real = points_real,
+	inau = points_inaugural,
 	top1 = rep(1,1),
@@
 	exp3 = 1/3^seq_len(top),
-	exp4 = 1/4^seq_len(top)
+	exp4 = 1/4^seq_len(top),
+	fis1 = c(1,1,1),
+	fis2 = c(1,1,0.9),
+	fis3 = c(1,1,1,0.1),
+	fis4 = c(1,1,0.9,0.1),
+	fis5 = c(1,1,0.5),
+	fis6 = c(1,1,1,0.5),
+	fis7 = c(1,1,0.5,0.5)
 )
@@
 ```{r,label=multiplemontecarloexecute, cache=TRUE,echo=FALSE}
+set.seed(9)
 f2017 <- read.table("formula1_2017.txt",header=TRUE)[seq_len(top),1:20]

### 2020-06-21 inst/f1points.Rmd
restructuring

@@
 abstract: >
-  The points scoring system of Formula 1 motor racing is a long-standing
+  The point scoring system of Formula 1 motor racing is a long-standing
   contention among fans and competitors, with changes to the points
@@
 Formula 1 racing is an important and prestigious motor sport
-\citep{moul2009}.  Season ranking is based on a points system; points
-are awarded to drivers and teams based on race order, and accumulate
-additively.  The current points system, for example, is
-(25,18,15,12,10,8,6,4,2,1) [that is, the winner is awarded 25 points,
-second place 18 points, and so on], although since 2019 the fastest
-lap attracts one point.
+\citep{moul2009}.  Season ranking is based on a points allocation
+system wherein competitors are awarded points based on race finishing
+order; points accumulate additively.  The overall competition winner
+is the competitor who accumulates the most points after the final
+race.
+
+However, in the case of Formula 1 Motor Racing, the points system is
+the subject of much controversy, having changed often since the
+competition's inauguration in 1950 when the points allocation, used
+from 1950 to 1959 was $(8,6,4,3,2)$, thus crediting only the first
+five finishers.  As of 2020, the current system of
+$(25,18,15,12,10,8,6,4,2,1)$ credits the first 10.  Arguably these two
+systems could introduce different rational behaviour under zero-sum
+assumptions: if, in a race, a driver knows he will place seventh under
+a low-risk strategy but may place sixth by dint of driving more
+aggressively, the low-risk strategy might be rational under the first
+points system (which does not reward the extra ranking), but not under
+the second, which does.  Still, it is reasonable to assume that each
+driver strives to maximise his rank.
 
@@
 
-The points system of Formula 1 Motor racing is a contentious topic;
-there have been many changes since systems has changed many times,
-although some changes involved only bonus points (typically, one point
-for fastest lap), which are not considered here.  The inaugural
-system, used from 1950 to 1959 was $(8,6,4,3,2)$, thus crediting only
-the first five finishers, and the current system of
-$(25,18,15,12,10,8,6,4,2,1)$ credits the first 10.  Arguably these two
-systems could introduce different rational behaviour under zero-sum
-assumptions: if, in a race, a driver knows he will place seventh under
-a low-risk strategy but may place sixth by dint of driving more
-aggressively, the low-risk strategy might be rational under the first
-points system (which does not reward the extra ranking), but not under
-the second, which does.  Still, it is reasonable to assume that each
-driver strives to maximise his rank.
-
 # Formula 1 dataset
@@
 top <- 11
+points_inaugural <- c(8,6,4,3,2)
 ```
@@
 
+## Points systems
+
 Here we follow the above protocol to assess different points systems.
@@
 \begin{itemize}
+\item Equal points to the top $n$ competitors
+$(\underbrace{1,1,\ldots ,1}_{r},0,0,\ldots)$, noting that
+$n=r$ corresponds to a ``winner takes all'' system
+\item Linear decrease $(r,r-1,r-2,\ldots,3,2,1,0)$ for integer $0<r<n$
 \item Exponential decay $(1,r,r^2,r^3,\ldots)$ for some fixed $r<1$
-\item Equal points to the top $n$ competitors
-$(\underbrace{1,1,\ldots ,1}_{n},0,0,\ldots)$, noting that
-$n=1$ corresponds to a ``winner takes all''
-\item Linear decrease $(r,r-1,r-2,\ldots,3,2,1,0)$ for integer $0<r<n$
 \item Zipf's law $(1,1/2,1/3,1/4,\ldots)$
 \item The current Formula 1 system $(25,18,15,12,10,8,6,4,2,1)$
+\item The inaugural Formula 1 system $(8,6,4,3,2)$
 \end{itemize}
 
-Applying Zipf's law, for example, to the 2017 table we would rank the
-drivers as follows:
+Applying the current points system, for example, to the 2017 table we
+would rank the drivers as follows:
+
+```{r,label=realpointsshower,echo=FALSE}
+real <- ordertable2points(ranktable_to_ordertable(wikitable_to_ranktable(F1_table_2017)),p)
+real <- round(real[seq_len(top)],2)
+real <- sort(real,decreasing=TRUE)
+real[] <- seq_along(real)
+real
+```
+
+(this happens to be identical to the ranking after the extra point was
+awarded for fastest lap).  However, if we were to use Zipf's law, the
+ranking would be:
 
@@
 
-compare the real:
+Thus these two systems agree on the first three places but fourth is
+awarded to Raikkonen under the F1 system and Ricciardo under Zipf.
 
-```{r,label=realpointsshower,echo=FALSE}
-real <- ordertable2points(ranktable_to_ordertable(wikitable_to_ranktable(F1_table_2017)),p)
-real <- round(real[seq_len(top)],2)
-real <- sort(real,decreasing=TRUE)
-real[] <- seq_along(real)
-real
+Now compare the likelihood ranking:
+
+```{r,label=likelihoodrankingshower,echo=FALSE}
+jj <- mH
+jj <- sort(jj,decreasing=TRUE)
+jj[] <- seq_along(jj)
+jj
 ```
-
-Thus these two systems agree on the first three places but fourth is
-awarded to Ricciardo and Raikkonen respectively.
 
@@
 We now assess the various points systems discussed above, using 1000
-_in silico_ trials:
+_in silico_ trials.
 
@@
 
-however,
+# Discussion
 
-`r sin(1)`
+Points systems are a source of lively debate from many perspectives.
+The point of a point allocation system is to rank competitors in an
+objective way; but
+
+Here I treat total points scored as a random variable and compare
+different point allocation schemes against both one another and
+
+but until now there is no objective way of comparing points systems.
+if one considers order tables to be a random variable.  Because the
+numerical value of points awarded for first, second, third etc, does
+not affect driver behaviour it is a reasonable requirement that
+
+Here, I assess
+several plausible points systems including the one actually used in
+Formula 1.  By several criteria, the
 

### 2020-06-23 inst/f1points.Rmd
more work

@@
 
-## Formula 1 racing: points systems
-
 # Formula 1 dataset
@@
 effectively ``steal'' strength from Vettel, and generally ``give'' it
-to Hamilton or Bottas.}.  Although it has many terms, the first few
-are
+to Hamilton or Bottas.}.  Although it has many terms, the overall loglikelihood 
+expression is of the general form
 
@@
 H <- ordertable2supp(F1_table_2017[seq_len(top),])
-head(H,5)
 ```
 
+\newcommand{\pham}{p_\mathrm{Ham}}
+\newcommand{\pvet}{p_\mathrm{Vet}}
+\newcommand{\pbot}{p_\mathrm{Bot}}
+\newcommand{\prai}{p_\mathrm{Rai}}
+\newcommand{\pric}{p_\mathrm{Ric}}
+\newcommand{\pver}{p_\mathrm{Ver}}
+\newcommand{\pper}{p_\mathrm{Per}}
+\newcommand{\poco}{p_\mathrm{Oco}}
+\newcommand{\psai}{p_\mathrm{Sai}}
+\newcommand{\phul}{p_\mathrm{Hul}}
+\newcommand{\pmas}{p_\mathrm{Mas}}
+
+\begin{equation}
+\frac{\pham^{20}\,\pmas^{16}\,\pbot^{19}\ldots
+}{\parbox{4in}{$(\pper+\poco)(\pver+\pper+\psai+\phul+\pmas)\\
+\rule{10mm}{0mm}(\pric+\pper+\pmas)(\psai+\pmas)\\ \rule{20mm}{0mm}
+         (\pver + \pper+ \poco + \psai+ \phul)\ldots
+	 $}}
+ \end{equation}
+
 Finding the maximum likelihood estimate for the players' strengths is
-straightforward:
+straightforward.  The `hyper2` package includes a suite of numerical
+optimization routines, and because they have access to derivatives,
+convergence is rapid.
 
@@
 mH <- maxp(H)
-print(mH)
+mH
 pie(maxp(H)[c(1,6,11,2,7,3,8,4,9,5,10)])
@@
 
-We see the Hamilton has the largest estimated strength, although it
-should be noted that a likelihood ratio test [`samep.test()`, supplied
-with the package] does not reject the null of
-$p_\mathrm{Hamilton}=p_\mathrm{Vettel}$ , with a likelihood ratio of
-$e^{1.5}\simeq 4.48$, corresponding to an asymptotic $p$-value of
-about $0.08$.
+We see the Hamilton has the largest estimated strength (at about
+0.39), followed by Vettel and then Bottas at about 0.2 and 0.13
+respectively.  However, it should be noted that a likelihood ratio
+test [`samep.test()`, supplied with the package] does not reject the
+null that Hamilton and Vettel have the same strength ($H_0\colon
+\pham= \pvet$) with a likelihood ratio of $e^{1.5}\simeq 4.48$,
+corresponding (by Wilks's theorem) to an asymptotic $p$-value of about
+$0.08$.
 
@@
 ```
+
+(note carefully that the historically correct points awarded to the
+drivers differs from that calculated here.  That is for two reasons:
+firstly, ``fastest lap'' points are not included here, and also the
+truncation of the order table to the first 11 drivers can increase the
+rank of a driver if non-first-11 drivers are placed).
 
@@
 
-Have a quick look at the result:
+now some summaries:
 
-```{r,label=quicklookatOO}
-dim(OO)
-rownames(OO) <- names(pointslist)
-plot(apply(OO,1,mean),ylab='mean number of ranks correct')
+```{r,label=somesummaries}
+summarytable <- data.frame(
+means = apply(OO,1,mean),
+winner_correct = rowSums(OO>1),
+all_correct = rowSums(OO==top))
+rownames(summarytable) <- names(pointslist)
+summarytable
 ```
 
-And more work:
-
-```{r,label=multiplemontecarlotabulation,echo=FALSE}
-M3 <- t(apply(OO,1,function(o){tabulate(o+1,nbins=12)}))
-dimnames(M3) <- list(system=names(pointslist),correct=0:11)
-head(M3)
-```
-
-now cumulative:
-
-```{r,label=cummulttab,echo=FALSE}
-Md <- t(apply(M3,1,cumsum))
-colnames(Md) <- paste("<=",colnames(Md),sep="")
-Md
+```{r,fig=TRUE}
+typecol <- c(rep("red",2),rep("orange",10),rep("blue",11),rep("orange",1),rep("black",7),rep("gray",7))
+plot(summarytable[,1],ylab='mean',pch=16,col=typecol)
+plot(summarytable[,2],ylab='prob of winner correct',pch=16,col=typecol)
+plot(summarytable[,3],ylab='prob. of total rank order correct',pch=16,col=typecol)
 ```
@@
 ```{r,label=plotcumulativeM,fig=TRUE}
-plot(Md[,1],ylab='number wrong')
-plot(1000-Md[,11],ylab='number perfect')
-matplot(t(Md),type="b",lty=1,pch=16,col=rainbow(nrow(Md)))
-legend("bottomright",lty=1,pch=16,col=rainbow(nrow(Md)),legend=letters[seq_len(nrow(Md))])
-plot(points_real)
-grid()
+o <- t(apply(1+OO,1,tabulate))
+#matplot(0:11,apply(o[,12:1],1,cumsum)/1e3,type='b',lty=1,pch=16,ylab='probability',axes=FALSE,xlab='number incorrect')
+jj <- c(0,2,4,6,8,11)
+#axis(1,pos=0,at=jj,labels=c("0",paste("<=",jj[-1],sep="")))
+#axis(2)
 ```
@@
 ```{r,makeatable}
-table(real_winner_ok=OO[1,]>0,wta_winner_ok=OO[3,]>0)
-table(real_all=OO[1,]==11,wta_all=OO[3,]==11)
-write.csv(OO,file="~/f.txt")
+table(real_winner_ok=OO[1,]>0,top3_winner_ok=OO[3,]>0)
+save(OO,file="~/f.Rd")
 ```
@@
 virtual races in which each scoring system gets the winner incorrect;
-a good points system has a low value here.  Note the peculiar finding
-that `top3` has by far the lowest number, a finding which I have no
-explanation for.
+a good points system has a low value here.  Note that `top3` has by
+far the lowest number, a finding which I have no explanation for.
 

### 2020-07-20 inst/f1points.Rmd
use figure labelling

@@
 optimization routines, and because they have access to derivatives,
-convergence is rapid.
+convergence is rapid.  A graphical illustration of the strengths is
+given in figure \ref{piechartstrength}
 
-```{r,label=printapiechart,echo=FALSE}
+```{r,label=printapiechart,echo=FALSE,fig.cap="Maximum likelihood estimates \\label{piechartstrength} of the strengths of the top-ranked 11 drivers in the 2017 Formula 1 season"}
 mH <- maxp(H)
@@
 
-```{r,label=likevspoints,cache=TRUE,echo=FALSE}
+```{r,label=likevspoints,cache=TRUE,echo=FALSE,fig.cap="Ordering of the \\label{orderingbypointsandlikelihood} top-ranked 11 drivers in the 2017 Formula 1 season.
+Horizontal axis gives official (points-based) order, and the vertical axis gives the likelihood horder.
+Thus the points-based ordering would be Hamilton, Vettel, R\"{a}ikk\"{o}nen YES I KNOW IT IS WRONG: RMARKDOWN IS DEFECTIVE while the likelihood ordering is (reading vertically) Hamilton, Vettel, Bottas"}
 par(pty='s')        # square plot
@@
 
-From the above, we see that the first two drivers (Hamilton, Vettel)
-are are respectively first and second according to both ranking
-ranking procedures; but R\"{a}ikk\"{o}nen *yes I know it is wrong:
-rmarkdown is defective* and Bottas are third and fourth, and fourth
-and third, according to points and likelihood respectively.  The
-degree of agreement between the two systems might be quantified as
-"two", for the positions of the top two drivers agree.
+From figure \r=ef{orderingbypointsandlikelihood}, we see that the
+first two drivers (Hamilton, Vettel) are are respectively first and
+second according to both ranking ranking procedures; but
+R\"{a}ikk\"{o}nen YES I KNOW IT IS WRONG: RMARKDOWN IS DEFECTIVE and
+Bottas are third and fourth, and fourth and third, according to points
+and likelihood respectively.  The degree of agreement between the two
+systems might be quantified as "two", for the positions of the top two
+drivers agree.
 

### 2020-07-20 inst/f1points.Rmd
typo

@@
 \begin{itemize}
-\item Equal points to the top $n$ competitors
+\item Equal points to the top $r$ competitors
 $(\underbrace{1,1,\ldots ,1}_{r},0,0,\ldots)$, noting that
-$n=r$ corresponds to a ``winner takes all'' system
+$r=1$ corresponds to a ``winner takes all'' system
 \item Linear decrease $(r,r-1,r-2,\ldots,3,2,1,0)$ for integer $0<r<n$

### 2020-07-20 inst/f1points.Rmd
typo

@@
 
-From figure \r=ef{orderingbypointsandlikelihood}, we see that the
+From figure \ref{orderingbypointsandlikelihood}, we see that the
 first two drivers (Hamilton, Vettel) are are respectively first and

### 2020-07-22 inst/f1points.Rmd
tiny changes to filename

@@
   plain:     [keywords, not capitalized, Java]
-preamble: >
+preamble: 
   \usepackage{amsmath}
@@
 table(real_winner_ok=OO[1,]>0,top3_winner_ok=OO[3,]>0)
-save(OO,file="~/f.Rd")
+save(OO,file="~/rstudio/hyper2/inst/cache_results_2017.Rd")
 ```

### 2020-07-22 inst/soling.Rmd
work
fixes #46

@@
 
-  \url{https://en.wikipedia.org/w/index.php?title=Sailing_at_the_2000_Summer_Olympics_\%E2\%80\%93_Soling&oldid=945362535}
+[https://en.wikipedia.org/w/index.php?title=Sailing_at_the_2000_Summer_Olympics_\%E2\%80\%93_Soling&oldid=945362535]
 
@@
 ```
+
+So `H2` is the _difference_ between `H`--- which is `soling2000` plus
+the observations from the roundrobins and quarter finals---and
+`soling2000`.  In other words, `H2` is the simply the information in
+the roundrobins and quarter finals.
+
+```{r,testH2equalp}
+equalp.test(H2)
+```

### 2020-07-22 inst/f1points.Rmd
cache for years 2016-9, heavy lifting done in f1points_Omaker.R

@@
 top <- 11
-H <- ordertable2supp(F1_table_2017[seq_len(top),])
+H2016 <- ordertable2supp(F1_table_2016[seq_len(top),])
+H2017 <- ordertable2supp(F1_table_2017[seq_len(top),])
+H2018 <- ordertable2supp(F1_table_2018[seq_len(top),])
+H2019 <- ordertable2supp(F1_table_2019[seq_len(top),])
 ```
@@
 
-```{r,label=printapiechart,echo=FALSE,fig.cap="Maximum likelihood estimates \\label{piechartstrength} of the strengths of the top-ranked 11 drivers in the 2017 Formula 1 season"}
-mH <- maxp(H)
-mH
-pie(maxp(H)[c(1,6,11,2,7,3,8,4,9,5,10)])
+```{r,label=maximumlikelihoodcalculations,cache=TRUE}
+mH2016 <- maxp(H2016)
+mH2017 <- maxp(H2017)
+mH2018 <- maxp(H2018)
+mH2019 <- maxp(H2019)
 ```
 
-We see the Hamilton has the largest estimated strength (at about
-0.39), followed by Vettel and then Bottas at about 0.2 and 0.13
-respectively.  However, it should be noted that a likelihood ratio
-test [`samep.test()`, supplied with the package] does not reject the
-null that Hamilton and Vettel have the same strength ($H_0\colon
-\pham= \pvet$) with a likelihood ratio of $e^{1.5}\simeq 4.48$,
-corresponding (by Wilks's theorem) to an asymptotic $p$-value of about
-$0.08$.
+```{r,label=printapiechart,echo=FALSE,fig.cap="Maximum likelihood estimates \\label{piechartstrength} of the strengths of the top-ranked 11 drivers in Formula 1 motor racing, seasons 2016-2019"}
+mH2016
+mH2017
+mH2018
+mH2019
+par(mfrow=c(2,2))
+par(mar = c(2,2,2,2)/2)
+pie(mH2016[c(1,6,11,2,7,3,8,4,9,5,10)],cex=0.7,main="2016")
+pie(mH2017[c(1,6,11,2,7,3,8,4,9,5,10)],cex=0.7,main="2017")
+pie(mH2018[c(1,6,11,2,7,3,8,4,9,5,10)],cex=0.7,main="2018")
+pie(mH2019[c(1,6,11,2,7,3,8,4,9,5,10)],cex=0.7,main="2019")
+```
+
+We see in figure \ref{piechartstrength} that the driver with the
+largest estimated strength was Rosberg in 2016 (at about 49\%) and in
+years 2017-8-9 was Hamilton (at about 29\% 35\% 50\ respectively).
+However, it should be noted that a likelihood ratio test
+[`samep.test()`, supplied with the package] does not reject, for
+example, the null that Hamilton and Vettel have the same strength in
+2017 ($H_0\colon \pham= \pvet$) with a likelihood ratio of
+$e^{1.5}\simeq 4.48$, corresponding (by Wilks's theorem) to an
+asymptotic $p$-value of about $0.08$.
 
 ```{r,label=dosameptest, echo=FALSE, print=FALSE, cache=TRUE}
-ignore <- samep.test(H,1:2)
+ignore <- samep.test(H2017,1:2)
 ```
@@
 dp[] <- seq_along(dp)
-m <- maxp(H)
+m <- maxp(H2017)
 m <- sort(m,decreasing=TRUE)
@@
 ```{r,label=likelihoodrankingshower,echo=FALSE}
-jj <- mH
+jj <- mH2017
 jj <- sort(jj,decreasing=TRUE)
@@
 ```{r,label=definemultiplemontecarlo, echo=FALSE}
-top <- 11
-f2017 <- read.table("formula1_2017.txt",header=TRUE)[seq_len(top),1:20]
-m <- maxp(ordertable2supp(as.ordertable(f2017)))
-
-## Now work on synthetic dataset generation:
-resampling_multiple <- function(m,f2017,pointslist){
-    random_table <- rrank(n=ncol(f2017), p=m)
-    rownames(random_table) <- colnames(f2017)
-    ## "random_table" is a random table; now calculate lstar and pstar for
-    ## likelihood ranks and points ranks respectively:
-
-    l_star <- maxp(ordertable2supp(ranktable_to_ordertable(random_table)))
-    l_star[] <- seq_along(l_star)
-    ## l_star is the likelihood order statistic.
-
-	goodnesses <- seq_along(pointslist)
-	for(i in seq_along(pointslist)){
-		p_star <- ordertable2points(as.ordertable(ranktable_to_ordertable(random_table)),pointslist[[i]])
-		p_star <- sort(p_star,decreasing=TRUE)
-		p_star[] <- seq_along(p_star)
-		goodness <- sum(cumprod(names(p_star)==names(l_star)))
-		goodnesses[i] <- goodness
-		}
-	return(goodnesses)
-}
-pointslist <- list(
-	real = points_real,
-	inau = points_inaugural,
-	top1 = rep(1,1),
-	top2 = rep(1,2),
-	top3 = rep(1,3),
-	top4 = rep(1,4),
-	top5 = rep(1,5),
-	top6 = rep(1,6),
-	top7 = rep(1,7),
-	top8 = rep(1,8),
-	top9 = rep(1,9),
-	topt = rep(1,10),  # effectively punishing the last finisher
-   	lin1 = 1,   # same as top1
-	lin2 = 2:1,
-	lin3 = 3:1,
-	lin4 = 4:1,
-	lin5 = 5:1,
-	lin6 = 6:1,
-	lin7 = 7:1,
-	lin8 = 8:1,
-	lin9 = 9:1,
-	lint = 10:1,
-	line = 11:1,
-	zipf = 1/seq_len(top),
-	expa = 1/1.01^seq_len(top),
-	expb = 1/1.10^seq_len(top),
-	expc = 1/1.20^seq_len(top),
-	expd = 1/1.50^seq_len(top),
-	exp2 = 1/2^seq_len(top),
-	exp3 = 1/3^seq_len(top),
-	exp4 = 1/4^seq_len(top),
-	fis1 = c(1,1,1),
-	fis2 = c(1,1,0.9),
-	fis3 = c(1,1,1,0.1),
-	fis4 = c(1,1,0.9,0.1),
-	fis5 = c(1,1,0.5),
-	fis6 = c(1,1,1,0.5),
-	fis7 = c(1,1,0.5,0.5)
-)
-```
-
-```{r,label=multiplemontecarloexecute, cache=TRUE,echo=FALSE}
-set.seed(9)
-f2017 <- read.table("formula1_2017.txt",header=TRUE)[seq_len(top),1:20]
-m <- maxp(ordertable2supp(as.ordertable(f2017)))
-OO <- replicate(1000,resampling_multiple(m,f2017,pointslist=pointslist))
+load("formula1_results_2017.rda") # this created by running f1points_Omaker.R; file contains R variables OO and pointslist
 ```

### 2020-07-22 inst/f1points_Omaker.R

@@ -1,0 +1,145 @@
+## This file creates 1000 (variable "howmany") simulated Formula 1
+## results and compares the simulated ranks with the actual
+## points-based ranks according to different points systems such as
+## the real one or Zipf's law.  The R commands here are copied from
+## f1points.Rmd.  Some of the lines are redundant.
+
+## The simulations take quite a long time to run and at the bottom the
+## results are saved to file
+"~/rstudio/hyper2/inst/formula1_results_2017.rda"
+
+## The "point" of this file is that you can open it in emacs and
+## globally replace "2017" with "2018" and everything still works
+## (note that this would save the results to file
+## "~/rstudio/hyper2/inst/formula1_results_2018.rda": the year in the
+## filename will change too.
+
+howmany <- 1000
+top <- 11
+points_inaugural <- c(8,6,4,3,2)
+
+library(hyper2,quietly=TRUE)
+H <- ordertable2supp(F1_table_2017[seq_len(top),])
+
+
+mH <- maxp(H)
+mH
+pie(maxp(H)[c(1,6,11,2,7,3,8,4,9,5,10)])
+
+
+ignore <- samep.test(H,1:2)
+
+p <- c(25, 18, 15, 12, 10, 8, 6, 4, 2, 1, 0, 0)
+points_real <- p
+points_inaugural <- c(8,6,4,3,2)
+dp <- ordertable2points(as.ordertable(F1_table_2017[seq_len(top),]),p)
+dp <- sort(dp,decreasing=TRUE)
+dp[] <- seq_along(dp)
+m <- maxp(H)
+m <- sort(m,decreasing=TRUE)
+m[] <- seq_along(m)
+
+par(pty='s')        # square plot
+ox <- dp
+oy <- ordertrans(m,dp)
+
+
+real <- ordertable2points(ranktable_to_ordertable(wikitable_to_ranktable(F1_table_2017)),p)
+real <- round(real[seq_len(top)],2)
+real <- sort(real,decreasing=TRUE)
+real[] <- seq_along(real)
+real
+
+zipf <- ordertable2points(ranktable_to_ordertable(wikitable_to_ranktable(F1_table_2017)),1/seq_len(top))
[…]

### 2020-07-22 inst/f1points.Rmd
name 2017 variables to foo2017

@@
 points_inaugural <- c(8,6,4,3,2)
-dp <- ordertable2points(as.ordertable(F1_table_2017[seq_len(top),]),p)
-dp <- sort(dp,decreasing=TRUE)
-dp[] <- seq_along(dp)
-m <- maxp(H2017)
-m <- sort(m,decreasing=TRUE)
-m[] <- seq_along(m)
+dp2017 <- ordertable2points(as.ordertable(F1_table_2017[seq_len(top),]),p)
+dp2017 <- sort(dp2017,decreasing=TRUE)
+dp2017[] <- seq_along(dp2017)
+m2017 <- maxp(H2017)
+m2017 <- sort(m2017,decreasing=TRUE)
+m2017[] <- seq_along(m2017)
 ```
@@
 par(pty='s')        # square plot
-ox <- dp
-oy <- ordertrans(m,dp)
+ox <- dp2017
+oy <- ordertrans(m2017,dp2017)
 plot(ox,oy,asp=1,pty='s',xlim=c(0,11.2),ylim=c(0,11.2),pch=16,

### 2020-07-23 inst/f1points.Rmd
more consistentification

--- 2020-07-22 inst/f1points.Rmd
@@
 doubles tennis teams; we may use a similar device to account for pole
-position in Formula 1.  Here I analyse racing results from the 2017
-season using the `hyper2` package \citep{hankin2017} which implements
-the Plackett-Luce likelihood function with additional reified
-entities.
+position in Formula 1.  Here I analyse seasons 2016-2019 using the
+`hyper2` package \citep{hankin2017} which implements the Plackett-Luce
+likelihood function with additional reified entities.
 
@@
 
-Here we use the results from 2017:
+Taking 2017 as an example, we see the rank table as follows:
 
@@
 largest estimated strength was Rosberg in 2016 (at about 49\%) and in
-years 2017-8-9 was Hamilton (at about 29\% 35\% 50\ respectively).
-However, it should be noted that a likelihood ratio test
-[`samep.test()`, supplied with the package] does not reject, for
+years 2017-8-9 was Hamilton (at about 29\%, 35\%, and 50\%
+respectively).  However, it should be noted that a likelihood ratio
+test [`samep.test()`, supplied with the package] does not reject, for
 example, the null that Hamilton and Vettel have the same strength in
@@
 points_inaugural <- c(8,6,4,3,2)
-dp <- ordertable2points(as.ordertable(F1_table_2017[seq_len(top),]),p)
-dp <- sort(dp,decreasing=TRUE)
-dp[] <- seq_along(dp)
-m <- maxp(H2017)
-m <- sort(m,decreasing=TRUE)
-m[] <- seq_along(m)
+dp2016 <- ordertable2points(as.ordertable(F1_table_2016[seq_len(top),]),p)
+dp2016 <- sort(dp2016,decreasing=TRUE)
+dp2016[] <- seq_along(dp2016)
+m2016 <- maxp(H2016)
+m2016 <- sort(m2016,decreasing=TRUE)
+m2016[] <- seq_along(m2016)
+
+dp2017 <- ordertable2points(as.ordertable(F1_table_2017[seq_len(top),]),p)
+dp2017 <- sort(dp2017,decreasing=TRUE)
+dp2017[] <- seq_along(dp2017)
+m2017 <- maxp(H2017)
+m2017 <- sort(m2017,decreasing=TRUE)
+m2017[] <- seq_along(m2017)
+
+dp2018 <- ordertable2points(as.ordertable(F1_table_2018[seq_len(top),]),p)
+dp2018 <- sort(dp2018,decreasing=TRUE)
+dp2018[] <- seq_along(dp2018)
+m2018 <- maxp(H2018)
+m2018 <- sort(m2018,decreasing=TRUE)
+m2018[] <- seq_along(m2018)
+
+dp2019 <- ordertable2points(as.ordertable(F1_table_2019[seq_len(top),]),p)
+dp2019 <- sort(dp2019,decreasing=TRUE)
+dp2019[] <- seq_along(dp2019)
+m2019 <- maxp(H2019)
+m2019 <- sort(m2019,decreasing=TRUE)
+m2019[] <- seq_along(m2019)
 ```
@@
 Horizontal axis gives official (points-based) order, and the vertical axis gives the likelihood horder.
-Thus the points-based ordering would be Hamilton, Vettel, R\"{a}ikk\"{o}nen YES I KNOW IT IS WRONG: RMARKDOWN IS DEFECTIVE while the likelihood ordering is (reading vertically) Hamilton, Vettel, Bottas"}
+Thus, taking 201 as an example,  the points-based ordering would be Hamilton first, then Vettel, then R\"{a}ikk\"{o}nen; YES I KNOW IT IS WRONG: RMARKDOWN IS DEFECTIVE while the likelihood ordering is (reading vertically) Hamilton, Vettel, Bottas ALSO R PLOTTING SUCKS HERE AND I WILL HAVE TO CHANGE THE PDF BY HAND WITH INKSCAPE TO MAKE IT LOOK GOOD"}
+par(mfrow=c(2,2))
+par(mar = c(3,3,3,3))
+
 par(pty='s')        # square plot
-ox <- dp
-oy <- ordertrans(m,dp)
+ox <- dp2016
+oy <- ordertrans(m2016,dp2016)
 plot(ox,oy,asp=1,pty='s',xlim=c(0,11.2),ylim=c(0,11.2),pch=16,
-    xlab="official order",ylab="my order",main='Formula 1, 2017 season')
+    xlab="points order",ylab="likelihood order",main='2016 season')
 par(xpd=TRUE)       # allow drivers' names to appear outside plotting region
-for(i in seq_along(ox)){text(ox[i],oy[i],names(ox)[i],pos=4,col='gray') }
+for(i in seq_along(ox)){text(ox[i],oy[i],names(ox)[i],pos=4,col='gray',cex=0.6) }
+par(xpd=FALSE)      # stop diagonal line from protruding beyond plotting region
+abline(0,1)
+
+par(pty='s')        # square plot
+ox <- dp2017
+oy <- ordertrans(m2017,dp2017)
+plot(ox,oy,asp=1,pty='s',xlim=c(0,11.2),ylim=c(0,11.2),pch=16,
+    xlab="points order",ylab="likelihood order",main='2017 season')
+par(xpd=TRUE)       # allow drivers' names to appear outside plotting region
+for(i in seq_along(ox)){text(ox[i],oy[i],names(ox)[i],pos=4,col='gray', cex=0.6) }
+par(xpd=FALSE)      # stop diagonal line from protruding beyond plotting region
+abline(0,1)
+
+par(pty='s')        # square plot
+ox <- dp2018
+oy <- ordertrans(m2018,dp2018)
+plot(ox,oy,asp=1,pty='s',xlim=c(0,11.2),ylim=c(0,11.2),pch=16,
+    xlab="points order",ylab="likelihood order",main='2018 season')
+par(xpd=TRUE)       # allow drivers' names to appear outside plotting region
+for(i in seq_along(ox)){text(ox[i],oy[i],names(ox)[i],pos=4,col='gray', cex=0.6) }
+par(xpd=FALSE)      # stop diagonal line from protruding beyond plotting region
+abline(0,1)
+
+par(pty='s')        # square plot
+ox <- dp2019
+oy <- ordertrans(m2019,dp2019)
+plot(ox,oy,asp=1,pty='s',xlim=c(0,11.2),ylim=c(0,11.2),pch=16,
+    xlab="points order",ylab="likelihood order",main='2019 season')
+par(xpd=TRUE)       # allow drivers' names to appear outside plotting region
+for(i in seq_along(ox)){text(ox[i],oy[i],names(ox)[i],pos=4,col='gray',cex=0.6) }
 par(xpd=FALSE)      # stop diagonal line from protruding beyond plotting region
@@
 
-From figure \ref{orderingbypointsandlikelihood}, we see that the
-first two drivers (Hamilton, Vettel) are are respectively first and
-second according to both ranking ranking procedures; but
-R\"{a}ikk\"{o}nen YES I KNOW IT IS WRONG: RMARKDOWN IS DEFECTIVE and
-Bottas are third and fourth, and fourth and third, according to points
-and likelihood respectively.  The degree of agreement between the two
-systems might be quantified as "two", for the positions of the top two
-drivers agree.
-
-Observing that both the likelihood ranking and the points ranking are
-random variables suggests a method whereby we can objectively assess a
-given points system.  Using sampling techniques we can repeatedly
-generate an order table \emph{in silico}, using estimated driver
-strengths from the observed 2017 table.  For each of, say, 1000 such
-synthetic tables, calculate drivers' maximum likelihood
-\citeauthor{plackett1975} strengths, and also their points awarded
-according to any given points system.  We then compare rankings
-generated by the \citeauthor{plackett1975} strengths and the points
-awarded and note the degree of agreement between the two.  This
-furnishes an objective assessment of the points system used.
+From figure \ref{orderingbypointsandlikelihood}, we see a comparison
+between the two ordering systems.  Taking 2017 as an example, drivers
+Hamilton and Vettel are are respectively first and second according to
+both ranking ranking procedures; but R\"{a}ikk\"{o}nen YES I KNOW IT
+IS WRONG: RMARKDOWN IS DEFECTIVE and Bottas are third and fourth, and
+fourth and third, according to points and likelihood respectively.
+The degree of agreement between the two systems might be quantified as
+"2", as the positions of the top two drivers agree.  Thus the degree
+of agreement for the years 2016-9 would be 8,2,6, and 2 respectively.
 
@@
 (this happens to be identical to the ranking after the extra point was
-awarded for fastest lap).  However, if we were to use Zipf's law, the
-ranking would be:
+awarded for fastest lap).  However, if we were award points according
+to Zipf's law we would have:
 
 ```{r,label=zipfpointsshower,echo=FALSE}
-zipf <- ordertable2points(ranktable_to_ordertable(wikitable_to_ranktable(F1_table_2017)),1/seq_len(top))
+zipf <- ordertable2points(ranktable_to_ordertable(wikitable_to_ranktable(F1_table_2017)),1)
 zipf <- round(zipf[seq_len(top)],2)
@@
 
-Now compare the likelihood ranking:
-
 ```{r,label=likelihoodrankingshower,echo=FALSE}
@@
 ```
+
+Thus, for 2017 at least, we see that the current poionts system agrees
+with likelihood ranking for the top four places, while Zipf's law
+agrees to three.
+
+Observing that both the likelihood ranking and the points ranking are
+random variables in this paradigm suggests a method whereby we can
+objectively assess a given points system.  Using sampling techniques
+we can repeatedly generate an order table \emph{in silico}, using
+estimated driver strengths from the observed tables.  For each of,
+say, 1000 such synthetic tables, calculate drivers' maximum likelihood
+\citeauthor{plackett1975} strengths, and also their points awarded
+according to any given points system.  We then compare rankings
+generated by the \citeauthor{plackett1975} strengths and the points
+awarded and note the degree of agreement between the two.  This
+furnishes an objective assessment of the points system used.
 

### 2020-07-26 inst/f1points.Rmd
bugfix, should *not* save!
typo

@@
 
-Thus, for 2017 at least, we see that the current poionts system agrees
+Thus, for 2017 at least, we see that the current points system agrees
 with likelihood ranking for the top four places, while Zipf's law
@@
 ```{r,label=definemultiplemontecarlo, echo=FALSE}
+load("formula1_results_2016.rda") # this created by running f1points_Omaker.R; file contains R variables OO and pointslist
+OO2016 <- OO
+
 load("formula1_results_2017.rda") # this created by running f1points_Omaker.R; file contains R variables OO and pointslist
+OO2017 <- OO
+
+load("formula1_results_2018.rda") # this created by running f1points_Omaker.R; file contains R variables OO and pointslist
+OO2018 <- OO
+
+load("formula1_results_2019.rda") # this created by running f1points_Omaker.R; file contains R variables OO and pointslist
+OO2019 <- OO
 ```
@@
 ```{r,label=somesummaries}
-summarytable <- data.frame(
+summarytablemaker <- function(OO){
+  out <- data.frame(
       means = apply(OO,1,mean),
-winner_correct = rowSums(OO>1),
-all_correct = rowSums(OO==top))
-rownames(summarytable) <- names(pointslist)
-summarytable
+      winner_correct = rowSums(OO>1)/ncol(OO),
+      all_correct = rowSums(OO==top)/ncol(OO)
+  )
+  rownames(out) <- names(pointslist)
+  return(out)
+}
+s2016 <- summarytablemaker(OO2016)
+s2017 <- summarytablemaker(OO2017)
+s2018 <- summarytablemaker(OO2018)
+s2019 <- summarytablemaker(OO2019)
 ```
@@
 ```{r,fig=TRUE}
-typecol <- c(rep("red",2),rep("orange",10),rep("blue",11),rep("orange",1),rep("black",7),rep("gray",7))
-plot(summarytable[,1],ylab='mean',pch=16,col=typecol)
-plot(summarytable[,2],ylab='prob of winner correct',pch=16,col=typecol)
-plot(summarytable[,3],ylab='prob. of total rank order correct',pch=16,col=typecol)
+typecol <- c(rep("red",2),rep("orange",10),rep("yellow",11),rep("green",1),rep("blue",7),rep("purple",7))
+jj <- cbind(s2016[,1],s2017[,1],s2018[,1],s2019[,1])  # mean
+matplot(jj,type="n",xlab='points system',ylab='mean correct places')
+points(s2016[,1],col=typecol,pch=1)
+points(s2017[,1],col=typecol,pch=2)
+points(s2018[,1],col=typecol,pch=3)
+points(s2019[,1],col=typecol,pch=4)
+legend("topright",lty=0,pch=1:4,legend=c("2016","2017","2018","2019"))
 ```
 
-and visual:
+```{r,fig=TRUE}
+jj <- cbind(s2016[,2],s2017[,2],s2018[,2],s2019[,2])  # prob of winner
+matplot(jj,type="n",xlab='points system',ylab='probability of winner correct')
+points(s2016[,2],col=typecol,pch=1)
+points(s2017[,2],col=typecol,pch=2)
+points(s2018[,2],col=typecol,pch=3)
+points(s2019[,2],col=typecol,pch=4)
+legend("topright",lty=0,pch=1:4,legend=c("2016","2017","2018","2019"))
+```
 
-```{r,label=plotcumulativeM,fig=TRUE}
-o <- t(apply(1+OO,1,tabulate))
-#matplot(0:11,apply(o[,12:1],1,cumsum)/1e3,type='b',lty=1,pch=16,ylab='probability',axes=FALSE,xlab='number incorrect')
-jj <- c(0,2,4,6,8,11)
-#axis(1,pos=0,at=jj,labels=c("0",paste("<=",jj[-1],sep="")))
-#axis(2)
+```{r,fig=TRUE}
+jj <- cbind(s2016[,3],s2017[,3],s2018[,3],s2019[,3])  # total rank order correct
+matplot(jj,type="n",xlab='points system',ylab='probability of total rank order correct')
+points(s2016[,3],col=typecol,pch=1)
+points(s2017[,3],col=typecol,pch=2)
+points(s2018[,3],col=typecol,pch=3)
+points(s2019[,3],col=typecol,pch=4)
+legend("topright",lty=0,pch=1:4,legend=c("2016","2017","2018","2019"))
 ```
@@
 table(real_winner_ok=OO[1,]>0,top3_winner_ok=OO[3,]>0)
-save(OO,file="~/rstudio/hyper2/inst/cache_results_2017.Rd")
 ```

### 2020-07-27 inst/f1points.Rmd
use formula1_points_system() instead

@@
 top <- 11
-points_inaugural <- c(8,6,4,3,2)
 ```
@@
 ```{r, label=calcpoints, echo=FALSE, cache=TRUE}
-p <- c(25, 18, 15, 12, 10, 8, 6, 4, 2, 1, 0, 0)
-points_real <- p
-points_inaugural <- c(8,6,4,3,2)
+p <- formula1_points_systems()[[1]]
 dp2016 <- ordertable2points(as.ordertable(F1_table_2016[seq_len(top),]),p)

### 2020-07-27 inst/f1points.Rmd
include zipf law rank comparison

@@
 
-The points scored by each driver give a ranking shown in the table;
-but Plackett-Luce likelihoods also give a ranking that does not depend
-on any arbitrary points system.  We can plot one ranking against the
-other:
-
-```{r, label=calcpoints, echo=FALSE, cache=TRUE}
-p <- formula1_points_systems()[[1]]
+```{r, label=calcpoints_real, echo=FALSE, cache=TRUE}
+p <- formula1_points_systems()[[1]]   # real points system
 dp2016 <- ordertable2points(as.ordertable(F1_table_2016[seq_len(top),]),p)
@@
 
-(note carefully that the historically correct points awarded to the
-drivers differs from that calculated here.  That is for two reasons:
-firstly, ``fastest lap'' points are not included here, and also the
-truncation of the order table to the first 11 drivers can increase the
-rank of a driver if non-first-11 drivers are placed).
-
-```{r,label=likevspoints,cache=TRUE,echo=FALSE,fig.cap="Ordering of the \\label{orderingbypointsandlikelihood} top-ranked 11 drivers in the 2017 Formula 1 season.
+```{r,label=likevspoints,cache=TRUE,echo=FALSE,fig.cap="Ordering of the \\label{orderingbypointsandlikelihood} top-ranked 11 drivers in Formula 1 seasons 2016-2019.
 Horizontal axis gives official (points-based) order, and the vertical axis gives the likelihood horder.
-Thus, taking 201 as an example,  the points-based ordering would be Hamilton first, then Vettel, then R\"{a}ikk\"{o}nen; YES I KNOW IT IS WRONG: RMARKDOWN IS DEFECTIVE while the likelihood ordering is (reading vertically) Hamilton, Vettel, Bottas ALSO R PLOTTING SUCKS HERE AND I WILL HAVE TO CHANGE THE PDF BY HAND WITH INKSCAPE TO MAKE IT LOOK GOOD"}
+Thus, taking 2017 as an example,  the points-based ordering would be Hamilton first, then Vettel, then R\"{a}ikk\"{o}nen; YES I KNOW IT IS WRONG: RMARKDOWN IS DEFECTIVE while the likelihood ordering is (reading vertically) Hamilton, Vettel, Bottas ALSO R PLOTTING SUCKS HERE AND I WILL HAVE TO CHANGE THE PDF BY HAND WITH INKSCAPE TO MAKE IT LOOK GOOD"}
 par(mfrow=c(2,2))
@@
 
-From figure \ref{orderingbypointsandlikelihood}, we see a comparison
-between the two ordering systems.  Taking 2017 as an example, drivers
-Hamilton and Vettel are are respectively first and second according to
-both ranking ranking procedures; but R\"{a}ikk\"{o}nen YES I KNOW IT
-IS WRONG: RMARKDOWN IS DEFECTIVE and Bottas are third and fourth, and
+The points scored by each driver give a ranking shown in the table;
+but Plackett-Luce likelihoods also give a ranking that does not depend
+on any arbitrary points system.  We can plot one ranking against the
+other, see Figure \ref{orderingbypointsandlikelihood}\footnote{The
+historically correct points awarded to the drivers differs from that
+calculated here.  That is for two reasons: firstly, ``fastest lap''
+points are not included here, and also the truncation of the order
+table to the first 11 drivers can increase the rank of a driver if
+non-first-11 drivers are placed}.  We thus see a comparison between
+the two ordering systems.  Taking 2017 as an example, drivers Hamilton
+and Vettel are are respectively first and second according to both
+ranking ranking procedures; but R\"{a}ikk\"{o}nen YES I KNOW IT IS
+WRONG: RMARKDOWN IS DEFECTIVE and Bottas are third and fourth, and
 fourth and third, according to points and likelihood respectively.
@@
 "2", as the positions of the top two drivers agree.  Thus the degree
-of agreement for the years 2016-9 would be 8,2,6, and 2 respectively.
+of agreement between likelihood ranking and points ranking for the
+years 2016-9 would be 8,2,6, and 2 respectively.
+
+However, the points system used is essentially arbitrary.  We could
+use, for example, Zipf's law to rank the drivers: award one point to
+the winner, half a point to second place, one third of a point to
+third, and so on; see figure \ref{orderingbypointsandlikelihood_zipf}
+in which we see generally poorer agreement between points-based ranks
+and likelihood-based ranks.
+
+```{r, label=calcpoints_zipf, echo=FALSE, cache=TRUE}
+p <- formula1_points_systems()$zipf
+dp2016 <- ordertable2points(as.ordertable(F1_table_2016[seq_len(top),]),p)
+dp2016 <- sort(dp2016,decreasing=TRUE)
+dp2016[] <- seq_along(dp2016)
+m2016 <- maxp(H2016)
+m2016 <- sort(m2016,decreasing=TRUE)
+m2016[] <- seq_along(m2016)
+
+dp2017 <- ordertable2points(as.ordertable(F1_table_2017[seq_len(top),]),p)
+dp2017 <- sort(dp2017,decreasing=TRUE)
+dp2017[] <- seq_along(dp2017)
+m2017 <- maxp(H2017)
+m2017 <- sort(m2017,decreasing=TRUE)
+m2017[] <- seq_along(m2017)
+
+dp2018 <- ordertable2points(as.ordertable(F1_table_2018[seq_len(top),]),p)
+dp2018 <- sort(dp2018,decreasing=TRUE)
+dp2018[] <- seq_along(dp2018)
+m2018 <- maxp(H2018)
+m2018 <- sort(m2018,decreasing=TRUE)
+m2018[] <- seq_along(m2018)
+
+dp2019 <- ordertable2points(as.ordertable(F1_table_2019[seq_len(top),]),p)
+dp2019 <- sort(dp2019,decreasing=TRUE)
+dp2019[] <- seq_along(dp2019)
+m2019 <- maxp(H2019)
+m2019 <- sort(m2019,decreasing=TRUE)
+m2019[] <- seq_along(m2019)
+```
+
+```{r,label=likevspointszipf,cache=TRUE,echo=FALSE,fig.cap="As for figure \ref{likevspoints} \\label{orderingbypointsandlikelihood_zipf} but points ranking calculated by Zipf's law.
+Note the generally poorer agreement between the two ranking systems ALSO R PLOTTING SUCKS HERE AND I WILL HAVE TO CHANGE THE PDF BY HAND WITH INKSCAPE TO MAKE IT LOOK GOOD"}
+par(mfrow=c(2,2))
+par(mar = c(3,3,3,3))
+
+par(pty='s')        # square plot
+ox <- dp2016
+oy <- ordertrans(m2016,dp2016)
+plot(ox,oy,asp=1,pty='s',xlim=c(0,11.2),ylim=c(0,11.2),pch=16,
+    xlab="points order",ylab="likelihood order",main='2016 season')
+par(xpd=TRUE)       # allow drivers' names to appear outside plotting region
+for(i in seq_along(ox)){text(ox[i],oy[i],names(ox)[i],pos=4,col='gray',cex=0.6) }
+par(xpd=FALSE)      # stop diagonal line from protruding beyond plotting region
+abline(0,1)
+
+par(pty='s')        # square plot
+ox <- dp2017
+oy <- ordertrans(m2017,dp2017)
+plot(ox,oy,asp=1,pty='s',xlim=c(0,11.2),ylim=c(0,11.2),pch=16,
+    xlab="points order",ylab="likelihood order",main='2017 season')
+par(xpd=TRUE)       # allow drivers' names to appear outside plotting region
+for(i in seq_along(ox)){text(ox[i],oy[i],names(ox)[i],pos=4,col='gray', cex=0.6) }
+par(xpd=FALSE)      # stop diagonal line from protruding beyond plotting region
+abline(0,1)
+
+par(pty='s')        # square plot
+ox <- dp2018
+oy <- ordertrans(m2018,dp2018)
+plot(ox,oy,asp=1,pty='s',xlim=c(0,11.2),ylim=c(0,11.2),pch=16,
+    xlab="points order",ylab="likelihood order",main='2018 season')
+par(xpd=TRUE)       # allow drivers' names to appear outside plotting region
+for(i in seq_along(ox)){text(ox[i],oy[i],names(ox)[i],pos=4,col='gray', cex=0.6) }
+par(xpd=FALSE)      # stop diagonal line from protruding beyond plotting region
+abline(0,1)
+
+par(pty='s')        # square plot
+ox <- dp2019
+oy <- ordertrans(m2019,dp2019)
+plot(ox,oy,asp=1,pty='s',xlim=c(0,11.2),ylim=c(0,11.2),pch=16,
+    xlab="points order",ylab="likelihood order",main='2019 season')
+par(xpd=TRUE)       # allow drivers' names to appear outside plotting region
+for(i in seq_along(ox)){text(ox[i],oy[i],names(ox)[i],pos=4,col='gray',cex=0.6) }
+par(xpd=FALSE)      # stop diagonal line from protruding beyond plotting region
+abline(0,1)
+```
 

### 2020-07-27 inst/f1points.Rmd
more reconfiguration

@@
 
+Applying the current points system, for example, to the 2017 results table we
+would rank the drivers as follows:
+
+```{r,label=realpointsshower,echo=FALSE}
+real <- ordertable2points(ranktable_to_ordertable(wikitable_to_ranktable(F1_table_2017)),formula1_points_systems(top)$real)
+real <- round(real[seq_len(top)],2)
+real <- sort(real,decreasing=TRUE)
+real[] <- seq_along(real)
+real
+```
+
+(this happens to be identical to the ranking after the extra point was
+awarded for fastest lap).  However, if we were award points according
+to Zipf's law we would have:
+
+```{r,label=zipfpointsshower,echo=FALSE}
+zipf <- ordertable2points(ranktable_to_ordertable(wikitable_to_ranktable(F1_table_2017)),formula1_points_systems(top)$zipf)
+zipf <- round(zipf[seq_len(top)],2)
+zipf <- sort(zipf,decreasing=TRUE)
+zipf[] <- seq_along(zipf)
+zipf
+```
+
+Thus these two systems agree on the first three places but fourth is
+awarded to Raikkonen under the F1 system and Ricciardo under Zipf.
+
+```{r,label=likelihoodrankingshower,echo=FALSE}
+jj <- mH2017
+jj <- sort(jj,decreasing=TRUE)
+jj[] <- seq_along(jj)
+jj
+```
+
+Thus, for 2017 at least, we see that the current points system agrees
+with likelihood ranking for the top four places, while Zipf's law
+agrees to three.
+
 ```{r, label=calcpoints_real, echo=FALSE, cache=TRUE}
@@
 
-The points scored by each driver give a ranking shown in the table;
-but Plackett-Luce likelihoods also give a ranking that does not depend
-on any arbitrary points system.  We can plot one ranking against the
-other, see Figure \ref{orderingbypointsandlikelihood}\footnote{The
-historically correct points awarded to the drivers differs from that
-calculated here.  That is for two reasons: firstly, ``fastest lap''
-points are not included here, and also the truncation of the order
-table to the first 11 drivers can increase the rank of a driver if
-non-first-11 drivers are placed}.  We thus see a comparison between
+We can plot one ranking against the other\footnote{The historically
+correct points awarded to the drivers differs from that calculated
+here.  That is for two reasons: firstly, ``fastest lap'' points are
+not included here, and also the truncation of the order table to the
+first 11 drivers can increase the rank of a driver if non-first-11
+drivers are placed} , shown in Figure
+\ref{orderingbypointsandlikelihood}.  We thus see a comparison between
 the two ordering systems.  Taking 2017 as an example, drivers Hamilton
@@
 ```{r, label=calcpoints_zipf, echo=FALSE, cache=TRUE}
-p <- formula1_points_systems()$zipf
-dp2016 <- ordertable2points(as.ordertable(F1_table_2016[seq_len(top),]),p)
-dp2016 <- sort(dp2016,decreasing=TRUE)
-dp2016[] <- seq_along(dp2016)
-m2016 <- maxp(H2016)
-m2016 <- sort(m2016,decreasing=TRUE)
-m2016[] <- seq_along(m2016)
+pz <- formula1_points_systems()$zipf
+dp2016z <- ordertable2points(as.ordertable(F1_table_2016[seq_len(top),]),pz)
+dp2016z <- sort(dp2016z,decreasing=TRUE)
+dp2016z[] <- seq_along(dp2016z)
 
-dp2017 <- ordertable2points(as.ordertable(F1_table_2017[seq_len(top),]),p)
-dp2017 <- sort(dp2017,decreasing=TRUE)
-dp2017[] <- seq_along(dp2017)
-m2017 <- maxp(H2017)
-m2017 <- sort(m2017,decreasing=TRUE)
-m2017[] <- seq_along(m2017)
+dp2017z <- ordertable2points(as.ordertable(F1_table_2017[seq_len(top),]),pz)
+dp2017z <- sort(dp2017z,decreasing=TRUE)
+dp2017z[] <- seq_along(dp2017z)
 
-dp2018 <- ordertable2points(as.ordertable(F1_table_2018[seq_len(top),]),p)
-dp2018 <- sort(dp2018,decreasing=TRUE)
-dp2018[] <- seq_along(dp2018)
-m2018 <- maxp(H2018)
-m2018 <- sort(m2018,decreasing=TRUE)
-m2018[] <- seq_along(m2018)
+dp2018z <- ordertable2points(as.ordertable(F1_table_2018[seq_len(top),]),pz)
+dp2018z <- sort(dp2018z,decreasing=TRUE)
+dp2018z[] <- seq_along(dp2018z)
 
-dp2019 <- ordertable2points(as.ordertable(F1_table_2019[seq_len(top),]),p)
-dp2019 <- sort(dp2019,decreasing=TRUE)
-dp2019[] <- seq_along(dp2019)
-m2019 <- maxp(H2019)
-m2019 <- sort(m2019,decreasing=TRUE)
-m2019[] <- seq_along(m2019)
+dp2019z <- ordertable2points(as.ordertable(F1_table_2019[seq_len(top),]),pz)
+dp2019z <- sort(dp2019z,decreasing=TRUE)
+dp2019z[] <- seq_along(dp2019z)
 ```
@@
 par(pty='s')        # square plot
-ox <- dp2016
-oy <- ordertrans(m2016,dp2016)
+ox <- dp2016z
+oy <- ordertrans(m2016,dp2016z)
 plot(ox,oy,asp=1,pty='s',xlim=c(0,11.2),ylim=c(0,11.2),pch=16,
@@
 par(pty='s')        # square plot
-ox <- dp2017
-oy <- ordertrans(m2017,dp2017)
+ox <- dp2017z
+oy <- ordertrans(m2017,dp2017z)
 plot(ox,oy,asp=1,pty='s',xlim=c(0,11.2),ylim=c(0,11.2),pch=16,
@@
 par(pty='s')        # square plot
-ox <- dp2018
-oy <- ordertrans(m2018,dp2018)
+ox <- dp2018z
+oy <- ordertrans(m2018,dp2018z)
 plot(ox,oy,asp=1,pty='s',xlim=c(0,11.2),ylim=c(0,11.2),pch=16,
@@
 par(pty='s')        # square plot
-ox <- dp2019
-oy <- ordertrans(m2019,dp2019)
+ox <- dp2019z
+oy <- ordertrans(m2019,dp2019z)
 plot(ox,oy,asp=1,pty='s',xlim=c(0,11.2),ylim=c(0,11.2),pch=16,
@@
 
-## Points systems
+## Possible points systems
 
-Here we follow the above protocol to assess different points systems.
 There are a number of plausible points systems that might be used:
@@
 \end{itemize}
-
-Applying the current points system, for example, to the 2017 table we
-would rank the drivers as follows:
-
-```{r,label=realpointsshower,echo=FALSE}
-real <- ordertable2points(ranktable_to_ordertable(wikitable_to_ranktable(F1_table_2017)),p)
-real <- round(real[seq_len(top)],2)
-real <- sort(real,decreasing=TRUE)
-real[] <- seq_along(real)
-real
-```
-
-(this happens to be identical to the ranking after the extra point was
-awarded for fastest lap).  However, if we were award points according
-to Zipf's law we would have:
-
-```{r,label=zipfpointsshower,echo=FALSE}
-zipf <- ordertable2points(ranktable_to_ordertable(wikitable_to_ranktable(F1_table_2017)),1)
-zipf <- round(zipf[seq_len(top)],2)
-zipf <- sort(zipf,decreasing=TRUE)
-zipf[] <- seq_along(zipf)
-zipf
-```
-
-Thus these two systems agree on the first three places but fourth is
-awarded to Raikkonen under the F1 system and Ricciardo under Zipf.
-
-```{r,label=likelihoodrankingshower,echo=FALSE}
-jj <- mH2017
-jj <- sort(jj,decreasing=TRUE)
-jj[] <- seq_along(jj)
-jj
-```
-
-Thus, for 2017 at least, we see that the current points system agrees
-with likelihood ranking for the top four places, while Zipf's law
-agrees to three.
 

### 2020-07-27 inst/f1points.Rmd
more work

@@
 
+\begin{multline}
+\mbox{Hamilton}\succ
+\mbox{Vettel}\succ
+\mbox{Bottas}\succ
+\mbox{Raikkonen}\succ
+\mbox{Ricciardo}\succ
+\mbox{Verstappen}\succ\\
+\mbox{Perez1}\succ
+\mbox{Ocon}\succ
+\mbox{Sainz}\succ
+\mbox{Hulkenberg}\succ
+\mbox{Massa}
+\end{multline}
+
 (this happens to be identical to the ranking after the extra point was
@@
 ```
+
+\begin{multline}
+\mbox{Hamilton}\succ
+\mbox{Vettel}\succ
+\mbox{Bottas}\succ
+\mbox{Ricciardo}\succ
+\mbox{Verstappen}\succ
+\mbox{Raikkonen}\succ\\
+\mbox{Perez}\succ
+\mbox{Ocon}\succ
+\mbox{Massa}\succ
+\mbox{Sainz}\succ
+\mbox{Hulkenberg}\succ
+\end{multline}
 
@@
 
+We can now calculate number of places agreed for each of the points systems:
+
+```{r, label=calcpoints_allpointsystems, echo=FALSE, cache=TRUE}
+agreement2016 <- rep(NA,length(formula1_points_systems(top)))
+agreement2017 <- rep(NA,length(formula1_points_systems(top)))
+agreement2018 <- rep(NA,length(formula1_points_systems(top)))
+agreement2019 <- rep(NA,length(formula1_points_systems(top)))
+
+for(i in seq_along(formula1_points_systems(top))){
+
+  pa <- formula1_points_systems(top)[[i]]
+
+  dp2016a <- ordertable2points(as.ordertable(F1_table_2016[seq_len(top),]),pa)
+  dp2016a <- sort(dp2016a,decreasing=TRUE)
+  dp2016a[] <- seq_along(dp2016a)
+  agreement2016[i] <-  sum(cumprod(names(dp2016a) == names(m2016)))
+
+  dp2017a <- ordertable2points(as.ordertable(F1_table_2017[seq_len(top),]),pa)
+  dp2017a <- sort(dp2017a,decreasing=TRUE)
+  dp2017a[] <- seq_along(dp2017a)
+  agreement2017[i] <-  sum(cumprod(names(dp2017a) == names(m2017)))
+
+  dp2018a <- ordertable2points(as.ordertable(F1_table_2018[seq_len(top),]),pa)
+  dp2018a <- sort(dp2018a,decreasing=TRUE)
+  dp2018a[] <- seq_along(dp2018a)
+  agreement2018[i] <-  sum(cumprod(names(dp2018a) == names(m2018)))
+
+  dp2019a <- ordertable2points(as.ordertable(F1_table_2019[seq_len(top),]),pa)
+  dp2019a <- sort(dp2019a,decreasing=TRUE)
+  dp2019a[] <- seq_along(dp2019a)
+  agreement2019[i] <-  sum(cumprod(names(dp2019a) == names(m2019)))
+
+}  # i loop closes
+
+agreement4yearsallp <- cbind(
+    agreement2016,
+    agreement2017,
+    agreement2018,
+    agreement2019
+)
+
+rownames(agreement4yearsallp) <- names(formula1_points_systems(top))
+colnames(agreement4yearsallp) <- 2016:2019
+```
+
+```{r}
+agreement4yearsallp
+```
+
 Observing that both the likelihood ranking and the points ranking are

### 2020-07-28 inst/f1points.Rmd
more work

@@
 
-```{r,label=maximumlikelihoodcalculations,cache=TRUE}
+```{r,label=maximumlikelihoodcalculations,cache=TRUE,echo=FALSE}
 mH2016 <- maxp(H2016)
@@
 ```{r,label=printapiechart,echo=FALSE,fig.cap="Maximum likelihood estimates \\label{piechartstrength} of the strengths of the top-ranked 11 drivers in Formula 1 motor racing, seasons 2016-2019"}
-mH2016
-mH2017
-mH2018
-mH2019
 par(mfrow=c(2,2))
@@
 real[] <- seq_along(real)
-real
 ```
@@
 zipf[] <- seq_along(zipf)
-zipf
 ```
@@
 awarded to Raikkonen under the F1 system and Ricciardo under Zipf.
+Compare the likelihood ranking:
 
@@
 jj[] <- seq_along(jj)
-jj
 ```
@@
 in which we see generally poorer agreement between points-based ranks
-and likelihood-based ranks.
+and likelihood-based ranks.  There are a number of plausible points
+systems that might be used:
 
@@
 
-## Possible points systems
-
-There are a number of plausible points systems that might be used:
-
 \begin{itemize}
-\item Equal points to the top $r$ competitors
-$(\underbrace{1,1,\ldots ,1}_{r},0,0,\ldots)$, noting that
-$r=1$ corresponds to a ``winner takes all'' system
-\item Linear decrease $(r,r-1,r-2,\ldots,3,2,1,0)$ for integer $0<r<n$
-\item Exponential decay $(1,r,r^2,r^3,\ldots)$ for some fixed $r<1$
-\item Zipf's law $(1,1/2,1/3,1/4,\ldots)$
 \item The current Formula 1 system $(25,18,15,12,10,8,6,4,2,1)$
 \item The inaugural Formula 1 system $(8,6,4,3,2)$
+\item Zipf's law $(1,\frac{1}{2},\frac{1}{3},\frac{1}{4},\ldots)$
+\item Linear decrease $(n,n-1,n-2,\ldots,3,2,1,0)$
+\item Halving system:  $(1,\frac{1}{2},\frac{1}{4},\frac{1}{8},\ldots)$
+\item A ``winner takes all'' system $(1,0,\ldots)$
 \end{itemize}
 
-We can now calculate number of places agreed for each of the points systems:
+We note that some of these may be generalized; we might consider a
+generalized linear decrease $(r,r-1,r-2,\ldots,3,2,1,0)$ for fixed
+integer $0<r<n$; the halving system can be generalized to a geometric
+distribution; and the winner takes all system can be replaced by
+giving equal points to the top $r$ competitors
+$(\underbrace{1,1,\ldots ,1}_{r},0,0,\ldots)$, for some integer $r<n$.
 
@@
 colnames(agreement4yearsallp) <- 2016:2019
+wanted <- c(1,2,24,23,3,29)
 ```
@@
 ```{r}
-agreement4yearsallp
+agreement4yearsallp[wanted,]
 ```
 
-Observing that both the likelihood ranking and the points ranking are
+There is no points system that is the best for all years.  However,
+observing that both the likelihood ranking and the points ranking are
 random variables in this paradigm suggests a method whereby we can
@@
 
-```{r,label=somesummaries}
+```{r,label=somesummaries,echo=FALSE}
 summarytablemaker <- function(OO){
@@
 
-```{r,fig=TRUE}
-typecol <- c(rep("red",2),rep("orange",10),rep("yellow",11),rep("green",1),rep("blue",7),rep("purple",7))
+```{r,label=alottaplotsmeancorrect,echo=FALSE,fig=TRUE}
 jj <- cbind(s2016[,1],s2017[,1],s2018[,1],s2019[,1])  # mean
-matplot(jj,type="n",xlab='points system',ylab='mean correct places')
-points(s2016[,1],col=typecol,pch=1)
-points(s2017[,1],col=typecol,pch=2)
-points(s2018[,1],col=typecol,pch=3)
-points(s2019[,1],col=typecol,pch=4)
+
+#typecol <- c(rep("red",2),rep("orange",10),rep("yellow",11),rep("green",1),rep("blue",7),rep("purple",7))
+#matplot(jj,type="n",xlab='points system',ylab='mean correct places')
+#points(s2016[,1],col=typecol,pch=1)
+#points(s2017[,1],col=typecol,pch=2)
+#points(s2018[,1],col=typecol,pch=3)
+#points(s2019[,1],col=typecol,pch=4)
+#legend("topright",lty=0,pch=1:4,legend=c("2016","2017","2018","2019"))
+
+matplot(jj[wanted,],type="n",xlab='points system',ylab='mean correct places',axes=FALSE)
+axis(1,at=1:6,labels=names(formula1_points_systems(top)[wanted]))
+axis(2,at=1:4)
+points(s2016[wanted,1],pch=1)
+points(s2017[wanted,1],pch=2)
+points(s2018[wanted,1],pch=3)
+points(s2019[wanted,1],pch=4)
 legend("topright",lty=0,pch=1:4,legend=c("2016","2017","2018","2019"))
@@
 
-```{r,fig=TRUE}
+```{r,label=alottaplotswinnercorrect,echo=FALSE,fig=TRUE}
 jj <- cbind(s2016[,2],s2017[,2],s2018[,2],s2019[,2])  # prob of winner
-matplot(jj,type="n",xlab='points system',ylab='probability of winner correct')
-points(s2016[,2],col=typecol,pch=1)
-points(s2017[,2],col=typecol,pch=2)
-points(s2018[,2],col=typecol,pch=3)
-points(s2019[,2],col=typecol,pch=4)
-legend("topright",lty=0,pch=1:4,legend=c("2016","2017","2018","2019"))
+#matplot(jj,type="n",xlab='points system',ylab='probability of winner correct')
+#points(s2016[,2],col=typecol,pch=1)
+#points(s2017[,2],col=typecol,pch=2)
+#points(s2018[,2],col=typecol,pch=3)
+#points(s2019[,2],col=typecol,pch=4)
+#legend("topright",lty=0,pch=1:4,legend=c("2016","2017","2018","2019"))
+
+matplot(jj[wanted,],type="n",xlab='points system',ylab='probability of winner correct',axes=FALSE)
+axis(1,at=1:6,labels=names(formula1_points_systems(top)[wanted]))
+axis(2)
+points(s2016[wanted,2],pch=1)
+points(s2017[wanted,2],pch=2)
+points(s2018[wanted,2],pch=3)
+points(s2019[wanted,2],pch=4)
+legend(x=3.4,y=0.67,lty=0,pch=1:4,legend=c("2016","2017","2018","2019"))
 ```
 
-```{r,fig=TRUE}
+```{r,label=alottaplotstotalrankorder,echo=FALSE,fig=TRUE}
 jj <- cbind(s2016[,3],s2017[,3],s2018[,3],s2019[,3])  # total rank order correct
-matplot(jj,type="n",xlab='points system',ylab='probability of total rank order correct')
-points(s2016[,3],col=typecol,pch=1)
-points(s2017[,3],col=typecol,pch=2)
-points(s2018[,3],col=typecol,pch=3)
-points(s2019[,3],col=typecol,pch=4)
-legend("topright",lty=0,pch=1:4,legend=c("2016","2017","2018","2019"))
+#matplot(jj,type="n",xlab='points system',ylab='probability of total rank order correct')
+#points(s2016[,3],col=typecol,pch=1)
+#points(s2017[,3],col=typecol,pch=2)
+#points(s2018[,3],col=typecol,pch=3)
+#points(s2019[,3],col=typecol,pch=4)
+#legend("topright",lty=0,pch=1:4,legend=c("2016","2017","2018","2019"))
+
+matplot(jj[wanted,],type="n",xlab='points system',ylab='probability of total rank order correct',axes=FALSE)
+axis(1,at=1:6,labels=names(formula1_points_systems(top)[wanted]))
+axis(2)
+points(s2016[wanted,3],pch=1)
+points(s2017[wanted,3],pch=2)
+points(s2018[wanted,3],pch=3)
+points(s2019[wanted,3],pch=4)
+legend("topleft",lty=0,pch=1:4,legend=c("2016","2017","2018","2019"))
 ```
-
-```{r,makeatable}
-table(real_winner_ok=OO[1,]>0,top3_winner_ok=OO[3,]>0)
-```
-
-Consider the first column of the table above which shows the number of
-virtual races in which each scoring system gets the winner incorrect;
-a good points system has a low value here.  Note that `top3` has by
-far the lowest number, a finding which I have no explanation for.
 

### 2020-07-29 inst/f1points.Rmd
work

@@
 \mbox{Bottas}\succ
-\mbox{Raikkonen}\succ
+\mbox{Raikkonen}\succ\mbox{YES I KNOW}
 \mbox{Ricciardo}\succ
@@
 \mbox{Sainz}\succ
-\mbox{Hulkenberg}\succ
+\mbox{Hulkenberg}\mbox{YES I KNOW}\succ
 \mbox{Massa}
@@
 \mbox{Verstappen}\succ
-\mbox{Raikkonen}\succ\\
+\mbox{Raikkonen}\succ\mbox{YES I KNOW}\\
 \mbox{Perez}\succ
@@
 \mbox{Sainz}\succ
-\mbox{Hulkenberg}\succ
+\mbox{Hulkenberg}
 \end{multline}
@@
 jj[] <- seq_along(jj)
+jj
 ```
+
+\begin{multline}
+\mbox{Hamilton}\succ
+\mbox{Vettel}\succ
+\mbox{Bottas}\succ
+\mbox{Raikkonen}\mbox{YES I KNOW}\succ
+\mbox{Ocon}\succ
+\mbox{Ricciardo}\succ\\
+\mbox{Verstappen}\succ
+\mbox{Perez}\succ
+\mbox{Massa}\succ
+\mbox{Sainz}\succ
+\mbox{Hulkenberg}\mbox{YES I KNOW}
+\end{multline}
 
@@
 first 11 drivers can increase the rank of a driver if non-first-11
-drivers are placed} , shown in Figure
+drivers are placed}, shown in Figure
 \ref{orderingbypointsandlikelihood}.  We thus see a comparison between
@@
 WRONG: RMARKDOWN IS DEFECTIVE and Bottas are third and fourth, and
-fourth and third, according to points and likelihood respectively.
-The degree of agreement between the two systems might be quantified as
-"2", as the positions of the top two drivers agree.  Thus the degree
-of agreement between likelihood ranking and points ranking for the
-years 2016-9 would be 8,2,6, and 2 respectively.
+fourth and third, according to points and likelihood respectively.  We
+define the \emph{degree of agreement} between the two ranking systems
+as the maximal value of $n$ such that places $1,2,\ldots n$ all match.
+Thus, from figure \ref{orderingbypointsandlikelihood}, the degree of
+agreement between likelihood ranking and points ranking for the years
+2016-2019 would be 8,2,6, and 2 respectively.
 
@@
 in which we see generally poorer agreement between points-based ranks
-and likelihood-based ranks.  There are a number of plausible points
-systems that might be used:
+and likelihood-based ranks, with a degree of agreement of 4,2,2,2 for
+the years 2016-2019 respectively.  This might be an indication that
+using a Zipf law points allocation is objectively worse than the
+current points system.   There are a number of plausible points systems
+that might be used:
 
@@
 
-```{r,label=likevspointszipf,cache=TRUE,echo=FALSE,fig.cap="As for figure \ref{likevspoints} \\label{orderingbypointsandlikelihood_zipf} but points ranking calculated by Zipf's law.
+```{r,label=likevspointszipf,cache=TRUE,echo=FALSE,fig.cap="As for figure \ref{orderingbypointsandlikelihood} \\label{orderingbypointsandlikelihood_zipf} but points ranking calculated by Zipf's law.
 Note the generally poorer agreement between the two ranking systems ALSO R PLOTTING SUCKS HERE AND I WILL HAVE TO CHANGE THE PDF BY HAND WITH INKSCAPE TO MAKE IT LOOK GOOD"}
@@
 We note that some of these may be generalized; we might consider a
-generalized linear decrease $(r,r-1,r-2,\ldots,3,2,1,0)$ for fixed
-integer $0<r<n$; the halving system can be generalized to a geometric
-distribution; and the winner takes all system can be replaced by
-giving equal points to the top $r$ competitors
+more general linear points system $(r,r-1,r-2,\ldots,3,2,1,0)$ for
+fixed integer $0<r<n$; the halving system can be generalized to a
+geometric distribution; and the winner takes all system can be
+replaced by giving equal points to the top $r$ competitors
 $(\underbrace{1,1,\ldots ,1}_{r},0,0,\ldots)$, for some integer $r<n$.
@@
 
+It is straightforward to calculate the degree of agreement for the
+observed rank table for years 2016-2019, shown in the following table:
+
 ```{r}
@@
 
-There is no points system that is the best for all years.  However,
-observing that both the likelihood ranking and the points ranking are
-random variables in this paradigm suggests a method whereby we can
-objectively assess a given points system.  Using sampling techniques
-we can repeatedly generate an order table \emph{in silico}, using
-estimated driver strengths from the observed tables.  For each of,
-say, 1000 such synthetic tables, calculate drivers' maximum likelihood
-\citeauthor{plackett1975} strengths, and also their points awarded
-according to any given points system.  We then compare rankings
-generated by the \citeauthor{plackett1975} strengths and the points
-awarded and note the degree of agreement between the two.  This
+It is clear that there is no points system that is the best for all
+four years.  However, observing that both the likelihood ranking and
+the points ranking are random variables in this paradigm suggests a
+method whereby we can objectively assess a given points system.  Using
+sampling techniques we can repeatedly generate an order table \emph{in
+silico}, using estimated driver strengths from the observed tables.
+For each of, say, 1000 such synthetic tables, calculate drivers'
+maximum likelihood \citeauthor{plackett1975} strengths, and also their
+points awarded according to any given points system.  We then compare
+rankings generated by the \citeauthor{plackett1975} strengths and the
+points awarded and note the degree of agreement between the two, as
+measured by the number of rankings correctly predicted.  This
 furnishes an objective assessment of the points system used.
 
-# Numerical results
-
-We now assess the various points systems discussed above, using 1000
-_in silico_ trials.
+## Numerical results
 
@@
 ```
-
-now some summaries:
 
@@
 
-```{r,label=alottaplotsmeancorrect,echo=FALSE,fig=TRUE}
+```{r,label=alottaplotsmeancorrect,echo=FALSE,fig=TRUE,fig.cap="Simulated Formula 1 races, seasons 2016-2019: mean number of agreeing places for each of six points \\label{simulated_mean} systems when compared with a Placket-Luce strength ordering"}
 jj <- cbind(s2016[,1],s2017[,1],s2018[,1],s2019[,1])  # mean
@@
 axis(2,at=1:4)
-points(s2016[wanted,1],pch=1)
-points(s2017[wanted,1],pch=2)
-points(s2018[wanted,1],pch=3)
-points(s2019[wanted,1],pch=4)
-legend("topright",lty=0,pch=1:4,legend=c("2016","2017","2018","2019"))
+points(s2016[wanted,1],pch=1,type='b')
+points(s2017[wanted,1],pch=2,type='b')
+points(s2018[wanted,1],pch=3,type='b')
+points(s2019[wanted,1],pch=4,type='b')
+#legend("topright",lty=0,pch=1:4,legend=c("2016","2017","2018","2019"))
+par(xpd=TRUE)
+text(x=6,y=s2016[wanted[6],1],pos=4,"2016")
+text(x=6,y=s2017[wanted[6],1],pos=4,"2017")
+text(x=6,y=s2018[wanted[6],1],pos=4,"2018")
+text(x=6,y=s2019[wanted[6],1],pos=4,"2019")
 ```
 
-```{r,label=alottaplotswinnercorrect,echo=FALSE,fig=TRUE}
+```{r,label=alottaplotswinnercorrect,echo=FALSE,fig=TRUE,fig.cap="Simulated Formula 1 races, seasons 2016-2019: probability of identifying the Plackett-Luce winner for each of six points \\label{simulated_winner_correct} systems"}
 jj <- cbind(s2016[,2],s2017[,2],s2018[,2],s2019[,2])  # prob of winner
@@
 axis(2)
-points(s2016[wanted,2],pch=1)
-points(s2017[wanted,2],pch=2)
-points(s2018[wanted,2],pch=3)
-points(s2019[wanted,2],pch=4)
-legend(x=3.4,y=0.67,lty=0,pch=1:4,legend=c("2016","2017","2018","2019"))
+points(s2016[wanted,2],pch=1,type='b')
+points(s2017[wanted,2],pch=2,type='b')
+points(s2018[wanted,2],pch=3,type='b')
+points(s2019[wanted,2],pch=4,type='b')
+#legend(x=3.4,y=0.67,lty=0,pch=1:4,legend=c("2016","2017","2018","2019"))
+par(xpd=TRUE)
+text(x=6,y=s2016[wanted[6],2],pos=4,"2016")
+text(x=6,y=s2017[wanted[6],2],pos=4,"2017")
+text(x=6,y=s2018[wanted[6],2],pos=4,"2018")
+text(x=6,y=s2019[wanted[6],2],pos=4,"2019")
 ```
 
-```{r,label=alottaplotstotalrankorder,echo=FALSE,fig=TRUE}
+```{r,label=alottaplotstotalrankorder,echo=FALSE,fig=TRUE,fig.cap="Simulated Formula 1 races, seasons 2016-2019: probability of complete agreement between Plackett-Luce ranks and ranks calculated by each of six points \\label{simulated_complete_agreement} systems"}
 jj <- cbind(s2016[,3],s2017[,3],s2018[,3],s2019[,3])  # total rank order correct
@@
 axis(2)
-points(s2016[wanted,3],pch=1)
-points(s2017[wanted,3],pch=2)
-points(s2018[wanted,3],pch=3)
-points(s2019[wanted,3],pch=4)
-legend("topleft",lty=0,pch=1:4,legend=c("2016","2017","2018","2019"))
+points(s2016[wanted,3],pch=1,type='b')
+points(s2017[wanted,3],pch=2,type='b')
+points(s2018[wanted,3],pch=3,type='b')
+points(s2019[wanted,3],pch=4,type='b')
+#legend("topleft",lty=0,pch=1:4,legend=c("2016","2017","2018","2019"))
+par(xpd=TRUE)
+text(x=5,y=s2016[wanted[5],3],pos=3,"2016")
+text(x=5,y=s2017[wanted[5],3]-0.01,pos=1,"2017")
+text(x=5,y=s2018[wanted[5],3],pos=3,"2018")
+text(x=5,y=s2019[wanted[5],3],pos=3,"2019")
 ```
+
+We now assess the six points systems using the methodology outlined
+above, using 1000 _in silico_ trials.  For each of the six points
+systems, each of the 1000 trials results in a single non-negative
+integer: the degree of agreement between the Plackett-Luce ranking and
+the rankings according to the points system considered.  There are
+three measures that might be used to assess the distribution of degree
+of agreement: (1), the mean degree of agreement $\overline{d}$; (2),
+the probability of correctly predicting the winner,
+$\operatorname{Pr}(d\geqslant 1)$; and (3), the probability of
+correctly predicting the complete order statistic
+$\operatorname{Pr}(D=n)$.  These summaries are shown for each of the
+six points systems in figures \ref{simulated_mean},
+\ref{simulated_winner_correct} and \ref{simulated_complete_agreement}
+respectively.
 

### 2020-07-29 inst/eurovision.Rmd
make eurovision hyper2 pnames be character, not factor
more detail

@@
 
-competitors <- jj[1:18]
+competitors <- as.character(jj[1:18])
 colnames(preference) <- jj      # voters; 20 countries (18 + DE + UK)
@@
 other countries (`CZ`, `BE`, `SW`, `CH`, `BG`, `PT`) did not receive
-any points from Montenegro.  We need to convert matrix `preference`
-into a likelihood function, here `eurovision2009`:
+any points from Montenegro.  The rows show the votes cast by each
+country; note the final two rows corresponding to votes cast by
+Germany and the UK, who did not compete and therefore cast votes for
+all competitors.
+
+We need to convert matrix `preference` into a likelihood function,
+here `eurovision2009`:
 

### 2020-07-29 inst/curling.Rmd
fixes  #58

@@
 
-```{r,definsupportb}
-H1 <- ordertable2supp(ct)
-head(H1)
-(mH1 <- maxp(H1))
+```{r,definsupportb,cache=TRUE}
+H1 <- ordertable2supp(ranktable_to_ordertable(wikitable_to_ranktable(ct)))
+H1
+mH1 <- maxp(H1)
 pie(mH1)
@@
 plot(log10(mH1),log10(mH2),asp=1,pch=16,xlim=c(-2.6,0),ylim=c(-2.6,0))
-posn <- c(Canada=1, China=2, Denmark=4, Germany=4, GB=4, Italy=1, Japan=4, Norway=4,
-          Russia=2, South_Korea=4, Sweden=2, Switzerland=2, US=4)
+posn <- c(Canada=2, China=4, Denmark=4, Germany=4, GB=4, Italy=1, Japan=2, Norway=2,
+          Russia=4, South_Korea=4, Sweden=4, Switzerland=2, US=4)
 for(i in seq_len(nrow(curling_table))){
@@
 
-Note that the two strongest competitors, Canada and Sweden, switch
-between first and second in the two systems.
+The two likelihood systems have very different evaluates.

### 2020-07-30 inst/f1points.Rmd
more work

@@
     affiliation: Auckland University of Technology
-    address: >
+    address:
       2-14 Wakefield Street
@@
 abstract: >
-  The point scoring system of Formula 1 motor racing is a long-standing
-  contention among fans and competitors, with changes to the points
-  scored occurring frequently.  Here, I apply generalized Plackett-Luce
-  likelihood functions to Formula 1 racing, and assess historical points
-  scoring systems objectively.  In a well-defined sense, the points
-  scoring system of 1985 is the best of the historical ones.  I make
-  similar observations about related competitive situations including
-  the Eurovision Song contest and XXX.
+  The points scoring system of Formula 1 motor racing is a long-standing
+  contention among fans and competitors, with changes being controversial.
+  The inaugural points system, in use from 1950
+  to 1959 was $(8,6,4,3,2)$: 8 points to the winner, 6 to second
+  place, and so on.  In 2020 the points system was changed to
+  $(25,18,15,12,10,8,6,4,2,1)$.  The overall ranking of the
+  competitors is determined by summing points awarded over a season;
+  but this system suffers from the defect that the points system
+  itself is intrinsically arbitrary.  Here, I apply generalized
+  Plackett-Luce likelihood functions to Formula 1 motor racing, and
+  use maximum likelihood techniques to identify a ranking for the competitors;
+  this method includes no use of an arbitrary points system.
+  Using this methodology, I assess a number of reasonable points
+  scoring systems objectively.  In a well-defined statistical sense, a
+  ``winner-takes-all'' points system outperforms the other points
+  systems considered.
 keywords: 
   # at least one keyword must be supplied
-  formatted: [keywords, not capitalized, "\\proglang{Java}"]
-  plain:     [keywords, not capitalized, Java]
+  formatted: [Bradley-Terry, Plackett-Luce, order statistics, rank
+    statistics, points systems, Formula 1]
+  plain:     [Bradley-Terry, Plackett-Luce, order statistics, rank
+    statistics, points systems, Formula 1]
 preamble:
   \usepackage{amsmath}
+  \usepackage{amssymb}
 output: rticles::jss_article
@@
 \ref{simulated_winner_correct} and \ref{simulated_complete_agreement}
-respectively.
+respectively.  These figures differ in detail but all have one
+feature: the winner-takes-all points system consistently outperforms
+the other systems.  This is particularly evident in figure
+\ref{simulated_complete_agreement}.
 
@@
 
-Points systems are a source of lively debate from many perspectives.
-The point of a point allocation system is to rank competitors in an
-objective way; but
+Many competitive situations involve ranking competitors and one common
+method for combining repeated observations is to assign points for the
+winner, second place, third place, etc.  The overall ranking is
+decided on the basis of accumulated points.  However, statistical
+analysis made difficult by the arbitrary values of the points given.
 
+The points system used in Formula 1 motor racing is a source of lively
+debate from many perspectives, with changes being controversial.  
 Here I treat total points scored as a random variable and compare
-different point allocation schemes against both one another and
-
-but until now there is no objective way of comparing points systems.
-if one considers order tables to be a random variable.  Because the
-numerical value of points awarded for first, second, third etc, does
-not affect driver behaviour it is a reasonable requirement that
-
-Here, I assess
-several plausible points systems including the one actually used in
-Formula 1.  By several criteria, the
+different point allocation schemes against objective Plackett-Luce
+ranks.  By several criteria, a ``winner takes all'' performs better
+than other reasonable points systems.  However, such a system
+furnishes little incentive for mid-ranking competitors and as such
+might result in duller races.
 

### 2020-08-02 inst/curling.Rmd
slightly more efficient idiom
clarification

@@
 
-Thus we see that Canada came first in 1998, third in 2002, and so on;
-in 1998, Canada came first, China did not compete [`NA` meaning did
-not compete], Denmark came second, and so on.
+Thus the first row shows that Canada came first in 1998, third in
+2002, and so on; the first column shows that in 1998, Canada came
+first, China did not compete [`NA` meaning did not compete], Denmark
+came second, and so on.
 
@@
 ```{r,definsupportb,cache=TRUE}
-H1 <- ordertable2supp(ranktable_to_ordertable(wikitable_to_ranktable(ct)))
+H1 <- ordertable2supp(ct)
 H1

### 2020-08-02 inst/jester.Rmd
say explicitly how jester and maxjest are calculated
label entries, also some rewriting

@@
 ```
+
+(this file creates objects `jester` and `maxjest`, which are datasets
+available in the `hyper2` package, and documented at `jester.Rd`.
+This file takes quite a long time to run).
 
@@
 
-```{r }
+```{r,label=kickoff}
 library("hyper2",quietly=TRUE)
@@
 
-Row 1 of `a` is displayed and shows that the first respondent rated
-joke 5 at -1.65, joke 7 at -0.78, and so on; `NA` means that that
-respondent did not rank that joke.  We can perform some summary
-statistics of `a`:
+Row 1 of `a` is displayed (most entries are `NA`, signifying that
+respondent 1 did not rank that particular joke).  It shows that the
+first respondent rated joke 5 at -1.65, joke 7 at -0.78, and so on.
+We can perform some summary statistics of `a`:
 
-```{r}
+```{r,label=plothowmanyjokes}
 plot(rowSums(!is.na(a)),xlab="respondent")
@@
 
-```{r}
+```{r,label=plothowmanyrespondents}
 plot(colSums(!is.na(a)),xlab="joke index")
@@
 
-```{r}
+```{r,label=removeunratedjokes}
 dim(a)
@@
 
-```{r}
+```{r,label=transformthedataset}
 f <- function(x){
@@
 
-```{r,calculate_jester,cache=TRUE}
+```{r,label=calculate_jester,cache=TRUE}
 jester <- hyper2(pnames=colnames(a))
@@
 
-```{r,cache=TRUE}
+```{r,label=findmaxp,cache=TRUE}
 system.time(mjest <- maxp(jester,n=1))
@@
 
-```{r}
+```{r,showmjest}
 mjest
@@
 
-```{r,cache=TRUE}
-equalp.test(H,startp=indep(mjest))
+```{r,label=equalptestdoer,cache=TRUE}
+equalp.test(jester,startp=indep(mjest))
 ```

### 2020-08-02 inst/f1points.Rmd
better writing and coloured lines

@@
 Noting that there is information in the whole of the finishing order,
-not just the first across the line, we can follow \cite{plackett1975}
-and consider the runner-up to be the winner among the remaining
-competitors; and so on down the finishing order.  Without loss of
-generality, if the order of finishing were $1,2,3,4,5$, then a suitable
-\citeauthor{plackett1975} likelihood function would be
+and not just the first across the line, we can follow
+\cite{plackett1975} and consider the runner-up to be the winner among
+the remaining competitors, and so on down the finishing order.
+Without loss of generality, if the order of finishing were
+$1,2,3,4,5$, then a suitable
+\citeauthor{plackett1975}-\citeauthor{luce1959} likelihood function
+would be
 
@@
 ```{r, label=calcpoints_real, echo=FALSE, cache=TRUE}
+```{r, label=calcpoints_real, echo=FALSE, cache=TRUE}
 p <- formula1_points_systems()[[1]]   # real points system
@@
 	```{r,label=likevspointszipf,cache=TRUE,echo=FALSE,fig.cap="As for figure \ref{orderingbypointsandlikelihood} \\label{orderingbypointsandlikelihood_zipf} but points ranking calculated by Zipf's law.
-Note the generally poorer agreement between the two ranking systems ALSO R PLOTTING SUCKS HERE AND I WILL HAVE TO CHANGE THE PDF BY HAND WITH INKSCAPE TO MAKE IT LOOK GOOD"}
+REF DOES NOT WORK HERE  Note the generally poorer agreement between the two ranking systems ALSO R PLOTTING SUCKS HERE AND I WILL HAVE TO CHANGE THE PDF BY HAND WITH INKSCAPE TO MAKE IT LOOK GOOD"}
 par(mfrow=c(2,2))
@@
 axis(2,at=1:4)
-points(s2016[wanted,1],pch=1,type='b')
-points(s2017[wanted,1],pch=2,type='b')
-points(s2018[wanted,1],pch=3,type='b')
-points(s2019[wanted,1],pch=4,type='b')
+points(s2016[wanted,1],pch=1,type='b',col='red')
+points(s2017[wanted,1],pch=2,type='b',col='orange')
+points(s2018[wanted,1],pch=3,type='b',col='green')
+points(s2019[wanted,1],pch=4,type='b',col='blue')
 #legend("topright",lty=0,pch=1:4,legend=c("2016","2017","2018","2019"))
 par(xpd=TRUE)
-text(x=6,y=s2016[wanted[6],1],pos=4,"2016")
-text(x=6,y=s2017[wanted[6],1],pos=4,"2017")
-text(x=6,y=s2018[wanted[6],1],pos=4,"2018")
-text(x=6,y=s2019[wanted[6],1],pos=4,"2019")
+text(x=6,y=s2016[wanted[6],1],pos=4,"2016",col='red')
+text(x=6,y=s2017[wanted[6],1],pos=4,"2017",col='orange')
+text(x=6,y=s2018[wanted[6],1],pos=4,"2018",col='green')
+text(x=6,y=s2019[wanted[6],1],pos=4,"2019",col='blue')
 ```
@@
 axis(2)
-points(s2016[wanted,2],pch=1,type='b')
-points(s2017[wanted,2],pch=2,type='b')
-points(s2018[wanted,2],pch=3,type='b')
-points(s2019[wanted,2],pch=4,type='b')
+points(s2016[wanted,2],pch=1,type='b',col='red')
+points(s2017[wanted,2],pch=2,type='b',col='orange')
+points(s2018[wanted,2],pch=3,type='b',col='green')
+points(s2019[wanted,2],pch=4,type='b',col='blue')
 #legend(x=3.4,y=0.67,lty=0,pch=1:4,legend=c("2016","2017","2018","2019"))
 par(xpd=TRUE)
-text(x=6,y=s2016[wanted[6],2],pos=4,"2016")
-text(x=6,y=s2017[wanted[6],2],pos=4,"2017")
-text(x=6,y=s2018[wanted[6],2],pos=4,"2018")
-text(x=6,y=s2019[wanted[6],2],pos=4,"2019")
+text(x=6,y=s2016[wanted[6],2],pos=4,"2016",col='red')
+text(x=6,y=s2017[wanted[6],2],pos=4,"2017",col='orange')
+text(x=6,y=s2018[wanted[6],2],pos=4,"2018",col='green')
+text(x=6,y=s2019[wanted[6],2],pos=4,"2019",col='blue')
 ```
@@
 axis(2)
-points(s2016[wanted,3],pch=1,type='b')
-points(s2017[wanted,3],pch=2,type='b')
-points(s2018[wanted,3],pch=3,type='b')
-points(s2019[wanted,3],pch=4,type='b')
+points(s2016[wanted,3],pch=1,type='b',col='red')
+points(s2017[wanted,3],pch=2,type='b',col='orange')
+points(s2018[wanted,3],pch=3,type='b',col='green')
+points(s2019[wanted,3],pch=4,type='b',col='blue')
 #legend("topleft",lty=0,pch=1:4,legend=c("2016","2017","2018","2019"))
 par(xpd=TRUE)
-text(x=5,y=s2016[wanted[5],3],pos=3,"2016")
-text(x=5,y=s2017[wanted[5],3]-0.01,pos=1,"2017")
-text(x=5,y=s2018[wanted[5],3],pos=3,"2018")
-text(x=5,y=s2019[wanted[5],3],pos=3,"2019")
+text(x=5,y=s2016[wanted[5],3],pos=3,"2016",col='red')
+text(x=5,y=s2017[wanted[5],3],pos=1,"2017",col='orange')
+text(x=5,y=s2018[wanted[5],3],pos=3,"2018",col='green')
+text(x=5,y=s2019[wanted[5],3],pos=3,"2019",col='blue')
 ```

### 2020-08-03 inst/f1points.Rmd
minor tweaks

@@
   but this system suffers from the defect that the points system
-  itself is intrinsically arbitrary.  Here, I apply generalized
-  Plackett-Luce likelihood functions to Formula 1 motor racing, and
-  use maximum likelihood techniques to identify a ranking for the competitors;
-  this method includes no use of an arbitrary points system.
-  Using this methodology, I assess a number of reasonable points
-  scoring systems objectively.  In a well-defined statistical sense, a
-  ``winner-takes-all'' points system outperforms the other points
-  systems considered.
+  itself is intrinsically arbitrary: the points system does not affect
+  the competitors' behaviour but can change the overall ranking.
+  Here, I calculate generalized Plackett-Luce likelihood functions for
+  Formula 1 motor racing, and use maximum likelihood techniques to
+  identify a ranking for the competitors; this method includes no
+  arbitrary points system. Using this methodology, I assess a number
+  of reasonable points scoring systems  objectively,  by comparing
+  likelihood-based rankings with points-based rankings.  In one
+  well-defined statistical sense, a linear  points system outperforms
+  the other points systems considered, and in another the ``winner
+  takes all'' system is the best.
 keywords: 
@@
 Formula 1 racing is an important and prestigious motor sport
-\citep{moul2009}.  Season ranking is based on a points allocation
+\citep{test}.  Season ranking is based on a points allocation
 system wherein competitors are awarded points based on race finishing
@@
 from 1950 to 1959 was $(8,6,4,3,2)$, thus crediting only the first
-five finishers.  As of 2020, the current system of
+five finishers.  As of 2020, the current points system of
 $(25,18,15,12,10,8,6,4,2,1)$ credits the first 10.  Arguably these two
@@
 the second, which does.  Still, it is reasonable to assume that each
-driver strives to maximise his rank.
+driver strives to maximise his rank and this will be done here.  If
+this is so, then changing the points system might change the
+competitors' rankings but not their behaviour: surely a defect of
+using points to rank the competitors.
 
@@
 \citet{luce1959}, in which the probability of competitor $i$ winning
-in a field of $j=1,\ldots, n$ is $\frac{p_i}{p_1+\cdots +p_n}$.
-Noting that there is information in the whole of the finishing order,
-and not just the first across the line, we can follow
-\cite{plackett1975} and consider the runner-up to be the winner among
-the remaining competitors, and so on down the finishing order.
-Without loss of generality, if the order of finishing were
-$1,2,3,4,5$, then a suitable
-\citeauthor{plackett1975}-\citeauthor{luce1959} likelihood function
-would be
+in a field of $1,\ldots, n$ is $\frac{p_i}{p_1+\cdots +p_n}$.  Noting
+that there is information in the whole of the finishing order, and not
+just the first across the line, we can follow \cite{plackett1975} and
+consider the runner-up to be the winner among the remaining
+competitors, and so on down the finishing order.  Without loss of
+generality, if the order of finishing were $1,2,3,4,5$, then a
+suitable \citeauthor{plackett1975}-\citeauthor{luce1959} likelihood
+function would be
 
@@
 We see in figure \ref{piechartstrength} that the driver with the
-largest estimated strength was Rosberg in 2016 (at about 49\%) and in
-years 2017-8-9 was Hamilton (at about 29\%, 35\%, and 50\%
-respectively).  However, it should be noted that a likelihood ratio
-test [`samep.test()`, supplied with the package] does not reject, for
-example, the null that Hamilton and Vettel have the same strength in
-2017 ($H_0\colon \pham= \pvet$) with a likelihood ratio of
-$e^{1.5}\simeq 4.48$, corresponding (by Wilks's theorem) to an
-asymptotic $p$-value of about $0.08$.
+largest estimated strength was Rosberg in 2016 at about 30\%, and in
+years 2017-8-9 was Hamilton (at about 29\%, 37\%, and 42\%
+respectively).  As an illustration of the value of likelihood methods
+(as opposed to points-based methods), a likelihood ratio test
+[`samep.test()`, supplied with the package] rejects
+the null that Hamilton and Vettel have the same strength in
+2018 ($H_0\colon \pham= \pvet$) with a likelihood ratio of
+$e^{2.76}\simeq 15.8$, corresponding (by Wilks's theorem) to an
+asymptotic $p$-value of about $0.02$.
 
 ```{r,label=dosameptest, echo=FALSE, print=FALSE, cache=TRUE}
-ignore <- samep.test(H2017,1:2)
+ignore <- samep.test(H2018,c("Hamilton","Vettel"))
 ```
-
 ## Likelihood scoring vs points scoring
@@
 define the \emph{degree of agreement} between the two ranking systems
-as the maximal value of $n$ such that places $1,2,\ldots n$ all match.
+as the maximal value of $r$ such that places $1,2,\ldots r$ all match.
 Thus, from figure \ref{orderingbypointsandlikelihood}, the degree of
@@
 \ref{simulated_winner_correct} and \ref{simulated_complete_agreement}
-respectively.  These figures differ in detail but all have one
-feature: the winner-takes-all points system consistently outperforms
-the other systems.  This is particularly evident in figure
-\ref{simulated_complete_agreement}.
+respectively.
+
+We see that the linear points system $(n,n-1,n-2,\ldots,3,2,1)$ gives
+the highest mean number of places, and also the highest probability of
+correctly predicting the winner.  However, the probability of
+predicting the complete order statistic is maximized using the winnner
+takes all system $(1,0,0,\ldots)$.
 
@@
 
-Many competitive situations involve ranking competitors and one common
-method for combining repeated observations is to assign points for the
-winner, second place, third place, etc.  The overall ranking is
-decided on the basis of accumulated points.  However, statistical
-analysis made difficult by the arbitrary values of the points given.
+Many competitive situations involve a ranking of the participants and
+one common method for combining repeated observations is to assign
+points for the winner, second place, third place, etc.  The overall
+ranking is decided on the basis of accumulated points.  However,
+points systems are unavoidably arbitrary.  Assuming that participants
+attempt to optimize their placing means that the arbitrary
+nature of any points system.
+
+Maximum likelihood estimation of participants' Plackett-Luce strengths
+furnishes a ranking system that does not suffer from the arbitrariness
+of a points system.
 
 The points system used in Formula 1 motor racing is a source of lively
-debate from many perspectives, with changes being controversial.  
-Here I treat total points scored as a random variable and compare
-different point allocation schemes against objective Plackett-Luce
-ranks.  By several criteria, a ``winner takes all'' performs better
-than other reasonable points systems.  However, such a system
-furnishes little incentive for mid-ranking competitors and as such
-might result in duller races.
+debate from many perspectives, with changes being controversial.  Here
+I treat total points scored as a random variable and compare different
+point allocation schemes against objective Plackett-Luce ranks.  Of
+the six points systems considered here, a linear system or a
+winner-takes-all-system appear to be closest to objective
+Plackett-Luce, depending on the exact definition of ``closest''.
 

### 2020-08-03 inst/retain.Rmd
bugfix, last column interpreted wrongly as points
points -> misslast
get rid of that stupid 'misslast' argument
update to reflect removal of keep()

@@
 
-Functions `keep()` and `keep_flawed()` can be confusing, here is a
-tiny example that should clarify things.  We will use a subset of
-`skating_table`:
+Function `keep_flawed()` can be confusing.  Originally, the package
+included a function `keep()` but this used flawed logic, hence the
+name of its replacement.  Here is a tiny example that should clarify
+things.  We will use a subset of `skating_table`:
 
@@
 
-Object `jj` does not function as a rank table because the entries are
-not sequential.  We need to convert it:
+Above, rows of object `jj` correspond to competitors and columns to
+judges.  Object `jj` does not function as a rank table because the
+entries are not sequential.
+
+We could convert explicitly:
 
@@
 ```{r}
-OT <- ordertable2supp(a,misslast=FALSE)
+OT <- ordertable2supp(a)
 OT
@@
 
-[note carefully how `misslast=FALSE` is needed here].  Now suppose we
-are interested only in the first two competitors, `a` and `b`.  There
-are two ways to discard competitor `c`, namely `keep()` and
-`keep_flawed()`, which do slightly different things.  First, `keep()`:
+(function `ordertable2supp()` does this automatically, so
+`ordertable2supp(jj)` would have given the same result).  Now suppose
+we are interested only in the first two competitors, `a` and `b`,
+which are competitors 1 and 2.
 
@@
 nevertheless interpreting joint strength term on the denominator [viz
-`a+c`] as a *loss* for `a`.  We will now use function `keep()`, which
-operates on order tables:
+`a+c`] as a *loss* for `a`.
+
+What happens if we keep only `a`?
 
 ```{r}
-RA <- keep(a,1:2)
-RA
-```
-
-Looking at object `ra` we see six comparisons, of which `a` wins 5 and
-`b` wins 1.
-
-```{r}
-OR <- ordertable2supp(RA,misslast=FALSE)
-OR
-```
-
-Above, object `or` is a perfectly sensible likelihood function.  What
-happens if we keep only `a`?
-
-```{r}
-keep_flawed(OR,1)
+keep_flawed(OT,1)
 ```

### 2020-08-03 inst/loser.Rmd
label chunks and cache the slow ones

@@
 
-```{r}
+```{r,label=useggrl}
 n <- 6 # number of competitors
@@
 
-```{r}
+```{r,label=executemaxp,cache=TRUE}
 (ans_unconstrained <- maxp(L,give=TRUE))
@@
 
-```{r}
+```{r,label=constrainthething,cache=TRUE}
 small <- 1e-2  # numerical necessity
@@
 
-```{r}
+```{r,label=likesingle}
 like_single_list(indep(equalp(L[[1]])),L)
@@
 
-```{r}
+```{r,label=useqchis}
 qchisq(0.95,1)

### 2020-08-03 inst/moto.Rmd
use cache

@@
 Each row is a driver (rider?) and each column (after the first) a
-venue.  Takin the first row we see that Marquez came second in Qatar,
+venue.  Taking the first row we see that Marquez came second in Qatar,
 first in Argentina, retired in Austin, and so on.  In the first
@@
 
-```{r,maxp2017}
+```{r,maxp2017,cache=TRUE}
 options(width = 100)

### 2020-08-10 inst/formula1.Rmd
typo
bugfix

@@
 Formula 1 racing is an important and prestigious motor sport.  Here I
-analyse a series of racing results using the `hyper` package which
+analyse a series of racing results using the `hyper2` package which
 implements the Plackett-Luce likelihood function.
@@
 ```{r,fouryearsdata, cache=TRUE}
-F1_2014 <- "formula1_2014.txt" %>% read.table(header=TRUE) %>% ordertable2supp
-F1_2015 <- "formula1_2015.txt" %>% read.table(header=TRUE) %>% ordertable2supp
-F1_2016 <- "formula1_2016.txt" %>% read.table(header=TRUE) %>% ordertable2supp
-F1_2017 <- "formula1_2017.txt" %>% read.table(header=TRUE) %>% ordertable2supp
+
+jj <- read.table("formula1_2014.txt",header=TRUE)
+F1_2014 <- ordertable2supp(jj[,seq_len(ncol(jj)-1)], incomplete = TRUE)
+
+jj <- read.table("formula1_2015.txt",header=TRUE)
+F1_2015 <- ordertable2supp(jj[,seq_len(ncol(jj)-1)], incomplete = TRUE)
+
+jj <- read.table("formula1_2016.txt",header=TRUE)
+F1_2016 <- ordertable2supp(jj[,seq_len(ncol(jj)-1)], incomplete = TRUE)
+
+jj <- read.table("formula1_2017.txt",header=TRUE)
+F1_2017 <- ordertable2supp(jj[,seq_len(ncol(jj)-1)], incomplete = TRUE)
+
 a <- list(F1_2014, F1_2015, F1_2016, F1_2017)

### 2020-08-10 inst/universities.Rmd
more work

@@
 
-```{r, coerceem}
-f <- function(y,supp=TRUE,...){
-  jj <- as.data.frame(a)
-  jj <- subset(jj,jj$year==y)
-  o <- function(v){
+```{r, label=definehelperfunctionf}
+f <- function(a,y){
+  a <- subset(a,a$year==y)
+  o <- function(v){  # 'v' is a character vector
     good <- v != "DNP"
-     v[good] <- order(v[good],decreasing=TRUE)
+    v[good] <- rank(-as.numeric(v[good]),ties.method="last")
     return(v)
   }
-  x <- apply(jj,2,o)
-  rownames(x) <- jj[,1]
-  x <- x[,3:6]
-  if(supp){
-    return(ordertable2supp(x,...))
-  } else {
-  return(as.data.frame(x))
+  out <- data.frame(o(a[,3]),o(a[,4]),o(a[,5]),o(a[,6]),o(a[,7]))
+  colnames(out) <- colnames(a)[3:7]
+  rownames(out) <- a[,1]
+  return(out)
 }
-}
-f(2012)
+f(a,2012)
 ```
@@
 ```{r,playerwithadvantage}
-H <- f(2013) + f(2014) + f(2015)                          # pre  2016
-H1 <- H + f(2016) + f(2017) + f(2018) + f(2019) + f(2020) # post 2016
+H <- hyper2(pnames=a[1:8,1])
+for(i in 2013:2020){ H %<>% `+`(ordertable2supp(f(a,i)))}
+H
 ```
 
-Object `H1` is simply the support function for the universities'
-strenghts for observations (assumed to be independent) 2013-2020:
+Object `H` is simply the support function for the universities'
+strengths for all years' observations (assumed to be independent)
+2013-2020:
 
-```{r}
-maxp(H)
+```{r,label=maxpcalc,cache=TRUE}
+m <- maxp(H)
+m
+pie(m)
 ```
@@
 
-```{r}
-pnames(H) <- c(pnames(H),"S")
-H2 <- H + pwa(f(2016),"AUT") + pwa(f(2017),"AUT") +
-          pwa(f(2018),"AUT") + pwa(f(2019),"AUT") +
-  	  pwa(f(2020),"AUT")
-head(H2)
-```
+```{r,label=preandpost}
 
-Object `H2` is a support function for the Bradley-Terry strengths of
-the universities, but with AUT having extra strength `S` post-2016.
-We can test the null that `S=0` using package idiom
-\code{specificp.test()}:
+H1 <- hyper2(pnames=c("AUT_pre",a[c(1:2,4:8),1]))
+for(i in 2012:2016){
+  b <- f(a,i)
+  rownames(b)[3] <- "AUT_pre"
+  H1 <- combine(H1,ordertable2supp(b))
+} # column loop closes
 
-```{r}
-specificp.test(H2,"S",0)
+H2 <- hyper2(pnames=c("AUT_post",a[c(1:2,4:8),1]))
+for(i in 2017:2020){
+  jj <- f(a,i)
+  rownames(jj)[3] <- "AUT_post"
+  H2 <- combine(H2,ordertable2supp(jj))
+}
+maxp(H2)
 ```
@@
 significant at 1\%.
+
+```{r}
+H <- combine(H1,H2)
+maxp(H)
+samep.test(H,c(1,9))
+```

### 2020-08-10 inst/f1points.Rmd
include Jenkins

@@
 Formula 1 racing is an important and prestigious motor sport
-\citep{test}.  Season ranking is based on a points allocation
-system wherein competitors are awarded points based on race finishing
-order; points accumulate additively.  The overall competition winner
-is the competitor who accumulates the most points after the final
-race.
+\citep{codling2017,jenkins2010}.  Season ranking is based on a points
+allocation system wherein competitors are awarded points based on race
+finishing order; points accumulate additively.  The overall
+competition winner is the competitor who accumulates the most points
+after the final race.
 
@@
 
-and a slight generalization of this would allow the incorporation of
-non-finishers (DNF etc).  If, say, competitors 4 and 5 did not finish,
-we would have
+and this would be a forward ranking Plackett-Luce model in the
+terminology of \cite{johnson2020}.  A slight generalization allows the
+incorporation of non-finishers (DNF etc).  If, say, competitors 4 and
+5 did not finish, we would have
 
@@
 optimization routines, and because they have access to derivatives,
-convergence is rapid.  A graphical illustration of the strengths is
-given in figure \ref{piechartstrength}
+convergence is rapid.  A graphical diagram of the strengths is given
+in figure \ref{piechartstrength}
 
@@
 (as opposed to points-based methods), a likelihood ratio test
-[`samep.test()`, supplied with the package] rejects
-the null that Hamilton and Vettel have the same strength in
-2018 ($H_0\colon \pham= \pvet$) with a likelihood ratio of
-$e^{2.76}\simeq 15.8$, corresponding (by Wilks's theorem) to an
-asymptotic $p$-value of about $0.02$.
+[`samep.test()`, supplied with the package] rejects the null that
+Hamilton and Vettel have the same strength in 2018 ($H_0\colon \pham=
+\pvet$) with a likelihood ratio of $e^{2.76}\simeq 15.8$,
+corresponding (by Wilks's theorem) to an asymptotic $p$-value of about
+$0.02$.  We may also use the reified entity concept to test the null
+hypothesis that Hamilton's strength remains constant 2016, where
+Rosberg had the highest estimated strength, and 2017-9, where Hamilton
+did; we fail to reject this null.
+
+```{r,label=combineallyearssetup,echo=FALSE,cache=TRUE}
+H2016a <- hyper2(brackets(H2016),d=powers(H2016),pnames=sub("Hamilton","Hamilton_pre", pnames(H2016)))
+H2017a <- hyper2(brackets(H2017),d=powers(H2017),pnames=sub("Hamilton","Hamilton_post",pnames(H2017)))
+H2018a <- hyper2(brackets(H2018),d=powers(H2018),pnames=sub("Hamilton","Hamilton_post",pnames(H2018)))
+H2019a <- hyper2(brackets(H2019),d=powers(H2019),pnames=sub("Hamilton","Hamilton_post",pnames(H2019)))
+Ha <- H2016a
+Ha <- combine(Ha,H2017a)
+Ha <- combine(Ha,H2018a) 
+Ha <- combine(Ha,H2019a)
+ignorethis <- samep.test(Ha,c("Hamilton_pre","Hamilton_post"))
+```
 

### 2020-08-11 inst/red_bus_blue_bus.Rmd
red bus, blue  bus

---
title: "Red bus, blue bus"
output: pdf_document
bibliography: chess.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The red bus-blue bus problem was originally stated by
@davidson1970.  The other one was the holiday problem.

A traveller has a choice of going to work by car or taking a bus.  The
bus may be red or blue.  Here, we assume that the traveller has strong
(but unknown) views on cars vs buses.  If considering a car or bus,
the colour of the bus is of no importance to him.  
However, when offered the choice between a red bus or a blue bus, the
colour becomes important (having nothing else to guide his decision).

A sensible likelihood function is shown in figure \ref{likefuncred}.

\begin{table}[h]
\centering
\caption{Likelihoods\label{likefuncred} for red bus, blue bus problem}
\begin{tabular}{|l|l|}
\hline
Choice set & likelihood  \\ \hline
$\left\lbrace C,RB  \right\rbrace$ & $\left(\frac{p_C   }{p_C+p_{B}   }\right)^a \left(\frac{p_{B}}{p_C+p_{B}}\right)^b $ \\ \hline
$\left\lbrace C,BB  \right\rbrace$ & $\left(\frac{p_C   }{p_C+p_{B}   }\right)^c \left(\frac{p_{B}}{p_C+p_{B}}\right)^d $ \\ \hline
$\left\lbrace RB,BB \right\rbrace$ & $\left(\frac{1}{2}\right)^e 
                                      \left(\frac{1}{2}\right)^f
$ \\ \hline
$\left\lbrace C,RB,BB\right\rbrace$ & $
\left(\frac{p_C}{p_C + 2p_{B}}\right)^g
\left(\frac{p_B}{p_C + 2p_{B}}\right)^h
\left(\frac{p_B}{p_C + 2p_{B}}\right)^i
$\\ \hline
\end{tabular}
\end{table}

\begin{table}[h]
\centering
\caption{Likelihoods\label{likefuncredghost} for red bus, blue bus problem}
\begin{tabular}{|l|l|}
\hline
Choice set & likelihood  \\ \hline
$\left\lbrace C,RB  \right\rbrace$ & $\left(\frac{p_C   }{p_C+p_{B}   }\right)^a \left(\frac{p_{B}}{p_C+p_{B}}\right)^b $ \\ \hline
$\left\lbrace C,BB  \right\rbrace$ & $\left(\frac{p_C   }{p_C+p_{B}   }\right)^c \left(\frac{p_{B}}{p_C+p_{B}}\right)^d $ \\ \hline
$\left\lbrace RB,BB \right\rbrace$ & $\left(\frac{p_{B}+p_{RG}}{2p_{B}+p_{RG}}\right)^e 
                                      \left(\frac{p_{B}       }{2p_{B}+p_{RG}}\right)^f
$ \\ \hline
$\left\lbrace C,RB,BB\right\rbrace$ & $
\left(\frac{p_C       }{p_C + 2p_{B}+p_{RG}}\right)^g
\left(\frac{p_B+p_{RG}}{p_C + 2p_{B}+p_{RG}}\right)^h
\left(\frac{p_B       }{p_C + 2p_{B}+p_{RG}}\right)^i
$\\ \hline
\end{tabular}
\end{table}

 @debreu1960 offers:

Let the set U have the following three elements: $D_c$, a recording of
the Debussy quartet by the $C$ quartet, $B_F$, a recording of the
eighth symphony of Beethoven by the $B$ orchestra conducted by $F$,
$B_K$, a recording of the eighth symphony of Beethoven by the $B$
orchestra conducted by $K$.  The subject will be presented with a
subset of $U$, will be asked to choose an element in that subset, and
will listen to the recording he has chosen.  When presented with
$\left\lbrace D_c, B_F\right\rbrace$, he chooses $D_c$ with
probability $\frac{3}{5}$.  When presented with $\lbrace BF,
BK\rbrace$ he chooses $B_F$ with probability 1/2. When presented with
$\left\lbrace DC, BK\right\rbrace$ he chooses $D_c$ with probability
$\frac{3}{5}$.  What happens if he is presented with $\lbrace D_c,
B_F, B_K\rbrace$?  According to the axiom, he must choose $D_c$ with
probability 3/7.  Thus if he can choose between $D_c$ and $B_F$, he
would rather have Debussy.  However if he can choose between $D_c$,
$B_F$, and $B_K$, while being indifferent between $B_F$ and $B_K$, he
would rather have Beethoven.

THERE IS ALSO THE THING WITH THE HOLIDAY IN CALIFORNIA AND THE HOLIDAY
IN FLORIDA AND THE HOLIDAY IN FLORIDA PLUS TEN BUCKS

```{r loadlib}
library("hyper2")
2+4
```

# References

### 2020-08-22 inst/T20.Rmd
replace triple backticks with single

---
title: "The impact of winning the coin toss on the Indian Premier League"
author: "Tanvi Chandel and Robin K. S. Hankin"
date: "`r Sys.Date()`"
vignette: |
  %\VignetteIndexEntry{T20}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# The Indian Premier League

This script creates and analyses hyper2 object `T20` which is a
likelihood function for the strengths of the competitors in the Indian
Premier League.

```{r}
library("hyper2",quietly=TRUE)
head(T20_table)
nrow(T20_table)
```

Dataframe `T20_table` has one row for each T20 IPL match in the
period 2008-2017 with the exception of three no-result matches and
seven tied matches, which were removed.  Object `T20` is a
likelihood function for the strengths of the 13 teams, and
`T20_toss` is a likelihood function that also includes a toss
strength term.

some details are given in the package help file
`T20.Rd` (type `?T20` at the R prompt).

## One-way univariate analysis

What is the probability of a team winning the match, given that they
won the toss?

```{r}
x <- table(T20_table$toss_winner==T20_table$match_winner)
x
binom.test(x,alternative="less")$p.value
```

Binomial test against a null of $p=0.5$ is thus not significant: there
is no evidence that winning the toss increases one's probability of
winning the match.  Now we consider the decision (to bat first or to
field first) made by the toss winner:

```{r}
table(T20_table$toss_decision)
```

a clear majority elect to field first. We can now ask what the
probability of the batting team winning the match is:

```{r}
attach(T20_table)
bat_first_wins <- 
((toss_winner==match_winner)&(toss_decision=='bat'  )) | 
((toss_winner!=match_winner)&(toss_decision=='field'))
sum(bat_first_wins)
field_first_wins <- 
((toss_winner==match_winner)&(toss_decision=='field'  )) | 
((toss_winner!=match_winner)&(toss_decision=='bat'))
sum(field_first_wins)
detach(T20_table)
```

as a consistency check we have

```{r}
sum(field_first_wins) + sum(bat_first_wins)
nrow(T20_table)
```
agreeing as expected.

## Two-way contingency tables

We can do some contingency tables:

```{r}
M <- table(
decision           = T20_table$toss_decision,
won_toss_won_match = T20_table$toss_winner==T20_table$match_winner)
dimnames(M) <- list(
decision=c('bat first','field first'),
won_match =c(FALSE,TRUE))
M
```

We may reject homogoneity of proportion (Fisher's exact test, $p= `r round(fisher.test(M)$p.value,3)`$).  Recalling that a slight majority
of toss winners elect to field first (361 out of 633), we see that this is a consistent result.

# Strengths of the individual teams

We can now consider the matches as Bernoulli trials and try to infer
the teams' strengths.  The first step is to calculate a likelihood
function for the strengths:

```{r}
attach(lapply(T20_table[, ], as.vector))  # vector access to team1, etc
T20 <- hyper2(pnames=c(levels(T20_table$team1)))
for(i in seq_along(team1)){
   T20[match_winner[i]] %<>% inc
   T20[c(team1[i],team2[i])] %<>% dec

}		       
T20
```

and maximization is straightforward:

```{r}
dotchart(maxp(T20))
```

So we can test the null that all the teams have the same strength:

```{r}
a1 <- maxp(T20,give=TRUE)$value        # likelihood at the evaluate
a2 <- loglik(indep(equalp(T20)),T20)  # likelihood at p1=p2=...=p14
a1-a2
```

and the above would correspond to a support $\Lambda$ of about 14.5,
or a likelihood ratio of about $e^\Lambda\simeq 2\times 10^6$.  This
is not significant by Edwards's criterion of two units of support per
degree of freedom (we have 13df here).  Alternatively, we might
observe that the statistic $2\Lambda$ is in the tail region of its
asymptotic distribution $\chi^2_{13}$ given by Wilks, with a p-value
of

```{r}
pchisq(2*14.5,df=13,lower.tail=FALSE)
```

showing a significant result although frankly Edwards's criterion is
somewhat flaky in this context.  Wilks's theorem is only asymptotic
and it is not clear that the actual sampling distribution is close to
its asymptotic limit of $\chi^2$.  However, it turns out that the
asymptotic approximation is good in this particular case:
`inst/T20_analysis.Rmd` presents some analysis (it takes a very
long time to run which is why it is not included in this vignette).

# The effect of the toss

We can now attempt to see if the toss makes a difference:

```{r}
T20_toss  <- hyper2(pnames=c('toss',levels(T20_table$team1)))
for(i in seq_along(team1)){
      players <- c(team1[i],team2[i])
      if(toss_winner[i] == match_winner[i]){ # win toss, win match
         T20_toss[c('toss',match_winner[i])] %<>% inc
      } else { # win toss, lose match
         T20_toss[match_winner[i]] %<>% inc
      }			 
   T20_toss[c('toss',players)] %<>% dec
   T20_toss
}			  
```
and again we can maximize the likelihood:

```{r}
dotchart(maxp(T20_toss))
```

Above, note the small estimated strength of the toss.

### 2020-08-22 inst/eurovision.Rmd
bugfix and named object for eurovision

@@
 
-Each row corresponds to a competitor and each column to a judge.  Note
+Each row corresponds to a contestant and each column to a judge.  Note
 that the matrix is not square: the last two columns correspond to
 Germany and the UK, who were judges but not competitors.  I have
-removed the first column in the Wikipedia table (which gave the total
-of the points), replaced self-voting entries with `NA`, and replaced
-blanks with `0`.  Thus the first row corresponds to votes cast for
-`ME`, Montenegro.  We see that `BE` (Belgium) awarded them three
-points, `am` (Armenia) five points, and so on.
+removed the first column from the Wikipedia table (which gave the
+total of the points), replaced self-voting entries with `NA`, and
+replaced blanks with `0`.  Thus the first row corresponds to votes
+cast for `ME`, Montenegro.  We see that `BY` (Belarus) awarded them
+three points, `AM` (Armenia) five points, and so on.  The first column
+corresponds to the points awarded by Monetenegro.  We see that they
+placed gave Belarus (`BY`) two points, Armenia (`AM`) four points, and
+so on.  Their favourite was Bosnia and Herzegovina (`BA`) to whom they
+gave twelve points.  The points system used was:
 
@@
 
-In the above, matrix `preference` records voters' first, second, third,
-etc choice.  A zero entry means no points (nul punkte!) and `NA` means
-that voter was forbidden from voting for that player (countries cannot
-vote for themselves).
+In the above, matrix `preference` records voters' first, second,
+third, etc choice.  A zero entry means no points (nul punkte!) and
+`NA` means that voter was forbidden from voting for that player
+(countries cannot vote for themselves).  The competitors were the
+first 18 countries (the last two countries,
+Germany and the UK, voted but did not compete).
 
-The competitors were the first 18 countries (the last two countries,
-Germany and the UK, voted but did not compete).  We need the voters'
-choices to be the *rows* for consistency with the rest of the package
-(and, for that matter, the aylmer package and indeed the emulator
-package), so we take the transpose:
-
-```{r,taketranspose}
-preference <- t(preference)
+```{r}
 preference
 ```
-
-Now, take the first row of `preference`.  This represents the votes
-cast *by* (sic) Montenegro (`"ME"`).  Their favourite was [last
-column] Bosnia & Herzegovina (`"BA"`), who they gave rank 1 to.  Their
-second favourite was Macedonia, their third was Turkey, and so on.
-They were not allowed to vote for themselves, which is why the first
-column is `NA`.  So the order was: `BA` (1), `MK` (2) , `TR` (3), `IS`
-(4), `RO` (5), `IL` (6), `AM` (7), `FI` (8), `BY` (9), `MT` (10).  The
-other countries (`CZ`, `BE`, `SW`, `CH`, `BG`, `PT`) did not receive
-any points from Montenegro.  The rows show the votes cast by each
-country; note the final two rows corresponding to votes cast by
-Germany and the UK, who did not compete and therefore cast votes for
-all competitors.
+Now, take the first column of `preference`.  This represents the order
+of preferences of Montenegro (`ME`).  Their favourite was [last row]
+Bosnia & Herzegovina (`BA`), who they gave rank 1 to (that is, 12
+points).  Their second favourite was Macedonia (`MK`), their third was
+Turkey (`TR`), and so on.  They were not allowed to vote for themselves,
+which is why the first row is `NA`.  So the order was: `BA` (first),
+`MK` (second) , `TR` (third), `IS` (fourth), `RO` (fifth), `IL`
+(sixth), `AM` (seventh), `FI` (eigthth), `BY` (ninth), `MT` (tenth).
+The other countries (`CZ`, `BE`, `SW`, `CH`, `BG`, `PT`) came joint
+last and did not receive any points from Montenegro.  Note the final
+two rows corresponding to votes cast by Germany and the UK, who did
+not compete and therefore cast votes for all competitors.
 
@@
 ```{r,morestuff}
-eurovision2009 <- hyper2(d=18)
+eurovision2009 <- hyper2()
 
-for(i in seq_len(nrow(preference))){   # cycle through the rows; each row is a voter
-    d <- preference[i,,drop=TRUE]
-    d[is.na(d)] <- -1  # kludge: make the voting country ineligible to vote.
-    while(any(d>0)){
-        eligible <- which(d>=0)   # This is why we set NA values to -1
-        eurovision2009[which(d==1)] %<>% "+"(1)
-                                            # The first choice among
-                                             # eligible players has +1
-                                              # power on the numerator
-
-        eurovision2009[eligible] %<>% dec()   # denominator of all eligible players
-
-        d[d==1] <- -1  # once you've won, you are ineligible to be chosen again
-                       # NB, pipe idiom is ugly:  d %<>% `[<-`(. == 1, -1)
-
-        d[d>0] %<>% dec()  # everyone moves down the list, so who
-                            # *was* second choice becomes first
-                            # choice, who *was* third choice becomes
-                            # second, and so on.
-
-    } # while() loop closes
+for(i in seq_len(ncol(preference))){   # cycle through the rows; each row is a voter
+    d <- preference[,i,drop=TRUE]
+	eurovision2009 <- 	eurovision2009  + ordervec2supp(d[!is.na(d)])
 } # i loop closes
-
-## syntatic sugar:
-pnames(eurovision2009) <- competitors
 ```
@@
 
-```{r,checknull}
-m <- maxp(eurovision2009)  # MLE
-print(loglik(indep(m),eurovision2009))
-print(loglik(indep(equalp(eurovision2009)),eurovision2009))
+```{r,checknull,cache=TRUE}
+m <- maxp(eurovision2009)
+m
+equalp.test(eurovision2009)
 ```
@@
 oy <- rank(-m)   # m = MLE
+oyp <- ordertrans(oy,names(ox))
+
 par(pty='s') # square plot
-plot(ox,oy,asp=1,pty='s',xlim=c(0,18),ylim=c(0,18),pch=16,
+plot(ox,oyp,asp=1,pty='s',xlim=c(0,18),ylim=c(0,18),pch=16,
 xlab="official order",ylab="my order",main="Eurovision 2009")
+
 par(xpd=TRUE) # allow country names to appear outside plotting region
-for(i in seq_along(ox)){text(ox[i],oy[i],names(m)[ox[i]],pos=4,col='gray') }
+for(i in seq_along(ox)){text(ox[i],oyp[i],names(ox)[i],pos=4,col='gray') }
 par(xpd=FALSE) # stop diagonal line from protruding beyond plotting region
@@
 ```
-
-In the graph above, we see that both systems agree on the first four places.
-We can also plot the numerical values:
-
-```{r}
-plot(log(rowSums(wiki_matrix,na.rm=TRUE)),log(m),
-     pch=16,xlab="points scored",ylab="MLE strength")
-```

### 2020-08-22 data/eurovision.R

@@ -1,160 +1,263 @@
 `eurovision2009` <-
- structure(list(brackets = list(1, c(1, 2, 3, 4, 5, 6, 7, 8, 9, 
- 10, 11, 12, 13, 14, 15, 16, 17), c(1, 2, 3, 4, 5, 6, 7, 8, 9, 
- 10, 11, 12, 13, 14, 15, 16, 17, 18), c(1, 2, 3, 4, 5, 6, 7, 8, 
- 9, 10, 11, 12, 13, 14, 15, 16, 18), c(1, 2, 3, 4, 5, 6, 7, 8, 
- 9, 10, 11, 12, 13, 14, 15, 17, 18), c(1, 2, 3, 4, 5, 6, 7, 8, 
- 9, 10, 11, 12, 13, 14, 16, 17, 18), c(1, 2, 3, 4, 5, 6, 7, 8, 
- 9, 10, 11, 12, 13, 15, 16, 17, 18), c(1, 2, 3, 4, 5, 6, 7, 8, 
- 9, 10, 11, 12, 14, 15, 16, 17, 18), c(1, 2, 3, 4, 5, 6, 7, 8, 
- 9, 10, 11, 13, 14, 15, 16, 17, 18), c(1, 2, 3, 4, 5, 6, 7, 8, 
- 9, 10, 11, 13, 14, 15, 16, 18), c(1, 2, 3, 4, 5, 6, 7, 8, 9, 
- 10, 11, 13, 14, 15, 17, 18), c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 
- 11, 13, 14, 16, 17, 18), c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 
- 13, 15, 17, 18), c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 12, 13, 14, 
- 15, 16, 17, 18), c(1, 2, 3, 4, 5, 6, 7, 8, 9, 11, 12, 13, 14, 
- 15, 16, 17, 18), c(1, 2, 3, 4, 5, 6, 7, 8, 9, 11, 13, 14, 15, 
- 16, 17, 18), c(1, 2, 3, 4, 5, 6, 7, 8, 10, 11, 12, 13, 14, 15, 
- 16, 17), c(1, 2, 3, 4, 5, 6, 7, 8, 10, 11, 12, 13, 14, 15, 16, 
[…]

### 2020-08-22 data/handover.R
fix handover for names

@@ -1,4 +1,5 @@
 `handover` <-
-structure(list(brackets = list(1, c(1, 2), c(1, 3), 2, c(2, 4
-), 3, c(3, 4), 4), powers = c(18, 4, 9, 1, 2, 5, 4, 26), pnames = c("t00", 
-"t10", "t01", "t11")), class = "hyper2")
+structure(list(brackets = list("t00", c("t00", "t01"), c("t00", 
+"t10"), "t01", c("t01", "t11"), "t10", c("t10", "t11"), "t11"), 
+    powers = c(18, 9, 4, 5, 4, 1, 2, 26), pnames = c("t00", "t01", 
+    "t10", "t11")), class = "hyper2")

### 2020-08-23 inst/soling.Rmd
fix soling for names

@@
 			if(M[i,j]){
-				winner <- i
-				loser  <- j
+				winner <- rownames(M)[i]
+				loser  <- rownames(M)[j]
 			} else {
-				winner <- j
-				loser <- i
+				winner <- rownames(M)[j]
+				loser <-  rownames(M)[i]
 			}

### 2020-08-23 data/soling.R

@@ -1,54 +1,131 @@
 `soling2000` <-
-    structure(list(
-        brackets = list(1, c(1, 2, 3, 4, 5, 6, 7, 8, 9, 
-10, 11, 12, 13, 14, 15, 16), c(1, 2, 3, 4, 5, 6, 7, 9, 10, 11, 
-12, 13, 14, 15, 16), c(1, 2, 3, 4, 5, 7, 8, 9, 10, 11, 12, 13, 
-14, 15, 16), c(1, 2, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 
-16), c(1, 2, 4, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16), c(1, 
-2, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16), c(1, 2, 5, 6, 
-7, 8, 10, 11, 12, 13, 14, 15, 16), c(1, 2, 5, 6, 8, 10, 11, 12, 
-13, 14, 15, 16), c(1, 3, 4, 5, 7, 8, 9, 10, 11, 12, 13, 14, 15, 
-16), c(1, 4, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16), c(1, 6, 
-7, 8, 9, 10, 11, 12, 13, 14, 15, 16), 2, c(2, 3, 4, 5, 6, 7, 
-8, 9, 10, 11, 12, 13, 14, 15, 16), c(2, 3, 4, 5, 6, 7, 9, 10, 
-11, 12, 13, 14, 15, 16), c(2, 3, 5, 6, 7, 8, 9, 10, 11, 12, 13, 
-14, 15, 16), c(2, 3, 5, 6, 7, 8, 9, 11, 12, 13, 14, 15, 16), 
-    c(2, 3, 6, 7, 8, 9, 11, 12, 13, 14, 15, 16), c(2, 3, 6, 7, 
-    8, 9, 12, 13, 14, 15, 16), c(2, 3, 7, 8, 9, 12, 13, 14, 15, 
-    16), c(2, 3, 7, 9, 12, 13, 14, 15, 16), c(2, 3, 9, 12, 13, 
-    14, 15, 16), c(2, 3, 9, 12, 14, 15, 16), c(2, 5, 6, 8, 10, 
-    11, 12, 13, 14, 15, 16), c(2, 6, 8, 10, 11, 12, 13, 14, 15, 
[…]

### 2020-08-23 inst/universities.Rmd
universities now named

@@
   rownames(b)[3] <- "AUT_pre"
-  H1 <- combine(H1,ordertable2supp(b))
+  H1 <- H1 + ordertable2supp(b)
 } # column loop closes
@@
   rownames(jj)[3] <- "AUT_post"
-  H2 <- combine(H2,ordertable2supp(jj))
+  H2 <- H2 + ordertable2supp(jj)
 }
@@
 ```{r}
-H <- combine(H1,H2)
+H <- H1 + H2
 maxp(H)

### 2020-08-23 data/universities.R

@@ -83,22 +83,78 @@ structure(list(uni = structure(c(6L, 5L, 1L, 2L, 3L, 4L, 7L,
 -72L))
 
 `universities` <- 
-structure(list(brackets = list(1, c(1, 2, 3, 4, 5, 6, 7, 8), 
-    c(1, 2, 4, 5, 6, 7, 8), c(1, 3, 4, 5, 6, 7, 8), c(1, 4, 5, 
-    6, 7, 8), 2, c(2, 3, 4, 5), c(2, 3, 4, 5, 6, 7, 8), c(2, 
-    3, 4, 5, 7, 8), c(2, 3, 4, 5, 8), c(2, 3, 5), c(2, 3, 5, 
-    6, 7, 8), c(2, 3, 5, 6, 8), 3, c(3, 4, 5), c(3, 4, 5, 6), 
-    c(3, 4, 5, 6, 7), c(3, 4, 5, 6, 7, 8), c(3, 4, 5, 6, 8), 
-    c(3, 4, 5, 7), c(3, 4, 5, 7, 8), c(3, 4, 5, 8), c(3, 4, 6, 
-    7), c(3, 4, 6, 7, 8), c(3, 5), c(3, 5, 6), c(3, 5, 6, 7), 
-    c(3, 5, 6, 7, 8), c(3, 5, 6, 8), c(3, 5, 7), c(3, 5, 8), 
-    c(3, 6, 7), c(3, 6, 7, 8), c(3, 7), 4, c(4, 5, 6, 7, 8), 
-    c(4, 5, 6, 8), c(4, 5, 7), c(4, 5, 7, 8), c(4, 6, 7), c(4, 
-    7), 5, c(5, 6), c(5, 6, 7), c(5, 6, 7, 8), c(5, 6, 8), c(5, 
-    7), 6, c(6, 8), 7, 8), powers = c(24, -24, -2, -4, -3, 24, 
--1, -18, -1, -1, -1, -2, -2, 14, -1, -1, -12, -18, -2, -2, -2, 
--1, -2, -1, -5, -1, -8, -1, -4, -5, -1, -2, -1, -2, 24, -3, -1, 
--1, -1, -1, -1, 6, -1, -3, -2, -2, -5, 17, -1, 20, 24), pnames = c("UoA", 
-"Otago", "AUT", "Canterbury", "Lincoln", "Massey", "Victoria", 
-"Waikato")), class = "hyper2")
-
+structure(list(brackets = list("AUT_post", c("AUT_post", "Canterbury"
[…]

### 2020-08-23 inst/notthewinner.Rmd
slight rewriting
works with names paradigm

@@
 
-Here we consider a situation in which we make an observation that is
-only very very weakly informative.  We consider the observation
-${\mathcal O}$ that a particular player *does not* win: the order
-statistic is unobserved and we know nothing about it except for the
-fact that (wlog) player 1 ($p_1$, aka "me") is not first placed.  We
-consider whether there is evidence to support two different assertions:
+(see also `inst/loser.Rmd`).
+Here we consider a situation in which we make
+an observation that is only very very weakly informative.  We consider
+the observation ${\mathcal O}$ that a particular player *does not*
+win: the order statistic is unobserved and we know nothing about it
+except for the fact that (wlog) player 1 ($p_1$, aka "me") is not
+first placed.  We consider whether there is evidence to support two
+different assertions:
 
@@
 
-Our alternative would be that $(p_1,\ldots,p_n) is unconstrained
+Our alternative would be that $(p_1,\ldots,p_n)$ is unconstrained
 except for the unit sum constraint:
@@
 ```{r}
-(H <- hyper2(pnames=paste("p",1:5,sep="")) +
-    rank_likelihood(c(5,2,1,4)) + 
-    rank_likelihood(c(2,4,1)) + 
-    rank_likelihood(c(4,3,1)) + 
-    rank_likelihood(c(3,5)))
+(H <- 
+	rank_likelihood(c(a=3,c=2,b=1,d=4)) + 
+	rank_likelihood(c(b=2,c=3,d=1    )) + 
+	rank_likelihood(c(d=2,a=3,c=1    )) + 
+	rank_likelihood(c(b=1,d=2        )) )
 ```
@@
 ```{r}
-n <- 5 # number of players
-small <- 1e-3  # numerical necessity
-(mH1 <- maxp(H,startp=c(small*2,rep(small,n-2)),fcm=rbind(c(1,0,0,-1)),fcv=0))
-loglik(indep(mH1),H)
-loglik(indep(mH0),H)-loglik(indep(mH1),H)   # support
+samep.test(H,c(1,4))
 ```
@@
 for(i in seq_len(ncol(M))){
-	SL[[i]] <- H + rank_likelihood(M[,i,drop=TRUE])
+	jj <- M[,i,drop=TRUE]
+	names(jj) <- letters[seq_along(jj)]
+	SL[[i]] <- H + rank_likelihood(jj)
 }
@@
 ```{r,mls0,cache=TRUE}
-mSL0 <- maxp(SL)
-mSL1 <- maxp(SL,startp=c(small*2,rep(small,n-2)),fcm=rbind(c(1,0,0,-1)),fcv=0)
+n <- 5 # number of players
+small <- 1e-4
+mSL0 <- maxplist(SL)
+mSL1 <- maxplist(SL,startp=c(small*2,rep(small,n-2)),fcm=rbind(c(1,0,0,-1)),fcv=0)
 mSL2 <- equalp(SL)
@@
 `mSL2` above.
-
-# Additional weakly informative observation
-
-Now we make another observation ${\mathcal O}$ which is independent of
-${\mathcal I}$: person `1` did not win.  It is clear that observation
-${\mathcal O}$ is only very weakly informative.  Package idiom is
-similar to the perfectly uninformative observation considered above,
-except that we remove every order that has person 1 winning:
-
-```{r,slacalc}
-M <- perms(5)
-M <- M[, M[1,]!=1]  # remove any column with "1" in first place
-SLa <- list()
-for(i in seq_len(ncol(M))){
-	SLa[[i]] <- H + rank_likelihood(M[,i,drop=TRUE])
-}
-SLa <- as.suplist(SLa)
-```
-
-So `SLa` is a compound object that gives a support function for the
-joint observation ${\mathcal I}$ and ${\mathcal O}$.  We can find the
-MLE:
-
-```{r,mlsa,cache=TRUE}
-mSLa0 <- maxp(SLa)
-mSLa1 <- maxp(SLa,startp=c(small*2,rep(small,n-2)),fcm=rbind(c(1,0,0,-1)),fcv=0)
-mSLa2 <- equalp(SLa)
-```
-
-```{r,suppacalc}
-Suppa <-
-      log(c(
-      H0 = like_single_list(indep(mSLa0),SLa),
-      H1 = like_single_list(indep(mSLa1),SLa),
-      H2 = like_single_list(indep(mSLa2),SLa)))
-(Suppa <- Suppa-max(Suppa))
-```
-
-We can now compare the support function for ${\mathcal I}$ with the
-support function for ${\mathcal I}$ and ${\mathcal H}$:
-
-```{r,comparesups}
-Supp2-Suppa
-```
-
-Thus ${\mathcal O}$ furnishes a tiny bit of extra evidence against
-$H_1$ (about $0.016$ units of support) and a little bit of extra
-evidence against $H_2$ (about $0.193$ units of support).

### 2020-08-24 inst/cook_repeated.Rmd
consistentify

@@
 cooking_table <- as.matrix(read.table("cooking_weekly.txt", header=TRUE))
-cooking_table
+head(cooking_table)
+tail(cooking_table)
 ```
@@
 ```{r}
-jj <- cooking_table[1,]
-H_row1 <- order_obs(hyper2(pnames=colnames(cooking_table)),jj[jj>0])
-H_row1
+(day1 <- cooking_table[1,])
+H_row1 <- rank_likelihood(day1[day1>0])
 ```
@@
 ```{r}
-p1 <- rdirichlet(1,dirichlet(rep(10,5)))
+p <- c(0.3,0.2,0.1,0.3)
 # Now create p2 which is the same but competitor 1 is stronger:
-p2 <- p1
-p2[1] <- p2[1] + 0.1     # person 1 becomes stronger...
-p2[-1] <- p2[-1] - 0.1/4 # ...and everyone else becomes weaker
+p2 <- p
+alpha <- 1.3
+beta <- (1-alpha*p[1])/(1-p[1])
+p2[1] <- p2[1]*alpha   # person 1 becomes stronger...
+p2[-1] <- p2[-1]*beta  # ...and everyone else becomes weaker
 
-loglik(indep(p1),H_row1)
+loglik(indep(p),H_row1)
 loglik(indep(p2),H_row1)
-```
-
-but more generally:
-
-```{r}
-(p <- rdirichlet(1,dirichlet(rep(1,5))))
-evaluate(brackets(H_row1), powers(H_row1), probs=p)
-evaluate(brackets(H_row1), powers(H_row1), probs=p*4)
 ```
@@
 ```{r}
-H <- hyper2(pnames=colnames(cooking_table))
+H <- hyper2()
 
 for(i in seq_len(nrow(cooking_table))){  # iterate over rows
-    x <- cooking_table[i,] 
-    names(x) <- colnames(cooking_table)
-    H %<>% order_obs(x[x>0])
+jj <- cooking_table[i,] 
+    H <- H + rank_likelihood(jj[jj>0])
 }
@@
 
-```{r}
+```{r,cache=TRUE}
 cooking_table <-
@@
 cooks <- colnames(cooking_table)
-P <- hyper2(pnames=c(cooks,"boy","girl"))
-P0 <- hyper2(pnames=cooks)
+P <- hyper2()
+P0 <- hyper2()
 for(i in seq_len(nrow(cooking_table))){  # iterate over rows
     x <- cooking_table[i,] 
-	print(x)
 	if(sum(x==0)>1){break}
@@
 		x <- x[x>0]  # remove the voter
-		Pjj <- order_obs(P0,x) # order statistic
-		Pjj %<>% pwa(names(x)[names(x) %in% boys], chameleon="boy")
-		pnames(Pjj) <- c(cooks,"boy","girl")
+		Pjj <- pwa(rank_likelihood(x),names(x)[names(x) %in% boys], chameleon="boy")
 	} else if (names(x[x==0]) %in% girls){ # voter a girl
 		x <- x[x>0]
-		Pjj <- order_obs(P0,x)
-		Pjj %<>% pwa(names(x)[names(x) %in% girls], chameleon="girl")
-		pnames(Pjj) <- c(cooks,"boy","girl")
+		Pjj <- pwa(rank_likelihood(x),names(x)[names(x) %in% girls], chameleon="girl")
 	} else {
@@
 }
-x
-Pjj
-P
-maxp(P)
+mP <- maxp(P)
 ```
+
+```{r}
+mP
+pie(mP)
+```
+
+Now test whether the girls preferentially vote:
+
+```{r,cache=TRUE}
+specificp.gt.test(P,"girl",0)
+```

### 2020-08-24 inst/jester.Rmd
names paradigm changes

@@
 ```{r,label=calculate_jester,cache=TRUE}
-jester <- hyper2(pnames=colnames(a))
-system.time(for(i in seq_len(nrow(a))){jester %<>% order_obs(f(a[i,]))})
+jester <- hyper2()
+system.time(for(i in seq_len(nrow(a))){jester <- jester + rank_likelihood(f(a[i,]))})
 ```

### 2020-08-26 inst/formula1_allyears.Rmd
new years and new analysis

---
title: "formula1_allyears"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library("hyper2")
```

## Introduction

Here I analyse Formula 1 results 2010-2019 using Plackett-Luce
likelihoods as implemented in the `hyper2` package.  Files like
`formula1_2017.txt` are directly copied from Wikipedia (with slight
whitespace changes).

```{r,label=f1yearsseparate,cache=TRUE}
f <- function(M){M[,-ncol(M)]} # removes 'points' column
o <- function(k){ordertable2supp(f(read.table(k)))}
F1_2010 <- o("formula1_2010.txt")
F1_2011 <- o("formula1_2011.txt")
F1_2012 <- o("formula1_2012.txt")
F1_2013 <- o("formula1_2013.txt")
F1_2014 <- o("formula1_2014.txt")
F1_2015 <- o("formula1_2015.txt")
F1_2016 <- o("formula1_2016.txt")
F1_2017 <- o("formula1_2017.txt")
F1_2018 <- o("formula1_2018.txt")
F1_2019 <- o("formula1_2019.txt")
```

```{r,label=combineallyears,cache=TRUE}
F1_list<- list(F1_2010, F1_2011, F1_2012, F1_2013, F1_2014, F1_2015, F1_2016, F1_2017, F1_2018, F1_2019)
F1_total <- F1_2010 + F1_2011 + F1_2012 + F1_2013 + F1_2014 + F1_2015 + F1_2016 + F1_2017 + F1_2018 + F1_2019
```

```{r,calculatemaxlikelhood,cache=TRUE}
m <- maxp(F1_total)
pie(m)
dotchart(m,pch=16,main='Formula 1, 2012-9')
```

```{r,workwithM}
M <- matrix(NA,length(pnames(F1_total)),2019-2010+1)
rownames(M) <- pnames(F1_total)
colnames(M) <- paste("y",10:19,sep="")

for(i in seq_len(nrow(M))){
    for(j in seq_len(ncol(M))){
        if(rownames(M)[i] %in% pnames(F1_list[[j]])){M[i,j] <- 1}
    }
}
jj <- order(rowSums(M,na.rm=T),decreasing=T)
M <- M[jj,]

image(x=2010:2019,y=seq_along(pnames(F1_total)),t(M))
```

### 2020-08-27 inst/formula1_allyears.Rmd
formula 1 all years just the top competitors

@@
 ## Introduction
+
+*Takes about 20 minutes to run with no cache*
 
@@
 ```
+
+```{r}
+howmanyraces <- sort(rowSums(M,na.rm=TRUE),decreasing=TRUE)
+plot(howmanyraces)
+```
+
+```{r}
+select <- function(F1tab,minraces=5){
+	wanted <- rownames(M)[howmanyraces >= minraces]
+	ordertable2supp(F1tab[rownames(F1tab) %in% wanted,])
+}
+```
+
+```{r,definefreqrace}
+freqracers <- function(minraces){
+(
+	select(f(read.table("formula1_2010.txt")),minraces) + 
+	select(f(read.table("formula1_2011.txt")),minraces) + 
+	select(f(read.table("formula1_2012.txt")),minraces) + 
+	select(f(read.table("formula1_2013.txt")),minraces) + 
+	select(f(read.table("formula1_2014.txt")),minraces) + 
+	select(f(read.table("formula1_2015.txt")),minraces) + 
+	select(f(read.table("formula1_2016.txt")),minraces) + 
+	select(f(read.table("formula1_2017.txt")),minraces) + 
+	select(f(read.table("formula1_2018.txt")),minraces) 
+	)
+}
+
+comp <- function(m,mr){
+	par(pty="s")
+#	plot(m,mr,asp=1,xlim=c(0,0.2),ylim=c(0,0.2))
+	plot(m,mr,asp=1)
+	abline(0,1)
+}
+```
+
+# All drivers
+
+```{r,freqtop1,cache=TRUE}
+F1_top1 <- freqracers(1)
+m1 <- maxp(F1_top1)
+pnames(F1_top1) <- rev(pnames(F1_top1))
+m1r <- rev(maxp(F1_top1)) # reverse of MLE with names reversed
+```
+
+```{r}
+pie(m1)
+comp(m1,m1r)
+```
+
+# Drivers with >= 2 races
+
+```{r,freqtop2,cache=TRUE}
+F1_top2 <- freqracers(2)
+m2 <- maxp(F1_top2)
+pnames(F1_top2) <- rev(pnames(F1_top2))
+m2r <- rev(maxp(F1_top2)) # reverse of MLE with names reversed
+```
+
+```{r}
+pie(m2)
+comp(m2,m2r)
+```
+
+# Drivers with >= 3 races
+
+```{r,freqtop3,cache=TRUE}
+F1_top3 <- freqracers(3)
+m3 <- maxp(F1_top3)
+pnames(F1_top3) <- rev(pnames(F1_top3))
+m3r <- rev(maxp(F1_top3)) # reverse of MLE with names reversed
+```
+
+```{r}
+pie(m3)
+comp(m3,m3r)
+```
+
+# Drivers with >= 4 races
+
+```{r,freqtop4,cache=TRUE}
+F1_top4 <- freqracers(4)
+m4 <- maxp(F1_top4)
+pnames(F1_top4) <- rev(pnames(F1_top4))
+m4r <- rev(maxp(F1_top4)) # reverse of MLE with names reversed
+```
+
+```{r}
+pie(m4)
+comp(m4,m4r)
+```
+
+# Drivers with >= 5 races
+
+```{r,freqtop5,cache=TRUE}
+F1_top5 <- freqracers(5)
+m5 <- maxp(F1_top5)
+pnames(F1_top5) <- rev(pnames(F1_top5))
+m5r <- rev(maxp(F1_top5)) # reverse of MLE with names reversed
+```
+
+```{r}
+pie(m5)
+comp(m5,m5r)
+```
+
+# Drivers with >= 6 races
+
+```{r,freqtop6,cache=TRUE}
+F1_top6 <- freqracers(6)
+m6 <- maxp(F1_top6)
+pnames(F1_top6) <- rev(pnames(F1_top6))
+m6r <- rev(maxp(F1_top6)) # reverse of MLE with names reversed
+```
+
+```{r}
+pie(m6)
+comp(m6,m6r)
+```

### 2020-08-28 inst/formula1_allyears.Rmd
consistent use of years

@@
 
-Here I analyse Formula 1 results 2010-2019 using Plackett-Luce
-likelihoods as implemented in the `hyper2` package.  Files like
-`formula1_2017.txt` are directly copied from Wikipedia (with slight
-whitespace changes).
+Here I analyse Formula 1 results in a range of years using
+Plackett-Luce likelihoods as implemented in the `hyper2` package.
+Files like `formula1_2017.txt` are directly copied from Wikipedia
+(with slight whitespace changes).
+
+```{r}
+years <- 2001:2015
+```
 
@@
 o <- function(k){ordertable2supp(f(read.table(k)))}
-F1_2010 <- o("formula1_2010.txt")
-F1_2011 <- o("formula1_2011.txt")
-F1_2012 <- o("formula1_2012.txt")
-F1_2013 <- o("formula1_2013.txt")
-F1_2014 <- o("formula1_2014.txt")
-F1_2015 <- o("formula1_2015.txt")
-F1_2016 <- o("formula1_2016.txt")
-F1_2017 <- o("formula1_2017.txt")
-F1_2018 <- o("formula1_2018.txt")
-F1_2019 <- o("formula1_2019.txt")
+filenamemaker <- function(year){paste("formula1_",year,".txt", sep="")}
+F1_list <- list()
+for(year in years){
+	filename <- filenamemaker(year)
+	jj <- paste("y",year,sep="")
+	F1_list[[jj]] <- o(filename)
+}
 ```
@@
 ```{r,label=combineallyears,cache=TRUE}
-F1_list<- list(F1_2010, F1_2011, F1_2012, F1_2013, F1_2014, F1_2015, F1_2016, F1_2017, F1_2018, F1_2019)
-F1_total <- F1_2010 + F1_2011 + F1_2012 + F1_2013 + F1_2014 + F1_2015 + F1_2016 + F1_2017 + F1_2018 + F1_2019
+F1_total <- hyper2()
+for(i in F1_list){F1_total <- F1_total + i}
 ```
@@
 pie(m)
-dotchart(m,pch=16,main='Formula 1, 2012-9')
+dotchart(m,pch=16,main=paste("Formula 1, years ",min(years)," to ",max(years)))
 ```
@@
 ```{r,workwithM}
-M <- matrix(NA,length(pnames(F1_total)),2019-2010+1)
+M <- matrix(NA,nrow=length(pnames(F1_total)),ncol=length(years))
 rownames(M) <- pnames(F1_total)
-colnames(M) <- paste("y",10:19,sep="")
+colnames(M) <- paste("y",years,sep="")
 
@@
 
-image(x=2010:2019,y=seq_along(pnames(F1_total)),t(M))
+image(x=years,y=seq_along(pnames(F1_total)),t(M))
 ```
@@
 ```{r,definefreqrace}
-freqracers <- function(minraces){
-(
-	select(f(read.table("formula1_2010.txt")),minraces) + 
-	select(f(read.table("formula1_2011.txt")),minraces) + 
-	select(f(read.table("formula1_2012.txt")),minraces) + 
-	select(f(read.table("formula1_2013.txt")),minraces) + 
-	select(f(read.table("formula1_2014.txt")),minraces) + 
-	select(f(read.table("formula1_2015.txt")),minraces) + 
-	select(f(read.table("formula1_2016.txt")),minraces) + 
-	select(f(read.table("formula1_2017.txt")),minraces) + 
-	select(f(read.table("formula1_2018.txt")),minraces) 
-	)
+freqracers <- function(minraces){ # minimum number of seasons
+	out <- hyper2()
+	for(year in years){
+		out <- out + select(f(read.table(filenamemaker(year),minraces)))
+	}	
+	return(out)
 }

### 2020-09-14 inst/jebo_chess_rebuttal.txt
revision of JEBO paper, ancillary files at ~/Dropbox/draft_papers/

“A generalization of the Bradley-Terry model for draws in chess”
JEBO-D-20-00916
Referee Report
This paper develops extensions to the Bradley-Terry model appropriate to Bernoulli trials
so that the model can accommodate the chess-specific needs of draws and the first-mover
advantage possessed by White. The author takes this extended model to various sets of
exclusively game-level outcomes and estimates time-invariant player strengths as well as the
importance of draws and first-mover advantage. After estimating common parameters for draws
and first-mover advantage, the author allows for (time-invariant) player-specific parameters on
these margins. The paper then concludes by considering distinct draw parameters for Soviet and
non-Soviet players. Using data from the top chess events of 1962 and 1963, the author
determines that Soviet players draw more frequently against each other than in other gamescenarios.
The paper does a nice job of clearly laying out the various reified “monsters” that extend
the original Bradley-Terry model to accommodate chess. This framework is plausible and
intuitive. I also appreciated the paper’s brevity in making its point. The allowance for playerspecific tendencies to draw or to gain excess-advantage from moving first also strikes me as an
innovation to the literature. Lastly, the paper is well-written and a fun read.
My comments on the rest of the paper are detailed below by decreasing importance.
1) The author never properly explores the paper’s identification strategy. As I read it, the
identifying assumption, besides the functional form assumption, is that player strengths
are constant over the duration of the considered data. This seems highly plausible within
a specific event (e.g., the Candidates tournament of 1963, the Interzonals of 1962), but
much less so with the Anand-Karpov-Kasparov example. The author does not specify the
timespan over which these 365 games were played, but I imagine that the timespan must
exceed 10 years. To the extent that all players have matured to full strength and none
have yet begun to decline, this assumption would be benign, but it should be supported.
Precise estimation will hinge on a large number of games, but this will be in tension with
the assumption of time-invariant player-strengths. The author should explore this tension.
2) The author neglects to motivate the paper beyond its application within the chess world.
The tournament literature’s applications lie primarily, though definitely not exclusively,
within compensation and labor. How could this extension be used within that context?
Where might economists see the first-mover advantage in action? Where might the
extension to allow draws be especially lucrative? Etc.
3) As the author notes, the work of Moul and Nye (2009) is directly on point. The author’s
discussion of this paper, however, is overly sparse and at times misleading. Moul and

Nye use existing observed player strengths (variations of commonly used scores based on
the Elo rating system that goes back to 1960). Moul and Nye then use these strengths
within an ordered probit to estimate parameters that relate back to draw likelihood and
White’s first-mover advantage. The sharp difference between that paper and this is then
that the author is estimating player relative strengths (conditional upon a pool of players)
while Moul and Nye use “observed” strengths.1 The latter approach is far less
parametrically demanding, but leans much more on the assumption of normally
distributed disturbances (basically the differing assumption of functional form). The latter
approach also allows individuals’ player strength to vary across time and is far more
flexible to changes in the composition of the pool of players.
Furthermore, Moul and Nye estimate several versions of their ordered probit without data
on the number of moves (ply-count). This is contrary to the implication of the paper (p. 6,
immediately below Table 2), which suggests that the ply-count data are essential to that
paper’s findings. All of that paper’s core results (significance of draws, first-mover
advantage, a distinct draw outcome for round-robin international tournament games that
involve two Soviets) are originally derived without ply-count data. The ply-count data are
incorporated to explore a potential mechanism by which players in a draw-cartel might
benefit.
4) The paper concludes that the estimates derived from the 1962 Interzonals and the 1963
Candidates strongly suggest that the “draw monster” won games involving two Soviet
players more frequently than games that involved at least one non-Soviet. This is
consistent with a Soviet cartel, but it could also be the result of Soviet-specific styles of
play. Moul and Nye (2009) test the Soviet style-of-play hypothesis by considering
international tournaments when the format changed from all-play-all to knockout and allSoviet tournaments. Both comparisons yield results consistent with a Soviet draw-cartel
in international all-play-all tournaments and rejecting the Soviet style-of-play hypothesis.
By only considering tournaments that were international and all-play-all, the author
forgoes this critical source of identification.
In the cartel example, the author also neglects the advantage that presumably accrued to
Soviet players when they played non-Soviet players in all-play-all international
tournaments. This would suggest another “monster” to be estimated, one that benefits the
Soviet player when playing a non-Soviet opponent. Without such an addition, Soviet
strengths will be overstated.

1

Those relative strengths are highly predictive, with t-statistics of around 20.

### 2020-09-14 inst/jebo_chess_revision2.tex
revision of JEBO paper, ancillary files at ~/Dropbox/draft_papers/

--- 2019-05-27 inst/collusion_chess_draws.Rnw
@@
-% -*- mode: noweb; noweb-default-code-mode: R-mode; -*-
-\documentclass[nojss]{jss}
+\documentclass[review]{elsarticle}
+
+\usepackage{lineno,hyperref}
 \usepackage{amsmath}
@@
 \usepackage{units}
+\usepackage{xcolor}
+\modulolinenumbers[5]
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\journal{Journal of Economic Behavior and Organization}
 
-%% just as usual
-\author{Robin K. S. Hankin\\Department of Mathematical Sciences\\Auckland University of Technology}
+%%%%%%%%%%%%%%%%%%%%%%%
+%% Elsevier bibliography styles
+%%%%%%%%%%%%%%%%%%%%%%%
+%% To change the style, put a % in front of the second line of the current style and
+%% remove the % from the second line of the style you would like to use.
+%%%%%%%%%%%%%%%%%%%%%%%
+
+%% Numbered
+%\bibliographystyle{model1-num-names}
+
+%% Numbered without titles
+%\bibliographystyle{model1a-num-names}
+
+%% Harvard
+\bibliographystyle{model2-names.bst}\biboptions{authoryear}
+
+%% Vancouver numbered
+%\usepackage{numcompress}\bibliographystyle{model3-num-names}
+
+%% Vancouver name/year
+%\usepackage{numcompress}\bibliographystyle{model4-names}\biboptions{authoryear}
+
+%% APA style
+%\bibliographystyle{model5-names}\biboptions{authoryear}
+
+%% AMA style
+%\usepackage{numcompress}\bibliographystyle{model6-num-names}
+
+%% `Elsevier LaTeX' style
+%\bibliographystyle{elsarticle-num}
+%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{document}
+
+\begin{frontmatter}
+
 \title{A generalization of the Bradley-Terry model for draws in chess}
 
-%\VignetteIndexEntry{chess draws}
+%% Group authors per affiliation:
+\author{Robin K. S. Hankin\fnref{myfootnote}}
+\address{AUT University, 2-14 Wakefield Street, Auckland, New Zealand}
 
-%% for pretty printing and a nice hypersummary also set:
-\Plainauthor{Robin K. S. Hankin}
-\Plaintitle{Chess Draws}
-\Shorttitle{Chess Draws}
-
-%% an abstract and keywords
-\Abstract{
-
+\begin{abstract}
 In inference problems where the dataset comprises Bernoulli outcomes
@@
 agreement with previous work that uses additional information such as
-historical results including number of moves.}
+historical results including number of moves.
+\end{abstract}
 
-\Keywords{Likelihood functions, method of support, \proglang{R},
-  constrained optimization, Bradley-Terry, Chess, Collusion, Chess
-  Draws}
-\Plainkeywords{Likelihood functions, method of support, R,
-  constrained optimization, Bradley-Terry, Chess, Collusion, Chess
-  Draws}
+\begin{keyword}
+  Chess draws\sep Bradley-Terry\sep Likelihood \sep Chess Collusion
+\end{keyword}
 
-%% publication information
-%% NOTE: This needs to filled out ONLY IF THE PAPER WAS ACCEPTED.
-%% If it was not (yet) accepted, leave them commented.
-%% \Volume{13}
-%% \Issue{9}
-%% \Month{September}
-%% \Year{2004}
-%% \Submitdate{2004-09-29}
-%% \Acceptdate{2004-09-29}
+\end{frontmatter}
 
-%% The address of (at least) one author should be given
-%% in the following format:
-\Address{
-  Robin K. S. Hankin\\
-  Auckland University of Technology\\
-  E-mail: \email{hankin.robin@gmail.com}
-}
-%% It is also possible to add a telephone and fax number
-%% before the e-mail in the following format:
-%% Telephone: +43/1/31336-5053
-%% Fax: +43/1/31336-734
-
-%% for those who use Sweave please include the following line (with % symbols):
-%% need no \usepackage{Sweave.sty}
-
-%% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\linenumbers
 
@@
 \newcommand{\white}{\ensuremath{\mathfrak{W}}}
-
-\SweaveOpts{}
-\begin{document}
-
-<<echo=FALSE,print=FALSE>>=
-ignore <- require(magrittr,quietly=TRUE)
-ignore <- require(hyper2,quietly=TRUE)
-options("width" = 90)
-@
+\newcommand{\Prob}{\operatorname{Prob}}
 
@@
 competitive situations such as tennis in which one player is
-identified as the winner.  The Bradley-Terry model~\citep{bradley1952}
-considers~$n$ players and assigns a non-negative number $p_i$ to player
-$i$, with $\sum p_i=1$.  Then player $i$ beats player $j$ with
-probability $\frac{p_i}{p_i+p_j}$.  This simple system has been
-modified and extended many times, for example to account for multi-way
-competitions such as {\em MasterChef Australia}~\citep{hankin2017},
-and non-competitive situations such as consumer
-behaviour~\citep{hankin2010} and mating preference~\citep{west2008}.
+identified as the winner.  The Bradley-Terry
+model~\citep{bradley1952,turner2012} considers~$n$ players and assigns
+a non-negative number $p_i$ to player $i$, with $\sum p_i=1$.  Then
+player $i$ beats player $j$ with probability $\frac{p_i}{p_i+p_j}$.
+This simple system has been modified and extended many times, for
+example to account for multi-way competitions such as {\em MasterChef
+  Australia}~\citep{hankin2017}, and non-competitive situations such
+as consumer behaviour~\citep{hankin2010} and mating
+preference~\citep{west2008}.
 
@@
 less material to discuss.  Players themselves do not seem to value
-draws in the wider context of the game: \citet{fischer1969}, for
-example, includes only 8 draws in his compilation of 60 memorable
+draws in the wider context of the game: \citet{fischer1969},
+for example, includes only 8 draws in his compilation of 60 memorable
 games, surely not representative of tournament play in which about
@@
 
-\cite{moul2009} conceptualise a player as being able to adopt a
+\citet{moul2009} conceptualise a player as being able to adopt a
 strategy on a spectrum ranging from conservative to aggressive, thus
@@
 trading expected payout for reduced variance.  Further,
-\cite{langer2013} point out that rational players in an ongoing
+\citet{langer2013} point out that rational players in an ongoing
 competitive situation modify their strategies depending on whether
@@
 However, interpretations of this statistic vary and the issue is hotly
-debated with opinions ranging from that of~\cite{adams2007}---``White
-to play and win'' to~\citet{adorjan1988}---``Black is OK!''.
+debated with opinions ranging from that of \citet{adams2007}---``White
+to play and win'' to \citet{adorjan1988}---``Black is OK!''.
 
@@
 
-where $W_{ij}$ is the number of times $i$ beats $j$ and $L_{ij}$ is
+\noindent where $W_{ij}$ is the number of times $i$ beats $j$ and $L_{ij}$ is
 the number of times $i$ loses to $j$.  Here I present a new
@@
 
-where $W_{ij}$ is the number of games won by $i$ when playing $j$,
-$L_{ij}$ is games lost and $D_{ij}$ is games drawn.  However, this
-approach ignores the extra strength conferred by the first-move
-advantage.  We thus consider another entity whose strength \white\ is
-added to the white player.  The strength of the white player is thus
-the sum of the player's strength and \white.  Recalling that a game
-might be a draw, we obtain the probabilities as given in
+\noindent where $D_{ij}$ is games drawn.  However, this approach
+ignores the extra strength conferred by the first-move advantage.  We
+thus consider another entity whose strength \white\ is added to that
+of the white player.  The strength of the white player is thus the sum
+of the player's strength and \white.  Recalling that a game might be a
+draw, we obtain the probabilities as given in
 Table~\ref{probabilities}.  The corresponding likelihood function is
@@
 \begin{table}
+  \caption{Probabilities \label{probabilities}of a win, draw, loss for the two players under a generalized Bradley-Terry model}
 \centering
+\fbox{%
 \begin{tabular}{l|cc}\\
@@
 $\Prob{(\mbox{$i$ loses})}$ & $\frac{p_j\oo}{p_i+p_j+\white+\draw}$   & $\frac{p_j+\white}{p_i+p_j+\white+\draw}$\\
-\end{tabular}
-\caption{Probabilities of a win, draw, loss for the two players under a generalized
-  Bradley-Terry model\label{probabilities}}
+\end{tabular}}
 \end{table}
@@
 
-where $W_{ij}^\mathrm{white}$ is the number of games won by $i$ when
-playing white against $j$, and so on.  Note that the power of the
-final term, $D_{ij}^\mathrm{white}+D_{ij}^\mathrm{black}$ is the total
-number of games drawn between $i$ and $j$ irrespective of colour.  We
-now apply this and related likelihood models to real chess datasets
-using the Method of Support~\citep{edwards1972}.  To test any
+\noindent where $W_{ij}^\mathrm{white}$ is the number of games won by
+$i$ when playing white against $j$, and so on.  Note that the power of
+the final term, $D_{ij}^\mathrm{white}+D_{ij}^\mathrm{black}$ is the
+total number of games drawn between $i$ and $j$ irrespective of
+colour.  We now apply this and related likelihood models to real chess
+datasets using the Method of Support~\citep{edwards1972}.  To test any
 hypothesis $H$ on likelihood function $\mathcal{L}$ one first
@@
 
-\cite{davidson1970} uses a threshhold approach to interpret draws.
+\citet{davidson1970} uses a threshhold approach to interpret draws.
 From a chess draw, he infers that the performance of the players is
@@
 difference in players' performance he uses, in current notation,
-$\draw=\nu\sqrt{p_ip_j}$.  However, \cite{joe1990} uses likelihood methods
-to show that this model does not fit data well.  Here we note that
-such a model automatically give stronger players a higher draw
+$\draw=\nu\sqrt{p_ip_j}$.  However, \citet{joe1990} uses likelihood
+methods to show that this model does not fit data well.  Here we note
+that such a model automatically give stronger players a higher draw
 potential, and also note that it is difficult to account for
@@
 constraint becomes nonlinear, making likelihood maximization
-difficult.  \cite{moul2009} consider chess draws and use sophisticated
-regression techniques to analyse games in terms of not only the
-result, but also the ply-count, data not used in the present analysis.
+difficult.  \citet{moul2009} consider chess draws and use
+sophisticated regression techniques to analyse games in terms of not
+only the result, but also the ply-count, data not used in the present
+analysis.
 
@@
 The analysis presented here uses the framework of the
-R~\citep{rcore2019} \pkg{hyper2} package~\citep{hankin2017} as a
+R~\citep{rcore2019} {\tt hyper2} package~\citep{hankin2017} as a
 convenient and unified interface.  Because gradients are available,
-convergence to the evaluate is generally rapid and hypothesis-specific
+convergence to the evaluate is generally rapid; hypothesis-specific
 constraints on parameter space are, in general, linear and
@@
 
-Grandmasters Anand, Karpov, and Kasparov have long-standing rivalry
+Grandmasters Anand, Karpov, and Kasparov have a long-standing rivalry
 and have played many games amongst themselves.  A summary may be found
@@
 \begin{table}
+\caption{Results of 365 chess games \label{kka} between Grandmaster
+  Anand, Grandmaster Karpov, and Grandmaster Kasparov.  For example,
+  Anand played white against Karpov a total of $18+20+5=43$ times and
+  won 18, drew 20, lost 5}
 \centering
+\fbox{%
 \begin{tabular}{@{} *{4}{c} @{}}
@@
   Ks &  2 &   7 &  -   \\
-\end{tabular}
-\caption{Results of 365 chess games \label{kka} between Grandmaster
-  Anand, Grandmaster Karpov, and Grandmaster Kasparov.  For example,
-  Anand played white against Karpov a total of $18+20+13=51$ times and
-  won 18, drew 20, lost 13}
+\end{tabular}}
 \end{table}
@@
 
-<<echo=FALSE,print=FALSE,results=hide>>=
-maxp(karpov_kasparov_anand)
-@
-
 \begin{figure}[t]
 \begin{center}
-<<ddc,echo=FALSE,fig=TRUE>>=
-dotchart(maxp(karpov_kasparov_anand),pch=16,xlim=c(0,0.55))
-@
+\makebox{\includegraphics{collusion_chess_draws-ddc.pdf}}
 \caption{Maximum likelihood estimate for the strengths of the 3 Grandmasters in Table~\ref{kka} 
@@
 
-<<maxlikekka,echo=FALSE,print=FALSE,results=hide>>=
-small <- 1e-4
-
-e <- function(white){
-    maxp(
-        karpov_kasparov_anand,
-        startp=c(1/3-small,1/3-small,1/3-small,small),
-        fcm=-cbind(0, 0, 0, 1),fcv= -white,give=TRUE)$value
-}
-ignore1 <- e(0.001)
-ignore2 <- e(0.002)
-ignore3 <- e(0.003)
-ignore4 <- maxp(karpov_kasparov_anand,give=TRUE)$value
-support <- 328.2975-335.6025
-@
-
-<<ignore34diff,echo=FALSE,print=FALSE,results=hide>>=
-ignore3
-ignore4
-ignore3-ignore4
-@
-
 The result of numerical optimization is about 7.01 units of support,
 or a likelihood ratio of $e^{7.01}\simeq 1107.7$: that is, well over
-the two-units-of-support per degree of freedom suggested by
-\citeauthor{edwards1972}; we can convincingly reject the hypothesis
-that $\white=0$ and infer that playing white confers a significant
-advantage.  This is consistent with expectation, but it is now
-possible to go significantly beyond mere existence of the effect and
-to quantify it, accounting for the differential strengths of the
-players.  Taking the MLE above, and considering a game between say
-Karpov and Anand, the strengths are Karpov=0.064, Anand=0.151, \white
-=0.104 and \draw =0.532.  Thus we would estimate that Karpov playing
-white against Kasparov would win with probability
+the two-units-of-support per degree of freedom suggested by Edwards;
+we can convincingly reject the hypothesis that $\white=0$ and infer
+that playing white confers a significant advantage,
+consistent with expectation.  It is now possible to go beyond
+merely establishing the existence of the effect, and to quantify it
+accounting for the differential strengths of the players.  Taking the
+MLE above, and considering a game between say Karpov and Anand, the
+strengths are Karpov=0.064, Anand=0.151, \white =0.104 and \draw
+=0.532.  Thus we would estimate that Karpov playing white against
+Kasparov would win with probability
 $\frac{0.064+0.104}{0.064+0.104+0.148+0.532}\simeq 0.198$, draw with
@@
 Karpov plays black).
-
-<<kka_equal_strength, print=FALSE,echo=FALSE>>=
-objective <- function(V){
-  x <- V[1]  # strength of all three players
-  w <- V[2]  # white player advantage
-  D <- 1-3*x-w  # Draw monster
-  return(  
-      -loglik(karpov_kasparov_anand,c(x,x,x,w))  # negative as we minimize
-  )
-}
-
-jj <-
-constrOptim(
-    theta=c(0.1,0.4),
-    f=objective,
-    grad=NULL,
-    ui = rbind(diag(2),-diag(2),c(-3,-1)),
-    ci = c(0,0,-1,-1,-1)
-)
-
-unconstrained <- maxp(karpov_kasparov_anand,give=TRUE)$value
-constrained <- jj$value
-
-extra_support <-   unconstrained + constrained  # negative used
-pvalue <- pchisq(2*extra_support,df=2,lower.tail=FALSE)
-@
 
@@
 \left(p_\mathrm{Kr} + p_\mathrm{An} + \white + \draw_\mathrm{Kr} + \draw_\mathrm{An}\right)^{92}\\ \rule{20mm}{0mm}
-\left(p_\mathrm{Ks} + p_\mathrm{An} + \white + \draw_\mathrm{Ks} + \draw_\mathrm{An}\right)^{80}
-$}}
+\left(p_\mathrm{Ks} + p_\mathrm{An} + \white + \draw_\mathrm{Ks} + \draw_\mathrm{An}\right)^{80}$}}
 \end{equation}
@@
 \begin{center}
-<<echo=FALSE,fig=TRUE>>=
-dotchart(maxp(kka_3draws),pch=16)
-@ 
-\caption{Maximum likelihood estimate for \label{kka_3draws} the support function given in
-  equation  \ref{KKA_personalised_drawmonster}
-} 
+\makebox{\includegraphics{collusion_chess_draws-007.pdf}}
+\caption{Maximum likelihood estimate for \label{kka_3draws} the support function given in equation  \ref{KKA_personalised_drawmonster}} 
   \end{center}
@@
 
-In Figure~\ref{kka_3draws}, we see considerable variability in
-proclivity to draw.  Again this may be tested with the Method of
-Support, which gives a $p$-value of about 0.039, borderline
-significant: an interesting result, given that in this case no draw
-monster appears alone anywhere in the likelihood function.
-
-<<echo=FALSE,print=FALSE,results=hide>>=
-objective <- function(V){
-  karpov <- V[1]
-  kasparov <- V[2]
-  anand <- V[3]
-  white <- V[4]
-
-  draw <- (1-karpov-kasparov-anand-white)/3
-
-    return(  
-        -loglik(kka_3draws,
-                c(karpov,kasparov,anand,white,draw,draw)
-                )
-        )
-}
-
-jj <-
-constrOptim(
-    theta=c(0.1,0.1,0.1,0.1),
-    f=objective,
-    grad=NULL,
-    ui <- rbind(
-        diag(nrow=4),
-        c(-1,-1,-1,-1)
-        ),
-    ci = c(
-        rep(0,4),
-        -1
-        )
-)
-
-unconstrained <- maxp(kka_3draws,give=TRUE)$value
-constrained <- jj$value
-extra_support <-   unconstrained + constrained
-pchisq(2*extra_support,df=2,lower.tail=FALSE)
-@
+Figure~\ref{kka_3draws} shows the maximum likelihood estimate for the
+terms of equation~\ref{KKA_personalised_drawmonster}.  We see
+considerable variability in proclivity to draw.  A null of equal
+proclivity to draw may be tested with the Method of Support, which
+gives a $p$-value of about 0.039, borderline significant: an
+interesting result, given that in this case no draw monster appears
+alone anywhere in the likelihood function.
 
@@
 \left(p_\mathrm{Kr} + p_\mathrm{Ks} + \white_\mathrm{Ks} +  \draw\right)^{94}
-$}}
+$}}.
 \end{equation}
@@
 perspective and present a detailed analysis based on number of moves
-in each game\footnote{The reasoning was that hard-fought games that
-  happened to end in a draw are more likely to last longer than
-  collusive or pre-arranged draws.}.  Their analysis incorporated a
-dataset of 7355 games played in the period 1940-1978, used to
-calibrate the ability of the players.  In contrast, the analysis
-presented here uses only outcome of games (win-loss-draw) in the
-competition itself.
+in each game, reasoning that hard-fought games which happened to end
+in a draw are more likely to last longer than collusive or
+pre-arranged draws.  Their analysis incorporated a dataset of 7355
+games played in the period 1940-1978, used to calibrate the ability of
+the players.  In contrast, the analysis presented here uses only game
+outcomes (win-loss-draw) in the competition itself.
 
  \begin{table}
- \centering
- \begin{tabular}{ll|cccccccc}
- &
+ \caption{Cura\c{c}ao 1962 \label{candidates} Candidates, World Chess Championship 1963}
+ \fbox{%
+ \begin{tabular}{ll|cccccccc}  &
  &%\rotatebox{-90}{Petrosian} 
@@
  Filip     &TCH     &             & -      & -     & -     & -     & -     & -     & -  \\
- \end{tabular}
- \caption{Cura\c{c}ao 1962 \label{candidates} Candidates, World Chess Championship 1963}
+ \end{tabular}}
  \end{table}
@@
 One plausible explanation for this result is that Soviet players have
-a higher propensity to draw than western players.  We can investigate
+a higher proclivity to draw than western players.  We can investigate
 this by considering the interzonal tournament which took place in
@@
 \newcommand{\halfw}{$\nicefrac{1}{2}$}
-\newcommand{\halfb}{\colorbox{black}{\textcolor{white}{$\nicefrac{{\bf 1}}{{\bf 2}}$}}}
+\newcommand{\halfb}{\colorbox{black!65}{\makebox[2em]{\strut\textcolor{white}{$\nicefrac{{\bf 1}}{{\bf 2}}$}}}}
 \newcommand{\zerow}{$0$}
-\newcommand{\zerob}{\colorbox{black}{\textcolor{white}{$0$}}}
+\newcommand{\zerob}{\colorbox{black!65}{\makebox[2em]{\strut\textcolor{white}{$0$}}}}
 \newcommand{\oneww}{$1$}
@@
 \begin{table}
+\caption{Subset of the Fifth Interzonal round-robin, Stockholm
+  1962.\label{interzonal} Results $\in\{0,\halfw,1\}$ and box colour
+  refer to row player.  Thus, for example, Fischer played black
+  against Geller and drew. 
+  The full dataset, not shown here, has 23
+  players}
 \centering
+\fbox{%
 \begin{tabular}{l|cccccc}
@@
 $\vdots$&&&&&\\
-\end{tabular}
-\caption{Subset of the Fifth Interzonal round-robin, Stockholm
-  1962.\label{interzonal} Results $\in\{0,\halfw,1\}$ and box colour
-  refer to row player.  Thus, for example, Fischer played black
-  against Geller and drew.  The full dataset, not shown here, has 23
-  players}
+\end{tabular}}
 \end{table}
@@
 one non-Soviet player.  Then $H_N\colon \draw_s=\draw$ corresponds to
-no collusion and the likelihood function for the data of
+a null of Soviet players having the same drawing proclivity as
+non-Soviets.  The likelihood function for the data of
 Table~\ref{interzonal} would be
@@
                                (K+L+\white+\draw)
- $}}
+ $}}.
 \end{equation}
@@
 
-<<echo=FALSE,print=FALSE>>=
-do_from_scratch  <- FALSE  # set to TRUE to recalculate (slow, maybe 2 mins)
-hotstart <- indep(maxp(interzonal_collusion))
-
-f <- function(LO,give=FALSE){
-    small <- 1e-5
-
-    eq <- hotstart
-    m <- sum(eq[2:3])
-    a <- exp(LO)
-    eq[2:3] <- m/(1+a)*c(1,a)  # exact
-
-    eq[2:3] <- eq[2:3] + c(-small,small)
-
-    out <-
-        maxp(collusion,
-             startp=eq,
-             fcm=c(0, -1, exp(LO),rep(0,22)),
-             fcv=0,
-             give=TRUE)
-
-    if(give){return(out)}else{return(out$value)}
-}
-
-x <- seq(from= 1.5,to = 2.7,len=32)
-if(do_from_scratch){
-    jj <- sapply(x,f)
-} else {
-    jj <-
-        c(-246.6939, -245.7192, -244.7981, -243.9319, -243.1223, -242.3704, 
-          -241.6776, -241.0452, -240.4744, -239.9661, -239.5216, -239.1417, 
-          -238.8274, -238.5795, -238.3987, -238.2858, -238.2414, -238.2660, 
-          -238.3601, -238.5240, -238.7582, -239.0629, -239.4382, -239.8843, 
-          -240.4013, -240.9891, -241.6476, -242.3767, -243.1762, -244.0458, 
-          -244.9853, -245.9941)
-}
-@
-
 \begin{figure}[htbp]
 \begin{center}
-<<proflike,echo=FALSE,fig=TRUE>>=
-plot(x,jj-max(jj),type='b',xlab="log-odds",ylab="support")
-abline(h=c(0,-2))
-@
+\makebox{\includegraphics{collusion_chess_draws-proflike.pdf}}
 \caption{Profile likelihood \label{proflike} for odds ratio
@@
 The present work details an approach to chess analysis based on
-reification of entities representing propensity to draw, and the
+reification of entities representing proclivity to draw, and the
 first-mover (or white player) advantage.  Likelihood functions for the
@@
 
-In this context we note, following the observation of~\citet{moul2009}
+In this context we note, following the observation of Moul and Nye
 that a player may well adopt a riskier strategy after a sequence of
@@
 
-\cite{moul2009} present detailed Monte Carlo simulations
-to assess the likely impact of any Soviet cartel on tournament outcomes.
-In principle, the present system of likelihood and reified entities
-would admit such simulations and it might be interesting to compare results.
+\cite{moul2009} present detailed Monte Carlo simulations to assess the
+likely impact of a putative Soviet cartel on tournament outcomes.  In
+principle, the present system of likelihood and reified entities would
+admit such simulations and it might be interesting to compare results.
 
@@
 R code for all statistical analysis given here is available at
-\url{https://github.com/RobinHankin/hyper2}.
 
-\bibliography{chess}
+{\tt https://github.com/RobinHankin/hyper2}.
+
+\section*{References}
+
+\bibliography{jebo_chess.bib}
+
 \end{document}

### 2020-09-14 inst/jebo_chess_rebuttal.txt
line breaks, also now seven-bit clean ascii

@@
-“A generalization of the Bradley-Terry model for draws in chess”
+"A generalization of the Bradley-Terry model for draws in chess"
+
 JEBO-D-20-00916
+
 Referee Report
-This paper develops extensions to the Bradley-Terry model appropriate to Bernoulli trials
-so that the model can accommodate the chess-specific needs of draws and the first-mover
-advantage possessed by White. The author takes this extended model to various sets of
-exclusively game-level outcomes and estimates time-invariant player strengths as well as the
-importance of draws and first-mover advantage. After estimating common parameters for draws
-and first-mover advantage, the author allows for (time-invariant) player-specific parameters on
-these margins. The paper then concludes by considering distinct draw parameters for Soviet and
-non-Soviet players. Using data from the top chess events of 1962 and 1963, the author
-determines that Soviet players draw more frequently against each other than in other gamescenarios.
-The paper does a nice job of clearly laying out the various reified “monsters” that extend
-the original Bradley-Terry model to accommodate chess. This framework is plausible and
-intuitive. I also appreciated the paper’s brevity in making its point. The allowance for playerspecific tendencies to draw or to gain excess-advantage from moving first also strikes me as an
-innovation to the literature. Lastly, the paper is well-written and a fun read.
-My comments on the rest of the paper are detailed below by decreasing importance.
-1) The author never properly explores the paper’s identification strategy. As I read it, the
-identifying assumption, besides the functional form assumption, is that player strengths
-are constant over the duration of the considered data. This seems highly plausible within
-a specific event (e.g., the Candidates tournament of 1963, the Interzonals of 1962), but
-much less so with the Anand-Karpov-Kasparov example. The author does not specify the
-timespan over which these 365 games were played, but I imagine that the timespan must
-exceed 10 years. To the extent that all players have matured to full strength and none
-have yet begun to decline, this assumption would be benign, but it should be supported.
-Precise estimation will hinge on a large number of games, but this will be in tension with
-the assumption of time-invariant player-strengths. The author should explore this tension.
-2) The author neglects to motivate the paper beyond its application within the chess world.
-The tournament literature’s applications lie primarily, though definitely not exclusively,
-within compensation and labor. How could this extension be used within that context?
-Where might economists see the first-mover advantage in action? Where might the
-extension to allow draws be especially lucrative? Etc.
-3) As the author notes, the work of Moul and Nye (2009) is directly on point. The author’s
-discussion of this paper, however, is overly sparse and at times misleading. Moul and
 
-Nye use existing observed player strengths (variations of commonly used scores based on
-the Elo rating system that goes back to 1960). Moul and Nye then use these strengths
-within an ordered probit to estimate parameters that relate back to draw likelihood and
-White’s first-mover advantage. The sharp difference between that paper and this is then
-that the author is estimating player relative strengths (conditional upon a pool of players)
-while Moul and Nye use “observed” strengths.1 The latter approach is far less
-parametrically demanding, but leans much more on the assumption of normally
-distributed disturbances (basically the differing assumption of functional form). The latter
-approach also allows individuals’ player strength to vary across time and is far more
-flexible to changes in the composition of the pool of players.
-Furthermore, Moul and Nye estimate several versions of their ordered probit without data
-on the number of moves (ply-count). This is contrary to the implication of the paper (p. 6,
-immediately below Table 2), which suggests that the ply-count data are essential to that
-paper’s findings. All of that paper’s core results (significance of draws, first-mover
-advantage, a distinct draw outcome for round-robin international tournament games that
-involve two Soviets) are originally derived without ply-count data. The ply-count data are
-incorporated to explore a potential mechanism by which players in a draw-cartel might
-benefit.
-4) The paper concludes that the estimates derived from the 1962 Interzonals and the 1963
-Candidates strongly suggest that the “draw monster” won games involving two Soviet
-players more frequently than games that involved at least one non-Soviet. This is
-consistent with a Soviet cartel, but it could also be the result of Soviet-specific styles of
-play. Moul and Nye (2009) test the Soviet style-of-play hypothesis by considering
-international tournaments when the format changed from all-play-all to knockout and allSoviet tournaments. Both comparisons yield results consistent with a Soviet draw-cartel
-in international all-play-all tournaments and rejecting the Soviet style-of-play hypothesis.
-By only considering tournaments that were international and all-play-all, the author
-forgoes this critical source of identification.
-In the cartel example, the author also neglects the advantage that presumably accrued to
-Soviet players when they played non-Soviet players in all-play-all international
-tournaments. This would suggest another “monster” to be estimated, one that benefits the
-Soviet player when playing a non-Soviet opponent. Without such an addition, Soviet
+This paper develops extensions to the Bradley-Terry model appropriate
+to Bernoulli trials so that the model can accommodate the
+chess-specific needs of draws and the first-mover advantage possessed
+by White. The author takes this extended model to various sets of
+exclusively game-level outcomes and estimates time-invariant player
+strengths as well as the importance of draws and first-mover
+advantage. After estimating common parameters for draws and
+first-mover advantage, the author allows for (time-invariant)
+player-specific parameters on these margins. The paper then concludes
+by considering distinct draw parameters for Soviet and non-Soviet
+players. Using data from the top chess events of 1962 and 1963, the
+author determines that Soviet players draw more frequently against
+each other than in other game scenarios.
+
+The paper does a nice job of clearly laying out the various reified
+"monsters" that extend the original Bradley-Terry model to accommodate
+chess. This framework is plausible and intuitive. I also appreciated
+the paper's brevity in making its point. The allowance for
+playerspecific tendencies to draw or to gain excess-advantage from
+moving first also strikes me as an innovation to the
+literature. Lastly, the paper is well-written and a fun read.
+
+My comments on the rest of the paper are detailed below by decreasing
+importance.
+
+1) The author never properly explores the paper's identification
+strategy. As I read it, the identifying assumption, besides the
+functional form assumption, is that player strengths are constant over
+the duration of the considered data. This seems highly plausible
+within a specific event (e.g., the Candidates tournament of 1963, the
+Interzonals of 1962), but much less so with the Anand-Karpov-Kasparov
+example. The author does not specify the timespan over which these 365
+games were played, but I imagine that the timespan must exceed 10
+years. To the extent that all players have matured to full strength
+and none have yet begun to decline, this assumption would be benign,
+but it should be supported.
+
+Precise estimation will hinge on a large number of games, but this
+will be in tension with the assumption of time-invariant
+player-strengths. The author should explore this tension.
+
+2) The author neglects to motivate the paper beyond its application
+within the chess world.  The tournament literature's applications lie
+primarily, though definitely not exclusively, within compensation and
+labor. How could this extension be used within that context?  Where
+might economists see the first-mover advantage in action? Where might
+the extension to allow draws be especially lucrative? Etc.
+
+3) As the author notes, the work of Moul and Nye (2009) is directly on
+point. The author's discussion of this paper, however, is overly
+sparse and at times misleading. Moul and Nye use existing observed
+player strengths (variations of commonly used scores based on the Elo
+rating system that goes back to 1960). Moul and Nye then use these
+strengths within an ordered probit to estimate parameters that relate
+back to draw likelihood and White's first-mover advantage. The sharp
+difference between that paper and this is then that the author is
+estimating player relative strengths (conditional upon a pool of
+players) while Moul and Nye use "observed" strengths [footnote: Those
+relative strengths are highly predictive, with t-statistics of around
+20]. The latter approach is far less parametrically demanding, but
+leans much more on the assumption of normally distributed disturbances
+(basically the differing assumption of functional form). The latter
+approach also allows individuals' player strength to vary across time
+and is far more flexible to changes in the composition of the pool of
+players.
+
+Furthermore, Moul and Nye estimate several versions of their ordered
+probit without data on the number of moves (ply-count). This is
+contrary to the implication of the paper (p. 6, immediately below
+Table 2), which suggests that the ply-count data are essential to that
+paper's findings. All of that paper's core results (significance of
+draws, first-mover advantage, a distinct draw outcome for round-robin
+international tournament games that involve two Soviets) are
+originally derived without ply-count data. The ply-count data are
+incorporated to explore a potential mechanism by which players in a
+draw-cartel might benefit.
+
+4) The paper concludes that the estimates derived from the 1962
+Interzonals and the 1963 Candidates strongly suggest that the "draw
+monster" won games involving two Soviet players more frequently than
+games that involved at least one non-Soviet. This is consistent with a
+Soviet cartel, but it could also be the result of Soviet-specific
+styles of play. Moul and Nye (2009) test the Soviet style-of-play
+hypothesis by considering international tournaments when the format
+changed from all-play-all to knockout and allSoviet tournaments. Both
+comparisons yield results consistent with a Soviet draw-cartel in
+international all-play-all tournaments and rejecting the Soviet
+style-of-play hypothesis.  By only considering tournaments that were
+international and all-play-all, the author forgoes this critical
+source of identification.
+
+In the cartel example, the author also neglects the advantage that
+presumably accrued to Soviet players when they played non-Soviet
+players in all-play-all international tournaments. This would suggest
+another "monster" to be estimated, one that benefits the Soviet player
+when playing a non-Soviet opponent. Without such an addition, Soviet
 strengths will be overstated.
-
-1
-
-Those relative strengths are highly predictive, with t-statistics of around 20.

### 2020-09-14 inst/jebo_chess_rebuttal.txt
whitespace

@@
 playerspecific tendencies to draw or to gain excess-advantage from
-moving first also strikes me as an innovation to the
-literature. Lastly, the paper is well-written and a fun read.
+moving first also strikes me as an innovation to the literature.
+Lastly, the paper is well-written and a fun read.
 
@@
 Interzonals of 1962), but much less so with the Anand-Karpov-Kasparov
-example. The author does not specify the timespan over which these 365
-games were played, but I imagine that the timespan must exceed 10
+example.  The author does not specify the timespan over which these
+365 games were played, but I imagine that the timespan must exceed 10
 years.  To the extent that all players have matured to full strength
@@
 monster" won games involving two Soviet players more frequently than
-games that involved at least one non-Soviet. This is consistent with a
-Soviet cartel, but it could also be the result of Soviet-specific
+games that involved at least one non-Soviet.  This is consistent with
+a Soviet cartel, but it could also be the result of Soviet-specific
 styles of play.  Moul and Nye (2009) test the Soviet style-of-play
 hypothesis by considering international tournaments when the format
-changed from all-play-all to knockout and allSoviet tournaments. Both
-comparisons yield results consistent with a Soviet draw-cartel in
+changed from all-play-all to knockout and all Soviet tournaments.
+Both comparisons yield results consistent with a Soviet draw-cartel in
 international all-play-all tournaments and rejecting the Soviet

### 2020-09-14 inst/jebo_chess_rebuttal.txt
insert chevrons to mark referee's comments

--- 2020-09-14 inst/jebo_chess_rebuttal.txt
@@
 
-This paper develops extensions to the Bradley-Terry model appropriate
-to Bernoulli trials so that the model can accommodate the
-chess-specific needs of draws and the first-mover advantage possessed
-by White. The author takes this extended model to various sets of
-exclusively game-level outcomes and estimates time-invariant player
-strengths as well as the importance of draws and first-mover
-advantage. After estimating common parameters for draws and
-first-mover advantage, the author allows for (time-invariant)
-player-specific parameters on these margins. The paper then concludes
-by considering distinct draw parameters for Soviet and non-Soviet
-players. Using data from the top chess events of 1962 and 1963, the
-author determines that Soviet players draw more frequently against
-each other than in other game scenarios.
-
-The paper does a nice job of clearly laying out the various reified
-"monsters" that extend the original Bradley-Terry model to accommodate
-chess. This framework is plausible and intuitive. I also appreciated
-the paper's brevity in making its point. The allowance for
-playerspecific tendencies to draw or to gain excess-advantage from
-moving first also strikes me as an innovation to the
-literature. Lastly, the paper is well-written and a fun read.
-
-My comments on the rest of the paper are detailed below by decreasing
-importance.
-
-1) The author never properly explores the paper's identification
-strategy. As I read it, the identifying assumption, besides the
-functional form assumption, is that player strengths are constant over
-the duration of the considered data. This seems highly plausible
-within a specific event (e.g., the Candidates tournament of 1963, the
-Interzonals of 1962), but much less so with the Anand-Karpov-Kasparov
-example. The author does not specify the timespan over which these 365
-games were played, but I imagine that the timespan must exceed 10
-years. To the extent that all players have matured to full strength
-and none have yet begun to decline, this assumption would be benign,
-but it should be supported.
-
-Precise estimation will hinge on a large number of games, but this
-will be in tension with the assumption of time-invariant
-player-strengths. The author should explore this tension.
-
-2) The author neglects to motivate the paper beyond its application
-within the chess world.  The tournament literature's applications lie
-primarily, though definitely not exclusively, within compensation and
-labor. How could this extension be used within that context?  Where
-might economists see the first-mover advantage in action? Where might
-the extension to allow draws be especially lucrative? Etc.
-
-3) As the author notes, the work of Moul and Nye (2009) is directly on
-point. The author's discussion of this paper, however, is overly
-sparse and at times misleading. Moul and Nye use existing observed
-player strengths (variations of commonly used scores based on the Elo
-rating system that goes back to 1960). Moul and Nye then use these
-strengths within an ordered probit to estimate parameters that relate
-back to draw likelihood and White's first-mover advantage. The sharp
-difference between that paper and this is then that the author is
-estimating player relative strengths (conditional upon a pool of
-players) while Moul and Nye use "observed" strengths [footnote: Those
-relative strengths are highly predictive, with t-statistics of around
-20]. The latter approach is far less parametrically demanding, but
-leans much more on the assumption of normally distributed disturbances
-(basically the differing assumption of functional form). The latter
-approach also allows individuals' player strength to vary across time
-and is far more flexible to changes in the composition of the pool of
-players.
-
-Furthermore, Moul and Nye estimate several versions of their ordered
-probit without data on the number of moves (ply-count). This is
-contrary to the implication of the paper (p. 6, immediately below
-Table 2), which suggests that the ply-count data are essential to that
-paper's findings. All of that paper's core results (significance of
-draws, first-mover advantage, a distinct draw outcome for round-robin
-international tournament games that involve two Soviets) are
-originally derived without ply-count data. The ply-count data are
-incorporated to explore a potential mechanism by which players in a
-draw-cartel might benefit.
-
-4) The paper concludes that the estimates derived from the 1962
-Interzonals and the 1963 Candidates strongly suggest that the "draw
-monster" won games involving two Soviet players more frequently than
-games that involved at least one non-Soviet. This is consistent with a
-Soviet cartel, but it could also be the result of Soviet-specific
-styles of play. Moul and Nye (2009) test the Soviet style-of-play
-hypothesis by considering international tournaments when the format
-changed from all-play-all to knockout and allSoviet tournaments. Both
-comparisons yield results consistent with a Soviet draw-cartel in
-international all-play-all tournaments and rejecting the Soviet
-style-of-play hypothesis.  By only considering tournaments that were
-international and all-play-all, the author forgoes this critical
-source of identification.
-
-In the cartel example, the author also neglects the advantage that
-presumably accrued to Soviet players when they played non-Soviet
-players in all-play-all international tournaments. This would suggest
-another "monster" to be estimated, one that benefits the Soviet player
-when playing a non-Soviet opponent. Without such an addition, Soviet
-strengths will be overstated.
+> This paper develops extensions to the Bradley-Terry model appropriate
+> to Bernoulli trials so that the model can accommodate the
+> chess-specific needs of draws and the first-mover advantage possessed
+> by White.  The author takes this extended model to various sets of
+> exclusively game-level outcomes and estimates time-invariant player
+> strengths as well as the importance of draws and first-mover
+> advantage.  After estimating common parameters for draws and
+> first-mover advantage, the author allows for (time-invariant)
+> player-specific parameters on these margins.  The paper then concludes
+> by considering distinct draw parameters for Soviet and non-Soviet
+> players.  Using data from the top chess events of 1962 and 1963, the
+> author determines that Soviet players draw more frequently against
+> each other than in other game scenarios.
+> 
+> The paper does a nice job of clearly laying out the various reified
+> "monsters" that extend the original Bradley-Terry model to accommodate
+> chess.  This framework is plausible and intuitive.  I also appreciated
+> the paper's brevity in making its point. The allowance for
+> playerspecific tendencies to draw or to gain excess-advantage from
+> moving first also strikes me as an innovation to the literature.
+> Lastly, the paper is well-written and a fun read.
+> 
+> 
+> My comments on the rest of the paper are detailed below by decreasing
+> importance.
+> 
+> 1) The author never properly explores the paper's identification
+> strategy.  As I read it, the identifying assumption, besides the
+> functional form assumption, is that player strengths are constant over
+> the duration of the considered data. This seems highly plausible
+> within a specific event (e.g., the Candidates tournament of 1963, the
+> Interzonals of 1962), but much less so with the Anand-Karpov-Kasparov
+> example.  The author does not specify the timespan over which these
+> 365 games were played, but I imagine that the timespan must exceed 10
+> years.  To the extent that all players have matured to full strength
+> and none have yet begun to decline, this assumption would be benign,
+> but it should be supported.
+> 
+> Precise estimation will hinge on a large number of games, but this
+> will be in tension with the assumption of time-invariant
+> player-strengths.  The author should explore this tension.
+> 
+> 2) The author neglects to motivate the paper beyond its application
+> within the chess world.  The tournament literature's applications lie
+> primarily, though definitely not exclusively, within compensation and
+> labor.  How could this extension be used within that context?  Where
+> might economists see the first-mover advantage in action? Where might
+> the extension to allow draws be especially lucrative? Etc.
+> 
+> 3) As the author notes, the work of Moul and Nye (2009) is directly on
+> point.  The author's discussion of this paper, however, is overly
+> sparse and at times misleading.  Moul and Nye use existing observed
+> player strengths (variations of commonly used scores based on the Elo
+> rating system that goes back to 1960).  Moul and Nye then use these
+> strengths within an ordered probit to estimate parameters that relate
+> back to draw likelihood and White's first-mover advantage.  The sharp
+> difference between that paper and this is then that the author is
+> estimating player relative strengths (conditional upon a pool of
+> players) while Moul and Nye use "observed" strengths [footnote: Those
+> relative strengths are highly predictive, with t-statistics of around
+> 20].  The latter approach is far less parametrically demanding, but
+> leans much more on the assumption of normally distributed disturbances
+> (basically the differing assumption of functional form).  The latter
+> approach also allows individuals' player strength to vary across time
+> and is far more flexible to changes in the composition of the pool of
+> players.
+> 
+> Furthermore, Moul and Nye estimate several versions of their ordered
+> probit without data on the number of moves (ply-count).  This is
+> contrary to the implication of the paper (p. 6, immediately below
+> Table 2), which suggests that the ply-count data are essential to that
+> paper's findings.  All of that paper's core results (significance of
+> draws, first-mover advantage, a distinct draw outcome for round-robin
+> international tournament games that involve two Soviets) are
+> originally derived without ply-count data.  The ply-count data are
+> incorporated to explore a potential mechanism by which players in a
+> draw-cartel might benefit.
+> 
+> 4) The paper concludes that the estimates derived from the 1962
+> Interzonals and the 1963 Candidates strongly suggest that the "draw
+> monster" won games involving two Soviet players more frequently than
+> games that involved at least one non-Soviet.  This is consistent with
+> a Soviet cartel, but it could also be the result of Soviet-specific
+> styles of play.  Moul and Nye (2009) test the Soviet style-of-play
+> hypothesis by considering international tournaments when the format
+> changed from all-play-all to knockout and all Soviet tournaments.
+> Both comparisons yield results consistent with a Soviet draw-cartel in
+> international all-play-all tournaments and rejecting the Soviet
+> style-of-play hypothesis.  By only considering tournaments that were
+> international and all-play-all, the author forgoes this critical
+> source of identification.
+> 
+> In the cartel example, the author also neglects the advantage that
+> presumably accrued to Soviet players when they played non-Soviet
+> players in all-play-all international tournaments.  This would suggest
+> another "monster" to be estimated, one that benefits the Soviet player
+> when playing a non-Soviet opponent.  Without such an addition, Soviet
+> strengths will be overstated.
+>

### 2020-09-14 inst/jebo_chess_rebuttal.txt
overall response

@@
 Referee Report
+
+The referee's comments are preceded by a chevron (">") and my response
+is included inline.  I have changed line-breaks in the interests of
+clarity.
+
+Overall Response
+
+I am delighted that this insightful referee "gets"
+the manuscript.  The referee makes a number of observations and
+suggestions, all of which I agree with, and I have made changes to the
+submission in response.  I have tried very hard to preserve the
+brevity and pleasing structure that the referee explicitly notes.  In
+short, the revised manuscript is a stronger and more widely applicable
+document and I commend it to you.
+
+Software implementation note
+
+In the interests of full disclosure, I have made a significant change
+to the C++ back end of the hyper2 package:
+
+https://github.com/RobinHankin/hyper2/commit/d9d16c45daa752d4687446d7df9805439f461451#diff-25d902c24283ab8cfbac54dfa101ad31
+
+The hyper2 package now uses player names as keys in the STL map class,
+rather than integers.  The change is a conceptual streamlining which
+results in simplified R idiom for a number of cases, but does not
+change the mathematical basis of the software or alter any numerical
+results.
+
+Detailed responses
 

### 2020-09-15 inst/jebo_chess_rebuttal.txt
begin addressing point 1

@@
 
-I am delighted that this insightful referee "gets"
-the manuscript.  The referee makes a number of observations and
-suggestions, all of which I agree with, and I have made changes to the
-submission in response.  I have tried very hard to preserve the
+I am delighted that this insightful referee "gets" the manuscript.
+The referee makes a number of observations and suggestions, all of
+which I agree with, and I have made changes to the submission in
+response, marked in red.  I have tried very hard to preserve the
 brevity and pleasing structure that the referee explicitly notes.  In
@@
 > but it should be supported.
+
+An excellent observation.  In the chess world the pre-eminent strength
+measure would be the Elo rating in which the winning player takes
+points from the loser based on the pre-match difference in Elo rating.
+Elo exhibits an emergent phenomenon in which sudden changes of ability
+are not reflected instantly but have an exponentially damped response.
+
+My view is that this damping feature is sometimes desirable (old games
+are surely less informative than more recent ones) but uncontrolled in
+the sense that the time-constant is not explicit or controllable, and
+the Elo system uses an arbitrary points system with arbitrary cut-offs
+[as opposed to the objective Bradley-Terry system].
+
 > 
@@
 >
+
+I am in total agreement with this observation.  My thinking on this
+issue has been strongly affected by my recent work applying
+Bradley-Terry to Formula 1 motor racing.  In F1, drivers generally
+start their careers as weak competitors, become stronger with
+experience, and finally become uncompetitive due to age [although
+frankly Lewis Hamilton at 35yo seems to buck this general trend, as
+recent events in Tuscany would indicate].  I would suggest that chess
+players have a similar career arc, although Formula 1 Teams typically
+take a dim view of underperforming drivers---whom they rapidly and
+unceremonially drop.  The career arc for these competitors would have
+a truncated right hand side, unlike chess players who typically
+decline gradually over decades [although I would observe that
+Kasparov's current FIDE rating of 2812 is not that much less than his
+peak of 2851].
+
+Specifically, the dataset in question stretches from 1981 to 2017, a
+very wide range.  Generalized Bradley-Terry, together with reified
+entities, is capable of handling strengths varying over a career: one
+would introduce a "mid-career monster", or perhaps an "experience
+monster".  I would add here that monsters' strength is inherently
+positive, so one could deal with waning strength with age by
+introducing a "youth monster" who helps younger players.  But such
+analysis is, I would suggest, firmly in the category of further work.
+I have added a discussion of this issue to the manuscript.
+
 > 2) The author neglects to motivate the paper beyond its application

### 2020-09-15 inst/jebo_chess_revision2.tex
begin addressing point 1

@@
 and have played many games amongst themselves.  A summary may be found
-in Table~\ref{kka}.  Equation~\ref{likelihood_WD} becomes
+in Table~\ref{kka}, \textcolor{red}{showing games played from 1981 to
+  2017.  During this time, the players' strengths might have changed
+  substantially.  Statistical assessment of time-varying strength is
+  possible with the reified entity technique and this is discussed
+  under further work below.  However, we may adopt one of two
+  approaches: either we make the working assumption that the strengths
+  do not change, or alternatively we wish to make inferences about the
+  players' strength {\em averaged over a playing career}.  Either
+  way,} Equation~\ref{likelihood_WD} becomes
 
@@
 
+\textcolor{red}{One strong assumption made in the
+  Kasparov-Karpov-Anand analysis was the constancy of the players'
+  strengths over a 36-year period.  Many sports have a career arc
+  whereby competitors start their careers weak, become stronger with
+  experience, and finally become uncompetitive due to age.  Further
+  work might include the introduction of additional reified entities
+  that account for such phenomena: we might have a ``mid career
+  monster'' who helps competitors at or near their career peak.}
+
 \cite{moul2009} present detailed Monte Carlo simulations to assess the

### 2020-09-15 inst/jebo_chess_rebuttal.txt
say that different comments have different colours

@@
 which I agree with, and I have made changes to the submission in
-response, marked in red.  I have tried very hard to preserve the
-brevity and pleasing structure that the referee explicitly notes.  In
-short, the revised manuscript is a stronger and more widely applicable
-document and I commend it to you.
+response, colour coded for ease of reference.  I have tried very hard
+to preserve the brevity and pleasing structure that the referee
+explicitly notes.  In short, the revised manuscript is a stronger and
+more widely applicable document and I commend it to you.
 
@@
 analysis is, I would suggest, firmly in the category of further work.
-I have added a discussion of this issue to the manuscript.
+I have added a discussion of this issue to the manuscript (RED), in
+two parts: firstly when introducing the dataset, and secondly under
+Further Work.
 

### 2020-09-15 inst/jebo_chess_revision2.tex
rephrasing

@@
   possible with the reified entity technique and this is discussed
-  under further work below.  However, we may adopt one of two
-  approaches: either we make the working assumption that the strengths
-  do not change, or alternatively we wish to make inferences about the
-  players' strength {\em averaged over a playing career}.  Either
-  way,} Equation~\ref{likelihood_WD} becomes
+  under further work below.  However, we may treat the strengths as
+  constant by adopting one of two approaches: either we assume that
+  the strengths do not change, or alternatively we make inferences
+  about the players' strength {\em averaged over a playing career}.
+  Either way,} Equation~\ref{likelihood_WD} becomes
 

### 2020-09-15 inst/jebo_chess_revision2.tex
much better phrasing

--- 2020-09-15 inst/jebo_chess_revision2.tex
@@
   possible with the reified entity technique and this is discussed
-  under further work below.  However, we may adopt one of two
-  approaches: either we make the working assumption that the strengths
-  do not change, or alternatively we wish to make inferences about the
-  players' strength {\em averaged over a playing career}.  Either
-  way,} Equation~\ref{likelihood_WD} becomes
+  under further work below.  However, we may treat the strengths as
+  constant by adopting one of two approaches: either we assume that
+  the strengths do not change, or alternatively we interpret the
+  strengths as measures of ability {\em averaged over a playing
+    career}.  Either way,} Equation~\ref{likelihood_WD} becomes
 

### 2020-09-15 inst/jebo_chess_rebuttal.txt
discussion of Elo rating
* inst/jebo_chess_rebuttal.txt:

@@
 
+But frankly I don't want to open the can of worms that would be
+comparison between Elo and Bradley-Terry.  I had a protracted argument
+on this issue with a chess enthusiast; let us say the discussion was
+intense and robust.  The problem is that Elo was designed specifically
+to downweight old data and this downweighting is what users of such
+metrics want.  Unfortunately, what my correspondent wanted from BT was
+pretty much Elo: a somewhat circular position.
+
 > 
@@
 >
+
 > 3) As the author notes, the work of Moul and Nye (2009) is directly on

### 2020-09-15 inst/jebo_chess_revision2.tex
discussion of Elo rating
* inst/jebo_chess_rebuttal.txt:

@@
   Kasparov-Karpov-Anand analysis was the constancy of the players'
-  strengths over a 36-year period.  Many sports have a career arc
+  strengths over a 36-year period.  One approach might be to consider
+  only matches played over a shorter span, but this would result in a
+  smaller and less informative dataset; the tension between timeliness
+  and informativeness is a delicate and difficult matter.}
+
+  \textcolor{orange}{In a wider context, many sports have a career arc
     whereby competitors start their careers weak, become stronger with

### 2020-09-16 inst/jebo_chess_rebuttal.txt
work on incumbent advantage

@@
 
+(manuscript changes addressing this comment are marked in ORANGE).
+
+The original submission discussed applications to MasterChef
+Australia, consumer behaviour and mating preference, but I was not
+aware of tournament literature as a distinct academic field.  I have
+been studying Lazear and Rosen and pondering how reified monsters
+could help understanding tournaments in a workplace remuneration
+context.  Lazear and Rosen supply a good rationale to consider only
+rank in situations such as Formula 1 motor racing
+
+First-mover advantage is closely related to incumbent advantage,
+something which had not occurred to me until now.  One of the papers
+cited [Hankin 2010] uses a variant of Bradley-Terry to assess
+incumbent advantage in biology (territory defense by lizards) and I
+have added a more explicit pointer to this.
+
+Returning to Lazear and Rosen, schemes that reward rank are common in
+sports (notably Formula 1, discussed above, and sculling, both good
+applications of the Plackett-Luce extension of Bradley-Terry).
+
+Also, I have discussed Formula 1 motor racing above under part 1.
+
 > 3) As the author notes, the work of Moul and Nye (2009) is directly on

### 2020-09-16 inst/jebo_chess_revision2.tex
work on incumbent advantage

@@
 secondly, many chess games, about 39.4\%, are drawn (figures from
-Chessbase).
+Chessbase).  \textcolor{orange}{First-mover advantage is closely
+  related to incumbent advantage, and the techniques presented here
+  have been applied to ecological incumbency~\cite{hankin2010}.}
 
@@
 
+\textcolor{orange}{The maximum likelihood estimate for strengths
+  furnishes a natural ranking for the competitors.  \cite{lazear1981}
+  observe that such ranking systems occur naturally in organizational
+  situations (tournament theory) and the analysis presented here might
+  be applicable to this situation, with the advantage of allowing for
+  both draws and incumbent advantage.}
+
 \cite{moul2009} present detailed Monte Carlo simulations to assess the

### 2020-09-18 inst/jebo_chess_rebuttal.txt
slight tweaks

@@
 to downweight old data and this downweighting is what users of such
-metrics want.  Unfortunately, what my correspondent wanted from BT was
-pretty much Elo: a somewhat circular position.
+metrics want.  Unfortunately, what my correspondent wanted from
+Bradley-Terry strengths was pretty much Elo: a somewhat circular
+position.
 
@@
 issue has been strongly affected by my recent work applying
-Bradley-Terry to Formula 1 motor racing.  In F1, drivers generally
-start their careers as weak competitors, become stronger with
-experience, and finally become uncompetitive due to age [although
-frankly Lewis Hamilton at 35yo seems to buck this general trend, as
-recent events in Tuscany would indicate].  I would suggest that chess
-players have a similar career arc, although Formula 1 Teams typically
-take a dim view of underperforming drivers---whom they rapidly and
-unceremonially drop.  The career arc for these competitors would have
-a truncated right hand side, unlike chess players who typically
-decline gradually over decades [although I would observe that
-Kasparov's current FIDE rating of 2812 is not that much less than his
-peak of 2851].
+Bradley-Terry to Formula 1 motor racing [I have a manuscript under
+review at JQAS].  In F1, drivers generally start their careers as weak
+competitors, become stronger with experience, and finally become
+uncompetitive due to age [although frankly Lewis Hamilton at 35yo
+seems to buck this general trend, as recent events in Tuscany would
+indicate].  I would suggest that chess players have a similar career
+arc, although Formula 1 Teams typically take a dim view of
+underperforming drivers---whom they rapidly and unceremonially drop.
+The career arc for these competitors would have a truncated right hand
+side, unlike chess players who typically decline gradually over
+decades [although I would observe that Kasparov's current FIDE rating
+of 2812 is not that much less than his peak of 2851].
 
@@
 would introduce a "mid-career monster", or perhaps an "experience
-monster".  I would add here that monsters' strength is inherently
+monster".  I would add here that monsters' strengths are inherently
 positive, so one could deal with waning strength with age by
@@
 
-The original submission discussed applications to MasterChef
+The submission discusses applications to MasterChef
 Australia, consumer behaviour and mating preference, but I was not
@@
 incumbent advantage in biology (territory defense by lizards) and I
-have added a more explicit pointer to this.
+have added a more explicit pointer to this.  One point of contact with
+tournament literature might be incumbent advantage possessed in a
+remuneration tournament by candidates who are working temporarily in
+the capacity being fought over, such as an 'acting CEO' or 'acting
+head of department'.  The appropriate incumbent monster
 
@@
 
-Also, I have discussed Formula 1 motor racing above under part 1.
+Also, I have discussed Formula 1 motor racing above under part 1 of
+the review.
 

### 2020-09-20 inst/jebo_chess_rebuttal.txt
more work

@@
 >
+
+This interpretation of Moul and Nye is very close to my own, and
+re-reading my work I realise that the sentence in question was poorly
+phrased and indeed unintentionally misleading.  What I wanted to say
+was that chess draws have been studied by many workers and Moul and
+Nye observe that
+
+I have rephrased the sentence in question, and added a few more points
+of reference to Mouland Nye, in GREEN.
+
 > 4) The paper concludes that the estimates derived from the 1962

### 2020-09-27 inst/jebo_chess_rebuttal.txt
more work

@@
 to preserve the brevity and pleasing structure that the referee
-explicitly notes.  In short, the revised manuscript is a stronger and
-more widely applicable document and I commend it to you.
+explicitly notes.  In short, the revised manuscript is a stronger,
+better argued, and more widely applicable document and I commend it to
+you.
 
@@
 are surely less informative than more recent ones) but uncontrolled in
-the sense that the time-constant is not explicit or controllable, and
-the Elo system uses an arbitrary points system with arbitrary cut-offs
-[as opposed to the objective Bradley-Terry system].
+the sense that the time-constant is not explicit or controllable.
+Also, the Elo system uses an arbitrary points system with arbitrary
+cut-offs [as opposed to the objective Bradley-Terry system] and has
+the undesirable feature that changing the points system or the
+cut-offs will change players' Elo ratings even if their playing
+performance is unchanged.
 
@@
 metrics want.  Unfortunately, what my correspondent wanted from
-Bradley-Terry strengths was pretty much Elo: a somewhat circular
-position.
+Bradley-Terry strengths was pretty much Elo with all its
+arbitrariness: a somewhat circular position.
 
@@
 context.  Lazear and Rosen supply a good rationale to consider only
-rank in situations such as Formula 1 motor racing
+rank in situations such as Formula 1 motor racing.
 
@@
 the capacity being fought over, such as an 'acting CEO' or 'acting
-head of department'.  The appropriate incumbent monster
+head of department'.  The appropriate incumbent monster would furnish
+a quantification of the incumbent advantage.
 
@@
 >
+
+This is an excellent suggestion, one which I feel nicely rounds out
+the analysis.  I have included a "rest monster" whose strength aids
+players following a collusive draw.  Reified Bradley-Terry allows one
+not only to establish the existence of such a rest monster, but also
+to quantify its strength and I have done just this in the revision.

### 2020-09-28 inst/curacao1962_threeplayers_rest_monster.Rmd
rest monster (flawed, not sure what is wrong)

---
title: "curacao_threeplayers_rest_monster"
author: "Robin Hankin"
date: "28/09/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This work extends that of `curacao1962_threeplayers.R` by including
a rest monster.

```{r}
library("hyper2")

a <- read.table("curacao1962_candidates.txt",header=FALSE)
colnames(a) <- c("W", "B", "result", "round")
head(a)
```

specify the players:

```{r}
  players <- 
    c("Petrosian", "Keres", "Geller", "Fischer", "Korchnoi",
      "Benko", "Tal", "Filip")
  names(players) <-   # accused of collusion or not
    c("coll", "coll", "coll", "USA", "USSR", "USA", "USSR", "TCH")
collusive_players <-     c("Petrosian", "Keres", "Geller")
 ```

```{r}
d <- as.matrix(cbind(as.character(a$V1),as.character(a$V2)))
colnames(d) <- c("white","black")

stopifnot(all(c(d) %in% players))

jj <- matrix(0,28,8)
colnames(jj) <- players

cumulative_matches <- jj
cumulative_collusive_matches <- jj
cumulative_noncollusive_matches <- jj

for(i in seq_len(28)){
  for(j in seq_along(players)){
    p <- players[j]
    jj <- subset(a,a$round < i)
    cm <- 0
    ncm <- 0
    for(k in seq_len(nrow(jj))){

      w <- jj$W[k]
      b <- jj$B[k]

      this_match <- c(w,b)
      if((players[j] %in% c(w,b))){
        if(players[j] %in% collusive_players){  # this player is collusive
          if((w %in% collusive_players) & (b %in% collusive_players)){  # both players collusive
            cm <- cm+1
          } else { # non-collusive match
            ncm <- ncm+1
          }
        } else { # noncollusive player
          ncm <- ncm+1
        }
      } 
    } # k loop closes
    cumulative_collusive_matches[i,j] <- cm
    cumulative_noncollusive_matches[i,j] <- ncm
    cumulative_matches[i,j] <- cm + ncm
  } # j loop closes
} # i loop closes
cumulative_matches
cumulative_collusive_matches
cumulative_noncollusive_matches
```

In the above three matrices, each row is a round, and the entries show
the number of matches, collusive matches, and noncollusive matches
played up to and including that round.

```{r}
a <- cbind(a,cncmp_white=0,cncmp_black=0)
for(i in seq_len(nrow(a))){
  r <- a$round[i]
  jj_w <- which(colnames(cumulative_noncollusive_matches)==a$W[i])
  jj_b <- which(colnames(cumulative_noncollusive_matches)==a$B[i])
  ncmp_w <- cumulative_noncollusive_matches[r, jj_w]
  ncmp_b <- cumulative_noncollusive_matches[r, jj_b]
  a$cncmp_white[i] <- ncmp_w
  a$cncmp_black[i] <- ncmp_b
}
head(a)
tail(a)
```

Thus the final match, between Petrosian and Filip was played in round
28, ended in a draw, and by that time Petrosian had played 18
noncollusive matches and Filip had played 26.

Now create a hyper2 object with appropriate reified monsters:

```{r}
H <- hyper2(pnames=c(players,"white","draw","cdraw","rest"))
for(i in seq_len(nrow(a))){
  white_player <- a[i,1]
  black_player <- a[i,2]

  collusive <- (white_player %in% collusive_players) & (black_player %in% collusive_players)
  if(collusive){
    drawmonster <- "cdraw"
  } else {
    drawmonster <- "draw"
  }

  white_player_rested <- a[i,5] > a[i,6]
  black_player_rested <- a[i,5] < a[i,6]

  result <- a[i,3]

    if(white_player_rested){
      if(result == "1-0"){   # white win
      H[c(white_player,"white","rest"                         )] %<>% inc
      H[c(white_player,"white","rest",black_player,drawmonster)] %<>% dec
    } else if(black_player_rested){
      H[c(white_player,"white"                                )] %<>% inc
      H[c(white_player,"white","rest",black_player,drawmonster)] %<>% dec
    } else { # neither player rested
      H[c(white_player,"white"                                )] %<>% inc
      H[c(white_player,"white",       black_player,drawmonster)] %<>% dec
    }
  } else if(result == "0-1"){  # black win
    if(white_player_rested){
      H[c(                            black_player            )] %<>% inc
      H[c(white_player,"white","rest",black_player,drawmonster)] %<>% dec
    } else if(black_player_rested){
      H[c(                     "rest",black_player            )] %<>% inc
      H[c(white_player,"white","rest",black_player,drawmonster)] %<>% dec
    } else { # neither player rested
      H[c(                            black_player            )] %<>% inc
      H[c(white_player,"white",       black_player,drawmonster)] %<>% dec
    }
  } else if(result == "1/2-1/2"){ # draw
      H[c(                                         drawmonster)] %<>% inc
      H[c(white_player,"white","rest",black_player,drawmonster)] %<>% dec
  } else { # not a white win or a black win or a draw
    stop("this cannot happen")
  }
}
H
pie(maxp(H))
```

### 2020-09-28 inst/curacao1962_threeplayers_rest_monster.Rmd
typo

@@
 
+  if(result == "1-0"){   # white win
     if(white_player_rested){
-      if(result == "1-0"){   # white win
       H[c(white_player,"white","rest"                         )] %<>% inc
@@
 pie(maxp(H))
+summary(H)
 ```

### 2020-09-28 inst/curacao1962_threeplayers_rest_monster.Rmd
more work

--- 2020-09-28 inst/curacao1962_threeplayers_rest_monster.Rmd
@@
 } # i loop closes
+```
+
+So we have three matrices `cumulative_matches`,
+`cumulative_collusive_matches` and `cumulative_noncollusive_matches`:
+
+```{r}
 cumulative_matches
@@
 
+We can see the difference between the two players:
+
+```{r}
+diff <- a$cncmp_white-a$cncmp_black  # difference in each game
+table(diff)
+table(abs(diff))
+cumsum(table(abs(diff)))
+```
+
 Now create a hyper2 object with appropriate reified monsters:
@@
 ```{r}
-H <- hyper2(pnames=c(players,"white","draw","cdraw","rest"))
+H <- hyper2(pnames=c(players,"white","draw","rest"))
 for(i in seq_len(nrow(a))){
@@
   if(collusive){
-    drawmonster <- "cdraw"
+    drawmonster <- "draw"
   } else {
@@
 
-  white_player_rested <- a[i,5] > a[i,6]
-  black_player_rested <- a[i,5] < a[i,6]
+  white_player_rested <- a[i,5] - a[i,6] > 0
+  black_player_rested <- a[i,6] - a[i,5] > 0
 
@@
 H
-pie(maxp(H))
+(mH <- maxp(H))
+pie(mH)
+summary(H)
 ```

### 2020-09-29 inst/curacao1962_threeplayers_rest_monster.Rmd
verification; reveals a flaw

@@
 H <- hyper2(pnames=c(players,"white","draw","rest"))
+prob <- function(...){stop("must be a white win or a black win or a draw")}
 for(i in seq_len(nrow(a))){
@@
       H[c(white_player,"white"                                )] %<>% inc
-      H[c(white_player,"white",       black_player,drawmonster)] %<>% dec
+      H[c(white_player,"white"       ,black_player,drawmonster)] %<>% dec
     }
@@
       H[c(                                         drawmonster)] %<>% inc
-      H[c(white_player,"white","rest",black_player,drawmonster)] %<>% dec
+      H[c(white_player,"white"       ,black_player,drawmonster)] %<>% dec
   } else { # not a white win or a black win or a draw
-    stop("this cannot happen")
+    prob()
   }
@@
 ```
+
+```{r,label=verify}
+I <- hyper2(pnames=c(players,"white","draw","rest"))
+for(i in seq_len(nrow(a))){
+  white_player <- a[i,1]
+  black_player <- a[i,2]
+
+  collusive <- (white_player %in% collusive_players) & (black_player %in% collusive_players)
+  if(collusive){
+    drawmonster <- "draw"
+  } else {
+    drawmonster <- "draw"
+  }
+
+  white_player_rested <- a[i,5] - a[i,6] > 0
+  black_player_rested <- a[i,6] - a[i,5] > 0
+
+  result <- a[i,3]
+
+  if(white_player_rested){  
+    if(       result == "1-0"){ 
+      I[c(white_player,             "white","rest"            )] %<>% inc
+    } else if(result == "0-1"){ 
+      I[c(             black_player                           )] %<>% inc
+    } else if(result == "1/2-1/2"){                         
+      I[c(                                         drawmonster)] %<>% inc
+    } else { prob() }
+      I[c(white_player,black_player,"white","rest",drawmonster)] %<>% dec
+  } else if(black_player_rested){
+    if(       result == "1-0"){ 
+      I[c(white_player,             "white"                   )] %<>% inc
+    } else if(result == "0-1"){ 
+      I[c(             black_player,        "rest"            )] %<>% inc
+    } else if(result == "1/2-1/2"){                         
+      I[c(                                         drawmonster)] %<>% inc
+    } else { prob() }
+      I[c(white_player,black_player,"white","rest",drawmonster)] %<>% dec
+  } else { # neither player rested
+    if(       result == "1-0"){ 
+      I[c(white_player,             "white"                   )] %<>% inc
+    } else if(result == "0-1"){ 
+      I[c(             black_player                           )] %<>% inc
+    } else if(result == "1/2-1/2"){                         
+      I[c(                                         drawmonster)] %<>% inc
+    } else { prob() }
+      I[c(white_player,black_player,"white",       drawmonster)] %<>% dec
+  }
+}
+
+I
+
+summary(I)
+maxp(I)
+I==H
+I + (-1)*H
+```

### 2020-09-29 inst/curacao1962_threeplayers_rest_monster.Rmd
clarification of monster's status

@@
       I[c(white_player,black_player,"white","rest",drawmonster)] %<>% dec
-  } else { # neither player rested
+  } else { # neither player rested, rest monster inactive
     if(       result == "1-0"){ 

### 2020-09-29 inst/curacao1962_threeplayers_rest_monster.Rmd
fixes #65

--- 2020-09-28 inst/curacao1962_threeplayers_rest_monster.Rmd
@@
 ```{r}
+problem <- function(...){stop("must be a white win or a black win or a draw")}
+
 H <- hyper2(pnames=c(players,"white","draw","rest"))
@@
 
-  if(result == "1-0"){   # white win
   if(white_player_rested){  
-      H[c(white_player,"white","rest"                         )] %<>% inc
-      H[c(white_player,"white","rest",black_player,drawmonster)] %<>% dec
+    if(       result == "1-0"){ 
+      H[c(white_player,             "white","rest"            )] %<>% inc
+    } else if(result == "0-1"){ 
+      H[c(             black_player                           )] %<>% inc
+    } else if(result == "1/2-1/2"){                         
+      H[c(                                         drawmonster)] %<>% inc
+    } else { problem() }
+      H[c(white_player,black_player,"white","rest",drawmonster)] %<>% dec
   } else if(black_player_rested){
-      H[c(white_player,"white"                                )] %<>% inc
-      H[c(white_player,"white","rest",black_player,drawmonster)] %<>% dec
-    } else { # neither player rested
-      H[c(white_player,"white"                                )] %<>% inc
-      H[c(white_player,"white",       black_player,drawmonster)] %<>% dec
-    }
-  } else if(result == "0-1"){  # black win
-    if(white_player_rested){
+    if(       result == "1-0"){ 
+      H[c(white_player,             "white"                   )] %<>% inc
+    } else if(result == "0-1"){ 
+      H[c(             black_player,        "rest"            )] %<>% inc
+    } else if(result == "1/2-1/2"){                         
+      H[c(                                         drawmonster)] %<>% inc
+    } else { problem() }
+      H[c(white_player,black_player,"white","rest",drawmonster)] %<>% dec
+  } else { # neither player rested, rest monster inactive
+    if(       result == "1-0"){ 
+      H[c(white_player,             "white"                   )] %<>% inc
+    } else if(result == "0-1"){ 
       H[c(             black_player                           )] %<>% inc
-      H[c(white_player,"white","rest",black_player,drawmonster)] %<>% dec
-    } else if(black_player_rested){
-      H[c(                     "rest",black_player            )] %<>% inc
-      H[c(white_player,"white","rest",black_player,drawmonster)] %<>% dec
-    } else { # neither player rested
-      H[c(                            black_player            )] %<>% inc
-      H[c(white_player,"white",       black_player,drawmonster)] %<>% dec
-    }
-  } else if(result == "1/2-1/2"){ # draw
+    } else if(result == "1/2-1/2"){                         
       H[c(                                         drawmonster)] %<>% inc
-      H[c(white_player,"white","rest",black_player,drawmonster)] %<>% dec
-  } else { # not a white win or a black win or a draw
-    stop("this cannot happen")
+    } else { problem() }
+      H[c(white_player,black_player,"white",       drawmonster)] %<>% dec
   }
 }
+
 H
-(mH <- maxp(H))
+summary(H)
+mH <- maxp(H)
+mH
 pie(mH)
-summary(H)
 ```

### 2020-09-29 inst/curacao1962_threeplayers_rest_monster.Rmd
lower-level (Fisher) analysis of rested players

@@
 
-  white_player_rested <- a[i,5] - a[i,6] > 0
-  black_player_rested <- a[i,6] - a[i,5] > 0
+  white_player_rested <- a[i,5] - a[i,6] > 1
+  black_player_rested <- a[i,6] - a[i,5] > 1
 
@@
 ```
+
+```{r}
+table(a[,3],a[,5]-a[,6])
+```
+
+A bit of tidying:
+
+```{r}
+aa <- subset(a,a[,5] != a[,6])
+aa <- subset(aa,aa[,3] != "1/2-1/2")
+
+M <- table(aa[,3],white_rested=aa[,5]<aa[,6])
+M
+fisher.test(M,alternative="greater")
+```

### 2020-09-30 inst/curacao1962_threeplayers_rest_monster.Rmd
bugfix
use specificp.gt.test()
slight rearrangement and better description of cncmp

@@
 the number of matches, collusive matches, and noncollusive matches
-played up to and including that round.
+played up to and including that round for each of the 8 players.  Each
+(named) column corresponds to a player.  Looking at the last row of
+`cumulative_noncollusive_matches` we see that in round 28, players
+Petrosian, Keres, and Geller had played a total of 18 noncollusive
+matches while the other players had played considerably more and would
+be expected to be more fatigued at this point in the tournament.  Note
+that Tal retired after round 21 due to illness, which explains why he
+had played fewer noncollusive games than Fischer, Korchnoi, Benko, or
+Filip at this point.
 
@@
 
-We can see the difference between the two players:
+We can see the differences between cumulative noncollusive games
+played between Black and White, summed over the whole tournament:
 
@@
 ```
+
+We can assess a null of no effect of fatigue:
+
+```{r}
+table(a[,3],a[,5]-a[,6])
+```
+
+Then Fisher test:
+```{r}
+aa <- subset(a,a[,5] != a[,6])
+aa <- subset(aa,aa[,3] != "1/2-1/2")
+
+M <- table(aa[,3],white_rested=aa[,5]<aa[,6])
+M
+fisher.test(M,alternative="greater")
+```
+
+Highly signficant.
 
@@
 
-  white_player_rested <- a[i,5] - a[i,6] > 1
-  black_player_rested <- a[i,6] - a[i,5] > 1
+  white_player_rested <- a[i,6] - a[i,5] > 1
+  black_player_rested <- a[i,5] - a[i,6] > 1
 
@@
 
-```{r}
-table(a[,3],a[,5]-a[,6])
+ Now a hypothesis test
+
+```{r,label=hyptest,cache=TRUE}
+specificp.gt.test(H,"rest",0)
 ```
-
-A bit of tidying:
-
-```{r}
-aa <- subset(a,a[,5] != a[,6])
-aa <- subset(aa,aa[,3] != "1/2-1/2")
-
-M <- table(aa[,3],white_rested=aa[,5]<aa[,6])
-M
-fisher.test(M,alternative="greater")
-```

### 2020-09-30 inst/curacao1962_threeplayers_rest_monster.Rmd
reconfiguration

@@
 ```{r}
-aa <- subset(a,a[,5] != a[,6])
-aa <- subset(aa,aa[,3] != "1/2-1/2")
+b <- subset(a,a[,5] != a[,6])
+b <- subset(b,abs(b[,5]-b[,6])>1)
+b <- subset(b,b[,3] != "1/2-1/2")
+b <- cbind(b,white_rested = b[,5] < b[,6])
+b
 
-M <- table(aa[,3],white_rested=aa[,5]<aa[,6])
+M <- table(b[,3],white_rested=b[,7])
 M
@@
 
-```{r,label=hyptest,cache=TRUE}
+```{r,label=hyptestrest,cache=TRUE}
 specificp.gt.test(H,"rest",0)
 ```
+
+```{r,label=hyptestdraw,cache=TRUE}
+specificp.gt.test(H,"draw",0)
+```
+
+```{r,label=hyptestwhite,cache=TRUE}
+specificp.gt.test(H,"white",0)
+```

### 2020-09-30 inst/jebo_chess_revision2.tex
more work
write up the rest monster work

@@
 constraint becomes nonlinear, making likelihood maximization
-difficult.  \citet{moul2009} consider chess draws and use
+difficult.
+
+The seminal statistical analysis of 
+\citet{moul2009} consider chess draws and use
 sophisticated regression techniques to analyse games in terms of not
@@
 
+\definecolor{DarkGreen}{RGB}{14,99,4}
+
+\textcolor{DarkGreen}{It is reasonable to suppose that collusion of
+  the type considered here is ``effective'' in the sense that
+  noncollusive games are detrimental to future play in a way that
+  collusively drawn games are not.  At the end of the tournament,
+  Petrosian, Keres and Geller had each played a total of 18
+  noncollusive matches while Fisher, Benko and Filip had played a
+  total of 26 or 27 (Tal, retiring mid tournament due to illness,
+  played only 21).  For each game, we count the noncollusive games
+  played up to that point in the match for White and Black.  If, in
+  that game, the number of noncollusive games differ by more than 1,
+  we posit a ``rest monster'' whose strength helps the player who has
+  played fewer noncollusive games.  Observing that the rest monster is
+  active in 49 of the 105 games of the tournament, we may consider
+  these 49 games and, discarding draws, test whether rested players
+  actually win more often than expected; observations are given in
+  Table~\ref{restmonsterfisher} .  We reject the null of rested
+  players having the same win probability as their more tired
+  opponents (one-sided Fisher's exact test, $p\simeq 10^{-4}$).}
+
+\begin{table}
+\textcolor{DarkGreen}{
+  \caption{Results for won games in Cura\c{c}ao 1962  between
+      rested and unrested players    \label{restmonsterfisher}}
+  \centering
+  \fbox{%
+    \begin{tabular}{l|cc}\\
+      & White rested & Black rested \\ \hline
+      White wins  & $9$   & $ 3$   \\
+      Black wins  & $0$   & $13$   \\
+  \end{tabular}}
+  }
+\end{table}
+
+\textcolor{DarkGreen}{The Bradley-Terry model together with a white
+  player advantage, draw monster, and rest monster, can quantify the
+  strength of the rest monster at about 26\% (compare the draw monster
+  at 24\%).  The null of the rest monster having zero strength is
+  rejected with a likelihood ratio of about $e^{11.9}$ or a $p$-value
+  of about $10^{-6}$.}
+
 \section{Conclusions and further work}
@@
 
+\subsection*{\textcolor{DarkGreen}{The ``rest monster'' is due to a suggestion by an insightful JEBO referee}.}
+
 R code for all statistical analysis given here is available at

### 2020-09-30 inst/jebo_chess_revision2.tex
typo --   Fisher is not Fischer!

@@
   Petrosian, Keres and Geller had each played a total of 18
-  noncollusive matches while Fisher, Benko and Filip had played a
+  noncollusive matches while Fischer, Benko and Filip had played a
   total of 26 or 27 (Tal, retiring mid tournament due to illness,

### 2020-09-30 inst/jebo_chess_revision2.tex
write up rest monster

--- 2020-09-16 inst/jebo_chess_revision2.tex
@@
 
+\definecolor{DarkGreen}{RGB}{14,159,4}
+
+\textcolor{DarkGreen}{It is reasonable to suppose that collusion of
+  the type considered here is ``effective'' in the sense that
+  noncollusive games are detrimental to future play in a way that
+  collusively drawn games are not.  At the end of the tournament,
+  Petrosian, Keres and Geller had each played a total of 18
+  noncollusive matches while Fischer, Benko and Filip had played a
+  total of 26 or 27 (Tal, retiring mid tournament due to illness,
+  played only 21).  For each game in the tournament, we count the
+  total noncollusive games played up to that point for both the white
+  and black player.  If, in that game, the number of previous
+  noncollusive games for the white and black player differs by more
+  than one, we posit a ``rest
+  monster'' whose strength helps the player who has played fewer
+  noncollusive games.}
+
+\textcolor{DarkGreen}{
+  Observing that the rest monster is
+  active in 49 of the 105 games of the tournament, we may consider
+  these 49 games and, discarding draws, test whether rested players
+  actually win more often than expected; observations are given in
+  Table~\ref{restmonsterfisher} .  We reject the null of rested
+  players having the same strength as their more tired opponents
+  (one-sided Fisher's exact test, $p\simeq 10^{-4}$).}
+
+\begin{table}
+\textcolor{DarkGreen}{
+  \caption{Results for won games in Cura\c{c}ao 1962  between
+      rested and unrested players    \label{restmonsterfisher}}
+  \centering
+  \fbox{%
+    \begin{tabular}{l|cc}\\
+      & White rested & Black rested \\ \hline
+      White wins  & $9$   & $ 3$   \\
+      Black wins  & $0$   & $13$   \\
+  \end{tabular}}
+  }
+\end{table}
+
+\textcolor{DarkGreen}{The Bradley-Terry model together with a white
+  player advantage, draw monster, and rest monster, estimates the
+  strength of the rest monster at about 26\% (compare the draw monster
+  at 24\% and the White advantage at about 3\%).  The null of the rest
+  monster having zero strength is rejected with a likelihood ratio of
+  about $e^{11.9}$ or a $p$-value of about $10^{-6}$.  It is an
+  interesting observation that neither Table~\ref{restmonsterfisher},
+  nor reified Bradley-Terry, admit a White advantage (one-sided
+  McNemar test, $p=\frac{1}{8}$, likelihood ratio $1.609$
+  respectively).  }
+
 \section{Conclusions and further work}

### 2020-09-30 inst/curacao1962_threeplayers_rest_monster.Rmd
footnote on difference in games played

@@
 ```{r}
+##  To justify the footnote "No game in the tournament is played
+##  between two players with cumulative game count (collusivitity
+##  ignored), differ by more than one" in jebo_chess_revision2.tex,
+##  replace `cumulative_noncollusive_matches` with
+##  `cumulative_matches` in this chunk.
+##
+
 a <- cbind(a,cncmp_white=0,cncmp_black=0)

### 2020-09-30 inst/jebo_chess_revision2.tex
footnote on difference in games played

@@
   noncollusive games for the white and black player differs by more
-  than one, we posit a ``rest
+  than one\footnote{\textcolor{DarkGreen}{If collusivity is ignored,
+    no game in the tournament is played between two players with
+    cumulative game count differ by more than one.}}, we posit a ``rest
   monster'' whose strength helps the player who has played fewer

### 2020-09-30 inst/jebo_chess_rebuttal.txt
consistentify colour coding

@@
 I have rephrased the sentence in question, and added a few more points
-of reference to Mouland Nye, in GREEN.
+of reference to Mouland Nye, in BLUE.
 
@@
 
+(comments addressing this comment in DARK GREEN)
 This is an excellent suggestion, one which I feel nicely rounds out

### 2020-10-01 inst/jebo_chess_rebuttal.txt
more accounting for comments on Moul and Nye (blue)

@@
 re-reading my work I realise that the sentence in question was poorly
-phrased and indeed unintentionally misleading.  What I wanted to say
-was that chess draws have been studied by many workers and Moul and
-Nye observe that
-
-I have rephrased the sentence in question, and added a few more points
-of reference to Mouland Nye, in BLUE.
+phrased and indeed clearly (but unintentionally) misleading.  I have
+rephrased the sentence in question, and added a few more points of
+reference to Moul and Nye, in BLUE.
 

### 2020-10-01 inst/jebo_chess_revision2.tex
more accounting for comments on Moul and Nye (blue)

@@
 
-The seminal statistical analysis of 
-\citet{moul2009} consider chess draws and use
-sophisticated regression techniques to analyse games in terms of not
-only the result, but also the ply-count, data not used in the present
-analysis.
+\textcolor{blue}{The seminal work of \citet{moul2009} presents
+  sophisticated regression techniques to analyse draws and first-mover
+  advantage.  They compare all-play-all events with knockout format
+  events to test---and reject---hypotheses of a distinct Soviet
+  style-of-play; the work I present also considers Soviet proclivity
+  to draw, using data from a single event, specifically the Stockholm
+  Interzonal tournament of (1962).}
 
@@
 
-\citet{moul2009} consider this accusation from a statistical
-perspective and present a detailed analysis based on number of moves
-in each game, reasoning that hard-fought games which happened to end
-in a draw are more likely to last longer than collusive or
-pre-arranged draws.  Their analysis incorporated a dataset of 7355
-games played in the period 1940-1978, used to calibrate the ability of
-the players.  In contrast, the analysis presented here uses only game
-outcomes (win-loss-draw) in the competition itself.
+\textcolor{blue}{\citet{moul2009} consider this accusation from a
+  statistical perspective.  Their analysis incorporated a dataset of
+  7355 games played in the period 1940-1978, used to calibrate the
+  ability of the players.  In contrast, the analysis presented here
+  uses only game outcomes (win-loss-draw) in the competition itself.}
 
@@
 admit such simulations and it might be interesting to compare results.
+\textcolor{blue}{\cite{moul2009} go on to compare round robin
+  tournaments to knockout format events, searching for evidence of a
+  Soviet draw cartel.  By only considering international tournaments
+  [where Soviets have incentive to play differently against fellow
+    Soviets or Westerners] the analysis presented here does not allow
+  me to do this but the technique could easily be applied to a more
+  informative dataset.}
 
-\subsection*{Supplementary material}
+\textcolor{DarkGreen}{The Bradley-Terry techniques used here furnish
+  evidence that collusive games played during during Cura\c{c}ao 1962
+  were less detrimental to future performance than noncollusive games.
+  However, there was no cognitively realistic model for mental fatigue
+  (for example, downweighting heavy workloads long in the past) and
+  the criterion for restedness itself was quite coarse. not allowing
+  It would be possible in principle to investigate the effect of
+  recovery using similar reified entities.}
 
-\subsection*{\textcolor{DarkGreen}{The ``rest monster'' is due to a suggestion by an insightful JEBO referee}.}
+\subsection*{\textcolor{DarkGreen}{Acknowledgement}}
+
+\textcolor{DarkGreen}{The ``rest monster'' concept is due to a
+  suggestion by an insightful JEBO referee}
+
+\subsection*{Computational resources, supplementary material}
 

### 2020-10-01 inst/jebo_chess_revision2.tex
citations on mental fatigue

@@
 \textcolor{DarkGreen}{The Bradley-Terry techniques used here furnish
-  evidence that collusive games played during during Cura\c{c}ao 1962
-  were less detrimental to future performance than noncollusive games.
-  However, there was no cognitively realistic model for mental fatigue
-  (for example, downweighting heavy workloads long in the past) and
-  the criterion for restedness itself was quite coarse. not allowing
-  It would be possible in principle to investigate the effect of
-  recovery using similar reified entities.}
+  evidence that collusive games played during Cura\c{c}ao 1962 were
+  less detrimental to future performance than noncollusive games.
+  However, no cognitively realistic model for mental
+  fatigue~\citep{boksem2008,linden2003} was used.  There was no
+  allowance for recovery, for example; and the criterion for
+  restedness itself was quite coarse.  It would be possible in
+  principle to investigate such mechanisms using further reified
+  entities.}
 

### 2020-10-04 inst/curacao1962_threeplayers_rest_monster.Rmd
consistentify notation (also 'match' -> 'game')

--- 2020-09-30 inst/curacao1962_threeplayers_rest_monster.Rmd
@@
 
-cumulative_matches <- jj
-cumulative_collusive_matches <- jj
-cumulative_noncollusive_matches <- jj
+cumulative_games <- jj
+cumulative_collusive_games <- jj
+cumulative_noncollusive_games <- jj
 
-for(i in seq_len(28)){
+for(i in seq_len(28)){  # 'i' loops through rounds 
   for(j in seq_along(players)){
-    p <- players[j]
-    jj <- subset(a,a$round < i)
-    cm <- 0
-    ncm <- 0
-    for(k in seq_len(nrow(jj))){
+    previous_games <- subset(a,a$round < i)
+    cg   <- 0   # cumulative games
+    cncg <- 0   # cumulative noncollusive games
+    for(k in seq_len(nrow(previous_games))){
 
-      w <- jj$W[k]
-      b <- jj$B[k]
+      w <- previous_games$W[k]  # white player
+      b <- previous_games$B[k]  # black player
 
-      this_match <- c(w,b)
-      if((players[j] %in% c(w,b))){
+      this_game <- c(w,b)
+      if((players[j] %in% this_game)){
         if(players[j] %in% collusive_players){  # this player is collusive
           if((w %in% collusive_players) & (b %in% collusive_players)){  # both players collusive
-            cm <- cm+1
-          } else { # non-collusive match
-            ncm <- ncm+1
+            cg <- cg+1   # increment cm  == cumulative games
+          } else { # non-collusive game
+            cncg <- cncg+1 # increment ncm == cumulative noncollusive games
           }
         } else { # noncollusive player
-          ncm <- ncm+1
+          cncg <- cncg+1   # increment ncm == cumulative noncollusive games
         }
@@
     } # k loop closes
-    cumulative_collusive_matches[i,j] <- cm
-    cumulative_noncollusive_matches[i,j] <- ncm
-    cumulative_matches[i,j] <- cm + ncm
+    cumulative_collusive_games[i,j] <- cg
+    cumulative_noncollusive_games[i,j] <- cncg
+    cumulative_games[i,j] <- cg + cncg
   } # j loop closes
@@
 
-So we have three matrices `cumulative_matches`,
-`cumulative_collusive_matches` and `cumulative_noncollusive_matches`:
+So we have three matrices `cumulative_games`,
+`cumulative_collusive_games` and `cumulative_noncollusive_games`:
 
 ```{r}
-cumulative_matches
-cumulative_collusive_matches
-cumulative_noncollusive_matches
+cumulative_games
+cumulative_collusive_games
+cumulative_noncollusive_games
 ```
@@
 In the above three matrices, each row is a round, and the entries show
-the number of matches, collusive matches, and noncollusive matches
-played up to and including that round for each of the 8 players.  Each
-(named) column corresponds to a player.  Looking at the last row of
-`cumulative_noncollusive_matches` we see that in round 28, players
+the number of games, collusive games, and noncollusive games played up
+to and including that round for each of the 8 players.  Each (named)
+column corresponds to a player.  Looking at the last row of
+`cumulative_noncollusive_games` we see that in round 28, players
 Petrosian, Keres, and Geller had played a total of 18 noncollusive
-matches while the other players had played considerably more and would
+games while the other players had played considerably more and would
 be expected to be more fatigued at this point in the tournament.  Note
@@
 ```{r}
+##  To justify the footnote "No game in the tournament is played
+##  between two players with cumulative game count (collusivitity
+##  ignored), differ by more than one" in jebo_chess_revision2.tex,
+##  replace `cumulative_noncollusive_games` with
+##  `cumulative_games` in this chunk.
+##
+
 a <- cbind(a,cncmp_white=0,cncmp_black=0)
@@
   r <- a$round[i]
-  jj_w <- which(colnames(cumulative_noncollusive_matches)==a$W[i])
-  jj_b <- which(colnames(cumulative_noncollusive_matches)==a$B[i])
-  ncmp_w <- cumulative_noncollusive_matches[r, jj_w]
-  ncmp_b <- cumulative_noncollusive_matches[r, jj_b]
+  jj_w <- which(colnames(cumulative_noncollusive_games)==a$W[i])
+  jj_b <- which(colnames(cumulative_noncollusive_games)==a$B[i])
+  ncmp_w <- cumulative_noncollusive_games[r, jj_w]
+  ncmp_b <- cumulative_noncollusive_games[r, jj_b]
   a$cncmp_white[i] <- ncmp_w
@@
 28, ended in a draw, and by that time Petrosian had played 18
-noncollusive matches and Filip had played 26.
+noncollusive games and Filip had played 26.
 

### 2020-10-04 inst/jebo_chess_rebuttal.txt
more accounting for referee comments

@@
 introducing a "youth monster" who helps younger players.  But such
-analysis is, I would suggest, firmly in the category of further work.
-I have added a discussion of this issue to the manuscript (RED), in
-two parts: firstly when introducing the dataset, and secondly under
-Further Work.
+analysis is, I would suggest, firmly in the category of further work
+(although the revision does address a time-dependent component in the
+form of a "rest monster").  I have added a discussion of the
+identification issue to the manuscript (RED), in two parts: firstly
+when introducing the dataset, and secondly under Further Work.
 
@@
 
-The submission discusses applications to MasterChef
-Australia, consumer behaviour and mating preference, but I was not
-aware of tournament literature as a distinct academic field.  I have
-been studying Lazear and Rosen and pondering how reified monsters
-could help understanding tournaments in a workplace remuneration
-context.  Lazear and Rosen supply a good rationale to consider only
-rank in situations such as Formula 1 motor racing.
+The submission discusses applications to MasterChef Australia,
+consumer behaviour and mating preference, but I was not aware of
+tournament literature as a distinct academic field.  I have been
+studying Lazear and Rosen, and Szymanksi 2003, and pondering how
+reified monsters could help understanding tournaments in a workplace
+remuneration context.  Lazear and Rosen supply a good rationale to
+consider only rank in situations such as Formula 1 motor racing.
 

### 2020-10-04 inst/jebo_chess_revision2.tex
more accounting for referee comments

@@
 (Cura\c{c}ao 1962) exhibits evidence of collusion ($p<10^{-5}$), in
-agreement with previous work that uses additional information such as
-historical results including number of moves.
+agreement with previous work.  \textcolor{blue}{I also present
+  scoreline evidence for the effectiveness of such a drawing cartel:
+  noncollusive games are less detrimental to future play than
+  collusive games ($p<10^{-6}$).}
 \end{abstract}
@@
   related to incumbent advantage, and the techniques presented here
-  have been applied to ecological incumbency~\cite{hankin2010}.}
+  have been applied to ecological incumbency~\cite{hankin2010}.  More
+  generally, \cite{szymanski2003} considers sporting contests as
+  examples of competitive situations in which a contribution of some
+  kind [effort, investment, attendance of the elite, etc] is elicited
+  from contestants who may as a result win a prize.
+  \citeauthor{szymanski2003} points out that this paradigm arguably
+  covers executive corporate promotion, interviews for a position in
+  an office, and most sporting tournaments.  As such, the phenomenon
+  of collusion and indeed drawing in pairwise comparison is surely an
+  interesting phenomenon.}
 
@@
 methods to show that this model does not fit data well.  Here we note
-that such a model automatically give stronger players a higher draw
+that such a model automatically gives stronger players a higher draw
 potential, and also note that it is difficult to account for
@@
 \textcolor{DarkGreen}{
-  \caption{Results for won games in Cura\c{c}ao 1962  between
-      rested and unrested players    \label{restmonsterfisher}}
-  \centering
-  \fbox{%
+  \caption{Results for the 25 won games (out of 49 played) in
+    Cura\c{c}ao 1962 between rested and unrested
+    players \label{restmonsterfisher}} \centering \fbox{%
     \begin{tabular}{l|cc}\\
@@
   tournaments to knockout format events, searching for evidence of a
-  Soviet draw cartel.  By only considering international tournaments
-  [where Soviets have incentive to play differently against fellow
-    Soviets or Westerners] the analysis presented here does not allow
-  me to do this but the technique could easily be applied to a more
+  Soviet draw cartel.  The analysis presented here, only considering
+  international tournaments [where Soviets have incentive to play
+    differently against fellow Soviets or Westerners] does not admit
+  this phenomenon, but the technique could easily be applied to a more
   informative dataset.}
@@
 
-\section*{References}
-
 \bibliography{jebo_chess.bib}

### 2020-10-04 inst/jebo_chess_revision2.tex
more blue text

@@
 The present work details an approach to chess analysis based on
-reification of entities representing proclivity to draw, and the
-first-mover (or white player) advantage.  Likelihood functions for the
-players' and reified entities' strengths are defined using only
-specific tournament outcomes; further, the analysis uses only game
-outcome, not length.  Many hypotheses of interest to the chess
-community may be analysed straightforwardly.
+reification of entities representing proclivity to draw, first-mover
+(or white player) advantage, and \textcolor{blue}{the additional
+  playing strength conferred by previous collusive draws.  Likelihood
+  functions for the players' and reified entities' strengths are
+  defined using only specific tournament outcomes.}  Many hypotheses
+of interest to the chess community may be analysed straightforwardly.
 

### 2020-10-04 inst/jebo_chess_revision2.tex
minor rephrase

@@
   situations (tournament theory) and the analysis presented here might
-  be applicable to this situation, with the advantage of allowing for
+  be applicable to chess, with the advantage of allowing for
   both draws and incumbent advantage.}

### 2020-10-04 inst/jebo_chess_revision2.tex
change some text from blue to darkgreen to match the rebuttal

@@
 \begin{abstract}
+\definecolor{DarkGreen}{RGB}{14,159,4}
 In inference problems where the dataset comprises Bernoulli outcomes
@@
 (Cura\c{c}ao 1962) exhibits evidence of collusion ($p<10^{-5}$), in
-agreement with previous work.  \textcolor{blue}{I also present
+agreement with previous work.  \textcolor{DarkGreen}{I also present
   scoreline evidence for the effectiveness of such a drawing cartel:
@@
 \newcommand{\Prob}{\operatorname{Prob}}
+\definecolor{DarkGreen}{RGB}{14,159,4}
 
@@
 
-\definecolor{DarkGreen}{RGB}{14,159,4}
-
 \textcolor{DarkGreen}{It is reasonable to suppose that collusion of
@@
 reification of entities representing proclivity to draw, first-mover
-(or white player) advantage, and \textcolor{blue}{the additional
+(or white player) advantage, and \textcolor{DarkGreen}{the additional
   playing strength conferred by previous collusive draws.  Likelihood

### 2020-10-05 inst/jebo_chess_rebuttal.txt
rewording and mention McNemar's test

@@
 measure would be the Elo rating in which the winning player takes
-points from the loser based on the pre-match difference in Elo rating.
+points from the loser based on the pre-match Elo rating difference.
 Elo exhibits an emergent phenomenon in which sudden changes of ability
@@
 
-(comments addressing this comment in DARK GREEN)
-This is an excellent suggestion, one which I feel nicely rounds out
-the analysis.  I have included a "rest monster" whose strength aids
-players following a collusive draw.  Reified Bradley-Terry allows one
-not only to establish the existence of such a rest monster, but also
-to quantify its strength and I have done just this in the revision.
+(comments addressing this comment in DARK GREEN) This is an excellent
+suggestion, one which I feel nicely rounds out the analysis.  I have
+included a "rest monster" whose strength aids players following a
+collusive draw.  Reified Bradley-Terry allows one not only to
+establish the existence of such a rest monster, but also to quantify
+its strength and I have done just this in the revision, and compared
+the findings with a classical McNemar's test.

### 2020-10-08 inst/jebo_chess_revision2.tex
correction of pvalues and rephrasing

@@
   noncollusive games are less detrimental to future play than
-  collusive games ($p<10^{-6}$).}
+  collusive games ($p<10^{-5}$).}
 \end{abstract}
@@
   monster having zero strength is rejected with a likelihood ratio of
-  about $e^{11.9}$ or a $p$-value of about $10^{-6}$.  It is an
-  interesting observation that neither Table~\ref{restmonsterfisher},
-  nor reified Bradley-Terry, admit a White advantage (one-sided
-  McNemar test, $p=\frac{1}{8}$, likelihood ratio $1.609$
-  respectively).  }
+  about $e^{11.9}$ or a $p$-value of about $7\times 10^{-6}$.  It is
+  an interesting observation that neither
+  Table~\ref{restmonsterfisher}, nor reified Bradley-Terry, admit a
+  White advantage (one-sided McNemar test, $p=\frac{1}{8}$, likelihood
+  ratio $1.609$ respectively).  }
 
@@
 \textcolor{DarkGreen}{The ``rest monster'' concept is due to a
-  suggestion by an insightful JEBO referee}
+  suggestion by an insightful JEBO referee.}
 

### 2020-10-11 inst/jebo_chess_revision2.tex
title change suggested by Daniel Houser for increased visibility
remove colour coding for upload to JEBO

@@
 
-\title{A generalization of the Bradley-Terry model for draws in chess}
+\title{A generalization of the Bradley-Terry model for draws in chess with an application to collusion}
 
@@
 (Cura\c{c}ao 1962) exhibits evidence of collusion ($p<10^{-5}$), in
-agreement with previous work.  \textcolor{DarkGreen}{I also present
+agreement with previous work.  I also present
   scoreline evidence for the effectiveness of such a drawing cartel:
   noncollusive games are less detrimental to future play than
-  collusive games ($p<10^{-5}$).}
+  collusive games ($p<10^{-5}$).
 \end{abstract}
@@
 secondly, many chess games, about 39.4\%, are drawn (figures from
-Chessbase).  \textcolor{orange}{First-mover advantage is closely
+Chessbase).  First-mover advantage is closely
   related to incumbent advantage, and the techniques presented here
@@
   of collusion and indeed drawing in pairwise comparison is surely an
-  interesting phenomenon.}
+  interesting phenomenon.
 
@@
 
-\textcolor{blue}{The seminal work of \citet{moul2009} presents
+The seminal work of \citet{moul2009} presents
   sophisticated regression techniques to analyse draws and first-mover
@@
   to draw, using data from a single event, specifically the Stockholm
-  Interzonal tournament of (1962).}
+  Interzonal tournament of (1962).
 
@@
 and have played many games amongst themselves.  A summary may be found
-in Table~\ref{kka}, \textcolor{red}{showing games played from 1981 to
+in Table~\ref{kka}, showing games played from 1981 to
   2017.  During this time, the players' strengths might have changed
@@
   strengths as measures of ability {\em averaged over a playing
-    career}.  Either way,} Equation~\ref{likelihood_WD} becomes
+    career}.  Either way, Equation~\ref{likelihood_WD} becomes
 
@@
 
-\textcolor{blue}{\citet{moul2009} consider this accusation from a
+  \citet{moul2009} consider this accusation from a
   statistical perspective.  Their analysis incorporated a dataset of
@@
   ability of the players.  In contrast, the analysis presented here
-  uses only game outcomes (win-loss-draw) in the competition itself.}
+  uses only game outcomes (win-loss-draw) in the competition itself.
 
@@
 
-\textcolor{DarkGreen}{It is reasonable to suppose that collusion of
+It is reasonable to suppose that collusion of
   the type considered here is ``effective'' in the sense that
@@
   noncollusive games for the white and black player differs by more
-  than one\footnote{\textcolor{DarkGreen}{If collusivity is ignored,
+  than one\footnote{If collusivity is ignored,
     no game in the tournament is played between two players with
-    cumulative game count differ by more than one.}}, we posit a ``rest
+    cumulative game count differ by more than one.}, we posit a ``rest
   monster'' whose strength helps the player who has played fewer
-  noncollusive games.}
+  noncollusive games.
 
-\textcolor{DarkGreen}{
   Observing that the rest monster is
@@
   players having the same strength as their more tired opponents
-  (one-sided Fisher's exact test, $p\simeq 10^{-4}$).}
+  (one-sided Fisher's exact test, $p\simeq 10^{-4}$).
 
 \begin{table}
-\textcolor{DarkGreen}{
   \caption{Results for the 25 won games (out of 49 played) in
@@
   \end{tabular}}
-  }
 \end{table}
 
-\textcolor{DarkGreen}{The Bradley-Terry model together with a white
+The Bradley-Terry model together with a white
   player advantage, draw monster, and rest monster, estimates the
@@
   White advantage (one-sided McNemar test, $p=\frac{1}{8}$, likelihood
-  ratio $1.609$ respectively).  }
+  ratio $1.609$ respectively).
 
@@
 reification of entities representing proclivity to draw, first-mover
-(or white player) advantage, and \textcolor{DarkGreen}{the additional
+(or white player) advantage, and the additional
   playing strength conferred by previous collusive draws.  Likelihood
   functions for the players' and reified entities' strengths are
-  defined using only specific tournament outcomes.}  Many hypotheses
+  defined using only specific tournament outcomes.  Many hypotheses
 of interest to the chess community may be analysed straightforwardly.
@@
 
-\textcolor{red}{One strong assumption made in the
+One strong assumption made in the
   Kasparov-Karpov-Anand analysis was the constancy of the players'
@@
   smaller and less informative dataset; the tension between timeliness
-  and informativeness is a delicate and difficult matter.}
+  and informativeness is a delicate and difficult matter.
 
-\textcolor{orange}{In a wider context, many sports have a career arc
+In a wider context, many sports have a career arc
   whereby competitors start their careers weak, become stronger with
@@
   that account for such phenomena: we might have a ``mid career
-  monster'' who helps competitors at or near their career peak.}
+  monster'' who helps competitors at or near their career peak.
 
-\textcolor{orange}{The maximum likelihood estimate for strengths
+The maximum likelihood estimate for strengths
   furnishes a natural ranking for the competitors.  \cite{lazear1981}
@@
   be applicable to chess, with the advantage of allowing for
-  both draws and incumbent advantage.}
+  both draws and incumbent advantage.
 
@@
 admit such simulations and it might be interesting to compare results.
-\textcolor{blue}{\cite{moul2009} go on to compare round robin
+\cite{moul2009} go on to compare round robin
   tournaments to knockout format events, searching for evidence of a
@@
   this phenomenon, but the technique could easily be applied to a more
-  informative dataset.}
+  informative dataset.
 
-\textcolor{DarkGreen}{The Bradley-Terry techniques used here furnish
+The Bradley-Terry techniques used here furnish
   evidence that collusive games played during Cura\c{c}ao 1962 were
@@
   principle to investigate such mechanisms using further reified
-  entities.}
+  entities.
 
-\subsection*{\textcolor{DarkGreen}{Acknowledgement}}
+\subsection*{Acknowledgement}
 

### 2020-10-11 inst/jebo_chess_revision2.tex
further tiny changes for JEBO

@@
   related to incumbent advantage, and the techniques presented here
-  have been applied to ecological incumbency~\cite{hankin2010}.  More
+  have been applied to ecological incumbency~\citep{hankin2010}.  More
   generally, \cite{szymanski2003} considers sporting contests as
@@
   to draw, using data from a single event, specifically the Stockholm
-  Interzonal tournament of (1962).
+  Interzonal tournament of 1962.
 
@@
 more frequently than non-Soviets.  Indeed we could go further and plot
-a profile likelihood curve for the log-odds ratio
+a profile likelihood curve for
 $\log\left(\draw_s/\draw\right)$ which, if Dirichlet, would be
@@
 \makebox{\includegraphics{collusion_chess_draws-proflike.pdf}}
-\caption{Profile likelihood \label{proflike} for odds ratio
+\caption{Profile likelihood \label{proflike} for 
   $\log\left(\draw_s/\draw\right)$: note the accurately parabolic figure in
@@
 
-\textcolor{DarkGreen}{The ``rest monster'' concept is due to a
-  suggestion by an insightful JEBO referee.}
+The ``rest monster'' concept is due to a suggestion by an insightful
+JEBO referee.
 

### 2020-10-26 inst/ordertable_to_ranktable.Rmd
more work and rephrasing

@@
 
-Object `a` is a straightforward order table, just like
-`skating_table`; we see, for example, that both judges put competitor
-`a` first, but put `b` second and fifth respectively.  The judges
-second choice was `b` and `c` respectively.  We can coerce order table
-`OT` to a rank table `RT`:
+Object `OT` is a straightforward order table, just like
+`skating_table` but smaller and simpler.  We have two columns, one for
+each judge (named `J1` and`J2`), and five competitors, `a` through
+`e`.  The first row refers to competitor `a`; both judges put
+competitor `a` first.  The second row shows that `J1` put `b` second
+and `J2` put competitor `b` fifth.  If instead we ask who was placed
+third by each judge, this is not directly given; but by looking for
+`3` in the table we see that `J1` put `c` third and `J2` put `d`
+third.  To answer this and similar questions directly, we can coerce
+order table `OT` to a rank table `RT`:
 
@@
 
-Object `RT` gives the same information but in a different form.  Note
-that above we see the print method in action, which uses
-`ranktable_to_printable_object()`.  However, it is possible to view
-`RT` directly by removing class `ranktable`:
+Object `RT` gives the same information as `OT` but in a different
+form.  Note that above we see the print method in action, which uses
+`ranktable_to_printable_object()`.  Here the _rows_ correspond to the
+two judges.  Each row is a rank vector.  Column headings refer to
+placings (`c1` is supposed to be read "came first", `c2` "came second"
+etc).  Looking at the fifth column (`c5`) we see that `J1` and `J2`
+placed `e` and `b` fifth respectively.
+
+Note that it is possible to view object `RT` directly by removing
+class `ranktable`:
 
@@
 read the second row of `RT` is as follows.  We see the entries `1 3 4
-5 2` indicating that the first choice was competitor 1 [viz `a`, the
-second choice competitor 3 [viz `c`], the third choice competitor 4
-[viz `d`], fourth choice competitor 5 [`e`] and fifth choice
-competitor 2 [`b`].  So the rows are *ranks* (see the discussion in
-`rrank.Rd`).  But it is very tempting to read the second row of `RT`
-as "judge 2 placed competitor `b` in third place", which is wrong
-because this is interpreting the row as an *order*.  This is why the
-print method is there.
+5 2` indicating that the first choice was competitor 1 [viz `a`], the
+second choice was competitor 3 [viz `c`], the third choice was
+competitor 4 [viz `d`], fourth choice was competitor 5 [`e`] and fifth
+choice was competitor 2 [`b`].  So the rows are *ranks* (see the
+discussion in `rrank.Rd`).  But it is very tempting to read the second
+row of `RT` as "judge 2 placed competitor `b` in third place", which
+is wrong because this is interpreting the row as an *order*.  This is
+why the print method is there.

### 2021-01-15 inst/ordertrans.Rmd
ordertransplot() work

---
title: "The ordertrans() function"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("magrittr")
```

# Introduction

The `ordertrans()` function can be difficult to understand and this
short document provides some sensible use-cases.

## Hello world

The manpage provides a very simple example:

```{r,label=simpexample}
x <- c(d=2,a=3,b=1,c=4)
x
```

In the above, object `x` is a named vector with elements
\code{seq_along(x)} in some order.  Technically `x` is an order vector
because it answers the question "where did the first-named competitor
come?"  (see also the discussion at `rrank.Rd`).  However, it might
equally be a rank vector because it answers the question "who came
first? who came second?".

But note that `x` is not helpfully structured to answer either of
these questions: you have to go searching through the names or the
ranks respectively---both of which may appear in any order---to find
the competitor or rank you are interested in.  What we want is to
rearrange the elements of `x` so that the names are in alphabetical
order.  That way we can see straighaway where competitor `a` placed
(in this case, third).  This nontrivial task is accomplished by
function `ordertrans()`:

```{r,label=useordertrans}
o <- ordertrans(x,letters[1:4])
o
```

Observe that objects `x` and `o` are equal in the sense that they are
a rearrangement of one another:

```{r,label=equalrearrangement}
identical(x, o[c(4,1,2,3)])
identical(o, x[c(2,3,4,1)])
```

One consequence of this is that the resulting Plackett-Luce support
functions will be the same:

```{r,showordervec}
ordervec2supp(x)
ordervec2supp(o)
```

Note carefully that the two support functions above can be
mathematically identical but not formally `identical()` because
neither the order of the brackets, nor the order of terms within a
bracket, is defined.  They look the same on my system but YMMV.  It is
possible that the two support functions might appear to be different
even though they are mathematically the same. We can verify that the
two are mathematically identical using package idiom `==`:

```{r}
ordervec2supp(x) == ordervec2supp(o)
```

## Skating

Let us consider the skating dataset and use `ordertrans()` to study
it.

```{r,label=lookatskating}
skating_table
```

We might ask how judges 1 and 2 compare to one another?  We need to
create vectors like `x` and `y` above:

```{r,makej1j2}
j1 <- skating_table[,1]  # column 1 is judge number 1
names(j1) <- rownames(skating_table)
j2 <- skating_table[,2]  # column 2 is judge number 2
names(j2) <- rownames(skating_table)
j1
j2
cbind(j1,j2)
```

In the above, see how objects `j1` and `j2` have identical names, in
the same order.  Observe that `hughes` is ranked 1 by judge 1, and 4
by judge 2.  This makes it sensible to plot `j1` against `j2`:

```{r}
par(pty='s')  # forces plot to be square
plot(j1,j2,asp=1,pty='s',xlim=c(0,25),ylim=c(0,25),pch=16,xlab='judge 1',ylab='judge 2')
abline(0,1)  # diagonal line
for(i in seq_along(j1)){text(j1[i],j2[i],names(j1)[i],pos=4,col='gray',cex=0.7)}
```

In the above plot we see general agreement but differences in detail.
For example, `hughes` is ranked first by judge 1 and fourth by judge
2.  However, other problems are not so easy.  Suppose we wish to
compare the ranks according to likelihood with the ranks according to
some points system.

```{r,label=maxlikeskating,cache=TRUE}
mL <- maxp(skating)  # 'L' for Likelihood
mL
```

Note that in the above, the competitors' names are in alphabetical
order.  We first need to convert strengths to ranks:

```{r,label=strengthstoranks}
mL[] <- rank(-mL)  # minus because ranks orders from weak to strong
mL
```

(note that the names are in the same order as before, alphabetical).
In the above we see that `slutskya` ranks first, `hughes` second, and
so on.  Another way of ranking the skaters is to use a Borda-type
system: essentially the rowsums of the table (and, of course, the
_lowest_ score wins):

```{r,label=pointsskating,cache=TRUE}
mP <- rowSums(skating_table)  # 'P' for Points
mP[] <- rank(mP,ties='first') # positive sign here
mP
```

It is not at all obvious how to compare `mP` and `mL`.  For example,
we might be interested in `hegel`.  It takes some effort to find that
her likelihood rank is 20 and her Borda rank is 19.  Function
`ordertrans()` facilitates this:

```{r,label=ordertransexample}
ordertrans(mP,names(mL))  
```

See above how `ordertrans()` shows the points-based ranks but in
alphabetical order, to facilitate comparison with `mL`.  We can now
plot these against one another:

```{r,label=crapplot}
plot(mL,ordertrans(mP,mL))
```

However, the plot above is a bit crude.  Function `ordertransplot()`
gives a more visually pleasing output:

```{r,label=showoffordertransplot}
ordertransplot(mL,mP,xlab="likelihood rank",ylab="Borda rank")
```

So now we may compare judge 1 against likelihood:
```{r,transplotjudge1}
ordertransplot(mL,j1,xlab="likelihood rank",ylab="Judge 1 rank")
```

In the above, looking at the lower-left corner, we see (reading
horizontally) that the likelihood method placed Slutskya first, then
Hughes second, then Kwan third; while (reading vertically) judge 1
placed Hughes first, then Kwan, then Slutskya.

### 2021-01-15 R/hyper2.R

@@ -939,3 +939,12 @@ setGeneric("pnames<-",function(x,value){standardGeneric("pnames<-")})
     x[apply(outer(players,names(x),`==`),1,which)]
 }
 
+`ordertransplot` <- function(ox,oy,...){
+  oyp <- ordertrans(oy,names(ox))
+  par(pty='s') # square plot
+  jj <- c(0,max(c(ox,oy)))
+  plot(ox,oyp,asp=1,pty='s',xlim=jj,ylim=jj,pch=16,...)
+  abline(0,1)
+  for(i in seq_along(ox)){text(ox[i],oyp[i],names(ox)[i],pos=4,col='gray',cex=0.7)}
+}
+

### 2021-01-15 inst/moto.Rmd
bugfix

@@
 includes six different classes of no-score such as "Ret" for retired,
-"WD" for withdrawn, and so on.  I now use bespoke R function
+"WD" for withdrawn, and so on.  I now use package function
 \code{ordertable2supp()}, that translates data of this type into a
@@
 ```{r,applydata}
+head(motoGP_2019_table)
+```
+
+In the package, object `motoGP_2019` is the corresponding
+Plackett-Luce likelihood function; it is equivalent to
+`ordertable2supp(motoGP_2019_table[,1:19])`:
+
+```{r,showmotolikelihoodhead}
 head(motoGP_2019)
@@
 
-(object `motoGP_2019` is equivalent to
-`ordertable2supp(motoGP_2019_table)`.  Above, we see a small part of
-the Placket-Luce likelihood function for the 2019 results.  We can
-find the maximum likelihood estimator by using the `maxp()` function:
+Above, we see a small part of the Placket-Luce likelihood function for
+the 2019 results.  We can find the maximum likelihood estimator by
+using the `maxp()` function:
 
@@
 
-In the above, note the dominance of Marquez.  It is sometimes
-preferable to plot the estimated strengths on a log scale:
+In the two figures above, note the dominance of Marquez.  It is
+sometimes preferable to plot the estimated strengths on a log scale:
 
@@
 
-(in the diagram above, we remove the bottom-ranked player, Lecuona,
-whose estimated strength is zero).
+(in the diagram above, observe thatthe bottom-ranked player, Lecuona,
+has a ML strength of zero but the estimate is slightly positive for
+numerical reasons).
 

### 2021-01-16 inst/ordertrans.Rmd
typos

@@
 In the above, object `x` is a named vector with elements
-\code{seq_along(x)} in some order.  Technically `x` is an order vector
+`seq_along(x)` in some order.  Technically `x` is an order vector
 because it answers the question "where did the first-named competitor
@@
 So now we may compare judge 1 against likelihood:
+
 ```{r,transplotjudge1}

### 2021-01-17 inst/ordertrans.Rmd
slight clarification

@@
 In the above, object `x` is a named vector with elements
-`seq_along(x)` in some order.  Technically `x` is an order vector
-because it answers the question "where did the first-named competitor
-come?"  (see also the discussion at `rrank.Rd`).  However, it might
-equally be a rank vector because it answers the question "who came
-first? who came second?".
+`seq_along(x)` in some order.  It means that competitor `d` came
+second, competitor `a` came third, `b` came first and `c` came fourth.
+Technically `x` is an order vector because it answers the question
+"where did a particular competitor come?"  However, it might equally
+be a rank vector because it answers the question "who came first? who
+came second?"  (see also the discussion at `rrank.Rd`).
 
@@
 ```{r,showordervec}
-ordervec2supp(x)
-ordervec2supp(o)
+(Sx <- ordervec2supp(x))
+(So <- ordervec2supp(o))
 ```
@@
 ```{r}
-ordervec2supp(x) == ordervec2supp(o)
+Sx==So
 ```

### 2021-01-18 inst/skating_analysis.Rmd
get rid of that stupid 'misslast' argument
use skating_analysis.Rmd to create data/skating.rda

@@
 `skating_table` is copied from Lock and Lock (see skating.Rd for more
-details).
+details).  It also creates file `data/skating.rda` which contains R objects `skating_table`,
 
@@
 set.seed(0)
-data("skating")
-skating_table
+skating_table <- read.table("skating.txt",header=TRUE)
 ```
@@
 ```{r,convertordertosupport}
-head(ordertable2supp(skating_table,misslast=FALSE))
+head(ordertable2supp(skating_table))
 ```
@@
 ```{r,maxshow, cache = TRUE}
-(m <- maxp(skating))
+skating <- ordertable2supp(skating_table)
+skating_maxp <- maxp(skating)
+m <- skating_maxp # for ease of  typing
 ```
@@
 ```{r}
-o <- ordertable2supp(o,misslast=FALSE)
+o <- ordertable2supp(o)
 o
@@
 ```
+
+### Package dataset
+
+Following lines create `skating.rda`, residing in the `data/` directory of the package.
+
+```{r,label=saveskatingdataset}
+save(skating_table,skating,skating_maxp,file="skating.rda")
+```

### 2021-01-18 inst/skating_analysis.Rmd
consistentification of skating dataset

@@
 
- maximum likelihood estimated strengths furnish an ordering for the competitors.
-We can compare ranking by strengths with the official point-tallying method:
+maximum likelihood estimated strengths furnish an ordering for the
+competitors.  We can compare ranking by strengths with the official
+point-tallying method:
 
 ```{r,dpi=72}
-par(pty='s')
-lik <- rank(-m)
-plot(1:23,lik,asp=1,pch=16,xlab='official order',ylab='likelihood order')
-text(1:23,lik,asp=1,pch=16,pos=c(rep(4,19),rep(2,4)),rownames(skating_table))
-abline(0,1)
+ rL <- sort(skating_maxp,decreasing=TRUE)
+ rL[] <- seq_along(rL)
+ rO <- seq_len(nrow(skating_table))
+ names(rO) <- rownames(skating_table)
+ ordertransplot(rO,rL,xlab="offical rank",ylab="likelihood rank",main="Ladies free skating, 2002 Winter Olympics")
 ```

### 2021-01-19 inst/rock_paper_scissors_monster.Rmd
new vignette

---
title: "Quantifying the rock-paper-scissors effect"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("magrittr")
abbreviated <- TRUE  # change to FALSE for full names
```

## Abstract

Here I quantify the strength of a circular rock-paper-scissors effect
using the `hyper2` package, improving on previous results which merely
established its existence.

## Rock Paper Scissors

Consider the following dataset:

```{r,chesstable}
o <- matrix(c(22,NA,8,13,23,NA,NA,12,10),3,3)
dimnames(o) <- list(
    match=c("T_vs_A","A_vs_K","T_vs_K"),
    player=c("Topalov","Anand","Karpov"))
o
```

This is documented in `chess.Rd` and shows that Topalov played Anand
$22+13=35$ times, winning 22 and losing 13 matches.  The corresponding
likelihood function is `chess`:

```{r,showchess}
chess
maxp(chess)
```

(we refer to this as the "TAK" model).  We might ask whether there is
evidence for the players having differing strengths:

```{r}
equalp.test(chess)
```

Thus we fail to reject $H_0\colon p_T=p_A=p_K$ under T-A-K as the
p-value is about 0.39.

## RPS in chess

The existence of a rock-paper-scissors relationship between the
players was discussed in Hankin (2010) and the null of no RPS
phenomena could be rejected using the Alymer test.

Here we improve on that work by introducing a monster, `M`, that
accounts for the rock-paper-scissors nature of the players.  The RPS
monster helps players that benefit from the RPS nature of the result.
The simplest nontrivial probability model is that all three real
players have the same strength: $p_T=p_A=p_K$, which we write $T=A=K$.
Under $T=A=K$, writing $p_T=p_A=p_K=\alpha$ we can use a binomial test
to calculate the p-value for the hypothesis that $p_M=0$:

```{r,binomtest}
binom.test(sum(diag(o)),sum(o,na.rm=TRUE),alternative="greater")
```

Estimating $p_M$ is also straightforward:

$$
Prob(\mathrm{procyclic}\ \mathrm{win})=\frac{\alpha+M}{2\alpha+M}
$$

(a "procyclic win" is a win that would be expected under a strong
monster).  Here we also have $3\alpha+M=0$, so to estimate the
strength of the monster we would solve

$$
\frac{\alpha+M}{2\alpha+M}=p,\qquad{3\alpha + M}=1
$$

for $\alpha$ and $M$, where $p$ is the probability of a procyclic win.
This gives us

$$
\alpha=\frac{1-p}{2-p},\qquad M=\frac{2p-1}{2-p}.
$$

If we use the maximum likelihood estimate value for a procyclic win of
$\frac{55}{88}=0.625$ we find that $\hat{M}=0.18$ approximately.
Further, we may express $p$, the probability of a pro-cyclic win, as
$p=\frac{1+2M}{2+M}$.  If we observe $a$ procyclic wins and $b$
countercyclic wins a likelihood function for $M$ would then be

$$
\mathcal{L}(M)= p^a(1-p)^b=\left(\frac{1+2M}{2+M}\right)^a\cdot
\left(\frac{1-M}{2+M}\right)^b
$$

(a graph is given below).  We may relax the assumption of equal
strength of the real players by using a more sophisticated hyper2
construction:

```{r,setupH}
H <- hyper2()
H[c("Topalov",                 "M")] %<>% inc(22   )
H[c(          "Anand"             )] %<>% inc(   13)
H[c("Topalov","Anand",         "M")] %<>% dec(22+13)

H[c(          "Anand",         "M")] %<>% inc(   23)
H[c(                  "Karpov"    )] %<>% inc(   12)
H[c(          "Anand","Karpov","M")] %<>% dec(23+12)

H[c("Topalov"                     )] %<>% inc( 8   )
H[c(                  "Karpov","M")] %<>% inc(   10)
H[c("Topalov",        "Karpov","M")] %<>% dec( 8+10)
H
```

We can call this the TAKM model, and find the evaluate:

```{r}
maxp(H)
```

We see that, under TAKM, the monster has strength of about 0.16,
slightly less than the 0.18 obtained for $M$ under the equality
assumption $p_T=p_A=p_K$.  Further, under TAKM, the three "real"
players are again all of approximately equal strength, and we may test
this:

```{r,testequalTAKM}
samep.test(H,c("Topalov","Anand","Karpov"))
```

(note the estimated monster strengths of 0.16 and 0.18 agree with the
previous estimate).  Thus the p-value of about 0.77 indicates failure
to reject the null (compare 0.39 under TAK).  Now we can test
$H_0\colon p_M=0$ under TAKM:

```{r}
specificp.gt.test(H,"M",0)
```

(in the test above, note that the players may appear in a different
order from previously).  The test shows that we may reject the
hypothesis that $p_M=0$ under TAKM with a p-value of $0.04$ (compare
0.012 from the binomial test).

### Profile likelihood

We may plot a profile likelihood curve for $M$:

```{r}
M <- seq(from=0.001,to=0.4,len=40)
L1 <- profsupp(H,"M",M)
L1 <- L1-max(L1)

L2 <- 55*log(1+2*M)+33*log(1-M)-88*log(2+M)
L2 <- L2-max(L2)

plot(M,L1,type="l",col="black",ylab="log likelihood",xlab="Monster strength")
points(M,L2,type="l",col="red")
abline(h=c(0,-2),lty=2)
legend("topright",col=c("black","red"),lty=1,legend=c("TAKM","T=A=K"))
```

Above we see support curves for RPS monster strength using TAKM and
$T=A=K$, showing very similar results.

### 2021-01-20 inst/rock_paper_scissors_monster.Rmd
generalized RPS

@@
 
-Here I quantify the strength of a circular rock-paper-scissors effect
-using the `hyper2` package, improving on previous results which merely
-established its existence.
+Here I use the `hyper2` package (Hankin 2017) to quantify the strength
+of a circular rock-paper-scissors effect using the `hyper2` package,
+improving on previous results (West and Hankin 2008) which merely
+established its existence.  I use two datasets, a three-player
+scoreline and a synthetic dataset referring to wear on a seven-tooth
+gear wheel.
 
@@
 
-We can call this the TAKM model, and find the evaluate:
+(in the above, a win by a player helped by the RPS effect has `M`
+added to his strength).  We can call this the TAKM model, and find the
+evaluate:
 
@@
 $T=A=K$, showing very similar results.
+
+# Generalized RPS
+
+West and Hankin (2008) consider a synthetic dataset referring to seven
+teeth of a gear wheel.  Consecutive pairs are repeatedly compared and
+the most worn tooth is nominated by independent judges.  Here a
+dataset of the same form will be used, but the entries have been
+changed for pedagogical reasons:
+
+```{r,echo=FALSE}
+t1 <- "t1"
+t2 <- "t2"
+t3 <- "t3"
+t4 <- "t4"
+t5 <- "t5"
+t6 <- "t6"
+t7 <- "t7"
+M <- "M"
+```
+
+```{r,label=setuphw}
+hw <- matrix(c(
+    16,05,NA,NA,NA,NA,NA,
+    NA,06,09,NA,NA,NA,NA,
+    NA,NA,05,11,NA,NA,NA,
+    NA,NA,NA,06,12,NA,NA,
+    NA,NA,NA,NA,05,09,NA,
+    NA,NA,NA,NA,NA,04,07,
+    13,NA,NA,NA,NA,NA,05),
+    7,7,byrow=TRUE)
+colnames(hw) <- paste("t",1:7,sep="")
+hw
+```
+
+Thus, considering the top row, we see that in a total of $16+5=21$
+independent comparisons between `t1` and `t2`, tooth `t1` was judged
+to be the more worn one 16 times, and `t2` was judged to be the more
+worn one $5$ times (NB: in the above, I have changed the dataset in
+order to make `t1` stronger than the other players while maintaining
+the RPS phenomenon).  We may convert this dataset to a generalized
+`hyper2` object.  First a standard Bradley-Terry:
+
+```{r,label=HWnomonster}
+HW <- hyper2()
+rs <- -rowSums(hw,na.rm=TRUE)
+cs <- colSums(hw,na.rm=TRUE)
+HW[t1] <- cs[1]
+HW[t2] <- cs[2]
+HW[t3] <- cs[3]
+HW[t4] <- cs[4]
+HW[t5] <- cs[5]
+HW[t6] <- cs[6]
+HW[t7] <- cs[7]
+HW[c(t1,t2)] <- rs[1]
+HW[c(t2,t3)] <- rs[2]
+HW[c(t3,t4)] <- rs[3]
+HW[c(t4,t5)] <- rs[4]
+HW[c(t5,t6)] <- rs[5]
+HW[c(t6,t7)] <- rs[6]
+HW[c(t7,t1)] <- rs[7]
+HW
+```
+
+(likelihood function `HW` has no monster).  We can then analyse this:
+
+```{r,label=testHW}
+maxp(HW)
+equalp.test(HW)
+```
+
+Thus according to this model we may reject the hypothesis of equal
+strength.  We now incorporate the monster:
+
+```{r,label=NWwithmonster}
+HWM <- hyper2()
+tt <- c("t1","t2","t3","t4","t5","t6","t7")
+for(i in 1:6){
+  HWM[c(tt[i]          )] <-  hw[i,i  ]
+  HWM[c(      tt[i+1],M)] <-  hw[i,i+1]
+  HWM[c(tt[i],tt[i+1],M)] <- rs[i]
+}
+HWM[c(tt[7]        )] <-  hw[7,7]
+HWM[c(      tt[1],M)] <-  hw[7,1]
+HWM[c(tt[7],tt[1],M)] <- rs[7]
+HWM
+```
+
+Then the same analysis applies:
+
+```{r,label=maxpHWM}
+maxp(HWM)
+```
+
+and we can test the hypothesis that $p_M=0$:
+
+```{r}
+specificp.gt.test(HWM,"M",0)
+```
+
+suggesting that we may reject the hypothesis of zero monster strength
+with a p-value of about 4.7\%, even after accounting for the different
+strengths of the teeth.
+
+## References
+
+* L. J. West and R. K. S. Hankin 2008.  "Exact tests for two-way
+  contingency tables with structural zeros".  _Journal of Statistical
+  Software_ 2008, volume 28 issue 11.
+* R. K. S. Hankin 2010.  "A generalization of the Dirichlet
+  distribution".  _Journal of Statistical Software_ 2010, volume 33
+  issue 11.
+* R. K. S. Hankin 2017.  "Partial rank data with the `hyper2` package:
+  likelihood functions for generalized Bradley-Terry models".  _The R
+  Journal_, volume 9, number 2.

### 2021-01-21 inst/curling.Rmd
changes to document curling dataset properly with an Rmd file

@@
 first, China did not compete [`NA` meaning did not compete], Denmark
-came second, and so on.
-
-First we will interpret `NA` in the same way as `DNS` in the context
-of formula 1 motor racing.  We need to define a new data frame, `b`:
+came second, and so on.  First we will interpret `NA` in the same way
+as `DNS` in the context of formula 1 motor racing.  We need to define
+a new data frame, `b`:
 
@@
 ```{r,definsupportb,cache=TRUE}
-H1 <- ordertable2supp(ct)
-H1
-mH1 <- maxp(H1)
-pie(mH1)
+curling1 <- ordertable2supp(ct)
+head(curling1)
+curling1_maxp <- maxp(curling1)
+pie(curling1_maxp)
 ```
@@
 ```{r,makehyper2curling}
-H2 <- hyper2(pnames=rownames(curling_table))
+curling2 <- hyper2()
 
@@
     names(x) <- rownames(curling_table)
-    H2 %<>% order_obs(x[!is.na(x)])
+    curling2 <- curling2 + ordervec2supp(x[!is.na(x)])
 }
-head(H2)
-(mH2 <- maxp(H2))
-pie(mH2)
+head(curling2)
+(curling2_maxp <- maxp(curling2))
+pie(curling2_maxp)
 ```
@@
 par(pty="s")
-plot(log10(mH1),log10(mH2),asp=1,pch=16,xlim=c(-2.6,0),ylim=c(-2.6,0))
+plot(log10(curling1_maxp),log10(curling2_maxp),asp=1,pch=16,xlim=c(-2.6,0),ylim=c(-2.6,0))
 posn <- c(Canada=2, China=4, Denmark=4, Germany=4, GB=4, Italy=1, Japan=2, Norway=2,
@@
 for(i in seq_len(nrow(curling_table))){
-	text(log10(mH1)[i],log10(mH2)[i],rownames(curling_table)[i],col="gray",pos=posn[i])
+	text(log10(curling1_maxp)[i],log10(curling2_maxp)[i],rownames(curling_table)[i],col="gray",pos=posn[i])
 }
@@
 The two likelihood systems have very different evaluates.
+
+### Package dataset
+
+Following lines create `curling.rda`, residing in the `data/` directory of the package.
+
+```{r,label=savecurlingdataset}
+save(curling_table,curling1,curling2,curling1_maxp,curling2_maxp,file="curling.rda")
+```

### 2021-01-21 inst/curling.Rmd
tidy URLs (which were wrong)

@@
 
-## Curling
-
-Curling is a sport with stones and brooms.  Following dataset taken
-from [wikipedia](https://www.google.com "Google's Homepage"), women's
-tournament:
+[Curling](https://en.wikipedia.org/wiki/Curling) is a sport with
+stones and brooms.  Following dataset taken from
+[wikipedia](https://en.wikipedia.org/w/index.php?title=Curling_at_the_Winter_Olympics&oldid=998891075),
+women's tournament:
 
@@
 
+### References
+
+Curling at the Winter Olympics, https://en.wikipedia.org/w/index.php?title=Curling_at_the_Winter_Olympics&oldid=998891075 (last visited Jan. 21, 2021).
+
 ### Package dataset

### 2021-01-21 inst/karate.Rmd
consistentify karate

---
title: "karate"
output: html_document
---

```{r setup, include=FALSE}
set.seed(0)
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("magrittr")
options("digits" = 5)
```

Here I analyse a [karate](https://en.wikipedia.org/wiki/Karate)
dataset taken from
[https://en.wikipedia.org/wiki/2018_World_Karate_Championships].

```{r readkaratetable}
karate_table  <- read.table("karate.txt",header=TRUE)
head(karate_table)
```

Converting it to a likelihood function is straightforward:

```{r converttolikelihood}
karate <- hyper2()
for(i in seq_len(nrow(karate_table))){

  p1 <- karate_table$karateka1[i]
  p2 <- karate_table$karateka2[i]

  w1 <- karate_table$wins1[i]
  w2 <- karate_table$wins2[i]

  karate[p1] %<>% inc(w1)
  karate[p2] %<>% inc(w2)
  karate[c(p1,p2)] %<>% dec(w1+w2)
}
karate
karate_maxp <- maxp(karate)
karate_maxp
```

### Package dataset

Following lines create `karate.rda`, residing in the `data/` directory
of the package.

```{r,label=savekaratedataset}
save(karate_table,karate,karate_maxp,file="karate.rda")
```

### 2021-01-21 inst/karate_analysis.R

@@ -1,34 +1,0 @@
-## Analysis of the 2018 World Karate Championships, men's 67kg.
-
-## Wikipedia contributors. (2020, February 2). 2018 World Karate
-## Championships – Men's 67 kg. In Wikipedia, The Free
-## Encyclopedia. Retrieved 00:21, February 24, 2020, from
-## https://en.wikipedia.org/w/index.php?title=2018_World_Karate_Championships_%E2%80%93_Men%27s_67_kg&oldid=938728758
-
-
-
-set.seed(0)
-library("hyper2")
-library("magrittr")
-
-a <- read.table("karate.txt",colClasses = "character",header=TRUE)
-a$wins1 <-  as.numeric(a$wins1)
-a$wins2 <-  as.numeric(a$wins2)
-
-H <- hyper2(pnames=sort(unique(c(a$karateka1,a$karateka2))))
-for(i in seq_len(nrow(a))){
-
-  p1 <- a$karateka1[i]
-  p2 <- a$karateka2[i]
-
-  w1 <- a$wins1[i]
-  w2 <- a$wins2[i]
-
-  
-  H[p1] %<>% inc(w1)
-  H[p2] %<>% inc(w2)
-  H[c(p1,p2)] %<>% dec(w1+w2)
-}
-
-
-

### 2021-01-21 inst/T20.Rmd
consistentify T20 dataset

@@
 library("hyper2",quietly=TRUE)
+T20_table <- read.table("T20.txt",header=TRUE)
 head(T20_table)
@@
 
-We may reject homogoneity of proportion (Fisher's exact test, $p= `r round(fisher.test(M)$p.value,3)`$).  Recalling that a slight majority
-of toss winners elect to field first (361 out of 633), we see that this is a consistent result.
+We may reject homogoneity of proportion (Fisher's exact test, $p= `r
+round(fisher.test(M)$p.value,3)`$).  Recalling that a slight majority
+of toss winners elect to field first (361 out of 633), we see that
+this is a consistent result.
 
@@
 attach(lapply(T20_table[, ], as.vector))  # vector access to team1, etc
-T20 <- hyper2(pnames=c(levels(T20_table$team1)))
+T20 <- hyper2()
 for(i in seq_along(team1)){
@@
 ```{r}
-dotchart(maxp(T20))
+T20_maxp <- maxp(T20)
+dotchart(T20_maxp)
 ```
@@
 ```{r}
-dotchart(maxp(T20_toss))
+T20_maxp_toss <- maxp(T20_toss)
+dotchart(T20_maxp_toss)
 ```
@@
 Above, note the small estimated strength of the toss.
+
+### Package dataset
+
+The following lines create `T20.rda`, residing in the `data/`
+directory of the package.
+
+```{r,label=saveT20dataset}
+save(T20_table,T20,T20_maxp,T20_toss,file="T20.rda")
+```

### 2021-01-21 data/T20.R

@@ -1,359 +1,0 @@
-`T20_table` <-
-structure(list(team1 = structure(c(13L, 8L, 4L, 11L, 12L, 4L, 
-7L, 12L, 3L, 13L, 5L, 12L, 11L, 7L, 3L, 4L, 11L, 3L, 13L, 12L, 
-13L, 5L, 7L, 8L, 13L, 5L, 7L, 11L, 11L, 12L, 3L, 13L, 11L, 4L, 
-3L, 13L, 12L, 4L, 13L, 7L, 4L, 5L, 11L, 8L, 12L, 5L, 8L, 5L, 
-4L, 5L, 3L, 4L, 8L, 5L, 12L, 11L, 13L, 7L, 8L, 7L, 1L, 10L, 8L, 
-2L, 5L, 2L, 1L, 2L, 5L, 12L, 7L, 8L, 3L, 1L, 7L, 3L, 2L, 10L, 
-1L, 2L, 5L, 8L, 1L, 12L, 1L, 10L, 3L, 7L, 2L, 12L, 1L, 7L, 3L, 
-12L, 7L, 1L, 5L, 3L, 7L, 3L, 10L, 8L, 7L, 12L, 7L, 5L, 12L, 2L, 
-8L, 10L, 12L, 5L, 8L, 2L, 10L, 5L, 1L, 8L, 12L, 5L, 7L, 1L, 5L, 
-2L, 3L, 10L, 12L, 2L, 12L, 5L, 1L, 8L, 3L, 7L, 5L, 2L, 1L, 8L, 
-12L, 2L, 1L, 7L, 8L, 1L, 10L, 7L, 2L, 12L, 1L, 8L, 2L, 10L, 8L, 
-7L, 2L, 7L, 5L, 3L, 1L, 10L, 3L, 8L, 7L, 5L, 3L, 1L, 3L, 10L, 
-1L, 8L, 12L, 3L, 1L, 2L, 7L, 8L, 5L, 12L, 2L, 10L, 5L, 1L, 8L, 
-10L, 3L, 2L, 10L, 8L, 2L, 5L, 7L, 12L, 10L, 1L, 2L, 7L, 3L, 10L, 
-8L, 3L, 5L, 12L, 3L, 7L, 5L, 1L, 8L, 7L, 3L, 10L, 1L, 5L, 7L, 
[…]

### 2021-01-21 inst/icons.Rmd
consistentify the icons dataset and analysis

---
title: "icons"
output: html_document
---

```{r setup, include=FALSE}
set.seed(0)
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("magrittr")
options("digits" = 5)
```

This minimal document reads file `icons.txt` and creates R objects
`icons`, `icons_matrix`, and `icons_maxp`.  The dataset is discussed
extensively in West and Hankin 2008, and Hankin 2010.

```{r readiconstable}
icons_table  <- as.matrix(read.table("icons.txt",header=TRUE))
icons_table
```

Thus the first row shows that a total of $5+3+4+3=15$ respondents were
shown icons `NB`, `L`, `THC`, and `WAIS`; 5 of them chose `NB` as
"most concerning", 3 chose `L`, 4 chose `THC` and 3 chose `WAIS`.

We can use function `saffy()` to convert the table into a `hyper2`
object:

```{r createhyper2}
icons <- saffy(icons_table)
icons
icons_maxp <- maxp(icons)
icons_maxp
```

This file exists for consistency; an extended discussion (and
statistical tests) are presented in the vignette `icons`.

### Package dataset

Following lines create `icons.rda`, residing in the `data/` directory of the package.

```{r,label=saveiconsdataset}
save(icons_table,icons,icons_maxp,file="icons.rda")
```

### 2021-01-21 inst/NBA.Rmd
consistentification

---
title: "Basketball: an analysis of a single game"
output: html_document
---

```{r setup, include=FALSE}
set.seed(0)
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("magrittr")
options("digits" = 5)
```

File `NBA.txt` contains a point-by-point analysis of a basketball
match between the Cleveland Cavaliers and the Golden State Warriors on
13 June 2017, data taken from
[here](https://www.espn.com/nba/playbyplay?gameId=400954514).

```{r readNBAtable}
NBA_table  <- read.csv("NBA.txt",header=TRUE)
head(NBA_table)
```

Each row corresponds to a point scored.  The first column is the time
of the score, the second is the number of points scored, the third
shows which team had possession at the start of play, and the fourth
shows which team scored.  The other columns show the players.  Table
entries show whether or not that particular player was on the pitch
when the point was scored.  Note that the two "ghost" players
represent the effect of having possession.

First some basics:

```{r,fisherbasic}
(TT <- table(NBA_table[,3:4]))
fisher.test(TT,alternative="greater")
```

Thus the identity of the scoring team is not independent of the
identity of the possessing team (as one might expect).  
We can convert `NBA_table` to a likelihood function:

```{r,definsupportb,cache=TRUE}
allplayers <- as.matrix(NBA_table[,5:23])
NBA <- hyper2()

## players 1-9 Cleveland
## players 10-19 (sic) Warriors:
C_onpitch <- allplayers[,1:9]
W_onpitch <- allplayers[,10:19]

possession <- NBA_table[,3]
scored <-    NBA_table[,4]

for(i in seq_len(nrow(NBA_table))){
  if(scored[i] == 'W'){  # Warriors score a point
    onpitch_scored  <- names(which(W_onpitch[i,]))
    onpitch_noscore <- names(which(C_onpitch[i,]))
  } else if(scored[i] == "C"){ # Cleveland score a point
    onpitch_scored  <- names(which(C_onpitch[i,]))
    onpitch_noscore <- names(which(W_onpitch[i,]))
  } else {
    stop("scorint team must be either W or C")
  }
  if(possession[i] == 'W'){
    onpitch_scored <- c(onpitch_scored,"W_possession")
  } else if (possession[i] == 'C'){
    onpitch_scored <- c(onpitch_scored,"C_possession")
  } else {
    stop("possession must be either W or C")
  }
  NBA[onpitch_scored] %<>% inc()
  NBA[c(onpitch_scored,onpitch_noscore)] %<>% dec()
}

NBA_maxp <- maxp(NBA)
NBA_maxp %<>% ordertrans(c(colnames(NBA_table)[-(1:4)],"C_possession","W_possession"))
dotchart(NBA_maxp,col=c(rep("black",9),rep("red",10),rep("blue",2)),pch=16)
```

# Statistical analysis

```{r,statsanal,cache=TRUE}
samep.test(NBA,c("C_possession","W_possession"))
```

We may thus assume that Cleveland having possession is the same
strength as the Warriors having possession.

Now, what is the probability of C scoring both with and without
possession?  We may condition on the estimated player and possession
strengths, and assume that only the 5 strongest players on each team
play.

```{r}
C_allstrengths <- sort(NBA_maxp[01:9],decreasing=TRUE)
W_allstrengths <- sort(NBA_maxp[10:19],decreasing=TRUE)
C_strongestteam <- sum(C_allstrengths[1:5])
W_strongestteam <- sum(W_allstrengths[1:5])
possession <- 0.2155786 # taken from samep.test() above

# probability of C scoring without possession:
(C_strongestteam           )/(C_strongestteam + W_strongestteam + possession)

# probability of C scoring with possession:
(C_strongestteam+possession)/(C_strongestteam + W_strongestteam + possession)
```

### Package dataset

Following lines create `NBA.rda`, residing in the `data/` directory of the package.

```{r,label=saveNBAdataset}
save(NBA_table,NBA,NBA_maxp,file="NBA.rda")
```

### 2021-01-21 data/NBA.R

@@ -1,349 +1,0 @@
-`NBA_likelihood` <-
-structure(list(brackets = list(c("Barnes", "Curry", "Iguodala", 
-"Irving", "Korver", "Livingston", "McCaw", "Shumpert", "Smith", 
-"W_possession", "Williams"), c("Curry", "Durant", "Green", "Iguodala", 
-"Irving", "James", "Jefferson", "Livingston", "Smith", "Thompson", 
-"W_possession"), c("Curry", "Durant", "Green", "Iguodala", "Irving", 
-"James", "Jefferson", "Smith", "Thompson", "Thompson.1", "W_possession"
-), c("Curry", "Durant", "Green", "Iguodala", "Irving", "James", 
-"Jefferson", "Smith", "Thompson", "W_possession", "West"), c("Curry", 
-"Durant", "Green", "Iguodala", "Irving", "James", "Korver", "Love", 
-"Smith", "Thompson.1", "W_possession"), c("Curry", "Durant", 
-"Green", "Iguodala", "Irving", "James", "Love", "Smith", "Thompson", 
-"Thompson.1", "W_possession"), c("Curry", "Durant", "Green", 
-"Iguodala", "Irving", "James", "Love", "Smith", "Thompson", "W_possession", 
-"West"), c("Curry", "Durant", "Green", "Irving", "James", "Jefferson", 
-"Pachiuila", "Smith", "Thompson", "Thompson.1", "W_possession"
-), c("Curry", "Durant", "Green", "Irving", "James", "Love", "Pachiuila", 
-"Smith", "Thompson", "Thompson.1", "W_possession"), c("Curry", 
[…]

### 2021-01-22 inst/NBA.Rmd
corrections

@@
 NBA <- hyper2()
+NBA2 <- hyper2()
 
@@
   }
+  print(onpitch_scored)
+  print(onpitch_noscore)
   if(possession[i] == 'W'){
-    onpitch_scored <- c(onpitch_scored,"W_possession")
+    onpitch_scored1 <- c(onpitch_scored,"possession")
+    onpitch_scored2 <- c(onpitch_scored,"W_possession")
   } else if (possession[i] == 'C'){
-    onpitch_scored <- c(onpitch_scored,"C_possession")
+    onpitch_scored1 <- c(onpitch_scored,"possession")
+    onpitch_scored2 <- c(onpitch_scored,"C_possession")
   } else {
@@
   }
-  NBA[onpitch_scored] %<>% inc()
-  NBA[c(onpitch_scored,onpitch_noscore)] %<>% dec()
+
+  NBA[onpitch_scored1] %<>% inc()
+  NBA[c(onpitch_scored1,onpitch_noscore)] %<>% dec()
+
+  NBA2[onpitch_scored2] %<>% inc()
+  NBA2[c(onpitch_scored,onpitch_noscore)] %<>% dec()
 }
 
+NBA2_maxp <- maxp(NBA2)
+NBA2_maxp %<>% ordertrans(c(colnames(NBA_table)[-(1:4)],"C_possession","W_possession"))
+dotchart(NBA2_maxp,col=c(rep("black",9),rep("red",10),rep("blue",2)),pch=16)
+
 NBA_maxp <- maxp(NBA)
-NBA_maxp %<>% ordertrans(c(colnames(NBA_table)[-(1:4)],"C_possession","W_possession"))
-dotchart(NBA_maxp,col=c(rep("black",9),rep("red",10),rep("blue",2)),pch=16)
+NBA_maxp %<>% ordertrans(c(colnames(NBA_table)[-(1:4)],"possession"))
+dotchart(NBA_maxp,col=c(rep("black",9),rep("red",10),rep("blue",1)),pch=16)
 ```
@@
 ```{r,statsanal,cache=TRUE}
-samep.test(NBA,c("C_possession","W_possession"))
+samep.test(NBA2,c("C_possession","W_possession"))
 ```
@@
 We may thus assume that Cleveland having possession is the same
-strength as the Warriors having possession.
+strength as the Warriors having possession, and from now on we work
+with `NBA` rather than `NBA2`.
+
+```{r,cache=TRUE}
+(C_allstrengths <- sort(NBA_maxp[01:9],decreasing=TRUE))
+(W_allstrengths <- sort(NBA_maxp[10:19],decreasing=TRUE))
+samep.test(NBA, names(C_allstrengths))
+samep.test(NBA, names(W_allstrengths))
+```
 
@@
 ```{r}
-C_allstrengths <- sort(NBA_maxp[01:9],decreasing=TRUE)
-W_allstrengths <- sort(NBA_maxp[10:19],decreasing=TRUE)
 C_strongestteam <- sum(C_allstrengths[1:5])

### 2021-01-24 inst/jester.Rmd
fixes #72

@@
 ```{r,label=findmaxp,cache=TRUE}
-system.time(mjest <- maxp(jester,n=1))
+system.time(jester_maxp <- maxp(jester,n=1))
 ```
@@
 
-```{r,showmjest}
-mjest
-plot(mjest)
+```{r,showjester_maxp}
+jester_maxp
+plot(jester_maxp)
 ```
@@
 ```{r,label=equalptestdoer,cache=TRUE}
-equalp.test(jester,startp=indep(mjest))
+equalp.test(jester,startp=indep(jester_maxp))
 ```
@@
 Retrieval_, 4(2), 133-151. July 2001.
+
+### Package dataset
+
+Following lines create `jester.rda`, residing in the `data/` directory of the package.
+
+```{r,label=savejesterdataset}
+save(jester,jester_maxp,file="jester.rda")
+```

### 2021-01-27 inst/counterstrike.Rmd
cache this computationally expensive result
Zachary wants to be called zac not zach

@@
 
-```{r, zachslist}
-zachslist <- list(
+```{r, zacslist}
+zacslist <- list(
     round1 = c("Skadoodle","olofmeister","tarik","GuardiaN","RUSH",
@@
 H <- hyper2(pnames=c(team1,team2))
-for(i in zachslist){
+for(i in zacslist){
   H <- H + counterstrike_maker(team1,team2, deathorder=i)
@@
 
-```{r, random2}
+```{r, random2,cache=TRUE}
 

### 2021-01-27 inst/counterstrike.Rmd
consistentification

@@
 ```{r, useit}
-H <- hyper2(pnames=c(team1,team2))
+counterstrike <- hyper2(pnames=c(team1,team2))
 for(i in zacslist){
-  H <- H + counterstrike_maker(team1,team2, deathorder=i)
+  counterstrike <- counterstrike + (counterstrike_maker(team1,team2, deathorder=i))
 }
@@
 ```{r, plotit}
-dotchart(maxp(H),pch=16,main='observed data')
+counterstrike_maxp <- maxp(counterstrike)
+dotchart(counterstrike_maxp,pch=16,main='observed data')
 ```
@@
 
-datapoint <- loglik(indep(maxp(H)),H) - loglik(indep(equalp(H)),H)
+datapoint <- loglik(indep(maxp(counterstrike)),counterstrike) - loglik(indep(equalp(counterstrike)),counterstrike)
 
@@
 	}
-  distrib[i] <- loglik(indep(maxp(Hrand)),Hrand) - loglik(indep(equalp(H)),Hrand)
+  distrib[i] <- loglik(indep(maxp(Hrand)),Hrand) - loglik(indep(equalp(counterstrike)),Hrand)
 }
@@
 perspective, the deathorder may as well be totally random.
+
+### Package dataset
+
+Following lines create `counterstrike.rda`, residing in the `data/` directory of the package.
+
+```{r,label=savecounterstrikedataset}
+save(counterstrike,counterstrike_maxp,zacslist,file="counterstrike.rda")
+```

### 2021-01-27 data/counterstrike.R

@@ -1,53 +1,0 @@
-`counterstrike` <-
-structure(list(brackets = list(c("GuardiaN", "NiKo"), c("GuardiaN", 
-"NiKo", "RUSH", "Skadoodle", "Stewie2k", "autimatic"), c("GuardiaN", 
-"NiKo", "RUSH", "Skadoodle", "Stewie2k", "autimatic", "karrigan", 
-"olofmeister"), c("GuardiaN", "NiKo", "RUSH", "Skadoodle", "Stewie2k", 
-"autimatic", "karrigan", "olofmeister", "rain", "tarik"), c("GuardiaN", 
-"NiKo", "RUSH", "Skadoodle", "Stewie2k", "autimatic", "karrigan", 
-"olofmeister", "tarik"), c("GuardiaN", "NiKo", "RUSH", "Skadoodle", 
-"Stewie2k", "autimatic", "karrigan", "rain", "tarik"), c("GuardiaN", 
-"NiKo", "RUSH", "Skadoodle", "Stewie2k", "autimatic", "karrigan", 
-"tarik"), c("GuardiaN", "NiKo", "RUSH", "Skadoodle", "Stewie2k", 
-"autimatic", "olofmeister", "rain", "tarik"), c("GuardiaN", "NiKo", 
-"RUSH", "Skadoodle", "Stewie2k", "autimatic", "tarik"), c("GuardiaN", 
-"NiKo", "RUSH", "Skadoodle", "Stewie2k", "karrigan", "olofmeister"
-), c("GuardiaN", "NiKo", "RUSH", "Skadoodle", "Stewie2k", "olofmeister"
-), c("GuardiaN", "NiKo", "RUSH", "Stewie2k", "autimatic", "karrigan", 
-"olofmeister", "rain", "tarik"), c("GuardiaN", "NiKo", "RUSH", 
-"Stewie2k", "autimatic", "karrigan", "rain"), c("GuardiaN", "NiKo", 
-"RUSH", "Stewie2k", "autimatic", "karrigan", "rain", "tarik"), 
-    c("GuardiaN", "NiKo", "Skadoodle", "Stewie2k", "olofmeister"
-    ), c("GuardiaN", "NiKo", "karrigan", "olofmeister"), c("GuardiaN", 
-    "NiKo", "karrigan", "olofmeister", "rain"), c("GuardiaN", 
-    "NiKo", "karrigan", "rain"), c("GuardiaN", "NiKo", "olofmeister"
[…]

### 2021-01-27 inst/NBA.Rmd
issue #73

@@
----
+--
 title: "Basketball: an analysis of a single game"
@@
 NBA2_maxp <- maxp(NBA2)
+
 NBA2_maxp %<>% ordertrans(c(colnames(NBA_table)[-(1:4)],"C_possession","W_possession"))
@@
 
+NBA_maxp_precomputed  <- c(
+                                        # precomputed; very very computer
+                                        # intensive but far higher likelihood
+                                        # than the result of maxp(NBA):
+
+    Barnes     = 0.150357987445096,   Curry       = 1.65443013016506e-06,
+    Durant     = 0.00856905879575095, Green       = 0.0253069705861388,
+    Iguodala   = 0.00916478949994613, Irving      = 0.0290988580288644, 
+    James      = 1.0000039853319e-06, Jefferson   = 0.00446265583190174, 
+    Korver     = 0.00959220615914446, Livingston  = 0.0775529739242179, 
+    Love       = 0.00253529174567821, McCaw       = 0.0189848298815197,
+    Pachiuila  = 0.0365345555046955,
+    possession = 0.431812626360354,
+    Shumpert   = 0.0556862911754396,   Smith      = 0.00064258717253104,
+    Thompson   = 6.86257295623038e-05, Thompson.1 = 0.00824420466131726,
+    West       = 0.0761409792438058,   Williams   = 0.0552418538199202)
 NBA_maxp <- maxp(NBA)
 NBA_maxp %<>% ordertrans(c(colnames(NBA_table)[-(1:4)],"possession"))
+
 dotchart(NBA_maxp,col=c(rep("black",9),rep("red",10),rep("blue",1)),pch=16)
@@
 ```{r,label=saveNBAdataset}
+NBA_maxp <- NBA_maxp_precomputed
 save(NBA_table,NBA,NBA_maxp,file="NBA.rda")

### 2021-01-28 inst/handover.Rmd
consistentification of handover work

---
title: "handover"
output: html_document
---

```{r setup, include=FALSE}
set.seed(0)
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("aylmer")
library("magrittr")
options("digits" = 5)
```

This short document is based on Altham and Hankin 2010, which
discussed an interesting dataset originally presented by Lin
et. al. (2009).  It uses more sophisticated idiom furnished by the
`hyper2` package, which is an improved and efficient upgrade of the
primitive `hyperdirichlet` package.  The dataset arose from 69 medical
malpractice claims and refers to the two surgeon–reviewers' answers to
the question "was there a communication breakdown in the hand-off
between physicians caring for the patient?".  The rows of Table 1
correspond to the answers that were given by reviewer 1, and the
columns to the answers given by reviewer 2.

```{r readcurlingtable}
handover_table <- as.matrix(read.table("handover.txt"))
names(dimnames(handover_table)) <- c("reviewer1","reviewer2")
handover_table[3,3] <- NA # structural zero
handover_table
```

Thus we see that in 26 cases, both reviewers answered "yes", in 18
both said "no".  The table also records partial responses that include
observations missed by one or both reviewers.  Responses missed by
both reviewers are uninformative and automatically rejected:
structural zeros, in `aylmer` terminology.  We first apply the Aylmer
test to assess whether the two reviewers give the same proportion of
"yes" responses:

```
> aylmer.test(a,alternative=function(a){a[1,2]-a[2,1]})

	Aylmer functional test for count data

data:  a
p-value = 0.169
alternative hypothesis: test function exceeds observed

> 
```

(Compare the McNemar test which gives an exact $p$-value of
$7/64\simeq 0.1094$).  We follow the previous workers and assume that
cases are missing at random and seek a likelihood function on
$\theta_{11},\theta_{10},\theta_{01},\theta_{00} subjet to
$\theta_{ij}\geq 0,\sum_{i,j \in\left\lbrace
0,1\right\rbrace}\theta_{ij}=1$.  Here "1" means "yes" and "0" means
"no", so $\theta_{01}$ is the probability of reviewer 1 responding
"no" and reviewer 2 responding "yes".

```{r}
handover <- hyper2()
handover["t11"] %<>% inc(handover_table[1,1])
handover["t10"] %<>% inc(handover_table[1,2])
handover["t01"] %<>% inc(handover_table[2,1])
handover["t00"] %<>% inc(handover_table[2,2])

handover[c("t11","t10")] %<>% inc(handover_table[1,3])  # yes-yes plus yes-no  = 2 
handover[c("t01","t00")] %<>% inc(handover_table[2,3])  #  no-yes plus  no-no  = 9 
handover[c("t11","t01")] %<>% inc(handover_table[3,1])  # yes-yes plus  no-yes = 4 
handover[c("t10","t00")] %<>% inc(handover_table[3,2])  # yes-no  plus  no-no  = 4 
handover
handover_maxp <- maxp(handover)
handover_maxp
```

We may simply test the hypothesis that $\theta_{01} = \theta_{10}$:

```{r}
samep.test(handover,c("t01","t10"))
```

Alternatively we might use the Bayesian paradigm and calculate the
probability of $\theta_{01} > \theta_{10}$:

```{r}
probability(handover,disallowed=function(p){p[2]<p[3]},tol=0.05)
```

the result differing slightly from the previous value, probably due to
numerical issues.

### References

* PME Altham and RKS Hankin 2010.  "Using recently developed software on
a 2x2 table of matched pairs with incompletely classified data".
_Journal of the Royal Statistical Society_, 59:2):377-379
* L West and RKS Hankin 2008. "Exact tests for two-way contingency
tables with structural zeros".  _Journal of Statistical Software_
28(11)
* RKS Hankin 2010. "A generalization of the Dirichlet distribution". 
 _Journal of Statistical Software_ 33(11)
* RKS Hankin 2017. "Partial rank data with the hyper2 package:
  likelihood functions for generalized Bradley-Terry models".  _The R
  Journal_ 9(2)

### Package dataset

Following lines create `handover.rda`, residing in the `data/` directory of the package.

```{r,label=savecurlingdataset}
save(handover_table,handover_maxp,handover,file="handover.rda")
```

### 2021-01-29 inst/handover.Rmd
consistentification

@@
 title: "handover"
-output: html_document
+output: bookdown::html_document2
 ---
@@
 Thus we see that in 26 cases, both reviewers answered "yes", in 18
-both said "no".  The table also records partial responses that include
+<both said "no".  The table also records partial responses that include
 observations missed by one or both reviewers.  Responses missed by
@@
 cases are missing at random and seek a likelihood function on
-$\theta_{11},\theta_{10},\theta_{01},\theta_{00} subjet to
+$\theta_{11},\theta_{10},\theta_{01},\theta_{00}$ subject to
 $\theta_{ij}\geq 0,\sum_{i,j \in\left\lbrace
-0,1\right\rbrace}\theta_{ij}=1$.  Here "1" means "yes" and "0" means
-"no", so $\theta_{01}$ is the probability of reviewer 1 responding
-"no" and reviewer 2 responding "yes".
+0,1\right\rbrace}\theta_{ij}=1$.  Here "1" means "yes" and "0"
+means "no", so $\theta_{01}$ is the probability of reviewer 1
+responding "no" and reviewer 2 responding "yes".
 
@@
 
-### References
+We can now consider the distribution of
+$\psi=\theta_{10}/(\theta_{10}+\theta_{01})$ under the posterior
+distribution induced by the complete cases:
+
+```{r,samplepsi,message=FALSE,cache=TRUE}
+jj <- rp(300000,H=handover)
+head(jj)
+psifun <- function(x){x[3]/(x[2]+x[3])}
+logoddsfun <- function(x){log(x[2]/x[3])}
+psi <- apply(jj,1,psifun)
+log_odds <- apply(jj,1,logoddsfun)
+psi_complete <- rbeta(length(psi),2,6)
+```
+
+We can now reproduce the four graphics given by Altham and Hankin:
+
+Sample of $\psi=\theta_{10}/(\theta_{10}+\theta_{01})$ under the
+posterior induced by the whole dataset (histogram), together with the
+posterior distribution of $\psi$ induced by the complete cases only
+(a beta distribution with parameters 2 and 6).
+
+```{r histpsi,echo=FALSE,fig.cap="Sample of $\\psi=\\theta_{10}/(\\theta_{10}+\\theta_{01})$ under the posterior induced  by the whole dataset (histogram), together with the posterior distribution of $\\psi$ induced by the complete cases only (a beta distribution with parameters 2 and 6)"}
+hist(psi,nclass=70,probability=TRUE,main=expression(paste("PDF of ",psi)))
+p <- seq(from=0,to=1,len=100)
+points(p,dbeta(p,2,6),type="l")
+abline(v=psifun(handover_maxp),col="red",lwd=5)
+segments(x0=0.5,y0=0,y1=1.5,col="blue",lwd=5)
+legend("topright",lty=1,lwd=c(1,5,5),col=c("black","red","blue"),
+       legend=c(
+           "complete cases",
+           expression(hat(psi) * " (complete cases)"),
+           expression(psi * " = 0.5")
+           ))
+```
+
+```{r qqplotpsi,echo=FALSE,fig.cap="Quantile-quantile plot of $\\psi=\\theta_{10}/(\\theta_{10}+\\theta_{01})$ under the posterior induced  by the whole dataset against the posterior distribution of $\\psi$ induced by the complete cases only (a beta distribution with parameters 2 and 6)",cache=TRUE}
+par(pty='s')
+qqplot(psi,psi_complete,asp=1,xlim=0:1,ylim=0:1)
+abline(0,1)
+```
+
+```{r ecdfpsi,echo=FALSE,fig.cap="Empirical cumulative distribution function of $\\psi=\\theta_{10}/(\\theta_{10}+\\theta_{01})$ under the posterior induced  by the whole dataset, and that of $\\psi$ induced by the complete cases only (a beta distribution with parameters 2 and 6)",cache=TRUE}
+plot(psi,seq_along(psi)/length(psi),type="l",col="black",xlab=expression(psi),ylab="quantile")
+points(psi_complete,seq_along(psi_complete)/length(psi_complete),type="l",col="red")
+legend("topleft",lty=1,col=c("black","red"),legend=c("whole dataset","complete cases"))
+```
+
+```{r logoddsqqplot,echo=FALSE,fig.cap="QQ plot of $\\phi=\\log(\\theta_{01}/\\theta_{10})$",cache=TRUE}
+qqnorm(log_odds,main="QQ plot of log-odds")
+abline(mean(log_odds),sd(log_odds))
+```
+
+Figures \@ref(fig:histpsi) to  \@ref(fig:logoddsqqplot) show different aspects of the dataset.
+
+### References {-}
 
@@
 
-### Package dataset
+### Package dataset {-}
 

### 2021-01-29 inst/handover.Rmd
typo
aylmer package not needed (it is deprecated anyway)

@@
 
-Thus we see that in 26 cases, both reviewers answered "yes", in 18
-<both said "no".  The table also records partial responses that include
+Thus we see that in 26 cases, both reviewers answered "yes", and in 18
+both said "no".  The table also records partial responses that include
 observations missed by one or both reviewers.  Responses missed by
@@
 ```{r,samplepsi,message=FALSE,cache=TRUE}
-jj <- rp(300000,H=handover)
+jj <- rp(3000,H=handover)
 head(jj)

### 2021-01-29 inst/handover.Rmd
consistentification

--- 2021-01-29 inst/handover.Rmd
@@
 library("hyper2")
-library("aylmer")
 library("magrittr")

### 2021-01-29 inst/moto.Rmd
consistentification

--- 2020-03-01 inst/moto.Rmd
@@
 title: "Grand Prix motorcycle racing"
-output: html_document
+output: bookdown::html_document2
 ---
@@
 Here I analyse a series of Grand Prix motorcycling race results using
-the `hyper` package which implements the Plackett-Luce likelihood
+the `hyper2` package which implements the Plackett-Luce likelihood
 function.  Consider the following dataset, taken from Wikipedia:
@@
 ```{r cars}
-data(motoGP)
-motoGP_2019_table
+moto_table <- read.table("motoGP_2019.txt")
+head(moto_table)
 ```
@@
 Each row is a driver (rider?) and each column (after the first) a
-venue.  Takin the first row we see that Marquez came second in Qatar,
-first in Argentina, retired in Austin, and so on.  In the first
+venue.  Taking the first row we see that Marquez came second in Qatar
+(QAT), first in Argentina, retired in Austin, and so on.  In the first
 column, we see the result from Qatar (QAT) in which Marquez came
@@
 includes six different classes of no-score such as "Ret" for retired,
-"WD" for withdrawn, and so on.  I now use bespoke R function
-\code{ordertable2supp()}, that translates data of this type into a
-log-likelihood.  The function is similar to, but slightly different
-from, the analysis in `eurovision.R`.
+"WD" for withdrawn, and so on.  I now use package function
+`ordertable2supp()`, that translates data of this type into a
+log-likelihood.  In the package, object `moto` is the corresponding
+Plackett-Luce likelihood function; we drop the final column which is
+the points scored by each rider.
 
-```{r,applydata}
-head(motoGP_2019)
+```{r,showmotolikelihoodhead}
+moto <- ordertable2supp(moto_table[,-ncol(moto_table)])
+head(moto)
 ```
 
-(object `motoGP_2019` is equivalent to
-`ordertable2supp(motoGP_2019_table)`.  Above, we see a small part of
-the Placket-Luce likelihood function for the 2019 results.  We can
-find the maximum likelihood estimator by using the `maxp()` function:
+Above, we see a small part of the Placket-Luce likelihood function for
+the 2019 results.  We can find the maximum likelihood estimator by
+using the `maxp()` function:
 
-```{r,maxp2017}
-options(width = 100)
-m <- maxp(motoGP_2019)
-m
-dotchart(m,pch=16)
-pie(m)
+```{r,maxpcalculator,cache=TRUE}
+moto_maxp <- maxp(moto)
+moto_maxp
 ```
 
-In the above, note the dominance of Marquez.  It is sometimes
-preferable to plot the estimated strengths on a log scale:
-
-```{r,maxp2017log}
-dotchart(log(m[1:27]),pch=16)
+```{r dotchartmoto, echo=TRUE,fig.cap="Dot chart of riders' strengths"}
+dotchart(moto_maxp,pch=16)
 ```
 
-(in the diagram above, we remove the bottom-ranked player, Lecuona,
-whose estimated strength is zero).
-
-# Likelihood scoring vs points scoring
-
-The championship winning order is determined by a points system: the
-winner is awarded 25 points, second place 18 points, and so on.  We
-can now check to see how the points total matches up with the
-likelihood estimation.
-
-```{r,pointsforwinning}
-(points_awarded <- c(`1st`=25, `2nd`=20, `3rd`=16, `4th`=13, `5th`= 11,
-`6th`=10, `7th`= 9, `8th`=8,`9th`=7, `10th`=6, `11th`=5, `12th`=4,
-`13th`=3, `14th`=2, `15th`=1))
+```{r piechartmoto,echo=TRUE,fig.cap="Pie chart of riders' strengths"}
+pie(moto_maxp)
 ```
 
-```{r,likelihoodandscore}
-points <- motoGP_2019_table$points
-names(points) <- motoGP_2019_table$driver
+In figures \@ref(fig:dotchartmoto) and \@ref(fig:piechartmoto), note
+the dominance of Marquez.  It is sometimes preferable to plot the
+estimated strengths on a log scale:
 
-ox <- rank(-points) # negative because number 1 = largest number of points
-oy <- rank(-m)      # negative because number 1 = largest strength
-par(pty='s')        # square plot
-plot(ox,oy,asp=1,pty='s',xlim=c(0,28),ylim=c(0,28),pch=16,
-xlab="official order",ylab="my order",main='motoGP, 2019 season')
-par(xpd=TRUE)       # allow drivers' names to appear outside plotting region
-for(i in seq_along(ox)){text(ox[i],oy[i],names(ox)[i],pos=4,col='gray') }
-par(xpd=FALSE)      # stop diagonal line from protruding beyond plotting region
-abline(0,1)
+```{r dotchartlogmoto, echo=TRUE,fig.cap="Dot chart of riders' strengths, log scale"}
+dotchart(log10(moto_maxp),pch=16)
 ```
 
-In the graph above, we see a certain amount of disagreement.  Only the
-top two riders, Marquez and Dovizioso, have rankings common to
-likelihood system and points system.  Considering the next four
-players (Vinales, Rins, Quartaro, Petrucci) we see that the
-points-based order is Vinales, Rins, Petrucci, but the likelihood
-order is Petrucci, Vinales, Rins (Rossi), Quartararo.
+(in figure \@ref(fig:dotchartlogmot) observe that the bottom-ranked
+player, Lecuona, has a ML strength of zero but the estimate is
+slightly positive for numerical reasons).
+
+### Package dataset {-}
+
+Following lines create `moto.rda`, residing in the `data/` directory
+of the package.
+
+```{r,label=savemotodataset}
+save(moto,moto_table,moto_maxp,file="moto.rda")
+```

### 2021-01-30 inst/pentathlon.Rmd
consistentify pentathlon

---
title: "Pentathlon"
output: bookdown::html_document2
---

```{r setup, include=FALSE}
set.seed(0)
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("magrittr")
options("digits" = 5)
```

The [modern
pentathlon](https://en.wikipedia.org/wiki/Modern_pentathlon) is an
Olympic sport that comprises five different events: fencing, swimming,
horse riding, and running.  This short document discusses a dataset
taken from the [2004 Summer
Olympics](https://en.wikipedia.org/wiki/Modern_pentathlon_at_the_2004_Summer_Olympics),
analysing the order statistic for the top seven competitors:

* Moiseev		
* Zadneprovskis	
* Capalini	
* Cerkovskis	
* Meliakh		
* Michalik	
* Walther

(I have removed diacritics from some of the names).

```{r readpentathlontable}
pentathlon_table <- as.matrix(read.table("pentathlon.txt"))
pentathlon_table
```

Thus the top line shows that Moiseev came fifth in shooting, first in
fencing and swimming, sixth in riding, and fifth in running.  The
first column shows that the shooting was won by Meliakh (who came
first), runner up was Machalik, and so on.

We can present the same information in another way:

```{r showranktable}
ordertable_to_ranktable(pentathlon_table)
```

```{r}
pentathlon <- ordertable2supp(pentathlon_table)
pentathlon_maxp <- maxp(pentathlon)
pentathlon_maxp
pie(pentathlon_maxp)
```

And we can test the null that the competitors all have the same
Plackett-Luce strengths:

```{r}
equalp.test(pentathlon)
```

### Package dataset {-}

Following lines create `pentathlon.rda`, residing in the `data/` directory of the package.

```{r,label=savecurlingdataset}
save(pentathlon_table,pentathlon,pentathlon_maxp,file="pentathlon.rda")
```

### 2021-01-30 data/pentathlon.R

@@ -1,34 +1,0 @@
-`pentathlon_ordertable` <-
-  structure(list(shooting = c(5L, 6L, 4L, 3L, 1L, 2L, 7L), fencing = c(1L, 
-2L, 6L, 3L, 7L, 4L, 5L), swimming = c(1L, 5L, 2L, 7L, 4L, 6L, 
-3L), riding = c(6L, 5L, 3L, 7L, 1L, 2L, 4L), running = c(5L, 
-1L, 4L, 2L, 6L, 7L, 3L)), class = "data.frame", row.names = c("Moiseev", 
-"Zadneprovskis", "Capalini", "Cerkovskis", "Meliakh", "Michalik", 
-"Walther"))
-
-`pentathlon` <-
-structure(list(brackets = list("Capalini", c("Capalini", "Cerkovskis", 
-"Meliakh", "Michalik", "Moiseev", "Walther"), c("Capalini", "Cerkovskis", 
-"Meliakh", "Michalik", "Moiseev", "Walther", "Zadneprovskis"), 
-    c("Capalini", "Cerkovskis", "Meliakh", "Michalik", "Walther"
-    ), c("Capalini", "Cerkovskis", "Meliakh", "Michalik", "Walther", 
-    "Zadneprovskis"), c("Capalini", "Cerkovskis", "Michalik", 
-    "Moiseev", "Walther", "Zadneprovskis"), c("Capalini", "Cerkovskis", 
-    "Moiseev", "Walther", "Zadneprovskis"), c("Capalini", "Meliakh"
-    ), c("Capalini", "Meliakh", "Michalik", "Moiseev"), c("Capalini", 
-    "Meliakh", "Michalik", "Moiseev", "Walther"), c("Capalini", 
-    "Meliakh", "Michalik", "Walther"), c("Capalini", "Meliakh", 
-    "Walther"), c("Capalini", "Moiseev", "Walther", "Zadneprovskis"
-    ), "Cerkovskis", c("Cerkovskis", "Meliakh", "Michalik", "Walther", 
-    "Zadneprovskis"), c("Cerkovskis", "Meliakh", "Michalik", 
-    "Zadneprovskis"), c("Cerkovskis", "Michalik"), c("Cerkovskis", 
-    "Michalik", "Zadneprovskis"), c("Cerkovskis", "Moiseev"), 
[…]

### 2021-01-30 inst/ordertrans.Rmd
better default for players argument to ordertrans()
bugfix and typos

@@
 ---
-title: "The ordertrans() function"
-output: html_document
+title: "The `ordertrans()` function"
+output: bookdown::html_document2
 ---
@@
 The `ordertrans()` function can be difficult to understand and this
-short document provides some sensible use-cases.
-
-## Hello world
-
-The manpage provides a very simple example:
+short document provides some sensible use-cases.  The manpage provides
+a very simple example, but here we are going to use an even simpler
+example:
 
@@
 ```{r,label=useordertrans}
-o <- ordertrans(x,letters[1:4])
+o <- ordertrans(x) # by default, sorts names() into alphabetical order
 o
@@
 
-Let us consider the skating dataset and use `ordertrans()` to study
-it.
+Let us consider the skating dataset and use `ordertrans()` to study it
+(note that the skating dataset is analysed in more depth in
+`skating.Rmd`).
 
@@
 
-We might ask how judges 1 and 2 compare to one another?  We need to
+We might ask how judges `J1` and `J2` compare to one another?  We need to
 create vectors like `x` and `y` above:
@@
 In the above, see how objects `j1` and `j2` have identical names, in
-the same order.  Observe that `hughes` is ranked 1 by judge 1, and 4
-by judge 2.  This makes it sensible to plot `j1` against `j2`:
+the same order.  Observe that `hughes` is ranked 1 (that is, first) by
+`J1`, and 4 (that is, 4th) by `J2`.  This makes it sensible to plot
+`j1` against `j2`:
 
-```{r}
+```{r j1vsj2,fig.cap="Judge 1 vs judge 2"}
 par(pty='s')  # forces plot to be square
@@
 
-In the above plot we see general agreement but differences in detail.
-For example, `hughes` is ranked first by judge 1 and fourth by judge
-2.  However, other problems are not so easy.  Suppose we wish to
-compare the ranks according to likelihood with the ranks according to
-some points system.
+In figure \@ref(fig:j1vsj2), we see general agreement but differences
+in detail.  For example, `hughes` is ranked first by judge 1 and
+fourth by judge 2.  However, other problems are not so easy.  Suppose
+we wish to compare the ranks according to likelihood with the ranks
+according to some points system.
 
@@
 
-```{r,label=crapplot}
-plot(mL,ordertrans(mP,mL))
+```{r,label=crapplot,fig.cap="points-based ranks vs likelihood ranks"}
+plot(mL,ordertrans(mP,names(mL)))
 ```
 
-However, the plot above is a bit crude.  Function `ordertransplot()`
-gives a more visually pleasing output:
+However, figure \@ref(fig:crapplot) is a bit crude.  Function
+`ordertransplot()` gives a more visually pleasing output, see figure
+\@ref(fig:showoffordertransplot).
 
-```{r,label=showoffordertransplot}
+```{r,label=showoffordertransplot,fig.cap="points=based rank vs likelihood rank using `ordertransplot()`"}
 ordertransplot(mL,mP,xlab="likelihood rank",ylab="Borda rank")
@@
 
-```{r,transplotjudge1}
+```{r transplotjudge1,fig.cap="Likelihood rank vs rank according to Judge 1"}
 ordertransplot(mL,j1,xlab="likelihood rank",ylab="Judge 1 rank")
@@
 
-In the above, looking at the lower-left corner, we see (reading
-horizontally) that the likelihood method placed Slutskya first, then
-Hughes second, then Kwan third; while (reading vertically) judge 1
-placed Hughes first, then Kwan, then Slutskya.
+In figure \@ref(fig:transplotjudge1), looking at the lower-left
+corner, we see (reading horizontally) that the likelihood method
+placed Slutskya first, then Hughes second, then Kwan third; while
+(reading vertically) judge 1 placed Hughes first, then Kwan, then
+Slutskya.

### 2021-01-30 inst/skating.Rmd
consistentification

--- 2021-01-18 inst/skating_analysis.Rmd
@@
-title: "Ladies' figure skating"
-author: "Robin K. S. Hankin"
-date: "`r Sys.Date()`"
-vignette: >
-  %\VignetteIndexEntry{figure_skating}
-  %\VignetteEngine{knitr::rmarkdown}
-  %\VignetteEncoding{UTF-8}
+---
+title: "skating"
+output: bookdown::html_document2
 ---
 
+```{r setup, include=FALSE}
+set.seed(0)
+knitr::opts_chunk$set(echo = TRUE)
+library("hyper2")
+library("magrittr")
+options("digits" = 5)
+```
+
 # Ladies' figure skating in the 2002 Winter Olympics
-
-```{r, echo=T, results = 'hide', message= FALSE}
-library("hyper2",quietly=TRUE)
-```
 
@@
 `skating_table` is copied from Lock and Lock (see skating.Rd for more
-details).  It also creates file `data/skating.rda` which contains R objects `skating_table`,
+details).  It also creates file `data/skating.rda` which contains R
+objects `skating`, `skating_table`, and `skating_maxp`.
 
@@
 skating_table <- read.table("skating.txt",header=TRUE)
+skating_table
 ```
 
-Object `skating_table` is an **order table**.  It is structured so
-that each competitor is a row, and each judge is a column.  Function
+Object `skating_table` is an order table.  It is structured so that
+each competitor is a row, and each judge is a column.  Function
 `rank_likelihood()` considers each row to be a race [or a judge], so
 we need to take a transpose.  The following R idiom is plausible but
-incorrect:
+**incorrect**:
 
 ```{r,skatingwrong}
-skating_incorrect <- rank_likelihood(t(skating_table))
+skating_incorrect <- rank_likelihood(t(skating_table)) # incorrect
 ```
@@
 The above is incorrect because we are applying a rank technique to an
-order table.  Correct analysis follows; firstly we coerce to a
-**rank** table:
+order table.  Correct analysis follows.  One good thing to do is to
+present the dataset as an order table;
 
 ```{r,coercetoranktable}
-(skating_order <- ordertable_to_ranktable(skating_table))
+ordertable_to_ranktable(skating_table)
 ```
@@
 "came second" (`c2`) and so on to "came 23rd" (`c23`).  From the first
-column, we see that Hughes came first 5 times and Slutskya came first
-four times, and from the last column we see that all judges except
-`J3` awarded Luca last place.  File `man/rrank.Rd` has a detailed
-discussion of the differences between rank and order.
+column, we see that Hughes came first 5 times [according to judges
+1,5,7,8,9] and Slutskya came first four times [judges 2,3,4,6].  From
+the last column we see that all judges except `J3` awarded Luca last
+place.  File `man/rrank.Rd` has a detailed discussion of the
+differences between rank and order.
 
@@
 Slutskaya second, five gave Kawn second, and one gave Cohen second.
+The first row corresponds to Hughes and we see that, of $5+1+1+2=9$
+judges, 5 placed Hughes first, one placed her second, one placed her
+third, and two placed her fourth.
 
@@
 
- maximum likelihood estimated strengths furnish an ordering for the competitors.
-We can compare ranking by strengths with the official point-tallying method:
+maximum likelihood estimated strengths furnish an ordering for the
+competitors.  We can compare ranking by strengths with the official
+point-tallying method:
 
-```{r,dpi=72}
-par(pty='s')
-lik <- rank(-m)
-plot(1:23,lik,asp=1,pch=16,xlab='official order',ylab='likelihood order')
-text(1:23,lik,asp=1,pch=16,pos=c(rep(4,19),rep(2,4)),rownames(skating_table))
-abline(0,1)
+```{r rLrO,fig.cap="Official ranks vs likelihood rank, Ladies free skating, 2002 Winter Olympics"}
+ rL <- sort(skating_maxp,decreasing=TRUE)
+ rL[] <- seq_along(rL)
+ rO <- seq_len(nrow(skating_table))
+ names(rO) <- rownames(skating_table)
+ ordertransplot(rO,rL,xlab="offical rank",ylab="likelihood rank",main="Ladies free skating, 2002 Winter Olympics")
 ```
 
-In the above diagram, there is close agreement between the two methods
-in the large, but differences in detail.  For example, the official
-ordering is hughes first, then slutskaya second, then kwan third;
-likelihood winner is slutskya, followed by hughes and then kwan.
-
-We can gain some understanding of this result by looking at the raw
-judging results for the three medallists:
+In figure \@ref(fig:rLrO), there is close agreement between the two
+methods in the large, but differences in detail.  For example, the
+official ordering is hughes first, then slutskaya second, then kwan
+third; likelihood winner is slutskya, followed by hughes and then
+kwan.  We can gain some understanding of this result by looking at the
+raw judging results for the three medallists:
 
@@
 ```{r}
-o <- maxp(o)
+(o <- maxp(o))
 pie(o)
@@
 
-### Package dataset
+### Package dataset {-}
 
-Following lines create `skating.rda`, residing in the `data/` directory of the package.
+Following lines create `skating.rda`, residing in the `data/`
+directory of the package.
 

### 2021-01-31 inst/soling.Rmd
consistentify soling

@@
 ---
-title: "Soling at the 2000 Summer Olympics"
-output: html_document
+title: "soling"
+output: bookdown::html_document2
 ---
@@
 ```{r setup, include=FALSE}
+set.seed(0)
 knitr::opts_chunk$set(echo = TRUE)
+library("hyper2")
+library("magrittr")
+options("digits" = 5)
 ```
 
-The Soling is a type of boat raced in the 2000 Summer Olympics.  Here I analyse race data taken from
-
-[https://en.wikipedia.org/w/index.php?title=Sailing_at_the_2000_Summer_Olympics_\%E2\%80\%93_Soling&oldid=945362535]
-
+The Soling is a type of boat raced in the 2000 Summer Olympics.  Here
+I analyse race data taken from
+[wikipedia](https://en.wikipedia.org/w/index.php?title=Sailing_at_the_2000_Summer_Olympics_\%E2\%80\%93_Soling&oldid=945362535)
 using a Plackett-Luce extension of the Bradley-Terry model.
@@
 ```{r loadlib}
-library("hyper2")
-data(soling)
-H <- soling_table_2000
-ordertable_to_ranktable(soling_table_2000)
-head(soling2000)
+soling_table <- read.table("soling.txt")
+soling_table
+```
+
+Thus we see, reading column 1, that in race 1, `norway` came fifth,
+`nz` third, `netherlands` first, and so on.  We can present this
+information in a rank table format:
+
+```{r interpretsoling}
+ordertable_to_ranktable(soling_table)
+```
+
+and also convert it to a likelihood function:
+
+```{r solingtolikelihood}
+soling <- ordertable2supp(soling_table)
+head(soling)
 ```
@@
 ```{r,rrqf}
-soling2000_rr1
-soling2000_rr2
-soling2000_qf
+soling_rr1 <- read.table("soling_rr1.txt")
+soling_rr2 <- read.table("soling_rr2.txt")
+soling_qf  <- read.table("soling_qf.txt" )
+soling_rr1
 ```
 
-Thus, for example, taking `soling2000_rr1[1,2] == TRUE` we see that
+Thus, for example, taking `soling_rr1[1,2] == TRUE` we see that
 Germany beat Sweden in round robin 1.  To incorporate this information
@@
 
-```{r}
-H <- soling2000
-before <- maxp(H)
-H %<>% rr(soling2000_rr1)
-H %<>% rr(soling2000_rr2)
-H %<>% rr(soling2000_qf)
-after <- maxp(H)
+```{r showdiff,fig.cap="Evaluate before and after incorporation round robin and quarter final information"}
+H <- soling  # before new information
+soling_maxp <- maxp(H) 
+H %<>% rr(soling_rr1)
+H %<>% rr(soling_rr2)
+H %<>% rr(soling_qf)
+soling_after <- H # after new information
+soling_after_maxp <- maxp(soling_after)
 par(pty="s")
-plot(before,after,asp=1,pch=16,xlim=c(0,0.3),ylim=c(0,0.3))
+plot(soling_maxp,soling_after_maxp,,asp=1,pch=16,xlim=c(0,0.3),ylim=c(0,0.3))
 abline(0,1)
@@
 
-The diagram above shows a modest change.  We might ask whether the
-additional observations are consistent with the original Placket-Luce
-likelihood:
+Figure \@ref(fig:showdiff) shows a modest change.  We might ask
+whether the additional observations are consistent with the original
+Placket-Luce likelihood:
 
 ```{r}
-(H2 <- H + (-1)*soling2000)
+(soling_diff <- soling_after + (-1)*soling)
 ```
 
-So `H2` is the _difference_ between `H`--- which is `soling2000` plus
-the observations from the roundrobins and quarter finals---and
-`soling2000`.  In other words, `H2` is the simply the information in
-the roundrobins and quarter finals.
+So `H2` is the _difference_ between `H`--- which is `soling` plus the
+observations from the roundrobins and quarter finals---and `soling`
+withouth the extra information.  In other words, `H2` is the simply
+the information in the roundrobins and quarter finals.
 
 ```{r,testH2equalp}
-equalp.test(H2)
+equalp.test(soling_diff)
 ```
+
+### Package dataset {-}
+
+Following lines create `soling.rda`, residing in the `data/`
+directory of the package.
+
+```{r,label=savesolingdataset}
+save(soling,soling_after, soling_maxp,soling_after_maxp,soling_table,
+     soling_rr1, soling_rr2, soling_qf,
+  file="soling.rda")
+```

### 2021-01-31 data/soling.R

@@ -1,184 +1,0 @@
-`soling2000` <-
-    structure(list(brackets = list("UK", c("UK", "USA", "australia", 
-"canada", "denmark", "finland", "france", "germany", "italy", 
-"netherlands", "norway", "nz", "russia", "spain", "sweden"), 
-    c("UK", "USA", "australia", "canada", "denmark", "finland", 
-    "france", "germany", "italy", "netherlands", "norway", "nz", 
-    "russia", "spain", "sweden", "ukraine"), c("UK", "USA", "australia", 
-    "canada", "denmark", "finland", "france", "germany", "italy", 
-    "netherlands", "nz", "russia", "spain", "sweden"), c("UK", 
-    "USA", "australia", "canada", "denmark", "finland", "france", 
-    "germany", "italy", "netherlands", "nz", "russia", "spain", 
-    "sweden", "ukraine"), c("UK", "USA", "australia", "canada", 
-    "denmark", "finland", "france", "germany", "italy", "netherlands", 
-    "russia", "spain", "sweden"), c("UK", "USA", "australia", 
-    "canada", "denmark", "finland", "france", "germany", "italy", 
-    "norway", "nz", "russia", "spain", "sweden", "ukraine"), 
-    c("UK", "USA", "australia", "canada", "denmark", "finland", 
-    "france", "germany", "italy", "norway", "nz", "spain", "sweden", 
-    "ukraine"), c("UK", "USA", "australia", "canada", "denmark", 
-    "finland", "france", "germany", "italy", "norway", "spain", 
-    "sweden", "ukraine"), c("UK", "USA", "australia", "canada", 
-    "denmark", "france", "germany", "italy", "netherlands", "russia", 
-    "spain", "sweden"), c("UK", "USA", "australia", "canada", 
-    "denmark", "france", "germany", "italy", "russia", "spain", 
-    "sweden"), c("UK", "USA", "australia", "canada", "denmark", 
-    "france", "germany", "russia", "spain", "sweden"), c("UK", 
-    "USA", "canada", "denmark", "finland", "france", "germany", 
-    "italy", "netherlands", "norway", "nz", "russia", "spain", 
[…]

### 2021-01-31 inst/rowing.Rmd
consistentify rowing dataset

---
title: "rowing: analysis of sculls in the 2016 Olympics"
output: bookdown::html_document2
---

```{r setup, include=FALSE}
set.seed(0)
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("magrittr")
options("digits" = 5)
```

This short document analyses the `rowing` dataset: sculls in the 2016
olympics.  It is not really standalone and is here as part of a
consistent dataset generation mechanism used in the `hyper2` package.
See the `hyper2` vignette for a more formal discussion.  In Olympic
sculling, up to six individual competitors race a small boat called a
scull over a course of length 2\,km; the object is to cross the
finishing line first.  Note that actual timings are irrelevant, given
the model, as the sufficient statistic is the order in which
competitors cross the finishing line.  The 2016 Summer Olympics is a
case in point: the gold and silver medallists finished less than~5
milliseconds apart, corresponding to a lead of~$\sim
2.5\,\mathrm{cm}$.  The dataset is as follows;

```{r}
show <- function(filename){
  unlist(lapply(strsplit(readLines(filename)," "),paste,collapse=" "))
}					       
show("rowing.txt")
```

The first line shows that `fournier` came first, `cabrera` second, and
so on.  We can convert this dataset into a Plackett-Luce likelihood
function as follows:

```{r definerowing, cache=TRUE}

f <- function(filename){
  o <- strsplit(readLines(filename)," ")

  rowers <- sort(unique(unlist(o)))
  H <- hyper2(list(),0,pnames=rowers)

  for(v in o){
      v <- rev(v)
      for(i in seq_along(v)){
          H[v[i]] %<>% inc           # H[v[i]] %<>% `+`(1)
          H[v[seq_len(i)]] %<>% dec  # H[v[seq_len(i)]] %<>% `-`(1)
      }
  }
  return(H)
}

rowing <- f("rowing.txt")
pnames(rowing)
head(rowing)
```

Object `rowing` is a loglikelihood function on the strengths of the 32
competitors.  Our first step is to find the evaluate:

```{r maxpcalc,cache=TRUE}
rowing_maxp <- maxp(rowing)
rowing_maxp
```

and we may display it using a dot chart:

```{r showrowingmaxppie, echo=FALSE, fig.cap="dot chart of results in the 2016 Olympic Sculls"}
dotchart(rowing_maxp)
```

```{r showrowingmaxppielog, echo=FALSE, fig.cap="dot chart of results in the 2016 Olympic Sculls (log scale)"}
dotchart(log10(rowing_maxp))
```

Figure \@ref(fig:showrowingmaxppie) shows that many competitors have
very small strength and figure \@ref(fig:showrowingmaxppielog) shows
that many competitors have zero strength to numerical precision.
Looking at the original dataset we can see why: observe that `gambour`
came last in every race except for race number 7 (won by `boudina`).
In race 7, he beat only `yaklovlev`.  But we see that `yaklovlev` came
last in every race except race 20, where he beat only `gambour`.  Thus
the maximum likelihood estimate for `gambour + yaklovlev` would be
zero, because these two competitors always lose to everyone else.  It
is thus reasonable to eliminate `gambour` and `yaklovlev` from the
dataset. We may apply this reasoning recursively and end up with the
following dataset:

```{r}
show("rowing_minimal.txt")
```

We may then analyse this:

```{r rowingminimaldef}
rowing_minimal <- f("rowing_minimal.txt")
head(rowing_minimal)
pnames(rowing_minimal)
```

This is a much more manageable dataset, much easier to study and less
cluttered, as competitors with zero estimated strength are not
present. *However, it is incorrect as `drysdale` does not appear.
Something is wrong.*

```{r minimalmaxp,cache=TRUE}
(rowing_minimal_maxp <- maxp(rowing_minimal))
```

```{r maxpcalcminimal,fig.cap="Minimal dataset, maximum likelihood estimate of strength"}
dotchart(rowing_minimal_maxp)
```

### Package dataset {-}

Following lines create `rowing.rda`, residing in the `data/` directory
of the package.

```{r,label=saverowingingdataset}
save(rowing,rowing_minimal,rowing_maxp,rowing_minimal_maxp,file="rowing.rda")
```

### 2021-01-31 data/rowing.R

@@ -1,102 +1,0 @@
-`sculls2016` <-
-structure(list(brackets = list("banna", c("banna", "boudina", 
-"cabrera", "molnar", "obreno", "rivarola"), c("banna", "boudina", 
-"cabrera", "molnar", "rivarola"), c("banna", "boudina", "molnar", 
-"rivarola"), c("banna", "cabrera", "campbell", "grant", "hoff"
-), c("banna", "cabrera", "campbell", "grant", "hoff", "szymczyk"
-), c("banna", "cabrera", "fournier", "hoff"), c("banna", "cabrera", 
-"fournier", "hoff", "martin"), c("banna", "cabrera", "fournier", 
-"hoff", "martin", "synek"), c("banna", "cabrera", "hoff"), c("banna", 
-"campbell", "grant", "hoff"), c("banna", "campbell", "hoff"), 
-    c("banna", "hoff"), c("banna", "hoff", "martin", "taieb", 
-    "yakovlev"), c("banna", "martin", "taieb", "yakovlev"), c("banna", 
-    "taieb", "yakovlev"), "bhokanal", c("bhokanal", "cabrera", 
-    "fournier", "kelmelis", "saensuk", "teilemb"), c("bhokanal", 
-    "cabrera", "kelmelis", "saensuk", "teilemb"), c("bhokanal", 
-    "campbell", "esquivel", "khafaji", "martin", "szymczyk"), 
-    c("bhokanal", "campbell", "esquivel", "khafaji", "szymczyk"
-    ), c("bhokanal", "dongyong", "esquivel", "memo", "molnar", 
-    "rosso"), c("bhokanal", "dongyong", "garcia", "kelmelis", 
-    "khafaji"), c("bhokanal", "dongyong", "garcia", "kelmelis", 
-    "khafaji", "molnar"), c("bhokanal", "esquivel", "khafaji"
-    ), c("bhokanal", "esquivel", "khafaji", "szymczyk"), c("bhokanal", 
-    "kelmelis", "saensuk", "teilemb"), "boudina", c("boudina", 
-    "esquivel", "kholmirzayev", "memo", "rivarola", "rosso"), 
-    c("boudina", "gambour", "garcia", "teilemb", "yakovlev"), 
-    c("boudina", "garcia", "kelmelis", "khafaji", "kholmirzayev", 
[…]

### 2021-02-01 inst/rowing.Rmd
correct removal of dominating or dominated rowers

@@
 
-We may then analyse this:
+(also, we have removed `drysdale`, as he won everything and his ML
+strength would be 1).  We may then analyse this:
 
@@
 cluttered, as competitors with zero estimated strength are not
-present. *However, it is incorrect as `drysdale` does not appear.
-Something is wrong.*
+present.
 
 ```{r minimalmaxp,cache=TRUE}
-(rowing_minimal_maxp <- maxp(rowing_minimal))
+rowing_minimal_maxp <- maxp(rowing_minimal)
+rowing_minimal_maxp
 ```
@@
 dotchart(rowing_minimal_maxp)
+dotchart(log10(rowing_minimal_maxp))
 ```

### 2021-02-01 inst/universities.Rmd
consistentify universities

@@
 title: "universities"
-output: html_document
+output: bookdown::html_document2
 ---
@@
 ```{r setup, include=FALSE}
+set.seed(0)
+knitr::opts_chunk$set(echo = TRUE)
 library("hyper2")
 library("magrittr")
-knitr::opts_chunk$set(echo = TRUE)
+options("digits" = 5)
 ```
@@
 ```{r cars}
-a <- read.table("universities.txt",header=TRUE)
-head(a,12)
+universities_table <- read.table("universities.txt",header=TRUE)
+head(universities_table,12)
 ```
@@
 }
-f(a,2012)
+f(universities_table,2012)
 ```
@@
 ```{r,playerwithadvantage}
-H <- hyper2(pnames=a[1:8,1])
-for(i in 2013:2020){ H %<>% `+`(ordertable2supp(f(a,i)))}
-H
+universities <- hyper2(pnames=universities_table[1:8,1])
+for(i in 2013:2020){universities %<>% `+`(ordertable2supp(f(universities_table,i)))}
+universities
 ```
 
-Object `H` is simply the support function for the universities'
-strengths for all years' observations (assumed to be independent)
-2013-2020:
+Object `universities` is simply the support function for the
+universities' strengths for all years' observations (assumed to be
+independent) 2013-2020:
 
 ```{r,label=maxpcalc,cache=TRUE}
-m <- maxp(H)
-m
-pie(m)
+universities_maxp <- maxp(universities)
+universities_maxp
+pie(universities_maxp)
 ```
@@
 
-H1 <- hyper2(pnames=c("AUT_pre",a[c(1:2,4:8),1]))
+H1 <- hyper2(pnames=c("AUT_pre",universities_table[c(1:2,4:8),1]))
 for(i in 2012:2016){
-  b <- f(a,i)
+  b <- f(universities_table,i)
   rownames(b)[3] <- "AUT_pre"
@@
 
-H2 <- hyper2(pnames=c("AUT_post",a[c(1:2,4:8),1]))
+H2 <- hyper2(pnames=c("AUT_post",universities_table[c(1:2,4:8),1]))
 for(i in 2017:2020){
-  jj <- f(a,i)
+  jj <- f(universities_table,i)
   rownames(jj)[3] <- "AUT_post"
@@
 
-See how we reject $H_0\colon S=0$ with a p-value of about 0.00676,
-significant at 1\%.
-
 ```{r}
@@
 maxp(H)
-samep.test(H,c(1,9))
 ```
+
+So from the above we see that the estimated post-2016 strength of AUT
+is larger than the pre-2016 strength.  Is this change significant?
+
+```{r}
+samep.test(H,c("AUT_pre","AUT_post"))
+```
+
+### Package dataset {-}
+
+Following lines create `universities.rda`, residing in the `data/`
+directory of the package.
+
+```{r,label=savecurlingdataset}
+save(universities_table,universities_maxp,universities,file="universities.rda")
+```

### 2021-02-01 data/universities.R

@@ -1,160 +1,0 @@
-`universities_table` <-
-structure(list(uni = structure(c(6L, 5L, 1L, 2L, 3L, 4L, 7L, 
-8L, 6L, 5L, 1L, 2L, 3L, 4L, 7L, 8L, 6L, 5L, 1L, 2L, 3L, 4L, 7L, 
-8L, 6L, 5L, 1L, 2L, 3L, 4L, 7L, 8L, 6L, 5L, 1L, 2L, 3L, 4L, 7L, 
-8L, 6L, 5L, 1L, 2L, 3L, 4L, 7L, 8L, 6L, 5L, 1L, 2L, 3L, 4L, 7L, 
-8L, 6L, 5L, 1L, 2L, 3L, 4L, 7L, 8L, 6L, 5L, 1L, 2L, 3L, 4L, 7L, 
-8L), .Label = c("AUT", "Canterbury", "Lincoln", "Massey", "Otago", 
-"UoA", "Victoria", "Waikato"), class = "factor"), year = c(2012L, 
-2012L, 2012L, 2012L, 2012L, 2012L, 2012L, 2012L, 2013L, 2013L, 
-2013L, 2013L, 2013L, 2013L, 2013L, 2013L, 2014L, 2014L, 2014L, 
-2014L, 2014L, 2014L, 2014L, 2014L, 2015L, 2015L, 2015L, 2015L, 
-2015L, 2015L, 2015L, 2015L, 2016L, 2016L, 2016L, 2016L, 2016L, 
-2016L, 2016L, 2016L, 2017L, 2017L, 2017L, 2017L, 2017L, 2017L, 
-2017L, 2017L, 2018L, 2018L, 2018L, 2018L, 2018L, 2018L, 2018L, 
-2018L, 2019L, 2019L, 2019L, 2019L, 2019L, 2019L, 2019L, 2019L, 
-2020L, 2020L, 2020L, 2020L, 2020L, 2020L, 2020L, 2020L), Teaching = structure(c(41L, 
-33L, 54L, 3L, 54L, 4L, 2L, 1L, 53L, 43L, 54L, 14L, 54L, 10L, 
-21L, 5L, 37L, 34L, 54L, 7L, 54L, 54L, 12L, 8L, 42L, 38L, 54L, 
-19L, 54L, 54L, 20L, 9L, 50L, 48L, 3L, 28L, 54L, 26L, 36L, 11L, 
[…]

### 2021-02-01 inst/eurovision.Rmd
inst/eurovision.Rmd

@@
 title: "Eurovision Song Contest: semifinal 1, 2009"
-output: html_document
+output: bookdown::html_document2
 ---
@@
 library("magrittr")
+options("digits" = 5)
 abbreviated <- TRUE  # change to FALSE for full names
@@
 
-## The Eurovision Song contest
-
-The dataset `wiki_matrix`, defined below, is copied from "Eurovision
+The file `eurovision.txt`, defined below, is copied from "Eurovision
 Song Contest 2009," Wikipedia, accessed May 13, 2018.  It refers to
@@
 ```{r, specifymatrix}
-data(eurovision)
-(wiki_matrix <- as.matrix(eurovision2009_votingtable))
+eurovision_table <- as.matrix(read.table("eurovision.txt"))
+eurovision_table
 ```
@@
 ```{r,prefsetter}
-preference <- wiki_matrix*0  
-for(i in seq_along(points)){ preference[wiki_matrix == points[i]] <- i }
+preference <- eurovision_table*0  
+for(i in seq_along(points)){ preference[eurovision_table == points[i]] <- i }
 
@@
 
-rownames(wiki_matrix) <- jj[1:18]
-colnames(wiki_matrix) <- jj
+rownames(eurovision_table) <- jj[1:18]
+colnames(eurovision_table) <- jj
 ```
@@
 We need to convert matrix `preference` into a likelihood function,
-here `eurovision2009`:
+here `eurovision`:
 
 ```{r,morestuff}
-eurovision2009 <- hyper2()
+eurovision <- hyper2()
 
@@
     d <- preference[,i,drop=TRUE]
-	eurovision2009 <- 	eurovision2009  + ordervec2supp(d[!is.na(d)])
+	eurovision <- eurovision + ordervec2supp(d[!is.na(d)])
 } # i loop closes
+```
+
+```{r,checknull,cache=TRUE}
+eurovision_maxp <- maxp(eurovision)
+eurovision_maxp
+```
+
+```{r,plotpieeurovision,fig.cap="Maximum likelihood strengths of competitors in the 2009 Eurovision song contest, semifinal 1"} 
+pie(eurovision_maxp)
 ```
@@
 
-```{r,checknull,cache=TRUE}
-m <- maxp(eurovision2009)
-m
-equalp.test(eurovision2009)
+'''{r,label=equalptesteurovision,cache=TRUE}
+equalp.test(eurovision)
 ```
@@
 ```{r,orderplot}
-points_scored <- rowSums(wiki_matrix,na.rm=TRUE)  # points
+points_scored <- rowSums(eurovision_table,na.rm=TRUE)  # points
 ox <- rank(-points_scored)
-oy <- rank(-m)   # m = MLE
+oy <- rank(-eurovision_maxp)   # m = MLE
 oyp <- ordertrans(oy,names(ox))
@@
 ```
+
+### Package dataset {-}
+
+Following lines create `eurovision.rda`, residing in the `data/` directory of the package.
+
+```{r,label=saveeurovisiondataset}
+save(eurovision_table,eurovision_maxp,eurovision,file="eurovision.rda")
+```

### 2021-02-01 inst/rowing.Rmd
race() and tidyup

@@
   o <- strsplit(readLines(filename)," ")
-
-  rowers <- sort(unique(unlist(o)))
-  H <- hyper2(list(),0,pnames=rowers)
-
-  for(v in o){
-      v <- rev(v)
-      for(i in seq_along(v)){
-          H[v[i]] %<>% inc           # H[v[i]] %<>% `+`(1)
-          H[v[seq_len(i)]] %<>% dec  # H[v[seq_len(i)]] %<>% `-`(1)
-      }
-  }
+  H <- hyper2()
+  for(v in o){H <- H + race(v)}
   return(H)

### 2021-02-01 R/hyper2.R

@@ -498,11 +498,18 @@ setGeneric("pnames<-",function(x,value){standardGeneric("pnames<-")})
   return(out)
 }
 
-`rankvec_likelihood` <- function(x){
-    jj <- seq_along(x)
-    names(jj)  <- x
-    return(rank_likelihood(jj))
+`rankvec_likelihood` <- function(v){
+  stopifnot(all(table(v)==1))
+  out <- hyper2()
+  v <- rev(v)   # first-placed competitor is the first element of v
+  for(i in seq_along(v)){
+    out[v[i]] <- out[v[i]] + 1  #  out[v[i]] %<>% inc
+    out[v[seq_len(i)]] <- out[v[seq_len(i)]] - 1
+  }
+  return(out)
 }
+
+`race` <- rankvec_likelihood
 
 #`addrank` <- function(H,ranks){
 #     .Defunct(new = 'rank_likelihood',msg='use H <- H+rank_likelihood(...)')
@@ -977,4 +984,3 @@ setGeneric("pnames<-",function(x,value){standardGeneric("pnames<-")})
   abline(0,1)
   for(i in seq_along(ox)){text(ox[i],oy[i],names(ox)[i],pos=4,col='gray',cex=0.7)}
 }
-

### 2021-02-01 inst/soling.Rmd
H1 - H2 now gives the difference in likelihood functions

@@
 par(pty="s")
-plot(soling_maxp,soling_after_maxp,,asp=1,pch=16,xlim=c(0,0.3),ylim=c(0,0.3))
+plot(soling_maxp,soling_after_maxp,asp=1,pch=16,xlim=c(0,0.3),ylim=c(0,0.3))
 abline(0,1)
@@
 ```{r}
-(soling_diff <- soling_after + (-1)*soling)
+(soling_diff <- soling_after - soling)
 ```
 
-So `H2` is the _difference_ between `H`--- which is `soling` plus the
-observations from the roundrobins and quarter finals---and `soling`
-withouth the extra information.  In other words, `H2` is the simply
-the information in the roundrobins and quarter finals.
+So `soling_diff` is the _difference_ between `H`--- which is `soling`
+plus the observations from the roundrobins and quarter finals---and
+`soling` without the extra information.  In other words, the
+information in the roundrobins and quarter finals.
 

### 2021-02-01 R/hyper2.R

@@ -179,13 +179,15 @@ setGeneric("pnames<-",function(x,value){standardGeneric("pnames<-")})
     stop("Unary operator '", .Generic, "' is not implemented for hyper2 objects")
     }
 
-  if (!is.element(.Generic, c("+", "-", "==", "!=", "*", "^" ))){
+  if (!is.element(.Generic, c("+", "==", "!=", "*", "^" ))){
     stop("Binary operator '", .Generic, "' is not implemented for hyper2 objects")
   }
   
   if(lclass && rclass){  
     if (.Generic == "+"){
       return(hyper2_add(e1,e2))
+    } else if (.Generic == "-"){
+      return(hyper2_add(e1,hyper2_prod(e2,-1)))
     } else if (.Generic == "==") {
       return(hyper2_equal(e1, e2))
     } else if (.Generic == "!=") {

### 2021-02-01 inst/cook.Rmd
fixes #75

@@
 library(hyper2)
-H <- hyper2(pnames=paste("p",1:5,sep=""))
-H
+
 ```
@@
 ```{r data}
-H <- H +
-    rank_likelihood(c(5,2,3,4)) + # person 1 says 5 is the best, 4 the worst.
-    rank_likelihood(c(1,5,3,4)) + # person 2 says 1 is the best, 4 the worst.
-    rank_likelihood(c(1,4,2,5)) + # person 3 says 1 is the best, 5 the worst.
-    rank_likelihood(c(1,2,3,5)) + # person 4 says 1 is the best, 5 the worst.
-    rank_likelihood(c(2,1,3,4))   # person 5 says 2 is the best, 4 the worst.
+
+H <- hyper2() + 
+  race(c("p5","p2","p3","p4")) + # person 1 says p5 is the best, p4 the worst.
+  race(c("p1","p5","p3","p4")) + # person 2 says p1 is the best, p4 the worst.
+  race(c("p1","p4","p2","p5")) + # person 3 says p1 is the best, p5 the worst.
+  race(c("p1","p2","p3","p5")) + # person 4 says p1 is the best, p5 the worst.
+  race(c("p2","p1","p3","p4"))   # person 5 says p2 is the best, p4 the worst.
 H
@@
 ```{r}
-L1 <- loglik(indep(equalp(H)),H)  # support at equal strengths
-L2 <- loglik(indep(maxp(H)),H)    # maximum support
-L2-L1
-pchisq(2*(L2-L1),df=1,lower.tail=FALSE) # p-value of asymptotic distribution
+equalp.test(H)
 ```
@@
 of the five.  We can follow the reasoning in the _icons_
-demonstration:
+demonstration and test $H_0\colon p_1=p_2$:
 
 ```{r}
-o <- function(Ul,Cl,startp,give=FALSE){
-    small <- 1e-4  #  ensure start at an interior point
-    if(missing(startp)){startp <- small*(1:4)+rep(0.1,4)}
-    out <- maxp(H, startp=small*(1:4)+rep(0.1,4), give=TRUE, fcm=Ul,fcv=Cl)
-		if(give){
-			return(out)
-		} else {
-			return(out$value)
-		}
-}
-
-p2max <- o(c(-1, 1, 0, 0), 0)
-p3max <- o(c(-1, 0, 1, 0), 0)
-p4max <- o(c(-1, 0, 0, 1), 0)
-p5max <- o(c(-2,-1,-1,-1),-1)
+samep.test(H,c("p1","p2"))
 ```
 
-(the final line is different because `p5` is the fillup value).
-
-```{r}
-(likes <- c(p2max,p3max,p4max,p5max))
-max(likes)
-```
-
-Observe that the hypothesis with the maximum likelihood is
-$p_2\geqslant p_1$, as might be expected on the grounds that in the
-unconstrained case we have
-$\hat{p_2}>\max\left(\hat{p_3},\hat{p_4},\hat{p_5}\right)$.  Now
-compare the constrained maximum with the unconstrained maximum:
-
-```{r}
-maxp(H,give=TRUE)$value
-```
-
-So the support loss by the constraint is
-
-```{r}
-(L <- maxp(H,give=TRUE)$value - max(likes))
-```
-
-short of Edwards's two-units-of-support criterion.  There is no strong
-evidence to support the assertion that person 1 is actually stronger
-than person 2 in the sense that $p_1>p_2$.
+There is no strong evidence to support the assertion that person 1 is
+actually stronger than person 2 in the sense that $p_1>p_2$.
 
@@
 
-Suppose that we subsequently observe order statistic `5,4,3,2,1`, that
-is, person 5 is the best, 4 is the second best, and so on to 1 being
-the worst.  Is this observation consistent with the previous dataset?
+Suppose that we subsequently observe order statistic `p5,p4,p3,p2,p1`,
+that is, person 5 is the best, 4 is the second best, and so on to 1
+being the worst.  Is this observation consistent with the previous
+dataset?
 
@@
 library(partitions)
+M <- perms(5)
+M[] <- paste("p",M,sep="")
 imH <- indep(maxp(H))
-f <- function(o){loglik(imH,rank_likelihood(o),log=FALSE)}
+f <- function(o){loglik(imH,race(o),log=FALSE)}
 LL <- apply(perms(5),2,f)
@@
 Hbefore <- H +
-    rank_likelihood(c(5,2,3,4,1)) + #  5 is the best, 4 the worst.
-    rank_likelihood(c(5,3,2,4,1)) + #  5 is the best, 1 the worst.
-    rank_likelihood(c(4,5,3,2,1)) + #  4 is the best, 1 the worst.
-    rank_likelihood(c(5,3,4,1,2)) + #  5 is the best, 2 the worst.
-    rank_likelihood(c(2,5,3,1,4))   #  2 is the best, 4 the worst.
+  race(c("p5","p2","p3","p4","p1")) + #  5 is the best, 4 the worst.
+  race(c("p5","p3","p2","p4","p1")) + #  5 is the best, 1 the worst.
+  race(c("p4","p5","p3","p2","p1")) + #  4 is the best, 1 the worst.
+  race(c("p5","p3","p4","p1","p2")) + #  5 is the best, 2 the worst.
+  race(c("p2","p5","p3","p1","p4"))   #  2 is the best, 4 the worst.
 Hbefore
@@
 
-Now, after the training we have two order observations: `1,2,3,4,5`
-and `3,1,5,2,4`.  Thus in the first observation, player 1 comes first
-and in the second observation, player 1 comes second second (after
-player 3).  The likelihood for the first observation would be
+Now, after the training we have two finishing order observations:
+`p1,p2,p3,p4,p5` and `p3,p1,p5,p2,p4`.  Thus in the first observation,
+`p1` comes first and in the second observation, `p1` comes second
+second (after `p3`).  The likelihood for the first observation would
+be
 
@@
 ```{r,usepwafortraining}
-Hafter <- rank_likelihood(1:5)+rank_likelihood(c(3,1,5,2,4))
-pnames(Hafter) <- pnames(Hbefore)
+Hafter <- race(c("p1","p2","p3","p4","p5")) + race(c("p3","p1","p5","p2","p4"))
 Hafter %<>%  pwa("p1")

### 2021-02-02 inst/volvo.Rmd
consistentification

---
title: "Race results from the twelfth edition of the
  round-the-world Volvo Ocean Race"
output: bookdown::html_document2
---

```{r setup, include=FALSE}
set.seed(0)
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("magrittr")
options("digits" = 5)
```

This short document creates the `volvo` suite of R objects and
presents some preliminary analysis.

```{r readvolvotable}
volvo_table <- read.table("volvo.txt")
volvo_table
```

Thus the first column refers to leg 1, in which competitor AbuDhavi
came first, Brunel third, and so on.  Here, `DNF` signifies "did not
finish" and `DNS` "did not start".  We can conver the table to a
support function with `ordertable2supp()`, remembering to specify the
noscore options:

```{r}
volvo <- ordertable2supp(volvo_table,noscore=c("DNF","DNS"))
(volvo_maxp <- maxp(volvo))
```

```{r volvomaxp, fig.cap="Pie chart of competitors' strengths"}
pie(volvo_maxp)
```

Figure \@ref(fig:volvomaxp) shows the maximum likelihood strengths
visually, and we may assess the null of equal strengths:

```{r,equalvolvostrengths}
equalp.test(volvo)
```

## References {-}

Wikipedia contributors, 2019.  \dQuote{2014-2015 Volvo Ocean Race}.
In \emph{Wikipedia, the free encyclopedia}.  Retrieved 22:21, February
28, 2020.
\url{https://en.wikipedia.org/w/index.php?title=2014\%E2\%80\%932015_Volvo_Ocean_Race&oldid=914916131},

### Package dataset {-}

Following lines create `volvo.rda`, residing in the `data/` directory of the package.

```{r,label=savecurlingdataset}
save(volvo_table,volvo_maxp,volvo,file="volvo.rda")
```

### 2021-02-02 data/volvo.R

@@ -1,45 +1,0 @@
-`volvo2014` <- 
-structure(list(brackets = list("AbuDhabi", c("AbuDhabi", "Alvimedica", 
-"Brunel", "Dongfeng", "MAPFRE"), c("AbuDhabi", "Alvimedica", 
-"Brunel", "Dongfeng", "MAPFRE", "SCA", "Vestas"), c("AbuDhabi", 
-"Alvimedica", "Brunel", "Dongfeng", "MAPFRE", "Vestas"), c("AbuDhabi", 
-"Alvimedica", "Brunel", "Dongfeng", "SCA", "Vestas"), c("AbuDhabi", 
-"Alvimedica", "Brunel", "MAPFRE", "SCA", "Vestas"), c("AbuDhabi", 
-"Alvimedica", "Dongfeng", "MAPFRE", "SCA", "Vestas"), c("AbuDhabi", 
-"Alvimedica", "Dongfeng", "SCA", "Vestas"), c("AbuDhabi", "Alvimedica", 
-"MAPFRE", "SCA", "Vestas"), c("AbuDhabi", "Brunel", "Dongfeng", 
-"MAPFRE", "SCA", "Vestas"), c("AbuDhabi", "Dongfeng", "MAPFRE", 
-"SCA", "Vestas"), c("AbuDhabi", "Dongfeng", "SCA", "Vestas"), 
-    c("AbuDhabi", "SCA", "Vestas"), "Alvimedica", c("Alvimedica", 
-    "Brunel", "Dongfeng"), c("Alvimedica", "Brunel", "Dongfeng", 
-    "MAPFRE"), c("Alvimedica", "Brunel", "Dongfeng", "MAPFRE", 
-    "SCA", "Vestas"), c("Alvimedica", "Brunel", "Dongfeng", "SCA", 
-    "Vestas"), c("Alvimedica", "Brunel", "MAPFRE", "SCA", "Vestas"
-    ), c("Alvimedica", "Brunel", "SCA", "Vestas"), c("Alvimedica", 
-    "Dongfeng"), c("Alvimedica", "MAPFRE", "SCA"), c("Alvimedica", 
-    "MAPFRE", "SCA", "Vestas"), c("Alvimedica", "SCA", "Vestas"
-    ), "Brunel", c("Brunel", "Dongfeng", "SCA", "Vestas"), c("Brunel", 
[…]

### 2021-02-02 inst/retain.Rmd
bugfix

@@
 ```{r}
-keep_flawed(OT,1:2)
+keep_flawed(OT, c("a","b"))
 ```
@@
 Above, function `keep_flawed()` has effectively taken `OT`, set `c=0`,
-and then discarded the `c^5` term.  This process, while natural, has
-the effect of replacing `(a+c)^-4` with `a^-4`, which has no natural
-probabilistic interpretation: we are ignoring `c`'s victories but
-nevertheless interpreting joint strength term on the denominator [viz
-`a+c`] as a *loss* for `a`.
+and then discarded the `c^5` term.  This process, while natural from a
+computing perspective, has the effect of replacing `(a+c)^-4` with
+`a^-4`, which has no natural probabilistic interpretation: we are
+ignoring `c`'s victories but nevertheless interpreting the joint
+strength term on the denominator [viz `a+c`] as a *loss* for `a`.
 
@@
 ```{r}
-keep_flawed(OT,1)
+keep_flawed(OT,"a")
 ```

### 2021-02-02 inst/cook.Rmd
bugfix

@@
 ```{r,beforetraining}
-H <- hyper2(pnames=c(paste("p",1:5,sep="")))  # 'S' the strength of training
 # observations before training:
-Hbefore <- H +
+Hbefore <- 
   race(c("p5","p2","p3","p4","p1")) + #  5 is the best, 4 the worst.
@@
 ```{r,addbeforeandafter}
-pnames(Hbefore) <- c(pnames(Hbefore),"S")
 maxp(Hbefore+Hafter)

### 2021-02-02 inst/order.Rmd
fixes #39
bugfix

@@
 library("hyper2")
-H <- order_likelihood(rbind(1:10))
+H <- race(letters[1:10])
 H
@@
 if we give the system a hand by starting at a point closer to the
-evaluate we can do much better:
+evaluate we can do slightly better:
 

### 2021-02-02 inst/moto.Rmd
typo

@@
 
-(in figure \@ref(fig:dotchartlogmot) observe that the bottom-ranked
+(in figure \@ref(fig:dotchartlogmoto) observe that the bottom-ranked
 player, Lecuona, has a ML strength of zero but the estimate is

### 2021-02-03 inst/moto.Rmd
typo

--- 2021-01-29 inst/moto.Rmd
@@
 
-(in figure \@ref(fig:dotchartlogmot) observe that the bottom-ranked
+(in figure \@ref(fig:dotchartlogmoto
+) observe that the bottom-ranked
 player, Lecuona, has a ML strength of zero but the estimate is

### 2021-02-03 inst/tennis.Rmd
consistentify tennis dataset

---
title: "Tennis doubles and ghosts"
output: bookdown::html_document2
---

```{r setup, include=FALSE}
set.seed(0)
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("magrittr")
options("digits" = 5)
```

This short document defines and analyses a small dataset originally
considered by Hankin (2010).  There, results from tennis games among
four players was considered:

```
       {p1,p2} vs {p3,p4}   9-2  
       {p1,p3} vs {p2,p4}   4-4  
       {p1,p4} vs {p2,p3}   6-7  
          {p1} vs {p3}     10-14 
          {p2} vs {p3}     12-14 
          {p1} vs {p4}     10-14 
          {p2} vs {p4}     11-10 
          {p3} vs {p4}     13-13

```

It is suspected that $p_1$ and $p_2$ have some form of team cohesion
and play better when paired than when either solo or with other
players.  As the scores show, each player and, apart from the first
line, each doubles partnership is of approximately the same strength.

We will create two likelihood functions: a regular one that assumes
the stength of a doubles partnership is the sum of the strength of the
players; and one that accounts for team cohesion.

```{r}
H <- hyper2()

## 1&3 vs 2&4, scoreline 4-4:
H[c("p1","p3")] %<>% inc(4)
H[c("p2","p4")] %<>% inc(4)
H[c("p1","p2","p3","p4")] %<>% dec(8)

## 1&4 vs 2&3, scoreline 6-7:
H[c("p1","p4")] %<>% inc(6)
H[c("p2","p3")] %<>% inc(7)
H[c("p1","p2","p3","p4")] %<>% dec(13)

## 1 vs 3, scoreline 10-14:
H["p1"] %<>% inc(10)
H["p3"] %<>% inc(14)
H[c("p1","p3")] %<>% dec(24)

## 2 vs 3, scoreline 12-14:
H["p2"] %<>% inc(12)
H["p3"] %<>% inc(14)
H[c("p2","p3")] %<>% dec(26)

## 1 vs 4, scoreline 10-14:
H["p1"] %<>% inc(10)
H["p4"] %<>% inc(14)
H[c("p1","p4")] %<>% dec(24)

## 2 vs 4, scoreline 11-10:
H["p2"] %<>% inc(11)
H["p4"] %<>% inc(10)
H[c("p2","p4")] %<>% dec(21)

## 3 vs 4, scoreline 13-13:
H["p3"] %<>% inc(13)
H["p4"] %<>% inc(13)
H[c("p3","p4")] %<>% dec(26)
```

Thus `H` is a hyper2 object corresponding to results except the first
line.  We include this in two ways.  Firstly, treating the first
scoreline like the others:

```{r}
tennis <- H

## 1&2 vs 3&4, scoreline 9-2:
tennis[c("p1","p2")] %<>% inc(9)
tennis[c("p3","p4")] %<>% inc(2)
tennis[c("p1","p2","p3","p4")] %<>% dec(11)

tennis
```

And secondly, accounting for any possible team cohesion by way of a
ghost player `G`:

```{r}
tennis_ghost <- H

## 1&2 vs 3&4 (NB: includes ghost!), scoreline 9-2 (again):
tennis_ghost[c("p1","p2","G")] %<>% inc(9)
tennis_ghost[c("p3","p4")] %<>% inc(2)
tennis_ghost[c("p1","p2","p3","p4","G")] %<>% dec(11)

tennis_ghost
```

standard evaluates:

```{r}
(tennis_maxp <- maxp(tennis))
(tennis_ghost_maxp <- maxp(tennis_ghost))
```

First test the null of equal strength:

```{r}
equalp.test(tennis)
```

not significant.  Is the ghost real?

```{r}
specificp.test(tennis_ghost,"G",0.0001)  # nonzero value needed to avoid triggering known R bug
```

so we can infer that the ghost is indeed real as we reject $H_0\colon
p_G=0$.

```{r}
(hess1 <- hessian(tennis,indep(tennis_maxp),border=TRUE))
(hess2 <- hessian(tennis,indep(tennis_maxp),border=FALSE))
eigen(hess2,TRUE,TRUE)
```

### References {-}

* RKS Hankin 2010. "A generalization of the Dirichlet distribution". 
 _Journal of Statistical Software_ 33(11)
* RKS Hankin 2017. "Partial rank data with the hyper2 package:
  likelihood functions for generalized Bradley-Terry models".  _The R
  Journal_ 9(2)

### Package dataset {-}

Following lines create `tennis.rda`, residing in the `data/` directory of the package.

```{r,label=savecurlingdataset}
save(tennis,tennis_maxp,tennis_ghost, tennis_ghost_maxp,file="tennis.rda")
```

### 2021-02-04 inst/tennis.Rmd
better docs

@@
 
+We can use the Hessian matrix:
+
 ```{r}
-(hess1 <- hessian(tennis,indep(tennis_maxp),border=TRUE))
-(hess2 <- hessian(tennis,indep(tennis_maxp),border=FALSE))
-eigen(hess2,TRUE,TRUE)
+(M <- hessian(tennis,indep(tennis_maxp),border=TRUE))
 ```
+
+See how the fillup value `p4` takes up the unit sum constraint [it is
+defined as $p_4=1-p_1-p_2-p_3$].  Matrix `M` should be
+negative-semidefinite:
+
+```{r}
+eigen(M,TRUE,TRUE)$values # one value is zero to numerical precision; the rest should be <0
+```
+
+See how one eigenvalue is zero (to numerical precision) and the rest
+are strictly negative, corresponding to the support function having a
+well-defined maximum.
 

### 2021-02-04 inst/volleyball.Rmd
consistentify volleyball

---
title: "volleyball dataset"
output: bookdown::html_document2
---

```{r setup, include=FALSE}
set.seed(0)
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("magrittr")
options("digits" = 5)
```

This short document discusses a dataset first presented by Hankin
(2010), although here only the first 52 (out of 4000) observations are
used.

```{r readcurlingtable}
volleyball_table <- as.matrix(read.table("volleyball.txt",header=TRUE))
nrow(volleyball_table)
head(volleyball_table)
```

The first row shows a game between a team comprising `p1`, `p4`, and
`p8` against a team comprising `p5` and `p6`.  Team `p1 p4 p8` won.
We may use function `volley()` to convert this to a likelihood
function:

```{r}
volleyball <- volley(volleyball_table)
(volleyball_maxp <- maxp(volleyball))
```

The original synthetic dataset was prepared using Zipf's law for the
players' strengths, so we may test the hypothesis that this is the
case; $H_0\colon p_i\propto i^{-1}$:

```{r}
zipf <- function(n){jj <- 1/(1:n); jj/sum(jj)}
zipf(9)
(null_support <- loglik(indep(zipf(9)),volleyball))
(alternative_support <- loglik(indep(volleyball_maxp),volleyball))
(Lambda <- 2*(alternative_support-null_support))
pchisq(Lambda,df=8,lower.tail=FALSE)
```

somewhat disappointingly rejecting the null with a $p$-value of about
4\% (and indeed--just---with a two units of support per degree of
freedom criterion).  However, it is not clear to me the extent to
which Wilks's theorem is applicable here [Wilks is an asymptotic
result; recall that we have only 52 observations here], nor whether
the support criterion is appropriate with 8 degrees of freedom.

### Package dataset {-}

Following lines create `volleyball.rda`, residing in the `data/` directory of the package.

```{r,label=savecurlingdataset}
save(volleyball_table,volleyball_maxp,volleyball,file="volleyball.rda")
```

### 2021-02-04 inst/gradient.Rmd
fixes #79

@@
 library("hyper2",quietly=TRUE)
+set.seed(0)
 ```
@@
 ```
+
+## Case study: icons
+
+Consider the `icons` likelihood function; we wonder whether the
+loglikelihood function has a well-defined maximum.
+
+```{r}
+(M <- hessian(icons))
+is_ok_hessian(M)
+```
+
+This would suggest that the maximum is defined; we may ask how sharp
+it is:
+
+```{r}
+is_ok_hessian(M,give=TRUE)
+```
+
+suggesting a reasonably sharp constrained maximum point.  We now
+demostrate that the Hessian matrix of second derivatives is
+numerically correct.  First the unconstrained problem, working with
+the first five elements.  The first step is to look at the likelihood,
+the gradient, and the Hessian, all evaluated at a certain point:
+
+```{r}
+(p <- indep(equalp(icons))) # NB not the evaluate
+(G <- gradient(icons,p))
+(M <- hessian(icons,p,border=FALSE))
+eigen(M,TRUE,TRUE)$values
+```
+
+```{r}
+dp <- rnorm(5)*1e-4        # small perturbation
+dL <- loglik(p+dp,icons) - loglik(p,icons)  # delta loglikelihood (exact)
+
+dL1 <- sum(gradient(icons,p)*dp)    # first order approximation
+dL2 <- dL1 + t(dp) %*% M %*% dp/2   # second order; should be quad.form(dp,M)
+
+c(dL,dL1,dL2)
+
+c(first_order_error=dL-dL1,second_order_error=dL-dL2)
+
+```
+
+In the above, see how the first order approximation is pretty good: it
+is of ${\mathcal O}(f''(x)\delta x^2)$ which here would be about
+$1000\times 10^{-8}=10^{-5}$ which is pretty much what we observe [the
+1000 is the typical size of the elements of the Hessian].  The
+second-order approximation is better, having a smaller error ... but
+not that much better.  I would expect the error to be ${\mathcal
+O}(f'''(x)\delta x^3)$ but maybe the third order derivatives are
+larger than one might expect.  We can perform a similar analysis but
+using bordered Hessians:
+
+```{r}
+
+dpc <- fillup(dp,0)   # constrained delta p
+
+MC <- hessian(icons,p,border=TRUE)
+dLC1 <- sum(gradientn(icons,fillup(p))*dpc)    # first order approximation
+dLC2 <- dLC1 + t(dpc) %*% MC %*% dpc/2     # second order
+
+c(dL,dLC1,dLC2)
+
+c(first_order_error=dL-dLC1,second_order_error=dL-dLC2)
+
+```
+
+showing similar results.

### 2021-02-04 inst/volleyball.Rmd
consistentification of volleyball

@@
 This short document discusses a dataset first presented by Hankin
-(2010), although here only the first 52 (out of 4000) observations are
-used.
+(2010), although here only the first 52 observations are used.  A
+volleyball \dfn{set} is a Bernoulli trial between two disjoint subsets
+of the players.  The two subsets are denoted (after the game) as the
+_winners_ and the _losers_: these are denoted by `1` and `0`
+respectively.
 
@@
 
-The first row shows a game between a team comprising `p1`, `p4`, and
-`p8` against a team comprising `p5` and `p6`.  Team `p1 p4 p8` won.
+Each row of `volleyball_table` is a set.  Thus the first line shows a
+game between a team comprising `p1`, `p4`, and `p8` against a team
+comprising `p5` and `p6`.  Player `p9` did not play; team `p1 p4 p8`
+won.
+
 We may use function `volley()` to convert this to a likelihood

### 2021-02-04 data/volleyball.R

@@ -1,96 +1,0 @@
-`volleyball_matrix` <- matrix(c(
- 1, 0,NA, 1, 0, 0,NA, 1,NA,
-NA,NA, 1, 1, 0, 1, 0, 0,NA,
-NA,NA, 1, 1, 0,NA, 1, 0,NA,
-NA, 1, 1, 0, 0,NA, 1, 1,NA,
- 1, 1, 1, 0, 0, 0,NA,NA,NA,
- 1, 1, 0,NA, 1, 1, 0, 0, 0,
- 0, 1, 0,NA, 1, 1, 0, 1, 0,
- 1, 1, 0,NA, 0, 1, 0, 0, 0,
- 1, 0, 1,NA, 0, 0, 1, 1, 0,
-NA,NA,NA,NA,NA, 0, 1, 1, 0,
- 0,NA, 1, 1, 0, 0, 1, 1, 0,
-NA,NA, 1, 0, 1, 0, 1, 1, 0,
-NA,NA, 1, 0, 1, 0, 1, 1, 0,
- 0, 1, 0, 1, 0,NA,NA,NA, 0,
- 1, 0, 0, 1, 1,NA,NA,NA, 0,
- 1, 0, 1, 0, 1, 0, 0, 1, 0,
- 1, 0, 1, 0, 1, 0, 0, 0, 0,
- 1, 0, 1, 0, 1, 0, 0, 1, 0,
- 1, 0, 1,NA,NA, 0, 0, 1, 0,
- 1, 0, 1,NA, 1, 0, 0, 1, 0,
- 1, 0, 1, 0, 1, 0,NA, 0, 1,
- 1, 0, 1, 0, 1, 0, 0,NA, 0,
-NA,NA,NA, 0,NA,NA, 0, 0, 1,
-NA,NA,NA, 0,NA,NA, 0, 0, 1,
-NA, 0,NA,NA,NA, 0, 0, 1, 1,
-NA,NA, 1, 0, 1, 0, 1, 1, 0,
-NA, 1,NA,NA,NA, 1, 0, 0, 1,
- 1, 1, 0, 0, 1, 0, 0, 0,NA,
- 1, 0, 0, 0, 1, 0, 0, 1, 1,
- 1, 0,NA, 0, 1, 0, 0, 1, 1,
- 1, 0,NA, 0, 1, 0, 0,NA, 1,
- 1, 0,NA, 1, 0, 0, 0,NA, 1,
[…]

### 2021-02-05 inst/masterchef.Rmd
consistentification

---
title: "Masterchef"
output: html_document
---

```{r setup, include=FALSE}
set.seed(0)
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("magrittr")
options("digits" = 5)
```

This document creates objects `masterchef`, `masterched_maxp`, and
`materchef_constrained_maxp`.

```{r}
do_from_scratch <- FALSE
```

```{r}

Hall <-
    hyper2(pnames=c("Amy", "Ben", "Brendan", "Brent", "Byron",
                    "Cecilia", "Colin", "Deepali", "Emelia", "Emily",
                    "Georgia", "Jamie", "Kira", "Laura", "Nick",
                    "Nicole", "Rachael", "Renae", "Sam", "Sarah",
                    "Scott", "Sean", "Steven", "Tash", "Tracy"))  # NB alphabetical order

# players left in after L7b

H <- hyper2(pnames= c("Amy", "Ben", "Brent", "Colin", "Emelia",
                      "Georgia", "Jamie", "Kira", "Laura", "Renae",
                      "Sarah", "Tash", "Tracy"))

allplayers <-
    c("Brent","Laura","Emelia","Jamie","Tracy","Ben","Amy", "Renae",
      "Sarah","Colin","Kira","Georgia","Tash", "Byron","Sam",
      "Rachael","Steven","Sean","Scott", "Emily","Nick","Nicole",
      "Deepali","Brendan","Cecilia"
      )  # NB game order: Brent won, Laura runner-up, through to
         # Cecilia, who withdrew

 ## variable 'doo' is a Boolean, with entries governing whether a
 ## particular round is included in L or not.

doo <- c(
    L1   = FALSE,
    L2a  = FALSE,
    L2b  = FALSE,
    L2c  = FALSE,
    L3a  = FALSE,
    L3b  = FALSE,
    L3c  = FALSE,
    L4a  = FALSE,
    L4b  = FALSE,
    L4c  = FALSE,
    L5a  = FALSE,
    L5b  = FALSE,
    L6a  = FALSE,
    L6b  = FALSE,
    L7a  = FALSE,
    L7b  = TRUE,
    L8a  = TRUE,
    L8b  = TRUE,
    L8c  = FALSE,  # second chance, can't include this with n=13 players
    L8d  = TRUE,
    L8e  = TRUE,
    L9a  = TRUE,
    L9b  = TRUE,
    L10a = TRUE,
    L10b = TRUE,
    L10c = TRUE,
    L11a = TRUE,
    L11b = TRUE,
    L11c = TRUE,
    L11d = TRUE,
    L11e = TRUE,
    L11f = TRUE,
    L12a = TRUE,
    L12b = TRUE,
    L12c = TRUE,
    L12d = TRUE,
    L12e = TRUE,
    L13a = TRUE,
    L13b = TRUE
    )

L <- list()  # overall list

## If a date is given in the comments to a multi-stage elimination
## dataset, then the date refers to the date on which a player was
## actually eliminated.  Note that a likelihood function may also
## include selection of contestants for pressure tests.

if(doo[["L1"]]){
L$week1 <-  # 8 May 2014
ggrl(H,
     top3       = c("Laura","Jamie","Sean"),
     btm21      = c("Emelia", "Tracy", "Sarah", "Colin", "Kira",
                    "Georgia", "Tash", "Byron", "Scott", "Emily",
                    "Nick", "Nicole"),
     btm9       = c("Brent", "Ben", "Amy", "Renae", "Sam", "Steven"),
     btm3       = c("Rachael", "Deepali"),
     eliminated = c("Brendan")
     )
}

if(doo[["L2a"]]){
L$week2a <- # elimination 4 chosen 11 May; Deepali eliminated 12 May 2014
ggrl(H,
            win = c("Sarah"),
             IN = c("Brent","Laura","Emelia","Tracy","Ben","Amy","Renae",
                    "Colin","Kira","Georgia","Tash","Byron","Sam","Rachael",
                    "Steven","Sean","Emily","Nicole"),
           btm4 = c("Jamie","Scott","Nick"),
     eliminated = c("Deepali")
     )

}

if(doo[["L2b"]]){  
L$week2b <- H  # 13 May 2014
L$week2b[c(    #  winning team (11 players)
    "Emelia","Jamie","Ben","Renae","Sarah","Colin","Kira","Georgia",
    "Sam","Sean","Nick"
)] <- 1

L$week2b[c( # winning team union losing team; 22 players
    "Brent","Laura","Emelia","Jamie","Tracy","Ben","Amy","Renae",
    "Sarah","Colin","Kira","Georgia","Tash","Byron","Sam","Rachael",
    "Steven","Sean","Scott","Emily","Nick","Nicole"
)] <- -1
}

if(doo[["L2c"]]){  # 14 May
L$week2c <- ggrl(H,  # elimination, field of losing team from 13 May.
            team_lose  = c("Laura","Amy","Tash","Rachael","Emily"),
            btm6       = c("Brent","Tracy","Byron","Steven","Scott"),
            eliminated = c("Nicole")
            )
}

if(doo[["L3a"]]){
L$week3a <-  # Nick eliminated 19 May, 
    ggrl(H,
         top3       = c("Laura","Tracy","Amy"),
         IN         = c("Brent","Emelia","Jamie","Ben","Renae","Colin",
                        "Kira","Georgia","Tash","Byron","Sam","Rachael",
                        "Steven","Scott","Emily"),
         btm3       = c("Sarah","Sean"),
         eliminated = c("Nick")
         )
}

if(doo[["L3b"]]){
L$week3b <- H # team challenge 21 May
L$week3b[c(  # Blue team wins:
    "Laura","Amy","Renae","Sarah","Colin","Kira",
    "Tash","Byron","Rachael","Scott"
    )] <- 1

L$week3b[c(  # winning team union losing team
    "Brent","Laura","Emelia","Jamie","Tracy","Ben", "Amy","Renae",
    "Sarah","Colin","Kira","Georgia", "Tash","Byron","Sam","Rachael",
    "Steven","Sean","Scott","Emily")] <- -1
}

if(doo[["L3c"]]){
L$week3c <- ggrl(H, # Emily eliminated 22 May
            team_lose  = c("Emelia","Tracy","Ben","Sam"),
            btm6       = c("Brent","Jamie","Georgia"),
            btm3       = c("Steven","Sean"),
            eliminated = c("Emily")
            )
}

if(doo[["L4a"]]){
L$week4a <- ggrl(H, # Scott eliminated 26 May
            top3       = c("Brent","Sarah","Tash"),
            IN         = c("Laura","Emelia","Tracy","Ben","Amy","Renae",
                           "Colin","Kira","Georgia","Byron","Sam",
                           "Rachael","Sean"),
            btm3       = c("Jamie","Steven"),
            eliminated = c("Scott")
            )
}

if(doo[["L4b"]]){
L$week4b <- H  # 
L$week4b[c(  # team challenge 28 May; Blue team wins
    "Laura","Emelia","Jamie","Tracy","Ben","Renae","Sarah",
    "Byron","Steven"
)] <- 1

L$week4b[c(  # winning team union losing team
    "Brent","Laura","Emelia","Jamie","Tracy","Ben","Amy",
    "Renae","Sarah","Colin","Kira","Georgia","Tash",
    "Byron","Sam","Rachael","Steven","Sean"
)] <- -1
}

if(doo[["L4c"]]){ # Sean eliminated 29 May; losing team chosen 28 May
L$week4c <- ggrl(H,
             team_lose  = c("Amy","Colin","Georgia",
                           "Tash","Sam","Rachael"),
             btm3       = c("Brent","Kira"),
             eliminated = c("Sean")
            )
}

if(doo[["L5a"]]){  # Steven eliminated 2 Jun; pressure test 1 Jun
L$week5a <- ggrl(H,
            top3       = c("Brent","Renae","Rachael"),
            IN         = c("Laura","Emelia","Jamie","Tracy",
                           "Amy","Colin","Kira","Georgia","Tash",
                           "Sam"),
            btm4       = c("Ben","Sarah","Byron"),
            eliminated = c("Steven")
            )
}

if(doo[["L5b"]]){
L$week5b <- H   
L$week5b[c(   # Team challenge  4 June; all players
    "Brent","Laura","Emelia","Jamie","Tracy",
    "Ben","Amy","Renae","Sarah","Colin","Kira",
    "Georgia","Tash","Byron","Sam","Rchael"
   )] <- -1

L$week5b[c(  
    "Tracy","Sarah","Kira","Sam"  # Yellow team wins
    )] <- 1

L$week5b[c(  # green team came second
    "Laura","Amy","Georgia","Byron"
)] <- 1

L$week5b[c(   # all players left after winning team sits out (denominator)
    "Brent","Laura","Emelia","Jamie",
    "Ben","Amy","Renae","Colin",
    "Georgia","Tash","Byron","Rachael"
)] <- -1

L$week5b[c( # third team
    "Emelia","Jamie","Renae","Colin"
)] <- 1

L$week5b[c(   # all players left after winning and second team sits
              # out (denominator)
    "Brent","Emelia","Jamie", "Ben","Renae","Colin",
    "Tash","Rachael"
   )] <- -1

L$week5b <- ggrl(L$week5b,  # Rachael eliminated 5 Jun
            btm4   = c("Brent","Ben","Tash"),
            eliminated = c("Rachel")
            )

}

if(doo[["L6a"]]){
L$week6a <- H  # Sam eliminated 9th Jun
L$week6a <- ggrl(H,
            top3       = c("Amy","Colin","Georgia"),
            IN         = c("Brent","Emelia","Jamie","Tracy",
                           "Ben","Renae","Sarah","Kira","Tash"),
            btm3       = c("Laura","Byron"),
            eliminated = c("Sam")
            )
}

if(doo[["L6b"]]){
L$week6b <- H

L$week6b[c(   # teams allocated randomly; red team wins 11 Jun
    "Laura","Ben","Amy","Renea","Kira","Georgia","Tash"
)] <- 1

L$week6b[c(   # all players (denominator)
    "Brent","Laura","Emelia","Jamie","Tracy","Ben","Amy","Renea","Sarah",
    "Colin", "Kira","Georgia","Tash","Byron"
)] <- 1

L$week6b <- ggrl(L$week6b,    # Sarah eliminated 12 June
            team_lose  = c("Brent","Emelia","Jamie","Byron"),
            btm3       = c("Tracy","Colin"),
            eliminated = c("Sarah")
            )
}

if(doo[["L7a"]]){  # Byron eliminated 16 Jun;  
L$week7a <- ggrl(H,
            top3       = c("Laura", "Tracy","Ben"),
            IN         = c("Brent","Emelia","Jamie","Renae",
                           "Colin","Kira","Georgia"),
            btm3       = c("Amy","Tash"),
            eliminated = c("Byron")
            )
}

 ## up to this point, following players are eliminated: Cecilia,
 ## Brendan, Deepali, Nicole, Nick, Emily, Scott, Sean, Steven,
 ## Rachael, Sam, Byron.

 ## Players left in are: Tash, Georgia, Kira, Colin, Sarah, Renae,
 ## Amy, Ben, Tracy, Jamie, Emelia, Laura, Brent

if(doo[["L7b"]]){
L$week7b <- H  # Team challenge 18 June; 
L$week7b[c(  # red team wins +=1
    "Jamie","Tracy","Ben","Amy","Renae","Georgia"
)] <- 1

L$week7b[c(  # all players -=1
    "Brent","Laura","Emelia","Jamie","Tracy",
    "Ben","Amy","Renae","Colin","Kira",  # NB no Sarah
    "Georgia","Tash"
    )] <- -1

L$week7b <- ggrl(L$week7b,  # Tash eliminated 19 Jun; no Laura (she used her immunity pin)
            team_lose  = c("Brent","Emelia","Colin","Kira"),
            eliminated = c("Tash")
    )
}

if(doo[["L8a"]]){
L$week8a <- ggrl(H,  # Georgia eliminated 23 June
            win       = c("Tracy"),
            top3      = c("Brent","Emelia","Jamie"),
            IN        = c("Ben","Colin","Kira"),
            btm3      = c("Laura","Amy","Renae"),  # Renae a special case;
            eliminated= c("Georgia")
            )
}

if(doo[["L8b"]]){

jj <- H  # team challenge 
jj[c(   # winners += 1
    "Brent","Emelia","Jamie","Amy","Renae"
)] <- 1

jj[c(   # all players -= 1
    "Brent","Laura","Emelia","Jamie","Tracy","Ben","Amy","Renae","Sarah","Colin","Kira"
)] <- -1

L$week8b <- list(jj)
}

if(doo[["L8c"]]){
jj <- H    # second chance; Sarah won so she is back in the contest
jj[c("Sarah")] <- 1
jj[c(
    "Sarah", "Georgia","Tash","Byron","Sam","Rachael","Steven",
    "Sean","Scott","Emily","Nick","Nicole","Deepali","Brendan",
    "Cecilia"
)] <- -1

L$week8c <- list(jj)
}

if(doo[["L8d"]]){
jj <- H     # team  challenge 25 Jun
jj[c(  # Blue team wins;  +=1
    "Brent","Emelia","Jamie","Ben","Amy", "Renae"
)] <- +1

jj[c(  # all players -= 1
    "Brent","Laura","Emelia","Jamie","Ben",
    "Amy","Renae","Sarah","Colin","Kira"
)] <- -1

L$week8d <- list(jj)
}

if(doo[["L8e"]]){
jj <- H    # Tracy wins power apron 26 Jun
jj[c("Tracy")] <- 1  # winner
jj[c("Brent","Emelia","Jamie","Tracy","Amy","Renae")] <- -1
L$week8e <- list(jj)
}

if(doo[["L9a"]]){    # Kira eliminated 30 Jun
L$week9a <- ggrl(H,  # NB: Tracy excluded due to the power apron
            IN         = c("Laura","Emelia","Amy","Renae","Sarah"),
            btm5       = c("Brent","Ben"),
            btm3       = c("Jamie","Colin"),
            eliminated = c("Kira")
            )
}

if(doo[["L9b"]]){
L$week9b <- H  # team challenge 2 Jul

L$week9b[c(  # all contestants -= 1
    "Brent","Laura","Emelia","Jamie","Tracy","Ben","Amy","Renae","Sarah","Colin"
    )] <- -1

L$week9b[c("Emelia","Jamie","Tracy","Ben","Sarah")] <- -1    # Red team wins; +=1

L$week9b <- ggrl(L$week9b,   # Colin eliminated 3 July
            team_lose = c("Brent","Laura","Amy"),
            btm2 = "Renae",
            eliminated = c("Colin")
            )
}

if(doo[["L10a"]]){  # Sarah eliminated 7 July
L$week10a <- ggrl(H,
             top3       = c("Laura","Ben","Amy"),
             IN         = c("Brent","Jamie"),
             btm4       = c("Renae"),
             btm3       = c("Emelia","Tracy"),
             eliminated = c("Sarah")
             )
}

if(doo[["L10b"]]){  # team challenge 9 July
jj <- H
jj[c("Laura","Jamie")] <- 1  # blue team wins; +=1
jj[c(                        # all players -=1
    "Brent","Laura","Emelia","Jamie",
    "Tracy","Ben","Amy","Renae"
)] <- -1

jj[c("Emelia","Amy")] <- 1   # yellow team came second
jj[c("Brent","Emelia","Tracy", # all remaining players sans blue
       "Ben", "Amy","Renae"
       )] <- -1

jj[c("Brent","Tracy")] <- 1 # green team came third
jj[c("Brent","Tracy","Ben","Renae")] <- -1 # remaining players

L$week10b <- list(jj)
}

if(doo[["L10c"]]){  # Renae eliminated 10 Jul
L$week10c <- ggrl(H,
             win        = c("Laura"),
             IN         = c("Brent","Tracy","Ben"),
             eliminated = c("Renae")
             )
}

if(doo[["L11a"]]){
L$week11a <- ggrl(H,  # Heston week, 13 July part 1
             top3=c("Laura","Jamie","Ben"),
             IN = c("Brent","Emelia","Tracy")
             )
}

if(doo[["L11b"]]){ # Heston week, 13 July part 2
L$week11b <- ggrl(H,
             win  = c("Brent"),
             IN   = c("Laura","Emelia","Jamie","Ben","Amy"),
             lose = c("Tracy")
             )
}

if(doo[["L11c"]]){ # Heston week, 14 July
L$week11c <- ggrl(H,
             win = c("Amy"),
             IN  = c("Brent","Laura","Emelia","Jamie"),
             lose = c("Ben")
             )
}

if(doo[["L11d"]]){
L$week11d <- ggrl(H,  # Heston week, 15 July
             win  = c("Amy"),
             IN   = c("Laura","Emelia","Jamie"),
             lose = c("Brent")
             )
}

if(doo[["L11e"]]){  # Heston week, 16 July
L$week11e <- ggrl(H,
             win  = c("Laura"),
             IN   = c("Emelia","Jamie"),
             lose = c("Amy")
             )
}

if(doo[["L11f"]]){
L$week11f <- ggrl(H,  # Amy eliminated 17 July
                    win = c("Laura"),
                   top2 = c("Jamie"),
                   btm4 = c("Brent","Tracy","Ben"),
             eliminated = c("Amy")
             )
}

if(doo[["L12a"]]){
L$week12a <- ggrl(H,  # Ben eliminated 20 July
                    win = c("Brent","Laura","Emelia","Tracy"),
                   btm2 = c("Jamie"),
             eliminated = c("Ben")
             )
}

if(doo[["L12b"]]){
L$week12b <- ggrl(H, # Tracy eliminated 21 July
                    win = c("Laura"),
                   top3 = c("Emelia","Jamie"),
                   btm2 = c("Brent"),
             eliminated = c("Tracy")
             )
}

if(doo[["L12c"]]){  # three-round duel challenge, 22 July
jj <- H
jj[c("Laura")] <- 1
jj[c("Brent","Laura")] <- -1

jj[c("Emelia")] <- 1
jj[c("Emelia","Jamie")] <- -1

jj[c("Laura")] <- powers(jj[c("Laura")]) + 1
jj[c("Emelia","Laura")] <- -1
L$week12c <- list(jj)
}

if(doo[["L12d"]]){  # 23 July?
L$week12d <- ggrl(H,
             win = c("Brent"),
             IN = c("Emelia","Jamie")
             )
}

if(doo[["L12e"]]){ # Jamie eliminated 24 Jul
L$week12e <- ggrl(H,
             btm3       = c("Brent","Emelia"),
             eliminated = c("Jamie")
             )
}

if(doo[["L13a"]]){ # Emelia eliminated 27 Jul
L$week13a <- ggrl(H,
                    win = c("Laura"),
                   btm2 = c("Brent"),
             eliminated = c("Emelia")
             )
}

if(doo[["L13b"]]){
jj <- H    # Final 28 July; Laura eliminated and Brent wins
jj[c("Brent")] <- 1
jj[c("Brent","Laura")] <- -1
L$week13b <- list(jj)
}

`rprop` <- function(n){
    out <- runif(n)
    out <- out/sum(out)
    out[-n]
}

n <- 13   # 13 players; now specify constraints:
UI <- rbind(diag(n-1),-1)
CI <- c(rep(0,n-1),-1)

startp <- function(n){rep(1/n,n-1)}

jj <- # hot-start
  c(0.108618206752482, 0.0745797046860904, 0.134355320426558,
    0.0281960563127382, 0.116976608268827, 6.85045511136332e-09,
    0.106541176599712, 0.0205579373409824, 0.275062072951276,
    0.0407064266223856, 0.0280389311011074, 1.14209384061803e-09)

if(do_from_scratch){
`masterchef_maxp` <-   # takes about an hour to run without hotstart
constrOptim(
    theta = jj,
    f = function(p){-like_series(p,L)},  # 'L' created sequentially above
    grad=NULL,
    ui = UI, ci=CI,
    control=list(trace=100,maxit=100000)
)
} else { # precomputed

`masterchef_maxp` <-	
structure(list(par = c(0.108618206752482, 0.0745797046860904, 
0.134355320426558, 0.0281960563127382, 0.116976608268827, 6.85045511136332e-09, 
0.106541176599712, 0.0205579373409824, 0.275062072951276, 0.0407064266223856, 
0.0280389311011074, 1.14209384061803e-09), value = 66.1965202262011, 
    counts = 0, convergence = 0L, message = NULL, outer.iterations = 1L, 
    barrier.value = 0.000214030311354918), .Names = c("par", 
"value", "counts", "convergence", "message", "outer.iterations", 
"barrier.value"))
}

## first implement the restriction that Brent >= Laura:
UI = rbind(UI,c(0,0,1,0,0,0,0,0,-1,0,0,0))  # Brent - Laura >= 0
CI <- c(CI,0)

## second, find a consistent starting value, swapping Laura's strength for that of Brent:
swap  <- jj[3]  # Brent == jj[3]
jj[3] <- jj[9]  # Laura == jj[9]
jj[9] <- swap

jj2 <- # hot-start
c(0.109667788153559, 0.0785631722927578, 0.188668727599298, 0.029077960904546, 
0.124873146008887, 2.63812282535037e-09, 0.115674972706005, 0.0213623680862838, 
0.188668726563236, 0.0421144946907343, 0.0307905358378776, 1.18098133183577e-09
)

if(do_from_scratch){
`masterchef_constrained_maxp` <- 
constrOptim(
    theta = jj2,
    f = function(p){-like_series(p,L)},  # 'L' created sequentially above
    grad=NULL,
    ui = UI, ci=CI,
    control=list(trace=100,maxit=10000000)
)
} else { # precomputed
`masterchef_constrained_maxp` <- 
structure(list(par = c(0.109667788153559, 0.0785631722927578, 
0.188668727599298, 0.029077960904546, 0.124873146008887, 2.63812282535037e-09, 
0.115674972706005, 0.0213623680862838, 0.188668726563236, 0.0421144946907343, 
0.0307905358378776, 1.18098133183577e-09), value = 67.3764248634724, 
    counts = 0, convergence = 0L, message = NULL, outer.iterations = 1L, 
    barrier.value = 0.000219352800726824), .Names = c("par", 
"value", "counts", "convergence", "message", "outer.iterations", 
"barrier.value"))
}

masterchef_maxp <- fillup(masterchef_maxp$par)
names(masterchef_maxp) <- pnames(H)

masterchef_constrained_maxp <- fillup(masterchef_constrained_maxp$par)
names(masterchef_constrained_maxp) <- pnames(H)

masterchef <- L

```

### Package dataset

Following lines create `masterchef.rda`, residing in the `data/` directory of the package.

```{r,label=savecurlingdataset}
save(masterchef,masterchef_maxp,masterchef_constrained_maxp,file="masterchef.rda")
```

### 2021-02-06 inst/powerboat.Rmd
consitentify powerboat

---
title: "powerboat"
output: html_document
---

```{r setup, include=FALSE}
set.seed(0)
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("magrittr")
options("digits" = 5)
```

[powerboat](https://en.wikipedia.org/w/index.php?title=2018_F1_Powerboat_World_Championship&oldid=920386507)
racing is an important and prestigious sport.  Here I analyse a
dataset taken from the 2018 F1 Powerboat world championships.

```{r readpowerboattable}
powerboat_table <- read.table("powerboat.txt",header=TRUE)
```

We need some data cleaning first; specifically a zero entry is treated
as "did not start":

```{r}
powerboat_table[powerboat_table==0] <- "DNS"
powerboat_table[20:21,8] <- 0  # above does not apply to points
powerboat_table
```

Each row is a driver and each column a venue (apart from the final
column, which is points scored).  Thus we see from the first column
that, in Portugal, Torrente came first, Stark second, T. AlQamzi
second, and so on.

In this form we can easily convert to a support function:

```{r,definsupportb,cache=TRUE}
powerboat <- ordertable2supp(powerboat_table[,1:7])
head(powerboat)
(powerboat_maxp <- maxp(powerboat))
pie(powerboat_maxp)
par(pty="s")
points <- as.numeric(powerboat_table[,8])
names(points) <- rownames(powerboat_table)
points
points <- points[order(names(points))]
points
plot(powerboat_maxp,points*max(powerboat_maxp)/max(points),xlab="likelihood",ylab="points",pch=16)
abline(0,1)
```

### Package dataset

Following lines create `powerboat.rda`, residing in the `data/`
directory of the package.

```{r,label=savepowerboatdataset}
save(powerboat_table,powerboat,powerboat_maxp,file="powerboat.rda")
```

### 2021-02-06 inst/rock_paper_scissors_monster.Rmd
consistentify chess

@@
 ```{r,chesstable}
-o <- matrix(c(22,NA,8,13,23,NA,NA,12,10),3,3)
-dimnames(o) <- list(
+chess_table <- as.matrix(read.table("chess.txt"))
+dimnames(chess_table) <- list(
     match=c("T_vs_A","A_vs_K","T_vs_K"),
     player=c("Topalov","Anand","Karpov"))
-o
+chess_table
 ```
@@
 ```{r,showchess}
+chess <- saffy(chess_table)
 chess
-maxp(chess)
+chess_maxp <- maxp(chess)
+chess_maxp
 ```
@@
 ```{r,binomtest}
-binom.test(sum(diag(o)),sum(o,na.rm=TRUE),alternative="greater")
+binom.test(sum(diag(chess_table)),sum(chess_table,na.rm=TRUE),alternative="greater")
 ```
@@
   Journal_, volume 9, number 2.
+
+### Package dataset
+
+Following lines create `chess.rda`, residing in the `data/` directory of the package.
+
+```{r,label=savechessdataset}
+save(chess_table,chess,chess_maxp,file="chess.rda")
+```

### 2021-02-06 data/chess.R

@@ -1,7 +1,0 @@
-`chess` <- 
-  structure(list(
-      brackets = list("Topalov",
-                      c("Topalov", "Anand"), c("Topalov", "Karpov"), "Anand", c("Anand", "Karpov"), "Karpov"),
-      powers = c(30, -35, -18, 36, -35, 22),
-      pnames   = c("Topalov", "Anand", "Karpov")),
-      class = "hyper2")

### 2021-02-06 inst/kka.Rmd
consistentify kka

---
title: "Karpov, Kasprov, Anand"
output: html_document
---

```{r setup, include=FALSE}
set.seed(0)
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("magrittr")
options("digits" = 5)
```

This file defines dataset `kka` which is the observed scorelines
between the three players.  It is used to calculate likelihood
functions: `karpov_kasparov_anand`, `kka_3whites` and `kka_3draws`
This Rmd file supercedes files `inst/karpov_kasparov_anand.R` and
`inst/kka_3draws.R` and `inst/kka_array.R`.  Some more detail is given
in `man/karpov_kasparov_anand.Rd`.

```{r setupkka}
kka <- c(
## Kasparov vs Karpov
karpov_plays_white_beats_kasparov = 18,  # 12 on p1, 6 on p2
kasparov_plays_white_beats_karpov = 30, # 13 on p1, 17 on p2
kasparov_plays_white_losesto_karpov = 07,
karpov_plays_white_losesto_kasparov = 09,
karpov_plays_white_draws_kasparov = 11+13+14+17+14+03,
kasparov_plays_white_draws_karpov = 14+12+11+08+11+01,

## Kasparov vs Anand
kasparov_plays_white_beats_anand = 15,
anand_plays_white_beats_kasparov = 06,
anand_plays_white_losesto_kasparov = 11,
kasparov_plays_white_losesto_anand = 02,
kasparov_plays_white_draws_anand = 26,
anand_plays_white_draws_kasparov = 20,

## Karpov vs Anand
karpov_plays_white_beats_anand = 07,
anand_plays_white_beats_karpov = 18,
anand_plays_white_losesto_karpov = 05,
karpov_plays_white_losesto_anand = 13,
karpov_plays_white_draws_anand = 29,
anand_plays_white_draws_karpov = 20
)
kka
```

Quite a lot of information here, but for example we might ask what the
scoreline is for Karpov playing white against Kasparov:

```{r kkaexample}
(a1 <- c(
kka["karpov_plays_white_beats_kasparov"],
kka["karpov_plays_white_draws_kasparov"],
kka["karpov_plays_white_losesto_kasparov"]
))
```

If we were to ask instead what happens if Karpov plays black against
Kasparov this would be:

```{r kkaexample2}
(a2 <- c(
kka["kasparov_plays_white_beats_karpov"],
kka["kasparov_plays_white_draws_karpov"],
kka["kasparov_plays_white_losesto_karpov"]
))
```

To investigate the effect of playing white we might do this:

```{r anamwa}
M <- rbind(a1,rev(a2))
dimnames(M) <- list(karpov_plays=c("White","Black"),result=c("karpov wins","karpov draws","kaprov loses"))
M
fisher.test(M)
```

but we need a likelihood function.  Here I define likelihood function
`karpov_kasparov_anand` but to save typing we work with temporary
object `H` and rename it at the end:

```{r defineH}

results <- as.list(kka)
attach(results)
H <- hyper2()

## Kasparov vs Karpov
karpov_vs_kasparov <- c("Karpov","Kasparov","white","draw")  # all "players", real and imaginary

H <- H + trial(c("Karpov"  ,"white"), karpov_vs_kasparov, karpov_plays_white_beats_kasparov)
H <- H + trial(c("Kasparov","white"), karpov_vs_kasparov, kasparov_plays_white_beats_karpov)
H <- H + trial(  "Karpov"           , karpov_vs_kasparov, kasparov_plays_white_losesto_karpov)
H <- H + trial(  "Kasparov"         , karpov_vs_kasparov, karpov_plays_white_losesto_kasparov)
H <- H + trial(  "draw"             , karpov_vs_kasparov, karpov_plays_white_draws_kasparov)
H <- H + trial(  "draw"             , karpov_vs_kasparov, kasparov_plays_white_draws_karpov)

## Kasparov vs Anand
kasparov_vs_anand <- c("Kasparov","Anand","white","draw")
H <- H + trial(c("Kasparov","white"), kasparov_vs_anand, kasparov_plays_white_beats_anand)
H <- H + trial(c("Anand"   ,"white"), kasparov_vs_anand, anand_plays_white_beats_kasparov)
H <- H + trial(c("Kasparov"        ), kasparov_vs_anand, anand_plays_white_losesto_kasparov)
H <- H + trial(c("Anand"           ), kasparov_vs_anand, kasparov_plays_white_losesto_anand)
H <- H + trial(  "draw"             , kasparov_vs_anand, kasparov_plays_white_draws_anand)
H <- H + trial(  "draw"             , kasparov_vs_anand, anand_plays_white_draws_kasparov)

## Karpov vs Anand
karpov_vs_anand <- c("Karpov","Anand","white","draw")
H <- H + trial(c("Karpov","white"), karpov_vs_anand, karpov_plays_white_beats_anand)
H <- H + trial(c("Anand" ,"white"), karpov_vs_anand, anand_plays_white_beats_karpov)
H <- H + trial(  "Karpov"         , karpov_vs_anand, anand_plays_white_losesto_karpov)
H <- H + trial(  "Anand"          , karpov_vs_anand, karpov_plays_white_losesto_anand)
H <- H + trial(  "draw"           , karpov_vs_anand, karpov_plays_white_draws_anand)
H <- H + trial(  "draw"           , karpov_vs_anand, anand_plays_white_draws_karpov)

detach(results)

karpov_kasparov_anand <- H
karpov_kasparov_anand
(karpov_kasparov_anand_maxp <- maxp(karpov_kasparov_anand))
```

Test the hypothesis that all three players have the same strength; first do the free optimization:

```{r testall3}
samep.test(karpov_kasparov_anand,c("Karpov","Kasparov","Anand"))
```

and we reject that null.  Now test the hypothesis that playing white confers no advantage:

```{r whitenoadvantage,cache=TRUE}
specificp.test(karpov_kasparov_anand,"white",0)
```

We now allow each player to have a personalised draw monster.  The
resulting likelihood function is called `kka_3draws` but the idiom
below uses `H` to save typing.

```{r kka3draws}
library("hyper2")
H <- hyper2()

results <- as.list(kka)
attach(results)

karpov_vs_kasparov <- c("Karpov","Kasparov","Karpov_draw","Kasparov_draw","white")
draw1 <- c("Karpov_draw","Kasparov_draw")

H <- H + trial(c("Karpov"  ,"white"), karpov_vs_kasparov, karpov_plays_white_beats_kasparov)
H <- H + trial(c("Kasparov","white"), karpov_vs_kasparov, kasparov_plays_white_beats_karpov)
H <- H + trial(c("Karpov"  )        , karpov_vs_kasparov, kasparov_plays_white_losesto_karpov)
H <- H + trial(c("Kasparov")        , karpov_vs_kasparov, karpov_plays_white_losesto_kasparov)
H <- H + trial(draw1                , karpov_vs_kasparov, karpov_plays_white_draws_kasparov)
H <- H + trial(draw1                , karpov_vs_kasparov, kasparov_plays_white_draws_karpov)

## Kasparov vs Anand
kasparov_vs_anand <- c("Kasparov","Anand","Kasparov_draw","Anand_draw","white")
draw2 <- c("Kasparov_draw","Anand_draw")
H <- H + trial(c("Kasparov","white"), kasparov_vs_anand, kasparov_plays_white_beats_anand)
H <- H + trial(c("Anand"   ,"white"), kasparov_vs_anand, anand_plays_white_beats_kasparov)
H <- H + trial(c("Kasparov"  )      , kasparov_vs_anand, anand_plays_white_losesto_kasparov)
H <- H + trial(c("Anand")           , kasparov_vs_anand, kasparov_plays_white_losesto_anand)
H <- H + trial(draw2                , kasparov_vs_anand, kasparov_plays_white_draws_anand)
H <- H + trial(draw2                , kasparov_vs_anand, anand_plays_white_draws_kasparov)

## Karpov vs Anand
karpov_vs_anand <- c("Karpov","Anand","Karpov_draw","Anand_draw","white")
draw3 <- c("Karpov_draw","Anand_draw")

H <- H + trial(c("Karpov","white"), karpov_vs_anand, karpov_plays_white_beats_anand)
H <- H + trial(c("Anand" ,"white"), karpov_vs_anand, anand_plays_white_beats_karpov)
H <- H + trial(c("Karpov"        ), karpov_vs_anand, anand_plays_white_losesto_karpov)
H <- H + trial(c("Anand"         ), karpov_vs_anand, karpov_plays_white_losesto_anand)
H <- H + trial(draw3              , karpov_vs_anand, karpov_plays_white_draws_anand) 
H <- H + trial(draw3              , karpov_vs_anand, anand_plays_white_draws_karpov)

detach(results)

kka_3draws <- H
kka_3draws 
```

and one natural test is that the three draw monsters are of equal strength:

```{r threedrawsequal,cache=TRUE}
(kka_3draws_maxp <- maxp(kka_3draws))
pie(kka_3draws_maxp)
samep.test(kka_3draws,c("Karpov_draw","Kasparov_draw","Anand_draw"))
```

Now allow each player to have distinct white strength.  We create
`kka_3whites`, again using temporary variable `H` to save typing:

```{r kka3whitedef}
H <- hyper2()

results <- as.list(kka)
attach(results)

D <- "draw"

## First: Karpov vs Kasparov
karpov_plays_white_vs_kasparov <- c("Karpov","Kasparov","Karpov_white","draw"  )  # "players" 
kasparov_plays_white_vs_karpov <- c("Karpov","Kasparov","Kasparov_white","draw")
H <- H + trial(c("Karpov"  ,"Karpov_white"  ), karpov_plays_white_vs_kasparov, karpov_plays_white_beats_kasparov  ) # Karpov wins playing white
H <- H + trial(c("Kasparov","Kasparov_white"), kasparov_plays_white_vs_karpov, kasparov_plays_white_beats_karpov  ) # Kasparov wins playing white
H <- H + trial(c("Kasparov")                 , karpov_plays_white_vs_kasparov, karpov_plays_white_losesto_kasparov) # Kasparov wins playing black
H <- H + trial(c("Karpov"  )                 , kasparov_plays_white_vs_karpov, kasparov_plays_white_losesto_karpov) # Karpov wins playing black
H <- H + trial(D                             , karpov_plays_white_vs_kasparov, karpov_plays_white_draws_kasparov  ) # Karpov white, draws
H <- H + trial(D                             , kasparov_plays_white_vs_karpov, kasparov_plays_white_draws_karpov  ) # Kasparov white, draws

## Second: Karpov vs Anand
karpov_plays_white_vs_anand <- c("Karpov","Anand","Karpov_white","draw"  )
anand_plays_white_vs_karpov <- c("Karpov","Anand","Anand_white","draw")
H <- H + trial(c("Karpov","Karpov_white"), karpov_plays_white_vs_anand, karpov_plays_white_beats_anand  ) # Karpov wins playing white
H <- H + trial(c("Anand" ,"Anand_white" ), anand_plays_white_vs_karpov, anand_plays_white_beats_karpov  ) # Anand wins playing white
H <- H + trial(c("Anand"                ), karpov_plays_white_vs_anand, karpov_plays_white_losesto_anand) # Anand wins playing black
H <- H + trial(c("Karpov"               ), anand_plays_white_vs_karpov, anand_plays_white_losesto_karpov) # Karpov wins playing black
H <- H + trial(D                         , karpov_plays_white_vs_anand, karpov_plays_white_draws_anand  ) # Karpov white, draws
H <- H + trial(D                         , anand_plays_white_vs_karpov, anand_plays_white_draws_karpov  ) # Anand white, draws

## Third: Kasparov vs Anand
anand_plays_white_vs_kasparov <- c("Anand","Kasparov","Anand_white","draw"   )
kasparov_plays_white_vs_anand <- c("Anand","Kasparov","Kasparov_white","draw")
H <- H + trial(c("Kasparov","Kasparov_white"), kasparov_plays_white_vs_anand, kasparov_plays_white_beats_anand  ) # Kasparov wins playing white
H <- H + trial(c("Anand"   ,"Anand_white"   ), anand_plays_white_vs_kasparov, anand_plays_white_beats_kasparov  ) # Anand wins playing white
H <- H + trial(c("Anand"                    ), kasparov_plays_white_vs_anand, kasparov_plays_white_losesto_anand) # Anand wins playing black
H <- H + trial(c("Kasparov"                 ), anand_plays_white_vs_kasparov, anand_plays_white_losesto_kasparov) # Kasparov wins playing black
H <- H + trial(D                             , kasparov_plays_white_vs_anand, kasparov_plays_white_draws_anand  ) # Kasparov white, draws
H <- H + trial(D                             , anand_plays_white_vs_kasparov, anand_plays_white_draws_kasparov  ) # Anand white, draws

detach(results)

kka_3whites <- H
kka_3whites
```

Natural test for equality of the three white monsters's strengths:

```{r threewhitesequal,cache=TRUE}
(kka_3whites_maxp <- maxp(kka_3whites))
pie(kka_3whites_maxp)
samep.test(kka_3whites,c("Karpov_white","Kasparov_white","Anand_white"))
```
(fail to reject the null).

We now create a 3x3x3 array of results for the dataset.

```{r kkaarray}

library("hyper2")
library("abind")
attach(as.list(kka))

players <- c("Anand","Karpov","Kasparov")

plays_white_wins <- matrix(NA,3,3)
dimnames(plays_white_wins) <- list(plays_white_wins=players,plays_black_loses=players)
plays_white_wins["Anand"   ,"Karpov"  ] <- anand_plays_white_beats_karpov
plays_white_wins["Anand"   ,"Kasparov"] <- anand_plays_white_beats_kasparov
plays_white_wins["Karpov"  ,"Anand"   ] <- karpov_plays_white_beats_anand
plays_white_wins["Karpov"  ,"Kasparov"] <- karpov_plays_white_beats_kasparov
plays_white_wins["Kasparov","Anand"   ] <- kasparov_plays_white_beats_anand
plays_white_wins["Kasparov","Karpov"  ] <- kasparov_plays_white_beats_karpov

plays_white_draws  <- matrix(NA,3,3)
dimnames(plays_white_draws) <- list(plays_white_draws=players,plays_black_draws=players)
plays_white_draws["Anand"   ,"Karpov"  ] <- anand_plays_white_draws_karpov
plays_white_draws["Anand"   ,"Kasparov"] <- anand_plays_white_draws_kasparov
plays_white_draws["Karpov"  ,"Anand"   ] <- karpov_plays_white_draws_anand
plays_white_draws["Karpov"  ,"Kasparov"] <- karpov_plays_white_draws_kasparov
plays_white_draws["Kasparov","Anand"   ] <- kasparov_plays_white_draws_anand
plays_white_draws["Kasparov","Karpov"  ] <- kasparov_plays_white_draws_karpov

plays_white_loses <- matrix(NA,3,3)
dimnames(plays_white_loses) <- list(plays_white_loses=players,plays_black_wins=players)
plays_white_loses["Karpov"   ,"Anand"   ] <- karpov_plays_white_losesto_anand
plays_white_loses["Kasparov" ,"Anand"   ] <- kasparov_plays_white_losesto_anand
plays_white_loses["Anand"    ,"Karpov"  ] <- anand_plays_white_losesto_karpov
plays_white_loses["Kasparov" ,"Karpov"  ] <- kasparov_plays_white_losesto_karpov
plays_white_loses["Anand"    ,"Kasparov"] <- anand_plays_white_losesto_kasparov
plays_white_loses["Karpov"   ,"Kasparov"] <- karpov_plays_white_losesto_kasparov
detach(as.list(kka))

    plays_white_wins
    plays_white_draws
    plays_white_loses

kka_array <- abind(
    plays_white_wins,
    plays_white_draws,
    plays_white_loses,
    along=3)
dimnames(kka_array)[[3]] <- c("plays_white_wins", "plays_white_draws", "plays_white_loses")
kka_array
```

### Package dataset

Following lines create `kka.rda`, residing in the `data/` directory of
the package.

```{r,label=savecurlingdataset,cache=TRUE}
save(kka,karpov_kasparov_anand,kka_3draws,kka_3whites,kka_array,file="kka.rda")
```

### 2021-02-06 inst/karpov_kasparov_anand.R

@@ -1,130 +1,0 @@
-## This file creates likelihood function 'H' which is identical to
-## data object 'karpov_kasparov_anand' in the package.
-
-## In this file, object 'kka' holds the chess playing results; see
-## karpov_kasparov_anand.Rd for more details.
-
-## Then object 'H' is created, and used to test two hypotheses:
-## firstly, that all three players have the same strength, and
-## secondly that playing white confers no strength.
-
-## In this file, all three players are assumed to have the same draw
-## strength.
-
-
-library("hyper2")
-results <- as.list(kka)
-attach(results)
-H <- hyper2()
-
-
-## Kasparov vs Karpov
-karpov_vs_kasparov <- c("Karpov","Kasparov","white","draw")  # all "players", real and imaginary
-
-H <- H + trial(c("Karpov"  ,"white"), karpov_vs_kasparov, karpov_plays_white_beats_kasparov)
-H <- H + trial(c("Kasparov","white"), karpov_vs_kasparov, kasparov_plays_white_beats_karpov)
-H <- H + trial(  "Karpov"           , karpov_vs_kasparov, kasparov_plays_white_losesto_karpov)
-H <- H + trial(  "Kasparov"         , karpov_vs_kasparov, karpov_plays_white_losesto_kasparov)
-H <- H + trial(  "draw"             , karpov_vs_kasparov, karpov_plays_white_draws_kasparov)
-H <- H + trial(  "draw"             , karpov_vs_kasparov, kasparov_plays_white_draws_karpov)
-
-## Kasparov vs Anand
-kasparov_vs_anand <- c("Kasparov","Anand","white","draw")
-H <- H + trial(c("Kasparov","white"), kasparov_vs_anand, kasparov_plays_white_beats_anand)
-H <- H + trial(c("Anand"   ,"white"), kasparov_vs_anand, anand_plays_white_beats_kasparov)
-H <- H + trial(c("Kasparov"        ), kasparov_vs_anand, anand_plays_white_losesto_kasparov)
-H <- H + trial(c("Anand"           ), kasparov_vs_anand, kasparov_plays_white_losesto_anand)
-H <- H + trial(  "draw"             , kasparov_vs_anand, kasparov_plays_white_draws_anand)
-H <- H + trial(  "draw"             , kasparov_vs_anand, anand_plays_white_draws_kasparov)
-
-
-## Karpov vs Anand
-karpov_vs_anand <- c("Karpov","Anand","white","draw")
[…]

### 2021-02-06 inst/kka_3draws.R

@@ -1,135 +1,0 @@
-## Analysis of three chess players (Kasparov, Karpov, Anand).  See
-## file karpov_kasparov_anand.R for details on the origin of the data.
-
-## This is a more sophisticated version (with more parameters) than
-## inst/karpov_kasparov.R: it allows for each player to have a
-## distinct draw monster.
-
-## This file creates likelihood function 'H' 
-
-
-library("hyper2")
-H <- hyper2()
-
-results <- as.list(kka)
-attach(results)
-
-karpov_vs_kasparov <- c("Karpov","Kasparov","Karpov_draw","Kasparov_draw","white")
-draw1 <- c("Karpov_draw","Kasparov_draw")
-
-H <- H + trial(c("Karpov"  ,"white"), karpov_vs_kasparov, karpov_plays_white_beats_kasparov)
-H <- H + trial(c("Kasparov","white"), karpov_vs_kasparov, kasparov_plays_white_beats_karpov)
-H <- H + trial(c("Karpov"  )        , karpov_vs_kasparov, kasparov_plays_white_losesto_karpov)
-H <- H + trial(c("Kasparov")        , karpov_vs_kasparov, karpov_plays_white_losesto_kasparov)
-H <- H + trial(draw1                , karpov_vs_kasparov, karpov_plays_white_draws_kasparov)
-H <- H + trial(draw1                , karpov_vs_kasparov, kasparov_plays_white_draws_karpov)
-
-## Kasparov vs Anand
-kasparov_vs_anand <- c("Kasparov","Anand","Kasparov_draw","Anand_draw","white")
-draw2 <- c("Kasparov_draw","Anand_draw")
-H <- H + trial(c("Kasparov","white"), kasparov_vs_anand, kasparov_plays_white_beats_anand)
-H <- H + trial(c("Anand"   ,"white"), kasparov_vs_anand, anand_plays_white_beats_kasparov)
-H <- H + trial(c("Kasparov"  )      , kasparov_vs_anand, anand_plays_white_losesto_kasparov)
-H <- H + trial(c("Anand")           , kasparov_vs_anand, kasparov_plays_white_losesto_anand)
-H <- H + trial(draw2                , kasparov_vs_anand, kasparov_plays_white_draws_anand)
-H <- H + trial(draw2                , kasparov_vs_anand, anand_plays_white_draws_kasparov)
-
-
-## Karpov vs Anand
[…]

### 2021-02-06 inst/kka_3whites.R

@@ -1,95 +1,0 @@
-## Analysis of three chess players (Kasparov, Karpov, Anand).  See
-## file karpov_kasparov_anand.R for details on the origin of the data.
-
-## This is a more sophisticated version (with more parameters) than
-## inst/karpov_kasparov.R: it allows each player to have a distinct
-## white strength.
-
-## This file creates likelihood function 'H', which is identical to
-## dataset kka_3whites in the package.
-
-
-library("hyper2")
-H <- hyper2()
-
-results <- as.list(kka)
-attach(results)
-
-D <- "draw"
-
-## First: Karpov vs Kasparov
-karpov_plays_white_vs_kasparov <- c("Karpov","Kasparov","Karpov_white","draw"  )  # "players" 
-kasparov_plays_white_vs_karpov <- c("Karpov","Kasparov","Kasparov_white","draw")
-H <- H + trial(c("Karpov"  ,"Karpov_white"  ), karpov_plays_white_vs_kasparov, karpov_plays_white_beats_kasparov  ) # Karpov wins playing white
-H <- H + trial(c("Kasparov","Kasparov_white"), kasparov_plays_white_vs_karpov, kasparov_plays_white_beats_karpov  ) # Kasparov wins playing white
-H <- H + trial(c("Kasparov")                 , karpov_plays_white_vs_kasparov, karpov_plays_white_losesto_kasparov) # Kasparov wins playing black
-H <- H + trial(c("Karpov"  )                 , kasparov_plays_white_vs_karpov, kasparov_plays_white_losesto_karpov) # Karpov wins playing black
-H <- H + trial(D                             , karpov_plays_white_vs_kasparov, karpov_plays_white_draws_kasparov  ) # Karpov white, draws
-H <- H + trial(D                             , kasparov_plays_white_vs_karpov, kasparov_plays_white_draws_karpov  ) # Kasparov white, draws
-
-## Second: Karpov vs Anand
-karpov_plays_white_vs_anand <- c("Karpov","Anand","Karpov_white","draw"  )
-anand_plays_white_vs_karpov <- c("Karpov","Anand","Anand_white","draw")
-H <- H + trial(c("Karpov","Karpov_white"), karpov_plays_white_vs_anand, karpov_plays_white_beats_anand  ) # Karpov wins playing white
-H <- H + trial(c("Anand" ,"Anand_white" ), anand_plays_white_vs_karpov, anand_plays_white_beats_karpov  ) # Anand wins playing white
[…]

### 2021-02-07 inst/table_tennis_serve.Rmd
consistentify table tennis

@@
 ```{r}
-H
+table_tennis <- H
+table_tennis
 ```
@@
 ```{r}
-p1 <- maxp(H)
-p1
-```
-
-And we can find the support at this point:
-
-```{r}
-maxlike_free <- loglik(indep(p1),H)
-maxlike_free
+table_tennis_maxp <- maxp(table_tennis)
+table_tennis_maxp
 ```
@@
 
-To do this we follow the protocol in the `icons` vignette.  First find
-the MLE conditional on the serve having zero strength:
-
-```{r}
-small <- 1e-3
-maxlike_constrained <-
-    maxp(
-        H,
-        startp=c(small/2, 1/3-small/2, 1/3-small/2),
-        fcm=c(-1,0,0),  fcv=-small,give=TRUE
-)
-
-maxlike_constrained
-```
-
-The difference in support is:
-
-```{r}    	    
-delta_support <- maxlike_free - maxlike_constrained$value
-delta_support
-```
-
-and the likelihood ratio is:
-```{r}
-LR <- exp(delta_support)
-LR
-```
-
-The difference in support, `delta_support`, is about 2.62 which
-exceeds Edwards's two-units-of-support criterion.  We can reject the
-null.  Alternatively, Wilks's theorem gives us that
-$\Lambda=2\log(LR)$ is distributed as chi-squared with one degree of
-freedom:
-
-```{r}
-Lambda <- 2*delta_support
-pval <- pchisq(Lambda,df=1,lower.tail=FALSE)
-pval 
+```{r testservezero}
+specificp.test(table_tennis,"S",0.01)
 ```
@@
 ```{r}
-maxp(H)
+maxp(table_tennis)
 ```
@@
 ```{r}
-profile_likelihood <- function(S,give=FALSE){
-    small <- 1e-2  # cannot make this too small, optimization routine needs wiggle room
-    out <- maxp(
-       H, startp=c(S+small/2 , small,small),
-       give=TRUE,
-       fcm=rbind(c(1,0,0),c(-1,0,0)),fcv=c(S,-S-small)  # S <= p[1] <= S+small
-       )
-    if(give){
-       return(out)
-      } else {
-       return(out$value)
-      }
-}
-```
-
-We can then evaluate the profile likelihood for a range of possible
-values of $S$:
-
-```{r}
 S_vals <- seq(from=0.02,to=0.85,len=30)  # possible values for S
-prof_like <- sapply(S_vals,profile_likelihood)  
-plot(S_vals, prof_like-max(prof_like),xlab="serve strength",ylab="profile likelihood")
-abline(h= -2)
+plot(S_vals,profsupp(table_tennis,"S",S_vals))
+abline(h=c(0,-2))
 ```
@@
 server and the value of $S$ is itself uncertain.
+
+### Package dataset
+
+Following lines create `table_tennis.rda`, residing in the `data/`
+directory of the package.
+
+```{r,label=savetabletennisdataset}
+save(table_tennis,file="table_tennis.rda")
+```

### 2021-02-07 data/table_tennis.R

@@ -1,5 +1,0 @@
-`table_tennis` <-
-structure(list(brackets = list(c("S", "a"), c("S", "a", "b"), 
-    c("S", "a", "c"), c("S", "b"), c("S", "c"), "a", "b", "c"), 
-    powers = c(9, -10, -8, 3, 2, 2, 1, 1), pnames = c("S", "a", 
-    "b", "c")), class = "hyper2")

### 2021-02-23 inst/karate.Rmd
rewording

@@
 dataset taken from
-[https://en.wikipedia.org/wiki/2018_World_Karate_Championships].
+[wikipedia](https://en.wikipedia.org/wiki/2018_World_Karate_Championships).
 

### 2021-02-23 inst/retain.Rmd
clarification

@@
 judges.  Object `jj` does not function as a rank table because the
-entries are not sequential.
+entries are not sequential (judge `J2`, for example, has no
+second-ranked competitor).
 
@@
 natural interpretation in probability, but if so I don't see it.
+
+###  Conclusions{-}
+
+It does not seem possible to take a `hyper2` object and discard
+certain players.  It seems that the only way to discard players is to
+work with the original observation, discard players in the dataset
+according to some criterion, and coerce to a likelihood function.

### 2021-02-23 inst/eurovision.Rmd
typos

@@
 
-The file `eurovision.txt`, defined below, is copied from "Eurovision
-Song Contest 2009," Wikipedia, accessed May 13, 2018.  It refers to
+The file `eurovision.txt`, used below, is copied from "Eurovision Song
+Contest 2009," Wikipedia, accessed May 13, 2018.  It refers to
 semi-final 1.  More documentation is given in `eurovision.Rd` [type
@@
 
-'''{r,label=equalptesteurovision,cache=TRUE}
+```{r,label=equalptesteurovision,cache=TRUE}
 equalp.test(eurovision)

### 2021-03-18 inst/fatigue.Rmd
interzonal 1973 dataset and analysis

---
title: "fatigue"
author: "Robin Hankin"
date: "16/03/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
```

## R Markdown

Analysis of  the 1973 interzonal competition

```{r cars}
jj <- read.table("interzonal1973_petropolis.txt",header=TRUE,skip=2)
a <- data.frame(day=jj$day + 31*(jj$month=="August"),white=jj$white,black=jj$black,result=jj$result,round=jj$round,moves=jj$moves)

plot(sort(unique(a$day)),type='b',pch=c(1,1,16))
tail(a)
```

```{r,label=define_H,cache=TRUE}
players <-  unique(sort(c(a$white,a$black)))

wm <- "wm" 
dm <- "dm"
rm <- "rm"

tab <- matrix(0,length(players),3)
colnames(tab) <- c("won","lost","drawn")
rownames(tab) <- players

H <- hyper2()
for(i in seq_len(nrow(a))){
  white_player <- a$white[i]
  black_player <- a$black[i]
  result <- a$result[i]
  H[c(white_player,wm,black_player,dm)] %<>% dec()
  if(result == "1-0"){ # white win, black loss
    H[c(white_player,wm)] %<>% inc()
    tab[white_player,] <- tab[white_player,] + c(1,0,0)
    tab[black_player,] <- tab[black_player,] + c(0,1,0)
  } else if(result == "0-1"){ # white loss, black win
    H[black_player] %<>% inc()
    tab[white_player,] <- tab[white_player,] + c(0,1,0)
    tab[black_player,] <- tab[black_player,] + c(1,0,0)
  } else if(result == "1/2-1/2"){  # draw
    H[dm] %<>% inc()
    tab[white_player,] <- tab[white_player,] + c(0,0,1)
    tab[black_player,] <- tab[black_player,] + c(0,0,1)

  } else {
    stop("this cannot happen")
  }
}
```

```{r,label=calculatemaxp,cache=TRUE}
mH <- maxp(H,n=100)
```

```{r,testwhitezero,cache=TRUE}
specificp.gt.test(H,"wm",0)
```

```{r,summaryofH}
head(H)
pnames(H)

pie(mH)
dotchart(mH)
mH
summary(H)
```

```{r}
tab <- cbind(tab,played=rowSums(tab))
tab
```

```{r}
table(a$result)
hist(a$moves)
plot(moves~as.factor(result),data=a)
result
moves_drawn_games <- a$moves[a$result == "1/2-1/2"]
moves_won_games   <- a$moves[a$result != "1/2-1/2"]
t.test(moves_drawn_games,moves_won_games)
```

```{r}
head(a,10)
o <- matrix(0,max(a[,5]),length(players))
dimnames(o) <- list(round=paste("r",seq_len(nrow(o)),sep=""),player=players)

for(i in seq_len(nrow(a))){
	o[a[i,5],which(players==a[i,2])] %<>% `+`(1)
	o[a[i,5],which(players==a[i,3])] %<>% `+`(1)
}
o
```

Now focus on pairs of successive rounds: 1,2; 4,5; 7,8; 10,11; 13,14; 16,17

```{r}
unlist(a[2,])
`fatigued` <- function(i){
  x <- unlist(a[i,])
  day <- as.numeric(a[i,1])
  white_player <- a[i,2]
  black_player <- a[i,3]
  result <- a[i,4]
  round <- as.numeric(a[i,5])
  out <- c(white_player_fatigued=FALSE,black_player_fatigued=FALSE)

  if(round%%3 == 1){return(out)}  # yesterday was a break day
  out <- c(white_player_fatigued=FALSE,black_player_fatigued=FALSE)
  b <- a[a$round==round-1,]  # b = yesterday's schedule
  yesterdays_white_players <- unique(sort(c(b[,2])))
  yesterdays_black_players <- unique(sort(c(b[,3])))

  if(white_player %in% yesterdays_white_players){  # he played white yesterday...
    if(b[which(white_player==b[,2]),4] !="1/2-1/2"){  # ...and didn't draw...
      out["white_player_fatigued"] <- TRUE # ... the white player is fatigued
    }
  }

  if(white_player %in% yesterdays_black_players){  # he played black yesterday...
    if(b[which(white_player==b[,3]),4] !="1/2-1/2"){  # ...and didn't draw...
      out["white_player_fatigued"] <- TRUE # ... the white player is fatigued
    }
  }

  if(black_player %in% yesterdays_white_players){  # he played white yesterday...
    if(b[which(black_player==b[,2]),4] !="1/2-1/2"){  # ...and didn't draw...
      out["black_player_fatigued"] <- TRUE # ... the black player is fatigued
    }
  }

  if(black_player %in% yesterdays_black_players){  # he played black yesterday...
    if(b[which(white_player==b[,3]),4] !="1/2-1/2"){  # ... and didn't draw...
      out["black_player_fatigued"] <- TRUE # ... the white player is fatigued
    }
  }

  return(out)
}

rm(H)
Hf <- hyper2() # H fatigue
for(i in seq_len(nrow(a))){
  white_player <- a[i,2]
  black_player <- a[i,3]
  result <- a[i,4]

  jj <- fatigued(i)
  white_player_fatigued <- jj[1]
  black_player_fatigued <- jj[2]

  if(white_player_fatigued & black_player_fatigued){  # both fatigued: no fatigue monster
    Hf[c(white_player,wm,black_player,dm)] %<>% dec()
    if(result == "1-0"){ # white win, black loss
      Hf[c(white_player,wm)] %<>% inc()
    } else if(result == "0-1"){ # white loss, black win
      Hf[c(black_player)] %<>% inc()
    } else if(result == "1/2-1/2"){ # draw
      Hf[dm] %<>% inc()
    } else {
      stop("result must be 0-1, 1-0, or 1/2-1/2")
    }
  } else if (white_player_fatigued & !black_player_fatigued){
    Hf[c(white_player,wm,black_player,dm,rm)] %<>% dec()
    if(result == "1-0"){ # white win, black loss
      Hf[c(white_player,wm)] %<>% inc()
    } else if(result == "0-1"){ # white loss, black win
      Hf[c(black_player,rm)] %<>% inc()
    } else if(result == "1/2-1/2"){ # draw
      Hf[dm] %<>% inc()
    } else {
      stop("result must be 0-1, 1-0, or 1/2-1/2")
    }
  } else if (!white_player_fatigued & black_player_fatigued){
    Hf[c(white_player,wm,black_player,dm,rm)] %<>% dec()
    if(result == "1-0"){ # white win, black loss
      Hf[c(white_player,wm,rm)] %<>% inc()
    } else if(result == "0-1"){ # white loss, black win
      Hf[c(black_player)] %<>% inc()
    } else if(result == "1/2-1/2"){ # draw
      Hf[dm] %<>% inc()
    } else {
      stop("result must be 0-1, 1-0, or 1/2-1/2")
    }
  } else if (!white_player_fatigued & !black_player_fatigued){ # neither fatigued
    Hf[c(white_player,wm,black_player,dm)] %<>% dec()
    if(result == "1-0"){ # white win, black loss
      Hf[c(white_player,wm)] %<>% inc()
    } else if(result == "0-1"){ # white loss, black win
      Hf[c(black_player)] %<>% inc()
    } else if(result == "1/2-1/2"){ # draw
      Hf[dm] %<>% inc()
    } else {
      stop("result must be 0-1, 1-0, or 1/2-1/2")
    }
  } else {
    stop("This is a logical impossibility")
  }
}
```

now play with it:

```{r,calchfmax,cache=TRUE}
Hf_maxp <- maxp(Hf)
````

```{r}
summary(Hf)
pnames(Hf)
Hf_maxp
pie(Hf_maxp)
```

```{r}
jj <- rep(0,nrow(a))
M <- data.frame(
white_player=rep(NA,nrow(a)),
black_player=rep(NA,nrow(a)),
white_player_fatigued=FALSE,
black_player_fatigued=FALSE
)
for(i in seq_len(nrow(a))){
	M[i,1] <- a[i,2]	
	M[i,2] <- a[i,3]	
	jj <- fatigued(i)
	M[i,3] <- jj[1]
	M[i,4] <- jj[2]
}
colnames(M) <- c("white","black","white_fatigued","black_fatigued")
tail(M)
table(M[,3])
table(M[,4])
```

### 2021-03-27 inst/formula1.Rmd
update link
remove points analysis (which was wrong): see formula1_points.Rmd

@@
 into a log-likelihood.  The function is similar to, but slightly
-different from, the analysis in `eurovision.R`.
+different from, the analysis in `eurovision.Rmd`.
 
@@
 which is even worse.
-
-# Likelihood scoring vs points scoring
-
-The championship winning order is determined by a points system: the
-winner is awarded 25 points, second place 18 points, and so on.  We
-can now check to see how the points total matches up with the
-likelihood estimation.
-
-```{r,likelihoodandscore}
-jj <- read.table("formula1_2017.txt",header=TRUE)
-points <- jj[,21]
-names(points) <- rownames(jj)
-
-m <- mL2017   # maxp(formula1)
-ox <- rank(-points) # negative because number 1 = largest number of points
-oy <- rank(-m)      # negative because number 1 = largest strength
-par(pty='s')        # square plot
-plot(ox,oy,asp=1,pty='s',xlim=c(0,25),ylim=c(0,25),pch=16,
-xlab="official order",ylab="my order",main='Formula 1, 2017 season')
-par(xpd=TRUE)       # allow drivers' names to appear outside plotting region
-for(i in seq_along(ox)){text(ox[i],oy[i],names(ox)[i],pos=4,col='gray') }
-par(xpd=FALSE)      # stop diagonal line from protruding beyond plotting region
-abline(0,1)
-```
-
-In the graph above, we see a certain amount of disagreement.  For
-example, using F1 points scored places Hamilton first (363), Vettel
-second (317), and Bottas third (305); while `hyper2` places Hamilton
-first (0.29), Bottas second (0.17), and Vettel third (0.14).  Note
-also the large number of players with zero points (who rank equal 23rd
-by the F1 points system) but are nevertheless separated by
-Bradley-Terry.  These players appear as a line of points at $x=23$:
-Resta to Ericcson.  Note also Palmer and Alonso who managed to score a
-F1 point but rank below Ericcson according to Bradley-Terry.
-
-But overall there
-seems to be some degree of correlation between the two, and
-differences in detail.
 
@@
 a <- list(F1_2014, F1_2015, F1_2016, F1_2017)
-alldrivers <- all_pnames(a)
+alldrivers <- unique(sort(unlist(lapply(a,pnames))))
 
-F1_total <- hyper2(pnames=alldrivers)
-
-F1_total %<>% add(change_pnames(F1_2014,alldrivers)) 
-F1_total %<>% add(change_pnames(F1_2015,alldrivers)) 
-F1_total %<>% add(change_pnames(F1_2016,alldrivers)) 
-F1_total %<>% add(change_pnames(F1_2017,alldrivers))
+F1_total <- F1_2014 + F1_2015 + F1_2016 + F1_2017
 
@@
 	}
-	L2017 %<>% order_obs(o)
+	L2017 <- L2017 + ordervec2supp(o)
 }  # for(i) loop closes

### 2021-03-28 inst/ordertable_to_ranktable.Rmd
issue #43 refers

@@
 
-## Rank tables
+# Introduction
 
-Rank tables can be confusing, here is a tiny example that should
-clarify things.
+Rank tables can be confusing.  The package includes a range of
+functionality for dealing with them, including a bespoke print method.
+Here, I discuss two cases: firstly, a tiny toy example, and secondly
+the volvo ocean racing dataset.
+
+# Toy dataset
 
@@
 why the print method is there.
+
+# Volvo ocean racing dataset
+
+(See `volvo.Rd` for more details).
+
+```{r}
+volvo_table
+```
+
+Thus we see from the first column that in leg 1, AbuDabhi came first,
+Brunel third, and so on.  From the first row we see that AbuDhabi came
+first in leg 1, third in leg 2, second in leg 3, and so on.  Suppose
+we wish to know who came first in each leg:
+
+```{r}
+x <- wikitable_to_ranktable(volvo_table)
+x
+```
+
+And we see from the first column that AbuDhabi came first in leg 1,
+Brunel came first in leg 2, Dongfeng came first in leg 3, and so on.
+The second column shows who came second in each leg.  Now suppose we
+are interested in only the first, second and third-placed teams:
+
+```{r}
+x[,1:3]
+```
+
+The above shows that the extraction method for `ranktable` objects is
+not yet written; maybe one day I will get round to implementing
+`[.ranktable()` (right now I am not 100% sure that this is a good
+idea).  Note that simply restoring the `ranktable` class can be
+misleading:
+
+```{r}
+xx <- x[,1:3]
+xx
+class(xx) <- "ranktable"
+xx
+```
+
+In the above, `NA`s are introduced because a `ranktable` object uses
+the column names to store names of the competitors, and of course
+extracting the first three columns means keeping only the first three
+column names.  Observe that extraction after passing to
+`ranktable_to_printable_object()` works nicely:
+
+```{r}
+ranktable_to_printable_object(x)[,1:3]
+```

### 2021-03-30 inst/formula1.Rmd
more detail

@@
 ## Formula 1
+
+**This Rmd file takes about two hours to run without cache.**
 
@@
 mL2017
+mL2017 <- mL2017[-which(names(mL2017) %in% c("Resta","Button"))]
 dotchart(mL2017,pch=16,main='2017 Formula 1')

### 2021-03-30 inst/eurodance.Rmd
Eurovision Dance contest 2008

---
title: "Eurovision **Dance** Contest 2008"
output: bookdown::html_document2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("magrittr")
options("digits" = 5)
abbreviated <- TRUE  # change to FALSE for full names
```

The file `eurodance_2008.txt`, used below, is copied from
"Eurovision Dance Contest 2008," Wikipedia, accessed March 31, 2021.
First we specify the matrix as appearing in the Wikipedia page:

```{r, specifymatrix}
eurodance_table <- as.matrix(read.table("eurovision_dance_2008.txt"))
eurodance_table
o <- eurodance_table # saves typing
```

Each row corresponds to a contestant and each column to a judge.  The
points system used was:

```{r,pointsforplaces}
points_voters <- c(12,10,8,7,6,5,4,3,2,1)
points_jury   <- 4*points_voters
ov <- o[,-(1:2)] # ='o' but just the voters: exclude total and jury
pref <- ov*0  # retains NA
for(i in seq_along(points_voters)){pref[ov == points_voters[i]] <- i}
pref
```

Now we treat each column as an order statistic:

```{r,cache=TRUE}
eurodance <- hyper2()
for(i in seq_len(ncol(pref))){
	x <- pref[,i]
	eurodance %<>% `+`(ordervec2supp(x[!is.na(x)]))
}  # i loop closes
eurodance_maxp <- maxp(eurodance)
eurodance_maxp
pie(eurodance_maxp)
```

Note the dominance of Poland.  Now we consider the jury's vote:

```{r}
(jj <- o[,2])
jj <- sort(jj,decreasing=TRUE)
jj[jj>0] <- seq_len(sum(jj>0))
jury <- jj
jury
```

Now the question is whether the partial rank conferred by the jury is
consistent with that of the voters.  The permutation test technique
used in `cook.Rmd` is not appropriate here as there are $14!/4!\simeq
3.6\times 10^9$ permutations to consider; we will use a randomised
sampling method.  Below, we use a resampling technique: 4000
randomised jury observation are created, and the likelihood [under the
Plackett-Luce likelihood function for the order statistics of the
nationals' voters] is calculated.  The $p$-value is then the
probability of observing the actual observation or an observation more
unlikely under this likelihood function; it is operationally identical
to a simulated Fisher's exact test.

```{r,cache=TRUE}
n <- 4000
LL <- rep(0,n)
for(i in seq_len(n)){
	jjstar <- sample(jury)
	names(jjstar) <- names(jury)
	LL[i] <- loglik(indep(eurodance_maxp),ordervec2supp(jjstar),log=FALSE)
}
```

```{r}
plot(sort(LL),log="y")
p <- loglik(indep(eurodance_maxp),ordervec2supp(jury),log=FALSE)
abline(h=p)
```

```{r}
sum(LL[LL<p])/sum(LL)
```

The $p$-value of about 2.4\% shows that there is strong evidence to
suggest that the expert jury's ordering differs from that of the
national popular vote.

### Package dataset {-}

Following lines create `eurodance.rda`, residing in the `data/` directory of the package.

```{r,label=saveeurodancedataset}
save(eurodance_table,eurodance_maxp,eurodance,file="eurodance.rda")
```

### 2021-04-29 inst/surfing.Rmd
surfing work by Tim Driver; some initial thoughts

---
title: "surfing"
output: bookdown::html_document2
---

```{r setup, include=FALSE}
set.seed(0)
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("magrittr")
options("digits" = 5)
```

(This document is in two parts.  The first is a slightly edited
version of Tim Driver's dissertation on surfing statistics, the second
a short analysis---using new software not available to Tim---of a
dataset he provided).

Abstract: As surfing continuous to grow as a professional sport, there
is a lack of analysis on competition results. This report applies
generalized Bradley-Terry models to get the maximum likelihood of
surfer’s strengths in competitions with powerful waves. By gathering
competition data, analysis can be conducted in R using the `hyper2`
package, resulting in the surfer’s strengths. Using this methodology,
weaker surfers were able to be identified when considering all
powerful wave locations, however, there was not enough data to get a
significant result for single events. Furthermore, it is hoped that a
result forecasting model can be produced using similar techniques.

Competitive surfing involves surfers competing against one or two
other surfers in heats lasting 30-50 minutes. Points are awarded on a
10-point scale, with [0.0- 1.9] a poor score, [2.0-3.9] a fair score,
[4.0 - 5.9] an average score, [6.0 – 7.9] a good score and finally
[8.0 – 10.0] an Excellent score. The highest two waves ridden by an
individual are then combined to produce an individual heat total, the
highest of which then wins the heat [1].

Five judges rate each of the waves ridden by a surfer and award points
based off of a range of criteria including commitment and degree of
difficulty, innovation and progress of manoeuvres, combination of
major manoeuvre, variety of manoeuvres and speed, power and flow
[1]. The top score and bottom score of the judges are removed and the
remaining three judges scores are averaged to give the final score for
the surfer.

Structure of the event is as follows, initially there is a seeding
round consisting of 12 heats with three surfers competing in each
heat. The top two surfers of each heat advance to the round of 32
while the 12 surfers who failed to finish first or second are placed
in an elimination round. The elimination round is again four,
three-man heats with last placed in each heat being eliminated from
the competition. Surfers are then in man on man heats, with the loser
of each heat eliminated from the competition. This format then repeats
with a round of 16, quarter finals, semi-finals and finally the
finals.

The World Surf League (WSL) is the main governing body of surfing
competitions. The WSL conducts a world tour that the best ranked 34
surfers compete in. At each tour venue, two wild cards also enter the
competition. The wildcard surfers are chosen from the main sponsors of
the event and the winner of a local competition that is held before
the main event starts. On occasion, if or more of the 34 surfers are
injured, an injury replacement surfer will be selected by the WSL to
be added to the competition to keep the heat format the same.

The 2019 tour saw competitive events held at eleven different surf
locations. In each of these locations the wave type of wave
differs. Some events are known to be more favourable to those who have
a high number of manoeuvres at their disposal and others are more
suited to surfers who regularly surf bigger waves. Of the 11 locations
in 2019, three competitions stand out as the waves at locations are
known for the size and power. At these locations, judges score riding
inside the wave ("tubed") higher than other turn or air style
manoeuvres. The locations are Pipeline in Hawaii, Teahupo’o in Tahiti
and Margaret River in Western Australia.

At these locations, ocean swell comes from deep water and hits shallow
reefs. The lower part of the wave slows down due to the interference
with the ocean floor while the top part keeps a similar velocity and
eventually the wave breaks. As the top part of the wave is moving
forward at a larger velocity, it breaks closer to the reef, causing
there to be a gap between the top part and the bottom close to the
surface of the water [2].  Surfing in this critical zone is when you
are getting tubed and subsequently scored highly, due to the
commitment and degree of difficulty involved.

Analysis on competitive surfing has not attracted much attention since
surfing became a professional sport 1960s [3]. With the recent
inclusion in the Olympic games, surfing continues to grow in
professionalism. High-level analysis of surfing needs to develop to
allow new insights to be discovered about surfers performance in
competitions to continue this trend of professionalism in the sport.

[discussion of Bradley-Terry elided]

# Literature Review

Currently relevant research into competitive surfing has been limited
to the scoring of ariel manoeuvres compared with other manoeuvres in
events [10] and analysis in the difference in average wave scores for
the top 10 and bottom 10 surfers using a One-way ANOVA method [11].

In the work discussed in the paper "Analysis of Manoeuvres and Scoring
in Competitive Surfing" [10], analysis was done on the waves ridden in
the quarter finals, semi-finals and final heats of the 2012 World
Championship Tour at all locations. On this year of the Tour, Western
Australia was not included and so Margaret River was not surfed. Waves
which involved a tube or arial manoeuvre were shown to score higher
than doing other turn style on the wave across all locations and have
a much lower completion rate than other manoeuvres. This shows the
importance for getting tubed at elite surfing levels to win heats and
thus win competitions. As this research piece is focusing on the Tour
locations that have waves with higher opportunities to get tubed, poor
results here are indicative of poor ability to get tubed.

In the research done in the paper "Scoring Analysis of the Men's 2013
World Championship Tour of Surfing" [11], scores from the top 10
surfers and bottom 10 surfers 2013 World Championship Tour are
analysed to determine if there is any difference in the average wave
score and average total heat score. A one-way ANOVA was performed on
the two groups to determine if there was any major discrepancies. It
was found that the top 10 surfers had a score of 14.7% higher than the
bottom 10 surfers and the standard deviation of the top surfers score
was 9.39% less than that of the bottom 10 surfers. The authors then
took this work to surf coaches to highlight the importance of
performing better as an increase in the average wave score by 10% for
a bottom ranked surfer would greatly improve their overall ranking on
the Tour.

The method of approach in this research report is to use Bradley-Terry
models [4] for paired comparisons to analyse the results of
heats.  Bradley-Terry models are popular for paired comparisons between
competitors.  As earlier heats in the competition have three man heats
a general Bradley Terry model will be used to look at this comparison.

Existing research has been conducted on other sports using
Bradley-Terry models, including tennis [12], baseball [13] and
American football [14].

The existing research conducted by McHale and Morton [12] titled "A
Bradley-Terry type model for forecasting tennis match results" looked
at the top tier results from the Association of Tennis Professionals
(ATP) for 9 seasons between 2000-2008. The data used was the amount of
games in sets won by competitors when playing against other
competitors. Information about the surface that the court was played
on, be it clay, grass or a hardcourt was also included and a form
decaying factor based of previous results. Once

the modified Bradley-Terry models were applied to the data, the model
produced an alternative ranking system. Five of the top 15 official
ATP ranking were not included in the top 15 players returned by the
model. They then go on to use this new ranking to forecast the results
of upcoming tennis matches.

The existing research by Butler and Whelan [14] describes modification
to the original Bradley-Terry model and applies the model to American
Football results in the round robin stages of the 1998 Western
Athletic conference. In the round robin stage, teams compete against
all the teams in their division, along with one team in the other
division. The divisions are separated geographically, and each
contained 8 teams. The model initially identified if any pairs of
teams i and j were strictly better than the other through analysing
previous matchups. If results existed that did not display
transitivity relationship, then Bradley-Terry models were applied to
these subsets in order to determine a superior team.

The original Bradley-Terry model developed only considered paired data
and did not describe a case with more than two competitors. This is
appropriate for the later heats but fails in the earlier three-man
heat competitions. A general model needs to be applied as described by
Luce [7], to consider the case for three or more.

The generalized Bradley-Terry model has been put into R under the
Hyper2 package by Robin K. S. Hankin [15] and has since been used for
a number of projects including analysis of competitive cooking and
rowing [9]. The research done on the rowing piece analysed the men’s
single scull in the 2016 Olympic games. 25 races were run at this
competition with 31 rowers entering, each with a maximum of six rowers
competing per. If a rower won the race, be it a millisecond or a
minute, they were still recorded the same as the only concern of the
model is the placing in whish the rowers finished. The maximum
likelihood estimate for each rower winning the event was produced,
showing an alternative ranking system to the knock out format at the
Olympics. The models ranking had Drysdale first, Synek second and
Martin third. The actual results in the final were Drysdale first,
Martin second and Synek Third.

# Research Question:

The purpose of this research report is to do analysis using
generalized Bradley-Terry models on the 2019 WSL Tour at competition
locations known to have powerful surf. These locations are Margaret
River, Teahupo'o and Pipeline. Generalized Bradley-Terry models will
be used on the surfer's heat results.

Analysis will be done on each competition individually, a combination
of Margaret River and Teahupo'o as well as a combination of all three
locations. Each individual location will be analysed to create an
alternative ranking system to the competition knockout format. All
three locations will be analysed together to determine a ranking
system for the best surfer in powerful waves, based off these
locations having waves of consequence. Margaret River and Teahupo’o
will be analysed together to investigate an early model used for
forecasting results at these locations, specifically can the results
from these two locations be used to forecast the pipeline competition
results.

[1] "Rules and Regulations" [Online]. Available:
https://www.worldsurfleague.com/pages/rules- and-regulations.

[2] M. Longuet-Higgins, "Progress Toward Understanding How Waves
Break," in Twenty-First Symposium on Naval Hydrodynamics, National
Research Council, Naval Studies Board, Division on Engineering and
Physical Sciences, Commission on Physical Sciences, Mathematics, and
Applications, 1997, pp. 7-29.

[3] D. Booth, "Ambiguities in Pleasure and Discipline: The Development
of Competitive Surfing," Journal of Sport HIstory, 1995.

[4] E. M. Terry and A. R. Bradley, "Rank Analysis of Incomplete
Block Designs: I. The Method of Paired Comparisons," Biometrika,
pp. 324-345, 1952.

[5] D. R. Hunter, "MM algorithms for generalized Bradley-Terry
models," The annals of Statistics, pp. 384-406, 2004.

[6] H. Turner and D. Firth, "Bradley-Terry Models in R: The
BradleyTerry2," Journal of Statistical Software, 2012.

[7] R. D. Luce, Individual choice behaviour: A theoretical analysis,
New York: Wiley, 1959.

[8] R. Plackett, "The Analysis of Permutations," Journal of the Royal
Statistical Society, pp. 193-202, 1975.

[9] R. K. S. Hankin, "Partial rank data with the hyper2 package:
likelihood functions for generalized Bradley-Terry," R Journal, 2017.

[10] L. Lundgren, R. Newton, T. Tran, M. Dunn, S. Nimphius and
J. Sheppard, "Analysis of Manoeuvres and Scoring in Competitive
Surfing," International Journal of Sports Science & Coachin, 2014.

[11] O. Farley, E. Raymond, J. Secomb, B. Ferrier and L. Lundgren,
"Scoring Analysis of the Men’s 2013 World Championship Tour of
Surfing," The International Journal of Aquatic Research and Education,
2015.

[12] I. McHale and A. Morton, "A Bradley-Terry type model for
forecasting tennis match results," International Journal of
Forecasting, pp. 619-630, 2011.

[13] H. Turner and D. Firth, "Bradley-Terry Models in R: The
BradleyTerry2 Package," Journal of Statistical Software, 2011.

[14] K. Butler and W. T. John , "The existence of maximum likelihood
estimates in the Bradley-Terry model and its extensions," 2004.

[15] R. K. S. Hankin, "A Generalization of the Dirichlet
Distribution," Journal of Statistical Software - Vol 33 - Issue 11,
2010.

[16] "Margaret River," World Surf League, 2019. [Online]. Available:
https://www.worldsurfleague.com/events/2019/mct/2913/margaret-river-pro.

[17] World Surf League, "Teahupo'o,"
August-September 2019. [Online]. Available:
https://www.worldsurfleague.com/events/2019/mct/2919/tahiti-pro-teahupoo.

[18] World Surf League, "Pipeline,"
December 2019. [Online]. Available:
https://www.worldsurfleague.com/events/2019/mct/2927/billabong-pipe-masters.

[19] "2019 Surfers," World Surf League, 2019. [Online]. Available:
https://www.worldsurfleague.com/athletes/tour/mct?year=2019.

[Tim's work ends, my work starts]

# Introduction

Competitive surfing involves surfers competing against one or two
other surfers in heats lasting 30-50 minutes.  Points are awarded on a
10-point scale: each of five judges awards points to a surfer riding a
particular wave.

The five judges rate each of the waves ridden by a surfer and award
points based on a range of subjective criteria including commitment
and degree of difficulty, innovation and progress of manoeuvres,
combination of major manoeuvre, variety of manoeuvres and speed, power
and flow.  The top score and bottom score among the five judges are
removed and the remaining three judges' scores are averaged to give
the final score for the surfer for that wave.

Each surfer's aggregate score is given by the average of the two
highest-scoring waves.  The winner of the tournament is the surfer
with the highest aggregate score.

The World Surf League (WSL) is the main governing body of surfing
competitions.  The WSL conducts a world tour in which the best ranked
34 surfers compete; in addition, at each tour venue, two "wildcard"
surfers also enter the competition.

The 2019 WSL tour saw competitive events held at eleven different surf
locations.  In each of these locations the wave type of wave
differs.  Some events are known to be more favourable to those who have
a high number of manoeuvres at their disposal and others are more
suited to surfers who regularly surf bigger waves.

Of the 11 locations
in 2019, three competitions stand out as the waves at locations are
known for the size and power.

At these locations, judges score riding
inside the wave (“tubed”) higher than other turn or air style
manoeuvres.  The locations are Pipeline in Hawaii, Teahupo’o in Tahiti
and Margaret River in Western Australia.

```{r}
show <- function(filename){
  unlist(lapply(strsplit(readLines(filename)," "),paste,collapse=" "))
}					       
show("surfing_minimal.txt")
```

The first line shows that `Florence` came first, `Andre` second, and
so on.  We can convert this dataset into a Plackett-Luce likelihood
function as follows:

```{r definerowing, cache=TRUE}

f <- function(filename){
  o <- strsplit(readLines(filename)," ")
  H <- hyper2()
  for(v in o){H <- H + race(v)}
  return(H)
}

surfing <- f("surfing_minimal.txt")
pnames(surfing)
head(surfing)
```

Object `surfing` is a loglikelihood function on the strengths of the
competitors.  Our first step is to find the evaluate:

```{r maxpcalc,cache=TRUE}
surfing_maxp <- maxp(surfing)
surfing_maxp
sort(surfing_maxp)
```

and we may display it using a dot chart:

```{r showsurfingmaxppie, echo=FALSE, fig.cap="dot chart of result"}
dotchart(surfing_maxp)
dotchart(log10(surfing_maxp))
pie(surfing_maxp)
```

```{r pval,cache=TRUE}
equalp.test(surfing)
```

### Package dataset {-}

Following lines create `surfing.rda`, residing in the `data/` directory
of the package.

```{r,label=savesurfiningingdataset}
save(surfing,surfing_maxp,file="surfing.rda")
```

### 2021-05-04 inst/surfing.Rmd
surfing

@@
 
-Of the 11 locations
-in 2019, three competitions stand out as the waves at locations are
-known for the size and power.
-
-At these locations, judges score riding
-inside the wave (“tubed”) higher than other turn or air style
-manoeuvres.  The locations are Pipeline in Hawaii, Teahupo’o in Tahiti
-and Margaret River in Western Australia.
+```{r}
+head(read.table("players.txt",header=TRUE))
+```
 
 ```{r}
-show <- function(filename){
-  unlist(lapply(strsplit(readLines(filename)," "),paste,collapse=" "))
-}					       
-show("surfing_minimal.txt")
+o <- strsplit(readLines("wct.txt")," ")
+jj <- unlist(lapply(o,paste,collapse=" "))
+head(jj)
+tail(jj)
 ```
 
-The first line shows that `Florence` came first, `Andre` second, and
-so on.  We can convert this dataset into a Plackett-Luce likelihood
+The first line shows that, at one of the heats in `wct01`, `Colapinto`
+came first, `Bailey` second, and `Wright` came third.  The last line
+shows that, at `wct11`, `Ferreira` came first and `Medina` came
+second.  We can convert this dataset into a Plackett-Luce likelihood
 function as follows:
@@
 
-f <- function(filename){
-  o <- strsplit(readLines(filename)," ")
-  H <- hyper2()
-  for(v in o){H <- H + race(v)}
-  return(H)
-}
-
-surfing <- f("surfing_minimal.txt")
-pnames(surfing)
-head(surfing)
+surfing <- hyper2()
+for(v in o){surfing <- surfing + race(v[-1])} # first entry is venue
+summary(surfing)
 ```

### 2021-05-04 inst/surfing.Rmd
more work, exclude wildcards and Vries

@@
 ```{r}
-head(read.table("players.txt",header=TRUE))
+players <- read.table("players.txt",header=TRUE)
+head(players)
+table(players$nation)
+players$name
+wildcards <- players$name[-(1:36)]
+wildcards
 ```
@@
 second.  We can convert this dataset into a Plackett-Luce likelihood
-function as follows:
+function but first have to remove the wildcards and also player Vries
+whose maximum likelihood strength is zero.
 
-```{r definerowing, cache=TRUE}
-
+```{r definesurfing, cache=TRUE}
+remove <- c(wildcards, "Vries")
 surfing <- hyper2()
-for(v in o){surfing <- surfing + race(v[-1])} # first entry is venue
+for(v in o){
+   r <- v[-1]  # first entry is venue
+   numberretained <- sum(!(r %in% remove))
+   if(numberretained > 1){
+     r <- r[!(r %in% remove)]
+     surfing <- surfing + race(r)
+    }
+}	
 summary(surfing)

### 2021-05-05 inst/eurodance.Rmd
more work

@@
 
-The file `eurodance_2008.txt`, used below, is copied from
-"Eurovision Dance Contest 2008," Wikipedia, accessed March 31, 2021.
+The file `eurodance2008.txt`, used below, is copied from
+"Eurovision Dance Contest 2008", Wikipedia, accessed March 31, 2021.
 First we specify the matrix as appearing in the Wikipedia page:
@@
 ```{r, specifymatrix}
-eurodance_table <- as.matrix(read.table("eurovision_dance_2008.txt"))
+eurodance_table <- as.matrix(read.table("eurodance2008.txt"))
 eurodance_table
@@
 
-Note the dominance of Poland.  Now we consider the jury's vote:
+Note the dominance of Poland.  We verify consistency:
+
+```{r,cache=TRUE}
+consistency(eurodance)
+```
+
+Now we consider the jury's vote:
 

### 2021-05-06 inst/eurovision.Rmd
consistency check

@@
 
+Observe the small estimated strength for Czechoslovakia (`CZ`) at
+about `1e-06`.  This is consistent with the second row of `preference`
+which shows that noone gave them any points.
+
+```{r,consistencycheck,cache=TRUE,fig.cap="consistency check: maximum likelihood obtained two different ways.
+Exact agreement corresponds to points lying on the diagonal line.
+Note the small estimated strength of Czechoslovakia"}
+consistency(eurovision)
+```
+
 Now we can check null of equal strengths:

### 2021-05-06 inst/surfing.Rmd
consistency check

@@
 second.  We can convert this dataset into a Plackett-Luce likelihood
-function but first have to remove the wildcards and also player Vries
-whose maximum likelihood strength is zero.
+function but first have to remove the wildcards and also player
+`Vries` whose maximum likelihood strength is zero.
 
@@
 
+and the second step is to establish its consistency:
+
+```{r consistencycheck,cache=TRUE}
+consistency(surfing)
+```
+
 and we may display it using a dot chart:

### 2021-05-07 inst/surfing.Rmd
better writing

@@
 
-Competitive surfing involves surfers competing against one or two
-other surfers in heats lasting 30-50 minutes.  Points are awarded on a
-10-point scale: each of five judges awards points to a surfer riding a
-particular wave.
+Surfing is a popular and growing activity in terms of both
+participation and viewing.  Competitive surfing involves surfers
+competing against one or two other surfers in heats lasting 30-50
+minutes.  Points are awarded on a 10-point scale: each of five judges
+awards points to a surfer riding a particular wave.
 
-The five judges rate each of the waves ridden by a surfer and award
-points based on a range of subjective criteria including commitment
+The typical format is as follows.  Up to four surfers---the
+competitors---are in the water simultaneously, watched by up to five
+judges.  Subject to surfing etiquette, competitors are free to catch a
+wave at their discresion within the heat's time window, typically
+20-30 minutes.  The judges rate each wave ride and award points to the
+surfer based on a range of subjective criteria including commitment
 and degree of difficulty, innovation and progress of manoeuvres,
-combination of major manoeuvre, variety of manoeuvres and speed, power
-and flow.  The top score and bottom score among the five judges are
-removed and the remaining three judges' scores are averaged to give
-the final score for the surfer for that wave.
+variety of manoeuvres and speed, power and flow.  The top score and
+bottom score among the five judges are removed and the remaining three
+judges' scores are averaged to give the final score for the surfer for
+that wave.  Each surfer's aggregate score is given by the average of
+the two highest-scoring waves.  The winner of the tournament is the
+surfer with the highest aggregate score.
 
-Each surfer's aggregate score is given by the average of the two
-highest-scoring waves.  The winner of the tournament is the surfer
-with the highest aggregate score.
+Such scoring systems are designed to account for the random nature of
+wave quality while reflecting competitors' abilities fairly.
+Differences in wave quality between successive days mean that direct
+comparison of scores between one day and another are not informative
+about competitors' abilities as they are strongly dependent on details
+of wave quality at the time of the competition.  However, for a
+particular heat, the order statistic---that is, which competitor
+scored most highly, second highest, and third---is informative about
+competitors' abilities: the wave environment is common to each surfer.
 
@@
 34 surfers compete; in addition, at each tour venue, two "wildcard"
-surfers also enter the competition.
-
-The 2019 WSL tour saw competitive events held at eleven different surf
-locations.  In each of these locations the wave type of wave
-differs.  Some events are known to be more favourable to those who have
-a high number of manoeuvres at their disposal and others are more
-suited to surfers who regularly surf bigger waves.
+surfers also enter the competition.  The 2019 WSL tour saw competitive
+events held at eleven different surf locations.  In each of these
+locations the wave type of wave differs.  Some events are known to be
+more favourable to those who have a high number of manoeuvres at their
+disposal and others are more suited to surfers who regularly surf
+bigger waves.
 
@@
 
-```{r showsurfingmaxppie, echo=FALSE, fig.cap="dot chart of result"}
+```{r showsurfingmaxppie, echo=FALSE, fig.height=8,fig.cap="dot chart of result"}
 dotchart(surfing_maxp)

### 2021-05-07 inst/surfing.Rmd
citations

@@
 output: bookdown::html_document2
+bibliography: chess.bib
 ---
@@
 ```
+
+@fischer1969
+
+[@fischer1969]
+
+[-@fischer1969]
 
@@
 minutes.  Points are awarded on a 10-point scale: each of five judges
-awards points to a surfer riding a particular wave.
+awards points to a surfer riding a particular wave.  The World Surf
+League (WSL) is the main governing body of surfing competitions.  The
+WSL conducts a world tour in which the best ranked 34 surfers compete;
+in addition, at each tour venue, two "wildcard" surfers also enter the
+competition, who are ignored here.  The 2019 WSL tour saw competitive
+events held at eleven different surf locations.
 
@@
 judges.  Subject to surfing etiquette, competitors are free to catch a
-wave at their discresion within the heat's time window, typically
+wave at their discretion within the heat's time window, typically
 20-30 minutes.  The judges rate each wave ride and award points to the
 surfer based on a range of subjective criteria including commitment
-and degree of difficulty, innovation and progress of manoeuvres,
-variety of manoeuvres and speed, power and flow.  The top score and
-bottom score among the five judges are removed and the remaining three
-judges' scores are averaged to give the final score for the surfer for
-that wave.  Each surfer's aggregate score is given by the average of
-the two highest-scoring waves.  The winner of the tournament is the
-surfer with the highest aggregate score.
+and degree of difficulty, innovation and variety of manoeuvres, and
+transition between manoeuvres.  The top score and bottom score among
+the five judges are removed and the remaining three judges' scores are
+averaged to give the final score for the surfer for that wave.  Each
+surfer's aggregate score is given by the average of the two
+highest-scoring waves.  The winner of the tournament is the surfer
+with the highest aggregate score.
 
@@
 
-The World Surf League (WSL) is the main governing body of surfing
-competitions.  The WSL conducts a world tour in which the best ranked
-34 surfers compete; in addition, at each tour venue, two "wildcard"
-surfers also enter the competition.  The 2019 WSL tour saw competitive
-events held at eleven different surf locations.  In each of these
-locations the wave type of wave differs.  Some events are known to be
-more favourable to those who have a high number of manoeuvres at their
-disposal and others are more suited to surfers who regularly surf
-bigger waves.
+# Bradley-Terry
+
+\subsection{Bradley-Terry and generalizations for rank
+statistics}\label{bradley-terry-and-generalizations-for-rank-statistics}
+
+The Bradley-Terry model [@bradley1954] assigns non-negative strengths
+$p_1,\ldots, p_n$ to each of $n$ competitors in such a way that the
+probability of $i$ beating $j\neq i$ in pairwise competition is
+$\frac{p_i}{p_i+p_j}$; it is conventional to normalize so that $\sum
+p_i=1$.  Further, we use a generalization due to [@luce1959], in which
+the probability of competitor~$i$ winning in a field of $\left\lbrace
+1,\ldots, n\right\rbrace$ is $\frac{p_i}{p_1+\cdots +p_n}$.  Noting
+that there is information in the whole of the finishing order, and not
+just the first across the line, we can follow [@plackett1975] and
+consider the runner-up to be the winner among the remaining
+competitors, and so on down the finishing order. Without loss of
+generality, if the order of finishing were $1,2,3,4,5$, then a
+suitable Plackett-Luce likelihood function would be
+
+\begin{equation}\label{competitors_1_to_5_likelihood}
+\frac{p_1}{p_1+p_2+p_3+p_4+p_5}\cdot
+\frac{p_2}{p_2+p_3+p_4+p_5}\cdot
+\frac{p_3}{p_3+p_4+p_5}\cdot
+\frac{p_4}{p_4+p_5}\cdot
+\frac{p_5}{p_5}
+\end{equation}
+
+and this would be a forward ranking Plackett-Luce model in the
+terminology of [@johnson2020].  A slight generalization allows the
+incorporation of non-finishers (DNF etc). If, say, competitors 4 and 5
+did not finish, we would have
+
+\begin{equation}\label{competitors_1_to_3_only_finished}
+\frac{p_1}{p_1+p_2+p_3+p_4+p_5}\cdot
+\frac{p_2}{p_2+p_3+p_4+p_5}\cdot
+\frac{p_3}{p_3+p_4+p_5}
+\end{equation}
+
+(observe how this likelihood function, while informative about
+$p_4+p_5$, is uninformative about $p_4\left|p_4+p_5\right.$).  We now
+use a technique due to Hankin (2010, 2017) and introduce
+fictional (reified) entities whose nonzero Bradley-Terry strength
+helps certain competitors or sets of competitors under certain
+conditions. The canonical example would be the home-ground advantage
+in association football.  If players (teams) $1,2$ with strengths
+$p_1,p_2$ compete, and if our observation were $a$ home wins and $b$
+away wins for team $1$, and $c$ home wins and $d$ away wins for
+team~$2$, then a suitable likelihood function would be
+
+\[
+\left(\frac{p_1+p_H}{p_1+p_2+p_H}\right)^a
+\left(\frac{p_1}{p_1+p_2+p_H}\right)^b
+\left(\frac{p_2+p_H}{p_1+p_2+p_H}\right)^c
+\left(\frac{p_2+p_H}{p_1+p_2+p_H}\right)^d,
+\]
+
+\noindent where $p_H$ is a quantification of the beneficial home
+ground effect.  Similar techniques have been used to account for the
+first-move advantage in chess, and effective coordination between
+members of doubles tennis teams; we may use a similar device to
+account for (e.g.) wet conditions in Formula 1.
+
+# Literature Review and previous work
+
+Most previous research into competitive surfing is limited to the
+scoring of ariel manoeuvres compared with other manoeuvres in events
+[10] and analysis in the difference in average wave scores for the top
+10 and bottom 10 surfers using a One-way ANOVA method [11].
 

### 2021-05-08 inst/surfing.Rmd
rewording

@@
 ```
-
-@fischer1969
-
-[@fischer1969]
-
-[-@fischer1969]
 
@@
 Surfing is a popular and growing activity in terms of both
-participation and viewing.  Competitive surfing involves surfers
-competing against one or two other surfers in heats lasting 30-50
-minutes.  Points are awarded on a 10-point scale: each of five judges
-awards points to a surfer riding a particular wave.  The World Surf
-League (WSL) is the main governing body of surfing competitions.  The
-WSL conducts a world tour in which the best ranked 34 surfers compete;
-in addition, at each tour venue, two "wildcard" surfers also enter the
-competition, who are ignored here.  The 2019 WSL tour saw competitive
-events held at eleven different surf locations.
+participation and viewing [@warshaw2010].  Competitive surfing
+involves surfers competing against one or two other surfers in heats
+lasting 30-50 minutes [@booth1995].  Points are awarded on a 10-point
+scale: each of five judges awards points to a surfer riding a
+particular wave.  The World Surf League (WSL) is the main governing
+body of surfing competitions [@wsl].  The WSL conducts a world tour in
+which the best ranked 34 surfers compete; in addition, at each tour
+venue, two "wildcard" surfers also enter the competition, who are
+ignored here.  The 2019 WSL tour saw competitive events held at eleven
+different surf locations.
 
@@
 20-30 minutes.  The judges rate each wave ride and award points to the
-surfer based on a range of subjective criteria including commitment
-and degree of difficulty, innovation and variety of manoeuvres, and
-transition between manoeuvres.  The top score and bottom score among
-the five judges are removed and the remaining three judges' scores are
-averaged to give the final score for the surfer for that wave.  Each
-surfer's aggregate score is given by the average of the two
-highest-scoring waves.  The winner of the tournament is the surfer
-with the highest aggregate score.
+surfer based on a range of subjective criteria such as stylish
+execution and transition of manoeuvres, but also credit elements such
+as the difficulty and novelty of the performance.  The top score and
+bottom score among the five judges are removed and the remaining three
+judges' scores are averaged to give the final score for the surfer for
+that wave.  Each surfer's aggregate score is given by the average of
+the two highest-scoring waves.  The winner is the surfer with the
+highest aggregate score.
 
@@
 ```
+
+# references

### 2021-05-09 inst/surfing.Rmd
rejigging for journal of sports analytics

@@
 ```
-
-(This document is in two parts.  The first is a slightly edited
-version of Tim Driver's dissertation on surfing statistics, the second
-a short analysis---using new software not available to Tim---of a
-dataset he provided).
-
-Abstract: As surfing continuous to grow as a professional sport, there
-is a lack of analysis on competition results. This report applies
-generalized Bradley-Terry models to get the maximum likelihood of
-surfer’s strengths in competitions with powerful waves. By gathering
-competition data, analysis can be conducted in R using the `hyper2`
-package, resulting in the surfer’s strengths. Using this methodology,
-weaker surfers were able to be identified when considering all
-powerful wave locations, however, there was not enough data to get a
-significant result for single events. Furthermore, it is hoped that a
-result forecasting model can be produced using similar techniques.
-
-Competitive surfing involves surfers competing against one or two
-other surfers in heats lasting 30-50 minutes. Points are awarded on a
-10-point scale, with [0.0- 1.9] a poor score, [2.0-3.9] a fair score,
-[4.0 - 5.9] an average score, [6.0 – 7.9] a good score and finally
-[8.0 – 10.0] an Excellent score. The highest two waves ridden by an
-individual are then combined to produce an individual heat total, the
-highest of which then wins the heat [1].
-
-Five judges rate each of the waves ridden by a surfer and award points
-based off of a range of criteria including commitment and degree of
-difficulty, innovation and progress of manoeuvres, combination of
-major manoeuvre, variety of manoeuvres and speed, power and flow
-[1]. The top score and bottom score of the judges are removed and the
-remaining three judges scores are averaged to give the final score for
-the surfer.
-
-Structure of the event is as follows, initially there is a seeding
-round consisting of 12 heats with three surfers competing in each
-heat. The top two surfers of each heat advance to the round of 32
-while the 12 surfers who failed to finish first or second are placed
-in an elimination round. The elimination round is again four,
-three-man heats with last placed in each heat being eliminated from
-the competition. Surfers are then in man on man heats, with the loser
-of each heat eliminated from the competition. This format then repeats
-with a round of 16, quarter finals, semi-finals and finally the
-finals.
-
-The World Surf League (WSL) is the main governing body of surfing
-competitions. The WSL conducts a world tour that the best ranked 34
-surfers compete in. At each tour venue, two wild cards also enter the
-competition. The wildcard surfers are chosen from the main sponsors of
-the event and the winner of a local competition that is held before
-the main event starts. On occasion, if or more of the 34 surfers are
-injured, an injury replacement surfer will be selected by the WSL to
-be added to the competition to keep the heat format the same.
-
-The 2019 tour saw competitive events held at eleven different surf
-locations. In each of these locations the wave type of wave
-differs. Some events are known to be more favourable to those who have
-a high number of manoeuvres at their disposal and others are more
-suited to surfers who regularly surf bigger waves. Of the 11 locations
-in 2019, three competitions stand out as the waves at locations are
-known for the size and power. At these locations, judges score riding
-inside the wave ("tubed") higher than other turn or air style
-manoeuvres. The locations are Pipeline in Hawaii, Teahupo’o in Tahiti
-and Margaret River in Western Australia.
-
-At these locations, ocean swell comes from deep water and hits shallow
-reefs. The lower part of the wave slows down due to the interference
-with the ocean floor while the top part keeps a similar velocity and
-eventually the wave breaks. As the top part of the wave is moving
-forward at a larger velocity, it breaks closer to the reef, causing
-there to be a gap between the top part and the bottom close to the
-surface of the water [2].  Surfing in this critical zone is when you
-are getting tubed and subsequently scored highly, due to the
-commitment and degree of difficulty involved.
-
-Analysis on competitive surfing has not attracted much attention since
-surfing became a professional sport 1960s [3]. With the recent
-inclusion in the Olympic games, surfing continues to grow in
-professionalism. High-level analysis of surfing needs to develop to
-allow new insights to be discovered about surfers performance in
-competitions to continue this trend of professionalism in the sport.
-
-[discussion of Bradley-Terry elided]
-
-# Literature Review
-
-Currently relevant research into competitive surfing has been limited
-to the scoring of ariel manoeuvres compared with other manoeuvres in
-events [10] and analysis in the difference in average wave scores for
-the top 10 and bottom 10 surfers using a One-way ANOVA method [11].
-
-In the work discussed in the paper "Analysis of Manoeuvres and Scoring
-in Competitive Surfing" [10], analysis was done on the waves ridden in
-the quarter finals, semi-finals and final heats of the 2012 World
-Championship Tour at all locations. On this year of the Tour, Western
-Australia was not included and so Margaret River was not surfed. Waves
-which involved a tube or arial manoeuvre were shown to score higher
-than doing other turn style on the wave across all locations and have
-a much lower completion rate than other manoeuvres. This shows the
-importance for getting tubed at elite surfing levels to win heats and
-thus win competitions. As this research piece is focusing on the Tour
-locations that have waves with higher opportunities to get tubed, poor
-results here are indicative of poor ability to get tubed.
-
-In the research done in the paper "Scoring Analysis of the Men's 2013
-World Championship Tour of Surfing" [11], scores from the top 10
-surfers and bottom 10 surfers 2013 World Championship Tour are
-analysed to determine if there is any difference in the average wave
-score and average total heat score. A one-way ANOVA was performed on
-the two groups to determine if there was any major discrepancies. It
-was found that the top 10 surfers had a score of 14.7% higher than the
-bottom 10 surfers and the standard deviation of the top surfers score
-was 9.39% less than that of the bottom 10 surfers. The authors then
-took this work to surf coaches to highlight the importance of
-performing better as an increase in the average wave score by 10% for
-a bottom ranked surfer would greatly improve their overall ranking on
-the Tour.
-
-The method of approach in this research report is to use Bradley-Terry
-models [4] for paired comparisons to analyse the results of
-heats.  Bradley-Terry models are popular for paired comparisons between
-competitors.  As earlier heats in the competition have three man heats
-a general Bradley Terry model will be used to look at this comparison.
-
-Existing research has been conducted on other sports using
-Bradley-Terry models, including tennis [12], baseball [13] and
-American football [14].
-
-The existing research conducted by McHale and Morton [12] titled "A
-Bradley-Terry type model for forecasting tennis match results" looked
-at the top tier results from the Association of Tennis Professionals
-(ATP) for 9 seasons between 2000-2008. The data used was the amount of
-games in sets won by competitors when playing against other
-competitors. Information about the surface that the court was played
-on, be it clay, grass or a hardcourt was also included and a form
-decaying factor based of previous results. Once
-
-the modified Bradley-Terry models were applied to the data, the model
-produced an alternative ranking system. Five of the top 15 official
-ATP ranking were not included in the top 15 players returned by the
-model. They then go on to use this new ranking to forecast the results
-of upcoming tennis matches.
-
-The existing research by Butler and Whelan [14] describes modification
-to the original Bradley-Terry model and applies the model to American
-Football results in the round robin stages of the 1998 Western
-Athletic conference. In the round robin stage, teams compete against
-all the teams in their division, along with one team in the other
-division. The divisions are separated geographically, and each
-contained 8 teams. The model initially identified if any pairs of
-teams i and j were strictly better than the other through analysing
-previous matchups. If results existed that did not display
-transitivity relationship, then Bradley-Terry models were applied to
-these subsets in order to determine a superior team.
-
-The original Bradley-Terry model developed only considered paired data
-and did not describe a case with more than two competitors. This is
-appropriate for the later heats but fails in the earlier three-man
-heat competitions. A general model needs to be applied as described by
-Luce [7], to consider the case for three or more.
-
-The generalized Bradley-Terry model has been put into R under the
-Hyper2 package by Robin K. S. Hankin [15] and has since been used for
-a number of projects including analysis of competitive cooking and
-rowing [9]. The research done on the rowing piece analysed the men’s
-single scull in the 2016 Olympic games. 25 races were run at this
-competition with 31 rowers entering, each with a maximum of six rowers
-competing per. If a rower won the race, be it a millisecond or a
-minute, they were still recorded the same as the only concern of the
-model is the placing in whish the rowers finished. The maximum
-likelihood estimate for each rower winning the event was produced,
-showing an alternative ranking system to the knock out format at the
-Olympics. The models ranking had Drysdale first, Synek second and
-Martin third. The actual results in the final were Drysdale first,
-Martin second and Synek Third.
-
-# Research Question:
-
-The purpose of this research report is to do analysis using
-generalized Bradley-Terry models on the 2019 WSL Tour at competition
-locations known to have powerful surf. These locations are Margaret
-River, Teahupo'o and Pipeline. Generalized Bradley-Terry models will
-be used on the surfer's heat results.
-
-Analysis will be done on each competition individually, a combination
-of Margaret River and Teahupo'o as well as a combination of all three
-locations. Each individual location will be analysed to create an
-alternative ranking system to the competition knockout format. All
-three locations will be analysed together to determine a ranking
-system for the best surfer in powerful waves, based off these
-locations having waves of consequence. Margaret River and Teahupo’o
-will be analysed together to investigate an early model used for
-forecasting results at these locations, specifically can the results
-from these two locations be used to forecast the pipeline competition
-results.
-
-[1] "Rules and Regulations" [Online]. Available:
-https://www.worldsurfleague.com/pages/rules- and-regulations.
-
-[2] M. Longuet-Higgins, "Progress Toward Understanding How Waves
-Break," in Twenty-First Symposium on Naval Hydrodynamics, National
-Research Council, Naval Studies Board, Division on Engineering and
-Physical Sciences, Commission on Physical Sciences, Mathematics, and
-Applications, 1997, pp. 7-29.
-
-[3] D. Booth, "Ambiguities in Pleasure and Discipline: The Development
-of Competitive Surfing," Journal of Sport HIstory, 1995.
-
-[4] E. M. Terry and A. R. Bradley, "Rank Analysis of Incomplete
-Block Designs: I. The Method of Paired Comparisons," Biometrika,
-pp. 324-345, 1952.
-
-[5] D. R. Hunter, "MM algorithms for generalized Bradley-Terry
-models," The annals of Statistics, pp. 384-406, 2004.
-
-[6] H. Turner and D. Firth, "Bradley-Terry Models in R: The
-BradleyTerry2," Journal of Statistical Software, 2012.
-
-[7] R. D. Luce, Individual choice behaviour: A theoretical analysis,
-New York: Wiley, 1959.
-
-[8] R. Plackett, "The Analysis of Permutations," Journal of the Royal
-Statistical Society, pp. 193-202, 1975.
-
-[9] R. K. S. Hankin, "Partial rank data with the hyper2 package:
-likelihood functions for generalized Bradley-Terry," R Journal, 2017.
-
-[10] L. Lundgren, R. Newton, T. Tran, M. Dunn, S. Nimphius and
-J. Sheppard, "Analysis of Manoeuvres and Scoring in Competitive
-Surfing," International Journal of Sports Science & Coachin, 2014.
-
-[11] O. Farley, E. Raymond, J. Secomb, B. Ferrier and L. Lundgren,
-"Scoring Analysis of the Men’s 2013 World Championship Tour of
-Surfing," The International Journal of Aquatic Research and Education,
-2015.
-
-[12] I. McHale and A. Morton, "A Bradley-Terry type model for
-forecasting tennis match results," International Journal of
-Forecasting, pp. 619-630, 2011.
-
-[13] H. Turner and D. Firth, "Bradley-Terry Models in R: The
-BradleyTerry2 Package," Journal of Statistical Software, 2011.
-
-[14] K. Butler and W. T. John , "The existence of maximum likelihood
-estimates in the Bradley-Terry model and its extensions," 2004.
-
-[15] R. K. S. Hankin, "A Generalization of the Dirichlet
-Distribution," Journal of Statistical Software - Vol 33 - Issue 11,
-2010.
-
-[16] "Margaret River," World Surf League, 2019. [Online]. Available:
-https://www.worldsurfleague.com/events/2019/mct/2913/margaret-river-pro.
-
-[17] World Surf League, "Teahupo'o,"
-August-September 2019. [Online]. Available:
-https://www.worldsurfleague.com/events/2019/mct/2919/tahiti-pro-teahupoo.
-
-[18] World Surf League, "Pipeline,"
-December 2019. [Online]. Available:
-https://www.worldsurfleague.com/events/2019/mct/2927/billabong-pipe-masters.
-
-[19] "2019 Surfers," World Surf League, 2019. [Online]. Available:
-https://www.worldsurfleague.com/athletes/tour/mct?year=2019.
-
-[Tim's work ends, my work starts]
 
@@
 
+# Previous related work
+
+@farley2013 analyse surfing results using one-way ANOVA techniques and
+report comparisons between the top 10 and bottom 10 surfers in the
+2013 World Championship Tour and conclude that the top-ranked athletes
+are more consistent than the lower-ranked group. They report that "To
+date, only a limited number of research studies have reported on the
+results of competitive surfing" [p44], and the analysis presented here
+is hoped to change that.
+
+Other statistical research into competitive surfing appears to focus
+on the scoring of aerial manoeuvres compared with other manoeuvres in
+events.  @lundgren2014, for example, report that aerial manoeuvres
+scored higher than other stunts but had a lower completion rate, a
+theme we return to below.
+
 # Bradley-Terry
-
-\subsection{Bradley-Terry and generalizations for rank
-statistics}\label{bradley-terry-and-generalizations-for-rank-statistics}
 
@@
 and this would be a forward ranking Plackett-Luce model in the
-terminology of [@johnson2020].  A slight generalization allows the
-incorporation of non-finishers (DNF etc). If, say, competitors 4 and 5
-did not finish, we would have
-
-\begin{equation}\label{competitors_1_to_3_only_finished}
-\frac{p_1}{p_1+p_2+p_3+p_4+p_5}\cdot
-\frac{p_2}{p_2+p_3+p_4+p_5}\cdot
-\frac{p_3}{p_3+p_4+p_5}
-\end{equation}
-
-(observe how this likelihood function, while informative about
-$p_4+p_5$, is uninformative about $p_4\left|p_4+p_5\right.$).  We now
-use a technique due to Hankin (2010, 2017) and introduce
-fictional (reified) entities whose nonzero Bradley-Terry strength
-helps certain competitors or sets of competitors under certain
-conditions. The canonical example would be the home-ground advantage
-in association football.  If players (teams) $1,2$ with strengths
-$p_1,p_2$ compete, and if our observation were $a$ home wins and $b$
-away wins for team $1$, and $c$ home wins and $d$ away wins for
-team~$2$, then a suitable likelihood function would be
+terminology of [@johnson2020].  We now use a technique due to Hankin
+(2010, 2017) and introduce fictional (reified) entities whose nonzero
+Bradley-Terry strength helps certain competitors or sets of
+competitors under certain conditions.  The original example was the
+home-ground advantage in football.  If players (teams) $1,2$ with
+strengths $p_1,p_2$ compete, and if our observation were $a$ home wins
+and $b$ away wins for team $1$, and $c$ home wins and $d$ away wins
+for team~$2$, then a suitable likelihood function would be
 
@@
 first-move advantage in chess, and effective coordination between
-members of doubles tennis teams; we may use a similar device to
-account for (e.g.) wet conditions in Formula 1.
-
-# Literature Review and previous work
-
-Most previous research into competitive surfing is limited to the
-scoring of ariel manoeuvres compared with other manoeuvres in events
-[10] and analysis in the difference in average wave scores for the top
-10 and bottom 10 surfers using a One-way ANOVA method [11].
+members of doubles tennis teams.
 

### 2021-05-09 inst/surfing.Rmd
consistentification

@@
 ```{r}
-players <- read.table("players.txt",header=TRUE)
+players <- read.table("wct_surfers_table.txt",header=TRUE)
 head(players)
@@
 ```{r,label=savesurfiningingdataset}
-save(surfing,surfing_maxp,file="surfing.rda")
+surfing_table <- players
+save(surfing,surfing_maxp,surfing_table,file="surfing.rda")
 ```

### 2021-05-10 inst/surfing.Rmd
further work on issue #94

--- 2021-05-09 inst/surfing.Rmd
@@
 ```{r}
-players <- read.table("players.txt",header=TRUE)
-head(players)
-table(players$nation)
-players$name
-wildcards <- players$name[-(1:36)]
+surfing_table <- read.table("wct_surfers_table.txt",header=TRUE)
+head(surfing_table)
+table(surfing_table$nation)
+surfing_table$name
+wildcards <- rownames(surfing_table)[-(1:36)]
 wildcards
@@
 
+Now compare the finishing order based on points vs the finishing order
+on likelhood:
+
+```{r ordertransplotplot,fig.cap="Official ranks vs likelihood rank, WCT"}
+ rL <- sort(surfing_maxp,decreasing=TRUE)
+ rL[] <- seq_along(rL)
+ rO <- seq_len(nrow(surfing_table))
+ names(rO) <- rownames(surfing_table)
+ rO
+ rL
+
+# ordertransplot(rO,rL,xlab="offical rank",ylab="likelihood rank",main="WCT 2019")
+```
+
 ### Package dataset {-}
@@
 ```{r,label=savesurfiningingdataset}
-save(surfing,surfing_maxp,file="surfing.rda")
+save(surfing,surfing_maxp,surfing_table,file="surfing.rda")
 ```

### 2021-05-10 inst/surfing.Rmd
ordertransplot

--- 2021-05-09 inst/surfing.Rmd
@@
 ```{r}
-players <- read.table("players.txt",header=TRUE)
-head(players)
-table(players$nation)
-players$name
-wildcards <- players$name[-(1:36)]
+surfing_table <- read.table("wct_surfers_table.txt",header=TRUE)
+surfing_table
+table(surfing_table$nation)
+surfing_table$name
+wildcards <- rownames(surfing_table)[-(1:36)]
 wildcards
@@
 
+Test the null of equal strengths:
+
 ```{r pval,cache=TRUE}
 equalp.test(surfing)
+```
+
+Now compare the finishing order based on points vs the finishing order
+on likelhood:
+
+```{r ordertransplotplot,fig.cap="Official ranks vs likelihood rank, WCT"}
+rL <- sort(surfing_maxp,decreasing=TRUE)
+rL[] <- seq_along(rL)    # likelihood-based rank
+rO <- seq_len(nrow(surfing_table))[1:36]
+names(rO) <- rownames(surfing_table)[1:36]
+rO
+rL
+sort(names(rO))
+sort(names(rO))
+sort(names(rO))==sort(names(rO))
+ordertransplot(rO,rL,xlab="offical rank",ylab="likelihood rank",main="WCT 2019")
 ```

### 2021-05-11 inst/surfing.Rmd
include wct08 (different format)

@@
 competitors' abilities: the wave environment is common to each surfer.
+
+We note in passing that `wct8`, hosted at Lemoore had a different
+format from the other venues in the championship, being held in an
+artificial wave pool in which every wave was essentially identical (up
+to a mirror reflection).  Statistically, there were seven heats of
+either 2, 4, 8, or 16 surfers.  We show below how to incorporate the
+information present in such observations with the remainder of the
+championship using a consistent and intuitive statistical model.
 
@@
 surfing_table <- read.table("wct_surfers_table.txt",header=TRUE)
+options(width=100)
 surfing_table
-table(surfing_table$nation)
-surfing_table$name
 wildcards <- rownames(surfing_table)[-(1:36)]
@@
 
+```{r directlogplotofstrenghts}
+jjO <- surfing_table$points[1:36]
+names(jjO) <- rownames(surfing_table)[1:36]
+jjL <- surfing_maxp
+jjO <- ordertrans(jjO)
+jjL <- ordertrans(jjL)
+plot(jjO,jjL,pch=16,log='xy')
+fit <- lm(log10(jjL)~log10(jjO))
+for(i in seq_along(jjO)){
+text(jjO[i],jjL[i],names(jjL)[i],col='gray',pos=4,cex=0.5)
+}
+abline(fit)
+```
+
 Now compare the finishing order based on points vs the finishing order
@@
 
+```{r dosameptest, cache=TRUE}
+samep.test(surfing, c("Ferreira","Medina"))
+samep.test(surfing, c("Ferreira","Florence"))
+samep.test(surfing, c("Florence","Medina"))
+samep.test(surfing, c("Florence","Medina","Ferreira"))
+```
+
 ### Package dataset {-}

### 2021-05-12 inst/surfing.Rmd
do not include QS surfers

@@
 wildcards
+QS <- rownames(surfing_table)[c(23:33,35:36)]
+QS
 ```
@@
 ```{r definesurfing, cache=TRUE}
-remove <- c(wildcards, "Vries")
+remove <- c(wildcards, QS, "Vries")
+wanted <- which(!(rownames(surfing_table) %in% remove)) # used at directlogplotofstrengths
 surfing <- hyper2()
@@
 ```{r directlogplotofstrenghts}
-jjO <- surfing_table$points[1:36]
-names(jjO) <- rownames(surfing_table)[1:36]
+jjO <- surfing_table$points[wanted]
+names(jjO) <- rownames(surfing_table)[wanted]
 jjL <- surfing_maxp
@@
 rL[] <- seq_along(rL)    # likelihood-based rank
-rO <- seq_len(nrow(surfing_table))[1:36]
-names(rO) <- rownames(surfing_table)[1:36]
+rO <- seq_len(nrow(surfing_table))[wanted]
+names(rO) <- rownames(surfing_table)[wanted]
 rO

### 2021-05-12 inst/surfing.Rmd
consistentify: put year in filename

@@
 ```{r}
-surfing_table <- read.table("wct_surfers_table.txt",header=TRUE)
+surfing_table <- read.table("wct_table2019.txt",header=TRUE)
 options(width=100)
@@
 ```{r}
-o <- strsplit(readLines("wct.txt")," ")
+o <- strsplit(readLines("wct2019.txt")," ")
 jj <- unlist(lapply(o,paste,collapse=" "))

### 2021-05-14 inst/surfing.Rmd
typo

@@
 rL[] <- seq_along(rL)    # likelihood-based rank
-rO <- seq_len(nrow(surfing_table))[wanted]
+rO <- seq_len(nrow(surfing_table[wanted,]))
 names(rO) <- rownames(surfing_table)[wanted]

### 2021-05-15 inst/surfing.Rmd
typos

@@
 surfing_table <- read.table("wct_table2019.txt",header=TRUE)
+nationality <- function(surfers){surfing_table$nation[rownames(surfing_table) %in% surfers]}
+```
+
+```{r}
+type <- read.table("venuetypes2019.txt",header=TRUE)
+type
+beachtype <- function(venue){type$beachtype[which(type$comp==venue)]}
+beachtype("wct05")
+beachtype("wct06")
+
 options(width=100)
@@
 
+# Brazilian null
+
+One suggestion is that Brazilians do worse on reef break and better on
+point and beach break beaches.  We now assess the null that Brazilians
+perform equally well on all types of beaches.
+
+```{r surfing2, cache=TRUE}
+surfing2 <- hyper2()
+for(v in o){
+   venue <- v[1]  # first entry is venue
+   r <- v[-1] 
+   numberretained <- sum(!(r %in% remove))
+   if(numberretained > 1){
+     r <- r[!(r %in% remove)]
+     jj <- race(r)
+     surfers <- pnames(jj)
+     if(any(nationality(surfers) == "BRA")){
+       if((beachtype(venue)=="point") | (beachtype(venue) == "beach")){
+         jj %<>% pwa(surfers[nationality(surfers)=="BRA"],"Brazilian_beachtype")
+       }    	     
+     } 
+     surfing2 %<>% sum(jj)
+   }
+}	
+summary(surfing2)
+```
+
+```{r surfing2max, cache=TRUE}
+maxp(surfing2)
+specificp.gt.test(surfing2,"Brazilian_beachtype",0)
+```
+
+```{r profsuppcalcsurfing2, cache=TRUE}
+p <- seq(from=0.001,by=0.001,len=10)
+l <- profsupp(surfing2,"Brazilian_beachtype",p)
+```
+
+```{r}
+plot(p,l)
+```
+
 ### Package dataset {-}

### 2021-05-15 inst/surfing.Rmd
GLO

@@
 ```{r surfing2max, cache=TRUE}
-maxp(surfing2)
+(surfing2_maxp <- maxp(surfing2))
 specificp.gt.test(surfing2,"Brazilian_beachtype",0)
@@
 ```{r profsuppcalcsurfing2, cache=TRUE}
-p <- seq(from=0.001,by=0.001,len=10)
-l <- profsupp(surfing2,"Brazilian_beachtype",p)
+p <- seq(from=0.0005,to=0.01,by=0.0005)
+l <- profsupp(surfing2,"Brazilian_beachtype",p,relative=FALSE)
+l0 <- loglik(indep(surfing_maxp),surfing)
 ```
@@
 ```{r}
-plot(p,l)
+plot(c(0,p),c(l0,l)-max(l),type='b',xlab="Brazilian beachtype",ylab="support")
+```
+
+```{r profsuppcalcsurfing2wide, cache=TRUE}
+p_wide <- seq(from=0.002,to=0.04,by=0.002)
+l_wide <- profsupp(surfing2,"Brazilian_beachtype",p_wide,relative=FALSE)
+```
+
+```{r}
+plot(c(0,p_wide),c(l0,l_wide)-max(l),type='b',xlab="Brazilian beachtype",ylab="support")
+```
+
+## Generalized log-odds ratios
+
+```{r}
+all_Brazilian_surfers <- rownames(surfing_table)[surfing_table$nation == "BRA"]
+all_Brazilian_surfers
+pnames(surfing2)
+pnames(surfing2) %in% all_Brazilian_surfers
+table(pnames(surfing2) %in% all_Brazilian_surfers)
+Brazilian_surfers <- pnames(surfing2)[pnames(surfing2) %in% all_Brazilian_surfers]
+Brazilian_surfers
+```
+
+```{r,cache=TRUE}
+f <- function(B,...){
+  jj <- specificp.test(surfing2, "Brazilian_beachtype", B)
+  mp <- jj$null_estimate
+  n <- sum(names(mp) %in% all_Brazilian_surfers)
+  GLO <- log(mp["Brazilian_beachtype"]^n/prod(mp[which(names(mp) %in% all_Brazilian_surfers)]))
+  out <- c(B,GLO, support=jj$null_support)
+  names(out) <- c("B","GLO","support")
+  return(out)
+}
+```
+
+```{r calcbrazodr,cache=TRUE}
+braz <- c(0.0003,0.0005,0.0008, 0.001,0.002,0.005,0.01,0.02,0.03,0.05)
+M <- do.call("rbind",lapply(braz,f))
+M[,3] <- M[,3]-max(M[,3])
+M
+```
+
+```{r}
+
+plot(M[,2],M[,3]-max(M[,3]),type='b')
+plot(M[,2],M[,3]-max(M[,3]),type='b',ylim=c(-0.3,0))
 ```
@@
 ```{r,label=savesurfiningingdataset}
-save(surfing,surfing_maxp,surfing_table,file="surfing.rda")
+save(surfing,surfing_maxp,type,surfing_table,file="surfing.rda")
 ```

### 2021-05-20 inst/surfing.Rmd
more work

@@
 ground effect.  Similar techniques have been used to account for the
-first-move advantage in chess, and effective coordination between
-members of doubles tennis teams.
+first-move advantage in chess, and here we apply it to assess the
+Brazilian  preference for reef-breaking waves.
 
@@
 point and beach break beaches.  We now assess the null that Brazilians
-perform equally well on all types of beaches.
+perform equally well on all types of beaches.  To that end, we create
+a reified surfer, "Brazilian_beachtype", whose strength helps
+Brazilian surfers when surfing on their (putative) preferred
+beachtype.
 
@@
 ```{r calcbrazodr,cache=TRUE}
-braz <- c(0.0003,0.0005,0.0008, 0.001,0.002,0.005,0.01,0.02,0.03,0.05)
+braz <- c(0.0003,0.0005,0.0008, 0.001,0.0015,0.002,0.003,0.004,0.005,0.007,0.01,0.015,0.02,0.03,0.05)
 M <- do.call("rbind",lapply(braz,f))
@@
 ```{r}
-
-plot(M[,2],M[,3]-max(M[,3]),type='b')
-plot(M[,2],M[,3]-max(M[,3]),type='b',ylim=c(-0.3,0))
+plot(M[,2],M[,3]-max(M[,3]),type='b',xlab="Brazilian beachtype (log-odds)",ylab="support")
+plot(M[,2],M[,3]-max(M[,3]),type='b',ylim=c(-0.3,0),xlab="Brazilian beachtype (log-odds)",ylab="support")
 ```

### 2021-06-28 inst/surfing.Rmd
begin compatriot analysis

@@
 
+# compatriot selection
+
+```{r calculate_compatriot_monster_LF,cache=TRUE}
+scomp <- hyper2()
+for(v in o){
+   r <- v[-1]  # first entry is venue
+   numberretained <- sum(!(r %in% remove))
+   if(numberretained > 1){
+     r <- r[!(r %in% remove)]
+     ## Now add a compatriot monster if needed
+
+     if(numberretained ==3){
+       nationalities <- c(nationality(r[1]),nationality(r[2]),nationality(r[3]))
+       if(length(unique(nationalities)) == 2){
+         print(nationalities)
+         if(sum(nationality(r[1]) %in% nationalities) == 1){
+           Hjj <- pwa(race(r),pwa=r[1])
+	   print("a")
+	   print(r)
+	   print(nationalities)
+	   print(Hjj)
+         } else if(sum(nationality(r[2]) %in% nationalities) == 1){
+           Hjj <- pwa(race(r),pwa=r[2])
+	   print("b")
+	   print(r)
+	   print(nationalities)
+	   print(Hjj)
+         } else if(sum(nationality(r[3]) %in% nationalities) == 1){
+	   Hjj <- pwa(race(r),pwa=r[3])
+	   print("c")
+	   print(r)
+	   print(nationalities)
+	   print(Hjj)
+         } else {
+           stop("this cannot happen")
+         }
+         scomp <- scomp + Hjj
+       } else {  # no compatriot monster
+         scomp <- scomp + race(r)
+       }
+     }
+   }
+}
+```
+
+```{r maximize_compatriot_monster,cache=TRUE}
+maxs <- maxp(scomp)
+maxs
+```
+
+```{r}
+pie(maxs)
+```
+
+```{r testcompatriotmonster,cache=TRUE}
+specificp.gt.test(scomp,"S",0)
+```
+
 ### Package dataset {-}

### 2021-06-29 inst/surfing.Rmd
compatriot work

--- 2021-05-20 inst/surfing.Rmd
@@
 
-options(width=100)
+options(width=120)
 surfing_table
@@
 ```{r profsuppcalcsurfing2, cache=TRUE}
-p <- seq(from=0.0005,to=0.01,by=0.0005)
+p <- seq(from=0.001,to=0.01,by=0.0005)
 l <- profsupp(surfing2,"Brazilian_beachtype",p,relative=FALSE)
@@
 
+# compatriot selection
+
+```{r calculate_compatriot_monster_LF,cache=FALSE}
+scomp <- hyper2()
+scomp3 <- hyper2()  # scomp3 is the likelihood function for just the races with 3 competitors, 2 of one nationality and one odd-man-out.
+for(v in o){
+   r <- v[-1]  # first entry is venue; "r" for "race"
+   numberretained <- sum(!(r %in% remove))
+   if(numberretained > 1){
+     r <- r[!(r %in% remove)]
+     if(numberretained ==3){
+       nationalities <- c(nationality(r[1]),nationality(r[2]),nationality(r[3]))
+       if(length(unique(nationalities)) == 2){
+         if(sum(nationalities %in% nationality(r[1])) == 1){ # winner is the odd one out
+           Hjj <- pwa(race(r),pwa=r[1])
+           print("odd one out is number 1")
+           print(100*surfing_maxp[r])
+         } else if(sum(nationalities %in% nationality(r[2]) ) == 1){ # second best is the odd one out
+           Hjj <- pwa(race(r),pwa=r[2])
+           print("odd one out is number 2")
+           print(100*surfing_maxp[r])
+         } else if(sum(nationalities %in% nationality(r[3])) == 1){  # third best is the odd one out
+	   Hjj <- pwa(race(r),pwa=r[3])
+           print("odd one out is number 3")
+           print(100*surfing_maxp[r])
+         } else {
+           stop("this cannot happen")
+         }
+         print(Hjj)
+         print("---")
+         scomp <- scomp + Hjj      # add compatriot-monster modified likelihood 
+         scomp3 <- scomp3 + Hjj
+       } else {                    # 3 in race, but no compatriot monster...
+         scomp <- scomp + race(r)  # ... so use regular Plackett-Luce for 3-competitor race
+       }
+     } else {                    # if(numberretained==3) closes...
+       scomp <- scomp + race(r)  # numberretained !=3; use regular Plackett-Luce
+     }
+  } # if(numberretained>1) closes
+} # for(v in o) loop closes
+```
+
+```{r maximize_compatriot_monster,cache=FALSE}
+maxs <- maxp(scomp)
+maxs
+```
+
+```{r}
+pie(maxs)
+```
+
+```{r testcompatriotmonster,cache=TRUE}
+specificp.gt.test(scomp,"S",0,n=100,maxtry=100)
+```
+
+Now profile likelihood:
+
+```{r profsuppcalcscompatriotmonster, cache=TRUE}
+p <- c(0.0002,seq(from=0.001,to=0.02,len=10))
+l <- profsupp(scomp,"S",p,relative=FALSE)
+```
+
+plot it:
+
+```{r}  
+plot(p,l,type='b',xlab="compatriot support",ylab="support")
+```
+
+```{r}
+scomp3
+maxp(scomp3)
+specificp.gt.test(scomp3,'S',0)
+````
+
+We can do a much more basic test on just the entries with three competitors comprising
+two of one nationality and one of a different nationality:
+
+```{r basicanalysis}
+##     Dora Ferreira   Slater 
+##   2.5741  12.5929   3.8933  (3)
+
+##     Smith    Buchan Freestone 
+##    5.2764    2.0552    1.8673  (1)
+
+##   Slater   Andino Callinan 
+##   3.8933   5.9190   1.8408  (3)
+
+## Crisanto   Wilson   Ibelli 
+##   2.1644   2.7100   2.2479  (2)
+
+##     Moniz    Wright Freestone 
+##    3.0250    3.6502    1.8673 (1)
+
+## Carmichael  Freestone     Flores 
+##     1.8855     1.8673     2.0480 (3)
+
+##    Bourez    Wright Freestone 
+##    1.9068    3.6502    1.8673 (1)
+
+## Colapinto    Flores    Bourez 
+##    2.9871    2.0480    1.9068  (1)
+
+## Ferreira   Morais   Ibelli 
+##  12.5929   1.0817   2.2479 (2)
+
+## Freestone    Flores    Bourez 
+##    1.8673    2.0480    1.9068  (1)
+
+##     Dora Ferreira   Morais 
+##   2.5741  12.5929   1.0817 (1)
+
+M <- matrix(c(
+  2.5741,12.5929,3.8933,3
+, 5.2764, 2.0552,1.8673,1
+, 3.8933, 5.9190,1.8408,3
+, 2.1644, 2.7100,2.2479,2
+, 3.0250, 3.6502,1.8673,1
+, 1.8855, 1.8673,2.0480,3
+, 1.9068, 3.6502,1.8673,1
+, 2.9871, 2.0480,1.9068,1
+,12.5929, 1.0817,2.2479,2
+, 1.8673, 2.0480,1.9068,1
+, 2.5741,12.5929,1.0817,1),byrow=TRUE, ncol=4)
+
+for(i in seq_len(nrow(M))){ M[i,1:3] <-   M[i,1:3]/sum(M[i,1:3])}
+print(M)  # rows 1-3 of M are strengths of the first, second, third
+          # competitors.  Row 4 is the place of the odd-man-out.  Thus
+          # on row 1, the odd-man-out came third, and on row 2, the
+          # odd-man-out came first.
+
+p <- M[cbind(1:11, M[,4])]
+print(p) 
+                                        ## element i of p is the probability of the odd-man-out
+                                        ## winning, given the maximum likelihood estimate.
+
+ww <-table(replicate(1e5,sum(rbinom(11,1,p))))
+print(ww)
+sum(ww[7:10])/sum(ww) # pvalue for observation of six wins by the odd-man-out
+```
+
 ### Package dataset {-}

### 2021-06-29 inst/surfing.Rmd
work

@@
 ```{r}  
-plot(p,l,type='b',xlab="compatriot support",ylab="support")
+plot(p,l,type='b',xlab="compatriot support",ylab="support",main="profile support for S, scomp")
 ```
 
-```{r}
+```{r cache=TRUE}
 scomp3
 maxp(scomp3)
-specificp.gt.test(scomp3,'S',0)
+specificp.gt.test(scomp3,'S',0,n=100)
 ````
+
+Now profile likelihood:
+
+```{r profsuppcalcscomp3atriotmonster, cache=TRUE}
+p <- c(0.0002,seq(from=0.001,to=0.02,len=10))
+l <- profsupp(scomp3,"S",p,relative=FALSE,n=100)
+```
+
+plot it:
+
+```{r}  
+plot(p,l,type='b',xlab="compatriot support",ylab="support",main="profile support for S, scomp3")
+```
 

### 2021-06-30 inst/surfing.Rmd
name chunks
whitespace

--- 2021-06-29 inst/surfing.Rmd
@@
 
-```{r}
+```{r plotpthink}
 plot(c(0,p),c(l0,l)-max(l),type='b',xlab="Brazilian beachtype",ylab="support")
@@
 
-```{r}
+```{r plotwidep}
 plot(c(0,p_wide),c(l0,l_wide)-max(l),type='b',xlab="Brazilian beachtype",ylab="support")
@@
 
-```{r}
+```{r allbrazsurf}
 all_Brazilian_surfers <- rownames(surfing_table)[surfing_table$nation == "BRA"]
@@
 
-```{r,cache=TRUE}
+```{r biglikefun,cache=TRUE}
 f <- function(B,...){
@@
 
-```{r}
+```{r plotBraz}
 plot(M[,2],M[,3]-max(M[,3]),type='b',xlab="Brazilian beachtype (log-odds)",ylab="support")
@@
 
-```{r}
+```{r plotpiemaxs}
 pie(maxs)
@@
 
-```{r}  
-plot(p,l,type='b',xlab="compatriot support",ylab="support")
+```{r plotprofilescomp}  
+plot(p,l,type='b',xlab="compatriot support",ylab="support",main="profile support for S, scomp")
 ```
 
-```{r}
+```{r scomp3maxp, cache=TRUE}
 scomp3
 maxp(scomp3)
-specificp.gt.test(scomp3,'S',0)
+specificp.gt.test(scomp3,'S',0,n=100)
 ````
+
+Now profile likelihood:
+
+```{r profsuppcalcscomp3atriotmonster, cache=TRUE}
+p <- c(0.0002,seq(from=0.001,to=0.02,len=10))
+l <- profsupp(scomp3,"S",p,relative=FALSE,n=100)
+```
+
+plot it:
+
+```{r plotprofilescomp3}  
+plot(p,l,type='b',xlab="compatriot support",ylab="support",main="profile support for S, scomp3")
+```
 

### 2021-07-14 inst/surfing.Rmd
document surfing dataset

@@
 ```{r}
-type <- read.table("venuetypes2019.txt",header=TRUE)
-type
-beachtype <- function(venue){type$beachtype[which(type$comp==venue)]}
+surfing_venuetypes <- read.table("venuetypes2019.txt",header=TRUE)
+surfing_venuetypes
+beachtype <- function(venue){surfing_venuetypes$beachtype[which(surfing_venuetypes$comp==venue)]}
 beachtype("wct05")
@@
 ```{r,label=savesurfiningingdataset}
-save(surfing,surfing_maxp,type,surfing_table,file="surfing.rda")
+save(surfing,surfing_maxp,surfing_venuetypes,surfing_table,file="surfing.rda")
 ```

### 2021-07-20 inst/surfing.Rmd
analyse using just the pairs and compare with whole dataset

@@
 surfing <- hyper2()
+surfing_pairs <- hyper2()
 for(v in o){
@@
      surfing <- surfing + race(r)
+     if(numberretained == 2){
+     r <- r[!(r %in% remove)]
+     surfing_pairs <- surfing_pairs + race(r)
+     }	      
     }
@@
 summary(surfing)
+summary(surfing_pairs)
+```
+
+```{r showsurfing_pairs}
+summary(surfing_pairs)
 ```
@@
 
+Now compare `surfing` with `surfing_pairs`.  First calculate
+
+```{r maxpcalcpairs,cache=TRUE}
+surfing_pairs_maxp <- maxp(surfing_pairs)
+```
+
+then plot
+
+```{r plotmaxppairs}
+surfing_pairs_maxp
+surfing_maxp
+par(pty='s')
+plot(surfing_pairs_maxp,surfing_maxp,asp=1,pch=16)
+abline(0,1)
+plot(surfing_pairs_maxp,surfing_maxp,asp=1,pch=16,log='xy',xlim=c(0.01,0.2),ylim=c(0.01,0.2))
+abline(0,1)
+plot(log(surfing_pairs_maxp/surfing_maxp),pch=16)
+abline(0,0)
+```
+
 and the second step is to establish its consistency:
@@
 consistency(surfing)
+consistency(surfing_pairs)
+```
+
+```{r findhessiansurfing,cache=TRUE}
+e_all <- eigen(hessian(surfing,indep(surfing_maxp)),T,T)$values
+e_pairs <- eigen(hessian(surfing_pairs,indep(surfing_pairs_maxp)),T,T)$values
+```
+
+```{r showhessiansurfing}
+e_all
+e_pairs
+f <- function(x){prod(sort(x)[-1])}
+f(e_all)
+f(e_pairs)
+f(e_all)/f(e_pairs)
 ```
@@
 and we may display it using a dot chart:
+
+```{r echo=FALSE}
+`mydotchart` <- 
+function (x, labels = NULL, groups = NULL, gdata = NULL, offset = 1/8, 
+    ann = par("ann"), xaxt = par("xaxt"), frame.plot = TRUE, 
+    log = "", cex = par("cex"), pt.cex = cex, pch = 21, gpch = 21, 
+    bg = par("bg"), color = par("fg"), gcolor = par("fg"), lcolor = "gray", 
+    xlim = range(x[is.finite(x)]), main = NULL, xlab = NULL, 
+    ylab = NULL, ...) 
+{
+    opar <- par("mai", "mar", "mgp", "cex", "yaxs")
+    on.exit(par(opar))
+    par(cex = cex, yaxs = "i")
+    if (!is.numeric(x)) 
+        stop("'x' must be a numeric vector or matrix")
+    n <- length(x)
+    if (is.matrix(x)) {
+        if (is.null(labels)) 
+            labels <- rownames(x)
+        if (is.null(labels)) 
+            labels <- as.character(seq_len(nrow(x)))
+        labels <- rep_len(labels, n)
+        if (is.null(groups)) 
+            groups <- col(x, as.factor = TRUE)
+        glabels <- levels(groups)
+    }
+    else {
+        if (is.null(labels)) 
+            labels <- names(x)
+        glabels <- if (!is.null(groups)) 
+            levels(groups)
+        if (!is.vector(x)) {
+            warning("'x' is neither a vector nor a matrix: using as.numeric(x)")
+            x <- as.numeric(x)
+        }
+    }
+    linch <- if (!is.null(labels)) 
+        max(strwidth(labels, "inch"), na.rm = TRUE)
+    else 0
+    if (is.null(glabels)) {
+        ginch <- 0
+        goffset <- 0
+    }
+    else {
+        ginch <- max(strwidth(glabels, "inch"), na.rm = TRUE)
+        goffset <- offset
+    }
+    nmai <- opar[["mai"]]
+    if (ann) 
+        nm.2 <- nmai[2L]
+    if (!(is.null(labels) && is.null(glabels))) {
+        yi <- if (is.null(ylab) || !ann) 
+            0
+        else offset
+        nm.2 <- nmai[4L] + max(yi + linch + goffset, ginch) + 
+            1/16
+        if (nmai[2L] < nm.2) {
+            nmai[2L] <- nm.2
+            par(mai = nmai)
+        }
+    }
+    if (is.null(groups)) {
+        o <- seq_len(n)
+        y <- o
+        ylim <- c(0, n + 1)
+    }
+    else {
+        o <- sort.list(as.numeric(groups), decreasing = TRUE)
+        x <- x[o]
+        groups <- groups[o]
+        color <- rep_len(color, length(groups))[o]
+        lcolor <- rep_len(lcolor, length(groups))[o]
+        pch <- rep_len(pch, length(groups))[o]
+        of.1 <- cumsum(c(0, diff(as.numeric(groups)) != 0))
+        y <- seq_len(n) + 2 * of.1
+        ylim <- range(0, y + 2)
+    }
+#    plot.window(xlim = xlim, ylim = ylim, log = log)
+    lheight <- par("csi")
+    if (!is.null(labels)) {
+        loffset <- (linch + 0.1)/lheight
+        mtext(labels[o], side = 2, line = loffset, at = y, adj = 0, 
+            col = color, las = 2, cex = cex, ...)
+    }
+    abline(h = y, lty = "dotted", col = lcolor)
+    points(x, y, pch = pch, col = color, bg = bg, cex = pt.cex/cex)
+    if (!is.null(groups)) {
+        gpos <- rev(cumsum(rev(tapply(groups, groups, length)) + 
+            2) - 1)
+        ginch <- max(strwidth(glabels, "inch"), na.rm = TRUE)
+        goffset <- (max(linch + offset, ginch, na.rm = TRUE) + 
+            1/16)/lheight
+        mtext(glabels, side = 2, line = goffset, at = gpos, adj = 0, 
+            col = gcolor, las = 2, cex = cex, ...)
+        if (!is.null(gdata)) {
+            abline(h = gpos, lty = "dotted")
+            points(gdata, gpos, pch = gpch, col = gcolor, bg = bg, 
+                cex = pt.cex/cex, ...)
+        }
+    }
+    axis(1, xaxt = xaxt)
+    if (frame.plot) 
+        box()
+    if (ann) {
+        title(main = main, xlab = xlab, ...)
+        mgp <- par("mgp")
+        par(mgp = c(max(mgp[1], nm.2/lheight - 1.5), mgp[-1]))
+        title(ylab = ylab, ...)
+    }
+    invisible()
+}
+```
 
@@
 pie(surfing_maxp)
+jj <- log(surfing_maxp)
+kk <- log(surfing_pairs_maxp)
+names(jj) <- NULL
+names(kk) <- NULL
+l <- c(-4.6, -1.5)
+dotchart(log(surfing_maxp),pch=NA,xlim=l)
+mydotchart(jj,col="red" ,pch=16,xlim=l,xaxt='n',frame.plot=FALSE)
+mydotchart(kk,col="blue",pch=16,xlim=l,xaxt='n',frame.plot=FALSE)
 ```
@@
 
-```{r calculate_compatriot_monster_LF,cache=FALSE}
+```{r calculate_compatriot_monster_LF,cache=TRUE}
 scomp <- hyper2()
@@
 
-```{r maximize_compatriot_monster,cache=FALSE}
+```{r maximize_compatriot_monster,cache=TRUE}
 maxs <- maxp(scomp)

### 2021-07-26 inst/ordertrans.Rmd
more explanation

@@
 library("magrittr")
+```
+
+```{r}
+ordertrans
 ```
@@
 ranks respectively---both of which may appear in any order---to find
-the competitor or rank you are interested in.  What we want is to
+the competitor or rank you are interested in.  To find the rank vector
+(that is, who came first, second etc), one would use `sort()`:
+
+```{r}
+sort(x)
+```
+
+But this is suboptimal to find the order vector ["where did a
+particular competitor come?"].  For the order vector, we want to
 rearrange the elements of `x` so that the names are in alphabetical

### 2021-08-10 inst/counterstrike.Rmd
use sticker in Rmd

@@
 ```
+
+<p style="text-align: right;">
+![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
+</p>
+
 # Abstract

### 2021-08-10 inst/surfing.Rmd
typo

@@
 
-```{r directlogplotofstrenghts}
+```{r directlogplotofstrengths,cache=TRUE}
 jjO <- surfing_table$points[wanted]
@@
 jjL <- ordertrans(jjL)
-plot(jjO,jjL,pch=16,log='xy')
+plot(jjO,jjL,pch=16,log='xy',xlab='points',ylab='maximum likelihood strength')
 fit <- lm(log10(jjL)~log10(jjO))
+summary(fit)
 for(i in seq_along(jjO)){

### 2021-08-10 inst/surfing_jsa.tex
start work on issue #105

\documentclass{article}
\usepackage{amssymb}
\usepackage{graphicx}
\begin{document}
\bibliographystyle{plain}

\title{Analysis of professional surfing tournaments with generalized Bradley-Terry}

\section{Introduction}

Surfing is a popular and growing activity in terms of both
participation and viewing \cite{warshaw2010}.  Competitive surfing
involves surfers competing against one or two other surfers in heats
lasting 30-50 minutes \cite{booth1995}.  Points are awarded on a
10-point scale: each of five judges awards points to a surfer riding a
particular wave.  The World Surf League (WSL) is the main governing
body of surfing competitions \cite{wsl}.  The WSL conducts a world tour in
which the best ranked 34 surfers compete; in addition, at each tour
venue, two "wildcard" surfers also enter the competition, who are
ignored here.  The 2019 WSL tour saw competitive events held at eleven
different surf locations.

The typical format is as follows.  Up to four surfers---the
competitors---are in the water simultaneously, watched by up to five
judges.  Subject to surfing etiquette, competitors are free to catch a
wave at their discretion within the heat's time window, typically
20-30 minutes.  The judges rate each wave ride and award points to the
surfer based on a range of subjective criteria such as stylish
execution and transition of manoeuvres, but also credit elements such
as the difficulty and novelty of the performance.  The top score and
bottom score among the five judges are removed and the remaining three
judges' scores are averaged to give the final score for the surfer for
that wave.  Each surfer's aggregate score is given by the average of
the two highest-scoring waves.  The winner is the surfer with the
highest aggregate score.

Such scoring systems are designed to account for the random nature of
wave quality while reflecting competitors' abilities fairly.
Differences in wave quality between successive days mean that direct
comparison of scores between one day and another are not informative
about competitors' abilities as they are strongly dependent on details
of wave quality at the time of the competition.  However, for a
particular heat, the order statistic---that is, which competitor
scored most highly, second highest, and third---is informative about
competitors' abilities: the wave environment is common to each surfer.

We note in passing that `wct8`, hosted at Lemoore had a different
format from the other venues in the championship, being held in an
artificial wave pool in which every wave was essentially identical (up
to a mirror reflection).  Statistically, there were seven heats of
either 2, 4, 8, or 16 surfers.  We show below how to incorporate the
information present in such observations with the remainder of the
championship using a consistent and intuitive statistical model.

\subsection{Previous related work}

\cite{farley2013} analyse surfing results using one-way ANOVA techniques and
report comparisons between the top 10 and bottom 10 surfers in the
2013 World Championship Tour and conclude that the top-ranked athletes
are more consistent than the lower-ranked group. They report that "To
date, only a limited number of research studies have reported on the
results of competitive surfing" [p44], and the analysis presented here
is hoped to change that.

Other statistical research into competitive surfing appears to focus
on the scoring of aerial manoeuvres compared with other manoeuvres in
events.  \cite{undgren2014}, for example, report that aerial manoeuvres
scored higher than other stunts but had a lower completion rate, a
theme we return to below.

\section{Bradley-Terry}

The Bradley-Terry model \cite{bradley1954} assigns non-negative strengths
$p_1,\ldots, p_n$ to each of $n$ competitors in such a way that the
probability of $i$ beating $j\neq i$ in pairwise competition is
$\frac{p_i}{p_i+p_j}$; it is conventional to normalize so that $\sum
p_i=1$.  Further, we use a generalization due to \cite{luce1959}, in which
the probability of competitor~$i$ winning in a field of $\left\lbrace
1,\ldots, n\right\rbrace$ is $\frac{p_i}{p_1+\cdots +p_n}$.  Noting
that there is information in the whole of the finishing order, and not
just the first across the line, we can follow \cite{plackett1975} and
consider the runner-up to be the winner among the remaining
competitors, and so on down the finishing order. Without loss of
generality, if the order of finishing were $1,2,3,4,5$, then a
suitable Plackett-Luce likelihood function would be

\begin{equation}\label{competitors_1_to_5_likelihood}
\frac{p_1}{p_1+p_2+p_3+p_4+p_5}\cdot
\frac{p_2}{p_2+p_3+p_4+p_5}\cdot
\frac{p_3}{p_3+p_4+p_5}\cdot
\frac{p_4}{p_4+p_5}\cdot
\frac{p_5}{p_5}
\end{equation}

and this would be a forward ranking Plackett-Luce model in the
terminology of \cite{johnson2020}.  We now use a technique due to Hankin
(2010, 2017) and introduce fictional (reified) entities whose nonzero
Bradley-Terry strength helps certain competitors or sets of
competitors under certain conditions.  The original example was the
home-ground advantage in football.  If players (teams) $1,2$ with
strengths $p_1,p_2$ compete, and if our observation were $a$ home wins
and $b$ away wins for team $1$, and $c$ home wins and $d$ away wins
for team~$2$, then a suitable likelihood function would be

\[
\left(\frac{p_1+p_H}{p_1+p_2+p_H}\right)^a
\left(\frac{p_1}{p_1+p_2+p_H}\right)^b
\left(\frac{p_2+p_H}{p_1+p_2+p_H}\right)^c
\left(\frac{p_2+p_H}{p_1+p_2+p_H}\right)^d,
\]

\noindent where $p_H$ is a quantification of the beneficial home
ground effect.  Similar techniques have been used to account for the
first-move advantage in chess, and here we apply it to assess the
supposed Brazilian preference for reef-breaking waves.

\section{Likelihood-based systems and points-based systems}

\begin{table}
\begin{tabular}{l|l|l|l|l|l|l|l|l|l|l|l|l|l}
competitor  & nationality&wct01  &wct02  &wct03  & \ldots &wct09&wct10&wct11&points\\ \hline
 Ferreira   &   BRA      &   1   &   5   &   17  & \ldots      &  2   &   1   &   1   &   59740\\
 Medina     &   BRA      &   5   &   5   &   17  & \ldots      &  9   &   9   &   2   &   56475\\
 Smith 	    &   ZAF      &   3   &   3   &   17  & \ldots      &  9   &   2   &   17  &   49985\\
 Toledo     &   BRA      &   9   &   2   &   5   & \ldots      &  17  &   5   &   17  &   49145\\
 Andino     &   USA      &   2   &   17  &   5   & \dots       &  5   &   5   &   9   &   46655
\end{tabular}
\caption{WCT tour results, conventional league table\label{resultstable}}
\end{table}

Table~\ref{resultstable} shows the official results table from the
tour.  We see that Ferreira places first with 59740 points, Medina
second with 56475, and so on.  The overall ranking of the surfers is
determined by summing the season's points.  However, this system
suffers from the defect that the points system itself is intrinsically
arbitrary: the details of the points system does not affect the
competitors' behaviour but can change the overall ranking.
Plackett-Luce likelihood functions are not subject to such a
phenomenon and furnish an objective and coherent method for ranking
competitors.  Further, likelihood-based methods offer a further
advantage over points-based systems: the ability to conduct
statistically rigorous tests of meaningful nulls.

\section{The dataset and likelihood function}

\begin{table}
\begin{verbatim}
wct01 Colapinto Bailey Wright
wct01 Freestone Lau Smith
wct01 Dora Ferreira Slater
wct01 Duru Toledo Ibelli
wct01 Moniz Heazlewood Wilson
...
wct11 Medina Florence
wct11 Colapinto Bourez
wct11 Ferreira Slater
wct11 Medina Colapinto
wct11 Ferreira Medina
\end{verbatim}
\caption{Raw results\label{rawresults}}
\end{table}

The first line of table \ref{rawresults} shows that, at one of the
heats in {\tt wct01}, Colapinto came first, Bailey second, and Wright
came third.  The last line shows that, at {\tt wct11}, Ferreira came
first and Medina second.  We can convert this dataset into a
Plackett-Luce likelihood function but first have to remove the
wildcards and also player Vries whose maximum likelihood strength is
zero.  For the first line we would have Plackett-Luce likelihood function as

\begin{equation}
\frac{p_\mathrm{Colapinto}}{p_\mathrm{Colapinto} + p_\mathrm{Bailey} + p_\mathrm{Wright}}\cdot
\frac{p_\mathrm{Bailey}}{p_\mathrm{Bailey} + p_\mathrm{Wright}}
\end{equation}

[observe that there is no requirement here for the strengths to have
unit sum].  The entire dataset has a likelihood function that
incorporates all 477 lines, but it includes:

\begin{equation}
\frac{
p_\mathrm{Andino}^{19}\, p_\mathrm{Callinan}^{19}\, p_\mathrm{Coffin}^{8}\ldots
}{
(p_\mathrm{Andino} + p_\mathrm{Bourez})  (p_\mathrm{Andino} + p_\mathrm{Buchan})(p_\mathrm{Bourez} + p_\mathrm{Flores} + p_\mathrm{Freestone})\ldots
}
\end{equation}

\begin{figure}
\includegraphics[width=4in]{surfing_files/figure-html/showsurfingmaxppie-3.png} % Pie chart of strengths
\caption{Pie chart\label{piechartmax} showing maximum likelihood strengths of the 23 competitors.  Note the
dominance of Ferreira, Florence, and Medina}
\end{figure}

Figure~\ref{piechartmax} shows the maximum likelihood estimate for the
23 competitors' strengths, which appears to show a wide range from
Florence at about 17.5\% down to Coffin at about 1.5\%.  The {\tt
hyper2} software may be used to formally assess one plausible null:
that the competitors all have equal skill, that is $H_0\colon
p_\mathrm{Andino} = p_\mathrm{Bourez}=\ldots=
p_\mathrm{Wright}=\frac{1}{23}$, and any differences in placings are
due to random variation.  The Method of Support \cite{edwards1973}
rejects the null (in favour of the evaluate) at $p=4\times 10^{-5}$.
There is thus strong evidence to suggest that the competitors are
indeed of different skills and the result is not random.  Note that
the competitors' points (that is, the final column of
table~\ref{resultstable}) does not allow one to come to this
conclusion as the points awarded are arbitrary.

However, points awarded may be compared with likelihood estimates:
Figure~\ref{compare_likelihood_points} shows a loose correlation
($R^2=0.64$; $p=3\times 10^{-6}$) and
Figure~\ref{compare_likelihood_points_rankings} shows how the rankings
differ when calculated by the points and likelihood systems.

\begin{figure}[h]
\includegraphics[width=4in]{surfing_files/figure-html/directlogplotofstrengths-1.png}
\caption{Scatterplot of points scored  \label{compare_likelihood_points} 
against maximum likelihood estimates}
\end{figure}

\begin{figure}[h]
\includegraphics[width=4in]{surfing_files/figure-html/ordertransplotplot-1.png}
\caption{Scatterplot of ranks \label{compare_likelihood_points_rankings} of points-based ranking
against likelihood-based ranking.  Note the anomalous position of
Florence}
\end{figure}

\bibliography{chess}
\end{document}

### 2021-08-10 inst/surfing_jsa.tex
rewording

@@
 the competitors' points (that is, the final column of
-table~\ref{resultstable}) does not allow one to come to this
-conclusion as the points awarded are arbitrary.
+table~\ref{resultstable}) does not admit this type of analysis, as the
+points awarded are arbitrary.
 
-However, points awarded may be compared with likelihood estimates:
-Figure~\ref{compare_likelihood_points} shows a loose correlation
-($R^2=0.64$; $p=3\times 10^{-6}$) and
+However, points awarded may be compared with likelihood estimates, and
+one would expect high points totals to be associated with high
+likelihoods.  Figure~\ref{compare_likelihood_points} shows a loose
+correlation ($R^2=0.64$; $p=3\times 10^{-6}$) and
 Figure~\ref{compare_likelihood_points_rankings} shows how the rankings
-differ when calculated by the points and likelihood systems.
+differ when calculated by the points and likelihood systems.  Note the
+anomalous position of Florence, who did not compete in five venues due
+to injury: this strongly affecting his points total but not his
+likelihood-based ranking.
 

### 2021-08-10 inst/skating.Rmd
add hex sticker to more Rmd files

@@
 ```
+<p style="text-align: right;">
+![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
+</p>
 

### 2021-08-10 inst/volvo.Rmd
tidy up reference
more detail

@@
 
+<p style="text-align: right;">
+![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
+</p>
+
 This short document creates the `volvo` suite of R objects and
-presents some preliminary analysis.
+presents some preliminary analysis.  It refers to the 12th edition of
+the Volvo Ocean Race, 2014-2015 in which seven racing teams competed
+over nine legs.  Dataset taken from the wikipedia page (see references
+below).
 
@@
 
-Wikipedia contributors, 2019.  \dQuote{2014-2015 Volvo Ocean Race}.
-In \emph{Wikipedia, the free encyclopedia}.  Retrieved 22:21, February
+Wikipedia contributors, 2019.  "2014-2015 Volvo Ocean Race".
+In _Wikipedia, the free encyclopedia_.  Retrieved 22:21, February
 28, 2020.
-\url{https://en.wikipedia.org/w/index.php?title=2014\%E2\%80\%932015_Volvo_Ocean_Race&oldid=914916131},
+(https://en.wikipedia.org/w/index.php?title=2014\%E2\%80\%932015_Volvo_Ocean_Race&oldid=914916131)
 

### 2021-08-12 inst/surfing.Rmd
more work on surfing paper

@@
 samep.test(surfing, c("Florence","Medina","Ferreira"))
+samep.test(surfing, c("Medina","Wilson"))
 ```
@@
 ```{r plotwidep}
-plot(c(0,p_wide),c(l0,l_wide)-max(l),type='b',xlab="Brazilian beachtype",ylab="support")
+plot(c(0,p_wide),c(l0,l_wide)-max(l),type='b',xlab="Brazilian beachtype wide",ylab="support")
+abline(h= -2,lty=2)
 ```

### 2021-08-12 inst/surfing_jsa.tex
more work on surfing paper

@@
 
-\[
+\begin{equation}\label{home_ground_advantage}
 \left(\frac{p_1+p_H}{p_1+p_2+p_H}\right)^a
@@
 \left(\frac{p_2+p_H}{p_1+p_2+p_H}\right)^d,
-\]
+\end{equation}
 
@@
 
+We may ask whether the top two officially ranked competitors, Medina
+and Ferreira, have different strengths.  To test the null of equal
+strength, we maximize the likelihood subject to the null, and compare
+with the global maximum likelihood~\cite{hankin2010}.  This gives us a
+support difference of about 0.05, well short of Edwards's 2 units of
+support criterion.  Alternatively, we could observe that the
+$-2\log\left(\frac{\Lambda}{\Lambda_0}\right)\simeq 0.1$ is not in the tail
+region of its asymptotic $\chi^2_1$ distribution.  We may thus assert
+that there is no strong evidence for Ferreira's actually having a
+competitive advantage over Medina.  {\bf Tim: how about comparing
+  Medina with, say, Wilson, would that make sense from a surfing
+  perspective? Tim can you say something about Wilson's rivalry with
+  Medina?}  We may similarly wonder whether Medina is equal in
+strength to Wilson; here we reject the null with support difference of
+about 2.33, corresponding to a $p$-value of $0.031$.
+
+\section{Brazilian beach type prefereneces}
+
+The statistical method described here can test many different
+suggestions drawn from surfing lore {\bf Tim is there a better way to
+  describe this?}.  For example, one frequent suggestion {\bf Tim I
+  need citations here} is that Brazilian surfers such as Ferreira and
+Toledo have surfing skillsets that favour point and beach break
+beaches, those types of beach being readily accessible to these
+surfers' native Brazil.  We now assess the null that Brazilians
+perform equally well on all types of beaches.  To that end, we create
+a reified surfer, ``Brazilian beachtype", whose strength helps all
+Brazilian surfers when surfing on their (putative) preferred
+beachtype, analogously to the ``home ground monster" who helps the
+home team as per equation~\ref{home_ground_advantage}.  The null would
+be that this reified entity has zero strength.  We fail to reject this
+null (differential support = 0.035; $p=0.79$) and conclude that there
+is no evidence to suggest that Brazilians perform is better on their
+preferred beach types than one would expect by chance.  We may go
+further and state that not only is the maximum likelihood estimate for
+Brazilian beachtype is very small (about 0.002), but we may rule out
+the suggestion that the strength of this entity exceeds about 0.025
+(Figure~\ref{brazilian_beachtype_support}).
+
+\begin{figure}
+  \includegraphics[width=4in]{surfing_files/figure-html/plotwidep-1.png}
+\caption{Support \label{brazilian_beachtype_support} for the Brazilian beachtype reified entity}
+\end{figure}
+
+\section{Compatriot selection}
+
 \bibliography{chess}

### 2021-08-12 inst/surfing.Rmd
table of number of surfers in the water

@@
 shows that, at `wct11`, `Ferreira` came first and `Medina` came
-second.  We can convert this dataset into a Plackett-Luce likelihood
+second.
+
+Following idiom creates a table:
+
+```{r}
+table(unlist(lapply(strsplit(jj, " "),length)))
+```
+
+This shows that object `jj` has 312 lines of 2 surfers, 160 lines with
+3 surfers, 2 lines with 4 surfers, 2 lines with 8 surfers, and one
+line with 16 surfers.
+
+We can convert the dataset into a Plackett-Luce likelihood
 function but first have to remove the wildcards and also player

### 2021-08-12 inst/surfing_jsa.tex
more work on details

@@
 \title{Analysis of professional surfing tournaments with generalized Bradley-Terry}
+
+\section*{Abstract}
+
+{\bf\large Tim we need to discuss this}
 
@@
 
-\cite{farley2013} analyse surfing results using one-way ANOVA techniques and
-report comparisons between the top 10 and bottom 10 surfers in the
-2013 World Championship Tour and conclude that the top-ranked athletes
-are more consistent than the lower-ranked group. They report that "To
-date, only a limited number of research studies have reported on the
-results of competitive surfing" [p44], and the analysis presented here
-is hoped to change that.
+Farley {\em et. al.} \cite{farley2013} analyse surfing results using
+one-way ANOVA techniques and report comparisons between the top 10 and
+bottom 10 surfers in the 2013 World Championship Tour and conclude
+that the top-ranked athletes are more consistent than the lower-ranked
+group.  They report that ``to date, only a limited number of research
+studies have reported on the results of competitive surfing" [p44],
+and the analysis presented here fills some of this gap.
 
@@
 on the scoring of aerial manoeuvres compared with other manoeuvres in
-events.  \cite{undgren2014}, for example, report that aerial manoeuvres
-scored higher than other stunts but had a lower completion rate, a
-theme we return to below.
+events.  Lundgren {\em et. al.} \cite{lundgren2014}, for example,
+report that aerial manoeuvres scored higher than other stunts but had
+a lower completion rate, a theme we return to below.
 
@@
 arbitrary: the details of the points system does not affect the
-competitors' behaviour but can change the overall ranking.
-Plackett-Luce likelihood functions are not subject to such a
-phenomenon and furnish an objective and coherent method for ranking
-competitors.  Further, likelihood-based methods offer a further
+competitors' behaviour but can change the overall ranking; this would
+suggest that more formal statistical methods are neede for analysis of
+such datasets.  The Plackett-Luce likelihood functions are not subject
+to such a phenomenon and furnish an objective and coherent method for
+ranking competitors.  Further, likelihood-based methods offer one
 advantage over points-based systems: the ability to conduct
@@
 
-\begin{equation}
+\begin{equation}\label{complete_likelihood_function}
 \frac{
@@
 
+\subsection{Bradley-Terry and pairs}
+
+Paired comparisons are particularly favourable to Bradley-Terry
+because the likelihood function is exponential family and as such is
+amenable to analysis via general linear models~\cite{turner2012}.
+However, this breaks down when more than two entities are compared and
+here {\bf clumsy wording} we see comparisons between three surfers
+and, in the case of wct08, many more.  It is natural to ask how much
+information is contained in the observations where more than two
+surfers are in the water.
+
+We can directly calculate the determinant of the observed information
+matrix for the likelihood function in
+equation~\ref{complete_likelihood_function} at about $4.3\times
+10^{71}$, as compared with~$2.9\times 10^{69}$ for the pairs
+considered in isolation, a ratio of about $146.4$.  We are thus
+motivated to consider the whole dataset, not just the part which
+involves paired comparisons.  {\bf clumsy wording}.
+
+\section{Results}
 \begin{figure}
@@
 p_\mathrm{Wright}=\frac{1}{23}$, and any differences in placings are
-due to random variation.  The Method of Support \cite{edwards1973}
+due to random variation.  The Method of Support \cite{edwards1992}
 rejects the null (in favour of the evaluate) at $p=4\times 10^{-5}$.
 There is thus strong evidence to suggest that the competitors are
-indeed of different skills and the result is not random.  Note that
-the competitors' points (that is, the final column of
-table~\ref{resultstable}) does not admit this type of analysis, as the
-points awarded are arbitrary.
+indeed of different skills and the result actually reflects genuine
+differences in abilities, rather than the result of random chance.
+Observe that the competitors' points (that is, the final column of
+table~\ref{resultstable}) cannot be used for this type of
+probabalistic analysis: the points awarded are arbitrary and changing
+the points system would change the points scored but not affect the
+competitors' behaviour.
 
@@
 Toledo have surfing skillsets that favour point and beach break
-beaches, those types of beach being readily accessible to these
-surfers' native Brazil.  We now assess the null that Brazilians
-perform equally well on all types of beaches.  To that end, we create
-a reified surfer, ``Brazilian beachtype", whose strength helps all
-Brazilian surfers when surfing on their (putative) preferred
-beachtype, analogously to the ``home ground monster" who helps the
-home team as per equation~\ref{home_ground_advantage}.  The null would
-be that this reified entity has zero strength.  We fail to reject this
-null (differential support = 0.035; $p=0.79$) and conclude that there
-is no evidence to suggest that Brazilians perform is better on their
+beaches {\bf Tim I need some help describing and defining these here},
+those types of beach being readily accessible to these surfers' native
+Brazil.  We now assess the null that Brazilians perform equally well
+on all types of beaches.  To that end, we create a reified surfer,
+``Brazilian beachtype", whose strength helps all Brazilian surfers
+when surfing on their (putative) preferred beachtype, analogously to
+the ``home ground monster" who helps the home team as per
+equation~\ref{home_ground_advantage}.  The null would be that this
+reified entity has zero strength.  We fail to reject this null
+(differential support = 0.035; $p=0.79$) and conclude that there is no
+evidence to suggest that Brazilians' performce is better on their
 preferred beach types than one would expect by chance.  We may go
-further and state that not only is the maximum likelihood estimate for
-Brazilian beachtype is very small (about 0.002), but we may rule out
-the suggestion that the strength of this entity exceeds about 0.025
-(Figure~\ref{brazilian_beachtype_support}).
+further using the method of support and state that not only is the
+maximum likelihood estimate for Brazilian beachtype is very small
+(about 0.002), but we may rule out the suggestion that the strength of
+this entity exceeds about 0.025 (Figure~\ref{brazilian_beachtype_support}).
 
@@
   \includegraphics[width=4in]{surfing_files/figure-html/plotwidep-1.png}
-\caption{Support \label{brazilian_beachtype_support} for the Brazilian beachtype reified entity}
+\caption{Profile support \label{brazilian_beachtype_support} for the
+  Brazilian beachtype reified entity.  The horizontal axis specifies a
+  particular strength for Brazilian beachtype, and the height of the
+  curve gives the support conditional on the x-axis}
 \end{figure}
@@
 
+{\bf\large Tim I am on thin ice here, I need to talk with you and
+  clarify this}
+
+One measure of the rising popularity of surfing is its inclusion in
+the 2020 Summer Olympics, and it is natural to ask whether this would
+affect the competition in some way.  One suggestion {\bf Tim?} would
+be that competitors would become more focussed on beating compatriots
+than surfers from other countries, the thinking being that admission
+to a country's Olympic squad would be of more importance to a
+competitor than their result in wct.  {\bf Tim this is clumsy
+  wording}.  We would expect competitors to value their ranking when
+restricted to within their country more than their ranking over all of
+wct.  If this is the case then we would expect competitors to change
+their behaviour if they are in the water with another surfer of the
+same nationality.  Specifically, if we have three surfers in the water
+of whom two are of one nationality and the third of another, we would
+expect the compatriots to adopt a style calculated to beat their
+countryman, possibly at the expense of performance relative to the odd
+man out.  {\bf Tim this is poor wording we need to rephrase}.
+
+Such hypotheses are difficult to test with conventional techniques as
+competitors's strengths are nuisance parameters ~\cite{basu1977}.
+However, the
+
+Consider, as an example, the case of Smith, Buchan, Freestone.  Buchan
+and Freestone are Australian, and Jordy Smith is South African.  It is
+reasonable to ask whether Buchan and Freestone, who are in direct
+competition with one another for a place in the Australian Olympic
+surfing squad, would be more motivated to beating each other than to
+come ahead of Smith.  This would be a rational response, but arguably
+noncompetitive.  How would such a strategy manifest itself in
+scorelines?
+
+Reified Bradley-Terry technique furnishes an intuitive statistical
+test for such a mechanism which controls for competitors' strengths.
+Noting that the compatriots effectively place themselves at a {\em
+  disadvantage} when compared with the odd man out, we introduce a
+compatriot monster whose strength $p_C$ helps the odd man out.
+Writing $p_1, p_2$ for the strengths of the compatriots and $q$ for
+the odd man out, we construct likelihood functions for the six
+possible orderings when two compatriots and an odd man out are in the
+water at the same time in Table~\ref{likelihoodcompatriot} (if we have
+either three different nationalities or three identical nationalities,
+we revert to the standard Plackett-Luce likelihood).  Note that all
+six likelihood functions in table~\ref{likelihoodcompatriot} are
+informative about $p_1,p_2,q$ in addition to being informative about
+the object of inference, $p_C$.
+
+\begin{table}
+\begin{tabular}{l|l}
+order  & ${\mathcal L}(p_1,p_2,q,p_C)$\\ \hline
+$p_1\succ p_2\succ q $ & $ \frac{p_1}{p_1+p_2+q+p_C}\cdot\frac{p_2}{p_2+q+p_C}$\\
+$p_2\succ p_1\succ q $ & $ \frac{p_2}{p_1+p_2+q+p_C}\cdot\frac{p_1}{p_1+q+p_C}$\\
+$p_1\succ q\succ p_2 $ & $ \frac{p_1}{p_1+p_2+q+p_C}\cdot\frac{q+p_C}{p_2+q+p_C}$\\
+$p_2\succ q\succ p_1 $ & $ \frac{p_2}{p_1+p_2+q+p_C}\cdot\frac{q+p_C}{p_1+q+p_C}$\\
+$q\succ p_1\succ p_2 $ & $ \frac{q+p_C}{p_1+p_2+q+p_C}\cdot\frac{p_1}{p_1+p_2}$\\
+$q\succ p_2\succ p_1 $ & $ \frac{q+p_C}{p_1+p_2+q+p_C}\cdot\frac{p_2}{p_1+p_2}$\\
+\end{tabular}
+\caption{The \label{likelihoodcompatriot} six possible finishing
+  orders (``$\succ$'' means ``scored more points than'') for the case
+  where two compatriot surfers and an odd man out are in the water at
+  the same time, with associated likelihood function for strengths
+  $p_1,p_2,q, p_C$.  Here $p_1,p_2$ are the strengths of two
+  compatriot surfers, $q$ the strength of the odd man out, $p_C$ the
+  strength of the compatriot monster. and nationality and a third are
+  in the water at the same time}
+\end{table}
+
+We can now apply the Method of Support to the hypothesis that $p_C=0$,
+that is, that this noncompetitive behaviour is absent.  We find that
+the maximum likelihood estimate for $p_C$ is about 0.015, weaker than
+the ``real'' surfers, but the support for a nonzero $p_C$ is about 1.8
+[asymptotic $p=0.28$], short of the two units of support required for
+rejection of our null.  We conclude that there is no strong evidence
+to suggest that such noncompetitive behaviour occurred, at least not
+from the order statistics we considered.
+
+\section{Conclusions and further work}
+
+{\bf\large Tim we need to discuss this}
+
 \bibliography{chess}

### 2021-08-13 inst/ordertrans.Rmd
named vector

@@
 ```
+
+## Another use-case
+
+How about this:
+
+```{r}
+(x <- c(d=2, a=3, c=4, b=3, e=6))
+(y <- c(e=3, c=2, a=4, b=5, d=1))
+x+y
+```
+
+Above, observe that R is behaving as intended $\ldots$ but is
+potentially confusing.  Take the first element.  This has value
+`2+5=7` and name `d` from the name of element 1 of `x`.  But it would
+be reasonable to ask for the first element to be the sum of element
+`d` of `x` and element `d` of `y`, which would be `2+1=3`.
+This can be done with `ordertrans()`:
+
+```{r}
+ordertrans(x) + ordertrans(y)
+```
+
+See how `ordertrans()` has rearranged both `x` and `y` so that the
+names are in alphabetical order.
+
+But this is not perfect.  Consider:
+
+```{r}
+(z <- c(f=3, g=2, h=4, a=5, b=1))  # names NOT letters[1:5]
+ordertrans(x) + ordertrans(z)  # arguably not well-defined
+```
+
+In this case the result is arguably incorrect.
 

### 2021-08-21 inst/surfing_jsa.tex
more work

@@
 
-\title{Analysis of professional surfing tournaments with generalized Bradley-Terry}
+\title{Analysis of competitive surfing tournaments with generalized Bradley-Terry}
 
 \section*{Abstract}
-
-{\bf\large Tim we need to discuss this}
+Here we apply a generalization of Bradley-Terry model to competitive
+surfing, specifically the 2018 World Surf League.  We use
+Bradley-Terry likelihoods to assess a number of hypotheses of interest
+to the surfing community with formal statistical methods.  We go on to
+study the ``Brazilian storm'' phenomenon using reified Bradley-Terry
+likelihood functions in two ways: firstly, the supposed Brazilian
+preference for point and beach break beaches; and secondly from the
+perspective of tournament theory in which we test for competitors
+modifying their strategy in the presence of compatriot rivals.
 
@@
 amenable to analysis via general linear models~\cite{turner2012}.
-However, this breaks down when more than two entities are compared and
-here {\bf clumsy wording} we see comparisons between three surfers
-and, in the case of wct08, many more.  It is natural to ask how much
-information is contained in the observations where more than two
-surfers are in the water.
+However, this is not applicable when more than two entities are
+compared; and indeed here we see comparisons between three surfers
+and, in the case of Jeep Surf Freshwater Pro (wct08), many more.  It
+is natural to ask how much information is contained in the
+observations where more than two surfers are in the water.
 
@@
 10^{71}$, as compared with~$2.9\times 10^{69}$ for the pairs
-considered in isolation, a ratio of about $146.4$.  We are thus
-motivated to consider the whole dataset, not just the part which
-involves paired comparisons.  {\bf clumsy wording}.
+considered in isolation, a ratio of about $146.4$.  This ratio shows
+that there is considerable information loss in considering only the
+paired comparisons.  We are thus motivated to consider the whole
+dataset, not just the part which involves paired comparisons.
 
@@
 
-\section{Brazilian beach type prefereneces}
+\section{The Brazilian storm}
+
+In surfing lore, the ``Brazilian storm'' refers to the emergence of
+Brazil as a dominant nation of elite surfers.  The term was first used
+to describe the fact six out of eight querterfinalists at 2011 Nike
+Lowers Pro were Brazilian; it rose in public conciousness following
+the victories of Medina and de Souza in the World Surf League
+championships in 2014 and 2015 respectively.  The storm shows no signs
+of abating, with commentators such as Douglas opining that this
+generation ``represents a seismic shift for Brazilian competitors onto
+an entirely new plane'' \cite{douglas2020}.
+
+Here, we study two aspects of this sporting phenomenon: the suggested
+Brazilian preference for beach break and point break reefs; and the
+noncompetitive effect of compatriot rivalry.
+
+\subsection{Brazilian beach type prefereneces}
 
@@
 countryman, possibly at the expense of performance relative to the odd
-man out.  {\bf Tim this is poor wording we need to rephrase}.
-
-Such hypotheses are difficult to test with conventional techniques as
-competitors's strengths are nuisance parameters ~\cite{basu1977}.
-However, the
+man out.  {\bf Tim this is poor wording we need to rephrase}.  Such
+hypotheses are difficult to test with conventional techniques as
+competitors's strengths themselves are nuisance
+parameters~\cite{basu1977}.
 

### 2021-08-21 inst/surfing_jsa.tex
tidyup

@@
 \section*{Abstract}
-Here we apply a generalization of Bradley-Terry model to competitive
-surfing, specifically the 2018 World Surf League.  We use
+Here we analyse competitive surfing, specifically the 2018 World Surf
+League, using formal statistical methods.  We use generalized
 Bradley-Terry likelihoods to assess a number of hypotheses of interest
-to the surfing community with formal statistical methods.  We go on to
-study the ``Brazilian storm'' phenomenon using reified Bradley-Terry
-likelihood functions in two ways: firstly, the supposed Brazilian
-preference for point and beach break beaches; and secondly from the
-perspective of tournament theory in which we test for competitors
-modifying their strategy in the presence of compatriot rivals.
+to the surfing community.  Specifically, we quantify the dominance of
+the top competitors using likelihood techniques, and go on to study
+the ``Brazilian storm'' phenomenon using reified entities in two ways.
+Firstly we assess the supposed Brazilian preference for point and
+beach break beaches; and secondly we consider tournament results from
+the perspective of tournament theory and test for competitors
+modifying their strategy in the presence of compatriot rivals.  We
+quantify the evidence for these commonly assumed features of
+contemporary competitive surfing and suggest further avenues of
+research.
 
@@
 
-We note in passing that `wct8`, hosted at Lemoore had a different
-format from the other venues in the championship, being held in an
-artificial wave pool in which every wave was essentially identical (up
-to a mirror reflection).  Statistically, there were seven heats of
-either 2, 4, 8, or 16 surfers.  We show below how to incorporate the
-information present in such observations with the remainder of the
-championship using a consistent and intuitive statistical model.
+We note in passing that Lemoore had a different format from the other
+venues in the championship, being held in an artificial wave pool in
+which every wave was essentially identical (up to a mirror
+reflection).  Statistically, there were seven heats of either 2, 4, 8,
+or 16 surfers.  We show below how to incorporate the information
+present in such observations with the remainder of the championship
+using a consistent and intuitive statistical model.
 
@@
 \end{verbatim}
-\caption{Raw results\label{rawresults}}
+\caption{Extract from observations: raw results\label{rawresults} from
+  WCT 2018}
 \end{table}
@@
 support criterion.  Alternatively, we could observe that the
-$-2\log\left(\frac{\Lambda}{\Lambda_0}\right)\simeq 0.1$ is not in the tail
-region of its asymptotic $\chi^2_1$ distribution.  We may thus assert
-that there is no strong evidence for Ferreira's actually having a
-competitive advantage over Medina.  {\bf Tim: how about comparing
-  Medina with, say, Wilson, would that make sense from a surfing
-  perspective? Tim can you say something about Wilson's rivalry with
-  Medina?}  We may similarly wonder whether Medina is equal in
-strength to Wilson; here we reject the null with support difference of
-about 2.33, corresponding to a $p$-value of $0.031$.
+$-2\log\left(\frac{\Lambda}{\Lambda_0}\right)\simeq 0.1$ is not in the
+tail region of its asymptotic $\chi^2_1$ distribution.  We may thus
+assert that there is no strong evidence for Ferreira's actually having
+a competitive advantage over Medina.  We may similarly wonder whether
+Medina is equal in strength to Wilson; here we reject the null with
+support difference of about 2.33, corresponding to a $p$-value of
+$0.031$.
 
@@
 The statistical method described here can test many different
-suggestions drawn from surfing lore {\bf Tim is there a better way to
-  describe this?}.  For example, one frequent suggestion {\bf Tim I
-  need citations here} is that Brazilian surfers such as Ferreira and
-Toledo have surfing skillsets that favour point and beach break
-beaches {\bf Tim I need some help describing and defining these here},
-those types of beach being readily accessible to these surfers' native
-Brazil.  We now assess the null that Brazilians perform equally well
-on all types of beaches.  To that end, we create a reified surfer,
-``Brazilian beachtype", whose strength helps all Brazilian surfers
-when surfing on their (putative) preferred beachtype, analogously to
-the ``home ground monster" who helps the home team as per
-equation~\ref{home_ground_advantage}.  The null would be that this
-reified entity has zero strength.  We fail to reject this null
-(differential support = 0.035; $p=0.79$) and conclude that there is no
-evidence to suggest that Brazilians' performce is better on their
-preferred beach types than one would expect by chance.  We may go
-further using the method of support and state that not only is the
+suggestions drawn from surfing lore.  For example, one frequent
+suggestion {\bf Tim I need citations here} is that Brazilian surfers
+such as Ferreira and Toledo have surfing skillsets that favour point
+and beach break beaches {\bf Tim I need some help describing and
+  defining these here}, those types of beach being readily accessible
+to these surfers' native Brazil.  We now assess the null that
+Brazilians perform equally well on all types of beaches.  To that end,
+we create a reified surfer, ``Brazilian beachtype", whose strength
+helps all Brazilian surfers when surfing on their (putative) preferred
+beachtype, analogously to the ``home ground monster" who helps the
+home team as per equation~\ref{home_ground_advantage}.  The null would
+be that this reified entity has zero strength.  We fail to reject this
+null (differential support = 0.035; $p=0.79$) and conclude that there
+is no evidence to suggest that Brazilians' performce is better on
+their preferred beach types than one would expect by chance.  We may
+go further using the method of support and state that not only is the
 maximum likelihood estimate for Brazilian beachtype is very small
 (about 0.002), but we may rule out the suggestion that the strength of
-this entity exceeds about 0.025 (Figure~\ref{brazilian_beachtype_support}).
+this entity exceeds about 0.025
+(Figure~\ref{brazilian_beachtype_support}).
 
@@
 
-{\bf\large Tim I am on thin ice here, I need to talk with you and
-  clarify this}
+One measure of the rising popularity of surfing is its inclusion in
+the 2020 Summer Olympics~\cite{tulloch2019}, and it is natural to ask
+whether this would affect the competition in some way~\cite{ho2021}.
+One suggestion would be that competitors would become more focussed on
+beating compatriots than surfers from other countries, the thinking
+being that admission to a country's Olympic squad would be of more
+importance to a competitor than their result in wct.  One might expect
+competitors to value their ranking when restricted to within their
+country more than their ranking over all of wct.  If this is the case
+then we would expect competitors to change their behaviour if they are
+in the water with another surfer of the same nationality.
+Specifically, if we have three surfers in the water of whom two are of
+one nationality and the third of another, we would expect the
+compatriots to adopt a style calculated to beat their countryman,
+possibly at the expense of performance relative to the odd man out.
 
-One measure of the rising popularity of surfing is its inclusion in
-the 2020 Summer Olympics, and it is natural to ask whether this would
-affect the competition in some way.  One suggestion {\bf Tim?} would
-be that competitors would become more focussed on beating compatriots
-than surfers from other countries, the thinking being that admission
-to a country's Olympic squad would be of more importance to a
-competitor than their result in wct.  {\bf Tim this is clumsy
-  wording}.  We would expect competitors to value their ranking when
-restricted to within their country more than their ranking over all of
-wct.  If this is the case then we would expect competitors to change
-their behaviour if they are in the water with another surfer of the
-same nationality.  Specifically, if we have three surfers in the water
-of whom two are of one nationality and the third of another, we would
-expect the compatriots to adopt a style calculated to beat their
-countryman, possibly at the expense of performance relative to the odd
-man out.  {\bf Tim this is poor wording we need to rephrase}.  Such
-hypotheses are difficult to test with conventional techniques as
-competitors's strengths themselves are nuisance
-parameters~\cite{basu1977}.
+Such hypotheses
+are difficult to test with conventional techniques as competitors's
+strengths themselves are nuisance parameters~\cite{basu1977}.
 

### 2021-08-22 inst/surfing.Rmd
issue #107 refers: now pwa() only operates for a single Brazilian

@@
      surfers <- pnames(jj)
-     if(any(nationality(surfers) == "BRA")){
+     if(sum(nationality(surfers)=="BRA")==1){	
        if((beachtype(venue)=="point") | (beachtype(venue) == "beach")){

### 2021-08-22 inst/surfing_jsa.tex
further dealing with issue #107

@@
 beachtype, analogously to the ``home ground monster" who helps the
-home team as per equation~\ref{home_ground_advantage}.  The null would
-be that this reified entity has zero strength.  We fail to reject this
-null (differential support = 0.035; $p=0.79$) and conclude that there
-is no evidence to suggest that Brazilians' performce is better on
-their preferred beach types than one would expect by chance.  We may
-go further using the method of support and state that not only is the
-maximum likelihood estimate for Brazilian beachtype is very small
-(about 0.002), but we may rule out the suggestion that the strength of
-this entity exceeds about 0.025
+home team as per equation~\ref{home_ground_advantage}.  Here we modify
+the likelihood function only when a single Brazilian is in the water,
+to avoid possible interaction with the compatriot issue discussed in
+section~\ref{compatriot}.  The null would be that this reified entity
+has zero strength.  We fail to reject this null (differential support
+= 0.035; $p=0.79$) and conclude that there is no evidence to suggest
+that Brazilians' performce is better on their preferred beach types
+than one would expect by chance.  We may go further using the method
+of support: not only is the maximum likelihood estimate for Brazilian
+beachtype very small (about 0.002), but in addition we may reject the
+the proposal that the strength of this entity exceeds about 0.025
 (Figure~\ref{brazilian_beachtype_support}).

### 2021-08-22 inst/surfing_jsa.tex
tweaks

@@
 Florence at about 17.5\% down to Coffin at about 1.5\%.  The {\tt
-  hyper2} software may be used to formally assess one plausible null:
-that the competitors all have equal skill, that is $H_0\colon
-p_\mathrm{Andino} = p_\mathrm{Bourez}=\ldots=
+  hyper2} software~\cite{hankin2017} may be used to formally assess
+one plausible null: that the competitors all have equal skill, that is
+$H_0\colon p_\mathrm{Andino} = p_\mathrm{Bourez}=\ldots=
 p_\mathrm{Wright}=\frac{1}{23}$, and any differences in placings are
@@
 differences in abilities, rather than the result of random chance.
-Observe that the competitors' points (that is, the final column of
+Alternatively, we may view the dataset as being sufficiently
+informative to reveal actual differences in surfing ability.  Observe
+that the competitors' points (that is, the final column of
 table~\ref{resultstable}) cannot be used for this type of
@@
 suggestions drawn from surfing lore.  For example, one frequent
-suggestion {\bf Tim I need citations here} is that Brazilian surfers
-such as Ferreira and Toledo have surfing skillsets that favour point
+suggestion~\cite{burgess2020,ho2021} is that Brazilian surfers such as
+Ferreira and Toledo tend to have surfing skillsets that favour point
 and beach break beaches {\bf Tim I need some help describing and
@@
 helps all Brazilian surfers when surfing on their (putative) preferred
-beachtype, analogously to the ``home ground monster" who helps the
+beach type, an analogue of the ``home ground monster" who helps the
 home team as per equation~\ref{home_ground_advantage}.  Here we modify
@@
 
-\section{Compatriot selection}
+\section{Differential \label{compatriot} behaviour toward compatriotic competitors}
 
@@
 whether this would affect the competition in some way~\cite{ho2021}.
-One suggestion would be that competitors would become more focussed on
+One suggestion would be that competitors become more focussed on
 beating compatriots than surfers from other countries, the thinking
 being that admission to a country's Olympic squad would be of more
-importance to a competitor than their result in wct.  One might expect
-competitors to value their ranking when restricted to within their
-country more than their ranking over all of wct.  If this is the case
-then we would expect competitors to change their behaviour if they are
-in the water with another surfer of the same nationality.
-Specifically, if we have three surfers in the water of whom two are of
-one nationality and the third of another, we would expect the
-compatriots to adopt a style calculated to beat their countryman,
-possibly at the expense of performance relative to the odd man out.
+importance to a competitor than their overall standing in the
+tournament.  One might expect competitors to value their ranking when
+restricted to within their country more than their international
+ranking.  If this is the case then we would expect competitors to
+change their behaviour if they are in the water with another surfer of
+the same nationality.  Specifically, if we have three surfers in the
+water of whom two are of one nationality and the third of another, we
+would expect the compatriots to adopt a style calculated to beat their
+countryman, possibly at the expense of performance relative to the odd
+man out.
 
-Such hypotheses
-are difficult to test with conventional techniques as competitors's
-strengths themselves are nuisance parameters~\cite{basu1977}.
+Such hypotheses are difficult to test with conventional techniques as
+competitors's strengths themselves are nuisance
+parameters~\cite{basu1977}.
 
-Consider, as an example, the case of Smith, Buchan, Freestone.  Buchan
+Consider, as an example, the case of Smith, Buchan, Freestone who were
+simultaneously in the water at Bells Beach, Australia.  Now, Buchan
 and Freestone are Australian, and Jordy Smith is South African.  It is
@@
 Reified Bradley-Terry technique furnishes an intuitive statistical
-test for such a mechanism which controls for competitors' strengths.
-Noting that the compatriots effectively place themselves at a {\em
+test for mechanisms that control for competitors' strengths.  Noting
+that the compatriots effectively place themselves at a {\em
   disadvantage} when compared with the odd man out, we introduce a
@@
 We can now apply the Method of Support to the hypothesis that $p_C=0$,
-that is, that this noncompetitive behaviour is absent.  We find that
+that is, that such noncompetitive behaviour is absent.  We find that
 the maximum likelihood estimate for $p_C$ is about 0.015, weaker than
 the ``real'' surfers, but the support for a nonzero $p_C$ is about 1.8
-[asymptotic $p=0.28$], short of the two units of support required for
-rejection of our null.  We conclude that there is no strong evidence
-to suggest that such noncompetitive behaviour occurred, at least not
-from the order statistics we considered.
+[asymptotic $p=0.28$], slightly short of the two units of support
+required for rejection of our null.  We conclude that there is
+evidence, but no strong evidence, to suggest that such noncompetitive
+behaviour occurred---at least not from the order statistics we
+considered.
 
@@
 
-{\bf\large Tim we need to discuss this}
+We analysed results from the 2018 World Surf League using reified
+Bradley-Terry likelihoods to assess a number of hypotheses of interest
+to the surfing community.  Reified Bradley-Terry allows the use of
+formal statistical methods in ways that analysing points scored does
+not.  For example, we can be confident that the results reflect
+differences in skill (rather than random variation); there is no
+statistical difference in strength between the top two officially
+ranked competitors, Medina and Ferreira; but we may reject the
+hypothesis that Medina and Wilson are equal in strength.
+
+We discuss and assess the ``Brazilian storm'' phenomenon using reified
+entities and show that there is no strong evidence to support the
+supposed Brazilian preference for point and beach break beaches.  We
+also find no evidence to suggest that surfing's being included in the
+2020 Summer Olympics results in modified strategies to bolster
+intranational ranking at the expense WCT performance.
+
+Further work might include  {\bf\large Tim we should discuss}
 

### 2021-08-23 inst/surfing_jsa.tex
work on Tim's suggestions

@@
 The statistical method described here can test many different
-suggestions drawn from surfing lore.  For example, one frequent
+hypotheses drawn from surfing lore.  For example, one frequent
 suggestion~\cite{burgess2020,ho2021} is that Brazilian surfers such as
-Ferreira and Toledo tend to have surfing skillsets that favour point
-and beach break beaches {\bf Tim I need some help describing and
-  defining these here}, those types of beach being readily accessible
-to these surfers' native Brazil.  We now assess the null that
+Ferreira and Toledo tend to have skillsets that favour point and beach
+break beaches, those types of beach being readily accessible to these
+surfers' native Brazil; see~\cite{scarfe2003} for a scientific
+overview of surf break characteristics, and~\cite{butt2004} for a
+surfer's perspective.
+
+We now assess the null that
 Brazilians perform equally well on all types of beaches.  To that end,
@@
 noncompetitive.  How would such a strategy manifest itself in
-scorelines?
+scorelines?  We would suggest that Smith is able to
+modify his strategy to benefit from the mutual rivalry between
+Buchan and Freestone; and indeed we see Smith winning that particular
+heat.
 
 Reified Bradley-Terry technique furnishes an intuitive statistical
-test for mechanisms that control for competitors' strengths.  Noting
+test for such mechanisms, that automatically controls for competitors'
+strengths.  Noting
 that the compatriots effectively place themselves at a {\em

### 2021-08-24 inst/surfing_jsa.tex
slight changes following conversation with Tim

@@
 Brazil as a dominant nation of elite surfers.  The term was first used
-to describe the fact six out of eight querterfinalists at 2011 Nike
+to describe the fact that six out of eight querterfinalists at the 2011 Nike
 Lowers Pro were Brazilian; it rose in public conciousness following
@@
 Here, we study two aspects of this sporting phenomenon: the suggested
-Brazilian preference for beach break and point break reefs; and the
+Brazilian preference for beach break and point break wave types; and the
 noncompetitive effect of compatriot rivalry.
 
-\subsection{Brazilian beach type prefereneces}
+\subsection{Brazilian wave type prefereneces}
 
 The statistical method described here can test many different
-hypotheses drawn from surfing lore.  For example, one frequent
-suggestion~\cite{burgess2020,ho2021} is that Brazilian surfers such as
-Ferreira and Toledo tend to have skillsets that favour point and beach
-break beaches, those types of beach being readily accessible to these
-surfers' native Brazil; see~\cite{scarfe2003} for a scientific
-overview of surf break characteristics, and~\cite{butt2004} for a
-surfer's perspective.
+hypotheses inspired by surfing lore.  For example, one frequent
+suggestion~\cite{burgess2020,ho2021} is that Brazilian surfers tend to
+have skillsets that favour point and beach break wave types, readily
+accessible to these surfers' native Brazil; see~\cite{scarfe2003} for
+a scientific overview of surf break characteristics,
+and~\cite{butt2004} for a surfer's perspective.
 
-We now assess the null that
-Brazilians perform equally well on all types of beaches.  To that end,
-we create a reified surfer, ``Brazilian beachtype", whose strength
-helps all Brazilian surfers when surfing on their (putative) preferred
-beach type, an analogue of the ``home ground monster" who helps the
-home team as per equation~\ref{home_ground_advantage}.  Here we modify
-the likelihood function only when a single Brazilian is in the water,
-to avoid possible interaction with the compatriot issue discussed in
+We now assess the null that Brazilians perform equally well on all
+types of wave types.  To that end, we create a reified surfer,
+``Brazilian wavetype", whose strength helps all Brazilian surfers
+when surfing on their (putative) preferred wave type, an analogue of
+the ``home ground monster" who helps the home team as per
+equation~\ref{home_ground_advantage}.  Here we modify the likelihood
+function only when a single Brazilian is in the water, thus avoiding
+possible interaction with the compatriot issue discussed in
 section~\ref{compatriot}.  The null would be that this reified entity
@@
 = 0.035; $p=0.79$) and conclude that there is no evidence to suggest
-that Brazilians' performce is better on their preferred beach types
+that Brazilians' performce is better on their preferred wave types
 than one would expect by chance.  We may go further using the method
 of support: not only is the maximum likelihood estimate for Brazilian
-beachtype very small (about 0.002), but in addition we may reject the
+wavetype very small (about 0.002), but in addition we may reject the
 the proposal that the strength of this entity exceeds about 0.025
-(Figure~\ref{brazilian_beachtype_support}).
+(Figure~\ref{brazilian_wavetype_support}).
 
@@
   \includegraphics[width=4in]{surfing_files/figure-html/plotwidep-1.png}
-\caption{Profile support \label{brazilian_beachtype_support} for the
-  Brazilian beachtype reified entity.  The horizontal axis specifies a
-  particular strength for Brazilian beachtype, and the height of the
+\caption{Profile support \label{brazilian_wavetype_support} for the
+  Brazilian wavetype reified entity.  The horizontal axis specifies a
+  particular strength for Brazilian wavetype, and the height of the
   curve gives the support conditional on the x-axis}
@@
 beating compatriots than surfers from other countries, the thinking
-being that admission to a country's Olympic squad would be of more
+being that admission to a country's Olympic team would be of more
 importance to a competitor than their overall standing in the
@@
 countryman, possibly at the expense of performance relative to the odd
-man out.
-
-Such hypotheses are difficult to test with conventional techniques as
-competitors's strengths themselves are nuisance
+man out.  Such hypotheses are difficult to test with conventional
+techniques as competitors's strengths themselves are nuisance
 parameters~\cite{basu1977}.
@@
 competition with one another for a place in the Australian Olympic
-surfing squad, would be more motivated to beating each other than to
+surfing team, would be more motivated to beating each other than to
 come ahead of Smith.  This would be a rational response, but arguably
 noncompetitive.  How would such a strategy manifest itself in
-scorelines?  We would suggest that Smith is able to
-modify his strategy to benefit from the mutual rivalry between
-Buchan and Freestone; and indeed we see Smith winning that particular
-heat.
+scorelines?  We would suggest that Smith is able to modify his
+strategy to benefit from the mutual rivalry between Buchan and
+Freestone; and indeed we see Smith winning that particular heat.  We
+tentatively interpret this as a Nash-type equilibrium: there is no
+incentive for either Buchan nor Freestone to deviate from their
+prioritization of defeating their compatriot; and indeed neither
+Buchan nor Freestone were selected for the 2020 Olympic team.
 
-Reified Bradley-Terry technique furnishes an intuitive statistical
+The reified Bradley-Terry technique furnishes an intuitive statistical
 test for such mechanisms, that automatically controls for competitors'
@@
 the maximum likelihood estimate for $p_C$ is about 0.015, weaker than
-the ``real'' surfers, but the support for a nonzero $p_C$ is about 1.8
-[asymptotic $p=0.28$], slightly short of the two units of support
-required for rejection of our null.  We conclude that there is
-evidence, but no strong evidence, to suggest that such noncompetitive
-behaviour occurred---at least not from the order statistics we
-considered.
+the ``real'' surfers, but the support for a nonzero $p_C$ is about
+0.59 [asymptotic $p=0.28$], insufficient to reject our null.  We
+conclude that there is no strong evidence to suggest that such
+noncompetitive behaviour occurred---at least not from the order
+statistics we considered.
 
@@
 We discuss and assess the ``Brazilian storm'' phenomenon using reified
-entities and show that there is no strong evidence to support the
-supposed Brazilian preference for point and beach break beaches.  We
-also find no evidence to suggest that surfing's being included in the
-2020 Summer Olympics results in modified strategies to bolster
-intranational ranking at the expense WCT performance.
+Bradley-Terry techniques, and show that there is no strong evidence to
+support the supposed Brazilian preference for point and beach break
+beaches.  We also find no evidence to suggest that surfing's being
+included in the 2020 Summer Olympics results in modified strategies to
+bolster intranational ranking at the expense WCT performance.
 
-Further work might include  {\bf\large Tim we should discuss}
+Further work might include statistical investigation of noncompetitive
+NASH STUFF compatriot priority infringements between compatriots.
 

### 2021-08-25 inst/surfing_jsa.tex
start work on JSA house style

@@
 \usepackage{graphicx}
-\begin{document}
-\bibliographystyle{plain}
+\usepackage{natbib}
+
+  \author{Driver, Timothy${}^{1,2}$ \and
+    Hankin, Robin K. S.${}^{2,3}$ \\
+    \texttt{hankin.robin@gmail.com}\\(corresponding author)}
+
+  \date{${}^1$ {\bf Tim what do you want here you can put anything}\\
+    ${}^{1,2}$ Auckland University of Technology
+    Wellesley Street, Auckland, New Zealand\\
+    +64-9-9219999x5106\\
+    ${}^3$ Corresponding author}
 
@@
 
-\section*{Abstract}
+\begin{document}
+
+  \maketitle
+
+\subsubsection*{Abstract}
    Here we analyse competitive surfing, specifically the 2018 World Surf
@@
 research.
+\newpage
 
@@
 Surfing is a popular and growing activity in terms of both
-participation and viewing \cite{warshaw2010}.  Competitive surfing
+participation and viewing \citep{warshaw2010}.  Competitive surfing
 involves surfers competing against one or two other surfers in heats
-lasting 30-50 minutes \cite{booth1995}.  Points are awarded on a
+lasting 30-50 minutes \citep{booth1995}.  Points are awarded on a
 10-point scale: each of five judges awards points to a surfer riding a
 particular wave.  The World Surf League (WSL) is the main governing
-body of surfing competitions \cite{wsl}.  The WSL conducts a world tour in
+body of surfing competitions \citep{wsl}.  The WSL conducts a world tour in
 which the best ranked 34 surfers compete; in addition, at each tour
-venue, two "wildcard" surfers also enter the competition, who are
+venue, two ``wildcard" surfers also enter the competition, who are
 ignored here.  The 2019 WSL tour saw competitive events held at eleven
@@
 
-Farley {\em et. al.} \cite{farley2013} analyse surfing results using
+\citet{farley2013} analyse surfing results using
 one-way ANOVA techniques and report comparisons between the top 10 and
@@
 on the scoring of aerial manoeuvres compared with other manoeuvres in
-events.  Lundgren {\em et. al.} \cite{lundgren2014}, for example,
+events.  \citet{lundgren2014}, for example,
 report that aerial manoeuvres scored higher than other stunts but had
@@
 
-The Bradley-Terry model \cite{bradley1954} assigns non-negative strengths
+The Bradley-Terry model \citep{bradley1954} assigns non-negative strengths
 $p_1,\ldots, p_n$ to each of $n$ competitors in such a way that the
@@
 $\frac{p_i}{p_i+p_j}$; it is conventional to normalize so that $\sum
-p_i=1$.  Further, we use a generalization due to \cite{luce1959}, in which
+p_i=1$.  Further, we use a generalization due to \citep{luce1959}, in which
 the probability of competitor~$i$ winning in a field of $\left\lbrace
@@
 that there is information in the whole of the finishing order, and not
-just the first across the line, we can follow \cite{plackett1975} and
+just the first across the line, we can follow \citet{plackett1975} and
 consider the runner-up to be the winner among the remaining
@@
 and this would be a forward ranking Plackett-Luce model in the
-terminology of \cite{johnson2020}.  We now use a technique due to Hankin
-(2010, 2017) and introduce fictional (reified) entities whose nonzero
-Bradley-Terry strength helps certain competitors or sets of
-competitors under certain conditions.  The original example was the
-home-ground advantage in football.  If players (teams) $1,2$ with
-strengths $p_1,p_2$ compete, and if our observation were $a$ home wins
-and $b$ away wins for team $1$, and $c$ home wins and $d$ away wins
-for team~$2$, then a suitable likelihood function would be
+terminology of \citet{johnson2020}.  We now use a technique due to
+\citet{hankin2010,hankin2017} and introduce fictional (reified)
+entities whose nonzero Bradley-Terry strength helps certain
+competitors or sets of competitors under certain conditions.  The
+original example was the home-ground advantage in football.  If
+players (teams) $1,2$ with strengths $p_1,p_2$ compete, and if our
+observation were $a$ home wins and $b$ away wins for team $1$, and $c$
+home wins and $d$ away wins for team~$2$, then a suitable likelihood
+function would be
 
@@
 ground effect.  Similar techniques have been used to account for the
-first-move advantage in chess, and here we apply it to assess the
-supposed Brazilian preference for reef-breaking waves.
+first-move advantage in chess~\citep{hankin2020}, and here we apply it
+to assess the supposed Brazilian preference for reef-breaking waves.
 
@@
 because the likelihood function is exponential family and as such is
-amenable to analysis via general linear models~\cite{turner2012}.
+amenable to analysis via general linear models~\citep{turner2012}.
 However, this is not applicable when more than two entities are
@@
 Florence at about 17.5\% down to Coffin at about 1.5\%.  The {\tt
-  hyper2} software~\cite{hankin2017} may be used to formally assess
+  hyper2} software~\citep{hankin2017} may be used to formally assess
 one plausible null: that the competitors all have equal skill, that is
@@
 p_\mathrm{Wright}=\frac{1}{23}$, and any differences in placings are
-due to random variation.  The Method of Support \cite{edwards1992}
+due to random variation.  The Method of Support \citep{edwards1992}
 rejects the null (in favour of the evaluate) at $p=4\times 10^{-5}$.
@@
 strength, we maximize the likelihood subject to the null, and compare
-with the global maximum likelihood~\cite{hankin2010}.  This gives us a
+with the global maximum likelihood~\citep{hankin2010}.  This gives us a
 support difference of about 0.05, well short of Edwards's 2 units of
@@
 generation ``represents a seismic shift for Brazilian competitors onto
-an entirely new plane'' \cite{douglas2020}.
+an entirely new plane'' \citep{douglas2020}.
 
@@
 hypotheses inspired by surfing lore.  For example, one frequent
-suggestion~\cite{burgess2020,ho2021} is that Brazilian surfers tend to
+suggestion~\citep{burgess2020,ho2021} is that Brazilian surfers tend to
 have skillsets that favour point and beach break wave types, readily
-accessible to these surfers' native Brazil; see~\cite{scarfe2003} for
+accessible to these surfers' native Brazil; see~\citet{scarfe2003} for
 a scientific overview of surf break characteristics,
-and~\cite{butt2004} for a surfer's perspective.
+and~\citet{butt2004} for a surfer's perspective.
 
@@
 One measure of the rising popularity of surfing is its inclusion in
-the 2020 Summer Olympics~\cite{tulloch2019}, and it is natural to ask
-whether this would affect the competition in some way~\cite{ho2021}.
+the 2020 Summer Olympics~\citep{tulloch2019}, and it is natural to ask
+whether this would affect the competition in some way~\citep{ho2021}.
 One suggestion would be that competitors become more focussed on
@@
 techniques as competitors's strengths themselves are nuisance
-parameters~\cite{basu1977}.
+parameters~\citep{basu1977}.
 
@@
 
+\bibliographystyle{agsm}
 \bibliography{chess}

### 2021-08-25 inst/surfing_jsa.tex
hyperref as per https://editorialexpress.com/e-editor/pdftips.html
appearance changes as per instructions to authors

@@
 \usepackage{natbib}
+\usepackage[pdftex]{hyperref}
 
@@
 
-\begin{table}
-\begin{tabular}{l|l|l|l|l|l|l|l|l|l|l|l|l|l}
-competitor  & nationality&wct01  &wct02  &wct03  & \ldots &wct09&wct10&wct11&points\\ \hline
- Ferreira   &   BRA      &   1   &   5   &   17  & \ldots      &  2   &   1   &   1   &   59740\\
- Medina     &   BRA      &   5   &   5   &   17  & \ldots      &  9   &   9   &   2   &   56475\\
- Smith 	    &   ZAF      &   3   &   3   &   17  & \ldots      &  9   &   2   &   17  &   49985\\
- Toledo     &   BRA      &   9   &   2   &   5   & \ldots      &  17  &   5   &   17  &   49145\\
- Andino     &   USA      &   2   &   17  &   5   & \dots       &  5   &   5   &   9   &   46655
-\end{tabular}
-\caption{WCT tour results, conventional league table\label{resultstable}}
-\end{table}
-
 Table~\ref{resultstable} shows the official results table from the
@@
 
-\begin{table}
-\begin{verbatim}
-wct01 Colapinto Bailey Wright
-wct01 Freestone Lau Smith
-wct01 Dora Ferreira Slater
-wct01 Duru Toledo Ibelli
-wct01 Moniz Heazlewood Wilson
-...
-wct11 Medina Florence
-wct11 Colapinto Bourez
-wct11 Ferreira Slater
-wct11 Medina Colapinto
-wct11 Ferreira Medina
-\end{verbatim}
-\caption{Extract from observations: raw results\label{rawresults} from
-  WCT 2018}
-\end{table}
-
-The first line of table \ref{rawresults} shows that, at one of the
-heats in {\tt wct01}, Colapinto came first, Bailey second, and Wright
-came third.  The last line shows that, at {\tt wct11}, Ferreira came
-first and Medina second.  We can convert this dataset into a
-Plackett-Luce likelihood function but first have to remove the
-wildcards and also player Vries whose maximum likelihood strength is
-zero.  For the first line we would have Plackett-Luce likelihood function as
+Table~\ref{rawresults} shows the first few and last few entries in the
+dataset we use.  The first line shows that, at one of the heats in
+{\tt wct01}, Colapinto came first, Bailey second, and Wright came
+third.  The last line shows that, at {\tt wct11}, Ferreira came first
+and Medina second.  We can convert this dataset into a Plackett-Luce
+likelihood function but first have to remove the wildcards and also
+player Vries whose maximum likelihood strength is zero.  For the first
+line we would have Plackett-Luce likelihood function as
 
@@
 \section{Results}
-\begin{figure}
-\includegraphics[width=4in]{surfing_files/figure-html/showsurfingmaxppie-3.png} % Pie chart of strengths
-\caption{Pie chart\label{piechartmax} showing maximum likelihood strengths of the 23 competitors.  Note the
-dominance of Ferreira, Florence, and Medina}
-\end{figure}
 
@@
 likelihood-based ranking.
-
-\begin{figure}[h]
-\includegraphics[width=4in]{surfing_files/figure-html/directlogplotofstrengths-1.png}
-\caption{Scatterplot of points scored  \label{compare_likelihood_points} 
-against maximum likelihood estimates}
-\end{figure}
-
-\begin{figure}[h]
-\includegraphics[width=4in]{surfing_files/figure-html/ordertransplotplot-1.png}
-\caption{Scatterplot of ranks \label{compare_likelihood_points_rankings} of points-based ranking
-against likelihood-based ranking.  Note the anomalous position of
-Florence}
-\end{figure}
 
@@
 
-\begin{figure}
-  \includegraphics[width=4in]{surfing_files/figure-html/plotwidep-1.png}
-\caption{Profile support \label{brazilian_wavetype_support} for the
-  Brazilian wavetype reified entity.  The horizontal axis specifies a
-  particular strength for Brazilian wavetype, and the height of the
-  curve gives the support conditional on the x-axis}
-\end{figure}
-
 \section{Differential \label{compatriot} behaviour toward compatriotic competitors}
@@
 
-\begin{table}
-\begin{tabular}{l|l}
-order  & ${\mathcal L}(p_1,p_2,q,p_C)$\\ \hline
-$p_1\succ p_2\succ q $ & $ \frac{p_1}{p_1+p_2+q+p_C}\cdot\frac{p_2}{p_2+q+p_C}$\\
-$p_2\succ p_1\succ q $ & $ \frac{p_2}{p_1+p_2+q+p_C}\cdot\frac{p_1}{p_1+q+p_C}$\\
-$p_1\succ q\succ p_2 $ & $ \frac{p_1}{p_1+p_2+q+p_C}\cdot\frac{q+p_C}{p_2+q+p_C}$\\
-$p_2\succ q\succ p_1 $ & $ \frac{p_2}{p_1+p_2+q+p_C}\cdot\frac{q+p_C}{p_1+q+p_C}$\\
-$q\succ p_1\succ p_2 $ & $ \frac{q+p_C}{p_1+p_2+q+p_C}\cdot\frac{p_1}{p_1+p_2}$\\
-$q\succ p_2\succ p_1 $ & $ \frac{q+p_C}{p_1+p_2+q+p_C}\cdot\frac{p_2}{p_1+p_2}$\\
-\end{tabular}
-\caption{The \label{likelihoodcompatriot} six possible finishing
-  orders (``$\succ$'' means ``scored more points than'') for the case
-  where two compatriot surfers and an odd man out are in the water at
-  the same time, with associated likelihood function for strengths
-  $p_1,p_2,q, p_C$.  Here $p_1,p_2$ are the strengths of two
-  compatriot surfers, $q$ the strength of the odd man out, $p_C$ the
-  strength of the compatriot monster. and nationality and a third are
-  in the water at the same time}
-\end{table}
-
 We can now apply the Method of Support to the hypothesis that $p_C=0$,
@@
 \bibliography{chess}
+
+\clearpage
+\newpage
+
+\section*{Tables}
+
+\begin{table}[h]
+\begin{tabular}{llllllllllllll}
+competitor  & nationality&wct01  &wct02  &wct03  & \ldots &wct09&wct10&wct11&points\\ \hline
+ Ferreira   &   BRA      &   1   &   5   &   17  & \ldots      &  2   &   1   &   1   &   59740\\
+ Medina     &   BRA      &   5   &   5   &   17  & \ldots      &  9   &   9   &   2   &   56475\\
+ Smith 	    &   ZAF      &   3   &   3   &   17  & \ldots      &  9   &   2   &   17  &   49985\\
+ Toledo     &   BRA      &   9   &   2   &   5   & \ldots      &  17  &   5   &   17  &   49145\\
+ Andino     &   USA      &   2   &   17  &   5   & \dots       &  5   &   5   &   9   &   46655
+\end{tabular}
+\caption{WCT tour results, conventional league table\label{resultstable}}
+\end{table}
+
+\begin{table}[h]
+\begin{verbatim}
+wct01 Colapinto Bailey Wright
+wct01 Freestone Lau Smith
+wct01 Dora Ferreira Slater
+wct01 Duru Toledo Ibelli
+wct01 Moniz Heazlewood Wilson
+...
+wct11 Medina Florence
+wct11 Colapinto Bourez
+wct11 Ferreira Slater
+wct11 Medina Colapinto
+wct11 Ferreira Medina
+\end{verbatim}
+\caption{Extract from observations: raw results\label{rawresults} from
+  WCT 2018}
+\end{table}
+
+\begin{table}[h]
+\begin{tabular}{l|l}
+order  & ${\mathcal L}(p_1,p_2,q,p_C)$\\ \hline
+$p_1\succ p_2\succ q $ & $ \frac{p_1}{p_1+p_2+q+p_C}\cdot\frac{p_2}{p_2+q+p_C}$\\
+$p_2\succ p_1\succ q $ & $ \frac{p_2}{p_1+p_2+q+p_C}\cdot\frac{p_1}{p_1+q+p_C}$\\
+$p_1\succ q\succ p_2 $ & $ \frac{p_1}{p_1+p_2+q+p_C}\cdot\frac{q+p_C}{p_2+q+p_C}$\\
+$p_2\succ q\succ p_1 $ & $ \frac{p_2}{p_1+p_2+q+p_C}\cdot\frac{q+p_C}{p_1+q+p_C}$\\
+$q\succ p_1\succ p_2 $ & $ \frac{q+p_C}{p_1+p_2+q+p_C}\cdot\frac{p_1}{p_1+p_2}$\\
+$q\succ p_2\succ p_1 $ & $ \frac{q+p_C}{p_1+p_2+q+p_C}\cdot\frac{p_2}{p_1+p_2}$\\
+\end{tabular}
+\caption{The \label{likelihoodcompatriot} six possible finishing
+  orders (``$\succ$'' means ``scored more points than'') for the case
+  where two compatriot surfers and an odd man out are in the water at
+  the same time, with associated likelihood function for strengths
+  $p_1,p_2,q, p_C$.  Here $p_1,p_2$ are the strengths of two
+  compatriot surfers, $q$ the strength of the odd man out, $p_C$ the
+  strength of the compatriot monster. and nationality and a third are
+  in the water at the same time}
+\end{table}
+
+\clearpage
+\newpage
+\section*{Figures}
+
+\begin{figure}[h]
+\includegraphics[width=4in]{surfing_files/figure-html/showsurfingmaxppie-3.png} % Pie chart of strengths
+\caption{Pie chart\label{piechartmax} showing maximum likelihood strengths of the 23 competitors.  Note the
+dominance of Ferreira, Florence, and Medina}
+\end{figure}
+
+\begin{figure}[h]
+\includegraphics[width=4in]{surfing_files/figure-html/directlogplotofstrengths-1.png}
+\caption{Scatterplot of points scored  \label{compare_likelihood_points} 
+against maximum likelihood estimates}
+\end{figure}
+
+\begin{figure}[h]
+\includegraphics[width=4in]{surfing_files/figure-html/ordertransplotplot-1.png}
+\caption{Scatterplot of ranks \label{compare_likelihood_points_rankings} of points-based ranking
+against likelihood-based ranking.  Note the anomalous position of
+Florence}
+\end{figure}
+
+\begin{figure}[h]
+  \includegraphics[width=4in]{surfing_files/figure-html/plotwidep-1.png}
+\caption{Profile support \label{brazilian_wavetype_support} for the
+  Brazilian wavetype reified entity.  The horizontal axis specifies a
+  particular strength for Brazilian wavetype, and the height of the
+  curve gives the support conditional on the x-axis}
+\end{figure}
+
 \end{document}

### 2021-08-25 inst/surfing.Rmd
use PDF not PNG in surfing_jsa [and create them in surfing.Rmd]

@@
 pie(surfing_maxp)
+pdf(file="showsurfingmaxppie.pdf")
+pie(surfing_maxp)
+dev.off()
 jj <- log(surfing_maxp)
@@
 abline(fit)
+
+pdf(file="directlogplotofstrengths.pdf")
+plot(jjO,jjL,pch=16,log='xy',xlab='points',ylab='maximum likelihood strength')
+fit <- lm(log10(jjL)~log10(jjO))
+for(i in seq_along(jjO)){
+text(jjO[i],jjL[i],names(jjL)[i],col='gray',pos=4,cex=0.5)
+}
+abline(fit)
+dev.off()
 ```
@@
 ordertransplot(rO,rL,xlab="offical rank",ylab="likelihood rank",main="WCT 2019")
+pdf(file="ordertransplotplot.pdf")
+ordertransplot(rO,rL,xlab="offical rank",ylab="likelihood rank",main="WCT 2019")
+dev.off()
 ```
@@
 abline(h= -2,lty=2)
+pdf(file="plotwidep.pdf")
+plot(c(0,p_wide),c(l0,l_wide)-max(l),type='b',xlab="Brazilian beachtype wide",ylab="support")
+abline(h= -2,lty=2)
+dev.off()
 ```

### 2021-08-25 inst/surfing_jsa.tex
use PDF not PNG in surfing_jsa [and create them in surfing.Rmd]

@@
 \begin{figure}[h]
-\includegraphics[width=4in]{surfing_files/figure-html/showsurfingmaxppie-3.png} % Pie chart of strengths
+\includegraphics[width=4in]{showsurfingmaxppie.pdf} % Pie chart of strengths
 \caption{Pie chart\label{piechartmax} showing maximum likelihood strengths of the 23 competitors.  Note the
@@
 \begin{figure}[h]
-\includegraphics[width=4in]{surfing_files/figure-html/directlogplotofstrengths-1.png}
+\includegraphics[width=4in]{directlogplotofstrengths.pdf}
 \caption{Scatterplot of points scored  \label{compare_likelihood_points} 
@@
 \begin{figure}[h]
-\includegraphics[width=4in]{surfing_files/figure-html/ordertransplotplot-1.png}
+\includegraphics[width=4in]{ordertransplotplot.pdf}
 \caption{Scatterplot of ranks \label{compare_likelihood_points_rankings} of points-based ranking
@@
 \begin{figure}[h]
-  \includegraphics[width=4in]{surfing_files/figure-html/plotwidep-1.png}
+  \includegraphics[width=4in]{plotwidep.pdf}
 \caption{Profile support \label{brazilian_wavetype_support} for the

### 2021-08-25 inst/surfing_jsa.tex
tweaks

@@
 included in the 2020 Summer Olympics results in modified strategies to
-bolster intranational ranking at the expense WCT performance.
+bolster intranational ranking at the expense of WCT performance.
 
-Further work might include statistical investigation of noncompetitive
-NASH STUFF compatriot priority infringements between compatriots.
+Further work might include incorporation of more tournament results
+into our likelihood function, and possibly to investigate priority
+infringements between compatriots.
 

### 2021-08-26 inst/surfing_jsa.tex
tweaks and spellcheck

@@
 
-  \author{Driver, Timothy${}^{1,2}$ \and
-    Hankin, Robin K. S.${}^{2,3}$ \\
-    \texttt{hankin.robin@gmail.com}\\(corresponding author)}
+  \author{Driver, Timothy C.${}^{1}$ \and
+    Hankin, Robin K. S.${}^{1,2}$}
 
-  \date{${}^1$ {\bf Tim what do you want here you can put anything}\\
-    ${}^{1,2}$ Auckland University of Technology
+  \date{${}^{1}$ Auckland University of Technology\\
     Wellesley Street, Auckland, New Zealand\\
     +64-9-9219999x5106\\
-    ${}^3$ Corresponding author}
+    ${}^2$ Corresponding author\\
+    \texttt{hankin.robin@gmail.com}
+  }
+
+  % Tim:   ORCID 0000-0002-3184-2118
+  % Robin: ORCID 0000-0001-5982-0415
 
@@
 \subsubsection*{Abstract}
-   Here we analyse competitive surfing, specifically the 2018 World Surf
+Here we analyse competitive surfing, specifically the 2019 World Surf
 League, using formal statistical methods.  We use generalized
@@
 the ``Brazilian storm'' phenomenon using reified entities in two ways.
-Firstly we assess the supposed Brazilian preference for point and
-beach break beaches; and secondly we consider tournament results from
-the perspective of tournament theory and test for competitors
-modifying their strategy in the presence of compatriot rivals.  We
-quantify the evidence for these commonly assumed features of
-contemporary competitive surfing and suggest further avenues of
-research.
+Firstly we assess the supposed Brazilian preference for beach break
+and point break wave types; and secondly we consider results from the
+perspective of tournament theory and test for competitors modifying
+their strategy in the presence of compatriot rivals.  We quantify the
+evidence for these commonly assumed features of contemporary
+competitive surfing and suggest further avenues of research.
 \newpage
@@
 judges.  Subject to surfing etiquette, competitors are free to catch a
-wave at their discretion within the heat's time window, typically
-20-30 minutes.  The judges rate each wave ride and award points to the
-surfer based on a range of subjective criteria such as stylish
-execution and transition of manoeuvres, but also credit elements such
-as the difficulty and novelty of the performance.  The top score and
-bottom score among the five judges are removed and the remaining three
-judges' scores are averaged to give the final score for the surfer for
-that wave.  Each surfer's aggregate score is given by the average of
-the two highest-scoring waves.  The winner is the surfer with the
-highest aggregate score.
+wave at their discretion within the heat's time window.  The judges
+rate each wave ride and award points to the surfer based on a range of
+subjective criteria such as stylish execution and transition of
+manoeuvres, but also credit elements such as the difficulty and
+novelty of the performance.  The top score and bottom score among the
+five judges are removed and the remaining three judges' scores are
+averaged to give the final score for the surfer for that wave.  Each
+surfer's aggregate score is given by the average of the two
+highest-scoring waves.  The winner is the surfer with the highest
+aggregate score.
 
@@
 
-The Bradley-Terry model \citep{bradley1954} assigns non-negative strengths
-$p_1,\ldots, p_n$ to each of $n$ competitors in such a way that the
-probability of $i$ beating $j\neq i$ in pairwise competition is
-$\frac{p_i}{p_i+p_j}$; it is conventional to normalize so that $\sum
-p_i=1$.  Further, we use a generalization due to \citep{luce1959}, in which
-the probability of competitor~$i$ winning in a field of $\left\lbrace
-1,\ldots, n\right\rbrace$ is $\frac{p_i}{p_1+\cdots +p_n}$.  Noting
-that there is information in the whole of the finishing order, and not
-just the first across the line, we can follow \citet{plackett1975} and
-consider the runner-up to be the winner among the remaining
-competitors, and so on down the finishing order. Without loss of
-generality, if the order of finishing were $1,2,3,4,5$, then a
+The Bradley-Terry model \citep{bradley1954} assigns non-negative
+strengths $p_1,\ldots, p_n$ to each of $n$ competitors in such a way
+that the probability of $i$ beating $j\neq i$ in pairwise competition
+is $\frac{p_i}{p_i+p_j}$; it is conventional to normalize so that
+$\sum p_i=1$.  Further, we use a generalization due to
+\citep{luce1959}, in which the probability of competitor~$i$ winning
+in a field of $\left\lbrace 1,\ldots, n\right\rbrace$ is
+$\frac{p_i}{p_1+\cdots +p_n}$.  Noting that there is information in
+the whole of the finishing order, and not just the first across the
+line, we can follow \citet{plackett1975} and consider the runner-up to
+be the winner among the remaining competitors, and so on down the
+finishing order. Without loss of generality, if the order of finishing
+were $1,2,3,4,5$ [we write $1\succ 2\succ 3\succ 4\succ5$], then a
 suitable Plackett-Luce likelihood function would be
@@
 competitors' behaviour but can change the overall ranking; this would
-suggest that more formal statistical methods are neede for analysis of
+suggest that more formal statistical methods are needed for analysis of
 such datasets.  The Plackett-Luce likelihood functions are not subject
@@
 dataset we use.  The first line shows that, at one of the heats in
-{\tt wct01}, Colapinto came first, Bailey second, and Wright came
-third.  The last line shows that, at {\tt wct11}, Ferreira came first
-and Medina second.  We can convert this dataset into a Plackett-Luce
-likelihood function but first have to remove the wildcards and also
-player Vries whose maximum likelihood strength is zero.  For the first
-line we would have Plackett-Luce likelihood function as
+WCT01 (Coolangatta, QLD), Colapinto came first, Bailey second, and
+Wright came third: we write
+$\mbox{Colapinto}\succ\mbox{Bailey}\succ\mbox{Wright}$.  The last line
+shows that, at WCT11 (Banzai Pipeline), Ferreira came first and Medina
+second: $\mbox{Ferreira}\succ\mbox{Medina}$.  We can convert this
+dataset into a Plackett-Luce likelihood function but first have to
+remove the wildcards and also player Vries whose maximum likelihood
+strength is zero.  For the first line we would have Plackett-Luce
+likelihood function as
 
@@
 
-[observe that there is no requirement here for the strengths to have
-unit sum].  The entire dataset has a likelihood function that
-incorporates all 477 lines, but it includes:
+\noindent [observe that there is no requirement here for the strengths
+  to have unit sum].  The entire dataset has a likelihood function
+that incorporates all 477 lines of Table~\ref{rawresults}, but it
+includes:
 
@@
 compared; and indeed here we see comparisons between three surfers
-and, in the case of Jeep Surf Freshwater Pro (wct08), many more.  It
+and, in the case of Jeep Surf Freshwater Pro WCT08, many more.  It
 is natural to ask how much information is contained in the
@@
 table~\ref{resultstable}) cannot be used for this type of
-probabalistic analysis: the points awarded are arbitrary and changing
+probabilistic analysis: the points awarded are arbitrary and changing
 the points system would change the points scored but not affect the
@@
 Brazil as a dominant nation of elite surfers.  The term was first used
-to describe the fact that six out of eight querterfinalists at the 2011 Nike
-Lowers Pro were Brazilian; it rose in public conciousness following
+to describe the fact that six out of eight quarterfinalists at the 2011 Nike
+Lowers Pro were Brazilian; it rose in public consciousness following
 the victories of Medina and de Souza in the World Surf League
@@
 
-\subsection{Brazilian wave type prefereneces}
+\subsection{Brazilian wave type preferences}
 
@@
 = 0.035; $p=0.79$) and conclude that there is no evidence to suggest
-that Brazilians' performce is better on their preferred wave types
+that Brazilians' performance is better on their preferred wave types
 than one would expect by chance.  We may go further using the method
@@
 man out.  Such hypotheses are difficult to test with conventional
-techniques as competitors's strengths themselves are nuisance
+techniques as the competitors' strengths themselves are nuisance
 parameters~\citep{basu1977}.
@@
 
-We analysed results from the 2018 World Surf League using reified
+We analysed results from the 2019 World Surf League using reified
 Bradley-Terry likelihoods to assess a number of hypotheses of interest
@@
 
-\section*{Tables}
-
 \begin{table}[h]
 \begin{tabular}{llllllllllllll}
-competitor  & nationality&wct01  &wct02  &wct03  & \ldots &wct09&wct10&wct11&points\\ \hline
+competitor  & nationality&WCT01  &WCT02  &WCT03  & \ldots &WCT09&WCT10&WCT11&points\\ \hline
  Ferreira   &   BRA      &   1   &   5   &   17  & \ldots      &  2   &   1   &   1   &   59740\\
@@
  Toledo     &   BRA      &   9   &   2   &   5   & \ldots      &  17  &   5   &   17  &   49145\\
- Andino     &   USA      &   2   &   17  &   5   & \dots       &  5   &   5   &   9   &   46655
+ Andino     &   USA      &   2   &   17  &   5   & \dots       &  5   &   5   &   9   &   46655\\
+ $\ldots$   &            &   &   &       &       & \dots       &      &       &       &
 \end{tabular}
-\caption{WCT tour results, conventional league table\label{resultstable}}
+\caption{Conventional league table for the top five competitors in WCT
+  2019 tour. \label{resultstable} Thus at WCT01 (Coolangatta, QLD) we see
+  Ferreira coming first, Andino second and Smith third}
 \end{table}
+
+\clearpage
+\newpage
 
@@
 \begin{verbatim}
-wct01 Colapinto Bailey Wright
-wct01 Freestone Lau Smith
-wct01 Dora Ferreira Slater
-wct01 Duru Toledo Ibelli
-wct01 Moniz Heazlewood Wilson
+WCT01 Colapinto Bailey Wright
+WCT01 Freestone Lau Smith
+WCT01 Dora Ferreira Slater
+WCT01 Duru Toledo Ibelli
+WCT01 Moniz Heazlewood Wilson
 ...
-wct11 Medina Florence
-wct11 Colapinto Bourez
-wct11 Ferreira Slater
-wct11 Medina Colapinto
-wct11 Ferreira Medina
+WCT11 Medina Florence
+WCT11 Colapinto Bourez
+WCT11 Ferreira Slater
+WCT11 Medina Colapinto
+WCT11 Ferreira Medina
 \end{verbatim}
 \caption{Extract from observations: raw results\label{rawresults} from
-  WCT 2018}
+  WCT 2019.  Thus the first heat at WCT01 (Coolangatta, QLD) shows
+  competitors Colapinto, Bailey, and Wright in the water at the same
+  time, the competition order being
+  $\mbox{Colapinto}\succ\mbox{Bailey}\succ\mbox{Wright}$; as discussed
+  in the text we use only the competition order for our likelihood
+  function}
 \end{table}
+
+\clearpage
+\newpage
 
@@
 \newpage
-\section*{Figures}
 
@@
 
+\clearpage
+\newpage
+
 \begin{figure}[h]
@@
 \end{figure}
+
+\clearpage
+\newpage
 

### 2021-08-26 inst/surfing_jsa.tex
better title
keywords

@@
 
-  \title{Analysis of competitive surfing tournaments with generalized Bradley-Terry}
+  \title{Analysis of competitive surfing tournaments with generalized Bradley-Terry likelihoods}
 
@@
 competitive surfing and suggest further avenues of research.
+
+keywords: Reified Bradley-Terry, surfing, competitive surfing, noncompetitive behaviour, likelihood methods, Brazilian storm
 \newpage

### 2021-08-26 inst/surfing_jsa.tex
more tweaks

@@
 competitors' behaviour but can change the overall ranking; this would
-suggest that more formal statistical methods are needed for analysis of
-such datasets.  The Plackett-Luce likelihood functions are not subject
-to such a phenomenon and furnish an objective and coherent method for
-ranking competitors.  Further, likelihood-based methods offer one
-advantage over points-based systems: the ability to conduct
-statistically rigorous tests of meaningful nulls.
+suggest that more formal statistical methods are needed for analysis
+of such datasets~\citep{aldous2017}.  The Plackett-Luce likelihood
+functions are not subject to such a phenomenon and furnish an
+objective and coherent method for ranking competitors.  Further,
+likelihood-based methods offer one advantage over points-based
+systems: the ability to conduct statistically rigorous tests of
+meaningful nulls.
 
@@
 countryman, possibly at the expense of performance relative to the odd
-man out.  Such hypotheses are difficult to test with conventional
-techniques as the competitors' strengths themselves are nuisance
-parameters~\citep{basu1977}.
+man out.  In this context, risk sensitivity is likely to be become
+important.  One example might be aerial manoeuvres which are risky
+but, if successful, attract high scores~\citep{lundgren2014}.  Such
+strategies are known to occur in chess~\citep{hankin2020}: players
+suffering a sequence of defeats adopt a high-risk (all-or-nothing)
+approach, while players in a winning streak hunker down, play
+defensively, and entrench their lead.  Such hypotheses are difficult
+to test with conventional techniques as the competitors' strengths
+themselves are nuisance parameters~\citep{basu1977}.
 

### 2021-10-12 inst/red_bus_blue_bus.Rmd
include notes

@@
 
-```{r loadlib}
-library("hyper2")
-2+4
+```
+a,b,c in a race; a and b have an advantage S
+
+finishing order a>b>c:   L = (a+S)/(a+b+c+2S) * (b+S)/(b+c+ S)
+finishing order a>c>b:   L = (a+S)/(a+b+c+2S) * (c  )/(a+b+2S)
+finishing order c>a>b:   L = (c  )/(a+b+c+2S) * (a+S)/(a+b+2S)
+
+a,b,c,d in teams {a,b} and {c,d}
+
+{a+b} wins, a and c have S:       (a+b+S)/(a+b+c+d+2S)
+{a+b} wins, c and d have S:       (a+b)/(a+b+c+d+2S)
+{a+b} wins, a and b have S:       (a+b+2S)/(a+b+c+d+2S)
+{a+b} wins, a and b and c have S:  (a+b+2S)/(a+b+c+d+3S)
+
+Car, red bus blue bus:
+
+C  wins:  2C/(2C+RB+BB)
+RB wins:  RB/(2C+RB+BB)
+BB wins:  BB/(2C+RB+BB)
+
+[NB if red and blue are really equivalent, then RB=BB=B, say and we have
+
+C wins:  C/(C+B)
+RB wins: B/(2C+2B) = 0.5*B/(C+B)
+BB wins: B/(2C+2B) = 0.5*B/(C+B)
+
+holidays:  Florida, California, California plus a nickel:
+
+{F,C}       :  (F,C)/(F+C)  # nickel ignored in likelihood function
+{F,CN}      :  (F,C)/(F+C)  # nickel ignored in likelihood function
+{FN,C}      :  (F,C)/(F+C)  # nickel ignored in likelihood function
+{FN,CN}     :  (F,C)/(F+C)  # nickel ignored in likelihood function
+
+{F,FN}      :  (F,F+N)/(2F+N)
+{C,CN}      :  (C,C+N)/(2C+N)
+
+{F,C,CN}    :  (F,C,C+N)/(F+2C+N)   # FN missing
+{F,FN,C}    :  (F,F+N,C)/(2F+C+N)   # CN missing
+
+{FN,C,CN}   :  (F,C,C+N)/(F+2C+N)   # F missing
+{F,FN,CN}   :  (F,F+N,C)/(2F+C+N)   # C missing
+
+{F,FN,C,CN} :  (F,F+N,C,C+N)/(2F+2C+2N)  ??????????????
+
+Last line (viz {F,FN,C,CN} is problematic.  Suppose that N is actually
+important [as in, eg {F,FN} and N=0.999] but also N is immaterial when
+comparing Florida with California.  This is a perfectly reasonable
+scenario.  Then in the last line we have pretty much 50-50 probability
+between {F+N,C+N}.  But this is not consistent with {F,C} which shows
+that the chooser is not indifferent between Florida and California.
+He might be 75% Florida and 25% California; and if this is the case we
+should have {F,C} = (0.75,0.25) and {FN,CN}= (0.75,0.25) [because the
+Nickel is immaterial when comparing F and C].  So what we need is for
+{F,F+N,C,C+N} and N=0.999 to reduce to {F+N,C+N} which itself should
+reduce to {F,C}.  And the last line above has {F,FN,C,CN} =
+(0,0.5,0,0.5) which is effectively stating that he is indifferent
+between Florida and California, or at least they have equal BT
+strength.  This probability model is flawed because the only effect of
+the nickel in {F,FN,C,CN} should be to reduce the probability of F and
+C and {FN,CN} should be (F,C).
+
+The correct probability model is to use a two-stage process as
+follows:
+
+{F,FN,FC,CN}  :  p(F)  = F/(F+C) * F/(F+N)
+              :  p(FN) = F/(F+C) * N/(F+N) 
+              :  p(C)  = C/(F+C) * C/(C+N) 
+              :  p(CN) = C/(F+C) * N/(C+N)
+
+Then
+
+{F,C,CN}    :  p(F)  = F/(F+C)
+            :  p(C)  = C/(F+C) * C/(C+N)
+            :  p(CN) = C/(F+C) * N/(C+N)
+
+etc.
+
+Try this approach with red bus blue bus:
+
+L = L(C,B,red,blue)
+
+{C,RB}    : p(C)  = C/(C+B)
+{C,RB}    : p(RB) = B/(C+B)
+
+{C,BB}    : p(C)  = C/(C+B)
+{C,BB}    : p(BB) = B/(C+B)  # (same as above)
+
+{RB,BB}   : p(RB) =  red/(red+blue)
+          : p(BB) = blue/(red+blue)
+
+{C,RB,BB} :  p(C)  = C/(C+B) 
+{C,RB,BB} :  p(RB) = B/(C+B) * red/(red+blue)
+{C,RB,BB} :  p(BB) = B/(C+B) * blue/(red+blue)
 ```

### 2021-10-20 inst/dg_template.tex
JQAS issue #111

%% This file is to be used as a template for your submission.
%% Rename this file and replace the text with the text of
%% your manuscript.
%%
%% The standard LaTeX document class "article" is recommended.
%% Use options letterpaper and 12pt.
\documentclass[letterpaper,12pt]{article}

%% This is the recommended preamble for your document.

%% Load De Gruyter specific settings
\usepackage{dgjournal}
\usepackage{amsmath}

%% The mathptmx package is recommended for Times compatible math symbols.
%% Use mtpro2 or mathtime instead of mathptmx if you have the commercially
%% available MathTime fonts.
%% Other options are txfonts (free) or belleek (free) or TM-Math (commercial)
\usepackage{mathptmx}

%% Use the graphics package to include figures
\usepackage{graphics}

%% Use natbib with these recommended options
\usepackage[authoryear,comma,longnamesfirst,sectionbib]{natbib}

%% Start your document body here
\begin{document}

%% Do NOT include any fronmatter information; including the title, author names,
%% institutes, acknowledgments and title footnotes (author information, funding
%% sources, etc.). Start the document with the first section or paragraph of
%% the article.

\newcommand{\raik}{R\"{a}ikk\"{o}nen}

\section*{Abstract}
The points scoring system of Formula 1 motor racing is a long-standing
contention among fans and competitors, with changes being
controversial.  The inaugural points system was $(8,6,4,3,2)$, that
is, 8 points to the winner, 6 to second place, and so on and in 2020
the points system was changed to $(25,18,15,12,10,8,6,4,2,1)$.  The
overall ranking of the competitors is determined by summing points
awarded over a season, but this system suffers from two defects:
firstly, the points are intrinsically arbitrary in the sense that the
points system does not affect the competitors' behaviour but can
change the overall ranking.  Secondly, it is difficult to assess
statistical nulls using accumulated points.  Here, I use Plackett-Luce
likelihood to identify a ranking for the competitors, this method
including no arbitrary points system and being amenable to statistical
testing of nulls.  I go on to assess a number of reasonable points
systems objectively, by comparing likelihood-based rankings with
points-based rankings.  In one well-defined statistical sense a Borda
points system is preferred, and in another a ``winner takes all''
system is best.

\section{Introduction}\label{introduction}
Formula 1 motor racing is an important and prestigious motor sport
\citep{codling2017,jenkins2010}.  Season ranking is based on a points
allocation system wherein competitors are awarded points based on race
finishing order; points accumulate additively.  The overall
competition winner is the competitor who accumulates the most points
after the final race.

However, in the case of Formula 1 motor racing, the points system is
the subject of much controversy, having changed often since the
competition's inauguration in 1950 when the points allocation was
$(8,6,4,3,2)$---eight points to the winner, six for second place, and
so on.  This system credits only the first five finishers.  As of
2020, the current points system of $(25,18,15,12,10,8,6,4,2,1)$
credits the first 10 (we ignore the bonus point awarded for fastest
lap and assume a strictly monotonic decrease).  Arguably these two
systems could introduce different rational behaviour under zero-sum
assumptions: if, in a race, a driver knows he will place seventh under
a low-risk strategy but may place sixth by dint of driving more
aggressively, the low-risk strategy might be rational under the first
points system (which does not reward the extra ranking), but not under
the second, which does.

Still, drivers have strong incentives to maximize their ranking
irrespective of any points that may be awarded: sponsors and teams
note details of drivers' performance and a great deal of personal
pride may be at stake~\citep{gayrees2019}.  It is therefore reasonable
to assume that each driver strives to maximise his rank and this will
be done here.  If this is so, then changing the points system might
change the competitors' rankings~\citep{wood2020} but not their
behaviour: surely a defect of using points to rank the competitors.

Given that racing is a zero-sum game---and that points are
monotonically decreasing---each player will try to get as high a rank
as possible regardless of the actual points system used.  However,
there are other consistent interpretations. \cite{bakhrankova2011},
for example, considers the possibility of inter-driver collusion, a
phenomenon not pursued here; and \cite{mastromarco2009} suggest that
the frequency of rule changes is driven by factors such as driver
safety and revenue optimization.

Points systems similar to that of Formula 1 are common in other racing
sports; all have the common feature of translating ranks into points
which combine additively to generate an overall ranking.  Further, we
see points systems used in the wider context of competitive situations
such as the Eurovision Song Contest and many other such group
tournaments.

\subsection{Bradley-Terry and generalizations for rank
statistics}\label{bradley-terry-and-generalizations-for-rank-statistics}

The Bradley-Terry model \citep{bradley1952} assigns non-negative
strengths $p_1,\ldots, p_n$ to each of $n$ competitors in such a way
that the probability of $i$ beating $j\neq i$ in pairwise competition
is $\frac{p_i}{p_i+p_j}$; it is conventional to normalize so that
$\sum p_i=1$. Further, we use a generalization due to
\citet{luce1959}, in which the probability of competitor~$i$ winning
in a field of $\left\lbrace 1,\ldots, n\right\rbrace$ is
$\frac{p_i}{p_1+\cdots +p_n}$.  Noting that there is information in the
whole of the finishing order, and not just the first across the line,
we can follow \cite{plackett1975} and consider the runner-up to be the
winner among the remaining competitors, and so on down the finishing
order. Without loss of generality, if the order of finishing were
$1,2,3,4,5$, then a suitable
\citeauthor{plackett1975}-\citeauthor{luce1959} likelihood function
would be

\begin{equation}\label{competitors_1_to_5_likelihood}
\frac{p_1}{p_1+p_2+p_3+p_4+p_5}\cdot
\frac{p_2}{p_2+p_3+p_4+p_5}\cdot
\frac{p_3}{p_3+p_4+p_5}\cdot
\frac{p_4}{p_4+p_5}\cdot
\frac{p_5}{p_5}
\end{equation}

and this would be a forward ranking Plackett-Luce model in the
terminology of \cite{johnson2020}. A slight generalization allows the
incorporation of non-finishers (DNF etc). If, say, competitors 4 and 5
did not finish, we would have

\begin{equation}\label{competitors_1_to_3_only_finished}
\frac{p_1}{p_1+p_2+p_3+p_4+p_5}\cdot
\frac{p_2}{p_2+p_3+p_4+p_5}\cdot
\frac{p_3}{p_3+p_4+p_5}
\end{equation}

(observe how this likelihood function, while informative about
$p_4+p_5$, is uninformative about $p_4\left|p_4+p_5\right.$). We now
use a technique due to \cite{hankin2010,hankin2017} and introduce
fictional (reified) entities whose nonzero Bradley-Terry strength
helps certain competitors or sets of competitors under certain
conditions. The canonical example would be the home-ground advantage
in association football.  If players (teams) $1,2$ with strengths
$p_1,p_2$ compete, and if our observation were $a$ home wins and $b$
away wins for team $1$, and $c$ home wins and $d$ away wins for
team~$2$, then a suitable likelihood function would be

\[
\left(\frac{p_1+p_H}{p_1+p_2+p_H}\right)^a
\left(\frac{p_1}{p_1+p_2+p_H}\right)^b
\left(\frac{p_2+p_H}{p_1+p_2+p_H}\right)^c
\left(\frac{p_2+p_H}{p_1+p_2+p_H}\right)^d,
\]

\noindent where $p_H$ is a quantification of the beneficial home
ground effect.  Similar techniques have been used to account for the
first-move advantage in chess, and effective coordination between
members of doubles tennis teams; we may use a similar device to
account for (e.g.) wet conditions in Formula 1.  Here I analyse
seasons 2016-2019 using the \texttt{hyper2} package \citep{hankin2017}
which implements the Plackett-Luce likelihood function with additional
reified entities.

One component of Formula 1 motor racing is the starting grid.  Placing
on the starting grid is determined by time trials usually driven the
day before the race itself.  Pole position is awarded to the driver
with the fastest qualifying time, and this confers a considerable
advantage to the sitter.  In this analysis we do not consider pole
position specifically, but attempt to make inferences about the time
trials and the race itself in combination (alternatively, we treat
grid placing and P-L strengths to be conditionally independent, given
race ranking).  Similarly, we treat the driver and the team as a
single entity about which we wish to make inferences.

\section{Formula 1 dataset}\label{formula-1-dataset}

\begin{table}
\centering
%\begin{tabular}{ |c|c|c|c|c|c|  ccc |c| c|c|}
\begin{tabular}{ lccccc  ccc c cc}
 \hline
          &AUS&CHN&BHR&RUS&ESP&MON&$\ldots$&BRA&ABU\\ \hline
Hamilton  &  2&  1&  2&  4&  1&  7&$\ldots$&  4&  2\\
Vettel    &  1&  2&  1&  2&  2&  1&$\ldots$&  1&  3\\
Bottas    &  3&  6&  3&  1&Ret&  4&$\ldots$&  2&  1\\
\raik &  4&  5&  4&  3&Ret&  2&$\ldots$&  3&  4\\
Ricciardo &Ret&  4&  5&Ret&  3&  3&$\ldots$&  6&Ret\\
$\ldots$&$\ldots$&$\ldots$&$\ldots$&$\ldots$&$\ldots$&$\ldots$&$\ldots$&$\ldots$&$\ldots$\\
Hartley   &  0&  0&  0&  0&  0&  0&$\ldots$&Ret& 15\\
Button    &  0&  0&  0&  0&  0&Ret&$\ldots$&  0&  0\\
Resta     &  0&  0&  0&  0&  0&  0&$\ldots$&  0&  0\\ \hline
\end{tabular}
\caption{2017 Season results\label{results2017} table.  Each row is a
  driver and each column (after the first) a venue. We see that
  Hamilton, the first row, came second in Australia, first in China,
  second in Bahrain, fourth in Russia, and so on (Hartley, Button, and
  Resta placed last).  In the first column, we see the result from
  Australia (AUS) in which Hamilton came second, Vettel first, Bottas
  third, and so on.  Here, ``Ret'' means retired and a zero entry
  means ``did not finish''}
\end{table}

Taking 2017 as an example, Table~\ref{results2017} shows the drivers'
ranks.  It is straightforward to translate this table into a
Plackett-Luce likelihood function using the \texttt{hyper2} package;
for simplicity we will consider only the 11 top-ranked drivers (in the
Plackett-Luce likelihood function, the performance of lower-ranked
players can be weakly informative about higher-ranked players'
strengths.  For example, we see that Vettel retired twice---in
Singapore and Japan---so any player who placed in those venues will
effectively ``steal'' strength from Vettel, and generally ``give'' it
to Hamilton or Bottas).  Although it has many terms, the overall
likelihood expression for the 2017 season is of the general form

\newcommand{\pham}{p_\mathrm{Ham}}
\newcommand{\pvet}{p_\mathrm{Vet}}
\newcommand{\pbot}{p_\mathrm{Bot}}
\newcommand{\prai}{p_\mathrm{Rai}}
\newcommand{\pric}{p_\mathrm{Ric}}
\newcommand{\pver}{p_\mathrm{Ver}}
\newcommand{\pper}{p_\mathrm{Per}}
\newcommand{\poco}{p_\mathrm{Oco}}
\newcommand{\psai}{p_\mathrm{Sai}}
\newcommand{\phul}{p_\mathrm{H\"{u}l}}
\newcommand{\pmas}{p_\mathrm{Mas}}

\begin{equation}
\frac{p_\mathrm{Ham}^{20}\,p_\mathrm{Mas}^{16}\,p_\mathrm{Bot}^{19}\ldots
}{\parbox{4in}{$(p_\mathrm{Per}+p_\mathrm{Oco})(p_\mathrm{Ver}+p_\mathrm{Per}+p_\mathrm{Sai}+p_\mathrm{Hul}+p_\mathrm{Mas})\\
    \rule{10mm}{0mm}(p_\mathrm{Ric}+p_\mathrm{Per}+p_\mathrm{Mas})(p_\mathrm{Sai}+p_\mathrm{Mas})\\ \rule{20mm}{0mm}
    (p_\mathrm{Ver}+ p_\mathrm{Per}+ p_\mathrm{Oco}+ p_\mathrm{Sai}+ p_\mathrm{Hul})\ldots
     $}}.
 \end{equation}

\begin{figure}
{\centering \includegraphics{printapiechart-1}}
\caption[Maximum likelihood estimates \label{piechartstrength} of the
  strengths of the top-ranked 11 drivers in Formula 1 motor racing,
  seasons 2016-2019]{Maximum likelihood
  estimates \label{piechartstrength} of the strengths of the
  top-ranked 11 drivers in Formula 1 motor racing, seasons
  2016-2019}\label{fig:printapiechart}
\end{figure}

Finding the maximum likelihood estimate for the players' strengths is
straightforward numerically. The \texttt{hyper2} package includes a
suite of numerical optimization routines, and because they have access
to derivatives, convergence is rapid.  A graphical diagram of the
strengths is given in Figure~\ref{piechartstrength}.  We see that in
2016 the driver with the largest estimated strength was Rosberg at
about 30\%, and in years 2017-2019 was Hamilton at about 29\%, 37\%,
and 42\% respectively. As an illustration of the value of likelihood
methods (as opposed to points-based methods), a likelihood ratio test
[\texttt{samep.test()}, supplied with the \texttt{hyper2} package]
rejects the null that Hamilton and Vettel have the same strength in
2018 $(H_0\colon p_\mathrm{Ham}= p_\mathrm{Vet}$) with a likelihood
ratio of $e^{2.76}\simeq 15.8$, corresponding by Wilks's theorem to an
asymptotic \(p\)-value of about \(0.02\). We may also use the reified
entity concept to test the null hypothesis that Hamilton's strength
was unchanged from 2016, where Rosberg had the highest estimated
strength, to 2017-2019, where Hamilton did; we fail to reject this
null.

\subsection{Likelihood scoring vs points
scoring}\label{likelihood-scoring-vs-points-scoring}

Applying the current points system, for example, to the 2017 results
table we would rank the drivers as follows:

\begin{multline}\nonumber
\mbox{Hamilton}\succ
\mbox{Vettel}\succ
\mbox{Bottas}\succ
\mbox{\raik}\succ
\mbox{Ricciardo}\succ\\
\mbox{Verstappen}\succ
\mbox{Perez}\succ
\mbox{Ocon}\succ
\mbox{Sainz}\succ
\mbox{H\"{u}lkenberg}\succ
\mbox{Massa}
\end{multline}

\noindent (this happens to be identical to the ranking after the extra
point was awarded for fastest lap).  However, if we were to follow
\citet{zipf1949} and adopt a points system of $(1,1/2,1/3,\ldots)$ we
would have

\begin{multline}\nonumber
\mbox{Hamilton}\succ
\mbox{Vettel}\succ
\mbox{Bottas}\succ
\mbox{Ricciardo}\succ
\mbox{Verstappen}\succ\\
\mbox{\raik}\succ
\mbox{Perez}\succ
\mbox{Ocon}\succ
\mbox{Massa}\succ
\mbox{Sainz}\succ
\mbox{H\"{u}lkenberg}.
\end{multline}

\noindent Thus, these two systems agree on the first three places but
fourth is awarded to \raik\ under the current F1 system and Ricciardo
under \citeauthor{zipf1949}.  Compare the likelihood ranking:

\begin{multline}\nonumber
\mbox{Hamilton}\succ
\mbox{Vettel}\succ
\mbox{Bottas}\succ
\mbox{\raik}\succ
\mbox{Ocon}\succ
\mbox{Ricciardo}\succ\\
\mbox{Verstappen}\succ
\mbox{Perez}\succ
\mbox{Massa}\succ
\mbox{Sainz}\succ
\mbox{H\"{u}lkenberg}.
\end{multline}

\begin{figure}
{\centering \includegraphics{likevspoints-1}}
\caption{Ordering of the \label{orderingbypointsandlikelihood}
  top-ranked 11 drivers in Formula 1 seasons 2016-2019.  Horizontal
  axis gives official (points-based) order, and the vertical axis
  gives the likelihood order.  Thus, taking 2017 as an example, the
  points-based ordering would be Hamilton first, then Vettel, then
  \raik; while the likelihood ordering is (reading
  vertically) Hamilton, Vettel, Bottas }\label{fig:likevspoints}
\end{figure}

\noindent So for 2017 at least, we see that the current points system
agrees with likelihood ranking for the top four places, while Zipf's
law agrees to three.  We can plot one ranking against the other, shown
in Figure~\ref{orderingbypointsandlikelihood}.  Note that the
historically correct points awarded to the drivers differs from that
calculated here.  That is for two reasons: firstly, ``fastest lap''
points are not included here, and also the truncation of the order
table to the first 11 drivers can increase the rank of a driver if
non-first-11 drivers are placed.

We thus see a comparison between two ordering systems.  Taking 2017 as
an example, drivers Hamilton and Vettel are respectively first and
second according to both ranking ranking procedures; but \raik\ and
Bottas are third and fourth, and fourth and third, according to points
and likelihood respectively. We define the \emph{degree of agreement}
between the two ranking systems as the maximal value of \(r\) such
that places \(1,2,\ldots, r\) all match.  Thus, from
Figure~\ref{orderingbypointsandlikelihood}, the degree of agreement
between likelihood ranking and points ranking for the years 2016-2019
would be 11,1,2,2 respectively.

However, the points system used is essentially arbitrary.  We could
use, for example, Zipf's law to rank the drivers: award one point to
the winner, half a point to second place, one third of a point to
third, and so on; see Figure~\ref{orderingbypointsandlikelihood_zipf}
in which we see generally poorer agreement between points-based ranks
and likelihood-based ranks, with a degree of agreement of 0,1,2,2 for
the years 2016-2019 respectively.  This might be an indication that
using a Zipf law points allocation is objectively worse than the
current points system.  There are a number of plausible points systems
that might be used:

\begin{figure}
{\centering \includegraphics{likevspointszipf-1}}
\caption[As for
  Figure~\ref{orderingbypointsandlikelihood} \label{orderingbypointsandlikelihood_zipf}
  but points ranking calculated by Zipf's law]{As for
  Figure~\ref{orderingbypointsandlikelihood} \label{orderingbypointsandlikelihood_zipf}
  but points ranking calculated by Zipf's law. Note the generally
  poorer agreement between the two ranking
  systems}\label{fig:likevspointszipf}
\end{figure}

\begin{itemize}
\item The current Formula 1 system $(25,18,15,12,10,8,6,4,2,1)$
\item The inaugural Formula 1 system $(8,6,4,3,2)$
\item Zipf's law $(1,\frac{1}{2},\frac{1}{3},\frac{1}{4},\ldots)$
\item Borda $(n,n-1,n-2,\ldots,3,2,1,0)$
\item Halving system:  $(1,\frac{1}{2},\frac{1}{4},\frac{1}{8},\ldots)$
\item A ``winner takes all'' system $(1,0,\ldots)$
\end{itemize}

We note that some of these may be generalized.  We might consider a
more general Borda-like points system $(r,r-1,r-2,\ldots,3,2,1,0)$ for
fixed integer $0<r<n$ \citep{emerson2007}; the halving system can be
generalized to a geometric distribution; and the winner takes all
system can be replaced by giving equal points to the top \(r\)
competitors $(\underbrace{1,1,\ldots ,1}_{r},0,0,\ldots)$, for some
integer~$r<n$.

\begin{table}
\centering
\begin{tabular}{ |c|c|c|c|c|}
 \hline
         &    2016 & 2017 & 2018 & 2019\\ \hline
current  &    11   & 1    & 2    & 2   \\
inaugural&     8   & 1    & 2    & 2   \\
Zipf     &     0   & 1    & 2    & 2   \\
Borda    &    11   & 4    & 6    & 2   \\
WTA      &     0   & 1    & 2    & 2   \\
exp2     &     0   & 1    & 2    & 2   \\
 \hline
\end{tabular}
\caption{Degree of agreement for seasons 2016-9 with different
  \label{doapoints} points systems}
\end{table}

It is straightforward to calculate the degree of agreement for the
observed rank table for years 2016-2019, shown in
Table~\ref{doapoints}.  It is clear that there is no points system
that is the best for all four years.  However, observing that both the
likelihood ranking and the points ranking are random variables in this
paradigm suggests a method whereby we can objectively assess a given
points system.  Using sampling techniques we can repeatedly generate
an order table \emph{in silico}, using estimated driver strengths from
the observed tables.  For each of, say, 1000 such synthetic tables,
calculate drivers' maximum likelihood \citeauthor{plackett1975}
strengths, and also their points awarded according to any given points
system.  We then compare rankings generated by the
\citeauthor{plackett1975} strengths and the points awarded and note
the degree of agreement between the two, as measured by the number of
rankings correctly predicted.  This furnishes an objective assessment
of the points system used.

\subsection{Numerical results}\label{numerical-results}

\begin{figure}
{\centering \includegraphics{alottaplotsmeancorrect-1}}
\caption[Simulated Formula 1 races, seasons 2016-2019]{Simulated
  Formula 1 races, seasons 2016-2019: mean number of agreeing places
  for each of six points \label{simulated_mean} systems when compared
  with a Plackett-Luce strength
  ordering}\label{fig:alottaplotsmeancorrect}
\end{figure}

\begin{figure}
{\centering \includegraphics{alottaplotswinnercorrect-1} }
\caption[Simulated Formula 1 races, seasons 2016-2019]{Simulated
  Formula 1 races, seasons 2016-2019: probability of identifying the
  Plackett-Luce winner for each of six
  points \label{simulated_winner_correct}
  systems}\label{fig:alottaplotswinnercorrect}
\end{figure}

\begin{figure}
{\centering \includegraphics{alottaplotstotalrankorder-1}}
\caption[Simulated Formula 1 races, seasons 2016-2019]{Simulated
  Formula 1 races, seasons 2016-2019: probability of complete
  agreement between Plackett-Luce ranks and ranks calculated by each
  of six points \label{simulated_complete_agreement}
  systems}\label{fig:alottaplotstotalrankorder}
\end{figure}

We now assess the six points systems using the methodology outlined
above, using 1000 \emph{in silico} trials.  For each of the six points
systems, each of the 1000 trials results in a single non-negative
integer: the degree of agreement between the Plackett-Luce ranking and
the rankings according to the points system considered.  There are
three measures that might be used to assess the distribution of degree
of agreement: (1), the mean degree of agreement \(\overline{d}\); (2),
the probability of correctly predicting the winner,
$\operatorname{Prob}(d\geq 1)$; and (3), the probability of correctly
predicting the complete order statistic $\operatorname{Prob}(D=n)$.
These summaries are shown for each of the six points systems in
Figures~\ref{simulated_mean}, \ref{simulated_winner_correct}
and~\ref{simulated_complete_agreement} respectively.

We see that the Borda points system $(n,n-1,n-2,\ldots,3,2,1)$ gives
the highest mean number of places, and also the highest probability of
correctly predicting the winner.  However, the probability of predicting
the complete order statistic is maximized using the winnner takes all
system \((1,0,0,\ldots)\).

\section{Conclusions}

Many competitive situations involve ranking the participants and one
way of doing this is to assign points for the winner, second place,
third place, etc.  The overall ranking for a sequence of observations
is decided on the basis of accumulated points.  Because changing the
points system can change participants' overall ranking but does not
affect their behaviour, accumulated points should not be used to make
inferences about participants' skills.  Maximum likelihood estimation
of Plackett-Luce strengths furnishes a ranking system that does not
suffer from the arbitrariness of a points system.  Further, this
allows one to conduct statistical tests on a range of interesting
nulls in the context of an established suite of software.

The points system used in Formula 1 motor racing is a source of lively
debate from many perspectives, with changes being controversial.  By
treating total points scored as a random variable, it is possible to
compare different point allocation schemes against objective
Plackett-Luce ranks.  Of the six points systems considered here, a
Borda system or a winner-takes-all-system appear to be closest to
objective Plackett-Luce, depending on the exact definition of
``closest''.

%% BibTeX support
\bibliographystyle{DeGruyter}
\bibliography{formula1}

\end{document}

### 2021-11-01 inst/pudding.Rmd
pudding

---
title: "Analysis of puddings following Davidson 1970"
output: bookdown::html_document2
---

```{r setup, include=FALSE}
set.seed(0)
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("magrittr")
library("PlackettLuce")
options("digits" = 5)
```

<p style="text-align: right;">
![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
</p>

This short document creates the `pudding` suite of R objects and
presents some preliminary analysis.  It follows Davidson 1970 and the
`PlackettLuce` package.

Here is the pudding result:

```{r}
pudding
```

Here is a method to convert the pudding matrix to a likelihood function:

```{r}

    numbers <- as.character(1:6)
    puddings <- paste("pudding",numbers,sep="_")
    tyers    <- paste("tie",numbers,sep="_")

`convert` <- function(M){

    H <- hyper2()
    I <- hyper2()
    for(i in seq_len(nrow(M))){
        r_ij <- M[i,3]
        w_ij <- M[i,4]
        w_ji <- M[i,5]
        t_ij <- M[i,6]
        H[c(puddings[c(M[i,1]       )]      )] %<>% inc(w_ij)
        H[c(puddings[c(       M[i,2])]      )] %<>% inc(w_ji)
        H[c(                           "tie")] %<>% inc(t_ij) 
        H[c(puddings[c(M[i,1],M[i,2])],"tie")] %<>% dec(r_ij)

        I[c(puddings[c(M[i,1]       )]      )] %<>% inc(w_ij)
        I[c(puddings[c(       M[i,2])]      )] %<>% inc(w_ji)
        I[c(tyers   [c(M[i,1],M[i,2])]      )] %<>% inc(t_ij)
        I[c(puddings[c(M[i,1],M[i,2])],tyers[c(M[i,1],M[i,2])])] %<>% dec(r_ij)    
    }

    return(list(H,I))
}
```

Now use it:

```{r calculatelikelihoodfunctionsandmaximize,cache=TRUE}
lf <- convert(pudding)
H1 <- lf[[1]]
H2 <- lf[[2]]
m1 <- maxp(H1)
m2 <- maxp(H2)
```

```{r}
m1
m2
```

We will analyse `m1` first:

```{r}
pie(m1)
```

First we use `H1` to test a null of equal pudding strengths:

```{r,m1test}
samep.test(H1,puddings)
```

so according to `samep.test()` the puddings are all the same strength.
Now some tests where the puddings' tying ability is allowed to vary:

```{r,m2test,cache=TRUE}
samep.test(H2,puddings)
samep.test(H2,tyers)
```

### 2021-11-02 inst/pudding.Rmd
more work

@@
 ---
-title: "Analysis of puddings following Davidson 1970"
+title: "Analysis of pudding following Davidson 1970"
 output: bookdown::html_document2
@@
 library("PlackettLuce")
+library("partitions")
 options("digits" = 5)
@@
 ```{r}
-pudding
+pudding_table <- pudding  # standard hyper2 terminology
+pudding_table
 ```
@@
 
-```{r}
+```{r,cache=TRUE}
 
@@
     I <- hyper2()
+    J <- hyper2()
     for(i in seq_len(nrow(M))){
@@
         I[c(puddings[c(M[i,1],M[i,2])],tyers[c(M[i,1],M[i,2])])] %<>% dec(r_ij)
+
+        J[c(puddings[c(M[i,1]       )],"first")] %<>% inc(w_ij)
+        J[c(puddings[c(       M[i,2])]        )] %<>% inc(w_ji)
+        J[c(tyers   [c(M[i,1],M[i,2])]        )] %<>% inc(t_ij)
+        J[c(puddings[c(M[i,1],M[i,2])],tyers[c(M[i,1],M[i,2])],"first")] %<>% dec(r_ij)
     }
 
-    return(list(H,I))
+    return(list(H,I,J))
 }
@@
 ```{r calculatelikelihoodfunctionsandmaximize,cache=TRUE}
-lf <- convert(pudding)
-H1 <- lf[[1]]
-H2 <- lf[[2]]
-m1 <- maxp(H1)
-m2 <- maxp(H2)
+lf <- convert(pudding_table)
+pudding <- lf[[1]] # standard hyper2 terminology
+pudding2 <- lf[[2]]
+pudding3 <- lf[[3]]
+pudding_maxp  <- maxp(pudding)
+pudding2_maxp <- maxp(pudding2)
+pudding3_maxp <- maxp(pudding3)
 ```
@@
 ```{r}
-m1
-m2
+pudding_maxp
+pudding2_maxp
+pudding3_maxp
 ```
 
-We will analyse `m1` first:
+# Pudding 1
+
+We will analyse `pudding` first:
 
 ```{r}
-pie(m1)
+summary(pudding)
+pie(pudding_maxp)
 ```
 
-First we use `H1` to test a null of equal pudding strengths:
+First we use `pudding` to test a null of equal pudding strengths:
 
-```{r,m1test}
-samep.test(H1,puddings)
+```{r,puddingtest,cache=TRUE}
+samep.test(pudding,puddings)
 ```
@@
 so according to `samep.test()` the puddings are all the same strength.
-Now some tests where the puddings' tying ability is allowed to vary:
+
+# Pudding 2
+
+Object `pudding2` includes a pudding-specific propensity to tie.  We
+can test different nulls:
 
 ```{r,m2test,cache=TRUE}
-samep.test(H2,puddings)
-samep.test(H2,tyers)
+samep.test(pudding2,puddings)
+samep.test(pudding2,tyers)
 ```
+
+# Pudding 3
+
+Object `pudding3` includes a first-mover advantage:
+
+```{r}
+summary(pudding3)
+pudding3_maxp
+```
+
+# Davidson's analysis
+
+Davidson (1970) considers the following probability model:
+
+\[
+\operatorname{Prob}(i\succ j) = \frac{\pi_i}{\pi_i+\pi_j+\nu\sqrt{\pi_i\pi_j}}\\
+\operatorname{Prob}(j\succ i) = \frac{\pi_j}{\pi_i+\pi_j+\nu\sqrt{\pi_i\pi_j}}\\
+\operatorname{Prob}(i\sim  j) = \frac{\nu\sqrt{\pi_i\pi_j}}{\pi_i+\pi_j+\nu\sqrt{\pi_i\pi_j}}
+\]
+
+and gives the following estimate for the pudding data:
+
+```{r}
+davidson <- c(0.1388005,0.1729985, 0.1617420, 0.1653930, 0.1586805, 0.2023855, 0.7468147)
+names(davidson) <- c(puddings,"tie")
+davidson
+par(pty='s')
+plot(davidson[1:6],pudding_maxp[1:6]/(1-pudding_maxp[7]),asp=1,xlim=c(0.13,0.2),ylim=c(0.13,0.2),pch=16)
+abline(0,1)
+```
+
+Although the probability model is different, we may compare the
+probabilities of $i\succ j$, $j\succ i$, and $i\sim j$ for $i<j$ with
+the two models.  This gives us a total of $3{6\choose 2}=45$
+(nonindependent) probabilities and we may plot a scattergraph:
+
+```{r,davidsonvshyper2,cache=TRUE}
+ph <- pudding_maxp # ph == 'pudding hankin'; saves typing
+pd <- davidson     # pd == 'pudding davidson'
+print(ph)
+print(pd)
+th <- ph[7]        # th == 'tie hankin'
+td <- pd[7]        # td == 'tie davidson'
+hankin <- matrix(NA,15,3)
+davidson <- matrix(NA,15,3)
+m <- allbinom(6,2)
+m
+
+colnames(hankin) <- c("i","j","tie")
+colnames(davidson) <- c("i","j","tie")
+rownames(hankin) <- apply(m,2,paste,collapse="")
+rownames(davidson) <- apply(m,2,paste,collapse="")
+for(n in seq_len(15)){
+      i <- m[1,n]
+      j <- m[2,n]
+      dh <- ph[i] + ph[j] + th    # dh == 'denominator hankin'
+      hankin[n,1] <- ph[i]/dh
+      hankin[n,2] <- ph[j]/dh
+      hankin[n,3] <- th   /dh
+
+      dd <- pd[i] + pd[j] + td*sqrt(pd[i]*pd[j])  # dd == 'denominator davidson'
+      davidson[n,1] <- pd[i]/dd
+      davidson[n,2] <- pd[j]/dd
+      davidson[n,3] <- td*sqrt(pd[i]*pd[j])/dd
+}
+```
+
+```{r}
+print(hankin)
+print(davidson)
+par(pty="s")
+plot(c(t(hankin)),c(t(davidson)),col=c('red','green','blue'),pch=16,asp=1,xlim=c(0.25,0.45),ylim=c(0.25,0.45),type='b')
+
+abline(0,1)
+```
+
+# Appendix: a synthetic dataset
+
+Above we see that, according to this model, there is no evidence to
+suggest that the puddings' strengths, or their tying ability, differ.
+But we can adduce a dataset for which the tying ability _does_ differ.
+Consider this:
+
+```{r, makem}
+p <- pudding_table
+p[1 ,4:6] <- c(10,10,37)  # 1v2
+p[2 ,4:6] <- c(10,10,27)  # 1v3
+p[4 ,4:6] <- c(10,10,24)  # 1v4
+p[7 ,4:6] <- c(10,10,20)  # 1v5
+p[11,4:6] <- c(10,10,31)  # 1v6
+p
+```
+
+(above, we doctor the dataset so that any comparison involving pudding
+1 has a high probability of tying).
+
+Then:
+
+```{r,tiepud,cache=TRUE}
+Hp <- convert(p)[[2]]
+mHp <- maxp(Hp)
+mHpt <- mHp[7:12]  #just the tyers
+```
+
+visualise:
+
+```{r}
+mHpt
+plot(mHpt)
+```
+
+(above we see that pudding 1 does indeed have a highing tie
+probability).  Test the null of equal tying probabilities:
+
+```{r,thptest,cache=TRUE}
+(tHp <- samep.test(Hp,tyers))
+```
+
+Further, test the null the pudding 1 has zero tying probability:
+
+```{r,thpzero,cache=TRUE}
+(tHpzero <- specificp.test(Hp,"tie_3"))
+```
+
+# References
+
+* R.
+D.
+Davidson 1970.
+"On extending the Bradley-Terry model to accommodate ties in paired comparison experiments".
+_Journal of the American Statistical Association_, Volume 65, Number 329, pp317--328
+
+### Package dataset
+
+Following lines create `pudding.rda`, residing in the `data/` directory of the package.
+
+```{r,label=savecurlingdataset}
+save(pudding_table,pudding,pudding_maxp,curling2_maxp,file="pudding.rda")
+```

### 2021-11-02 inst/pudding.Rmd
scatterplot

@@
 par(pty="s")
-plot(c(t(hankin)),c(t(davidson)),col=c('red','green','blue'),pch=16,asp=1,xlim=c(0.25,0.45),ylim=c(0.25,0.45),type='b')
+plot(c(t(hankin)),c(t(davidson)),col=c('red','green','blue'),pch=16,asp=1,xlim=c(0.25,0.45),ylim=c(0.25,0.45))
+for(i in seq_len(nrow(hankin))){
+text(hankin[i,1],davidson[i,1],rownames(hankin)[i],pos=4)
+text(hankin[i,2],davidson[i,2],rownames(hankin)[i],pos=4)
+text(hankin[i,3],davidson[i,3],rownames(hankin)[i],pos=4)
+
+segments(x0=hankin[i,1],y0=davidson[i,1], x1=hankin[i,2],y1=davidson[i,2])
+}
 

### 2021-11-04 inst/pudding.Rmd
more work

--- 2021-11-02 inst/pudding.Rmd
@@
 print(pd)
-th <- ph[7]        # th == 'tie hankin'
-td <- pd[7]        # td == 'tie davidson'
-hankin <- matrix(NA,15,3)
-davidson <- matrix(NA,15,3)
-m <- allbinom(6,2)
-m
 
-colnames(hankin) <- c("i","j","tie")
-colnames(davidson) <- c("i","j","tie")
-rownames(hankin) <- apply(m,2,paste,collapse="")
-rownames(davidson) <- apply(m,2,paste,collapse="")
-for(n in seq_len(15)){
-      i <- m[1,n]
-      j <- m[2,n]
+`hankin` <- function(ph){
+  ##  ph a vector of length n+1; first n entries are strengths, last one
+  ##  propensity to tie
+  n <- length(ph)-1
+  m <- allbinom(n,2)
+  out <- matrix(NA,nrow=ncol(m),ncol=3)
+  colnames(out) <- c("i","j","tie")
+  rownames(out) <- apply(m,2,paste,collapse="")
+  th <- ph[n+1]
+  for(o in seq_len(ncol(m))){
+      i <- m[1,o]
+      j <- m[2,o]
       dh <- ph[i] + ph[j] + th    # dh == 'denominator hankin'
-      hankin[n,1] <- ph[i]/dh
-      hankin[n,2] <- ph[j]/dh
-      hankin[n,3] <- th   /dh
+      out[o,1] <- ph[i]/dh
+      out[o,2] <- ph[j]/dh
+      out[o,3] <- th/dh
+  }
+  return(out)
+}
 
-      dd <- pd[i] + pd[j] + td*sqrt(pd[i]*pd[j])  # dd == 'denominator davidson'
-      davidson[n,1] <- pd[i]/dd
-      davidson[n,2] <- pd[j]/dd
-      davidson[n,3] <- td*sqrt(pd[i]*pd[j])/dd
+`davidson` <- function(pd){
+
+  ##  pd a vector of length n+1; first n entries are strengths, last one
+  ##  nu
+  n <- length(pd)-1  # how many puddings
+  nu <- pd[n+1]  
+  m <- allbinom(n,2)
+  out <- matrix(NA,nrow=ncol(m),ncol=3)
+  colnames(out) <- c("i","j","tie")
+
+  rownames(out) <- apply(m,2,paste,collapse="")
+
+  nu <- pd[n+1]
+
+  for(o in seq_len(ncol(m))){
+      i <- m[1,o]
+      j <- m[2,o]
+      dd <- pd[i] + pd[j] + nu*sqrt(pd[i]*pd[j])  # dd == 'denominator davidson'
+      out[o,1] <- pd[i]/dd
+      out[o,2] <- pd[j]/dd
+      out[o,3] <- nu*sqrt(pd[i]*pd[j])/dd
+  }
+  return(out)
+}
+
+plotter <- function(ph,pd,...){
+  h <- hankin(ph)
+  d <- davidson(pd)
+  nh <- nrow(h)
+  plot(c(h),c(d),pch=16,asp=1,col=c(rep("red",nh),rep("blue",nh),rep("green",nh)),...)
 }
@@
 ```{r}
-print(hankin)
-print(davidson)
-par(pty="s")
-plot(c(t(hankin)),c(t(davidson)),col=c('red','green','blue'),pch=16,asp=1,xlim=c(0.25,0.45),ylim=c(0.25,0.45),type='b')
+print(ph)
+print(hankin(ph))
+print(pd)
+print(davidson(pd))
+jj <- c(2,5)/10
+par(pty='s')
+plotter(ph,pd,xlim=jj,ylim=jj)
+grid()
+```
 
-abline(0,1)
+Now calculate likelihoods:
+
+```{r}
+n <- pudding_table[,-(1:3),]
+h <- hankin(ph)
+d <- davidson(pd)
+n
+h
+d
+sum(n*log(h))
+sum(n*log(d))
+sum(n*log(h))-sum(n*log(d))
 ```

### 2021-11-04 inst/pudding.Rmd
more work

@@
 
-# Appendix: a synthetic dataset
+# Appendix:  synthetic datasets
 
@@
 
-(above we see that pudding 1 does indeed have a highing tie
+(above we see that pudding 1 does indeed have a higher tie
 probability).  Test the null of equal tying probabilities:
@@
 ```
+
+## Test of Davidson vs Hankin
+
+Although the likelihood for Davidson's model was higher than mine for
+the original dataset, we can create a synthetic dataset which is a
+random observation drawn from a reified Bradley-Terry probability
+model.  I will then test the hypothesis that the data is in fact drawn
+from an RBT distribution by calculating a likelihood ratio.
+
+```{r,synthmaker}
+set.seed(2)
+f <- function(i){rmultinom(1,50,prob=h[i,])} # f(i) = sample from row i of h
+synth <- pudding_table
+synth[,-(1:3)] <- t(sapply(1:15,f))  # random sample
+synth[,3] <- 50
+rownames(synth) <- rownames(p)
+colnames(synth) <- colnames(p)
+synth
+```
+
+Above, `synth` is a dataset drawn from an RBT model with maximum
+likelihood parameters obtained from the original dataset.  We now use
+both methods.  First reified Bradley-Terry:
+
+```{r,useboth,cache=TRUE}
+(H_synth <- convert(synth)[[1]])
+ph_synth <- maxp(H_synth)
+```
+
+```{r}
+ph_synth
+```
+
+The above strengths are different from (but not too different from)
+the value obtaine from `maxp(H)` above.  Now analyse the same dataset
+but using the method given in the `PlackettLuce` vignette:
+
+```{r}
+PlackettLuce_coef <- function(pudding_table){
+i_wins <- data.frame(Winner = pudding_table$i, Loser = pudding_table$j)
+j_wins <- data.frame(Winner = pudding_table$j, Loser = pudding_table$i)
+  ties <- data.frame(Winner = asplit(pudding_table[c("i", "j")], 1),
+                     Loser = rep(NA, 15))
+R <- as.rankings(rbind(i_wins, j_wins, ties),
+                 input = "orderings")
+w <- unlist(pudding_table[c("w_ij", "w_ji", "t_ij")])
+mod <- PlackettLuce(R, weights = w, npseudo = 0, maxit = 7)
+return(coef(mod,log=FALSE))
+}	
+```
+
+As a consistency check, try verifying the original result:
+
+```{r,verifypl,cache=TRUE}
+PlackettLuce_coef(pudding_table)
+```
+
+Above, we see values that match those in the `PlackettLuce` vignette.
+Now apply the same method to the synthetic dataset `synth`:
+
+```{r,calcpdsynth,cache=TRUE}
+(pd_synth <- PlackettLuce_coef(synth))
+```
+
+Now we need to calculate the probabilities:
+
+```{r}
+(h_synth <- hankin(ph_synth))
+(d_synth <- davidson(pd_synth))
+```
+
+And calculate a likelihood ratio for the two hypotheses, RBT and
+Davidson:
+
+```{r}
+(n <- synth[,-(1:3),])
+(lh_synth <- sum(n*log(h_synth)))
+(ld_synth <- sum(n*log(d_synth)))
+lh_synth - ld_synth
+```
+
+So reified Bradley-Terry is better than Davidson, but (in this case at
+least) not by much.
 

### 2021-11-04 inst/pudding.Rmd
work

@@
 options("digits" = 5)
+showsome <- function(D){
+jj <- round(D[c(1:4,nrow(D)),],4)
+jj[4,] <- "..."
+rownames(jj)[4] <- "..."
+noquote(jj)
+}
 ```
@@
 pudding_table <- pudding  # standard hyper2 terminology
+rownames(pudding_table) <- apply(allbinom(6,2),2,paste,collapse="v")
 pudding_table
@@
   colnames(out) <- c("i","j","tie")
-  rownames(out) <- apply(m,2,paste,collapse="")
+  rownames(out) <- apply(m,2,paste,collapse="v")
   th <- ph[n+1]
@@
   colnames(out) <- c("i","j","tie")
-
-  rownames(out) <- apply(m,2,paste,collapse="")
-
+  rownames(out) <- apply(m,2,paste,collapse="v")
   nu <- pd[n+1]
-
   for(o in seq_len(ncol(m))){
@@
 ```{r}
-n <- pudding_table[,-(1:3),]
+n <- pudding_table[,-(1:3)]
 h <- hankin(ph)
 d <- davidson(pd)
-n
-h
-d
+showsome(n)
+showsome(h)
+showsome(d)
 sum(n*log(h))
 sum(n*log(d))
-sum(n*log(h))-sum(n*log(d))
+(suppdiff <- sum(n*log(d))-sum(n*log(h)))
+```
+
+Above we see that Davidson's method has almost 3 units of support more
+than RBT.  However, we note that Davidson's model has seven degrees of
+freedom while RBT has only six (because of the the unit sum
+constraint).  Here Wilks's theorem is not applicable because the two
+models are not nested but even so the likelihood ratio is a meaningful
+statistic.
+
+```{r}
+pchisq(suppdiff,df=1,lower.tail=FALSE)
 ```
@@
 (above, we doctor the dataset so that any comparison involving pudding
-1 has a high probability of tying).
-
-Then:
+1 has a high probability of tying).  Then:
 
@@
 
-Further, test the null the pudding 1 has zero tying probability:
-
-```{r,thpzero,cache=TRUE}
-(tHpzero <- specificp.test(Hp,"tie_3"))
-```
+Above analysis shows that we can reject the null that all puddings
+have the same tying propensity.
 
@@
 ```{r,synthmaker}
-set.seed(2)
+set.seed(0)
 f <- function(i){rmultinom(1,50,prob=h[i,])} # f(i) = sample from row i of h
@@
 The above strengths are different from (but not too different from)
-the value obtaine from `maxp(H)` above.  Now analyse the same dataset
+the value obtained from `maxp(H)` above.  Now analyse the same dataset
 but using the method given in the `PlackettLuce` vignette:
@@
 w <- unlist(pudding_table[c("w_ij", "w_ji", "t_ij")])
-mod <- PlackettLuce(R, weights = w, npseudo = 0, maxit = 7)
+mod <- PlackettLuce(R, weights = w, npseudo = 0, maxit = 70)
 return(coef(mod,log=FALSE))
@@
 ```{r}
-(h_synth <- hankin(ph_synth))
-(d_synth <- davidson(pd_synth))
+showsome(h_synth <- hankin(ph_synth))
+showsome(d_synth <- davidson(pd_synth))
 ```
@@
 ```{r}
-(n <- synth[,-(1:3),])
-(lh_synth <- sum(n*log(h_synth)))
-(ld_synth <- sum(n*log(d_synth)))
+showsome(n <- synth[,-(1:3),])
+(lh_synth <- sum(n*log(h_synth)))  # Hankin; RBT
+(ld_synth <- sum(n*log(d_synth)))  # Davidson
 lh_synth - ld_synth
@@
 So reified Bradley-Terry is better than Davidson, but (in this case at
-least) not by much.
+least) not by much.  The difference is small because the puddings have
+very similar strengths.  We can repeat the analysis but with puddings
+of different strengths:
+
+```{r}
+dpudstrengths <- c(zipf(6)*0.9,tie=0.1)  # different pudding strengths
+names(dpudstrengths) <- c(puddings,"tie")
+dpudstrengths
+```
+
+Above we see that `pudding_1` is six times stronger than `pudding_6`
+in accordance with Zipf's law.  Same analysis as before:
+
+```{r}
+showsome(h2 <- hankin(dpudstrengths))
+f <- function(i){rmultinom(1,50,prob=h2[i,])} 
+synth2 <- pudding_table
+synth2[,-(1:3)] <- t(sapply(1:15,f))  # random sample
+synth2[,3] <- 50
+rownames(synth2) <- rownames(p)
+colnames(synth2) <- colnames(p)
+showsome(synth2)
+```
+
+OK so `synth2` is a different synthetic dataset but this time with
+pudding strengths following Zipf.  As before, create a `hyper2`
+object and find the evaluate:
+
+```{r,synth2maxp,cache=TRUE}
+H_synth2 <- convert(synth2)[[1]] # a hyper2 likelihood function
+ph_synth2 <- maxp(H_synth2)
+```
+
+```{r}
+ph_synth2
+```
+
+As a consistency check compare estimate with true value:
+
+```{r}
+ph_synth2-dpudstrengths  # difference is small
+```
+
+Now Davidson estimate using `synth2`:
+
+```{r,synth2david,cache=TRUE}
+(pd_synth2 <- PlackettLuce_coef(synth2))
+```
+
+```{r}
+h_synth2 <- hankin(ph_synth2)
+d_synth2 <- davidson(pd_synth2)
+showsome(h_synth2)  # same as before
+```
+
+And calculate a support function:
+
+```{r}
+n <- synth2[,-(1:3),]  # as before, extract just the counts
+(lh_synth2 <- sum(n*log(h_synth2)))
+(ld_synth2 <- sum(n*log(d_synth2)))
+lh_synth2 - ld_synth2
+```
+
+So this time the difference in support is larger, due to the puddings
+having different strengths.
 

### 2021-11-05 inst/pudding.Rmd
description of return value of convert()

@@
 This short document creates the `pudding` suite of R objects and
-presents some preliminary analysis.  It follows Davidson 1970 and the
-`PlackettLuce` package.
+presents some preliminary analysis.  It is not particularly
+well-structured or terse and is essentially a sequence of notes.  It
+follows Davidson 1970 and the `PlackettLuce` package.
 
@@
 
-Here is a method to convert the pudding matrix to a likelihood function:
+Here is a method to convert the pudding matrix to a `hyper2`
+likelihood function:
 
 ```{r,cache=TRUE}
-
 numbers <- as.character(1:6)
@@
 
+Function `convert()` gives a list of three `hyper2` objects.  The
+first assumes that each pudding has a strength $p_i$, $1\leq i\leq 6$
+and
+
+\[
+\operatorname{Prob}(i\succ j) = \frac{\pi_i}{\pi_i+\pi_j+D}\\
+\operatorname{Prob}(j\succ i) = \frac{\pi_j}{\pi_i+\pi_j+D}\\
+\operatorname{Prob}(i\sim  j) = \frac{D}{\pi_i+\pi_j+D}
+\]
+
+where $p_i,D\geq 0$ and $\sum p_i+D=1$ (here $i\succ j$ means $i$ is
+preferred to $j$ and $i\sim j$ means no preference).  The second
+element is a generalization of the first, allowing the tying
+propensities to differ:
+
+\[
+\operatorname{Prob}(i\succ j) = \frac{\pi_i}{\pi_i+\pi_j+D_i+D_j}\\
+\operatorname{Prob}(j\succ i) = \frac{\pi_j}{\pi_i+\pi_j+D_i+D_j}\\
+\operatorname{Prob}(i\sim  j) = \frac{D}{\pi_i+\pi_j+D_i+D_j}
+\]
+
+where $\sum p_i+D_i=1$.  The third element is a generalization of the
+second, giving the first pudding a "first-mover" advantage analogous
+to the home-team advantage in football, or playing white in chess:
+
 Now use it:
+
+\[
+\operatorname{Prob}(i\succ j) = \frac{\pi_i+F}{\pi_i+\pi_j+D_i+D_j+F}\\
+\operatorname{Prob}(j\succ i) = \frac{\pi_j}{\pi_i+\pi_j+D_i+D_j+F}\\
+\operatorname{Prob}(i\sim  j) = \frac{D}{\pi_i+\pi_j+D_i+D_j+F}
+\]
+
+Above, the first-mover advantage is given to pudding $i$.  We can use
+these now:
 

### 2021-11-05 inst/pudding.Rmd
more work

@@
 
-Now use it:
-
 \[
@@
 so according to `samep.test()` the puddings are all the same strength.
+We note that the propensity to tie cannot possibly be zero but we can
+give a profile likelihood curve:
+
+```{r,calcptie,cache=TRUE}
+ptie <- seq(from=0.08,len=20,to=0.15)
+l <- profsupp(pudding,"tie",ptie)
+```
+
+```{r}
+plot(ptie,l,type='b',lwd=2)
+abline(h=c(0,-2))
+#abline(v=c(0.095,0.127))
+grid()
+```
+
+and we see that a credible interval for the proclivity to tie would be
+$(0.095,0.127)$.
 
@@
 ```
+
+We can see above that the first-mover advantage is small (and indeed a
+"second-mover advantage") is also very small.
 
@@
 colnames(synth) <- colnames(p)
-synth
+showsome(synth)
 ```

### 2021-11-06 inst/karate.Rmd
use sticker in Rmd
work on Zermelo's iterative scheme for pairwise comparisons

@@
 
+<p style="text-align: right;">
+![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
+</p>
+
 Here I analyse a [karate](https://en.wikipedia.org/wiki/Karate)
@@
 [wikipedia](https://en.wikipedia.org/wiki/2018_World_Karate_Championships).
+Short story: the dataset has too many degrees of freedom to be easily
+analysed with `hyper2` and Zermelo's iteration scheme as implemented
+by the `BradleyTerry` package exploits the special pairwise structure
+of the likelihood function very effectively.
 
@@
 
-```{r converttolikelihood}
+```{r converttolikelihood,cache=TRUE}
 karate <- hyper2()
@@
 }
-karate
+summary(karate)
+```
+
+In the summary above (in the summary of bracket lengths), we see how
+the likelihood function comprises only singletons [which appear on the
+numerator] and pairs [on the denominator].
+
+```{r}
 karate_maxp <- maxp(karate)
-karate_maxp
+```
+
+```{r}
+dotchart(log10(karate_maxp))
+```
+
+# MM analysis
+
+We will define a matrix of win-loss statistics:
+
+```{r}
+pk <- pnames(karate)
+M <- matrix(0,length(pk),length(pk))
+rownames(M) <- pk
+colnames(M) <- pk
+o <- karate_table # saves typing
+for(i in seq_len(nrow(o))){
+  M[o[i,1],o[i,3]] %<>% `+`(o[i,2])
+  M[o[i,3],o[i,1]] %<>% `+`(o[i,4])
+}
+```
+
+```{r,cache=TRUE}
+pM <- pairwise(M)
+pM == karate  # consistency check; should be TRUE
+```
+
+Now use Zermelo's iteration method and compare with `hyper2`'s `maxp()`:
+
+```{r}
+iM <- zermelo(M)
+loglik(iM,karate)
+loglik(karate_maxp,karate)
+```
+
+Above we see that Zermelo gives a much more likely point in parameter
+space than `maxp()`.
+
+```{r}
+par(pty='s')
+plot(log10(iM),log10(karate_maxp),asp=1,xlim=c(-4,0),ylim=c(-4,0))
+abline(0,1)
+grid()
 ```

### 2021-11-06 inst/zermelo.Rmd
work on Zermelo's iterative scheme for pairwise comparisons

---
title: "Some thoughts on Zermelo's iteration process in the context of the hyper2 package"
output: bookdown::html_document2
---

```{r setup, include=FALSE}
set.seed(0)
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("magrittr")
options("digits" = 5)
```

<p style="text-align: right;">
![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
</p>

This short document sets out a use-case for functions `zermelo()` and
`pairwise()`.  First I create `n=10` competitors with Zipf strengths:

```{r}
n <- 10
p <- zipf(n)
names(p) <- letters[seq_len(n)]
p
```

Now I create `M`, a matrix holding pairwise comparisons between the
competitors:

```{r}
M <- matrix(0,n,n)
rownames(M) <- names(p)
colnames(M) <- names(p)
index <- which(upper.tri(M),arr.ind=TRUE)
for(o in seq_len(nrow(index))){
  i <- index[o,1]
  j <- index[o,2]
  prob <- p[i]/(p[i] + p[j]) # Bradley-Terry
  nobs <- 80 
  i_wins <- rbinom(1,nobs,prob)
  j_wins <- nobs - i_wins
  M[i,j] <- i_wins
  M[j,i] <- j_wins
}
M
```

Above we see that `a` plays `b` a total of $48+32=80$ times, winning
$48$ and losing $32$.  The probability of `a` winning is
$\frac{p_1}{p_1+p_2}=\frac{1}{1+\frac{1}{2}}=\frac{2}{3}$ so we would
expect $80\cdot\frac{2}{3}=53.3$ victories.  We can estimate the
strengths of the competitors using either Zermelo's iteration scheme
or function `maxp()`.  First Zermelo:

```{r}
(pZ <- zermelo(M))
```

And now hyper2:

```{r,cache=TRUE}
(H <- pairwise(M))
(pH <- maxp(H))
```

We can compare the two using scattergraphs and likelihood:

```{r}
par(pty='s')
plot(pZ,pH,asp=1,log='xy')
abline(0,1)
```

```{r}
plot(log(pZ),log(abs(pH-pZ)/pZ))
abline(h=0)
```

```{r}
loglik(pZ,H)-loglik(pH,H)
```

Above we see that the two methods agree closely in both point and
likelihood.  Also, both evaluates are reasonably close to the true Zipf
strengths:

```{r}
par(pty='s')
plot(zipf(n),log(abs(zipf(n)-pZ)/zipf(n)),pch=16,col='black',log='x')
```

Now we can have a bit of fun with Zermelo:

```{r}
o <- zermelo(M,give=TRUE)
head(o)
l <- apply(o,1,loglik,H)
plot(log(abs(l-max(l))))
```

### 2021-11-06 R/hyper2.R

@@ -1068,3 +1068,43 @@ setGeneric("pnames<-",function(x,value){standardGeneric("pnames<-")})
   }
 }
 
+`zermelo` <- function(M,maxit=100,start,tol=1e-10,give=FALSE){
+    diag(M) <- 0  # usual convention is NA on the diagonal
+    rM <- rowSums(M)  # only need to calculate this once
+    M <- M+t(M)
+    if(missing(start)){
+        p <- rep(1/nrow(M),nrow(M))  # starting point for iteration
+    } else {
+        p <- start
+    }
+    if(give){
+        pout <- matrix(0,maxit+1,ncol(M))
+        colnames(pout) <- colnames(M)
+        pout[1,] <- p
+    }
+    for(i in seq_len(maxit)){
+        pnew <- rM/colSums(M/outer(p,p,`+`))  # the meat
+        pnew[is.nan(pnew)] <- 0
+        pnew <- pnew/sum(pnew)  # normalize
+        if(give){pout[i+1,] <- pnew}
+        if(all(abs(p-pnew)<tol)){break}else{p <- pnew}  # maybe finish early
+    }
+    if(give){
+        return(pout[seq_len(i+1),])
+    } else {
+        return(pnew)
+    }
+}
+
+`pairwise` <- function(M){
+    diag(M) <- 0
+    if(is.null(rownames(M))){
+        rownames(M) <- paste("p",seq_len(nrow(M)),sep="_")
+        colnames(M) <- rownames(M)
+    }
+    jj <- M + t(M)
+    
+    index <- which(upper.tri(M),arr.ind=TRUE)
+    dirichlet(rowSums(M)) - hyper2(apply(index,1,function(x){rownames(M)[x]},simplify=FALSE),jj[index])
+}
+

### 2021-11-06 inst/karate.Rmd
consistentification

@@
 ```{r}
-iM <- zermelo(M)
-loglik(iM,karate)
+karate_zermelo <- zermelo(M)
+loglik(karate_zermelo,karate)
 loglik(karate_maxp,karate)
@@
 par(pty='s')
-plot(log10(iM),log10(karate_maxp),asp=1,xlim=c(-4,0),ylim=c(-4,0))
+plot(log10(karate_zermelo),log10(karate_maxp),asp=1,xlim=c(-4,0),ylim=c(-4,0))
 abline(0,1)
@@
 ```{r,label=savekaratedataset}
-save(karate_table,karate,karate_maxp,file="karate.rda")
+save(karate_table,karate,karate_maxp,karate_zermelo,file="karate.rda")
 ```

### 2021-11-09 inst/hunter.Rmd
discussion of issue #118

---
title: "Iteration techniques for Plackett-Luce likelihood functions and the `hyper2` package"
output: bookdown::html_document2
---

```{r setup, include=FALSE}
set.seed(0)
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("magrittr")
options("digits" = 5)
```

<p style="text-align: right;">
![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
</p>

Here I discuss why implementing equation 30 of Hunter (2004) for
`hyper2` objects representing Plackett-Luce likelihood functions is
difficult.  Consider the following rank observations: `724356` [that
is, 7 came first, then 2, then 4 then 3 then 5 then 6] and `1372`
[that is, 1 came first, then 3 then 7 then 2].  This would have a
Plackett-Luce likelihood function of

\[
\left(
\frac{[7]}{[234567]}\cdot\frac{[2]}{[23456]}\cdot\frac{[4]}{[3456]}\cdot\frac{[3]}{[356]}\cdot\frac{[5]}{[56]}
\right)\left(\frac{[1]}{[1237]}\cdot\frac{[3]}{[237]}\cdot\frac{[7]}{[27]}\right)
\]

(where $[1234]$ is short for $p_1+p_2+p_3+p_4$).  Then Hunter's
iteration formula gives, for example:

\[
p_3\longrightarrow\frac{p_3}{(p_7+p_2+p_4+p_3)^{-1}+(p_1+p_3)^{-1}}=
\frac{[3]}{[7243]^{-1}+[13]^{-1}}
\]

However, the `hyper2` object corresponding to the PL likelihood function would be

\[
\frac{
  [1][2][3]^2[4][5][7]^2
}{
  [1237]\cdot [23456]\cdot [234567]\cdot [237]\cdot [27]
\cdot [3456] \cdot [356]\cdot[56]
}
\]

(Note that the brackets have been rearranged and also the brackets
themselves.  Also we have no idea which term on the numerator goes
with which one on the denominator).  The desired iterative step would
be

$p_1 \longrightarrow [1]$

$p_2 \longrightarrow\frac{[2]}{[27]^{-1}+[1237]^{-1}}$

$p_3 \longrightarrow\frac{[3]}{[2347]^{-1}+[13]^{-1}}$

$p_4 \longrightarrow\frac{[4]}{[247]^{-1}}$

$p_5 \longrightarrow\frac{[5]}{[23457]^{-1}}$

$p_6 \longrightarrow\frac{[6]}{[234567]^{-1}}$

$p_7 \longrightarrow\frac{[7]}{[137]^{-1}}$

and it does not seem at all easy to convert the `hyper2` object to
Hunter's iteration scheme.  To update $p_3$, for example, we would
start with the denominator $[1237]\cdot[23456]\cdot[234567]\cdot
[237]\cdot[27]\cdot[3456]\cdot[356]\cdot[56]$ and then extract just
the brackets containing 3 to get $[1237]\cdot[23456]\cdot[234567]\cdot
[237]\cdot[3456]\cdot[356]$ and then somehow identify two increasing chains

$[27]\longrightarrow[237]\longrightarrow [1237]$

and

$[356]\longrightarrow[3456]\longrightarrow[23456]\longrightarrow[234567]$

and _then_ somehow identify from these two chains that 1 beats 3
(leading to $[13]$ in the iteration) and also that $\left\lbrace
2,4,7\right\rbrace$ beat 3 (leading to $[2347]$).

### Reference {-}

* D. R. Hunter 2004.  "MM algorithms for generalized Bradley-Terry
  models".  _The Annals of Statistics_, 32(1):384

### 2021-11-09 inst/rowing.Rmd
consistency check
add hex sticker to more Rmd files
minor typesetting

@@
 
+<p style="text-align: right;">
+![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
+</p>
+
 This short document analyses the `rowing` dataset: sculls in the 2016
@@
 sculling, up to six individual competitors race a small boat called a
-scull over a course of length 2\,km; the object is to cross the
-finishing line first.  Note that actual timings are irrelevant, given
-the model, as the sufficient statistic is the order in which
+scull over a course of length $2\,\mathrm{km}$; the object is to cross
+the finishing line first.  Note that actual timings are irrelevant,
+given the model, as the sufficient statistic is the order in which
 competitors cross the finishing line.  The 2016 Summer Olympics is a
-case in point: the gold and silver medallists finished less than~5
-milliseconds apart, corresponding to a lead of~$\sim
+case in point: the gold and silver medallists finished less than $\sim
+5$ milliseconds apart, corresponding to a lead of $\sim
 2.5\,\mathrm{cm}$.  The dataset is as follows;
@@
 dotchart(log10(rowing_maxp))
+```
+
+```{r conscheck, cache=TRUE,fig.cap="consistency check for rowing function"}
+consistency(rowing)
 ```

### 2022-03-13 inst/surfing_jsa_rebuttal.tex
codify referees' comments

Referee 1

To the authors,

An interesting paper submitted, thank you. My only
real concerns are your actual comments related to the sport of
competitive surfing.

Firstly, some confusion as you initially state heats involve one or
two other competitiors. However, you begin the second paragraph in the
introduction with a statement saying 'up to four surfers', so which is
it?

Second statement is that the World Surf League is not the main
governing body of surfing competition. What about the International
Surfing Association, who organised and ran the Olympics, also run the
World Surfing games and national level competitions. The WSL really
only run professional level competition.

34 surfers are for the mens competition.

Why are you ignoring the "wildcards". No real justification for
this. Especially as some of the wildcards competed in more than one of
the competitions. So some form of statement to validate why you omit
the wildcards would be good.

One other comment is based around the statements around judging in the
second last paragraph before section 1.1. Judging is based on the
heats conditions, as surfing is performed within an ever changing
environment. Therefore, to try and compare between surfers throughout
a competition is really not applicable. As each heat, and each wave is
different from the previous. So a direct comparison throughout a
competition is somewhat unfair for the competitors. As for the claim
around abilities, is this not subjective, as each performer is judged
upon their surfing prowess on that particular wave in that particular
heat.

Your hypothesis on Brazilians preference for reef breaking waves is based on what?

Statement on the points system not affecting competitiors
behaviour. How could you say this, as depending on the finishing spot
at each competition, the athlete is awarded a set number of points. So
a higher placing in a competition will award you a greater number of
points. Impacting upon your placing in the rankings for the subsequent
competition. And therefore, your seeding for the next competition.

Other than that, an interesting paper. Just feel you need to clarify a
few of the points above to provide readers a clearer viewpoint of the
sport and understanding to the importance of your study. As I found it
quite interesting. Just some claims and anomalies related to
competitive surfing.

Referee 2

The World Surf League (WSL) is the main governing
body of surfing competitions (World Surf League 2021).

Not exactly. What about ISA (world surfing games, Olympics, national
championships, regional championships).

The WSL conducts a
world tour in which the best ranked 34 surfers compete;

This only constitutes for male competition

in addition, at each tour venue, two "wildcard" surfers also enter the
competition, who are ignored here

Why are they ignored? Especially as some of the wildcards competed in multiple events.

Up to four surfers--the competitors--are in the water simultaneously,
watched by up to five judges.

You stated one or two other competitors above. So which is it?

However, for a particular heat, the order statistic--that is, which
competitor scored most highly, second highest, and third--is
informative about competitors abilities: the wave environment is
common to each surfer

But judging is based on the heats conditions, as surfing is performed
within an ever changing environment. Therefore, to try and compare
between surfers throughout a competition is really not applicable. As
each heat, and each wave is different from the previous. So a direct
comparison is somewhat unfair for the competitors.

and here we apply it to assess the supposed Brazilian preference for
reef-breaking waves.

Based on what hypothesis?

the points system itself is intrinsically arbitrary: the details of
the points system does not affect the competitors' behaviour but can
change the overall ranking;

How could you say this, as depending on finishing spot at each
competition, the athlete is awarded a set number of points. So a
higher placing in a competition will award you a greater number of
points. Impacting upon your placing in the rankings for the subsequent
competition. And therefore, your seeding for the next competition.

### 2022-03-13 inst/zermelo.R

@@ -1,0 +1,18 @@
+library(hyper2)
+n <- 10
+p <- zipf(n)
+names(p) <- letters[seq_len(n)]
+M <- matrix(0,n,n)
+rownames(M) <- names(p)
+colnames(M) <- names(p)
+index <- which(upper.tri(M),arr.ind=TRUE)
+for(o in seq_len(nrow(index))){
+  i <- index[o,1]
+  j <- index[o,2]
+  prob <- p[i]/(p[i] + p[j]) # Bradley-Terry
+  nobs <- 80 
+  i_wins <- rbinom(1,nobs,prob)
+  j_wins <- nobs - i_wins
+  M[i,j] <- i_wins
+  M[j,i] <- j_wins
+}

### 2022-03-17 inst/surfing_jsa_rebuttal.tex
zoom conversation with Tim
latex-ready rebuttal

@@
-Referee 1
+\documentclass[12pt]{article}
+\usepackage{xcolor}
+\begin{document}
+
+\section*{MS596 ``Analysis of competitive surfing tournaments with generalized Bradley-Terry likelihoods" by Driver and Hankin: rebuttal}
+
+Below, the reviewer's comments are in black, and our replies to the
+issues are in \textcolor{blue}{blue}.  We have indicated changes to
+the manuscript where appropriate.
+
+Short story: we have accommodated all the comments with rewording and
+clarification.  The resulting document is, I believe, stronger and
+more scholarly than before and I recommend it to you.
+
+\section*{Detailed rebuttal: Reviewer \#1}
 
@@
 
+{\bf intent: } we should remove "up to four surfers"
+
 Second statement is that the World Surf League is not the main
@@
 
+{\bf intent: }  WSL is the *professional* body organisation.
+
 34 surfers are for the mens competition.
+
+{\bf intent: } We focus on the male competition for the moment
 
@@
 
+{\bf intent:}  Will put some statistical comment in here.
+
 Your hypothesis on Brazilians preference for reef breaking waves is based on what?
+
+{\bf intent: } We tested for Brazilians NOT liking reef breaking waves
+and liking point and beach breaking waves.  consistent with 6.1
 
@@
 
+{\bf intent: } need to rephrase
+
 Other than that, an interesting paper. Just feel you need to clarify a
@@
 
-Referee 2
+\section*{Detailed rebuttal: Reviewer \#2}
 
@@
 
+{\bf intent}  dealt with above
+
 Not exactly. What about ISA (world surfing games, Olympics, national
 championships, regional championships).
+
+{\bf intent}  ditto
 
@@
 
+{\bf intent}  sorted above
+
 in addition, at each tour venue, two "wildcard" surfers also enter the
@@
 Why are they ignored? Especially as some of the wildcards competed in multiple events.
+
+{\bf intent}  sorted above.
 
@@
 competition. And therefore, your seeding for the next competition.
+\end{document}

### 2022-03-17 inst/zermelo.R

@@ -1,18 +1,0 @@
-library(hyper2)
-n <- 10
-p <- zipf(n)
-names(p) <- letters[seq_len(n)]
-M <- matrix(0,n,n)
-rownames(M) <- names(p)
-colnames(M) <- names(p)
-index <- which(upper.tri(M),arr.ind=TRUE)
-for(o in seq_len(nrow(index))){
-  i <- index[o,1]
-  j <- index[o,2]
-  prob <- p[i]/(p[i] + p[j]) # Bradley-Terry
-  nobs <- 80 
-  i_wins <- rbinom(1,nobs,prob)
-  j_wins <- nobs - i_wins
-  M[i,j] <- i_wins
-  M[j,i] <- j_wins
-}

### 2022-03-18 inst/surfing_jsa_rebuttal.tex
add results table so we can discuss wildcards

@@
 
+\subsection*{Number of competitors}
+
 To the authors,
@@
 
+\subsection*{WSL not the main governing body}
+
 Second statement is that the World Surf League is not the main
@@
 
+\subsection*{Male vs female competition issues}
+
 34 surfers are for the mens competition.
@@
 {\bf intent: } We focus on the male competition for the moment
+
+\subsection*{Ignoring wildcards}
 
@@
 the wildcards would be good.
+
+\subsection*{Direct comparison on particular waves}
 
@@
 
+\subsection*{Brazilian preference}
+
 Your hypothesis on Brazilians preference for reef breaking waves is based on what?
@@
 and liking point and beach breaking waves.  consistent with 6.1
+
+\subsection*{Points system afffecting behaviour}
 
@@
 {\bf intent: } need to rephrase
+
+\subsection*{Clarification}
 
@@
 competition. And therefore, your seeding for the next competition.
+
+\begin{verbatim}
+             nation wct01 wct02 wct03 wct04 wct05 wct06 wct07 wct08 wct09 wct10 wct11 points
+ Ferreira       BRA     1     5    17     5    17     2    17     9     2     1     1  59740
+ Medina         BRA     5     5    17    17     5     1     2     1     9     9     2  56475
+ Smith          ZAF     3     3    17     5     2     9     3     5     9     2    17  49985
+ Toledo         BRA     9     2     5    17     1     3     9     2    17     5    17  49145
+ Andino         USA     2    17     5     2     3     3    17    17     5     5     9  46655
+ Igarashi       JPN     9     9     1     9     5     5    17     9    17     3    17  40185
+ Florence       HAW     3     1    17     1     5   INJ   INJ   INJ   INJ   INJ     5  37700
+ Slater         USA    33     5     3     9     9     9    17     9    17     9     3  34845
+ Wright         AUS     9     9    17     9    17     5     1     3    17    17    17  34780
+ Flores         FRA    17     9     2    17    33    17     5    17     1    17    33  32515
+ Wilson         AUS    17    17     9     3     5    17     9     5     9    17     9  31515
+ Moniz          HAW     5     9    17     5    17    17     3    17     9    33     9  29525
+ Bourez         FRA     9    17     9     9     9     9     9    33     9    17     5  29315
+ Callinan       AUS    17     3     9     5    17     9    17    17     5    33    17  27535
+ Freestone      AUS    17    17     9    17    17    17     9    17     3     5     5  27535
+ Colapinto      USA    17   INJ    17    17     9    17     9     3    17     9     3  27450
+ Ibelli         BRA    33    17    17     3    17    17     5    17    17     3     9  26885
+ Carmichael     AUS     5    17     5    33     9    17    17     9     9     9    17  26760
+ Buchan         AUS    17    17     5    33    33     5    17     5     5    17    17  25630
+ Coffin         USA     5     9     9     9    17    17    17    17    17     9    17  23345
+ Crisanto       BRA    17     9    17     9    33     9    17    33    17     5     9  23345
+ Dora           BRA     9    17    17    17    17    17    17     5     9    17     5  22780
+ Silva          BRA    17     9    17    17     9     9     9     9    33    17    17  21920
+ Cardoso        BRA     9     9    17    17    17     9    17     9    17    17    17  19930
+ Mendes         BRA    17    33     9    17     9    33    17    17    17     9     9  19930
+ Rodrigues      BRA    17    17     3    17     9    17    33    33    17     9    33  19640
+ Zietz          HAW    17   PAR    33     9    17     5    17     9    33    17    17  18300
+ Duru           FRA    17    33     9    17     9    17     9    17    17    17    17  17940
+ Lau            HAW    17    17    33    17    17     9    17     9     9    33    17  17940
+ Bailey         AUS    17    17    33    17    17    17    17    17    17     9     9  15950
+ Fioravanti     ITA    33    17     9    17   INJ   INJ   INJ   INJ     3    17    33  14455
+ Andre          BRA    33    17    17    17    17    33     5    17    33    17    17  14320
+ Christie       NZL    17    17    17    17    17    17    17    17    33    33     9  13960
+ Morais         PRT     -     -     -    33     3    17    33     -    17    17    33  10870
+ deSouza        BRA   INJ   INJ   INJ   INJ    17    17     5   INJ   INJ   INJ   INJ   8995
+ Wright_M       AUS     9    17    17   INJ   INJ   INJ   INJ   INJ   INJ   INJ   INJ   7570
+ Willcox        AUS     -     5    33    33     -     -     -     -     -     -     -   5275
+ Lacomare       FRA     -     -     -     -     -     -     -     -     5     -     -   4745
+ Heazlewood     AUS     9    17     -     -     -     -     -     -     -     -     -   4650
+ Robinson       AUS     -     -     -     9     -     -     -     -     -     -     -   3320
+ Vaast          FRA     -     -     -     -     -     -     9     -     -     -     -   3320
+ Colapinto_C    USA     -     -     -     -     -     -     -    17     -    17     -   2660
+ Herdy          BRA    17     -     -     -   INJ     -     -    33     -     -     -   1860
+ Couzinet       FRA     -     -     -     -     -    33     -     -    17     -     -   1595
+ Waida          INA     -     -    17     -     -     -     -     -     -     -     -   1330
+ Kymerson       BRA     -     -     -     -    17     -     -     -     -     -     -   1330
+ February       ZAF     -     -     -     -     -    17     -     -     -     -     -   1330
+ Mamiya         HAW     -     -     -     -     -     -     -    17     -     -     -   1330
+ Matson         USA     -     -     -     -     -     -     -    17     -     -     -   1330
+ Schilling      USA     -     -     -     -     -     -     -    17     -     -     -   1330
+ Mignot         FRA     -     -     -     -     -     -     -     -    17     -     -   1330
+ Ribeiro_V      PRT     -     -     -     -     -     -     -     -     -    17     -   1330
+ Blanco         PRT     -     -     -     -     -     -     -     -     -    17     -   1330
+ deVault        HAW     -     -     -     -     -     -     -     -     -     -    17   1330
+ Kemper         HAW     -     -     -     -     -     -     -     -     -     -    17   1330
+ Mann           AUS     -    33     -     -     -     -     -     -     -     -     -    265
+ Huxtable       AUS     -    33     -     -     -     -     -     -     -     -     -    265
+ Ribeiro_A      BRA     -     -     -     -    33     -     -     -     -     -     -    265
+ DeVries        ZAF     -     -     -     -     -    33     -     -     -     -     -    265
+ Newton         HAW     -     -     -     -     -     -    33     -     -     -     -    265
+ Drollet        PYF     -     -     -     -     -     -    33     -     -     -     -    265
+\end{verbatim}
 \end{document}

### 2022-03-19 inst/surfing_jsa_rebuttal.tex
referee: why ignore wildcards?

@@
 the wildcards would be good.
+
+\textcolor{blue}{See the table of results at the end of this document.
+  We see that the wildcards participate in very few venues: of the 25
+  wildcards, 20 appear at only a single venue, four at two venues, and
+  only one (Willcox) at three venues.  Further, the wildcards perform
+  poorly: of the 31 ranked scores, only 5 are better than last or
+  second to last.  Maximum likelihood Bradley-Terry strengths for such
+  competitors are either exactly zero or very small (here $\sim
+  10^{-5}$) and are neither interesting from a surfing lore
+  perspective, nor susceptible to meaningful statistical
+  inference.\newline Quite apart from the fact that our focus was on
+  the top-performing competitors, including such datapoints is
+  problematic for several reasons.  Firstly, numerical optimization
+  techniques tend to struggle as one increases the number of players:
+  one is necessarily working with a higher-dimensional problem,
+  leading to poorly defined maxima [the Jacobian becomes very small];
+  in addition, local maxima become more numerous and difficult to
+  identify.\newline We have added a brief summary of this to the
+  manuscript.}
 
@@
 
+\textcolor{blue}{Discussed above}
+
 {\bf intent}  sorted above.
@@
 
+{\footnotesize
 \begin{verbatim}
@@
 \end{verbatim}
+}
+The wildcards are from Willcox onwards.
+
 \end{document}

### 2022-03-19 inst/surfing.Rmd
referee: why ignore wildcards?

@@
 ```
+
+Some quick analysis of the wildcard surfers.  First, how did the wildcards perform?
+
+```{r}
+wilds <- as.matrix(surfing_table[-(1:36),2:12])
+wilds[wilds == "INJ"] <- NA
+wilds[wilds == "-"] <- NA
+table(wilds)
+```
+
+Above we see that 10 wildcards placed "33" [meaning last place], 16
+placed "17" [second to last] three placed "9" and two placed "5".
+[the placings are treated as character strings which is why they are
+not in numerical orer].  Next, how many venues did each wildcard
+participate at?
+
+```{r}
+table(apply(wilds,1,function(x){sum(!is.na(x))}))
+```
+
+We see that 20 wildcards appeared at 1 venue, 4 at two venues, and one
+(Willcox) at three venues.q
 

### 2022-03-19 inst/surfing_jsa.tex
referee: why ignore wildcards?

@@
 particular wave.  The World Surf League (WSL) is the main governing
-body of surfing competitions \citep{wsl}.  The WSL conducts a world tour in
-which the best ranked 34 surfers compete; in addition, at each tour
-venue, two ``wildcard" surfers also enter the competition, who are
-ignored here.  The 2019 WSL tour saw competitive events held at eleven
+body of surfing competitions \citep{wsl}.  The WSL conducts a world
+tour in which the best ranked 34 surfers compete; in addition, at each
+tour venue, two ``wildcard" surfers also enter the competition, who
+are ignored here due to their generally poor performance and low
+participation.
+
+The 2019 WSL tour saw competitive events held at eleven
 different surf locations.

### 2022-03-19 inst/surfing_jsa_rebuttal.tex
referee: "is it up to four surfers"
typo
referee: men's championship

@@
 
-{\bf intent: } we should remove "up to four surfers"
+\textcolor{blue}{We have removed the inconsistency (and made the text
+  more precise) by rephrasing the introduction which now reads:
+  ``Competitive surfing involves surfers competing against one {\em or
+    more} other surfers in heats\ldots''}
 
@@
 
-{\bf intent: } We focus on the male competition for the moment
+\textcolor{blue}{We have clarified that we are studying the men's
+  championship}
 
@@
 
-{\bf intent}  sorted above
+\textcolor{blue}{We have clarified that we are studying the men's
+  championship}
 
@@
 
-{\bf intent}  sorted above.
-
 Up to four surfers--the competitors--are in the water simultaneously,
@@
 You stated one or two other competitors above. So which is it?
+
+\textcolor{blue}{Discussed above}
 

### 2022-03-19 inst/surfing_jsa.tex
referee: "is it up to four surfers"
referee: men's championship

@@
 \subsubsection*{Abstract}
-Here we analyse competitive surfing, specifically the 2019 World Surf
+Here we analyse competitive surfing, specifically the 2019 Men's World Surf
 League, using formal statistical methods.  We use generalized
@@
 participation and viewing \citep{warshaw2010}.  Competitive surfing
-involves surfers competing against one or two other surfers in heats
+involves surfers competing against one or multiple other surfers in heats
 lasting 30-50 minutes \citep{booth1995}.  Points are awarded on a

### 2022-03-19 inst/surfing_jsa_rebuttal.tex
refree: WSL is for professional surfers
referee: brazilians do not like reef breaking waves

@@
 
-{\bf intent: }  WSL is the *professional* body organisation.
+\textcolor{blue}{We have clarified that the WSL is the body for professional surfers}
 
@@
 
-{\bf intent: } We tested for Brazilians NOT liking reef breaking waves
-and liking point and beach breaking waves.  consistent with 6.1
+\textcolor{blue}{We have clarified the issue and corrected an
+  unfortunate typographical error.  The manuscript reads: ``one
+  suggestion (Burgess 2020, Ho 2021) is that Brazilian surfers tend to
+  have skillsets that favour point and beach break wave types, readily
+  accessible to these surfers' native Brazil; see (Scarfe 2003) for a
+  scientific overview of surf break characteristics, and (Butt 2004)
+  for a surfer's perspective''.  In short, we suggest that Brazilians
+  {\em do not} like reef breaking waves.}
 
@@
 
-{\bf intent}  dealt with above
-
 Not exactly. What about ISA (world surfing games, Olympics, national
@@
 
-{\bf intent}  ditto
+\textcolor{blue}{We have clarified that the WSL is the body for professional surfers}
 
@@
 Based on what hypothesis?
+
+\textcolor{blue}{Discussed above}
 

### 2022-03-19 inst/surfing_jsa.tex
refree: WSL is for professional surfers
referee: brazilians do not like reef breaking waves

@@
 particular wave.  The World Surf League (WSL) is the main governing
-body of surfing competitions \citep{wsl}.  The WSL conducts a world
+body for professional surfers \citep{wsl}.  The WSL conducts a world
 tour in which the best ranked 34 surfers compete; in addition, at each
@@
 first-move advantage in chess~\citep{hankin2020}, and here we apply it
-to assess the supposed Brazilian preference for reef-breaking waves.
+to assess the supposed Brazilian preference for particular wave types.
 

### 2022-03-19 inst/surfing_jsa_rebuttal.tex
referee: points system is arbitrary

@@
 
-the points system itself is intrinsically arbitrary: the details of
-the points system does not affect the competitors' behaviour but can
-change the overall ranking;
-
-How could you say this, as depending on finishing spot at each
+Responding to ``the points system itself is intrinsically arbitrary:
+the details of the points system does not affect the competitors'
+behaviour but can change the overall ranking\ldots'' the referee asks:
+``How could you say this, as depending on finishing spot at each
 competition, the athlete is awarded a set number of points. So a
@@
 points. Impacting upon your placing in the rankings for the subsequent
-competition. And therefore, your seeding for the next competition.
+competition. And therefore, your seeding for the next competition.''
+
+\textcolor{blue}{We agree that the original wording was confusing.
+  Our point was that competitors are motivated to win, and if the
+  points system awards (strictly) more points for higher ranking, then
+  the competitors' behaviour is effectively driven by the ranking and
+  not the points awarded.  They will be motivated to come first,
+  regardless of how many points coming first is awarded (as long as
+  there are more points for coming first than second etc).  }
+
+\textcolor{blue}{
+  So, for example, consider two points systems: System A awards 10
+  points for a win, 7 for coming second and 1 for third; and system B
+  awards 3 for a win, 2 for second and 1 for third.  Competitors will
+  be equally motivated to be first, second, and third in the two
+  systems, for in each system they score more points for a higher
+  ranking.  But of course, if we have repeated trials (as in WSL) the
+  two systems could easily result in different points-based rankings.
+  We see a similar phenomenon in Formula 1 motor racing.}
+
+\textcolor{blue}{ After discussion we believe that the manuscript
+  would be improved by simply omitting the text under consideration.
+  It now reads: ``\ldots the points system itself is intrinsically
+  arbitrary: this would suggest that more formal statistical methods
+  are needed for analysis of such datasets\ldots''.  Our emphasis was
+  not on the disadvantages of the points systems used by WSL; we were
+  more interested in analysing the results formally.
+}
 

### 2022-03-19 inst/surfing_jsa.tex
referee: points system is arbitrary

@@
 suffers from the defect that the points system itself is intrinsically
-arbitrary: the details of the points system does not affect the
-competitors' behaviour but can change the overall ranking; this would
-suggest that more formal statistical methods are needed for analysis
-of such datasets~\citep{aldous2017}.  The Plackett-Luce likelihood
-functions are not subject to such a phenomenon and furnish an
-objective and coherent method for ranking competitors.  Further,
-likelihood-based methods offer one advantage over points-based
-systems: the ability to conduct statistically rigorous tests of
-meaningful nulls.
+arbitrary: this would suggest that more formal statistical methods are
+needed for analysis of such datasets~\citep{aldous2017}.  We suggest
+here that Plackett-Luce likelihood functions furnish an objective and
+coherent method for ranking competitors.  Further, likelihood-based
+methods offer one advantage over points-based systems: the ability to
+conduct statistically rigorous tests of meaningful nulls.
 

### 2022-03-21 inst/surfing_jsa_rebuttal.tex
more work on the rebuttal

@@
 
-{\bf intent:}  Will put some statistical comment in here.
+\textcolor{blue}{Competition is indeed performed within an ever
+  changing environment.  Each wave is indeed different from the
+  previous.  We believe that these facts are the reason for the
+  scoring system adopted in WSL tournaments.  We point out that the
+  environment during any individual heat is common to all competitors
+  in the water, so comparison between the competitors is fair, in the
+  sense that no competitor can justifiably claim that he was placed at
+  a disadvantage by wave quality.  All the competitors in the water
+  have the same opportunity to catch the same waves (subject to wave
+  priority rules, but these rules are fair in the same sense).  We
+  would contend that a competitor's ability to choose an opportune
+  wave is part of the skill of surfing, and is certainly necessary for
+  tournament success.}
+
+\textcolor{blue}{We use the word ``subjective" to mean ``personal
+  aesthetic valuation"; we wished to emphasise that pursuits such as
+  surfing are scored on judgements, rather than objective physical
+  measures such as timings or heights as seen in, for example, 100m
+  sprinting or archery.}
 
@@
 
-Your hypothesis on Brazilians preference for reef breaking waves is based on what?
+Your hypothesis on Brazilians preference for reef breaking waves is
+based on what?
 
@@
 
-\subsection*{Points system afffecting behaviour}
+\subsection*{Points system affecting behaviour}
 
@@
 
-{\bf intent: } need to rephrase
+\textcolor{blue}{Discussed below under ``points system"}
 
@@
 \textcolor{blue}{Discussed above}
+
+\subsection*{Points system}
 

### 2022-03-21 inst/surfing_jsa_rebuttal.tex
further work

@@
 Short story: we have accommodated all the comments with rewording and
-clarification.  The resulting document is, I believe, stronger and
-more scholarly than before and I recommend it to you.
+clarification.  The resulting document is, we believe, stronger and
+more scholarly than before and we recommend it to you.
 
@@
 \textcolor{blue}{See the table of results at the end of this document.
-  We see that the wildcards participate in very few venues: of the 25
-  wildcards, 20 appear at only a single venue, four at two venues, and
-  only one (Willcox) at three venues.  Further, the wildcards perform
-  poorly: of the 31 ranked scores, only 5 are better than last or
-  second to last.  Maximum likelihood Bradley-Terry strengths for such
-  competitors are either exactly zero or very small (here $\sim
-  10^{-5}$) and are neither interesting from a surfing lore
-  perspective, nor susceptible to meaningful statistical
-  inference.\newline Quite apart from the fact that our focus was on
-  the top-performing competitors, including such datapoints is
-  problematic for several reasons.  Firstly, numerical optimization
-  techniques tend to struggle as one increases the number of players:
-  one is necessarily working with a higher-dimensional problem,
-  leading to poorly defined maxima [the Jacobian becomes very small];
-  in addition, local maxima become more numerous and difficult to
-  identify.\newline We have added a brief summary of this to the
-  manuscript.}
+  We see that in general, a wildcard competitor participates in very
+  few venues: of the 25 wildcards, 20 appear at only a single venue,
+  four at two venues, and only one (Willcox) at three venues.
+  Further, the wildcards perform poorly: of the 31 ranked scores, only
+  5 are better than last or second to last.  Maximum likelihood
+  Bradley-Terry strengths for such competitors are either exactly zero
+  or very small (here $\sim 10^{-5}$) and are neither interesting from
+  a surfing lore perspective, nor susceptible to meaningful
+  statistical inference.\newline Quite apart from the fact that our
+  focus was on the top-performing competitors, including such
+  datapoints is problematic for several reasons.  Firstly, numerical
+  optimization techniques tend to struggle as one increases the number
+  of players: one is necessarily working with a higher-dimensional
+  problem, leading to poorly defined maxima [the Jacobian becomes very
+    small]; in addition, local maxima become more numerous and
+  difficult to identify.\newline We have added a brief summary of this
+  to the manuscript.}
 
@@
   surfing are scored on judgements, rather than objective physical
-  measures such as timings or heights as seen in, for example, 100m
+  measures such as timings or distances as seen in, for example, 100m
   sprinting or archery.}
@@
 \textcolor{blue}{We have clarified the issue and corrected an
-  unfortunate typographical error.  The manuscript reads: ``one
+  unfortunate typographical error.  The new manuscript reads: ``one
   suggestion (Burgess 2020, Ho 2021) is that Brazilian surfers tend to
@@
 
-The World Surf League (WSL) is the main governing
+{\bf manuscript:} The World Surf League (WSL) is the main governing
   body of surfing competitions (World Surf League 2021).
 
-Not exactly. What about ISA (world surfing games, Olympics, national
+{\bf comment:} Not exactly. What about ISA (world surfing games, Olympics, national
 championships, regional championships).
 
-\textcolor{blue}{We have clarified that the WSL is the body for professional surfers}
+\textcolor{blue}{{\bf response:} We have clarified that the WSL is the body for professional surfers}
 
-The WSL conducts a
+\rule{0mm}{10mm}
+
+{\bf manuscript:} The WSL conducts a
 world tour in which the best ranked 34 surfers compete;
 
-This only constitutes for male competition
+{\bf comment:} This only constitutes for male competition
 
-\textcolor{blue}{We have clarified that we are studying the men's
+\textcolor{blue}{{\bf response:} We have clarified that we are studying the men's
   championship}
 
-in addition, at each tour venue, two "wildcard" surfers also enter the
+\rule{0mm}{10mm}
+
+{\bf manuscript: } in addition, at each tour venue, two "wildcard" surfers also enter the
 competition, who are ignored here
 
-Why are they ignored? Especially as some of the wildcards competed in multiple events.
+{\bf comment: } Why are they ignored? Especially as some of the wildcards competed in multiple events.
 
-\textcolor{blue}{Discussed above}
+\textcolor{blue}{{\bf response: } Discussed above}
 
-Up to four surfers--the competitors--are in the water simultaneously,
+\rule{0mm}{10mm}
+
+{\bf manuscript: } Up to four surfers--the competitors--are in the water simultaneously,
 watched by up to five judges.
 
-You stated one or two other competitors above. So which is it?
+{\bf comment: } You stated one or two other competitors above. So which is it?
 
-\textcolor{blue}{Discussed above}
+\textcolor{blue}{{\bf response: } Discussed above}
 
-However, for a particular heat, the order statistic--that is, which
+\rule{0mm}{10mm}
+
+{\bf manuscript: } However, for a particular heat, the order statistic--that is, which
 competitor scored most highly, second highest, and third--is
@@
 
-But judging is based on the heats conditions, as surfing is performed
+{\bf  comment: } But judging is based on the heats conditions, as surfing is performed
 within an ever changing environment. Therefore, to try and compare
@@
 
-and here we apply it to assess the supposed Brazilian preference for
+\textcolor{blue}{{\bf response: } Discussed above}
+
+\rule{0mm}{10mm}
+
+{\bf manuscript: }  and here we apply it to assess the supposed Brazilian preference for
 reef-breaking waves.
 
-Based on what hypothesis?
+{\bf comment: } Based on what hypothesis?
 
-\textcolor{blue}{Discussed above}
+\textcolor{blue}{{\bf response: } Discussed above}
 

### 2022-03-21 inst/surfing_jsa.tex
further work

@@
 participation and viewing \citep{warshaw2010}.  Competitive surfing
-involves surfers competing against one or multiple other surfers in heats
+involves surfers competing against one or more other surfers in heats
 lasting 30-50 minutes \citep{booth1995}.  Points are awarded on a
@@
 tour venue, two ``wildcard" surfers also enter the competition, who
-are ignored here due to their generally poor performance and low
-participation.
+are ignored here due to their generally poor performance.
 

### 2022-03-22 inst/rowing.Rmd
slight rewording

@@
 5$ milliseconds apart, corresponding to a lead of $\sim
-2.5\,\mathrm{cm}$.  The dataset is as follows;
+2.5\,\mathrm{cm}$.  The package dataset is file `rowing.txt`, here are
+the first three lines:
 
 ```{r}
-show <- function(filename){
-  unlist(lapply(strsplit(readLines(filename)," "),paste,collapse=" "))
-}					       
-show("rowing.txt")
+readLines("rowing.txt")[1:3]
+```
+
+We can process this file using standard R functions.
+
+```{r}
+o <- sapply(readLines("rowing.txt"),strsplit," ")
+names(o) <- NULL  # names are pointless here
+o[1:3]
 ```
@@
 ```{r definerowing, cache=TRUE}
-
-f <- function(filename){
-  o <- strsplit(readLines(filename)," ")
-  H <- hyper2()
-  for(v in o){H <- H + race(v)}
-  return(H)
-}
-
-rowing <- f("rowing.txt")
-pnames(rowing)
-head(rowing)
+rowing <- Reduce("+",sapply(o,race))
+summary(rowing)
 ```
@@
 ```{r rowingminimaldef}
-rowing_minimal <- f("rowing_minimal.txt")
-head(rowing_minimal)
-pnames(rowing_minimal)
+o <- sapply(readLines("rowing_minimal.txt"),strsplit," ")
+names(o) <- NULL  
+rowing_minimal  <- Reduce("+",sapply(o,race))
+summary(rowing_minimal)
 ```

### 2022-03-30 inst/surfing_jsa.tex
tiny typos

@@
 Bradley-Terry likelihoods to assess a number of hypotheses of interest
-to the surfing community.  Specifically, we quantify the dominance of
+to the surfing community.  We quantify the dominance of
 the top competitors using likelihood techniques, and go on to study

### 2022-03-30 inst/surfing_jsa_rebuttal.tex
tiny typos
anonymise rebuttal

@@
 
-\section*{MS596 ``Analysis of competitive surfing tournaments with generalized Bradley-Terry likelihoods" by Driver and Hankin: rebuttal}
+\section*{MS596 ``Analysis of competitive surfing tournaments with generalized Bradley-Terry likelihoods": rebuttal}
 
-Below, the reviewer's comments are in black, and our replies to the
+Below, we reproduce the two reviewers' comments; our replies to the
 issues are in \textcolor{blue}{blue}.  We have indicated changes to

### 2022-04-11 inst/grad_weight.Rmd
gradient for hyper3 Rmd file

---
title: "grad_weight"
author: "Robin KS Hankin"
date: "10/04/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("disordR")
```

## Differentiate with respect to weights for a `hyper3` object

We will start with a simple example.  Consider two players $p_1$ and
$p_2$ who play tennis say, $a+b$ times with $p_1$ winning $a$ times
and $p_2$ winning $b$ times.  We use the following likelihood function:

$$
\frac{(wp_1)^ap_2^b}{(wp_1+p_2)^{a+b}}
$$

where $w$ represents some form of competitive advantage present in the
$a+b$ trials but perhaps absent in subsequent trials [this might be
home ground advantage in soccer or the first-mover advantage in
chess].  We may represent the log-likelihood in `hyper` idiom as
follows:

```{r,label=definefw}
f <- function(w,a=5,b=5){
    H <- hyper3()
    H[c(p1=w)] <- a
    H[c(p2=1)] <- b
    H[c(p2=1,p1=w)] <- -(a+b)
    H
}
```

Thus

```{r,label=showfinuse}
f(w=1.1)
```

Here I consider how we may differentiate this support function with respect to $w$:

```{r,label=definegws}
grad_weight_single <- function(H, probs, wpv, wanted){
    jj <- H[wpv]
    alpha <- drop(powers(jj))
    powers(jj) <- 1
print(wanted)
    alpha*sum(probs[names(probs) %in% wanted])/loglik(probs,jj,log=FALSE)
}
```

Function `grad_weight_single()` takes a `hyper3` object `H`, a vector
of non-negative probabilities `p` with unit sum, a named vector `wpv`
corresponding to one of the brackets of `H`, and a character vector
`wanted` that specifies which term(s) in bracket `wpv` have a weight
that is to be differentiated with respect to.  If the weight appears
in multiple brackets we need to call `grad_weight_single()` once for
each bracket the weight appears in, and add the results.  For example:

```{r,label=testthederiv}
w <- 1.1
dw <- 1e-4
H <- f(w)
probs <- c(p1=0.35, p2=0.65)

wpv <- c(p1=w)
wanted <- "p1"
d1 <- grad_weight_single(H, probs, wpv, wanted)

wpv <- c(p2=1,p1=w)
wanted <- "p1"
d2 <- grad_weight_single(H, probs, wpv, wanted)

print(c(d1,d2))

LHS <- (d1+d2)                                          # differential from grad_weight_single()
RHS <- (loglik(probs,f(w+dw)) - loglik(probs,f(w)))/dw  # differential via numerics

print(c(lhs=LHS,rhs=RHS,diff=RHS-LHS))
```

above we see very close agreement.

## A slightly more complicated example

Two chess players, 1, and 2, play chess.  Our observations (chess
games, say) comprise $a+b+c$ games in which $p_1$ played White,
winning $a$ games, drawing $b$ and losing $c$ [we write $+a=b-c$]; and
$d+e+f$ games in which $p_2$ played White, with $+d=e-f$.  Consider
the following likelihood function:

$$
\frac{(w p_1)^a\cdot(D(p_1+p_2))^b\cdot p_2^c}{(w p_1+D(p_1+p_2)+p_2)^{a+b+c}}
\frac{(w p_2)^d\cdot(D(p_1+p_2))^e\cdot p_1^f}{(w p_2+D(p_1+p_2)+p_1)^{d+e+f}}
$$

```{r}
grad_weight_single

w <- 1.2
D <- 0.3

brackfunc <- function(w,D){
  list(
  b1 = c(p1=w),
  b2 = c(p1=D, p2=D),	
  b3 = c(p2=1),
  b4 = c(p1=w+D,p2=1+D),
  b5 = c(p2=w),
  b6 = c(p1=1),
  b7 = c(p1=1+D,p2=w+D)
)
}
brackfunc(w,D)

wanted <- list(NULL,c("p1","p2"),NULL,c("p1","p2"), NULL,NULL,c("p1","p2"))
wanted

`f2` <- function(w,D, a=3,b=12,c=1,d=5,e=6,f=6){
  H <- hyper3()	      
  brackets <- brackfunc(w,D)
  H[brackets[[1]]] %<>% inc(a)	
  H[brackets[[2]]] %<>% inc(b)	
  H[brackets[[3]]] %<>% inc(c)	
  H[brackets[[4]]] %<>% dec(a+b+c)
  H[brackets[[5]]] %<>% inc(d)	
  H[brackets[[2]]] %<>% inc(e)	 # sic, the draw brackets are the same
  H[brackets[[6]]] %<>% inc(f)	
  H[brackets[[7]]] %<>% dec(d+e+f)
  return(H)
}

H <- f2(w=w, D=D)
H
```

$$
\frac{(w p_1)^3\cdot(D(p_1+p_2))^{12}\cdot p_2^1}{(w p_1+D(p_1+p_2)+p_2)^{16}}
\frac{(w p_2)^5\cdot(D(p_1+p_2))^6\cdot p_1^6}{(w p_2+D(p_1+p_2)+p_1)^{17}}
$$

Differentiate with respect to `D`:

```{r}
diff <- 0
dw <- 1e-4
probs <- c(p1=0.35 , p2=0.65)
B <- brackfunc(w,D)
for(i in seq_along(wanted)){
                                   diff <- diff +  grad_weight_single(H, probs, wpv=B[[i]],wanted=wanted[[i]])
                         print(grad_weight_single(H, probs, wpv=B[[i]],wanted=wanted[[i]]))

}

LHS <- diff
RHS <-  (loglik(probs,f2(w=w+dw,D=D)) - loglik(probs,f2(w=w,D=D)))/dw  # differential via numerics
c(LHS,RHS,LHS-RHS)
```

### 2022-04-12 inst/grad_weight.Rmd
more examples (there is a bug somewhere)

@@
 
-## A slightly more complicated example
+# Slightly more complicated example
 
-Two chess players, 1, and 2, play chess.  Our observations (chess
+```{r}
+H <- hyper3()
+f <- function(w){
+  H <- hyper3()
+  H[c(p1=1 , p2=w , p3=w , p4=1)] <- 5
+  H
+}
+w <- 1.4
+dw <- 1e-6
+HH <- f(w)
+HH
+probs <- c(p1=0.1,p2=0.2,p3=0.3,p4=0.4)
+wpv <- c(p1=1 , p2=w , p3=w , p4=1)
+LHS <- grad_weight_single(HH, probs, wpv, c("p2","p3"))
+RHS <-  (loglik(probs,f(w=w+dw)) - loglik(probs,f(w=w)))/dw
+c(RHS,LHS,LHS-RHS)
+```
+
+## An example from chess
+
+Two chess players, 1, 2, and 3, play chess.  Our observations (chess
 games, say) comprise $a+b+c$ games in which $p_1$ played White,
 winning $a$ games, drawing $b$ and losing $c$ [we write $+a=b-c$]; and
-$d+e+f$ games in which $p_2$ played White, with $+d=e-f$.  Consider
-the following likelihood function:
+$d+e+f$ games in which $p_2$ played White, with $+d=e-f$.  There is
+also a single game between $p_3$ (white) who beat $p_1$ (black).
+Consider the following likelihood function:
 
@@
 \frac{(w p_2)^d\cdot(D(p_1+p_2))^e\cdot p_1^f}{(w p_2+D(p_1+p_2)+p_1)^{d+e+f}}
+\frac{ w p_3}{wp_3+p_1}
 $$
@@
   b6 = c(p1=1),
-  b7 = c(p1=1+D,p2=w+D)
+  b7 = c(p1=1+D,p2=w+D),
+  b8 = c(p3=w),
+  b9 = c(p3=w,p1=1)
 )
@@
 
-wanted <- list(NULL,c("p1","p2"),NULL,c("p1","p2"), NULL,NULL,c("p1","p2"))
+wanted <- list(NULL,c("p1","p2"),NULL,c("p1","p2"), NULL,NULL,c("p1","p2"),NULL,NULL)
 wanted
@@
   H[brackets[[7]]] %<>% dec(d+e+f)
+  H[brackets[[8]]] %<>% inc
+  H[brackets[[9]]] %<>% dec
   return(H)
@@
 $$
-\frac{(w p_1)^3\cdot(D(p_1+p_2))^{12}\cdot p_2^1}{(w p_1+D(p_1+p_2)+p_2)^{16}}
-\frac{(w p_2)^5\cdot(D(p_1+p_2))^6\cdot p_1^6}{(w p_2+D(p_1+p_2)+p_1)^{17}}
+\frac{(w p_1)^a\cdot(D(p_1+p_2))^b\cdot p_2^c}{(w p_1+D(p_1+p_2)+p_2)^{a+b+c}}
+\frac{(w p_2)^d\cdot(D(p_1+p_2))^e\cdot p_1^f}{(w p_2+D(p_1+p_2)+p_1)^{d+e+f}}
+\frac{ w p_3}{wp_3+p_1}
 $$
@@
 dw <- 1e-4
-probs <- c(p1=0.35 , p2=0.65)
+probs <- c(p1=0.15 , p2=0.45, p3=0.4)
 B <- brackfunc(w,D)

### 2022-04-12 inst/grad_weight.Rmd
bug fixed: differentiate WRT w, not D

@@
----
+<---
 title: "grad_weight"
@@
     powers(jj) <- 1
-print(wanted)
     alpha*sum(probs[names(probs) %in% wanted])/loglik(probs,jj,log=FALSE)
@@
 ```{r}
-grad_weight_single
 
@@
 }
-brackfunc(w,D)
 
-wanted <- list(NULL,c("p1","p2"),NULL,c("p1","p2"), NULL,NULL,c("p1","p2"),NULL,NULL)
+wanted <- list("p1",NULL,NULL,"p1","p2",NULL,"p2","p3","p3")
 wanted
@@
 
-Differentiate with respect to `D`:
+Differentiate with respect to `w`:
 
@@
    diff <- diff +  grad_weight_single(H, probs, wpv=B[[i]],wanted=wanted[[i]])
-                         print(grad_weight_single(H, probs, wpv=B[[i]],wanted=wanted[[i]]))
-
 }
 
-LHS <- diff
+LHS <- diff  # using grad_weight_single()
 RHS <- (loglik(probs,f2(w=w+dw,D=D)) - loglik(probs,f2(w=w,D=D)))/dw  # differential via numerics
@@
 ```
+
+We may even be a little clever  and use a central difference:
+
+```{r}
+RHS2 <- (loglik(probs,f2(w=w+dw/2,D=D)) - loglik(probs,f2(w=w-dw/2,D=D)))/dw  # differential via numerics
+c(LHS,RHS2,LHS-RHS2)
+```

### 2022-04-12 inst/grad_weight.Rmd
differentiate WRT D

@@
 wanted <- list("p1",NULL,NULL,"p1","p2",NULL,"p2","p3","p3")
-wanted
 
@@
 ```
+
+If we wish to differentiate with respect to `D` we need to change the `wanted` list:
+
+```{r}
+o <- c("p1","p2")
+n <- NULL
+wantedD <- list(n,o,n,o,n,n,o,n,n)
+
+diff <- 0
+dD <- 1e-5
+B <- brackfunc(w,D)
+for(i in seq_along(wanted)){
+   diff <- diff +  grad_weight_single(H, probs, wpv=B[[i]],wanted=wantedD[[i]])
+}
+
+LHS <- diff  # using grad_weight_single()
+RHS <- (loglik(probs,f2(w,D=D+dD)) - loglik(probs,f2(w,D)))/dD  # differential via numerics
+c(LHS,RHS,LHS-RHS)
+```

### 2022-04-16 inst/race3.Rmd
more work on order statistics in hyper3

---
title: "Races with hyper3 objects"
author: "Robin KS Hankin"
date: "10/04/2022"
output: html_document
---

```{r setup, include=FALSE}
set.seed(1)
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("disordR")
```

Consider a race in which there are six runners 1-6 but we happen to
know that three of the runners (1,2,3) are clones of strength $p_a$,
say, two of the runners (4,5) have strength $p_b$, and the final
runner (6) is of strength $p_c$.  The runners race and the finishing order is:

$$a\succ c\succ b\succ a\succ a \succ b$$

Thus the winner was $a$, second place for $b$, third for $a$, and so
on.  Alternatively we might say that $a$ came first, fourth, and
fifth; $b$ came third and sixth, and $c$ came second.  The
Plackett-Luce likelihood function for $p_a,p_b,p_c$ would be

$$
\frac{p_a}{3p_a+2p_b+p_c}\cdot
\frac{p_c}{2p_a+2p_b+p_c}\cdot
\frac{p_b}{2p_a+2p_b    }\cdot
\frac{p_a}{2p_a+ p_b    }\cdot
\frac{p_a}{ p_a+ p_b    }\cdot
\frac{p_b}{      p_b    },\qquad p_a+p_b+p_c=1
$$

The likelihood function is easily specified in package idiom:

```{r, label=definehyper3}
H <- ordervec2supp3(c("a","c","b","a","a","b"))
H
```

```{r, label=maximizeprobs, cache=TRUE}
mH  <-  maxp(H)
mH
```

We can test various hypotheses, for example that the competitors are of equal strength:

```{r testequality, cache=TRUE}
equalp.test(H)
```

showing that this small dataset does not allow one to say that the
competitors differ.  However, it is easy to generate synthetic order
statistics where one can reject the null of equality of strength:

```{r,biggerrace,cache=TRUE}
o <- rrace3(pn=c(a=20,b=11,c=7),ps=c(a=0.7,b=0.2,c=0.1))
o
H <- ordervec2supp3(o)
H
```

Above, we specify that there are 20 $a$'s, 11 $b$'s and 7 $c$'s with
strength $70\%$, $20\%$, and $10\%$ respectively.  Recall this means
that, in a head-to-head between $a$ and $b$ we would have $a$ winning
with probability $0.7/(0.7+0.2)=\frac{7}{9}$.

```{r,bigraceestimation,cache=TRUE}
mH <- maxp(H)
mH
pTRUE <- c(a=0.7,b=0.2,c=0.1)   # true values 
mH - pTRUE # residual
```

Above we see a reasonably close agreement between the true values and the estimate.

```{r,dothetest,cache=TRUE}
equalp.test(H)
```

Above we see strong evidence to reject the hypothesis that
$p_a=p_b=p_c=\frac{1}{3}$.  We may also test the hypothesis that the
strength vector is equal to its true value:

```{r,isittrue,cache=TRUE}
knownp.test(H,pTRUE)
```

And above we see that there is no evidence to reject the hypothesis
that the strengths are in fact the specified strengths in `pTRUE`.

# Some simple cases

The simplest nontrivial case is three competitors with strengths
$a,a,b$ and finishing ordr $a\succ b\succ a$.  The likelihood
function would be

$$
\frac{a}{2a+b}\cdot\frac{b}{a+b}
$$

In this case we know that $a+b=1$ so this is equal to
$\mathcal{L}=\mathcal{L}(a)=\frac{a(1-a)}{1+a}$.

$$
\frac{d\mathcal{L}}{da}=\frac{(1+a)(1-2a)-a(1-a)}{(1+a)^2}=
\frac{1-2a+a-2a^2 -a + a^2}{(1+a)^2}=
\frac{1-2a-a^2}{(1+a)^2}
$$

This will be zero at $\sqrt{2}\pm 1$; we also note that
$d^2\mathcal{L}/da^2=-4(1+a)^{-3}$, manifestly strictly negative for
$0\leq a\leq 1$: the root is a maximum.

```{r label=maxaba,cache=TRUE}
maxp(ordervec2supp3(c("a","b","a")))
```

above, we see close agreement with the theoretical value of
$(\sqrt{2}-1,1-\sqrt{2})\simeq (0.414,0.586)$

With four competitors we have five nontrivial order statistics which,
without loss of generality, are $(a,b,a,a)$ and $(a,a,b,a)$.
$(a,b,a,b)$, $(a,b,b,a)$, $(a,b,c,a)$. Taking $(a,b,a,b)$ as an
example we have

$$
\mathcal{L}(a)=
\frac{a}{2a+2b}\cdot
\frac{b}{ a+2b}\cdot
\frac{a}{ a+ b}\cdot
\frac{b}{    b}=\frac{a^2-a^3}{4-2a}
$$

the evaluate would be a root of $2a^2-7a+4$, or $(7-\sqrt{17})/4\simeq
0.72$:

```{r label=maxabab,cache=TRUE}
maxp(ordervec2supp3(c("a","b","a","b")))
```

If we allow non-finishers there is another nontrivial order statistic,
viz $\left(a,b,\left\lbrace a,b\right\rbrace\right)$:

# The constructors championship

Here we define  likelihood function for the constructors' championship.

```{r,dotheconst}
jj <- read.table("constructor_2021.txt",header=TRUE)
constructor_table_2021 <- jj[,-ncol(jj)]
constructor_table_2021[,1:10]
```

```{r,maketheconst,cache=TRUE}
constructor_2021 <- ordertable2supp3(constructor_table_2021)
```

Then

```{r,label=maxconstr,cache=TRUE}
mc <- maxp(constructor_2021)
mc
```

```{r}
pie(mc)
```

### 2022-04-16 inst/race3.Rmd
more work on constructors

@@
 ```
+
+(takes about 15 minutes to process)
 
@@
 
-# The constructors championship
+$$
+\mathcal{L}(a)=
+\frac{a}{2a+2b}\cdot
+\frac{b}{ a+2b}\propto\frac{a(1-a)}{2-a}
+$$
+
+(see how the likelihood function is actually simpler than for the
+complete order statistic).  The evaluate would be $2-\sqrt{2}\simeq
+0.586$:
+
+```{r label=maxnonfinishers,cache=TRUE}
+o <- ordervec2supp3(c("a","b"),nonfinishers=c("a","b"))
+o
+maxp(o)
+```
+
+# The constructors' championship
 
@@
 ```
+
+Looking at the first column we see that the constructors are Mercedes
+("`Merc`"), Red Bull Racing Honda ("`RBRH`"), and so on.  Each
+constructor fields two cars in each race.  Looking at the next column,
+we see the order statistic for Bahrain (`BHR`).  We see that Mercedes
+came first and third, Red Bull came second and fifth, Ferrari sixth
+and eighth, and so on.
 
@@
 
-```{r}
+```{r,label=drawaconstructorpie}
 pie(mc)
 ```
+
+We may test the null of equal strengths:
+
+```{r,label=testequalityconstructor,cache=TRUE}
+equalp.test(constructor_2021)
+```
+
+and see that there is (very) strong evidence to reject this null.
+Further, we may test the hypothesis that the big four (viz Mercedies,
+Ferrari, Red Bull Racing Honda, and McLaren-Mercedes) are equal:
+
+```{r,label=testsameconstructor,cache=TRUE}
+samep.test(constructor_2021,c("Merc","Ferrari","RBRH","MM")) # ~10 mins
+```
+
+and fail to reject that null.

### 2022-04-17 inst/race3.Rmd
pnames<-.hyper3()

@@
 
+Now consider
+
+$$a\succ b\succ c\succ a$$
+
+with likelihood function
+
+$$
+\mathcal{L} =
+\frac{a}{2a+b+c}\cdot
+\frac{b}{ a+b+c}\cdot
+\frac{c}{ a  +c},\qquad a+b+c=1$$
+
+It can be shown that this has a maximum if $a^3+4a^2+4a-1=0$.
+Although cubic equations have analytical roots, the expression is
+complicated, and equal to about `c(a=0.2056,b=0.5466,c=0.2478)`:
+
+```{r,label=abca,cache=TRUE}
+maxp(ordervec2supp3(c("a","b","c","a")))
+```
+
 # The constructors' championship
@@
 jj <- read.table("constructor_2021.txt",header=TRUE)
+i <- seq(from=1,by=2,to=nrow(jj)) # extract every second entry for points
+constructor_points_2021 <- jj[i,ncol(jj)]
+names(constructor_points_2021) <- jj[i,1]
+
+constructor_points_2021 <- sort(constructor_points_2021,dec=TRUE)
+constructor_points_2021
+
 constructor_table_2021 <- jj[,-ncol(jj)]
@@
 constructor_2021 <- ordertable2supp3(constructor_table_2021)
+pnames(constructor_2021) <- names(constructor_points_2021)
 ```
@@
 ```{r,label=maxconstr,cache=TRUE}
-mc <- maxp(constructor_2021)
-mc
+constructor_2021_maxp <-  maxp(constructor_2021)
+mc <- constructor_2021_maxp  # saves typing
 ```
@@
 pie(mc)
+dotchart(mc,pch=16)
+ordertransplot(rank(-mc),rank(-constructor_points_2021),xlab="likelihood rank",ylab="points rank")
 ```
@@
 
-and fail to reject that null.
+and fail to reject that null.  How about the mid-range constructors:
+
+```{r,label=testsameconstructorlowscores,cache=TRUE}
+samep.test(constructor_2021,c("ATH","ARRF","AR","AMM")) # ~ 10 mins
+```
+
+### Package dataset {-}
+
+Following lines create `constructor.rda`, residing in the `data/`
+directory of the package.
+
+```{r,label=saveconstructordataset}
+save(constructor_2021_maxp,constructor_2021,constructor_table_2021,file="constructor.rda")
+```

### 2022-04-17 R/hyper3.R

@@ -331,7 +331,7 @@ char2nv <- function(x){
        
 `rhyper3` <- function(n=5,s=3,type='race',...){
     switch(type,
-           race = rracehyer3(n=n,size=s,...),
+           race = rracehyper3(n=n,size=s,...),
            pair = rpair3(n=n,s=s,...)
            )
     }
@@ -343,7 +343,6 @@ stop("not yet written")
 
 
 }
-
 
 
 `num3` <- function(v){  # numerators
@@ -409,7 +408,15 @@ stop("not yet written")
     }
     return(out)
 }
-    
+
+`pnames<-.hyper3` <- function(x,value){
+    hyper3(
+        elements(brackets(x)),
+        elements(weights(x)),
+        elements(powers(x)),
+        pnames=value)
+}
+
 ## a <- read.table("constructor_2021.txt",header=TRUE)
 ## constructor(a[,-ncol(a)])  # final column is points
 

### 2022-04-17 inst/race3.Rmd
constructor's championship 2021

@@
 constructor_table_2021[,1:10]
+constructor_2021_table <- constructor_table_2021
 ```
@@
 constructor_2021 <- ordertable2supp3(constructor_table_2021)
-pnames(constructor_2021) <- names(constructor_points_2021)
 ```
@@
 ```{r,label=saveconstructordataset}
-save(constructor_2021_maxp,constructor_2021,constructor_table_2021,file="constructor.rda")
+save(
+constructor_2021_maxp,
+constructor_2021,
+constructor_2021_table,
+file="constructor.rda")
 ```

### 2022-04-17 inst/javelin.Rmd
very rough work on javelins

---
title: "Javelins with hyper3 objects"
author: "Robin KS Hankin"
date: "10/04/2022"
output: html_document
---

```{r setup, include=FALSE}
set.seed(1)
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("disordR")
```

`https://en.wikipedia.org/w/index.php?title=Athletics_at_the_2020_Summer_Olympics_%E2%80%93_Men%27s_javelin_throw&oldid=1051014985`

```{r}
a <- read.table("javelin.txt",header=TRUE)
a
a <- t(a)
a
o <- NULL
for(i in seq_len(ncol(a))){
  jj <- as.numeric(a[,i])
  names(jj) <- rep(colnames(a)[i],length(jj))
  o <- c(o,jj)
}
o <-sort(o,na.last=TRUE,decreasing=TRUE)
o
names(o)
plot(o)
```

Now process it:

```{r,label=dothething1,cache=TRUE}    	   
H1 <- ordervec2supp3(v=names(o)[!is.na(o)],nonfinishers=names(o)[ is.na(o)])
m1 <- maxp(H1)
m1
```

```{r,label=dothething2,cache=TRUE}    	   
H2 <- ordervec2supp3(v=names(o)[!is.na(o)])
m2 <- maxp(H2)
m2
```

```{r,plotm}
pie(m1)
pie(m2)
dotchart(m1,pch=16)
dotchart(m2,pch=16)
par(pty="s")
plot(m1,m2,pch=16,type="p",asp=1)
abline(0,1)
```

### 2022-04-17 inst/javelin.Rmd
tidyup and vectorization

@@
 a
-a <- t(a)
-a
-o <- NULL
-for(i in seq_len(ncol(a))){
-  jj <- as.numeric(a[,i])
-  names(jj) <- rep(colnames(a)[i],length(jj))
-  o <- c(o,jj)
-}
-o <-sort(o,na.last=TRUE,decreasing=TRUE)
+
+o <- c(matrix(suppressWarnings(as.numeric(c(a,recursive=TRUE))),nrow(a),ncol(a)))
+names(o) <- c(matrix(rownames(a),nrow(a),ncol(a)))
+o <- sort(o,na.last=TRUE,decreasing=TRUE)
 o
 names(o)
-plot(o)
+n <- length(levels(as.factor(names(o))))
+plot(o,pch=16,col=rainbow(n))
+legend("topright",pch=16,col=rainbow(n),legend=levels(as.factor(names(o))))
 ```

### 2022-04-18 inst/javelin.Rmd
consistentification of attemptstable2supp3()

--- 2022-04-17 inst/javelin.Rmd
@@
 
-```{r}
+```{r,label=loadjavelinandconvert}
 a <- read.table("javelin.txt",header=TRUE)
 a
-a <- t(a)
-a
-o <- NULL
-for(i in seq_len(ncol(a))){
-  jj <- as.numeric(a[,i])
-  names(jj) <- rep(colnames(a)[i],length(jj))
-  o <- c(o,jj)
-}
-o <-sort(o,na.last=TRUE,decreasing=TRUE)
+o <- attemptstable2supp3(a,decreasing=TRUE,give.supp=FALSE)
 o
-names(o)
-plot(o)
+```
+
+Plot the data in visual form:
+
+```{r, label=plotjav}
+n <- length(levels(as.factor(names(o))))
+plot(o,pch=16,col=rainbow(n)[as.factor(names(o))])
+legend("topright",pch=16,col=rainbow(n),legend=levels(as.factor(names(o))))
 ```

### 2022-04-18 R/hyper3.R

@@ -435,3 +435,27 @@ stop("not yet written")
     }
     return(out)
 }
+
+
+wtest <- function(formula,data){
+    stopifnot(is.formula(formula))
+    
+
+
+}
+
+`attemptstable2supp3` <- function(a, decreasing, give.supp=TRUE, dnf.last=TRUE){
+    o <- c(matrix(suppressWarnings(as.numeric(c(a,recursive=TRUE))),nrow(a),ncol(a)))
+    names(o) <- c(matrix(rownames(a),nrow(a),ncol(a)))
+    o <- sort(o,na.last=TRUE,decreasing=TRUE)
+    if(give.supp){
+        if(dnf.last){
+            nf <- names(o)[is.na(o)]
+        } else {
+            nf <- NULL
+        }
+        return(ordervec2supp3(v=names(o)[!is.na(o)], nonfinishers=nf))
+    } else {
+        return(o)
+    }
+}

### 2022-04-18 inst/javelin.Rmd
javelin and attempts table work

@@
 ```{r,label=loadjavelinandconvert}
-a <- read.table("javelin.txt",header=TRUE)
-a
+javelin_table <- read.table("javelin.txt",header=TRUE)
+a <- javelin_table # saves typing
+
 o <- attemptstable2supp3(a,decreasing=TRUE,give.supp=FALSE)
-o
 ```
@@
 ```{r,label=dothething1,cache=TRUE}    	   
-H1 <- ordervec2supp3(v=names(o)[!is.na(o)],nonfinishers=names(o)[ is.na(o)])
-m1 <- maxp(H1)
-m1
+javelin <- attemptstable2supp3(a,decreasing=TRUE,give.supp=TRUE)
+H1 <- javelin # saves typing
+javelin1_maxp <- maxp(H1)
+m1 <- javelin1_maxp
 ```
@@
 H2 <- ordervec2supp3(v=names(o)[!is.na(o)])
-m2 <- maxp(H2)
-m2
+javelin2_maxp  <- maxp(H2)
+m2 <- javelin2_maxp
 ```
@@
 ```
+
+### Package dataset {-}
+
+Following lines create `javelin.rda`, residing in the `data/`
+directory of the package.
+
+```{r,label=savejavelindataset}
+save(javelin,javelin_table,javelin1_maxp,javelin2_maxp,  file="javelin.rda")
+```

### 2022-04-18 inst/javelin.Rmd
clarification

@@
 a <- javelin_table # saves typing
-
-o <- attemptstable2supp3(a,decreasing=TRUE,give.supp=FALSE)
+javelin_table
 ```
 
-Plot the data in visual form:
+Above, we see that Chopra threw 87.03m on his first throw, 87.58m on
+his second, and so on.  Vadlejch threw 83.98m on his first throw, etc.
+We may convert this to a named vector with elements being the throw
+distances, and names being the competitors, using
+`attemptstable2supp3()`:
+
+```{r,label=converttosupp3,cache=TRUE}
+javelin_vector <- attemptstable2supp3(a,decreasing=TRUE,give.supp=FALSE)
+javelin_vector
+o <- javelin_vector # saves typing
+```
+
+Above we see that Chopra threw the longest and second-longest throws
+of 87.58m and 87.03 respectively; Vadlejch threw the third-longest
+throw of 86.67m, and so on.  It is easy to plot the data in visual
+form:
 
@@
 
-Now process it:
+Now convert the attempts table to a `hyper3` object, again using
+function `attemptstable2supp3()` but this time pass `give.supp=TRUE`.
+However, note that there are two reasonable interpretations: firstly,
+we simply ignore no-throws [in the package docs, these are sometimes
+referred to as "DNF", or "Did not finish", following Formula 1
+terminology]:
 
-```{r,label=dothething1,cache=TRUE}    	   
-javelin <- attemptstable2supp3(a,decreasing=TRUE,give.supp=TRUE)
-H1 <- javelin # saves typing
+```{r,label=dothething2,cache=TRUE}    	   
+javelin1 <- ordervec2supp3(v=names(o)[!is.na(o)])
+H1 <- javelin1 # saves typing
 javelin1_maxp  <- maxp(H1)
@@
 
-```{r,label=dothething2,cache=TRUE}    	   
-H2 <- ordervec2supp3(v=names(o)[!is.na(o)])
+Secondly, we may treat no-throws as being inferior to any actual
+measured throw.  To do this it is easiest to pass `give.supp=TRUE` to
+`attemptstable2supp3()`:
+
+```{r,label=dothething1,cache=TRUE}    	   
+javelin2 <- attemptstable2supp3(a,decreasing=TRUE,give.supp=TRUE)
+H2 <- javelin2 # saves typing
 javelin2_maxp <- maxp(H2)
@@
 
+Inveestigation:
+
 ```{r,plotm}
+rbind(one=m1,two=m2)
 pie(m1)

### 2022-04-19 inst/javelin.Rmd
slight changes

@@
 
-`https://en.wikipedia.org/w/index.php?title=Athletics_at_the_2020_Summer_Olympics_%E2%80%93_Men%27s_javelin_throw&oldid=1051014985`
+We analyse results from the men's javelin, 2020 Summer Olympics.
 
@@
 
-Inveestigation:
+Investigation:
 
@@
 
+Reference
+
+`https://en.wikipedia.org/w/index.php?title=Athletics_at_the_2020_Summer_Olympics_%E2%80%93_Men%27s_javelin_throw&oldid=1051014985`
+
 ### Package dataset {-}
@@
 ```{r,label=savejavelindataset}
-save(javelin,javelin_table,javelin1_maxp,javelin2_maxp,  file="javelin.rda")
+save(javelin1,javelin2,javelin_table,javelin_vector,javelin1_maxp,javelin2_maxp,  file="javelin.rda")
 ```

### 2022-04-19 inst/javelin.Rmd
slight improvements

@@
 of 87.58m and 87.03 respectively; Vadlejch threw the third-longest
-throw of 86.67m, and so on.  It is easy to plot the data in visual
-form:
+throw of 86.67m, and so on.  `NA` entries correspond to no-throws.  It
+is easy to plot the data in visual form:
 
@@
 dotchart(m2,pch=16)
-par(pty="s")
-plot(m1,m2,pch=16,type="p",asp=1)
 abline(0,1)
+ordertransplot(m1,m2)
 ```

### 2022-04-20 inst/home_advantage.Rmd
use sticker in Rmd
label each chunk, and use cache

@@
 
+<p style="text-align: right;">
+![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
+</p>
+
+(takes about ten minutes to run without cache)
+
 Here I analyse a small synthetic dataset for home ground advantage.
@@
 First load the package:
-```{r}
+```{r,label=libload}
 library(hyper2)
@@
 
-```{r}
+```{r,label=genthedata}
 home_games_won <- matrix(c(
@@
 
-```{r}
+```{r,label=makehyper2}
 H <- hyper2(pnames=c(teams,"home"))
@@
 
-```{r}
+```{r,label=includehome}
 for(i in seq_len(nrow(home_games_won))){
@@
 
-```{r}
+```{r,label=awaygameswon}
 for(i in seq_len(nrow(away_games_won))){
@@
 
-```{r}
+```{r,label=showH}
 H
@@
 
-```{r}
+```{r,label=showvisH}
 dotchart(maxp(H))
@@
 
-```{r}
+```{r,label=nullofzerohomestrength}
 f <- function(home){
@@
 
-```{r}
+```{r,label=showLambda,cache=TRUE}
 Lambda <- -2*(f(0.01)-maxp(H,give=TRUE)$value)
@@
 
-```{r}
+```{r,calcprofilelikelihood}
 g <- function(home){
@@
 
-```{r}
+```{r,showprofilelikelihood,cache=TRUE}  # this one takes the time
 home_strength <- seq(from=0.2,to=0.6,length=30)

### 2022-04-20 inst/home_advantage.Rmd
update home advantage work

@@
 
-```{r,label=makehyper2}
-H <- hyper2(pnames=c(teams,"home"))
+```{r,label=makehyper2,cache=TRUE}
+H <- home_away(home_games_won,away_games_won)
+H
+options("use_alabama" = TRUE) # to avoid the stupid wmmin bug
+specificp.gt.test(H,"home",0)
 ```
 
-First include the home games won:
-
-```{r,label=includehome}
-for(i in seq_len(nrow(home_games_won))){
-  for(j in seq_len(ncol(home_games_won))){
-    if(i != j){  # diagonal means a team plays itself, meaningless
-      game_winner <- teams[i]
-      game_loser  <- teams[j]
-      no_of_matches <- home_games_won[i,j] # won with home strength helping
-      H[c(game_winner,           "home")] %<>% inc(no_of_matches)  # won with home strength helping
-      H[c(game_winner,game_loser,"home")] %<>% dec(no_of_matches)
-    }
-  }
-}
-```
-
-Now away games won:
-
-```{r,label=awaygameswon}
-for(i in seq_len(nrow(away_games_won))){
-  for(j in seq_len(ncol(away_games_won))){
-    if(i != j){
-      game_winner <- teams[i]
-      game_loser  <- teams[j]
-      no_of_matches <- away_games_won[i,j] # won without the benefit of home strength
-      H[c(game_winner                  )] %<>% inc(no_of_matches)  # won without home ground helping...
-      H[c(game_winner,game_loser,"home")] %<>% dec(no_of_matches)  # home strength nevertheless on denominator
-    }
-  }
-}
-```
-
-Thus
-
-```{r,label=showH}
-H
-```
-
+We see strong evidence to support the contention that home advantage is real.
 Visually:
@@
 
-Now a likelihood ratio for the hypothesis that home strength (or home
-advantage) is zero.  First define a function `f()` that maximises the
-likelihood subject to the home strength being *less than* its
-argument:
+A profile likelihood diagram:
 
-```{r,label=nullofzerohomestrength}
-f <- function(home){
-	small <- 1e-3
-	maxp(H,startp=rep((1-small)/4,4),give=TRUE,fcm=t(rep(1,4)),fcv=1-home)$value
-}
-```
-
-See how the `fcm` and `fcv` constraints make the sum of a,b,c,d,
-greater than `1-home`.  Thus 'home' [which is the fillup value,
-$p_\mathrm{home}=1-(p_A+p_B+p_C+p_D)$] is constrained to be less than
-argument `home`.  Using this function, the likelihood ratio test is
-straightforward:
-
-```{r,label=showLambda,cache=TRUE}
-Lambda <- -2*(f(0.01)-maxp(H,give=TRUE)$value)
-Lambda
-```
-
-Under the null, Wilks's theorem gives us that Lambda is distributed as
-a chi-square with 1 degree of freedom (clearly rejected).  However, it
-is always interesting to plot likelihood functions.  To plot a profile
-likelihood we can constrain the home to be greater than a particular
-value:
-
-```{r,calcprofilelikelihood}
-g <- function(home){
-	small <- 1e-3
-	maxp(H,startp=rep(small/4,4),give=TRUE,fcm=-t(rep(1,4)),fcv=home-1)$value
-}
-```
-
-```{r,showprofilelikelihood,cache=TRUE}  # this one takes the time
-home_strength <- seq(from=0.2,to=0.6,length=30)
-support <- pmin(sapply(home_strength,f),sapply(home_strength,g))
-plot(home_strength,support-max(support),
+```{r,label=showprofilelikelihood, cache=TRUE}
+home_strength <- seq(from=0.2,to=0.6,length=20)
+plot(home_strength,profsupp(H,"home",home_strength),type="b",pch=16,
      xlab="home strength",ylab="support",main="profile likelihood for home advantage")

### 2022-04-20 R/hyper2.R

@@ -1181,3 +1181,38 @@ setGeneric("pnames<-",function(x,value){standardGeneric("pnames<-")})
     dirichlet(rowSums(M)) - hyper2(apply(index,1,function(x){rownames(M)[x]},simplify=FALSE),jj[index])
 }
 
+`home_away` <- function(home_games_won,away_games_won){
+    if(is.complex(home_games_won)){
+        away_games_won <- Im(home_games_won)
+        home_games_won <- Re(home_games_won) # note, away_games_won is assigned first
+    }
+    teams <- rownames(home_games_won)
+    stopifnot(identical(teams,colnames(home_games_won)))
+    stopifnot(identical(teams,rownames(away_games_won)))
+    stopifnot(identical(teams,colnames(away_games_won)))
+    teams <- c(teams,"home")
+    
+    H <- hyper2(pnames=teams)
+
+    for(i in seq_len(nrow(home_games_won))){
+        for(j in seq_len(ncol(home_games_won))){
+            if(i != j){  # diagonal means a team plays itself, meaningless.
+                ## First deal with home_games:
+                game_winner <- teams[i]
+                game_loser  <- teams[j]
+                no_of_matches <- home_games_won[i,j] # won with home strength helping
+                H[c(game_winner,           "home")] %<>% inc(no_of_matches)  # won with home strength helping
+                H[c(game_winner,game_loser,"home")] %<>% dec(no_of_matches)
+
+                ## now away games:
+                no_of_matches <- away_games_won[i,j] # won without the benefit of home strength
+                H[c(game_winner                  )] %<>% inc(no_of_matches)  # won without home ground helping...
+                H[c(game_winner,game_loser,"home")] %<>% dec(no_of_matches)  # home strength nevertheless on denominator
+                                
+            } # if(i != j) closes
+        } # j loop closes
+    } # i loop closes
+    return(H)
+}
+
+

### 2022-04-21 inst/home_advantage.Rmd
more work

@@
 
-(takes about ten minutes to run without cache)
+(takes about fifteen minutes to run without cache)
 
-Here I analyse a small synthetic dataset for home ground advantage.
+Here I analyse a small synthetic dataset for home ground advantage, then go on to discuss
+a dataset presented by Agresti.
 
@@
 
-We see strong evidence to support the contention that home advantage is real.
-Visually:
+We see strong evidence to support the contention that home advantage
+is real.  Further, we may test the hypothesis that all the teams have
+the same strength, after accounting for the home team advantage:
+
+```{r,label=testhomegroundadvantageAgresti,cache=TRUE}
+samep.test(H,teams)
+```
+
+Above, we see no evidence for a difference in team strengths.  Visually:
 
@@
 ground advantage.
+
+# Example in Agresti
+
+Agresti considers a dataset of seven baseball teams who play one
+another repeatedly.  Each game is played at the home ground of one
+team, the other playing away.  The dataset is as follows:
+
+```{r,label=makeAgrestimatrix}
+baseball_table <- matrix(c(
+NA,   4+3i, 4+2i, 4+3i, 6+1i, 4+2i, 6+0i,
+3+3i, NA  , 4+2i, 4+3i, 6+0i, 6+1i, 4+3i,
+2+5i, 4+3i, NA  , 2+4i, 4+3i, 4+2i, 6+0i,
+3+3i, 5+1i, 2+5i, NA  , 4+3i, 4+2i, 6+1i,
+5+1i, 2+5i, 3+3i, 4+2i, NA  , 5+2i, 6+0i,
+2+5i, 3+3i, 3+4i, 4+3i, 4+2i, NA  , 2+4i,
+2+5i, 1+5i, 1+6i, 2+4i, 1+6i, 3+4i, NA),7,7)
+teams <- c("Miulwaukee", "Detroit", "Toronto", "New York", "Boston","Cleveland","Baltimore")
+rownames(baseball_table) <- teams
+colnames(baseball_table) <- teams
+M <- baseball_table # saves typing
+```
+
+Above, we represent home team wins with the real part of the matrix,
+the imaginary component being away wins.  Now process it:
+
+```{r,label=makesuppfrommatrix}
+baseball <- home_away(M)
+baseball
+```
+
+Maximum likelihood:
+
+```{r,label=calculatemaxpAgresti,cache=TRUE}
+baseball_maxp <- maxp(baseball)
+baseball_maxp
+```
+
+Visually:
+
+```{r,label=showmaxpAgresti}
+dotchart(baseball_maxp,pch=16)
+pie(baseball_maxp)
+```
+
+Above, we see that the teams are all of roughly equal strength, and
+the home advantage is small.  However, we may easily assess the null
+that the home strength is zero:
+
+```{r,analyseAgresti,cache=TRUE}
+specificp.gt.test(baseball,"home",0)
+```
+
+```{r,Agrestiprofsupp,cache=TRUE}
+home_strength <- seq(from=0.005,to=0.1,length=20)
+plot(home_strength,profsupp(baseball,"home",home_strength),type="b",pch=16,
+     xlab="home strength",ylab="support",main="profile likelihood for home advantage")
+abline(h=c(0,-2))
+```
+
+Further, we may test the hypothesis that the baseball teams have the same strength:
+
+```{r,label=testhomegroundadvantage,cache=TRUE}
+samep.test(baseball,teams)
+```
+
+Above, we see no evidence for the teams having differential strengths.
+
+### Some further thoughts on the home ground monster and Agresti's analysis.
+
+Considering the simplest case of two teams, a and b, with results:
+
+* $a$ at home, $b$ away, $a,b$ wins with probability $\frac{\lambda a,b}{\lambda a+b}$ (Agresti), or $\frac{a+H,b}{a+b+H}$ (me). 
+* $a$ away, $b$ at home, $a,b$ wins with probability $\frac{a,\lambda b}{\lambda a+b}$ (Agresti), or $\frac{a,b+H}{a+b+H}$ (me).
+
+Matching the probabilities for $a$ at home gives us $H=a(\lambda-1)$
+and matching the probabilities for $a$ away gives $H=b(\lambda-1)$.
+These cannot both be true unless $a=b$ or $\lambda=1$.  But we can, in principle,
+test which one is a better model.
+
+### Package dataset {-}
+
+Following lines create `home_advantage.rda`, residing in the `data/`
+directory of the package.
+
+```{r,label=savesolingdataset}
+save(baseball_table, baseball, baseball_maxp, 
+  file="baseball.rda")
+```

### 2022-04-21 inst/home_advantage.Rmd
typos

@@
 
-```{r,label=testhomegroundadvantageAgresti,cache=TRUE}
+```{r,label=homegroundAgresti,cache=TRUE}
 samep.test(H,teams)
@@
 2+5i, 1+5i, 1+6i, 2+4i, 1+6i, 3+4i, NA),7,7)
-teams <- c("Miulwaukee", "Detroit", "Toronto", "New York", "Boston","Cleveland","Baltimore")
+teams <- c("Milwaukee", "Detroit", "Toronto", "New York", "Boston","Cleveland","Baltimore")
 rownames(baseball_table) <- teams
@@
 
+As an interesting aside, suppose we have some additional data from
+games in which (for some reason), the home advantage was not
+operating.
+
+```{r,label=newinfotobaseball}
+M2 <- matrix(c(0,1,2,0),2,2)
+dimnames(M2) <- list(winner=teams[1:2],loser=teams[1:2])
+M2
+```
+
+Thus we have additional data from three games in which Milwaukee won
+twice and lost once.  This additional information may easily be
+incorporated into our likelihood function, at least on the assumption
+of independence:
+
+```{r,label=addsomenewinfotobaseball}
+baseball <- baseball + pairwise(M2)
+```
+
+We may assess the difference that the new information makes:
+
+```{r,label=baseballtransplot,cache=TRUE}
+ordertransplot(baseball_maxp[1:7],maxp(baseball)[1:7],xlab="old",ylab="new",plotlims=c(0.1,0.16))
+```
+
+Above, we see very little difference in the two evaluates, except for
+Detroit and Milwaukee, as expected.
+
 ### Some further thoughts on the home ground monster and Agresti's analysis.
@@
 and matching the probabilities for $a$ away gives $H=b(\lambda-1)$.
-These cannot both be true unless $a=b$ or $\lambda=1$.  But we can, in principle,
-test which one is a better model.
+These cannot both be true unless $a=b$ or $\lambda=1$.  But we can, in
+principle, test which one is a better model.
 

### 2022-04-21 inst/surfing_jsa_rebuttal_referee2.tex
start work on second referee's comments

\documentclass[12pt]{article}
\usepackage{xcolor}
\begin{document}

\section*{MS596 ``Analysis of competitive surfing tournaments
  with generalized Bradley-Terry likelihoods" by Driver and
  Hankin: second rebuttal}

Below, the reviewer's comments are in black, and our replies to the
issues are in \textcolor{blue}{blue}.  We have indicated changes to
the manuscript where appropriate.

Short story: we have accommodated all the comments with rewording and
clarification.  We have included an extended discussion of some of the
points below.  The resulting document is, we believe, stronger and
more scholarly than before and we recommend it to you.

\section*{Referee's comments}

The paper does a good job of outlining the basics of surfing in terms
of competition rules and structure, and frames up the problem at
hand. However, the paper itself feels rather light on content. The
authors jump almost immediately to the results section and don't go
into very much detail in terms of how or why Bradley-Terry models are
used, other than the fact that they are used. The authors are guilty
of referring to various numerical quantities throughout the paper,
without expanding on how these values were calculated, and what
implications the values have. The paper would benefit from more
explicit equations throughout; on several occasions the authors refer
to modifying or changing their likelihood function to suit their
problem at hand, but it is difficult for the reader (or at least
myself!) to intuitively understand what this modified function would
look like, without writing it down or thinking hard about
it. Throughout the paper I found I had to do a lot of the mathemati
cal and abstract thinking myself, which would have been made much
easier with more equations and explicit definitions and derivations of
parameters, likelihood functions, and important quantities that are
referred to in Sections 5, 6, and 7.

The figures and tables are poorly presented and would benefit from
significant rework in terms of formatting.

Section 7, while an interesting topic, feels like it falls more into
the realm of speculation than concrete analysis in the way it is
presented. The methodology is not very clear and it would be nice to
see a more explicit definition and derivation of pc

Overall, it feels like the main conclusion is "not all surfers have
the same skill". This is not insignificant, however, it is not
particularly surprising and is one that could have been reached using
a number of different methods. Had the authors done a more complete
job of fleshing out why the Bradley-Terry approach was better for
reaching this conclusion than others, I would be more satisfied, but
as it stands it falls a little flat.

From a style perspective, the paper is well-written and enjoyable to
read, which makes the lack of substantial content a bit of a shame, as
the authors are clearly able to write a paper worth reading. Overall,
it sounds like the authors do have the data and analysis available to
warrant a more complete paper, however, what they have submitted feels
rushed and rather rough around the edges. While possible, it would
take a fairly significant overhaul/expansion to the current version of
the paper to get it into a publication-ready format.

\end{document}

### 2022-04-22 inst/surfing_jsa_rebuttal_referee2.tex
work on issue #129

@@
 into very much detail in terms of how or why Bradley-Terry models are
-used, other than the fact that they are used. The authors are guilty
+used, other than the fact that they are used.
+
+\textcolor{blue}{We have re-read our manuscript and indeed
+  Bradley-Terry (BT) does make a rather abrupt entrance.  However, we
+  do describe and justify and motivate the use of BT in section three.
+  On reflection we believe we can address this comment by rearranging
+  sections 2 and 3: this gives a gentler and more appropriate
+  introduction to BT and indeed is a more appropriate structure for a
+  manuscript.  In line with this, we have changed the section heading
+  from ``previous related work'' to ``Previous statistical analysis of
+  competitive surfing'' which is a better description, we feel.}
+
+The authors are guilty
 of referring to various numerical quantities throughout the paper,

### 2022-04-22 inst/surfing_jsa.tex
work on issue #129

@@
 
-\subsection{Previous related work}
+\section{Previous statistical analysis of competitive surfing}
 
@@
 
-\section{Bradley-Terry}
+\section{Likelihood-based systems and points-based systems}
+
+Table~\ref{resultstable} shows the official results table from the
+tour.  We see that Ferreira places first with 59740 points, Medina
+second with 56475, and so on.  The overall ranking of the surfers is
+determined by summing the season's points.
+
+As in many subjective sports, points are awarded to competitors on the
+basis that a better performance attracts a higher number of points.
+However, such systems are difficult to analyse statistically, because
+the points awarded are intrinsically arbitrary: provided that better
+performances attract more points, the precise number of points awarded
+does not change the competitors' behaviour.  Competitors will
+endeavour to earn the maximum number of points possible, and to do
+this will deliver the best performance they are able to give.
+
+As discussed above, analysis of order statistics does not suffer from
+this particular defect~\citep{aldous2017}.  Order statistics can be
+analysed using a number of formal statistical methods, but here we
+focus on one method that has been successfully applied in many
+competitive sporting contexts: here we suggest that Plackett-Luce
+likelihood functions furnish an objective and coherent method for
+ranking competitors.  Further, likelihood-based methods offer one
+signal advantage over points-based systems: the ability to conduct
+statistically rigorous tests of meaningful nulls.
+
+\section{The Bradley-Terry class of probability models for order statistics}
 
@@
 to assess the supposed Brazilian preference for particular wave types.
-
-\section{Likelihood-based systems and points-based systems}
-
-Table~\ref{resultstable} shows the official results table from the
-tour.  We see that Ferreira places first with 59740 points, Medina
-second with 56475, and so on.  The overall ranking of the surfers is
-determined by summing the season's points.  However, this system
-suffers from the defect that the points system itself is intrinsically
-arbitrary: this would suggest that more formal statistical methods are
-needed for analysis of such datasets~\citep{aldous2017}.  We suggest
-here that Plackett-Luce likelihood functions furnish an objective and
-coherent method for ranking competitors.  Further, likelihood-based
-methods offer one advantage over points-based systems: the ability to
-conduct statistically rigorous tests of meaningful nulls.
 

### 2022-04-22 inst/surfing_jsa_rebuttal_referee2.tex
further work on issue #129

@@
 
+\textcolor{blue}{These comments are entirely reasonable and, on
+  re-reading the manuscript, we agree that we could have done a much
+  better job describing the mathematical and statistical basis for our
+  conclusions, especially the motivation for the likelihood equations.
+  In the resubmission we have added a large amount of explanation that
+  hopefully makes our reasoning more clear.\\ \\
+  However, we observe that {\em all} our calculations are available
+  for public inspection on gihub at our repo {\tt RobinHankin/hyper2}.
+  In particular, file {\tt inst/surfing.Rmd} can reproduce all our
+  findings.  However, the nitty-gritty detail of our R idiom is, we
+  suggest, not really suitable for a JSA manuscript.  We believe that
+  our enhanced reporting in the revision is a good compromise between
+  not enough detail and too much.
+  \begin{itemize}
+  \item We have provided more detail (mathematical and computational)
+    for the Fisher information argument for retaining the data with
+    three surfers, as opposed to retaining just the pairs
+  \item Added detail on the computational infrastructure for locating
+    maximum likelihood estimates
+  \item Referenced the computational method [{\tt equalp.test()} for
+    testing the null of equal competitive abilities
+  \item Introduced a new table that explicitly sets out the likelihood
+    function for the reified entity Brazilian beachtype reified entity
+  \end{itemize}
+}
+
 The figures and tables are poorly presented and would benefit from

### 2022-04-22 inst/surfing_jsa.tex
further work on issue #129

@@
 \usepackage{amssymb}
+\usepackage{amsmath}
 \usepackage{graphicx}
@@
 
-We can directly calculate the determinant of the observed information
-matrix for the likelihood function in
-equation~\ref{complete_likelihood_function} at about $4.3\times
-10^{71}$, as compared with~$2.9\times 10^{69}$ for the pairs
-considered in isolation, a ratio of about $146.4$.  This ratio shows
-that there is considerable information loss in considering only the
-paired comparisons.  We are thus motivated to consider the whole
-dataset, not just the part which involves paired comparisons.
+To answer this, we directly calculate the observed information matrix
+for the log-likelihood function in
+equation~\ref{complete_likelihood_function}.  Observing that
+differentiation is linear, we may calculate second derivatives by
+summing individual terms of the support function and further observing
+that, for a typical term of the form $\log\left(\sum_{k\in K}
+p_k\right)^n$ [where the sum is over some subset $K$ of strengths] we
+have
+
+\begin{equation}
+  \frac{\partial^2}{\partial p_i\partial p_j}
+  \log\left(\sum p_k\right)^n=
+  \begin{cases}
+    -n/\left(p_i+p_j+\sum p_k\right)^2\qquad\mbox{if $i,j\in K$}\\
+    0\qquad\mbox{otherwise.}
+    \end{cases}
+\end{equation}
+
+The {\tt hyper2} package includes function {\tt hessian()} which
+returns this matrix for any given support function.  For
+equation~\ref{complete_likelihood_function} we find that the
+determinant is about $4.3\times 10^{71}$.  Compare this with
+~$2.9\times 10^{69}$ for the pairs considered in isolation, a ratio of
+about $146.4$.  This ratio shows that there is considerable
+information loss in considering only the paired comparisons.  We are
+thus motivated to consider the whole dataset, not just the part which
+involves paired comparisons.
 
 \section{Results}
+
+For any likelihood function, one important diagnostic is the maximum
+likelihood estimate.  The {\tt hyper2} package facilitates the finding
+of maximum likelihood estimates using the bespoke {\tt maxp()}
+function, which uses standard optimization techniques.  Derivatives
+are available which ensures rapid convergence to the evaluate.
 
@@
 p_\mathrm{Wright}=\frac{1}{23}$, and any differences in placings are
-due to random variation.  The Method of Support \citep{edwards1992}
-rejects the null (in favour of the evaluate) at $p=4\times 10^{-5}$.
-There is thus strong evidence to suggest that the competitors are
-indeed of different skills and the result actually reflects genuine
-differences in abilities, rather than the result of random chance.
-Alternatively, we may view the dataset as being sufficiently
-informative to reveal actual differences in surfing ability.  Observe
-that the competitors' points (that is, the final column of
-table~\ref{resultstable}) cannot be used for this type of
-probabilistic analysis: the points awarded are arbitrary and changing
-the points system would change the points scored but not affect the
-competitors' behaviour.
+due to random variation.  This test is performed by function {\tt
+  equalp.test()} in the package which uses the method of support
+\citep{edwards1992}.  The test rejects the null (in favour of the
+evaluate) at $p=4\times 10^{-5}$.  There is thus strong evidence to
+suggest that the competitors are indeed of different skills and the
+result actually reflects genuine differences in abilities, rather than
+the result of random chance.  Alternatively, we may view the dataset
+as being sufficiently informative to reveal actual differences in
+surfing ability.  Observe that the competitors' points (that is, the
+final column of table~\ref{resultstable}) cannot be used for this type
+of probabilistic analysis: the points awarded are arbitrary and
+changing the points system would change the points scored but not
+affect the competitors' behaviour.
 
@@
 strength, we maximize the likelihood subject to the null, and compare
-with the global maximum likelihood~\citep{hankin2010}.  This gives us a
-support difference of about 0.05, well short of Edwards's 2 units of
-support criterion.  Alternatively, we could observe that the
+with the global maximum likelihood~\citep{hankin2010}; package idiom
+is to use {\tt equalp.test()} which reports its findings in a
+standardised format.  This gives us a support difference of about
+0.05, well short of Edwards's 2 units of support criterion.
+Alternatively, we could observe that the
 $-2\log\left(\frac{\Lambda}{\Lambda_0}\right)\simeq 0.1$ is not in the
@@
 possible interaction with the compatriot issue discussed in
-section~\ref{compatriot}.  The null would be that this reified entity
-has zero strength.  We fail to reject this null (differential support
-= 0.035; $p=0.79$) and conclude that there is no evidence to suggest
-that Brazilians' performance is better on their preferred wave types
-than one would expect by chance.  We may go further using the method
-of support: not only is the maximum likelihood estimate for Brazilian
-wavetype very small (about 0.002), but in addition we may reject the
-the proposal that the strength of this entity exceeds about 0.025
-(Figure~\ref{brazilian_wavetype_support}).
+section~\ref{compatriot}.
+
+To illustrate the concept, we show a simple likelihood function in
+Table~\ref{probabilitybrazilian}.  The null would be that this reified
+entity has zero strength.  We fail to reject this null (differential
+support = 0.035; $p=0.79$) and conclude that there is no evidence to
+suggest that Brazilians' performance is better on their preferred wave
+types than one would expect by chance.  We may go further using the
+method of support: not only is the maximum likelihood estimate for
+Brazilian wavetype very small (about 0.002), but in addition we may
+reject the the proposal that the strength of this entity exceeds about
+0.025 (Figure~\ref{brazilian_wavetype_support}).
 
@@
 
+\begin{table}[h]
+\begin{tabular}{l|l|l}
+wave type  & $\operatorname{Prob}(p_1\succ p_2)$ & $\operatorname{Prob}(p_2\succ p_1)$\\ \hline
+beach & $\frac{p_1+B}{p_1+p_2+B}$ & $\frac{p_2}{p_1+p_2+B}$\\
+point & $\frac{p_1+B}{p_1+p_2+B}$ & $\frac{p_2}{p_1+p_2+B}$\\
+reef  & $\frac{p_1\vphantom{B}}{p_1+p_2}$ & $\frac{p_2}{p_1+p_2}$\\
+\end{tabular}
+\caption{Likelihood function for results between surfer 1 (Brazilian)
+  and surfer 2 (non-Brazilian) with \label{probabilitybrazilian} Bradley-Terry
+  strengths $p_1,p_2$ respectively, and a ``Brazilian wavetype" entity with
+  strength $B$ that quantifies the enhanced Brazilian performance on reef wave types}
+\end{table}
+
+\clearpage
+\newpage
+
 \begin{figure}[h]

### 2022-04-23 inst/surfing_jsa_rebuttal_referee2.tex
"speculation [rather] than concrete analysis" referee's comment

@@
 
-Overall, it feels like the main conclusion is "not all surfers have
+\textcolor{blue}{We agree that this section is more speculative than
+  the remainder of the manuscript.  However, we would reiterate that
+  the topic is clearly interesting and worth investigating, as both
+  referees note.  Non-competitive behaviour of the sort discussed here
+  is notoriously difficult to study; and in particular, formal
+  statistical analysis of non-competitive behaviour is almost
+  non-existant in the literature.  We are not aware of any other
+  probability model that is able to even establish the existence of
+  non-competitive behaviour, let alone quantifiy it.  We believe that
+  our approach is new, and our probability model for non-competitive
+  behaviour appropriate and plausible.\\ But, on re-reading the
+  section, we agree that the presentation is indeed somewhat obscure.
+  We have extensively re-written the section and believe it is much
+  improved.  We include a new table motivating and discussing
+  non-competitive behaviour using a particular example from the
+  tournament.  We note that we presented a likelihood function for the
+  three-competitor case in the first draft (it is now Table 3), and we
+  consider this to be as clear as reasonably practicable.}
+
+Overall, it feels like the main conclusion is ``not all surfers have
 the same skill". This is not insignificant, however, it is not

### 2022-04-23 inst/surfing_jsa.tex
"speculation [rather] than concrete analysis" referee's comment

@@
 \usepackage{natbib}
+\usepackage{diagbox}
 \usepackage[pdftex]{hyperref}
@@
 test for such mechanisms, that automatically controls for competitors'
-strengths.  Noting
-that the compatriots effectively place themselves at a {\em
-  disadvantage} when compared with the odd man out, we introduce a
-compatriot monster whose strength $p_C$ helps the odd man out.
-Writing $p_1, p_2$ for the strengths of the compatriots and $q$ for
-the odd man out, we construct likelihood functions for the six
+strengths.  Noting that the compatriots effectively place themselves
+at a {\em disadvantage} when compared with the odd man out, we
+introduce a compatriot monster whose strength $p_C$ helps the odd man
+out.  Writing $p_1, p_2$ for the strengths of the compatriots and $q$
+for the odd man out, we construct likelihood functions for the six
 possible orderings when two compatriots and an odd man out are in the
@@
 \end{table}
+\clearpage
+\newpage
+
+\begin{table}[h]
+  \begin{tabular}{l|l|l|l}\\
+    \backslashbox{order}{value}          & B & F & S\\ \hline
+    $\mbox{B}\succ\mbox{F}\succ\mbox{S}$ & 1 & 0 & 0\\
+    $\mbox{B}\succ\mbox{S}\succ\mbox{F}$ & 1 & 0 & 1\\
+    $\mbox{F}\succ\mbox{B}\succ\mbox{S}$ & 0 & 1 & 0\\
+    $\mbox{F}\succ\mbox{S}\succ\mbox{B}$ & 0 & 1 & 1\\
+    $\mbox{S}\succ\mbox{B}\succ\mbox{F}$ & 1 & 0 & 2\\
+    $\mbox{S}\succ\mbox{F}\succ\mbox{B}$ & 0 & 1 & 2\\
+  \end{tabular}
+  \caption{Simplified payoff table\label{valuetable} for Buchan (B),
+    Freestone (F), and Smith (S).  Buchan and Freestone, being
+    Australian, have a payoff that reflects their likelihood of being
+    picked for the national Olympic squad: they are interested in
+    whether or not they beat their countryman.  Thus Buchan's payoff
+    shows whether or not he beat Freestone, and vice versa.  Smith,
+    being South African, is motivated by his ranking against his
+    competitors and here we show Smith with a Borda-type payoff.
+    Smith observes that {\em both} Buchan and Freestone are
+    indifferent between outcomes $\left\lbrace
+    \mbox{B}\succ\mbox{F}\succ\mbox{S},
+    \mbox{B}\succ\mbox{S}\succ\mbox{F},
+    \mbox{S}\succ\mbox{B}\succ\mbox{F}\right\rbrace$, and may be able
+    to exploit this indifference by strategies increasing the
+    probability of his desired outcome,
+    $\mbox{S}\succ\mbox{B}\succ\mbox{F}$}
+\end{table}
 

### 2022-04-24 inst/surfing.Rmd
more work on referee 2

@@
 
+# Medina's performance at the Surf Ranch
+
+We will test the null that Medina's performance at the Surf Ranch wa
+the same as his performance in the remainder of the tourament.
+
+```{r label=medinaattheranchlike, cache=TRUE}
+surfing_wct08     <- hyper2()
+surfing_elsewhere <- hyper2()
+
+for(v in o){
+   venue <- v[1]
+   r <- v[-1]  # first entry is venue
+
+   r <- r[!(r %in% remove)]
+   if(length(r)>1){
+       if(venue == "wct08"){
+           rmod <- r
+           rmod[rmod == "Medina"] <- "Medina_wct08"
+           surfing_wct08 <- surfing_wct08 + race(rmod)
+       } else {
+           surfing_elsewhere <- surfing_elsewhere + race(r)
+       } 
+   }
+} # v loop closes	
+both_together <- surfing_wct08 + surfing_elsewhere
+```
+
+Now test the null that Medina's performance is identical at wct08 and
+elsewhere:
+
+```{r, label=testmedinasranchperformance,cache=TRUE}
+samep.test(both_together,c("Medina","Medina_wct08"))
+```
+
 # Brazilian null

### 2022-04-24 inst/surfing_jsa.tex
more work on referee 2

@@
 
+Medina's performance was startlingly good at Freshwater Pro ({\tt
+  wct08}).  Indeed, Medina came first at only two venues (the other
+was at Jeffries Bay, South Africa) and it is natural to enquire
+whether there is statistical evidence to support the assertion that he
+performed better under the more controlled conditions of the Lemoore
+Surf Ranch.  To do this, we create an artificial surfer, {\tt
+  Medina\_wct08}, who competes at Lemoore, and test the null that {\tt
+  Medina} has the same Plackett-Luce strength as {\tt Medina\_wct08}.
+We reject this null at a $p$-value of 0.043 and conclude that Medina's
+performance at Lemoore is significantly better than his performance at
+natural venues.
+
 \section{The Brazilian storm}

### 2022-04-24 inst/surfing_jsa_rebuttal_referee2.tex
more rebutalling

@@
 
+\textcolor{blue}{We regard the novelty of our computational approach
+  as being access to the full likelihood function over allowable
+  Bradley-Terry strengths (that is, not restricted to the MLE or point
+  hypotheses, but over the entire simplex, $\left\lbrace
+  p_1,\ldots,p_n\left|p_i\geq 0,\sum p_j=1\right.\right\rbrace$) and
+  also the furnishing of compact and natural R idiom to allow the
+  creation of such likelihood functions.  As such, there are many many
+  statistical tests that could be conceived and executed using our
+  method.  Our original priority was to submit a brief and compact
+  manuscript but on reflection the referee seems to be saying that we
+  have erred on the side of brevity.  We have added a topical
+  statistical test, assessing Medina's performance at Lemoore as an
+  example of the power of our method.  This test returned a
+  significant p-value, a result not achievable with other methods.  We
+  feel that this new analysis complements the existing analysis
+  comparing Medina with Ferreira (and which also rejected the null);
+  we have added a short discussion to the manuscript.  \\ \\ Reading
+  between the lines of the comment, we detect a certain disappointment
+  that some of the results were non-significant.  We too felt a little
+  chagrined that our method returned a non-significant result for the
+  compatriot analysis.  However, we would argue that it is important
+  to report all findings, significant or not---especially for
+  difficult or otherwise challenging hypotheses not accessible to
+  conventional techniques.  We would also point out that even for
+  nonsiginficant results, our methods were considerably more
+  informative than ``failure to reject a null'': as we stated in the
+  submission, we present a novel quantification of noncompetitiveness
+  (viz, $p_C$), and further, presented a nonzero maximum likelihood
+  estimate for $p_C$ which, we suggest, is a worthy and interesting
+  finding.  We also point out that such noncompetitiveness is a
+  general phenomenon in the more general context of tournaments.  We
+  present our work as the first application of a novel and very
+  general method: we hope that subsequent analysis would cite this
+  work.}
+
 From a style perspective, the paper is well-written and enjoyable to

### 2022-04-24 inst/surfing_jsa.tex
more rebuttal

@@
   Medina\_wct08}, who competes at Lemoore, and test the null that {\tt
-  Medina} has the same Plackett-Luce strength as {\tt Medina\_wct08}.
+  Medina} has the same competitive strength as {\tt Medina\_wct08}.
 We reject this null at a $p$-value of 0.043 and conclude that Medina's
 performance at Lemoore is significantly better than his performance at
-natural venues.
+natural venues.  From an inferential perspective, this test is
+interesting because competitor {\tt Medina} never faces {\tt
+  Medina\_wct08}.  It is as though competitor A consistently beats B,
+and B consistently beats C in pairwise competition, but A has never
+directly competed against C.  Nevertheless we may assert with some
+confidence that the Bradley-Terry strength of A exceeds that of C in
+the absence of nontransitive effects~\citep{west2008}, via indirect
+comparison.
 

### 2022-04-25 inst/surfing_jsa_rebuttal_referee2.tex
appearance of the graphs

@@
 Short story: we have accommodated all the comments with rewording and
-clarification.  We have included an extended discussion of some of the
-points below.  The resulting document is, we believe, stronger and
-more scholarly than before and we recommend it to you.
+clarification.
+
+Slightly longer story: the referee makes a wide range of observations,
+some good and some bad.  We have overhauled much of the writing, added
+descriptions of the computational and mathematical methodology,
+clarified ambiguous wording, produced new statistical tests, and
+improved the visuals.
+
+We have included an extended discussion of some of the points below.
+The resulting document is, we believe, stronger and more scholarly
+than before and we recommend it to you.
 
@@
 significant rework in terms of formatting.
+
+\textcolor{blue}{We have reworked the figures:
+  \begin{itemize}
+  \item replaced the pie chart with a dotchart (log scale)
+  \item retouched the fine-scale detail of the scatterplot of points scored 
+    against maximum likelihood estimates
+  \item changed the caption of the ``Profile support for the Brazilian
+    wavetype reified entity" diagram to be more informative.  We
+    suspect that the low information density of this graph contributed
+    to overall impression of the presentation alluded to by the
+    comment above.  We would observe, however, that this particular
+    diagram is worth retaining despite its plain appearance on two
+    grounds: firstly, it supports our contention that the whole
+    likelihood function is informative; and secondly, the smoothly
+    varying curve is strong evidence that numerical optimization
+    techniques are working correctly.
+  \end{itemize}
+  For the tables, we are happy to take specific advice, but in the
+  meantime we have improved  some of the mathematical sections [using
+    {\tt vphantom}] which reduces the cluttered appearance.
+}
 
@@
 
+\textcolor{blue}{As discussed above, we have added new statistical
+  tests and intepretations of existing tests.  We have rewritten and
+  overhauled large sections of the writing.}
+
 \end{document}

### 2022-04-25 inst/surfing.Rmd
appearance of the graphs

@@
 mydotchart(kk,col="blue",pch=16,xlim=l,xaxt='n',frame.plot=FALSE)
+
+pdf(file="showsurfingmaxpdotchart.pdf")
+jj <- log10(surfing_maxp)
+dotchart(jj,pch=16,xlab="Log of Bradley-Terry strength of competitors (base 10)")
+dev.off()
 ```
@@
 pdf(file="directlogplotofstrengths.pdf")
-plot(jjO,jjL,pch=16,log='xy',xlab='points',ylab='maximum likelihood strength')
+par(xpd=NA)
+plot(jjO,jjL,pch=16,log='xy',xlab='points',ylab='maximum likelihood strength',cex=1.5)
 fit <- lm(log10(jjL)~log10(jjO))
 for(i in seq_along(jjO)){
-text(jjO[i],jjL[i],names(jjL)[i],col='gray',pos=4,cex=0.5)
+text(jjO[i],jjL[i],names(jjL)[i],col='gray',pos=4,cex=0.8)
 }
+par(xpd=FALSE)
 abline(fit)
@@
 pdf(file="plotwidep.pdf")
-plot(c(0,p_wide),c(l0,l_wide)-max(l),type='b',xlab="Brazilian beachtype wide",ylab="support")
+plot(c(0,p_wide),c(l0,l_wide)-max(l),type='b',xlab="Bradley-Terry strength of the Brazilian beachtype reified entity",ylab="support")
 abline(h= -2,lty=2)

### 2022-04-25 inst/surfing_jsa.tex
appearance of the graphs

@@
 
-Figure~\ref{piechartmax} shows the maximum likelihood estimate for the
+Figure~\ref{dotchartmax} shows the maximum likelihood estimate for the
 23 competitors' strengths, which appears to show a wide range from
@@
 \begin{tabular}{llllllllllllll}
-competitor  & nationality&WCT01  &WCT02  &WCT03  & \ldots &WCT09&WCT10&WCT11&points\\ \hline
- Ferreira   &   BRA      &   1   &   5   &   17  & \ldots      &  2   &   1   &   1   &   59740\\
- Medina     &   BRA      &   5   &   5   &   17  & \ldots      &  9   &   9   &   2   &   56475\\
- Smith 	    &   ZAF      &   3   &   3   &   17  & \ldots      &  9   &   2   &   17  &   49985\\
- Toledo     &   BRA      &   9   &   2   &   5   & \ldots      &  17  &   5   &   17  &   49145\\
- Andino     &   USA      &   2   &   17  &   5   & \dots       &  5   &   5   &   9   &   46655\\
- $\ldots$   &            &   &   &       &       & \dots       &      &       &       &
+competitor  & nationality&WCT01  &WCT02  &WCT03  & \ldots  &WCT10&WCT11&points\\ \hline
+ Ferreira   &   BRA      &   1   &   5   &   17  & \ldots  &   1   &   1   &   59740\\
+ Medina     &   BRA      &   5   &   5   &   17  & \ldots  &   9   &   2   &   56475\\
+ Smith 	    &   ZAF      &   3   &   3   &   17  & \ldots  &   2   &   17  &   49985\\
+ Toledo     &   BRA      &   9   &   2   &   5   & \ldots  &   5   &   17  &   49145\\
+ Andino     &   USA      &   2   &   17  &   5   & \dots   &   5   &   9   &   46655\\
+ $\vdots$   &            &   &   &       &       & $\vdots$&       &       &
 \end{tabular}
@@
 order  & ${\mathcal L}(p_1,p_2,q,p_C)$\\ \hline
-$p_1\succ p_2\succ q $ & $ \frac{p_1}{p_1+p_2+q+p_C}\cdot\frac{p_2}{p_2+q+p_C}$\\
-$p_2\succ p_1\succ q $ & $ \frac{p_2}{p_1+p_2+q+p_C}\cdot\frac{p_1}{p_1+q+p_C}$\\
-$p_1\succ q\succ p_2 $ & $ \frac{p_1}{p_1+p_2+q+p_C}\cdot\frac{q+p_C}{p_2+q+p_C}$\\
-$p_2\succ q\succ p_1 $ & $ \frac{p_2}{p_1+p_2+q+p_C}\cdot\frac{q+p_C}{p_1+q+p_C}$\\
-$q\succ p_1\succ p_2 $ & $ \frac{q+p_C}{p_1+p_2+q+p_C}\cdot\frac{p_1}{p_1+p_2}$\\
-$q\succ p_2\succ p_1 $ & $ \frac{q+p_C}{p_1+p_2+q+p_C}\cdot\frac{p_2}{p_1+p_2}$\\
+$p_1\succ p_2\succ q $ & $ \frac{p_1}{p_1+p_2+q+p_C}\cdot\frac{p_2^{\vphantom{h}}}{p_2+q+p_C}$\\
+$p_2\succ p_1\succ q $ & $ \frac{p_2}{p_1+p_2+q+p_C}\cdot\frac{p_1^{\vphantom{h}}}{p_1+q+p_C}$\\
+$p_1\succ q\succ p_2 $ & $ \frac{p_1}{p_1+p_2+q+p_C}\cdot\frac{q+p_C^{\vphantom{h}}}{p_2+q+p_C}$\\
+$p_2\succ q\succ p_1 $ & $ \frac{p_2}{p_1+p_2+q+p_C}\cdot\frac{q+p_C^{\vphantom{h}}}{p_1+q+p_C}$\\
+$q\succ p_1\succ p_2 $ & $ \frac{q+p_C}{p_1+p_2+q+p_C}\cdot\frac{p_1^{\vphantom{h}}}{p_1+p_2}$\\
+$q\succ p_2\succ p_1 $ & $ \frac{q+p_C}{p_1+p_2+q+p_C}\cdot\frac{p_2^{\vphantom{h}}}{p_1+p_2}$\\
 \end{tabular}
@@
 \begin{tabular}{l|l|l}
-wave type  & $\operatorname{Prob}(p_1\succ p_2)$ & $\operatorname{Prob}(p_2\succ p_1)$\\ \hline
-beach & $\frac{p_1+B}{p_1+p_2+B}$ & $\frac{p_2}{p_1+p_2+B}$\\
-point & $\frac{p_1+B}{p_1+p_2+B}$ & $\frac{p_2}{p_1+p_2+B}$\\
-reef  & $\frac{p_1\vphantom{B}}{p_1+p_2}$ & $\frac{p_2}{p_1+p_2}$\\
+wave type  & $\operatorname{Prob}(p_1\succ p_2)$ & $\operatorname{Prob}(p_2\succ p_1^{\vphantom{h}})$\\ \hline
+beach & $\frac{p_1+B}{p_1+p_2^{\vphantom{h}}+B}$ & $\frac{p_2^{\vphantom{h}}}{p_1+p_2^{\vphantom{h}}+B}$\\
+point & $\frac{p_1+B}{p_1+p_2^{\vphantom{h}}+B}$ & $\frac{p_2^{\vphantom{h}}}{p_1+p_2^{\vphantom{h}}+B}$\\
+reef  & $\frac{p_1\vphantom{B}}{p_1+p_2}$ & $\frac{p_2^{\vphantom{h}}}{p_1^{\vphantom{h}}+p_2}$\\
 \end{tabular}
@@
 \begin{figure}[h]
-\includegraphics[width=4in]{showsurfingmaxppie.pdf} % Pie chart of strengths
-\caption{Pie chart\label{piechartmax} showing maximum likelihood strengths of the 23 competitors.  Note the
-dominance of Ferreira, Florence, and Medina}
+\includegraphics[width=4in]{showsurfingmaxpdotchart.pdf} % pdf produced by surfing.Rmd
+\caption{Dot chart\label{dotchartmax} showing maximum likelihood
+  strengths of the 23 competitors in alphabetic order, logarithmic
+  axis.  Note the dominance of Ferreira, Florence, and Medina}
 \end{figure}
@@
 \begin{figure}[h]
-\includegraphics[width=4in]{directlogplotofstrengths.pdf}
+\includegraphics[width=4in]{directlogplotofstrengths_postinkscape.pdf}
 \caption{Scatterplot of points scored  \label{compare_likelihood_points}

### 2022-04-26 inst/surfing_jsa_rebuttal_referee2.tex
typos

@@
     maximum likelihood estimates
-  \item Referenced the computational method [{\tt equalp.test()} for
+  \item Referenced the computational method [{\tt equalp.test()}] for
     testing the null of equal competitive abilities
   \item Introduced a new table that explicitly sets out the likelihood
-    function for the reified entity Brazilian beachtype reified entity
+    function for the {\em Brazilian beachtype} reified entity
   \end{itemize}
@@
     suspect that the low information density of this graph contributed
-    to overall impression of the presentation alluded to by the
+    towards overall impression of the presentation alluded to by the
     comment above.  We would observe, however, that this particular
@@
   probability model that is able to even establish the existence of
-  non-competitive behaviour, let alone quantifiy it.  We believe that
+  non-competitive behaviour, let alone quantify it.  We believe that
   our approach is new, and our probability model for non-competitive
@@
   conventional techniques.  We would also point out that even for
-  nonsiginficant results, our methods were considerably more
+  nonsignificant results, our methods were considerably more
   informative than ``failure to reject a null'': as we stated in the

### 2022-04-26 inst/surfing_jsa.tex
typos

@@
 23 competitors' strengths, which appears to show a wide range from
-Florence at about 17.5\% down to Coffin at about 1.5\%.  The {\tt
+Florence at about 17.5\% down to Morais at about 1\%.  The {\tt
   hyper2} software~\citep{hankin2017} may be used to formally assess
@@
 that is, that such noncompetitive behaviour is absent.  We find that
-the maximum likelihood estimate for $p_C$ is about 0.015, weaker than
+the maximum likelihood estimate for $p_C$ is about 0.017, weaker than
 the ``real'' surfers, but the support for a nonzero $p_C$ is about
@@
 support the supposed Brazilian preference for point and beach break
-beaches.  We also find no evidence to suggest that surfing's being
-included in the 2020 Summer Olympics results in modified strategies to
-bolster intranational ranking at the expense of WCT performance.
+beaches.  We also suggest that surfing's being included in the 2020
+Summer Olympics might result in modified strategies to bolster
+intranational ranking at the expense of WCT performance; we quantify
+the strength of such a phenomenon as the strength of a reified
+Bradley-Terry entity: a ``compatriot monster'' with an estimated
+strength of about $0.017$.
 

### 2022-04-26 inst/surfing_jsa_rebuttal_referee2.tex
tweaks

@@
 Slightly longer story: the referee makes a wide range of observations,
-some good and some bad.  We have overhauled much of the writing, added
-descriptions of the computational and mathematical methodology,
-clarified ambiguous wording, produced new statistical tests, and
-improved the visuals.
+all which we have engaged with.  We have overhauled much of the
+writing, added descriptions of the computational and mathematical
+methodology, clarified ambiguous wording, produced new statistical
+tests, and improved the figures.  We have tried hard to retain the
+many positive aspects of the manuscript that the referee explicitly
+notes.
 
@@
   from ``previous related work'' to ``Previous statistical analysis of
-  competitive surfing'' which is a better description, we feel.}
+  competitive surfing'' which is a better description, we feel.  We
+  have added several new statistical tests and given brief outlines of
+  their significance.}
 
@@
 
-\textcolor{blue}{As discussed above, we have added new statistical
-  tests and intepretations of existing tests.  We have rewritten and
-  overhauled large sections of the writing.}
+\textcolor{blue}{We believe that this referee's insightful and
+  constructive comments have led to a much improved submission.}
 

### 2022-04-26 inst/surfing_jsa.tex
tweaks

@@
 
-The 2019 WSL tour saw competitive events held at eleven
-different surf locations.
-
-The typical format is as follows.  Up to four surfers---the
-competitors---are in the water simultaneously, watched by up to five
-judges.  Subject to surfing etiquette, competitors are free to catch a
-wave at their discretion within the heat's time window.  The judges
-rate each wave ride and award points to the surfer based on a range of
-subjective criteria such as stylish execution and transition of
-manoeuvres, but also credit elements such as the difficulty and
-novelty of the performance.  The top score and bottom score among the
-five judges are removed and the remaining three judges' scores are
-averaged to give the final score for the surfer for that wave.  Each
-surfer's aggregate score is given by the average of the two
-highest-scoring waves.  The winner is the surfer with the highest
-aggregate score.
+The 2019 WSL tour saw competitive events held at eleven different surf
+locations.  The typical format is as follows.  Up to four
+surfers---the competitors---are in the water simultaneously, watched
+by up to five judges.  Subject to surfing etiquette, competitors are
+free to catch a wave at their discretion within the heat's time
+window.  The judges rate each wave ride and award points to the surfer
+based on a range of subjective criteria such as stylish execution and
+transition of manoeuvres, but also credit elements such as the
+difficulty and novelty of the performance.  The top score and bottom
+score among the five judges are removed and the remaining three
+judges' scores are averaged to give the final score for the surfer for
+that wave.  Each surfer's aggregate score is given by the average of
+the two highest-scoring waves.  The winner is the surfer with the
+highest aggregate score.
 
@@
   \begin{cases}
-    -n/\left(p_i+p_j+\sum p_k\right)^2\qquad\mbox{if $i,j\in K$}\\
+    -n/\left(\sum p_k\right)^2\qquad\mbox{if $i,j\in K$}\\
     0\qquad\mbox{otherwise.}
@@
 competition with one another for a place in the Australian Olympic
-surfing team, would be more motivated to beating each other than to
+surfing team, would be more motivated to beat each other than to
 come ahead of Smith.  This would be a rational response, but arguably

### 2022-04-28 inst/kka.Rmd
hyper3 likelihoods for chess with draws

@@
 ```
+
+<p style="text-align: right;">
+![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
+</p>
 
@@
 
-```{r defineH}
-
+```{r defineH,cache=TRUE}
 results <- as.list(kka)
@@
 
-```{r testall3}
+```{r testall3,cache=TRUE}
 samep.test(karpov_kasparov_anand,c("Karpov","Kasparov","Anand"))
@@
 
+# Some ideas for `hyper3` analysis.
+
+First we ignore the draws and create a matrix with real parts
+representing white wins (home wins in the context of football) and
+imaginary parts representing black (away) wins.  A pairwise likelihood function would be
+
+$$
+\left(\frac{\lambda p_1}{\lambda p_1+p_2}\right)^a
+\left(\frac{        p_2}{\lambda p_1+p_2}\right)^b
+\left(\frac{        p_1}{p_1+\lambda p_2}\right)^c
+\left(\frac{\lambda p_1}{p_1+\lambda p_2}\right)^d
+$$
+
+for the observation of $+a\,=0\,-b$ for $p_1$ playing white and
+$+c\,=0\,-d$ for $p_2$ playing white.  We then translate this into a
+`hyper3` likelihood function as follows
+
+```{r,label=kka_hyper3,cache=TRUE}
+(M <- kka_array[,,1] + 1i*kka_array[,,3])
+home_away3(M,lambda=1.1)
+```
+
+Above we use $\lambda=1.1$, as a specific numeric value is required by
+the idiom.
+
+```{r, showhyper3kkamaxlikeexists,cache=TRUE}
+l <- function(lambda){
+  H <- home_away3(M,lambda=lambda)
+  loglik(maxp(H),H)
+}
+lam <- seq(from=1.2,to=3.5,len=10)
+like <- sapply(lam,l)
+```
+
+and plot it:
+
+```{r,label=plothyper3likecurve}
+plot(lam,like-max(like),type="b",pch=16)
+abline(h=c(0,-2),lty=2)
+```
+
+Above we see a credible interval for $\lambda$ being about
+$(1.4,2.9)$.  Of course, this is crude and it would be betterto have
+more transparent package idiom.  But I'm working on this.
+
+We can now introduce draws but first a change of notation.  The
+previous likelihood function can be represented thus:
+
+$$
+\left(\frac{\lambda p_1, p_2}{\lambda p_1+p_2}\right)^{(a,b)}
+\left(\frac{p_1, \lambda p_2}{p_1+\lambda p_2}\right)^{(c,d)}
+$$
+
+Draws would change this by introducing a weighted Draw monster D$:
+
+$$
+\left(\frac{\lambda p_1, D(p_1+p_2),p_2}{\lambda p_1+D(p_1+p_2)+p_2}\right)^{(a,b,c)}
+\left(\frac{p_1, D(p_1+p_2),\lambda p_2}{p_1+D(p_1+p_2)+\lambda p_2}\right)^{(d,e,f)}
+$$
+
+or possibly an individual Draw monster:
+
+$$
+\left(\frac{\lambda p_1, D_1p_1+D_2p_2,p_2}{\lambda p_1+D_1p_1+D_2p_2+p_2}\right)^{(a,b,c)}
+\left(\frac{p_1, D_1p_1+D_2p_2),\lambda p_2}{p_1+D_1p_1+D_2p_2+\lambda p_2}\right)^{(d,e,f)}
+$$
+
+Note that in neither of these models do we have the probability of a
+draw for $p_1$ playing white being the same as for $p_2$ playing
+white.  In the first case the two probabilities are
+
+$$
+\frac{D(p_1+p_2)}{(\lambda +D)p_1 + (1+D)p_2},\qquad
+\frac{D(p_1+p_2)}{(1+D)p_1 + (\lambda+D)p_2}
+$$
+
+and in the second they are
+
+$$
+\frac{D_1p_1+D_2p_2}{(\lambda +D_1)p_1 + (1+D_2)p_2},\qquad
+\frac{D_1p_1+D_2p_2}{(1 +D_1)p_1 + (\lambda+D_2)p_2}
+$$
+
 ### Package dataset

### 2022-04-28 inst/kka.Rmd
white_draw3()

@@
 
-Draws would change this by introducing a weighted Draw monster D$:
+Draws would change this by introducing a weighted Draw monster $D$:
 
@@
 
+With strictly positive $D,D_1,D_2$ these are equal if and only if
+$(\lambda-1)(p_1-p_2)=0$, that is, if the white advantage is
+nonexistant, or if the players have the same strength.
+
+We can use the bespoke function `white_draw3()` to make a likelihood
+function:
+
+```{r,makewhitedrawlike,cache=TRUE}
+white_draw3(kka_array, 1.1, 2.3)
+```
+
+Then make a wrapper:
+
+```{r,maxewhitedrawlike,cache=TRUE}
+f <- function(v){
+    lambda <- v[1]
+    D      <- v[2]
+    H <- white_draw3(kka_array, lambda,D)
+    max(maxp(H,justlikes=TRUE))
+}	
+f(c(1.5,2.3))
+```
+
+```{r,contourlikewhite,cache=TRUE}
+n <- 10
+lambda <- seq(from=1.4,to=2.9,len=n)
+D <- seq(from=1.6,to=3.4,len=n)	
+M <- as.matrix(expand.grid(lambda=lambda,D=D))
+l <- matrix(apply(M,1,f),n,n)
+```
+
+```{r}
+l <- l-max(l)
+contour(lambda,D,l,nlevels=30,xlab="lambda",ylab="D")
+abline(v=lambda,col='gray',lwd=0.2)
+abline(h=D,col='gray',lwd=0.2)
+lambda
+D
+```
+
 ### Package dataset

### 2022-04-28 R/hyper3.R

@@ -516,5 +516,46 @@ stop("not yet written")
     return(H)
 }
 
+## kka_array is arranged differently from the previous; it uses
+## white_wins and white_losses, not white wins and white losses
 
-
+`white_draw3` <- function(A,lambda,D){ # white_draw3(kka_array,1.1,0.2)
+    stopifnot(identical(dimnames(A)[[1]],dimnames(A)[[2]]))
+    players <- dimnames(A)[[1]] 
+    H <- hyper3(pnames=c(players,"lambda","D"))
+    d <- dim(A)
+    
+    for(i in seq_len(d[1])){
+        for(j in seq_len(d[2])[-i]){
+            white_player <- players[i]
+            black_player <- players[j]
+            
+            white_player_wins  <- A[i,j,1]
+            white_player_draws <- A[i,j,2]
+            white_player_loses <- A[i,j,3]
+            
+            total_matches <- white_player_wins + white_player_draws + white_player_loses 
+            
+            ## white player wins:
+            jj <- lambda
+            names(jj) <- white_player
+            H[jj] %<>% inc(white_player_wins)
+            
+            ## white player loses (so the black player wins):
+            jj <- 1
+            names(jj) <- black_player
+            H[jj] %<>% inc(white_player_loses)
+            
+            ## draw:
+            jj <- c(D,D)
+            names(jj) <- c(white_player,black_player)
+            H[jj] %<>% inc(white_player_draws)
+            
+            ## lastly, change the denominator so the powers sum to zero:
+            jj <- c(lambda+D , 1+D)
+            names(jj) <- c(white_player,black_player)
+            H[jj] %<>% dec(total_matches)
+        } # j loop closes
+    } # i loop closes
+    return(H)
+}

### 2022-05-10 inst/race3.Rmd
typos
slight rewording and tidyup

@@
 
-(takes about 15 minutes to process)
+(takes about an hour to process)
 
@@
 
-# Some simple cases
+# Some simple cases: likelihood functions for order statistics with clones
 
-The simplest nontrivial case is three competitors with strengths
-$a,a,b$ and finishing ordr $a\succ b\succ a$.  The likelihood
-function would be
+Here I consider some order statistics with nontrivial maximum
+likelihood Bradley-Terry strength.   The simplest nontrivial case is
+three competitors with strengths $a,a,b$ and finishing order $a\succ
+b\succ a$.  The likelihood function would be
 
@@
 
+# Some simple cases: Fisher information for simple cases
+
+(work in progress).
+
 # The constructors' championship
@@
 constructor_table_2021 <- jj[,-ncol(jj)]
-constructor_table_2021[,1:10]
+constructor_table_2021[,1:17]
 constructor_2021_table <- constructor_table_2021
@@
 and see that there is (very) strong evidence to reject this null.
-Further, we may test the hypothesis that the big four (viz Mercedies,
+Further, we may test the hypothesis that the big four (viz Mercedes,
 Ferrari, Red Bull Racing Honda, and McLaren-Mercedes) are equal:

### 2022-05-10 inst/kka.Rmd
typo

@@
 M <- rbind(a1,rev(a2))
-dimnames(M) <- list(karpov_plays=c("White","Black"),result=c("karpov wins","karpov draws","kaprov loses"))
+dimnames(M) <- list(karpov_plays=c("White","Black"),result=c("karpov wins","karpov draws","karpov loses"))
 M

### 2022-05-11 inst/carcinoma.Rmd
further work on issue #131

---
title: "Carcinoma"
output: bookdown::html_document2
---

```{r setup, include=FALSE}
set.seed(0)
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("magrittr")
options("digits" = 5)
```

<p style="text-align: right;">
![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
</p>

```{r loadlib}
carcinoma_table <- read.table("carcinoma.txt",header=TRUE)
carcinoma_table[,1:7] <- carcinoma_table[,1:7] > 0
wanted <- !(apply(carcinoma_table[,1:7],1,sum) %in% c(0,7))
carcinoma_table <- carcinoma_table[wanted,]
carcinoma_table
```

```{r makeggrllist,cache=TRUE}
W <- hyper2(pnames=LETTERS[1:7])
Wi <- list()

for(i in seq_len(nrow(carcinoma_table))){
      jj <- unlist(carcinoma_table[i,1:7])
      negative <- names(jj)[jj==0]
      positive <- names(jj)[jj==1]
      print(negative)
      print(positive)
      Wi[[i]] <- ggrl(W,positive,negative)
}
HL <- lsl(Wi,powers= carcinoma_table$n)
```

Now

```{r,label=calccloglikzipf,cache=TRUE}
z <- zipf(7)
names(z) <- LETTERS[1:7]
system.time(lz <- loglik_lsl(z,HL))
lz
```

```{r,label-calcloglikequalp,cache=TRUE}
e <- equalp(W)
system.time(le <- loglik_lsl(e,HL))
le
```

Agresti, table 13.1, p542.

Landis, J. R., and G. G. Koch. 1977. An application of hierarchical
kappa-type statistics in the assessment of majority agreement among
multiple observers. Biometrics 33: 363-374.

### Package dataset {-}

Following lines create `carcinoma.rda`, residing in the `data/`
directory of the package.

```{r}
carcinoma <- 5

```{r,label=savecarcinomadataset}
save(carcinoma_table,carcinoma, file="carcinoma.rda")
```

### 2022-05-12 inst/rock_paper_scissors_monster.Rmd
fixes issue #133

@@
 
-```{r}
+```{r equalptestchunk,cache=TRUE}
 equalp.test(chess)
@@
 (a "procyclic win" is a win that would be expected under a strong
-monster).  Here we also have $3\alpha+M=0$, so to estimate the
+monster).  Here we also have $3\alpha+M=1$, so to estimate the
 strength of the monster we would solve
@@
 
-(a graph is given below).  We may relax the assumption of equal
-strength of the real players by using a more sophisticated hyper2
-construction:
+and here we have $a=22+23+10=55$ and $b=13+12+8=33$ (a graph is given
+below).  It is straightforward but not trivial to translate this into
+`hyper3` idiom.  Noting that $3\alpha+M=1$ we define $\beta=3\alpha$
+and then the unit sum constraint is $\beta+M=1$ and our probabilities
+are
+
+$$
+\frac{\alpha+M;\alpha}{2\alpha+M   } =
+\frac{\beta/3+M;\beta/3}{2\beta/3+M} =
+\frac{1+2M;1-M}{2+M}
+$$
+
+(where the last expression is obtained using $\beta+M=1$).  We may set
+up a log-likelihood function of $a,b$ using `hyper3` idiom as follows:
+
+```{r,usehyper3}
+TqAqK <- function(a,b){
+  H3 <- hyper3(pnames=c("beta","M"))
+  H3[c(beta=1/3,M=1)] %<>% inc(a  )
+  H3[c(beta=1/3    )] %<>% inc(  b)
+  H3[c(beta=2/3,M=1)] %<>% dec(a+b)
+  return(H3)
+}
+H3 <- TqAqK(55,33)
+loglik(c(M=0.1,beta=0.9),H3)
+loglik(c(beta=0.9,M=0.1),H3)  #  NB: identical!
+```
+
+[above, we note that $\beta+M=3\alpha+M=1$, so the monster is the
+fillup value with respect to $\beta$].  This is easy to maximize:
+
+```{r,maximizeTqAqK}
+optimize(function(M){loglik(c(M=M,beta=1-M),H3)},c(0.01,0.99),maximum=TRUE)
+```
+
+This gives maximum likelihood estimates of $M=0.18$ as would be
+expected from the simple binomia model; the probability of a procyclic
+win is about $0.625$.
+
+We may relax the assumption of equal strength of the real
+players by using a more sophisticated hyper2 construction:
 
@@
 
-```{r,testequalTAKM}
+```{r,testequalTAKM,cache=TRUE}
 samep.test(H,c("Topalov","Anand","Karpov"))
@@
 
-```{r}
+```{r,specmtest,cache=TRUE}
 specificp.gt.test(H,"M",0)
@@
 
-```{r}
+```{r,label=plotproflik,cache=TRUE}
 M <- seq(from=0.001,to=0.4,len=40)
@@
 Above we see support curves for RPS monster strength using TAKM and
-$T=A=K$, showing very similar results.
+$T=A=K$, showing very similar results.  Note that the T=A=K curve
+could have been obtained more nicely with function `TqAqK` but I
+haven't got round to changing it yet.
 
@@
 
-```{r,label=testHW}
+```{r,label=testHW,cache=TRUE}
 maxp(HW)
@@
 
-```{r,label=maxpHWM}
+```{r,label=maxpHWM,cache=TRUE}
 maxp(HWM)
@@
 
-```{r}
+```{r,label=specicifgtest,cache=TRUE}
 specificp.gt.test(HWM,"M",0)

### 2022-05-15 inst/carcinoma.Rmd
carcinoma analysis

@@
 </p>
+
+This is a short analysis of the carcinoma dataset presented by
+Agresti, table 13.1, p542.  Seven pathologists classified each of 118
+slides on the presence or absence of carcinoma in the uterine
+cervix.
+
+The probability model is as follows.  To each pathologist $i,1\leq i
+\leq 7$, we assign a non-negative Bradley-Terry strength and require
+$\sum p_i=1$.  Given that two pathologists $i$ and $j$ differ in their
+diagnosis then the probability of $i$ diagnosing the presence, and $j$
+diagnosing the absence, of carcinoma is given by
+
+\begin{equation}\tag{1}
+i^+,j^-\frac{p_i}{p_i+p_j},\qquad i^-,j^+\frac{p_j}{p_i+p_j}
+\end{equation}
+
+Operationally the way we implement this in `hyper2` idiom is to posit
+an (unobservable) ranking on the clinicians: we place the clinicians
+in order from most convinced of the presence of carcinoma to the least
+convinced.  Thus a ranking of `DACBEGF` means that clinician `D` was
+most convinced of the presense of carcinoma and `F` least convinced.
+Given this, and given a single slide, our actual observation might be
+that `D` and `A` diagnosed presence, and the others diagosed absence,
+of carcinoma.  The corresponding entry would be
+
+```
+A B C D E F G
+T F F T F F F
+```
+
+We ask what rankings are consistent with this observation.  Any
+ranking with `A` and `D` ahead of `BCEFG` would be admissible.  In
+`hyper2` notation we would say
+
+$$\left\lbrace A,D\right\rbrace\succ\left\lbrace B,C,E,F,G\right\rbrace$$
+
+Any ranking where `A` and `D` occupy the first two ranks and
+`B`,`C`,`E`,`F`,`G` ranks three to seven, would be admissible.  We
+would have $2!5!=240$ rankings.  The probability of the actual ranking
+being one of these is simply the sum of these 48 as the rankings are
+mutually exclusive.
+
+Each ranking has a Plackett-Luce likelihood function which is created
+in `hyper2` idiom using the bespoke `race()` function.  Here, we use
+generalized grouped rank likelihood (`ggrl()`) to create an `lsl`
+object, as used by Hankin 2019 for the MasterChef analysis.  This
+furnishes a Bradley-Terry model for the clinicians' strength that
+recovers equation (1).
 
@@
 }
-HL <- lsl(Wi,powers= carcinoma_table$n)
+carcinoma <- lsl(Wi,powers= carcinoma_table$n)
 ```
 
-Now
+To maximize the likelihood we need a good starting point and rather
+than use the default `equalp()` we could simply count how many times
+each clinician diagnoses carcinoma:
+
+```{r,label=countcarcinoma}
+count <- colSums(sweep(carcinoma_table[,1:7],1,carcinoma_table$n,"*"))
+names(count) <- colnames(carcinoma_table)[1:7]
+count <- count/sum(count)
+count
+pie(count)
+carcinoma_count <- count
+```
+
+We can compare this to other reasonable starting points:
+
+```{r,label=calccloglikcount,cache=TRUE}
+system.time(lc <- loglik_lsl(count,carcinoma))
+```
 
@@
 names(z) <- LETTERS[1:7]
-system.time(lz <- loglik_lsl(z,HL))
-lz
+system.time(lz <- loglik_lsl(z,carcinoma))
 ```
@@
 e <- equalp(W)
-system.time(le <- loglik_lsl(e,HL))
-le
+system.time(le <- loglik_lsl(e,carcinoma))
 ```
+
+```{r,label=showthreemethods}
+c(count=lc, zipf=lz, equal=le)
+```
+
+To find the maximum likelihood estimate we would use:
+
+`maxp_lsl(carcinoma,startp=indep(carcinoma_count), control=list(trace=100))`
+
+but this will take many hours (a week?) to run.
 
@@
 
-```{r}
-carcinoma <- 5
-
 ```{r,label=savecarcinomadataset}
-save(carcinoma_table,carcinoma, file="carcinoma.rda")
+save(carcinoma_table,carcinoma, carcinoma_count, file="carcinoma.rda")
 ```

### 2022-05-15 inst/carcinoma.Rmd
more work on maximum likelihood in carcinoma

@@
 
-To maximize the likelihood we need a good starting point and rather
-than use the default `equalp()` we could simply count how many times
-each clinician diagnoses carcinoma:
+There are several natural BT strengths to consider.  First, we could
+simply count how many times each clinician diagnoses carcinoma:
 
-```{r,label=countcarcinoma}
-count <- colSums(sweep(carcinoma_table[,1:7],1,carcinoma_table$n,"*"))
-names(count) <- colnames(carcinoma_table)[1:7]
-count <- count/sum(count)
-count
-pie(count)
-carcinoma_count <- count
+```{r,label=countcarcinoma,cache=TRUE}
+carcinoma_count <- colSums(sweep(carcinoma_table[,1:7],1,carcinoma_table$n,"*"))
+names(carcinoma_count) <- colnames(carcinoma_table)[1:7]
+carcinoma_count <- carcinoma_count/sum(carcinoma_count)
+carcinoma_count
+system.time(lc <- loglik_lsl(carcinoma_count,carcinoma))
 ```
 
-We can compare this to other reasonable starting points:
-
-```{r,label=calccloglikcount,cache=TRUE}
-system.time(lc <- loglik_lsl(count,carcinoma))
-```
+Secondly, Zipf:
 
@@
 
-```{r,label-calcloglikequalp,cache=TRUE}
+Thirdly, equal probabilities:
+
+```{r,label=calcloglikequalp,cache=TRUE}
 e <- equalp(W)
@@
 
-```{r,label=showthreemethods}
-c(count=lc, zipf=lz, equal=le)
-```
-
-To find the maximum likelihood estimate we would use:
+The maximum likelihood estimate is somewhat more demanding.  To find
+the maximum likelihood estimate we would use:
 
@@
 
-but this will take many hours (a week?) to run.
+but this takes many hours to run.  It gives the following:
+
+```{r,label=calculateloglikformaxp, cache=TRUE}
+carcinoma_maxp <- 
+c(A = 0.118778759012282, B = 0.542873969223425, C = 0.00654044305270073, 
+D = 0.00242693714895262, E = 0.202821434317779, F = 0.00130596252816986, 
+G = 0.125252494716691)
+system.time(lmax <- loglik_lsl(carcinoma_maxp,carcinoma))
+```
+
+We are now in a position to compare  these three points:
+
+```{r,label=showthreemethods}
+results <- c(count=lc, zipf=lz, equal=le,maxp=lmax)
+results <- results - max(results)
+results
+```
 
@@
 ```{r,label=savecarcinomadataset}
-save(carcinoma_table,carcinoma, carcinoma_count, file="carcinoma.rda")
+save(carcinoma_table,carcinoma, carcinoma_count, carcinoma_maxp,file="carcinoma.rda")
 ```

### 2022-05-16 inst/carcinoma.Rmd
minor rewriting

@@
 
+(takes maybe twenty minutes to run without cache.  The value of
+`carcinoma_maxp` is calculated separately [discussed below], and that
+takes a few days to run).
+
 This is a short analysis of the carcinoma dataset presented by
@@
 \begin{equation}\tag{1}
-i^+,j^-\frac{p_i}{p_i+p_j},\qquad i^-,j^+\frac{p_j}{p_i+p_j}
+\operatorname{Prob}(i^+,j^-) = \frac{p_i}{p_i+p_j},\qquad
+\operatorname{Prob}(i^-,j^+) = \frac{p_j}{p_i+p_j}
 \end{equation}
 
-Operationally the way we implement this in `hyper2` idiom is to posit
-an (unobservable) ranking on the clinicians: we place the clinicians
-in order from most convinced of the presence of carcinoma to the least
-convinced.  Thus a ranking of `DACBEGF` means that clinician `D` was
-most convinced of the presense of carcinoma and `F` least convinced.
-Given this, and given a single slide, our actual observation might be
-that `D` and `A` diagnosed presence, and the others diagosed absence,
-of carcinoma.  The corresponding entry would be
+(here superscripts $+$ and $-$ refer to a positive and negative
+diagnosis respectively).  Operationally the way we implement this in
+`hyper2` idiom is to posit an (unobservable) ranking on the
+clinicians: we place the clinicians in order from most convinced of
+the presence of carcinoma to the least convinced.  Thus a ranking of
+`DACBEGF` means that clinician `D` was most convinced of the presense
+of carcinoma and `F` least convinced.  Given this, and given a single
+slide, our actual observation might be that `D` and `A` diagnosed
+presence, and the others diagosed absence, of carcinoma.  The
+corresponding entry would be
 
@@
       positive <- names(jj)[jj==1]
-      print(negative)
-      print(positive)
       Wi[[i]] <- ggrl(W,positive,negative)
@@
 carcinoma_count
-system.time(lc <- loglik_lsl(carcinoma_count,carcinoma))
+lc <- loglik_lsl(carcinoma_count,carcinoma)
 ```
@@
 names(z) <- LETTERS[1:7]
-system.time(lz <- loglik_lsl(z,carcinoma))
+z
+lz <- loglik_lsl(z,carcinoma)  # takes about 30 seconds to run
 ```
@@
 e <- equalp(W)
-system.time(le <- loglik_lsl(e,carcinoma))
+e
+le <- loglik_lsl(e,carcinoma) # takes about 30 seconds to run
 ```
@@
 G = 0.125252494716691)
-system.time(lmax <- loglik_lsl(carcinoma_maxp,carcinoma))
+carcinoma_maxp
+lmax <- loglik_lsl(carcinoma_maxp,carcinoma)
 ```
 
-We are now in a position to compare  these three points:
+[of course, if `carcinoma_maxp` is available, we would use this as a
+starting point, viz
+`maxp_lsl(carcinoma,startp=indep(carcinoma_maxp))`].  With this, we
+are now in a position to compare these four points:
 
 ```{r,label=showthreemethods}
-results <- c(count=lc, zipf=lz, equal=le,maxp=lmax)
+(results <- c(count=lc, zipf=lz, equal=le,maxp=lmax))
 results <- results - max(results)
 results
+```
+
+We can use Wilks here to assess the null of `carcinoma_count`.  We
+have $-2\log(\Lambda/\Lambda_0)\sim\chi^2_6$, giving us a $p$-value of
+
+```{r,label=calculatechisquaredpvalueforcount}
+pchisq(-2*results[1],df=6,lower.tail=FALSE)
+```
+
+What is the difference between `carcinoma_count` and `carcinoma_maxp`?
+
+```{r}
+carcinoma_count
+carcinoma_maxp
+ordertransplot(carcinoma_count,carcinoma_maxp)
+ordertransplot(log(carcinoma_count),log(carcinoma_maxp),plotlims=c(-8,0))
 ```

### 2022-05-16 inst/hepatitis.Rmd
fixes #130

---
title: "Hepatitis"
output: bookdown::html_document2
---

```{r setup, include=FALSE}
set.seed(0)
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("magrittr")
options("digits" = 5)
```

<p style="text-align: right;">
![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
</p>

```{r loadlib}
hepatitis_table <- read.table("hepatitis.txt",header=TRUE)
hepatitis_table[,1:3] <- hepatitis_table[,1:3] > 0
wanted <- !(apply(hepatitis_table[,1:3],1,sum) %in% c(0,3))
hepatitis_table <- hepatitis_table[wanted,]
hepatitis_table
```

```{r makeggrllist,cache=TRUE}
W <- hyper2(pnames=c("P","Q","E"))
Wi <- list()
for(i in seq_len(nrow(hepatitis_table))){
      jj <- unlist(hepatitis_table[i,1:3])
      negative <- names(jj)[jj==0]
      positive <- names(jj)[jj==1]
      Wi[[i]] <- ggrl(W,positive,negative)
}
hepatitis <- lsl(Wi,powers= hepatitis_table$OC)
```

```{r}
hepatitis
```

There are several natural BT strengths to consider.  First, we could
simply count how many times each clinician gives a positive diagnosis:

```{r,label=counthep,cache=TRUE}
hepatitis_count <- colSums(sweep(hepatitis_table[,1:3],1,hepatitis_table$OC,"*"))
names(hepatitis_count) <- colnames(hepatitis_table)[1:3]
hepatitis_count <- hepatitis_count/sum(hepatitis_count)
hepatitis_count
```

Secondly, Zipf:

```{r,label=calccloglikzipf,cache=TRUE}
z <- zipf(3)
names(z) <- c("P","Q","E")
z
```

Thirdly, equal probabilities:

```{r,label=calcloglikequalp,cache=TRUE}
e <- equalp(W)
e
```

To find the maximum likelihood estimate we would use:

```{r,label=calculatehepmaxl,cache=TRUE}
hepatitis_maxp <- maxp_lsl(hepatitis,startp=indep(hepatitis_count), control=list(trace=110))
```

```{r,label=calculateloglikformaxp, cache=TRUE}
hepatitis_maxp
```

With this, we are now in a position to compare these four points:

```{r,label=showthreemethods}
lz <- loglik_lsl(z,hepatitis) 
le <- loglik_lsl(e,hepatitis)
lm <- loglik_lsl(hepatitis_maxp,hepatitis)
lc <- loglik_lsl(hepatitis_count,hepatitis)
(results <- c(zipf=lz, equal=le, maxlike=lm, count=lc))
results <- results - max(results)
results
```

We can use Wilks here to assess the null of `hepatitis_count`.  We
have $-2\log(\Lambda/\Lambda_0)\sim\chi^2_6$, giving us a $p$-value of

```{r,label=calculatechisquaredpvalueforcount}
pchisq(-2*results[4],df=2,lower.tail=FALSE)
```

Agresti, table 13.1, p542.

### Package dataset {-}

Following lines create `hepatitis.rda`, residing in the `data/`
directory of the package.

```{r,label=savehepatitisdataset}
save(hepatitis_table, hepatitis, hepatitis_count, hepatitis_maxp, file="hepatitis.rda")
```

### 2022-05-16 inst/carcinoma.Rmd
fixes #135

@@
 Agresti, table 13.1, p542.  Seven pathologists classified each of 118
-slides on the presence or absence of carcinoma in the uterine
-cervix.
+slides on the presence or absence of carcinoma in the uterine cervix.
+
+Here, the object of inference is the set of pathologists.  Informative
 
@@
 carcinoma_table[,1:7] <- carcinoma_table[,1:7] > 0
+carcinoma_table
+```
+
+```{r makeshorttable}
 wanted <- !(apply(carcinoma_table[,1:7],1,sum) %in% c(0,7))
-carcinoma_table <- carcinoma_table[wanted,]
-carcinoma_table
+carcinoma_table_short <- carcinoma_table[wanted,]
 ```
@@
 
-for(i in seq_len(nrow(carcinoma_table))){
-      jj <- unlist(carcinoma_table[i,1:7])
+for(i in seq_len(nrow(carcinoma_table_short))){
+      jj <- unlist(carcinoma_table_short[i,1:7])
       negative <- names(jj)[jj==0]
@@
 }
-carcinoma <- lsl(Wi,powers= carcinoma_table$n)
+carcinoma <- lsl(Wi,powers= carcinoma_table_short$n)
 ```
@@
 ```{r,label=countcarcinoma,cache=TRUE}
-carcinoma_count <- colSums(sweep(carcinoma_table[,1:7],1,carcinoma_table$n,"*"))
-names(carcinoma_count) <- colnames(carcinoma_table)[1:7]
+carcinoma_count <- colSums(sweep(carcinoma_table_short[,1:7],1,carcinoma_table_short$n,"*"))
+names(carcinoma_count) <- colnames(carcinoma_table_short)[1:7]
 carcinoma_count <- carcinoma_count/sum(carcinoma_count)

### 2022-05-17 inst/plackett_luce_monster.Rmd
start work on tethered runners

---
title: "plackett_luce_monster"
output: html_document
date: '2022-05-16'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("partitions")
```

We will calculate a Plackett-Luce likelihood function for four runners 1,2,3,4  but with a reified
Bradley-Terry monster that corresponds
to having 1 and 2 tethered by an elastic string and 3 and 4 also tethered by an elastic string.  The string 
acts to bring the two runners together and makes it more likely that such attached runners will place consecutively.

 first do three: 1 and 2 tethered,  3 not tethered to anyone

 123: $\frac{1}{1+2+3}\cdot\frac{2+M}{2+3+M}$

 132: $\frac{1}{1+3+2}\cdot\frac{3}{3+2+M}$

 213: $\frac{2}{2+1+3}\cdot\frac{1+M}{1+3+M}$

 231: $\frac{2}{2+3+1}\cdot\frac{3}{3+1+M}$

 312: $\frac{3}{3+1+2}\cdot\frac{1}{2+1}$

 321: $\frac{3}{3+2+1}\cdot\frac{2}{2+1}$

new stuff here

new stuff here

now 1 and 2 tethered, 3 and 4 tethered.

 1234: $\frac{1}{1+2+3+4}\cdot\frac{2+M}{2+3+4+M}\cdot\frac{3}{3+4}$

 1243: $\frac{1}{1+2+4+3}\cdot\frac{2+M}{2+4+3+M}\cdot\frac{4}{4+3}$

 1324: $\frac{1}{1+3+2+4}\cdot\frac{3}{3+2+4+M}\cdot\frac{2}{2+4+M}$

 1342: $\frac{1}{1+3+4+2}\cdot\frac{3}{3+4+2+M}\cdot\frac{4+M}{2+4+M}$

 1423: $\frac{1}{1+4+2+3}\cdot\frac{4}{4+2+3+M}\cdot\frac{2}{2+3}$

 1432: $\frac{1}{1+4+3+2}\cdot\frac{4}{4+3+2+M}\cdot\frac{3}{2+3}$

 2134: $\frac{2}{2+1+3+4}\cdot\frac{1+M}{1+3+4+M}\cdot\frac{3}{3+4}$

 2143: $\frac{2}{2+1+4+3}\cdot\frac{1+M}{1+4+3+M}\cdot\frac{4}{4+3}$

 2314: $\frac{2}{2+3+1+4}\cdot\frac{3}{3+1+4+M}\cdot\frac{1}{1+4+M}$

 2341: $\frac{2}{2+3+4+1}\cdot\frac{3}{3+4+1+M}\cdot\frac{4+M}{1+4+M}$

 2413: $\frac{2}{2+4+1+3}\cdot\frac{4}{4+1+3+M}\cdot\frac{1}{1+3}$

 2431: $\frac{2}{2+4+3+1}\cdot\frac{4}{4+3+1+M}\cdot\frac{3}{1+3}$

etc.

Now 123 tethered, 45 tethered:

12345: $\frac{1}{1+2+3+4+5}\cdot\frac{2+M}{2+3+4+5+M}\cdot\frac{3+M}{3+4+5+M}\cdot\frac{4}{4+5}$

12354: $\frac{1}{1+2+3+5+4}\cdot\frac{2+M}{2+3+5+4+M}\cdot\frac{3+M}{3+5+4+M}\cdot\frac{5}{5+4}$

12435: $\frac{1}{1+2+3+5+4}\cdot\frac{2+M}{2+3+5+4+M}\cdot\frac{4}{3+5+4+M}\cdot\frac{3}{3+5+M}$

12453: $\frac{1}{1+2+3+5+4}\cdot\frac{2+M}{2+3+5+4+M}\cdot\frac{4}{3+5+4+M}\cdot\frac{5+M}{5+3+M}$

12534: $\frac{1}{1+2+3+5+4}\cdot\frac{2+M}{2+5+3+4+M}\cdot\frac{5}{3+5+4+M}\cdot\frac{3}{3+4+M}$

12543: $\frac{1}{1+2+3+5+4}\cdot\frac{2+M}{2+5+4+3+M}\cdot\frac{5}{3+5+4+M}\cdot\frac{4+M}{4+3+M}$

Might be better to consider ${5\choose 2\,3}$ arrangements of $aaabb\ldots bbaaa$.

aaabb: $\frac{a}{a+a+a+b+b}\cdot\frac{a+M}{a+a+b+b+M}\cdot\frac{a+M}{a+b+b+M}\cdot\frac{b}{b+b+M}$

aabab: $\frac{a}{a+a+a+b+b}\cdot\frac{a+M}{a+a+b+b+M}\cdot\frac{b}{a+b+b+M}\cdot\frac{a}{a+b+M}$

aabba: $\frac{a}{a+a+a+b+b}\cdot\frac{a+M}{a+a+b+b+M}\cdot\frac{b}{a+b+b+M}\cdot\frac{b+M}{b+a+M}$

abaab: $\frac{a}{a+a+a+b+b}\cdot\frac{b}{a+a+b+b+M}\cdot\frac{a}{a+a+b+M}\cdot\frac{a+M}{a+b+M}$

ababa: $\frac{a}{a+a+a+b+b}\cdot\frac{b}{a+a+b+b+M}\cdot\frac{a}{a+a+b+M}\cdot\frac{b}{b+a+M}$

abbaa: $\frac{a}{a+a+a+b+b}\cdot\frac{b}{a+a+b+b+M}\cdot\frac{b+M}{b+a+a+M}\cdot\frac{a}{a+a+M}$

baaab: $\frac{b}{a+a+a+b+b}\cdot\frac{a}{a+a+a+b+M}\cdot\frac{a+M}{a+a+b+M}\cdot\frac{a+M}{a+b+M}$

baaba: $\frac{b}{a+a+a+b+b}\cdot\frac{a}{a+a+a+b+M}\cdot\frac{a+M}{a+a+b+M}\cdot\frac{b}{a+b+M}$

babaa: $\frac{b}{a+a+a+b+b}\cdot\frac{a}{a+a+a+b+M}\cdot\frac{b}{b+a+a+M}\cdot\frac{a}{a+a}$

bbaaa: $\frac{b}{a+a+a+b+b}\cdot\frac{b+M}{a+a+a+b+M}\cdot\frac{a}{a+a+a}\cdot\frac{a}{a+a}$

```{r,label=definefunctionf}
f <- function (a,b,M){
return(factorial(3)*factorial(2)*(
(a)/(a+a+a+b+b)*(a+M)/(a+a+b+b+M)*(a+M)/(a+b+b+M)*(b)/(b+b+M)+
(a)/(a+a+a+b+b)*(a+M)/(a+a+b+b+M)*(b)/(a+b+b+M)*(a)/(a+b+M)+
(a)/(a+a+a+b+b)*(a+M)/(a+a+b+b+M)*(b)/(a+b+b+M)*(b+M)/(b+a+M)+
(a)/(a+a+a+b+b)*(b)  /(a+a+b+b+M)*(a)/(a+a+b+M)*(a+M)/(a+b+M)+
(a)/(a+a+a+b+b)*(b)  /(a+a+b+b+M)*(a)/(a+a+b+M)*(b)/(b+a+M)+
(a)/(a+a+a+b+b)*(b)  /(a+a+b+b+M)*(b+M)/(b+a+a+M)*(a)/(a+a+M)+
(b)/(a+a+a+b+b)*(a)  /(a+a+a+b+M)*(a+M)/(a+a+b+M)*(a+M)/(a+b+M)+
(b)/(a+a+a+b+b)*(a)  /(a+a+a+b+M)*(a+M)/(a+a+b+M)*(b)/(a+b+M)+
(b)/(a+a+a+b+b)*(a)  /(a+a+a+b+M)*(b)/(b+a+a)*(a)/(a+a)+
(b)/(a+a+a+b+b)*(b+M)/(a+a+a+b+M)*(a)/(a+a+a)*(a)/(a+a)))
}
```

```{r}
f(4.4,3.3, 0.2)*12
f(4.4,3.3, 0.0)*12
```

### 2022-05-20 inst/rock_paper_scissors_monster.Rmd
typo

@@
 This gives maximum likelihood estimates of $M=0.18$ as would be
-expected from the simple binomia model; the probability of a procyclic
-win is about $0.625$.
+expected from the simple binomial model; the probability of a
+procyclic win is about $0.625$.
 

### 2022-05-21 inst/race3.Rmd
more algebra

@@
 
+At the evaluate of $\hat{a}=\sqrt{2}-1$ we can show that
+$\log\mathcal{l}(\hat{a})=\log\left(3-2\sqrt{2}\right)\simeq -1.76$.
+Defining $\mathcal{S}=\log\mathcal{L}$ as the support [log-likelihood]
+we have
+
+$$\mathcal{S}=\mathcal{S}(a)=\log\left(\frac{a(1-a)}{1+a}\right)-\log\left(3-2\sqrt{2}\right)$$
+
+as a standard support function which has a maximum value of zero.
+Further, we can use either analytical or numerical methods to find a
+credible interval.  With an $n$-unit of support criterion the
+analytical solution to $\mathcal{S}(p)=-2$ is given by defining
+$X=\log(3-2\sqrt{2})-n$ and solving $p(1-p)/(1+p)=X, or
+$\left(1-X\pm\sqrt{1+4X+X^2}\right)/2$, the two roots being the lower
+and upper limits of the credible interval.  Graphically:
+
+```{r}
+a <- seq(from=0,by=0.005,to=1)
+S <- function(a){log(a*(1-a)/((1+a)*(3-2*sqrt(2))))}
+plot(a,S(a),type='b')
+abline(h=c(0,-2))
+abline(v=c(0.02438102,0.9524271),col='red')
+abline(v=sqrt(2)-1)
+```
+
 With four competitors we have five nontrivial order statistics which,
@@
 
-# Some simple cases: Fisher information for simple cases
+# Fisher information for simple cases for aab
 
-(work in progress).
+We have three possible finishing orders:
+
+$b\succ a\succ a$, probability $\frac{b}{2a+b}=\frac{1-a}{1+a}$.  And
+$\frac{\partial\log\mathcal{L}}{\partial a}=\frac{2}{p^2-1}$.
+
+$a\succ b\succ a$, probability $\frac{a}{2a+b}\cdot\frac{b}{a+b}=\frac{a(1-a)}{1+a}$. And
+$\frac{\partial\log\mathcal{L}}{\partial a}=\frac{a^2+2a-1}{a^3-a}$.
+
+$a\succ a\succ b$, probability $\frac{a}{2a+b}\cdot\frac{a}{a+b}=\frac{a^2}{1+a}$.  And
+$\frac{\partial\log\mathcal{L}}{\partial a}=\frac{a+2}{a^2+a}$.
+
+Thus the Fisher information is
+
+$$
+\frac{1-a}{1+a}\cdot\left(\frac{2}{a^2-1}\right)^2
++
+\frac{a(1-a)}{1+a}\cdot\left(\frac{a^2+2a-1}{a^3-a}\right)^2
++\frac{a^2}{1+a}\cdot\left(\frac{a+2}{a^2+a}\right)^2
+$$
+
+Simplifying,
+
+$$\mathcal{I}(a)=\frac{1+16a+18a^2-4a^3-15a^4-7a^5-a^6}{a(1-a)(1+a)^3}$$
 

### 2022-05-22 inst/race3.Rmd
more work and algebra

@@
 With four competitors we have five nontrivial order statistics which,
-without loss of generality, are $(a,b,a,a)$ and $(a,a,b,a)$.
-$(a,b,a,b)$, $(a,b,b,a)$, $(a,b,c,a)$. Taking $(a,b,a,b)$ as an
-example we have
+without loss of generality, are:
+
+* $a\succ b\succ a\succ a$
+* $a\succ a\succ b\succ a$
+* $a\succ b\succ a\succ b$
+* $a\succ b\succ b\succ a$
+* $a\succ b\succ c\succ a$.
+
+Taking $a\succ b\succ a\succ b$ as an example we have
 
 $$
-\mathcal{L}(a)=
+\mathcal{L}_{a\succ b\succ a\succ b}(a)=
 \frac{a}{2a+2b}\cdot
@@
 If we allow non-finishers there is another nontrivial order statistic,
-viz $\left(a,b,\left\lbrace a,b\right\rbrace\right)$:
+viz $a\succ b\succ\left\lbrace a,b\right\rbrace$ [thus one of the two
+$a$'s won, one of the $b$'s came second, and one of each of $a$ and
+$b$ failed to finish].
 
@@
 $$\mathcal{I}(a)=\frac{1+16a+18a^2-4a^3-15a^4-7a^5-a^6}{a(1-a)(1+a)^3}$$
+
+```{r}
+fisher <- function(a){(1+16*a+18*a^2-4*a^3-15*a^4-7*a^5-a^6)/(a*(1-a)*(1+a)^3)}
+a <- seq(from=0.05,to=0.95,by=0.01)
+plot(a,fisher(a))
+```
 

### 2022-05-22 inst/race3.Rmd
clarification and test H0: p=0.5

@@
 above, we see close agreement with the theoretical value of
-$(\sqrt{2}-1,1-\sqrt{2})\simeq (0.414,0.586)$
-
-At the evaluate of $\hat{a}=\sqrt{2}-1$ we can show that
-$\log\mathcal{l}(\hat{a})=\log\left(3-2\sqrt{2}\right)\simeq -1.76$.
+$(\sqrt{2}-1,1-\sqrt{2})\simeq (0.414,0.586)$.  Observe that the
+maximum likelihood estimate for $a$ is strictly less than 0.5, even
+though the finishing order is symmetric.  Using
+$\mathcal{L}(a)=\frac{a(1-a)}{1+a}$, we can show that
+$\log\mathcal{L}(\hat{a})=\log\left(3-2\sqrt{2}\right)\simeq -1.76$,
+where $\hat{a}=\sqrt{2}-1$ is the maximum likelihood estimate for $a$.
 Defining $\mathcal{S}=\log\mathcal{L}$ as the support [log-likelihood]
@@
 
-as a standard support function which has a maximum value of zero.
+as a standard support function which has a maximum value of zero when
+evaluated at $\hat{a]=\sqrt{2}-1$.  For example, we can test the null
+that $a=b=\frac{1}{2}$, the statement that the competitors have equal
+Bradley-Terry strengths:
+
+```{r}
+a <- 1/2 # null
+(S_delta <- log( a*(1-a)/(1+a))- log(3-2*sqrt(2)))
+```
+
+Thus the additional support gained in moving from $a=\frac{1}{2}$ to
+the evaluate of $a=\sqrt{2}-1$ is 0.029, rather small [as might be
+expected given that we have only one rather uninformative observation,
+and also given that the maximum likelihood estimate ($\simeq 0.41$) is
+quite close to the null ($0.5$).  Nevertheless we can apply Wilks's
+theorem for a $p$-value:
+
+```{r}
+pchisq(-2*S_delta,df=1,lower.tail=FALSE)
+```
+
+The $p$-value is about 0.81, exceeding 0.05; thus we have no strong
+evidence to reject the null of $a=\frac{1}{2}.
+
 Further, we can use either analytical or numerical methods to find a
@@
 analytical solution to $\mathcal{S}(p)=-2$ is given by defining
-$X=\log(3-2\sqrt{2})-n$ and solving $p(1-p)/(1+p)=X, or
-$\left(1-X\pm\sqrt{1+4X+X^2}\right)/2$, the two roots being the lower
-and upper limits of the credible interval.  Graphically:
+$X=\log(3-2\sqrt{2})-n$ and solving $p(1-p)/(1+p)=X$, or
+$p_\pm=\left(1-X\pm\sqrt{1+4X+X^2}\right)/2$, the two roots being the
+lower and upper limits of the credible interval.  Graphically:
 

### 2022-05-27 inst/plackett_luce_monster.Rmd
change from elastic to cheering

@@
 
-We will calculate a Plackett-Luce likelihood function for four runners 1,2,3,4  but with a reified
-Bradley-Terry monster that corresponds
-to having 1 and 2 tethered by an elastic string and 3 and 4 also tethered by an elastic string.  The string 
-acts to bring the two runners together and makes it more likely that such attached runners will place consecutively.
+We will calculate a Plackett-Luce likelihood function for four runners
+$\left\lbrace 1,2,3,4\right\rbrace$ but with a reified Bradley-Terry
+monster that corresponds to having subsets of mutual friends who
+support one another.  For example, we might have runners 1 and 2 being
+friends, and runners 3 and 4 being friends: we would represent this as
+$\left\lbrace\left\lbrace 1,2\right\rbrace,\left\lbrace
+3,4\right\rbrace\right\rbrace$.
 
- first do three: 1 and 2 tethered,  3 not tethered to anyone
+The statistical model is as follows.  If a runner finishes, then he
+cheers on all his friends who are still running, thereby adding to
+their strength.  This is a very simple form of non-independence.
 
- 123: $\frac{1}{1+2+3}\cdot\frac{2+M}{2+3+M}$
+First do three runners, and suppose the equivalence classes are
+$\left\lbrace\left\lbrace 1,2\right\rbrace,\left\lbrace
+3\right\rbrace\right\rbrace$.
 
- 132: $\frac{1}{1+3+2}\cdot\frac{3}{3+2+M}$
+ 123: $\frac{p_1}{p_1+p_2+p_3}\cdot\frac{p_2+M}{p_2+p_3+M}\qquad\frac{p_1}{p_1+p_2+p_3}\cdot\frac{\lambda p_2}{\lambda p_2+p_3}$
 
- 213: $\frac{2}{2+1+3}\cdot\frac{1+M}{1+3+M}$
+ 132: $\frac{p_1}{p_1+p_3+p_2}\cdot\frac{p_3}{p_3+p_2+M}\qquad\frac{p_1}{p_1+p_3+p_2}\cdot\frac{p_3}{p_3+\lambda p_2}$
 
- 231: $\frac{2}{2+3+1}\cdot\frac{3}{3+1+M}$
+ 213: $\frac{p_2}{p_2+p_1+p_3}\cdot\frac{p_1+M}{p_1+p_3+M}\qquad\frac{p_2}{p_2+p_1+p_3}\cdot\frac{\lambda p_1}{\lambda p_1+p_3}$
 
- 312: $\frac{3}{3+1+2}\cdot\frac{1}{2+1}$
+ 231: $\frac{p_2}{p_2+p_3+p_1}\cdot\frac{p_3}{p_3+p_1+M}$
 
- 321: $\frac{3}{3+2+1}\cdot\frac{2}{2+1}$
+ 312: $\frac{p_3}{p_3+p_1+p_2}\cdot\frac{p_1}{p_2+p_1}$
 
-new stuff here
+ 321: $\frac{p_3}{p_3+p_2+p_1}\cdot\frac{p_2}{p_2+p_1}$
 
-new stuff here
-
-now 1 and 2 tethered, 3 and 4 tethered.
+Now four runners with classes $\left\lbrace\left\lbrace
+1,2\right\rbrace,\left\lbrace 3,4\right\rbrace\right\rbrace$.
 

### 2022-05-30 inst/kka.Rmd
chess3

@@
     H <- white_draw3(kka_array, lambda,D)
-    max(maxp(H,justlikes=TRUE))
+    max(maxp(H,n=1,justlikes=TRUE))
 }	
@@
 
+Optimize:
+
+```{r,label=useoptimopencloseforf,cache=TRUE}
+(o <- optim(par=c(2.1,2.4),fn=f,control=list(fnscale = -1)))
+```
+
+```{r,createchess3}
+(chess3 <- f(o$par))
+```
+
 ### Package dataset
@@
 ```{r,label=savecurlingdataset,cache=TRUE}
-save(kka,karpov_kasparov_anand,kka_3draws,kka_3whites,kka_array,file="kka.rda")
+save(kka,karpov_kasparov_anand,kka_3draws,kka_3whites,kka_array,chess3,file="kka.rda")
 ```

### 2022-05-30 inst/kka.Rmd
bugfix

@@
 ```{r,createchess3}
-(chess3 <- f(o$par))
+jj <- round(o$par,2)
+(chess3 <-  white_draw3(kka_array, lambda=jj[1],D=jj[2]))
 ```
@@
 
-```{r,label=savecurlingdataset,cache=TRUE}
+```{r,label=savecurlingdataset,cache=FALSE}
 save(kka,karpov_kasparov_anand,kka_3draws,kka_3whites,kka_array,chess3,file="kka.rda")

### 2022-05-30 inst/kka.Rmd
chess3_maxp

--- 2022-05-30 inst/kka.Rmd
@@
 ```{r,createchess3}
-(chess3 <- f(o$par))
+jj <- round(o$par,2)
+(chess3 <-  white_draw3(kka_array, lambda=jj[1],D=jj[2]))
+(chess3_maxp <-  maxp(chess3))
 ```
@@
 
-```{r,label=savecurlingdataset,cache=TRUE}
-save(kka,karpov_kasparov_anand,kka_3draws,kka_3whites,kka_array,chess3,file="kka.rda")
+```{r,label=savecurlingdataset,cache=FALSE}
+save(kka,karpov_kasparov_anand,kka_3draws,kka_3whites,kka_array,chess3,chess3_maxp,file="kka.rda")
 ```

### 2022-05-31 inst/kka.Rmd
update text

@@
 ---
-title: "Karpov, Kasprov, Anand"
+title: "Karpov, Kasparov, Anand"
 output: html_document
@@
 
+(takes about 45 minutes to process without cache)
+
 <p style="text-align: right;">
@@
 This file defines dataset `kka` which is the observed scorelines
-between the three players.  It is used to calculate likelihood
-functions: `karpov_kasparov_anand`, `kka_3whites` and `kka_3draws`
-This Rmd file supercedes files `inst/karpov_kasparov_anand.R` and
-`inst/kka_3draws.R` and `inst/kka_array.R`.  Some more detail is given
-in `man/karpov_kasparov_anand.Rd`.
+between players Karpov, Kasparov, and Anand.  It is used to calculate
+likelihood functions: `karpov_kasparov_anand`, `kka_3whites` and
+`kka_3draws`.  This Rmd file supercedes files
+`inst/karpov_kasparov_anand.R` and `inst/kka_3draws.R` and
+`inst/kka_array.R`.  Some more detail is given in
+`man/kka.Rd`.
 
@@
 
-for the observation of $+a\,=0\,-b$ for $p_1$ playing white and
-$+c\,=0\,-d$ for $p_2$ playing white.  We then translate this into a
-`hyper3` likelihood function as follows
+for the observation of $+a$ $\text{=}0$ $-b$ for $p_1$ playing white
+and $+c$ $\text{=}0$ $-d$ for $p_2$ playing white.  We then translate
+this into a `hyper3` likelihood function as follows
 
@@
 Above we see a credible interval for $\lambda$ being about
-$(1.4,2.9)$.  Of course, this is crude and it would be betterto have
+$(1.4,2.9)$.  Of course, this is crude and it would be better to have
 more transparent package idiom.  But I'm working on this.
@@
 
-```{r,createchess3}
+```{r,createchess3,cache=TRUE}
 jj <- round(o$par,2)

### 2022-06-02 inst/universities.Rmd
minor changes

@@
 ---
-title: "universities"
+title: "university ranking data"
 output: bookdown::html_document2
@@
 
-## University ranking data
+<p style="text-align: right;">
+![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
+</p>
 
-First load the data:
+Here I analyse a dataset on NZ university rankings, taken from the
+Times Higher Educational Supplement.  First load the data:
 
-```{r cars}
+```{r loadthetable}
 universities_table <- read.table("universities.txt",header=TRUE)

### 2022-06-02 inst/NBA.Rmd
minor changes

@@
---
-title: "Basketball: an analysis of a single game"
+---
+title: "An analysis of a single basketball game"
 output: html_document
@@
   }
-  print(onpitch_scored)
-  print(onpitch_noscore)
   if(possession[i] == 'W'){

### 2022-06-20 inst/grad_weight.Rmd
typos
include sticker

@@
-<---
+---
 title: "grad_weight"

### 2022-06-20 inst/race3.Rmd
slight typos

@@
 
-(takes about an hour to process)
+(takes about an hour to process without cache)
+
+<p style="text-align: right;">
+![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
+</p>
 
@@
 as a standard support function which has a maximum value of zero when
-evaluated at $\hat{a]=\sqrt{2}-1$.  For example, we can test the null
+evaluated at $\hat{a}=\sqrt{2}-1$.  For example, we can test the null
 that $a=b=\frac{1}{2}$, the statement that the competitors have equal
@@
 The $p$-value is about 0.81, exceeding 0.05; thus we have no strong
-evidence to reject the null of $a=\frac{1}{2}.
-
-Further, we can use either analytical or numerical methods to find a
-credible interval.  With an $n$-unit of support criterion the
-analytical solution to $\mathcal{S}(p)=-2$ is given by defining
-$X=\log(3-2\sqrt{2})-n$ and solving $p(1-p)/(1+p)=X$, or
+evidence to reject the null of $a=\frac{1}{2}$.  Further, we can use
+either analytical or numerical methods to find a credible interval.
+With an $n$-unit of support criterion the analytical solution to
+$\mathcal{S}(p)=-2$ is given by defining $X=\log(3-2\sqrt{2})-n$ and
+solving $p(1-p)/(1+p)=X$, or
 $p_\pm=\left(1-X\pm\sqrt{1+4X+X^2}\right)/2$, the two roots being the

### 2022-06-23 inst/tennis.Rmd
consistentification

@@
----
+s---
 title: "Tennis doubles and ghosts"
@@
 
-```{r}
+```{r defineHhyper2}
 H <- hyper2()
@@
 
-```{r}
+```{r addtohyper2H}
 tennis <- H
@@
 
-```{r}
+```{r doghoststuff}
 tennis_ghost <- H
@@
 
-```{r}
+```{r maximizeghostlike,cache=TRUE}
 (tennis_maxp <- maxp(tennis))
@@
 
-```{r}
+```{r doequalptest,cache=TRUE}
 equalp.test(tennis)
@@
 
-```{r}
+```{r specificptesthere,cache=TRUE}
 specificp.test(tennis_ghost,"G",0.0001)  # nonzero value needed to avoid triggering known R bug
@@
 
-```{r}
+```{r hessianmatrix}
 (M <- hessian(tennis,indep(tennis_maxp),border=TRUE))
@@
 
-```{r}
+```{r eigenwork}
 eigen(M,TRUE,TRUE)$values # one value is zero to numerical precision; the rest should be <0
@@
 
-```{r,label=savecurlingdataset}
+```{r,label=saveghosttennisdataset}
 save(tennis,tennis_maxp,tennis_ghost, tennis_ghost_maxp,file="tennis.rda")

### 2022-06-23 inst/kka.Rmd

@@
 
-We can now introduce draws but first a change of notation.  The
-previous likelihood function can be represented thus:
+We can now introduce draws but first a change of notation.
+Considering repeated trials between entities $1,2,\ldots, r$ with
+Bradley-Terry strengths $p_1,\ldots, p_r$ and counts $n_1,\ldots, n_r$ a
+suitable likelihood function would be
+
+$$
+\left(\frac{p_1}{\sum p_i}\right)^{n_1}\cdot
+\left(\frac{p_2}{\sum p_i}\right)^{n_2}\ldots
+\left(\frac{p_r}{\sum p_i}\right)^{n_r}
+$$
+
+but we may represent this in the following compact form:
+
+$$
+\left(\frac{p_1,p_2,\ldots, p_r}{\sum p_i}\right)^{\left(n_1,\ldots n_r\right)}
+$$
+
+Or, noting that the sum of the numerators is equal to the common
+denominator, even more compactly as $\left(p_1,p_2,\ldots, p_r\right)^{\left(n_1,\ldots n_r\right)}$.
+
+The previous likelihood function can be represented thus:
 
@@
 \left(\frac{p_1, D_1p_1+D_2p_2),\lambda p_2}{p_1+D_1p_1+D_2p_2+\lambda p_2}\right)^{(d,e,f)}
+$$
+
+Package idiom for this would require the following form:
+
+$$
+\left(\frac{\lambda p_1, D_1p_1+D_2p_2,p_2}{(\lambda+D_1)p_1+(1+D_2)p_2}\right)^{(a,b,c)}
+\left(\frac{p_1, D_1p_1+D_2p_2,\lambda p_2}{(1+D_1)p_1+(\lambda+D_2)p_2}\right)^{(d,e,f)}
 $$

### 2022-06-25 inst/table_tennis_serve.Rmd
labels and cache

@@
 
-```{r,message=FALSE}
+```{r,label=defineH,message=FALSE}
 library("hyper2",quietly=TRUE)
@@
 ### A vs B, A serves, score 5-1:
-```{r}
+```{r,label=abaserve}
 H[c("a","S")]     %<>% "+"(5  ) # a+S win 5 games
@@
 ### A vs B, B serves, score 1-3:
-```{r}
+```{r,label=abbserve}
 H[c("b","S")]     %<>% "+"(3  ) # b+S win 3 games
@@
 ### A vs C, A serves, score 4-1:
-```{r}
+```{r,label=acaserve}
 H[c("a","S")]     %<>% "+"(4  ) # a+S win 4 games
@@
 ### A vs C, C serves, score 1-2:
-```{r}
+```{r,label=accserve}
 H["a"]            %<>% "+"(1  ) # a wins 1 game
@@
 
-```{r}
+```{r,savetime}
 table_tennis <- H
@@
 
-```{r}
+```{r,label=maxptennis,cache=TRUE}
 table_tennis_maxp <- maxp(table_tennis)
@@
 
-```{r testservezero}
+```{r, label=testservezero,cache=TRUE}
 specificp.test(table_tennis,"S",0.01)
@@
 
-```{r}
+```{r,label=maxtab,cache=TRUE}
 maxp(table_tennis)
@@
 
-```{r}
+```{r,label=svals,cache=TRUE}
 S_vals <- seq(from=0.02,to=0.85,len=30)  # possible values for S
@@
 
-```{r,dpi=72}
+```{r,label=proffli,dpi=72,cache=TRUE}
 proflike <- function(bc,give=FALSE){
@@
 
-```{r}
+```{r,label=logcont}
 logcontrast <- apply(jj,1,function(x){log(x[1]/x[2])})

### 2022-06-26 inst/race3.Rmd
another figure

--- 2022-05-22 inst/race3.Rmd
@@
 (takes about an hour to process)
+
+<p style="text-align: right;">
+![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
+</p>
 
@@
 
+## Three competitors
+
 Here I consider some order statistics with nontrivial maximum
@@
 ```
+
+### all results for AAB
+
+* $A\succ A\succ B\longrightarrow\frac{a^2}{1+a}$
+* $A\succ B\succ A\longrightarrow\frac{1-2a-a^2}{(1+a)^2}$
+* $B\succ A\succ A\longrightarrow\frac{1-a}{1+a}$
+
+Graphically
+
+```{r}
+f_aab <- function(a){a^2/(1+a)}
+f_aba <- function(a){a*(1-a)/(1+a)^2}
+f_baa <- function(a){(1-a)/(1+a)}
+
+p <- function(f,...){points(a,f(a)/max(f(a)),...)}
+plot(0:1,0:1,xlab="p(a)",ylab="Likelihood",type="n")
+p(f_aab,type="l",col="black")
+p(f_aba,type="l",col="red")
+p(f_baa,type="l",col="blue")
+legend(x=0.85,y=0.7,col=c("black","red","blue"),lty=1,pch=NA,legend=c("AAB","ABA","BAA"))
+abline(h=exp(-2),lty=2)
+```
+
+## Four competitors
 

### 2022-06-26 inst/race3.Rmd
consistent likelihood function

@@
 
+## Results for AAAB
+
+```{r,plotaaab,cache=TRUE}
+Hab <- function(i,n=4){
+  jj <- rep("a",n)
+  jj[i] <- "b"
+  ordervec2supp3(jj)
+}
+
+small <- 1e-5
+a <- seq(from=small,to=1-small,len=100)
+plot(0:1,c(0,1),type="n",xlab="p[a]",ylab="likelihood")
+dop <- function(w, ...){
+  Hjj <- Hab(w)
+  y <- sapply(a,function(p){loglik(c(a=p,b=1-p),Hjj,log=FALSE)})
+  points(a,y/max(y),type="l",...)
+}
+dop(1,col="black")
+dop(2,col="red")
+dop(3,col="blue")
+dop(4,col="green")
+
+```
+
 # Fisher information for simple cases for aab

### 2022-06-27 inst/rock_paper_scissors_monster.Rmd
hyper3 analysis

@@
 ```
+
+(takes about an hour to run without cache)
+
+<p style="text-align: right;">
+![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
+</p>
 
@@
 ```{r,label=maxpHWM,cache=TRUE}
-maxp(HWM)
+maxHW <- maxp(HWM)
+maxHW
 ```
@@
 
-```{r,label=specicifgtest,cache=TRUE}
+```{r,label=specificgtesthwm,cache=TRUE}
 specificp.gt.test(HWM,"M",0)
@@
 strengths of the teeth.
+
+### Analysis of seven-tooth gear with `hyper3`
+
+We can now use `hyper3` to analyse the same dataset.  Here the
+likelihood function for pairwise competition between $p_1$ and $p_2$,
+with $a$ procyclic wins and $b$ anticyclic wins, is
+
+$$
+\left(\frac{\lambda p_1}{\lambda p_1 + p_2}\right)^a
+\cdot
+\left(\frac{p_2}{\lambda p_1 + p_2}\right)^b
+$$
+
+where $\lambda$ is a quantification of the asymmetry.  We can modify
+the `hyper2` idiom to make a `hyper3` object instead:
+
+```{r,label=defineh3seven}
+H3seven <- function(lambda){
+  H3 <- hyper3()
+  for(i in 1:6){
+     jj <- 1  # diagonal players do not have lambda-advantage
+     names(jj) <- tt[i]
+     H3[jj] %<>% inc(hw[i,i])
+
+     jj <- lambda
+     names(jj) <- tt[i+1]
+     H3[jj] %<>% inc(hw[i,i+1])
+
+     jj <- c(1,lambda)
+     names(jj) <- c(tt[i],tt[i+1])
+     H3[jj] %<>% dec(hw[i,i] + hw[i,i+1])
+  }  # i loop closes
+
+  ## Now do final one:
+  jj <- 1
+  names(jj) <- tt[7]
+  H3[jj] %<>% inc(hw[7,7])
+
+  jj <- lambda
+  names(jj) <- tt[1]
+  H3[jj] %<>% inc(hw[7,1])
+
+  jj <- c(1,lambda)
+  names(jj) <- c(tt[7],tt[1])
+  H3[jj] %<>% dec(hw[7,7] + hw[7,1])
+
+  return(H3)
+}  # function H3seven() defn closes
+```
+
+Now we can execute it:
+
+```{r,label=executeh3seven}
+H3 <- H3seven(lambda=1.66)
+H3
+```
+
+And try a profile likelihood:
+
+```{r,label=defineproflikeh3}
+profl <- function(lambda){
+  H <- H3seven(lambda)
+  return(loglik(maxp(H),H))
+}
+```
+
+```{r,label=proflikh3,cache=TRUE}
+lam <- seq(from=0.8,to=2.6,len=15)
+like3 <- sapply(lam,profl)
+like3 <- like3-max(like3)
+```
+
+```{r,plotproflike3}
+plot(lam,like3,type='b')
+abline(h=c(0,-2))
+abline(v=1)
+```
 

### 2022-06-27 inst/race3.Rmd
more likelihood functions

--- 2022-06-26 inst/race3.Rmd
@@
 
+## Results for AAAB
+
+```{r,plotaaab,cache=TRUE}
+Hab <- function(i,n=4){
+  jj <- rep("a",n)
+  jj[i] <- "b"
+  ordervec2supp3(jj)
+}
+
+small <- 1e-5
+a <- seq(from=small,to=1-small,len=400)
+
+plot(0:1,c(0,1),type="n",xlab="p[a]",ylab="likelihood")
+dop <- function(w, n, ...){
+  Hjj <- Hab(w,n)
+  y <- sapply(a,function(p){loglik(c(a=p,b=1-p),Hjj,log=FALSE)})
+  points(a,y/max(y),type="l",...)
+}
+dop(1,4,col="black")
+dop(2,4,col="red")
+dop(3,4,col="blue")
+dop(4,4,col="green")
+```
+
+```{r,plotaaaaaaab,cache=TRUE}
+plot(0:1,c(0,1),type="n",xlab="p[a]",ylab="likelihood")
+for(i in 1:10){dop(i,10,col=rainbow(10)[i])}
+```
+
+```{r,plot40ab,cache=TRUE}
+plot(0:1,c(0,1),type="n",xlab="p[a]",ylab="likelihood")
+for(i in 1:40){dop(i,40,col=rainbow(40)[i])}
+```
+
 # Fisher information for simple cases for aab

### 2022-06-28 inst/rock_paper_scissors_monster.Rmd
more likelihood plots

@@
 
-And try a profile likelihood:
-
 ```{r,label=defineproflikeh3}
@@
 
+First optimize the likelihood:
+
+```{r,optlike,cache=TRUE}
+o <- optimize(profl,c(1.4,1.6),maximum=TRUE)
+```
+
+```{r}
+o
+```
+
+So now plot a profile likelihood curve:
+
 ```{r,label=proflikh3,cache=TRUE}
-lam <- seq(from=0.8,to=2.6,len=15)
+lam <- seq(from=0.8,to=2.6,len=30)
 like3 <- sapply(lam,profl)
-like3 <- like3-max(like3)
 ```
@@
 ```{r,plotproflike3}
-plot(lam,like3,type='b')
+plot(lam,like3-o$objective,type='b',ylab="likelihood",xlab="lambda")
 abline(h=c(0,-2))
-abline(v=1)
+abline(v=c(1,o$maximum))
 ```

### 2022-06-28 inst/tennis.Rmd

@@
-s---
+---
 title: "Tennis doubles and ghosts"
@@
 
+### Analysis with `hyper3` {-}
+
+The likelihood for the 9-2 scoreline between $\left\lbrace
+p1,p2\right\rbrace$ and $\left\lbrace p3,p4\right\rbrace$ would have a
+likelihood function of
+
+$$
+\frac{
+\left(\lambda p_1 +\lambda p_2\right)^{9}
+\left(p_2 + p_3\right)^{2}
+}{
+\left(\lambda p_1 +\lambda p_2+p_2 + p_3\right)^{11}
+}
+$$
+
+```{r printH}
+H3 <- as.hyper3(H)
+H3
+
+## 1&2 vs 3&4;  scoreline 9-2 (again), with lambda=1.2
+
+f <- function(lambda){
+  H3[c(p1=lambda,p2=lambda)] %<>% inc(9)
+  H3[c(p3=1,p2=1)] %<>% dec(2)
+  H3[c(p1=lambda,p2=lambda,p3=1,p2=1)] %<>% dec(11)
+  return(H3)
+}
+f(1.1443345)
+f(1.888888)
+```
+
+maximize:
+
+```{r,maxH3,cache=TRUE}
+tennis_H3_maxp <- maxp(H3)
+tennis_H3_maxp
+```
+
+Compare the `hyper2` evaluates:
+
+```{r}
+tennis_maxp
+tennis_ghost_maxp
+```
+
 ### References {-}

### 2022-06-28 inst/race3.Rmd
point to constructor_names.txt

@@
 Looking at the first column we see that the constructors are Mercedes
-("`Merc`"), Red Bull Racing Honda ("`RBRH`"), and so on.  Each
+("`Merc`"), Red Bull Racing Honda ("`RBRH`"), and so on (see file
+`inst/constructor_names.txt` for a list of abbreviations).  Each
 constructor fields two cars in each race.  Looking at the next column,

### 2022-06-29 inst/race3.Rmd
remove second sticker
start work on comparison with Wilcoxon Mann Whitney

@@
 (takes about an hour to process without cache)
-
-<p style="text-align: right;">
-![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
-</p>
 
@@
 
+# Power test for `hyper3` objects compared with Mann-Whitney-Wilcox
+
+First do some stuff where the null is true:
+
+```{r,label=definepower}
+wilcox_pval <- function(a,b,pa,pb){
+  o <- rrace3(pn=c(a=a,b=b),ps=c(a=pa,b=pb))
+  wilcox.test(which(o=='a'),which(o=='b'))$p.value
+}
+
+hyper3_pval <- function(a,b,pa,pb){
+  o <- rrace3(pn=c(a=a,b=b),ps=c(a=pa,b=pb))
+  equalp.test(ordervec2supp3(o))$p.value
+}
+```
+
+```{r,label=dopowernulltrue,cache=TRUE}
+wil_p_h0 <- replicate(100,wilcox_pval(10,10,0.5,0.5))
+hyp_p_h0 <- replicate(100,hyper3_pval(10,10,0.5,0.5))
+```
+
+```{r,label=plotpvalnulltrue}
+plot(ecdf(wil_p_h0),col="black")
+plot(ecdf(hyp_p_h0),col="red",add=TRUE)
+legend("topleft",pch=16,lty=1,col=c("black","red"),legend=c("Wilcox","hyper3"))
+```
+
+Now some stuff where the null is false:
+
+```{r,label=dopower,cache=TRUE}
+wil_p <- replicate(100,wilcox_pval(10,10,0.7,0.3))
+hyp_p <- replicate(100,hyper3_pval(10,10,0.7,0.3))
+```
+
+```{r,label=plotpower}
+hist(wil_p)
+hist(hyp_p)
+par(pty="s")
+qqplot(wil_p,hyp_p,asp=1,xlim=c(0,1),ylim=c(0,1),pty="m")
+abline(0,1)
+qqplot(log(wil_p),log(hyp_p),asp=1,xlim=c(-6,1),ylim=c(-6,1),pty="m")
+abline(0,1)
+ks.test(wil_p,hyp_p)
+```
+
 # The constructors' championship

### 2022-06-30 inst/race3.Rmd
more work on false null
typo

@@
 With an $n$-unit of support criterion the analytical solution to
-$\mathcal{S}(p)=-2$ is given by defining $X=\log(3-2\sqrt{2})-n$ and
+$\mathcal{S}(p)=-n$ is given by defining $X=\log(3-2\sqrt{2})-n$ and
 solving $p(1-p)/(1+p)=X$, or
@@
 
+# Estimation of strength
+
+We will try some random simulation where the strengths of the competitors is known:
+
+```{r,label=randomsimknownstrength,cache=TRUE}
+
+es <- replicate(100,maxp(ordervec2supp3(rrace3(pn=c(a=10,b=10),ps=c(a=1/2,b=1/2))))[1])
+```
+
+```{r,label=showestimates}
+hist(es)
+abline(v=1/2,lwd=5)
+```
+
 # Power test for `hyper3` objects compared with Mann-Whitney-Wilcox

### 2022-06-30 inst/volvo.Rmd
typo

@@
 
-Thus the first column refers to leg 1, in which competitor AbuDhavi
+Thus the first column refers to leg 1, in which competitor Abu Dhabi
 came first, Brunel third, and so on.  Here, `DNF` signifies "did not
-finish" and `DNS` "did not start".  We can conver the table to a
+finish" and `DNS` "did not start".  We can convert the table to a
 support function with `ordertable2supp()`, remembering to specify the

### 2022-07-01 inst/NBA.Rmd
use cache and label chunks

@@
 
-```{r,definsupportb,cache=TRUE}
+```{r,label=definesupportb,cache=TRUE}
 allplayers <- as.matrix(NBA_table[,5:23])
@@
 
-dotchart(NBA_maxp,col=c(rep("black",9),rep("red",10),rep("blue",1)),pch=16)
+Dotchart(NBA_maxp,col=c(rep("black",9),rep("red",10),rep("blue",1)),pch=16)
 ```
@@
 
-```{r,statsanal,cache=TRUE}
+```{r,label=statsanal,cache=TRUE}
 samep.test(NBA2,c("C_possession","W_possession"))
@@
 
-```{r,cache=TRUE}
+```{r,label=doCWtest, cache=TRUE}
 (C_allstrengths <- sort(NBA_maxp[01:9],decreasing=TRUE))

### 2022-07-01 inst/cook.Rmd
use sticker in Rmd
labels and cache

@@
 
-```{r}
+```{r,label=maxH,cache=TRUE}
 maxp(H)
@@
 
-```{r maxlike, echo=FALSE}
+```{r maxlike, echo=FALSE,cache=TRUE}
 dotchart(maxp(H),pch=16,main="probabilities")
@@
 
-```{r}
+```{r,equaltest,cache=TRUE}
 equalp.test(H)
@@
 
-```{r}
+```{r,samtest,cache=TRUE}
 samep.test(H,c("p1","p2"))
@@
 
-```{r}
+```{r,label=obscons,cache=TRUE}
 library(partitions)
@@
 
-```{r}
+```{r,label=obspv}
 obs <- c(5,4,3,2,1)
@@
 
-## Changes in cooking strength.
+## Changes in cooking strength
 
@@
 
-```{r,beforetraining}
+```{r,label=beforetraining,cache=TRUE}
 # observations before training:
@@
 
-```{r,addbeforeandafter}
+```{r,addbeforeandafter,cache=TRUE}
 maxp(Hbefore+Hafter)
@@
 
-```{r,testnullzerotrainingeffect}
+```{r,testnullzerotrainingeffect,cache=TRUE}
 specificp.gt.test(Hbefore+Hafter, "S", 0)

### 2022-07-01 inst/cook.Rmd
hyper3 analysis

@@
 \frac{p_4  }{              p_4+p_5}\cdot
-\frac{p_5  }{                  p_5}\cdot
+\frac{p_5  }{                  p_5}
 \]
@@
 thus we have strong evidence for the training being effective.
+
+# Analysis with `hyper3` objects
+
+We can take the "changes in cooking strength" augmented Plackett-Luce
+likelihood functions above, and apply `hyper3` idiom to it.  Instead
+of using a reified entity $S$ we have likelihood functions
+
+\[
+\frac{\lambda p_1}{\lambda p_1+p_2+p_3+p_4+p_5}\cdot
+\frac{p_2  }{                  p_2+p_3+p_4+p_5}\cdot
+\frac{p_3  }{                      p_3+p_4+p_5}\cdot
+\frac{p_4  }{                          p_4+p_5}\cdot
+\frac{p_5  }{                              p_5}
+\]
+
+\[
+\frac{p_3  }{\lambda p_1+p_2+p_3+p_4+p_5}\cdot
+\frac{\lambda p_1}{\lambda p_1+p_2+    p_4+p_5}\cdot
+\frac{p_5  }{                  p_2+    p_4+p_5}\cdot
+\frac{p_2  }{                  p_2+    p_4    }\cdot
+\frac{p_4  }{                          p_4    }\cdot
+\]
+
+and then test $H_0\colon\lambda=1$.  The `hyper3` likelihood function
+is mathematically the same as the reified `hyper2` BT likelihood
+function provided $p_1+S=\lambda p_1$ (the likelihood functions differ
+only if more than one player has increased strength); note that
+$\lambda < 1$ is not possible with reified BT because it implies
+negative $S$.  We will create a `hyper3` object for the post-training
+observations:
+
+```{r,hyper3post}
+Hpost1 <- function(lambda){
+   out <- hyper3()
+   out[c(p1=lambda)] %<>% inc
+   out[c(p1=lambda,p2=1,p3=1,p4=1,p5=1)] %<>% dec
+
+   out[c(p2=1)] %<>% inc
+   out[c(p2=1,p3=1,p4=1,p5=1)] %<>% dec
+
+   out[c(p3=1)] %<>% inc
+   out[c(p3=1,p4=1,p5=1)] %<>% dec
+
+   out[c(p4=1)] %<>% inc
+   out[c(p4=1,p5=1)] %<>% dec
+   return(out)
+}
+
+Hpost2 <- function(lambda){
+   out <- hyper3()
+   out[c(p3=1)] %<>% inc
+   out[c(p1=lambda,p2=1,p3=1,p4=1,p5=1)] %<>% dec
+
+   out[c(p1=lambda)] %<>% inc
+   out[c(p1=lambda,p2=1,p4=1,p5=1)] %<>% dec
+
+   out[c(p5=1)] %<>% inc
+   out[c(p2=1,p4=1,p5=1)] %<>% dec
+
+   out[c(p2=1)] %<>% inc
+   out[c(p2=1,p4=1)] %<>% dec
+   return(out)
+}
+```
+
+Now The likelihood functions for the "before"  observations simply add, so for example
+
+```{r,showclever}
+as.hyper3(Hbefore) + Hpost1(1.8888) + Hpost2(1.8888)
+```
+
+Then we can calculate a profile likelihood curve:
+
+```{r,label=calculatesupportlambda,cache=TRUE}
+H3 <- function(lambda){as.hyper3(Hbefore) + Hpost1(lambda) + Hpost2(lambda)}
+supplam <- function(lambda){maxp(H3(lambda),n=1,give=TRUE)$likes}
+lam <- exp(seq(from=log(0.8),to=log(100),len=50))
+L <- sapply(lam,supplam)
+```
+
+```{r,label=plotsupp}
+plot(lam,L-max(L),type="l",lwd=2)
+abline(h=c(0,-2),col='gray')
+abline(v=1,col='red')
+text(0,-0.2,"lambda=1",col="red",pos=4)
+```
+
+We see that we can reject $H_0\colon\lambda=1$ but for this dataset
+there does not appear to be a clear upper bound on $\lambda$.

### 2022-07-03 inst/plackett_luce_monster.Rmd
add sticker
transfer tikz picture to race3.Rmd

@@
 ```
+
+```{tikz, fig.cap = "Partial probability tree for five competitors $a$-$e$ with mutually supporting subsets $(abc)$ and $(de)$, reified approach", fig.ext = 'png', echo=FALSE}
+\usetikzlibrary{arrows}
+\usetikzlibrary{patterns}
+\begin{tikzpicture}[line cap=round,line join=round,>=triangle 45,x=1cm,y=1cm]
+\fill (0,0) circle[radius=2pt]; % root; paths abcde
+\draw (0,0) -- (2,3); 
+\draw (0,0) -- (2,2);
+\draw (0,0) -- (2,1);
+\draw (0,0) -- (2,0);
+\draw (0,0) -- (2,-1);
+\node at (0,2.5) (eq1) {$(abc)(de)$};
+
+\node at (1.5,2.5) {$a$};
+\node at (1.5,1.7) {$b$};
+\node at (1.5,1.0) {$c$};
+\node at (1.5,0.2) {$d$};
+\node at (1.5,-0.5) {$e$};
+
+\node at (3, 3)  {$\frac{a}{a+b+c+d+e}$};
+\node at (3, 2)  {$\frac{b}{a+b+c+d+e}$};
+\node at (3, 1)  {$\frac{c}{a+b+c+d+e}$};
+\node at (3, 0)  {$\frac{d}{a+b+c+d+e}$};
+\node at (3,-1)  {$\frac{e}{a+b+c+d+e}$};
+
+\fill  (4, 3) circle[radius=2pt];  % a finishes; paths bcde
+\fill  (4, 2) circle[radius=2pt];  % terminal node
+\fill  (4, 1) circle[radius=2pt];  % terminal node
+\fill  (4, 0) circle[radius=2pt];  % d finishes; paths abce
+\fill  (4,-1) circle[radius=2pt];  % terminal node
+
+\draw (4,3) -- (5,5); 
+\draw (4,3) -- (5,4); 
+\draw (4,3) -- (5,3); 
+\draw (4,3) -- (5,2);
+
+\node at (4.5,4.6) {$b$};
+\node at (4.5,3.7) {$c$};
+\node at (4.5,3.2) {$d$};
+\node at (4.5,2.7) {$e$};
+
+\node at (6, 5)  {$\frac{b+M}{b+c+d+e+2M}$};
+\node at (6, 4)  {$\frac{c+M}{b+c+d+e+2M}$};
+\node at (6, 3)  {$\frac{d}{b+c+d+e+2M}$};
+\node at (6, 2)  {$\frac{e}{b+c+d+e+2M}$};
+
+\fill (7.4, 5) circle[radius=2pt];  % ab finishes; pahts cde
+\draw (7.4,5) -- (8.4,6);
+\draw (7.4,5) -- (8.4,5);
+\draw (7.4,5) -- (8.4,4);
+
+\node at (7.9,5.7) {$c$};
+\node at (7.9,5.2) {$d$};
+\node at (7.9,4.7) {$e$};
+
+\node at (9.3, 6) {$\frac{c+M}{c+d+e+M}$};
+\node at (9.3, 5) {$\frac{d}{c+d+e+M}$};
+\node at (9.3, 4) {$\frac{e}{c+d+e+M}$};
+
+\fill (10.2, 6) circle[radius=2pt];  % abc finishes; paths de
+\draw (10.2, 6) -- (11.2,7);
+\node at (10.7,6.8) {$d$};
+
+\node at (11.8, 7) {$\frac{d}{d+e}$};
+
+\draw (10.2, 6) -- (11.2,6.2);
+\node at (11.8, 6.2) {$\frac{e}{d+e}$};
+\node at (10.7,6.3) {$e$};
+
+\fill (10.2, 5) circle[radius=2pt];
+\draw (10.2, 5) -- (11.2,5.3);
+\node at (11.8, 5.3) {$\frac{c}{c+e+M}$};
+\node at (10.7,5.3) {$c$};
+
+\draw (10.2, 5) -- (11.2,4.7);
+\node at (11.8, 4.7) {$\frac{e+M}{c+e+M}$};
+\node at (10.7,4.7) {$e$};
+
+\fill (10.2, 4) circle[radius=2pt];
+\draw (10.2, 4) -- (11.2,3.9);
+\node at (11.8, 3.9) {$\frac{c}{c+e+M}$};
+\node at (10.7,4.1) {$c$};
+
+\draw (10.2, 4) -- (11.2,3.1);
+\node at (11.8, 3.1) {$\frac{e+M}{c+e+M}$};
+\node at (10.7,3.7) {$e$};
+
+\draw (4, 0) -- (5,1);
+\node at (6, 1)  {$\frac{a}{a+b+c+e+M}$};
+
+\draw (4, 0) -- (5,0);
+\node at (6, 0)  {$\frac{b}{a+b+c+e+M}$};
+
+\draw (4, 0) -- (5,-1);
+\node at (6, -1)  {$\frac{c}{a+b+c+e+M}$};
+
+\draw (4, 0) -- (5,-2);
+\node at (6, -2)  {$\frac{e+M}{a+b+c+e+M}$};
+
+\node at (4.5,0.7) {$a$};
+\node at (4.5,0.2) {$b$};
+\node at (4.5,-0.3) {$c$};
+\node at (4.5,-0.8) {$e$};
+
+\fill (7.4, -2) circle[radius=2pt];  % de finishes; paths abc
+\draw (7.4, -2) -- (8.4,-1);
+\draw (7.4, -2) -- (8.4,-2);
+\draw (7.4, -2) -- (8.4,-3);
+
+\node at (7.9,-1.2) {$a$};
+\node at (7.9,-1.8) {$b$};
+\node at (7.9,-2.3) {$c$};
+
+\node at (9.3, -1) {$\frac{a}{a+b+c}$};
+\node at (9.3, -2) {$\frac{b}{a+b+c}$};
+\node at (9.3, -3) {$\frac{c}{a+b+c}$};
+
+\fill (10.2, -1) circle[radius=2pt];  % dea finishes; paths bc
+\fill (10.2, -2) circle[radius=2pt];  % deb finishes; paths ac
+\fill (10.2, -3) circle[radius=2pt];  % dec finishes; paths ab
+
+\draw (10.2, -1) -- (11.2,-0);
+\node at (10.6,-0.4) {$b$};
+\node at (10.6,-0.8) {$c$};
+
+\node at (11.8, 0) {$\frac{b+M}{b+c+2M}$};
+\draw (10.2, -1) -- (11.2,-1);
+
+\node at (11.8, -1) {$\frac{c+M}{b+c+2M}$};
+
+\draw (10.2, -2) -- (11.2,-1.6);
+\node at (11.8, -1.6) {$\frac{a+M}{a+c+2M}$};
+
+\node at (10.6,-1.7) {$a$};
+\node at (10.6,-2) {$c$};
+
+\draw (10.2, -2) -- (11.2,-2.2);
+\node at (11.8, -2.2) {$\frac{c+M}{a+c+2M}$};
+
+\draw (10.2, -3) -- (11.2,-2.9);
+\node at (11.8, -2.9) {$\frac{a+M}{a+b+2M}$};
+
+\draw (10.2, -3) -- (11.2,-3.4);
+\node at (11.8, -3.4) {$\frac{b+M}{a+b+2M}$};
+
+\node at (10.6,-2.8) {$a$};
+\node at (10.6,-3.3) {$b$};
+
+\fill (7.4, 1) circle[radius=2pt];  % da finishes; paths bce
+
+\draw (7.4, 1) -- (8.4,2);
+\draw (7.4, 1) -- (8.4,1);
+\draw (7.4, 1) -- (8.4,0);
+
+\node at (7.9,1.7) {$b$};
+\node at (7.9,1.2) {$c$};
+\node at (7.9,0.7) {$e$};
+
+\node at (9.3, 2) {$\frac{b+M}{b+c+e+2M}$};
+\node at (9.3, 1) {$\frac{c+M}{b+c+e+2M}$};
+\node at (9.3, 0) {$\frac{e}{b+c+e+2M}$};
+
+\fill (10.2, 1) circle[radius=2pt];  % da finishes; paths bce
+\draw (10.2, 1) -- (11.2,2);
+\node at (11.8, 2) {$\frac{b+M}{b+e+M}$};
+\draw (10.2, 1) -- (11.2,1);
+\node at (11.8, 1) {$\frac{e}{b+e+M}$};
+
+\end{tikzpicture}
+```
+
+```{tikz, fig.cap = "Partial probability tree for five competitors $a$-$e$ with mutually supporting subsets $(abc)$ and $(de)$, hyper3 approach", fig.ext = 'png', echo=FALSE}
+\usetikzlibrary{arrows}
+\usetikzlibrary{patterns}
+\begin{tikzpicture}[line cap=round,line join=round,>=triangle 45,x=1cm,y=1cm]
+\fill (0,0) circle[radius=2pt]; % root; paths abcde
+\draw (0,0) -- (2,3); 
+\draw (0,0) -- (2,2);
+\draw (0,0) -- (2,1);
+\draw (0,0) -- (2,0);
+\draw (0,0) -- (2,-1);
+\node at (0,2.5) (eq1) {$(abc)(de)$};
+
+\node at (1.5,2.5) {$a$};
+\node at (1.5,1.7) {$b$};
+\node at (1.5,1.0) {$c$};
+\node at (1.5,0.2) {$d$};
+\node at (1.5,-0.5) {$e$};
+
+\node at (3, 3)  {$\frac{a}{a+b+c+d+e}$};
+\node at (3, 2)  {$\frac{b}{a+b+c+d+e}$};
+\node at (3, 1)  {$\frac{c}{a+b+c+d+e}$};
+\node at (3, 0)  {$\frac{d}{a+b+c+d+e}$};
+\node at (3,-1)  {$\frac{e}{a+b+c+d+e}$};
+
+\fill  (4, 3) circle[radius=2pt];  % a finishes; paths bcde
+\fill  (4, 2) circle[radius=2pt];  % terminal node
+\fill  (4, 1) circle[radius=2pt];  % terminal node
+\fill  (4, 0) circle[radius=2pt];  % d finishes; paths abce
+\fill  (4,-1) circle[radius=2pt];  % terminal node
+
+\draw (4,3) -- (5,5); 
+\draw (4,3) -- (5,4); 
+\draw (4,3) -- (5,3); 
+\draw (4,3) -- (5,2);
+
+\node at (4.5,4.6) {$b$};
+\node at (4.5,3.7) {$c$};
+\node at (4.5,3.2) {$d$};
+\node at (4.5,2.7) {$e$};
+
+\newcommand{\lb}{\lambda b }
+\newcommand{\lc}{\lambda c }
+\newcommand{\lee}{\lambda e }
+\node at (6, 5)  {$\frac{\lb}{\lb+\lc+d+e}$};
+\node at (6, 4)  {$\frac{\lc}{\lb+\lc+d+e}$};
+\node at (6, 3)  {$\frac{d  }{\lb+\lc+d+e}$};
+\node at (6, 2)  {$\frac{e  }{\lb+\lc+d+e}$};
+
+\fill (7.4, 5) circle[radius=2pt];  % ab finishes; pahts cde
+\draw (7.4,5) -- (8.4,6);
+\draw (7.4,5) -- (8.4,5);
+\draw (7.4,5) -- (8.4,4);
+
+\node at (7.9,5.7) {$c$};
+\node at (7.9,5.2) {$d$};
+\node at (7.9,4.7) {$e$};
+\node at (9.3, 6) {$\frac{\lc}{\lc+d+e}$};
+\node at (9.3, 5) {$\frac{d  }{\lc+d+e}$};
+\node at (9.3, 4) {$\frac{e  }{\lc+d+e}$};
+
+\fill (10.2, 6) circle[radius=2pt];  % abc finishes; paths de
+\draw (10.2, 6) -- (11.2,7);
+\node at (10.7,6.8) {$d$};
+
+\node at (11.8, 7) {$\frac{d}{d+e}$};
+
+\draw (10.2, 6) -- (11.2,6.2);
+\node at (11.8, 6.2) {$\frac{e}{d+e}$};
+\node at (10.7,6.3) {$e$};
+
+\fill (10.2, 5) circle[radius=2pt];
+\draw (10.2, 5) -- (11.2,5.3);
+\node at (11.8, 5.3) {$\frac{c}{c+\lee}$};
+\node at (10.7,5.3) {$c$};
+
+\draw (10.2, 5) -- (11.2,4.7);
+\node at (11.8, 4.7) {$\frac{\lee}{c+\lee}$};
+\node at (10.7,4.7) {$e$};
+
+\fill (10.2, 4) circle[radius=2pt];
+\draw (10.2, 4) -- (11.2,3.9);
+\node at (11.8, 3.9) {$\frac{c}{c+\lee}$};
+\node at (10.7,4.1) {$c$};
+
+\draw (10.2, 4) -- (11.2,3.1);
+\node at (11.8, 3.1) {$\frac{\lee}{c+\lee}$};
+\node at (10.7,3.7) {$e$};
+
+\draw (4, 0) -- (5,1);
+\node at (6, 1)  {$\frac{a}{a+b+c+\lee}$};
+
+\draw (4, 0) -- (5,0);
+\node at (6, 0)  {$\frac{b}{a+b+c+\lee}$};
+
+\draw (4, 0) -- (5,-1);
+\node at (6, -1)  {$\frac{c}{a+b+c+\lee}$};
+
+\draw (4, 0) -- (5,-2);
+\node at (6, -2)  {$\frac{\lee}{a+b+c+\lee}$};
+
+\node at (4.5,0.7) {$a$};
+\node at (4.5,0.2) {$b$};
+\node at (4.5,-0.3) {$c$};
+\node at (4.5,-0.8) {$e$};
+
+\fill (7.4, -2) circle[radius=2pt];  % de finishes; paths abc
+\draw (7.4, -2) -- (8.4,-1);
+\draw (7.4, -2) -- (8.4,-2);
+\draw (7.4, -2) -- (8.4,-3);
+
+\node at (7.9,-1.2) {$a$};
+\node at (7.9,-1.8) {$b$};
+\node at (7.9,-2.3) {$c$};
+
+\node at (9.3, -1) {$\frac{a}{a+b+c}$};
+\node at (9.3, -2) {$\frac{b}{a+b+c}$};
+\node at (9.3, -3) {$\frac{c}{a+b+c}$};
+
+\fill (10.2, -1) circle[radius=2pt];  % dea finishes; paths bc
+\fill (10.2, -2) circle[radius=2pt];  % deb finishes; paths ac
+\fill (10.2, -3) circle[radius=2pt];  % dec finishes; paths ab
+
+\draw (10.2, -1) -- (11.2,-0);
+\node at (10.6,-0.4) {$b$};
+\node at (10.6,-0.8) {$c$};
+
+\node at (11.8, 0) {$\frac{b}{b+c}$};
+\draw (10.2, -1) -- (11.2,-1);
+
+\node at (11.8, -1) {$\frac{c}{b+c}$};
+
+\draw (10.2, -2) -- (11.2,-1.6);
+\node at (11.8, -1.6) {$\frac{a}{a+c}$};
+
+\node at (10.6,-1.7) {$a$};
+\node at (10.6,-2) {$c$};
+
+\draw (10.2, -2) -- (11.2,-2.2);
+\node at (11.8, -2.2) {$\frac{c}{a+c}$};
+
+\draw (10.2, -3) -- (11.2,-2.9);
+\node at (11.8, -2.9) {$\frac{a}{a+b}$};
+
+\draw (10.2, -3) -- (11.2,-3.4);
+\node at (11.8, -3.4) {$\frac{b}{a+b}$};
+
+\node at (10.6,-2.8) {$a$};
+\node at (10.6,-3.3) {$b$};
+
+\fill (7.4, 1) circle[radius=2pt];  % da finishes; paths bce
+
+\draw (7.4, 1) -- (8.4,2);
+\draw (7.4, 1) -- (8.4,1);
+\draw (7.4, 1) -- (8.4,0);
+
+\node at (7.9,1.7) {$b$};
+\node at (7.9,1.2) {$c$};
+\node at (7.9,0.7) {$e$};
+
+\node at (9.3, 2) {$\frac{\lb}{\lb+\lc+e}$};
+\node at (9.3, 1) {$\frac{\lc}{\lb+\lc+e}$};
+\node at (9.3, 0) {$\frac{e  }{\lb+\lc+e}$};
+
+\fill (10.2, 1) circle[radius=2pt];  % da finishes; paths bce
+\draw (10.2, 1) -- (11.2,2);
+\node at (11.8, 2) {$\frac{\lb}{\lb+e}$};
+\draw (10.2, 1) -- (11.2,1);
+\node at (11.8, 1) {$\frac{e}{\lb+e}$};
+
+\end{tikzpicture}
+```

### 2022-07-04 inst/plackett_luce_monster.Rmd
further thoughts on nonindependence

@@
 
-The statistical model is as follows.  If a runner finishes, then he
-cheers on all his friends who are still running, thereby adding to
-their strength.  This is a very simple form of non-independence.
+There seem to be a few distinct statistical models for nonindependence:
+
+* Runner $i$ might wish to support $j$.  If runner $i$ finishes, once
+  his rank is established he might be able to help $j$ in some way.
+  He might cheer his still-running friend.  In general, we can
+  consider $n$ ordered pairs $(i_r,j_r)$, $1\leq r\leq n$, with runner
+  $i_1$ helping runner $j_1$, runner $i_2$ helping $j_2$, and so on.
+  One simple case would be equivalence classes of
+  $[n]=\{1,2,\ldots,n\}$ with each equivalence class comprising
+  mutually supporting runners.  In this model, with a pair of mutually
+  supporting runners only the higher-placed competitor can affect the
+  other.  The causation is not symmetrical.
+* Runner $i$ might hate runner $j$.  If $i$ finishes, he might expend
+  his spare energy hinderig $j$ (by throwing stones, for example).
+  Note that equivalence classes of mutual hatred might exist.  Again
+  note that the link is not symmetrical.
+* The help or hindrance discussed above might be contingent on
+  placing: perhaps only the winnner (that is, first-placing
+  competitor) would be able to influence other competitors.  Or maybe
+  competitors with any placing could influence other competitors.
+* Runners $i$ and $j$ might elect to cross the finishing line holding
+  hands, with the apparent tie broken randomly.  This would be a
+  symmetrical relationship.  In general, we could consider
+  hand-holding equivalence classes.
+* We might have a resource allocation problem, with runners having a
+  certain [possibly unknown] amount of "strength" but also the ability
+  to expend this strength on objectives other than ranking in the
+  race.  A runner might be able to expend just enough energy to remain
+  slightly ahead of a rival, reserving energy for other purposes.  If
+  runner $i$ wishes to beat runner $j$, and has sufficient resources
+  to do so, then runners $i$ and $j$ will finish consecutively with
+  $i$ ahead of $j$.
 

### 2022-07-04 inst/carcinoma.Rmd

@@
 
+(takes maybe twenty minutes to run without cache.  The value of
+`carcinoma_maxp` is calculated separately [discussed below], and that
+takes a few days to run).
+
 <p style="text-align: right;">
@@
 </p>
-
-(takes maybe twenty minutes to run without cache.  The value of
-`carcinoma_maxp` is calculated separately [discussed below], and that
-takes a few days to run).
 
@@
 
-Here, the object of inference is the set of pathologists.  Informative
-
-The probability model is as follows.  To each pathologist $i,1\leq i
-\leq 7$, we assign a non-negative Bradley-Terry strength and require
-$\sum p_i=1$.  Given that two pathologists $i$ and $j$ differ in their
+Here, the object of inference is the set of pathologists.  The
+probability model is as follows.  To each pathologist $i,1\leq i \leq
+7$, we assign a non-negative Bradley-Terry strength and require $\sum
+p_i=1$.  Given that two pathologists $i$ and $j$ differ in their
 diagnosis then the probability of $i$ diagnosing the presence, and $j$
@@
 the presence of carcinoma to the least convinced.  Thus a ranking of
-`DACBEGF` means that clinician `D` was most convinced of the presense
+`DACBEGF` means that clinician `D` was most convinced of the presence
 of carcinoma and `F` least convinced.  Given this, and given a single
 slide, our actual observation might be that `D` and `A` diagnosed
-presence, and the others diagosed absence, of carcinoma.  The
+presence, and the others diagnosed absence, of carcinoma.  The
 corresponding entry would be

### 2022-07-04 inst/surfing_jsa.tex
documentation for directlogplotofstrengths_postinkscape.pdf

@@
 \begin{figure}[h]
-\includegraphics[width=4in]{directlogplotofstrengths_postinkscape.pdf}
+  \includegraphics[width=4in]{directlogplotofstrengths_postinkscape.pdf}  % to create this file, open directlogplotofstrengths_postinkscape.svg in inkscape and export to PDF.
+File directlogplotofstrengths_postinkscape.svg is under version control in the hyper2 github repo.
+It is a slight modification of directlogplotofstrengths.pdf, with slight optical glitches removed.
+File directelogplotofstrengths.pdf is produced by surfing.Rmd.
 \caption{Scatterplot of points scored  \label{compare_likelihood_points}

### 2022-07-06 inst/plackett_luce_monster.Rmd
more thoughts

@@
   supporting runners only the higher-placed competitor can affect the
-  other.  The causation is not symmetrical.
+  other.  Each equivalence class might have a different supportive
+  proclivity.
+
 * Runner $i$ might hate runner $j$.  If $i$ finishes, he might expend
-  his spare energy hinderig $j$ (by throwing stones, for example).
-  Note that equivalence classes of mutual hatred might exist.  Again
-  note that the link is not symmetrical.
+  his spare energy hindering $j$ (by throwing stones or booing , for
+  example).  Note that equivalence classes of mutual hatred might
+  exist.
+
+* Also, the help or hindrance might have a finite lifetime.  Under one
+  model, if $i$ finishes then he can stay helping (or hindering) $j$
+  wherever $j$ places.  Alternatively, suppose $i$ places $n$th.  Then
+  perhaps he can affect $j$ only place $n+1$, and after a competitor
+  finishes in place $n+1$, then $i$ can exert no further influence.
+
 * The help or hindrance discussed above might be contingent on
@@
   competitors with any placing could influence other competitors.
+
 * Runners $i$ and $j$ might elect to cross the finishing line holding
@@
   hand-holding equivalence classes.
+
 * We might have a resource allocation problem, with runners having a
@@
 
-```{tikz, fig.cap = "Partial probability tree for five competitors $a$-$e$ with mutually supporting subsets $(abc)$ and $(de)$, reified approach", fig.ext = 'png', echo=FALSE}
+```{tikz, fig.cap = "Partial probability tree for five competitors $a-e$ with $a$ supporting $b$, hyper3 approach", fig.ext = 'png', echo=FALSE}
 \usetikzlibrary{arrows}
@@
 \draw (0,0) -- (2,-1);
-\node at (0,2.5) (eq1) {$(abc)(de)$};
-
-\node at (1.5,2.5) {$a$};
-\node at (1.5,1.7) {$b$};
-\node at (1.5,1.0) {$c$};
-\node at (1.5,0.2) {$d$};
-\node at (1.5,-0.5) {$e$};
-
-\node at (3, 3)  {$\frac{a}{a+b+c+d+e}$};
-\node at (3, 2)  {$\frac{b}{a+b+c+d+e}$};
-\node at (3, 1)  {$\frac{c}{a+b+c+d+e}$};
-\node at (3, 0)  {$\frac{d}{a+b+c+d+e}$};
-\node at (3,-1)  {$\frac{e}{a+b+c+d+e}$};
-
-\fill  (4, 3) circle[radius=2pt];  % a finishes; paths bcde
-\fill  (4, 2) circle[radius=2pt];  % terminal node
-\fill  (4, 1) circle[radius=2pt];  % terminal node
-\fill  (4, 0) circle[radius=2pt];  % d finishes; paths abce
-\fill  (4,-1) circle[radius=2pt];  % terminal node
-
-\draw (4,3) -- (5,5); 
-\draw (4,3) -- (5,4); 
-\draw (4,3) -- (5,3); 
-\draw (4,3) -- (5,2);
-
-\node at (4.5,4.6) {$b$};
-\node at (4.5,3.7) {$c$};
-\node at (4.5,3.2) {$d$};
-\node at (4.5,2.7) {$e$};
-
-\node at (6, 5)  {$\frac{b+M}{b+c+d+e+2M}$};
-\node at (6, 4)  {$\frac{c+M}{b+c+d+e+2M}$};
-\node at (6, 3)  {$\frac{d}{b+c+d+e+2M}$};
-\node at (6, 2)  {$\frac{e}{b+c+d+e+2M}$};
-
-\fill (7.4, 5) circle[radius=2pt];  % ab finishes; pahts cde
-\draw (7.4,5) -- (8.4,6);
-\draw (7.4,5) -- (8.4,5);
-\draw (7.4,5) -- (8.4,4);
-
-\node at (7.9,5.7) {$c$};
-\node at (7.9,5.2) {$d$};
-\node at (7.9,4.7) {$e$};
-
-\node at (9.3, 6) {$\frac{c+M}{c+d+e+M}$};
-\node at (9.3, 5) {$\frac{d}{c+d+e+M}$};
-\node at (9.3, 4) {$\frac{e}{c+d+e+M}$};
-
-\fill (10.2, 6) circle[radius=2pt];  % abc finishes; paths de
-\draw (10.2, 6) -- (11.2,7);
-\node at (10.7,6.8) {$d$};
-
-\node at (11.8, 7) {$\frac{d}{d+e}$};
-
-\draw (10.2, 6) -- (11.2,6.2);
-\node at (11.8, 6.2) {$\frac{e}{d+e}$};
-\node at (10.7,6.3) {$e$};
-
-\fill (10.2, 5) circle[radius=2pt];
-\draw (10.2, 5) -- (11.2,5.3);
-\node at (11.8, 5.3) {$\frac{c}{c+e+M}$};
-\node at (10.7,5.3) {$c$};
-
-\draw (10.2, 5) -- (11.2,4.7);
-\node at (11.8, 4.7) {$\frac{e+M}{c+e+M}$};
-\node at (10.7,4.7) {$e$};
-
-\fill (10.2, 4) circle[radius=2pt];
-\draw (10.2, 4) -- (11.2,3.9);
-\node at (11.8, 3.9) {$\frac{c}{c+e+M}$};
-\node at (10.7,4.1) {$c$};
-
-\draw (10.2, 4) -- (11.2,3.1);
-\node at (11.8, 3.1) {$\frac{e+M}{c+e+M}$};
-\node at (10.7,3.7) {$e$};
-
-\draw (4, 0) -- (5,1);
-\node at (6, 1)  {$\frac{a}{a+b+c+e+M}$};
-
-\draw (4, 0) -- (5,0);
-\node at (6, 0)  {$\frac{b}{a+b+c+e+M}$};
-
-\draw (4, 0) -- (5,-1);
-\node at (6, -1)  {$\frac{c}{a+b+c+e+M}$};
-
-\draw (4, 0) -- (5,-2);
-\node at (6, -2)  {$\frac{e+M}{a+b+c+e+M}$};
-
-\node at (4.5,0.7) {$a$};
-\node at (4.5,0.2) {$b$};
-\node at (4.5,-0.3) {$c$};
-\node at (4.5,-0.8) {$e$};
-
-\fill (7.4, -2) circle[radius=2pt];  % de finishes; paths abc
-\draw (7.4, -2) -- (8.4,-1);
-\draw (7.4, -2) -- (8.4,-2);
-\draw (7.4, -2) -- (8.4,-3);
-
-\node at (7.9,-1.2) {$a$};
-\node at (7.9,-1.8) {$b$};
-\node at (7.9,-2.3) {$c$};
-
-\node at (9.3, -1) {$\frac{a}{a+b+c}$};
-\node at (9.3, -2) {$\frac{b}{a+b+c}$};
-\node at (9.3, -3) {$\frac{c}{a+b+c}$};
-
-\fill (10.2, -1) circle[radius=2pt];  % dea finishes; paths bc
-\fill (10.2, -2) circle[radius=2pt];  % deb finishes; paths ac
-\fill (10.2, -3) circle[radius=2pt];  % dec finishes; paths ab
-
-\draw (10.2, -1) -- (11.2,-0);
-\node at (10.6,-0.4) {$b$};
-\node at (10.6,-0.8) {$c$};
-
-\node at (11.8, 0) {$\frac{b+M}{b+c+2M}$};
-\draw (10.2, -1) -- (11.2,-1);
-
-\node at (11.8, -1) {$\frac{c+M}{b+c+2M}$};
-
-\draw (10.2, -2) -- (11.2,-1.6);
-\node at (11.8, -1.6) {$\frac{a+M}{a+c+2M}$};
-
-\node at (10.6,-1.7) {$a$};
-\node at (10.6,-2) {$c$};
-
-\draw (10.2, -2) -- (11.2,-2.2);
-\node at (11.8, -2.2) {$\frac{c+M}{a+c+2M}$};
-
-\draw (10.2, -3) -- (11.2,-2.9);
-\node at (11.8, -2.9) {$\frac{a+M}{a+b+2M}$};
-
-\draw (10.2, -3) -- (11.2,-3.4);
-\node at (11.8, -3.4) {$\frac{b+M}{a+b+2M}$};
-
-\node at (10.6,-2.8) {$a$};
-\node at (10.6,-3.3) {$b$};
-
-\fill (7.4, 1) circle[radius=2pt];  % da finishes; paths bce
-
-\draw (7.4, 1) -- (8.4,2);
-\draw (7.4, 1) -- (8.4,1);
-\draw (7.4, 1) -- (8.4,0);
-
-\node at (7.9,1.7) {$b$};
-\node at (7.9,1.2) {$c$};
-\node at (7.9,0.7) {$e$};
-
-\node at (9.3, 2) {$\frac{b+M}{b+c+e+2M}$};
-\node at (9.3, 1) {$\frac{c+M}{b+c+e+2M}$};
-\node at (9.3, 0) {$\frac{e}{b+c+e+2M}$};
-
-\fill (10.2, 1) circle[radius=2pt];  % da finishes; paths bce
-\draw (10.2, 1) -- (11.2,2);
-\node at (11.8, 2) {$\frac{b+M}{b+e+M}$};
-\draw (10.2, 1) -- (11.2,1);
-\node at (11.8, 1) {$\frac{e}{b+e+M}$};
-
-\end{tikzpicture}
-```
-
-```{tikz, fig.cap = "Partial probability tree for five competitors $a$-$e$ with mutually supporting subsets $(abc)$ and $(de)$, hyper3 approach", fig.ext = 'png', echo=FALSE}
-\usetikzlibrary{arrows}
-\usetikzlibrary{patterns}
-\begin{tikzpicture}[line cap=round,line join=round,>=triangle 45,x=1cm,y=1cm]
-\fill (0,0) circle[radius=2pt]; % root; paths abcde
-\draw (0,0) -- (2,3); 
-\draw (0,0) -- (2,2);
-\draw (0,0) -- (2,1);
-\draw (0,0) -- (2,0);
-\draw (0,0) -- (2,-1);
-\node at (0,2.5) (eq1) {$(abc)(de)$};
+\node at (0,2.5) (eq1) {$a\longrightarrow b$};
 
@@
 \newcommand{\lee}{\lambda e }
-\node at (6, 5)  {$\frac{\lb}{\lb+\lc+d+e}$};
-\node at (6, 4)  {$\frac{\lc}{\lb+\lc+d+e}$};
-\node at (6, 3)  {$\frac{d  }{\lb+\lc+d+e}$};
-\node at (6, 2)  {$\frac{e  }{\lb+\lc+d+e}$};
+\node at (6, 5)  {$\frac{\lb}{\lb+c+d+e}$};
+\node at (6, 4)  {$\frac{c}{\lb+c+d+e}$};
+\node at (6, 3)  {$\frac{d  }{\lb+c+d+e}$};
+\node at (6, 2)  {$\frac{e  }{\lb+c+d+e}$};
 
-\fill (7.4, 5) circle[radius=2pt];  % ab finishes; pahts cde
+\fill (7.4, 5) circle[radius=2pt];  % ab finishes; paths cde
 \draw (7.4,5) -- (8.4,6);
@@
 \node at (7.9,4.7) {$e$};
-\node at (9.3, 6) {$\frac{\lc}{\lc+d+e}$};
-\node at (9.3, 5) {$\frac{d  }{\lc+d+e}$};
-\node at (9.3, 4) {$\frac{e  }{\lc+d+e}$};
+\node at (9.3, 6) {$\frac{c}{c+d+e}$};
+\node at (9.3, 5) {$\frac{d  }{c+d+e}$};
+\node at (9.3, 4) {$\frac{e  }{c+d+e}$};
 
@@
 \draw (10.2, 5) -- (11.2,5.3);
-\node at (11.8, 5.3) {$\frac{c}{c+\lee}$};
+\node at (11.8, 5.3) {$\frac{c}{c+e}$};
 \node at (10.7,5.3) {$c$};
@@
 \draw (10.2, 5) -- (11.2,4.7);
-\node at (11.8, 4.7) {$\frac{\lee}{c+\lee}$};
+\node at (11.8, 4.7) {$\frac{e}{c+e}$};
 \node at (10.7,4.7) {$e$};
@@
 \draw (10.2, 4) -- (11.2,3.9);
-\node at (11.8, 3.9) {$\frac{c}{c+\lee}$};
+\node at (11.8, 3.9) {$\frac{c}{c+e}$};
 \node at (10.7,4.1) {$c$};
@@
 \draw (10.2, 4) -- (11.2,3.1);
-\node at (11.8, 3.1) {$\frac{\lee}{c+\lee}$};
+\node at (11.8, 3.1) {$\frac{e}{c+e}$};
 \node at (10.7,3.7) {$e$};
@@
 \draw (4, 0) -- (5,1);
-\node at (6, 1)  {$\frac{a}{a+b+c+\lee}$};
+\node at (6, 1)  {$\frac{a}{a+b+c+e}$};
 
 \draw (4, 0) -- (5,0);
-\node at (6, 0)  {$\frac{b}{a+b+c+\lee}$};
+\node at (6, 0)  {$\frac{b}{a+b+c+e}$};
 
 \draw (4, 0) -- (5,-1);
-\node at (6, -1)  {$\frac{c}{a+b+c+\lee}$};
+\node at (6, -1)  {$\frac{c}{a+b+c+e}$};
 
 \draw (4, 0) -- (5,-2);
-\node at (6, -2)  {$\frac{\lee}{a+b+c+\lee}$};
+\node at (6, -2)  {$\frac{e}{a+b+c+e}$};
 
@@
 
-\node at (11.8, 0) {$\frac{b}{b+c}$};
+\node at (11.8, 0) {$\frac{\lb}{\lb+c}$};
 \draw (10.2, -1) -- (11.2,-1);
 
-\node at (11.8, -1) {$\frac{c}{b+c}$};
+\node at (11.8, -1) {$\frac{c}{\lb+c}$};
 
@@
 
+\node at (9.3, 2) {$\frac{\lb}{\lb+c+e}$};
+\node at (9.3, 1) {$\frac{  c}{\lb+c+e}$};
+\node at (9.3, 0) {$\frac{e  }{\lb+c+e}$};
+
+\fill (10.2, 1) circle[radius=2pt];  % da finishes; paths bce
+\draw (10.2, 1) -- (11.2,2);
+\node at (11.8, 2) {$\frac{\lb}{\lb+e}$};
+\draw (10.2, 1) -- (11.2,1);
+\node at (11.8, 1) {$\frac{e}{\lb+e}$};
+
+\end{tikzpicture}
+```
+
+```{tikz, fig.cap = "Partial probability tree for five competitors $a-e$ with mutually supporting  group $abc$, hyper3 approach.
+If any competitor in the set {$a,b,c$} finishes, they lend their strength to the other competitors in the set who are still running", fig.ext = 'png', echo=FALSE}
+\usetikzlibrary{arrows}
+\usetikzlibrary{patterns}
+\begin{tikzpicture}[line cap=round,line join=round,>=triangle 45,x=1cm,y=1cm]
+\fill (0,0) circle[radius=2pt]; % root; paths abcde
+\draw (0,0) -- (2,3); 
+\draw (0,0) -- (2,2);
+\draw (0,0) -- (2,1);
+\draw (0,0) -- (2,0);
+\draw (0,0) -- (2,-1);
+\node at (0,2.5) (eq1) {$\underbrace{\left\lbrace a,b,c\right\rbrace}_{\lambda}$};
+
+\node at (1.5,2.5) {$a$};
+\node at (1.5,1.7) {$b$};
+\node at (1.5,1.0) {$c$};
+\node at (1.5,0.2) {$d$};
+\node at (1.5,-0.5) {$e$};
+
+\node at (3, 3)  {$\frac{a}{a+b+c+d+e}$};
+\node at (3, 2)  {$\frac{b}{a+b+c+d+e}$};
+\node at (3, 1)  {$\frac{c}{a+b+c+d+e}$};
+\node at (3, 0)  {$\frac{d}{a+b+c+d+e}$};
+\node at (3,-1)  {$\frac{e}{a+b+c+d+e}$};
+
+\fill  (4, 3) circle[radius=2pt];  % a finishes; paths bcde
+\fill  (4, 2) circle[radius=2pt];  % terminal node
+\fill  (4, 1) circle[radius=2pt];  % terminal node
+\fill  (4, 0) circle[radius=2pt];  % d finishes; paths abce
+\fill  (4,-1) circle[radius=2pt];  % terminal node
+
+\draw (4,3) -- (5,5); 
+\draw (4,3) -- (5,4); 
+\draw (4,3) -- (5,3); 
+\draw (4,3) -- (5,2);
+
+\node at (4.5,4.6) {$b$};
+\node at (4.5,3.7) {$c$};
+\node at (4.5,3.2) {$d$};
+\node at (4.5,2.7) {$e$};
+
+\newcommand{\la}{\lambda a }
+\newcommand{\lb}{\lambda b }
+\newcommand{\lc}{\lambda c }
+\newcommand{\lee}{\lambda e }
+ \node at (6, 5)  {$\frac{\lb}{\lb+\lc+d+e}$};
+ \node at (6, 4)  {$\frac{\lc}{\lb+\lc+d+e}$};
+ \node at (6, 3)  {$\frac{d  }{\lb+\lc+d+e}$};
+ \node at (6, 2)  {$\frac{e  }{\lb+\lc+d+e}$};
+
+ \fill (7.4, 5) circle[radius=2pt];  % ab finishes; paths cde
+ \draw (7.4,5) -- (8.4,6);
+ \draw (7.4,5) -- (8.4,5);
+ \draw (7.4,5) -- (8.4,4);
+
+ \node at (7.9,5.7) {$c$};
+ \node at (7.9,5.2) {$d$};
+ \node at (7.9,4.7) {$e$};
+ \node at (9.3, 6) {$\frac{\lc}{\lc+d+e}$};
+ \node at (9.3, 5) {$\frac{d  }{\lc+d+e}$};
+ \node at (9.3, 4) {$\frac{e  }{\lc+d+e}$};
+
+ \fill (10.2, 6) circle[radius=2pt];  % abc finishes; paths de
+ \draw (10.2, 6) -- (11.2,7);
+ \node at (10.7,6.8) {$d$};
+
+ \node at (11.8, 7) {$\frac{d}{d+e}$};
+
+ \draw (10.2, 6) -- (11.2,6.2);
+ \node at (11.8, 6.2) {$\frac{e}{d+e}$};
+ \node at (10.7,6.3) {$e$};
+
+ \fill (10.2, 5) circle[radius=2pt];
+ \draw (10.2, 5) -- (11.2,5.3);
+ \node at (11.8, 5.3) {$\frac{\lc}{\lc+e}$};
+ \node at (10.7,5.3) {$c$};
+
+ \draw (10.2, 5) -- (11.2,4.7);
+ \node at (11.8, 4.7) {$\frac{e}{\lc+e}$};
+ \node at (10.7,4.7) {$e$};
+
+ \fill (10.2, 4) circle[radius=2pt];
+ \draw (10.2, 4) -- (11.2,3.9);
+ \node at (11.8, 3.9) {$\frac{\lc}{\lc+e}$};
+ \node at (10.7,4.1) {$c$};
+
+ \draw (10.2, 4) -- (11.2,3.1);
+ \node at (11.8, 3.1) {$\frac{e}{\lc+e}$};
+ \node at (10.7,3.7) {$e$};
+
+ \draw (4, 0) -- (5,1);
+ \node at (6, 1)  {$\frac{a}{a+b+c+e}$};
+
+ \draw (4, 0) -- (5,0);
+ \node at (6, 0)  {$\frac{b}{a+b+c+e}$};
+
+ \draw (4, 0) -- (5,-1);
+ \node at (6, -1)  {$\frac{c}{a+b+c+e}$};
+
+ \draw (4, 0) -- (5,-2);
+ \node at (6, -2)  {$\frac{e}{a+b+c+e}$};
+
+ \node at (4.5,0.7) {$a$};
+ \node at (4.5,0.2) {$b$};
+ \node at (4.5,-0.3) {$c$};
+ \node at (4.5,-0.8) {$e$};
+
+ \fill (7.4, -2) circle[radius=2pt];  % de finishes; paths abc
+ \draw (7.4, -2) -- (8.4,-1);
+ \draw (7.4, -2) -- (8.4,-2);
+ \draw (7.4, -2) -- (8.4,-3);
+
+ \node at (7.9,-1.2) {$a$};
+ \node at (7.9,-1.8) {$b$};
+ \node at (7.9,-2.3) {$c$};
+
+ \node at (9.3, -1) {$\frac{a}{a+b+c}$};
+ \node at (9.3, -2) {$\frac{b}{a+b+c}$};
+ \node at (9.3, -3) {$\frac{c}{a+b+c}$};
+
+ \fill (10.2, -1) circle[radius=2pt];  % dea finishes; paths bc
+ \fill (10.2, -2) circle[radius=2pt];  % deb finishes; paths ac
+ \fill (10.2, -3) circle[radius=2pt];  % dec finishes; paths ab
+
+ \draw (10.2, -1) -- (11.2,-0);
+ \node at (10.6,-0.4) {$b$};
+ \node at (10.6,-0.8) {$c$};
+
+ \node at (11.8, 0) {$\frac{\lb}{\lb+\lc}$};
+ \draw (10.2, -1) -- (11.2,-1);
+
+ \node at (11.8, -1) {$\frac{c}{\lb+\lc}$};
+
+ \draw (10.2, -2) -- (11.2,-1.6);
+ \node at (11.8, -1.6) {$\frac{\la}{\la+\lc}$};
+
+ \node at (10.6,-1.7) {$a$};
+ \node at (10.6,-2) {$c$};
+
+ \draw (10.2, -2) -- (11.2,-2.2);
+ \node at (11.8, -2.2) {$\frac{\lc}{\la+\lc}$};
+
+ \draw (10.2, -3) -- (11.2,-2.9);
+ \node at (11.8, -2.9) {$\frac{\la}{\la+\lb}$};
+
+ \draw (10.2, -3) -- (11.2,-3.4);
+ \node at (11.8, -3.4) {$\frac{\lb}{\la+\lb}$};
+
+ \node at (10.6,-2.8) {$a$};
+ \node at (10.6,-3.3) {$b$};
+
+ \fill (7.4, 1) circle[radius=2pt];  % da finishes; paths bce
+
+ \draw (7.4, 1) -- (8.4,2);
+ \draw (7.4, 1) -- (8.4,1);
+ \draw (7.4, 1) -- (8.4,0);
+
+ \node at (7.9,1.7) {$b$};
+ \node at (7.9,1.2) {$c$};
+ \node at (7.9,0.7) {$e$};
+
  \node at (9.3, 2) {$\frac{\lb}{\lb+\lc+e}$};
@@
 ```
+
+```{tikz, fig.cap = "Partial probability tree for five competitors $a$-$e$ with mutually supporting subsets $\\left\\lbrace a,b,c\\right\\rbrace$ [with support term $\\lambda$] and $(de)$ [with support term $\\mu$], hyper3 approach", fig.ext = 'png', echo=FALSE}
+\usetikzlibrary{arrows}
+\usetikzlibrary{patterns}
+\begin{tikzpicture}[line cap=round,line join=round,>=triangle 45,x=1cm,y=1cm]
+\fill (0,0) circle[radius=2pt]; % root; paths abcde
+\draw (0,0) -- (2,3); 
+\draw (0,0) -- (2,2);
+\draw (0,0) -- (2,1);
+\draw (0,0) -- (2,0);
+\draw (0,0) -- (2,-1);
+\node at (0,2.5) (eq1) {$\underbrace{\left\lbrace a,b,c\right\rbrace}_{\lambda}\underbrace{\left\lbrace d,e\right\rbrace}_{\mu}$};
+
+\node at (1.5,2.5) {$a$};
+\node at (1.5,1.7) {$b$};
+\node at (1.5,1.0) {$c$};
+\node at (1.5,0.2) {$d$};
+\node at (1.5,-0.5) {$e$};
+
+\node at (3, 3)  {$\frac{a}{a+b+c+d+e}$};
+\node at (3, 2)  {$\frac{b}{a+b+c+d+e}$};
+\node at (3, 1)  {$\frac{c}{a+b+c+d+e}$};
+\node at (3, 0)  {$\frac{d}{a+b+c+d+e}$};
+\node at (3,-1)  {$\frac{e}{a+b+c+d+e}$};
+
+\fill  (4, 3) circle[radius=2pt];  % a finishes; paths bcde
+\fill  (4, 2) circle[radius=2pt];  % terminal node
+\fill  (4, 1) circle[radius=2pt];  % terminal node
+\fill  (4, 0) circle[radius=2pt];  % d finishes; paths abce
+\fill  (4,-1) circle[radius=2pt];  % terminal node
+
+\draw (4,3) -- (5,5); 
+\draw (4,3) -- (5,4); 
+\draw (4,3) -- (5,3); 
+\draw (4,3) -- (5,2);
+
+\node at (4.5,4.6) {$b$};
+\node at (4.5,3.7) {$c$};
+\node at (4.5,3.2) {$d$};
+\node at (4.5,2.7) {$e$};
+
+\newcommand{\la}[1]{\lambda{#1}}
+\newcommand{\ld}[1]{\mu    {#1}}
+\node at (6, 5)  {$\frac{\la{b}}{\la{b}+\la{c}+d+e}$};
+\node at (6, 4)  {$\frac{\la{c}}{\la{b}+\la{c}+d+e}$};
+\node at (6, 3)  {$\frac{d     }{\la{b}+\la{c}+d+e}$};
+\node at (6, 2)  {$\frac{e     }{\la{b}+\la{c}+d+e}$};
+
+\fill (7.4, 5) circle[radius=2pt];  % ab finishes; pahts cde
+\draw (7.4,5) -- (8.4,6);
+\draw (7.4,5) -- (8.4,5);
+\draw (7.4,5) -- (8.4,4);
+
+\node at (7.9,5.7) {$c$};
+\node at (7.9,5.2) {$d$};
+\node at (7.9,4.7) {$e$};
+\node at (9.3, 6) {$\frac{\la{c}}{\la{c}+d+e}$};
+\node at (9.3, 5) {$\frac{d  }{\la{c}+d+e}$};
+\node at (9.3, 4) {$\frac{e  }{\la{c}+d+e}$};
+
+\fill (10.2, 6) circle[radius=2pt];  % abc finishes; paths de
+\draw (10.2, 6) -- (11.2,7);
+\node at (10.7,6.8) {$d$};
+
+\node at (11.8, 7) {$\frac{d}{d+e}$};
+
+\draw (10.2, 6) -- (11.2,6.2);
+\node at (11.8, 6.2) {$\frac{e}{d+e}$};
+\node at (10.7,6.3) {$e$};
+
+\fill (10.2, 5) circle[radius=2pt];
+\draw (10.2, 5) -- (11.2,5.3);
+\node at (11.8, 5.3) {$\frac{\la{c}}{\la{c}+\ld{e}}$};
+\node at (10.7,5.3) {$c$};
+
+\draw (10.2, 5) -- (11.2,4.7);
+\node at (11.8, 4.7) {$\frac{\ld{e}}{\la{c}+\ld{e}}$};
+\node at (10.7,4.7) {$e$};
+
+\fill (10.2, 4) circle[radius=2pt];
+\draw (10.2, 4) -- (11.2,3.9);
+\node at (11.8, 3.9) {$\frac{\la{c}}{\la{c}+\ld{e}}$};
+\node at (10.7,4.1) {$c$};
+
+\draw (10.2, 4) -- (11.2,3.1);
+\node at (11.8, 3.1) {$\frac{\ld{e}}{\la{c}+\ld{e}}$};
+\node at (10.7,3.7) {$e$};
+
+\draw (4, 0) -- (5,1);
+\node at (6, 1)  {$\frac{a}{a+b+c+\ld{e}}$};
+
+\draw (4, 0) -- (5,0);
+\node at (6, 0)  {$\frac{b}{a+b+c+\ld{e}}$};
+
+\draw (4, 0) -- (5,-1);
+\node at (6, -1)  {$\frac{c}{a+b+c+\ld{e}}$};
+
+\draw (4, 0) -- (5,-2);
+\node at (6, -2)  {$\frac{\ld{e}}{a+b+c+\ld{e}}$};
+
+\node at (4.5,0.7) {$a$};
+\node at (4.5,0.2) {$b$};
+\node at (4.5,-0.3) {$c$};
+\node at (4.5,-0.8) {$e$};
+
+\fill (7.4, -2) circle[radius=2pt];  % de finishes; paths abc
+\draw (7.4, -2) -- (8.4,-1);
+\draw (7.4, -2) -- (8.4,-2);
+\draw (7.4, -2) -- (8.4,-3);
+
+\node at (7.9,-1.2) {$a$};
+\node at (7.9,-1.8) {$b$};
+\node at (7.9,-2.3) {$c$};
+
+\node at (9.3, -1) {$\frac{a}{a+b+c}$};
+\node at (9.3, -2) {$\frac{b}{a+b+c}$};
+\node at (9.3, -3) {$\frac{c}{a+b+c}$};
+
+\fill (10.2, -1) circle[radius=2pt];  % dea finishes; paths bc
+\fill (10.2, -2) circle[radius=2pt];  % deb finishes; paths ac
+\fill (10.2, -3) circle[radius=2pt];  % dec finishes; paths ab
+
+\draw (10.2, -1) -- (11.2,-0);
+\node at (10.6,-0.4) {$b$};
+\node at (10.6,-0.8) {$c$};
+
+\node at (11.8, 0) {$\frac{\la{b}}{\la{b}+\la{c}}$};
+\draw (10.2, -1) -- (11.2,-1);
+
+\node at (11.8, -1) {$\frac{\la{c}}{\la{b}+\la{c}}$};
+
+\draw (10.2, -2) -- (11.2,-1.6);
+\node at (11.8, -1.6) {$\frac{\la{a}}{\la{a}+\la{c}}$};
+
+\node at (10.6,-1.7) {$a$};
+\node at (10.6,-2) {$c$};
+
+\draw (10.2, -2) -- (11.2,-2.2);
+\node at (11.8, -2.2) {$\frac{\la{c}}{\la{a}+\la{c}}$};
+
+\draw (10.2, -3) -- (11.2,-2.9);
+\node at (11.8, -2.9) {$\frac{\la{a}}{\la{a}+\la{b}}$};
+
+\draw (10.2, -3) -- (11.2,-3.4);
+\node at (11.8, -3.4) {$\frac{\la{b}}{\la{a}+\la{b}}$};
+
+\node at (10.6,-2.8) {$a$};
+\node at (10.6,-3.3) {$b$};
+
+\fill (7.4, 1) circle[radius=2pt];  % da finishes; paths bce
+
+\draw (7.4, 1) -- (8.4,2);
+\draw (7.4, 1) -- (8.4,1);
+\draw (7.4, 1) -- (8.4,0);
+
+\node at (7.9,1.7) {$b$};
+\node at (7.9,1.2) {$c$};
+\node at (7.9,0.7) {$e$};
+
+\node at (9.3, 2) {$\frac{\la{b}}{\la{b}+\la{c}+e}$};
+\node at (9.3, 1) {$\frac{\la{c}}{\la{b}+\la{c}+e}$};
+\node at (9.3, 0) {$\frac{e     }{\la{b}+\la{c}+e}$};
+
+\fill (10.2, 1) circle[radius=2pt];  % da finishes; paths bce
+\draw (10.2, 1) -- (11.2,2);
+\node at (11.8, 2) {$\frac{\la{b}}{\la{b}+\ld{e}}$};
+\draw (10.2, 1) -- (11.2,1);
+\node at (11.8, 1) {$\frac{\ld{e}}{\la{b}+e}$};
+
+\end{tikzpicture}
+```

### 2022-07-07 inst/red_bus_blue_bus.Rmd
red bus blue bus references
more work on red bus/blue bus

@@
 ---
-title: "Red bus, blue bus"
+title: "Red bus, blue bus; and the choice between Florida, California, and Florida plus a nickel"
 output: pdf_document
@@
 
-The red bus-blue bus problem was originally stated by
-@davidson1970.  The other one was the holiday problem.
+In this short discussion document I discuss three related problems:
+
+* The red bus-blue bus problem
+* Debussy-Beethoven problem
+* Holiday problem
+
+The red bus-blue bus problem was originally stated by @chipman1960 and
+discussed by @davidson1970 and @alvo2010 (page 153).
+Debussy-Beethoven was @debreu1960.  I have lost the original reference
+to the Holiday problem.
 
@@
 
-THERE IS ALSO THE THING WITH THE HOLIDAY IN CALIFORNIA AND THE HOLIDAY
-IN FLORIDA AND THE HOLIDAY IN FLORIDA PLUS TEN BUCKS
-
 ```
@@
 {F,FN,C,CN} :  (F,F+N,C,C+N)/(2F+2C+2N)  ??????????????
+```
 
@@
 
+```
 {F,FN,FC,CN}  :  p(F)  = F/(F+C) * F/(F+N)
@@
               :  p(CN) = C/(F+C) * N/(C+N)
+```
 
@@
 
+```
 {F,C,CN}    :  p(F)  = F/(F+C)
@@
             :  p(CN) = C/(F+C) * N/(C+N)
+```
 
@@
 
+```
 L = L(C,B,red,blue)

### 2022-07-07 inst/plackett_luce_monster.Rmd
more work on red bus/blue bus
cheering and an application to red bus-blue bus

@@
 * Runner $i$ might hate runner $j$.  If $i$ finishes, he might expend
-  his spare energy hindering $j$ (by throwing stones or booing , for
+  his spare energy hindering $j$ (by throwing stones or booing, for
   example).  Note that equivalence classes of mutual hatred might
@@
 
+The runner metaphor is not the only one.  We can apply the resulting
+likelihood functions to other cases including the red bus-blue bus
+problem (see `inst/red_bus_blue_bus.Rmd` for more discussion).
+
 First do three runners, and suppose the equivalence classes are
@@
 
- 231: $\frac{p_2}{p_2+p_3+p_1}\cdot\frac{p_3}{p_3+p_1+M}$
+ 231: $\frac{p_2}{p_2+p_3+p_1}\cdot\frac{p_3}{p_3+p_1+M}\qquad\frac{p_2}{p_2+p_1+p_3}\cdot\frac{\lambda p_1}{p_3+\lambda p_1}$
 
- 312: $\frac{p_3}{p_3+p_1+p_2}\cdot\frac{p_1}{p_2+p_1}$
+ 312: $\frac{p_3}{p_3+p_1+p_2}\cdot\frac{p_1}{p_2+p_1}\qquad\frac{p_3}{p_3+p_1+p_2}\cdot\frac{\lambda p_1}{p_3+p_1}$
 
- 321: $\frac{p_3}{p_3+p_2+p_1}\cdot\frac{p_2}{p_2+p_1}$
+ 321: $\frac{p_3}{p_3+p_2+p_1}\cdot\frac{p_2}{p_2+p_1}\qquad\frac{p_3}{p_3+p_2+p_1}\cdot\frac{p_2}{p_2+p_1}$
 
@@
 ```
+
+# Package idiom
+
+We can investigate red bus-blue bus phenomenon (as discussed, in a
+slightly different context, in `inst/red_bus_blue_bus.Rmd`).  Here, we
+consider a person who is given the choice of five transport methods:
+
+* `C`, car
+* `T`, train
+* `RB` a red bus
+* `BB` a blue bus
+* `W` walking
+
+Now, he does not really care what colour the bus is.  If we ask him to
+rank his options, it is highly probable that he will put `RB` and `BB`
+consecutively (because they are essentially indistinguishable).  Can
+we quantify the strength of this effect?  To do this, we define a
+bespoke function `RB_BB_LF()` which returns a `hyper3` log-likelihood
+function corresponding to repeated observations of our commuter's
+reported ranks for the five options:
+
+```{r,defineredbuslf}
+`RB_BB_LF` <- function(s){
+    ec <- c(C=1,T=2,RB=3,BB=3,W=4) # equivalence classes
+    h <- c(1,1,s,1)                # strength of support
+    (
+        cheering3(v=c("RB","BB","C" ,"T","W"),e=ec,h=h)*7 + 
+        cheering3(v=c("BB","RB","T" ,"C","W"),e=ec,h=h)*4 + 
+        cheering3(v=c("T" ,"BB","RB","C","W"),e=ec,h=h)*5 + 
+        cheering3(v=c("W" ,"BB","RB","T","C"),e=ec,h=h)*8 + 
+        cheering3(v=c("C" ,"RB","BB","W","T"),e=ec,h=h)*9 + 
+        cheering3(v=c("BB","C" ,"RB","T","W"),e=ec,h=h)*5
+    )
+}
+```
+
+Above, we see from the function body that he reported `RB,BB,C,T,W`
+seven times, `BB,RB,T,C,W` four times, and so on; perhaps his ranking
+depends on the weather or how tired he is on any given day.  Observe
+that in almost every case he ranks `RB` and `BB` consecutively.
+Function `RB_BB_LF()` takes argument `s` that quantifies the perceived
+similarity between `RB` and `BB`.  For example:
+
+```{r,examplerbbblf,cache=TRUE}
+(H <- RB_BB_LF(s=1.8888))
+(mH <- maxp(H,n=1))
+```
+
+Now to find a profile likelihood function for `s`:
+
+```{r,label=proflikeredbus,cache=TRUE}
+o <- function(s){maxp(RB_BB_LF(s),give=TRUE,n=1)$likes} # optimand
+s <- exp(seq(from=log(1.3),to=log(37),len=17)) # putative similarity measures
+L <- sapply(s,o)
+L <- L-max(L)
+```
+
+We can plot these:
+
+```{r,plotproflike}
+plot(s,L,type="b")
+abline(h=c(0,-2))
+abline(v=1)
+plot(log(s),L,type="b")
+abline(h=c(0,-2))
+abline(v=0)
+```
+
+And formally maximize the likelihood:
+
+```{r,label=maxlikesim,cache=TRUE}
+(osup <- optimize(o,c(6,10),maximum=TRUE))
+```
+
+So a likelihood ratio test of the null that $S=1$ would be:
+
+```{r,label=LRT_RB,cache=TRUE}
+(suppdif <- o(osup$maximum) - o(1))
+```
+
+Falling just shy of Edwards's two-units-of-support criterion.

### 2022-07-07 inst/cheering.R

@@ -1,84 +1,0 @@
-`race_cheering3` <- function(v,e,help){
-  ## competitors abcdef, with (abc) 1.1 mutual friends with helpfulness 1.1 and (de) helpfulness 1.3:
-  ## race_cheering3(c("a","b","c","d","e","f"),c(a=1,b=1,c=1,d=2,e=2,f=3), help=c(1.1,1.3))
-
-  stopifnot(all(table(v)==1))
-  stopifnot(all(table(names(e))==1))
-  stopifnot(all(sort(names(e)) == sort(v)))
-    
-  out <- hyper3()
-
-  finished <- NULL # these are competitors who have finished and are cheering their friends
-  still_running <- v
-  while(length(still_running) > 0){
-    winner <- still_running[1]  # winner is the first element of the field
-    winners_friends <- names(e[e==e[winner]])
-    jjn <- 1  # numerator
-    names(jjn) <- winner
-
-    winner_ec_number <- e[names(e)==winner]  # this would be an integer; "ec" means "equivalence class"
-    finished_ec_number <- e[names(e) %in% finished]
-    if(winner_ec_number %in% finished_ec_number){  # that is, if the winner has a friend who has finished...
-      jjn[] <- help[winner_ec_number]   # ...then he gets extra strength.
-    }
-
-    jjd <- rep(1,length(still_running)) # denominator
-    names(jjd) <- still_running
-    for(i in seq_along(still_running)){
-      runner_ec_number <- e[names(e) == still_running[i]]
-      if(length(finished_ec_number)>0){
-        if(length(runner_ec_number) ==0){browser()}
-        out
-        if(runner_ec_number %in% finished_ec_number){
-          jjd[i] <- help[runner_ec_number]
-        }
-      }
-    }
-    out[jjn] %<>% inc   # numerator
-    out[jjd] %<>% dec   # denominator
-    
-    finished <- c(finished,winner)     # the winner is now finished ...
-    still_running <- still_running[-1] # ... and not still running
-  } #while loop closes
-  return(out)
-}
-
-H3maker  <- function(help){
-    (
-        race_cheering3(c("a","b","c"),c(a=1,b=1,c=2), help=help)*7    +
-        race_cheering3(c("a","c","b"),c(a=1,b=1,c=2), help=help)*5    +
[…]

### 2022-07-07 inst/plackett_luce_monster.Rmd
slight change to observation

@@
     (
-        cheering3(v=c("RB","BB","C" ,"T","W"),e=ec,h=h)*7 + 
-        cheering3(v=c("BB","RB","T" ,"C","W"),e=ec,h=h)*4 + 
-        cheering3(v=c("T" ,"BB","RB","C","W"),e=ec,h=h)*5 + 
-        cheering3(v=c("W" ,"BB","RB","T","C"),e=ec,h=h)*8 + 
-        cheering3(v=c("C" ,"RB","BB","W","T"),e=ec,h=h)*9 + 
-        cheering3(v=c("BB","C" ,"RB","T","W"),e=ec,h=h)*5
+        cheering3(v=c("RB","BB","C" ,"T","W"),e=ec,h=h)*3 + 
+        cheering3(v=c("BB","RB","T" ,"C","W"),e=ec,h=h)*2 + 
+        cheering3(v=c("T" ,"BB","RB","C","W"),e=ec,h=h)*2 + 
+        cheering3(v=c("W" ,"BB","RB","T","C"),e=ec,h=h)*4 + 
+        cheering3(v=c("C" ,"RB","BB","W","T"),e=ec,h=h)*4 + 
+        cheering3(v=c("BB","C" ,"RB","T","W"),e=ec,h=h)*3
     )
@@
 Above, we see from the function body that he reported `RB,BB,C,T,W`
-seven times, `BB,RB,T,C,W` four times, and so on; perhaps his ranking
-depends on the weather or how tired he is on any given day.  Observe
-that in almost every case he ranks `RB` and `BB` consecutively.
-Function `RB_BB_LF()` takes argument `s` that quantifies the perceived
-similarity between `RB` and `BB`.  For example:
+three times[first row], `BB,RB,T,C,W` twice [second row], and so on;
+perhaps his ranking depends on the weather or how tired he is on any
+given day.  Observe that in almost every case he ranks `RB` and `BB`
+consecutively.  Function `RB_BB_LF()` takes argument `s` that
+quantifies the perceived similarity between `RB` and `BB`.  For
+example:
 
@@
 o <- function(s){maxp(RB_BB_LF(s),give=TRUE,n=1)$likes} # optimand
-s <- exp(seq(from=log(1.3),to=log(37),len=17)) # putative similarity measures
+s <- exp(seq(from=log(1.3),to=log(47),len=17)) # putative similarity measures
 L <- sapply(s,o)
@@
 ```{r,label=LRT_RB,cache=TRUE}
-(suppdif <- o(osup$maximum) - o(1))
+(suppdiff <- o(osup$maximum) - o(1))
 ```
 
-Falling just shy of Edwards's two-units-of-support criterion.
+Easily satisfying Edwards's two-units-of-support criterion; Wilks gives us an asymptotic $p$-value:
+
+```{r,label=RBBBpval}
+pchisq(suppdiff*2,df=1,lower.tail=FALSE)
+```

### 2022-07-08 inst/home_advantage.Rmd
nicer printing, do not echo messages
home advantage via hpyer3

@@
 
+```{r,label=libload,echo=FALSE,print=FALSE,warning=FALSE,include=FALSE}
+defaultW <- getOption("warn") 
+options(warn = -1) 
+library("hyper2",quietly=TRUE)
+options(warn = defaultW)
+```
+
 (takes about fifteen minutes to run without cache)
 
-Here I analyse a small synthetic dataset for home ground advantage, then go on to discuss
-a dataset presented by Agresti.
-
-First load the package:
-```{r,label=libload}
-library(hyper2)
-```
-
-Now generate the dataset:
+Here I analyse a small synthetic dataset for home ground advantage,
+then go on to discuss a dataset presented by Agresti.  First generate
+the dataset by hand:
 
@@
 
+# hyper3 idiom
+
+We may apply `hyper3` idiom to the dataset.  First we show
+`home_away3()` being used with a specific value for $\lambda$:
+
+```{r, label=hyper3_homeaway,cache=TRUE}
+(o <- home_games_won + 1i*away_games_won)
+f <- function(l){ maxp(home_away3(o,l),give=TRUE,n=1)$likes}
+l <- exp(seq(from=log(2),to=log(6),len=10))
+L <- sapply(l,f)
+L <- L - max(L)
+```
+
+```{r,label=plothyper3home}
+plot(l,L,type="b")
+abline(h=c(0,-2))
+plot(log(l),L,type="b")
+abline(h=c(0,-2))
+```
+
+```{r, label=hyper3_baseball,cache=TRUE}
+baseball_table
+f <- function(l){ maxp(home_away3(baseball_table,l),give=TRUE,n=1)$likes}
+l <- exp(seq(from=log(0.8),to=log(1.9),len=14))
+L <- sapply(l,f)
+L <- L - max(L)
+```
+
+```{r,label=plothyper3baseball}
+plot(log(l),L,type="b")
+abline(h=c(0,-2))
+abline(v=0)
+```
+
+Now a formal test:
+
+```{r,label=optimizebasebasll,cache=TRUE}
+o <- optimize(f,c(0.2,0.4),maximum=TRUE)
+o
+```
+
+```{r}
+f(o$maximum) - f(0)
+```
+
 ### Package dataset {-}

### 2022-07-09 inst/plackett_luce_monster.Rmd
further work on university ranking

@@
 library("partitions")
+library("pdftools")
 ```
@@
 </p>
+
+%
 
@@
 ```
+
+# University ranking analysis
+
+Here we use a dataset of university rankings, `timesData.csv`, taken
+from
+`https://github.com/arnaudbenard/university-ranking/blob/master/timesData.csv`.
+
+```{r}
+a <- read.table("timesData.csv",sep=",", header=TRUE)
+wanted <- c("California Institute of Technology", "Harvard University", 
+"Massachusetts Institute of Technology", "Princeton University", 
+"Stanford University", "University of Cambridge", "University of Oxford")
+names(wanted) <- c("cal","harv","mass","prin","stan","cam","ox")
+
+a <- a[a$university_name %in% wanted,]
+a <- cbind(a,"top7rank"=0)
+for(y in unique(a$year)){
+    a[a$year==y,"top7rank"] <- order(
+     			  as.numeric(a[a$year==y,"world_rank"]) + 
+                                     a[a$year==y,"research"]/1e6,
+decreasing=TRUE)}
+a <- a[,c("top7rank","university_name","year")]
+a <- reshape(a,idvar="university_name",timevar="year",direction="wide")
+for(i in seq_len(nrow(a))){
+   a$university_name[i] <- names(which(wanted == a$university_name[i]))
+}
+rownames(a) <- a$university_name
+a <- a[,-1]
+colnames(a) <- paste("Y",2011:2016,sep="")
+a
+H <- ordertable2supp(a)
+```
+
+```{r,label=testuniv,cache=TRUE}
+equalp.test(H)
+samep.test(H,c("ox","cam"))
+```
+
+Start to use `hyper3` idiom:
+
+```{r}
+H3 <- hyper3()
+for(i in seq_len(ncol(a))){
+  jj <- rep("",nrow(a))
+  jj[a[,i]] <- rownames(a)
+  print(jj)
+}
+```

### 2022-07-09 inst/plackett_luce_monster.Rmd
more work

@@
 
-```{r}
-H3 <- hyper3()
+```{r,label=defineoxcam}
+H3 <- function(oxcam){
+  out <- hyper3()
   for(i in seq_len(ncol(a))){
@@
     jj[a[,i]] <- rownames(a)
-  print(jj)
+    out <- out + cheering3(v=jj,e=c(ox=1,cam=1,prin=2, stan=3, mass=4, harv=5, cal=6), help=c(oxcam,1,1,1,1,1))
+  }
+  return(out)
 }
 ```
+
+```{r,label=calculateoxcam,cache=TRUE}
+maxp(H3(oxcam=1.8888),give=TRUE,n=6)$likes
+```

### 2022-07-09 inst/plackett_luce_monster.Rmd
nice profile likelihood graph

@@
 ```{r,label=calculateoxcam,cache=TRUE}
-maxp(H3(oxcam=1.8888),give=TRUE,n=6)$likes
+o <- function(oxcam){maxp(H3(oxcam),give=TRUE,n=1)$likes}
+oc <- exp(seq(from=log(0.5),to=log(5),len=15))
+L <- sapply(oc,o)
+L <- L - max(L)
 ```
+
+```{r,label=plotoxcam}
+plot(log(oc),L,type="b")
+abline(v=0)
+```

### 2022-07-11 inst/home_advantage.Rmd
hyper3 analysis of home advantage

--- 2022-04-21 inst/home_advantage.Rmd
@@
 
+# hyper3 idiom
+
+We may apply `hyper3` idiom to the dataset. First a formal test:
+
+```{r,label=optimizebaseball,cache=TRUE}
+baseball_table
+f <- function(l){ maxp(home_away3(baseball_table,l),give=TRUE,n=1)$likes}
+o <- optimize(f,c(0.8,1.6),maximum=TRUE)
+o
+maxL <- f(o$maximum)
+W <- maxL - f(1)    # f(1) being the likelihood of the null
+W
+pchisq(2*W,df=1,lower.tail=FALSE)
+```
+
+The pvalue is significant.  Now plot a profile likelihood (support)
+curve:
+
+```{r, label=hyper3_baseball,cache=TRUE}
+jj1  <- seq(from=log(o$maximum),to=log(1.9),len=7)
+jj2  <- seq(from=log(o$maximum),to=log(0.8),len=7)
+l <- sort(unique(exp(c(jj1,jj2))))
+lam  <- sort(c(l,o$maximum))
+L <- sapply(lam,f) - maxL
+```
+
+```{r,label=plothyper3baseball}
+plot(log(lam),L,type="b")
+abline(h=c(0,-2))
+abline(v=0)
+
+plot(lam,L,type="b")
+abline(h=c(0,-2))
+```
+
 ### Package dataset {-}

### 2022-07-13 inst/eurovision.Rmd
use sticker in Rmd
start work on Eurovision

@@
 
-```{r,morestuff}
+```{r,morestuff,cache=TRUE}
 eurovision <- hyper2()
@@
 
+## Testing nonindependence using `hyper3` idiom
+
+First we use `hyper2`:
+
+```{r,label=loadeurohyper2,cache=TRUE}
+a <- read.table("euro_ranks.txt",header=TRUE)
+jj <- table(a$country)
+wanted_countries <- names(jj[jj >= 10])
+a <- a[a$country %in% wanted_countries,]
+H <- hyper2()
+for(y in unique(a$year)){
+   H <- H + race((a$country)[a$year == y])
+}
+H
+```
+
+Find the evaluate:
+
+```{r,findhyper2euroevaluate,cache=TRUE}
+mH <- maxp(H)
+```
+
+```{r,plotmH}
+dotchart(mH)
+```
+
+Now use `hyper3` and follow `RB_BB_LF()` in
+`plackett_luce_monster.Rmd`
+
 ### Package dataset {-}

### 2022-07-13 inst/plackett_luce_monster.Rmd
typo

@@
 Above, we see from the function body that he reported `RB,BB,C,T,W`
-three times[first row], `BB,RB,T,C,W` twice [second row], and so on;
+three times [first row], `BB,RB,T,C,W` twice [second row], and so on;
 perhaps his ranking depends on the weather or how tired he is on any

### 2022-07-14 inst/eurovision.Rmd
fixes #146

@@
 
-```{r}
+```{r,printpref}
 preference
@@
 }
-H
+summary(H)
 ```
@@
 Now use `hyper3` and follow `RB_BB_LF()` in
-`plackett_luce_monster.Rmd`
+`plackett_luce_monster.Rmd`.  The blocs follow those of Price 2020.
+
+```{r,plm}
+scandinavia <- c("Denmark","Finland","Iceland","Norway","Sweden")
+
+balkans  <- c("Italy","Serbia","Albania","BosHerz") # red
+eastern  <- c("Russia","Ukraine","Moldova","Armenia","Azerbaijan", "Malta","Israel") # purple
+black    <- c("Hungary","Romania","Greece","Cyprus") # orange
+iberia   <- c("Spain","Germany") # blue
+northern <- c("Denmark", "Estonia", "Finland", "France", "Iceland", "Lithuania", "Netherlands", "Norway", "Sweden", "UK") # green
+
+l <- list(balkans=balkans,eastern=eastern,black=black,iberia=iberia,northern=northern)
+
+ab <- c(balkans,eastern,black,iberia,northern)
+stopifnot(all(ab %in% wanted_countries))
+stopifnot(all(wanted_countries %in% ab))
+
+list2ec <- function(l){
+  jj <- stack(setNames(l,seq_along(l)))
+  out <- as.integer(c(jj$ind))
+  names(out) <- jj$values
+  return(out)
+}
+
+print(a$year)
+
+bloc <- function(lam){
+  out <- hyper3()
+  e <- list2ec(l)
+  for(y in unique(a$year)){
+     v <- a$country[a$year==y]
+     ec <- e[names(e) %in% v]
+     out <- out + cheering3(v=v, e=ec,h=rep(lam,length(unique(e))))
+   } 
+  return(out)
+}
+
+Ha <- bloc(1.888)
+Hb <- bloc(1.9)
+```
+
+Now maximize it:
+
+```{r,maxbloc,cache=TRUE}
+maxp(Ha,n=2,give=TRUE)
+maxp(Hb,n=2,give=TRUE)
+```
+
+```{r,maxhelp,cache=TRUE}
+o <- function(lam){
+   H <- bloc(lam)
+   return(maxp(Hb,give=TRUE,n=1)$likes)
+}
+
+o(1)
+o(2)
+
+lam <- seq(from=0.6,to=1.5,len=4)
+```
+
+# References {-}
+
+A. Price 2020. "Identifying voting blocs in the Eurovision Song Contest".  Towards Data Science,
+`https://medium.com/towards-data-science/identifying-voting-blocs-in-the-eurovision-song-contest-4792065fc337`
 

### 2022-07-14 R/cheering.R

@@ -1,17 +1,26 @@
 `cheering3` <- function(v,e,help){
 
-  ## competitors abcdef, with two groups of mutual friends: (abc) with
-  ## mutual helpfulness 1.1, and (de) with mutual helpfulness 1.3:
+  ## competitors abcdef, with two groups of mutual friends: (abd) with
+  ## mutual helpfulness 1.1, and (cf) with mutual helpfulness 1.3;
+  ## competitors e and g are in groups of 1.
 
   ## use-case:
-  ## cheering3(c("a","b","c","d","e","f"),c(a=1,b=1,c=1,d=2,e=2,f=3),help=c(1.1,1.3))
-
-  
+  ## cheering3(v=letters[1:7],e=c(a=1,b=1,c=2,d=1,f=2),help=c(1.88,1.66))
 
 
   stopifnot(all(table(names(e))==1))
-  stopifnot(all(sort(names(e)) == sort(unique(v))))
-    
+  stopifnot(all(e>0))
+  stopifnot(all(e==round(e)))
+  stopifnot(all(table(v)==1))
+  stopifnot(all(names(e) %in% v)) 
+
+  no_group <- !(v %in% names(e))
+  if(any(no_group)){
+    jj <- seq(from=max(e)+1,len=sum(no_group))
+    names(jj) <- v[no_group]   # a character vector
+    e <- c(e,jj)
+  }
+
   out <- hyper3()
 
   finished <- NULL # these are competitors who have finished and are cheering their friends
@@ -24,6 +33,7 @@
 
     winner_ec_number <- e[names(e)==winner]  # this would be an integer; "ec" means "equivalence class"
     finished_ec_number <- e[names(e) %in% finished]
+
     if(winner_ec_number %in% finished_ec_number){  # that is, if the winner has a friend who has finished...
       jjn[] <- help[winner_ec_number]   # ...then he gets extra strength.
     }
@@ -34,7 +44,6 @@
       runner_ec_number <- e[names(e) == still_running[i]]
       if(length(finished_ec_number)>0){
         if(length(runner_ec_number) ==0){browser()}
-        out
         if(runner_ec_number %in% finished_ec_number){
           jjd[i] <- help[runner_ec_number]
         }

### 2022-07-14 inst/eurovision.Rmd
more work

@@
 
-print(a$year)
-
 bloc <- function(lam){
@@
 }
-
-Ha <- bloc(1.888)
-Hb <- bloc(1.9)
 ```
 
-Now maximize it:
+```{r,defineoptimand}
+o <- function(lam){  # maximize function o()
+   H <- bloc(lam)
+   return(maxp(H,give=TRUE,n=1)$likes)
+}
 
-```{r,maxbloc,cache=TRUE}
-maxp(Ha,n=2,give=TRUE)
-maxp(Hb,n=2,give=TRUE)
+```{r,maximi,cache=TRUE}
+lam <- seq(from=0.5,to=2,len=10)
+L <- sapply(lam,o)
 ```
 
-```{r,maxhelp,cache=TRUE}
-o <- function(lam){
-   H <- bloc(lam)
-   return(maxp(Hb,give=TRUE,n=1)$likes)
-}
-
-o(1)
-o(2)
-
-lam <- seq(from=0.6,to=1.5,len=4)
+```{r,plotmaximi}
+plot(lam,L-max(L),type="b",pch=16)
 ```

### 2022-07-14 inst/eurovision.Rmd
improvements to graph

@@
 ```{r,maximi,cache=TRUE}
-lam <- seq(from=0.5,to=2,len=10)
+lam <- seq(from=0.6,to=2,len=20)
 L <- sapply(lam,o)
@@
 plot(lam,L-max(L),type="b",pch=16)
+abline(h=c(0,-2))
+abline(v=1)
 ```

### 2022-07-15 inst/plackett_luce_monster.Rmd
generalized generalized Plackett-Luce

@@
 ```
+
+# Five nations championship
+
+The five nations rugby championship was held from 1910 to 1999 and
+file five_nations.txt` shows the order statistic for England (E),
+Scotland (S), Ireland (I), Wales (W), and France (F).
+
+`https://en.wikipedia.org/wiki/Six_Nations_Championship`
+
+Here is `hyper2` analysis:
+
+```{r,load5nationsdata,cache=TRUE}
+a <- as.matrix(read.table("five_nations.txt",header=FALSE))
+head(a)
+H <- hyper2()
+for(i in seq_len(nrow(a))){
+  H <- H + race(a[i,-1])
+} 
+mH <- maxp(H)
+pie(mH)
+equalp.test(H)
+```
+
+Now use `hyper3` to see whether teams do better following a win:
+
+```{r,rugby3,cache=TRUE}
+rugby <- function(lambda){
+  H3a <- hyper3()
+  for(i in seq(from=2,to=nrow(a))){
+        last_year_winner <- a[i-1,2] # e.g. "W" or "E"
+        H3a <- H3a + ordervec2supp3a(a[i,-1],nonfinishers=NULL,helped=last_year_winner,lambda=lambda)
+  }
+  return(H3a)
+}
+rugby(1.888)
+rugby(1.111111)
+```
+
+```{r,rugby3max, cache=TRUE}
+maxp(rugby(1.8),n=1,give=TRUE)
+maxp(rugby(1.9),n=1,give=TRUE)
+```

### 2022-07-19 inst/global_liveability_ranking.Rmd
global liveability ranking work

---
title: "Global liveability ranking"
output: html_document
---

```{r setup, include=FALSE}
set.seed(0)
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("magrittr")
options("digits" = 5)
```

<p style="text-align: right;">
![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
</p>

[The global liveability ranking](https://en.wikipedia.org/wiki/Global_Liveability_Ranking)
shows where you would like to live.

```{r showglrtable}
readLines("global_liveability_ranking.txt")
```

First we will turn it into a `hyper2` object:

```{r readglrtable}
o <- readLines("global_liveability_ranking.txt") %>% strsplit(" ") %>% lapply(function(l){l[-1]})
H <- hyper2()
for(i in seq_along(o)){
  H <- H  + race(o[[i]])
}
```

And do some tests:

```{r testgrl, cache=TRUE}
mH <- maxp(H)
pie(mH)
equalp.test(H)
```

Now we tether some of the cities together:

```{r, tether}
elastic <- list(
Austria = c("Vienna"),
Denmark = c("Copenhagen"),
Switzerland = c("Zurich","Geneva"),
Canada = c("Calgary","Vancouver","Toronto"),
Germany = c("Frankfurt","Hamburg"),
Netherlands = c("Amsterdam"),
Japan = c("Osaka","Tokyo"),
Australia = c("Melbourne","Adelaide","Perth","Brisbane","Sydney"),
NZ = c("Auckland","Wellington"),
Finland = c("Helsinki")
)
e <- NULL
for(i in seq_along(elastic)){
  jj <- rep(i,length(elastic[[i]]))
  names(jj) <- elastic[[i]]
  e <- c(e,jj)
}
e 
```

Now create a `hyper3` object:

```{r,createhyper3func}
ML <- function(h){
   H3 <- hyper3()
   for(r in o){ H3 <- H3 + cheering3(r,e=e,help=rep(h,10))}
   return(H3)
}
f <- function(h){maxp(ML(h),give=TRUE,n=1)$likes}
```

And plot a profile likelihood

```{r,plliveable,cache=TRUE}
l <- seq(from=1,to=5,len=10)
L <- sapply(l,f)
```

and plot it:

```{r}
plot(log(l),L-max(L),type='b')
```

### 2022-07-20 inst/global_liveability_ranking.Rmd
more work

@@
 ```{r showglrtable}
-readLines("global_liveability_ranking.txt")
+(glr <- readLines("global_liveability_ranking.txt"))
 ```
@@
 ```{r readglrtable}
-o <- readLines("global_liveability_ranking.txt") %>% strsplit(" ") %>% lapply(function(l){l[-1]})
+(o <- glr  %>% strsplit(" ") %>% lapply(function(l){l[-1]}))
+```
+
+Will remove Helsinki, Brisbane, and Hamburg on account of their having
+very low rank scores (if we leave them in, the Hessian becomes
+non-positive-definite at the evaluate).
+
+```{r,removeHelsinki_and_Hamburg}
+o %<>%  lapply(function(x){x[!(x %in% c("Hamburg","Helsinki","Brisbane"))]})
+o
+```
+
+```{r,turnintohyper2,cache=TRUE}
 H <- hyper2()
@@
 ```{r testgrl, cache=TRUE}
-mH <- maxp(H)
+(mH <- maxp(H))
 pie(mH)
 equalp.test(H)
+```
+
+```{r,calculatemoreinfo,cache=TRUE}
+jjm <- maxp(H,n=1,give=TRUE,hessian=TRUE)
+```
+
+```{r}
+jjm$hessian
+diag(jjm$hessian)
+eigen(jjm$hessian)$values
+```
+
+So the Hessian is positive-definite [all the eigenvalues are
+positive].  Also, the evaluate is robust, as indicated by
+`consistency()`:
+
+```{r,doconsis,cache=TRUE}
+consistency(H)
 ```
@@
 ```
+
+# A More coarse division
+
+We will try dividing the countries into continents:
+
+```{r,label=elasticcontinent}
+elastic_continent <- list(
+Europe      = c("Vienna","Copenhagen","Zurich","Geneva","Frankfurt","Hamburg","Amsterdam","Helsinki"),
+America     = c("Calgary","Vancouver","Toronto"),
+Asia        = c("Osaka","Tokyo"),
+Australasia = c("Melbourne","Adelaide","Perth","Brisbane","Sydney","Auckland","Wellington")
+)
+
+ec <- NULL
+for(i in seq_along(elastic_continent)){
+  jj <- rep(i,length(elastic_continent[[i]]))
+  names(jj) <- elastic_continent[[i]]
+  ec <- c(ec,jj)
+}
+ec 
+MLc <- function(h){
+   H3 <- hyper3()
+   for(r in o){ H3 <- H3 + cheering3(r,e=ec,help=rep(h,4))}
+   return(H3)
+}
+fc <- function(h){maxp(ML(h),give=TRUE,n=1)$likes}
+```
+
+```{r,evaluatefc,cache=TRUE}
+l <- seq(from=0.8,to=3,len=10)
+L <- sapply(l,fc)
+```
+
+```{r,plotfc}
+plot(l,L-max(L),type="b")
+```

### 2022-07-20 inst/global_liveability_ranking.Rmd
list2nv()

@@
 )
-e <- NULL
-for(i in seq_along(elastic)){
-  jj <- rep(i,length(elastic[[i]]))
-  names(jj) <- elastic[[i]]
-  e <- c(e,jj)
-}
-e 
 ```
@@
    H3 <- hyper3()
-   for(r in o){ H3 <- H3 + cheering3(r,e=e,help=rep(h,10))}
+   for(r in o){ H3 <- H3 + cheering3(r,e=list2nv(elastic),help=rep(h,10))}
    return(H3)
@@
 ```
+
+## Country-level analysis
+
+```{r}
+glr
+```

### 2022-07-20 R/hyper3.R

@@ -578,3 +578,10 @@ stop("not yet written")
     } # i loop closes
     return(H)
 }
+
+
+`list2nv` <- function(L){
+  out <- rep(seq_along(L),times=unlist(lapply(L,length)))
+  names(out) <- c(L,recursive=TRUE)
+  return(out)
+}

### 2022-07-24 inst/global_liveability_ranking.Rmd
not a useful analysis, reasons discussed

@@
 
+_This analysis is essentially without value, see the very end for why_
+
+(takes about an hour to process without cache)
+
 <p style="text-align: right;">
@@
 
-[The global liveability ranking](https://en.wikipedia.org/wiki/Global_Liveability_Ranking)
-shows where you would like to live.
+[The global liveability
+ranking](https://en.wikipedia.org/wiki/Global_Liveability_Ranking)
+shows the nicest places to live.
 
@@
 
-# A More coarse division
+# A coarser division
 
@@
 plot(l,L-max(L),type="b")
+abline(v=1)
 ```
@@
 
+We need to replace each city with its country:
+
+```{r,label=replacecity}
+elastic
+(jj <- list2nv(elastic,FALSE))
+o_countries <- lapply(o, function(x){jj[names(jj) %in% x]})
+(o_countries <- lapply(o_countries,as.vector))
+```
+
+We see that NZ and Australia are dominated so we should remove them:
+
 ```{r}
-glr
+(o_countries <- lapply(o_countries,function(x){x[!(x %in% c("NZ","Australia","Japan"))]}))
 ```
+
+But hang on, if we continue to remove dominated countries we
+eventually remove every country that is not Austria.  So that was a big
+fat waste of time.

### 2022-07-25 inst/race3.Rmd
nonfinishers in cheering3()

@@
 
+## Constructors championship with `cheering` or `tether` functionality
+
+```{r,Bahrain2021_to_support3}
+constructor_2021_table
+(cons <- constructor_2021_table[,1])
+(BHR <- as.numeric(constructor_2021_table[,2]))  # Bahrain
+names(BHR) <- cons
+BHR
+finishers <- BHR[!is.na(BHR)]
+finishers <- names(finishers[order(finishers)])  # in order
+nonfinishers <- names(BHR[is.na(BHR)])  # no particular order
+finishers
+nonfinishers
+u <- unique(finishers)
+e <- seq_along(u)
+names(e) <- u
+e
+help <- rep(1.8,length(u))
+e
+help
+cheering3(v=finishers,e=e,help=help,nonfinishers=nonfinishers)
+```
+
+OK so now generalize it:
+
+```{r,definevenue3}
+venue3 <- function(ven,cons,help){
+   suppressWarnings(names(ven) <- cons)
+   finishers <- ven[!is.na(ven)]
+   finishers <- names(finishers[order(finishers)])  # in order
+   nonfinishers <- names(ven[is.na(ven)])  # no particular order
+   u <- unique(finishers)
+   e <- seq_along(u)
+   names(e) <- u
+   help <- rep(help,length(u))
+   cheering3(v=finishers,e=e,help=help,nonfinishers=nonfinishers)
+}  # venue3() definition closes
+```
+
+Now use `venue3()` for all the venues in the 2021 season, with a common help:
+
+```{r,defineallvenues}
+allvenues <- function(h){
+  H3 <- hyper3()
+  cons <- constructor_2021_table[,1]
+  for(i in seq_len(ncol(constructor_2021_table)-1)){
+     H3 <- H3 + venue3(ven=as.numeric(constructor_2021_table[,i]), cons=cons,help=h)
+  }
+  return(H3)
+}
+```
+
+And maximize it for different help values:
+
+```{r,helpseqsupport,cache=TRUE}
+o <- function(h){maxp(allvenues(h),give=TRUE,n=1)$likes}
+l <- seq(from=0.5,to=4,len=6)
+L <- sapply(l,o)
+```
+
+```{r,plotsupport}
+plot(l,L-max(L),type='b')
+```
+
+Above, we are testing (and rejecting) the hypothesis that the various
+constructors are independent of one another.  We reject the hypothesis
+that $h=1$.  Noting that we control both for constructors'
+Plackett-Luce strengths, and on the assumption that the two drivers of
+any given constructor are of equal strengths, we still reject the
+hypothesis of conditional independence.
+
+```{r,label=maxwithfour,cache=TRUE}
+o <- maxp(allvenues(4),give=TRUE,n=1,hessian=TRUE)
+```
+
+```{r,analyse4}
+pie(o$par)
+eigen(o$hessian)$values
+```
+
 ### Package dataset {-}

### 2022-07-25 R/cheering.R

@@ -1,4 +1,4 @@
-`cheering3` <- function(v,e,help){
+`cheering3` <- function(v,e,help,nonfinishers=NULL){
 
   ## competitors abcdef, with two groups of mutual friends: (abd) with
   ## mutual helpfulness 1.1, and (cf) with mutual helpfulness 1.3;
@@ -36,23 +36,20 @@
       jjn[] <- help[winner_ec_number]   # ...then he gets extra strength.
     }
 
-    jjd <- rep(1,length(still_running)) # denominator
-    names(jjd) <- still_running
-    for(i in seq_along(still_running)){
-      runner_ec_number <- e[names(e) == still_running[i]]
+    jjd <- rep(1,length(c(still_running,nonfinishers))) # denominator
+    names(jjd) <- c(still_running,nonfinishers)
+    for(i in seq_along(c(still_running))){
+      runner_ec_number <- e[names(e) == c(still_running,nonfinishers)[i]]
       if(length(finished_ec_number)>0){
         if(length(runner_ec_number) ==0){browser()}
-        if(runner_ec_number %in% finished_ec_number){
-          jjd[i] <- help[runner_ec_number]
-        }
+        if(runner_ec_number %in% finished_ec_number){ jjd[i] <- help[runner_ec_number] }
       }
     }
     out[jjn] %<>% inc   # numerator
     out[jjd] %<>% dec   # denominator
-    
     finished <- c(finished,winner)     # the winner is now finished ...
     still_running <- still_running[-1] # ... and not still running
-  } #while loop closes
+  } # while(length(still_running)>0) loop closes
   return(out)
 }
 

### 2022-07-25 inst/plackett_luce_monster.Rmd
typo

@@
 The five nations rugby championship was held from 1910 to 1999 and
-file five_nations.txt` shows the order statistic for England (E),
+file `five_nations.txt` shows the order statistic for England (E),
 Scotland (S), Ireland (I), Wales (W), and France (F).

### 2022-07-28 inst/race3.Rmd
bugfix

--- 2022-06-30 inst/race3.Rmd
@@
 
+## Constructors championship with `cheering` or `tether` functionality
+
+```{r,Bahrain2021_to_support3}
+constructor_2021_table
+(cons <- constructor_2021_table[,1])
+(BHR <- as.numeric(constructor_2021_table[,2]))  # Bahrain
+names(BHR) <- cons
+BHR
+finishers <- BHR[!is.na(BHR)]
+finishers <- names(finishers[order(finishers)])  # in order
+nonfinishers <- names(BHR[is.na(BHR)])  # no particular order
+finishers
+nonfinishers
+u <- unique(c(finishers,nonfinishers))
+e <- seq_along(u)
+names(e) <- u
+e
+help <- rep(1.8,length(u))
+e
+help
+cheering3(v=finishers,e=e,help=help,nonfinishers=nonfinishers)
+```
+
+OK so now generalize it:
+
+```{r,definevenue3}
+venue3 <- function(ven,cons,help){
+   suppressWarnings(names(ven) <- cons)
+   finishers <- ven[!is.na(ven)]
+   finishers <- names(finishers[order(finishers)])  # in order
+   nonfinishers <- names(ven[is.na(ven)])  # no particular order
+   u <- unique(c(finishers,nonfinishers))
+   e <- seq_along(u)
+   names(e) <- u
+   help <- rep(help,length(u))
+   cheering3(v=finishers,e=e,help=help,nonfinishers=nonfinishers)
+}  # venue3() definition closes
+```
+
+Now use `venue3()` for all the venues in the 2021 season, with a common help:
+
+```{r,defineallvenues,cache=TRUE}
+set.seed(9)
+allvenues <- function(h){
+  H3 <- hyper3()
+  cons <- constructor_2021_table[,1]
+  for(i in seq(from=2,to=ncol(constructor_2021_table),by=1)){
+     H3 <- H3 + venue3(ven=as.numeric(constructor_2021_table[,i]), cons=cons,help=h)
+  }
+  return(H3)
+}
+## And maximize it for different help values:
+
+o <- function(h){maxp(allvenues(h),give=TRUE,n=1)$likes}
+l <- seq(from=0.9,to=4,len=11)
+L <- sapply(l,o)
+```
+
+```{r,plotsupport}
+plot(l,L-max(L),type='b')
+```
+
+Above, we are testing (and rejecting) the hypothesis that the various
+constructors are independent of one another.  We reject the hypothesis
+that $h=1$.  Noting that we control both for constructors'
+Plackett-Luce strengths, and on the assumption that the two drivers of
+any given constructor are of equal strengths, we still reject the
+hypothesis of conditional independence.
+
+```{r,label=maxwithfour,cache=TRUE}
+o <- maxp(allvenues(4),give=TRUE,n=1,hessian=TRUE)
+```
+
+```{r,analyse4}
+pie(o$par)
+eigen(o$hessian)$values
+```
+
 ### Package dataset {-}

### 2022-07-28 inst/race3.Rmd
more cool work

@@
 o <- function(h){maxp(allvenues(h),give=TRUE,n=1)$likes}
-l <- seq(from=0.9,to=4,len=11)
+l <- seq(from=0.9,to=2.5,len=11)
 L <- sapply(l,o)
@@
 
+```{r,findmaxlikeusingoptim,cache=TRUE}
+jj <- optimize(o,c(1.3,1.5),maximum=TRUE)
+jj
+```
+
+```{r,likelihoodnull,cache=TRUE}
+(jjnull <- o(1))
+```
+
+```{r}
+pchisq(2*(jj$objective-jjnull),df=1,lower.tail=FALSE)
+```
+
 ```{r,label=maxwithfour,cache=TRUE}
-o <- maxp(allvenues(4),give=TRUE,n=1,hessian=TRUE)
+mv_max  <- maxp(allvenues(jj$maximum),give=TRUE,n=1)
+mv_null <- maxp(allvenues(1),give=TRUE,n=1)
 ```
@@
 ```{r,analyse4}
-pie(o$par)
-eigen(o$hessian)$values
+x <- mv_max$par
+y <- mv_null$par
+par(pty='s')
+plot(x,y,asp=1,pch=16,xlim=c(0,0.2),ylim=c(0,0.2),xlab="null",ylab="max like")
+for(i in seq_along(x)){text(x[i],y[i],names(x)[i],pos=4,cex=0.7,col='gray')}
+abline(0,1)
 ```

### 2022-07-29 inst/setpartitions_hyper2.Rmd
start work on issue # 148

---
title: "setpartitions_hyper2"
output: html_document
date: "2022-07-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("partitions")
```

<p style="text-align: right;">
![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
</p>

Here we will sum likelihoods across a sample space and verify that the total is constant.

```{r}
M <- apply(setparts(c(2,2,1)),2,function(x){letters[x]})
M
```

```{r}
f <- function(strength,help,vec){loglik(strength,cheering3(v=vec[1:4],e=c(a=1,b=2,c=3),help=help,vec[5]))}
help <- c(1.11,2.222,1.7345)
strength <- 1
```

### 2022-07-29 inst/setpartitions_hyper2.Rmd
slight reworking

@@
 
-Here we will sum likelihoods across a sample space and verify that the total is constant.
+Here we will sum likelihoods across a sample space and verify that the
+total is constant.
 
@@
 ```{r}
-f <- function(strength,help,vec){loglik(strength,cheering3(v=vec[1:4],e=c(a=1,b=2,c=3),help=help,vec[5]))}
-help <- c(1.11,2.222,1.7345)
-strength <- 1
+f <- function(strength,help,vec){
+loglik(strength,cheering3(v=vec[1:4],e=c(a=1,b=2,c=3),help=help,vec[5]),log=FALSE)
+}
+
+tfun <- function(help,strength){
+   out <- 0
+   for(i in seq_len(ncol(M))){
+      out <- out + f(strength,help,vec=M[i,])
+   } 
+   return(out)
+}
+
+print(tfun(help=c(1.11,2.222,1.7345), strength= c(0.1,0.3,0.6)))
+print(tfun(help=c(1.,2.2,1.2), strength= c(0.3,0.2,0.5)))
 ```

### 2022-07-30 inst/setpartitions_hyper2.Rmd
multiset() not setparts()

--- 2022-07-29 inst/setpartitions_hyper2.Rmd
@@
 
-Here we will sum likelihoods across a sample space and verify that the total is constant.
+Here we will sum likelihoods across a sample space and verify that the
+total is constant.
 
 ```{r}
-M <- apply(setparts(c(2,2,1)),2,function(x){letters[x]})
-M
+M <- apply(multiset(c(1,1,2,2,3)),2,function(x){letters[x]})
+noquote(M)
 ```
@@
 ```{r}
-f <- function(strength,help,vec){loglik(strength,cheering3(v=vec[1:4],e=c(a=1,b=2,c=3),help=help,vec[5]))}
-help <- c(1.11,2.222,1.7345)
-strength <- 1
+tfun <- function(help,strength){
+   out <- 0
+     for(i in seq_len(ncol(M))){
+     vec <- M[,i]
+     out <- out +
+       loglik(strength,cheering3(v=vec,e=c(a=1,b=2,c=3),help=help),log=FALSE)
+   }
+   return(out)
+}
+
+h <- c(1.11,2.222,1.7345)
+print(tfun(help=(1:3)/10, strength= c(a=0.1,b=0.3,c=0.6)))
+print(tfun(help=(7:5)/22, strength= c(a=0.3,b=0.2,c=0.5)))
+
 ```

### 2022-07-30 inst/multiset_hyper2.Rmd
move file

---
title: "multiset_hyper2"
output: html_document
date: "2022-07-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("partitions")
```

<p style="text-align: right;">
![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
</p>

Here we will sum probabilities across a sample space and verify that
the total is constant.  In theory the sum of the probabilities of all
members of the sample space is 1, but in practice the likelihood
calculation includes an indeterminate constant which means that all we
can say is that the total probability is not a function of the
parameters of the distribution.

```{r defineM}
M <- apply(multiset(c(1,1,2,2,3)),2,function(x){letters[x]})
noquote(M)
```

```{r definetfun}
tfun <- function(help,strength){
  out <- 0
  for(i in seq_len(ncol(M))){
     vec <- M[,i]
     out <- out +
       loglik(strength,cheering3(v=vec,e=c(a=1,b=2,c=3),help=help),log=FALSE)
  }
  return(out)
}
```

```{r showtisconstant,cache=TRUE}
print(tfun(help=(1:3)/10, strength= c(a=0.1,b=0.3,c=0.6)))
print(tfun(help=(7:5)/22, strength= c(a=0.3,b=0.2,c=0.5)))
```

Above we see that the sum of the likelihoods is constant.  The value
turns out to be $\frac{1}{1!2!2!}=0.25$.  This is because the
likelihood function (being Plackett-Luce) considers the observation to
be one of the $5!=120$ ways of ordering five distinct competitors.
Matrix `M` has only $\frac{5!}{2!2!1!}=30$ columns, each one of which
corresponds to four different observations, all of which have the same
probability.  Much of the confusion arises from the practice of
identifying a competitor with his strength; we write "$p_a$'' [or
sometimes just "$a$''] to signify both a particular competitor, and
his Bradley-Terry strength.

To explain the factor of $1/4$ discussed above we need some new
notation.  We have five competitors who I will call $\mathbf{do}$,
$\mathbf{re}$, $\mathbf{me}$, $\mathbf{fa}$, and $\mathbf{sol}$.  Now
$p_X$ means "the Bradley Terry strength of $X$''.  The setup here is
that $\mathbf{do}$ and $\mathbf{re}$ are twins, and that are twins
(and $\mathbf{sol}$ is a singleton).  We have that the Bradley-Terry
stregth of $\mathbf{do}$ and $\mathbf{re}$ is `a`, the strength of
$\mathbf{me}$ and $\mathbf{fa}$ is `b`, and the strength of
$\mathbf{sol}$ is `c`.

Consider, for example, the first column of `M` which is `a c b b a`.
In theory, this means the competitor who came first had strength `a`,
the competitor who came second had strength `c`.  So the order
statistic is one of

$$\mathbf{do}\succ\mathbf{sol}\succ\mathbf{me}\succ\mathbf{fa}\succ\mathbf{re}$$
$$\mathbf{do}\succ\mathbf{sol}\succ\mathbf{fa}\succ\mathbf{me}\succ\mathbf{re}$$
$$\mathbf{re}\succ\mathbf{sol}\succ\mathbf{me}\succ\mathbf{fa}\succ\mathbf{do}$$
$$\mathbf{re}\succ\mathbf{sol}\succ\mathbf{fa}\succ\mathbf{me}\succ\mathbf{do}$$

Above we see that $\mathbf{do}$ and $\mathbf{re}$ come first and last
(or last and first), and $\mathbf{me}$ and $\mathbf{fa}$ come third
and fourth (or fourth and third), while $\mathbf{sol}$ comes second.
All four possibilities have the same probability and all four are
admissible order statistics for the five solfeggio syllables.

But the `hyper2` package [specifically the `loglik()` and
`cheering3()` functions] consider `a c b b a` to mean something
different.  It considers it to be an observation of

$$\mathbf{do}\succ\mathbf{sol}\succ\mathbf{me}\succ\mathbf{fa}\succ\mathbf{re}$$

followed by an identification of $\mathbf{do}$ with (BT strength) `a`,
$\mathbf{sol}$ with strength `c`, and so on.  Any of the four
possibilities above would be equally admissible, and they all have the
same probability.  So to get "real'' probabilities [real in the sense
that summing the probabilities over the columns in matrix `M`], we
have to multiply by four to include all the admissible order
statistics.

The difference between these two interpretations does not matter from
a likelihood perspective as likelihood is defined only up to a
multiplicative constant.

### 2022-08-01 inst/multiset_hyper2.Rmd
more exacting test

@@
 M <- apply(multiset(c(1,1,2,2,3)),2,function(x){letters[x]})
-noquote(M)
+colnames(M) <- rep("",ncol(M))   # tidier and more compact appearance
+M <- noquote(M)
+M
 ```
 
+Above, we use `multiset()` to enumerate the sample space of five
+competitors comprising two twins denoted `a`, two twins denoted `b`,
+and a singleton `c`.  This document discusses the issue of whether the
+twins are identical twins and hence indistinguishable.  Matrix `M` has
+${5\choose2\,2\,1}=30$ columns, one per possible result.  We will
+calculate the likelihood of each column and calculate their sum:
+
 ```{r definetfun}
-tfun <- function(help,strength){
+tfun <- function(help,strength,M){
   out <- 0
@@
 ```{r showtisconstant,cache=TRUE}
-print(tfun(help=(1:3)/10, strength= c(a=0.1,b=0.3,c=0.6)))
-print(tfun(help=(7:5)/22, strength= c(a=0.3,b=0.2,c=0.5)))
+print(tfun(help=(1:3)/10, strength= c(a=0.1,b=0.3,c=0.6),M))
+print(tfun(help=(7:5)/22, strength= c(a=0.3,b=0.2,c=0.5),M))
 ```
@@
 
-Above we see that $\mathbf{do}$ and $\mathbf{re}$ come first and last
+heAbove we see that $\mathbf{do}$ and $\mathbf{re}$ come first and last
 (or last and first), and $\mathbf{me}$ and $\mathbf{fa}$ come third
@@
 multiplicative constant.
+
+## A more exacting numerical test
+
+```{r,exactingtest,cache=TRUE}
+rs <- fillup(rp(100,dirichlet(c(a=1,b=1,c=1)))) # random strengths
+colnames(rs) <- letters[1:3]
+head(rs)
+f <- function(s){
+  jj <- s
+  names(jj) <- letters[1:3]
+  tfun(help=1:3,strength=jj,M=M)
+}
+x <- apply(rs,1,f)
+```
+
+```{r}
+hist(x-mean(x))
+```
+
+Above we see very tight clustering about the mean value of 0.25

### 2022-08-02 inst/jss_hyper3.Rnw
typoe

\documentclass[article]{jss}

\author{Robin K. S. Hankin\\Auckland University of Technology}
\title{Generalized Plackett-Luce Likelihoods}
\Plainauthor{Robin K. S. Hankin} %% comma-separated
\Plaintitle{Generalized Plackett-Luce Likelihoods}

%% an abstract and keywords
\Abstract{
  The \code{hyper2} package is improved here to include \code{hyper3}
  objects.
}
\Keywords{Plackett-Luce, Bradley-Terry, Mann-Whitney}
\Plainkeywords{Plackett-Luce, Bradley-Terry, Mann-Whitney}

%% publication information
%% NOTE: Typically, this can be left commented and will be filled out by the technical editor
%% \Volume{50}
%% \Issue{9}
%% \Month{June}
%% \Year{2012}
%% \Submitdate{2012-06-04}
%% \Acceptdate{2012-06-04}

%% The address of (at least) one author should be given
%% in the following format:
\Address{
  Robin K. S. Hankin\\
  School of Engineering and Mathematical Sciences\\
  Auckland University of Technology\\
  Wellesley Street\\
  Auckland, New Zealand\\
  E-mail: \email{hankin.robin@gmail.com}\\
  URL: \url{https://academics.aut.ac.nz/robin.hankin}
}

%% for those who use Sweave please include the following line (with % symbols):
%% need no \usepackage{Sweave.sty}

%% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do
eiusmod tempor incididunt ut labore et dolore magna aliqua.  Ut enim
ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut
aliquip ex ea commodo consequat.  Duis aute irure dolor in
reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla
pariatur.  Excepteur sint occaecat cupidatat non proident, sunt in
culpa qui officia deserunt mollit anim id est laborum

The \code{hyper2} package~\cite{hankin2017} furnishes computational
support for generalized Plackett-Luce~\citep{plackett1975} likelihood
functions.  Our preferred interpretation is a race (as in track and
field atheletics): given six competitors $1-6$, we ascribe to them
nonnegative strengths $p_1\ldots p_6$; the probability that $i$ beats
$j$ is $p_i/(p_i+p_j)$.  It is conventional to normalise so that the
total strength is unity, and to identify a competitor with his
strength.  Given an order statistic, say $p_1\succ p_2\succ p_3\succ
p_4$, the Plackett-Luce likelihood function would be

\begin{equation}
  \frac{p_1}{p_1+p_2+p_3+p_4}\cdot
  \frac{p_2}{    p_2+p_3+p_4}\cdot
  \frac{p_3}{        p_3+p_4}\cdot
  \frac{p_4}{            p_4}.
\end{equation}

The \code{hyper2} package generalized this likelihood function to
functions of ${\mathbf p}=(p_1,\ldots,p_n)$ with

\begin{equation}\label{hyper2likelihood}
\mathcal{L}\left(\mathbf{p}\right)=
\prod_{s\in \mathcal{O}}\left({\sum_{i\in s}}p_i\right)^{n_s}
\end{equation}

\noindent
where~$\mathcal{O}$ is a set of observations and~$s$ a subset
of~$\left[n\right]=\left\{1,2,\ldots,n\right\}$; numbers~$n_s$ are
integers which may be positive or negative.  The approach adopted by
the~\pkg{hyperdirichlet} package~ is to sptore each of the~$2^n$
possible subsets of~$\left[n\right]$ together with an exponent:

\end{document}

### 2022-08-03 inst/jss_hyper3.Rnw
switch to Rnw format for possible inclusion as a vignette
more work

@@
+% -*- mode: noweb; noweb-default-code-mode: R-mode; -*-
 \documentclass[article]{jss}
@@
 
-The \code{hyper2} package~\cite{hankin2017} furnishes computational
-support for generalized Plackett-Luce~\citep{plackett1975} likelihood
-functions.  Our preferred interpretation is a race (as in track and
-field atheletics): given six competitors $1-6$, we ascribe to them
-nonnegative strengths $p_1\ldots p_6$; the probability that $i$ beats
-$j$ is $p_i/(p_i+p_j)$.  It is conventional to normalise so that the
-total strength is unity, and to identify a competitor with his
-strength.  Given an order statistic, say $p_1\succ p_2\succ p_3\succ
-p_4$, the Plackett-Luce likelihood function would be
+The \code{hyper2} package~\citep{hankin2009,hankin2017} furnishes
+computational support for generalized
+Plackett-Luce~\citep{plackett1975} likelihood functions.  Our
+preferred interpretation is a race (as in track and field atheletics):
+given six competitors $1-6$, we ascribe to them nonnegative strengths
+$p_1\ldots p_6$; the probability that $i$ beats $j$ is
+$p_i/(p_i+p_j)$.  It is conventional to normalise so that the total
+strength is unity, and to identify a competitor with his strength.
+Given an order statistic, say $p_1\succ p_2\succ p_3\succ p_4$, the
+Plackett-Luce likelihood function would be
 
@@
 
-The \code{hyper2} package generalized this likelihood function to
+The \code{hyper2} package generalizes this likelihood function to
 functions of ${\mathbf p}=(p_1,\ldots,p_n)$ with
@@
 
-\noindent
-where~$\mathcal{O}$ is a set of observations and~$s$ a subset
-of~$\left[n\right]=\left\{1,2,\ldots,n\right\}$; numbers~$n_s$ are
-integers which may be positive or negative.  The approach adopted by
-the~\pkg{hyperdirichlet} package~ is to sptore each of the~$2^n$
-possible subsets of~$\left[n\right]$ together with an exponent:
+\noindent where~$\mathcal{O}$ is a set of observations and~$s$ a
+subset of~$\left[n\right]=\left\{1,2,\ldots,n\right\}$; numbers~$n_s$
+are integers which may be positive or negative.  It becomes possible
+to analyse a wider range of observations such as incomplete order
+statistics or teams; and the introduction of reified entities allows
+one to consider draws~\citep{hankin2009}, noncompetitive tactics such
+as collusion~\citep{hankin2020}, and the phenomenon of team cohesion
+wherein the team becomes stronger than the sum of its parts.
+
+However, consider a race in which there are six runners 1-6 but we
+happen to know that three of the runners (1,2,3) are clones of
+strength $p_a$, say, two of the runners (4,5) have strength $p_b$, and
+the final runner (6) is of strength $p_c$.  The runners race and the
+finishing order is:
+
+$$a\succ c\succ b\succ a\succ a \succ b$$
+
+Thus the winner was $a$, second place for $b$, third for $a$, and so
+on.  Alternatively we might say that $a$ came first, fourth, and
+fifth; $b$ came third and sixth, and $c$ came second.  The
+Plackett-Luce likelihood function for $p_a,p_b,p_c$ would be
+
+\begin{equation}
+\frac{p_a}{3p_a+2p_b+p_c}\cdot
+\frac{p_c}{2p_a+2p_b+p_c}\cdot
+\frac{p_b}{2p_a+2p_b    }\cdot
+\frac{p_a}{2p_a+ p_b    }\cdot
+\frac{p_a}{ p_a+ p_b    }\cdot
+\frac{p_b}{      p_b    },\qquad p_a+p_b+p_c=1
+\end{equation}
+
+Existing \code{hyper2} formalism as described by \citet{hankin2017}
+cannot represent this likelihood function, because
+Equation~\ref{hyper2likelihood} uses sets as the indexing elements,
+and in this case we need multisets\footnote{Note that the version of
+\code{hyper2} presented by \citet{hankin2017} used integer-valued sets
+together with a print method that used a complicated mapping system
+from integers to competitor names.  Current methodology [following
+commit \code{51a8b46}] is to use sets of character strings which
+represent the competitors directly; this allows for easier combination
+of observations including different competitors.}.
+
+<<>>=
+2+2
+@
+
+\bibliography{chess}
 

### 2022-08-04 inst/jss_hyper3.Rnw
further work on JSS manuscript

@@
 
-Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do
-eiusmod tempor incididunt ut labore et dolore magna aliqua.  Ut enim
-ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut
-aliquip ex ea commodo consequat.  Duis aute irure dolor in
-reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla
-pariatur.  Excepteur sint occaecat cupidatat non proident, sunt in
-culpa qui officia deserunt mollit anim id est laborum
+<<setup, cache=FALSE,echo=FALSE>>=
+library("knitr")
+opts_chunk$set(cache=TRUE, autodep=TRUE)
+@
+
+\section{introduction}
 
@@
 given six competitors $1-6$, we ascribe to them nonnegative strengths
-$p_1\ldots p_6$; the probability that $i$ beats $j$ is
+$p_1,\ldots, p_6$; the probability that $i$ beats $j$ is
 $p_i/(p_i+p_j)$.  It is conventional to normalise so that the total
@@
 
-\begin{equation}
+\begin{equation}\label{PL_like}
   \frac{p_1}{p_1+p_2+p_3+p_4}\cdot
@@
 happen to know that three of the runners (1,2,3) are clones of
-strength $p_a$, say, two of the runners (4,5) have strength $p_b$, and
-the final runner (6) is of strength $p_c$.  The runners race and the
-finishing order is:
+strength $p_a$, two of the runners (4,5) have strength $p_b$, and the
+final runner (6) is of strength $p_c$.  We normalise so
+$p_a+p_b+p_c=1$.  The runners race and the finishing order is:
 
@@
 
-\begin{equation}
+\begin{equation}\label{plackettluce}
 \frac{p_a}{3p_a+2p_b+p_c}\cdot
@@
 represent the competitors directly; this allows for easier combination
-of observations including different competitors.}.
+of observations including different competitors.}.  The declarations
+
+\begin{verbatim}
+typedef map<string, long double> weightedplayervector;
+typedef map<weightedplayervector, long double> hyper3;
+\end{verbatim}
+
+show how the \proglang{map} class of the Standard Template Library are
+used: \proglang{weightedplayervector} objects map strings to long
+doubles (specifically, mapping player names to their multiplicities),
+and objects of class \proglang{hyper3} are maps from a
+\proglang{weightedplayervector} object to long doubles.  One advantage
+of this is efficiency: Search, removal, and insertion operations have
+logarithmic complexity.  As an example, the following \proglang{C++}
+pseudo code would create a log-likelihood function for the first term
+in Equation~\ref{plackettluce}:
+
+\begin{verbatim}
+weightedplayervector n,d;
+n["a"] = 1; // n for numerator
+d["a"] = 3; // d for denominator
+d["b"] = 2;
+d["c"] = 1;
+
+hyper3 L;
+L[n] = 1;
+L[d] = -1;
+\end{verbatim}
+
+Above, object \code{L} is an object of class \code{hyper3}; it may be
+evaluated at points in probability space [that is, a vector
+\code{[a,b,c]} of nonnegative values with unit sum] using standard
+\proglang{R} idiom wrapping \proglang{C++} back end.
+
+The package includes an \proglang{S4} class \code{hyper3} for this
+type of object.  Package idiom for creating such objects uses named
+vectors:
 
 <<>>=
-2+2
+library("hyper2")
+H <- hyper3()
+H[c(a=1)] <- 1
+H[c(a=3,b=2,c=1)] <- -1
+H
 @
+
+Above, we see object \code{H} is a log-likelihood function of the
+players' strengths, which may be evaluated at specified points in
+probability space.  If, for example, we wish to assess $H_1\colon
+p_a=0.9,p_b=0.05,p_c=0.05$ and $H_2\colon p_a=0.01,p_b=0.01,p_c=0.98$,
+we can evaluate these hypotheses using generic function
+\code{loglik()}:
+
+<<>>=
+loglik(c(a=0.01,b=0.01,c=0.98),H)
+loglik(c(a=0.9,b=0.05,c=0.05),H)
+@
+
+Thus we prefer $H_1$ over $H_2$ with about 3.5 units of support,
+satisfying the standard two units of support
+criterion~\citep{edwards1972}, and we conclude that our observation
+[in this case, that one of the three clones of player $a$ beat the $b$
+twins and the singleton $c$] furnishes strong support against $H_2$ in
+favour of $H_1$.
+
+The package includes many helper functions to work with order
+statistics of this type.  Function \code{ordervec2supp3()}, for
+example, can be used to generate a loglikelihood function for
+Equation~\ref{PL_like}:
+
+<<>>=
+(H <- ordervec2supp3(c("a","c","b","a","a","b")))
+@
+
+We may find a maximum likelihood estimate for the players' strengths:
+
+<<maxexamp,cache=TRUE>>=
+(mH <- maxp(H))
+@
+
+We may use a selection of statistical tests on the order statistic,
+modified from~\citep{hankin2017} to work with \code{hyper3} objects.
+For example, we may assess the hypothesis that all three players are
+of equal strength [viz $H_0\colon p_a=p_b=p_c=\frac{1}{3}$]:
+
+<<testequality,cache=TRUE>>=
+equalp.test(H)
+@
+
+showing, perhaps unsurprisingly, that this small dataset is consistent
+with $H_0$.
+
+\section{Some simple cases}
+
+Here I consider some order statistics with nontrivial maximum
+likelihood Bradley-Terry strength.  The simplest nontrivial case would
+be three competitors with strengths $a,a,b$ and finishing order
+$a\succ b\succ a$.  The likelihood function would be
+
+\begin{equation}\label{aba}
+\frac{a}{2a+b}\cdot\frac{b}{a+b}
+\end{equation}
+
+In this case we know that $a+b=1$ so this is equal to
+$\mathcal{L}=\mathcal{L}(a)=\frac{a(1-a)}{1+a}$.
+
+\begin{equation}\label{aba_mle}
+\frac{d\mathcal{L}}{da}=\frac{(1+a)(1-2a)-a(1-a)}{(1+a)^2}=
+\frac{1-2a+a-2a^2 -a + a^2}{(1+a)^2}=
+\frac{1-2a-a^2}{(1+a)^2}
+\end{equation}
+
+This will be zero at $\sqrt{2}\pm 1$; we also note that
+$d^2\mathcal{L}/da^2=-4(1+a)^{-3}$, manifestly strictly negative for
+$0\leq a\leq 1$: the root is a maximum.
+
+<<maxpaba,cache=TRUE>>=
+maxp(ordervec2supp3(c("a","b","a")))
+@
+
+above, we see close agreement with the theoretical value of
+$(\sqrt{2}-1,1-\sqrt{2})\simeq (0.414,0.586)$.  Observe that the
+maximum likelihood estimate for $a$ is strictly less than 0.5, even
+though the finishing order is symmetric.  Using
+$\mathcal{L}(a)=\frac{a(1-a)}{1+a}$, we can show that
+$\log\mathcal{L}(\hat{a})=\log\left(3-2\sqrt{2}\right)\simeq -1.76$,
+where $\hat{a}=\sqrt{2}-1$ is the maximum likelihood estimate for $a$.
+Defining $\mathcal{S}=\log\mathcal{L}$ as the support [log-likelihood]
+we have
+
+\begin{equation}
+\mathcal{S}=\mathcal{S}(a)=\log\left(\frac{a(1-a)}{1+a}\right)-\log\left(3-2\sqrt{2}\right)
+\end{equation}
+
+as a standard support function which has a maximum value of zero when
+evaluated at $\hat{a}=\sqrt{2}-1$.  For example, we can test the null
+that $a=b=\frac{1}{2}$, the statement that the competitors have equal
+Bradley-Terry strengths:
+
+<<>>=
+a <- 1/2 # null
+(S_delta <- log( a*(1-a)/(1+a))- log(3-2*sqrt(2)))
+@
+
+Thus the additional support gained in moving from $a=\frac{1}{2}$ to
+the evaluate of $a=\sqrt{2}-1$ is 0.029, rather small [as might be
+expected given that we have only one rather uninformative observation,
+and also given that the maximum likelihood estimate ($\simeq 0.41$) is
+quite close to the null of $0.5$].  Nevertheless we can apply Wilks's
+theorem for a $p$-value:
+
+<<>>=
+pchisq(-2*S_delta,df=1,lower.tail=FALSE)
+@
+
+The $p$-value is about 0.81, exceeding 0.05; thus we have no strong
+evidence to reject the null of $a=\frac{1}{2}$.  The observation is
+informative, in the sense that we can find a credible interval for
+$a$.  With an $n$-unit of support criterion the analytical solution to
+$\mathcal{S}(p)=-n$ is given by defining $X=\log(3-2\sqrt{2})-n$ and
+solving $p(1-p)/(1+p)=X$, or
+$p_\pm=\left(1-X\pm\sqrt{1+4X+X^2}\right)/2$, the two roots being the
+lower and upper limits of the credible interval;
+Figure~\ref{ABA_likelihood}.
+
+\begin{figure}[htbp]
+  \begin{center}
+<<figaba>>=
+a <- seq(from=0,by=0.005,to=1)
+S <- function(a){log(a*(1-a)/((1+a)*(3-2*sqrt(2))))}
+plot(a,S(a),type='b')
+abline(h=c(0,-2))
+abline(v=c(0.02438102,0.9524271),col='red')
+abline(v=sqrt(2)-1)
+@
+\caption{A\label{ABA_likelihood} support function for $p_a$ with observation $a\succ b\succ a$}
+\end{center}
+\end{figure}
+
+\subsection{All results for AAB}
+
+\begin{itemize}
+\item $A\succ A\succ B\longrightarrow\frac{a^2}{1+a}$
+\item $A\succ B\succ A\longrightarrow\frac{1-2a-a^2}{(1+a)^2}$
+\item $B\succ A\succ A\longrightarrow\frac{1-a}{1+a}$
+\end{itemize}
+
+Likelihood functions for these order statistics are given in
+Figure~\ref{ABAetc_likelihood}.  It can be shown that the Fisher
+information for such observations is
+
+\begin{equation}
+\mathcal{I}(a)=\frac{1+16a+18a^2-4a^3-15a^4-7a^5-a^6}{a(1-a)(1+a)^3}
+\end{equation}
+
+which has a minimum at about $0.575$.
+
+\begin{figure}[htbp]
+  \begin{center}
+<<figabbabb,echo=FALSE>>=
+f_aab <- function(a){a^2/(1+a)}
+f_aba <- function(a){a*(1-a)/(1+a)^2}
+f_baa <- function(a){(1-a)/(1+a)}
+
+p <- function(f,...){points(a,f(a)/max(f(a)),...)}
+plot(0:1,0:1,xlab="p(a)",ylab="Likelihood",type="n")
+p(f_aab,type="l",col="black")
+p(f_aba,type="l",col="red")
+p(f_baa,type="l",col="blue")
+legend(x=0.85,y=0.7,col=c("black","red","blue"),lty=1,pch=NA,legend=c("AAB","ABA","BAA"))
+abline(h=exp(-2),lty=2)
+@
+\caption{Likelihood \label{ABAetc_likelihood} functions for
+observations $a\succ a\succ b$, $a\succ b\succ a$, $b\succ a\succ a$}
+\end{center}
+\end{figure}
+
+\subsection{Nonfinishers}
+
+If we allow non-finishers [that is, a subset of competitors who are
+classified as being beaten by any finishing competitors], there is
+another nontrivial order statistic, viz $a\succ b\succ\left\lbrace
+a,b\right\rbrace$ [thus one of the two $a$'s won, one of the $b$'s
+came second, and one of each of $a$ and $b$ failed to finish].
+
+\begin{equation}
+\mathcal{L}(a)=
+\frac{a}{2a+2b}\cdot
+\frac{b}{ a+2b}\propto\frac{a(1-a)}{2-a}
+\end{equation}
+
+(see how the likelihood function is actually simpler than for the
+complete order statistic).  The evaluate would be $2-\sqrt{2}\simeq
+0.586$:
+
+<<cache=TRUE>>=
+maxp(ordervec2supp3(c("a","b"),nonfinishers=c("a","b")))
+@
+
+\section{Mann-Whitney test}
+
+The ideas presented above can easily be extended to arbitrarily large
+numbers of competitors.  Given an order statistic of the type
+considered above, the Mann-Whitney-Wilcoxon test assesses a null of
+identity of underlying distributions.  Consider the fly spray dataset
+presented by \cite{wilcoxon1945}:
+
+<<defineAB>>=
+A <- c(68,68,59,72,64,67,70,74)
+B <- c(60,67,61,62,67,63,56,58)
+@
+
+Above we see the percentage mortality of two preparations, A and B,
+each applied eight times.  We may easily convert this dataset to a
+rank order statistic:
+
+<<makeABorderstatistic>>=
+names(A) <- rep("a",length(A))
+names(B) <- rep("b",length(B))
+sort(c(A,B))
+names(sort(c(A,B)))
+@
+
+And thence to a likelihood function:
+
+<<definehyper3AB,cache=TRUE>>=
+(Hab <- ordervec2supp3(names(sort(c(A,B)))))
+@
+
+Package idiom for testing the null of equality is straightforward:
+
+<<testABforequality,cache=TRUE>>=
+equalp.test(Hab)
+@
+
+Compare \citeauthor{wilcoxon1945}'s $p$-value of $1.77\%$.  However,
+observe that the \code{hyper3} likelihood approach gives more
+information than Wilcoxon's analysis: Firstly, we see from the
+evaluate that the probability of a single application of preparation A
+killing more flies than B is about 15\%.  Further, we may plot a
+support curve for this dataset as above:
+
+\begin{figure}[htbp]
+  \begin{center}
+<<plotwilcoxlike,cache=TRUE>>=
+a <- seq(from=0.02,to=0.5,len=50)
+L <- sapply(a,function(p){loglik(p,Hab)})
+plot(a,L-max(L),type='b')
+abline(h=c(0,-2))
+abline(v=0.1558)
+@
+\caption{A\label{wilcox_likelihood} support for the Bradley-Terry
+strength $p_a$ of preparation A~\citep{wilcoxon1945}.  The evaluate of
+0.1558 is shown; and the two-units-of support credible interval, which
+excludes $p_a=0.5$}
+\end{center}
+\end{figure}
+
+\section{Multiple competitors: javelin}
+
+The ideas presented above are easily extended to multiple types of
+competitors.  Consider the following table, drawn from the men's
+javelin, 2020 Olympics:
+
+<<javelintable>>=
+javelin_table
+@
+
+This Chopra threw 87.03m on his first throw, 87.58m on his second, and
+so on.  No-throws, ignored here, are indicated with a \code{X}.  We
+may convert this to a named vector with elements being the throw
+distances, and names being the competitors, using
+`attemptstable2supp3()`:
+
+<<label=converttosupp3,cache=TRUE>>=
+javelin_vector <- attemptstable2supp3(javelin_table,decreasing=TRUE,give.supp=FALSE)
+javelin_vector
+@
+
+Above we see that Chopra threw the longest and second-longest throws
+of 87.58m and 87.03 respectively; Vadlejch threw the third-longest
+throw of 86.67m, and so on (\code{NA} entries correspond to
+no-throws.)  Now convert the attempts table to a \code{hyper3} object,
+again using function \code{attemptstable2supp3()} but this time pass
+\code{give.supp=TRUE}:
+
+<<dothething2,cache=TRUE>>=
+javelin <- ordervec2supp3(v=names(javelin_vector)[!is.na(javelin_vector)])
+@
+
+Above, object \code{javelin} is a \code{hyper3} likelihood function,
+so one has access to the standard likelihood-based methods, such as
+finding and displaying the maximum likelihood estimate:
+
+<<setdigits,echo=FALSE>>=
+options(digits=3)
+@
+
+\begin{figure}[htbp]
+  \begin{center}
+<<testthejav,cache=TRUE>>=
+(mj <- maxp(javelin))
+dotchart(mj,pch=16)
+@
+\caption{Maximum likelihood estimate for javelin throwers' Bradley-Terry strengths}
+\end{center}
+\end{figure}
+
+Above, we see that Vadlejch has the highest estimated Bradley-Terry
+strength, but further analysis with \code{equalp.test()} reveals that
+there is no strong evidence in the datast to reject the hypothesis of
+equal competitive strength ($p=0.26$), or that Vadlejch has a strength
+higher than the null value of $\frac{1}{8}$ ($p=0.1$).
+
+\section{Formula 1 motor racing: the constructors' championship}
+
+In Formula 1 motor racing, the constructors' championship takes place
+between {\em manufacturers} of racing cars (compare the drivers'
+championship).  Here, we consider the constructor to be the object of
+inference.  Each constructor typically fields two cars, each of which
+separately accumulates ranking-based points at each venue.  Here we
+use generalized Plackett-Luce to assess the contructors' performance.
+The following table shows rankings for the first 9 venues of the 2021
+season:
+
+<<showconstructortable>>=
+constructor_2021_table[,1:9]
+@
+
+Above, we see that Mercedes (``\code{Merc}'') came first and third at
+Bahrain (\code{BHR}, and at Emilia Romagna (\code{EMI} came second and
+retired (full details of the notation and conventions is given in the
+package at \code{constructor.Rd}).  The identity of the driver is not
+admissible information and indeed may change during a season.
+Alternatively, we may regard the driver and the constructor as a joint
+entity, with the constructor's ability to attract and retain a skilled
+driver being part of the object of inference.
+
+The associated generalized Plackett-Luce \code{hyper3} object is as follows:
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlstd{constructor_2021}
+\end{alltt}
+\begin{verbatim}
+## log( (AMM=1)^38 * (AMM=1, AR=1)^-1 * (AMM=1, AR=1, ARRF=1, ATH=1, HF=2,
+## MM=1, Merc=1, WM=2)^-1 * (AMM=1, AR=1, ARRF=1, ATH=1, HF=2, MM=1,
+## WM=2)^-2 * (AMM=1, AR=1, ARRF=1, ATH=1, HF=2, MM=2, Merc=1, WM=2)^-1 *
+## (AMM=1, AR=1, ARRF=1, ATH=1, HF=2, RBRH=1, WM=1)^-1 * (AMM=1, AR=1,
+## ARRF=1, ATH=1, HF=2, RBRH=1, WM=2)^-1 * (AMM=1, AR=1, ARRF=1, ATH=1,
+## HF=2, WM=2)^-1 * (AMM=1, AR=1, ARRF=1, ATH=2, Ferrari=1, HF=2, WM=1)^-1
+## * (AMM=1, AR=1, ARRF=1, HF=2, Merc=2, RBRH=1, WM=2)^-1 * (AMM=1, AR=1,
+## ARRF=1, HF=2, WM=1)^-1 * (AMM=1, AR=1, ARRF=2, ATH=1, Ferrari=1, HF=2,
+## MM=1, Merc=1, RBRH=1)^-1 * (AMM=1, AR=1, ARRF=2, ATH=1, Ferrari=1,
+## HF=2, MM=1, Merc=1, RBRH=1, WM=1)^-1 * (AMM=1, AR=1, ARRF=2, ATH=1,
+## Ferrari=1, HF=2, MM=1, Merc=2, RBRH=1, WM=2)^-1 * (AMM=1, AR=1, ARRF=2,
+...
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
 

### 2022-08-04 inst/jss_hyper3.Rnw
cite Mann and Whitney 1947

@@
 numbers of competitors.  Given an order statistic of the type
-considered above, the Mann-Whitney-Wilcoxon test assesses a null of
-identity of underlying distributions.  Consider the fly spray dataset
-presented by \cite{wilcoxon1945}:
+considered above, the Mann-Whitney-Wilcoxon
+test~\citep{mann1947,wilcoxon1945} assesses a null of identity of
+underlying distributions.  Consider the fly spray dataset presented by
+\citeauthor{wilcoxon1945}:
 

### 2022-08-04 inst/mann_whitney_wilcoxon.Rmd
work on Mann-Whitney-Wilcox test (as a source for jss_hyper3.Rnw)

---
title: "Mann-Whitney-Wilcoxon test with `hyper3` objects"
author: "Robin KS Hankin"
date: "10/04/2022"
output: html_document
---

```{r setup, include=FALSE}
set.seed(1)
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
```

<p style="text-align: right;">
![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
</p>

We do some work with Mann-Whitney test, using the chorioamnion dataset at `wilcox.test.Rd`:

```{r,label=defineAB}
x <- c(0.80, 0.83, 1.89, 1.04, 1.45, 1.38, 1.91, 1.64, 0.73, 1.46)
y <- c(1.15, 0.88, 0.90, 0.74, 1.21)
```

First do the Mann-Whitney-Wilcox test:

```{r,domwwtest}
wilcox.test(x,y)
```

Now use `hyper3`:

```{r,hyper3AB,cache=TRUE}
names(x) <- rep("x",length(x))
names(y) <- rep("y",length(y))
equalp.test(ordervec2supp3(names(sort(c(x,y)))))
```

Now some permutation tests under the null:

```{r,repeat_mww_H0,cache=TRUE}
n <- 10
mww_pval <- rep(NA,n)
for(i in seq_len(n)){
      jj <- sample(c(x,y))
      xstar <- jj[seq(from=1,length=length(x))]
      ystar <- jj[seq(from=length(x)+1,to=length(x)+length(y))]
      mww_pval[i] <- suppressWarnings(wilcox.test(xstar,ystar)$p.value)
}
```

```{r,histmww}
mww_pval
hist(mww_pval)
```

```{r,repeat_hyper3_H0,cache=TRUE}
n <- 10
hyper3_pval <- rep(NA,n)
for(i in seq_len(n)){
      jj <- sample(c(x,y))
      xstar <- jj[seq(from=1,length=length(x))]
      ystar <- jj[seq(from=length(x)+1,to=length(x)+length(y))]
      hyper3_pval[i] <- equalp.test(ordervec2supp3(args2ordervec(x=xstar,y=ystar)))$p.value
}
```

```{r,histhyper3}
hyper3_pval
hist(hyper3_pval)
```

### 2022-08-05 inst/mann_whitney_wilcoxon.Rmd
more work

@@
 ```{r,repeat_mww_H0,cache=TRUE}
-n <- 10
+n <- 100
 mww_pval <- rep(NA,n)
@@
 ```{r,repeat_hyper3_H0,cache=TRUE}
-n <- 10
+n <- 100
 hyper3_pval <- rep(NA,n)
@@
 ```{r,histhyper3}
-hyper3_pval
 hist(hyper3_pval)
+ks.test(hyper3_pval,"punif")
+
+plot(ecdf(hyper3_pval))
+par(pty="s")
+qqplot(mww_pval,hyper3_pval,asp=1)
+abline(0,1)
 ```
+
+```{r,dosomepvals,cache=TRUE}
+n <- 100
+a <- matrix(0,n,2)
+colnames(a) <- c("mww","hyper3")
+for(i in seq_len(n)){
+  d <- rrace3(pn=c(a=10,b=10),ps=c(a=0.2,b=0.8))
+  jj <- seq_along(d)
+  a[i,1] <- wilcox.test(jj[d=="a"],jj[d=="b"])$p.value
+  a[i,2] <- equalp.test(ordervec2supp3(d))$p.value
+}
+```
+
+```{r,showpvals}
+a
+plot(a)
+table(a[,1]<0.05,a[,2]<0.05)
+```

### 2022-08-05 inst/jss_hyper3.Rnw
typo

@@
 If we allow non-finishers [that is, a subset of competitors who are
-classified as being beaten by any finishing competitors], there is
-another nontrivial order statistic, viz $a\succ b\succ\left\lbrace
-a,b\right\rbrace$ [thus one of the two $a$'s won, one of the $b$'s
-came second, and one of each of $a$ and $b$ failed to finish].
+beaten by all the ranked competitors], there is another nontrivial
+order statistic, viz $a\succ b\succ\left\lbrace a,b\right\rbrace$
+[thus one of the two $a$'s won, one of the $b$'s came second, and one
+of each of $a$ and $b$ failed to finish].
 

### 2022-08-07 inst/mann_whitney_wilcoxon.Rmd
more work

@@
 
-Now some permutation tests under the null:
+#  Some permutation tests under the null:
 
-```{r,repeat_mww_H0,cache=TRUE}
+```{r,dosomepvalsNULL,cache=TRUE}
 n <- 100
-mww_pval <- rep(NA,n)
+a <- matrix(0,n,2)
+colnames(a) <- c("mww","hyper3")
 for(i in seq_len(n)){
-      jj <- sample(c(x,y))
-      xstar <- jj[seq(from=1,length=length(x))]
-      ystar <- jj[seq(from=length(x)+1,to=length(x)+length(y))]
-      mww_pval[i] <- suppressWarnings(wilcox.test(xstar,ystar)$p.value)
+  d <- rrace3(pn=c(a=10,b=10),ps=c(a=0.5,b=0.5))  #null 
+  jj <- seq_along(d)
+  a[i,1] <- wilcox.test(jj[d=="a"],jj[d=="b"])$p.value
+  a[i,2] <- equalp.test(ordervec2supp3(d))$p.value
 }
@@
 
-```{r,histmww}
-mww_pval
-hist(mww_pval)
+```{r,showpvalsNULL}
+head(a)
+par(pty="s")
+plot(log10(a),asp=1,xlim=c(-2,0),ylim=c(-2,0),main="p-values under the NULL (log scale)",
+xlab="log(p),Wilcox",ylab="log(p), hyper3")
+abline(0,1)
+plot(a,asp=1,xlim=c(0,1),ylim=c(0,1),main="p-values under the NULL (linear scale)",xlab="p,Wilcox",ylab="p,hyper3")
+abline(0,1)
+(size_wilcox <- sum(a[,1]<0.05)/nrow(a))
+(size_hyper3 <- sum(a[,2]<0.05)/nrow(a))
+
+table(type_I_wilcox=a[,1]<0.05,type_I_hyper3=a[,2]<0.05)
 ```
 
-```{r,repeat_hyper3_H0,cache=TRUE}
-n <- 100
-hyper3_pval <- rep(NA,n)
-for(i in seq_len(n)){
-      jj <- sample(c(x,y))
-      xstar <- jj[seq(from=1,length=length(x))]
-      ystar <- jj[seq(from=length(x)+1,to=length(x)+length(y))]
-      hyper3_pval[i] <- equalp.test(ordervec2supp3(args2ordervec(x=xstar,y=ystar)))$p.value
-}
-```
-
-```{r,histhyper3}
-hist(hyper3_pval)
-ks.test(hyper3_pval,"punif")
-
-plot(ecdf(hyper3_pval))
-par(pty="s")
-qqplot(mww_pval,hyper3_pval,asp=1)
-abline(0,1)
-```
+# Some permutation tests under the alternative
 
@@
 ```{r,showpvals}
-a
-plot(a)
-table(a[,1]<0.05,a[,2]<0.05)
+head(a)
+par(pty="s")
+plot(log10(a),asp=1,xlim=c(-4,0),ylim=c(-4,0))
+abline(0,1)
+(power_wilcox <- sum(a[,1]<0.05)/nrow(a))
+(power_hyper3 <- sum(a[,2]<0.05)/nrow(a))
+
+table(wilcox=a[,1]<0.05,hyper3=a[,2]<0.05)
 ```

### 2022-08-08 inst/race3.Rmd
point to the javelin analysis
more work on constructors comparing mercedes from 2020-2021

@@
 
+# Constructors' championship 2020
+
+```{r,label=constructor2020,cache=TRUE}
+jj <- read.table("constructor_2020.txt",header=TRUE)
+i <- seq(from=1,by=2,to=nrow(jj)) # extract every second entry for points
+constructor_points_2020 <- jj[i,ncol(jj)]
+names(constructor_points_2020) <- jj[i,1]
+
+constructor_points_2020 <- sort(constructor_points_2020,dec=TRUE)
+constructor_points_2020
+
+constructor_table_2020 <- jj[,-ncol(jj)]
+constructor_table_2020[,1:17]
+constructor_2020_table <- constructor_table_2020
+constructor_2020 <- ordertable2supp3(constructor_table_2020)
+```
+
+```{r,label=maxpconstructor2020,cache=TRUE}
+constructor_2020_maxp <- maxp(constructor_2020,n=1)
+constructor_2020_maxp <- sort(constructor_2020_maxp,dec=TRUE)
+constructor_2021_maxp <- sort(constructor_2021_maxp,dec=TRUE)
+constructor_2020_maxp
+constructor_2021_maxp
+```
+
+```{r,label=changenames,cache=TRUE}
+constructor_2020_and_2021 <- psubs(constructor_2020,"Merc","Merc2020") + psubs(constructor_2021,"Merc","Merc2021")
+```
+
+```{r,doconstructor20202021test,cache=TRUE}
+ans <- samep.test(constructor_2020_and_2021,c("Merc2020","Merc2021"))
+```
+
+```{r,show20201test}
+ans
+```
+
+Note that the analysis in `javelin.Rmd` is appropriate here too, a
+nice example.
+
 ### Package dataset {-}
@@
 save(
-constructor_2021_maxp,
-constructor_2021,
-constructor_2021_table,
+constructor_2020_maxp,constructor_2020,constructor_2020_table,
+constructor_2021_maxp,constructor_2021,constructor_2021_table,
 file="constructor.rda")

### 2022-08-09 inst/jss_hyper3.Rnw
change insect example to chorioamnion dataset from wilcox.Rd

@@
 test~\citep{mann1947,wilcoxon1945} assesses a null of identity of
-underlying distributions.  Consider the fly spray dataset presented by
-\citeauthor{wilcoxon1945}:
+underlying distributions.  Consider the chorioamnion dataset at
+\code{wilcox.test.Rd}:
 
-<<defineAB>>=
-A <- c(68,68,59,72,64,67,70,74)
-B <- c(60,67,61,62,67,63,56,58)
+<<define_xy_wilcox>>=
+x <- c(0.80, 0.83, 1.89, 1.04, 1.45, 1.38, 1.91, 1.64, 0.73, 1.46)
+y <- c(1.15, 0.88, 0.90, 0.74, 1.21)
 @
 
-Above we see the percentage mortality of two preparations, A and B,
-each applied eight times.  We may easily convert this dataset to a
-rank order statistic:
+Here we see permeability constants of the human chorioamnion at term
+(\code{x}) and between 12 to 26 weeks gestational age (\code{y}).  The
+order statistic is straightforward to calculate:
 
-<<makeABorderstatistic>>=
-names(A) <- rep("a",length(A))
-names(B) <- rep("b",length(B))
-sort(c(A,B))
-names(sort(c(A,B)))
+<<hyper3osdef,cache=TRUE>>=
+names(x) <- rep("x",length(x))
+names(y) <- rep("y",length(y))
+(os <- names(sort(c(x,y))))
 @
 
-And thence to a likelihood function:
+Then object \code{os} is converted to a \code{hyper3} object, again
+with \code{ordervec2supp3()}, which may be assessed using the Method
+of Support:
 
-<<definehyper3AB,cache=TRUE>>=
-(Hab <- ordervec2supp3(names(sort(c(A,B)))))
+<<hyper3xytest,cache=TRUE>>=
+Hxy <- ordervec2supp3(os)
+equalp.test(Hxy)
 @
 
-Package idiom for testing the null of equality is straightforward:
-
-<<testABforequality,cache=TRUE>>=
-equalp.test(Hab)
-@
-
-Compare \citeauthor{wilcoxon1945}'s $p$-value of $1.77\%$.  However,
-observe that the \code{hyper3} likelihood approach gives more
-information than Wilcoxon's analysis: Firstly, we see from the
-evaluate that the probability of a single application of preparation A
-killing more flies than B is about 15\%.  Further, we may plot a
-support curve for this dataset as above:
+Above, we use generic function \code{equalp.test()} to test the null
+that the permeability of the two groups both have Bradley-Terry
+strength of $0.5$.  We see a p-value of about 0.09; compare 0.25 from
+\code{wilcox.test()}.  However, observe that the \code{hyper3}
+likelihood approach gives more information than Wilcoxon's analysis:
+Firstly, we see that the maximum likelihood estimate for the
+Bradley-Terry strength of \code{x} is about 0.24, considerably less
+than the null of 0.5; further, we may plot a support curve for this
+dataset, given in Figure~\ref{wilcox_likelihood}.
 
@@
 <<plotwilcoxlike,cache=TRUE>>=
-a <- seq(from=0.02,to=0.5,len=50)
-L <- sapply(a,function(p){loglik(p,Hab)})
+a <- seq(from=0.02,to=0.8,len=40)
+L <- sapply(a,function(p){loglik(p,Hxy)})
 plot(a,L-max(L),type='b')
 abline(h=c(0,-2))
-abline(v=0.1558)
+abline(v=c(0.24))
+abline(v=c(0.5),lty=2)
 @
 \caption{A\label{wilcox_likelihood} support for the Bradley-Terry
-strength $p_a$ of preparation A~\citep{wilcoxon1945}.  The evaluate of
-0.1558 is shown; and the two-units-of support credible interval, which
-excludes $p_a=0.5$}
+strength $p_a$ of permeability at term.  The evaluate of 0.24 is
+shown; and the two-units-of support credible interval, which does not
+exclude $H_0\colon p_a=0.5$ (dotted line), is also shown}
 \end{center}
@@
 
-The ideas presented above are easily extended to multiple types of
+The ideas presented above may be extended to more than two types of
 competitors.  Consider the following table, drawn from the men's
@@
 distances, and names being the competitors, using
-`attemptstable2supp3()`:
+\code{attemptstable2supp3()}:
 
@@
 use generalized Plackett-Luce to assess the contructors' performance.
-The following table shows rankings for the first 9 venues of the 2021
-season:
+The following table, included in the \code{hyper2} package as a
+dataset, shows rankings for the first 9 venues of the 2021 season:
 
@@
 retired (full details of the notation and conventions is given in the
-package at \code{constructor.Rd}).  The identity of the driver is not
-admissible information and indeed may change during a season.
-Alternatively, we may regard the driver and the constructor as a joint
-entity, with the constructor's ability to attract and retain a skilled
-driver being part of the object of inference.
+package at \code{constructor.Rd}).  The identity of the driver is
+viewed as inadmissible information and indeed may change during a
+season.  Alternatively, we may regard the driver and the constructor
+as a joint entity, with the constructor's ability to attract and
+retain a skilled driver being part of the object of inference.  The
+associated generalized Plackett-Luce \code{hyper3} object is easily
+constructed using package idiom, in this case
+\code{ordertable2supp3()}, and we may use this to assess the
+Plackett-Luce strengths of the constructors:
 
-The associated generalized Plackett-Luce \code{hyper3} object is as follows:
+<<maxpconstructor2021, cache=TRUE>>=
+const2020 <- ordertable2supp3(constructor_2020_table)
+const2021 <- ordertable2supp3(constructor_2021_table)
+maxp(const2020,n=1)
+maxp(const2021,n=1)
+@
 
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlstd{constructor_2021}
-\end{alltt}
-\begin{verbatim}
-## log( (AMM=1)^38 * (AMM=1, AR=1)^-1 * (AMM=1, AR=1, ARRF=1, ATH=1, HF=2,
-## MM=1, Merc=1, WM=2)^-1 * (AMM=1, AR=1, ARRF=1, ATH=1, HF=2, MM=1,
-## WM=2)^-2 * (AMM=1, AR=1, ARRF=1, ATH=1, HF=2, MM=2, Merc=1, WM=2)^-1 *
-## (AMM=1, AR=1, ARRF=1, ATH=1, HF=2, RBRH=1, WM=1)^-1 * (AMM=1, AR=1,
-## ARRF=1, ATH=1, HF=2, RBRH=1, WM=2)^-1 * (AMM=1, AR=1, ARRF=1, ATH=1,
-## HF=2, WM=2)^-1 * (AMM=1, AR=1, ARRF=1, ATH=2, Ferrari=1, HF=2, WM=1)^-1
-## * (AMM=1, AR=1, ARRF=1, HF=2, Merc=2, RBRH=1, WM=2)^-1 * (AMM=1, AR=1,
-## ARRF=1, HF=2, WM=1)^-1 * (AMM=1, AR=1, ARRF=2, ATH=1, Ferrari=1, HF=2,
-## MM=1, Merc=1, RBRH=1)^-1 * (AMM=1, AR=1, ARRF=2, ATH=1, Ferrari=1,
-## HF=2, MM=1, Merc=1, RBRH=1, WM=1)^-1 * (AMM=1, AR=1, ARRF=2, ATH=1,
-## Ferrari=1, HF=2, MM=1, Merc=2, RBRH=1, WM=2)^-1 * (AMM=1, AR=1, ARRF=2,
-...
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
+Above, we see the strength of Mercedes falling from about 0.38 in 2020
+to less than 0.20 in 2021 and it is natural to wonder whether this can
+be ascribed to random variation.  Observe that testing such a
+hypothesis is complicated by the fact that constructors field multiple
+cars, and also that constructors come and go, with two 2020 teams
+dropping out between years and two joining.  Package idiom to test
+this statistically would be to define a combined likelihood function
+for both years, keeping track of the year:
+
+<<definecombinedlikelihoodfunction,cache=TRUE>>=
+H <- psubs(constructor_2020,"Merc","Merc2020") + psubs(constructor_2021,"Merc","Merc2021")
+@
+
+Above, we use generic function \code{psubs()} to change the name of
+Mercedes from \code{Merc} to \code{Merc2020} and \code{Merc2021}
+respectively.  Note the use of ``\code{+}" to represent addition of
+log-likelihoods, corresponding to the assumption of conditional
+independence of results.  The null would be simply that the strengths
+of \code{Merc2020} and of \code{Merc2021} are identical.  Package
+idiom would be to use generic function \code{samep.test()}:
+
+<<usesamep,cache=TRUE>>=
+samep.test(H,c("Merc2020","Merc2021"))
+@
+
+Above, we see strong evidence for a real decrease in the strength of
+the Mercedes team from 2020 to 2021, with $p=0.002$.
 

### 2022-08-09 inst/jss_hyper3.Rnw
include an abstract

@@
 \Abstract{
-  The \code{hyper2} package is improved here to include \code{hyper3}
-  objects.
+
+  The \code{hyper2} package furnishes extensive idiom to work with
+  extensions of the Bradley-Terry probability model such as
+  Plackett-Luce likelihood functions, team strengths, and reified
+  entities (monsters).  Here, I present a generalization of
+  \code{hyper2} in which multiple entities are constrained to have
+  identical Bradley-Terry strengths.  A new \proglang{S3} class
+  \code{hyper3}, along with associated methods, is motivated and
+  introduced.  Four datasets are analysed, each analysis furnishing
+  new insight, and each highlighting different capabilities of the
+  package.
 }
@@
   \begin{center}
-<<plotwilcoxlike,cache=TRUE>>=
+<<plotwilcoxlike,echo=FALSE>>=
 a <- seq(from=0.02,to=0.8,len=40)
@@
 throw of 86.67m, and so on (\code{NA} entries correspond to
-no-throws.)  Now convert the attempts table to a \code{hyper3} object,
-again using function \code{attemptstable2supp3()} but this time pass
-\code{give.supp=TRUE}:
+no-throws.)  The attempts table may be converted to a \code{hyper3}
+object, again using function \code{attemptstable2supp3()} but this
+time passing \code{give.supp=TRUE}:
 

### 2022-08-10 inst/jss_hyper3.Rnw
abstract

@@
 library("knitr")
+library("hyper2")
 opts_chunk$set(cache=TRUE, autodep=TRUE)
@@
 
-show how the \proglang{map} class of the Standard Template Library are
+show how the \proglang{map} class of the Standard Template Library is
 used: \proglang{weightedplayervector} objects map strings to long
@@
 <<>>=
-library("hyper2")
 H <- hyper3()
@@
 
+\section{Conclusions and further work}
+
+Plackett-Luce likelihood functions for rank datasets have been
+generalized to impose identity of Bradley-Terry strengths for certain
+groups; the preferred interpretation is a running race in which the
+competitors are split into equivalence classes of clones.
+Implementing this in \proglang{R} is accomplished via a \proglang{C++}
+back-end making use of the \proglang{STL} ``map" class which offers
+efficiency advantages, especially for large objects.
+
+New likelihood functions for simple cases with three or four
+competitors were presented, and extending to larger numbers furnishes
+a generalization of the Mann-Whitney-Wilcoxon test that offers a
+specific alternative (Bradley-Terry strength) with a clear operational
+definition.  Further generalizations allow the analysis of more than
+two groups, here applied to Olympic Javelin throw distances.
+Generalized Plackett-Luce likelihood functions were used to assess the
+constructors' championship and a reasonable null.  Specifically, the
+hypothesis that the strength of the Mercedes team remained unchanged
+between 2020 and 2021 was tested and rejected.
+
+Further work might include a systematic comparison between
+\code{hyper3} approach and the Mann-Whitney-Wilcoxon test, including
+the characterisation of the power function of both tests.  Further,
+the package could easily be extended to allow non-integer
+multiplicities, which might prove useful in the context of reified
+Bradley Terry techniques~\citep{hankin2020}.
+
 \bibliography{chess}
-
 \end{document}

### 2022-08-10 inst/jss_hyper3.Rnw
rewording

@@
 and also given that the maximum likelihood estimate ($\simeq 0.41$) is
-quite close to the null of $0.5$].  Nevertheless we can apply Wilks's
-theorem for a $p$-value:
+quite close to the null of $0.5$].  Nevertheless we can follow
+\cite{edwards1972} and apply Wilks's theorem for a $p$-value:
 
@@
 The ideas presented above can easily be extended to arbitrarily large
-numbers of competitors.  Given an order statistic of the type
-considered above, the Mann-Whitney-Wilcoxon
+numbers of competitors, although the analytical expressions tend to be
+intractable and numerical methods must be used.  Given an order
+statistic of the type considered above, the Mann-Whitney-Wilcoxon
 test~\citep{mann1947,wilcoxon1945} assesses a null of identity of

### 2022-08-11 inst/jss_hyper3.Rnw
more rewriting and detail

@@
 
-This Chopra threw 87.03m on his first throw, 87.58m on his second, and
-so on.  No-throws, ignored here, are indicated with a \code{X}.  We
+Thus Chopra threw 87.03m on his first throw, 87.58m on his second, and
+so on.  No-throws, ignored here, are indicated with an \code{X}.  We
 may convert this to a named vector with elements being the throw
@@
 
+A particularly attractive feature of this analysis is that the
+Bradley-Terry strengths have direct operational significance: If two
+competitors, say Vadlejch and Vesely, were to throw a javelin, then we
+would estimate the probability that competitors, say Vadlejch would
+throw further than Vesely at $\displaystyle p_{\mbox{\tiny
+Vad}}/\left(p_{\mbox{\tiny Vad}} + p_{\mbox{\tiny Ves}}\right)\simeq
+0.74$.  Indeed, from a training or selection perspective we might
+follow \cite{hankin2017} and observe that log-contrasts
+~\citep{ohagan2004} $\log\left(p_{\mbox{\tiny Vad}}/ p_{\mbox{\tiny
+Ves}}\right)$ appear to have approximately Gaussian likelihood
+functions for observations of the type considered here.  Profile
+log-likelihood curves for log-contrasts are easily produced by the
+package, Figure~\ref{profliklogcont}.
+
+<<havealook,echo=FALSE,cache=TRUE>>=
+f <- function(s){
+  jj <- specificp.test(javelin,"Vesely",s,n=2)
+  p <- jj$null_estimate
+ return(c(log(p[6]/p[7]),jj$null_support))
+}
+Ves <- seq(from=0.0199,to=0.33,len=16)
+M <- sapply(Ves,f)
+M[2,] <- M[2,]-max(M[2,])
+rownames(M) <- c("logcontrast","support")
+@
+
+\begin{figure}[htbp]
+\begin{center}
+<<plottheloglikcont,echo=FALSE>>=
+colnames(M) <-  as.character(Ves)
+plot(t(M),type="b")
+abline(h=c(0,-2))
+abline(v=0,lty=2)
+abline(v=log(0.32062833/0.11402735)) # these from mp
+@
+\caption{Profile \label{profliklogcont} likelihood for log-contrast
+$\log\left(p_{\mbox{\tiny Vad}}/ p_{\mbox{\tiny Ves}}\right)$.  Null
+of $p_{\mbox{\tiny Vad}}=p_{\mbox{\tiny Ves}}$ indicated with a dotted
+line, and two-units-of-support limit indicated with horizontal lines;
+thus the null is not rejected}
+\end{center}
+\end{figure}
+
 \section{Formula 1 motor racing: the constructors' championship}
@@
 between {\em manufacturers} of racing cars (compare the drivers'
-championship).  Here, we consider the constructor to be the object of
-inference.  Each constructor typically fields two cars, each of which
-separately accumulates ranking-based points at each venue.  Here we
-use generalized Plackett-Luce to assess the contructors' performance.
-The following table, included in the \code{hyper2} package as a
-dataset, shows rankings for the first 9 venues of the 2021 season:
+championship, which is between drivers).  Here, we consider the
+constructor to be the object of inference.  Each constructor typically
+fields two cars, each of which separately accumulates ranking-based
+points at each venue.  Here we use generalized Plackett-Luce to assess
+the contructors' performance.  The following table, included in the
+\code{hyper2} package as a dataset, shows rankings for the first 9
+venues of the 2021 season:
 
@@
 Above, we see that Mercedes (``\code{Merc}'') came first and third at
-Bahrain (\code{BHR}, and at Emilia Romagna (\code{EMI} came second and
-retired (full details of the notation and conventions is given in the
-package at \code{constructor.Rd}).  The identity of the driver is
-viewed as inadmissible information and indeed may change during a
+Bahrain (\code{BHR}); and came second and retired at Emilia Romagna
+(\code{EMI}); full details of the notation and conventions are given
+in the package at \code{constructor.Rd}.  The identity of the driver
+is viewed as inadmissible information and indeed may change during a
 season.  Alternatively, we may regard the driver and the constructor

### 2022-08-11 inst/jss_hyper3.Rnw
avoid stupid messages

@@
 <<setup, cache=FALSE,echo=FALSE>>=
-library("knitr")
-library("hyper2")
+source("loadlibrary.R",echo=FALSE)  # to avoid stupid messages 
 opts_chunk$set(cache=TRUE, autodep=TRUE)
@@
 preferred interpretation is a race (as in track and field atheletics):
-given six competitors $1-6$, we ascribe to them nonnegative strengths
+Given six competitors $1-6$, we ascribe to them nonnegative strengths
 $p_1,\ldots, p_6$; the probability that $i$ beats $j$ is
@@
 \frac{d\mathcal{L}}{da}=\frac{(1+a)(1-2a)-a(1-a)}{(1+a)^2}=
-\frac{1-2a+a-2a^2 -a + a^2}{(1+a)^2}=
 \frac{1-2a-a^2}{(1+a)^2}
@@
 above, we see close agreement with the theoretical value of
-$(\sqrt{2}-1,1-\sqrt{2})\simeq (0.414,0.586)$.  Observe that the
+$(\sqrt{2}-1,2-\sqrt{2})\simeq (0.414,0.586)$.  Observe that the
 maximum likelihood estimate for $a$ is strictly less than 0.5, even

### 2022-08-11 inst/loadlibrary.R

@@ -1,0 +1,2 @@
+suppressMessages(library("knitr",quietly=TRUE,verbose=FALSE))
+suppressMessages(library("hyper2",quietly=TRUE,verbose=FALSE))

### 2022-08-12 inst/jss_hyper3.Rnw
point to github repo
rewording

@@
 numbers of competitors, although the analytical expressions tend to be
-intractable and numerical methods must be used.  Given an order
-statistic of the type considered above, the Mann-Whitney-Wilcoxon
-test~\citep{mann1947,wilcoxon1945} assesses a null of identity of
-underlying distributions.  Consider the chorioamnion dataset at
-\code{wilcox.test.Rd}:
+intractable and numerical methods must be used.  All results and
+datasets presented here are maintained under version control and
+available at \url{https:github.com/RobinHankin/hyper2}.  Given an
+order statistic of the type considered above, the
+Mann-Whitney-Wilcoxon test~\citep{mann1947,wilcoxon1945} assesses a
+null of identity of underlying distributions.  Consider the
+chorioamnion dataset at \code{wilcox.test.Rd}:
 
@@
 competitors, say Vadlejch and Vesely, were to throw a javelin, then we
-would estimate the probability that competitors, say Vadlejch would
-throw further than Vesely at $\displaystyle p_{\mbox{\tiny
-Vad}}/\left(p_{\mbox{\tiny Vad}} + p_{\mbox{\tiny Ves}}\right)\simeq
-0.74$.  Indeed, from a training or selection perspective we might
-follow \cite{hankin2017} and observe that log-contrasts
-~\citep{ohagan2004} $\log\left(p_{\mbox{\tiny Vad}}/ p_{\mbox{\tiny
-Ves}}\right)$ appear to have approximately Gaussian likelihood
-functions for observations of the type considered here.  Profile
-log-likelihood curves for log-contrasts are easily produced by the
-package, Figure~\ref{profliklogcont}.
+would estimate the probability that Vadlejch would throw further than
+Vesely at $\displaystyle p_{\mbox{\tiny Vad}}/\left(p_{\mbox{\tiny
+Vad}} + p_{\mbox{\tiny Ves}}\right)\simeq 0.74$.  Indeed, from a
+training or selection perspective we might follow \cite{hankin2017}
+and observe that log-contrasts~\citep{ohagan2004}
+$\log\left(p_{\mbox{\tiny Vad}}/ p_{\mbox{\tiny Ves}}\right)$ appear
+to have approximately Gaussian likelihood functions for observations
+of the type considered here.  Profile log-likelihood curves for
+log-contrasts are easily produced by the package,
+Figure~\ref{profliklogcont}.
 

### 2022-08-12 inst/jss_hyper3.Rnw
slight rewording
cite some relevant work

@@
 
-\section{introduction}
+\section{Introduction}
 
@@
 Plackett-Luce~\citep{plackett1975} likelihood functions.  Our
-preferred interpretation is a race (as in track and field atheletics):
+preferred interpretation is a race (as in track and field athletics):
 Given six competitors $1-6$, we ascribe to them nonnegative strengths
@@
 
-The \code{hyper2} package generalizes this likelihood function to
-functions of ${\mathbf p}=(p_1,\ldots,p_n)$ with
+\citet{mollica2014} call this a forward ranking process on the grounds
+that the best (preferred; fastest; chosen) entities are identified in
+the same sequence as their rank.  The \code{hyper2} package
+generalizes this likelihood function to functions of ${\mathbf
+  p}=(p_1,\ldots,p_n)$ with
 
@@
 are integers which may be positive or negative.  It becomes possible
-to analyse a wider range of observations such as incomplete order
-statistics or teams; and the introduction of reified entities allows
-one to consider draws~\citep{hankin2009}, noncompetitive tactics such
-as collusion~\citep{hankin2020}, and the phenomenon of team cohesion
+to analyse a wider range of likelihood functions than standard
+Plackett-Luce~\citep{turner2020}.  For example, results involving
+incomplete order statistics or teams are tractable. Further, the
+introduction of reified entities allows one to consider
+draws~\citep{hankin2009}, noncompetitive tactics such as
+collusion~\citep{hankin2020}, and the phenomenon of team cohesion
 wherein the team becomes stronger than the sum of its parts.
 
-However, consider a race in which there are six runners 1-6 but we
-happen to know that three of the runners (1,2,3) are clones of
-strength $p_a$, two of the runners (4,5) have strength $p_b$, and the
-final runner (6) is of strength $p_c$.  We normalise so
-$p_a+p_b+p_c=1$.  The runners race and the finishing order is:
+Here I present a different generalization.  Consider a race in which
+there are six runners 1-6 but we happen to know that three of the
+runners (1,2,3) are clones of strength $p_a$, two of the runners (4,5)
+have strength $p_b$, and the final runner (6) is of strength $p_c$.
+We normalise so $p_a+p_b+p_c=1$.  The runners race and the finishing
+order is:
 
@@
 between {\em manufacturers} of racing cars (compare the drivers'
-championship, which is between drivers).  Here, we consider the
-constructor to be the object of inference.  Each constructor typically
+championship, which is between drivers).  In this analysis, the
+constructor is the object of inference.  Each constructor typically
 fields two cars, each of which separately accumulates ranking-based

### 2022-08-13 inst/jss_hyper3.Rnw
more rewording

@@
 
-If we allow non-finishers [that is, a subset of competitors who are
-beaten by all the ranked competitors], there is another nontrivial
-order statistic, viz $a\succ b\succ\left\lbrace a,b\right\rbrace$
-[thus one of the two $a$'s won, one of the $b$'s came second, and one
-of each of $a$ and $b$ failed to finish].
+If we allow non-finishers---that is, a subset of competitors who are
+beaten by all the ranked competitors: \cite{turner2020} call this a
+{\em top $n$ ranking}), there is another nontrivial order statistic,
+viz $a\succ b\succ\left\lbrace a,b\right\rbrace$ [thus one of the two
+  $a$'s won, one of the $b$'s came second, and one of each of $a$ and
+  $b$ failed to finish].
 
@@
 
+Draws are not considered in the present work but in principle may be
+accommodated, either using likelihoods comprising sums of
+Plackett-Luce probabilities~\citep{hankin2017}; or the introduction of
+a reified draw entity~\citep{hankin2020}.
+
 Further work might include a systematic comparison between
 \code{hyper3} approach and the Mann-Whitney-Wilcoxon test, including
-the characterisation of the power function of both tests.  Further,
-the package could easily be extended to allow non-integer
-multiplicities, which might prove useful in the context of reified
-Bradley Terry techniques~\citep{hankin2020}.
+the characterization of the power function of both tests.  The package
+could easily be extended to allow non-integer multiplicities, which
+might prove useful in the context of reified Bradley Terry
+techniques~\citep{hankin2020}.
 

### 2022-08-13 inst/jss_hyper3.Rnw
rewording

@@
 and in this case we need multisets\footnote{Note that the version of
-\code{hyper2} presented by \citet{hankin2017} used integer-valued sets
-together with a print method that used a complicated mapping system
-from integers to competitor names.  Current methodology [following
-commit \code{51a8b46}] is to use sets of character strings which
-represent the competitors directly; this allows for easier combination
-of observations including different competitors.}.  The declarations
+\code{hyper2} presented by \citet{hankin2017} and reviewed
+by~\cite{turner2020} used integer-valued sets together with a print
+method that used a complicated mapping system from integers to
+competitor names.  Current methodology [following commit
+  \code{51a8b46}] is to use sets of character strings which represent
+the competitors directly; this allows for easier combination of
+observations including different competitors.}.  The declarations
 
@@
 \code{hyper3} approach and the Mann-Whitney-Wilcoxon test, including
-the characterization of the power function of both tests.  The package
+the characterisation of the power function of both tests.  The package
 could easily be extended to allow non-integer multiplicities, which

### 2022-08-13 inst/jss_hyper3.Rnw
slight rewording

@@
 
-\begin{itemize}
-\item $A\succ A\succ B\longrightarrow\frac{a^2}{1+a}$
-\item $A\succ B\succ A\longrightarrow\frac{1-2a-a^2}{(1+a)^2}$
-\item $B\succ A\succ A\longrightarrow\frac{1-a}{1+a}$
-\end{itemize}
-
-Likelihood functions for these order statistics are given in
-Figure~\ref{ABAetc_likelihood}.  It can be shown that the Fisher
-information for such observations is
+If we have two clones of $a$ and a singleton $b$, then there are three
+possible rank statistics: (i), $a\succ a\succ b$ with probability
+$\frac{a^2}{1+a}$; (ii), $a\succ b\succ a$, with
+$\frac{1-2a-a^2}{(1+a)^2}$, (iii), $b\succ a\succ a$ at
+$\frac{1-a}{1+a}$.  Likelihood functions for these order statistics
+are given in Figure~\ref{ABAetc_likelihood}.  It can be shown that the
+Fisher information for such observations is
 
@@
 
-which has a minimum at about $0.575$.
+which has a minimum of about $13.9$ at at about $a=0.575$.
 
@@
 If we allow non-finishers---that is, a subset of competitors who are
-beaten by all the ranked competitors: \cite{turner2020} call this a
+beaten by all the ranked competitors (\cite{turner2020} call this a
 {\em top $n$ ranking}), there is another nontrivial order statistic,
@@
   $a$'s won, one of the $b$'s came second, and one of each of $a$ and
-  $b$ failed to finish].
+  $b$ failed to finish].  Now
 

### 2022-08-14 inst/jss_hyper3.Rnw
slight reorganisation

@@
 
-\subsection{All results for AAB}
-
+\subsubsection{Fisher information}
 If we have two clones of $a$ and a singleton $b$, then there are three

### 2022-08-14 inst/jss_hyper3.Rnw
minor tidyup

@@
 <<>>=
-H <- hyper3()
-H[c(a=1)] <- 1
-H[c(a=3,b=2,c=1)] <- -1
-H
+LL <- hyper3()
+LL[c(a=1)] <- 1
+LL[c(a=3,b=2,c=1)] <- -1
+LL
 @
 
-Above, we see object \code{H} is a log-likelihood function of the
+Above, we see object \code{LL} is a log-likelihood function of the
 players' strengths, which may be evaluated at specified points in
@@
 <<>>=
-loglik(c(a=0.01,b=0.01,c=0.98),H)
-loglik(c(a=0.9,b=0.05,c=0.05),H)
+loglik(c(a=0.01,b=0.01,c=0.98),LL)
+loglik(c(a=0.9,b=0.05,c=0.05),LL)
 @
@@
 
-We may use a selection of statistical tests on the order statistic,
-modified from~\citep{hankin2017} to work with \code{hyper3} objects.
-For example, we may assess the hypothesis that all three players are
-of equal strength [viz $H_0\colon p_a=p_b=p_c=\frac{1}{3}$]:
+The package provides a number of statistical tests on likelihood
+functions, modified from~\citep{hankin2017} to work with \code{hyper3}
+objects.  For example, we may assess the hypothesis that all three
+players are of equal strength [viz $H_0\colon
+p_a=p_b=p_c=\frac{1}{3}$]:
 
@@
 Here I consider some order statistics with nontrivial maximum
-likelihood Bradley-Terry strength.  The simplest nontrivial case would
-be three competitors with strengths $a,a,b$ and finishing order
+likelihood Bradley-Terry strengths.  The simplest nontrivial case
+would be three competitors with strengths $a,a,b$ and finishing order
 $a\succ b\succ a$.  The likelihood function would be
@@
 
-This will be zero at $\sqrt{2}\pm 1$; we also note that
+This will be zero at $\sqrt{2}-1$; we also note that
 $d^2\mathcal{L}/da^2=-4(1+a)^{-3}$, manifestly strictly negative for
@@
 
-above, we see close agreement with the theoretical value of
+Above, we see close agreement with the theoretical value of
 $(\sqrt{2}-1,2-\sqrt{2})\simeq (0.414,0.586)$.  Observe that the
@@
 informative, in the sense that we can find a credible interval for
-$a$.  With an $n$-unit of support criterion the analytical solution to
+$a$.  With an $n$-units of support criterion the analytical solution to
 $\mathcal{S}(p)=-n$ is given by defining $X=\log(3-2\sqrt{2})-n$ and
@@
 null of identity of underlying distributions.  Consider the
-chorioamnion dataset at \code{wilcox.test.Rd}:
+chorioamnion dataset~\citep{hollander2013}, used in
+\code{wilcox.test.Rd}:
 
@@
 
-Here we see permeability constants of the human chorioamnion at term
-(\code{x}) and between 12 to 26 weeks gestational age (\code{y}).  The
-order statistic is straightforward to calculate:
+Here we see a measure of permeability of the human placenta at term
+(\code{x}) and between 3 and 6 months' gestational age (\code{y}).
+The order statistic is straightforward to calculate:
 
@@
 training or selection perspective we might follow \cite{hankin2017}
-and observe that log-contrasts~\citep{ohagan2004}
-$\log\left(p_{\mbox{\tiny Vad}}/ p_{\mbox{\tiny Ves}}\right)$ appear
-to have approximately Gaussian likelihood functions for observations
-of the type considered here.  Profile log-likelihood curves for
-log-contrasts are easily produced by the package,
-Figure~\ref{profliklogcont}.
+and observe that log-contrasts~\citep{ohagan2004} appear to have
+approximately Gaussian likelihood functions for observations of the
+type considered here.  Profile log-likelihood curves for log-contrasts
+are easily produced by the package, Figure~\ref{profliklogcont}.  We
+see that the credible range for~$\log\left(p_{\mbox{\tiny Vad}}/
+p_{\mbox{\tiny Ves}}\right)$ includes zero and we have no strong
+evidence for these athletes having different (Bradley-Terry)
+strengths.
 
@@
 
-In Formula 1 motor racing, the constructors' championship takes place
-between {\em manufacturers} of racing cars (compare the drivers'
-championship, which is between drivers).  In this analysis, the
-constructor is the object of inference.  Each constructor typically
-fields two cars, each of which separately accumulates ranking-based
-points at each venue.  Here we use generalized Plackett-Luce to assess
-the contructors' performance.  The following table, included in the
-\code{hyper2} package as a dataset, shows rankings for the first 9
-venues of the 2021 season:
+Formula 1 motor racing is an important and prestigious motor sport
+\citep{codling2017,jenkins2010}.  In Formula 1 Grand Prix, the
+constructors' championship takes place between {\em manufacturers} of
+racing cars (compare the drivers' championship, which is between
+drivers).  In this analysis, the constructor is the object of
+inference.  Each constructor typically fields two cars, each of which
+separately accumulates ranking-based points at each venue.  Here we
+use generalized Plackett-Luce to assess the contructors' performance.
+The following table, included in the \code{hyper2} package as a
+dataset, shows rankings for the first 9 venues of the 2021 season:
 
@@
 definition.  Further generalizations allow the analysis of more than
-two groups, here applied to Olympic Javelin throw distances.
+two groups, here applied to Olympic javelin throw distances.
 Generalized Plackett-Luce likelihood functions were used to assess the
-constructors' championship and a reasonable null.  Specifically, the
-hypothesis that the strength of the Mercedes team remained unchanged
-between 2020 and 2021 was tested and rejected.
+Grand Prix constructors' championship and a reasonable null.
+Specifically, the hypothesis that the strength of the Mercedes team
+remained unchanged between 2020 and 2021 was tested and rejected.
 

### 2022-08-14 inst/jss_hyper3.Rnw
typos

@@
 strength, but further analysis with \code{equalp.test()} reveals that
-there is no strong evidence in the datast to reject the hypothesis of
+there is no strong evidence in the dataset to reject the hypothesis of
 equal competitive strength ($p=0.26$), or that Vadlejch has a strength
@@
 separately accumulates ranking-based points at each venue.  Here we
-use generalized Plackett-Luce to assess the contructors' performance.
+use generalized Plackett-Luce to assess the constructors' performance.
 The following table, included in the \code{hyper2} package as a

### 2022-08-15 inst/jss_hyper3.Rnw
typos

@@
 
-Thus the winner was $a$, second place for $b$, third for $a$, and so
+Thus the winner was $a$, second place for $c$, third for $b$, and so
 on.  Alternatively we might say that $a$ came first, fourth, and
@@
 \code{[a,b,c]} of nonnegative values with unit sum] using standard
-\proglang{R} idiom wrapping \proglang{C++} back end.
-
-The package includes an \proglang{S4} class \code{hyper3} for this
-type of object.  Package idiom for creating such objects uses named
-vectors:
+\proglang{R} idiom wrapping \proglang{C++} back end.  The package
+includes an \proglang{S3} class \code{hyper3} for this type of object.
+Package idiom for creating such objects uses named vectors:
 
@@
 
-We may find a maximum likelihood estimate for the players' strengths:
+We may find a maximum likelihood estimate for the players' strengths,
+using generic function \code{maxp()}:
 

### 2022-08-16 inst/jss_hyper3.Rnw
slight modifications including citing hankin 2010

@@
 collusion~\citep{hankin2020}, and the phenomenon of team cohesion
-wherein the team becomes stronger than the sum of its parts.
+wherein the team becomes stronger than the sum of its
+parts~\citep{hankin2010}.
 
@@
 would be three competitors with strengths $a,a,b$ and finishing order
-$a\succ b\succ a$.  The likelihood function would be
+$a\succ b\succ a$.  The Plackett-Luce likelihood function would be
 
@@
 
-In this case we know that $a+b=1$ so this is equal to
-$\mathcal{L}=\mathcal{L}(a)=\frac{a(1-a)}{1+a}$.
+and in this case we know that $a+b=1$ so this is equal to
+$\mathcal{L}=\mathcal{L}(a)=\frac{a(1-a)}{1+a}$.  The score would be
+given by
 
@@
 
-This will be zero at $\sqrt{2}-1$; we also note that
+and this will be zero at $\sqrt{2}-1$; we also note that
 $d^2\mathcal{L}/da^2=-4(1+a)^{-3}$, manifestly strictly negative for
@@
 
-which has a minimum of about $13.9$ at at about $a=0.575$.
+which has a minimum of about $13.9$ at at about $a=0.575$.  This is
+intermediate between the two-runner case where the Fisher information
+exceeds 4 (minimum at $p_a=\frac{1}{2}$), and the case of three
+distinct runners, where the Fisher information exceeds
+$\frac{1323}{16}\simeq 82.7$ (minimum at $p_a=p_b=p_c=\frac{1}{3}$); a
+brief discussion is given at \code{inst/fisher\_inf\_PL3.Rmd}.
 
@@
 Mann-Whitney-Wilcoxon test~\citep{mann1947,wilcoxon1945} assesses a
-null of identity of underlying distributions.  Consider the
-chorioamnion dataset~\citep{hollander2013}, used in
+null of identity of underlying distributions~\citep{ahmad1996}.
+Consider the chorioamnion dataset~\citep{hollander2013}, used in
 \code{wilcox.test.Rd}:
@@
 Plackett-Luce probabilities~\citep{hankin2017}; or the introduction of
-a reified draw entity~\citep{hankin2020}.
+a reified draw entity~\citep{hankin2010}.
 

### 2022-08-16 inst/jss_hyper3.Rnw
more careful wording

@@
 intermediate between the two-runner case where the Fisher information
-exceeds 4 (minimum at $p_a=\frac{1}{2}$), and the case of three
-distinct runners, where the Fisher information exceeds
-$\frac{1323}{16}\simeq 82.7$ (minimum at $p_a=p_b=p_c=\frac{1}{3}$); a
-brief discussion is given at \code{inst/fisher\_inf\_PL3.Rmd}.
+has a minimum of 4 at $p_a=p_b=\frac{1}{2}$, and the case of three
+distinct runners $a,b,c$, where the Fisher information has a minimum
+of $\frac{1323}{16}\simeq 82.7$ at $p_a=p_b=p_c=\frac{1}{3}$; a brief
+discussion is given at \code{inst/fisher\_inf\_PL3.Rmd}.
 

### 2022-08-17 inst/jss_hyper3.Rnw
the loser

--- 2022-08-16 inst/jss_hyper3.Rnw
@@
 
-which has a minimum of about $13.9$ at at about $a=0.575$.  This is
-intermediate between the two-runner case where the Fisher information
-exceeds 4 (minimum at $p_a=\frac{1}{2}$), and the case of three
-distinct runners, where the Fisher information exceeds
-$\frac{1323}{16}\simeq 82.7$ (minimum at $p_a=p_b=p_c=\frac{1}{3}$); a
-brief discussion is given at \code{inst/fisher\_inf\_PL3.Rmd}.
+which has a minimum of about $13.9$ at at about $a=0.575$.  We can
+compare this with the Fisher information ${\mathcal I}$ for the case
+of three distinct runners $a,b,c$, evaluated at its minimum of
+$p_a=p_b=p_c=\frac{1}{3}$.  If we observe the complete order
+statistic, ${\mathcal I}=\frac{1323}{16}\simeq 82.7$; if we observe
+just the winner, ${\mathcal I}=27$, and if we observe just the loser
+we have ${\mathcal I}=\frac{16875}{256}\simeq 65.9$.  A brief
+discussion is given at \code{inst/fisher\_inf\_PL3.Rmd}.
 

### 2022-08-22 inst/jss_hyper3.Rnw
disordR and Fisher information for nonfinishers, a>b>{a,b} etc

@@
 \proglang{R} idiom wrapping \proglang{C++} back end.  The package
-includes an \proglang{S3} class \code{hyper3} for this type of object.
-Package idiom for creating such objects uses named vectors:
+includes an \proglang{S3} class \code{hyper3} for this type of object;
+extraction and replacement methods use \pkg{disordR}
+discipline~\citep{hankin2022}.  Package idiom for creating such
+objects uses named vectors:
 
@@
 which has a minimum of about $13.9$ at at about $a=0.575$.  We can
-compare this with the Fisher information ${\mathcal I}$ for the case
-of three distinct runners $a,b,c$, evaluated at its minimum of
-$p_a=p_b=p_c=\frac{1}{3}$.  If we observe the complete order
-statistic, ${\mathcal I}=\frac{1323}{16}\simeq 82.7$; if we observe
-just the winner, ${\mathcal I}=27$, and if we observe just the loser
-we have ${\mathcal I}=\frac{16875}{256}\simeq 65.9$.  A brief
-discussion is given at \code{inst/fisher\_inf\_PL3.Rmd}.
+compare this with the Fisher information matrix ${\mathcal I}$, for
+the case of three distinct runners $a,b,c$, evaluated at its minimum
+of $p_a=p_b=p_c=\frac{1}{3}$.  If we observe the complete order
+statistic, $\left|{\mathcal I}\right| =\frac{1323}{16}\simeq 82.7$; if
+we observe just the winner, $\left|{\mathcal I}\right|=27$, and if we
+observe just the loser we have $\left|{\mathcal
+I}\right|=\frac{16875}{256}\simeq 65.9$.  A brief discussion is given
+at \code{inst/fisher\_inf\_PL3.Rmd}.
 
@@
 @
+
+The Fisher information for such observations has a minimum of
+$\frac{68}{9}\simeq 7.56$ at $a=\frac{1}{2}$.  An inference problem
+for a dataset including nonfinishers will be given below in
+section~\ref{f1}.
 
@@
 
-\section{Formula 1 motor racing: the constructors' championship}
+\section{Formula 1 motor racing: the constructors' championship}\label{f1}
 

### 2022-08-22 inst/jss_hyper3.Rnw
spaces after commas

@@
 LL[c(a=1)] <- 1
-LL[c(a=3,b=2,c=1)] <- -1
+LL[c(a=3, b=2, c=1)] <- -1
 LL
@@
 <<>>=
-loglik(c(a=0.01,b=0.01,c=0.98),LL)
-loglik(c(a=0.9,b=0.05,c=0.05),LL)
+loglik(c(a=0.01, b=0.01, c=0.98),LL)
+loglik(c(a=0.90, b=0.05, c=0.05),LL)
 @
@@
 <<>>=
-(H <- ordervec2supp3(c("a","c","b","a","a","b")))
+(H <- ordervec2supp3(c("a", "c", "b", "a", "a", "b")))
 @
@@
 <<maxpaba,cache=TRUE>>=
-maxp(ordervec2supp3(c("a","b","a")))
+maxp(ordervec2supp3(c("a", "b", "a")))
 @
@@
 <<figabbabb,echo=FALSE>>=
-f_aab <- function(a){a^2/(1+a)}
-f_aba <- function(a){a*(1-a)/(1+a)^2}
-f_baa <- function(a){(1-a)/(1+a)}
+f_aab <- function(a){a^2 / (1 + a)}
+f_aba <- function(a){a * (1 - a) / (1 + a)^2}
+f_baa <- function(a){(1 - a) / (1 + a)}
 
-p <- function(f,...){points(a,f(a)/max(f(a)),...)}
-plot(0:1,0:1,xlab="p(a)",ylab="Likelihood",type="n")
-p(f_aab,type="l",col="black")
-p(f_aba,type="l",col="red")
-p(f_baa,type="l",col="blue")
-legend(x=0.85,y=0.7,col=c("black","red","blue"),lty=1,pch=NA,legend=c("AAB","ABA","BAA"))
-abline(h=exp(-2),lty=2)
+p <- function(f, ...){points(a, f(a) / max(f(a)), ...)}
+plot(0:1, 0:1, xlab="p(a)", ylab="Likelihood", type="n")
+p(f_aab, type="l", col="black")
+p(f_aba, type="l", col="red")
+p(f_baa, type="l", col="blue")
+legend(x=0.85, y=0.7, col=c("black", "red", "blue"), lty=1, pch=NA, legend=c("AAB", "ABA", "BAA"))
+abline(h=exp(-2), lty=2)
 @
@@
 <<cache=TRUE>>=
-maxp(ordervec2supp3(c("a","b"),nonfinishers=c("a","b")))
+maxp(ordervec2supp3(c("a", "b"), nonfinishers=c("a", "b")))
 @
@@
 <<hyper3osdef,cache=TRUE>>=
-names(x) <- rep("x",length(x))
-names(y) <- rep("y",length(y))
-(os <- names(sort(c(x,y))))
+names(x) <- rep("x", length(x))
+names(y) <- rep("y", length(y))
+(os <- names(sort(c(x, y))))
 @
@@
 <<plotwilcoxlike,echo=FALSE>>=
-a <- seq(from=0.02,to=0.8,len=40)
-L <- sapply(a,function(p){loglik(p,Hxy)})
-plot(a,L-max(L),type='b')
-abline(h=c(0,-2))
+a <- seq(from=0.02, to=0.8, len=40)
+L <- sapply(a, function(p){loglik(p, Hxy)})
+plot(a, L - max(L),type='b')
+abline(h=c(0, -2))
 abline(v=c(0.24))
-abline(v=c(0.5),lty=2)
+abline(v=c(0.5), lty=2)
 @
@@
 <<label=converttosupp3,cache=TRUE>>=
-javelin_vector <- attemptstable2supp3(javelin_table,decreasing=TRUE,give.supp=FALSE)
+javelin_vector <- attemptstable2supp3(javelin_table, decreasing=TRUE, give.supp=FALSE)
 javelin_vector
@@
 (mj <- maxp(javelin))
-dotchart(mj,pch=16)
+dotchart(mj, pch=16)
 @
@@
 f <- function(s){
-  jj <- specificp.test(javelin,"Vesely",s,n=2)
+  jj <- specificp.test(javelin, "Vesely", s, n=2)
   p <- jj$null_estimate
- return(c(log(p[6]/p[7]),jj$null_support))
+ return(c(log(p[6] / p[7]), jj$null_support))
 }
-Ves <- seq(from=0.0199,to=0.33,len=16)
-M <- sapply(Ves,f)
-M[2,] <- M[2,]-max(M[2,])
-rownames(M) <- c("logcontrast","support")
+Ves <- seq(from=0.0199, to=0.33, len=16)
+M <- sapply(Ves, f)
+M[2,] <- M[2,] - max(M[2,])
+rownames(M) <- c("logcontrast", "support")
 @
@@
 colnames(M) <-  as.character(Ves)
-plot(t(M),type="b")
-abline(h=c(0,-2))
-abline(v=0,lty=2)
+plot(t(M), type="b")
+abline(h=c(0, -2))
+abline(v=0, lty=2)
 abline(v=log(0.32062833/0.11402735)) # these from mp
@@
 const2021 <- ordertable2supp3(constructor_2021_table)
-maxp(const2020,n=1)
-maxp(const2021,n=1)
+maxp(const2020, n=1)
+maxp(const2021, n=1)
 @
@@
 <<definecombinedlikelihoodfunction,cache=TRUE>>=
-H <- psubs(constructor_2020,"Merc","Merc2020") + psubs(constructor_2021,"Merc","Merc2021")
+H <- psubs(constructor_2020, "Merc", "Merc2020") + psubs(constructor_2021, "Merc", "Merc2021")
 @
@@
 <<usesamep,cache=TRUE>>=
-samep.test(H,c("Merc2020","Merc2021"))
+samep.test(H, c("Merc2020", "Merc2021"))
 @

### 2022-08-25 inst/collusion3.Rmd
try collusion with hyper3 objects

---
title: "Chess collusion with hyper3 objects"
author: "Robin KS Hankin"
output: html_document
---

```{r setup, include=FALSE}
set.seed(1)
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
```

<p style="text-align: right;">
![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
</p>

We analyse results from Curacao 1962 using `hyper3` formalism using
the weighted Draw monster appraoch outlined in `kka.Rmd`:

$$
\left(\frac{\lambda p_1, D(p_1+p_2),p_2}{\lambda p_1+D(p_1+p_2)+p_2}\right)^{(a,b,c)}
\left(\frac{p_1, D(p_1+p_2),\lambda p_2}{p_1+D(p_1+p_2)+\lambda p_2}\right)^{(d,e,f)}
$$

\newcommand{\wld}[3]{\mathord{+}#1\,\mathord{=}#2\,\mathord{-}#3}

Above, $p_1$ plays $p_2$ a total of $a+b+c+d+e+f$ times; we have
$a+b+c$ matches with $p_1$ playing white and $p_2$ black, with $a$
wins, $b$ draws and $c$ losses [write $\wld{a}{b}{c}$], and $d+e+f$
matches with $p_1$ playing black and $p_2$ playing white with
$\wld{d}{e}{f}$.

```{r,label=loadjavelinandconvert}
cur_matches <- read.table("stockholm1962_matches.txt",header=FALSE)
head(cur_matches)
nrow(cur_matches)
```

Above, we see on the first line that Aaron (white) lost to Barcza
(black).  We will set up `hyper2` and `hyper3` objects corresponding
to likelihood functions for these observations.

```{r,label=nat}
jj <- read.table("stockholm1962.txt",header=FALSE)[,1:2]
(players <- jj$V1)
(nationalities <- jj$V2)
```

```{r,label=H2H3}
H2 <- hyper2()
H3 <- hyper3()
f <- function(lambda,D){
  for(i in seq_len(nrow(cur_matches))){
    white_player <- cur_matches[i,1]
    black_player <- cur_matches[i,2]
    result <- cur_matches[i,3]
    if(result == "1-0"){ # white player wins
       num <- lambda 
       names(num) <- white_player
    } else if(result == "0-1"){  # black player wins
       num <- 1
       names(num) <- black_player
    } else if (result == "1/2-1/2"){ # draw, duh
      num <- c(D,D)
      names(num) <- c(white_player,black_player)
    } else {
      stop("this cannot happen")
    }

    H3[num] %<>% inc()       
    den <- c(lambda+D,1+D)
    names(den) <- c(white_player,black_player)
    H3[den] %<>% dec()
  }
  return(H3)
}
```

Now use it

```{r,try11and1888,cache=TRUE}
H3 <- f(1.1,1.888)
```

```{r,showH3}
H3
summary(H3)
```

### 2022-08-29 inst/collusion3.Rmd
profile likelihood
defect of the model

@@
 
+# Short story
+
+With this model, whatever the value of $D$ and $\lambda$, there is a
+pathology which BT cannot deal with.  Still not quite 100% sure but it
+seems that the maximum likelihood estimate of the players' strengths
+is zero except for Fisher and Petrosian; and indeed we see Fisher at
+57% and Petrosian at 42% [with $\lambda = 1.1, D=1.888$].  This on the
+grounds that these players never lost playing black (there are some
+$2\times 3$ tables at the end).  This might be considered a defect of
+the probability model.
+
+# Longer story
+
 ```{r,label=loadjavelinandconvert}
@@
 ```{r,label=H2H3}
-H2 <- hyper2()
+f <- function(lambda,D){
   H3 <- hyper3()
-f <- function(lambda,D){
   for(i in seq_len(nrow(cur_matches))){
@@
 ```
+
+```{r,testH3,cache=TRUE}
+equalp.test(H3,n=1)
+```
+
+```{r,comparelikelihoods,cache=TRUE}
+maxp(f(1.1,2.1),n=1,give=TRUE)
+maxp(f(1.1,2.3),n=1,give=TRUE)
+```
+
+A break here.
+
+```{r,label=defineprofilelike,cache=FALSE}
+profl <- function(v){
+    lambda <- v[1]
+    D <- v[2]
+    H <- f(lambda,v)
+    return(loglik(maxp(H,n=1),H))
+}
+```
+
+Now apply `profl()`:
+
+```{r,calcproflike,cache=TRUE}
+profl(c(1.1,2.1))
+#profl(c(1.2,2.1))
+```
+
+```{r,calcmanyproflike,cache=TRUE} 
+np <- 4
+lambda <- seq(from=0.1,to=1.5,len=np)
+D <- seq(from=0.1,to=0.4,len=np)
+jj <- as.matrix(expand.grid(lambda,D))
+LL <- apply(jj,1,profl)
+LL
+```
+
+```{r,plotll}
+LL <- LL - max(LL)
+m <- matrix(LL,np,np)
+m
+filled.contour(lambda,D,m,xlab="lambda",ylab="D",nlevels=40)
+```
+
+```{r,plotagain}
+res <- cur_matches$V3
+res
+table(res)
+white_wins <- cur_matches[res=="1-0"    ,]
+black_wins <- cur_matches[res=="0-1"    ,]
+draws      <- cur_matches[res=="1/2-1/2",]
+(players <- unique(sort(c(cur_matches[,1],cur_matches[,2]))))
+jj <- rep(0,length(players))
+names(jj) <- players
+jj
+plays_white_wins <- jj
+plays_white_lose <- jj
+plays_white_draw <- jj
+plays_black_wins <- jj
+plays_black_lose <- jj
+plays_black_draw <- jj
+for(i in seq_len(nrow(cur_matches))){
+  white_player <- cur_matches[i,1]
+  black_player <- cur_matches[i,2]
+  result       <- cur_matches[i,3]
+
+  if(result == "1-0"){
+      plays_white_wins[white_player] %<>% inc
+      plays_black_lose[black_player] %<>% inc
+  } else if(result == "0-1"){
+      plays_white_lose[white_player] %<>% inc
+      plays_black_wins[black_player] %<>% inc
+  } else if(result == "1/2-1/2"){
+      plays_white_draw[white_player] %<>% inc
+      plays_black_draw[black_player] %<>% inc
+  }
+}
+
+plays_white_wins
+plays_white_lose
+plays_black_wins
+plays_black_lose
+```
+
+Now we need some way to describe each player's wins/losses playing
+white and black:
+
+```{r}
+f <- function(player){
+out <- matrix(c(
+  plays_white_wins[player], plays_white_lose[player], plays_white_draw[player],
+  plays_black_wins[player], plays_black_lose[player], plays_black_draw[player]
+  ),byrow=TRUE,2,3)
+  dimnames(out) <- list(colour=c("white","black"),match=c("win","lose","draw"))
+  return(out)
+}
+f("Fischer")
+f("Korchnoi")
+f("Petrosian")
+M <- f("Aaron")*0
+for(p in players){ M <- M + f(p) }
+M
+```

### 2022-09-01 inst/interzonal.Rmd

---
title: "Analysis of the 1962 world chess championship, Curacao"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("magrittr")
```

<p style="text-align: right;">
![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
</p>

This file creates likelihood functions `H` and `H_coll`, which appear in
the package as `interzonal` and `collusion` respectively.

```{r readtabs}
jj <- read.table("stockholm1962.txt",header=FALSE)
results <- as.matrix(jj[,-(1:2)])
players <- as.character(jj$V1)
nationality <- as.character(jj$V2)
rownames(results) <- players
colnames(results) <- players
points <- rowSums(results,na.rm=TRUE)
```

Now some comparison between `stockholm1962.txt` (`results`) and
`stockholm1962_matches.txt` (`results2`):

```{r makeH2,cache=TRUE}
H <- hyper2(pnames=c("white","draw",players))
H_coll <- hyper2(pnames=c("white","draw","collusion",players))

restab <- read.table("stockholm1962_matches.txt",header=FALSE)
stopifnot(all(unique(sort(c(as.character(restab$V1),as.character(restab$V2)))) == sort(players)))
results2 <- matrix(0,length(players),length(players))
rownames(results2) <- players
colnames(results2) <- players
diag(results2) <- NA

for(i in seq_len(nrow(restab))){
    white_player <- as.character(restab[i,1])
    black_player <- as.character(restab[i,2])
    match_result <- as.character(restab[i,3])

    nw <- which(players==white_player)
    nb <- which(players==black_player)

    if(match_result == "1-0"){ # white win
        results2[nw,nb] <- 2
        results2[nb,nw] <- 0

        H[c(white_player             ,"white"       )] %<>% inc()
        H[c(white_player,black_player,"white","draw")] %<>% dec()

        if(nationality[nw]=="USSR" & nationality[nb]=="USSR"){
            H_coll[c(white_player             ,"white"       )] %<>% inc()
            H_coll[c(white_player,black_player,"white","collusion")] %<>% dec()
        } else {
            H_coll[c(white_player             ,"white"       )] %<>% inc()
            H_coll[c(white_player,black_player,"white","draw")] %<>% dec()
        }
    } else if(match_result == "0-1"){ # black wins
        results2[nw,nb] <- 0
        results2[nb,nw] <- 2
        H[c(             black_player               )] %<>% inc()
        H[c(white_player,black_player,"white","draw")] %<>% dec()

        if(nationality[nw]=="USSR" & nationality[nb]=="USSR"){
            H_coll[c(             black_player                    )] %<>% inc()
            H_coll[c(white_player,black_player,"white","collusion")] %<>% dec()
        } else {  # collusion not playing
            H_coll[c(             black_player               )] %<>% inc()
            H_coll[c(white_player,black_player,"white","draw")] %<>% dec()
        }

    } else if (match_result == "1/2-1/2"){ # draw
        results2[nw,nb] <- 1
        results2[nb,nw] <- 1
        H[c(                                  "draw")] %<>% inc()
        H[c(white_player,black_player,"white","draw")] %<>% dec()

        if(nationality[nw]=="USSR" & nationality[nb]=="USSR"){
            H_coll[c(                                  "collusion")] %<>% inc()
            H_coll[c(white_player,black_player,"white","collusion")] %<>% dec()
        } else {  # collusion not playing
            H_coll[c(                                  "draw")] %<>% inc()
            H_coll[c(white_player,black_player,"white","draw")] %<>% dec()
        }
    } else {
        stop("not possible")
    }
}  # i loop closes
```

So, `results` [from the square matrix] and `results2` [from the
3-column dataframe] should match:

```{r checksquareandtab}
stopifnot(all(results[!is.na(results)] == results2[!is.na(results2)]))
```

Now some maximum likelihood estimation.

```{r,hmaxpcalc,cache=TRUE}
H_maxp <- maxp(H)
H_maxp
```

```{r ml1,cache=TRUE}
small <- 0.01

maxlike_free <- maxp(H_coll,startp=c(small,small,small/2,rep(small,22)),give=TRUE)
print("l")
dput(maxlike_free$value)
```

```{r ml2,cache=TRUE}
jj <- maxlike_free$par
maxlike_free <- maxp(H_coll,startp=jj,give=TRUE)
dput(maxlike_free$value)
```

```{r}
H_coll_maxp <- maxlike_free$value
```

```{r dosometests,cache=TRUE}
samep.test(H_coll,c("draw","collusion"))
```

### Package dataset {-}

Following lines create `interzonal.rda`, residing in the `data/`
directory of the package.

```{r,label=saveinterzonaldataset}
interzonal <- H
interzonal_maxp <- H_maxp
collusion <- H_coll
collusion_maxp <- H_coll_maxp
save(interzonal,collusion,interzonal_maxp,collusion_maxp,
file="interzonal.rda")
```

# References

C. C. Moul and J. V. C. Nye 2009. "Did the Soviets collude? A
 statistical analysis of championship chess 1940-1978.  Journal of
 Economic Behaviour and Organization 70(2009) 10-21

### 2022-09-22 inst/jss_4382.tex
fixes #156

\documentclass[12pt]{article}
\usepackage{xcolor}
\begin{document}

\section*{JSS4832: Generalized Plackett-Luce Likelihoods by Robin
K. S. Hankin}

Below, the reviewer's comments are in black, and the replies to the
issues are in \textcolor{blue}{blue}.  I have indicated changes to
the manuscript where appropriate.

Short story: I have accommodated all the comments with rewording and
clarification.

\subsection*{From the editorial team:}

The manuscript currently lacks a thorough software overview allowing
the reader to assess which software packages (for R but also other
software environments) are available providing related functionality
and indicating how the presented functionality provides additional
functionality.

\textcolor{blue}{Response: A more detailed review of existing software
  is provided, although the key innovation (which is Plackett-Luce
  likelihood functions including competitors of identical
  Bradley-Terry strength) is not available elsewhere.}

The manuscript could benefit from some restructuring to better guide
the reader on available methods and implementations, the statistical
methodology covered and extended and the specific computational
implementation made.  It would seem beneficial to entangle these
different aspects as different readers might have differnt foci of
interest and could thus better select on which parts of the manuscript
to focus.  The sections also currently seem to be a mix between
different methods and illustrative applications. Again a thorough
overview on the statistical functionality provided by the package in a
separate section might ease accessibility for readers. Disussing
specific computational aspects of the implementation in a separate
section is certainly also of interest.

\textcolor{blue}{Response}

Note that in R many functions providing ML estimates do not only
return point estimates but either an object where then in an
additional step, e.g., the Hessian can be determined as well as
further information provided using summary / print etc., or directly
have already an object returned which contains more information on the
fit including also the log-likelihood attained, the model fitted, etc.

\textcolor{blue}{Response:}

As an example, methods(class = "PLADMM") in the PlackettLuce package
lists: anova, deviance, fitted, itempar, logLik, predict, print,
summary, vcov. The anova/fitted/predict might be specific to the
regression case but the others might be useful for hyper3 as
well. Similarly for btmodel() in the psychotools package and the eba
package while BradleyTerry2, BTLLasso, PLMIX, and prefmod provide
fewer methods.

The NAMESPACE contains

\verb=exportPattern("^[[:alpha:]]+")=

Please avoid this and check if this can be removed or how it could be
dropped.

It would seem that currently functions are exported which are not
supposed to be directly called by the user, e.g., no argument
checking. In this way they could be kept internal and documentation if
desired could be provided with keyword internal to avoid cluttering
the output of

\verb+help(package = "hyper2")+

\textcolor{blue}{Response:}

Manuscript style comments:

Code should have enough spaces to facilitate reading.  Please include
spaces before and after operators and after commas (unless spaces have
syntactical meaning).

\textcolor{blue}{Response:}

The rule for capitalizing the starting letters of Figure, Section and
Table is as follows: If you are referring to a particular
figure/section/table then capitalize the first letter, otherwise use a
lower-case first letter. For example, something shown in Section 4
vs. there are three sections in this paper.

\textcolor{blue}{Response:}

In all cases, code input/output must fit within the normal text
width of the manuscript.  Thus, code input should have appropriate
line breaks and code output should preferably be generated with a
suitable width (or otherwise edited).

\textcolor{blue}{Response:}

As a reminder, please make sure that: - \verb+\proglang, \pkg+ and
\verb+\code+ have been used for highlighting throughout the paper
(including titles and references), except where explicitly escaped.

\textcolor{blue}{Response:}

References:

o Please make sure that all software packages are \cite{}'d properly.

o All references should be in title style.

o See FAQ for specific reference instructions.

Code:

o Please make sure that the files needed to replicate all
code/examples within the manuscript are included in a standalone
replication script.

\textcolor{blue}{We believe that this referee's insightful and
  constructive comments have led to a much improved submission.}

\end{document}

### 2022-09-22 inst/jss_hyper3.Rnw
typo
fixes #156

@@
   \code{hyper3}, along with associated methods, is motivated and
-  introduced.  Four datasets are analysed, each analysis furnishing
+  introduced.  Three datasets are analysed, each analysis furnishing
   new insight, and each highlighting different capabilities of the
@@
 that the best (preferred; fastest; chosen) entities are identified in
-the same sequence as their rank.  The \code{hyper2} package
+the same sequence as their rank.
+
+Computational support for Bradley-Terry likelihood functions is
+available in a range of languages.  \cite{hunter2004}, for example,
+presents results in \proglang{matlab} [although he works with a
+nonlinear extension to account for ties]; \cite{allison1994} present
+related work for ranking statistics in \proglang{SAS} and
+\cite{maystre2022} has released a \proglang{python} package for
+Luce-type choice datasets.
+
+However, the majority of software is written in the \proglang{R}
+computer language~\citep{rcore2022}, which includes extensive
+functionality for working with such likelihood functions:
+\cite{turner2020} discuss several implementations from a computational
+perspective.  The \pkg{BradleyTerry} package \citep{firth2005}
+considers pairwise comparisons using \proglang{glm} but cannot deal
+with ties or player-specific predictors; the \pkg{BradleyTerry2}
+package \citep{turner2012} implements a flexible user interface and
+wider range of models to be fitted to pairwise comparison datasets,
+specifically simple random effects.  The \pkg{PlackettLuce} package
+\citep{turner2020} generalizes this to likelihood functions of the
+form of Equation~\ref{PL_like} and applies the Poisson transformation
+of~\cite{baker1994} to express the problem as a log-linear model.  The
+\pkg{hyper2} package, in contrast, gives a consistent language
+interface to create and manipulate likelihood functions over the
+simplex ${S}_n=\left\lbrace\left(p_1,\ldots,p_n\right)\left|p_i\geq
+0,\sum p_i=1\right.\right\rbrace$.  A further extension in the package
 generalizes this likelihood function to functions of ${\mathbf

### 2022-09-25 inst/surfing_jsa_rebuttal_minorcomments.tex
start work on #166

\documentclass[12pt]{article}
\usepackage{xcolor}
\begin{document}

\section*{MS596 ``Analysis of competitive surfing tournaments with generalized Bradley-Terry likelihoods": response to minor comments}

Below, we reproduce the two reviewers' minor comments; our replies to
the issues are in \textcolor{blue}{blue}.  We have indicated changes
to the manuscript where appropriate.

Short story: each of the  comments is perfectly reasonable and we have 
changed the manuscript as suggested.

\subsection*{placeholder comment}

Foo should be bar

\textcolor{blue}{generic placeholder response}

\end{document}

### 2022-09-25 inst/jss_hyper3.Rnw

@@
 \code{[a,b,c]} of nonnegative values with unit sum] using standard
-\proglang{R} idiom wrapping \proglang{C++} back end.  The package
-includes an \proglang{S3} class \code{hyper3} for this type of object;
-extraction and replacement methods use \pkg{disordR}
+\proglang{R} idiom wrapping \proglang{C++} back end.
+
+\subsection{Package implementation}
+
+The package includes an \proglang{S3} class \code{hyper3} for this
+type of object; extraction and replacement methods use \pkg{disordR}
 discipline~\citep{hankin2022}.  Package idiom for creating such
@@
 players' strengths, which may be evaluated at specified points in
-probability space.  If, for example, we wish to assess $H_1\colon
+probability space.  A typical use-case would be to assess $H_1\colon
 p_a=0.9,p_b=0.05,p_c=0.05$ and $H_2\colon p_a=0.01,p_b=0.01,p_c=0.98$,
@@
 We may find a maximum likelihood estimate for the players' strengths,
-using generic function \code{maxp()}:
+using generic function \code{maxp()}, dispatching to a specialist
+\code{hyper3} method:
 
@@
 with $H_0$.
+
+\code{Package helper functions}
+
+Arithmetic operations are implemented for \code{hyper3} objects in
+much the same way as for \code{hyper2} objects: independent
+observations may be combined using the overloaded \code{+} operator;
+an example is given below.
+
+The original motivation for \code{hyper3} was the analysis of Formula
+1 motor racing, and the package accordingly includes wrappers for
+\code{ordervec2supp()} such as \code{ordertable2supp3()} and
+\code{attemptstable2supp3()} which facilitate the analysis of commonly
+encountered result formats; an example is given below.
 

### 2022-09-25 inst/surfing_jsa_rebuttal_minorcomments.tex
fixes #167

@@
 
-Below, we reproduce the two reviewers' minor comments; our replies to
+Below, we reproduce the reviewers' minor comments; our replies to
 the issues are in \textcolor{blue}{blue}.  We have indicated changes
@@
 
-\subsection*{placeholder comment}
+Overview I am pleased to see the authors have addressed the issues
+identified in the first review of the paper, or in one or two cases,
+have provided reasonable justification for their approach. The changes
+to the general structure have improved the paper’s overall readability
+and flow, and I commend the inclusion of more explicit derivations and
+references to numerical quantities. As such, I am satisfied with the
+revisions the authors have made and I am happy to recommend the paper
+be accepted, conditional on some minor revisions, noted below:
 
-Foo should be bar
+Contextual changes and clarifications
 
-\textcolor{blue}{generic placeholder response}
+Page 2: The first reference to the ``World Championship Tour'' would
+benefit from including a one-off abbreviation (WCT), which is referred
+to throughout the data set.
+
+\textcolor{blue}{Done; WSL used throughout for ``World Surf League''
+  and minor rewording ensures that the phrase ``World Championship
+  Tour'' does not appear}
+
+Page 4: It is mentioned in the abstract that the analysis is performed on data from the 2019 WSL,
+however a reference in the main text at the top of Section 5 would help clarify this.
+Page 4: The authors mention the surfer Vries – who does not appear to be a wildcard – whose
+maximum likelihood strength is zero. Why is this? My guess is that Vries has never won a competitive
+surfing heat (and therefore there is no information on his true, underlying ability), however a brief
+clarification here would be useful.
+Page 6 / Figure 1: I’m pleased to see the authors replace the old pie chart in Figure 1 with a dot chart
+– this is a much clearer visualization of the results. On Page 6 the authors state: “Figure 1 shows the
+maximum likelihood estimate for the 23 competitors’ strengths, which appears to show a wide range
+from Florence at about 17.5% down to Morais at about 1%”. What exactly do the quantities 17.5%
+and 1% refer to/how are they computed? A quick clarification of these quantities would help explain
+the practical implications of the findings.
+Page 7/Figure 2: The authors suggest there is a loose positive correlation between maximum likelihood
+strength and total points, which is undoubtedly true. However, the x-axis on Figure 2 itself has been
+distorted, which makes the trend appear more linear than it really is. I’d prefer to see the x-axis
+undistorted, unless the authors are explicitly trying to make a point about the trend being log-linear,
+in which case I’d like to see this briefly discussed.
+Page 7: The authors suggest we may reject the hypothesis that the strength of the entity exceeds
+about 0.025. Does this assertation come with a supporting p-value? It is unclear from Figure 4 how
+this conclusion is reached.
+
+Grammatical changes
+Page 3: Change “Further, we use a generalization due to (Luce, 1959)...” to “Further, we use a
+generalization due to Luce (1959)...”.
+Page 7: Change “..., but in addition we may reject the the proposal that...” to “..., but in addition we
+may reject the proposal that...”.
 

### 2022-09-25 inst/surfing_jsa.tex
fixes #167

@@
 one-way ANOVA techniques and report comparisons between the top 10 and
-bottom 10 surfers in the 2013 World Championship Tour and conclude
+bottom 10 surfers in the 2013 WSL tour and conclude
 that the top-ranked athletes are more consistent than the lower-ranked
@@
 terminology of \citet{johnson2020}.  We now use a technique due to
-\citet{hankin2010,hankin2017} and introduce fictional (reified)
+\citet{hankin2010,hankin2017_nomarkup} and introduce fictional (reified)
 entities whose nonzero Bradley-Terry strength helps certain
@@
 Florence at about 17.5\% down to Morais at about 1\%.  The {\tt
-  hyper2} software~\citep{hankin2017} may be used to formally assess
+  hyper2} software~\citep{hankin2017_nomarkup} may be used to formally assess
 one plausible null: that the competitors all have equal skill, that is
@@
 Lowers Pro were Brazilian; it rose in public consciousness following
-the victories of Medina and de Souza in the World Surf League
+the victories of Medina and de Souza in the WSL
 championships in 2014 and 2015 respectively.  The storm shows no signs
@@
 
-We analysed results from the 2019 World Surf League using reified
+We analysed results from the 2019 WSL using reified
 Bradley-Terry likelihoods to assess a number of hypotheses of interest
@@
 \end{tabular}
-\caption{Conventional league table for the top five competitors in WCT
+\caption{Conventional league table for the top five competitors in WSL
   2019 tour. \label{resultstable} Thus at WCT01 (Coolangatta, QLD) we see

### 2022-09-25 inst/surfing_jsa_rebuttal_minorcomments.tex
fixes #168
fixes #169

@@
 
-Page 4: It is mentioned in the abstract that the analysis is performed on data from the 2019 WSL,
-however a reference in the main text at the top of Section 5 would help clarify this.
-Page 4: The authors mention the surfer Vries – who does not appear to be a wildcard – whose
-maximum likelihood strength is zero. Why is this? My guess is that Vries has never won a competitive
-surfing heat (and therefore there is no information on his true, underlying ability), however a brief
-clarification here would be useful.
+Page 4: It is mentioned in the abstract that the analysis is performed
+on data from the 2019 WSL, however a reference in the main text at the
+top of Section 5 would help clarify this.
+
+\textcolor{blue}{Done}
+
+Page 4: The authors mention the surfer Vries---who does not appear to
+be a wildcard---whose maximum likelihood strength is zero. Why is
+this? My guess is that Vries has never won a competitive surfing heat
+(and therefore there is no information on his true, underlying
+ability), however a brief clarification here woud be useful.
+
+\textcolor{blue}{The referee is quite right: Vries competed only twice
+  in 2019, coming last (second and third) in each of his two sessions.
+  Brief explanation added to he mauscript.}
+
 Page 6 / Figure 1: I’m pleased to see the authors replace the old pie chart in Figure 1 with a dot chart

### 2022-09-25 inst/surfing_jsa.tex
fixes #168
fixes #169

@@
 Table~\ref{rawresults} shows the first few and last few entries in the
-dataset we use.  The first line shows that, at one of the heats in
-WCT01 (Coolangatta, QLD), Colapinto came first, Bailey second, and
-Wright came third: we write
+dataset we use, drawn from 2019 WSL.  The first line shows that, at
+one of the heats in WCT01 (Coolangatta, QLD), Colapinto came first,
+Bailey second, and Wright came third: we write
 $\mbox{Colapinto}\succ\mbox{Bailey}\succ\mbox{Wright}$.  The last line
@@
 dataset into a Plackett-Luce likelihood function but first have to
-remove the wildcards and also player Vries whose maximum likelihood
-strength is zero.  For the first line we would have Plackett-Luce
-likelihood function as
+remove the wildcards whose strength is typically very low.  We also
+remove De Vries who, while not a wildcard, competed only twice in 2019
+and came last each time, thus having zero maximum likelihood strength.
+For the first line we would have Plackett-Luce likelihood function as
 

### 2022-09-25 inst/surfing_jsa_rebuttal_minorcomments.tex
fixes #170

@@
 
-Page 6 / Figure 1: I’m pleased to see the authors replace the old pie chart in Figure 1 with a dot chart
-– this is a much clearer visualization of the results. On Page 6 the authors state: “Figure 1 shows the
-maximum likelihood estimate for the 23 competitors’ strengths, which appears to show a wide range
-from Florence at about 17.5% down to Morais at about 1%”. What exactly do the quantities 17.5%
-and 1% refer to/how are they computed? A quick clarification of these quantities would help explain
-the practical implications of the findings.
+Page 6 / Figure 1: I'm pleased to see the authors replace the old pie
+chart in Figure 1 with a dot chart---this is a much clearer
+visualization of the results.  On Page 6 the authors state: ``Figure 1
+shows the maximum likelihood estimate for the 23 competitors'
+strengths, which appears to show a wide range from Florence at about
+$17.5\%$ down to Morais at about $1\%$". What exactly do the
+quantities $17.5\%$ and $1\%$ refer to/how are they computed? A quick
+clarification of these quantities would help explain the practical
+implications of the findings.
+
+\textcolor{blue}{The figures are, as stated, the maximum likelihood
+  estimates of the competitors' strengths.  We think that our use of
+  percentages (which add to 100) instead of strengths (which add to 1)
+  was the cause of the confusion and we have amended the manuscript to
+  use strengths, not percentages.}
+
 Page 7/Figure 2: The authors suggest there is a loose positive correlation between maximum likelihood

### 2022-09-25 inst/surfing_jsa.tex
fixes #170

@@
 23 competitors' strengths, which appears to show a wide range from
-Florence at about 17.5\% down to Morais at about 1\%.  The {\tt
+Florence at about $0.175$ down to Morais at about $0.01$.  The {\tt
   hyper2} software~\citep{hankin2017_nomarkup} may be used to formally assess

### 2022-09-25 inst/surfing.Rmd
work on #171

@@
 
+```{r, directplotofstengths}
+pdf(file="directplotofstrengths.pdf")
+par(xpd=NA)
+plot(jjO,jjL,pch=16,log='y',xlab='points',ylab='maximum likelihood strength',cex=1.5)
+fit <- lm(log10(jjL)~(jjO))
+for(i in seq_along(jjO)){
+text(jjO[i],jjL[i],names(jjL)[i],col='gray',pos=4,cex=0.8)
+}
+par(xpd=FALSE)
+abline(fit)
+dev.off()
+```
+
 Now compare the finishing order based on points vs the finishing order

### 2022-09-25 inst/surfing_jsa_rebuttal_minorcomments.tex
fixes #171

@@
 
-Page 7/Figure 2: The authors suggest there is a loose positive correlation between maximum likelihood
-strength and total points, which is undoubtedly true. However, the x-axis on Figure 2 itself has been
-distorted, which makes the trend appear more linear than it really is. I’d prefer to see the x-axis
-undistorted, unless the authors are explicitly trying to make a point about the trend being log-linear,
-in which case I’d like to see this briefly discussed.
+Page 7/Figure 2: The authors suggest there is a loose positive
+correlation between maximum likelihood strength and total points,
+which is undoubtedly true. However, the x-axis on Figure 2 itself has
+been distorted, which makes the trend appear more linear than it
+really is. I'd prefer to see the x-axis undistorted, unless the
+authors are explicitly trying to make a point about the trend being
+log-linear, in which case I'd like to see this briefly discussed.
+
+\textcolor{blue}{Horizontal axis now linear; no other change except
+  for tiny movement of labels so they do not obscure one another.  We
+  are mortified to see that the linearity is more apparent without the
+  log-transform.}
+
 Page 7: The authors suggest we may reject the hypothesis that the strength of the entity exceeds

### 2022-09-25 inst/surfing_jsa.tex
fixes #171

@@
 
-\begin{figure}[h]
-  \includegraphics[width=4in]{directlogplotofstrengths_postinkscape.pdf}  % to create this file, open directlogplotofstrengths_postinkscape.svg in inkscape and export to PDF.
+%\begin{figure}[h]
+%  \includegraphics[width=4in]{directlogplotofstrengths_postinkscape.pdf}  % to create this file, open directlogplotofstrengths_postinkscape.svg in inkscape and export to PDF.
 File directlogplotofstrengths_postinkscape.svg is under version control in the hyper2 github repo.
@@
 File directelogplotofstrengths.pdf is produced by surfing.Rmd.
+%\caption{Scatterplot of points scored  \label{compare_likelihood_points}
+%against maximum likelihood estimates}
+%\end{figure}
+
+%\clearpage
+%\newpage
+
+\begin{figure}[h]
+  \includegraphics[width=4in]{directplotofstrengths_postinkscape.pdf}  % to create this file, open directplotofstrengths_postinkscape.svg in inkscape and export to PDF.
+File directplotofstrengths_postinkscape.svg is under version control in the hyper2 github repo.
+It is a slight modification of directplotofstrengths.pdf, with slight optical glitches removed.
+File directplotofstrengths.pdf is produced by surfing.Rmd.
 \caption{Scatterplot of points scored  \label{compare_likelihood_points}

### 2022-09-25 inst/surfing.Rmd
fixes #171

--- 2022-04-25 inst/surfing.Rmd
@@
 
+```{r, directplotofstengths}
+pdf(file="directplotofstrengths.pdf")
+par(xpd=NA)
+plot(jjO,jjL,pch=16,log='y',xlab='points',ylab='maximum likelihood strength',cex=1.5)
+fit <- lm(log10(jjL)~(jjO))
+summary(fit)
+for(i in seq_along(jjO)){
+text(jjO[i],jjL[i],names(jjL)[i],col='gray',pos=4,cex=0.8)
+}
+par(xpd=FALSE)
+abline(fit)
+dev.off()
+```
+
 Now compare the finishing order based on points vs the finishing order

### 2022-09-25 inst/surfing_jsa.tex
fixes #171

@@
 likelihoods.  Figure~\ref{compare_likelihood_points} shows a loose
-correlation ($R^2=0.64$; $p=3\times 10^{-6}$) and
+correlation ($R^2=0.71$; $p=4\times 10^{-7}$) and
 Figure~\ref{compare_likelihood_points_rankings} shows how the rankings

### 2022-09-25 inst/surfing_jsa.tex
fixes #172

@@
 reject the the proposal that the strength of this entity exceeds about
-0.025 (Figure~\ref{brazilian_wavetype_support}).
+0.015 (Figure~\ref{brazilian_wavetype_support}): at this value of the
+strength or greater, we may gain two or more units of support by
+moving to the evaluate~\citep{edwards1992}.
 
@@
   particular strength for Brazilian wavetype, and the height of the
-  curve gives the support conditional on the x-axis}
+  curve gives the support conditional on the x-axis.  At a strength of
+  0.015 we see the support is $-2$ and we could thus be reasonably
+  confident that the strength does not exceed this value.}
 \end{figure}

### 2022-09-25 inst/surfing_jsa_rebuttal_minorcomments.tex
fixes #172
fixes #173

@@
 
-Page 7: The authors suggest we may reject the hypothesis that the strength of the entity exceeds
-about 0.025. Does this assertation come with a supporting p-value? It is unclear from Figure 4 how
-this conclusion is reached.
+Page 7: The authors suggest we may reject the hypothesis that the
+strength of the entity exceeds about 0.025. Does this assertation come
+with a supporting p-value? It is unclear from Figure 4 how this
+conclusion is reached.
+
+\textcolor{blue}{This was a (devil-inspired) typo.  The value should
+  read 0.015.  The reasoning was that at 0.015 the support was -2.  We
+  have ddded a sentence of explanation on p7 and also the figure
+  caption.}
 
 Grammatical changes
-Page 3: Change “Further, we use a generalization due to (Luce, 1959)...” to “Further, we use a
-generalization due to Luce (1959)...”.
-Page 7: Change “..., but in addition we may reject the the proposal that...” to “..., but in addition we
-may reject the proposal that...”.
+
+Page 3: Change ``Further, we use a generalization due to (Luce,
+1959)..." to ``Further, we use a generalization due to Luce
+(1959)...".
+
+Page 7: Change ``..., but in addition we may reject the
+the proposal that..." to ``..., but in addition we may reject the
+proposal that...".
+
+\textcolor{blue}{Done}
 

### 2022-09-25 inst/surfing_jsa.tex
fixes #173

@@
 $\sum p_i=1$.  Further, we use a generalization due to
-\citep{luce1959}, in which the probability of competitor~$i$ winning
+\citet{luce1959}, in which the probability of competitor~$i$ winning
 in a field of $\left\lbrace 1,\ldots, n\right\rbrace$ is
@@
 Brazilian wavetype very small (about 0.002), but in addition we may
-reject the the proposal that the strength of this entity exceeds about
+reject the proposal that the strength of this entity exceeds about
 0.015 (Figure~\ref{brazilian_wavetype_support}): at this value of the

### 2022-09-25 inst/surfing_jsa.tex
fixes #175

@@
 \usepackage{diagbox}
+\usepackage{multirow}
 \usepackage[pdftex]{hyperref}
@@
 
+% use   \begin{tabular}{@{}c@{}}\small{WCT} \\ 01\end{tabular} for WCT\\01
 \begin{table}[h]
 \begin{tabular}{llllllllllllll}
-competitor  & nationality&WCT01  &WCT02  &WCT03  & \ldots  &WCT10&WCT11&points\\ \hline
+competitor  & nationality & \multirow{ 5}{*}{1} &  &  & WCT & & &points\\
+            &             &  01   & 02& 03& \ldots& 10 & 11 & \\ \hline
  Ferreira   &   BRA       &   1   &   5   &   17  & \ldots  &   1   &   1   &   59740\\

### 2022-09-26 inst/surfing_jsa.tex
fixes #177

@@
 File directplotofstrengths.pdf is produced by surfing.Rmd.
-\caption{Scatterplot of points scored  \label{compare_likelihood_points}
-against maximum likelihood estimates}
+\caption{Scatterplot of points scored (horizontal
+  axis) \label{compare_likelihood_points} against maximum likelihood
+  estimates (vertical axis).  One would expect a positive correlation
+  as high Bradley-Terry strength would lead to a high probability of
+  earning more points; with these axes the correlation is moderate at
+  about $R^2=0.71$.  Florence, at $(37700,0.175)$, is something of an
+  anomaly. He did not compete at five venues due to injury, this
+  affecting his points total but not his likelihood-based
+  ranking.}
 \end{figure}

### 2022-09-26 inst/surfing.Rmd
fixes #178

@@
 sort(names(rO))==sort(names(rO))
-ordertransplot(rO,rL,xlab="offical rank",ylab="likelihood rank",main="WCT 2019")
+ordertransplot(rO,rL,xlab="offical rank",ylab="likelihood rank",main="")
 pdf(file="ordertransplotplot.pdf")
-ordertransplot(rO,rL,xlab="offical rank",ylab="likelihood rank",main="WCT 2019")
+ordertransplot(rO,rL,xlab="offical rank",ylab="likelihood rank",main="")
 dev.off()

### 2022-09-26 inst/surfing_jsa_rebuttal_minorcomments.tex
fixes #175
issue #174
issue #177

@@
 
+Page 10: References with URL's should be split over multiple lines to avoid being cut-off the page.
+
+\textcolor{blue}{Done}
+
+Page 11: Table 1 could be formatted to fit within the page margins.
+
+\textcolor{blue}{Done}
+
+Page 17: Capitalisation of Figure 2 axis titles.
+
+Page 17: Figure 2 caption is a little brief and could better explain
+what is observed in the plot (e.g. the positive correlation between
+variables). Especially as the authors have distorted the x-axis and
+fitted a linear trend through the data. What is the reason for this?
+As mentioned in a previous comment, I’d prefer to see the x-axis in
+its raw form, unless there is a reason it has been distorted.
+
+\textcolor{blue}{More detail added in caption}
+
+Page 18: Figure 3 title – this doesn’t provide the reader with any
+additional information. It does seem a little odd that Figure 2 does
+not have a title, Figure 3 does, and Figure 4 does not. For
+consistency, the authors could simply remove the title from Figure 33,
+as it does not really add any value.
+
+Page 19: Figure 4 Capitalisation of y-axis title.  General comment:
+should the Table and Figure captions have a full-stop at the end of
+the last sentence?
+
+General comment
+As part of a potential future publication in the field, I’d be curious to see the methodology in Section
+7 applied to World Surf League data from events that occurred prior to the announcement that surfing
+would be in the Olympics. Comparing these results with those in the paper that use data from after
+this announcement (perhaps supplemented with additional data and results from post-2019), would
+help in our understanding whether there truly has been an observable change in non-competitive
+behaviour between compatriots. Any such analysis is clearly well beyond the realm of minor revisions
+and is not something I expect of the authors prior to final submission; however, I thought it could be
+an interesting next step for the application of the methodology outlined in the paper!
+
 \end{document}

### 2022-09-26 inst/surfing.Rmd
fixes #179

@@
 pdf(file="plotwidep.pdf")
-plot(c(0,p_wide),c(l0,l_wide)-max(l),type='b',xlab="Bradley-Terry strength of the Brazilian beachtype reified entity",ylab="support")
+plot(c(0,p_wide),c(l0,l_wide)-max(l),type='b',xlab="Bradley-Terry strength of the Brazilian beachtype reified entity",ylab="Support")
 abline(h= -2,lty=2)

### 2022-09-26 inst/surfing_jsa.tex
issue #179

@@
   affecting his points total but not his likelihood-based
-  ranking.}
+  ranking}
 \end{figure}
@@
   0.015 we see the support is $-2$ and we could thus be reasonably
-  confident that the strength does not exceed this value.}
+  confident that the strength does not exceed this value}
 \end{figure}

### 2022-09-26 inst/surfing_jsa_rebuttal_minorcomments.tex
issue #176
issue #178
fixes #179
fixes #180

@@
 
+\textcolor{blue}{Done}
+
 Page 17: Figure 2 caption is a little brief and could better explain
@@
 
+\textcolor{blue}{Done (title removed from figure 3 as suggested)}
+
 Page 19: Figure 4 Capitalisation of y-axis title.  General comment:
@@
 
-General comment
-As part of a potential future publication in the field, I’d be curious to see the methodology in Section
-7 applied to World Surf League data from events that occurred prior to the announcement that surfing
-would be in the Olympics. Comparing these results with those in the paper that use data from after
-this announcement (perhaps supplemented with additional data and results from post-2019), would
-help in our understanding whether there truly has been an observable change in non-competitive
-behaviour between compatriots. Any such analysis is clearly well beyond the realm of minor revisions
-and is not something I expect of the authors prior to final submission; however, I thought it could be
-an interesting next step for the application of the methodology outlined in the paper!
+\textcolor{blue}{Done.  Looking at recent JSA articles the captions do
+  not have full-stops and we just copied that style}
+
+General comment: As part of a potential future publication in the
+field, I'd be curious to see the methodology in Section 7 applied to
+World Surf League data from events that occurred prior to the
+announcement that surfing would be in the Olympics.  Comparing these
+results with those in the paper that use data from after this
+announcement (perhaps supplemented with additional data and results
+from post-2019), would help in our understanding whether there truly
+has been an observable change in non-competitive behaviour between
+compatriots.  Any such analysis is clearly well beyond the realm of
+minor revisions and is not something I expect of the authors prior to
+final submission; however, I thought it could be an interesting next
+step for the application of the methodology outlined in the paper!
+
+\textcolor{blue}{This suggestion is an excellent one (which had not
+  occurred to us); as remarked, it is not in scope for the current
+  paper.  We do plan to work on this idea in the future and have added
+  a brief discussion to the ``Future work'' section
+}
 

### 2022-09-26 inst/surfing_jsa.tex
fixes #180
issue #167

@@
 one-way ANOVA techniques and report comparisons between the top 10 and
-bottom 10 surfers in the 2013 WSL tour and conclude
+bottom 10 surfers in the 2013 World Championship Tour (WCT) and conclude
 that the top-ranked athletes are more consistent than the lower-ranked
@@
 into our likelihood function, and possibly to investigate priority
-infringements between compatriots.
+infringements between compatriots.  One suggestion, made by an
+anonymous JSA referee, was to apply the methodology used in Section 7
+to events that occurred prior to the announcement that surfing would
+be in the Olympics.  This would allow the quantification of any {\em
+  change} in non-competitive behaviour between compatriots.
 

### 2022-09-26 inst/surfing_jsa_rebuttal_minorcomments.tex
issue #167
minor mucking about for issue #166

@@
 
-Overview I am pleased to see the authors have addressed the issues
-identified in the first review of the paper, or in one or two cases,
-have provided reasonable justification for their approach. The changes
-to the general structure have improved the paper’s overall readability
-and flow, and I commend the inclusion of more explicit derivations and
-references to numerical quantities. As such, I am satisfied with the
-revisions the authors have made and I am happy to recommend the paper
-be accepted, conditional on some minor revisions, noted below:
-
-Contextual changes and clarifications
+\subsubsection*{Referee comments}
 
@@
 
-\textcolor{blue}{Done; WSL used throughout for ``World Surf League''
-  and minor rewording ensures that the phrase ``World Championship
-  Tour'' does not appear}
+\textcolor{blue}{Done}\\
 
@@
 
-\textcolor{blue}{Done}
+\textcolor{blue}{Done}\\
 
@@
   in 2019, coming last (second and third) in each of his two sessions.
-  Brief explanation added to he mauscript.}
+  Brief explanation added to he mauscript.}\\
 
@@
   was the cause of the confusion and we have amended the manuscript to
-  use strengths, not percentages.}
+  use strengths, not percentages.}\\
 
@@
   are mortified to see that the linearity is more apparent without the
-  log-transform.}
+  log-transform.}\\
 
@@
   have ddded a sentence of explanation on p7 and also the figure
-  caption.}
+  caption.}\\
 
@@
 
-\textcolor{blue}{More detail added in caption}
+\textcolor{blue}{More detail added in caption}\\
 
@@
 
-\textcolor{blue}{Done (title removed from figure 3 as suggested)}
+\textcolor{blue}{Done (title removed from figure 3 as suggested)}\\
 
@@
 \textcolor{blue}{Done.  Looking at recent JSA articles the captions do
-  not have full-stops and we just copied that style}
+  not have full-stops and we just copied that style}\\
 
@@
   paper.  We do plan to work on this idea in the future and have added
-  a brief discussion to the ``Future work'' section
+  a brief discussion to the ``Future work'' section/.
 }

### 2022-09-26 inst/surfing_jsa.tex
very minor typesetting pursuant to issue #166

@@
 Table~\ref{rawresults} shows the first few and last few entries in the
-dataset we use, drawn from 2019 WSL.  The first line shows that, at
+dataset we use, drawn from 2019 WCT.  The first line shows that, at
 one of the heats in WCT01 (Coolangatta, QLD), Colapinto came first,
@@
 and came last each time, thus having zero maximum likelihood strength.
-For the first line we would have Plackett-Luce likelihood function as
+For the first line we would have a Plackett-Luce likelihood function of
 
@@
 is to use {\tt equalp.test()} which reports its findings in a
-standardised format.  This gives us a support difference of about
+standardised format.  This gives a support difference of about
 0.05, well short of Edwards's 2 units of support criterion.
-Alternatively, we could observe that the
-$-2\log\left(\frac{\Lambda}{\Lambda_0}\right)\simeq 0.1$ is not in the
+Alternatively, we could observe that~$\mathcal{S}=
+-2\log\left(\frac{\Lambda}{\Lambda_0}\right)\simeq 0.1$ is not in the
 tail region of its asymptotic $\chi^2_1$ distribution.  We may thus
-assert that there is no strong evidence for Ferreira's actually having
+assert that there is no strong evidence for Ferreira actually having
 a competitive advantage over Medina.  We may similarly wonder whether
@@
 
-Medina's performance was startlingly good at Freshwater Pro ({\tt
-  wct08}).  Indeed, Medina came first at only two venues (the other
+Medina's performance was startlingly good at Freshwater Pro (WCT08).
+Indeed, Medina came first at only two venues (the other
 was at Jeffries Bay, South Africa) and it is natural to enquire
@@
 
-We analysed results from the 2019 WSL using reified
+We analysed results from the 2019 WCT using reified
 Bradley-Terry likelihoods to assess a number of hypotheses of interest
@@
 support the supposed Brazilian preference for point and beach break
-beaches.  We also suggest that surfing's being included in the 2020
+wave types.  We also suggest that surfing's being included in the 2020
 Summer Olympics might result in modified strategies to bolster
 intranational ranking at the expense of WCT performance; we quantify
-the strength of such a phenomenon as the strength of a reified
+this phenomenon by introducing a reified
 Bradley-Terry entity: a ``compatriot monster'' with an estimated
@@
 \end{tabular}
-\caption{Conventional league table for the top five competitors in WSL
+\caption{Conventional league table for the top five competitors in the WCT
   2019 tour. \label{resultstable} Thus at WCT01 (Coolangatta, QLD) we see

### 2022-09-27 inst/jss_4382.tex
fixes #158

@@
 
-\textcolor{blue}{Response:}
-
 As an example, methods(class = "PLADMM") in the PlackettLuce package
@@
 fewer methods.
+
+\textcolor{blue}{Response: Function {\tt maxp()} can return many
+  additional objects in addition to the evaluate; standard idiom is to
+  pass argument {\tt give = TRUE}.  In particular, a list with
+  elements
+  \begin{itemize}
+  \item {\tt \$hessian} the Hessian matrix: passing argument {\tt give
+    = TRUE, hessian = TRUE} returns the Hessian matrix.  However, this
+    is restricted to {\tt hyper2} objects and I would add that {\tt
+      hyper3} objects are much, much, harder to differentiate than
+    {\tt hyper2} objects (let alone second derivatives).
+    \item {\tt \$likes} likelihoods
+    \item {\tt \$par} the evaluate
+  \end{itemize}
+  The package includes summary methods and print method for summary
+  objects.\\ Some of the suggestions are good for regression-type
+  analysis, but do not really carry over to the current
+  likelihood-based approach.  The {\tt hyper2} package is predicated
+  on straightforward idiom to create and manipulate likelihood
+  functions; statistical tests are (to me) just one application of
+  likelihood functions.}
 

### 2022-09-27 inst/jss_4382.tex
fixes #159

@@
 
-\textcolor{blue}{Response:}
+\textcolor{blue}{Response: the pattern has been removed as suggested,
+  all functions exported are exported explicitly.  However, it is not
+  clear who ``the user" is.  Currently, the users are restricted to
+  myself and a bunch of graduate students, who are mostly working on
+  applying hyper2 constructions to their own pet datasets; this group
+  of users needs to see similar work that they can modify for their
+  own projects.  But the output of \verb+help(package = "hyper2")+ is
+  now restricted only to functions that might be of interest to users.
+  }
 

### 2022-09-27 inst/jss_hyper3.Rnw
issue #161 issue #162

@@
 LL <- hyper3()
-LL[c(a=1)] <- 1
-LL[c(a=3, b=2, c=1)] <- -1
+LL[c(a = 1)] <- 1
+LL[c(a = 3, b = 2, c = 1)] <- -1
 LL
@@
 a <- 1/2 # null
-(S_delta <- log( a*(1-a)/(1+a))- log(3-2*sqrt(2)))
+(S_delta <- log(a * (1 - a)/(1 + a)) - log(3 - 2 * sqrt(2)))
 @
@@
 <<>>=
-pchisq(-2*S_delta,df=1,lower.tail=FALSE)
+pchisq(-2 * S_delta, df=1, lower.tail=FALSE)
 @
@@
 <<figaba>>=
-a <- seq(from=0,by=0.005,to=1)
-S <- function(a){log(a*(1-a)/((1+a)*(3-2*sqrt(2))))}
-plot(a,S(a),type='b')
-abline(h=c(0,-2))
-abline(v=c(0.02438102,0.9524271),col='red')
-abline(v=sqrt(2)-1)
+a <- seq(from=0, by=0.005, to=1)
+S <- function(a){log(a * (1 - a) / ((1 + a) * (3 - 2 * sqrt(2))))}
+plot(a, S(a), type='b')
+abline(h=c(0, -2))
+abline(v=c(0.02438102, 0.9524271), col='red')
+abline(v=sqrt(2) - 1)
 @
@@
 
-<<label=converttosupp3,cache=TRUE>>=
+<<label=converttosupp3, cache=TRUE>>=
 javelin_vector <- attemptstable2supp3(javelin_table, decreasing=TRUE, give.supp=FALSE)
@@
   \begin{center}
-<<testthejav,cache=TRUE>>=
+<<testthejav, cache=TRUE>>=
 (mj <- maxp(javelin))
@@
 abline(v=0, lty=2)
-abline(v=log(0.32062833/0.11402735)) # these from mp
+abline(v=log(0.32062833 / 0.11402735)) # these from mp
 @
@@
 <<showconstructortable>>=
-constructor_2021_table[,1:9]
+constructor_2021_table[, 1:9]
 @
@@
 <<definecombinedlikelihoodfunction,cache=TRUE>>=
-H <- psubs(constructor_2020, "Merc", "Merc2020") + psubs(constructor_2021, "Merc", "Merc2021")
+H <- (
+      psubs(constructor_2020, "Merc", "Merc2020") +
+      psubs(constructor_2021, "Merc", "Merc2021")
+     )
 @

### 2022-09-27 inst/jss_hyper3.Rnw
issue #162

@@
 <<label=converttosupp3, cache=TRUE>>=
-javelin_vector <- attemptstable2supp3(javelin_table, decreasing=TRUE, give.supp=FALSE)
+javelin_vector <- attemptstable2supp3(javelin_table,
+       decreasing=TRUE, give.supp=FALSE)
+options(width=60)
 javelin_vector
@@
 const2021 <- ordertable2supp3(constructor_2021_table)
+options(digits=4)
 maxp(const2020, n=1)
@@
 <<usesamep,cache=TRUE>>=
+options(digits=4)
 samep.test(H, c("Merc2020", "Merc2021"))

### 2022-09-27 inst/jss_hyper3.Rnw
fixes #164
fixes #161

@@
 for a dataset including nonfinishers will be given below in
-section~\ref{f1}.
+Section~\ref{f1}.
 
@@
 The ideas presented above may be extended to more than two types of
-competitors.  Consider the following table, drawn from the men's
+competitors.  Consider the following Table, drawn from the men's
 javelin, 2020 Olympics:
@@
 throw of 86.67m, and so on (\code{NA} entries correspond to
-no-throws.)  The attempts table may be converted to a \code{hyper3}
+no-throws.)  The attempts Table may be converted to a \code{hyper3}
 object, again using function \code{attemptstable2supp3()} but this
@@
 use generalized Plackett-Luce to assess the constructors' performance.
-The following table, included in the \code{hyper2} package as a
+The following Table, included in the \code{hyper2} package as a
 dataset, shows rankings for the first 9 venues of the 2021 season:
@@
 
-\bibliography{chess}
+\bibliography{jss_hyper3}
 \end{document}

### 2022-09-27 inst/jss_4382.tex
issue #159 issue #160 issue #161

--- 2022-09-27 inst/jss_4382.tex
@@
 
-\textcolor{blue}{Response:}
+\textcolor{blue}{Response: the pattern has been removed as suggested,
+  all functions exported are exported explicitly.  However, it is not
+  clear who ``the user" is.  Apart from myself, it seems that most
+  users of the package are my graduate students who are mostly working
+  on applying hyper2 constructions to their own pet datasets; this
+  group of users needs to see similar work that they can modify for
+  their own projects.  The package was reviewed by Turner in 2020 but
+  it is not clear whether she was a user in the sense of the
+  comment. In any event, the output of {\tt help(package = "hyper2")}
+  is now restricted only to functions that might be of interest to
+  users.  }
 
@@
 
-\textcolor{blue}{Response:}
+\textcolor{blue}{Response: spaces added as appropriate}
 
@@
 
-\textcolor{blue}{Response:}
+\textcolor{blue}{Response: done}
 

### 2022-09-28 inst/jss_hyper3.Rnw
fixes #160
further work on issue #157

--- 2022-09-27 inst/jss_hyper3.Rnw
@@
 
+\subsection{Computational methodology for generalized Plackett-Luce likelihood functions}
+
 Existing \code{hyper2} formalism as described by \citet{hankin2017}
-cannot represent this likelihood function, because
+cannot represent Equation~\ref{plackettluce}, because
 Equation~\ref{hyper2likelihood} uses sets as the indexing elements,

### 2022-09-28 inst/jss_4382.tex
further work on issue #157
issue #153

@@
 
-\textcolor{blue}{Response}
+\textcolor{blue}{Response: new sections 1.1 and 1.2 added discussing
+  the new functionality in the package}
 
@@
   additional objects in addition to the evaluate; standard idiom is to
-  pass argument {\tt give = TRUE}.  In particular, a list with
-  elements
+  pass argument {\tt give = TRUE}.  In particular, a it returns a list with
+  elements including
   \begin{itemize}
@@
   all functions exported are exported explicitly.  However, it is not
-  clear who ``the user" is.  Apart from myself, it seems that most
-  users of the package are my graduate students who are mostly working
-  on applying hyper2 constructions to their own pet datasets; this
-  group of users needs to see similar work that they can modify for
-  their own projects.  The package was reviewed by Turner in 2020 but
-  it is not clear whether she was a user in the sense of the
-  comment. In any event, the output of {\tt help(package = "hyper2")}
-  is now restricted only to functions that might be of interest to
-  users.  }
+  clear who ``the user" is.  The package is still under active
+  development.  Apart from myself, it seems that most users of the
+  package are my graduate students who are mostly working on applying
+  hyper2 constructions to their own pet datasets; this group of users
+  needs to see similar work that they can modify for their own
+  projects.  The package was reviewed by Turner in 2020 but it is not
+  clear whether she was a user in the sense of the comment. In any
+  event, the output of {\tt help(package = "hyper2")} is now
+  restricted only to functions that might be of interest to users.  }
 
@@
 
-o Please make sure that all software packages are \cite{}'d properly.
+o Please make sure that all software packages are cited properly.
 
@@
 o See FAQ for specific reference instructions.
+
+\textcolor{blue}{Response: done}
 
@@
 
-\textcolor{blue}{We believe that this referee's insightful and
+\textcolor{blue}{Summary: this referee's insightful and
   constructive comments have led to a much improved submission.}

### 2022-09-28 inst/jss_hyper3.Rnw
issue #153

@@
 <<>>=
-loglik(c(a=0.01, b=0.01, c=0.98),LL)
-loglik(c(a=0.90, b=0.05, c=0.05),LL)
+loglik(c(a = 0.01, b = 0.01, c = 0.98), LL)
+loglik(c(a = 0.90, b = 0.05, c = 0.05), LL)
 @
@@
 <<>>=
-pchisq(-2 * S_delta, df=1, lower.tail=FALSE)
+pchisq(-2 * S_delta, df = 1, lower.tail = FALSE)
 @
@@
 <<figaba>>=
-a <- seq(from=0, by=0.005, to=1)
+a <- seq(from = 0, by = 0.005, to = 1)
 S <- function(a){log(a * (1 - a) / ((1 + a) * (3 - 2 * sqrt(2))))}
-plot(a, S(a), type='b')
-abline(h=c(0, -2))
-abline(v=c(0.02438102, 0.9524271), col='red')
-abline(v=sqrt(2) - 1)
+plot(a, S(a), type = 'b')
+abline(h = c(0, -2))
+abline(v = c(0.02438102, 0.9524271), col = 'red')
+abline(v = sqrt(2) - 1)
 @
-\caption{A\label{ABA_likelihood} support function for $p_a$ with observation $a\succ b\succ a$}
+\caption{A\label{ABA_likelihood} support function for $p_a$ with observation $a\succ b\succ a$.}
 \end{center}
@@
 p <- function(f, ...){points(a, f(a) / max(f(a)), ...)}
-plot(0:1, 0:1, xlab="p(a)", ylab="Likelihood", type="n")
-p(f_aab, type="l", col="black")
-p(f_aba, type="l", col="red")
-p(f_baa, type="l", col="blue")
-legend(x=0.85, y=0.7, col=c("black", "red", "blue"), lty=1, pch=NA, legend=c("AAB", "ABA", "BAA"))
-abline(h=exp(-2), lty=2)
+plot(0:1, 0:1, xlab = "p(a)", ylab = "Likelihood", type = "n")
+p(f_aab, type = "l", col = "black")
+p(f_aba, type = "l", col = "red")
+p(f_baa, type = "l", col = "blue")
+legend(x=0.85, y=0.7, col = c("black", "red", "blue"), lty=1, pch=NA,
+       legend=c("AAB", "ABA", "BAA"))
+abline(h = exp(-2), lty = 2)
 @
 \caption{Likelihood \label{ABAetc_likelihood} functions for
-observations $a\succ a\succ b$, $a\succ b\succ a$, $b\succ a\succ a$}
+observations $a\succ a\succ b$, $a\succ b\succ a$, $b\succ a\succ a$.}
 \end{center}
@@
 <<plotwilcoxlike,echo=FALSE>>=
-a <- seq(from=0.02, to=0.8, len=40)
+a <- seq(from = 0.02, to = 0.8, len = 40)
 L <- sapply(a, function(p){loglik(p, Hxy)})
-plot(a, L - max(L),type='b')
-abline(h=c(0, -2))
-abline(v=c(0.24))
+plot(a, L - max(L), type = 'b')
+abline(h = c(0, -2))
+abline(v = c(0.24))
 abline(v=c(0.5), lty=2)
 @
-\caption{A\label{wilcox_likelihood} support for the Bradley-Terry
+\caption{A\label{wilcox_likelihood} support function for the Bradley-Terry
 strength $p_a$ of permeability at term.  The evaluate of 0.24 is
 shown; and the two-units-of support credible interval, which does not
-exclude $H_0\colon p_a=0.5$ (dotted line), is also shown}
+exclude $H_0\colon p_a=0.5$ (dotted line), is also shown.}
 \end{center}
@@
 
-\section{Multiple competitors: javelin}
+\section{Multiple competitors: Javelin}
 
@@
 javelin_vector <- attemptstable2supp3(javelin_table,
-       decreasing=TRUE, give.supp=FALSE)
-options(width=60)
+       decreasing = TRUE, give.supp = FALSE)
+options(width = 60)
 javelin_vector
@@
 <<dothething2,cache=TRUE>>=
-javelin <- ordervec2supp3(v=names(javelin_vector)[!is.na(javelin_vector)])
+javelin <- ordervec2supp3(v = names(javelin_vector)[!is.na(javelin_vector)])
 @
@@
 <<setdigits,echo=FALSE>>=
-options(digits=3)
+options(digits = 3)
 @
@@
 (mj <- maxp(javelin))
-dotchart(mj, pch=16)
+dotchart(mj, pch = 16)
 @
-\caption{Maximum likelihood estimate for javelin throwers' Bradley-Terry strengths}
+\caption{Maximum likelihood estimate for javelin throwers' Bradley-Terry strengths.}
 \end{center}
@@
 f <- function(s){
-  jj <- specificp.test(javelin, "Vesely", s, n=2)
+  jj <- specificp.test(javelin, "Vesely", s, n = 2)
   p <- jj$null_estimate
@@
 }
-Ves <- seq(from=0.0199, to=0.33, len=16)
+Ves <- seq(from = 0.0199, to = 0.33, len = 16)
 M <- sapply(Ves, f)
@@
 colnames(M) <-  as.character(Ves)
-plot(t(M), type="b")
-abline(h=c(0, -2))
-abline(v=0, lty=2)
-abline(v=log(0.32062833 / 0.11402735)) # these from mp
+plot(t(M), type = "b")
+abline(h = c(0, -2))
+abline(v = 0, lty = 2)
+abline(v = log(0.32062833 / 0.11402735)) # these from mp
 @
@@
 line, and two-units-of-support limit indicated with horizontal lines;
-thus the null is not rejected}
+thus the null is not rejected.}
 \end{center}
@@
 
-\section{Formula 1 motor racing: the constructors' championship}\label{f1}
+\section{Formula 1 motor racing: the Constructors' Championship}\label{f1}
 
@@
 const2021 <- ordertable2supp3(constructor_2021_table)
-options(digits=4)
-maxp(const2020, n=1)
-maxp(const2021, n=1)
+options(digits = 4)
+maxp(const2020, n = 1)
+maxp(const2021, n = 1)
 @
@@
 <<usesamep,cache=TRUE>>=
-options(digits=4)
+options(digits = 4)
 samep.test(H, c("Merc2020", "Merc2021"))

### 2022-09-28 inst/jss_hyper3.Rnw
house style for p-value and edition

@@
 quite close to the null of $0.5$].  Nevertheless we can follow
-\cite{edwards1972} and apply Wilks's theorem for a $p$-value:
+\cite{edwards1972} and apply Wilks's theorem for a $p$~value:
 
@@
 
-The $p$-value is about 0.81, exceeding 0.05; thus we have no strong
+The $p$~value is about 0.81, exceeding 0.05; thus we have no strong
 evidence to reject the null of $a=\frac{1}{2}$.  The observation is
@@
 that the permeability of the two groups both have Bradley-Terry
-strength of $0.5$.  We see a p-value of about 0.09; compare 0.25 from
+strength of $0.5$.  We see a $p$~value of about 0.09; compare 0.25 from
 \code{wilcox.test()}.  However, observe that the \code{hyper3}

### 2022-09-28 inst/jss_hyper3.Rnw
minor house style

@@
 
-\section{Formula 1 motor racing: the Constructors' Championship}\label{f1}
+\section{Formula 1 motor racing: The Constructors' Championship}\label{f1}
 

### 2022-09-29 inst/j_sports_econ_F1points.tex
house style for Journal of Sports Economics, issue #181

--- 2021-10-20 inst/dg_template.tex
@@
 %% Use options letterpaper and 12pt.
-\documentclass[letterpaper,12pt]{article}
+\documentclass[a4,12pt]{article}
 
@@
 %% Load De Gruyter specific settings
-\usepackage{dgjournal}
+
 \usepackage{amsmath}
@@
 
+\usepackage{setspace}\doublespacing
+
 %% Start your document body here
@@
 
-%% Do NOT include any fronmatter information; including the title, author names,
-%% institutes, acknowledgments and title footnotes (author information, funding
-%% sources, etc.). Start the document with the first section or paragraph of
-%% the article.
+\begin{titlepage}
+  \begin{center}{\Large
+      An objective assessment of point scoring systems in Formula 1 motor racing}\\
+    Robin K. S. Hankin\\
+{\tt hankin.robin@gmail.com}\\
+Auckland University of Technology
+\end{center}
+
+\end{titlepage}
 
@@
 
+\newpage
+\section*{An objective assessment of point scoring systems in Formula 1 motor racing}
 \section*{Abstract}
@@
 system is best.
+\newpage
 
-\section{Introduction}\label{introduction}
+\section{\centering Introduction}\label{introduction}
 Formula 1 motor racing is an important and prestigious motor sport
@@
 and this would be a forward ranking Plackett-Luce model in the
-terminology of \cite{johnson2020}. A slight generalization allows the
+terminology of \cite{mollica2014}. A slight generalization allows the
 incorporation of non-finishers (DNF etc). If, say, competitors 4 and 5
@@
 
-\section{Formula 1 dataset}\label{formula-1-dataset}
+\section{\centering Formula 1 dataset}\label{formula-1-dataset}
 
@@
 \end{tabular}
-\caption{2017 Season results\label{results2017} table.  Each row is a
-  driver and each column (after the first) a venue. We see that
-  Hamilton, the first row, came second in Australia, first in China,
-  second in Bahrain, fourth in Russia, and so on (Hartley, Button, and
-  Resta placed last).  In the first column, we see the result from
-  Australia (AUS) in which Hamilton came second, Vettel first, Bottas
-  third, and so on.  Here, ``Ret'' means retired and a zero entry
-  means ``did not finish''}
+\caption{\doublespacing 2017 Season results\label{results2017} table.
+  Each row is a driver and each column (after the first) a venue. We
+  see that Hamilton, the first row, came second in Australia, first in
+  China, second in Bahrain, fourth in Russia, and so on (Hartley,
+  Button, and Resta placed last).  In the first column, we see the
+  result from Australia (AUS) in which Hamilton came second, Vettel
+  first, Bottas third, and so on.  Here, ``Ret'' means retired and a
+  zero entry means ``did not finish''}
 \end{table}
@@
 {\centering \includegraphics{printapiechart-1}}
-\caption[Maximum likelihood estimates \label{piechartstrength} of the
+\caption[\doublespacing Maximum likelihood estimates \label{piechartstrength} of the
   strengths of the top-ranked 11 drivers in Formula 1 motor racing,
-  seasons 2016-2019]{Maximum likelihood
+  seasons 2016-2019]{\doublespacing Maximum likelihood
   estimates \label{piechartstrength} of the strengths of the
@@
 {\centering \includegraphics{likevspoints-1}}
-\caption{Ordering of the \label{orderingbypointsandlikelihood}
-  top-ranked 11 drivers in Formula 1 seasons 2016-2019.  Horizontal
-  axis gives official (points-based) order, and the vertical axis
-  gives the likelihood order.  Thus, taking 2017 as an example, the
-  points-based ordering would be Hamilton first, then Vettel, then
-  \raik; while the likelihood ordering is (reading
-  vertically) Hamilton, Vettel, Bottas }\label{fig:likevspoints}
+\caption{\doublespacing Ordering of
+  the \label{orderingbypointsandlikelihood} top-ranked 11 drivers in
+  Formula 1 seasons 2016-2019.  Horizontal axis gives official
+  (points-based) order, and the vertical axis gives the likelihood
+  order.  Thus, taking 2017 as an example, the points-based ordering
+  would be Hamilton first, then Vettel, then \raik; while the
+  likelihood ordering is (reading vertically) Hamilton, Vettel, Bottas
+}\label{fig:likevspoints}
 \end{figure}
@@
 {\centering \includegraphics{likevspointszipf-1}}
-\caption[As for
+\caption[\doublespacing As for
   Figure~\ref{orderingbypointsandlikelihood} \label{orderingbypointsandlikelihood_zipf}
-  but points ranking calculated by Zipf's law]{As for
+  but points ranking calculated by Zipf's law]{\doublespacing As for
   Figure~\ref{orderingbypointsandlikelihood} \label{orderingbypointsandlikelihood_zipf}
@@
 \end{tabular}
-\caption{Degree of agreement for seasons 2016-9 with different
-  \label{doapoints} points systems}
+\caption{\doublespacing Degree of agreement for seasons 2016-9 with
+  different \label{doapoints} points systems}
 \end{table}
@@
 {\centering \includegraphics{alottaplotsmeancorrect-1}}
-\caption[Simulated Formula 1 races, seasons 2016-2019]{Simulated
-  Formula 1 races, seasons 2016-2019: mean number of agreeing places
-  for each of six points \label{simulated_mean} systems when compared
-  with a Plackett-Luce strength
-  ordering}\label{fig:alottaplotsmeancorrect}
+\caption[\doublespacing Simulated Formula 1 races, seasons
+  2016-2019]{\doublespacing Simulated Formula 1 races, seasons
+  2016-2019: mean number of agreeing places for each of six
+  points \label{simulated_mean} systems when compared with a
+  Plackett-Luce strength ordering}\label{fig:alottaplotsmeancorrect}
 \end{figure}
@@
 {\centering \includegraphics{alottaplotswinnercorrect-1} }
-\caption[Simulated Formula 1 races, seasons 2016-2019]{Simulated
-  Formula 1 races, seasons 2016-2019: probability of identifying the
-  Plackett-Luce winner for each of six
-  points \label{simulated_winner_correct}
+\caption[\doublespacing Simulated Formula 1 races, seasons
+  2016-2019]{\doublespacing Simulated Formula 1 races, seasons
+  2016-2019: probability of identifying the Plackett-Luce winner for
+  each of six points \label{simulated_winner_correct}
   systems}\label{fig:alottaplotswinnercorrect}
@@
 {\centering \includegraphics{alottaplotstotalrankorder-1}}
-\caption[Simulated Formula 1 races, seasons 2016-2019]{Simulated
-  Formula 1 races, seasons 2016-2019: probability of complete
-  agreement between Plackett-Luce ranks and ranks calculated by each
-  of six points \label{simulated_complete_agreement}
+\caption[\doublespacing Simulated Formula 1 races, seasons
+  2016-2019]{\doublespacing Simulated Formula 1 races, seasons
+  2016-2019: probability of complete agreement between Plackett-Luce
+  ranks and ranks calculated by each of six
+  points \label{simulated_complete_agreement}
   systems}\label{fig:alottaplotstotalrankorder}
@@
 
-\section{Conclusions}
+\section{\centering Conclusions}
 
@@
 %% BibTeX support
-\bibliographystyle{DeGruyter}
+\bibliographystyle{apalike2}
 \bibliography{formula1}

### 2022-09-29 inst/j_sports_econ_F1points.tex
abstract must be <= 100 words according to submission website

@@
 \section*{Abstract}
+%The points scoring system of Formula 1 motor racing is a long-standing
+%contention among fans and competitors, with changes being
+%controversial.  The inaugural points system was $(8,6,4,3,2)$, that
+%is, 8 points to the winner, 6 to second place, and so on and in 2020
+%the points system was changed to $(25,18,15,12,10,8,6,4,2,1)$.  The
+%overall ranking of the competitors is determined by summing points
+%awarded over a season, but this system suffers from two defects:
+%firstly, the points are intrinsically arbitrary in the sense that the
+%points system does not affect the competitors' behaviour but can
+%change the overall ranking.  Secondly, it is difficult to assess
+%statistical nulls using accumulated points.  Here, I use Plackett-Luce
+%likelihood to identify a ranking for the competitors, this method
+%including no arbitrary points system and being amenable to statistical
+%testing of nulls.  I go on to assess a number of reasonable points
+%systems objectively, by comparing likelihood-based rankings with
+%points-based rankings.  In one well-defined statistical sense a Borda
+%points system is preferred, and in another a ``winner takes all''
+%system is best.
+
 The points scoring system of Formula 1 motor racing is a long-standing
-contention among fans and competitors, with changes being
-controversial.  The inaugural points system was $(8,6,4,3,2)$, that
-is, 8 points to the winner, 6 to second place, and so on and in 2020
-the points system was changed to $(25,18,15,12,10,8,6,4,2,1)$.  The
-overall ranking of the competitors is determined by summing points
-awarded over a season, but this system suffers from two defects:
-firstly, the points are intrinsically arbitrary in the sense that the
-points system does not affect the competitors' behaviour but can
-change the overall ranking.  Secondly, it is difficult to assess
+contention among fans and competitors.  The inaugural points system
+was $(8,6,4,3,2)$, that is, 8 points to the winner, 6 to second place,
+and so on; in 2020 the points system was changed to
+$(25,18,15,12,10,8,6,4,2,1)$.  However, it is difficult to assess
 statistical nulls using accumulated points.  Here, I use Plackett-Luce
 likelihood to identify a ranking for the competitors, this method
-including no arbitrary points system and being amenable to statistical
-testing of nulls.  I go on to assess a number of reasonable points
-systems objectively, by comparing likelihood-based rankings with
-points-based rankings.  In one well-defined statistical sense a Borda
-points system is preferred, and in another a ``winner takes all''
-system is best.
+being amenable to statistical testing.  I go on to assess a
+number of reasonable points systems objectively: in one well-defined
+statistical sense a Borda points system is optimal.
+
 \newpage
@@
 an example, drivers Hamilton and Vettel are respectively first and
+
 second according to both ranking ranking procedures; but \raik\ and

### 2022-10-10 inst/j_sports_econ_F1points.tex
issue #181 refers

@@
       An objective assessment of point scoring systems in Formula 1 motor racing}\\
-    Robin K. S. Hankin\\
-{\tt hankin.robin@gmail.com}\\
-Auckland University of Technology
+    name redacted\\
+{\tt email redacted }\\
+affiliation redacted
 \end{center}
@@
 
-\section{\centering Introduction}\label{introduction}
+\section{Introduction}\label{introduction}
 Formula 1 motor racing is an important and prestigious motor sport
@@
 
-\section{\centering Formula 1 dataset}\label{formula-1-dataset}
+\section{Formula 1 dataset}\label{formula-1-dataset}
 
@@
 
-\section{\centering Conclusions}
+\section{Conclusions}
 
@@
 \bibliographystyle{apalike2}
-\bibliography{formula1}
+\begin{thebibliography}{}
+
+\bibitem[Bakhrankova, 2011]{bakhrankova2011}
+Bakhrankova, K. (2011).
+\newblock Points scoring systems in {F}ormula 1: A game theoretic perspective.
+\newblock In D. Klatte (Ed.), {\em Operations Research Proceedings}  (pp.\
+  211--216).: Springer-Verlag.
+
+\bibitem[Bradley, 1952]{bradley1952}
+Bradley, R.~A. (1952).
+\newblock Incomplete block rank analysis: On the appropriateness of the model
+  for a method of paired comparisons.
+\newblock {\em Biometrics}, 10(3), 375--390.
+
+\bibitem[Codling, 2017]{codling2017}
+Codling, S. (2017).
+\newblock {\em Speed read {F1}}.
+\newblock Motorbooks.
+\newblock ISBN 978-0-7603-5562-6.
+
+\bibitem[Emerson, 2007]{emerson2007}
+Emerson, P. (2007).
+\newblock {\em Designing an All-Inclusive Democracy---Consensual Voting
+  Procedures for use in Parliaments, Councils and Committees}.
+\newblock Springer-Verlag.
+\newblock {I}{S}{B}{N} 978-3-540-33163-6.
+
+\bibitem[Gay-Rees, 2019]{gayrees2019}
+Gay-Rees, J. (2019).
+\newblock {F}ormula 1: drive to survive.
+\newblock Netflix documentary.
+
+\bibitem[Hankin, 2010]{hankin2010}
+Hankin, R. K.~S. (2010).
+\newblock A generalization of the {D}irichlet distribution.
+\newblock {\em Journal of Statistical Software}, 33(11), 1--18.
+
+\bibitem[Hankin, 2017]{hankin2017}
+Hankin, R. K.~S. (2017).
+\newblock Partial rank data with the {hyper2} package: likelihood functions for
+  generalized {B}radley-{T}erry models.
+\newblock {\em The R Journal}, 9(2), 429--439.
+
+\bibitem[Jenkins, 2010]{jenkins2010}
+Jenkins, M. (2010).
+\newblock Technological discontinuities and competitive advantage: a historical
+  perspective on {F}ormula 1 motor racing 1950-2006.
+\newblock {\em Journal of Management Studies}, 47(5), 884--910.
+
+\bibitem[Luce, 1959]{luce1959}
+Luce, R. (1959).
+\newblock {\em Individual choice behaviour: A theoretical analysis}.
+\newblock New York: Wiley.
+
+\bibitem[Mastromarco \& Runkel, 2009]{mastromarco2009}
+Mastromarco, C. \& Runkel, M. (2009).
+\newblock Rule changes and competitive balance in {F}ormula {O}ne motor racing.
+\newblock {\em Applied Economics}, 41(23), 3003--3014.
+
+\bibitem[Mollica \& Tardella, 2014]{mollica2014}
+Mollica, C. \& Tardella, L. (2014).
+\newblock Epitope profiling via mixture modeling of ranked data.
+\newblock {\em Statistics in Medicine}, 33, 3738--3758.
+
+\bibitem[Plackett, 1975]{plackett1975}
+Plackett, R.~L. (1975).
+\newblock The analysis of permutations.
+\newblock {\em Applied Statistics}, 24, 193--202.
+
+\bibitem[Wood, 2020]{wood2020}
+Wood, E. (2020).
+\newblock {F}ormula {S}cout (published online).
+
+\bibitem[Zipf, 1949]{zipf1949}
+Zipf, G. (1949).
+\newblock {\em Human Behavior and the Principle of Least Effort}.
+\newblock Addison-Wesley.
+
+\end{thebibliography}
 

### 2022-10-12 inst/volleyball.Rmd
labels

@@
 
-```{r}
+```{r calcmaxpvolley}
 volleyball <- volley(volleyball_table)
@@
 
-```{r}
+```{r zipflikelihood}
 zipf <- function(n){jj <- 1/(1:n); jj/sum(jj)}
@@
 
-```{r,label=savecurlingdataset}
+```{r,label=savevolleyballdataset}
 save(volleyball_table,volleyball_maxp,volleyball,file="volleyball.rda")

### 2022-10-17 inst/skating.Rmd
name every chunk, also change title

@@
 ---
-title: "skating"
+title: "Ladies' figure skating in the 2002 Winter Olympics"
 output: bookdown::html_document2
@@
 
-# Ladies' figure skating in the 2002 Winter Olympics
-
-This document analyses strengths of the competitors in the Ladies'
-Figure Skating at the 2002 Winter Olympics.  In the package, dataframe
-`skating_table` is copied from Lock and Lock (see skating.Rd for more
-details).  It also creates file `data/skating.rda` which contains R
-objects `skating`, `skating_table`, and `skating_maxp`.
+This document analyses Bradley-Terry strengths of the competitors in
+the Ladies' Figure Skating at the 2002 Winter Olympics.  In the
+package, dataframe `skating_table` is copied from Lock and Lock (see
+skating.Rd for more details).  It also creates file `data/skating.rda`
+which contains R objects `skating`, `skating_table`, and
+`skating_maxp`.
 
@@
 
-```{r,dpi=72}
+```{r,dotchartm,dpi=72}
 dotchart(m)
@@
 
-```{r}
+```{r subtractmaxfrommgv}
 mgv  - max(out[4:6])
@@
 
-```{r}
+```{r plotrestrictedlikes}
 plot(out,ylab="log likelihood",axes=FALSE)
@@
 
-```{r}
+```{r justtop7}
 skating_table
@@
 
-```{r}
+```{r setclasstomatrix}
 class(o) <- "matrix"
@@
 Select just the top seven:
-```{r}
+```{r selectthetop7}
 o <- o[,1:7]
@@
 
-and now replace ranks with internal ranks (that is, rank among the top seven):
+and now replace ranks with internal ranks (that is, rank among the top
+seven):
 
-```{r}
+```{r replaceranks}
 o <- t(apply(o,1,rank))
@@
 
-```{r}
+```{r coercetoranktble}
 class(o) <- "ranktable"
@@
 
-```{r}
+```{r thenceto}
 o <- ranktable_to_ordertable(o)
@@
 
-```{r}
+```{r makelikef}
 o <- ordertable2supp(o)
@@
 
-```{r}
+```{r plotapieofo}
 (o <- maxp(o))

### 2022-10-18 inst/skating.Rmd
loads of fixes, accounts for name ordering this time

@@
 m <- skating_maxp # for ease of  typing
+m
+```
+
+```{r equalskatingptest}
+equalp.test(skating)
 ```
@@
 
+```{r}
+f <- function(smaller,bigger){
+  jj <- rep(0,22)
+  names(jj) <- pnames(skating)[-23]  #  no fillup
+  jj[smaller] <- -1
+  jj[bigger ] <- +1
+  return(jj)
+}
+
+problem_constraints <-  NULL
+problem_constraints %<>% rbind(f("hughes"   ,"cohen" ))
+problem_constraints %<>% rbind(f("slutskaya","cohen" ))
+problem_constraints %<>% rbind(f("kwan"     ,"cohen" ))
+problem_constraints %<>% rbind(f("hughes"   ,"suguri"))
+problem_constraints %<>% rbind(f("slutskaya","suguri"))
+problem_constraints %<>% rbind(f("kwan"     ,"suguri"))
+
+rownames(problem_constraints) <-
+c("hug < coh", "slu < coh", "kwa < coh", "hug < sug", "slu < sug", "kwa < sug" )
+colnames(problem_constraints) %<>% substr(1,3)
+problem_constraints
+
+small <- 0.01
+big <- 0.1
+start_vector <- rep(small,22)
+names(start_vector) <- pnames(skating)[-23]  #  no fillup
+start_vector["cohen"] <- big
+start_vector["suguri"] <- big
+```
+
 ```{r,longthing,cache=TRUE}
-jj <-
-    matrix(c(
-        -1, 0, 0,   1,0,  # Hughes   < Cohen
-         0,-1, 0,   1,0,  # Slutskya < Cohen
-         0, 0,-1,   1,0,  # Kwan     < Cohen
-
-        -1, 0, 0,   0,1,  # Hughes   < Suguri
-         0,-1, 0,   0,1,  # Slutskya < Suguri
-         0, 0,-1,   0,1), # Kwan     < Suguri
-        byrow=TRUE,ncol=5)
-
-problem_constraints <-  # fill with zeros for other competitors
-    cbind(jj,matrix(0,nrow(jj),size(skating)-ncol(jj)-1))
-
-small <- 1e-4  # need a sensible start value satisfying the constraints
-start <- c(rep(2*small,3),rep(3*small,2),rep(small,17))
 
@@
 
+mle_skaters <- NULL
+
 for(i in seq_len(nrow(problem_constraints))){
-  jj <- maxp(skating, startp=start, give=TRUE,fcm=problem_constraints[i,], fcv=0,n=1,SMALL=1e-5)
+  jj <- maxp(skating, startp=start_vector, give=TRUE,fcm=problem_constraints[i,], fcv=0,n=10,SMALL=1e-5)
    fullout[[i]] <- jj
    out[i] <- jj$value
+   mle_skaters <- cbind(mle_skaters,jj$par)
+
 }
 out
+out-max(out)
+colnames(mle_skaters) <- rownames(problem_constraints)
+mle_skaters
 ```
@@
 mgv <- maxp(skating, give=TRUE, n=10)$value
+mgv
 mgv  - out
@@
 
-that is, a little over 1 unit of support, falling short of the two
-units suggested by Edwards.  Observe that the maximum likelihood among
-the six alternative hypotheses is that of number 3, in which the
-maximization was constrained to obey `Kwan < Cohen`.
-
-If, instead, we ask whether there is evidence that Suguri should not
-have been a medallist, we find
+that is, a little over 7 units of support, clearly exceeding the two
+units suggested by Edwards.  We have strong evidence to suggest that
+the medallists were indeed the strongest three competitors.  Observe
+that the maximum likelihood among the six alternative hypotheses is
+that of number 3, in which the maximization was constrained to obey
+`Kwan < Cohen`.  If we _have_ to change the medallists, then
+exchanging the order of Kwan and Cohen incurs the lowest likelihood
+penalty.  If, instead, we ask whether there is evidence that Suguri
+[who was much weaker than Cohen] should not have been a medallist, we
+find
 
@@
 
-We can be more specific with the likelihoods:
+Much stronger.
 

### 2022-10-18 inst/salad.Rmd
salad

---
title: "salad"
output: bookdown::html_document2
---

```{r setup, include=FALSE}
set.seed(0)
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("magrittr")
library("prefmod")
library("PlackettLuce")
options("digits" = 5)
```

<p style="text-align: right;">
![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
</p>

This short document shows how to apply `hyper2` idiom to the `salad`
dataset of the `prefmod` package.  From `salad.Rd`:

```
The dataset contains the rankings of four salad dressings concerning
tartness by 32 judges, with values ranging from 1 (most tart) to 4
(least tart).
```

```{r showsalad}
head(salad)
nrow(salad)
```

From row 3, for exampl,e we see that salad `A` was ranked as second
most tart, `B` the most tart, `C` the third most tart, and `D` the
least tart of the four.  We may process this using two methods,
explicit and slick.  First, explicit:

```{r useexplicitmethod,cache=TRUE}
H1 <- hyper2()
for(i in seq_len(nrow(salad))){
   H1 <-  H1 + ordervec2supp(as.matrix(salad)[i,])
}
H1
```

And second, slick:

```{r useslickmethod,cache=TRUE}
(H2 <- ordertable2supp(t(salad)))
```

Just to check:

```{r checkequalityofsalads}
H1 == H2
```

```{r equalsaladtart, cache=TRUE}
equalp.test(H1)
```

The null estimate agrees to six places of decimals with that
presented by Turner:

```{r usepladmm, cache=TRUE}
standardPL_PlackettLuce <- PlackettLuce(salad, npseudo = 0)
(p1 <- itempar(standardPL_PlackettLuce))
(p2 <- maxp(H1))
p1-p2
```

# References

Turner HL, van Etten J, Firth D, Kosmidis I (2020).  "Modelling
Rankings in R: The PlackettLuce Package."  _Computational Statistics_,
35, 1027–1057. doi:10.1007/s00180-020-00959-3,
\url{https://doi.org/10.1007/s00180-020-00959-3}

Yildiz, Ilkay, Jennifer Dy, Deniz Erdogmus, Jayashree Kalpathy-Cramer,
Susan Ostmo, J. Peter Campbell, Michael F. Chiang, and Stratis
Ioannidis. 2020. "Fast and Accurate Ranking Regression."  In
Proceedings of the Twenty Third International Conference on Artificial
Intelligence and Statistics, edited by Silvia Chiappa and Roberto
Calandra, 108:77–88. Proceedings of Machine Learning
Research. \url{http://proceedings.mlr.press/v108/yildiz20a.html}

### 2022-10-18 inst/collusion3.Rmd
minor rephrasing

@@
 We analyse results from Curacao 1962 using `hyper3` formalism using
-the weighted Draw monster appraoch outlined in `kka.Rmd`:
+the weighted Draw monster approach outlined in `kka.Rmd`:
 
@@
 
-With this model, whatever the value of $D$ and $\lambda$, there is a
-pathology which BT cannot deal with.  Still not quite 100% sure but it
-seems that the maximum likelihood estimate of the players' strengths
-is zero except for Fisher and Petrosian; and indeed we see Fisher at
-57% and Petrosian at 42% [with $\lambda = 1.1, D=1.888$].  This on the
-grounds that these players never lost playing black (there are some
-$2\times 3$ tables at the end).  This might be considered a defect of
-the probability model.
+With this model [discussed in a broader context at `kka.Rmd`],
+whatever the value of $D$ and $\lambda$, there is a pathology which BT
+cannot deal with.  Still not quite 100% sure but it seems that the
+maximum likelihood estimate of the players' strengths is zero except
+for Fisher and Petrosian; and indeed we see Fisher at 57% and
+Petrosian at 42% [with $\lambda = 1.1, D=1.888$].  This on the grounds
+that these players never lost playing black (there are some $2\times
+3$ tables at the end).  This might be considered a defect of the
+probability model.
 
@@
 cur_matches <- read.table("stockholm1962_matches.txt",header=FALSE)
+colnames(cur_matches) <- c("white","black","result")
 head(cur_matches)
@@
 ```{r,comparelikelihoods,cache=TRUE}
-maxp(f(1.1,2.1),n=1,give=TRUE)
-maxp(f(1.1,2.3),n=1,give=TRUE)
+m1121 <- maxp(f(1.1,2.1),n=1,give=TRUE)
+m1123 <- maxp(f(1.1,2.3),n=1,give=TRUE)
+```
+
+```{r,showmaxpwithlam}
+pie(m1121$par)
 ```

### 2022-10-25 inst/skating.Rmd
slight improvement on skating_maxp

@@
 skating <- ordertable2supp(skating_table)
-skating_maxp <- maxp(skating)
+options("use_alabama" = TRUE)
+skating_maxp <- maxp(skating,startp=
+c(5.0493913251e-06, 0.0053037393083, 0.090240695999, 
+0.00036893464900, 1.1397045995e-05, 8.3424027249e-05, 
+9.8129903007e-06, 0.00026960143514, 0.29831585124, 
+0.001376333090, 2.9434081237e-06, 0.26550514587, 
+0.00051946513092, 1.7e-06, 0.00080745068338, 
+0.00029358481485, 0.00048852120227, 0.0053075302458, 
+0.0025877438737, 0.30889220253, 1.2869403350e-05, 
+0.0184563624305147))    # values from previous repeated runs of maxp()
 m <- skating_maxp # for ease of  typing
@@
 
-```{r equalskatingptest}
+```{r equalpskatingtest,cache=TRUE}
 equalp.test(skating)
@@
 
+We may test the hypothesis that Slutskaya == Kwan == Hughes:
+
+```{r,testtop3,cache=TRUE}
+samep.test(skating,c("slutskaya","kwan","hughes"))
+```
+
 ## Ranking methodologies.
@@
 
-```{r plotapieofo}
+```{r plotapieofo,cache=TRUE}
 (o <- maxp(o))

### 2022-10-25 inst/f1points.Rmd
whitespace
more work on new parametrization

@@
 
-```{r,label=showanordertableandhead,echo=FALSE,print=FALSE}
-library(hyper2,quietly=TRUE)
+```{r,label=loadlibs,echo=FALSE,print=FALSE}
+library("hyper2",quietly=TRUE)
+library("magrittr",quietly=TRUE)
+```
+
+```{r,label=showanordertableandhead,echo=FALSE,print=FALSE,cache=TRUE}
 top <- 11
@@
 ```{r,label=combineallyearssetup,echo=FALSE,cache=TRUE}
-H2016a <- hyper2(brackets(H2016),d=powers(H2016),pnames=sub("Hamilton","Hamilton_pre", pnames(H2016)))
-H2017a <- hyper2(brackets(H2017),d=powers(H2017),pnames=sub("Hamilton","Hamilton_post",pnames(H2017)))
-H2018a <- hyper2(brackets(H2018),d=powers(H2018),pnames=sub("Hamilton","Hamilton_post",pnames(H2018)))
-H2019a <- hyper2(brackets(H2019),d=powers(H2019),pnames=sub("Hamilton","Hamilton_post",pnames(H2019)))
-Ha <- H2016a
-Ha <- combine(Ha,H2017a)
-Ha <- combine(Ha,H2018a) 
-Ha <- combine(Ha,H2019a)
+H2016a <- H2016 %>% psubs("Hamilton","Hamilton_pre")
+H2017a <- H2017 %>% psubs("Hamilton","Hamilton_post")
+H2018a <- H2018 %>% psubs("Hamilton","Hamilton_post")
+H2019a <- H2019 %>% psubs("Hamilton","Hamilton_post")
+Ha <-  H2016a + H2017a + H2018a + H2019a
+```
+
+Do a statistical test:
+
+```{r,testhamprepost,cache=TRUE}
 ignorethis <- samep.test(Ha,c("Hamilton_pre","Hamilton_post"))
+ignorethis
 ```
+
+Do another statistical test:
 
@@
 
-```{r, label=calcpoints_real, echo=FALSE, cache=TRUE}
-```{r, label=calcpoints_real, echo=FALSE, cache=TRUE}
+```{r, label=calcpoints_real_2016, echo=FALSE, cache=TRUE}
 p <- formula1_points_systems()[[1]]   # real points system
@@
 m2016[] <- seq_along(m2016)
+```
 
+```{r, label=calcpoints_real_2017, echo=FALSE, cache=TRUE}
 dp2017 <- ordertable2points(as.ordertable(F1_table_2017[seq_len(top),]),p)
@@
 
+```{r, label=calcpoints_real_2018, echo=FALSE, cache=TRUE}
 dp2018 <- ordertable2points(as.ordertable(F1_table_2018[seq_len(top),]),p)
@@
 m2018[] <- seq_along(m2018)
+```
 
+```{r, label=calcpoints_real_2019, echo=FALSE, cache=TRUE}
 dp2019 <- ordertable2points(as.ordertable(F1_table_2019[seq_len(top),]),p)
@@
 ox <- dp2016
-oy <- ordertrans(m2016,dp2016)
+
+print("m2016")
+dput(m2016)
+print("dp2016")
+dput(dp2016)
+
+oy <- ordertrans(m2016,names(dp2016))
 plot(ox,oy,asp=1,pty='s',xlim=c(0,11.2),ylim=c(0,11.2),pch=16,
@@
 ox <- dp2017
-oy <- ordertrans(m2017,dp2017)
+oy <- ordertrans(m2017,names(dp2017))
 plot(ox,oy,asp=1,pty='s',xlim=c(0,11.2),ylim=c(0,11.2),pch=16,
@@
 ox <- dp2018
-oy <- ordertrans(m2018,dp2018)
+oy <- ordertrans(m2018,names(dp2018))
 plot(ox,oy,asp=1,pty='s',xlim=c(0,11.2),ylim=c(0,11.2),pch=16,
@@
 ox <- dp2019
-oy <- ordertrans(m2019,dp2019)
+oy <- ordertrans(m2019,names(dp2019))
 plot(ox,oy,asp=1,pty='s',xlim=c(0,11.2),ylim=c(0,11.2),pch=16,
@@
 ox <- dp2016z
-oy <- ordertrans(m2016,dp2016z)
+oy <- ordertrans(m2016,names(dp2016z))
 plot(ox,oy,asp=1,pty='s',xlim=c(0,11.2),ylim=c(0,11.2),pch=16,
@@
 ox <- dp2017z
-oy <- ordertrans(m2017,dp2017z)
+oy <- ordertrans(m2017,names(dp2017z))
 plot(ox,oy,asp=1,pty='s',xlim=c(0,11.2),ylim=c(0,11.2),pch=16,
@@
 ox <- dp2018z
-oy <- ordertrans(m2018,dp2018z)
+oy <- ordertrans(m2018,names(dp2018z))
 plot(ox,oy,asp=1,pty='s',xlim=c(0,11.2),ylim=c(0,11.2),pch=16,
@@
 ox <- dp2019z
-oy <- ordertrans(m2019,dp2019z)
+oy <- ordertrans(m2019,names(dp2019z))
 plot(ox,oy,asp=1,pty='s',xlim=c(0,11.2),ylim=c(0,11.2),pch=16,

### 2022-10-25 inst/cook.Rmd
more work

@@
 library(hyper2)
-
 ```
@@
 
+```{r,label=plotsupplog}
+plot(log(lam),L-max(L),type="l",lwd=2)
+abline(h=c(0,-2),col='gray')
+abline(v=1,col='red')
+text(0,-0.2,"lambda=1",col="red",pos=4)
+```
+
 We see that we can reject $H_0\colon\lambda=1$ but for this dataset
 there does not appear to be a clear upper bound on $\lambda$.
+
+We can calculate the maximum likelihood estimate as follows:
+
+```{r,label=maxlikelambda,cache=TRUE}
+o <- optimize(supplam,c(40,70),maximum=TRUE)
+o
+```
+
+To calculate a formal likelihood ratio we need to calculate the
+support for $\lambda=1$ and compare it with the support for
+$\lambda=\hat{\lambda}$
+
+```{r,supmax,cache=TRUE}
+supplam(o$maximum)
+supplam(1)
+```
+
+and we get a $p$-value for $H_0\colon\lambda=1$ using Wilks,
+specifically that
+$2\log\left(\frac{\mathcal{L}_{\lambda=\hat{\lambda}}}{\mathcal{L}_{\lambda=1}}\right)\sim\chi^2_1$
+
+```{r pvalueforlambdaequalsone,cache=TRUE}
+pchisq(2*(supplam(o$maximum)-supplam(1)),lower.tail=FALSE,df=1)
+```
+
+Actually we should be able to analyse $\lim_{\lambda\longrightarrow\infty}H_\lambda$:
+
+```{r,lambigcalc,cache=TRUE}
+lambig <- 10^(0:6)
+LL <- sapply(lambig,supplam)
+```
+
+```{r,plotlambdiff,echo=F}
+plot(log(lambig),LL-max(LL),pch=16)
+for(i in seq(from=0,to=8,by=2)){abline(i,-1,col='gray',lwd=0.5)}
+abline(v=log(o$maximum))
+```
+
+Looking at the two parts of the likelihood function that includes
+lambda, and thinking about $\lim_{\lambda\longrightarrow\infty}$ we
+see that the first product is approximately constant and the second is
+$\propto\lambda^{-1}$.  This analysis is a little bit naive because as
+$\lambda$ grows very large, we might expect $\hat{p_3}$ to grow too.  But:
+
+```{r,maxplam1110,cache=TRUE}
+M <- rbind(
+maxp(H3(10^0    ),n=1,give=TRUE)$par,
+maxp(H3(10^1    ),n=1,give=TRUE)$par,
+maxp(H3(10^2    ),n=1,give=TRUE)$par,
+maxp(H3(10^3    ),n=1,give=TRUE)$par,
+maxp(H3(10^4    ),n=1,give=TRUE)$par,
+maxp(H3(10^5    ),n=1,give=TRUE)$par,
+maxp(H3(10^6    ),n=1,give=TRUE)$par
+)
+M
+```
+
+```{r}
+matplot(M,type='b')
+```
+
+Above we see that the evaluate hardly changes when $\lambda > 10^4$,
+and indeed from the figure we see that the slope of the support graph
+is about $-1$ for large values of $\lambda$ [the points follow a line
+parallel to one of the slanted gray lines, which have a slope of
+exactly $-1$; note that the aspect ratio of the graph is not unity]

### 2022-11-21 inst/ternaryplot_hyper2.Rmd
last part of issue #187

---
title: "ternary_hyper2"
author: "Robin Hankin"
date: "2022-11-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Ternary plots for hyper2 samples

Here I show how to use the excellent `Ternary` package to generate triangular plots for `hyper2` likelihood functions with size 3.  First load the `Ternary` and `hyper2` packages:

```{r loadlibraries}
library("hyper2")
library("Ternary")
```

Now generate a synthetic dataset.  Here we will sample from the posterior of a uniform prior with a likelihood function specified by the `chess` object (this takes quite a long time):

```{r generatedata, cache=TRUE}
set.seed(0)
xx <- rp(1000,chess)
```

Using the `TernaryPlot()` and `TernaryPoints()` functions is straightforward:
```{r plotdata}
jj <- pnames(chess)
TernaryPlot(atip=jj[1],btip=jj[2],ctip=jj[3])
TernaryPoints(xx,cex=0.5)
TernaryPoints(chess_maxp,pch=16,col="red")
```

Above we see black circles being samples from the posterior and the red point is the maximum likelihood estimate for the three players' strengths.

We can adapt this technique to higher-dimensional `hyper2` objects.

### 2022-11-21 inst/ternaryplot_hyper2.Rmd
use the icons dataset in a ternary plot (issue #187)

@@
 
-Here I show how to use the excellent `Ternary` package to generate triangular plots for `hyper2` likelihood functions with size 3.  First load the `Ternary` and `hyper2` packages:
+Here I show how to use the excellent `Ternary` package to generate
+triangular plots for `hyper2` likelihood functions with size 3.  First
+load the `Ternary` and `hyper2` packages:
 
@@
 
-Now generate a synthetic dataset.  Here we will sample from the posterior of a uniform prior with a likelihood function specified by the `chess` object (this takes quite a long time):
+## The `chess` dataset
+
+Now generate a synthetic dataset.  Here we will sample from the
+posterior of a uniform prior with a likelihood function specified by
+the `chess` object (this takes quite a long time):
 
@@
 
-Above we see black circles being samples from the posterior and the red point is the maximum likelihood estimate for the three players' strengths.
+Above we see black circles being samples from the posterior and the
+red point is the maximum likelihood estimate for the three players'
+strengths.
 
 We can adapt this technique to higher-dimensional `hyper2` objects.
+
+## The `icons` dataset
+
+Consider the `icons` likelihood function:
+
+```{r showicons}
+icons
+```
+
+We can sample from this object and plot it in different ways:
+
+```{r samplefromicons,cache=TRUE}
+n <- 10000
+x <- rp(n,icons)
+x[sample(n,5),]
+```
+
+```{r, iconsmatplot}
+matplot(x,type="l",lty=1)
+legend("topright",col=1:6,lty=1,legend=pnames(icons))
+```
+
+```{r, selectjustbigones}
+wanted <- c(1,3) # NB,PB
+xx <- cbind(x[,wanted],fillup=rowSums(x[,-wanted]))
+TernaryPlot(atip="NB",btip="PB",ctip="other")
+TernaryPoints(xx,cex=0.5)
+```

### 2022-11-22 inst/rowing.Rmd
rowing_table

@@
 ```{r}
-o <- sapply(readLines("rowing.txt"),strsplit," ")
-names(o) <- NULL  # names are pointless here
-o[1:3]
+rowing_table <- sapply(readLines("rowing.txt"),strsplit," ")
+names(rowing_table) <- NULL  # names are pointless here
+rowing_table[1:3]
 ```
@@
 ```{r definerowing, cache=TRUE}
-rowing <- Reduce("+",sapply(o,race))
+rowing <- Reduce("+",sapply(rowing_table,race))
 summary(rowing)
@@
 ```{r,label=saverowingingdataset}
-save(rowing,rowing_minimal,rowing_maxp,rowing_minimal_maxp,file="rowing.rda")
+save(rowing,rowing_minimal,rowing_maxp,rowing_minimal_maxp,rowing_table,file="rowing.rda")
 ```

### 2022-11-22 inst/rowing.Rmd
rowing_minimal_table

@@
 ```{r rowingminimaldef}
-o <- sapply(readLines("rowing_minimal.txt"),strsplit," ")
-names(o) <- NULL  
-rowing_minimal  <- Reduce("+",sapply(o,race))
+rowing_minimal_table <- sapply(readLines("rowing_minimal.txt"),strsplit," ")
+names(rowing_minimal_table) <- NULL  
+rowing_minimal  <- Reduce("+",sapply(rowing_minimal_table,race))
 summary(rowing_minimal)
@@
 ```{r,label=saverowingingdataset}
-save(rowing,rowing_minimal,rowing_maxp,rowing_minimal_maxp,rowing_table,file="rowing.rda")
+save(rowing,rowing_minimal,rowing_maxp,rowing_minimal_maxp,rowing_table,rowing_minimal_table,file="rowing.rda")
 ```

### 2022-11-28 inst/jebo_F1points.tex
issue #185

--- 2022-09-29 inst/j_sports_econ_F1points.tex
@@
-%% This file is to be used as a template for your submission.
-%% Rename this file and replace the text with the text of
-%% your manuscript.
-%%
-%% The standard LaTeX document class "article" is recommended.
-%% Use options letterpaper and 12pt.
-\documentclass[a4,12pt]{article}
+\documentclass{elsarticle}
 
-%% This is the recommended preamble for your document.
+\usepackage{lineno,hyperref}
+\usepackage{amsmath}
+\usepackage{mathptmx}
+\usepackage{graphics}
+\usepackage{setspace}\doublespacing
+\modulolinenumbers[5]
 
-%% Load De Gruyter specific settings
+\journal{Journal of Economic Behavior and Organization}
 
-\usepackage{amsmath}
-
-%% The mathptmx package is recommended for Times compatible math symbols.
-%% Use mtpro2 or mathtime instead of mathptmx if you have the commercially
-%% available MathTime fonts.
-%% Other options are txfonts (free) or belleek (free) or TM-Math (commercial)
-\usepackage{mathptmx}
-
-%% Use the graphics package to include figures
-\usepackage{graphics}
-
-%% Use natbib with these recommended options
-\usepackage[authoryear,comma,longnamesfirst,sectionbib]{natbib}
-
-\usepackage{setspace}\doublespacing
+\bibliographystyle{model2-names.bst}\biboptions{authoryear}
 
@@
 
-\begin{titlepage}
-  \begin{center}{\Large
-      An objective assessment of point scoring systems in Formula 1 motor racing}\\
-    Robin K. S. Hankin\\
-{\tt hankin.robin@gmail.com}\\
-Auckland University of Technology
-\end{center}
+\begin{frontmatter}
+  \title{An objective assessment of point scoring systems in Formula 1 motor racing}
+  \author{Robin K. S. Hankin\fnref{myfootnote}}
+\address{AUT University, 55 Wellesley Street, Auckland 1010, New Zealand}
 
-\end{titlepage}
-
-\newcommand{\raik}{R\"{a}ikk\"{o}nen}
-
-\newpage
-\section*{An objective assessment of point scoring systems in Formula 1 motor racing}
-\section*{Abstract}
-%The points scoring system of Formula 1 motor racing is a long-standing
-%contention among fans and competitors, with changes being
-%controversial.  The inaugural points system was $(8,6,4,3,2)$, that
-%is, 8 points to the winner, 6 to second place, and so on and in 2020
-%the points system was changed to $(25,18,15,12,10,8,6,4,2,1)$.  The
-%overall ranking of the competitors is determined by summing points
-%awarded over a season, but this system suffers from two defects:
-%firstly, the points are intrinsically arbitrary in the sense that the
-%points system does not affect the competitors' behaviour but can
-%change the overall ranking.  Secondly, it is difficult to assess
-%statistical nulls using accumulated points.  Here, I use Plackett-Luce
-%likelihood to identify a ranking for the competitors, this method
-%including no arbitrary points system and being amenable to statistical
-%testing of nulls.  I go on to assess a number of reasonable points
-%systems objectively, by comparing likelihood-based rankings with
-%points-based rankings.  In one well-defined statistical sense a Borda
-%points system is preferred, and in another a ``winner takes all''
-%system is best.
-
+\begin{abstract}
 The points scoring system of Formula 1 motor racing is a long-standing
@@
 statistical nulls using accumulated points.  Here, I use Plackett-Luce
-likelihood to identify a ranking for the competitors, this method
-being amenable to statistical testing.  I go on to assess a
-number of reasonable points systems objectively: in one well-defined
-statistical sense a Borda points system is optimal.
+likelihood to identify a ranking for the competitors from the
+perspective of tournament theory, this method being amenable to
+statistical testing.  I go on to assess a number of reasonable points
+systems objectively: in one well-defined statistical sense a Borda
+points system is optimal.
+\end{abstract}
+\begin{keyword}
+Reified Bradley-Terry\sep Likelihood\sep Formula 1 motor racing\sep Points systems
+\end{keyword}
 
-\newpage
+\end{frontmatter}
+\linenumbers
 
-\section{\centering Introduction}\label{introduction}
+\newcommand{\raik}{R\"{a}ikk\"{o}nen}
+
+\section{Introduction}\label{introduction}
 Formula 1 motor racing is an important and prestigious motor sport
@@
 competition winner is the competitor who accumulates the most points
-after the final race.
+after the final race.  The intent of the points system is to
+incentivize competitors, stimulate innovation, and to create an
+exciting sporting spectacle: as such, its study is a practical
+application of tournament theory~\citep{lazear1981}.
 
@@
 $p_4+p_5$, is uninformative about $p_4\left|p_4+p_5\right.$). We now
-use a technique due to \cite{hankin2010,hankin2017} and introduce
+use a technique due to \cite{hankin2010,hankin2020} and introduce
 fictional (reified) entities whose nonzero Bradley-Terry strength
@@
 which implements the Plackett-Luce likelihood function with additional
-reified entities.
+reified entities~\cite{hankin2020}.
 
@@
 %% BibTeX support
-\bibliographystyle{apalike2}
 \bibliography{formula1}

### 2022-11-30 inst/pentathlon.Rmd
typo

@@
 
-```{r,label=savecurlingdataset}
+```{r,label=savepentathlondataset}
 save(pentathlon_table,pentathlon,pentathlon_maxp,file="pentathlon.rda")

### 2022-12-01 inst/skeleton.Rmd
nice analysis of skeleton

---
title: "skeleton"
output: bookdown::html_document2
---

```{r setup, include=FALSE}
set.seed(0)
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("magrittr")
library("igraph")
options("digits" = 5)
```

<p style="text-align: right;">
![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
</p>

The [skeleton](https://en.wikipedia.org/wiki/Skeleton_(sport)) is a
winter sliding sport.  This short document discusses a dataset taken
from the [2018 Winter
Olympics](https://en.wikipedia.org/wiki/Skeleton_at_the_2018_Winter_Olympics_%E2%80%93_Men%27s)
and analyses the order statistic for the competitors.

```{r readskeletontable}
(skeleton_table <- read.table("skeleton2022.txt",header=TRUE))
```

Thus the top line shows that Grotheer came first in runs 1-3 and sixth
in run 4.  The first column shows the results for run 1: we see that
Grotheer came first, Jungk came fifth, and so on.  There is in fact no
dominated set: Timmings [place 25] beats Enzie [place 19] (in run 4),
Enzie [19] beats Crumpton [22] (run 1), Crumpton [17] beats Wyatt [18]
(2), Wyatt [16] beats Maier [15] (1), Maier [11] beats Bagnis [12]
(4), Bagnis [10] beasts Seunggi [11] (1), Seunggi [9] beats T. Dukurs
[10] (3 and 4), Dukurs [7] beats Tretyakov [8] (2), Tretyakov [2]
beats Wengang [3] (1), and Wengang beats everybody (4).

We can *try* to automate the above cycle as follows.  The first step
is to convert `skeleton_table` to a numeric matrix, which proves to be
a very awkward process

NB: `data.matrix()` fucks things up:

```{r,label=datamatrixfucksthingsup}
head(data.matrix(skeleton_table))
```

Above, we see that Grotheer apparently came 17th in run 4, clearly
some weird infelicity.  OK, moving swiftly on, can we coerce our
dataframe to a numeric matrix?  Yes, we can, but it's klunky:

```{r,showextremerank}
st <- matrix(as.numeric(c(as.matrix(skeleton_table))),ncol=4)
rownames(st) <- rownames(skeleton_table)
colnames(st) <- colnames(skeleton_table)
st
st[is.na(st)] <- 999
```

Taking Maier as an example (a mid-range competitor) we want to
identify competitors who beat Maier at least once, and competitors who
Maier beats at least once (these groups are not necessarily distinct)

```{r}
bm1 <- st[,1] < st["Maier",1]  # beats Maier in run 1
bm2 <- st[,2] < st["Maier",2]  # beats Maier in run 2
bm3 <- st[,3] < st["Maier",3]  # beats Maier in run 3
bm4 <- st[,4] < st["Maier",4]  # beats Maier in run 4
bm1 | bm2 | bm3 | bm4
st[bm1 | bm2 | bm3 | bm4,]
```

Above we see all competitors who beat `Maier` at least once.  Who is
the worst of these, as measured by his worst performance?  This would
be `Weston`, who ranked 14 in run 1, compared to `Maier` who ranked
15.  Idiomatically:

```{r}
jj <- st[bm1 | bm2 | bm3 | bm4,]
which(jj==max(jj),arr.ind=TRUE)
rownames(jj)[which(jj==max(jj),arr.ind=TRUE)[1]]
```

We can automate this process as follows:

```{r}
wf <- function(p){
 beats <- function(run){st[,run] < st[p,run]}
 jj <- st[apply(sapply(seq_len(ncol(st)),beats),1,any),]
 rownames(jj)[which(jj==max(jj),arr.ind=TRUE)[1]]
}
wf("Maier")
```

To check:

```{r}
wf("Zheng")
```

Above we see that `Dukurs_M` beats `Zheng` in run 7, but `Dukurs_M`
ranked only 13 in run 4.

```{r}
loop <- c("Grotheer")
for(i in 1:20){loop <- c(loop,wf(loop[length(loop)]))}
loop
```

Above we see an endless repetition of `Blaser` -> `Timmings` -> `Blaser`.

```{r}
loop <- c("Jungk")
for(i in 1:20){loop <- c(loop,wf(loop[length(loop)]))}
loop
```

```{r}
loop <- c("Wengang")
for(i in 1:20){loop <- c(loop,wf(loop[length(loop)]))}
loop
```

```{r}
forw <- function(start){
  loop <- start
  while(all(table(loop)==1)){ loop <- c(loop,wf(loop[length(loop)])) }
  return(loop)
}
forw("Wengang")
```

We can do the process in reverse:

```{r}
fw <- function(p){
 beatenby <- function(run){st[,run] > st[p,run]}
 jj <- st[apply(sapply(seq_len(ncol(st)),beatenby),1,any),]
 rownames(jj)[which(jj==min(jj),arr.ind=TRUE)[1]]
}
fw("Maier")
```

Above, we see that `Dukurs_M` is the best (ranking 3 in run 2) of
competitors who lose to `Maier` [11 vs 13 in run 4].

```{r}
back <- function(start){
  loop <- start
  while(all(table(loop)==1)){ loop <- c(loop,fw(loop[length(loop)])) }
  return(loop)
}
back("Wengang")
```

```{r}
gd <- graph(c(1,2, 2,3, 1,2, 2,4, 1,4, 5,5, 3,6))
plot(gd)
path <- function(a){c(t(cbind(a[-length(a)],a[-1])))}

gpath <- NULL
jjf <- sapply(rownames(st),forw)
jjb <- sapply(rownames(st),back)
for(i in seq_along(jjf)){gpath <- c(gpath,path(jjf[[i]]))}
for(i in seq_along(jjb)){gpath <- c(gpath,rev(path(jjb[[i]])))}
gpath <- substr(gpath,1,3)
plot(graph(gpath[1:20]))
plot(graph(gpath[1:40]))
plot(graph(gpath[1:80]))
```

# Rank table analysis {-}

We can present the result table in another way:

```{r showranktable}
wikitable_to_ranktable(skeleton_table)
```

Above we see, in the first column, that Grotheer came first ("`c1`")
in runs 1-3 and Wengang came first in run 4.  From the first column we
see that in run 1, Grotheer came fist, Tretyakov came second, Wengang
came third, and so on.

```{r,label=processskeleton,cache=TRUE}
skeleton <- ordertable2supp(skeleton_table)
skeleton_maxp <- maxp(skeleton)
skeleton_maxp
pie(skeleton_maxp)
```

And we can test the null that the competitors all have the same
Plackett-Luce strengths:

```{r,label=testskeleton,cache=TRUE}
equalp.test(skeleton)
```

We see very strong evidence for different competitor strengths, at
$p=2\times 10^{-33}$.  How strong is the evidence that Grotheer is
better than Jungk?

```{r,label=testgrotheerandjungk,cache=TRUE}
samep.test(skeleton,c("Grotheer","Jungk"))
```

### Package dataset {-}

Following lines create `skeleton.rda`, which is not part of the CRAN version of the package.

```{r,label=savecurlingdataset}
save(skeleton_table,skeleton,skeleton_maxp,file="skeleton.rda")
```

### 2022-12-02 inst/skeleton.Rmd
typos

@@
 is to convert `skeleton_table` to a numeric matrix, which proves to be
-a very awkward process
-
-NB: `data.matrix()` fucks things up:
+a very awkward process.  NB: `data.matrix()` buggers things up:
 
@@
 Above, we see that Grotheer apparently came 17th in run 4, clearly
-some weird infelicity.  OK, moving swiftly on, can we coerce our
-dataframe to a numeric matrix?  Yes, we can, but it's klunky:
+some weird infelicity.  OK, how to coerce our dataframe to a numeric
+matrix?  Possible but _extremely_  klunky:
 
@@
 Taking Maier as an example (a mid-range competitor) we want to
-identify competitors who beat Maier at least once, and competitors who
-Maier beats at least once (these groups are not necessarily distinct)
+identify competitors who Maier beats at least once:
 
 ```{r}
-bm1 <- st[,1] < st["Maier",1]  # beats Maier in run 1
-bm2 <- st[,2] < st["Maier",2]  # beats Maier in run 2
-bm3 <- st[,3] < st["Maier",3]  # beats Maier in run 3
-bm4 <- st[,4] < st["Maier",4]  # beats Maier in run 4
-bm1 | bm2 | bm3 | bm4
+bm1 <- st[,1] > st["Maier",1]  # lost to Maier in run 1
+bm2 <- st[,2] > st["Maier",2]  # lost to Maier in run 2
+bm3 <- st[,3] > st["Maier",3]  # lost to Maier in run 3
+bm4 <- st[,4] > st["Maier",4]  # lsot to Maier in run 4
 st[bm1 | bm2 | bm3 | bm4,]
@@
 
-Above we see all competitors who beat `Maier` at least once.  Who is
-the worst of these, as measured by his worst performance?  This would
-be `Weston`, who ranked 14 in run 1, compared to `Maier` who ranked
-15.  Idiomatically:
+Above we see a subset of rows of  `skeleton_table` corresponding to
+competitors who lost to Maier.
 
 ```{r}
-jj <- st[bm1 | bm2 | bm3 | bm4,]
-which(jj==max(jj),arr.ind=TRUE)
-rownames(jj)[which(jj==max(jj),arr.ind=TRUE)[1]]
+rownames(st[bm1 | bm2 | bm3 | bm4,])
+jj <- cbind(won="Maier",lost=rownames(st[bm1 | bm2 | bm3 | bm4,]))
+noquote(jj)
+plot(graph(c(t(jj))))
 ```
 
-We can automate this process as follows:
+Above we see a directed graph, the arrow means "beats", so for example
+Maier beats Timmings (2 o'clock).  We can functionify the calculation
+and define `beats()`:
 
 ```{r}
-wf <- function(p){
- beats <- function(run){st[,run] < st[p,run]}
- jj <- st[apply(sapply(seq_len(ncol(st)),beats),1,any),]
- rownames(jj)[which(jj==max(jj),arr.ind=TRUE)[1]]
+beats <- function(p){   # returns everyone whom competitor 'p' beats
+  bm1 <- st[,1] > st[p,1]  
+  bm2 <- st[,2] > st[p,2]  
+  bm3 <- st[,3] > st[p,3] 
+  bm4 <- st[,4] > st[p,4] 
+  jj <- cbind(p,rownames(st[bm1 | bm2 | bm3 | bm4,]))
+  return(c(t(jj)))
 }
-wf("Maier")
+bb <- c()
+for(i in rownames(st)){bb <- c(bb,beats(i))}
+graph(bb)
+plot(graph(bb))
 ```
 
-To check:
+Above we see a directed graph.  The aim was to detect dominated sets,
+that is, a subset of competitors who are always beaten by everyone
+outside the group.  But although there is no such group in
+`skeleton_table`, this fact is not immmediately obvious from the
+graph.  We can however consider just the top and bottom few
+competitors and see if that makes a difference.
 
 ```{r}
-wf("Zheng")
+(st2 <- st[-(4:19),])  # exclude a bunch of mid-range competitors
 ```
 
-Above we see that `Dukurs_M` beats `Zheng` in run 7, but `Dukurs_M`
-ranked only 13 in run 4.
+Above, it is pretty clear that there is a dominated set of everyone
+except the top 3.
 
 ```{r}
-loop <- c("Grotheer")
-for(i in 1:20){loop <- c(loop,wf(loop[length(loop)]))}
-loop
+bb <- c()
+for(i in rownames(st2)){bb <- c(bb,beats(i))}
+graph(bb)
+plot(graph(bb),layout = layout.fruchterman.reingold)
 ```
 
-Above we see an endless repetition of `Blaser` -> `Timmings` -> `Blaser`.
-
-```{r}
-loop <- c("Jungk")
-for(i in 1:20){loop <- c(loop,wf(loop[length(loop)]))}
-loop
-```
-
-```{r}
-loop <- c("Wengang")
-for(i in 1:20){loop <- c(loop,wf(loop[length(loop)]))}
-loop
-```
-
-```{r}
-forw <- function(start){
-  loop <- start
-  while(all(table(loop)==1)){ loop <- c(loop,wf(loop[length(loop)])) }
-  return(loop)
-}
-forw("Wengang")
-```
-
-We can do the process in reverse:
-
-```{r}
-fw <- function(p){
- beatenby <- function(run){st[,run] > st[p,run]}
- jj <- st[apply(sapply(seq_len(ncol(st)),beatenby),1,any),]
- rownames(jj)[which(jj==min(jj),arr.ind=TRUE)[1]]
-}
-fw("Maier")
-```
-
-Above, we see that `Dukurs_M` is the best (ranking 3 in run 2) of
-competitors who lose to `Maier` [11 vs 13 in run 4].
-
-```{r}
-back <- function(start){
-  loop <- start
-  while(all(table(loop)==1)){ loop <- c(loop,fw(loop[length(loop)])) }
-  return(loop)
-}
-back("Wengang")
-```
-
-```{r}
-gd <- graph(c(1,2, 2,3, 1,2, 2,4, 1,4, 5,5, 3,6))
-plot(gd)
-path <- function(a){c(t(cbind(a[-length(a)],a[-1])))}
-
-gpath <- NULL
-jjf <- sapply(rownames(st),forw)
-jjb <- sapply(rownames(st),back)
-for(i in seq_along(jjf)){gpath <- c(gpath,path(jjf[[i]]))}
-for(i in seq_along(jjb)){gpath <- c(gpath,rev(path(jjb[[i]])))}
-gpath <- substr(gpath,1,3)
-plot(graph(gpath[1:20]))
-plot(graph(gpath[1:40]))
-plot(graph(gpath[1:80]))
-```
+Above we see that the existence of the dominated set is not obvious
+without close scrutiny.  Observing that Grotheer, Jungk, and Wengang
+dominate the other competitors, we can see that these three have
+arrows pointing to each other within the group, but noone outside the
+group has an arrow pointing to any member of the set.  But this is not
+obvious from inspection.
 
@@
 pie(skeleton_maxp)
+dotchart(skeleton_maxp)
+dotchart(log(skeleton_maxp))
 ```
@@
 
+Not very strong.
+
+```{r,label=dgdtert,cache=TRUE}
+samep.test(skeleton,c("Grotheer","Jungk"))$p.value
+```
+
+```{r,label=dgdtkkkert,cache=TRUE}
+p <- rownames(st)
+pvals <- seq_along(p)
+for(i in seq(from=2,to=length(p))){ pvals[i] <- samep.test(skeleton,c(p[1],p[i]))$p.value }
+names(pvals) <- p
+pvals
+```
+
+(NB the above chunk takes about an hour to run, even with `n=1`).
+
+```{r}
+plot(pvals,log='y',pch=16)
+for(i in seq_along(pvals)){text(i,pvals[i],names(pvals)[i],pos=4,col='gray')}
+```
+
 ### Package dataset {-}

### 2022-12-02 inst/distance_matrix_idea.Rmd
new file, distance matrix idea

---
title: "icons: some experimental results"
output: html_document
---

```{r setup, include=FALSE}
set.seed(0)
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("magrittr")
library("ape")
library("phangorn")
options("digits" = 5)
```

<p style="text-align: right;">
![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
</p>

This file, `icons2.Rmd`, presents some speculative experimental work
which is under development.  Do not confuse this file with
`icons.Rmd`, which creates the `icons` objects in the package.

```{r,label=showicons}
icons
icons_maxp <- sort(icons_maxp,decreasing=TRUE)  # sort in decreasing order
icons_maxp
```

Recall `samep.test()`:

```{r,showsameptest,cache=TRUE}
samep.test(icons,c("NB","THC"))
```

I will calculate all pairwise distances between the icons, as measured
by the support difference [penultimate line of the output above, named
`statistic` in the object for consistency with other tests in R]

```{r,label=calculateM,cache=TRUE}
M <- matrix(0,6,6)
p <- names(icons_maxp)
rownames(M) <- p
colnames(M) <- p
for(i in seq_len(6)){
  for(j in seq_len(6)){
    if(i != j){M[i,j] <- samep.test(icons,p[c(i,j)])$statistic}
  }
}
```

```{r}
M
(M <- round(M*100))
```

Observations: the triangle inequality is broken, consider `NB`-`L`
(81), `L`-`PB` (38) and `PB`-`NB` (8)

```{r}
plot(phangorn::upgma(as.dist(M)))
```

### 2022-12-22 inst/cook_repeated.Rmd
use sticker in Rmd
consider ordertable2supp() [it is not correct]

@@
 
+<p style="text-align: right;">
+![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
+</p>
+
+(takes about five minutes without cache).
 Suppose there are five people in a family.  Each day for five days,
@@
 
-```{r}
+```{r lookatfirstrow}
 (day1 <- cooking_table[1,])
@@
 
-```{r}
+```{r uninformativeaboutcomp1}
 p <- c(0.3,0.2,0.1,0.3)
@@
 
-```{r}
+```{r iterateoverrows,cache=TRUE}
 H <- hyper2()
@@
 
+*NOTE*: converting `cooking_table` to an order table is _not_ correct,
+for the zero is interpreted as "did not finish", but the correct
+interpretation is "did not take part": the difference is that DNF
+implies lower trength, but the likelihood function is uninformative
+about DNTP players.  Even so, we may use `ordertable2supp()`:
+
+```{r showtrahnsposeisanordertable,cache=TRUE}
+tc <- t(cooking_table)
+tc[,1:20]
+```
+
 (compare file `curling.Rmd` in which we iterate over columns not rows).
@@
 
-```{r}
+```{r equalptest,cache=TRUE}
 equalp.test(H)
@@
 
-```{r}
+```{r trysimplethings}
 a <- cooking_table # saves typing
@@
 
-```{r}
+```{r tryfishertest}
 fisher.test(jj,alternative="less")
@@
 
-```{r}
+```{r tryfishertestzac}
 a <- cooking_table # saves typing
@@
 
-```{r,cache=TRUE}
+```{r girlsvotinggirls,cache=TRUE}
 cooking_table <-
@@
 
-```{r}
+```{r plotapie}
 mP
@@
 
-```{r,cache=TRUE}
+```{r girlspref, cache=TRUE}
 specificp.gt.test(P,"girl",0)

### 2022-12-22 inst/salad.Rmd
rename columns in the interests of clarity

@@
 
-And second, slick:
+And second, a slick method.  We take the transpose of `salad` (things
+are clearer if we name the rows):
+
+```{r}
+rownames(salad) <-
+   paste("j",formatC(seq_len(nrow(salad)),width=2,format="d",flag="0"),sep="")
+ts <- t(salad)
+ts
+```
+
+Above, we see that object `ts` is an order table.  It is just like the
+formula 1 result table `F1_table_2016`, except that venues are judges
+and drivers are the salads [which are "competing" for the "who has the
+most tart dressing" prize].  Converting this to a support function is
+accomplished with `ordertable2supp()`:
 
 ```{r useslickmethod,cache=TRUE}
-(H2 <- ordertable2supp(t(salad)))
+(H2 <- ordertable2supp(ts))
 ```

### 2022-12-24 inst/surfing_jsa.tex
issue #166

@@
 \usepackage{graphicx}
-\usepackage{natbib}
+\usepackage[comma]{natbib}
 \usepackage{diagbox}
@@
 
-Figure~\ref{dotchartmax} shows the maximum likelihood estimate for the
+Fig.~\ref{dotchartmax} shows the maximum likelihood estimate for the
 23 competitors' strengths, which appears to show a wide range from
@@
 surfing ability.  Observe that the competitors' points (that is, the
-final column of table~\ref{resultstable}) cannot be used for this type
+final column of Table~\ref{resultstable}) cannot be used for this type
 of probabilistic analysis: the points awarded are arbitrary and
@@
 one would expect high points totals to be associated with high
-likelihoods.  Figure~\ref{compare_likelihood_points} shows a loose
+likelihoods.  Fig.~\ref{compare_likelihood_points} shows a loose
 correlation ($R^2=0.71$; $p=4\times 10^{-7}$) and
-Figure~\ref{compare_likelihood_points_rankings} shows how the rankings
+Fig.~\ref{compare_likelihood_points_rankings} shows how the rankings
 differ when calculated by the points and likelihood systems.  Note the
@@
 reject the proposal that the strength of this entity exceeds about
-0.015 (Figure~\ref{brazilian_wavetype_support}): at this value of the
+0.015 (Fig.~\ref{brazilian_wavetype_support}): at this value of the
 strength or greater, we may gain two or more units of support by
@@
 we revert to the standard Plackett-Luce likelihood).  Note that all
-six likelihood functions in table~\ref{likelihoodcompatriot} are
+six likelihood functions in Table~\ref{likelihoodcompatriot} are
 informative about $p_1,p_2,q$ in addition to being informative about
@@
 \begin{table}[h]
-\begin{tabular}{l|l}
+\begin{tabular}{ll}
 order  & ${\mathcal L}(p_1,p_2,q,p_C)$\\ \hline
-$p_1\succ p_2\succ q $ & $ \frac{p_1}{p_1+p_2+q+p_C}\cdot\frac{p_2^{\vphantom{h}}}{p_2+q+p_C}$\\
+$p_1\succ p_2\succ q $ \hphantom{www}
+                       & $ \frac{p_1}{p_1+p_2+q+p_C}\cdot\frac{p_2^{\vphantom{h}}}{p_2+q+p_C}$\\
 $p_2\succ p_1\succ q $ & $ \frac{p_2}{p_1+p_2+q+p_C}\cdot\frac{p_1^{\vphantom{h}}}{p_1+q+p_C}$\\
@@
 \begin{table}[h]
-\begin{tabular}{l|l|l}
+\begin{tabular}{lll}
 wave type  & $\operatorname{Prob}(p_1\succ p_2)$ & $\operatorname{Prob}(p_2\succ p_1^{\vphantom{h}})$\\ \hline
@@
 \begin{table}[h]
-  \begin{tabular}{l|l|l|l}\\
+  \begin{tabular}{llll}\\
     \backslashbox{order}{value}          & B & F & S\\ \hline

### 2023-01-05 inst/multiset_hyper2.Rmd
label all chunks
reflect change to how rp() works, now fillup() not needed

@@
 ```{r,exactingtest,cache=TRUE}
-rs <- fillup(rp(100,dirichlet(c(a=1,b=1,c=1)))) # random strengths
+rs <- rp(100,dirichlet(c(a=1,b=1,c=1))) # random strengths
 colnames(rs) <- letters[1:3]
@@
 
-```{r}
+```{r,plotdifferences}
 hist(x-mean(x))

### 2023-01-05 inst/multiset_hyper2.Rmd
clarifications and rewording

@@
 ---
-title: "multiset_hyper2"
+title: "The `cheering3()` probability model for order statistics: a
+verification with the `multiset()` function of the `partitions` package"
 output: html_document
@@
 ```{r defineM}
-M <- apply(multiset(c(1,1,2,2,3)),2,function(x){letters[x]})
+M <- apply(partitions::multiset(c(1,1,2,2,3)),2,function(x){letters[x]})
 colnames(M) <- rep("",ncol(M))   # tidier and more compact appearance
@@
 ${5\choose2\,2\,1}=30$ columns, one per possible result.  We will
-calculate the likelihood of each column and calculate their sum:
+calculate the likelihood of each column and sum them to get the
+probability of observing any one of the 30 order statistics [the
+observations are mutually exclusive, so
+$\operatorname{Prob}\left(x_1\cup\cdots\cup x_n\right) =
+\operatorname{Prob}\left(x_1\right)+\cdots
++\operatorname{Prob}\left(x_n\right)$]:
 
@@
 $p_X$ means "the Bradley Terry strength of $X$''.  The setup here is
-that $\mathbf{do}$ and $\mathbf{re}$ are twins, and that are twins
-(and $\mathbf{sol}$ is a singleton).  We have that the Bradley-Terry
-stregth of $\mathbf{do}$ and $\mathbf{re}$ is `a`, the strength of
-$\mathbf{me}$ and $\mathbf{fa}$ is `b`, and the strength of
-$\mathbf{sol}$ is `c`.
+that $\mathbf{do}$ and $\mathbf{re}$ are twins, and that $\mathbf{me}$
+and $\mathbf{fa}$ are twins (and $\mathbf{sol}$ is a singleton).  We
+have that the Bradley-Terry strength of $\mathbf{do}$ and
+$\mathbf{re}$ is `a`, the strength of $\mathbf{me}$ and $\mathbf{fa}$
+is `b`, and the strength of $\mathbf{sol}$ is `c`.
 
-Consider, for example, the first column of `M` which is `a c b b a`.
-In theory, this means the competitor who came first had strength `a`,
-the competitor who came second had strength `c`.  So the order
-statistic is one of
+Consider, for example, the column of `M` reading `a c b b a`.  In
+theory, this means the competitor who came first had strength `a`, the
+competitor who came second had strength `c`.  So the order statistic
+is one of the $2!2!1!=4$ following orders:
 
@@
 
-heAbove we see that $\mathbf{do}$ and $\mathbf{re}$ come first and last
+Above we see that $\mathbf{do}$ and $\mathbf{re}$ come first and last
 (or last and first), and $\mathbf{me}$ and $\mathbf{fa}$ come third

### 2023-01-05 inst/salad.Rmd
use samep.test()

@@
 
-From row 3, for exampl,e we see that salad `A` was ranked as second
+From row 3, for example, we see that salad `A` was ranked as second
 most tart, `B` the most tart, `C` the third most tart, and `D` the
@@
 
+We may use `hyper2` to test some hypotheses.  Firstly, $H_0\colon
+p_B=\frac{1}{4}$, which we test with `specificp.test()`:
+
+```{r specificptestB,cache=TRUE}
+specificp.test(H2,"B",1/4)
+```
+
+Above we see a $p$-value of about $10^{-7}$, clearly significant.
+However, we may also test the hypothesis that $p_B$ is less than the
+second strongest, $p_C$:
+
+```{r sameptestBC,cache=TRUE}
+samep.test(H2,c("B","C"))
+```
+
+We see a $p$-value of about $7\times 10^{-4}$, much higher than the
+value from `specificp.test()` of $H_0\colon p_B=\frac{1}{4}$.  This is
+because `samep.test()` allows one to change the value of $p_C$ and we
+would expect a less significant result.
+
 # References

### 2023-01-06 inst/plackett_luce_monster.Rmd
rewording

@@
 ---
-title: "plackett_luce_monster"
+title: "A generalization of Plackett-Luce likelihoods to incorporate non-independence of competitors"
 output: html_document
@@
 
-%
-
-We will calculate a Plackett-Luce likelihood function for four runners
-$\left\lbrace 1,2,3,4\right\rbrace$ but with a reified Bradley-Terry
-monster that corresponds to having subsets of mutual friends who
-support one another.  For example, we might have runners 1 and 2 being
-friends, and runners 3 and 4 being friends: we would represent this as
-$\left\lbrace\left\lbrace 1,2\right\rbrace,\left\lbrace
-3,4\right\rbrace\right\rbrace$.
-
-There seem to be a few distinct statistical models for nonindependence:
+Here I modify standard Plackett-Luce likelihood functions to include
+an element of non-independence among the competitors.  There seem to
+be a few distinct statistical models for nonindependence:
 
@@
   $i$ ahead of $j$.
+
+We will calculate a Plackett-Luce likelihood function for four runners
+$\left\lbrace 1,2,3,4\right\rbrace$ but with a reified Bradley-Terry
+monster that corresponds to having subsets of mutual friends who
+support one another.  For example, we might have runners 1 and 2 being
+friends, and runners 3 and 4 being friends: we would represent this as
+$\left\lbrace\left\lbrace 1,2\right\rbrace,\left\lbrace
+3,4\right\rbrace\right\rbrace$.
 

### 2023-01-06 inst/plackett_luce_monster.Rmd
whitespace

@@
 f <- function (a,b,M){
-return(factorial(3)*factorial(2)*(
-(a)/(a+a+a+b+b)*(a+M)/(a+a+b+b+M)*(a+M)/(a+b+b+M)*(b)/(b+b+M)+
-(a)/(a+a+a+b+b)*(a+M)/(a+a+b+b+M)*(b)/(a+b+b+M)*(a)/(a+b+M)+
-(a)/(a+a+a+b+b)*(a+M)/(a+a+b+b+M)*(b)/(a+b+b+M)*(b+M)/(b+a+M)+
-(a)/(a+a+a+b+b)*(b)  /(a+a+b+b+M)*(a)/(a+a+b+M)*(a+M)/(a+b+M)+
-(a)/(a+a+a+b+b)*(b)  /(a+a+b+b+M)*(a)/(a+a+b+M)*(b)/(b+a+M)+
-(a)/(a+a+a+b+b)*(b)  /(a+a+b+b+M)*(b+M)/(b+a+a+M)*(a)/(a+a+M)+
-(b)/(a+a+a+b+b)*(a)  /(a+a+a+b+M)*(a+M)/(a+a+b+M)*(a+M)/(a+b+M)+
-(b)/(a+a+a+b+b)*(a)  /(a+a+a+b+M)*(a+M)/(a+a+b+M)*(b)/(a+b+M)+
-(b)/(a+a+a+b+b)*(a)  /(a+a+a+b+M)*(b)/(b+a+a)*(a)/(a+a)+
-(b)/(a+a+a+b+b)*(b+M)/(a+a+a+b+M)*(a)/(a+a+a)*(a)/(a+a)))
+return(
+  factorial(3)*factorial(2)*(
+   a/(a+a+a+b+b)*(a+M)/(a+a+b+b+M)*(a+M)/(a+b+b+M)*(b  )/(b+b+M) +
+   a/(a+a+a+b+b)*(a+M)/(a+a+b+b+M)*(b  )/(a+b+b+M)*(a  )/(a+b+M) +
+   a/(a+a+a+b+b)*(a+M)/(a+a+b+b+M)*(b  )/(a+b+b+M)*(b+M)/(b+a+M) +
+   a/(a+a+a+b+b)*(b  )/(a+a+b+b+M)*(a  )/(a+a+b+M)*(a+M)/(a+b+M) +
+   a/(a+a+a+b+b)*(b  )/(a+a+b+b+M)*(a  )/(a+a+b+M)*(b  )/(b+a+M) +
+   a/(a+a+a+b+b)*(b  )/(a+a+b+b+M)*(b+M)/(b+a+a+M)*(a  )/(a+a+M) +
+   b/(a+a+a+b+b)*(a  )/(a+a+a+b+M)*(a+M)/(a+a+b+M)*(a+M)/(a+b+M) +
+   b/(a+a+a+b+b)*(a  )/(a+a+a+b+M)*(a+M)/(a+a+b+M)*(b  )/(a+b+M) +
+   b/(a+a+a+b+b)*(a  )/(a+a+a+b+M)*(b  )/(b+a+a  )*(a  )/(a+a  ) +
+   b/(a+a+a+b+b)*(b+M)/(a+a+a+b+M)*(a  )/(a+a+a  )*(a  )/(a+a  )
+   ) 
+   )
 }

### 2023-01-06 inst/plackett_luce_monster.Rmd
clarification and partial correction

@@
   his rank is established he might be able to help $j$ in some way.
-  He might cheer his still-running friend.  In general, we can
-  consider $n$ ordered pairs $(i_r,j_r)$, $1\leq r\leq n$, with runner
-  $i_1$ helping runner $j_1$, runner $i_2$ helping $j_2$, and so on.
-  One simple case would be equivalence classes of
-  $[n]=\{1,2,\ldots,n\}$ with each equivalence class comprising
-  mutually supporting runners.  In this model, with a pair of mutually
-  supporting runners only the higher-placed competitor can affect the
-  other.  Each equivalence class might have a different supportive
-  proclivity.
+  He might cheer his still-running friend, spurring him to run faster
+  and augmenting his competitive strength.
+
+* In general, we can consider $n$ ordered pairs $(i_r,j_r)$, $1\leq
+  r\leq n$, with runner $i_1$ helping runner $j_1$, runner $i_2$
+  helping $j_2$, and so on.  One simple case would be equivalence
+  classes of $[n]=\{1,2,\ldots,n\}$ with each equivalence class
+  comprising mutually supporting runners.  In this model, with a pair
+  of mutually supporting runners only the higher-placed competitor can
+  affect the other.  Each equivalence class might have a different
+  supportive proclivity.
 

### 2023-01-06 inst/plackett_luce_monster.Rmd
better (start simple!) examples

@@
   $i$ ahead of $j$.
+
+# Start simple
+
+The simplest nontrivial case is three runners, $\left\lbrace
+1,2,3\right\rbrace$ where runner 1 helps runner 2 [and that is the
+only non PL interaction].  Note that this is _not_ a symmetric
+interaction: 1 helps 2, but 2 does not help 1.  The model is that 1
+helps 2 if and only if 1 has finished and 2 has not.
+
+123: $\frac{p_1}{p_1+p_2+p_3}\cdot\frac{\lambda p_2}{\lambda p_2+p_3}$
+
+132: $\frac{p_1}{p_1+p_2+p_3}\cdot\frac{p_3}{\lambda p_2+p_3}$
+
+213: $\frac{p_2}{p_1+p_2+p_3}\cdot\frac{p_1}{p_1+p_3}$
+
+231: $\frac{p_2}{p_1+p_2+p_3}\cdot\frac{p_1}{p_1+p_3}$
+
+312: $\frac{p_3}{p_1+p_2+p_3}\cdot\frac{p_1}{p_1+p_2}$
+
+321: $\frac{p_3}{p_1+p_2+p_3}\cdot\frac{p_2}{p_1+p_2}$
+
+(Above, the probabilities clearly sum to 1).  Now the same but with
+four runners $\left\lbrace 1,2,3\right\rbrace$, again with runner 1
+helping runner 2.  Thus, in a situation where $p_1$ has finished and
+$p_2$ is still running, we have $p_2\longrightarrow\lambda p_2$
+[although the $\lambda$ cancels if 2 is the only competitor still
+running, that is, if *all* the remaining competitors are helped]
+
+1234: $\frac{p_1}{p_1+p_2+p_3+p_4}\cdot\frac{\lambda p_2}{\lambda p_2+p_3+p_4}\cdot\frac{p_3}{p_3+p_4}$
+
+1243: $\frac{p_1}{p_1+p_2+p_3+p_4}\cdot\frac{\lambda p_2}{\lambda p_2+p_3+p_4}\cdot\frac{p_4}{p_3+p_4}$
+
+1324: $\frac{p_1}{p_1+p_2+p_3+p_4}\cdot\frac{p_3}{\lambda p_2+p_3+p_4}\cdot\frac{\lambda p_2}{\lambda p_2+p_4}$
+
+1342: $\frac{p_1}{p_1+p_2+p_3+p_4}\cdot\frac{p_3}{\lambda p_2+p_3+p_4}\cdot\frac{p_4}{\lambda p_2+p_4}$
+
+1324: $\frac{p_1}{p_1+p_2+p_3+p_4}\cdot\frac{p_3}{\lambda p_2+p_3+p_4}\cdot\frac{\lambda p_2}{\lambda p_2+p_4}$
+
+1423: $\frac{p_1}{p_1+p_2+p_3+p_4}\cdot\frac{p_4}{p_1+\lambda p_2+p_3+p_4}\cdot\frac{\lambda p_2}{\lambda p_2+p_3}$
+
+1432: $\frac{p_1}{p_1+p_2+p_3+p_4}\cdot\frac{p_4}{p_1+\lambda p_2+p_3+p_4}\cdot\frac{p_3}{\lambda p_2+p_3}$
+
+2xxx: P-L [no help given by 2 to 1]
+
+3124: $\frac{p_3}{p_1+p_2+p_3+p_4}\cdot\frac{p_1}{p_1+p_2+p_4}\cdot\frac{\lambda p_2}{\lambda p_2+p_4}$
+
+3142: $\frac{p_3}{p_1+p_2+p_3+p_4}\cdot\frac{p_1}{p_1+p_2+p_4}\cdot\frac{p_4}{\lambda p_2+p_4}$
+
+32xx: P-L
+
+3412: P-L [1 helps 2 but when 1 crosses the finishing line, only 2 is
+still running; there are no non-helped competitors]
+
+4123: $\frac{p_4}{p_1+p_2+p_3+p_4}\cdot\frac{p_1}{p_1+p_2+p_3}\cdot\frac{\lambda p_2}{\lambda p_2+p_3}$
+
+4132: $\frac{p_4}{p_1+p_2+p_3+p_4}\cdot\frac{p_1}{p_1+p_2+p_3}\cdot\frac{p_3}{\lambda p_2+p_3}$
+
+42xx: P-L
+
+43xx: P-L
+
+# Next simplest: 1 and 2 mutually support each other:
+
+123: $\frac{p_1}{p_1+p_2+p_3}\cdot\frac{\lambda 2}{\lambda p_2+p_3}$
+
+132: $\frac{p_1}{p_1+p_2+p_3}\cdot\frac{p_3}{\lambda p_2+p_3}$
+
+213: $\frac{p_2}{p_1+p_2+p_3}\cdot\frac{\lambda p_1}{p_1+p_3}$
+
+231: $\frac{p_2}{p_1+p_2+p_3}\cdot\frac{p_3}{\lambda p_1+p_3}$
+
+312: $\frac{p_3}{p_1+p_2+p_3}\cdot\frac{p_1}{p_1+p_2}$
+
+321: $\frac{p_3}{p_1+p_2+p_3}\cdot\frac{p_2}{p_1+p_2}$
+
+# Four runners
 
@@
 
-Might be better to consider ${5\choose 2\,3}$ arrangements of $aaabb\ldots bbaaa$.
+Might be better to consider ${5\choose 2\,3}$ arrangements of
+$aaabb\ldots bbaaa$.  FOLLOWING IS NOT RIGHT: THE PROBABILITY MODEL IS
+NOT PRECISELY DEFINED.
 

### 2023-01-08 inst/plackett_luce_monster.Rmd

@@
 
-# Start simple
+# Start simple: three runners
 
 The simplest nontrivial case is three runners, $\left\lbrace
-1,2,3\right\rbrace$ where runner 1 helps runner 2 [and that is the
-only non PL interaction].  Note that this is _not_ a symmetric
-interaction: 1 helps 2, but 2 does not help 1.  The model is that 1
-helps 2 if and only if 1 has finished and 2 has not.
+1,2,3\right\rbrace$ where runner 1 helps runner 2 with multiplier
+$\lambda$ [and that is the only non PL interaction].  Note that this
+is _not_ a symmetric interaction: 1 helps 2, but 2 does not help 1.
+The model is that 1 helps 2 if and only if 1 has finished and 2 has
+not.
 
@@
 
-231: $\frac{p_2}{p_1+p_2+p_3}\cdot\frac{p_1}{p_1+p_3}$
+231: $\frac{p_2}{p_1+p_2+p_3}\cdot\frac{p_3}{p_1+p_3}$
 
@@
 
-(Above, the probabilities clearly sum to 1).  Now the same but with
-four runners $\left\lbrace 1,2,3\right\rbrace$, again with runner 1
-helping runner 2.  Thus, in a situation where $p_1$ has finished and
-$p_2$ is still running, we have $p_2\longrightarrow\lambda p_2$
-[although the $\lambda$ cancels if 2 is the only competitor still
-running, that is, if *all* the remaining competitors are helped]
+(Above, the probabilities clearly sum to 1).  Next simplest: 1 and 2
+mutually support each other with $\lambda$:
+
+123: $\frac{p_1}{p_1+p_2+p_3}\cdot\frac{\lambda p_2}{\lambda p_2+p_3}$
+
+132: $\frac{p_1}{p_1+p_2+p_3}\cdot\frac{p_3}{\lambda p_2+p_3}$
+
+213: $\frac{p_2}{p_1+p_2+p_3}\cdot\frac{\lambda p_1}{p_1+p_3}$
+
+231: $\frac{p_2}{p_1+p_2+p_3}\cdot\frac{p_3}{\lambda p_1+p_3}$
+
+312: $\frac{p_3}{p_1+p_2+p_3}\cdot\frac{p_1}{p_1+p_2}$
+
+321: $\frac{p_3}{p_1+p_2+p_3}\cdot\frac{p_2}{p_1+p_2}$
+
+Now we retain the mutual support between 1 and 2 but allow an
+asymmetry: 1 helps 2 with $\lambda$ as before, but now in addition, 2
+helps 1 with $\mu$:
+
+123: $\frac{p_1}{p_1+p_2+p_3}\cdot\frac{\lambda p_2}{\lambda p_2+p_3}$
+
+132: $\frac{p_1}{p_1+p_2+p_3}\cdot\frac{p_3}{\lambda p_2+p_3}$
+
+213: $\frac{p_2}{p_1+p_2+p_3}\cdot\frac{\mu p_1}{\mu p_1+p_3}$
+
+231: $\frac{p_2}{p_1+p_2+p_3}\cdot\frac{p_3}{\mu p_1+p_3}$
+
+312: $\frac{p_3}{p_1+p_2+p_3}\cdot\frac{p_1}{p_1+p_2}$
+
+321: $\frac{p_3}{p_1+p_2+p_3}\cdot\frac{p_2}{p_1+p_2}$
+
+Now how about 1 must beat 2.  He does not care whether he comes first
+or second, only that he must beat 2.  Player 1 has sufficient energy
+to win if he wants, but is subject to some sort of resource constraint
+(money, perhaps) and wishes to expend only enough money to beat 2.  So
+conventional P-L likelihoods are modified for 1, but only if 2 is
+_still running_.
+
+123: $\frac{\lambda p_1}{\lambda p_1+p_2+p_3}\cdot\frac{p_2}{p_2+p_3}$
+
+132: $\frac{\lambda p_1}{\lambda p_1+p_2+p_3}\cdot\frac{p_3}{p_2+p_3}$
+
+213: $\frac{p_2}{\lambda p_1+p_2+p_3}\cdot\frac{p_1}{p_1+p_3}$
+
+231: $\frac{p_2}{\lambda p_1+p_2+p_3}\cdot\frac{p_3}{p_1+p_3}$
+
+312: $\frac{p_3}{\lambda p_1+p_2+p_3}\cdot\frac{\lambda p_1}{\lambda p_1+p_2}$
+
+321: $\frac{p_3}{\lambda p_1+p_2+p_3}\cdot\frac{p_2}{\lambda p_1+p_2}$
+
+## Four runners
+
+Now we have four runners $\left\lbrace 1,2,3,4\right\rbrace$, again
+with the only interaction being runner 1 helping runner 2.  Thus, in a
+situation where $p_1$ has finished and $p_2$ is still running, we have
+$p_2\longrightarrow\lambda p_2$ [although the $\lambda$ cancels if 2
+is the only competitor still running, that is, if *all* the remaining
+competitors are helped]
 
@@
 
-# Next simplest: 1 and 2 mutually support each other:
-
-123: $\frac{p_1}{p_1+p_2+p_3}\cdot\frac{\lambda 2}{\lambda p_2+p_3}$
-
-132: $\frac{p_1}{p_1+p_2+p_3}\cdot\frac{p_3}{\lambda p_2+p_3}$
-
-213: $\frac{p_2}{p_1+p_2+p_3}\cdot\frac{\lambda p_1}{p_1+p_3}$
-
-231: $\frac{p_2}{p_1+p_2+p_3}\cdot\frac{p_3}{\lambda p_1+p_3}$
-
-312: $\frac{p_3}{p_1+p_2+p_3}\cdot\frac{p_1}{p_1+p_2}$
-
-321: $\frac{p_3}{p_1+p_2+p_3}\cdot\frac{p_2}{p_1+p_2}$
-
-# Four runners
+Now 1 and 2 mutually support one another by $\lambda$.  Here I will
+put the last term in, to show explicitly that the $\lambda$ term
+remains in the expressions for numerator and denominator.
+
+1234: $\frac{p_1}{p_1+p_2+p_3+p_4}\cdot\frac{\lambda p_2}{\lambda
+p_2+p_3+p_4}\cdot\frac{p_3}{p_3+p_4}\cdot\frac{p_4}{p_4}$
+
+1243: $\frac{p_1}{p_1+p_2+p_3+p_4}\cdot\frac{\lambda p_2}{\lambda
+p_2+p_3+p_4}\cdot\frac{p_4}{p_3+p_4}\cdot\frac{p_3}{p_3}$
+
+1324: $\frac{p_1}{p_1+p_2+p_3+p_4}\cdot\frac{p_3}{\lambda
+p_2+p_3+p_4}\cdot\frac{\lambda p_2}{\lambda
+p_2+p_4}\cdot\frac{p_4}{p_4}$
+
+1342: $\frac{p_1}{p_1+p_2+p_3+p_4}\cdot\frac{p_3}{\lambda
+p_2+p_3+p_4}\cdot\frac{p_4}{\lambda p_2+p_4}\cdot\frac{\lambda
+p_2}{\lambda p_2}$
+
+1423: $\frac{p_1}{p_1+p_2+p_3+p_4}\cdot\frac{p_4}{\lambda
+p_2+p_3+p_4}\cdot\frac{\lambda p_2}{\lambda
+p_2+p_3}\cdot\frac{p_3}{p_3}$
+
+1432: $\frac{p_1}{p_1+p_2+p_3+p_4}\cdot\frac{p_4}{\lambda
+p_2+p_3+p_4}\cdot\frac{p_3}{\lambda p_2+p_3}\cdot\frac{\lambda
+p_2}{\lambda p_2}$
+
+2134: $\frac{p_2}{p_1+p_2+p_3+p_4}\cdot\frac{\lambda p_1}{\lambda
+p_1+p_3+p_4}\cdot\frac{p_3}{p_3+p_4}\cdot\frac{p_4}{p_4}$
+
+2143: $\frac{p_2}{p_1+p_2+p_3+p_4}\cdot\frac{\lambda p_1}{\lambda
+p_1+p_3+p_4}\cdot\frac{p_4}{p_3+p_4}\cdot\frac{p_3}{p_3}$
+
+2314: $\frac{p_2}{p_1+p_2+p_3+p_4}\cdot\frac{\lambda p_3}{\lambda
+p_1+p_3+p_4}\cdot\frac{\lambda p_1}{\lambda
+p_1+p_4}\cdot\frac{p_4}{p_4}$
+
+2341: $\frac{p_2}{p_1+p_2+p_3+p_4}\cdot\frac{\lambda p_3}{\lambda
+p_1+p_3+p_4}\cdot\frac{p_4}{\lambda p_1+p_4}\cdot\frac{\lambda
+p_1}{\lambda p_1}$
 

### 2023-01-10 inst/formula1_top9.Rmd
Fixes #91

@@
 	print(o)
-	L2017 %<>% order_obs(o)
+	L2017  <- L2017 + ordervec2supp(o) # NB zero  -> "Did not finish"
 }  # for(i) loop closes

### 2023-01-15 inst/NBA.Rmd
fixes #73

@@
 
-    Barnes     = 0.150357987445096,   Curry       = 1.65443013016506e-06,
-    Durant     = 0.00856905879575095, Green       = 0.0253069705861388,
-    Iguodala   = 0.00916478949994613, Irving      = 0.0290988580288644, 
-    James      = 1.0000039853319e-06, Jefferson   = 0.00446265583190174, 
-    Korver     = 0.00959220615914446, Livingston  = 0.0775529739242179, 
-    Love       = 0.00253529174567821, McCaw       = 0.0189848298815197,
-    Pachiuila  = 0.0365345555046955,
-    possession = 0.431812626360354,
-    Shumpert   = 0.0556862911754396,   Smith      = 0.00064258717253104,
-    Thompson   = 6.86257295623038e-05, Thompson.1 = 0.00824420466131726,
-    West       = 0.0761409792438058,   Williams   = 0.0552418538199202)
+Barnes    = 0.165210241492775,    Curry      = 1.11314508611164e-06,
+Durant    = 0.00538619616133955,  Green      = 0.0110738194130432,
+Iguodala  = 1.0005765074368e-06,  Irving     = 0.0198810257932021,
+James     = 1.00000582238398e-06, Jefferson  = 0.00990927936847305,
+Korver    = 0.00843806151505123,  Livingston = 0.075108213128189,
+Love      = 0.00175976502052894,  McCaw      = 0.0119305671842558,
+Pachiuila = 0.0496281731046277,   possession = 0.479366441714251,
+Shumpert  = 0.0666196978946451,   Smith      = 1.0008827880116e-06,
+Thompson  = 0.00159564073991825,  Thompson.1 = 0.00135198197477646,
+West      = 0.0927348818403838,   Williams   = 1.89904433611776e-06)
+
 NBA_maxp <- maxp(NBA)
@@
 
-Dotchart(NBA_maxp,col=c(rep("black",9),rep("red",10),rep("blue",1)),pch=16)
+dotchart(NBA_maxp,col=c(rep("black",9),rep("red",10),rep("blue",1)),pch=16)
 ```

### 2023-01-15 R/tests.R

@@ -358,7 +358,7 @@
         jj <- paste(paste("p_",i," = ",sep=""),collapse="")
         null_hypothesis <- substr(jj,1,nchar(jj)-3)
     }
-    SMALL <- 1e-7
+    SMALL <- 1e-6
 
     m_alternative <- maxp(H, startp=startp, ..., give=TRUE)  # free optimization
     alternative_support <- m_alternative$value

### 2023-01-17 inst/home_advantage.Rmd
new ref
typo

@@
 
-We may apply `hyper3` idiom to the dataset. First a formal test:
+We may apply `hyper3` idiom to the dataset, using the likelihood
+function of Davidson and Beaver (1977).  First a formal test:
 
@@
 
+###  References
+
+* R. R. Davidson and R. J. Beaver 1977. "On extending the
+  Bradley-Terry model to incorporate within-pair order effects"
+  _Biometrics_, 33:693--702
+
 ### Package dataset {-}
 
-Following lines create `home_advantage.rda`, residing in the `data/`
+Following lines create `baseball.rda`, residing in the `data/`
 directory of the package.

### 2023-01-18 inst/monster_vs_lambda.Rmd
monster v lambda

---
title: "Monster or Lambda?  A likelihood analysis"
author: "Robin KS Hankin"
output: html_document
---

```{r setup, include=FALSE}
set.seed(1)
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
```

(takes about an hour to process without cache)

<p style="text-align: right;">
![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
</p>

Here I consider pairwise comparisons with a home-ground advantage and
wish to compare two likelihood functions.  First, the monster parametrization:

$$
\mathcal{L_M}\left(p_1,p_2,M\right)=
\left(
\frac{p_1+M;p_2}{p_1+p_2+M}
\right)^{a;b}\cdot
\left(
\frac{p_1;p_2+M}{p_1+p_2+M}
\right)^{c;d}
$$

[here the first expression is for $p_1$ at home and $p_2$ away; the
second is $p_1$ away and $p_2$ at home; $a$ is number of home wins for
$p_1$, $b$ number of away wins for $p_2$, $c$ number of away wins for
$p_1$ and $d$ number of home wins for $p_2$].   Secondly the monster
parameterization:

$$
\mathcal{L_\lambda}\left(p_1,p_2,\lambda\right)=
\left(
\frac{\lambda p_1;p_2}{\lambda p_1+p_2}
\right)^{a;b}\cdot
\left(
\frac{p_1;\lambda p_2}{p_1+\lambda p_2}
\right)^{c;d}
$$

A suitable generalized function would be

$$
\mathcal{L}\left(p_1,p_2,\lambda,M\right)=
\left(
\frac{\lambda p_1+M;p_2}{\lambda p_1+p_2+M}
\right)^{a;b}\cdot
\left(
\frac{p_1;\lambda p_2+M}{p_1+\lambda p_2+M}
\right)^{c;d}
$$

First we need to define a slight generalization of `home_away3()`:

```{r}
`ha3` <- function(home_games_won,away_games_won,lambda){

    if(is.complex(home_games_won)){
        if(missing(lambda)){lambda <- away_games_won}
        away_games_won <- Im(home_games_won)
        home_games_won <- Re(home_games_won) 
    }

    teams <- rownames(home_games_won)
    stopifnot(identical(teams,colnames(home_games_won)))
    stopifnot(identical(teams,rownames(away_games_won)))
    stopifnot(identical(teams,colnames(away_games_won)))

    H <- hyper3(pnames=c(teams,"M","lambda"))

    for(i in seq_len(nrow(home_games_won))){
        for(j in seq_len(ncol(home_games_won))){
            if(i != j){  # diagonal means a team plays itself, meaningless.
                ## First deal with home_games:
                game_winner <- teams[i]
                game_loser  <- teams[j]
                no_of_matches <- home_games_won[i,j] # won with home strength helping

                jj <- c(lambda,1)
                names(jj) <- c(game_winner,"M")
                H[jj] %<>% inc(no_of_matches)  # won with home strength helping

                jj <- c(lambda,1,1)
                names(jj) <- c(game_winner,game_loser,"M")
                H[jj] %<>% dec(no_of_matches)

                ## now away games:
                no_of_matches <- away_games_won[i,j] # won without the benefit of home strength
                jj <- 1
                names(jj) <- game_winner
                H[jj] %<>% inc(no_of_matches)  # won without home ground helping...

                jj <- c(1,lambda,1)
                names(jj) <- c(game_winner,game_loser,"M")
                H[jj] %<>% dec(no_of_matches)  # home strength nevertheless on denominator

            } # if(i != j) closes
        } # j loop closes
    } # i loop closes
    return(H)
}
```

(maybe one day I will include this in the `hyper2` package).  Copying
from `home_advantage.Rmd`:

```{r label=definehomeawaymatrix}
D  <- matrix(c(
    NA, 16, 12, 11, 19, NA, 19, 16,
    17, 12, NA, 11, 11, 12, 12, NA),
    nrow=4,ncol=4,byrow=TRUE) + 1i* matrix(c(
    NA, 05, 02, 02, 02, NA, 10, 02,
    00, 04, NA, 07, 02, 06, 04, NA),
    nrow=4,ncol=4,byrow=TRUE)

teams <- LETTERS[1:4]
dimnames(D) <- list("@home" = teams,"@away"=teams)
D
```

```{r,tryha3,cache=TRUE}
H <- ha3(D,lambda=1.88)
s <- specificp.gt.test(H,"M",0,n=1)  # NB lambda = 1.88
s
```

```{r}
s$null_support
```

Above but in one line:

```{r,label=trytwovalues,cache=TRUE}
ll <- function(lambda){specificp.gt.test(ha3(D,lambda=lambda),"M",0,n=1)$null_support}
ll(1.22)
ll(1.88)
```

```{r,label=calclam,cache=TRUE}
lam <- seq(from=2.5,to=6,len=15)
supp <- sapply(lam,ll)
```

```{r}
plot(lam,supp-max(supp),type='b')
```

### 2023-01-24 inst/home_advantage.Rmd
show complex form of same dataset for reference
fixes #196

@@
 Here I analyse a small synthetic dataset for home ground advantage,
-then go on to discuss a dataset presented by Agresti.  First generate
-the dataset by hand:
+then go on to discuss a dataset presented by Agresti.  But first, the
+smallest nontrivial example is a 2x2 case:
+
+```{r,label=smallestexample}
+M <- matrix(c(
+    NA, 1+5i, 3+11i, NA),
+    nrow=2,ncol=2,byrow=TRUE)
+
+teams <- LETTERS[seq_len(2)]
+dimnames(M) <- list("@home" = teams,"@away"=teams)
+dimnames(M) <- list("@home" = teams,"@away"=teams)
+M
+home_away(M)
+```
+
+Above we see (from `M[1,2]`) that A played at home $1+5=6$ times,
+winning 1 and losing 5; and B played at home $3+11=14$ times, winning
+3 and losing 11.   A slightly more complicated example:
 
@@
     NA, 05, 02, 02,
-    02, NA, 10, 02,
-    00, 04, NA, 07,
-    02, 06, 04, NA),
+     9, NA, 10, 02,
+     3, 04, NA, 07,
+     8, 06, 04, NA),
     nrow=4,ncol=4,byrow=TRUE)
@@
 times and lost 2 times; and B played at home 19+5=24 times and won 19
-times and lost 5 times.
+times and lost 5 times.  Alternatively we may use complex numbers to
+represent the same dataset:
+
+```{r,label=showhomeawayusingcomplex}
+home_games_won + 1i*away_games_won
+```
 
@@
 H
+options("use_alabama" = FALSE) # to avoid the stupid wmmin bug
+specificp.gt.test(H,"home",0)
 options("use_alabama" = TRUE) # to avoid the stupid wmmin bug
-specificp.gt.test(H,"home",0)
 ```
@@
 
-```{r,label=showvisH}
-dotchart(maxp(H))
+```{r,label=showvisH,cache=TRUE}
+(mt <- maxp(H))
 ```
@@
 ```{r,label=showprofilelikelihood, cache=TRUE}
-home_strength <- seq(from=0.2,to=0.6,length=20)
+home_strength <- seq(from=0.15,to=0.45,length=13)
 plot(home_strength,profsupp(H,"home",home_strength),type="b",pch=16,
@@
 ```{r,label=baseballtransplot,cache=TRUE}
-ordertransplot(baseball_maxp[1:7],maxp(baseball)[1:7],xlab="old",ylab="new",plotlims=c(0.1,0.16))
+ordertransplot(baseball_maxp[1:7],maxp(baseball)[1:7],xlab="old",ylab="new",plotlims=c(0.01,0.16))
 ```

### 2023-01-24 inst/monster_vs_lambda.Rmd
fixes #196

@@
 
-```{r}
-`ha3` <- function(home_games_won,away_games_won,lambda){
+```{r defineha3}
+`ha3` <- function(home_games_won,away_games_won,lambda){  # based on home_away3()
 
@@
 
-    H <- hyper3(pnames=c(teams,"M","lambda"))
+    H <- hyper3(pnames=c(teams,"lambda","M"))
 
@@
         for(j in seq_len(ncol(home_games_won))){
-            if(i != j){  # diagonal means a team plays itself, meaningless.
-                ## First deal with home_games:
-                game_winner <- teams[i]
-                game_loser  <- teams[j]
-                no_of_matches <- home_games_won[i,j] # won with home strength helping
+            if(i != j){  
+                home_team <- teams[i]
+                away_team <- teams[j]
 
+		home_wins <- home_games_won[i,j]
+		away_wins <- away_games_won[i,j] 
+                no_of_matches <- home_wins + away_wins
+
+                ## home wins:
                 jj <- c(lambda,1)
-                names(jj) <- c(game_winner,"M")
-                H[jj] %<>% inc(no_of_matches)  # won with home strength helping
+                names(jj) <- c(home_team,"M")
+                H[jj] %<>% inc(home_wins)
 
+                ## away wins:
+                jj <- 1
+                names(jj) <- away_team
+                H[jj] %<>% inc(away_wins)
+
+                ## denominator
                 jj <- c(lambda,1,1)
-                names(jj) <- c(game_winner,game_loser,"M")
+                names(jj) <- c(home_team,away_team,"M")
                 H[jj] %<>% dec(no_of_matches)
-
-                ## now away games:
-                no_of_matches <- away_games_won[i,j] # won without the benefit of home strength
-                jj <- 1
-                names(jj) <- game_winner
-                H[jj] %<>% inc(no_of_matches)  # won without home ground helping...
-
-                jj <- c(1,lambda,1)
-                names(jj) <- c(game_winner,game_loser,"M")
-                H[jj] %<>% dec(no_of_matches)  # home strength nevertheless on denominator
-
             } # if(i != j) closes
@@
 
-(maybe one day I will include this in the `hyper2` package).  Copying
-from `home_advantage.Rmd`:
+We will try on a simple case:
+
+```{r,label=try2x2}
+M <- matrix(c(
+    NA, 1+5i, 3+11i, NA),
+    nrow=2,ncol=2,byrow=TRUE)
+
+teams <- LETTERS[seq_len(2)]
+dimnames(M) <- list("@home" = teams,"@away"=teams)
+dimnames(M) <- list("@home" = teams,"@away"=teams)
+M
+home_away(M)
+ha3(M,lambda=1.88)
+```
+
+(maybe one day I will include this function in the `hyper2` package).
+Copying from `home_advantage.Rmd`:
 
 ```{r label=definehomeawaymatrix}
-D  <- matrix(c(
-    NA, 16, 12, 11, 19, NA, 19, 16,
-    17, 12, NA, 11, 11, 12, 12, NA),
-    nrow=4,ncol=4,byrow=TRUE) + 1i* matrix(c(
-    NA, 05, 02, 02, 02, NA, 10, 02,
-    00, 04, NA, 07, 02, 06, 04, NA),
+home_games_won <- matrix(c(
+    NA, 16, 12, 11,
+    19, NA, 19, 16,
+    17, 12, NA, 11,
+    11, 12, 12, NA),
+    nrow=4,ncol=4,byrow=TRUE)
+
+away_games_won <- matrix(c(
+    NA, 05, 02, 02,
+     9, NA, 10, 02,
+     3, 04, NA, 07,
+     8, 06, 04, NA),
     nrow=4,ncol=4,byrow=TRUE)
@@
 teams <- LETTERS[1:4]
-dimnames(D) <- list("@home" = teams,"@away"=teams)
+dimnames(home_games_won) <- list("@home" = teams,"@away"=teams)
+dimnames(away_games_won) <- list("@home" = teams,"@away"=teams)
+
+D <- home_games_won + 1i*away_games_won
 D
@@
 
+Above we see, from row 1 column 2 [`36+10i`], that A played at home
+against B a total of $36+10=46$ times, winning $36$ and losing $10$;
+From row 2 column 1 [`11+10i`] we see that B played at home against A
+a total of 21 times, winning 11 and losing 10.  We can use the Monster
+likelihood function and test $H_0\colon M=0$:
+
 ```{r,tryha3,cache=TRUE}
-H <- ha3(D,lambda=1.88)
-s <- specificp.gt.test(H,"M",0,n=1)  # NB lambda = 1.88
+H <- ha3(D,lambda=1.01)
+print(H)
+s <- specificp.gt.test(H,"M",0,n=10)  # NB lambda = 1.88
 s
 ```
+
+We see strong evidence that $M>0$.
 
@@
 
-Above but in one line:
+We want to plot a profile log-likelihood function, that is, a
+likelihood function corresponding to the evaluate constrained to
+having a particular value of $M$:
 
-```{r,label=trytwovalues,cache=TRUE}
-ll <- function(lambda){specificp.gt.test(ha3(D,lambda=lambda),"M",0,n=1)$null_support}
-ll(1.22)
-ll(1.88)
+```{r,label=marginallambdaprofilelike,cache=TRUE}
+f <- function(lambda){
+ write(date(),file="marginallambdaprofilelike",append=TRUE)  # so we can check its progress
+ return(maxp(ha3(D,lambda=lambda),give=TRUE)$value)
+}
+
+lam <- seq(from=1.2,to=4.6,len=15)
+supp_M_eq_0 <- sapply(lam,f)
 ```
 
-```{r,label=calclam,cache=TRUE}
-lam <- seq(from=2.5,to=6,len=15)
-supp <- sapply(lam,ll)
+```{r,plotprofilelikem0}
+plot(lam,supp_M_eq_0,type="b",xlab="lambda")
 ```
 
-```{r}
-plot(lam,supp-max(supp),type='b')
+Now enforce the constraint $\lambda=1$ [and allow $M$ to vary],
+thereby reducing the likelihood function from a `hyper3` to a `hyper2`
+object:
+
+```{r,tryiconprofsupp,cache=TRUE}
+(H <- home_away(D))  # observe that the monster is 'home'
+home <- seq(from=0.15,to=0.4,len=15)
+supp_lambda_eq_0 <- profsupp(H,"home",home,relative=FALSE)
+plot(home,supp_lambda_eq_0 - max(supp_lambda_eq_0),type='b')
+abline(h=c(0,-2))
 ```
+
+```{r,showmaxplam1,cache=TRUE}
+maxp(H)
+```
+
+```{r,label=plotordinarysupport}
+plot(lam,supp_M_eq_0-max(supp_M_eq_0),type='b',main='Profile likelihood with M=0')
+grid()
+plot(home,supp_lambda_eq_0 - max(supp_lambda_eq_0),type='b')
+```
+
+```{r,label=plotbothsupports}
+both <- c(supp_lambda_eq_0,NA,supp_M_eq_0)
+plot(both-max(both,na.rm=TRUE),type='b',col=c(rep("black",length(supp_lambda_eq_0)),"green",rep("red",length(supp_M_eq_0))))
+abline(h=c(0,-2))
+```
+
+Support difference:
+
+```{r,label=supportdifference}
+max(supp_M_eq_0)-max(supp_lambda_eq_0)
+```
+
+exceeding Edwards's two-units-of-support criterion by a comfortable
+margin.  Note that both have equal numbers of degrees of freedom.
+
+# Create a new dataset by direct simulation
+
+```{r,syntheticdataP}
+set.seed(0)
+P <- matrix(0,4,4)
+teams <- LETTERS[1:4]
+dimnames(P) <- list("@home" = teams,"@away"=teams)
+P
+n <- 2000 # number of matches between a pair 
+M <- 0.0005  # home monster
+strengths <- 1/(1:4) # Zipf
+strengths <- strengths/sum(strengths)
+names(strengths) <- teams
+strengths
+for(home_team in 1:4){
+  for(away_team in 1:4){
+    if(home_team != away_team){
+      prob_home_win <- (strengths[home_team] + M)/(strengths[home_team] + strengths[away_team] + M)
+      home_wins <- rbinom(1,n,prob_home_win)
+      home_losses <- n - home_wins
+      P[home_team,away_team] <- home_wins + 1i*home_losses
+    }
+  } 
+}
+diag(P) <- NA
+P
+HP <- home_away(P)
+```
+
+```{r,label=showlike}
+HP
+```
+
+```{r,label=maxpmonsteqq,cache=TRUE}
+mH <- maxp(HP,n=3,give=T)
+mH
+```

### 2023-01-24 R/hyper2.R

@@ -1223,38 +1223,34 @@ setGeneric("pnames<-",function(x,value){standardGeneric("pnames<-")})
     dirichlet(rowSums(M)) - hyper2(apply(index,1,function(x){rownames(M)[x]},simplify=FALSE),jj[index])
 }
 
-`home_away` <- function(home_games_won,away_games_won){
-    if(is.complex(home_games_won)){
+`home_away` <- function (home_games_won, away_games_won){
+    if (is.complex(home_games_won)) {
         away_games_won <- Im(home_games_won)
-        home_games_won <- Re(home_games_won) # note, away_games_won is assigned first
+        home_games_won <- Re(home_games_won)
     }
-    teams <- rownames(home_games_won)
-    stopifnot(identical(teams,colnames(home_games_won)))
-    stopifnot(identical(teams,rownames(away_games_won)))
-    stopifnot(identical(teams,colnames(away_games_won)))
-    teams <- c(teams,"home")
     
-    H <- hyper2(pnames=teams)
+    teams <- rownames(home_games_won)
+    stopifnot(identical(teams, colnames(home_games_won)))
+    stopifnot(identical(teams, rownames(away_games_won)))
+    stopifnot(identical(teams, colnames(away_games_won)))
+    teams <- c(teams, "home")
+    H <- hyper2(pnames = teams)
+    for (i in seq_len(nrow(home_games_won))) {
+        for (j in seq_len(ncol(home_games_won))) {
+            if (i != j) {
+                home_team <- teams[i]
+                away_team <- teams[j]
 
-    for(i in seq_len(nrow(home_games_won))){
-        for(j in seq_len(ncol(home_games_won))){
-            if(i != j){  # diagonal means a team plays itself, meaningless.
-                ## First deal with home_games:
-                game_winner <- teams[i]
-                game_loser  <- teams[j]
-                no_of_matches <- home_games_won[i,j] # won with home strength helping
-                H[c(game_winner,           "home")] %<>% inc(no_of_matches)  # won with home strength helping
-                H[c(game_winner,game_loser,"home")] %<>% dec(no_of_matches)
-
-                ## now away games:
-                no_of_matches <- away_games_won[i,j] # won without the benefit of home strength
-                H[c(game_winner                  )] %<>% inc(no_of_matches)  # won without home ground helping...
-                H[c(game_winner,game_loser,"home")] %<>% dec(no_of_matches)  # home strength nevertheless on denominator
[…]

### 2023-01-24 R/hyper3.R

@@ -523,31 +523,28 @@ stop("not yet written")
 
     for(i in seq_len(nrow(home_games_won))){
         for(j in seq_len(ncol(home_games_won))){
-            if(i != j){  # diagonal means a team plays itself, meaningless.
-                ## First deal with home_games:
-                game_winner <- teams[i]
-                game_loser  <- teams[j]
-                no_of_matches <- home_games_won[i,j] # won with home strength helping
+            if(i != j){  
+                home_team <- teams[i]
+                away_team <- teams[j]
 
+		home_wins <- home_games_won[i,j]
+		away_wins <- away_games_won[i,j] 
+                no_of_matches <- home_wins + away_wins 
+
+                ## home wins:
                 jj <- lambda
-                names(jj) <- game_winner
-                H[jj] %<>% inc(no_of_matches)  # won with home strength helping
-
-                jj <- c(lambda,1)
-                names(jj) <- c(game_winner,game_loser)
-                H[jj] %<>% dec(no_of_matches)
-
-
-                ## now away games:
-                no_of_matches <- away_games_won[i,j] # won without the benefit of home strength
+                names(jj) <- home_team
+                H[jj] %<>% inc(home_wins)
+                
+                ## away wins:
                 jj <- 1
-                names(jj) <- game_winner
-                H[jj] %<>% inc(no_of_matches)  # won without home ground helping...
+                names(jj) <- away_team
+                H[jj] %<>% inc(away_wins)
 
-                jj <- c(1,lambda)
-                names(jj) <- c(game_winner,game_loser)
-                H[jj] %<>% dec(no_of_matches)  # home strength nevertheless on denominator
-                                
+                ## denominator
+                jj <- c(lambda,1)
+                names(jj) <- c(home_team,away_team)
+                H[jj] %<>% dec(no_of_matches)
             } # if(i != j) closes
         } # j loop closes
     } # i loop closes

### 2023-01-25 inst/monster_vs_lambda.Rmd
better parameter ranges

@@
 print(H)
+```
+
+```{r,maxha3,cache=TRUE}
+maxp(H)
+```
+
+```{r,testha3,cache=TRUE}
 s <- specificp.gt.test(H,"M",0,n=10)  # NB lambda = 1.88
@@
 ```{r,label=marginallambdaprofilelike,cache=TRUE}
+write(date(),file="marginallambdaprofilelike",append=FALSE)
 f <- function(lambda){
  write(date(),file="marginallambdaprofilelike",append=TRUE)  # so we can check its progress
- return(maxp(ha3(D,lambda=lambda),give=TRUE)$value)
+ return(maxp(ha3(D,lambda=lambda),n=1,give=TRUE)$value)
 }
 
-lam <- seq(from=1.2,to=4.6,len=15)
+lam <- seq(from=0.1,to=4.2,len=5)
 supp_M_eq_0 <- sapply(lam,f)
@@
 plot(lam,supp_M_eq_0,type="b",xlab="lambda")
+plot(lam,supp_M_eq_0-max(supp_M_eq_0),type="b",xlab="lambda")
+grid()
+abline(h=c(0,-2))
 ```
@@
 
-```{r,syntheticdataP}
+```{r,syntheticdatapp}
 set.seed(0)
@@
 P
-n <- 2000 # number of matches between a pair 
-M <- 0.0005  # home monster
+n <- 400 # number of matches between a pair 
+M <- 0.15  # home monster
 strengths <- 1/(1:4) # Zipf
@@
 
-```{r,label=maxpmonsteqq,cache=TRUE}
+```{r,label=maxpmonsteq1,cache=TRUE}
 mH <- maxp(HP,n=3,give=T)
@@
 ```
+
+```{r proflikesyntha,cache=TRUE}
+home <- seq(from=0.12,to=0.16,len=15)
+supp_lambda_eq_0 <- profsupp(HP,"home",home,relative=FALSE)
+```
+
+```{r plotsynthprof}
+plot(home,supp_lambda_eq_0,type='b')
+plot(home,supp_lambda_eq_0 - max(supp_lambda_eq_0),type='b')
+abline(h=c(0,-2))
+
+```{r marginalsynthcalct,cache=TRUE}
+f <- function(lambda){
+ return(maxp(ha3(P,lambda=lambda),n=1,give=TRUE)$value)
+}
+
+lam <- seq(from=0.5,to=1.3,len=7)
+supp_M_eq_0 <- sapply(lam,f)
+```
+
+```{r plotsynthma}
+plot(lam,supp_M_eq_0,type='b')
+plot(lam,supp_M_eq_0-max(supp_M_eq_0),type='b')
+abline(h=c(0,-2))
+```
+
+```{r plotbothtogether}
+jj <- c(supp_M_eq_0,NA,supp_lambda_eq_0)
+plot(jj-max(jj,na.rm=TRUE),type="b")
+abline(h=c(0,-2))
+```

### 2023-01-26 inst/monster_vs_lambda.Rmd
reorganisation: start simple

@@
 
-First we need to define a slight generalization of `home_away3()`:
+## Start simple: two teams
+
+```{r,echo=FALSE}
+jj <- matrix(c(NA, 1+2i ,4+5i, NA),2,2,byrow=TRUE)
+teams <- LETTERS[1:2]
+dimnames(jj) <- list("@home" = teams,"@away"=teams)
+dimnames(jj) <- list("@home" = teams,"@away"=teams)
+print(jj)
+```
+
+Above we see that A played at home 3 times, winning 1 and losing 2
+[row 1 column 2 is $1+2i$]; B played at home 9 times, winning 4 and
+losing 5 [row 2 column 1 is $4+5i$].  In its most general form, the
+$2\times 2$ result matrix would be
+
+$$
+\begin{bmatrix}
+-&\alpha+\beta i \\
+\gamma + \delta i&-\\
+\end{bmatrix}
+$$
+
+and the log-likelihood function would be, algebraically,
+
+$$
+\mathcal{S}=\log\mathcal{L}=
+\log\left(\frac{A^\delta\cdot(A + M)^\alpha\cdot B^\beta
+(B + M)^\gamma
+}{
+(A + B + M)^{\alpha+\beta+\gamma+\delta}
+}\right)
+=
+\delta\log A + \alpha\log(1-B) + \beta\log B + \gamma\log(1-A)
+$$
+
+where $M$ is the strength of the home monster `home` and $A+B+M=1$, so
+we understand that $M=1-A-B$.
+
+$$
+\frac{\partial\mathcal{S}}{\partial A}=\frac{\delta}{A}-\frac{\gamma}{1-A}\qquad
+\frac{\partial\mathcal{S}}{\partial B}=\frac{\beta }{B}-\frac{\alpha}{1-A}
+$$
+
+At a maximum, therefore,
+
+$$
+\hat{A}=\frac{\delta}{\delta + \gamma}\qquad
+\hat{B}=\frac{\beta }{\alpha  + \beta}
+$$
+
+Just to check, we evaluate the Hessian matrix and find
+
+$$H=-
+\left(\begin{array}{cc} 
+\frac{\partial^2\mathcal{S}}{\partial A\partial A} & 
+\frac{\partial^2\mathcal{S}}{\partial A\partial B} \\
+\frac{\partial^2\mathcal{S}}{\partial B\partial A} & 
+\frac{\partial^2\mathcal{S}}{\partial b\partial B} 
+\end{array}\right)
+=
+\left(\begin{array}{cc}
+\delta A^{-2} + \gamma(1-A)^{-2}&0\\
+0&\beta B^{-2} + \alpha(1-B)^{-2}
+\end{array}\right)
+$$
+
+which is clearly strictly positive-definite if $0<A,B<1$ and
+$\alpha+\beta>0$, $\gamma+\delta>0$.
+
+We will calculate numerical estimates using the
+monster-parameterization for a simple case:
+
+```{r,label=try2x2}
+M <- matrix(c(
+    NA, 5+1i, 9+2i, NA),
+    nrow=2,ncol=2,byrow=TRUE)
+teams <- LETTERS[seq_len(2)]
+dimnames(M) <- list("@home" = teams,"@away"=teams)
+dimnames(M) <- list("@home" = teams,"@away"=teams)
+M
+(Hsimp <- home_away(M))
+maxp(Hsimp)
+```
+
+Compare the theoretical value of
+
+```{r}
+theoretical <- c(A=2/(2+9),B=1/(1+5))
+fillup(theoretical,Hsimp)
+```
+
+showing good agreement.  We note that the model predicts a probability
+of $\hat{A}/(\hat{A}+\hat{B})=\frac{12}{23}\simeq 0.52$ for $A$
+winning in a match between $A$ and $B$ where the home team advantage
+is not operating (on neutral ground, for example).
+
+Now the $\lambda$ parameterization.  The likelihood function for the
+same dataset is
+
+$$
+\mathcal{S}=\log\mathcal{L}=
+\log\left(
+\frac{(\lambda A)^\alpha B^\beta}{(\lambda A+B)^{\alpha+\beta}}
+\cdot
+\frac{A^\gamma (\lambda B)^\delta}{(A+\lambda B)^{\gamma+\delta}}
+\right)$$
+
+This is equal to $(\alpha+\delta)\log\lambda +(\alpha+\gamma)\log
+A+(\beta+\delta)\log B -(\alpha+\beta)\log(\lambda A+B)
+-(\gamma+\delta)\log(A+\lambda B)$ and even after simplifying [we may
+assume $A+B=1$] the result does not appear to have a simple analytical
+solution; we must proceed numerically.  To do so,
+
+we need to define a
+slight generalization of `home_away3()`:
 
@@
 
-We will try on a simple case:
+(maybe one day I will include function `ha3()` in the `hyper2`
+package).
 
-```{r,label=try2x2}
-M <- matrix(c(
-    NA, 1+5i, 3+11i, NA),
-    nrow=2,ncol=2,byrow=TRUE)
-
-teams <- LETTERS[seq_len(2)]
-dimnames(M) <- list("@home" = teams,"@away"=teams)
-dimnames(M) <- list("@home" = teams,"@away"=teams)
-M
-home_away(M)
-ha3(M,lambda=1.88)
-```
-
-(maybe one day I will include this function in the `hyper2` package).
-Copying from `home_advantage.Rmd`:
+Now a more complicated example, a
+modified version of the 4x4 matrix on `home_advantage.Rmd`:
 
@@
 dimnames(P) <- list("@home" = teams,"@away"=teams)
-P
-n <- 400 # number of matches between a pair 
+n <- 90 # number of matches between a pair 
 M <- 0.15  # home monster
-strengths <- 1/(1:4) # Zipf
+strengths <- (4:1)/10
 strengths <- strengths/sum(strengths)
@@
 
-```{r,label=maxpmonsteq1,cache=TRUE}
+```{r,label=maxpmonstereq,cache=TRUE}
 mH <- maxp(HP,n=3,give=T)
@@
 
-```{r proflikesyntha,cache=TRUE}
+```{r proflikesynthcc,cache=TRUE}
 home <- seq(from=0.12,to=0.16,len=15)
-supp_lambda_eq_0 <- profsupp(HP,"home",home,relative=FALSE)
+supp_lambda_eq_1 <- profsupp(HP,"home",home,relative=FALSE)
 ```
 
-```{r plotsynthprof}
-plot(home,supp_lambda_eq_0,type='b')
-plot(home,supp_lambda_eq_0 - max(supp_lambda_eq_0),type='b')
+```{r label=matchershower}
+jjH <- ha3(P,lambda=1) # should match HP above
+jjH
+as.hyper3(HP) - jjH
+```
+
+```{r plotsynthprofa}
+plot(home,supp_lambda_eq_1,type='b')
+plot(home,supp_lambda_eq_1 - max(supp_lambda_eq_1),type='b')
 abline(h=c(0,-2))
+```
 
-```{r marginalsynthcalct,cache=TRUE}
+```{r showresult,cache=TRUE}
+specificp.gt.test(ha3(P,lambda=1.4),"M",0)
+specificp.gt.test(ha3(P,lambda=1.4),"M",1e-4)
+```
+
+```{r marginalsynthcale,cache=TRUE}
 f <- function(lambda){
- return(maxp(ha3(P,lambda=lambda),n=1,give=TRUE)$value)
+   o <- ha3(P,lambda=lambda)
+   return(specificp.gt.test(o,"M",0)$null_support)
 }
 
-lam <- seq(from=0.5,to=1.3,len=7)
+lam <- seq(from=1.9,to=2.4,len=5)
 supp_M_eq_0 <- sapply(lam,f)
@@
 ```{r plotbothtogether}
-jj <- c(supp_M_eq_0,NA,supp_lambda_eq_0)
+jj <- c(supp_M_eq_0,NA,supp_lambda_eq_1)
 plot(jj-max(jj,na.rm=TRUE),type="b")

### 2023-01-27 inst/home_advantage3.Rmd
consistentification

---
title: "Quantifying the home advantage with `hyper3` objects"
author: "Robin K. S. Hankin"
date: "`r Sys.Date()`"
vignette: |
  %\VignetteIndexEntry{hyper3}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

<p style="text-align: right;">
![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
</p>

Objects of class \code{hyper3} are a generalization of \code{hyper2}
objects that allow the brackets to contain weighted probabilities.
Likelihood functions are defined on non-negative $p_1,\ldots, p_n$
subject to the unit-sum constraint $\sum p_i=1$.  Given known weights
$w^i_j$ with $1\leq i\leq j$ we have

$$\mathcal{L}\left(p_1,\ldots, p_n\right)=\prod_j\left(\sum_{i=1}^n w^i_jp_i\right)^{n_j}.$$

As a motivating example, suppose two teams (players) with
Bradley-Terry strengths $p_1,p_2$ play football where we quantify the
home-ground advantage with a term $\lambda$.  If $p_1$ plays at home
$a+b$ times with $a$ wins and $b$ losses, and plays away [so $p_2$ is
at home] $c+d$ times with $c$ wins and $d$ losses, then a sensible
likelihood function might be

$$\mathcal{L}(p_1,p_2,\lambda;A,B,C,D)=
\left(\frac{\lambda p_1}{\lambda p_1 + p_2}\right)^{A}
\left(\frac{p_2        }{\lambda p_1 + p_2}\right)^{B}
\left(\frac{p_1        }{p_1 + \lambda p_2}\right)^{C}
\left(\frac{\lambda p_2}{p_1 + \lambda p_2}\right)^{D}.
$$

where we understand that $p_1+p_2=1$, $p_1,p_2,\lambda\geqslant 0$.
Elementary techniques allow us to identify a maximum and we find

$$
\hat{p_1}=\frac{\sqrt{AC}}{\sqrt{AC}+\sqrt{BD}}\qquad
\hat{p_2}=\frac{\sqrt{BD}}{\sqrt{AC}+\sqrt{BD}}\qquad
\hat{\lambda}=\sqrt{\frac{AD}{BC}}
$$

```{r label=defineABCD}
A <- 9
B <- 2
C <- 3
D <- 5
```

$$
\hat{p_1}=\frac{\sqrt{27}}{\sqrt{27} + \sqrt{10}}\simeq 0.62\qquad
\hat{p_2}=\frac{\sqrt{15}}{\sqrt{27} + \sqrt{10}}\simeq 0.38\qquad
\hat{\lambda}=\sqrt{\frac{AD}{BC}}\simeq 2.73$$

Keeping $A=x,B=x,C=x,D=xxxxxx$ and $\lambda=2.73$ [assumed for the moment
to be known, estimation techniques are discussed later], appropriate
package idiom might be to use bespoke function `home_away3()`:

```{r,label=usehomeaway3a}
library("hyper2",quietly=TRUE)
M <- matrix(c(NA, A+B*1i ,D+C*1i, NA),2,2,byrow=TRUE)
teams <- c("p1","p2")
dimnames(M) <- list("@home" = teams,"@away"=teams)
dimnames(M) <- list("@home" = teams,"@away"=teams)
print(M)
(H <- home_away3(M,lambda=2.73))
```

Above, obect `M` has real parts being home wins and imaginary parts
being away wins.

We may estimate $p_1$ and $p_2$ using maximum likelihood, `maxp()` in
package idiom:

```{r,label=findtheevaluate3,cache=TRUE}
H
maxp(H)
```

Further, we can test whether the players are in fact of equal strength:

```{r,label=testequalitynow,cache=TRUE}
equalp.test(H)
```

Showing convincingly that we may reject the null that $p_1=p_2$ and
assert that the players do in fact differ in strength, at least with
this value of $\lambda$.  Observe that a naive analysis would have
$p_1$ winning $w=A+C=`r A+C`$ games and losing $l=B+D=`r B+D`$ games,
from which we would obtain $\hat{p_1}=`r A+C`/`r A+B+C+D`$; a
likelihood ratio for $H_0\colon p_1=0.5$ would be

$$\frac
{
  {{w+l}\choose {w\, l}}\left(\frac{w}{w+l}\right)^w\left(\frac{l}{w+l}\right)^l
}{
  {{w+l}\choose {w\, l}}\left(\frac{1}{2}\right)^w\left(\frac{1}{2}\right)^l
}=\frac{2^{w+l}w^wl^l}{(w+l)^{w+l}}\simeq 1.94,
$$

not significant.

Now, how to estimate $\lambda$?

```{r,label=estimatelambda,cache=TRUE}
f <- function(lambda){maxp(home_away3(M,lambda=lambda),give=TRUE)$value}
f(2)
f(2.1)
lam <- seq(from=1,to=10,len=17)
Supp <- sapply(lam,f)
```

```{r,label=plotsup}
plot(log(lam),Supp-max(Supp),type="b")
```

We can proceed in two ways.  The flaw in this analysis is that it is
conditional on the estimated value of $\lambda$ [which is not too bad,
IMO].  Secondly

```{r,label=loopysuppcalc,cache=TRUE}
probs <- seq(from=0.1,to=0.8,len=17)
jj <- as.matrix(expand.grid(lam,probs))
S <- rep(NA,nrow(jj))
for(i in seq_len(nrow(jj))){
    lambda <- jj[i,1]
    p1 <- jj[i,2]
    p <- c(p1,1-p1)
    names(p) <- c("p1","p2")
   S[i] <- loglik(p=p,H=home_away3(M,lambda=lambda),log=TRUE)
}
S <- matrix(S,length(lam),length(probs))
```

```{r}
contour(lam,probs,S,levels=seq(from=-20,to=-10,by=0.5),xlab="lambda",ylab="p1")
abline(h=0.5)
```

### 2023-01-27 inst/hyper3.Rmd
consistentification

---
title: "The weighted hyperdirichlet distribution: the `hyper3` class"
author: "Robin K. S. Hankin"
date: "`r Sys.Date()`"
vignette: |
  %\VignetteIndexEntry{hyper3}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

<p style="text-align: right;">
![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
</p>

Objects of class \code{hyper3} are a generalization of \code{hyper2}
objects that allow the brackets to contain weighted probabilities.
Likelihood functions are defined on non-negative $p_1,\ldots, p_n$
subject to the unit-sum constraint $\sum p_i=1$.  Given known weights
$w^i_j$ with $1\leq i\leq j$ we have

$$\mathcal{L}\left(p_1,\ldots, p_n\right)=\prod_j\left(\sum_{i=1}^n w^i_jp_i\right)^{n_j}.$$

As a motivating example, suppose two teams (players) with
Bradley-Terry strengths $p_1,p_2$ play football where we quantify the
home-ground advantage with a term $\lambda$.  If $p_1$ plays at home
$a+b$ times with $a$ wins and $b$ losses, and plays away [so $p_2$ is
at home] $c+d$ times with $c$ wins and $d$ losses, then a sensible
likelihood function might be

$$\mathcal{L}(p_1,p_2,\lambda;A,B,C,D)=
\left(\frac{\lambda p_1}{\lambda p_1 + p_2}\right)^{A}
\left(\frac{p_2        }{\lambda p_1 + p_2}\right)^{B}
\left(\frac{p_1        }{p_1 + \lambda p_2}\right)^{C}
\left(\frac{\lambda p_2}{p_1 + \lambda p_2}\right)^{D}.
$$

where we understand that $p_1+p_2=1$, $p_1,p_2,\lambda\geqslant 0$.
Elementary techniques allow us to identify a maximum and we find

$$
\hat{p_1}=\frac{\sqrt{AC}}{\sqrt{AC}+\sqrt{BD}}\qquad
\hat{p_2}=\frac{\sqrt{BD}}{\sqrt{AC}+\sqrt{BD}}\qquad
\hat{\lambda}=\sqrt{\frac{AD}{BC}}
$$

If $A=x,B=x,C=x,D=x$ and $\lambda=1.3$ [assumed for the moment to be
known], appropriate package idiom might be:

```{r,label=usenamedvectorsforH}
library("hyper2",quietly=TRUE)
A <- 6 ; B <- 2 ; C <- 7 ; D <- 1
H <- hyper3()
H[c(p1=1.3)]      %<>% inc(A) 
H[c(p2=1)]        %<>% inc(B) 
H[c(p1=1.3,p2=1)] %<>% dec(A+B)
H[c(p1=1)]        %<>% inc(C) 
H[c(p2=1.3)]      %<>% inc(D) 
H[c(p1=1,p2=1.3)] %<>% dec(C+D)
H
```

Bespoke function `home_away3()` achieves the same result with a more
convenient input:

```{r,label=usehomeaway3}
M <- matrix(c(NA, 6+2i ,1+7i, NA),2,2,byrow=TRUE)
teams <- c("p1","p2")
dimnames(M) <- list("@home" = teams,"@away"=teams)
dimnames(M) <- list("@home" = teams,"@away"=teams)
print(M)
home_away3(M,lambda=1.3)
```

Above, obect `M` has real parts being home wins and imaginary parts
being away wins.

We may estimate $p_1$ and $p_2$ using maximum likelihood, `maxp()` in
package idiom:

```{r,label=findtheevaluate,cache=TRUE}
maxp(H)
```

Further, we can test whether the players are in fact of equal strength:

```{r,label=testequality,cache=TRUE}
equalp.test(H)
```

Showing convincingly that we may reject the null that $p_1=p_2$ and
assert that the players do in fact differ in strength.

Another motivating example might come from Plackett-Luce likelihood
functions for order statistics.  Suppose we have three players with
nonnegative strengths $p_\mathrm{M},p_\mathrm{RB},p_\mathrm{F}$, with
unit sum.  These players might be conceputalised as F1 constructors
(here Mercedes (M), Red Bull (RB), and Ferrari (F) for the sake of
argument).  Each constructor fields two cars and the order statistic
might be

\newcommand{\pr}{p_\mathrm{RB}}
\newcommand{\pm}{p_\mathrm{M}}
\newcommand{\pf}{p_\mathrm{F}}

$$
RB\succ M\succ F\succ F\succ RB\succ M$$

indicating that the finishing order was Red Bull first, Mercedes
second, Ferrari third, and so on (this was in fact observed at the
2021 Emilia Romagna Grand Prix). The Placket-Luce likelihood function
would be

$$
\frac{\pr}{\pr+\pm+\pf+\pf+\pr+\pm}\cdot
\frac{\pm}{    \pm+\pf+\pf+\pr+\pm}\cdot
\frac{\pf}{        \pf+\pf+\pr+\pm}\cdot
\frac{\pf}{            \pf+\pr+\pm}\cdot
\frac{\pr}{                \pr+\pm}\cdot
\frac{\pm}{                    \pm}
$$

or, in a form more suitable for the package,

$$
\frac{\pr}{2\pr + 2\pm + 2\pf}\cdot
\frac{\pm}{ \pr + 2\pm + 2\pf}\cdot
\frac{\pf}{ \pr +  \pm + 2\pf}\cdot
\frac{\pf}{ \pr +  \pm  + \pf}\cdot
\frac{\pr}{ \pr  + \pm       }$$

We could proceed as follows:

```{r,label=workouta}
a <- hyper3()
a[c(R=1)] %<>% inc
a[c(R=2,M=2,F=2)] %<>% dec
a[c(M=1)] %<>% inc
a[c(R=1,M=2,F=2)] %<>% dec
a[c(F=1)] %<>% inc
a[c(R=1,M=1,F=2)] %<>% dec
a[c(F=1)] %<>% inc
a[c(R=1,M=1,F=1)] %<>% dec
a[c(R=1)] %<>% inc
a[c(R=1,M=1    )] %<>% dec
a
```

Or maybe more logically

```{r,label=usebetteridiomforb}
b <- hyper3()
b["R"] %<>% inc ; b[c("R","M","F","F","R","M")] %<>% dec
b["M"] %<>% inc ; b[c(    "M","F","F","R","M")] %<>% dec
b["F"] %<>% inc ; b[c(        "F","F","R","M")] %<>% dec
b["F"] %<>% inc ; b[c(            "F","R","M")] %<>% dec
b["R"] %<>% inc ; b[c(                "R","M")] %<>% dec

a==b
```

However, bespoke function `ordervec2supp3()` performs all this in one
step:

```{r, label=showbespoke}
ordervec2supp3(c("RB","M","F","F","RB","M"))
```

And then

```{r, label=maxpa,cache=TRUE}
ma <- maxp(a)
ma
```

Evidence for different strengths?

```{r,aretheydifferent,cache=TRUE}
equalp.test(a)
```

No evidence agains the null of equal strengths. Further:

```{r,label=showgradissmall}
gradient(a,ma)  # should be small at the evaluate
```

### 2023-01-27 inst/home_advantage3.Rmd
further likelihood analysis

@@
 
-We can proceed in two ways.  The flaw in this analysis is that it is
-conditional on the estimated value of $\lambda$ [which is not too bad,
-IMO].  Secondly
+Or even
+
+```{r,label=optimlam,cache=TRUE}
+op <- optimize(f,interval=c(2,3),maximum=TRUE)
+op
+```
+
+We can proceed in two ways.  Firstly we can use $\hat{\lambda}$ directly:
+
+```{r,ept,cache=TRUE}
+equalp.test(home_away3(M,lambda=op$maximum))
+```
+
+The flaw in this analysis is that it is conditional on the estimated
+value of $\lambda$ [which is not too bad, IMO].  Secondly we can
+perform a fully 2D analysis, which allows for the fact that the
+evaluate might have a different value of $\lambda$ from the
+$\hat{\lambda}$:
 

### 2023-01-30 inst/home_advantage3.Rmd
rewrite

@@
 $$\mathcal{L}(p_1,p_2,\lambda;A,B,C,D)=
-\left(\frac{\lambda p_1}{\lambda p_1 + p_2}\right)^{A}
-\left(\frac{p_2        }{\lambda p_1 + p_2}\right)^{B}
-\left(\frac{p_1        }{p_1 + \lambda p_2}\right)^{C}
-\left(\frac{\lambda p_2}{p_1 + \lambda p_2}\right)^{D}.
+\underbrace{
+\overbrace{\left(\frac{\lambda p_1}{\lambda p_1 + p_2}\right)^{A}}^{\mbox{$A$ home wins for $p_1$}}\quad
+\overbrace{\left(\frac{p_2        }{\lambda p_1 + p_2}\right)^{B}}^{\mbox{$B$ away wins for $p_2$}}
+}_\mbox{$p_1$ plays at home to $p_2$ $A+B$ times}\quad
+\underbrace{
+\overbrace{\left(\frac{\lambda p_2}{p_1 + \lambda p_2}\right)^{C}}^{\mbox{$C$ home wins for $p_2$}}\quad
+\overbrace{\left(\frac{        p_1}{p_1 + \lambda p_2}\right)^{D}}^{\mbox{$D$ away wins for $p_1$}}
+}_\mbox{$p_2$ plays at home to $p_1$ $C+D$ times}
 $$
@@
 $$
-\hat{p_1}=\frac{\sqrt{AC}}{\sqrt{AC}+\sqrt{BD}}\qquad
-\hat{p_2}=\frac{\sqrt{BD}}{\sqrt{AC}+\sqrt{BD}}\qquad
-\hat{\lambda}=\sqrt{\frac{AD}{BC}}
+\hat{p_1}=\frac{\sqrt{AD}}{\sqrt{AD}+\sqrt{BC}}\qquad
+\hat{p_2}=\frac{\sqrt{AD}}{\sqrt{AD}+\sqrt{BC}}\qquad
+\hat{\lambda}=\sqrt{\frac{AC}{BD}}
 $$
+
+Note that the evaluate is unchanged if $A,B,C,D$ are all multiplied by
+a positive constant.
+
+## Aside: `hyper2` analysis of the same scoreline
+
+Now the likelihood function would be
+
+$$\mathcal{L}(p_1,p_2,\lambda;A,B,C,D)=
+\underbrace{
+\overbrace{\left(\frac{p_1+M}{p_1 + p_2+M}\right)^{A}}^{\mbox{$A$ home wins for $p_1$}}\quad
+\overbrace{\left(\frac{p_2  }{p_1 + p_2+M}\right)^{B}}^{\mbox{$B$ away wins for $p_2$}}
+}_\mbox{$p_1$ plays at home to $p_2$ $A+B$ times}\quad
+\underbrace{
+\overbrace{\left(\frac{p_2+M}{p_1 + p_2+M}\right)^{C}}^{\mbox{$C$ home wins for $p_2$}}\quad
+\overbrace{\left(\frac{p_1  }{p_1 + p_2+M}\right)^{D}}^{\mbox{$D$ away wins for $p_1$}}
+}_\mbox{$p_2$ plays at home to $p_1$ $C+D$ times}
+$$
+
+The maximum would be
+
+$$
+\hat{p_1}=\frac{C}{C+D}\qquad
+\hat{p_2}=\frac{A}{A+B}\qquad
+M=\frac{AD-BC}{(A+B)(B+C)}
+$$
+
+although it is not clear to me what the _constrained_ optimal point
+[that is, requiring $M\geq 0$] is.
+
+## Numerical analysis
 
@@
 A <- 9
-B <- 2
-C <- 3
-D <- 5
+B <- 5
+C <- 7
+D <- 3
+
+(p1hat <- sqrt(A*D)/(sqrt(A*D) + sqrt(B*C)))
+(p2hat <- sqrt(B*C)/(sqrt(A*D) + sqrt(B*C)))
+
+(lam_hat <- sqrt((A*C)/(B*D)))
 ```
@@
 $$
-\hat{p_1}=\frac{\sqrt{27}}{\sqrt{27} + \sqrt{10}}\simeq 0.62\qquad
-\hat{p_2}=\frac{\sqrt{15}}{\sqrt{27} + \sqrt{10}}\simeq 0.38\qquad
-\hat{\lambda}=\sqrt{\frac{AD}{BC}}\simeq 2.73$$
+\hat{p_1}=\frac{\sqrt{21}}{\sqrt{21} + \sqrt{35}}\simeq 0.47\qquad
+\hat{p_2}=\frac{\sqrt{63}}{\sqrt{21} + \sqrt{35}}\simeq 0.53\qquad
+\hat{\lambda}=\sqrt{\frac{AD}{BC}}=2.05.
+$$
 
-Keeping $A=x,B=x,C=x,D=xxxxxx$ and $\lambda=2.73$ [assumed for the moment
-to be known, estimation techniques are discussed later], appropriate
-package idiom might be to use bespoke function `home_away3()`:
+Keeping $A=9,B=5, C=7,D=3$, and $\lambda=2.05$ [assumed for the moment
+to be known, estimation techniques are discussed later], we can
+represent this information in a standard format:
 
@@
 library("hyper2",quietly=TRUE)
-M <- matrix(c(NA, A+B*1i ,D+C*1i, NA),2,2,byrow=TRUE)
+M <- matrix(c(NA, A+B*1i ,C+D*1i, NA),2,2,byrow=TRUE)
 teams <- c("p1","p2")
@@
 print(M)
-(H <- home_away3(M,lambda=2.73))
 ```
 
-Above, obect `M` has real parts being home wins and imaginary parts
-being away wins.
+Above, object `M` has real parts being home wins and imaginary parts
+being away wins.  Appropriate package idiom to specify a likelihood
+function might be to use bespoke function `home_away3()`:
 
-We may estimate $p_1$ and $p_2$ using maximum likelihood, `maxp()` in
-package idiom:
+```{r,label=ha3calc,cache=TRUE}
+(H <- home_away3(M,lambda=2.05))
+```
 
-```{r,label=findtheevaluate3,cache=TRUE}
+Keeping $A=9,B=5, C=7,D=3$, and $\lambda=2.05$ [assumed for the moment
+to be known, estimation techniques are discussed later], we may
+estimate $p_1$ and $p_2$ using maximum likelihood, `maxp()` in package
+idiom:
+
+```{r,label=findtheevaluateh3,cache=TRUE}
 H
@@
 
-```{r,label=estimatelambda,cache=TRUE}
+```{r,label=estimatelambda3,cache=TRUE}
 f <- function(lambda){maxp(home_away3(M,lambda=lambda),give=TRUE)$value}
@@
 
-```{r,label=optimlam,cache=TRUE}
+```{r,label=optimlambda,cache=TRUE}
 op <- optimize(f,interval=c(2,3),maximum=TRUE)
@@
 
-```{r,ept,cache=TRUE}
+```{r,ept3,cache=TRUE}
 equalp.test(home_away3(M,lambda=op$maximum))

### 2023-02-07 inst/j_sports_sci_F1_points.tex
minor journal-specific tweaks
unemphasise tournament theory

--- 2022-11-28 inst/jebo_F1points.tex
@@
 
-\journal{Journal of Economic Behavior and Organization}
+\journal{Journal of Sports Sciences}
 
@@
   \title{An objective assessment of point scoring systems in Formula 1 motor racing}
-  \author{Robin K. S. Hankin\fnref{myfootnote}}
-\address{AUT University, 55 Wellesley Street, Auckland 1010, New Zealand}
+  \author{Robin K. S. Hankin\fnref{fn1}}
+  \address{AUT University, 55 Wellesley Street, Auckland 1010, New Zealand\\
+  }
+
+\fntext[fn1]{\url{https://orcid.org/0000-0001-5982-0415}; {\tt hankin.robin@gmail.com}}
 
@@
 statistical nulls using accumulated points.  Here, I use Plackett-Luce
-likelihood to identify a ranking for the competitors from the
-perspective of tournament theory, this method being amenable to
-statistical testing.  I go on to assess a number of reasonable points
-systems objectively: in one well-defined statistical sense a Borda
-points system is optimal.
+likelihood to identify a ranking for the competitors, this method
+being amenable to statistical testing.  I go on to assess a number of
+reasonable points systems objectively: in one well-defined statistical
+sense a Borda points system is optimal.
 \end{abstract}

### 2023-02-08 inst/home_advantage3.Rmd
typos and mathematica transcript

@@
 \hat{p_1}=\frac{\sqrt{AD}}{\sqrt{AD}+\sqrt{BC}}\qquad
-\hat{p_2}=\frac{\sqrt{AD}}{\sqrt{AD}+\sqrt{BC}}\qquad
+\hat{p_2}=\frac{\sqrt{BC}}{\sqrt{AD}+\sqrt{BC}}\qquad
 \hat{\lambda}=\sqrt{\frac{AC}{BD}}
@@
 \hat{p_1}=\frac{C}{C+D}\qquad
-\hat{p_2}=\frac{A}{A+B}\qquad
-M=\frac{AD-BC}{(A+B)(B+C)}
+\hat{p_2}=\frac{B}{A+B}\qquad
+M=\frac{AC-BD}{(A+B)(C+D)}
 $$
@@
 \hat{p_2}=\frac{\sqrt{63}}{\sqrt{21} + \sqrt{35}}\simeq 0.53\qquad
-\hat{\lambda}=\sqrt{\frac{AD}{BC}}=2.05.
+\hat{\lambda}=\sqrt{\frac{AC}{BD}}=2.05.
 $$
@@
 ```
+
+# Appendix
+
+Here is a transcript of a mathematica session which establishes the
+estimates above.  First the `hyper2` likelihood:
+
+```
+rhankin@rhrpi4:~ $ wolfram
+Mathematica 12.2.0 Kernel for Linux ARM (32-bit)
+Copyright 1988-2021 Wolfram Research, Inc.
+
+In[1]:= L = Log[M+p1]*A +Log[p2]*B- Log[M+p1+p2]*(A+B+C+D) + Log[p2+M]*C + Log[p1]*D
+
+Out[1]= D Log[p1] + A Log[M + p1] + B Log[p2] + C Log[M + p2] - (A + B + C + D) Log[M + p1 + p2]
+
+In[2]:= LL = L /. {p2 -> 1-p1-M}
+
+Out[2]= C Log[1 - p1] + B Log[1 - M - p1] + D Log[p1] + A Log[M + p1]
+
+In[3]:= dp = D[LL,p1]//FullSimplify
+
+            C      D         B          A
+Out[3]=  ------- + -- + ----------- + ------
+         -1 + p1   p1   -1 + M + p1   M + p1
+
+In[4]:= dM = D[LL,M]//FullSimplify
+
+              B          A
+Out[4]= ----------- + ------
+         -1 + M + p1   M + p1
+
+In[5]:= Solve[{dp==0,dM==0},{p1,M}]
+
+                   D            A C - B D
+Out[5]=  {{p1 -> -----, M -> ---------------}}
+                 C + D       (A + B) (C + D)
+
+In[6]:= p2 = 1-p1-M /. {%} //FullSimplify
+
+             B
+Out[6]=  {{-----}}
+           A + B
+
+```
+
+Now the `hyper3` likelihood:
+
+```
+rhankin@rhrpi4:~ $ wolfram
+Mathematica 12.2.0 Kernel for Linux ARM (32-bit)
+Copyright 1988-2021 Wolfram Research, Inc.
+
+In[1]:= L = A*Log[lambda*p1] + B*Log[p2] -(A+B)*Log[lambda*p1 + p2] + C*Log[lambda*p2] + D*Log[p1] -(C+D)*Log[p1 + lambda*p2]
+
+Out[1]= D Log[p1] + A Log[lambda p1] + B Log[p2] + C Log[lambda p2] - (A + B) Log[lambda p1 + p2] - (C + D) Log[p1 + lambda p2]
+
+In[2]:= LL = L /. {p2 -> 1-p1}
+
+Out[2]= B Log[1 - p1] + C Log[lambda (1 - p1)] + D Log[p1] + A Log[lambda p1] - (C + D) Log[lambda (1 - p1) + p1] -
+
+>    (A + B) Log[1 - p1 + lambda p1]
+
+In[3]:= dp1 = D[LL,p1] //FullSimplify
+
+         B + C    A + D   (A + B) (-1 + lambda)    (C + D) (-1 + lambda)
+Out[3]= ------- + ----- - --------------------- + -----------------------
+        -1 + p1    p1     1 + (-1 + lambda) p1    lambda + p1 - lambda p1
+
+In[4]:= dlam = D[LL,lambda]//FullSimplify
+
+          A        C           (A + B) p1           (C + D) (-1 + p1)
+Out[4]= ------ + ------ - -------------------- + -----------------------
+        lambda   lambda   1 + (-1 + lambda) p1   lambda + p1 - lambda p1
+
+In[5]:= Solve[{dp1==0,dlam==0},{p1,lambda}]//FullSimplify
+
+[snip]
+
+In[6]:= %[[1]]
+
+                        1                     Sqrt[A] Sqrt[C]
+Out[6]= {p1 -> -------------------, lambda -> ---------------}
+                   Sqrt[B] Sqrt[C]            Sqrt[B] Sqrt[D]
+               1 + ---------------
+                   Sqrt[A] Sqrt[D]
+
+```

### 2023-02-08 inst/home_advantage3.Rmd
ternary plot

@@
 $$
-\hat{p_1}=\frac{C}{C+D}\qquad
+\hat{p_1}=\frac{D}{C+D}\qquad
 \hat{p_2}=\frac{B}{A+B}\qquad
-M=\frac{AC-BD}{(A+B)(C+D)}
+\hat{M}=\frac{AC-BD}{(A+B)(C+D)}
 $$
@@
 \hat{\lambda}=\sqrt{\frac{AC}{BD}}=2.05.
+$$
+
+Moving to the monster formulation we have
+
+$$
+\hat{p_1}=\frac{3}{7+3} = 0.3\qquad
+\hat{p_2}=\frac{5}{5+9}\simeq  0.36\qquad
+\hat{M}=\frac{63-21}{14\cdot 10} = 0.3
 $$
@@
 
+#  Ternary representation
+
+```{r,label=useternary}
+library(Ternary)
+
+par(mar = rep(0.2, 4))
+TernaryPlot(alab = "p1", blab = "p2", clab = "M")
+
+FunctionToContour <- function(p1,p2,M) {
+pmax(-20,log(M+p1)*A + log(p2)*B - log(M+p1+p2)*(A+B+C+D) + log(p2+M)*C + log(p1)*D )
+}
+
+# Compute and plot colour tiles
+values <- TernaryPointValues(FunctionToContour, resolution = 24L)
+ColourTernary(values)
+
+# Add contour lines
+TernaryContour(FunctionToContour, resolution = 36L)
+```
+
 # Appendix

### 2023-02-09 inst/podium.Rmd
new analysis of F1, podium effect

---
title: "Analysis of pudding following Davidson 1970"
output: bookdown::html_document2
---

```{r setup, include=FALSE}
set.seed(0)
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
options("digits" = 5)
```

<p style="text-align: right;">
![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
</p>

This short document defines the `podium()` function which tests the
hypothesis that a Formula 1 driver's being on the podium at one venue
affects his performance on the next.

```{r,label=definepodiumfun}
podium <- function (x, lambda,print=FALSE){
     noscore <- c("Ret", "WD", "DNS", "DSQ", "DNP", "NC", "DNQ", "EX", "Sick")
     venues <- colnames(x)
     jj <- apply(x, 2, function(y) {
         if (any(y %in% noscore)) {
             y[y %in% noscore] <- 0
         }
         return(y)
     })
     fmat <- matrix(as.numeric(jj), byrow = TRUE, ncol = nrow(x))
     colnames(fmat) <- rownames(x)
     rownames(fmat) <- venues

     o <- fmat[1, , drop = TRUE]
     o[o > 0] <- rank(o[o > 0])
     out <- as.hyper3(ordervec2supp(o))
     for (i in seq(from=2,to=nrow(fmat))){
       if(print){cat(paste(i,"/",nrow(fmat),"\n"))}
       yesterday <- fmat[i-1, , drop = TRUE]
       podium_yesterday <- names(which((yesterday <= 3) & (yesterday>0)))
       print(podium_yesterday)

       d <- fmat[i, , drop = TRUE] # today
       print(d)
       ## Following lines lifted from ordervec2supp()
       nd <- names(d)
       while (any(d > 0)) {
         eligible <- which(d >= 0)
         winner <- nd[d == 1]  # that is, the winner of those still racing
         if(winner %in% podium_yesterday){
           jj <- lambda
         } else {
           jj <- 1
         }
         names(jj) <- winner
         out[jj] %<>% inc    # numerator

         ## Now denominator
         jj <- rep(1,length(eligible))
         jj[nd[eligible] %in% podium_yesterday] <- lambda
         names(jj) <- names(eligible)
         out[jj] %<>% dec  # denominator

         d[d == 1] <- -1
         d[d > 0] %<>% dec
       } # ordervec2supp() lookalike closes
     }  # i loop closes
     return(out)
}
```

Now a dataset:

```{r,load2007}
a <- read.table("formula1_2007.txt",header=TRUE)
a <- a[,seq_len(ncol(a)-1)]
a
```

Define a subset of it:

```{r, definesubset}
b <- a[1:6,1:5]
b[,1] <- 1:6
b[,2] <- c(1,3,4,2,5,6)
b[,3] <- c(5,1,2,3,4,6)

b[,5] <- 1:6
b
```

```{r,usepodium,cache=TRUE}
Ha <- podium(b,1.4)
Hb <- podium(b,1.5)
ma <- maxp(Ha,give=TRUE,n=1)
mb <- maxp(Hb,give=TRUE,n=1)
```

```{r}
Ha
Hb
ma
mb
```

```{r,usefun,cache=TRUE}
fun <- function(lambda){maxp(podium(b,lambda),give=TRUE)$value}
lam <- seq(from=0.5,to=2,len=6)
like <- lapply(lam,fun)
```

```{r, plotlik}
plot(lam,like,type='b')
```

### 2023-02-10 inst/podium.Rmd
rejig of ordertable2supp(), also use it in podium.Rmd

@@
 ---
-title: "Analysis of pudding following Davidson 1970"
+title: "Does being on the podium affect Formula 1 drivers' performance in the next race?"
 output: bookdown::html_document2
@@
 ```{r,label=definepodiumfun}
-podium <- function (x, lambda,print=FALSE){
+podium <- function (lambda,x,print=FALSE){
      noscore <- c("Ret", "WD", "DNS", "DSQ", "DNP", "NC", "DNQ", "EX", "Sick")
@@
        yesterday <- fmat[i-1, , drop = TRUE]
+       yesterday[yesterday > 0] <- rank(yesterday[yesterday > 0])  # 'incomplete' functionality of ordertable2supp()
        podium_yesterday <- names(which((yesterday <= 3) & (yesterday>0)))
-       print(podium_yesterday)
-
        d <- fmat[i, , drop = TRUE] # today
-       print(d)
+       d[d > 0] <- rank(d[d > 0])  # This is the 'incomplete' functionality of ordertable2supp()
        ## Following lines lifted from ordervec2supp()
@@
 
-Now a dataset:
-
-```{r,load2007}
-a <- read.table("formula1_2007.txt",header=TRUE)
-a <- a[,seq_len(ncol(a)-1)]
-a
-```
-
 Define a subset of it:
@@
 ```{r, definesubset}
+a <- read.table("formula1_2007.txt",header=TRUE)
+a <- a[,seq_len(ncol(a)-1)]
 b <- a[1:6,1:5]
 b[,1] <- 1:6
-b[,2] <- c(1,3,4,2,5,6)
-b[,3] <- c(5,1,2,3,4,6)
+b[,2] <- c(4,5,6,1,2,3)
+b[,3] <- c(5,6,4,2,3,1)
 
-b[,5] <- 1:6
+b[,5] <- c(1,2,4,3,8,6)  # tests the 'incomplete' functionality of podium()
 b
@@
 
-```{r,usepodium,cache=TRUE}
-Ha <- podium(b,1.4)
-Hb <- podium(b,1.5)
+```{r,usepodiu,cache=TRUE}
+Ha <- podium(1.4,b)
+Hb <- podium(1.5,b)
 ma <- maxp(Ha,give=TRUE,n=1)
@@
 
-```{r,usefun,cache=TRUE}
-fun <- function(lambda){maxp(podium(b,lambda),give=TRUE)$value}
+```{r,usefuns,cache=TRUE}
+fun <- function(lambda){maxp(podium(lambda,b),give=TRUE)$value}
 lam <- seq(from=0.5,to=2,len=6)
@@
 ```{r, plotlik}
-plot(lam,like,type='b')
+like <- unlist(like)
+plot(lam,like-max(like),type='b')
+like
 ```
+
+Now a real dataset, Formula 1 2007 season:
+
+```{r,load2007}
+
+a <- read.table("formula1_2007.txt",header=TRUE)
+a <- a[1:10,seq_len(ncol(a)-1)]  # rows Raikkonen (1) through Coulthard (10)
+options(width=90)
+a
+```
+
+```{r,definepodlike}
+podlike <- function(lambda){maxp(podium(lambda=lambda,a),n=1,give=TRUE)$value}
+```
+
+```{r,usepodlike10,cache=TRUE}
+podlike(1.8)
+podlike(1.9)
+```
+
+```{r,podlikevec7,cache=TRUE}
+date()
+lam <- seq(from=0.2,to=1.2,len=7)
+like <- lapply(lam,podlike)
+date()
+```
+
+```{r,plotpodlike}
+like <- unlist(like)
+plot(lam,like-max(like),type="b")
+abline(h=c(0,-2))
+abline(v=1)
+```

### 2023-02-10 R/hyper2.R

@@ -996,6 +996,39 @@ setGeneric("pnames<-",function(x,value){standardGeneric("pnames<-")})
     return(out)
 }
 
+
+`complete` <- function(a,noscore=NULL,check=FALSE){
+  if(is.null(noscore)){
+    noscore <- c("Ret", "WD", "DNS", "DSQ", "DNP", "NC", "DNQ", "EX", "Sick")
+  }
+  
+  out <- apply(a,2,
+               function(v){
+                 v[v %in% noscore] <- "0"
+                 v <- as.numeric(v)
+                 if(all(sort(v[v>0]) == seq_along(v))){
+                   v[v>0] <- rank(v[v>0])
+                   return(v)
+                 } else {
+                   stop("some ranks missing")
+                 }
+               } )
+  rownames(out) <- rownames(a)
+  return(out)
+}
+
+`ordertable2supp_new` <- function(x, noscore=NULL, check=FALSE){
+    x <- complete(x,noscore=noscore,check=check)
+    out <- hyper2()
+    
+    ## Now cycle through the rows; each row is a venue [voter]
+    for(i in seq_len(ncol(x))){
+        o <- x[,i,drop=TRUE]
+        out %<>% `+`(ordervec2supp(o))
+    } # i loop closes
+    return(out)
+}
+
 `ordertable2supp` <- function(x, noscore, incomplete=TRUE){
     if(missing(noscore)){
         noscore <- c("Ret", "WD", "DNS", "DSQ", "DNP", "NC", "DNQ", "EX", "Sick")

### 2023-02-17 inst/monster_vs_lambda.Rmd
likelihood contours

@@
 ```{r,echo=FALSE}
-jj <- matrix(c(NA, 1+2i ,4+5i, NA),2,2,byrow=TRUE)
+jj <- matrix(c(NA, 6+2i ,6+5i, NA),2,2,byrow=TRUE)
 teams <- LETTERS[1:2]
@@
 
+# Likelihood as a function of $M,\lambda$
+
+```{r,label=deflikef2}
+likef <- function(M,lambda,...){specificp.test(ha3(D,lambda=lambda),"M",M,...)$null_support}
+like2 <- function(theta,...){likef(M=theta[1],lambda=theta[2],n=1,...)}
+```
+
+Use it:
+
+```{r,label=ttrydefflikefvc,cache=TRUE}
+Mval <- seq(from=0.0,to=0.3,len=5)
+lval <- seq(from=0.1,to=5,len=5)
+jj <- as.matrix(expand.grid(M=Mval,lam=lval))
+L <- apply(jj,1,like2)
+```
+
+```{r}
+L <- matrix(L,length(Mval),length(lval))
+L <- L-max(L)
+L
+L <- pmax(L,-10)
+L
+Mval
+lval
+contour(Mval,lval,L,xlab='M',ylab='lambda')
+filled.contour(Mval,lval,L,xlab='M',ylab='lambda')
+```
+
 # Create a new dataset by direct simulation

### 2023-03-05 inst/jss_hyper3.Rnw
force it to include the figures explicitly

@@
   \begin{center}
-<<figaba>>=
+<<figaba, fig=TRUE>>=
 a <- seq(from = 0, by = 0.005, to = 1)
@@
   \begin{center}
-<<figabbabb,echo=FALSE>>=
+<<figabbabb,echo=FALSE,fig=TRUE>>=
 f_aab <- function(a){a^2 / (1 + a)}
@@
   \begin{center}
-<<plotwilcoxlike,echo=FALSE>>=
+<<plotwilcoxlike,echo=FALSE,fig=TRUE>>=
 a <- seq(from = 0.02, to = 0.8, len = 40)
@@
   \begin{center}
-<<testthejav, cache=TRUE>>=
+<<testthejav, cache=TRUE, fig=TRUE>>=
 (mj <- maxp(javelin))
@@
 \begin{center}
-<<plottheloglikcont,echo=FALSE>>=
+<<plottheloglikcont,echo=FALSE,fig=TRUE>>=
 colnames(M) <-  as.character(Ves)

### 2023-03-06 README.Rmd
fixes #19

---
title: "The hyper 2 package"
output:
  github_document:
    pandoc_args: --webtex
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

<img src="man/figures/hyper2.png" width = "150" align="right" />

<!-- badges: start -->
[![Total Downloads](https://cranlogs.r-pkg.org/badges/grand-total/freealg)](https://CRAN.R-project.org/package=freealg)
[![Build Status](https://travis-ci.org/RobinHankin/freealg.svg?branch=master)](https://travis-ci.org/RobinHankin/freealg)
[![CRAN_Status_Badge](https://www.r-pkg.org/badges/version/freealg)](https://cran.r-project.org/package=freealg)
<!-- badges: end -->

# Overview

The `hyper2` package gives a suite of methods for creating and
analysing likelihood functions generated from a generalization of the
Bradley-Terry probability model.  The package uses the `STL` map class
of `C++` for efficiency; it conforms to `disordR` discipline.

# Installation

The released version of `hyper2` on [CRAN](https://CRAN.R-project.org)
may be installed with:

```{r, message=FALSE}
# install.packages("hyper2")  # uncomment this to install the package
library("hyper2")
```

# Further information

For more detail, see the package vignette

`vignette("freealg")`

# Citation

To cite the `hyper2` package in publications, please use

Robin K. S. Hankin 2017.  "Partial rank data with the `hyper2`
package: Likelihood functions for generalized Bradley-Terry models".
_The R Journal_, 9(2):429--439

### 2023-03-21 inst/monster_vs_lambda.Rmd
reworking of limits

@@
 ```{r,label=ttrydefflikefvc,cache=TRUE}
-Mval <- seq(from=0.0,to=0.3,len=5)
-lval <- seq(from=0.1,to=5,len=5)
+Mval <- seq(from=0.0,to=0.65,len=15)
+lval <- seq(from=0.2,to=5,by = 0.4) # we need lambda=1 (this is the null)
 jj <- as.matrix(expand.grid(M=Mval,lam=lval))
+options(use_alabama = FALSE)
 L <- apply(jj,1,like2)
@@
 ```{r}
+options(digits=2)
+options(width=150)
 L <- matrix(L,length(Mval),length(lval))
 L <- L-max(L)
+rownames(L) <- paste("M = ",round(Mval,2))
+colnames(L) <- paste("lam=",lval,sep="")
 L
-L <- pmax(L,-10)
+L <- pmax(L,-9)
 L
@@
 contour(Mval,lval,L,xlab='M',ylab='lambda')
+abline(h=1,lwd=5)
+abline(v=0,lwd=5)
 filled.contour(Mval,lval,L,xlab='M',ylab='lambda')
+matplot(Mval,L,type='b')
+matplot(lval,t(L),type='b')
 ```

### 2023-03-27 inst/monster_vs_lambda.Rmd
option "use_alabama" gives an initial value not feasible error

@@
 jj <- as.matrix(expand.grid(M=Mval,lam=lval))
-options(use_alabama = FALSE)
+options(use_alabama = FALSE) # setting to TRUE gives a "initial value not feasible" error, go figure
 L <- apply(jj,1,like2)

### 2023-04-17 inst/ijmm_rebuttal.tex
start on rebuttal as per issue #197

\documentclass[12pt]{article}
\usepackage{xcolor}
\begin{document}

\section*{An objective assessment of point scoring systems in Formula 1 motor racing: rebuttal}

Below, I reproduce the reviewers comments; my replies to the issues
are in \textcolor{blue}{blue}.  I have indicated changes to the
manuscript where appropriate.

Short story: I have accommodated all the comments with rewording and
clarification.  The resulting document is, I believe, stronger and
more scholarly than before and I recommend it to you.

\subsection*{Detailed response to comments}

Conclusion in Introduction: The final paragraph of section 1
(Introduction) seems to need a conclusion.

Points systems similar to that of Formula 1 are common in other racing
sports; all have the common feature of translating ranks into points
which combine additively to generate an overall ranking.  Further, we
see points systems used in the wider context of competitive situations
such as the Eurovision Song Contest and many other such group
tournaments.  \mbox{Therefore, ...}

\textcolor{blue}{Response to comment here}\\ \\

Addition of literature review of Zipf (1949)

\textcolor{blue}{Response to comment here}\\ \\

Many of our readers are unfamiliar with this work.  Thus, a short
passage will help in understanding the significance of the inclusion
of Zipf in the manuscript.

\textcolor{blue}{Response to comment here}\\ \\

The charts need to be larger for easier reading.  Please place figures
in full-page format as appendices. Share the actual charts rather than
pictures of the charts. Doing so will allow us to retain image quality
as figures are enlarged.

\textcolor{blue}{Response to comment here}\\ \\

Replace Pie Charts with Pareto Charts

\textcolor{blue}{Response to comment here}\\ \\

The pie charts are difficult to distinguish between components. For
example, in Figure 1 for 2017, the reviewers could not see a
difference between Bottas and Vettel.  The Pareto Chart style would
improve that aspect.

\textcolor{blue}{Response to comment here}\\ \\

\section*{Rankings in Section 2.1}

Place the comparisons of the rankings systems in a table.  Readers will
be able to make comparisons more easily.

\textcolor{blue}{Response to comment here}\\ \\

Table 2 and Figures 4-6 Consistency

Please use WTA or Top1 consistently in these items.

\textcolor{blue}{Response to comment here}\\ \\

Section 2.2 Numerical Results

Explain how the WTA system maximizes the probability of the complete
order statistic.

\textcolor{blue}{Response to comment here}\\ \\

Comment for Practitioners

Please include a summary of how the points system could benefit the
managers of Formula 1 and other series in the conclusion.

\textcolor{blue}{Response to comment here}\\ \\

\end{document}

### 2023-04-19 inst/ijmm_rebuttal.tex
work on issue #197

@@
 such as the Eurovision Song Contest and many other such group
-tournaments.  \mbox{Therefore, ...}
+tournaments.  \mbox{Therefore, \ldots}\\ \\
 
-\textcolor{blue}{Response to comment here}\\ \\
+\textcolor{blue}{\noindent I have added a sentence to the paragraph as
+  suggested, which now reads:\\ \\ \\Points systems similar to that of
+  Formula 1 are common in other racing sports; all have the common
+  feature of translating ranks into points which combine additively to
+  generate an overall ranking. Further, we see points systems used in
+  the wider context of competitive situations such as the Eurovision
+  Song Contest and many other such group tournaments.  {\bf Therefore,
+    the ideas used here for analysis of motorsports are directly
+    applicable to a broad range of competitive situations in which
+    points are used to rank competitors.}}\\ \\
 
 Addition of literature review of Zipf (1949)
-
-\textcolor{blue}{Response to comment here}\\ \\
 
@@
 
-\textcolor{blue}{Response to comment here}\\ \\
+\textcolor{blue}{I have realised that referring to a points system
+  $\left(1,\frac{1}{2},\frac{1}{3},\frac{1}{4},\ldots\right)$ as
+  ``Zipf's law'' is not really appropriate: I am not using it as a
+  probability mass function but rather as a simple monotonically
+  decreasing sequence.  I have removed references to ``Zipf's law'' in
+  favour of the more descriptive ``Zipfian''.  I have added a sentence
+  to this effect in the manuscript}\\ \\
 
 The charts need to be larger for easier reading.  Please place figures
-in full-page format as appendices. Share the actual charts rather than
-pictures of the charts. Doing so will allow us to retain image quality
-as figures are enlarged.
+in full-page format as appendices.  Share the actual charts rather
+than pictures of the charts.  Doing so will allow us to retain image
+quality as figures are enlarged.
 
@@
 Replace Pie Charts with Pareto Charts
-
-\textcolor{blue}{Response to comment here}\\ \\
 

### 2023-04-19 inst/f1points.Rmd
work on issue #197

@@
 pie(mH2019[c(1,6,11,2,7,3,8,4,9,5,10)],cex=0.7,main="2019")
+```
+
+```{r label=loadqcc}
+library("qcc")
+```
+
+```{r,label=printaparetochart2016,echo=FALSE, print=FALSE,fig.cap="Maximum likelihood estimates \\label{paretochartstrength2016} of the strengths of the top-ranked 11 drivers in Formula 1 motor racing,  2016"}
+pareto.chart(mH2016[c(1,6,11,2,7,3,8,4,9,5,10)],main="2016")
+```
+
+```{r,label=printaparetochart2017,echo=FALSE, print=FALSE,fig.cap="Maximum likelihood estimates \\label{paretochartstrength2017} of the strengths of the top-ranked 11 drivers in Formula 1 motor racing, 2017"}
+pareto.chart(mH2017[c(1,6,11,2,7,3,8,4,9,5,10)],main="2017")
+```
+
+```{r,label=printaparetochart2018,echo=FALSE, print=FALSE,fig.cap="Maximum likelihood estimates \\label{paretochartstrength2018} of the strengths of the top-ranked 11 drivers in Formula 1 motor racing, 2018"}
+pareto.chart(mH2018[c(1,6,11,2,7,3,8,4,9,5,10)],main="2018")
+```
+
+```{r,label=printaparetochart2019,echo=FALSE, print=FALSE,fig.cap="Maximum likelihood estimates \\label{paretochartstrength2019} of the strengths of the top-ranked 11 drivers in Formula 1 motor racing, 2019"}
+pareto.chart(mH2019[c(1,6,11,2,7,3,8,4,9,5,10)],main="2019")
 ```

### 2023-04-21 inst/f1points.Rmd
relabel axes

--- 2022-10-25 inst/f1points.Rmd
@@
 pie(mH2019[c(1,6,11,2,7,3,8,4,9,5,10)],cex=0.7,main="2019")
+```
+
+```{r label=loadqcc}
+library("qcc")
+```
+
+```{r,label=printaparetochart2016,echo=FALSE, print=FALSE,fig.cap="Maximum likelihood estimates \\label{paretochartstrength2016} of the strengths of the top-ranked 11 drivers in Formula 1 motor racing,  2016"}
+pareto.chart(mH2016[c(1,6,11,2,7,3,8,4,9,5,10)],main="2016",ylab="strength")
+```
+
+```{r,label=printaparetochart2017,echo=FALSE, print=FALSE,fig.cap="Maximum likelihood estimates \\label{paretochartstrength2017} of the strengths of the top-ranked 11 drivers in Formula 1 motor racing, 2017"}
+pareto.chart(mH2017[c(1,6,11,2,7,3,8,4,9,5,10)],main="2017",ylab="strength")
+```
+
+```{r,label=printaparetochart2018,echo=FALSE, print=FALSE,fig.cap="Maximum likelihood estimates \\label{paretochartstrength2018} of the strengths of the top-ranked 11 drivers in Formula 1 motor racing, 2018"}
+pareto.chart(mH2018[c(1,6,11,2,7,3,8,4,9,5,10)],main="2018",ylab="strength")
+```
+
+```{r,label=printaparetochart2019,echo=FALSE, print=FALSE,fig.cap="Maximum likelihood estimates \\label{paretochartstrength2019} of the strengths of the top-ranked 11 drivers in Formula 1 motor racing, 2019"}
+pareto.chart(mH2019[c(1,6,11,2,7,3,8,4,9,5,10)],main="2019",ylab="strength")
 ```

### 2023-04-21 inst/f1points.Rmd
add individual ordertransplots

@@
 
+```{r label=points_order_vs_likelihood_order_2016,cache=TRUE}
+par(pty="s")
+oy <- ordertrans(m2016,names(dp2016))
+plot(ox,oy,asp=1,pty='s',xlim=c(0,11.2),ylim=c(0,11.2),pch=16,
+    xlab="points order",ylab="likelihood order",main='2016 season')
+par(xpd=TRUE)       # allow drivers' names to appear outside plotting region
+for(i in seq_along(ox)){text(ox[i],oy[i],names(ox)[i],pos=4,col='gray',cex=0.6) }
+par(xpd=FALSE)      # stop diagonal line from protruding beyond plotting region
+abline(0,1)
+```
+
+```{r label=points_order_vs_likelihood_order_2017,cache=TRUE}
+par(pty="s")
+oy <- ordertrans(m2017,names(dp2017))
+plot(ox,oy,asp=1,pty='s',xlim=c(0,11.2),ylim=c(0,11.2),pch=16,
+    xlab="points order",ylab="likelihood order",main='2017 season')
+par(xpd=TRUE)       # allow drivers' names to appear outside plotting region
+for(i in seq_along(ox)){text(ox[i],oy[i],names(ox)[i],pos=4,col='gray',cex=0.6) }
+par(xpd=FALSE)      # stop diagonal line from protruding beyond plotting region
+abline(0,1)
+```
+
+```{r label=points_order_vs_likelihood_order_2018,cache=TRUE}
+par(pty="s")
+oy <- ordertrans(m2018,names(dp2018))
+plot(ox,oy,asp=1,pty='s',xlim=c(0,11.2),ylim=c(0,11.2),pch=16,
+    xlab="points order",ylab="likelihood order",main='2018 season')
+par(xpd=TRUE)       # allow drivers' names to appear outside plotting region
+for(i in seq_along(ox)){text(ox[i],oy[i],names(ox)[i],pos=4,col='gray',cex=0.6) }
+par(xpd=FALSE)      # stop diagonal line from protruding beyond plotting region
+abline(0,1)
+```
+
+```{r label=points_order_vs_likelihood_order_2019,cache=TRUE}
+par(pty="s")
+oy <- ordertrans(m2019,names(dp2019))
+plot(ox,oy,asp=1,pty='s',xlim=c(0,11.2),ylim=c(0,11.2),pch=16,
+    xlab="points order",ylab="likelihood order",main='2019 season')
+par(xpd=TRUE)       # allow drivers' names to appear outside plotting region
+for(i in seq_along(ox)){text(ox[i],oy[i],names(ox)[i],pos=4,col='gray',cex=0.6) }
+par(xpd=FALSE)      # stop diagonal line from protruding beyond plotting region
+abline(0,1)
+```
+
 We can plot one ranking against the other\footnote{The historically

### 2023-04-23 inst/f1points.Rmd
work on rebuttal (constency wta vs top1), issue #197

@@
 m2017[] <- seq_along(m2017)
+```
 
@@
 
+```{r label=zipf_order_vs_likelihood_order_2016,cache=TRUE}
+par(pty='s')        # square plot
+ox <- dp2016z
+oy <- ordertrans(m2016,names(dp2016z))
+plot(ox,oy,asp=1,pty='s',xlim=c(0,11.2),ylim=c(0,11.2),pch=16,
+    xlab="points order (Zipf)",ylab="likelihood order",main='2016 season')
+par(xpd=TRUE)       # allow drivers' names to appear outside plotting region
+for(i in seq_along(ox)){text(ox[i],oy[i],names(ox)[i],pos=4,col='gray',cex=0.6) }
+par(xpd=FALSE)      # stop diagonal line from protruding beyond plotting region
+abline(0,1)
+```
+
+```{r label=zipf_order_vs_likelihood_order_2017,cache=TRUE}
+par(pty='s')        # square plot
+ox <- dp2017z
+oy <- ordertrans(m2017,names(dp2017z))
+plot(ox,oy,asp=1,pty='s',xlim=c(0,11.2),ylim=c(0,11.2),pch=16,
+    xlab="points order (Zipf)",ylab="likelihood order",main='2017 season')
+par(xpd=TRUE)       # allow drivers' names to appear outside plotting region
+for(i in seq_along(ox)){text(ox[i],oy[i],names(ox)[i],pos=4,col='gray', cex=0.6) }
+par(xpd=FALSE)      # stop diagonal line from protruding beyond plotting region
+abline(0,1)
+```
+
+```{r label=zipf_order_vs_likelihood_order_2018,cache=TRUE}
+par(pty='s')        # square plot
+ox <- dp2018z
+oy <- ordertrans(m2018,names(dp2018z))
+plot(ox,oy,asp=1,pty='s',xlim=c(0,11.2),ylim=c(0,11.2),pch=16,
+    xlab="points order (Zipf)",ylab="likelihood order",main='2018 season')
+par(xpd=TRUE)       # allow drivers' names to appear outside plotting region
+for(i in seq_along(ox)){text(ox[i],oy[i],names(ox)[i],pos=4,col='gray', cex=0.6) }
+par(xpd=FALSE)      # stop diagonal line from protruding beyond plotting region
+abline(0,1)
+```
+
+```{r label=zipf_order_vs_likelihood_order_2019,cache=TRUE}
+par(pty='s')        # square plot
+ox <- dp2019z
+oy <- ordertrans(m2019,names(dp2019z))
+plot(ox,oy,asp=1,pty='s',xlim=c(0,11.2),ylim=c(0,11.2),pch=16,
+    xlab="points order (Zipf)",ylab="likelihood order",main='2019 season')
+par(xpd=TRUE)       # allow drivers' names to appear outside plotting region
+for(i in seq_along(ox)){text(ox[i],oy[i],names(ox)[i],pos=4,col='gray',cex=0.6) }
+par(xpd=FALSE)      # stop diagonal line from protruding beyond plotting region
+abline(0,1)
+```
+
 \begin{itemize}
@@
 load("formula1_results_2016.rda") # this created by running f1points_Omaker.R; file contains R variables OO and pointslist
+rownames(OO)[3] <- "wta"
 OO2016 <- OO
@@
 load("formula1_results_2017.rda") # this created by running f1points_Omaker.R; file contains R variables OO and pointslist
+rownames(OO)[3] <- "wta"
 OO2017 <- OO
@@
 load("formula1_results_2018.rda") # this created by running f1points_Omaker.R; file contains R variables OO and pointslist
+rownames(OO)[3] <- "wta"
 OO2018 <- OO
@@
 load("formula1_results_2019.rda") # this created by running f1points_Omaker.R; file contains R variables OO and pointslist
+rownames(OO)[3] <- "wta"
 OO2019 <- OO
@@
 matplot(jj[wanted,],type="n",xlab='points system',ylab='mean correct places',axes=FALSE)
-axis(1,at=1:6,labels=names(formula1_points_systems(top)[wanted]))
+jjlabels <- names(formula1_points_systems(top)[wanted])
+jjlabels[5] <- "wta"
+axis(1,at=1:6,labels=jjlabels)
 axis(2,at=1:4)
@@
 matplot(jj[wanted,],type="n",xlab='points system',ylab='probability of winner correct',axes=FALSE)
-axis(1,at=1:6,labels=names(formula1_points_systems(top)[wanted]))
+axis(1,at=1:6,labels=jjlabels)
 axis(2)
@@
 matplot(jj[wanted,],type="n",xlab='points system',ylab='probability of total rank order correct',axes=FALSE)
-axis(1,at=1:6,labels=names(formula1_points_systems(top)[wanted]))
+axis(1,at=1:6,labels=jjlabels)
 axis(2)

### 2023-04-24 inst/f1points.Rmd
bugfix (drivers' names were wrong)

@@
 Thus, taking 2017 as an example, the points-based ordering would be Hamilton first, then Vettel, then R\"{a}ikk\"{o}nen; YES I KNOW IT IS WRONG: RMARKDOWN IS DEFECTIVE while the likelihood ordering is (reading vertically) Hamilton, Vettel, Bottas ALSO R PLOTTING SUCKS HERE AND I WILL HAVE TO CHANGE THE PDF BY HAND WITH INKSCAPE TO MAKE IT LOOK GOOD"}
+
 par(mfrow=c(2,2))
@@
 par(pty="s")
+ox <- dp2016
 oy <- ordertrans(m2016,names(dp2016))
@@
 par(pty="s")
+ox <- dp2017
 oy <- ordertrans(m2017,names(dp2017))
@@
 par(pty="s")
+ox <- dp2018
 oy <- ordertrans(m2018,names(dp2018))
@@
 par(pty="s")
+ox <- dp2019
 oy <- ordertrans(m2019,names(dp2019))

### 2023-04-24 inst/ijmm_rebuttal.tex
further work on issue #197

@@
 
-\textcolor{blue}{Response to comment here}\\ \\
+\textcolor{blue}{Done.  Pie charts replaced with Pareto charts}\\ \\
 
@@
 
-\textcolor{blue}{Response to comment here}\\ \\
+\textcolor{blue}{Done.}\\ \\
 
@@
 
-\textcolor{blue}{Response to comment here}\\ \\
+\textcolor{blue}{Done, wta used consistently now}\\ \\
 
@@
 
-\textcolor{blue}{Response to comment here}\\ \\
+\textcolor{blue}{Brief sentence added summarizing the comparison
+  method}\\ \\
 
@@
 
+\textcolor{blue}{Summary added to the conclusions section.}
+
+Overall, your analysis of Formula 1 points systems and the comparison
+of points systems to an objective rank ordering has the potential to
+be a valuable resource for practitioners in the motorsports industry,
+helping to inform strategy, evaluate fairness, and guide
+decision-making.
+
 \textcolor{blue}{Response to comment here}\\ \\

### 2023-05-01 inst/ijmm_rebuttal.tex
more work on issue #197

@@
 
-Below, I reproduce the reviewers comments; my replies to the issues
-are in \textcolor{blue}{blue}.  I have indicated changes to the
-manuscript where appropriate.
-
+\textcolor{blue}{
+Below, I reproduce the reviewers comments; my writing, including
+replies to the reviewer issues are in blue.  I have indicated changes
+to the manuscript where appropriate.\\ \\
 Short story: I have accommodated all the comments with rewording and
@@
 more scholarly than before and I recommend it to you.
+}
 
@@
 
-\textcolor{blue}{Response to comment here}\\ \\
+\textcolor{blue}{Charts now enlarged, also original PDFs supplied as
+  requested}\\ \\
 
@@
 
-\textcolor{blue}{Summary added to the conclusions section.}
+\textcolor{blue}{Summary added to the conclusions section.}\\ \\
 
@@
 
-\textcolor{blue}{Response to comment here}\\ \\
+\textcolor{blue}{Sentence added at the end of the document to this
+  effect, slightly reworded from the reviewer's suggestion to fit
+  better in context.}
 

### 2023-05-01 inst/jester.Rmd
use sticker in Rmd
consistentify jester to match the formula 1 data

@@
 
+## Create a table like formula 1 results table
+
+```{r makejestertable}
+jester_table <- a
+for(i in seq_len(nrow(jester_table))){
+  x <- jester_table[i,]
+  x[!is.na(x)] <- rank(-x[!is.na(x)],ties.method="first")
+  jester_table[i,] <- x
+}
+jester_table <- t(jester_table)
+colnames(jester_table) <-paste("resp",seq_len(ncol(jester_table)),sep="_")
+jester_table[1:6,1:10]
+```
+
+Above we see that respondent 1 ranked joke 8 as the funniest and joke 7 as the 15th funniest.
+Comparing with `formula1_2022.txt`, for example, we see that respondents correspond to venues and jokes correspond to drivers.
+We have to be a little bit careful because `NA` means "not rated", not "did not finish" as in the Formula 1 datasets.
+
+```{r meanjokes}
+plot(rowMeans(jester_table,na.rm=TRUE))
+abline(v=17)
+```
+
 ### Package dataset
@@
 ```{r,label=savejesterdataset}
-save(jester,jester_maxp,file="jester.rda")
+save(jester,jester_table,jester_maxp,file="jester.rda")
 ```

### 2023-05-02 inst/simplified_likelihood.Rmd
simplified likelihood

---
title: "simplified_likelihood"
author: "Robin Hankin"
date: "2023-05-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Consider a race between $n+2$ competitors of strength $p_1$, $p_2$, and $n$ clones, all of strength $p=1-p_1-p_2$.  Our observation is that $p_1$ places $n_1$, and $p_2$ places $n_2>n_1$.  The order statistic is thus

$$
\underbrace{p\succ\cdots\succ p}_{\mbox{$n_1-1$ terms}}\succ p_1\succ\underbrace{p\succ\cdots\succ p}_{n_2-n_1-1}\succ p_2\succ\underbrace{p\succ\cdots\succ p}_{n-n_2+2}
$$

The Plackett-Luce likelihood function would be as follows

$$
\begin{eqnarray}
&{}&
\underbrace{
\frac{p}{p_1+p_2+np}\cdot
\frac{p}{p_1+p_2+(n-1)p}
\cdots
\frac{p}{p_1+p_2+(n-n_1+2)p}
}_{n_1-1}\cdot\\&{}&
\frac{p_1}{p_1+p_2+(n-n_1+1)p}\cdot\\&{}&
\underbrace{
\frac{p}{p_2+(n-n_1+1)p}\cdot
\frac{p}{p_2+(n-n_1)p}\cdots
\frac{p}{p_2+(n-n_2+3)p}
}_{n_2-n_1-1}\cdot\\&{}&
\frac{p_2}{p_2+(n-n_2+2)p}\cdot\\&{}&
\underbrace{
\frac{p}{(n-n_2+2)p}\cdot\frac{p}{(n-n_2)p}
\cdots\frac{p}{2p}\cdot\frac{p}{p}
}_{n-n_2+2}
\end{eqnarray}
$$

Observing that the fifth line may be neglected, R idiom for this would be as follows

```{r}
library("hyper2")
simplike <- function(n1,n2,n){
  H <- hyper3()
  H[c(p =1)] <- n2-2
  H[c(p1=1)] <- 1
  H[c(p2=1)] <- 1
  for(i in (n-n1+1):n)     { H[c(p1=1,p2=1,p=i)]  %<>% dec }
  for(i in (n-n2+2):(n-n1+1)){ H[c(p2=1,   p=i)]  %<>% dec }
  return(H)
} 
simplike(2,6,11)
simplike(1,6,11)
simplike(1,2,11)
```

### 2023-05-02 inst/simplified_likelihood.Rmd
better idiom

@@
 library("hyper2")
-simplike <- function(n1,n2,n){
+simplike_lowlevel <- function(n1,n2,n,p1="p1",p2="p2",other="other"){
+  stopifnot(n1<n2)
   H <- hyper3()
-  H[c(p =1)] <- n2-2
-  H[c(p1=1)] <- 1
-  H[c(p2=1)] <- 1
-  for(i in (n-n1+1):n)     { H[c(p1=1,p2=1,p=i)]  %<>% dec }
-  for(i in (n-n2+2):(n-n1+1)){ H[c(p2=1,   p=i)]  %<>% dec }
+  H[setNames(1,other)] <- n2-2
+  H[setNames(1,p1)] <- 1
+  H[setNames(1,p2)] <- 1
+  for(i in (n-n1+1):n){H[setNames(c(1,1,i),c(p1,p2,other))]  %<>% dec }
+  for(i in (n-n2+2):(n-n1+1)){H[setNames(c(1,i),c(p2,other))] %<>% dec }
   return(H)
 }
-simplike(2,6,11)
-simplike(1,6,11)
-simplike(1,2,11)
+
+twodrivers <- function(...,n){ # twodrivers(ham=3, bottas=5,n=20)
+  L <- list(...)
+  stopifnot(length(L)==2)
+  if(L[[1]] > L[[2]]){ # wrong way round
+    tmp <- L[[2]]
+    L[[2]] <- L[[1]]
+    L[[1]] <- tmp
+    names(L) <- rev(names(L))
+  }
+  simplike_lowlevel(as.numeric(L[[1]]),as.numeric(L[[2]]),n,names(L)[1],names(L)[2])
+}
+
+twodrivers(o=2,XX=6,n=11)
 ```
+
+Now
+
+```{r readtable}
+a <- read.table("formula1_2019.txt")[1:2,1:21]
+a
+f <- function(i){
+  jj <- as.list(a[,i])
+  names(jj) <- rownames(a)
+  return(jj)
+}
+
+H <- hyper3()
+for(i in seq_len(ncol(a))){
+  jj <- f(i)
+  if(all(!is.na(as.numeric(unlist(jj))))){
+    H <- H + do.call(twodrivers,c(jj,n=20))
+  }
+}
+H
+```
+
+```{r maxh,cache=TRUE}
+maxp(H)
+```
+
+```{r equaltest,cache=TRUE}
+samep.test(H,c("Hamilton","Bottas"))
+```

### 2023-05-03 inst/jss_4382_rebuttal.tex
start work on issue #153

\documentclass[12pt]{article}
\usepackage{xcolor}
\begin{document}

\section*{JSS4832: Generalized Plackett-Luce Likelihoods by Robin
K. S. Hankin}

Below, the reviewer's comments are in black, and the replies to the
issues are in \textcolor{blue}{blue}.  I have indicated changes to
the manuscript where appropriate.

Short story: I have accommodated all the comments with rewording and
clarification.

\subsection*{Reviewer's comments, 15 April 2023}

Dear Robin K.S. Hankin,

We have reached a decision regarding your submission to Journal of
Statistical Software, #4832 “Generalized Plackett-Luce Likelihoods”.

Unfortunately we have to tell you that we decided to reject in its current state.

However, we encourage you to resubmit the article after detailed
revision. Please check the journal management system for any
additional feedback:
https://www.jstatsoft.org/index.php/jss/authorDashboard/submission/4832. We
expect you to provide a point-by-point response to this feedback in
your revision.

Your sincerely,

The editorial team

The manuscript describes functionality of the hyper2 package for
Fitting and testing variants of the Placket-Luce ranking model.

1.1 Recommendation

The manuscript may be publishable in the journal after revisions have been
made.

1.2 General comments

G1 The introduction should provide an overview of the contents of the manuscript.
\textcolor{blue}{Response here}\\

In addition, the section headlines could be made more informative. Currently, they are:
2. "Some simple cases"
3. "Mann-Whitney test"
4. "Multiple competitors: javelin"
5. "Formula 1 motor racing: the constructors' championship"

Maybe the headlines could provide more of a summary of what the section
contains.
\textcolor{blue}{Response here}\\ \\

G2 There should be a comparison with other software. Although the author
states in the introduction that with hyper2 "It becomes possible to analyse
a wider range of likelihood functions than standard Plackett-Luce", it is
not clear to me which of the features described later on are unique to the
hyper2 package and which can also be obtained from other packages (like,
e.g., PlackettLuce).
\textcolor{blue}{Response here}\\ \\

G3 The manuscript should comment on the timing of the functions. For the
simple example on P. 4, the maxp(H) call takes more than 8 seconds on
my computer. This makes me wonder how timings would be for realistic
examples.
\textcolor{blue}{Response here}\\ \\

1.3 Specific comments\\
S1 I'm not claiming any expertise in English language, but occasionally, I found
the manuscript difficult to understand. For example, in the abstract it
says: "The package . . . furnishes idiom" Couldn't that simply be: "The
package . . . provides functionality"? On P. 13, it says "Here we use generalized
Plackett-Luce to assess the constructors' performance". Isn't there
a "model" missing? (Plackett-Luce is just a name and cannot be used).
\textcolor{blue}{Response here}\\ \\

S2 Figure 2: For readability, maybe use xlab = expression(p[a]) and maybe
instead of a legend, write AAB, ABA, BAA next to the line in the color
of that line.
\textcolor{blue}{Response here}\\ \\

S3 P. 7: "https:github.com/RobinHankin/hyper2" should be
"https://github.com/RobinHankin/hyper2".
\textcolor{blue}{Response here}\\ \\

S4 P. 7: "Wolf" should be "Wolfe".
\textcolor{blue}{Response here}\\ \\

S5 P. 15, maxp(const2020, n=1): What does n do?
\textcolor{blue}{Response here}\\ \\

S6 P. 16, "we see strong evidence for a real decrease in the strength": I'd tone
that down. The model is almost certainly wrong, so it does not measure
the "real decrease." Also, power is unknown, so it is difficult to say what
"evidence" there is against H0. (See, e.g., Gelman, Regression an other
stories, Chapter 4: Time to unlearn what you thought you knew about
statistics. https://statmodeling.stat.columbia.edu/2022/01/27/regression-
and-other-stories-free-pdf/)
\textcolor{blue}{Response here}\\ \\

S7 The references are lacking their DOIs.
\textcolor{blue}{Response here}\\ \\

2 Review of software

The package includes many functions with a similar name. E.g., ordervec2supp,
ordervec2supp3, ordervec2supp3a, ordertable2supp, ordertable2supp3. Maybe
there could be some overview of which to use for what purpose.

\textcolor{blue}{Response here}\\ \\

Section editor comments:

The author proposes an extension of the alreadty existing hyper2 package.
This new S3 class allows for some extended likelihood specifications.

\textcolor{blue}{Response here}\\ \\

Even though this is just a code snippet and the hyper2 package was already
published elsewhere, the author should provide some introductory elaborations
on the Plackett-Luce model. Along these lines it would also be helpful to give
a quick overview of the hyper2 package, and mention other related packages. As it
is now it is difficult for a reader who's not super familiar with PL-models and
hyper2, respectively, to follow the paper. Anyway, I find the runners example
as motivating example very illustrative, and actually like the examples throughout
the paper.

\textcolor{blue}{Response here}\\ \\

Some more aesthetic effort should be made in the code file (e.g., providing
comments). The code looks like a simple purl() extraction.

\textcolor{blue}{Response here}\\ \\

\textcolor{blue}{Summary: this referee's insightful and
  constructive comments have led to a much improved submission.}

\end{document}

### 2023-05-06 inst/eurodance.Rmd
use sticker in Rmd
typo

@@
 sampling method.  Below, we use a resampling technique: 4000
-randomised jury observation are created, and the likelihood [under the
-Plackett-Luce likelihood function for the order statistics of the
+randomised jury observations are created, and the likelihood [under
+the Plackett-Luce likelihood function for the order statistics of the
 nationals' voters] is calculated.  The $p$-value is then the

### 2023-05-08 inst/jss_4382_rebuttal.tex
more work on issue #153

@@
 We have reached a decision regarding your submission to Journal of
-Statistical Software, #4832 “Generalized Plackett-Luce Likelihoods”.
+Statistical Software, \verb+#4832+, ``Generalized Plackett-Luce
+Likelihoods".
 
@@
 
-G1 The introduction should provide an overview of the contents of the manuscript.
-\textcolor{blue}{Response here}\\
+G1 The introduction should provide an overview of the contents of the
+manuscript. \\  \textcolor{blue}{New sentence added to manuscript:\\ \\
+Here I consider such generalized Plackett-Luce likelihood functions,
+and give an exact analysis of several simple cases.  I then show how
+this class of likelihood functions may be applied to a range of
+inference problems involving order statistics.  Illustrative examples,
+drawn Formula 1 motor racing, and track-and-field athletics, are
+given.}\\ \\
 
-In addition, the section headlines could be made more informative. Currently, they are:
-2. "Some simple cases"
-3. "Mann-Whitney test"
-4. "Multiple competitors: javelin"
-5. "Formula 1 motor racing: the constructors' championship"
+In addition, the section headlines could be made more
+informative. Currently, they are:
 
-Maybe the headlines could provide more of a summary of what the section
-contains.
-\textcolor{blue}{Response here}\\ \\
+2. ``Some simple cases"
+3. ``Mann-Whitney test"
+4. ``Multiple competitors: javelin"
+5. ``Formula 1 motor racing: the constructors' championship"
 
-G2 There should be a comparison with other software. Although the author
-states in the introduction that with hyper2 "It becomes possible to analyse
-a wider range of likelihood functions than standard Plackett-Luce", it is
-not clear to me which of the features described later on are unique to the
-hyper2 package and which can also be obtained from other packages (like,
-e.g., PlackettLuce).
-\textcolor{blue}{Response here}\\ \\
+Maybe the headlines could provide more of a summary of what the
+section contains.  \textcolor{blue}{The headings are indeed somewhat
+  uninformative and are now more in line with the remainder of the
+  manuscript.  I have changed what was section 4 to a {\em sub}section
+  with heading ``A generalization of the Mann-Whitney test using
+  generalized Plackett-Luce likelihood'' which I think is in keeping
+  with this comment.  The headings are now
+  \begin{itemize}
+  \item Exact analytical solution for some simple generalized
+    Plackett-Luce likelihood functions
+  \item An alternative to the Mann-Whitney test using generalized
+    Plackett-Luce likelihood
+  \item A generalization of the Mann-Whitney test using
+    generalized Plackett-Luce likelihood (now a subsection)
+  \item Formula 1 motor racing: the constructors' championship
+\end{itemize}}
+\rule{0mm}{10mm}
+
+G2 There should be a comparison with other software.  Although the
+author states in the introduction that with hyper2 ``It becomes
+possible to analyse a wider range of likelihood functions than
+standard Plackett-Luce", it is not clear to me which of the features
+described later on are unique to the \verb+hyper2+ package and which
+can also be obtained from other packages (like, e.g., PlackettLuce).
+
+\textcolor{blue}{Although the hyper2 package has functionality not
+  present in other implementations, the current manuscript focuses on
+  the new generalization represented in the package by the hyper3
+  class (that is, generalized Plackett-Luce likelihoods for order
+  statistics with clones of identical Bradle-Terry strength).  Turner
+  et. al. (2020), which I have cited, does a good job of setting out
+  the functionality of hyper2 not present in other packages.  I think
+  that an extended discussion of hyper2 unique functionality would be
+  a distraction from the focus of the manuscript.  But I agree that a
+  short statement documenting the unique aspects of the hyper2 package
+  would be useful to JSS readers, and I have added a short paragraph
+  to this effect.  Actually (in the spirit of full disclosure), I have
+  added quite a lot of machinery to the package since Turner's 2020
+  review and this gives me a welcome opportunity to mention some of
+  it.}
 
@@
 examples.
-\textcolor{blue}{Response here}\\ \\
 
-1.3 Specific comments\\
-S1 I'm not claiming any expertise in English language, but occasionally, I found
-the manuscript difficult to understand. For example, in the abstract it
-says: "The package . . . furnishes idiom" Couldn't that simply be: "The
-package . . . provides functionality"? On P. 13, it says "Here we use generalized
-Plackett-Luce to assess the constructors' performance". Isn't there
-a "model" missing? (Plackett-Luce is just a name and cannot be used).
-\textcolor{blue}{Response here}\\ \\
+\textcolor{blue}{Any optimization technique becomes slow when used
+  over a high-dimensional space and {\tt maxp()} is no exception.  The
+  examples chosen were well within its capabilities.  In addition, the
+  default implementation of {\tt maxp()} in the package uses a *very*
+  conservative set of assumptions, for example the default is to use
+  $n=10$ slightly differing start points and perform (by default)
+  $n-10$ independent optimizations, to check for
+  convergence.  I have come to realise that this is absurdly
+  over-engineered for the examples considered in the manuscript.  If
+  one is sure that the convergence is guaranteed (as it is in the
+  examples considered here), $n=1$ is perfectly adequate.  For
+  example, using the case given on P4:
+  {\tt\\ \\
+> (system.time(mH <- maxp(H,n=1)))\\
+   user  system elapsed \\
+  0.355   0.003   0.357 \\
+> maxp(H,n=1)\\
+         a          b          c \\
+0.21324090 0.08724824 0.69951086 \\
+  }\\ \\
+  The latest package version on github includes a discussion of this
+  issue and I have added a brief discussion (``function {\tt maxp()}
+  uses standard optimization techniques to locate the evaluate; it has
+  access to first derivatives of the log-likelihood and as such has
+  rapid convergence, if its objective function is reasonably
+  smooth").}\\ \\
+
+1.3 Specific comments\\ S1 I'm not claiming any expertise in English
+language, but occasionally, I found the manuscript difficult to
+understand. For example, in the abstract it says: "The package
+. . . furnishes idiom" Couldn't that simply be: "The package
+. . . provides functionality"?
+
+\textcolor{blue}{The original motivation for the package occurred when
+  I had a large complicated set of Bradley-Terry observations which I
+  wanted to translate into a likelihood function.  The R code became
+  extremely long and unweildy, and was almost impossible to check.  It
+  also became very slow to run.  What I wanted was a nice way to enter
+  this kind of observations, in such a way as to maintain both
+  computational efficiency [hence the use of the {\tt STL} map class]
+  {\bf and} human readability.  I think that the phrase ``furnishes
+  idiom" expresses this reasonably clearly.  But on reflection it does
+  sound a bit odd and I have rephrased the abstract accordingly.}\\ \\
+
+On P. 13, it says "Here we use generalized Plackett-Luce to assess the
+constructors' performance". Isn't there a "model" missing?
+(Plackett-Luce is just a name and cannot be used).
+\textcolor{blue}{The original text was too terse and I have
+  reworded.}\\ \\
 

### 2023-05-08 inst/jss_hyper3.Rnw
typo
Acked-by: Robin Hankin <hankin.robin@gmail.com>
more work on issue #153

@@
 
-  The \code{hyper2} package furnishes extensive idiom to work with
+  The \code{hyper2} package provides functionality to work with
   extensions of the Bradley-Terry probability model such as
-  Plackett-Luce likelihood functions, team strengths, and reified
-  entities (monsters).  Here, I present a generalization of
-  \code{hyper2} in which multiple entities are constrained to have
-  identical Bradley-Terry strengths.  A new \proglang{S3} class
-  \code{hyper3}, along with associated methods, is motivated and
-  introduced.  Three datasets are analysed, each analysis furnishing
-  new insight, and each highlighting different capabilities of the
-  package.
-}
-\Keywords{Plackett-Luce, Bradley-Terry, Mann-Whitney}
-\Plainkeywords{Plackett-Luce, Bradley-Terry, Mann-Whitney}
+  Plackett-Luce likelihood including team strengths and reified
+  entities (monsters).  The package allows one to use relatively
+  natural R idiom to manipulate such likelihood functions.  Here, I
+  present a generalization of \code{hyper2} in which multiple entities
+  are constrained to have identical Bradley-Terry strengths.  A new
+  \proglang{S3} class \code{hyper3}, along with associated methods, is
+  motivated and introduced.  Three datasets are analysed, each
+  analysis furnishing new insight, and each highlighting different
+  capabilities of the package.  } \Keywords{Plackett-Luce,
+  Bradley-Terry, Mann-Whitney} \Plainkeywords{Plackett-Luce,
+  Bradley-Terry, Mann-Whitney}
 
@@
 subset of~$\left[n\right]=\left\{1,2,\ldots,n\right\}$; numbers~$n_s$
-are integers which may be positive or negative.  It becomes possible
-to analyse a wider range of likelihood functions than standard
-Plackett-Luce~\citep{turner2020}.  For example, results involving
-incomplete order statistics or teams are tractable. Further, the
-introduction of reified entities allows one to consider
-draws~\citep{hankin2009}, noncompetitive tactics such as
+are integers which may be positive or negative.  The \pkg{hyper2}
+package has the ability to evaluate such likelihood functions at any
+point in $S_n$, thereby admitting a wide range of non-standard nulls
+such as order statistics on the $p_i$~\cite{hankin2020}.  It becomes
+possible to analyse a wider range of likelihood functions than
+standard Plackett-Luce~\citep{turner2020}.  For example, results
+involving incomplete order statistics or teams are tractable.
+Further, the introduction of reified entities (monsters) allows one to
+consider draws~\citep{hankin2009}, noncompetitive tactics such as
 collusion~\citep{hankin2020}, and the phenomenon of team cohesion
 wherein the team becomes stronger than the sum of its
-parts~\citep{hankin2010}.
+parts~\citep{hankin2010}.  Recent versions of the package include
+experimental functionality (\code{cheering()}) to investigate the
+relaxing of the assumption of conditional independence of the
+forward-ranking process.
 
@@
 \end{equation}
+
+Here I consider such generalized Plackett-Luce likelihood functions,
+and give an exact analysis of several simple cases.  I then show how
+this class of likelihood functions may be applied to a range of
+inference problems involving order statistics.  Illustrative examples,
+drawn Formula 1 motor racing, and track-and-field athletics, are
+given.
 
@@
 
+(function \code{maxp()} uses standard optimization techniques to
+locate the evaluate; it has access to first derivatives of the
+log-likelihood and as such has rapid convergence, if its objective
+function is reasonably smooth").
+
 The package provides a number of statistical tests on likelihood
@@
 
-\code{Package helper functions}
+\subsection{Package helper functions}
 
@@
 
-\section{Some simple cases}
+\section{Exact analytical solution for some simple generalized
+Plackett-Luce likelihood functions}
 
@@
 f_baa <- function(a){(1 - a) / (1 + a)}
-
-p <- function(f, ...){points(a, f(a) / max(f(a)), ...)}
-plot(0:1, 0:1, xlab = "p(a)", ylab = "Likelihood", type = "n")
+p <- function(f, ...){
+  a <- seq(from = 0, by = 0.005, to = 1)
+  points(a, f(a) / max(f(a)), ...)
+  }
+plot(0:1, 0:1, xlab = expression(p[a]), ylab = "Likelihood", type = "n")
 p(f_aab, type = "l", col = "black")
@@
 p(f_baa, type = "l", col = "blue")
-legend(x=0.85, y=0.7, col = c("black", "red", "blue"), lty=1, pch=NA,
-       legend=c("AAB", "ABA", "BAA"))
+text(0.8,0.8,"AAB")
+text(0.8,0.5,"ABA",col="red")
+text(0.8,0.15,"BAA",col="blue")
 abline(h = exp(-2), lty = 2)
@@
 
-\section{Mann-Whitney test}
+\section{An alternative to the Mann-Whitney test using generalized
+Plackett-Luce likelihood}
 
@@
   \begin{center}
-<<plotwilcoxlike,echo=FALSE,fig=TRUE>>=
+<<plotwilcoxlike,echo=FALSE,fig=TRUE,cache=TRUE>>=
 a <- seq(from = 0.02, to = 0.8, len = 40)
@@
 
-\section{Multiple competitors: Javelin}
+\subsection{A generalization of the Mann-Whitney test using
+generalized Plackett-Luce likelihood}
 
@@
 \begin{center}
-<<plottheloglikcont,echo=FALSE,fig=TRUE>>=
+<<plottheloglikcont,echo=FALSE,fig=TRUE,cache=TRUE>>=
 colnames(M) <-  as.character(Ves)
@@
 separately accumulates ranking-based points at each venue.  Here we
-use generalized Plackett-Luce to assess the constructors' performance.
-The following table, included in the \code{hyper2} package as a
-dataset, shows rankings for the first 9 venues of the 2021 season:
+use generalized a Plackett-Luce model to assess the constructors'
+performance.  The following table, included in the \code{hyper2}
+package as a dataset, shows rankings for the first 9 venues of the
+2021 season:
 

### 2023-05-09 inst/jss_4382_rebuttal.tex
more work on issue #153

@@
 of that line.
-\textcolor{blue}{Response here}\\ \\
+\textcolor{blue}{Done, figure much improved}\\ \\
 
@@
 "https://github.com/RobinHankin/hyper2".
-\textcolor{blue}{Response here}\\ \\
+\textcolor{blue}{Corrected}\\ \\
 
 S4 P. 7: "Wolf" should be "Wolfe".
-\textcolor{blue}{Response here}\\ \\
+\textcolor{blue}{Corrected.}\\ \\
 
 S5 P. 15, maxp(const2020, n=1): What does n do?
-\textcolor{blue}{Response here}\\ \\
+\textcolor{blue}{Argument $n$ discussed above under Timing}\\ \\
 
@@
 statistics. https://statmodeling.stat.columbia.edu/2022/01/27/regression-
-and-other-stories-free-pdf/)
-\textcolor{blue}{Response here}\\ \\
+and-other-stories-free-pdf/)\\
+
+\textcolor{blue}{I have not seen ROS before and have been studying
+  chapter 4 as suggested.  I make some discursive responses
+  here.\\ \\ I was fascinated by ``Researcher degrees of freedom''; I
+  am familiar with the concept but have not seen this particular, very
+  apposite, phrase used to describe it.  I would observe that my work
+  is not in the same class as discussed in ``solving the Bible code
+  puzzle'' [in which an unacceptably high amount of researcher DoF was
+    documented].  My own work is, for example, written to
+  reproducibility standards and all data and indeed code held on
+  public repositories [github and CRAN], open to scrutiny.  There is
+  very little ``wiggle room'' in my approach: the data is unambiguous,
+  and the model (Plackett-Luce) is clear and objective.\\ \\ I would
+  observe that Gelman uses the frequentist paradigm uncritically, a
+  position adopted by a decreasing minority of researchers.  For
+  example, the notion of ``bias'' is discussed; but many workers
+  (myself included) regard the bias of an estimator as unimportant: a
+  biased estimator with a small mean square error is not without
+  value.\\ \\I am not sure what relevance ``power'' has here: this is
+  the probability of {\em correctly} failing to reject the null when
+  it is false (and here, I did not fail to reject my null: perhaps we
+  should include an ensemble of Earths and visit a random one on which
+  ``I'' failed to reject the null).\\ \\ The comment ``the model is
+  almost certainly wrong'' is, I would argue, misleading.  All models
+  are wrong.  But some are useful.  My view is that the truth of a
+  statistical model [for some theological definition of truth] is a
+  poor criterion: no random variable is truly Gaussian, for example;
+  but the Student $t$-test is not useless as a result.  The issue is
+  whether non-normality materially affects the properties of the test.
+  Often this is not the case and the Gaussian approximation is
+  adequate: the $t$-test is sufficiently robust to be useful in a wide
+  range of situations.  True? No. Useful?  Consistent? Reliable?
+  Trustworthy?  Easy to use?  Familiar?  Well-understood? Standard?
+  Effective?  Powerful?  Yes.  The Student $t$-test has value, and so
+  does Plackett-Luce.  My position is that the framework set out in
+  the manuscript is plausible and intuitive.  I make no claim for
+  absolute correctness.\\ \\ But in answer to your question, the small
+  $p$-value and large likelihood ratio statistic certainly {\em is}
+  evidence for a real decrease.  The hypothesis of no real strength
+  change has low likelihood compared with the alternative of a drop in
+  Bradley-Terry strength.  Either Plackett-Luce is seriously wrong, or
+  PL is correct but some low-probability event has occurred which is
+  misleading us.  Alternatively, as I suggest, Mercedes really is
+  suffering from some genuine race-affecting issue [porpoising? Lewis
+    Hamilton getting older?  unaccustomed F1 budget caps?].  The {\em
+    strength} of this evidence is quantified by the small $p$-value
+  (here less than 0.002) or alternatively a likelihood ratio of
+  $e^{4.722}\simeq 112.4$, surely worthy of the epithet
+  ``strong''.}\\ \\
 

### 2023-05-09 inst/jss_hyper3.Rnw
more work on issue #153

@@
 \caption{Likelihood \label{ABAetc_likelihood} functions for
-observations $a\succ a\succ b$, $a\succ b\succ a$, $b\succ a\succ a$.}
+observations $a\succ a\succ b$, $a\succ b\succ a$, $b\succ a\succ a$.  Horizontal dotted
+line represents two units of support}
 \end{center}
@@
 datasets presented here are maintained under version control and
-available at \url{https:github.com/RobinHankin/hyper2}.  Given an
+available at \url{https://github.com/RobinHankin/hyper2}.  Given an
 order statistic of the type considered above, the

### 2023-05-10 README.Rmd
typo

@@
 <!-- badges: start -->
-[![Total Downloads](https://cranlogs.r-pkg.org/badges/grand-total/freealg)](https://CRAN.R-project.org/package=freealg)
-[![Build Status](https://travis-ci.org/RobinHankin/freealg.svg?branch=master)](https://travis-ci.org/RobinHankin/freealg)
-[![CRAN_Status_Badge](https://www.r-pkg.org/badges/version/freealg)](https://cran.r-project.org/package=freealg)
+[![Total Downloads](https://cranlogs.r-pkg.org/badges/grand-total/hyper2)](https://CRAN.R-project.org/package=hyper2)
+[![CRAN_Status_Badge](https://www.r-pkg.org/badges/version/hyper2)](https://cran.r-project.org/package=hyper2)
 <!-- badges: end -->
@@
 
-For more detail, see the package vignette
+For more detail, see the package vignette; type
 
-`vignette("freealg")`
+`vignette("hyper2")`
+
+at the R command line.
 

### 2023-05-10 inst/jss_4382_rebuttal.tex
work on issue #153

@@
   Often this is not the case and the Gaussian approximation is
-  adequate: the $t$-test is sufficiently robust to be useful in a wide
-  range of situations.  True? No. Useful?  Consistent? Reliable?
-  Trustworthy?  Easy to use?  Familiar?  Well-understood? Standard?
-  Effective?  Powerful?  Yes.  The Student $t$-test has value, and so
-  does Plackett-Luce.  My position is that the framework set out in
-  the manuscript is plausible and intuitive.  I make no claim for
-  absolute correctness.\\ \\ But in answer to your question, the small
+  adequate: the $t$-test is sufficiently robust to non-normality to be
+  useful in a wide range of situations.  Is the $t$-test True?  No.
+  Is it Useful?  Yes.  Consistent?  Yes.  Reliable?  Yes.  Easy to
+  use?  Yes.  Familiar?  Yes.  Well-understood?  Yes.  Powerful?  Yes.
+  The Student $t$-test has value, and so does Plackett-Luce.  My
+  position is that the framework set out in the manuscript is
+  plausible and intuitive.  I make no claim for absolute
+  correctness.\\ \\ But in answer to your question, the small
   $p$-value and large likelihood ratio statistic certainly {\em is}
@@
 
-S7 The references are lacking their DOIs.
-\textcolor{blue}{Response here}\\ \\
+S7 The references are lacking their DOIs.\\ \textcolor{blue}{Done
+  (except for two books---for which I have followed APA6---and R
+  itself, for which I used the result of {\tt citation()})} \\ \\
 
-2 Review of software
+2 Review of software\\
 
-The package includes many functions with a similar name. E.g., ordervec2supp,
-ordervec2supp3, ordervec2supp3a, ordertable2supp, ordertable2supp3. Maybe
-there could be some overview of which to use for what purpose.
-
-\textcolor{blue}{Response here}\\ \\
+The package includes many functions with a similar name. E.g., {\tt
+  ordervec2supp()}, {\tt ordervec2supp3()}, {\tt ordervec2supp3a()},
+{\tt ordertable2supp()} , {\tt ordertable2supp3()}. Maybe there could
+be some overview of which to use for what
+purpose.\\ \textcolor{blue}{An overview is given in {\tt
+    ordervec2supp()} and {\tt ordervec2supp3()}, which are
+  cross-referenced to each other.  Order tables are documented at {\tt
+    ordertable.Rd}; order vectors at {\tt ordervec2supp.Rd}.  The
+  functions mentioned by the referee all follow a consistent
+  nomenclature: as it says in {\tt hyper3.Rd}, ``Functionality for
+  {\tt hyper3} objects is generally indicated by adding a 3 to
+  function names, eg {\tt gradient()} goes to {\tt gradient3()}''.}
 
@@
 
-\textcolor{blue}{Response here}\\ \\
+\textcolor{blue}{I would agree with this.}\\ \\
 
-Even though this is just a code snippet and the hyper2 package was already
-published elsewhere, the author should provide some introductory elaborations
-on the Plackett-Luce model. Along these lines it would also be helpful to give
-a quick overview of the hyper2 package, and mention other related packages. As it
-is now it is difficult for a reader who's not super familiar with PL-models and
-hyper2, respectively, to follow the paper. Anyway, I find the runners example
-as motivating example very illustrative, and actually like the examples throughout
-the paper.
+Even though this is just a code snippet and the hyper2 package was
+already published elsewhere, the author should provide some
+introductory elaborations on the Plackett-Luce model. Along these
+lines it would also be helpful to give a quick overview of the hyper2
+package, and mention other related packages. As it is now it is
+difficult for a reader who's not super familiar with PL-models and
+hyper2, respectively, to follow the paper. Anyway, I find the runners
+example as motivating example very illustrative, and actually like the
+examples throughout the paper.\\ \textcolor{blue}{Thank you!  I too
+  have enjoyed the runner paradigm.  But I am aware that the new
+  likelihood function is novel and likely to be challenging for those
+  not familiar with {\tt hyper2}.  I have added some discussion of
+  {\tt hyper2} and its design principles.\\ \\I would not agree with
+  the characterization as a code snippet.  The {\tt C++} code base for
+  {\tt hyper3} functionality (318 lines) is similar to that for {\tt
+    hyper2} (396 lines).  But admittedly {\tt hyper3} functionality
+  does leverage many of the S3 methods used in {\tt hyper2}.}\\ \\
 
-\textcolor{blue}{Response here}\\ \\
-
-Some more aesthetic effort should be made in the code file (e.g., providing
-comments). The code looks like a simple purl() extraction.
-
-\textcolor{blue}{Response here}\\ \\
+Some more aesthetic effort should be made in the code file (e.g.,
+providing comments). The code looks like a simple purl()
+extraction.\\ \textcolor{blue}{The JSS style file forbids comments
+  within the code but encourages authors to comment in the normal
+  LaTeX text.  Each code chunk is now either preceded by a description
+  of its content, or followed with ``Above, we see\ldots'' or
+  equivalent.}
 

### 2023-05-10 inst/jss_hyper3.Rnw
work on issue #153

@@
 However, the majority of software is written in the \proglang{R}
-computer language~\citep{rcore2022}, which includes extensive
+computer language~\citep{rcore2023}, which includes extensive
 functionality for working with such likelihood functions:

### 2023-05-15 inst/jss_4382_rebuttal.tex
more work on issue #153

@@
 
-On P. 13, it says "Here we use generalized Plackett-Luce to assess the
-constructors' performance". Isn't there a "model" missing?
+On P. 13, it says ``Here we use generalized Plackett-Luce to assess
+the constructors' performance''. Isn't there a ``model'' missing?
 (Plackett-Luce is just a name and cannot be used).
@@
 
-S3 P. 7: "https:github.com/RobinHankin/hyper2" should be
-"https://github.com/RobinHankin/hyper2".
+S3 P. 7: ``https:github.com/RobinHankin/hyper2'' should be
+``https://github.com/RobinHankin/hyper2''.
 \textcolor{blue}{Corrected}\\ \\
 
-S4 P. 7: "Wolf" should be "Wolfe".
+S4 P. 7: ``Wolf'' should be ``Wolfe''.
 \textcolor{blue}{Corrected.}\\ \\
@@
 
-S6 P. 16, "we see strong evidence for a real decrease in the strength": I'd tone
-that down. The model is almost certainly wrong, so it does not measure
-the "real decrease." Also, power is unknown, so it is difficult to say what
-"evidence" there is against H0. (See, e.g., Gelman, Regression an other
-stories, Chapter 4: Time to unlearn what you thought you knew about
+S6 P. 16, ``we see strong evidence for a real decrease in the
+strength'': I'd tone that down.  The model is almost certainly wrong,
+so it does not measure the ``real decrease.'' Also, power is unknown,
+so it is difficult to say what ``evidence'' there is against H0. (See,
+e.g., Gelman, Regression an other stories, Chapter 4: Time to unlearn
+what you thought you knew about
 statistics. https://statmodeling.stat.columbia.edu/2022/01/27/regression-
@@
 
-\textcolor{blue}{I have not seen ROS before and have been studying
-  chapter 4 as suggested.  I make some discursive responses
-  here.\\ \\ I was fascinated by ``Researcher degrees of freedom''; I
-  am familiar with the concept but have not seen this particular, very
-  apposite, phrase used to describe it.  I would observe that my work
-  is not in the same class as discussed in ``solving the Bible code
-  puzzle'' [in which an unacceptably high amount of researcher DoF was
-    documented].  My own work is, for example, written to
+\textcolor{blue}{I have not seen {\em Regression and other stories}
+  before.  It is very good.  I have been studying chapter 4 as
+  suggested.  I make some discursive responses here.\\ \\ I was
+  fascinated by ``Researcher degrees of freedom''; I am familiar with
+  the concept but have not seen this particular, very apposite, phrase
+  used to describe it.  I would observe that my work is not in the
+  same class as discussed in ``solving the Bible code puzzle'' [in
+    which an unacceptably high amount of researcher DoF was
+    documented].  The current manuscript is, for example, written to
   reproducibility standards and all data and indeed code held on
@@
   and the model (Plackett-Luce) is clear and objective.\\ \\ I would
-  observe that Gelman uses the frequentist paradigm uncritically, a
+  observe that ROS uses the frequentist paradigm uncritically, a
   position adopted by a decreasing minority of researchers.  For
@@
   biased estimator with a small mean square error is not without
-  value.\\ \\I am not sure what relevance ``power'' has here: this is
-  the probability of {\em correctly} failing to reject the null when
-  it is false (and here, I did not fail to reject my null: perhaps we
-  should include an ensemble of Earths and visit a random one on which
-  ``I'' failed to reject the null).\\ \\ The comment ``the model is
-  almost certainly wrong'' is, I would argue, misleading.  All models
-  are wrong.  But some are useful.  My view is that the truth of a
-  statistical model [for some theological definition of truth] is a
-  poor criterion: no random variable is truly Gaussian, for example;
-  but the Student $t$-test is not useless as a result.  The issue is
-  whether non-normality materially affects the properties of the test.
-  Often this is not the case and the Gaussian approximation is
-  adequate: the $t$-test is sufficiently robust to non-normality to be
-  useful in a wide range of situations.  Is the $t$-test True?  No.
-  Is it Useful?  Yes.  Consistent?  Yes.  Reliable?  Yes.  Easy to
-  use?  Yes.  Familiar?  Yes.  Well-understood?  Yes.  Powerful?  Yes.
-  The Student $t$-test has value, and so does Plackett-Luce.  My
-  position is that the framework set out in the manuscript is
-  plausible and intuitive.  I make no claim for absolute
-  correctness.\\ \\ But in answer to your question, the small
-  $p$-value and large likelihood ratio statistic certainly {\em is}
-  evidence for a real decrease.  The hypothesis of no real strength
-  change has low likelihood compared with the alternative of a drop in
-  Bradley-Terry strength.  Either Plackett-Luce is seriously wrong, or
-  PL is correct but some low-probability event has occurred which is
-  misleading us.  Alternatively, as I suggest, Mercedes really is
-  suffering from some genuine race-affecting issue [porpoising? Lewis
-    Hamilton getting older?  unaccustomed F1 budget caps?].  The {\em
-    strength} of this evidence is quantified by the small $p$-value
-  (here less than 0.002) or alternatively a likelihood ratio of
-  $e^{4.722}\simeq 112.4$, surely worthy of the epithet
-  ``strong''.}\\ \\
+  value.  I would observe that the concept of bias does not arise in
+  either the Bayesian or likelihood paradigms as used in the current
+  manuscript.\\ \\I am not sure what relevance ``power'' has here:
+  this is the probability of {\em correctly} failing to reject the
+  null when it is false (and here, I did not fail to reject my null:
+  perhaps we should include an ensemble of Earths and visit a random
+  one on which ``I'' failed to reject the null).\\ \\ The comment
+  ``the model is almost certainly wrong'' is, I would argue,
+  misleading.  All models are wrong.  But some are useful.  My view is
+  that the truth of a statistical model [for some theological
+    definition of truth] is a poor criterion: no random variable is
+  truly Gaussian, for example; but the Student $t$-test is not useless
+  as a result.  The issue is whether non-normality materially affects
+  the properties of the test.  Often this is not the case and the
+  Gaussian approximation is adequate: the $t$-test is sufficiently
+  robust to non-normality to be useful in a wide range of situations.
+  Is the $t$-test True?  No.  Is it Useful?  Yes.  Consistent?  Yes.
+  Reliable?  Yes.  Easy to use?  Yes.  Familiar?  Yes.
+  Well-understood?  Yes.  Powerful?  Yes.  The Student $t$-test has
+  value, and so does Plackett-Luce.  My position is that the framework
+  set out in the manuscript is plausible and intuitive.  I make no
+  claim for absolute correctness.\\ \\ But in answer to your question,
+  the small $p$-value and large likelihood ratio statistic certainly
+  {\em is} evidence for a real decrease.  The hypothesis of no real
+  strength change has low likelihood compared with the alternative of
+  a drop in Bradley-Terry strength.  Either Plackett-Luce is seriously
+  wrong, or PL is at least approximately correct, and some
+  low-probability event has occurred which is misleading us.
+  Alternatively, as I suggest, Mercedes really is suffering from some
+  genuine race-affecting issue [porpoising? Lewis Hamilton getting
+    older?  unaccustomed F1 budget caps?].  The {\em strength} of this
+  evidence is quantified by the small $p$-value (here less than 0.002)
+  or alternatively a likelihood ratio of $e^{4.722}\simeq 112.4$,
+  surely worthy of the epithet ``strong''.}\\ \\
 

### 2023-05-15 inst/jss_4382_rebuttal.tex
further work on issue #153

@@
 
-Short story: I have accommodated all the comments with rewording and
-clarification.
+\textcolor{blue}{Short story: I have accommodated all the comments with
+rewording and clarification.}
 
@@
 inference problems involving order statistics.  Illustrative examples,
-drawn Formula 1 motor racing, and track-and-field athletics, are
+drawn from Formula 1 motor racing, and track-and-field athletics, are
 given.}\\ \\
@@
 Maybe the headlines could provide more of a summary of what the
-section contains.  \textcolor{blue}{The headings are indeed somewhat
+section contains. \\ \\ \textcolor{blue}{The headings are indeed somewhat
   uninformative and are now more in line with the remainder of the
@@
   class (that is, generalized Plackett-Luce likelihoods for order
-  statistics with clones of identical Bradle-Terry strength).  Turner
+  statistics with clones of identical Bradley-Terry strength).  Turner
   et. al. (2020), which I have cited, does a good job of setting out
@@
 
-\textcolor{blue}{Any optimization technique becomes slow when used
-  over a high-dimensional space and {\tt maxp()} is no exception.  The
+\textcolor{blue}{All optimization techniques become slow when used
+  over high-dimensional spaces and {\tt maxp()} is no exception.  The
   examples chosen were well within its capabilities.  In addition, the
@@
   $n=10$ slightly differing start points and perform (by default)
-  $n-10$ independent optimizations, to check for
-  convergence.  I have come to realise that this is absurdly
-  over-engineered for the examples considered in the manuscript.  If
-  one is sure that the convergence is guaranteed (as it is in the
-  examples considered here), $n=1$ is perfectly adequate.  For
-  example, using the case given on P4:
+  $n-10$ independent optimizations, to check for convergence.  My
+  thinking has matured over the last few months, and I have come to
+  realise that this is absurdly over-engineered for the examples
+  considered in the manuscript.  If one is sure that the convergence
+  is guaranteed (as it is in the examples considered here), $n=1$ is
+  perfectly adequate.  For example, using the case given on P4:
   {\tt\\ \\
@@
 language, but occasionally, I found the manuscript difficult to
-understand. For example, in the abstract it says: "The package
-. . . furnishes idiom" Couldn't that simply be: "The package
-. . . provides functionality"?
+understand. For example, in the abstract it says: ``The package
+. . . furnishes idiom'' Couldn't that simply be: ``The package
+. . . provides functionality''?
 
@@
   and the model (Plackett-Luce) is clear and objective.\\ \\ I would
-  observe that ROS uses the frequentist paradigm uncritically, a
-  position adopted by a decreasing minority of researchers.  For
+  observe that ROS uses the frequentist paradigm, a
+  position adopted by a decreasing proportion of researchers.  For
   example, the notion of ``bias'' is discussed; but many workers
@@
   this is the probability of {\em correctly} failing to reject the
-  null when it is false (and here, I did not fail to reject my null:
-  perhaps we should include an ensemble of Earths and visit a random
-  one on which ``I'' failed to reject the null).\\ \\ The comment
-  ``the model is almost certainly wrong'' is, I would argue,
-  misleading.  All models are wrong.  But some are useful.  My view is
-  that the truth of a statistical model [for some theological
-    definition of truth] is a poor criterion: no random variable is
-  truly Gaussian, for example; but the Student $t$-test is not useless
-  as a result.  The issue is whether non-normality materially affects
-  the properties of the test.  Often this is not the case and the
-  Gaussian approximation is adequate: the $t$-test is sufficiently
-  robust to non-normality to be useful in a wide range of situations.
-  Is the $t$-test True?  No.  Is it Useful?  Yes.  Consistent?  Yes.
-  Reliable?  Yes.  Easy to use?  Yes.  Familiar?  Yes.
-  Well-understood?  Yes.  Powerful?  Yes.  The Student $t$-test has
-  value, and so does Plackett-Luce.  My position is that the framework
-  set out in the manuscript is plausible and intuitive.  I make no
-  claim for absolute correctness.\\ \\ But in answer to your question,
-  the small $p$-value and large likelihood ratio statistic certainly
-  {\em is} evidence for a real decrease.  The hypothesis of no real
-  strength change has low likelihood compared with the alternative of
-  a drop in Bradley-Terry strength.  Either Plackett-Luce is seriously
-  wrong, or PL is at least approximately correct, and some
-  low-probability event has occurred which is misleading us.
-  Alternatively, as I suggest, Mercedes really is suffering from some
-  genuine race-affecting issue [porpoising? Lewis Hamilton getting
-    older?  unaccustomed F1 budget caps?].  The {\em strength} of this
+  null when it is false.  In this case the null was not rejected, a
+  type II error was not committed.  It is not clear whether a
+  preferred alternative hypothesis exists (and further, there seems to
+  be no natural ensemble to consider).\\ \\ The comment ``the model is
+  almost certainly wrong'' is, I would argue, misleading.  All models
+  are wrong.  But some are useful.  My view is that the truth of a
+  statistical model [for some theological definition of truth] is a
+  poor criterion: no random variable is truly Gaussian, for example;
+  but the Student $t$-test is not useless as a result.  The issue is
+  whether non-normality materially affects the properties of the test.
+  Often this is not the case and the Gaussian approximation is
+  adequate: the $t$-test is sufficiently robust to non-normality to be
+  useful in a wide range of situations.  Is the $t$-test True?  No.
+  Is it Useful?  Yes.  Consistent?  Yes.  Reliable?  Yes.  Easy to
+  use?  Yes.  Familiar?  Yes.  Well-understood?  Yes.  Powerful?  Yes.
+  The Student $t$-test has value, and so does Plackett-Luce.  My
+  position is that the framework set out in the manuscript is
+  plausible and intuitive.  I make no claim for absolute
+  correctness.\\ \\ But in answer to your question, the small
+  $p$-value and large likelihood ratio statistic certainly {\em is}
+  evidence for a real decrease.  The hypothesis of no real strength
+  change has low likelihood compared with the alternative of a drop in
+  Bradley-Terry strength.  Either Plackett-Luce is seriously wrong, or
+  PL is at least approximately correct, and some low-probability event
+  has occurred which is misleading us.  Alternatively, as I suggest,
+  Mercedes really is suffering disproportionately from some genuine
+  race-affecting issue [porpoising? Lewis Hamilton getting older?
+    unaccustomed F1 budget caps?].  The {\em strength} of this
   evidence is quantified by the small $p$-value (here less than 0.002)
@@
   {\tt hyper3} objects is generally indicated by adding a 3 to
-  function names, eg {\tt gradient()} goes to {\tt gradient3()}''.}
+  function names, eg {\tt gradient()} goes to {\tt gradient3()}''.  I
+  have added a brief pointer in the manuscript to these package
+  files.}
 
@@
 
-\textcolor{blue}{Summary: this referee's insightful and
-  constructive comments have led to a much improved submission.}
+\textcolor{blue}{Summary: the insightful and constructive comments of
+  the review have led to a much improved submission.}
 

### 2023-05-15 inst/jss_hyper3.Rnw
further work on issue #153

@@
 computational support for generalized
-Plackett-Luce~\citep{plackett1975} likelihood functions.  Our
+Plackett-Luce~\citep{plackett1975} likelihood functions.  The
 preferred interpretation is a race (as in track and field athletics):
@@
 \noindent where~$\mathcal{O}$ is a set of observations and~$s$ a
-subset of~$\left[n\right]=\left\{1,2,\ldots,n\right\}$; numbers~$n_s$
-are integers which may be positive or negative.  The \pkg{hyper2}
-package has the ability to evaluate such likelihood functions at any
-point in $S_n$, thereby admitting a wide range of non-standard nulls
-such as order statistics on the $p_i$~\cite{hankin2020}.  It becomes
-possible to analyse a wider range of likelihood functions than
-standard Plackett-Luce~\citep{turner2020}.  For example, results
-involving incomplete order statistics or teams are tractable.
-Further, the introduction of reified entities (monsters) allows one to
-consider draws~\citep{hankin2009}, noncompetitive tactics such as
+subset of~$\left\{1,2,\ldots,n\right\}$; numbers~$n_s$ are integers
+which may be positive or negative.  The \pkg{hyper2} package has the
+ability to evaluate such likelihood functions at any point in $S_n$,
+thereby admitting a wide range of non-standard nulls such as order
+statistics on the $p_i$~\citep{hankin2017}.  It becomes possible to
+analyse a wider range of likelihood functions than standard
+Plackett-Luce~\citep{turner2020}.  For example, results involving
+incomplete order statistics or teams are tractable.  Further, the
+introduction of reified entities (monsters) allows one to consider
+draws~\citep{hankin2009}, noncompetitive tactics such as
 collusion~\citep{hankin2020}, and the phenomenon of team cohesion
@@
 \frac{p_a}{ p_a+ p_b    }\cdot
-\frac{p_b}{      p_b    },\qquad p_a+p_b+p_c=1
+\frac{p_b}{      p_b    },\qquad p_a+p_b+p_c=1.
 \end{equation}
@@
 inference problems involving order statistics.  Illustrative examples,
-drawn Formula 1 motor racing, and track-and-field athletics, are
+drawn from Formula 1 motor racing, and track-and-field athletics, are
 given.
@@
 show how the \proglang{map} class of the Standard Template Library is
-used: \proglang{weightedplayervector} objects map strings to long
-doubles (specifically, mapping player names to their multiplicities),
-and objects of class \proglang{hyper3} are maps from a
-\proglang{weightedplayervector} object to long doubles.  One advantage
-of this is efficiency: Search, removal, and insertion operations have
-logarithmic complexity.  As an example, the following \proglang{C++}
-pseudo code would create a log-likelihood function for the first term
-in Equation~\ref{plackettluce}:
+used with \proglang{weightedplayervector} objects mapping strings to
+long doubles (specifically, mapping player names to their
+multiplicities), and objects of class \proglang{hyper3} are maps from
+a \proglang{weightedplayervector} object to long doubles.  One
+advantage of this is efficiency: Search, removal, and insertion
+operations have logarithmic complexity.  As an example, the following
+\proglang{C++} pseudo code would create a log-likelihood function for
+the first term in Equation~\ref{plackettluce}:
 
@@
 p_a=0.9,p_b=0.05,p_c=0.05$ and $H_2\colon p_a=0.01,p_b=0.01,p_c=0.98$,
-we can evaluate these hypotheses using generic function
+and we may evaluate these hypotheses using generic function
 \code{loglik()}:
@@
 statistics of this type.  Function \code{ordervec2supp3()}, for
-example, can be used to generate a loglikelihood function for
+example, can be used to generate a log-likelihood function for
 Equation~\ref{PL_like}:
@@
 
-We may find a maximum likelihood estimate for the players' strengths,
-using generic function \code{maxp()}, dispatching to a specialist
-\code{hyper3} method:
+(the package gives extensive documentation at
+\code{ordervec2supp.Rd}).  We may find a maximum likelihood estimate
+for the players' strengths, using generic function \code{maxp()},
+dispatching to a specialist \code{hyper3} method:
 
@@
 log-likelihood and as such has rapid convergence, if its objective
-function is reasonably smooth").
+function is reasonably smooth).
 
 The package provides a number of statistical tests on likelihood
-functions, modified from~\citep{hankin2017} to work with \code{hyper3}
+functions, modified from~\citet{hankin2017} to work with \code{hyper3}
 objects.  For example, we may assess the hypothesis that all three
@@
 \code{attemptstable2supp3()} which facilitate the analysis of commonly
-encountered result formats; an example is given below.
+encountered result formats.  Package documentation for order tables is
+given at \code{ordertable.Rd} and an example is given below.
 
@@
 S <- function(a){log(a * (1 - a) / ((1 + a) * (3 - 2 * sqrt(2))))}
-plot(a, S(a), type = 'b')
+plot(a, S(a), type = 'b',xlab=expression(p[a]),ylab="support")
 abline(h = c(0, -2))
@@
 L <- sapply(a, function(p){loglik(p, Hxy)})
-plot(a, L - max(L), type = 'b')
+plot(a, L - max(L), type = 'b',xlab=expression(p[a]),ylab="likelihood")
 abline(h = c(0, -2))
@@
 so one has access to the standard likelihood-based methods, such as
-finding and displaying the maximum likelihood estimate:
+finding and displaying the maximum likelihood estimate, shown in
+Figure~\ref{javelinbradleyterry}.  From this, we see that Vadlejch has
+the highest estimated Bradley-Terry strength, but further analysis
+with \code{equalp.test()} reveals that there is no strong evidence in
+the dataset to reject the hypothesis of equal competitive strength
+($p=0.26$), or that Vadlejch has a strength higher than the null value
+of $\frac{1}{8}$ ($p=0.1$).
 
@@
 (mj <- maxp(javelin))
-dotchart(mj, pch = 16)
+dotchart(mj, pch = 16,xlab="Estimated Bradley-Terry strength")
 @
-\caption{Maximum likelihood estimate for javelin throwers' Bradley-Terry strengths.}
+\caption{Maximum \label{javelinbradleyterry} likelihood estimate for javelin throwers' Bradley-Terry strengths.}
 \end{center}
 \end{figure}
-
-Above, we see that Vadlejch has the highest estimated Bradley-Terry
-strength, but further analysis with \code{equalp.test()} reveals that
-there is no strong evidence in the dataset to reject the hypothesis of
-equal competitive strength ($p=0.26$), or that Vadlejch has a strength
-higher than the null value of $\frac{1}{8}$ ($p=0.1$).
 
@@
 separately accumulates ranking-based points at each venue.  Here we
-use generalized a Plackett-Luce model to assess the constructors'
+use a generalized Plackett-Luce model to assess the constructors'
 performance.  The following table, included in the \code{hyper2}
@@
 cars, and also that constructors come and go, with two 2020 teams
-dropping out between years and two joining.  Package idiom to test
-this statistically would be to define a combined likelihood function
-for both years, keeping track of the year:
+dropping out between years and two joining.  We may test this
+statistically by defining a combined likelihood function for both
+years, keeping track of the year:
 

### 2023-05-16 inst/jss_hyper3.Rnw
fixes #205

@@
 possible rank statistics: (i), $a\succ a\succ b$ with probability
-$\frac{a^2}{1+a}$; (ii), $a\succ b\succ a$, with
-$\frac{1-2a-a^2}{(1+a)^2}$, (iii), $b\succ a\succ a$ at
+$\frac{2a^2}{1+a}$; (ii), $a\succ b\succ a$, with
+$\frac{2a(1-a)}{(1+a)}$, (iii), $b\succ a\succ a$ at
 $\frac{1-a}{1+a}$.  Likelihood functions for these order statistics
@@
 \begin{equation}
-\mathcal{I}(a)=\frac{1+16a+18a^2-4a^3-15a^4-7a^5-a^6}{a(1-a)(1+a)^3}
+\mathcal{I}(a)=2\frac{1+a+a^2}{a(1-a)(1+a)^2}
 \end{equation}
 
-which has a minimum of about $13.9$ at at about $a=0.575$.  We can
+which has a minimum of about $6.21$ at at about $a=0.522$.  We can
 compare this with the Fisher information matrix ${\mathcal I}$, for
@@
 f_aab <- function(a){a^2 / (1 + a)}
-f_aba <- function(a){a * (1 - a) / (1 + a)^2}
+f_aba <- function(a){a * (1 - a) / (1 + a)}
 f_baa <- function(a){(1 - a) / (1 + a)}
@@
 text(0.8,0.8,"AAB")
-text(0.8,0.5,"ABA",col="red")
-text(0.8,0.15,"BAA",col="blue")
+text(0.8,0.4,"ABA",col="red")
+text(0.8,0.05,"BAA",col="blue")
 abline(h = exp(-2), lty = 2)

### 2023-05-22 inst/monster_vs_lambda.Rmd
slight change

--- 2023-03-21 inst/monster_vs_lambda.Rmd
@@
 ```{r,label=ttrydefflikefvc,cache=TRUE}
-Mval <- seq(from=0.0,to=0.65,len=15)
+Mval <- seq(from=0.01,to=0.65,len=15)
 lval <- seq(from=0.2,to=5,by = 0.4) # we need lambda=1 (this is the null)
 jj <- as.matrix(expand.grid(M=Mval,lam=lval))
-options(use_alabama = FALSE)
+options(use_alabama = FALSE) # setting to TRUE gives a "initial value not feasible" error, go figure
 L <- apply(jj,1,like2)

### 2023-05-22 inst/simplified_likelihood.Rmd
correct some stupid mistakes

--- 2023-05-02 inst/simplified_likelihood.Rmd
@@
 ```{r setup, include=FALSE}
+{ # this line vandalises this file: nonfinishers are ignored and should not be!
 knitr::opts_chunk$set(echo = TRUE)
+library("hyper2")
 ```
 
-Consider a race between $n+2$ competitors of strength $p_1$, $p_2$, and $n$ clones, all of strength $p=1-p_1-p_2$.
-Our observation is that $p_1$ places $n_1$, and $p_2$ places $n_2>n_1$.
-The order statistic is thus
+Consider a race between $n+2$ competitors of strength $p_1$, $p_2$,
+and $n$ clones, all of strength $p=1-p_1-p_2$.  Our observation is
+that $p_1$ places $n_1$, and $p_2$ places $n_2>n_1$.  The order
+statistic is thus
 
@@
 
-Observing that the fifth line may be neglected, R idiom for this would be as follows
+```{r}
+named_drivers <- function(n,...){ # twodrivers(20, ham=3, bottas=5)  # NB repeats allowed!
+   a <- list(...)
+   stopifnot(all(table(as.vector(unlist(a)))==1)) # stops two players having the same result
+   s <- rep("other",n)
+   for(i in seq_along(a)){
+     s[a[[i]]] <- names(a)[i]   # ham=3 -> s[3] <- "ham"
+   }
+   return(ordervec2supp3(s))  
+}
+```
 
 ```{r}
-library("hyper2")
-simplike <- function(n1,n2,n){
+(H <- named_drivers(20,ham=3,bott=5))
+```
+
+Thus `H` is a likelihood function on the assumption that the other
+drivers are all clones of identical strength.  Maximizing this:
+
+```{r readtable}
+a <- read.table("formula1_2019.txt")[1:2,1:21]
+a
+f <- function(i){
+  jj <- as.list(a[,i])
+  suppressWarnings(jj <- as.numeric(jj))
+  names(jj) <- rownames(a)
+  jj[!is.na(jj)]   # NA entries ignored - THESE SHOULD BE NONFINISHERS
+}
+
 H <- hyper3()
-  H[c(p =1)] <- n2-2
-  H[c(p1=1)] <- 1
-  H[c(p2=1)] <- 1
-  for(i in (n-n1+1):n)     { H[c(p1=1,p2=1,p=i)]  %<>% dec }
-  for(i in (n-n2+2):(n-n1+1)){ H[c(p2=1,   p=i)]  %<>% dec }
-  return(H)
+for(i in seq_len(ncol(a))){
+  jj <- f(i)
+  if(all(!is.na(as.numeric(unlist(jj))))){
   }
-simplike(2,6,11)
-simplike(1,6,11)
-simplike(1,2,11)
+}
+H
 ```
+
+```{r maxh,cache=TRUE}
+maxp(H)
+```
+Thus we narrowly fail to reject the null at $\alpha = 0.05$.  Now
+compare with the actual 2019 season:
+
+```{r actual2019,cache=TRUE}
+F1_2019 <- ordertable2supp(read.table("formula1_2019.txt",header=TRUE)[,1:21])
+samep.test(F1_2019,c("Hamilton","Bottas"))
+```
+
+We see a much smaller p-value when using the complete dataset.  It is
+not clear whether we would expect a larger or a smaller p-value here,
+but it is not surprising to see a difference.  On the simplified case,
+the estimated strength of `other` is about 0.016 (under the
+alternative); but in the full case, most of the other drivers have
+strength considerably more thn this.  I do not have a good way of
+thinking about this at this time.  One other feature is that the
+simplified likelihood results in different MLEs:
+
+* simplified: $\widehat{p_\mbox{Ham}} = 0.49$, $\widehat{p_\mbox{Bot}}=0.31$
+* full: $\widehat{p_\mbox{Ham}} = 0.38$, $\widehat{p_\mbox{Bot}}=0.13$
+
+Consider the log-contrasts, $\log(p_1/p_2)$ giving

### 2023-07-24 inst/jss_4832_minorcomments_rebuttal.tex
almost finished issue #153

\documentclass[12pt]{article}
\usepackage{xcolor}
\begin{document}

\section*{JSS4832: Generalized Plackett-Luce Likelihoods by Robin
K. S. Hankin}

Below, the reviewer's comments are in black, and the replies to the
issues are in \textcolor{blue}{blue}.  I have indicated changes to
the manuscript where appropriate.

\textcolor{blue}{Short story: I have accommodated all the comments with
rewording and clarification.}

From the editorial team:

I had a closer look at the revision of 4832 and, as far as I'm
concerned, the paper can be accepted as it is. However, I wanted to
draw your attention to the author's rebuttal file, p.7. It seems that
he doesn't agree that this paper should be published as code snippet,
as opposed to a regular paper. I'm not sure whether this was
previously communicated to the author by the editorial team. Anyway, I
think it would be good if the editorial team could clarify this issue
directly with the author.

\textcolor{blue}{Happy to classify as a code snippet if the editors
  wish, although I stand by my previous comments.  Happy either way.}

Also, I think he misunderstood my concerns about the supplemental R
code file (see bottom of p.7 in the rebuttal file). What I'm asking
for is to make the R code look nicer with proper comments, as opposed
to just pull it from the Rnw file. In addition, the code starts with

\begin{verbatim}
source("loadlibrary.R",echo=FALSE) # to avoid stupid messages
opts_chunk$set(cache=TRUE, autodep=TRUE)
\end{verbatim}

This is not reproducible. Also, the author doesn't even load the hyper2 package.

\textcolor{blue}{Fixed, code fully reproducible}

So, while I think the paper is in good shape, the code file is not. But this can be easily be fixed.

Manuscript style comments:

\begin{itemize}
  \item Please use "R>" as the command prompt, rather than ">".
\end{itemize}

\textcolor{blue}{fixed}

\begin{itemize}
\item The code presented in the manuscript should not contain comments
within the verbatim code. Instead the comments should be made in the
normal LaTeX text.
\end{itemize}

\textcolor{blue}{fixed}

\begin{itemize}
  \item For the code layout in R publications, we typically distinguish
input/output using Sinput/Soutput (or equivalently
CodeInput/CodeOutput). Unless there are special reasons to format it
differently, the input should use the text width (up to 76 or 77
characters) and be indented by two spaces, e.g.,

\begin{verbatim}
\begin{Sinput}
R> example_model <- lm(response ~ variable1 + variable2 + variable3,

weights = w, data = mydata)
\end{Sinput}
\end{verbatim}
\end{itemize}

\textcolor{blue}{fixed}

 As a reminder, please make sure that: \proglang, \pkg and \code have
 been used for highlighting throughout the paper (including titles and
 references), except where explicitly escaped.

\textcolor{blue}{fixed}

References:

\begin{itemize}
      \item MATLAB (not: Matlab, matlab)
guages. Hunter (2004), for example, presents results in matlab [although he works with a
\end{itemize}
\textcolor{blue}{fixed}

\begin{itemize}
\item As a reminder,
- Please make sure that all software packages are \cite{}'d properly.

- All references should be in title style.

- See FAQ for specific reference instructions.
Code:
\end{itemize}

\begin{itemize}
\item As a reminder, please make sure that the files needed to replicate
all code/examples within the manuscript are included in a standalone
replication script.
\end{itemize}
\textcolor{blue}{fixed}

\end{document}

### 2023-08-23 inst/jss_4832_minorcomments_rebuttal.tex
minor changes

@@
 
-\textcolor{blue}{Short story: I have accommodated all the comments with
-rewording and clarification.}
+\textcolor{blue}{Short story: I have accommodated all the comments and
+  in particular have refactored the replication script as requested.}
 
@@
 \begin{itemize}
-  \item Please use "R>" as the command prompt, rather than ">".
+  \item Please use \verb+"R>"+ as the command prompt, rather than \verb+">"+.
 \end{itemize}
@@
 
- As a reminder, please make sure that: \proglang, \pkg and \code have
- been used for highlighting throughout the paper (including titles and
- references), except where explicitly escaped.
-
+\begin{itemize}
+  \item
+ As a reminder, please make sure that: \verb+\proglang+, \verb+\pkg+
+ and \verb+\code+ have been used for highlighting throughout the paper
+ (including titles and references), except where explicitly escaped.
+\end{itemize}
 \textcolor{blue}{fixed}
@@
 \end{itemize}
+
 \textcolor{blue}{fixed}
@@
 \item As a reminder,
-- Please make sure that all software packages are \cite{}'d properly.
+- Please make sure that all software packages are \verb+\cite{}+'d properly.
 
@@
 - See FAQ for specific reference instructions.
-Code:
 \end{itemize}
+
+\textcolor{blue}{fixed}
 

### 2023-08-23 inst/jss_hyper3.Rnw
minor changes

@@
 opts_chunk$set(cache=TRUE, autodep=TRUE)
+options(prompt = "R> ")
 @
@@
 available in a range of languages.  \cite{hunter2004}, for example,
-presents results in \proglang{matlab} [although he works with a
+presents results in \proglang{MATLAB} [although he works with a
 nonlinear extension to account for ties]; \cite{allison1994} present
@@
 weightedplayervector n,d;
-n["a"] = 1; // n for numerator
-d["a"] = 3; // d for denominator
+n["a"] = 1; 
+d["a"] = 3; 
 d["b"] = 2;
@@
 
-Above, object \code{L} is an object of class \code{hyper3}; it may be
-evaluated at points in probability space [that is, a vector
-\code{[a,b,c]} of nonnegative values with unit sum] using standard
-\proglang{R} idiom wrapping \proglang{C++} back end.
+Above, we understand \code{n} and \code{d} to represent numerator and
+denominator respectively.  Object \code{L} is an object of class
+\code{hyper3}; it may be evaluated at points in probability space
+[that is, a vector \code{[a,b,c]} of nonnegative values with unit sum]
+using standard \proglang{R} idiom wrapping \proglang{C++} back end.
 

### 2023-08-23 inst/rock_paper_scissors_monster.Rmd
consistentification

@@
      H3[jj] %<>% inc(hw[i,i])
+     H3[setNames(1,tt[i])] %<>% inc(hw[i,i])
 

### 2023-10-29 inst/ternaryplot_hyper2.Rmd
change to include hex sticker, also cite Ternary package

@@
 ---
-title: "ternary_hyper2"
+title: "Ternary plots for hyper2 samples"
 author: "Robin Hankin"
-date: "2022-11-21"
 output: html_document
+bibliography: chess.bib
 ---
@@
 
-## Ternary plots for hyper2 samples
+```{r, label=hexstickerplotter, out.width='15%', out.extra='style="float:right; padding:10px"',echo=FALSE}
+knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+```
 
-Here I show how to use the excellent `Ternary` package to generate
-triangular plots for `hyper2` likelihood functions with size 3.  First
-load the `Ternary` and `hyper2` packages:
+To cite the `hyper2` package in publications please use @hankin2017.
+Here I show how to use the excellent `Ternary` package [@smith2017] to
+generate triangular plots for `hyper2` likelihood functions with size
+3.  First load the `Ternary` and `hyper2` packages:
 
@@
 red point is the maximum likelihood estimate for the three players'
-strengths.
-
-We can adapt this technique to higher-dimensional `hyper2` objects.
+strengths.  We can adapt this technique to higher-dimensional `hyper2`
+objects.
 

### 2023-10-29 inst/jss4832.tex
various changes for JSS editorial policy

--- 2023-08-23 inst/jss_hyper3.Rnw
@@
 \begin{document}
-
-<<setup, cache=FALSE,echo=FALSE>>=
-source("loadlibrary.R",echo=FALSE)  # to avoid stupid messages 
-opts_chunk$set(cache=TRUE, autodep=TRUE)
-options(prompt = "R> ")
-@
 
@@
 
-<<>>=
-LL <- hyper3()
-LL[c(a = 1)] <- 1
-LL[c(a = 3, b = 2, c = 1)] <- -1
-LL
-@
+\begin{Schunk}
+\begin{Sinput}
+R> LL <- hyper3()
+R> LL[c(a = 1)] <- 1
+R> LL[c(a = 3, b = 2, c = 1)] <- -1
+R> LL
+\end{Sinput}
+\begin{Soutput}
+log( (a=1)^1 * (a=3, b=2, c=1)^-1)
+\end{Soutput}
+\end{Schunk}
 
@@
 
-<<>>=
-loglik(c(a = 0.01, b = 0.01, c = 0.98), LL)
-loglik(c(a = 0.90, b = 0.05, c = 0.05), LL)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> loglik(c(a = 0.01, b = 0.01, c = 0.98), LL)
+\end{Sinput}
+\begin{Soutput}
+[1] -4.634729
+\end{Soutput}
+\begin{Sinput}
+R> loglik(c(a = 0.90, b = 0.05, c = 0.05), LL)
+\end{Sinput}
+\begin{Soutput}
+[1] -1.15268
+\end{Soutput}
+\end{Schunk}
 
@@
 
-<<>>=
-(H <- ordervec2supp3(c("a", "c", "b", "a", "a", "b")))
-@
+\begin{Schunk}
+\begin{Sinput}
+R> (H <- ordervec2supp3(c("a", "c", "b", "a", "a", "b")))
+\end{Sinput}
+\begin{Soutput}
+log( (a=1)^3 * (a=1, b=1)^-1 * (a=2, b=1)^-1 * (a=2, b=2)^-1 * (a=2,
+b=2, c=1)^-1 * (a=3, b=2, c=1)^-1 * (b=1)^1 * (c=1)^1)
+\end{Soutput}
+\end{Schunk}
 
@@
 
-<<maxexamp,cache=TRUE>>=
-(mH <- maxp(H))
-@
+\begin{Schunk}
+\begin{Sinput}
+R> (mH <- maxp(H))
+\end{Sinput}
+\begin{Soutput}
+         a          b          c 
+0.21324090 0.08724824 0.69951086 
+\end{Soutput}
+\end{Schunk}
 
@@
 
-<<testequality,cache=TRUE>>=
-equalp.test(H)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> equalp.test(H)
+\end{Sinput}
+\begin{Soutput}
+	Constrained support maximization
+
+data:  H
+null hypothesis: a = b = c
+null estimate:
+        a         b         c 
+0.3333333 0.3333333 0.3333333 
+(argmax, constrained optimization)
+Support for null:  -6.579251 + K
+
+alternative hypothesis:  sum p_i=1 
+alternative estimate:
+         a          b          c 
+0.21324090 0.08724824 0.69951086 
+(argmax, free optimization)
+Support for alternative:  -5.73209 + K
+
+degrees of freedom: 2
+support difference = 0.8471613
+p-value: 0.42863 
+\end{Soutput}
+\end{Schunk}
 
@@
 
-<<maxpaba,cache=TRUE>>=
-maxp(ordervec2supp3(c("a", "b", "a")))
-@
+\begin{Schunk}
+\begin{Sinput}
+R> maxp(ordervec2supp3(c("a", "b", "a")))
+\end{Sinput}
+\begin{Soutput}
+        a         b 
+0.4142108 0.5857892 
+\end{Soutput}
+\end{Schunk}
 
@@
 
-<<>>=
-a <- 1/2 # null
-(S_delta <- log(a * (1 - a)/(1 + a)) - log(3 - 2 * sqrt(2)))
-@
+\begin{Schunk}
+\begin{Sinput}
+R> a <- 1/2 # null
+R> (S_delta <- log(a * (1 - a)/(1 + a)) - log(3 - 2 * sqrt(2)))
+\end{Sinput}
+\begin{Soutput}
+[1] -0.0290123
+\end{Soutput}
+\end{Schunk}
 
@@
 
-<<>>=
-pchisq(-2 * S_delta, df = 1, lower.tail = FALSE)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> pchisq(-2 * S_delta, df = 1, lower.tail = FALSE)
+\end{Sinput}
+\begin{Soutput}
+[1] 0.8096458
+\end{Soutput}
+\end{Schunk}
 
@@
   \begin{center}
-<<figaba, fig=TRUE>>=
-a <- seq(from = 0, by = 0.005, to = 1)
-S <- function(a){log(a * (1 - a) / ((1 + a) * (3 - 2 * sqrt(2))))}
-plot(a, S(a), type = 'b',xlab=expression(p[a]),ylab="support")
-abline(h = c(0, -2))
-abline(v = c(0.02438102, 0.9524271), col = 'red')
-abline(v = sqrt(2) - 1)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> a <- seq(from = 0, by = 0.005, to = 1)
+R> S <- function(a){log(a * (1 - a) / ((1 + a) * (3 - 2 * sqrt(2))))}
+R> plot(a, S(a), type = 'b',xlab=expression(p[a]),ylab="support")
+R> abline(h = c(0, -2))
+R> abline(v = c(0.02438102, 0.9524271), col = 'red')
+R> abline(v = sqrt(2) - 1)
+\end{Sinput}
+\end{Schunk}
+\includegraphics{jss_hyper3-figaba}
 \caption{A\label{ABA_likelihood} support function for $p_a$ with observation $a\succ b\succ a$.}
@@
   \begin{center}
-<<figabbabb,echo=FALSE,fig=TRUE>>=
-f_aab <- function(a){a^2 / (1 + a)}
-f_aba <- function(a){a * (1 - a) / (1 + a)}
-f_baa <- function(a){(1 - a) / (1 + a)}
-p <- function(f, ...){
-  a <- seq(from = 0, by = 0.005, to = 1)
-  points(a, f(a) / max(f(a)), ...)
-  }
-plot(0:1, 0:1, xlab = expression(p[a]), ylab = "Likelihood", type = "n")
-p(f_aab, type = "l", col = "black")
-p(f_aba, type = "l", col = "red")
-p(f_baa, type = "l", col = "blue")
-text(0.8,0.8,"AAB")
-text(0.8,0.4,"ABA",col="red")
-text(0.8,0.05,"BAA",col="blue")
-abline(h = exp(-2), lty = 2)
-@
+\includegraphics{jss_hyper3-figabbabb}
 \caption{Likelihood \label{ABAetc_likelihood} functions for
@@
 
-<<cache=TRUE>>=
-maxp(ordervec2supp3(c("a", "b"), nonfinishers=c("a", "b")))
-@
+\begin{Schunk}
+\begin{Sinput}
+R> maxp(ordervec2supp3(c("a", "b"), nonfinishers=c("a", "b")))
+\end{Sinput}
+\begin{Soutput}
+        a         b 
+0.5857892 0.4142108 
+\end{Soutput}
+\end{Schunk}
 
@@
 
-<<define_xy_wilcox>>=
-x <- c(0.80, 0.83, 1.89, 1.04, 1.45, 1.38, 1.91, 1.64, 0.73, 1.46)
-y <- c(1.15, 0.88, 0.90, 0.74, 1.21)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> x <- c(0.80, 0.83, 1.89, 1.04, 1.45, 1.38, 1.91, 1.64, 0.73, 1.46)
+R> y <- c(1.15, 0.88, 0.90, 0.74, 1.21)
+\end{Sinput}
+\end{Schunk}
 
@@
 
-<<hyper3osdef,cache=TRUE>>=
-names(x) <- rep("x", length(x))
-names(y) <- rep("y", length(y))
-(os <- names(sort(c(x, y))))
-@
+\begin{Schunk}
+\begin{Sinput}
+R> names(x) <- rep("x", length(x))
+R> names(y) <- rep("y", length(y))
+R> (os <- names(sort(c(x, y))))
+\end{Sinput}
+\begin{Soutput}
+ [1] "x" "y" "x" "x" "y" "y" "x" "y" "y" "x" "x" "x" "x" "x" "x"
+\end{Soutput}
+\end{Schunk}
 
@@
 
-<<hyper3xytest,cache=TRUE>>=
-Hxy <- ordervec2supp3(os)
-equalp.test(Hxy)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> Hxy <- ordervec2supp3(os)
+R> equalp.test(Hxy)
+\end{Sinput}
+\begin{Soutput}
+	Constrained support maximization
+
+data:  Hxy
+null hypothesis: x = y
+null estimate:
+  x   y 
+0.5 0.5 
+(argmax, constrained optimization)
+Support for null:  -27.89927 + K
+
+alternative hypothesis:  sum p_i=1 
+alternative estimate:
+        x         y 
+0.2401539 0.7598461 
+(argmax, free optimization)
+Support for alternative:  -26.48443 + K
+
+degrees of freedom: 1
+support difference = 1.414837
+p-value: 0.09253716 
+\end{Soutput}
+\end{Schunk}
 
@@
   \begin{center}
-<<plotwilcoxlike,echo=FALSE,fig=TRUE,cache=TRUE>>=
-a <- seq(from = 0.02, to = 0.8, len = 40)
-L <- sapply(a, function(p){loglik(p, Hxy)})
-plot(a, L - max(L), type = 'b',xlab=expression(p[a]),ylab="likelihood")
-abline(h = c(0, -2))
-abline(v = c(0.24))
-abline(v=c(0.5), lty=2)
-@
+\includegraphics{jss_hyper3-plotwilcoxlike}
 \caption{A\label{wilcox_likelihood} support function for the Bradley-Terry
@@
 
-<<javelintable>>=
-javelin_table
-@
+\begin{Schunk}
+\begin{Sinput}
+R> javelin_table
+\end{Sinput}
+\begin{Soutput}
+          throw1 throw2 throw3 throw4 throw5 throw6
+Chopra     87.03  87.58  76.79      X      X  84.24
+Vadlejch   83.98      X      X  82.86  86.67      X
+Vesely     79.73  80.30  85.44      X  84.98      X
+Weber      85.30  77.90  78.00  83.10  85.15  75.72
+Nadeem     82.40      X  84.62  82.91  81.98      X
+Katkavets  82.49  81.03  83.71  79.24      X      X
+Mardare    81.16  81.73  82.84  81.90  83.30  81.09
+Etelatalo  78.43  76.59  83.28  79.20  79.99  83.05
+\end{Soutput}
+\end{Schunk}
 
@@
 
-<<label=converttosupp3, cache=TRUE>>=
-javelin_vector <- attemptstable2supp3(javelin_table,
-       decreasing = TRUE, give.supp = FALSE)
-options(width = 60)
-javelin_vector
-@
+\begin{Schunk}
+\begin{Sinput}
+R> javelin_vector <- attemptstable2supp3(javelin_table,
++        decreasing = TRUE, give.supp = FALSE)
+R> options(width = 60)
+R> javelin_vector
+\end{Sinput}
+\begin{Soutput}
+   Chopra    Chopra  Vadlejch    Vesely     Weber     Weber 
+    87.58     87.03     86.67     85.44     85.30     85.15 
+   Vesely    Nadeem    Chopra  Vadlejch Katkavets   Mardare 
+    84.98     84.62     84.24     83.98     83.71     83.30 
+Etelatalo     Weber Etelatalo    Nadeem  Vadlejch   Mardare 
+    83.28     83.10     83.05     82.91     82.86     82.84 
+Katkavets    Nadeem    Nadeem   Mardare   Mardare   Mardare 
+    82.49     82.40     81.98     81.90     81.73     81.16 
+  Mardare Katkavets    Vesely Etelatalo    Vesely Katkavets 
+    81.09     81.03     80.30     79.99     79.73     79.24 
+Etelatalo Etelatalo     Weber     Weber    Chopra Etelatalo 
+    79.20     78.43     78.00     77.90     76.79     76.59 
+    Weber  Vadlejch    Nadeem  Vadlejch    Chopra    Vesely 
+    75.72        NA        NA        NA        NA        NA 
+   Chopra Katkavets  Vadlejch    Vesely    Nadeem Katkavets 
+       NA        NA        NA        NA        NA        NA 
+\end{Soutput}
+\end{Schunk}
 
@@
 
-<<dothething2,cache=TRUE>>=
-javelin <- ordervec2supp3(v = names(javelin_vector)[!is.na(javelin_vector)])
-@
+\begin{Schunk}
+\begin{Sinput}
+R> javelin <- ordervec2supp3(v = names(javelin_vector)[!is.na(javelin_vector)])
+\end{Sinput}
+\end{Schunk}
 
@@
 
-<<setdigits,echo=FALSE>>=
-options(digits = 3)
-@
-
 \begin{figure}[htbp]
   \begin{center}
-<<testthejav, cache=TRUE, fig=TRUE>>=
-(mj <- maxp(javelin))
-dotchart(mj, pch = 16,xlab="Estimated Bradley-Terry strength")
-@
+\begin{Schunk}
+\begin{Sinput}
+R> (mj <- maxp(javelin))
+\end{Sinput}
+\begin{Soutput}
+   Chopra Etelatalo Katkavets   Mardare    Nadeem  Vadlejch 
+   0.0930    0.0482    0.0929    0.1173    0.1730    0.3206 
+   Vesely     Weber 
+   0.1140    0.0409 
+\end{Soutput}
+\begin{Sinput}
+R> dotchart(mj, pch = 16,xlab="Estimated Bradley-Terry strength")
+\end{Sinput}
+\end{Schunk}
+\includegraphics{jss_hyper3-testthejav}
 \caption{Maximum \label{javelinbradleyterry} likelihood estimate for javelin throwers' Bradley-Terry strengths.}
@@
 
-<<havealook,echo=FALSE,cache=TRUE>>=
-f <- function(s){
-  jj <- specificp.test(javelin, "Vesely", s, n = 2)
-  p <- jj$null_estimate
- return(c(log(p[6] / p[7]), jj$null_support))
-}
-Ves <- seq(from = 0.0199, to = 0.33, len = 16)
-M <- sapply(Ves, f)
-M[2,] <- M[2,] - max(M[2,])
-rownames(M) <- c("logcontrast", "support")
-@
-
 \begin{figure}[htbp]
 \begin{center}
-<<plottheloglikcont,echo=FALSE,fig=TRUE,cache=TRUE>>=
-colnames(M) <-  as.character(Ves)
-plot(t(M), type = "b")
-abline(h = c(0, -2))
-abline(v = 0, lty = 2)
-abline(v = log(0.32062833 / 0.11402735)) # these from mp
-@
+\includegraphics{jss_hyper3-plottheloglikcont}
 \caption{Profile \label{profliklogcont} likelihood for log-contrast
@@
 
-<<showconstructortable>>=
-constructor_2021_table[, 1:9]
-@
+\begin{Schunk}
+\begin{Sinput}
+R> constructor_2021_table[, 1:9]
+\end{Sinput}
+\begin{Soutput}
+   Constructor BHR EMI POR ESP  MON AZE FRA STY
+1         Merc   1   2   1   1    7  12   2   2
+2         Merc   3 Ret   3   3  Ret  15   4   3
+3         RBRH   2   1   2   2    1   1   1   1
+4         RBRH   5  11   4   5    4  18   3   4
+5      Ferrari   6   4   6   4    2   4  11   6
+6      Ferrari   8   5  11   7 DNSP   8  16   7
+7           MM   4   3   5   6    3   5   5   5
+8           MM   7   6   9   8   12   9   6  13
+9           AR  13   9   7   9    9   6   8   9
+10          AR Ret  10   8  17   13 Ret  14  14
+11         ATH   9   7  10  10    6   3   7  10
+12         ATH  17  12  15 Ret   16   7  13 Ret
+13         AMM  10   8  13  11    5   2   9   8
+14         AMM  15  15  14  13    8 Ret  10  12
+15          WM  14 Ret  16  14   14  16  12  17
+16          WM  18 Ret  18  16   15  17  18 Ret
+17        ARRF  11  13  12  12   10  10  15  11
+18        ARRF  12  14 Ret  15   11  11  17  15
+19          HF  16  16  17  18   17  13  19  16
+20          HF Ret  17  19  19   18  14  20  18
+\end{Soutput}
+\end{Schunk}
 
@@
 
-<<maxpconstructor2021, cache=TRUE>>=
-const2020 <- ordertable2supp3(constructor_2020_table)
-const2021 <- ordertable2supp3(constructor_2021_table)
-options(digits = 4)
-maxp(const2020, n = 1)
-maxp(const2021, n = 1)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> const2020 <- ordertable2supp3(constructor_2020_table)
+R> const2021 <- ordertable2supp3(constructor_2021_table)
+R> options(digits = 4)
+R> maxp(const2020, n = 1)
+\end{Sinput}
+\begin{Soutput}
+   ARRF     ATH Ferrari      HF    Merc      MR       R 
+0.04530 0.06807 0.06063 0.02623 0.37783 0.10026 0.09767 
+   RBRH  RPBWTM      WM 
+0.12072 0.08055 0.02273 
+\end{Soutput}
+\begin{Sinput}
+R> maxp(const2021, n = 1)
+\end{Sinput}
+\begin{Soutput}
+    AMM      AR    ARRF     ATH Ferrari      HF    Merc 
+0.05942 0.07543 0.06238 0.05611 0.16939 0.02023 0.19395 
+     MM    RBRH      WM 
+0.14126 0.18334 0.03848 
+\end{Soutput}
+\end{Schunk}
 
@@
 
-<<definecombinedlikelihoodfunction,cache=TRUE>>=
-H <- (
-      psubs(constructor_2020, "Merc", "Merc2020") +
-      psubs(constructor_2021, "Merc", "Merc2021")
-     )
-@
+\begin{Schunk}
+\begin{Sinput}
+R> H <- (
++       psubs(constructor_2020, "Merc", "Merc2020") +
++       psubs(constructor_2021, "Merc", "Merc2021")
++      )
+\end{Sinput}
+\end{Schunk}
 
@@
 
-<<usesamep,cache=TRUE>>=
-options(digits = 4)
-samep.test(H, c("Merc2020", "Merc2021"))
-@
+\begin{Schunk}
+\begin{Sinput}
+R> options(digits = 4)
+R> samep.test(H, c("Merc2020", "Merc2021"))
+\end{Sinput}
+\begin{Soutput}
+	Constrained support maximization
+
+data:  H
+null hypothesis: Merc2020 = Merc2021
+null estimate:
+     AMM       AR     ARRF      ATH  Ferrari       HF 
+ 0.04239  0.05413  0.04677  0.04374  0.07568  0.02323 
+Merc2020 Merc2021       MM       MR        R     RBRH 
+ 0.13903  0.13903  0.09016  0.07944  0.07475  0.10024 
+  RPBWTM       WM 
+ 0.06235  0.02905 
+(argmax, constrained optimization)
+Support for null:  -1189 + K
+
+alternative hypothesis:  sum p_i=1 
+alternative estimate:
+     AMM       AR     ARRF      ATH  Ferrari       HF 
+ 0.03766  0.04824  0.04333  0.04060  0.07036  0.02132 
+Merc2020 Merc2021       MM       MR        R     RBRH 
+ 0.23135  0.09216  0.07893  0.07973  0.07455  0.09322 
+  RPBWTM       WM 
+ 0.06177  0.02679 
+(argmax, free optimization)
+Support for alternative:  -1184 + K
+
+degrees of freedom: 1
+support difference = 4.722
+p-value: 0.002119 
+\end{Soutput}
+\end{Schunk}
 
@@
 
-\bibliography{jss_hyper3}
+\bibliography{jss4832}
 \end{document}

### 2023-10-29 inst/eurodance.Rmd
label each chunk

@@
 
-```{r,cache=TRUE}
+```{r,label=eachcolisorder,cache=TRUE}
 eurodance <- hyper2()
@@
 
-```{r,cache=TRUE}
+```{r,label=verifyconsistency,cache=TRUE}
 consistency(eurodance)
@@
 
-```{r}
+```{r,label=considerjury}
 (jj <- o[,2])
@@
 
-```{r,cache=TRUE}
+```{r,label=simfish,cache=TRUE}
 n <- 4000
@@
 
-```{r}
+```{r,label=comparelogliks}
 plot(sort(LL),log="y")
@@
 
-```{r}
+```{r,label=sumll}
 sum(LL[LL<p])/sum(LL)

### 2023-10-29 inst/fatigue.Rmd
bugfix
use sticker in Rmd
typo

@@
 tab <- cbind(tab,played=rowSums(tab))
+tab <- cbind(tab,points=tab[,1] + tab[,3]/2)
 tab
@@
 Hf_maxp <- maxp(Hf)
-````
+```
 

### 2023-10-30 inst/curacao1962_threeplayers_rest_monster.Rmd
typo
label all chunks

@@
 
-```{r}
+```{r, label=loadlibrary}
 library("hyper2")
@@
 
-```{r}
+```{r,label=specplay}
   players <- 
@@
 
-```{r}
+```{r cumandcons}
 cumulative_games
@@
 
-```{r}
+```{r calcdiff}
 diff <- a$cncmp_white-a$cncmp_black  # difference in each game
@@
 
-```{r}
+```{r assessnullnofat}
 table(a[,3],a[,5]-a[,6])
@@
 Then Fisher test:
-```{r}
+```{r thenfish}
 b <- subset(a,a[,5] != a[,6])
@@
 
-```{r}
+```{r hyper2reified}
 problem <- function(...){stop("must be a white win or a black win or a draw")}

### 2023-10-30 inst/fatigue.Rmd
label all chunks

@@
 
-```{r cars}
+```{r readdata}
 jj <- read.table("interzonal1973_petropolis.txt",header=TRUE,skip=2)
@@
 
-```{r}
+```{r maketabs}
 tab <- cbind(tab,played=rowSums(tab))
@@
 
-```{r}
+```{r dostudentttest}
 table(a$result)
@@
 
-```{r}
+```{r makeo}
 head(a,10)
@@
 
-```{r}
+```{r justsuccessive}
 unlist(a[2,])
@@
 
-```{r}
+```{r showsum}
 summary(Hf)
@@
 
-```{r}
+```{r maketable}
 jj <- rep(0,nrow(a))

### 2023-10-30 inst/T20.Rmd
add hex sticker to more Rmd files
needs to be a single line for knitr
name all chunks and cache the slow ones

@@
 
+<p style="text-align: right;">
+![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
+</p>
+
 # The Indian Premier League
@@
 
-```{r}
+```{r kickoff}
 library("hyper2",quietly=TRUE)
@@
 
-```{r}
+```{r onewayanal}
 x <- table(T20_table$toss_winner==T20_table$match_winner)
@@
 
-```{r}
+```{r showtable}
 table(T20_table$toss_decision)
@@
 
-```{r}
+```{r pbtw}
 attach(T20_table)
@@
 
-```{r}
+```{r conscheck}
 sum(field_first_wins) + sum(bat_first_wins)
@@
 
-```{r}
+```{r twowaycontables}
 M <- table(
@@
 
-We may reject homogoneity of proportion (Fisher's exact test, $p= `r
-round(fisher.test(M)$p.value,3)`$).  Recalling that a slight majority
+We may reject homogeneity of proportion (Fisher's exact test, $p= `r round(fisher.test(M)$p.value,3)`$).  Recalling that a slight majority
 of toss winners elect to field first (361 out of 633), we see that
@@
 
-```{r}
+```{r indstrength}
 attach(lapply(T20_table[, ], as.vector))  # vector access to team1, etc
@@
 
-```{r}
+```{r maxisstraight,cache=TRUE}
 T20_maxp <- maxp(T20)
@@
 
-```{r}
+```{r nullallteams}
 a1 <- maxp(T20,give=TRUE)$value        # likelihood at the evaluate
@@
 
-```{r}
+```{r wilkstest}
 pchisq(2*14.5,df=13,lower.tail=FALSE)
@@
 
-```{r}
+```{r tossdiff}
 T20_toss  <- hyper2(pnames=c('toss',levels(T20_table$team1)))
@@
 
-```{r}
+```{r maxlike,cache=TRUE}
 T20_maxp_toss <- maxp(T20_toss)

### 2023-10-30 inst/formula1.Rmd
name all chunks and cache the slow ones

@@
 
-```{r,cache=TRUE}
+```{r twonatural,cache=TRUE}
 mf <- maxp(formula1)
@@
 
-```{r}
+```{r ranklike}
 mf
@@
 
-```{r,testwetstrengthzero}
+```{r,testwetstrengthzero,cache=TRUE}
 pnames(H)

### 2023-10-30 inst/grad_weight.Rmd
name all chunks and cache the slow ones

@@
 
-```{r}
+```{r slightlymorecomplicated}
 H <- hyper3()
@@
 
-```{r}
+```{r chessexample}
 
@@
 
-```{r}
+```{r diffwrtw}
 diff <- 0
@@
 
-```{r}
+```{r cleverusecentral}
 RHS2 <- (loglik(probs,f2(w=w+dw/2,D=D)) - loglik(probs,f2(w=w-dw/2,D=D)))/dw  # differential via numerics
@@
 
-```{r}
+```{r diffwrtD}
 o <- c("p1","p2")

### 2023-10-30 inst/gradient.Rmd
name all chunks and cache the slow ones

@@
 
-```{r}
+```{r showchess}
 chess
@@
 
-```{r}
+```{r chessgrad}
 chess_gradient_theoretical <- function(x){
@@
 
-```{r}
+```{r pchessagree}
 p_chess <- 
@@
 
-```{r}
+```{r showicons}
 icons
@@
 
-```{r}
+```{r samethingalgebra}
 icons_gradient_theoretical <- function(x){
@@
 
-```{r}
+```{r isicons}
 p_icons <- c(NB=0.3, L=0.1, PB=0.2, THC=0.15, OA=0.15, WAIS=0.1)
@@
 
-```{r}
+```{r comparegradientwithwithout}
 icons_gradient_theoretical(p_icons)
@@
 
-```{r}
+```{r lookatchess}
 rbind(indep(p_chess), gradient(chess,indep(p_chess)))
@@
 
-```{r}
+```{r isderivzero}
 sum(powers(chess))
@@
 
-```{r}
+```{r seesumpower}
 sum(powers(icons))
@@
 
-```{r}
+```{r ishessian}
 (M <- hessian(icons))
@@
 
-```{r}
+```{r isnothessian}
 is_ok_hessian(M,give=TRUE)
@@
 
-```{r}
+```{r evaluatehessian}
 (p <- indep(equalp(icons))) # NB not the evaluate
@@
 
-```{r}
+```{r smallpert}
 dp <- rnorm(5)*1e-4        # small perturbation
@@
 
-```{r}
+```{r useborderedhess}
 

### 2023-10-30 inst/karate.Rmd
name all chunks and cache the slow ones

@@
 
-```{r}
+```{r  calcmaxp,cache=TRUE}
 karate_maxp <- maxp(karate)
@@
 
-```{r}
+```{r showdot}
 dotchart(log10(karate_maxp))
@@
 
-```{r}
+```{r defmatwinloss}
 pk <- pnames(karate)
@@
 
-```{r,cache=TRUE}
+```{r,label=pairwisem,cache=TRUE}
 pM <- pairwise(M)
@@
 
-```{r}
+```{r usezerm}
 karate_zermelo <- zermelo(M)
@@
 
-```{r}
+```{r plotzerm}
 par(pty='s')

### 2023-10-30 inst/kka.Rmd
name all chunks and cache the slow ones

@@
 
-```{r}
+```{r contourplotter}
 l <- l-max(l)

### 2023-10-30 inst/skeleton.Rmd
name all chunks and cache the slow ones

@@
 
-```{r}
+```{r identifycompetitors}
 bm1 <- st[,1] > st["Maier",1]  # lost to Maier in run 1
@@
 
-```{r}
+```{r losttomaier}
 rownames(st[bm1 | bm2 | bm3 | bm4,])
@@
 
-```{r}
+```{r definebeats}
 beats <- function(p){   # returns everyone whom competitor 'p' beats
@@
 
-```{r}
+```{r excludemidrange}
 (st2 <- st[-(4:19),])  # exclude a bunch of mid-range competitors
@@
 
-```{r}
+```{r justtop3}
 bb <- c()
@@
 
-```{r}
+```{r showpvals}
 plot(pvals,log='y',pch=16)
@@
 
-```{r,label=savecurlingdataset}
+```{r,label=saveskeletondataset}
 save(skeleton_table,skeleton,skeleton_maxp,file="skeleton.rda")

### 2023-10-30 inst/T20.Rmd
label all chunks, cache the slow ones

@@
 
-```{r indstrength}
+```{r indstrength,cache=TRUE}
 attach(lapply(T20_table[, ], as.vector))  # vector access to team1, etc
@@
 
-```{r nullallteams}
+```{r nullallteams,cache=TRUE}
 a1 <- maxp(T20,give=TRUE)$value        # likelihood at the evaluate
@@
 
-```{r tossdiff}
+```{r tossdiff,cache=TRUE}
 T20_toss  <- hyper2(pnames=c('toss',levels(T20_table$team1)))

### 2023-10-30 inst/formula1_points.Rmd
minor typos
label all chunks, cache the slow ones

@@
 
-```{r}
+```{r processtogetranks}
 f <- function(x){
@@
 
-```{r}
+```{r calcthepoints}
 points <- c(25,18,15,12,10,8,6,4,2,1,0,0,0,0,0,0,0,0,0,0)
@@
 
-```{r}
+```{r userandomsamp}
 random_table <- rrank(n=ncol(f2017), p=mL2017)
@@
 
-```{r}
+```{r howclosetoreality,cache=TRUE}
 maxp(rank_likelihood(random_table)) # should be close to mL2017
@@
 
-```{r}
+```{r checkagain}
 a <- random_table
@@
 
-```{r}
+```{r thuswesee}
 jj <- sort(p,decreasing=TRUE)
@@
 predicted by the points system.
+
+# Now account for pole position
+
+Now do the same thing again but include pole position.

### 2023-10-30 inst/handover.Rmd
use sticker in Rmd
label all chunks, cache the slow ones

@@
 
-```{r}
+```{r setuphandover,cache=TRUE}
 handover <- hyper2()
@@
 
-```{r}
+```{r testtheta01,cache=TRUE}
 samep.test(handover,c("t01","t10"))
@@
 
-```{r}
+```{r probofhandover}
 probability(handover,disallowed=function(p){p[2]<p[3]},tol=0.05)

### 2023-10-30 inst/loser.Rmd
slight rewriting
label all chunks, cache the slow ones

@@
 
+<p style="text-align: right;">
+![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
+</p>
+
 # Partial rank observations: who lost?
 
+(see also `inst/notthewinner.Rmd`).
 Here we consider a situation in which we make only a rather
@@
 
-```{r,label=useggrl}
+```{r,label=useggrl,cache=TRUE}
 n <- 6 # number of competitors

### 2023-10-30 inst/order.Rmd
label all chunks, cache the slow ones

@@
 
-```{r}
+```{r,label=maxwithevalp,cache=TRUE}
 evaluate1 <- maxp(H)
@@
 
-```{r}
+```{r bettereval,cache=TRUE}
 f <- function(p,n){c(p,rep((1-p)/n,n-1))}
@@
 
-```{r}
+```{r showmaxlike}
 dotchart(evaluate2,pch=16)
@@
 
-```{r}
+```{r diffbetweenlikelihoods}
 maxlike2-maxlike1
@@
 
-```{r}
+```{r mimiclimit}
 vsssp <- function(x,n){
@@
 
-```{r}
+```{r evaluatetwice,cache=TRUE}
 evaluate3 <- maxp(H,vsssp(2,10))
@@
 
-```{r}
+```{r veryslightlybetter}
 maxlike3-maxlike2

### 2023-10-30 inst/pentathlon.Rmd
label all chunks, cache the slow ones

@@
 
-```{r}
+```{r showpiechart,cache=TRUE}
 pentathlon <- ordertable2supp(pentathlon_table)
@@
 
-```{r}
+```{r testequal,cache=TRUE}
 equalp.test(pentathlon)

### 2023-10-30 inst/zermelo.Rmd
label all chunks, cache the slow ones

@@
 
-```{r}
+```{r maketencompetitors}
 n <- 10
@@
 
-```{r}
+```{r makeMwithpairwise}
 M <- matrix(0,n,n)
@@
 
-```{r}
+```{r usezerm}
 (pZ <- zermelo(M))
@@
 
-```{r,cache=TRUE}
+```{r,label=usehyper2,cache=TRUE}
 (H <- pairwise(M))
@@
 
-```{r}
+```{r comparewithscattergraph}
 par(pty='s')
@@
 
-```{r}
+```{r nowplotlog}
 plot(log(pZ),log(abs(pH-pZ)/pZ))
@@
 
-```{r}
+```{r calcloglik}
 loglik(pZ,H)-loglik(pH,H)
@@
 
-```{r}
+```{r arebothclose}
 par(pty='s')
@@
 
-```{r}
+```{r funwithzermelo}
 o <- zermelo(M,give=TRUE)

### 2023-10-30 inst/f1points.Rmd
consistentification

@@
 $p_4+p_5$, is uninformative about $p_4\left|p_4+p_5\right.$).  We now
-use a technique due to \cite{hankin2010,hankin2017} and introduce
+use a technique due to \cite{hankin2010,hankin2017a} and introduce
 fictional (reified) entities whose nonzero Bradley-Terry strength
@@
 position in Formula 1.  Here I analyse seasons 2016-2019 using the
-`hyper2` package \citep{hankin2017} which implements the Plackett-Luce
+`hyper2` package \citep{hankin2017a} which implements the Plackett-Luce
 likelihood function with additional reified entities.

### 2023-10-30 inst/mann_whitney_wilcoxon.Rmd
consistentification

@@
 ---
-title: "Mann-Whitney-Wilcoxon test with `hyper3` objects"
-author: "Robin KS Hankin"
-date: "10/04/2022"
-output: html_document
+title: "Generalizing the Mann-Whitney-Wilcoxon test with `hyper3` objects"
+author: "Robin K. S. Hankin"
+output: bookdown::html_document2
+bibliography: jss_hyper3.bib
 ---
@@
 
-<p style="text-align: right;">
-![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
-</p>
+```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
+knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+```
 
-We do some work with Mann-Whitney test, using the chorioamnion dataset at `wilcox.test.Rd`:
+To cite the `hyper2` package in publications, please use @hankin2017a.
+Here I discuss the `hyper2` generalization of the
+Mann-Whitney-Wilcoxon test, applying it to the chorioamnion dataset at
+`wilcox.test.Rd`:
 
@@
 ```
+
+# References

### 2023-10-30 inst/rowing.Rmd
consistentification

@@
 title: "rowing: analysis of sculls in the 2016 Olympics"
+author: "R. K. S. Hankin"
 output: bookdown::html_document2
+bibliography: jss_hyper3.bib
 ---
@@
 
-<p style="text-align: right;">
-![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
-</p>
+```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
+knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+```
 
+To cite the `hyper2` package in publications, please use @hankin2017a.
 This short document analyses the `rowing` dataset: sculls in the 2016
@@
 ```
+
+## References {-}

### 2023-10-30 inst/salad.Rmd
consistentification

@@
 output: bookdown::html_document2
+author: "R. K. S. Hankin"
+bibliography: jss_hyper3.bib
 ---
@@
 
-<p style="text-align: right;">
-![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
-</p>
+```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
+knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+```
 
+To cite the `hyper2` package in publications, please use @hankin2017a.
 This short document shows how to apply `hyper2` idiom to the `salad`
@@
 ```
-The dataset contains the rankings of four salad dressings concerning
-tartness by 32 judges, with values ranging from 1 (most tart) to 4
-(least tart).
+The dataset contains the rankings of four salad dressings
+concerning tartness by 32 judges, with values ranging
+from 1 (most tart) to 4 (least tart).
 ```

### 2023-10-30 inst/skating.Rmd
consistentification

@@
 title: "Ladies' figure skating in the 2002 Winter Olympics"
+author: "R. K. S. Hankin"
 output: bookdown::html_document2
+bibliography: jss_hyper3.bib
 ---
@@
 ```
-<p style="text-align: right;">
-![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
-</p>
 
+```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
+knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+```
+
+To cite the `hyper2` package in publications, please use @hankin2017a.
 This document analyses Bradley-Terry strengths of the competitors in
@@
 ```
+
+### References {-}

### 2023-10-30 inst/skeleton.Rmd
consistentification

@@
 ---
-title: "skeleton"
+title: "skeleton racing with the hyper2 package"
+author: "R. K. S. Hankin"
 output: bookdown::html_document2
+bibliography: jss_hyper3.bib
 ---
@@
 
-<p style="text-align: right;">
-![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
-</p>
+```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
+knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+```
 
+To cite the `hyper2` package in publications, please use @hankin2017a.
 The [skeleton](https://en.wikipedia.org/wiki/Skeleton_(sport)) is a

### 2023-10-30 inst/soling.Rmd
add hex sticker to more Rmd files
consistentification

@@
 title: "soling"
+author: "R. K. S. Hankin"
 output: bookdown::html_document2
+bibliography: jss_hyper3.bib
 ---
@@
 
-The Soling is a type of boat raced in the 2000 Summer Olympics.  Here
-I analyse race data taken from
+```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
+knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+```
+
+To cite the `hyper2` package in publications, please use @hankin2017a.
+Soling is a type of boat raced in the 2000 Summer Olympics.  Here I
+analyse race data taken from
 [wikipedia](https://en.wikipedia.org/w/index.php?title=Sailing_at_the_2000_Summer_Olympics_\%E2\%80\%93_Soling&oldid=945362535)
@@
 
-```{r}
+```{r solingdiff}
 (soling_diff <- soling_after - soling)

### 2023-10-30 inst/table_tennis_serve.Rmd
consistentification

@@
 title: "The importance of serve in table tennis"
-author: "Robin K. S. Hankin"
-date: "`r Sys.Date()`"
-vignette: >
-  %\VignetteIndexEntry{table_tennis_serve}
-  %\VignetteEngine{knitr::rmarkdown}
-  %\VignetteEncoding{UTF-8}
+author: "R. K. S. Hankin"
+output: bookdown::html_document2
+bibliography: jss_hyper3.bib
 ---
 
-# The importance of serve in table tennis.
+```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
+knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+```
 
-(In this vignette, object `H` is created.  It is identical to
-object `table_tennis` in the package; to see it type
-`data(table_tennis)` at the command line, and to see the help
-page, type `help(table_tennis)`).
-
+To cite the `hyper2` package in publications, please use @hankin2017a.
+In this vignette, object `H` is created.  It is identical to object
+`table_tennis` in the package; to see it type `data(table_tennis)` at
+the command line, and to see the help page, type `help(table_tennis)`.
 Suppose we have three players, "A","B", and "C" who play table tennis.

### 2023-10-30 inst/tennis.Rmd
consistentification

@@
 title: "Tennis doubles and ghosts"
+author: "R. K. S. Hankin"
 output: bookdown::html_document2
+bibliography: jss_hyper3.bib
 ---
@@
 
-<p style="text-align: right;">
-![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
-</p>
+```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
+knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+```
 
-This short document defines and analyses a small dataset originally
-considered by Hankin (2010).  There, results from tennis games among
-four players was considered:
+To cite the `hyper2` package in publications, please use @hankin2017a.
+This short document discusses a small dataset first presented by
+@hankin2010.  There, results from tennis games among four players was
+considered:
 
@@
 The likelihood for the 9-2 scoreline between $\left\lbrace
-p1,p2\right\rbrace$ and $\left\lbrace p3,p4\right\rbrace$ would have a
-likelihood function of
+p_1,p_2\right\rbrace$ and $\left\lbrace p_3,p_4\right\rbrace$ would
+have a likelihood function of
 
@@
 
-### References {-}
-
-* RKS Hankin 2010. "A generalization of the Dirichlet distribution". 
- _Journal of Statistical Software_ 33(11)
-* RKS Hankin 2017. "Partial rank data with the hyper2 package:
-  likelihood functions for generalized Bradley-Terry models".  _The R
-  Journal_ 9(2)
-
 ### Package dataset {-}

### 2023-10-30 inst/universities.Rmd
name all chunks and cache the slow ones
consistentification

--- 2021-02-01 inst/universities.Rmd
@@
 ---
-title: "universities"
+title: "university ranking data"
 output: bookdown::html_document2
+author: "R. K. S. Hankin"
+bibliography: jss_hyper3.bib
 ---
@@
 
-## University ranking data
+```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
+knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+```
 
-First load the data:
+To cite the `hyper2` package in publications, please use @hankin2017a.
+Here I analyse a dataset on NZ university rankings, taken from the
+Times Higher Educational Supplement.  First load the data:
 
-```{r cars}
-universities_table <- read.table("universities.txt",header=TRUE)
+```{r loadthetable}
+uni_table <- read.table("universities.txt",header=TRUE)
+universities_table <- uni_table
 head(universities_table,12)
@@
 
-```{r,label=preandpost}
+```{r,label=preandpost,cache=TRUE}
 
@@
 
-```{r}
+```{r maxph1h2,cache=TRUE}
 H <- H1 + H2
@@
 
-```{r}
+```{r sameptesthon2,cache=TRUE}
 samep.test(H,c("AUT_pre","AUT_post"))

### 2023-10-30 inst/volleyball.Rmd
name all chunks and cache the slow ones
consistentification

@@
 title: "volleyball dataset"
+author: "R. K. S. Hankin"
 output: bookdown::html_document2
+bibliography: jss_hyper3.bib
 ---
@@
 ```
-<p style="text-align: right;">
-![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
-</p>
 
-This short document discusses a dataset first presented by Hankin
-(2010), although here only the first 52 observations are used.  A
-volleyball \dfn{set} is a Bernoulli trial between two disjoint subsets
+```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
+knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+```
+
+To cite the `hyper2` package in publications, please use @hankin2017a.
+This short document discusses a dataset first presented by @hankin2010,
+although here only the first 52 observations are used.  A
+volleyball _set_ is a Bernoulli trial between two disjoint subsets
 of the players.  The two subsets are denoted (after the game) as the
@@
 comprising `p5` and `p6`.  Player `p9` did not play; team `p1 p4 p8`
-won.
-
-We may use function `volley()` to convert this to a likelihood
+won.  We may use function `volley()` to convert this to a likelihood
 function:
 
-```{r calcmaxpvolley}
+```{r calcmaxpvolley,cache=TRUE}
 volleyball <- volley(volleyball_table)

### 2023-10-30 inst/volvo.Rmd
label all chunks, cache the slow ones
consistentification

@@
   round-the-world Volvo Ocean Race"
+author: "Robin K. S. Hankin"
 output: bookdown::html_document2
+bibliography: jss_hyper3.bib
 ---
@@
 
-<p style="text-align: right;">
-![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
-</p>
+```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
+knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+```
 
+To cite the `hyper2` package in publications, please use @hankin2017a.
 This short document creates the `volvo` suite of R objects and
@@
 
-```{r}
+```{r specifynoscore,cache=TRUE}
 volvo <- ordertable2supp(volvo_table,noscore=c("DNF","DNS"))
@@
 
-```{r,equalvolvostrengths}
+```{r,label=equalvolvostrengths,cache=TRUE}
 equalp.test(volvo)
+```
+
+### Package dataset {-}
+
+Following lines create `volvo.rda`, residing in the `data/` directory of the package.
+
+```{r,label=savecurlingdataset}
+save(volvo_table,volvo_maxp,volvo,file="volvo.rda")
 ```
@@
 (https://en.wikipedia.org/w/index.php?title=2014\%E2\%80\%932015_Volvo_Ocean_Race&oldid=914916131)
-
-### Package dataset {-}
-
-Following lines create `volvo.rda`, residing in the `data/` directory of the package.
-
-```{r,label=savecurlingdataset}
-save(volvo_table,volvo_maxp,volvo,file="volvo.rda")
-```

### 2023-10-30 inst/zermelo.Rmd
consistentification

@@
 title: "Some thoughts on Zermelo's iteration process in the context of the hyper2 package"
+author: "Robin K. S. Hankin"
 output: bookdown::html_document2
+bibliography: jss_hyper3.bib
 ---
@@
 
-<p style="text-align: right;">
-![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
-</p>
+```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
+knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+```
 
+To cite the `hyper2` package in publications, please use @hankin2017a.
 This short document sets out a use-case for functions `zermelo()` and

### 2023-10-31 inst/carcinoma.Rmd
consistentification

@@
 ---
-title: "Carcinoma"
+title: "Analysing a carcinoma dataset with the hyper2 package"
+author: "R. K. S. Hankin"
 output: bookdown::html_document2
+bibliography: jss_hyper3.bib
 ---
@@
 
+```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
+knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+```
+
+To cite the `hyper2` package in publications, please use @hankin2017a
 (takes maybe twenty minutes to run without cache.  The value of
@@
 takes a few days to run).
-
-<p style="text-align: right;">
-![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
-</p>
 
@@
 
-Agresti, table 13.1, p542.
-
-Landis, J. R., and G. G. Koch. 1977. An application of hierarchical
-kappa-type statistics in the assessment of majority agreement among
-multiple observers. Biometrics 33: 363-374.
+```{r,label=savecarcinomadataset}
+save(carcinoma_table,carcinoma, carcinoma_count, carcinoma_maxp,file="carcinoma.rda")
+```
 
@@
 
-```{r,label=savecarcinomadataset}
-save(carcinoma_table,carcinoma, carcinoma_count, carcinoma_maxp,file="carcinoma.rda")
-```
+### References {-}
+
+Agresti, table 13.1, p542.
+
+Landis, J. R., and G. G. Koch. 1977. An application of hierarchical
+kappa-type statistics in the assessment of majority agreement among
+multiple observers. Biometrics 33: 363-374.

### 2023-10-31 inst/collusion3.Rmd
consistentification

@@
 title: "Chess collusion with hyper3 objects"
-author: "Robin KS Hankin"
-output: html_document
+author: "R. K. S. Hankin"
+output: bookdown::html_document2
+bibliography: jss_hyper3.bib
 ---
@@
 
-<p style="text-align: right;">
-![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
-</p>
+```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
+knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+```
 
-We analyse results from Curacao 1962 using `hyper3` formalism using
-the weighted Draw monster approach outlined in `kka.Rmd`:
+To cite the `hyper2` package in publications, please use @hankin2017a.
+Here I analyse results from Curacao 1962 using `hyper3` formalism
+using the weighted Draw monster approach outlined in `kka.Rmd`:
 

### 2023-10-31 inst/cook.Rmd
consistentification

@@
 title: "Domestic cooking and likelihood"
-output: html_document
+author: "R. K. S. Hankin"
+output: bookdown::html_document2
+bibliography: jss_hyper3.bib
+
 ---
@@
 
-<p style="text-align: right;">
-![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
-</p>
+```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
+knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+```
 
+To cite the `hyper2` package in publications, please use @hankin2017a.
 Suppose there are five people in a family.  Each day for five days,
@@
 exactly $-1$; note that the aspect ratio of the graph is not unity]
+
+### References {-}

### 2023-10-31 inst/curling.Rmd
use sticker in Rmd
name all chunks and cache the slow ones
consistentification

@@
 ---
-title: "curling"
-output: html_document
+title: "curling with hyper2 objects"
+author: "R. K. S. Hankin"
+output: bookdown::html_document2
+bibliography: jss_hyper3.bib
+
 ---
@@
 
+```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
+knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+```
+
+To cite the `hyper2` package in publications, please use @hankin2017a.
 [Curling](https://en.wikipedia.org/wiki/Curling) is a sport with
@@
 
-```{r,makehyper2curling}
+```{r,makehyper2curling,cache=TRUE}
 curling2 <- hyper2()
@@
 
-### References
-
-Curling at the Winter Olympics, https://en.wikipedia.org/w/index.php?title=Curling_at_the_Winter_Olympics&oldid=998891075 (last visited Jan. 21, 2021).
-
-### Package dataset
+### Package dataset {-}
 
@@
 ```
+
+### References {-}
+
+Curling at the Winter Olympics, https://en.wikipedia.org/w/index.php?title=Curling_at_the_Winter_Olympics&oldid=998891075 (last visited Jan. 21, 2021).

### 2023-10-31 inst/icons.Rmd
use sticker in Rmd
consistentification

@@
 ---
-title: "icons"
-output: html_document
+title: "Analysing icons for climate change and with `hyper3` objects"
+author: "R. K. S. Hankin"
+output: bookdown::html_document2
+bibliography: jss_hyper3.bib
 ---
@@
 
+```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
+knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+```
+
+To cite the `hyper2` package in publications, please use @hankin2017a.
 This minimal document reads file `icons.txt` and creates R objects
 `icons`, `icons_matrix`, and `icons_maxp`.  The dataset is discussed
-extensively in West and Hankin 2008, and Hankin 2010.
+extensively in @west2008 and @hankin2010.
 

### 2023-10-31 inst/javelin.Rmd
statistical tests
include sticker
consistentification

@@
 title: "Javelins with hyper3 objects"
-author: "Robin KS Hankin"
-date: "10/04/2022"
-output: html_document
+author: "R. K. S. Hankin"
+output: bookdown::html_document2
+bibliography: jss_hyper3.bib
 ---
@@
 
-We analyse results from the men's javelin, 2020 Summer Olympics.
+```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
+knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+```
+
+To cite the `hyper2` package in publications, please use @hankin2017a.
+Here we analyse results from the men's javelin, 2020 Summer Olympics.
 
@@
 
-Reference
-
-`https://en.wikipedia.org/w/index.php?title=Athletics_at_the_2020_Summer_Olympics_%E2%80%93_Men%27s_javelin_throw&oldid=1051014985`
+```{r,dothetests,cache=TRUE}
+equalp.test(javelin1)
+equalp.test(javelin2)
+```
 
@@
 ```
+
+## References {-}
+
+`https://en.wikipedia.org/w/index.php?title=Athletics_at_the_2020_Summer_Olympics_%E2%80%93_Men%27s_javelin_throw&oldid=1051014985`

### 2023-10-31 inst/jester.Rmd
consistentification

@@
 title: "The Jester joke recommender system"
-output: html_document
+author: "R. K. S. Hankin"
+output: bookdown::html_document2
+bibliography: jss_hyper3.bib
 ---
@@
 
-<p style="text-align: right;">
-![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
-</p>
+```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
+knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+```
 
-(this file creates objects `jester` and `maxjest`, which are datasets
+To cite the `hyper2` package in publications, please use @hankin2017a.
+This file creates objects `jester` and `maxjest`, which are datasets
 available in the `hyper2` package, and documented at `jester.Rd`.
-This file takes quite a long time to run).
-
-Goldberg et al present a dataset in which respondents rated a number
-of jokes.  Here, I analyse a small portion of this dataset using the
-`hyper2` package.  This document is intended to illustrate an
-extremely challenging application of the `hyper2` package and (without
-cache) takes a long time to process.  Goldberg's dataset has 24938
-lines, one per respondent, and 101 columns, one per joke (the first
-column shows the number of jokes rated by each respondent); here I use
-150 lines and 99 jokes (the 100th joke was not funny).
+This file takes quite a long time to run.  Goldberg et al present a
+dataset in which respondents rated a number of jokes.  Here, I analyse
+a small portion of this dataset using the `hyper2` package.  This
+document is intended to illustrate an extremely challenging
+application of the `hyper2` package and (without cache) takes a long
+time to process.  Goldberg's dataset has 24938 lines, one per
+respondent, and 101 columns, one per joke (the first column shows the
+number of jokes rated by each respondent); here I use 150 lines and 99
+jokes (the 100th joke was not funny).
 
@@
 ```
+
+ ### References

### 2023-10-31 inst/loser.Rmd
consistentification

@@
 title: "Weakly informative observations: identifying the weakest player"
-output: html_document
+author: "R. K. S. Hankin"
+output: bookdown::html_document2
+bibliography: jss_hyper3.bib
 ---
@@
 
-<p style="text-align: right;">
-![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
-</p>
+```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
+knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+```
 
-# Partial rank observations: who lost?
-
-(see also `inst/notthewinner.Rmd`).
-Here we consider a situation in which we make only a rather
-uninformative observation: namely, the identity of the _loser_ (which
-is here taken to be me, player `a`).  We have six competitors
-`a,b,c,d,e,f` with nonnegative strengths $p_1,p_2,p_3,p_4,p_5,p_6,
-\sum p_i=1$.  Competitor `a` has strength $p_1$, and loses.  We are
-interested in how informative this observation is with regard to
-various assertions about the strength of `a`.  We have three
-hypotheses:
+To cite the `hyper2` package in publications, please use @hankin2017a;
+see also `inst/notthewinner.Rmd`.  Here we consider a situation in
+which we make only a rather uninformative observation: namely, the
+identity of the _loser_ (which is here taken to be me, player `a`).
+We have six competitors `a,b,c,d,e,f` with nonnegative strengths
+$p_1,p_2,p_3,p_4,p_5,p_6, \sum p_i=1$.  Competitor `a` has strength
+$p_1$, and loses.  We are interested in how informative this
+observation is with regard to various assertions about the strength of
+`a`.  We have three hypotheses:
 
@@
 considered separately in `inst/notthewinner.Rmd`.
+
+### References {-}

### 2023-10-31 inst/ordertable_to_ranktable.Rmd
minor improvements
consistentification

@@
 title: "The print method for class ranktable"
-output: html_document
+author: "R. K. S. Hankin"
+output: bookdown::html_document2
+bibliography: jss_hyper3.bib
 ---
@@
 
-# Introduction
+```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
+knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+```
 
+To cite the `hyper2` package in publications, please use @hankin2017a.
 Rank tables can be confusing.  The package includes a range of
@@
 
-```{r}
+```{r makeatoy}
 OT <- cbind(1:5,c(1,5,2:4))
@@
 
-```{r}
+```{r useordertorank}
 RT <- ordertable_to_ranktable(OT)
@@
 
-```{r}
+```{r viewdirect}
 class(RT)
@@
 
-```{r}
+```{r showvolvotable}
 volvo_table
@@
 
-```{r}
+```{r nowusewikitorank}
 x <- wikitable_to_ranktable(volvo_table)
@@
 
-```{r}
+```{r just3}
 x[,1:3]
@@
 
-```{r}
+```{r nowmakerank3}
 xx <- x[,1:3]
@@
 
-```{r}
+```{r showprintableobject}
 ranktable_to_printable_object(x)[,1:3]
 ```
+
+## References {-}

### 2023-10-31 inst/pentathlon.Rmd
consistentification

@@
 title: "Pentathlon"
+author: "R. K. S. Hankin"
 output: bookdown::html_document2
+bibliography: jss_hyper3.bib
 ---
@@
 
-<p style="text-align: right;">
-![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
-</p>
+```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
+knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+```
 
+To cite the `hyper2` package in publications, please use @hankin2017a.
 The [modern

### 2023-10-31 inst/pudding.Rmd
consistentification

@@
 title: "Analysis of pudding following Davidson 1970"
+author: "R. K. S. Hankin"
 output: bookdown::html_document2
+bibliography: jss_hyper3.bib
 ---
@@
 
-<p style="text-align: right;">
-![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
-</p>
+```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
+knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+```
 
+To cite the `hyper2` package in publications, please use @hankin2017a.
 This short document creates the `pudding` suite of R objects and
@@
 
-# References
+### Package dataset {-}
+
+Following lines create `pudding.rda`, residing in the `data/` directory of the package.
+
+```{r,label=savecurlingdataset}
+save(pudding_table,pudding,pudding_maxp,curling2_maxp,file="pudding.rda")
+```
+
+### References {-}
 
@@
 _Journal of the American Statistical Association_, Volume 65, Number 329, pp317--328
-
-### Package dataset
-
-Following lines create `pudding.rda`, residing in the `data/` directory of the package.
-
-```{r,label=savecurlingdataset}
-save(pudding_table,pudding,pudding_maxp,curling2_maxp,file="pudding.rda")
-```

### 2023-10-31 inst/NBA.Rmd
consistentification

@@
 title: "An analysis of a single basketball game"
-output: html_document
+author: "R. K. S. Hankin"
+output: bookdown::html_document2
+bibliography: jss_hyper3.bib
 ---
@@
 
-<p style="text-align: right;">
-![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
-</p>
+```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
+knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+```
 
-File `NBA.txt` contains a point-by-point analysis of a basketball
-match between the Cleveland Cavaliers and the Golden State Warriors on
-13 June 2017, data taken from
-[here](https://www.espn.com/nba/playbyplay?gameId=400954514).
+To cite the `hyper2` package in publications, please use @hankin2017a.
+Here I consider a single basketball match from the perspective of the
+Bradley-Terry model.  File `NBA.txt` contains a point-by-point
+analysis of a basketball match between the Cleveland Cavaliers and the
+Golden State Warriors on 13 June 2017, data taken from
+[here](https://www.espn.com/nba/playbyplay?gameId=400954514).  Some
+further documentation is given at `inst/NBA.Rd` in the package.
 

### 2023-10-31 inst/T20.Rmd
consistentification

@@
 author: "Tanvi Chandel and Robin K. S. Hankin"
-date: "`r Sys.Date()`"
-vignette: |
-  %\VignetteIndexEntry{T20}
-  %\VignetteEngine{knitr::rmarkdown}
-  %\VignetteEncoding{UTF-8}
+output: bookdown::html_document2
+bibliography: jss_hyper3.bib
 ---
 
-<p style="text-align: right;">
-![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
-</p>
+```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
+knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+```
 
-# The Indian Premier League
-
+To cite the `hyper2` package in publications, please use @hankin2017a.
 This script creates and analyses hyper2 object `T20` which is a
@@
 Premier League.
+Dataframe `T20_table` has one row for each T20 IPL match in the
+period 2008-2017 with the exception of three no-result matches and
+seven tied matches, which were removed.
 
@@
 
-Dataframe `T20_table` has one row for each T20 IPL match in the
-period 2008-2017 with the exception of three no-result matches and
-seven tied matches, which were removed.  Object `T20` is a
-likelihood function for the strengths of the 13 teams, and
-`T20_toss` is a likelihood function that also includes a toss
-strength term.
-
-some details are given in the package help file
+Object `T20` is a likelihood function for the strengths of the 13
+teams, and `T20_toss` is a likelihood function that also includes a
+toss strength term.  Some details are given in the package help file
 `T20.Rd` (type `?T20` at the R prompt).

### 2023-10-31 inst/cook_repeated.Rmd
consistentification

@@
 title: "Judging domestic cooking ability: repeated incomplete rank observations"
-output: html_document
+author: "R. K. S. Hankin"
+output: bookdown::html_document2
+bibliography: jss_hyper3.bib
 ---
@@
 ```{r setup, include=FALSE}
+set.seed(1)
 knitr::opts_chunk$set(echo = TRUE)
+library("hyper2")
 ```
 
-<p style="text-align: right;">
-![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
-</p>
+```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
+knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+```
 
-(takes about five minutes without cache).
-Suppose there are five people in a family.  Each day for five days,
-one person cooks a meal.  After the fifth day, each person considers
-the four meals cooked by other family members, and puts them in order
-of preference.  This process is repeated and here I consider how to
-interpret such data.
+To cite the `hyper2` package in publications, please use @hankin2017a;
+takes about five minutes without cache.  Suppose there are five people
+in a family.  Each day for five days, one person cooks a meal.  After
+the fifth day, each person considers the four meals cooked by other
+family members, and puts them in order of preference.  This process is
+repeated and here I consider how to interpret such data.
 
@@
 method.  First we will consider the support function for the first row only:
-
-### First row of the dataset
 
@@
 
-## Complete dataset
+## Complete dataset {-}
 
@@
 
-# Preferential voting
+# Preferential voting {-}
 

### 2023-10-31 inst/counterstrike.Rmd
use cache
consistentification

@@
 title: "A statistical analysis of a counterstrike game: Boston 2018"
-output: html_document
+author: "R. K. S. Hankin"
+output: bookdown::html_document2
+bibliography: jss_hyper3.bib
 ---
@@
 
-<p style="text-align: right;">
-![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
-</p>
+```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
+knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+```
 
-# Abstract
-
+To cite the `hyper2` package in publications, please use @hankin2017a.
 In this short vignette I use the `hyper2` package to analyse an
@@
 
-```{r, plotit}
+```{r, calcmax,cache=TRUE}
 counterstrike_maxp <- maxp(counterstrike)
+```
+
+```{r, dotchartmaxp}
 dotchart(counterstrike_maxp,pch=16,main='observed data')
@@
 
-```{r, randomdata}
+```{r, label=randomdata,cache=TRUE}
 

### 2023-10-31 inst/distance_matrix_idea.Rmd
typo
name all chunks and cache the slow ones
consistentification

@@
 title: "icons: some experimental results"
-output: html_document
+author: "R. K. S. Hankin"
+output: bookdown::html_document2
+bibliography: jss_hyper3.bib
 ---
@@
 
-<p style="text-align: right;">
-![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
-</p>
+```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
+knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+```
 
-This file, `icons2.Rmd`, presents some speculative experimental work
-which is under development.  Do not confuse this file with
-`icons.Rmd`, which creates the `icons` objects in the package.
+To cite the `hyper2` package in publications, please use @hankin2017a.
+This file presents some _speculative_ experimental work which is under
+development.  Do not confuse this file with `icons.Rmd`, which creates
+the `icons` objects in the package.
 
@@
 
-```{r}
+```{r roundm}
 M
@@
 
-```{r}
+```{r plotphangorn}
 plot(phangorn::upgma(as.dist(M)))
 ```
+
+### References {-}

### 2023-10-31 inst/eurodance.Rmd
say how long it takes to run
consistentification

@@
 title: "Eurovision **Dance** Contest 2008"
+author: "R. K. S. Hankin"
 output: bookdown::html_document2
+bibliography: jss_hyper3.bib
 ---
@@
 
-<p style="text-align: right;">
-![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
-</p>
+```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
+knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+```
 
-The file `eurodance2008.txt`, used below, is copied from
-"Eurovision Dance Contest 2008", Wikipedia, accessed March 31, 2021.
-First we specify the matrix as appearing in the Wikipedia page:
+To cite the `hyper2` package in publications, please use @hankin2017a;
+takes about 20 minutes to process without cache.  The file
+`eurodance2008.txt`, used below, is copied from "Eurovision Dance
+Contest 2008", Wikipedia, accessed March 31, 2021.  First we specify
+the matrix as appearing in the Wikipedia page:
 
@@
 ```
+
+###R References {-}

### 2023-10-31 inst/eurovision.Rmd
consistentification

@@
 output: bookdown::html_document2
+author: "R. K. S. Hankin"
+output: bookdown::html_document2
+bibliography: jss_hyper3.bib
 ---
@@
 
-<p style="text-align: right;">
-![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
-</p>
+```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
+knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+```
 
+To cite the `hyper2` package in publications, please use @hankin2017a.
 The file `eurovision.txt`, used below, is copied from "Eurovision Song
@@
 
-# References {-}
-
-A. Price 2020. "Identifying voting blocs in the Eurovision Song Contest".  Towards Data Science,
-`https://medium.com/towards-data-science/identifying-voting-blocs-in-the-eurovision-song-contest-4792065fc337`
-
 ### Package dataset {-}
@@
 ```
+
+### References {-}
+
+A. Price 2020. "Identifying voting blocs in the Eurovision Song Contest".  Towards Data Science,
+`https://medium.com/towards-data-science/identifying-voting-blocs-in-the-eurovision-song-contest-4792065fc337`

### 2023-10-31 inst/fatigue.Rmd
consistentification

@@
 title: "fatigue"
-author: "Robin Hankin"
-date: "16/03/2021"
-output: html_document
+author: "R. K. S. Hankin"
+output: bookdown::html_document2
+bibliography: jss_hyper3.bib
 ---
@@
 ```{r setup, include=FALSE}
+set.seed(1)
 knitr::opts_chunk$set(echo = TRUE)
@@
 
-<p style="text-align: right;">
-![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
-</p>
+```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
+knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+```
 
-## R Markdown
-
-Analysis of  the 1973 interzonal competition
+To cite the `hyper2` package in publications, please use @hankin2017a.
+Analysis of the 1973 interzonal competition.
 

### 2023-10-31 inst/global_liveability_ranking.Rmd
consistentification

@@
 title: "Global liveability ranking"
-output: html_document
+author: "R. K. S. Hankin"
+output: bookdown::html_document2
+bibliography: jss_hyper3.bib
 ---
@@
 
+```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
+knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+```
+
+To cite the `hyper2` package in publications, please use @hankin2017a;
+takes about an hour to process without cache.
+
 _This analysis is essentially without value, see the very end for why_
-
-(takes about an hour to process without cache)
-
-<p style="text-align: right;">
-![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
-</p>
 
@@
 fat waste of time.
+
+## References {-]

### 2023-10-31 inst/grad_weight.Rmd
consistentification

@@
 ---
-title: "grad_weight"
-author: "Robin KS Hankin"
-date: "10/04/2022"
-output: html_document
+title: "Differentiation with respect to weights for a `hyper3` object"
+author: "R. K. S. Hankin"
+output: bookdown::html_document2
+bibliography: jss_hyper3.bib
 ---
@@
 
-<p style="text-align: right;">
-![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
-</p>
+```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
+knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+```
 
-## Differentiate with respect to weights for a `hyper3` object
-
+To cite the `hyper2` package in publications, please use @hankin2017a.
 We will start with a simple example.  Consider two players $p_1$ and

### 2023-10-31 inst/handover.Rmd
consistentification

@@
 ---
-title: "handover"
+title: "handover issues in medical care"
+author: "R. K. S. Hankin"
 output: bookdown::html_document2
+bibliography: jss_hyper3.bib
 ---
@@
 
-<p style="text-align: right;">
-![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
-</p>
+```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
+knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+```
 
-This short document is based on Altham and Hankin 2010, which
+To cite the `hyper2` package in publications, please use @hankin2017a.
+This short document is based on @altham2010, which
 discussed an interesting dataset originally presented by Lin
@@
 
-### References {-}
-
-* PME Altham and RKS Hankin 2010.  "Using recently developed software on
-a 2x2 table of matched pairs with incompletely classified data".
-_Journal of the Royal Statistical Society_, 59:2):377-379
-* L West and RKS Hankin 2008. "Exact tests for two-way contingency
-tables with structural zeros".  _Journal of Statistical Software_
-28(11)
-* RKS Hankin 2010. "A generalization of the Dirichlet distribution". 
- _Journal of Statistical Software_ 33(11)
-* RKS Hankin 2017. "Partial rank data with the hyper2 package:
-  likelihood functions for generalized Bradley-Terry models".  _The R
-  Journal_ 9(2)
-
 ### Package dataset {-}

### 2023-10-31 inst/hepatitis.Rmd
consistentification

@@
 title: "Hepatitis"
+author: "R. K. S. Hankin"
 output: bookdown::html_document2
+bibliography: jss_hyper3.bib
 ---
@@
 
-<p style="text-align: right;">
-![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
-</p>
+```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
+knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+```
+
+To cite the `hyper2` package in publications, please use @hankin2017a.
 

### 2023-10-31 inst/home_advantage.Rmd
consistentification

@@
 ---
-title: "Quantifying the home ground advantage"
-author: "Robin K. S. Hankin"
-date: "`r Sys.Date()`"
-vignette: >
-  %\VignetteIndexEntry{home_ground_advantage}
-  %\VignetteEngine{knitr::rmarkdown}
-  %\VignetteEncoding{UTF-8}
+title: "Quantifying the home ground advantage with `hyper2` objects"
+author: "R. K. S. Hankin"
+output: bookdown::html_document2
+bibliography: jss_hyper3.bib
 ---
-
-<p style="text-align: right;">
-![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
-</p>
 
@@
 
-(takes about fifteen minutes to run without cache)
+```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
+knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+```
 
-Here I analyse a small synthetic dataset for home ground advantage,
-then go on to discuss a dataset presented by Agresti.  But first, the
-smallest nontrivial example is a 2x2 case:
+To cite the `hyper2` package in publications, please use @hankin2017a
+(takes about fifteen minutes to run without cache).  Here I analyse a
+small synthetic dataset for home ground advantage, then go on to
+discuss a dataset presented by Agresti.  But first, the smallest
+nontrivial example is a 2x2 case:
 

### 2023-10-31 inst/home_advantage3.Rmd
consistentification

@@
 title: "Quantifying the home advantage with `hyper3` objects"
-author: "Robin K. S. Hankin"
-date: "`r Sys.Date()`"
-vignette: |
-  %\VignetteIndexEntry{hyper3}
-  %\VignetteEncoding{UTF-8}
-  %\VignetteEngine{knitr::rmarkdown}
+author: "R. K. S. Hankin"
+output: bookdown::html_document2
+bibliography: jss_hyper3.bib
 ---
 
-<p style="text-align: right;">
-![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
-</p>
+```{r setup, include=FALSE}
+set.seed(1)
+knitr::opts_chunk$set(echo = TRUE)
+library("hyper2")
+```
 
+```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
+knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+```
+
+To cite the `hyper2` package in publications, please use @hankin2017a.
 Objects of class \code{hyper3} are a generalization of \code{hyper2}
@@
 
-## Aside: `hyper2` analysis of the same scoreline
+# Aside: `hyper2` analysis of the same scoreline
 

### 2023-10-31 inst/hunter.Rmd
correct many mistakes
consistentification

@@
 title: "Iteration techniques for Plackett-Luce likelihood functions and the `hyper2` package"
+author: "R. K. S. Hankin"
 output: bookdown::html_document2
+bibliography: jss_hyper3.bib
 ---
@@
 
-<p style="text-align: right;">
-![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
-</p>
+```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
+knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+```
 
-Here I discuss why implementing equation 30 of Hunter (2004) for
+To cite the `hyper2` package in publications, please use @hankin2017a.
+Here I discuss why implementing equation 30 of [@hunter2004] for
 `hyper2` objects representing Plackett-Luce likelihood functions is
@@
 is, 7 came first, then 2, then 4 then 3 then 5 then 6] and `1372`
-[that is, 1 came first, then 3 then 7 then 2].  This would have a
-Plackett-Luce likelihood function of
+[that is, 1 came first, then 3 then 7 then 2].  If these two
+observations are independent, we would have a Plackett-Luce likelihood
+function of
 
@@
 \[
-p_3\longrightarrow\frac{p_3}{(p_7+p_2+p_4+p_3)^{-1}+(p_1+p_3)^{-1}}=
-\frac{[3]}{[7243]^{-1}+[13]^{-1}}
+p_3\longrightarrow\frac{2}{
+(p_7+p_2+p_4+p_3+p_5+p_6)^{-1}+
+(    p_2+p_4+p_3+p_5+p_6)^{-1}+
+(        p_4+p_3+p_5+p_6)^{-1}+
+(            p_3+p_5+p_6)^{-1}+
+(p_1+p_3+p_7+p_2)^{-1}+
+(    p_3+p_7+p_2)^{-1}
+}\]
+
+or, more compactly,
+\[p_3\longrightarrow
+\frac{2}{[724356]^{-1}+[24356]^{-1}+[4356]^{-1}+[356]^{-1}+[1372]^{-1}+[372]^{-1}}.
 \]
@@
 
-$p_1 \longrightarrow [1]$
+$p_1\longrightarrow\frac{1}{[1237]^{-1}}$
 
-$p_2 \longrightarrow\frac{[2]}{[27]^{-1}+[1237]^{-1}}$
+$p_2\longrightarrow\frac{1}{[234567]^{-1}+[23456]^{-1}}$
 
-$p_3 \longrightarrow\frac{[3]}{[2347]^{-1}+[13]^{-1}}$
+$p_3\longrightarrow
+\frac{2}{[273456]^{-1}+[23456]^{-1}+[3456]^{-1}+[356]^{-1}+[1237]^{-1}+[237]^{-1}}$
 
-$p_4 \longrightarrow\frac{[4]}{[247]^{-1}}$
+$p_4 \longrightarrow\frac{1}{[724356]^{-1}+[24356]^{-1}+[4356]^{-1}}$
 
-$p_5 \longrightarrow\frac{[5]}{[23457]^{-1}}$
+$p_5\longrightarrow\frac{1}{[273456]^{-1}+[23456]^{-1}+[3456]^{-1}+[356]^{-1}+[56]^{-1}}$
 
-$p_6 \longrightarrow\frac{[6]}{[234567]^{-1}}$
+$p_6 \longrightarrow\frac{0}{[6]^{-1}}$
 
-$p_7 \longrightarrow\frac{[7]}{[137]^{-1}}$
+$p_7 \longrightarrow\frac{2}{[273456]^{-1}+[27]^{-1}}$
 
 and it does not seem at all easy to convert the `hyper2` object to
-Hunter's iteration scheme.  To update $p_3$, for example, we would
-start with the denominator $[1237]\cdot[23456]\cdot[234567]\cdot
+Hunter's iteration scheme.  The numerators of the fractions are just
+the powers of the singleton terms in the likelihood function.  But the
+denominators are harder.  To update $p_3$, for example, we would start
+with the denominator $[1237]\cdot[23456]\cdot[234567]\cdot
 [237]\cdot[27]\cdot[3456]\cdot[356]\cdot[56]$ and then extract just
 the brackets containing 3 to get $[1237]\cdot[23456]\cdot[234567]\cdot
-[237]\cdot[3456]\cdot[356]$ and then somehow identify two increasing chains
+[237]\cdot[3456]\cdot[356]$ and then somehow identify two increasing
+chains
 
@@
 
-and _then_ somehow identify from these two chains that 1 beats 3
-(leading to $[13]$ in the iteration) and also that $\left\lbrace
-2,4,7\right\rbrace$ beat 3 (leading to $[2347]$).
-
-### Reference {-}
-
-* D. R. Hunter 2004.  "MM algorithms for generalized Bradley-Terry
-  models".  _The Annals of Statistics_, 32(1):384
+### References {-}

### 2023-10-31 inst/hyper3.Rmd
consistentification

@@
 title: "The weighted hyperdirichlet distribution: the `hyper3` class"
-author: "Robin K. S. Hankin"
-date: "`r Sys.Date()`"
-vignette: |
-  %\VignetteIndexEntry{hyper3}
-  %\VignetteEncoding{UTF-8}
-  %\VignetteEngine{knitr::rmarkdown}
+author: "R. K. S. Hankin"
+output: bookdown::html_document2
+bibliography: jss_hyper3.bib
 ---
 
-<p style="text-align: right;">
-![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
-</p>
+```{r setup, include=FALSE}
+set.seed(1)
+knitr::opts_chunk$set(echo = TRUE)
+library("hyper2")
+```
 
+```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
+knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+```
+
+To cite the `hyper2` package in publications, please use @hankin2017a.
 Objects of class \code{hyper3} are a generalization of \code{hyper2}

### 2023-10-31 inst/interzonal.Rmd
consistentification

@@
 title: "Analysis of the 1962 world chess championship, Curacao"
-output: html_document
+author: "R. K. S. Hankin"
+output: bookdown::html_document2
+bibliography: jss_hyper3.bib
 ---
@@
 
-<p style="text-align: right;">
-![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
-</p>
+```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
+knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+```
 
-This file creates likelihood functions `H` and `H_coll`, which appear in
-the package as `interzonal` and `collusion` respectively.
+To cite the `hyper2` package in publications, please use @hankin2017a.
+This file creates likelihood functions `H` and `H_coll`, which appear
+in the package as `interzonal` and `collusion` respectively.  It also
+presents some analysis of the datasets using the `hyper2` package.
 
@@
 
-# References
-
-C. C. Moul and J. V. C. Nye 2009. "Did the Soviets collude? A
- statistical analysis of championship chess 1940-1978.  Journal of
- Economic Behaviour and Organization 70(2009) 10-21
+# References {-}

### 2023-10-31 inst/karate.Rmd
consistentification

@@
 title: "karate"
-output: html_document
+author: "R. K. S. Hankin"
+output: bookdown::html_document2
+bibliography: jss_hyper3.bib
 ---
@@
 
-<p style="text-align: right;">
-![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
-</p>
+```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
+knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+```
 
+To cite the `hyper2` package in publications, please use @hankin2017a.
 Here I analyse a [karate](https://en.wikipedia.org/wiki/Karate)

### 2023-10-31 inst/kka.Rmd
consistentification

@@
 ---
-title: "Karpov, Kasparov, Anand"
-output: html_document
+title: "A Bradley-Terry analysis of three grandmasters: Karpov, Kasparov, Anand"
+author: "R. K. S. Hankin"
+output: bookdown::html_document2
+bibliography: jss_hyper3.bib
 ---
@@
 
-(takes about 45 minutes to process without cache)
+```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
+knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+```
 
-<p style="text-align: right;">
-![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
-</p>
-
-This file defines dataset `kka` which is the observed scorelines
-between players Karpov, Kasparov, and Anand.  It is used to calculate
+To cite the `hyper2` package in publications, please use @hankin2017a;
+this `Rmd` file takes about 45 minutes to process without cache.  Here
+we define dataset `kka` which is the observed scorelines between
+players Karpov, Kasparov, and Anand.  It is used to calculate
 likelihood functions: `karpov_kasparov_anand`, `kka_3whites` and
@@
 `inst/karpov_kasparov_anand.R` and `inst/kka_3draws.R` and
-`inst/kka_array.R`.  Some more detail is given in
-`man/kka.Rd`.
+`inst/kka_array.R`.  Some more detail is given in `man/kka.Rd`.
 

### 2023-10-31 inst/monster_vs_lambda.Rmd
consistentification

@@
 title: "Monster or Lambda?  A likelihood analysis"
-author: "Robin KS Hankin"
-output: html_document
+author: "R. K. S. Hankin"
+output: bookdown::html_document2
+bibliography: jss_hyper3.bib
 ---
@@
 
-(takes about an hour to process without cache)
+```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
+knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+```
 
-<p style="text-align: right;">
-![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
-</p>
-
-Here I consider pairwise comparisons with a home-ground advantage and
-wish to compare two likelihood functions.  First, the monster parametrization:
+To cite the `hyper2` package in publications, please use @hankin2017a;
+takes about an hour to process without cache.  Here I consider
+pairwise comparisons with a home-ground advantage and wish to compare
+two likelihood functions.  First, the monster parametrization:
 

### 2023-10-31 inst/multiset_hyper2.Rmd
consistentification

@@
 verification with the `multiset()` function of the `partitions` package"
-output: html_document
-date: "2022-07-28"
+author: "R. K. S. Hankin"
+output: bookdown::html_document2
+bibliography: jss_hyper3.bib
 ---
@@
 
-<p style="text-align: right;">
-![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
-</p>
+```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
+knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+```
 
+To cite the `hyper2` package in publications, please use @hankin2017a.
 Here we will sum probabilities across a sample space and verify that
 the total is constant.  In theory the sum of the probabilities of all
-members of the sample space is 1, but in practice the likelihood
-calculation includes an indeterminate constant which means that all we
-can say is that the total probability is not a function of the
-parameters of the distribution.
+members of the sample space is 1, but in practice [and certainly in
+`hyper2` idiom], the likelihood calculation includes an indeterminate
+constant which means that all we can say is that the total probability
+is not a function of the parameters of the distribution.
+
+Here I present some relatively informal and unstructured thoughts on
+this issue from a numerical and `hyper2` perspective.
 
@@
 
-Above we see very tight clustering about the mean value of 0.25
+Above we see very tight clustering about the mean value of 0.25.
+
+### References {-}

### 2023-10-31 inst/order.Rmd
consistentification

@@
 title: "Order statistics and numerical optimization: the case of a single observation"
-output: html_document
+author: "R. K. S. Hankin"
+output: bookdown::html_document2
+bibliography: jss_hyper3.bib
 ---
@@
 
-<p style="text-align: right;">
-![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
-</p>
+```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
+knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+```
 
-## The `hyper2` package
-
-We consider the likelihood function given by a single observation of
-an order statistic.  Here we investigate some numerical problems
-arising from the use of `optim()` and `constrOptim()`.  We will
-consider $n=10$ competitors and without loss of generality, the
-competitors finish in order $1,2,3,\ldots,10$.  The likelihood
-function for this observation is:
+To cite the `hyper2` package in publications, please use @hankin2017a.
+Here I consider the likelihood function given by a single observation
+of an order statistic, which is problematic in the sense that the
+maximum likelihood estimate for the BT strengths of the competitors is
+$(1,0,\ldots,0)$.  Here we investigate some numerical problems arising
+from the use of `optim()` and `constrOptim()`.  We will consider
+$n=10$ competitors and without loss of generality, the competitors
+finish in order $1,2,3,\ldots,10$.  The likelihood function for this
+observation is:
 
@@
 (given that $p_{10}$ is at its minimum value).
+
+## References {-}

### 2023-10-31 inst/plackett_luce_monster.Rmd
label all chunks, cache the slow ones
consistentification

@@
 title: "A generalization of Plackett-Luce likelihoods to incorporate non-independence of competitors"
-output: html_document
-date: '2022-05-16'
+author: "R. K. S. Hankin"
+output: bookdown::html_document2
+bibliography: jss_hyper3.bib
 ---
@@
 
-<p style="text-align: right;">
-![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
-</p>
-
+```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
+knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+```
+
+To cite the `hyper2` package in publications, please use @hankin2017a.
 Here I modify standard Plackett-Luce likelihood functions to include
@@
 
-```{r}
+```{r evalat44}
 f(4.4,3.3, 0.2)*12
@@
 
-```{tikz, fig.cap = "Partial probability tree for five competitors $a-e$ with $a$ supporting $b$, hyper3 approach", fig.ext = 'png', echo=FALSE}
+```{tikz, label=tikzpartprobtree1,fig.cap = "Partial probability tree for five competitors $a-e$ with $a$ supporting $b$, hyper3 approach", fig.ext = 'png', echo=FALSE}
 \usetikzlibrary{arrows}
@@
 
-```{tikz, fig.cap = "Partial probability tree for five competitors $a-e$ with mutually supporting  group $abc$, hyper3 approach.
+```{tikz, label=tikzpic2, fig.cap = "Partial probability tree for five competitors $a-e$ with mutually supporting  group $abc$, hyper3 approach.
 If any competitor in the set {$a,b,c$} finishes, they lend their strength to the other competitors in the set who are still running", fig.ext = 'png', echo=FALSE}
@@
 
-```{tikz, fig.cap = "Partial probability tree for five competitors $a$-$e$ with mutually supporting subsets $\\left\\lbrace a,b,c\\right\\rbrace$ [with support term $\\lambda$] and $(de)$ [with support term $\\mu$], hyper3 approach", fig.ext = 'png', echo=FALSE}
+```{tikz, label=tikz3,fig.cap = "Partial probability tree for five competitors $a$-$e$ with mutually supporting subsets $\\left\\lbrace a,b,c\\right\\rbrace$ [with support term $\\lambda$] and $(de)$ [with support term $\\mu$], hyper3 approach", fig.ext = 'png', echo=FALSE}
 \usetikzlibrary{arrows}
@@
 
-```{r}
+```{r, label=readtableandcalc,cache=TRUE}
 a <- read.table("timesData.csv",sep=",", header=TRUE)

### 2023-10-31 inst/podium.Rmd
consistentification

@@
 title: "Does being on the podium affect Formula 1 drivers' performance in the next race?"
+author: "R. K. S. Hankin"
 output: bookdown::html_document2
+bibliography: jss_hyper3.bib
 ---
@@
 
-<p style="text-align: right;">
-![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
-</p>
+```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
+knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+```
 
+To cite the `hyper2` package in publications, please use @hankin2017a.
 This short document defines the `podium()` function which tests the
 hypothesis that a Formula 1 driver's being on the podium at one venue
-affects his performance on the next.
+affects his performance on the next.  Data from the 2007 season are
+used.  It uses `hyper3` objects from the `hyper2` package.  A detailed
+use-case for the top six drivers (Alonso, Hamilton, Heidfeld, Kubica,
+Massa, and Raikkonen), and only the first five venues (Austria,
+Malaysia, Bahrain, Spain, and Monaco) is given.
+
+***TLDR***: Statistically significant evidence for the podium effect
+was found, with about 2.7 units of support for the alternative.
+
+Function `podium()`, defined below, takes a race table and a value of
+$\lambda$ and returns a `hyper3` support function.
 

### 2023-10-31 inst/powerboat.Rmd
consistentification

--- 2021-02-06 inst/powerboat.Rmd
@@
 ---
-title: "powerboat"
-output: html_document
+title: "powerboats with `hyper2`"
+author: "R. K. S. Hankin"
+output: bookdown::html_document2
+bibliography: jss_hyper3.bib
+---
+
 ---
@@
 
+```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
+knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+```
+
+To cite the `hyper2` package in publications, please use @hankin2017a.
 [powerboat](https://en.wikipedia.org/w/index.php?title=2018_F1_Powerboat_World_Championship&oldid=920386507)

### 2023-10-31 inst/race3.Rmd
do not need to load disordR
consistentification

@@
 title: "Races with hyper3 objects"
-author: "Robin KS Hankin"
-date: "10/04/2022"
-output: html_document
+author: "R. K. S. Hankin"
+output: bookdown::html_document2
+bibliography: jss_hyper3.bib
 ---
@@
 library("hyper2")
-library("disordR")
 ```
 
-(takes about an hour to process without cache)
+```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
+knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+```
 
-<p style="text-align: right;">
-![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
-</p>
-
-Consider a race in which there are six runners 1-6 but we happen to
-know that three of the runners (1,2,3) are clones of strength $p_a$,
-say, two of the runners (4,5) have strength $p_b$, and the final
-runner (6) is of strength $p_c$.  The runners race and the finishing order is:
+(takes about an hour to process without cache).  To cite the `hyper2`
+package in publications, please use @hankin2017a.  Consider a race in
+which there are six runners 1-6 but we happen to know that three of
+the runners (1,2,3) are clones of strength $p_a$, say, two of the
+runners (4,5) have strength $p_b$, and the final runner (6) is of
+strength $p_c$.  The runners race and the finishing order is:
 

### 2023-10-31 inst/retain.Rmd
consistentification

@@
 ---
-title: "Keeping and discarding players"
-output: html_document
+title: "Keeping and discarding players: functions `keep()` and `keep_flawed()`"
+author: "R. K. S. Hankin"
+output: bookdown::html_document2
+bibliography: jss_hyper3.bib
 ---
@@
 
-## Functions keep() and keep_flawed()
+```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
+knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+```
 
+To cite the `hyper2` package in publications, please use @hankin2017a.
 Function `keep_flawed()` can be confusing.  Originally, the package
@@
 according to some criterion, and coerce to a likelihood function.
+
+### References {-}

### 2023-10-31 inst/rock_paper_scissors_monster.Rmd
label all chunks, cache the slow ones
consistentification

@@
 title: "Quantifying the rock-paper-scissors effect"
-output: html_document
+author: "R. K. S. Hankin"
+output: bookdown::html_document2
+bibliography: jss_hyper3.bib
 ---
@@
 
-(takes about an hour to run without cache)
+```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
+knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+```
 
-<p style="text-align: right;">
-![](`r system.file("help/figures/hyper2.png", package = "hyper2")`){width=10%}
-</p>
-
-## Abstract
-
-Here I use the `hyper2` package (Hankin 2017) to quantify the strength
-of a circular rock-paper-scissors effect using the `hyper2` package,
-improving on previous results (West and Hankin 2008) which merely
-established its existence.  I use two datasets, a three-player
-scoreline and a synthetic dataset referring to wear on a seven-tooth
-gear wheel.
+To cite the `hyper2` package in publications, please use @hankin2017a;
+takes about an hour to run without cache.  Here I use the `hyper2`
+package to quantify the strength of a circular rock-paper-scissors
+effect using the `hyper2` package, improving on previous results
+[@west2008] which merely established its existence.  I use two
+datasets, a three-player scoreline and a synthetic dataset referring
+to wear on a seven-tooth gear wheel.
 
@@
 The existence of a rock-paper-scissors relationship between the
-players was discussed in Hankin (2010) and the null of no RPS
-phenomena could be rejected using the Alymer test.
+players was discussed in @hankin2010 and the null of no RPS phenomena
+could be rejected using the Alymer test.
 
@@
 
-```{r}
+```{r,calcmaxp,cache=TRUE}
 maxp(H)
@@
 
-## References
-
-* L. J. West and R. K. S. Hankin 2008.  "Exact tests for two-way
-  contingency tables with structural zeros".  _Journal of Statistical
-  Software_ 2008, volume 28 issue 11.
-* R. K. S. Hankin 2010.  "A generalization of the Dirichlet
-  distribution".  _Journal of Statistical Software_ 2010, volume 33
-  issue 11.
-* R. K. S. Hankin 2017.  "Partial rank data with the `hyper2` package:
-  likelihood functions for generalized Bradley-Terry models".  _The R
-  Journal_, volume 9, number 2.
-
-### Package dataset
+### Package dataset {-}
 

### 2023-11-01 inst/mann_whitney_wilcoxon.Rmd
consistentification

@@
 title: "Generalizing the Mann-Whitney-Wilcoxon test with `hyper3` objects"
-author: "Robin K. S. Hankin"
+author: "R. K. S. Hankin"
 output: bookdown::html_document2
-bibliography: jss_hyper3.bib
+bibliography: hyper2.bib
 ---

### 2023-11-01 inst/red_bus_blue_bus.Rmd
consistentification

@@
 output: pdf_document
-bibliography: chess.bib
+bibliography: hyper2.bib
 ---

### 2023-11-01 inst/surfing.Rmd
name all chunks and cache the slow ones
consistentification

@@
 output: bookdown::html_document2
-bibliography: chess.bib
+bibliography: hyper2.bib
 ---
@@
 specificp.gt.test(scomp3,'S',0,n=100)
-````
+```
 

### 2023-11-01 inst/carcinoma.Rmd
consistentification

@@
 output: bookdown::html_document2
-bibliography: jss_hyper3.bib
+bibliography: hyper2.bib
 ---
@@
 
-To cite the `hyper2` package in publications, please use @hankin2017a
+To cite the `hyper2` package in publications, please use @hankin2017
 (takes maybe twenty minutes to run without cache.  The value of

### 2023-11-01 inst/cook_repeated.Rmd
consistentification

@@
 
-To cite the `hyper2` package in publications, please use @hankin2017a;
+To cite the `hyper2` package in publications, please use @hankin2017;
 takes about five minutes without cache.  Suppose there are five people

### 2023-11-01 inst/eurodance.Rmd
consistentification

@@
 
-To cite the `hyper2` package in publications, please use @hankin2017a;
+To cite the `hyper2` package in publications, please use @hankin2017;
 takes about 20 minutes to process without cache.  The file

### 2023-11-01 inst/global_liveability_ranking.Rmd
consistentification

@@
 
-To cite the `hyper2` package in publications, please use @hankin2017a;
+To cite the `hyper2` package in publications, please use @hankin2017;
 takes about an hour to process without cache.

### 2023-11-01 inst/home_advantage.Rmd
consistentification

@@
 
-To cite the `hyper2` package in publications, please use @hankin2017a
+To cite the `hyper2` package in publications, please use @hankin2017
 (takes about fifteen minutes to run without cache).  Here I analyse a

### 2023-11-01 inst/kka.Rmd
consistentification

@@
 
-To cite the `hyper2` package in publications, please use @hankin2017a;
+To cite the `hyper2` package in publications, please use @hankin2017;
 this `Rmd` file takes about 45 minutes to process without cache.  Here

### 2023-11-01 inst/loser.Rmd
consistentification

@@
 
-To cite the `hyper2` package in publications, please use @hankin2017a;
+To cite the `hyper2` package in publications, please use @hankin2017;
 see also `inst/notthewinner.Rmd`.  Here we consider a situation in

### 2023-11-01 inst/monster_vs_lambda.Rmd
consistentification

@@
 
-To cite the `hyper2` package in publications, please use @hankin2017a;
+To cite the `hyper2` package in publications, please use @hankin2017;
 takes about an hour to process without cache.  Here I consider

### 2023-11-01 inst/rock_paper_scissors_monster.Rmd
consistentification

@@
 
-To cite the `hyper2` package in publications, please use @hankin2017a;
+To cite the `hyper2` package in publications, please use @hankin2017;
 takes about an hour to run without cache.  Here I use the `hyper2`

### 2023-11-09 inst/salad.Rmd
consistentification
fixes #155

@@
 author: "R. K. S. Hankin"
-bibliography: jss_hyper3.bib
+bibliography: hyper2.bib
 ---
@@
 
-To cite the `hyper2` package in publications, please use @hankin2017a.
+To cite the `hyper2` package in publications, please use @hankin2017.
 This short document shows how to apply `hyper2` idiom to the `salad`
-dataset of the `prefmod` package.  From `salad.Rd`:
+dataset of the `prefmod` package [@hatzinger2012].  From `salad.Rd`:
 
@@
 The null estimate agrees to six places of decimals with that
-presented by Turner:
+presented by @turner2020_nomarkup:
 
@@
 # References
-
-Turner HL, van Etten J, Firth D, Kosmidis I (2020).  "Modelling
-Rankings in R: The PlackettLuce Package."  _Computational Statistics_,
-35, 1027–1057. doi:10.1007/s00180-020-00959-3,
-\url{https://doi.org/10.1007/s00180-020-00959-3}
-
-Yildiz, Ilkay, Jennifer Dy, Deniz Erdogmus, Jayashree Kalpathy-Cramer,
-Susan Ostmo, J. Peter Campbell, Michael F. Chiang, and Stratis
-Ioannidis. 2020. "Fast and Accurate Ranking Regression."  In
-Proceedings of the Twenty Third International Conference on Artificial
-Intelligence and Statistics, edited by Silvia Chiappa and Roberto
-Calandra, 108:77–88. Proceedings of Machine Learning
-Research. \url{http://proceedings.mlr.press/v108/yildiz20a.html}

### 2023-11-22 inst/very_simplified_likelihood.Rmd
new idea

---
title: "Very simplified likelihood"
author: "Robin Hankin"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
```

```{r hexsticker, out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
```

Consider a race between $n+1$ competitors of strength $a,b,\ldots,b$
where we require the total strength to be unity: $a+nb=1$.  Suppose
our bro comes first:

\[\mathcal{L}(a)=a\]

Comes second:

\[
\mathcal{L}(a)=b\cdot\frac{a}{1-b}
\]

Third:

\[
\mathcal{L}(a)=b\cdot\frac{b}{1-b}\cdot\frac{a}{1-2b}
\]

Fourth:

\[
\mathcal{L}(a)=b\cdot\frac{b}{1-b}\cdot\frac{b}{1-2b}\cdot\frac{a}{1-3b}
\]

Placing after $r$ of the $b$ clones finish:

\[
\mathcal{L}(a)=\frac{b^ra}{(1-b)(1-2b)\cdots(1-rb)}=(1-nb)\frac{(-1)^r\Gamma(1-1/b)}{\Gamma(1+r-1/b)}
\]

```{r tryit}
small <- 1e-3
f_gamma <- function(a,r,n,log=FALSE){ # vectorised in 'a', but screws up for b=1/integer
  b <- (1-a)/n   # 0 <= b <= 1/n
  out <- log(a) + lgamma(1-1/b) - lgamma(1+r-1/b)
  if(!log){out <- exp(out)}
  return(out)
}

f_single_direct <- function(a,r,n,log=FALSE){ # not vectorised in anything
  stopifnot(length(a)==1)
  stopifnot(length(r)==1)
  stopifnot(length(n)==1)
  b <- (1-a)/n   # 0 <= b <= 1/n
  if(log){
      return(log(a)+r*log(b)-sum(log(1-seq_len(r)*b)))
     } else {
      return(a * b^r / prod(1-seq_len(r)*b))
     }
}

f <- function(a,r,n,log=FALSE){  # vectorized in a
out <- a*0
if(log){
    for(rr in r){  out <- out + f_single_direct(a,rr,n,log=TRUE)}
  } else {
    out[] <- 1
    for(rr in r){  out <- out * f_single_direct(a,rr,n,log=FALSE)}
  }
  return(out)
}

n <- 10
a <- seq(from=small,to=1-small,len=100)
plot(a,f(a,r=3,n=n))
L <- f(a,r=3,n=n,log=TRUE)
L <- L-max(L)
plot(a,L,ylim=c(-5,0))
f(0.9,3,10)
```

### 2023-11-22 inst/very_simplified_likelihood.Rmd
kickoff for simplified likelihood

@@
 small <- 1e-3
+
+f_single_direct <- function(a,r,n,log=FALSE){  # not vectorised in anything
+  b <- (1-a)/n   # 0 <= b <= 1/n
+  if(log){
+      return(log(1-n*b)+r*log(b)-sum(log(1-seq_len(r)*b)))
+     } else {
+      return((1-n*b)*b^r-prod(1-(1:3)*b))
+     }
+}
+
+f <- function(a,r,n,log=FALSE){
+out <- a*0
+if(log){
+    for(rr in r){  out <- out + f_single(a,rr,n,log=TRUE)}
+  } else {
+    out[] <- 1
+    for(rr in r){  out <- out * f_single(a,rr,n,log=FALSE)}
+  }
+  return(out)
+}
+
 f_gamma <- function(a,r,n,log=FALSE){ # vectorised in 'a', but screws up for b=1/integer
@@
 }
+```
 
-f_single_direct <- function(a,r,n,log=FALSE){ # not vectorised in anything
-  stopifnot(length(a)==1)
-  stopifnot(length(r)==1)
-  stopifnot(length(n)==1)
-  b <- (1-a)/n   # 0 <= b <= 1/n
-  if(log){
-      return(log(a)+r*log(b)-sum(log(1-seq_len(r)*b)))
-     } else {
-      return(a * b^r / prod(1-seq_len(r)*b))
-     }
-}
+Now use it:
 
-f <- function(a,r,n,log=FALSE){  # vectorized in a
-out <- a*0
-if(log){
-    for(rr in r){  out <- out + f_single_direct(a,rr,n,log=TRUE)}
-  } else {
-    out[] <- 1
-    for(rr in r){  out <- out * f_single_direct(a,rr,n,log=FALSE)}
-  }
-  return(out)
-}
-
-n <- 10
-a <- seq(from=small,to=1-small,len=100)
-plot(a,f(a,r=3,n=n))
-L <- f(a,r=3,n=n,log=TRUE)
-L <- L-max(L)
-plot(a,L,ylim=c(-5,0))
-f(0.9,3,10)
+```{r}
+ a <- seq(from=0.01,to=0.99,by=0.01)
+ L <- f(a,r=c(3,4,5),10)
+plot(a,L-max(L))
 ```

### 2023-11-23 inst/very_simplified_likelihood.Rmd
more work

@@
   if(log){
-      return(log(1-n*b)+r*log(b)-sum(log(1-seq_len(r)*b)))
+      return(log(a) + r*log(b) - sum(log(1-seq_len(r)*b)))
      } else {
-      return((1-n*b)*b^r-prod(1-(1:3)*b))
+      return(a * b^r/prod(1-seq_len(r)*b))
      }
+}
+
+`f_vectorised_direct` <- function(a,r,n,log=FALSE){  # vectorised in 'a', not in 'r'
+  b <- (1-a)/n   # 0 <= b <= 1/n
+  out <- r*log(b) + log(a) -rowSums(log(1-outer(b,seq_len(r))))
+  if(log){return(out)} else {return(exp(out))}
 }
@@
 if(log){
-    for(rr in r){  out <- out + f_single(a,rr,n,log=TRUE)}
+    for(rr in r){  out <- out + f_vectorised_direct(a,rr,n,log=TRUE)}
   } else {
     out[] <- 1
-    for(rr in r){  out <- out * f_single(a,rr,n,log=FALSE)}
+    for(rr in r){  out <- out * f_vectorised_direct(a,rr,n,log=FALSE)}
   }
@@
 ```{r}
- a <- seq(from=0.01,to=0.99,by=0.01)
- L <- f(a,r=c(3,4,5),10)
+ a <- seq(from=0.05,to=0.45,by=0.01)
+ L <- f(a,r=c(3,4,3,3,1,3,3,3,1,1,1,5),10,log=T)  # 10+1==11 competitors in total
 plot(a,L-max(L))
+abline(v=a[which.max(L)])
+abline(v=1/11)
+abline(h=c(0,-2))
 ```
+
+Now skating:
+
+```{r showhughes}
+skating_table
+hughes <- unlist(skating_table[1,])
+a <- seq(from=0.05,to=0.55,by=0.01)
+L <- f(a,hughes,23,log=TRUE)
+L <- L - max(L)
+plot(a,L,type='b')
+abline(v=1/23)
+```
+
+```{r showall}
+a <- seq(from=0.001,to=0.55,by=0.001)
+fun <- function(i){
+  out <- f(a,unlist(skating_table[i,]),23,log=TRUE)
+  return(out-max(out))
+}
+
+plot(NA,xlim=c(0,1),ylim=c(-10,0),type='n')
+abline(v=1/23)
+for(i in seq_len(9)){points(a,fun(i),type='b')}
+
+plot(NA,xlim=c(-5,1),ylim=c(-5,0),type='n',xlab="log-odds")
+abline(v=-log(22))
+for(i in seq_len(23)){points(log(a/(1-a)),fun(i),type='l')}
+```
+
+# Formula 1
+
+```{r showformula1}
+x <- read.table("formula1_2022.txt",header=TRUE)
+x <- x[,-ncol(x)]
+x
+```
+
+```{r analyseform1}
+fun <- function(i){
+  suppressWarnings(jj <- as.numeric(unlist(x[i,])))
+  jj[is.na(jj)] <- 23
+  L <- f(a,jj,23,log=TRUE)
+  return(L - max(L))
+}
+
+plot(NA,xlim=c(-5,1),ylim=c(-5,1),type='n',xlab="log-odds")
+abline(v=-log(22))
+for(i in seq_len(20)){points(log(a/(1-a)),fun(i),type='l')}
+
+```
+
+# Mathematics Olympiad
+
+```{r}
+x <- as.matrix(read.table("olympiad.txt",header=TRUE))
+x <- x[apply(x,1,max) <= 100,]
+fun <- function(i){
+ jj <- as.numeric(unlist(x[i,]))
+  L <- f(a,jj,100,log=TRUE)
+  return(L - max(L))
+}
+plot(NA,xlim=c(-5,1),ylim=c(-5,1),type='n',xlab="log-odds")
+abline(v=-log(99))
+for(i in seq_len(40)){points(log(a/(1-a)),fun(i),type='l')}
+
+```{r}
+x <- as.matrix(read.table("ol.txt",header=TRUE))[,1:4]
+x <- x[apply(x,1,max) <= 100,]
+fun <- function(i){
+ jj <- as.numeric(unlist(x[i,]))
+  L <- f(a,jj,100,log=TRUE)
+  return(L - max(L))
+}
+plot(NA,xlim=c(-5,1),ylim=c(-5,1),type='n',xlab="log-odds")
+abline(v=-log(99))
+for(i in seq_len(40)){points(log(a/(1-a)),fun(i),type='l')}
+```
+
+```{r}
+x <- as.matrix(read.table("ol.txt",header=TRUE))[,1:4]
+x <- x[apply(x,1,max) <= 100,]
+x[] <- sample(c(x))
+x
+rowSums(x)
+fun <- function(i){
+ jj <- as.numeric(unlist(x[i,]))
+  L <- f(a,jj,100,log=TRUE)
+  return(L - max(L))
+}
+plot(NA,xlim=c(-5,1),ylim=c(-5,1),type='n',xlab="log-odds")
+abline(v=-log(99))
+for(i in seq_len(40)){points(log(a/(1-a)),fun(i),type='l')}

### 2023-11-24 inst/very_simplified_likelihood.Rmd
more work including a vs b a/(a+b) horizontal axis

@@
 library("hyper2")
+library("pracma")
 ```
@@
  a <- seq(from=0.05,to=0.45,by=0.01)
- L <- f(a,r=c(3,4,3,3,1,3,3,3,1,1,1,5),10,log=T)  # 10+1==11 competitors in total
+ L <- f(a,r=5,10,log=T)  # 10+1==11 competitors in total
 plot(a,L-max(L))
@@
 plot(NA,xlim=c(0,1),ylim=c(-10,0),type='n')
-abline(v=1/23)
 for(i in seq_len(9)){points(a,fun(i),type='b')}
 
-plot(NA,xlim=c(-5,1),ylim=c(-5,0),type='n',xlab="log-odds")
-abline(v=-log(22))
-for(i in seq_len(23)){points(log(a/(1-a)),fun(i),type='l')}
+plot(NA,xlim=c(-3,3),ylim=c(-5,0),type='n',xlab="log-odds")
+for(i in seq_len(23)){points(log(22*a/(1-a)),fun(i),type='l')}
 ```
@@
 plot(NA,xlim=c(-5,1),ylim=c(-5,1),type='n',xlab="log-odds")
-abline(v=-log(22))
-for(i in seq_len(20)){points(log(a/(1-a)),fun(i),type='l')}
+for(i in seq_len(20)){points(log(22*a/(1-a)),fun(i),type='l')}
 
@@
 ```{r}
+n <- 82 # 83 competitors altogether
 x <- as.matrix(read.table("olympiad.txt",header=TRUE))
@@
 }
-plot(NA,xlim=c(-5,1),ylim=c(-5,1),type='n',xlab="log-odds")
-abline(v=-log(99))
-for(i in seq_len(40)){points(log(a/(1-a)),fun(i),type='l')}
+
+plot(NA,xlim=c(-5,5),ylim=c(-5,1),type='n',xlab="log-odds")
+for(i in seq_len(40)){points(log(n*a/(1-a)),fun(i),type='l')}
+abline(v=0)
+```
 
 ```{r}
-x <- as.matrix(read.table("ol.txt",header=TRUE))[,1:4]
+x <- as.matrix(read.table("olympiad.txt",header=TRUE))[,1:4]
 x <- x[apply(x,1,max) <= 100,]
@@
 }
-plot(NA,xlim=c(-5,1),ylim=c(-5,1),type='n',xlab="log-odds")
-abline(v=-log(99))
-for(i in seq_len(40)){points(log(a/(1-a)),fun(i),type='l')}
+plot(NA,xlim=c(-5,5),ylim=c(-5,1),type='n',xlab="log-odds")
+for(i in seq_len(40)){points(log(n*a/(1-a)),fun(i),type='l')}
 ```
@@
 ```{r}
-x <- as.matrix(read.table("ol.txt",header=TRUE))[,1:4]
+x <- as.matrix(read.table("olympiad.txt",header=TRUE))[,1:4]
 x <- x[apply(x,1,max) <= 100,]
 x[] <- sample(c(x))
-x
+head(x)
 rowSums(x)
@@
 }
-plot(NA,xlim=c(-5,1),ylim=c(-5,1),type='n',xlab="log-odds")
-abline(v=-log(99))
-for(i in seq_len(40)){points(log(a/(1-a)),fun(i),type='l')}
+plot(NA,xlim=c(-5,5),ylim=c(-5,1),type='n',xlab="log-odds")
+for(i in seq_len(40)){points(log(n*a/(1-a)),fun(i),type='l')}
+```
+
+In principle we can find the optimum point with calculus and indeed we
+have
+
+\[
+\frac{d}{db}\mathcal{L} = \frac{\psi(1-1/b)-\psi(r+1-1/b)}{b(b+\psi(1-1/b)-\psi(r+1-1/b))}-n
+\]
+
+but the implementation is difficult because the expression is not
+defined for many sensible values of $b$.  We can use `optimize()`
+though:
+
+```{r solvescore}
+  MLE <- function(r,n){
+  small <- 1e-6
+  a <- optimize(function(b){
+       f_single_direct(b,r=r,n=n,log=TRUE)},
+       maximum=TRUE, c(small,1-small))$maximum
+  b <- (1-a)/n
+  print(c(a,b))
+  return(a/(a+b))
+}
+
+MLE(0,10)
+MLE(1,10)
+MLE(2,10)
+MLE(3,10)
+MLE(4,10)
+MLE(5,10)
+MLE(6,10)
+
+MLE(0,11)
+MLE(1,11)
+MLE(2,11)
+MLE(3,11)
+MLE(4,11)
+MLE(5,11)
+MLE(6,11)
+
+```

### 2023-11-26 inst/very_simplified_likelihood.Rmd
consistentification with jss_hyper3

@@
 
-Consider a race between $n+1$ competitors of strength $a,b,\ldots,b$
-where we require the total strength to be unity: $a+nb=1$.  Suppose
-our bro comes first:
+Consider a race between a competitor of Bradley-Terry strength $a$ and
+$n$ competitors each of strength $b$; we require $a+b=1$.  An
+observation is indexed by $r$, the number of $b$'s finishing ahead of
+$a$, so $0\leqslant r\leqslant n$.
+
+### Likelihood functions
+
+Suppose our bro comes first (zero $b$'s ahead of him):
 
@@
 
-Comes second:
+Comes second, one $b$ ahead of him:
 
 \[
-\mathcal{L}(a)=b\cdot\frac{a}{1-b}
+\mathcal{L}(a)=b\cdot\frac{a}{1-b}=b
 \]
 
-Third:
+Third, two ahead of him:
 
 \[
-\mathcal{L}(a)=b\cdot\frac{b}{1-b}\cdot\frac{a}{1-2b}
+\mathcal{L}(a)=\frac{b}{1}\cdot\frac{b}{1-b}\cdot\frac{a}{1-2b}=\frac{b^2}{1-2b}
 \]
 
-Fourth:
+Fourth, three ahead:
 
 \[
-\mathcal{L}(a)=b\cdot\frac{b}{1-b}\cdot\frac{b}{1-2b}\cdot\frac{a}{1-3b}
+\mathcal{L}(a)=b\cdot\frac{b}{1-b}\cdot\frac{b}{1-2b}\cdot\frac{a}{1-3b}=\frac{b^3}{(1-2b)(1-3b)}
 \]
 
-Placing after $r$ of the $b$ clones finish:
+Fifth, four ahead:
 
 \[
-\mathcal{L}(a)=\frac{b^ra}{(1-b)(1-2b)\cdots(1-rb)}=(1-nb)\frac{(-1)^r\Gamma(1-1/b)}{\Gamma(1+r-1/b)}
+\mathcal{L}(a)=b\cdot\frac{b}{1-b}\cdot\frac{b}{1-2b}\cdot\frac{b}{1-3b}\cdot\frac{a}{1-4b}=\frac{b^4}{(1-2b)(1-3b)(1-4b)}
 \]
+
+Placing after $r$ of the $n$ clones of strength $b$ finish:
+
+\[
+\mathcal{L}(b)=\frac{b^ra}{(1-b)(1-2b)\cdots(1-rb)}=\frac{b^r}{(1-2b)\cdots(1-rb)},\qquad r\geqslant 2
+\]
+
+\[
+\mathcal{L}(b)=\frac{b^{r+1}}{(1-2b)\cdots(1-rb)}=\frac{(-1)^{r+1}b\Gamma(2-1/b)}{\Gamma(1+r-1/b)}
+\]
+
+The case $n=2$ was considered by Hankin 2023.
 
@@
   MLE <- function(r,n){
+  stopifnot(length(r)==1)
   small <- 1e-6
@@
   b <- (1-a)/n
-  print(c(a,b))
-  return(a/(a+b))
+#  print(c(a,b))
+  return(c(a/(a+b),b/(a+b)))
 }
+```
 
-MLE(0,10)
-MLE(1,10)
-MLE(2,10)
-MLE(3,10)
-MLE(4,10)
-MLE(5,10)
-MLE(6,10)
+```{r rbindmle}
+o <-
+rbind(
+r0=MLE(0,9),
+r1=MLE(1,9),
+r2=MLE(2,9),
+r3=MLE(3,9),
+r4=MLE(4,9),
+r5=MLE(5,9),
+r6=MLE(6,9),
+r7=MLE(7,9),
+r8=MLE(8,9),
+r9=MLE(9,9))
+o
+matplot(o)
+```
 
-MLE(0,11)
-MLE(1,11)
-MLE(2,11)
-MLE(3,11)
-MLE(4,11)
-MLE(5,11)
-MLE(6,11)
+## Compare New Zealand with Australia
 
+```{r nzaus}
+x <- as.matrix(read.table("olympiad.txt",header=TRUE))
+x <- x[c("AUS","NZL"),]
+x
 ```

### 2023-11-28 inst/very_simplified_likelihood.Rmd
more work

--- 2023-11-23 inst/very_simplified_likelihood.Rmd
@@
 
-Consider a race between $n+1$ competitors of strength $a,b,\ldots,b$
-where we require the total strength to be unity: $a+nb=1$.  Suppose
-our bro comes first:
+Consider a race between a competitor of Bradley-Terry strength $a$ and
+$n$ competitors each of strength $b$; we require $a+b=1$.  An
+observation is indexed by $r$, the number of $b$ clones finishing
+ahead of $a$, so $0\leqslant r\leqslant n$.  The initial field
+strength is $a+nb=n+(n-1)a$.
 
-\[\mathcal{L}(a)=a\]
+### Likelihood functions
 
-Comes second:
+Suppose our bro comes first (zero $b$'s ahead of him):
+
+\[\mathcal{L}(a)=\frac{a}{a+nb}=\frac{a}{n+(n-1)a}\]
+
+Comes second, one $b$ ahead of him:
 
 \[
-\mathcal{L}(a)=b\cdot\frac{a}{1-b}
+\mathcal{L}(a)=\frac{b}{a+nb}\cdot\frac{a}{a+(n-1)b}
 \]
 
-Third:
+Third, two ahead of him:
 
 \[
-\mathcal{L}(a)=b\cdot\frac{b}{1-b}\cdot\frac{a}{1-2b}
+\mathcal{L}(a)=
+\frac{b}{a+ n   b}\cdot
+\frac{b}{a+(n-1)b}\cdot
+\frac{a}{a+(n-2)b}
 \]
 
-Fourth:
+Fourth, three ahead:
 
 \[
-\mathcal{L}(a)=b\cdot\frac{b}{1-b}\cdot\frac{b}{1-2b}\cdot\frac{a}{1-3b}
+\mathcal{L}(a)=
+\frac{b}{a+ n   b}\cdot
+\frac{b}{a+(n-1)b}\cdot
+\frac{b}{a+(n-2)b}\cdot
+\frac{a}{a+(n-3)b}
 \]
 
-Placing after $r$ of the $b$ clones finish:
+Placing after $r$ of the $n$ clones of strength $b$ finish:
 
 \[
-\mathcal{L}(a)=\frac{b^ra}{(1-b)(1-2b)\cdots(1-rb)}=(1-nb)\frac{(-1)^r\Gamma(1-1/b)}{\Gamma(1+r-1/b)}
+\mathcal{L}(a)=
+\frac{b^ra}{
+(a+nb)(a+(n-1)b)(a+(n-2)b)\cdots(a+(n-r)b)}
+=
+\frac{B-1}{(B+n-1)(B+n-2)\cdots(B+n-r-1)}
 \]
 
-```{r tryit}
-small <- 1e-3
+where $B=1/b$.  In R:
 
-f_single_direct <- function(a,r,n,log=FALSE){  # not vectorised in anything
-  b <- (1-a)/n   # 0 <= b <= 1/n
+```{r}
+f_single <- function(a,r,n,log=FALSE){  # not vectorised
+  B <- 1/(1-a)
   if(log){
-      return(log(a) + r*log(b) - sum(log(1-seq_len(r)*b)))
+     out <- log(B-1) - sum(log(B + B+(n-r-1):(n-1)))
    } else {
-      return(a * b^r/prod(1-seq_len(r)*b))
-     }
-}
-
-`f_vectorised_direct` <- function(a,r,n,log=FALSE){  # vectorised in 'a', not in 'r'
-  b <- (1-a)/n   # 0 <= b <= 1/n
-  out <- r*log(b) + log(a) -rowSums(log(1-outer(b,seq_len(r))))
-  if(log){return(out)} else {return(exp(out))}
-}
-
-f <- function(a,r,n,log=FALSE){
-out <- a*0
-if(log){
-    for(rr in r){  out <- out + f_vectorised_direct(a,rr,n,log=TRUE)}
-  } else {
-    out[] <- 1
-    for(rr in r){  out <- out * f_vectorised_direct(a,rr,n,log=FALSE)}
+     out <- (B-1)/prod(B+(n-r-1):(n-1))
    }
@@
 
-f_gamma <- function(a,r,n,log=FALSE){ # vectorised in 'a', but screws up for b=1/integer
-  b <- (1-a)/n   # 0 <= b <= 1/n
-  out <- log(a) + lgamma(1-1/b) - lgamma(1+r-1/b)
-  if(!log){out <- exp(out)}
+f_vec_a <- function(a,r,n, log=FALSE){  # vectorised in 'a' but not in 'r'
+ sapply(a,function(a){f_single(a,r=r,n=n,log=log)})
+}
+
+f <- function(a,r,n,log=FALSE){ # vectorized in 'a', treats 'r' as a vector of independent observations
+
+  if(log){
+    out <- 0
+    for(rr in r){out <- out + f_vec_a(a,rr,n,log=TRUE)}
+  } else {
+    out <- 1
+    for(rr in r){out <- out * f_vec_a(a,rr,n,log=FALSE)}
+  }
   return(out)
 }
+
+MLE <- function(r,n){
+    d <- 1e-6
+   optimize(f,c(d,1-d),r=r,n=n,log=TRUE,maximum=TRUE)
+}
+MLE(3,9)
 ```
 
-Now use it:
+Now verification
 
 ```{r}
- a <- seq(from=0.05,to=0.45,by=0.01)
- L <- f(a,r=c(3,4,3,3,1,3,3,3,1,1,1,5),10,log=T)  # 10+1==11 competitors in total
-plot(a,L-max(L))
-abline(v=a[which.max(L)])
-abline(v=1/11)
-abline(h=c(0,-2))
+a <- 0.23423434
+b <- 1-a
+B <- 1/b
+n <- 9
+r <- 3
+c(
+b^3*a/((a+9*b) * (a+8*b) * (a+7*b) * (a+6*b)),
+(B-1)/prod(B+(5:8)),
+f(a,r,n,log=FALSE))
 ```
 
-Now skating:
+Now some visuals.  Suppose $r=4,n=8$:
 
-```{r showhughes}
-skating_table
-hughes <- unlist(skating_table[1,])
-a <- seq(from=0.05,to=0.55,by=0.01)
-L <- f(a,hughes,23,log=TRUE)
-L <- L - max(L)
-plot(a,L,type='b')
-abline(v=1/23)
+```{r}
+a <- seq(from=0,to=1,by=0.01)
+plot(a,f(a,3,8,log=FALSE))
 ```
 
-```{r showall}
-a <- seq(from=0.001,to=0.55,by=0.001)
-fun <- function(i){
-  out <- f(a,unlist(skating_table[i,]),23,log=TRUE)
-  return(out-max(out))
-}
+# Formula 1.
 
-plot(NA,xlim=c(0,1),ylim=c(-10,0),type='n')
-abline(v=1/23)
-for(i in seq_len(9)){points(a,fun(i),type='b')}
-
-plot(NA,xlim=c(-5,1),ylim=c(-5,0),type='n',xlab="log-odds")
-abline(v=-log(22))
-for(i in seq_len(23)){points(log(a/(1-a)),fun(i),type='l')}
+```{r maxvscheco}
+a <- read.table("formula1_2023.txt",header=TRUE)
+a <- a[,seq_len(ncol(a)-1)]
+max <- as.numeric(as.matrix(a)[1,])
+checo <- as.numeric(as.matrix(a)[2,])
+checo[is.na(checo)] <- 22
+max
+checo
+MLE(max,22)
+MLE(checo,22)
 ```
-
-# Formula 1
-
-```{r showformula1}
-x <- read.table("formula1_2022.txt",header=TRUE)
-x <- x[,-ncol(x)]
-x
-```
-
-```{r analyseform1}
-fun <- function(i){
-  suppressWarnings(jj <- as.numeric(unlist(x[i,])))
-  jj[is.na(jj)] <- 23
-  L <- f(a,jj,23,log=TRUE)
-  return(L - max(L))
-}
-
-plot(NA,xlim=c(-5,1),ylim=c(-5,1),type='n',xlab="log-odds")
-abline(v=-log(22))
-for(i in seq_len(20)){points(log(a/(1-a)),fun(i),type='l')}
-
-```
-
-# Mathematics Olympiad
-
-```{r}
-x <- as.matrix(read.table("olympiad.txt",header=TRUE))
-x <- x[apply(x,1,max) <= 100,]
-fun <- function(i){
- jj <- as.numeric(unlist(x[i,]))
-  L <- f(a,jj,100,log=TRUE)
-  return(L - max(L))
-}
-plot(NA,xlim=c(-5,1),ylim=c(-5,1),type='n',xlab="log-odds")
-abline(v=-log(99))
-for(i in seq_len(40)){points(log(a/(1-a)),fun(i),type='l')}
-
-```{r}
-x <- as.matrix(read.table("ol.txt",header=TRUE))[,1:4]
-x <- x[apply(x,1,max) <= 100,]
-fun <- function(i){
- jj <- as.numeric(unlist(x[i,]))
-  L <- f(a,jj,100,log=TRUE)
-  return(L - max(L))
-}
-plot(NA,xlim=c(-5,1),ylim=c(-5,1),type='n',xlab="log-odds")
-abline(v=-log(99))
-for(i in seq_len(40)){points(log(a/(1-a)),fun(i),type='l')}
-```
-
-```{r}
-x <- as.matrix(read.table("ol.txt",header=TRUE))[,1:4]
-x <- x[apply(x,1,max) <= 100,]
-x[] <- sample(c(x))
-x
-rowSums(x)
-fun <- function(i){
- jj <- as.numeric(unlist(x[i,]))
-  L <- f(a,jj,100,log=TRUE)
-  return(L - max(L))
-}
-plot(NA,xlim=c(-5,1),ylim=c(-5,1),type='n',xlab="log-odds")
-abline(v=-log(99))
-for(i in seq_len(40)){points(log(a/(1-a)),fun(i),type='l')}

### 2023-11-28 inst/very_simplified_likelihood.Rmd
bespoke statistical test

@@
 a <- a[,seq_len(ncol(a)-1)]
-max <- as.numeric(as.matrix(a)[1,])
-checo <- as.numeric(as.matrix(a)[2,])
-checo[is.na(checo)] <- 22
-max
-checo
-MLE(max,22)
-MLE(checo,22)
+d1 <- as.numeric(as.matrix(a)["Ocon",])
+d2 <- as.numeric(as.matrix(a)["Gasly",])
+d1[is.na(d1)] <- 22
+d2[is.na(d2)] <- 22
+d1
+d2
+MLE(d1,22)
+MLE(d2,22)
 ```
+
+two sample test:
+
+```{r 2samptest}
+both <- c(d1,d2)
+bothmax <- MLE(both,22)$maximum
+
+d1max <- MLE(d1,22)$maximum
+d2max <- MLE(d2,22)$maximum
+f(bothmax,d1,22,log=TRUE)
+f(bothmax,d2,22,log=TRUE)
+f(d1max,d1,22,log=TRUE)
+f(d2max,d2,22,log=TRUE)
+
+Lambda <- (f(d2max,d1,22,log=TRUE)+f(d2max,d2,22,log=TRUE)) -(f(bothmax,d1,22,log=TRUE)+f(bothmax,d2,22,log=TRUE))
+Lambda
+pchisq(-2*Lambda,df=1,lower.tail=FALSE)
+```
+
+## Mathematical Olympiad
+
+```{r readoly}
+sametest <- function(d1,d2,n){
+  both <- c(d1,d2)
+  bothmax <- MLE(both,n)$maximum
+
+  d1max <- MLE(d1,n)$maximum
+  d2max <- MLE(d2,n)$maximum
+  Lambda <- (f(d2max,d1,n,log=TRUE)+f(d2max,d2,n,log=TRUE)) -(f(bothmax,d1,n,log=TRUE)+f(bothmax,d2,n,log=TRUE))
+  pchisq(-2*Lambda,df=1,lower.tail=FALSE)
+}
+
+a <- as.matrix(read.table("olympiad.txt"))
+nrow(a)
+d1 <- a["AUS",]
+d2 <- a["NZL",]
+d1
+d2
+sametest(d1,d2,100)
+a["USA",]
+a["CHN",]
+sametest(a["USA",],a["CHN",],100)
+sametest(a["USA",],a["CHN",],5)
+```

### 2023-12-01 inst/very_simplified_likelihood.Rmd
more work on probs

@@
 strength is $a+nb=n+(n-1)a$.
+
+Probability of the focal competitor coming first:
+
+\[
+\frac{a}{a+nb}=\frac{a}{n - a(n-1)a}=\frac{1-b}{1+(n-1)b}
+\]
+
+Coming second:
+\[
+\frac{b}{a+nb}\cdot\frac{a}{a+(n-1)b}
+\]
+
+Third:
+
+\[
+\frac{b}{a+nb}\cdot\frac{b}{a+(n-1)b}\cdot\frac{a}{a+(n-2)b}
+\]
 
@@
 
-Placing after $r$ of the $n$ clones of strength $b$ finish:
+Placing after $r<n$ of the $n$ clones of strength $b$ finish:
 
@@
 
-where $B=1/b$.  In R:
+where $B=1/b$.
+
+## Placing last
+
+Placing last is finishing  after $n$ clones finish:
+
+\[
+n=1\longrightarrow\mathcal{L}=b\]
+
+\[
+n=2\longrightarrow\mathcal{L}=\frac{b}{a+2b}\cdot\frac{b}{a+b}=\frac{b^2}{1+b}
+\]
+
+\[
+n=3\longrightarrow\mathcal{L}=\frac{b}{a+3b}\cdot\frac{b}{a+2b}\cdot\frac{b}{a+b}=\frac{b^3}{(1+b)(1+2b)}
+\]
+
+\[
+n=4\longrightarrow\mathcal{L}=\frac{b}{a+4b}\cdot\frac{b}{a+3b}\cdot\frac{b}{a+2b}\cdot\frac{b}{a+b}=
+\frac{b^4}{(1+b)(1+2b)(1+3b)}
+\]
+
+For $n=4$ we have equivalently
+
+\[
+\frac{1}{B(B+1)(B+2)(B+3)}=\frac{(B-1)!}{(B+3)!}=\frac{\Gamma(B)}{\Gamma(B+4)}
+\]
+
+\[
+\frac{B-1}{(B+n-1)(B+n-2)\cdots(B+1)\cdot B\cdot(B-1)}
+\]
+
+(denominator has $n+1$ terms).  In R:
 
@@
   if(log){
+     if(r==n){
+       out <- lgamma(B)-lgamma(B+n)
+     } else {
        out <- log(B-1) - sum(log(B + B+(n-r-1):(n-1)))
+     }
+   } else {  # not-log 
+     if(r==n){
+      out <- gamma(B)/gamma(B+n)
      } else {
@@
      }
+   }	
    return(out)
-}
+}x
 

### 2024-01-05 inst/very_simplified_likelihood.Rmd
distinguish prob from likelihood

@@
 Consider a race between a competitor of Bradley-Terry strength $a$ and
-$n$ competitors each of strength $b$; we require $a+b=1$.  An
-observation is indexed by $r$, the number of $b$ clones finishing
+$n$ competitors each of strength $b$; we require $a+b=1$.  NB we have
+$n+1$ competitors in total, 1 of strength $a$ and $n$ of strength $b$.
+An observation is indexed by $r$, the number of $b$ clones finishing
 ahead of $a$, so $0\leqslant r\leqslant n$.  The initial field
@@
 
-where $B=1/b$.
+where $B=1/b$ [also using the fact that $a+b=1$].
 
@@
 
-Placing last is finishing  after $n$ clones finish:
+Placing last is finishing  after $n=r$ clones finish:
 
@@
 \[
-n=3\longrightarrow\mathcal{L}=\frac{b}{a+3b}\cdot\frac{b}{a+2b}\cdot\frac{b}{a+b}=\frac{b^3}{(1+b)(1+2b)}
+n=3\longrightarrow\mathcal{L}=\frac{b}{a+3b}\cdot\frac{b}{a+2b}\cdot\frac{b}{a+b}=\frac{b^3}{(1+2b)(1+b)}
 \]
@@
 n=4\longrightarrow\mathcal{L}=\frac{b}{a+4b}\cdot\frac{b}{a+3b}\cdot\frac{b}{a+2b}\cdot\frac{b}{a+b}=
-\frac{b^4}{(1+b)(1+2b)(1+3b)}
+\frac{b^4}{(1+3b)(1+2b)(1+b)}
 \]
@@
 \[
-\frac{1}{B(B+1)(B+2)(B+3)}=\frac{(B-1)!}{(B+3)!}=\frac{\Gamma(B)}{\Gamma(B+4)}
+\frac{1}{(B+3)(B+2)(B+1)B}=\frac{(B-1)!}{(B+3)!}=\frac{\Gamma(B)}{\Gamma(B+4)}
 \]
 
+In general we would have
+
 \[
-\frac{B-1}{(B+n-1)(B+n-2)\cdots(B+1)\cdot B\cdot(B-1)}
+\frac{1}{(B+n)(B+n-1)\cdots(B+1)B}=\frac{(B-1)!}{(B+n)!}=\frac{\Gamma(B)}{\Gamma(B+n+1)}
 \]
@@
   if(log){
-     if(r==n){
+     if(r==n){  # last place
        out <- lgamma(B)-lgamma(B+n)
-     } else {
+     } else {   # not last
        out <- log(B-1) - sum(log(B + B+(n-r-1):(n-1)))
@@
    return(out)
-}x
+}
 
@@
 
-Now some visuals.  Suppose $r=4,n=8$:
+## Note on the differences between probability and likelihood
+
+The likelihood calculations above can be confusing becuase likelihood
+is not the same as probability.
+
+Consider, for example, $n=2$ so we have 3 competitors.
+
+probability of focal competitor coming first [i.e. $r=0$]:
+
+\[\frac{a}{a+2b}=\frac{1-b}{1+b}\]
+
+Now what is the probability of coming second, that is, $r=1$?
+
+My first thought was:
+
+\[
+\mbox{prob of coming second}   =
+\frac{b}{a+2b}\cdot\frac{a}{a+b}=\frac{b(1-b)}{1+b}\qquad\mbox{WRONG}
+\]
+
+but this is incorrect: there are two clones of (strength) $b$, and the
+likelihood arguments above do not distinguish between their finishing
+order.  We can name the competitors $a$, $b_1$ and $b_2$ and specify
+that $b_1$ and $b_2$ both have strength $b$.  Then "coming second"
+means that the finishing order was $b_1,a,b_2$ or $b_2,a,b_1$.  Noting
+that these events are disjoint, the probability would be
+
+\[\mbox{prob of coming second}   =
+\operatorname{P}(b_1,a,b_2)      +
+\operatorname{P}(b_2,a,b_1)      =
+\frac{b}{a+2b}\cdot\frac{a}{a+b} +
+\frac{b}{a+2b}\cdot\frac{a}{a+b} = \frac{2(1-b)}{1+b}
+\]
+
+Similarly for the focal competitor coming last we need to sum the
+probabilities of finishing orders $b_1,b_2,a$ and $b_2,b_1,a$:
+
+\[\mbox{prob of coming third}=
+\operatorname{P}(b_1,b_2,a) +
+\operatorname{P}(b_2,b_1,a) =
+\frac{b}{a+2b}\cdot\frac{b}{a+b}+
+\frac{b}{a+2b}\cdot\frac{b}{a+b}=
+\frac{2b^2}{1+b}.
+\]
+
+Above we see a factor of two difference between the probabilities and
+the likelihoods presented previously.  Note that there is no such
+factor for the probability of the focal competitor finishing first:
+
+\[
+\mbox{prob of coming first}      =
+\operatorname{P}(a,b_1,b_2)      +
+\operatorname{P}(a,b_2,b_1)      =
+\frac{b}{a+2b}\cdot\frac{b}{b+b} +
+\frac{b}{a+2b}\cdot\frac{b}{b+b} = \frac{1-b}{1+b}
+\]
+
+Now we see that the probabilities sum to
+
+\[
+\frac{1-b}{1+b} + \frac{2b^2}{1+b} + \frac{2(1-b)}{1+b}=1
+\]
+
+Now try $r=3$, 4 competitors.   Taking things one step at a time:
+
+\[
+\mbox{prob of coming first}      =
+3!\cdot\frac{a}{a+3b}\cdot\frac{b}{3b}\cdot\frac{b}{2b}\cdot\frac{b}{b}
+\]
+
+## Visuals
+
+Now some likelihood visuals.  Suppose $r=4,n=8$:
 

### 2024-01-10 inst/hyper3.Rmd
consistentification
typos

@@
 
-To cite the `hyper2` package in publications, please use @hankin2017a.
-Objects of class \code{hyper3} are a generalization of \code{hyper2}
+To cite the `hyper2` package in publications, please use @hankin2017.
+Objects of class `hyper3` are a generalization of `hyper2`
 objects that allow the brackets to contain weighted probabilities.
-Likelihood functions are defined on non-negative $p_1,\ldots, p_n$
+Likelihood functions are defined on non-negative numbers $p_1,\ldots, p_n$
 subject to the unit-sum constraint $\sum p_i=1$.  Given known weights
@@
 
-$$\mathcal{L}(p_1,p_2,\lambda;A,B,C,D)=
-\left(\frac{\lambda p_1}{\lambda p_1 + p_2}\right)^{A}
-\left(\frac{p_2        }{\lambda p_1 + p_2}\right)^{B}
-\left(\frac{p_1        }{p_1 + \lambda p_2}\right)^{C}
-\left(\frac{\lambda p_2}{p_1 + \lambda p_2}\right)^{D}.
+$$\mathcal{L}(p_1,p_2,\lambda;a,b,c,d)=
+\left(\frac{\lambda p_1}{\lambda p_1 + p_2}\right)^{a}
+\left(\frac{p_2        }{\lambda p_1 + p_2}\right)^{b}
+\left(\frac{p_1        }{p_1 + \lambda p_2}\right)^{c}
+\left(\frac{\lambda p_2}{p_1 + \lambda p_2}\right)^{d}.
 $$
@@
 $$
-\hat{p_1}=\frac{\sqrt{AC}}{\sqrt{AC}+\sqrt{BD}}\qquad
-\hat{p_2}=\frac{\sqrt{BD}}{\sqrt{AC}+\sqrt{BD}}\qquad
-\hat{\lambda}=\sqrt{\frac{AD}{BC}}
+\hat{p_1}=\frac{\sqrt{ac}}{\sqrt{ac}+\sqrt{bd}}\qquad
+\hat{p_2}=\frac{\sqrt{bd}}{\sqrt{ac}+\sqrt{bd}}\qquad
+\hat{\lambda}=\sqrt{\frac{ad}{bc}}
 $$
+Likelihoods of this form are easily specified using package idiom.
 
-If $A=x,B=x,C=x,D=x$ and $\lambda=1.3$ [assumed for the moment to be
+If $a=6,b=2,c=7,d=1$ and $\lambda=1.3$ [assumed for the moment to be
 known], appropriate package idiom might be:
@@
 library("hyper2",quietly=TRUE)
-A <- 6 ; B <- 2 ; C <- 7 ; D <- 1
+A <- 6 ; B <- 2 ; C <- 7 ; D <- 1 # upper-case to avoid conflicts
 H <- hyper3()
@@
 
-Above, obect `M` has real parts being home wins and imaginary parts
-being away wins.
-
-We may estimate $p_1$ and $p_2$ using maximum likelihood, `maxp()` in
-package idiom:
+Above, object `M` is a $2\times 2$ complex matrix, with real parts being home
+wins and imaginary parts being away wins.  See how `home_away3(M)` returns the
+same likelihood function as the explicit calculation performed by hand
+previously.  Now, we can calculate maximum likelihood estimates
+$\hat{p_1},\hat{p_2}$ of $p_1$ and $p_2$ using numerical optimization
+techniques, `maxp()` in package idiom:
 
@@
 nonnegative strengths $p_\mathrm{M},p_\mathrm{RB},p_\mathrm{F}$, with
-unit sum.  These players might be conceputalised as F1 constructors
+unit sum.  These players might be conceptualised as F1 constructors
 (here Mercedes (M), Red Bull (RB), and Ferrari (F) for the sake of

### 2024-01-19 R/hyper2.R
fixes #123

@@ -850,9 +850,9 @@ rorder_single <- function(p){
         powers <- alpha-1
     }
 
-    if(is.null(names(powers))){stop("supply a named vector")}
-    
-    hyper2(as.list(names(powers)),d=powers)
+    np <- names(powers)
+    if(is.null(np)){stop("supply a named vector")}
+    hyper2(as.list(np),d=powers) - hyper2(list(np),d=sum(powers))
 }
 
 `GD` <- function(alpha, beta, beta0=0){
@@ -1110,7 +1110,24 @@ rorder_single <- function(p){
   ordertable_to_ranktable(xx)
 }
 
-`is.dirichlet` <- function(H){all(lapply(brackets(H),length)==1)}
+`is.dirichlet` <- function(H){
+    len <- unlist(elements(lapply(brackets(H), length)))
+    pow <- elements(powers(H))
+    bra <- elements(brackets(H))
+    if (sum(len > 1) > 1) {
+        print("brackets must all be of length 1 [the numerators] except one [the denominator]")
+        return(FALSE)
+    }
+    denominator_bracket <- unlist(bra[len > 1])
+    numerator_brackets <- unlist(bra[len == 1])
+    if (all(numerator_brackets %in% denominator_bracket)) {
+        return(TRUE)
+    }
+    else {
+        print("Dirichlet distribution requires denominator to be all competitors")
+        return(FALSE)
+    }
+}
 
 `rdirichlet` <- function(n, H){
     if(is.hyper2(H)){

### 2024-01-21 inst/handover.Rmd
consistentification
fixes #218

@@
 
-To cite the `hyper2` package in publications, please use @hankin2017a.
+To cite the `hyper2` package in publications, please use @hankin2017.
 This short document is based on @altham2010, which
@@
 handover[c("t10","t00")] %<>% inc(handover_table[3,2])  # yes-no  plus  no-no  = 4 
+handover <- balance(handover)
 handover

### 2024-01-21 R/hyper2.R

@@ -157,6 +157,7 @@ setGeneric("pnames<-",function(x,value){standardGeneric("pnames<-")})
 }
 
 `loglik_single` <- function(p,H,log=TRUE){
+  stopifnot(sum(powers(H))==0)
   stopifnot(all(p>=0))
   if(length(p) == size(H)-1){
     stopifnot(sum(p)<=1)
@@ -1311,3 +1312,7 @@ rorder_single <- function(p){
     return(H)
 }
 
+`balance` <- function(H){
+    H[pnames(H)] <- -sum(powers(H))
+    return(H)
+}

### 2024-01-21 inst/very_simplified_likelihood.Rmd
further work

@@
 
-Probability of the focal competitor coming first:
+Suppose our bro comes first (zero $b$'s ahead of him):
 
 \[
-\frac{a}{a+nb}=\frac{a}{n - a(n-1)a}=\frac{1-b}{1+(n-1)b}
+\mathcal{L}(a)=\frac{a}{a+nb}=\frac{1-b}{1+(n-1)b}
 \]
-
-Coming second:
-\[
-\frac{b}{a+nb}\cdot\frac{a}{a+(n-1)b}
-\]
-
-Third:
-
-\[
-\frac{b}{a+nb}\cdot\frac{b}{a+(n-1)b}\cdot\frac{a}{a+(n-2)b}
-\]
-
-### Likelihood functions
-
-Suppose our bro comes first (zero $b$'s ahead of him):
-
-\[\mathcal{L}(a)=\frac{a}{a+nb}=\frac{a}{n+(n-1)a}\]
 
@@
 n=2\longrightarrow\mathcal{L}=\frac{b}{a+2b}\cdot\frac{b}{a+b}=\frac{b^2}{1+b}
+=\frac{1}{B(B+1)}
 \]
@@
 \[
-n=3\longrightarrow\mathcal{L}=\frac{b}{a+3b}\cdot\frac{b}{a+2b}\cdot\frac{b}{a+b}=\frac{b^3}{(1+2b)(1+b)}
+n=3\longrightarrow\mathcal{L}=\frac{b}{a+3b}\cdot\frac{b}{a+2b}\cdot\frac{b}{a+b}=\frac{b^3}{(1+2b)(1+b)}=\frac{1}{B(B+1)(B+2)}
 \]
@@
 n=4\longrightarrow\mathcal{L}=\frac{b}{a+4b}\cdot\frac{b}{a+3b}\cdot\frac{b}{a+2b}\cdot\frac{b}{a+b}=
-\frac{b^4}{(1+3b)(1+2b)(1+b)}
+\frac{b^4}{(1+3b)(1+2b)(1+b)}=
+\frac{1}{B(B+1)(B+2)(B+3)}
 \]
@@
 \[
-\frac{1}{(B+n)(B+n-1)\cdots(B+1)B}=\frac{(B-1)!}{(B+n)!}=\frac{\Gamma(B)}{\Gamma(B+n+1)}
+\frac{1}{(B+n-1)(B+n-2)\cdots(B+1)B}=\frac{(B-1)!}{(B+n)!}=\frac{\Gamma(B)}{\Gamma(B+n+1)}
 \]
@@
 
-Now try $r=3$, 4 competitors.   Taking things one step at a time:
+Now try $r=3$ [i.e. 4 competitors].  Taking things one step at a time:
 
@@
 \mbox{prob of coming first} =
-3!\cdot\frac{a}{a+3b}\cdot\frac{b}{3b}\cdot\frac{b}{2b}\cdot\frac{b}{b}
+3!\cdot\frac{a}{a+3b}\cdot\frac{b}{3b}\cdot\frac{b}{2b}\cdot\frac{b}{b}=
+\frac{1-b}{1+2b}
+\]
+
+\[
+\mbox{prob of coming second} =
+3!\cdot\frac{b}{a+3b}\cdot\frac{a}{a+2b}\cdot\frac{b}{2b}\cdot\frac{b}{b}=
+3\cdot\frac{b(1-b)}{(1+2b)(1+b)}
+\]
+
+\[
+\mbox{prob of coming third} =
+3!\cdot\frac{b}{a+3b}\cdot\frac{b}{a+2b}\cdot\frac{a}{a+b}\cdot\frac{b}{b}
+=
+6\cdot\frac{b^2(1-b)}{(1+2b)(1+b)}
+\]
+\[
+\mbox{prob of coming fourth (last)} =
+3!\cdot\frac{b}{a+3b}\cdot\frac{b}{a+2b}\cdot\frac{b}{a+b}\cdot\frac{a}{a}
+=
+6\cdot\frac{b^3}{(1+2b)(1+b)}
+\]
+
+The sum would be $[(1+b)(1-b) + 3b(1-b) + 6b^2(1-b) +
+6b^3]/[(1+b)(1+2b)]=1$.
+
+Now consider the general case.  Again, $n$ clones of strength $b$ and
+one of strength $a=1-b$, $n+1$ competitors.
+
+\[
+\mbox{prob of coming first} =
+n!\cdot\frac{b}{a+nb}\cdot\frac{b}{nb}\cdot\frac{b}{(n-1)b}\cdots\frac{b}{b}
+=
+\frac{b}{1+(n-1)b}
+\]
+
+\[
+\mbox{prob of coming second} =
+n!\cdot\frac{b}{a+nb}\cdot\frac{a}{a+(n-1)b}\cdot\frac{b}{(n-1)b}\cdot\frac{b}{(n-2)b}\cdots\frac{b}{2b}\cdot\frac{b}{b}
+=n\cdot X
 \]
@@
 ```
+
+## Masterchef
+
+Look at the dotchart in the _MasterChef_ part of the `hyper2`
+vignette.  We could use a Very Simplified Likelihood to estimate, say,
+Laura's BT strength assuming that the other competitors are clones.

### 2024-01-21 inst/race3.Rmd
consistentification
rewording

@@
 (takes about an hour to process without cache).  To cite the `hyper2`
-package in publications, please use @hankin2017a.  Consider a race in
+package in publications, please use @hankin2017.  Consider a race in
 which there are six runners 1-6 but we happen to know that three of
@@
 
-Above we see strong evidence to reject the hypothesis that
-$p_a=p_b=p_c=\frac{1}{3}$.  We may also test the hypothesis that the
-strength vector is equal to its true value:
+Above we see a $p$-value of $<10^{-6}$, strong evidence to reject the
+hypothesis that $p_a=p_b=p_c=\frac{1}{3}$.  We may also test the
+hypothesis that the strength vector is equal to its true value:
 

### 2024-01-26 inst/NBA.Rmd
consistentification
distinguish hankin2017 from hankin2017_rnw

@@
 output: bookdown::html_document2
-bibliography: jss_hyper3.bib
+bibliography: hyper2.bib  # copy vignettes/hyper2.bib to inst/
 ---
@@
 
-To cite the `hyper2` package in publications, please use @hankin2017a.
+To cite the `hyper2` package in publications, please use @hankin2017_rmd.
 Here I consider a single basketball match from the perspective of the

### 2024-01-26 inst/T20.Rmd
consistentification
distinguish hankin2017 from hankin2017_rnw

@@
 output: bookdown::html_document2
-bibliography: jss_hyper3.bib
+bibliography: hyper2.bib  # copy hyper2.bib from the vignettes/ directory
 ---
@@
 
-To cite the `hyper2` package in publications, please use @hankin2017a.
-This script creates and analyses hyper2 object `T20` which is a
-likelihood function for the strengths of the competitors in the Indian
-Premier League.
-Dataframe `T20_table` has one row for each T20 IPL match in the
-period 2008-2017 with the exception of three no-result matches and
-seven tied matches, which were removed.
+To cite the `hyper2` package in publications, please use
+@hankin2017_rmd.  This script creates and analyses hyper2 object `T20`
+which is a likelihood function for the strengths of the competitors in
+the Indian Premier League.  Dataframe `T20_table` has one row for each
+T20 IPL match in the period 2008-2017 with the exception of three
+no-result matches and seven tied matches, which were removed.
 
@@
 library("hyper2",quietly=TRUE)
+library("magrittr",quietly=TRUE)
 T20_table <- read.table("T20.txt",header=TRUE)

### 2024-01-26 inst/carcinoma.Rmd
distinguish hankin2017 from hankin2017_rnw

@@
 
-To cite the `hyper2` package in publications, please use @hankin2017
-(takes maybe twenty minutes to run without cache.  The value of
+To cite the `hyper2` package in publications, please use @hankin2017_rmd; 
+takes maybe twenty minutes to run without cache.  The value of
 `carcinoma_maxp` is calculated separately [discussed below], and that
-takes a few days to run).
+takes a few days to run.
 

### 2024-01-26 inst/collusion3.Rmd
consistentification
distinguish hankin2017 from hankin2017_rnw

@@
 output: bookdown::html_document2
-bibliography: jss_hyper3.bib
+bibliography: hyper2.bib  # copy from vignettes/
 ---
@@
 
-To cite the `hyper2` package in publications, please use @hankin2017a.
+To cite the `hyper2` package in publications, please use @hankin2017_rmd.
 Here I analyse results from Curacao 1962 using `hyper3` formalism

### 2024-01-26 inst/cook_repeated.Rmd
distinguish hankin2017 from hankin2017_rnw

@@
 output: bookdown::html_document2
-bibliography: hyper2.bib
+bibliography: hyper2.bib # copy from vignettes/
 ---
@@
 
-To cite the `hyper2` package in publications, please use @hankin2017;
+To cite the `hyper2` package in publications, please use @hankin2017_rmd;
 takes about five minutes without cache.  Suppose there are five people

### 2024-01-26 inst/curling.Rmd
consistentification
distinguish hankin2017 from hankin2017_rnw

@@
 output: bookdown::html_document2
-bibliography: jss_hyper3.bib
+bibliography: hyper2.bib  # copy from vignettes/
 
@@
 
-To cite the `hyper2` package in publications, please use @hankin2017a.
+To cite the `hyper2` package in publications, please use @hankin2017_rmd.
 [Curling](https://en.wikipedia.org/wiki/Curling) is a sport with

### 2024-01-26 inst/distance_matrix_idea.Rmd
consistentification
distinguish hankin2017 from hankin2017_rnw

@@
 output: bookdown::html_document2
-bibliography: jss_hyper3.bib
+bibliography: hyper2.bib # copy from vignettes/
 ---
@@
 
-To cite the `hyper2` package in publications, please use @hankin2017a.
+To cite the `hyper2` package in publications, please use @hankin2017_rmd.
 This file presents some _speculative_ experimental work which is under

### 2024-01-26 inst/eurodance.Rmd
distinguish hankin2017 from hankin2017_rnw

@@
 output: bookdown::html_document2
-bibliography: hyper2.bib
+bibliography: hyper2.bib  # copy from vignettes/
 ---
@@
 
-To cite the `hyper2` package in publications, please use @hankin2017;
+To cite the `hyper2` package in publications, please use @hankin2017_rmd;
 takes about 20 minutes to process without cache.  The file

### 2024-01-26 inst/fatigue.Rmd
consistentification
distinguish hankin2017 from hankin2017_rnw

@@
 
-To cite the `hyper2` package in publications, please use @hankin2017a.
+To cite the `hyper2` package in publications, please use @hankin2017_rmd.
 Analysis of the 1973 interzonal competition.

### 2024-01-26 inst/global_liveability_ranking.Rmd
distinguish hankin2017 from hankin2017_rnw

@@
 output: bookdown::html_document2
-bibliography: hyper2.bib
+bibliography: hyper2.bib  # copy vignettes/hyper2.bib to inst/
 ---
@@
 
-To cite the `hyper2` package in publications, please use @hankin2017;
+To cite the `hyper2` package in publications, please use @hankin2017_rmd;
 takes about an hour to process without cache.

### 2024-01-26 inst/grad_weight.Rmd
consistentification
distinguish hankin2017 from hankin2017_rnw

@@
 
-To cite the `hyper2` package in publications, please use @hankin2017a.
+To cite the `hyper2` package in publications, please use @hankin2017_rmd.
 We will start with a simple example.  Consider two players $p_1$ and

### 2024-01-26 inst/handover.Rmd
distinguish hankin2017 from hankin2017_rnw

@@
 
-To cite the `hyper2` package in publications, please use @hankin2017.
+To cite the `hyper2` package in publications, please use @hankin2017_rmd.
 This short document is based on @altham2010, which

### 2024-01-26 inst/hepatitis.Rmd
consistentification
distinguish hankin2017 from hankin2017_rnw

@@
 
-To cite the `hyper2` package in publications, please use @hankin2017a.
+To cite the `hyper2` package in publications, please use @hankin2017_rmd.
 

### 2024-01-26 inst/home_advantage.Rmd
distinguish hankin2017 from hankin2017_rnw

@@
 
-To cite the `hyper2` package in publications, please use @hankin2017
+To cite the `hyper2` package in publications, please use @hankin2017_rmd
 (takes about fifteen minutes to run without cache).  Here I analyse a

### 2024-01-26 inst/home_advantage3.Rmd
consistentification
distinguish hankin2017 from hankin2017_rnw

@@
 
-To cite the `hyper2` package in publications, please use @hankin2017a.
+To cite the `hyper2` package in publications, please use @hankin2017_rmd.
 Objects of class \code{hyper3} are a generalization of \code{hyper2}

### 2024-01-26 inst/hunter.Rmd
consistentification
distinguish hankin2017 from hankin2017_rnw

@@
 
-To cite the `hyper2` package in publications, please use @hankin2017a.
+To cite the `hyper2` package in publications, please use @hankin2017_rmd.
 Here I discuss why implementing equation 30 of [@hunter2004] for

### 2024-01-26 inst/hyper3.Rmd
distinguish hankin2017 from hankin2017_rnw

@@
 
-To cite the `hyper2` package in publications, please use @hankin2017.
+To cite the `hyper2` package in publications, please use @hankin2017_rmd.
 Objects of class `hyper3` are a generalization of `hyper2`

### 2024-01-26 inst/icons.Rmd
consistentification
distinguish hankin2017 from hankin2017_rnw

@@
 
-To cite the `hyper2` package in publications, please use @hankin2017a.
+To cite the `hyper2` package in publications, please use @hankin2017_rmd.
 This minimal document reads file `icons.txt` and creates R objects

### 2024-01-26 inst/interzonal.Rmd
consistentification
distinguish hankin2017 from hankin2017_rnw

@@
 
-To cite the `hyper2` package in publications, please use @hankin2017a.
+To cite the `hyper2` package in publications, please use @hankin2017_rmd.
 This file creates likelihood functions `H` and `H_coll`, which appear

### 2024-01-26 inst/javelin.Rmd
consistentification
distinguish hankin2017 from hankin2017_rnw

@@
 
-To cite the `hyper2` package in publications, please use @hankin2017a.
+To cite the `hyper2` package in publications, please use @hankin2017_rmd.
 Here we analyse results from the men's javelin, 2020 Summer Olympics.

### 2024-01-26 inst/jester.Rmd
consistentification
distinguish hankin2017 from hankin2017_rnw

@@
 
-To cite the `hyper2` package in publications, please use @hankin2017a.
+To cite the `hyper2` package in publications, please use @hankin2017_rmd.
 This file creates objects `jester` and `maxjest`, which are datasets

### 2024-01-26 inst/karate.Rmd
consistentification
distinguish hankin2017 from hankin2017_rnw

@@
 output: bookdown::html_document2  
-bibliography: jss_hyper3.bib
+bibliography: hyper2.bib # copy vignettes/hyper2.bib to inst/
 ---
@@
 
-To cite the `hyper2` package in publications, please use @hankin2017a.
+To cite the `hyper2` package in publications, please use @hankin2017_rmd.
 Here I analyse a [karate](https://en.wikipedia.org/wiki/Karate)

### 2024-01-26 inst/kka.Rmd
distinguish hankin2017 from hankin2017_rnw

@@
 
-To cite the `hyper2` package in publications, please use @hankin2017;
+To cite the `hyper2` package in publications, please use @hankin2017_rmd;
 this `Rmd` file takes about 45 minutes to process without cache.  Here

### 2024-01-26 inst/loser.Rmd
distinguish hankin2017 from hankin2017_rnw

@@
 
-To cite the `hyper2` package in publications, please use @hankin2017;
+To cite the `hyper2` package in publications, please use @hankin2017_rmd;
 see also `inst/notthewinner.Rmd`.  Here we consider a situation in

### 2024-01-26 inst/mann_whitney_wilcoxon.Rmd
consistentification
distinguish hankin2017 from hankin2017_rnw

@@
 
-To cite the `hyper2` package in publications, please use @hankin2017a.
+To cite the `hyper2` package in publications, please use @hankin2017_rmd.
 Here I discuss the `hyper2` generalization of the

### 2024-01-26 inst/monster_vs_lambda.Rmd
distinguish hankin2017 from hankin2017_rnw

@@
 
-To cite the `hyper2` package in publications, please use @hankin2017;
+To cite the `hyper2` package in publications, please use @hankin2017_rmd;
 takes about an hour to process without cache.  Here I consider

### 2024-01-26 inst/multiset_hyper2.Rmd
consistentification
distinguish hankin2017 from hankin2017_rnw

@@
 
-To cite the `hyper2` package in publications, please use @hankin2017a.
+To cite the `hyper2` package in publications, please use @hankin2017_rmd.
 Here we will sum probabilities across a sample space and verify that

### 2024-01-26 inst/order.Rmd
consistentification
distinguish hankin2017 from hankin2017_rnw

@@
 
-To cite the `hyper2` package in publications, please use @hankin2017a.
+To cite the `hyper2` package in publications, please use @hankin2017_rmd.
 Here I consider the likelihood function given by a single observation

### 2024-01-26 inst/ordertable_to_ranktable.Rmd
consistentification
distinguish hankin2017 from hankin2017_rnw

@@
 
-To cite the `hyper2` package in publications, please use @hankin2017a.
+To cite the `hyper2` package in publications, please use @hankin2017_rmd.
 Rank tables can be confusing.  The package includes a range of

### 2024-01-26 inst/pentathlon.Rmd
consistentification
distinguish hankin2017 from hankin2017_rnw

@@
 
-To cite the `hyper2` package in publications, please use @hankin2017a.
+To cite the `hyper2` package in publications, please use @hankin2017_rmd.
 The [modern

### 2024-01-26 inst/plackett_luce_monster.Rmd
consistentification
distinguish hankin2017 from hankin2017_rnw

@@
 
-To cite the `hyper2` package in publications, please use @hankin2017a.
+To cite the `hyper2` package in publications, please use @hankin2017_rmd.
 Here I modify standard Plackett-Luce likelihood functions to include

### 2024-01-26 inst/podium.Rmd
consistentification
distinguish hankin2017 from hankin2017_rnw

@@
 
-To cite the `hyper2` package in publications, please use @hankin2017a.
+To cite the `hyper2` package in publications, please use @hankin2017_rmd.
 This short document defines the `podium()` function which tests the

### 2024-01-26 inst/powerboat.Rmd
consistentification
distinguish hankin2017 from hankin2017_rnw

@@
 
-To cite the `hyper2` package in publications, please use @hankin2017a.
+To cite the `hyper2` package in publications, please use @hankin2017_rmd.
 [powerboat](https://en.wikipedia.org/w/index.php?title=2018_F1_Powerboat_World_Championship&oldid=920386507)

### 2024-01-26 inst/pudding.Rmd
consistentification
distinguish hankin2017 from hankin2017_rnw

@@
 
-To cite the `hyper2` package in publications, please use @hankin2017a.
+To cite the `hyper2` package in publications, please use @hankin2017_rmd.
 This short document creates the `pudding` suite of R objects and

### 2024-01-26 inst/race3.Rmd
distinguish hankin2017 from hankin2017_rnw

@@
 (takes about an hour to process without cache).  To cite the `hyper2`
-package in publications, please use @hankin2017.  Consider a race in
+package in publications, please use @hankin2017_rmd.  Consider a race in
 which there are six runners 1-6 but we happen to know that three of

### 2024-01-26 inst/retain.Rmd
consistentification
distinguish hankin2017 from hankin2017_rnw

@@
 
-To cite the `hyper2` package in publications, please use @hankin2017a.
+To cite the `hyper2` package in publications, please use @hankin2017_rmd.
 Function `keep_flawed()` can be confusing.  Originally, the package

### 2024-01-26 inst/rock_paper_scissors_monster.Rmd
distinguish hankin2017 from hankin2017_rnw

@@
 
-To cite the `hyper2` package in publications, please use @hankin2017;
+To cite the `hyper2` package in publications, please use @hankin2017_rmd;
 takes about an hour to run without cache.  Here I use the `hyper2`

### 2024-01-26 inst/rowing.Rmd
consistentification
distinguish hankin2017 from hankin2017_rnw

@@
 
-To cite the `hyper2` package in publications, please use @hankin2017a.
+To cite the `hyper2` package in publications, please use @hankin2017_rmd.
 This short document analyses the `rowing` dataset: sculls in the 2016

### 2024-01-26 inst/skating.Rmd
consistentification
distinguish hankin2017 from hankin2017_rnw

@@
 
-To cite the `hyper2` package in publications, please use @hankin2017a.
+To cite the `hyper2` package in publications, please use @hankin2017_rmd.
 This document analyses Bradley-Terry strengths of the competitors in

### 2024-01-26 inst/skeleton.Rmd
consistentification
distinguish hankin2017 from hankin2017_rnw

@@
 
-To cite the `hyper2` package in publications, please use @hankin2017a.
+To cite the `hyper2` package in publications, please use @hankin2017_rmd.
 The [skeleton](https://en.wikipedia.org/wiki/Skeleton_(sport)) is a

### 2024-01-26 inst/soling.Rmd
consistentification
distinguish hankin2017 from hankin2017_rnw

@@
 ```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
-knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+knitr::includme_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
 ```
 
-To cite the `hyper2` package in publications, please use @hankin2017a.
+To cite the `hyper2` package in publications, please use @hankin2017_rmd.
 Soling is a type of boat raced in the 2000 Summer Olympics.  Here I

### 2024-01-26 inst/table_tennis_serve.Rmd
consistentification
distinguish hankin2017 from hankin2017_rnw

@@
 
-To cite the `hyper2` package in publications, please use @hankin2017a.
+To cite the `hyper2` package in publications, please use @hankin2017_rmd.
 In this vignette, object `H` is created.  It is identical to object

### 2024-01-26 inst/tennis.Rmd
consistentification
distinguish hankin2017 from hankin2017_rnw

@@
 
-To cite the `hyper2` package in publications, please use @hankin2017a.
+To cite the `hyper2` package in publications, please use @hankin2017_rmd.
 This short document discusses a small dataset first presented by

### 2024-01-26 inst/ternaryplot_hyper2.Rmd
consistentification
distinguish hankin2017 from hankin2017_rnw

@@
 output: html_document
-bibliography: chess.bib
+bibliography: hyper2.bib  # copy vignettes/hyper2.bib to inst/
 ---
@@
 
-To cite the `hyper2` package in publications please use @hankin2017.
+To cite the `hyper2` package in publications please use @hankin2017_rmd.
 Here I show how to use the excellent `Ternary` package [@smith2017] to

### 2024-01-26 inst/volleyball.Rmd
consistentification
distinguish hankin2017 from hankin2017_rnw

@@
 
-To cite the `hyper2` package in publications, please use @hankin2017a.
+To cite the `hyper2` package in publications, please use @hankin2017_rmd.
 This short document discusses a dataset first presented by @hankin2010,

### 2024-01-26 inst/volvo.Rmd
consistentification
distinguish hankin2017 from hankin2017_rnw

@@
 
-To cite the `hyper2` package in publications, please use @hankin2017a.
+To cite the `hyper2` package in publications, please use @hankin2017_rmd.
 This short document creates the `volvo` suite of R objects and

### 2024-01-26 inst/zermelo.Rmd
consistentification
distinguish hankin2017 from hankin2017_rnw

@@
 
-To cite the `hyper2` package in publications, please use @hankin2017a.
+To cite the `hyper2` package in publications, please use @hankin2017_rmd.
 This short document sets out a use-case for functions `zermelo()` and

### 2024-02-02 inst/ejs-sample.tex
typo

\documentclass[ejs]{imsart}

%% Packages
\RequirePackage{amsthm,amsmath,amsfonts,amssymb}
\RequirePackage[numbers]{natbib}
%\RequirePackage[authoryear]{natbib}%% uncomment this for author-year citations
\RequirePackage[colorlinks,citecolor=blue,urlcolor=blue]{hyperref}
\RequirePackage{graphicx}

\arxiv{2010.00000}
\startlocaldefs
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Uncomment next line to change            %%
%% the type of equation numbering           %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\numberwithin{equation}{section}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% For Axiom, Claim, Corollary, Hypothesis, %%
%% Lemma, Theorem, Proposition              %%
%% use \theoremstyle{plain}                 %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\theoremstyle{plain}
\newtheorem{axiom}{Axiom}
\newtheorem{claim}[axiom]{Claim}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% For Assumption, Definition, Example,     %%
%% Notation, Property, Remark, Fact         %%
%% use \theoremstyle{definition}            %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem*{example}{Example}
\newtheorem*{fact}{Fact}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% For Case use \theoremstyle{remark}       %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\theoremstyle{remark}
\newtheorem{case}{Case}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Please put your definitions here:        %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\endlocaldefs

\begin{document}
\begin{frontmatter}
\title{A sample article title}
%\title{A sample article title with some additional note\thanksref{t1}}
\runtitle{A sample running head title}
%\thankstext{T1}{A sample additional note to the title.}

\begin{aug}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Only one address is permitted per author. %%
%% Only division, organization and e-mail is %%
%% included in the address.                  %%
%% Additional information can be included in %%
%% the Acknowledgments section if necessary. %%
%% ORCID can be inserted by command:         %%
%% \orcid{0000-0000-0000-0000}               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\author[A]{\fnms{Robin}~\snm{Hankin}\ead[label=e1]{hankin.robin@gmail.com}},
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Addresses                                %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\address[A]{Department,
University or Company Name\printead[presep={,\ }]{e1}}

\end{aug}

\begin{abstract}

Here a simplification of the Plackett-Luce likelihood functions for
order statistics is proposed.  We consider a field of $n+1$ entities
that are ordered in some way [the preferred example is competitors in
  a running race, ordered by time of crossing the finishing line].
One entity is of particular interest to us: the ``focal entity'', or
focal competitor.  We assign Placket-Luce strength of $a$ to the focal
competitor and $b$ to all others; we require $a+b=1$.  This device
furnishes a reasonable, intuitive, and computationally tractable
likelihood function that I apply to a range of applications including
Formula 1 motor racing and the International Mathematical Olympiad.
The results are difficult to obtain with other methods.
\end{abstract}

\begin{keyword}[class=MSC]
\kwd[Primary ]{00X00}
\kwd{00X00}
\kwd[; secondary ]{00X00}
\end{keyword}

\begin{keyword}
\kwd{First keyword}
\kwd{second keyword}
\end{keyword}

\end{frontmatter}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Please use \tableofcontents for articles %%
%% with 50 pages and more                   %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\tableofcontents

\section{Introduction}

\cite{hankin2017_rmd}

This sample helps you to create a properly formatted \LaTeXe\ manuscript.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% `\ ' is used here because TeX ignores    %%
%% spaces after text commands.              %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Prepare your paper in the same style as used in this sample .pdf file.
Try to avoid excessive use of italics and bold face.
Please do not use any \LaTeXe\ or \TeX\ commands that affect the layout
or formatting of your document (i.e., commands like \verb|\textheight|,
\verb|\textwidth|, etc.).

\section{Section headings}
Here are some sub-sections:
\subsection{A sub-section}
Regular text.
\subsubsection{A sub-sub-section}
Regular text.

\section{Text}
\subsection{Lists}

The following is an example of an \emph{itemized} list,
two levels deep.
\begin{itemize}
\item
This is the first item of an itemized list.  Each item
in the list is marked with a ``tick.''  The document
style determines what kind of tick mark is used.
\item
This is the second item of the list.  It contains another
list nested inside it.
\begin{itemize}
\item This is the first item of an itemized list that
is nested within the itemized list.
\item This is the second item of the inner list.  \LaTeX\
allows you to nest lists deeper than you really should.
\end{itemize}
This is the rest of the second item of the outer list.
\item
This is the third item of the list.
\end{itemize}

The following is an example of an \emph{enumerated} list of one level.

\begin{enumerate}
\item This is the first item of an enumerated list.
\item This is the second item of an enumerated list.
\end{enumerate}

The following is an example of an \emph{enumerated} list, two levels deep.
\begin{enumerate}
\item
This is the first item of an enumerated list.  Each item
in the list is marked with a ``tick.''  The document
style determines what kind of tick mark is used.
\item
This is the second item of the list.  It contains another
list nested inside of it.
\begin{enumerate}
\item
This is the first item of an enumerated list that
is nested within.
\item
This is the second item of the inner list.  \LaTeX\
allows you to nest lists deeper than you really should.
\end{enumerate}
This is the rest of the second item of the outer list.
\item
This is the third item of the list.
\end{enumerate}

\subsection{Punctuation}
Dashes come in three sizes: a hyphen, an intra-word dash like ``$U$-statistics'' or ``the time-homogeneous model'';
a medium dash (also called an ``en-dash'') for number ranges or between two equal entities like ``1--2'' or ``Cauchy--Schwarz inequality'';
and a punctuation dash (also called an ``em-dash'') in place of a comma, semicolon,
colon or parentheses---like this.

Generating an ellipsis \ldots\ with the right spacing
around the periods requires a special command.

\section{Fonts}
Please use text fonts in text mode, e.g.:
\begin{itemize}
\item[]\textrm{Roman}
\item[]\textit{Italic}
\item[]\textbf{Bold}
\item[]\textsc{Small Caps}
\item[]\textsf{Sans serif}
\item[]\texttt{Typewriter}
\end{itemize}
Please use mathematical fonts in mathematical mode, e.g.:
\begin{itemize}
\item[] $\mathrm{ABCabc123}$
\item[] $\mathit{ABCabc123}$
\item[] $\mathbf{ABCabc123}$
\item[] $\boldsymbol{ABCabc123\alpha\beta\gamma}$
\item[] $\mathcal{ABC}$
\item[] $\mathbb{ABC}$
\item[] $\mathsf{ABCabc123}$
\item[] $\mathtt{ABCabc123}$
\item[] $\mathfrak{ABCabc123}$
\end{itemize}
Note that \verb|\mathcal, \mathbb| belongs to capital letters-only font typefaces.

\section{Notes}
Footnotes\footnote{This is an example of a footnote.}
pose no problem.\footnote{Note that footnote number is after punctuation.}

\section{Quotations}

Text is displayed by indenting it from the left margin. There are short quotations
\begin{quote}
This is a short quotation.  It consists of a
single paragraph of text.  There is no paragraph
indentation.
\end{quote}
and longer ones.
\begin{quotation}
This is a longer quotation.  It consists of two paragraphs
of text.  The beginning of each paragraph is indicated
by an extra indentation.

This is the second paragraph of the quotation.  It is just
as dull as the first paragraph.
\end{quotation}

\section{Environments}

\subsection{Examples for \emph{\texttt{plain}}-style environments}
\begin{axiom}\label{ax1}
This is the body of Axiom \ref{ax1}.
\end{axiom}

\begin{proof}
This is the body of the proof of the axiom above.
\end{proof}

\begin{claim}\label{cl1}
This is the body of Claim \ref{cl1}. Claim \ref{cl1} is numbered after
Axiom \ref{ax1} because we used \verb|[axiom]| in \verb|\newtheorem|.
\end{claim}

\begin{theorem}\label{th1}
This is the body of Theorem \ref{th1}. Theorem \ref{th1} numbering is
dependent on section because we used \verb|[section]| after \verb|\newtheorem|.
\end{theorem}

\begin{theorem}[Title of the theorem]\label{th2}
This is the body of Theorem \ref{th2}. Theorem \ref{th2} has additional title.
\end{theorem}

\begin{lemma}\label{le1}
This is the body of Lemma \ref{le1}. Lemma \ref{le1} is numbered after
Theorem \ref{th2} because we used \verb|[theorem]| in \verb|\newtheorem|.
\end{lemma}

\begin{proof}[Proof of Lemma \ref{le1}]
This is the body of the proof of Lemma \ref{le1}.
\end{proof}

\subsection{Examples for \emph{\texttt{definition}}-style environments}
\begin{definition}\label{de1}
This is the body of Definition \ref{de1}. Definition \ref{de1} is numbered after
Lemma \ref{le1} because we used \verb|[theorem]| in \verb|\newtheorem|.
\end{definition}

\begin{example}
This is the body of the example. Example is unnumbered because we used \verb|\newtheorem*|
instead of \verb|\newtheorem|.
\end{example}

\begin{fact}
This is the body of the fact. Fact is unnumbered because we used \verb|\newtheorem*|
instead of \verb|\newtheorem|.
\end{fact}

\subsection{Example for \emph{\texttt{remark}}-style environments}
\begin{case}\label{case}
This is the body of Case \ref{case}.
\end{case}

\section{Tables and figures}
Cross-references to labeled tables: As you can see in Table~\ref{sphericcase}
and also in Table~\ref{parset}.

\begin{table*}[t]
\caption{The spherical case ($I_1=0$, $I_2=0$)}
\label{sphericcase}
\begin{tabular}{@{}lrrrrc@{}}
\hline
Equil. \\
points & \multicolumn{1}{c}{$x$}
& \multicolumn{1}{c}{$y$} & \multicolumn{1}{c}{$z$}
& \multicolumn{1}{c}{$C$} & S \\
\hline
$L_1$    & $-$2.485252241 & 0.000000000  & 0.017100631  & 8.230711648  & U \\
$L_2$    & 0.000000000  & 0.000000000  & 3.068883732  & 0.000000000  & S \\
$L_3$    & 0.009869059  & 0.000000000  & 4.756386544  & $-$0.000057922 & U \\
$L_4$    & 0.210589855  & 0.000000000  & $-$0.007021459 & 9.440510897  & U \\
$L_5$    & 0.455926604  & 0.000000000  & $-$0.212446624 & 7.586126667  & U \\
$L_6$    & 0.667031314  & 0.000000000  & 0.529879957  & 3.497660052  & U \\
$L_7$    & 2.164386674  & 0.000000000  & $-$0.169308438 & 6.866562449  & U \\
$L_8$    & 0.560414471  & 0.421735658  & $-$0.093667445 & 9.241525367  & U \\
$L_9$    & 0.560414471  & $-$0.421735658 & $-$0.093667445 & 9.241525367  & U \\
$L_{10}$ & 1.472523232  & 1.393484549  & $-$0.083801333 & 6.733436505  & U \\
$L_{11}$ & 1.472523232  & $-$1.393484549 & $-$0.083801333 & 6.733436505  & U \\
\hline
\end{tabular}
\end{table*}

\begin{table}[t]
\caption{Sample posterior estimates for each model}
\label{parset}
%
\begin{tabular}{@{}lcrcrrr@{}}
\hline
&& & &\multicolumn{3}{c}{Quantile} \\
\cline{5-7}
Model &Parameter &
\multicolumn{1}{c}{Mean} &
Std. dev.&
\multicolumn{1}{c}{2.5\%} &
\multicolumn{1}{c}{50\%}&
\multicolumn{1}{c@{}}{97.5\%} \\
\hline
{Model 0} & $\beta_0$ & $-$12.29 & 2.29 & $-$18.04 & $-$11.99 & $-$8.56 \\
          & $\beta_1$  & 0.10   & 0.07 & $-$0.05  & 0.10   & 0.26  \\
          & $\beta_2$   & 0.01   & 0.09 & $-$0.22  & 0.02   & 0.16  \\[6pt]
{Model 1} & $\beta_0$   & $-$4.58  & 3.04 & $-$11.00 & $-$4.44  & 1.06  \\
          & $\beta_1$   & 0.79   & 0.21 & 0.38   & 0.78   & 1.20  \\
          & $\beta_2$   & $-$0.28  & 0.10 & $-$0.48  & $-$0.28  & $-$0.07 \\[6pt]
{Model 2} & $\beta_0$   & $-$11.85 & 2.24 & $-$17.34 & $-$11.60 & $-$7.85 \\
          & $\beta_1$   & 0.73   & 0.21 & 0.32   & 0.73   & 1.16  \\
          & $\beta_2$   & $-$0.60  & 0.14 & $-$0.88  & $-$0.60  & $-$0.34 \\
          & $\beta_3$   & 0.22   & 0.17 & $-$0.10  & 0.22   & 0.55  \\
\hline
\end{tabular}
%
\end{table}

\begin{figure}[t]
\includegraphics{figure1}
\caption{Pathway of the penicillin G biosynthesis.}
\label{penG}
\end{figure}

Sample of cross-reference to figure.
Figure~\ref{penG} shows that it is not easy to get something on paper.

\section{Equations and the like}

Two equations:
\begin{equation}
    C_{s}  =  K_{M} \frac{\mu/\mu_{x}}{1-\mu/\mu_{x}} \label{ccs}
\end{equation}
and
\begin{equation}
    G = \frac{P_{\mathrm{opt}} - P_{\mathrm{ref}}}{P_{\mathrm{ref}}}  100(\%).
\end{equation}

Equation arrays:
\begin{eqnarray}
  \frac{dS}{dt} & = & - \sigma X + s_{F} F,\\
  \frac{dX}{dt} & = &   \mu    X,\\
  \frac{dP}{dt} & = &   \pi    X - k_{h} P,\\
  \frac{dV}{dt} & = &   F.
\end{eqnarray}
One long equation:
\begin{eqnarray}
 \mu_{\text{normal}} & = & \mu_{x} \frac{C_{s}}{K_{x}C_{x}+C_{s}}  \nonumber\\
                     & = & \mu_{\text{normal}} - Y_{x/s}\bigl(1-H(C_{s})\bigr)(m_{s}+\pi /Y_{p/s})\\
                     & = & \mu_{\text{normal}}/Y_{x/s}+ H(C_{s}) (m_{s}+ \pi /Y_{p/s}).\nonumber
\end{eqnarray}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Example with single Appendix:            %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{appendix}
\section*{Title}\label{appn} %% if no title is needed, leave empty \section*{}.
Appendices should be provided in \verb|{appendix}| environment,
before Acknowledgements.

If there is only one appendix,
then please refer to it in text as \ldots\ in the \hyperref[appn]{Appendix}.
\end{appendix}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Example with multiple Appendixes:        %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{appendix}
\section{Title of the first appendix}\label{appA}
If there are more than one appendix, then please refer to it
as \ldots\ in Appendix \ref{appA}, Appendix \ref{appB}, etc.

\section{Title of the second appendix}\label{appB}
\subsection{First subsection of Appendix \protect\ref{appB}}

Use the standard \LaTeX\ commands for headings in \verb|{appendix}|.
Headings and other objects will be numbered automatically.
\begin{equation}
\mathcal{P}=(j_{k,1},j_{k,2},\dots,j_{k,m(k)}). \label{path}
\end{equation}

Sample of cross-reference to the formula (\ref{path}) in Appendix \ref{appB}.
\end{appendix}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Acknowledgements                         %%
%% should be provided in the                %%
%% Acknowledgements section.                %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{acks}[Acknowledgments]
The authors would like to thank the anonymous referees, an Associate
Editor and the Editor for their constructive comments that improved the
quality of this paper.
\end{acks}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Funding information, if any,             %%
%% should be provided in the                %%
%% funding section.                         %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{funding}
The first author was supported by NSF Grant DMS-??-??????.

The second author was supported in part by NIH Grant ???????????.
\end{funding}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Supplementary Material, including data   %%
%% sets and code, should be provided in     %%
%% {supplement} environment with title      %%
%% and short description. It cannot be      %%
%% available exclusively as external link.  %%
%% All Supplementary Material must be       %%
%% available to the reader on Project       %%
%% Euclid with the published article.       %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{supplement}
\stitle{Title of Supplement A}
\sdescription{Short description of Supplement A.}
\end{supplement}
\begin{supplement}
\stitle{Title of Supplement B}
\sdescription{Short description of Supplement B.}
\end{supplement}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                  The Bibliography                       %%
%%                                                         %%
%%  imsart-???.bst  will be used to                        %%
%%  create a .BBL file for submission.                     %%
%%                                                         %%
%%  Note that the displayed Bibliography will not          %%
%%  necessarily be rendered by Latex exactly as specified  %%
%%  in the online Instructions for Authors.                %%
%%                                                         %%
%%  MR numbers will be added by VTeX.                      %%
%%                                                         %%
%%  Use \cite{...} to cite references in text.             %%
%%                                                         %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% if your bibliography is in bibtex format, uncomment commands:
\bibliographystyle{imsart-number} % Style BST file (imsart-number.bst or imsart-nameyear.bst)
\bibliography{hyper2}       % Bibliography file (usually '*.bib')

\end{document}

### 2024-02-02 inst/ejs-sample.tex
more work

@@
 \begin{frontmatter}
-\title{A sample article title}
-%\title{A sample article title with some additional note\thanksref{t1}}
-\runtitle{A sample running head title}
+\title{A simplified Plackett-Luce likelihood function for rank orders}
+\runtitle{Simplified Plackett-Luce likelihood functions}
 %\thankstext{T1}{A sample additional note to the title.}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\address[A]{Department,
-University or Company Name\printead[presep={,\ }]{e1}}
+\address[A]{Computer Science and Mathematics,
+University of Stirling\printead[presep={,\ }]{e1}}
 
@@
 \begin{keyword}[class=MSC]
-\kwd[Primary ]{00X00}
+\kwd[Primary ]{6202}
 \kwd{00X00}
-\kwd[; secondary ]{00X00}
+\kwd[; secondary ]{62F30}
 \end{keyword}
@@
 
-\cite{hankin2017_rmd}
+Ranked data arises whenever multiple entities are sorted by some
+criterion; the preferred interpretation is a race in which the
+sufficient statistic is the order of competitors crossing the
+finishing line.  The Plackett-Luce likelihood
+function~\cite{luce1959,plackett1975} is a generalization of the
+Bradley-Terry~\cite{bradley1952} in which non-negative strengths are
+assigned to each competitor; without loss of generality, if the
+finishing order is $1,2,\ldots,n$ and competitor $i$ has strength
+$p_i\geqslant 0$, then the Plackett-Luce likelihood function will be
+proportional to
 
-This sample helps you to create a properly formatted \LaTeXe\ manuscript.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% `\ ' is used here because TeX ignores    %%
-%% spaces after text commands.              %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-Prepare your paper in the same style as used in this sample .pdf file.
-Try to avoid excessive use of italics and bold face.
-Please do not use any \LaTeXe\ or \TeX\ commands that affect the layout
-or formatting of your document (i.e., commands like \verb|\textheight|,
-\verb|\textwidth|, etc.).
+\begin{equation}\label{plackettluce}
+\prod_{i=1}^n\frac{p_i}{\sum_{j=i}^np_i}
+\end{equation}
 
-\section{Section headings}
-Here are some sub-sections:
-\subsection{A sub-section}
-Regular text.
-\subsubsection{A sub-sub-section}
-Regular text.
+\noindent where the strengths are normalised so $\sum p_i=1$.
+However, estimation of the strengths is somewhat difficult if $n$ is
+even moderately large: even with as few as 20 competitors following
+Zipf's law~\cite{zipf1949} [a typical assumption for Bradley-Terry
+  strengths], Hankin~\cite{hankin2017_rmd} shows that maximum
+likelihood strengths identify the correct ranking with a probability
+of essentially zero.  The situation is worse in cases such as the
+International Mathematical Olympiad, which requires us to consider
+over 100 competitors.
 
-\section{Text}
-\subsection{Lists}
+Consider a race between a competitor of Bradley-Terry strength $a$ and
+$n$ cloned competitors, each of strength $b=1-a$.  An observation is
+indexed by~$r$, the number of strength $b$ clones finishing ahead
+of~$a$, and because we have $n+1$ competitors $0\leqslant r\leqslant
+n$.  The initial field strength is $a+nb=n+(n-1)a$.  The Plackett-Luce
+likelihood function~\cite{plackett1975,luce1959} for this situation
+would be
 
-The following is an example of an \emph{itemized} list,
-two levels deep.
-\begin{itemize}
-\item
-This is the first item of an itemized list.  Each item
-in the list is marked with a ``tick.''  The document
-style determines what kind of tick mark is used.
-\item
-This is the second item of the list.  It contains another
-list nested inside it.
-\begin{itemize}
-\item This is the first item of an itemized list that
-is nested within the itemized list.
-\item This is the second item of the inner list.  \LaTeX\
-allows you to nest lists deeper than you really should.
-\end{itemize}
-This is the rest of the second item of the outer list.
-\item
-This is the third item of the list.
-\end{itemize}
+\begin{equation}\label{likeforrn}
+  \mathcal{L}_r(a)=\begin{cases}
+\frac{b}{a+ n   b}\cdot
+\frac{b}{a+(n-1)b}\cdot
+\frac{b}{a+(n-2)b}\cdots
+\frac{a}{a+(n-r)b}\qquad &{r<n}\\
+\frac{b}{a+ n   b}\cdot
+\frac{b}{a+(n-1)b}\cdot
+\frac{b}{a+(n-2)b}\cdots
+\frac{b}{a+b}\qquad &{r=n}
+  \end{cases}
+\end{equation}
 
-The following is an example of an \emph{enumerated} list of one level.
-
-\begin{enumerate}
-\item This is the first item of an enumerated list.
-\item This is the second item of an enumerated list.
-\end{enumerate}
-
-The following is an example of an \emph{enumerated} list, two levels deep.
-\begin{enumerate}
-\item
-This is the first item of an enumerated list.  Each item
-in the list is marked with a ``tick.''  The document
-style determines what kind of tick mark is used.
-\item
-This is the second item of the list.  It contains another
-list nested inside of it.
-\begin{enumerate}
-\item
-This is the first item of an enumerated list that
-is nested within.
-\item
-This is the second item of the inner list.  \LaTeX\
-allows you to nest lists deeper than you really should.
-\end{enumerate}
-This is the rest of the second item of the outer list.
-\item
-This is the third item of the list.
-\end{enumerate}
-
-\subsection{Punctuation}
-Dashes come in three sizes: a hyphen, an intra-word dash like ``$U$-statistics'' or ``the time-homogeneous model'';
-a medium dash (also called an ``en-dash'') for number ranges or between two equal entities like ``1--2'' or ``Cauchy--Schwarz inequality'';
-and a punctuation dash (also called an ``em-dash'') in place of a comma, semicolon,
-colon or parentheses---like this.
-
-Generating an ellipsis \ldots\ with the right spacing
-around the periods requires a special command.
-
-\section{Fonts}
-Please use text fonts in text mode, e.g.:
-\begin{itemize}
-\item[]\textrm{Roman}
-\item[]\textit{Italic}
-\item[]\textbf{Bold}
-\item[]\textsc{Small Caps}
-\item[]\textsf{Sans serif}
-\item[]\texttt{Typewriter}
-\end{itemize}
-Please use mathematical fonts in mathematical mode, e.g.:
-\begin{itemize}
-\item[] $\mathrm{ABCabc123}$
-\item[] $\mathit{ABCabc123}$
-\item[] $\mathbf{ABCabc123}$
-\item[] $\boldsymbol{ABCabc123\alpha\beta\gamma}$
-\item[] $\mathcal{ABC}$
-\item[] $\mathbb{ABC}$
-\item[] $\mathsf{ABCabc123}$
-\item[] $\mathtt{ABCabc123}$
-\item[] $\mathfrak{ABCabc123}$
-\end{itemize}
-Note that \verb|\mathcal, \mathbb| belongs to capital letters-only font typefaces.
-
-\section{Notes}
-Footnotes\footnote{This is an example of a footnote.}
-pose no problem.\footnote{Note that footnote number is after punctuation.}
-
-\section{Quotations}
-
-Text is displayed by indenting it from the left margin. There are short quotations
-\begin{quote}
-This is a short quotation.  It consists of a
-single paragraph of text.  There is no paragraph
-indentation.
-\end{quote}
-and longer ones.
-\begin{quotation}
-This is a longer quotation.  It consists of two paragraphs
-of text.  The beginning of each paragraph is indicated
-by an extra indentation.
-
-This is the second paragraph of the quotation.  It is just
-as dull as the first paragraph.
-\end{quotation}
-
-\section{Environments}
-
-\subsection{Examples for \emph{\texttt{plain}}-style environments}
-\begin{axiom}\label{ax1}
-This is the body of Axiom \ref{ax1}.
-\end{axiom}
-
-\begin{proof}
-This is the body of the proof of the axiom above.
-\end{proof}
-
-\begin{claim}\label{cl1}
-This is the body of Claim \ref{cl1}. Claim \ref{cl1} is numbered after
-Axiom \ref{ax1} because we used \verb|[axiom]| in \verb|\newtheorem|.
-\end{claim}
-
-\begin{theorem}\label{th1}
-This is the body of Theorem \ref{th1}. Theorem \ref{th1} numbering is
-dependent on section because we used \verb|[section]| after \verb|\newtheorem|.
-\end{theorem}
-
-\begin{theorem}[Title of the theorem]\label{th2}
-This is the body of Theorem \ref{th2}. Theorem \ref{th2} has additional title.
-\end{theorem}
-
-\begin{lemma}\label{le1}
-This is the body of Lemma \ref{le1}. Lemma \ref{le1} is numbered after
-Theorem \ref{th2} because we used \verb|[theorem]| in \verb|\newtheorem|.
-\end{lemma}
-
-\begin{proof}[Proof of Lemma \ref{le1}]
-This is the body of the proof of Lemma \ref{le1}.
-\end{proof}
-
-\subsection{Examples for \emph{\texttt{definition}}-style environments}
-\begin{definition}\label{de1}
-This is the body of Definition \ref{de1}. Definition \ref{de1} is numbered after
-Lemma \ref{le1} because we used \verb|[theorem]| in \verb|\newtheorem|.
-\end{definition}
-
-\begin{example}
-This is the body of the example. Example is unnumbered because we used \verb|\newtheorem*|
-instead of \verb|\newtheorem|.
-\end{example}
-
-\begin{fact}
-This is the body of the fact. Fact is unnumbered because we used \verb|\newtheorem*|
-instead of \verb|\newtheorem|.
-\end{fact}
-
-\subsection{Example for \emph{\texttt{remark}}-style environments}
-\begin{case}\label{case}
-This is the body of Case \ref{case}.
-\end{case}
-
-\section{Tables and figures}
-Cross-references to labeled tables: As you can see in Table~\ref{sphericcase}
-and also in Table~\ref{parset}.
-
-\begin{table*}[t]
-\caption{The spherical case ($I_1=0$, $I_2=0$)}
-\label{sphericcase}
-\begin{tabular}{@{}lrrrrc@{}}
-\hline
-Equil. \\
-points & \multicolumn{1}{c}{$x$}
-& \multicolumn{1}{c}{$y$} & \multicolumn{1}{c}{$z$}
-& \multicolumn{1}{c}{$C$} & S \\
-\hline
-$L_1$    & $-$2.485252241 & 0.000000000  & 0.017100631  & 8.230711648  & U \\
-$L_2$    & 0.000000000  & 0.000000000  & 3.068883732  & 0.000000000  & S \\
-$L_3$    & 0.009869059  & 0.000000000  & 4.756386544  & $-$0.000057922 & U \\
-$L_4$    & 0.210589855  & 0.000000000  & $-$0.007021459 & 9.440510897  & U \\
-$L_5$    & 0.455926604  & 0.000000000  & $-$0.212446624 & 7.586126667  & U \\
-$L_6$    & 0.667031314  & 0.000000000  & 0.529879957  & 3.497660052  & U \\
-$L_7$    & 2.164386674  & 0.000000000  & $-$0.169308438 & 6.866562449  & U \\
-$L_8$    & 0.560414471  & 0.421735658  & $-$0.093667445 & 9.241525367  & U \\
-$L_9$    & 0.560414471  & $-$0.421735658 & $-$0.093667445 & 9.241525367  & U \\
-$L_{10}$ & 1.472523232  & 1.393484549  & $-$0.083801333 & 6.733436505  & U \\
-$L_{11}$ & 1.472523232  & $-$1.393484549 & $-$0.083801333 & 6.733436505  & U \\
-\hline
-\end{tabular}
-\end{table*}
-
-\begin{table}[t]
-\caption{Sample posterior estimates for each model}
-\label{parset}
-%
-\begin{tabular}{@{}lcrcrrr@{}}
-\hline
-&& & &\multicolumn{3}{c}{Quantile} \\
-\cline{5-7}
-Model &Parameter &
-\multicolumn{1}{c}{Mean} &
-Std. dev.&
-\multicolumn{1}{c}{2.5\%} &
-\multicolumn{1}{c}{50\%}&
-\multicolumn{1}{c@{}}{97.5\%} \\
-\hline
-{Model 0} & $\beta_0$ & $-$12.29 & 2.29 & $-$18.04 & $-$11.99 & $-$8.56 \\
-          & $\beta_1$  & 0.10   & 0.07 & $-$0.05  & 0.10   & 0.26  \\
-          & $\beta_2$   & 0.01   & 0.09 & $-$0.22  & 0.02   & 0.16  \\[6pt]
-{Model 1} & $\beta_0$   & $-$4.58  & 3.04 & $-$11.00 & $-$4.44  & 1.06  \\
-          & $\beta_1$   & 0.79   & 0.21 & 0.38   & 0.78   & 1.20  \\
-          & $\beta_2$   & $-$0.28  & 0.10 & $-$0.48  & $-$0.28  & $-$0.07 \\[6pt]
-{Model 2} & $\beta_0$   & $-$11.85 & 2.24 & $-$17.34 & $-$11.60 & $-$7.85 \\
-          & $\beta_1$   & 0.73   & 0.21 & 0.32   & 0.73   & 1.16  \\
-          & $\beta_2$   & $-$0.60  & 0.14 & $-$0.88  & $-$0.60  & $-$0.34 \\
-          & $\beta_3$   & 0.22   & 0.17 & $-$0.10  & 0.22   & 0.55  \\
-\hline
-\end{tabular}
-%
-\end{table}
+Although these expressions do have closed-form analytical forms, it is
+not helpful in numerical work because they tend to be very unstable,
+especially where $kb\simeq n$ for some integer $k$, as limiting forms
+have to be used.  Note carefully that the expressions in
+equation~\ref{likeforrn} are likelihoods, not probabilities, and do
+not sum to 1.
 
 \begin{figure}[t]
-\includegraphics{figure1}
-\caption{Pathway of the penicillin G biosynthesis.}
-\label{penG}
+\includegraphics[width=4in]{ninelikes}  % produced by very_simplified_likelihood.Rmd
+\caption{Likelihood functions for $r=0(1)8$\label{ninelikes}}
 \end{figure}
 
-Sample of cross-reference to figure.
-Figure~\ref{penG} shows that it is not easy to get something on paper.
+Previous related work presented the case $n=2$ [in current notation].
+Considering the case $n=7$ gives us $7+1=8$ distinct likelihood
+functions, Figure~\ref{ninelikes}.  We see an unanticipated asymmetry
+between $r$ and $n-r$, and in particular that
+$\hat{a}\geqslant\frac{r}{n}$, with equality only in the edge-cases
+$r=0$ or $n$.  For $r=3$ one might expect that $\hat{a}=\frac{1}{2}$
+on the grounds that the focal competitor's performance was median.
+However, $\hat{a}\simeq 0.514$.
 
-\section{Equations and the like}
+\begin{figure}[t]
+\includegraphics[width=4in]{dotprobs}  % produced by very_simplified_likelihood.Rmd
+\caption{Maximum likelihood estimator $\hat{a}$ for $n=1(1)10$ and $r=0(1)n$
+  \label{dotprobs}}
+\end{figure}
 
-Two equations:
-\begin{equation}
-    C_{s}  =  K_{M} \frac{\mu/\mu_{x}}{1-\mu/\mu_{x}} \label{ccs}
-\end{equation}
-and
-\begin{equation}
-    G = \frac{P_{\mathrm{opt}} - P_{\mathrm{ref}}}{P_{\mathrm{ref}}}  100(\%).
-\end{equation}
+\section{Use-cases}
 
-Equation arrays:
-\begin{eqnarray}
-  \frac{dS}{dt} & = & - \sigma X + s_{F} F,\\
-  \frac{dX}{dt} & = &   \mu    X,\\
-  \frac{dP}{dt} & = &   \pi    X - k_{h} P,\\
-  \frac{dV}{dt} & = &   F.
-\end{eqnarray}
-One long equation:
-\begin{eqnarray}
- \mu_{\text{normal}} & = & \mu_{x} \frac{C_{s}}{K_{x}C_{x}+C_{s}}  \nonumber\\
-                     & = & \mu_{\text{normal}} - Y_{x/s}\bigl(1-H(C_{s})\bigr)(m_{s}+\pi /Y_{p/s})\\
-                     & = & \mu_{\text{normal}}/Y_{x/s}+ H(C_{s}) (m_{s}+ \pi /Y_{p/s}).\nonumber
-\end{eqnarray}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Example with single Appendix:            %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{appendix}
-\section*{Title}\label{appn} %% if no title is needed, leave empty \section*{}.
-Appendices should be provided in \verb|{appendix}| environment,
-before Acknowledgements.
+Formula 1 motor racing is an important and prestigious motor sport
+\citep{codling2017,jenkins2010}.  Season ranking is based on a points
+allocation system wherein competitors are awarded points based on race
+finishing order; points accumulate additively.  The overall
+competition winner is the competitor who accumulates the most points
+after the final race.  However, as argued by Hankin~\cite{hankin2023},
+the drivers' {\em ranks} are statistically sufficient for analysis of
 
-If there is only one appendix,
-then please refer to it in text as \ldots\ in the \hyperref[appn]{Appendix}.
-\end{appendix}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Example with multiple Appendixes:        %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{appendix}
-\section{Title of the first appendix}\label{appA}
-If there are more than one appendix, then please refer to it
-as \ldots\ in Appendix \ref{appA}, Appendix \ref{appB}, etc.
+The methods discussed above may be used to assess Bradley-Terry
+strengths of Formula 1 racing drivers.  Let us consider 2018 season as
+an example and first consider Lewis Hamilton.
 
-\section{Title of the second appendix}\label{appB}
-\subsection{First subsection of Appendix \protect\ref{appB}}
+Lewis Hamilton is a
 
-Use the standard \LaTeX\ commands for headings in \verb|{appendix}|.
-Headings and other objects will be numbered automatically.
-\begin{equation}
-\mathcal{P}=(j_{k,1},j_{k,2},\dots,j_{k,m(k)}). \label{path}
-\end{equation}
+Pierre Gasly and Esteban Ocon have a long-standing rivalry and
+comparisons between them typically compare disparate metrics such as
+number of podiums, and highest grid position in a season.
 
-Sample of cross-reference to the formula (\ref{path}) in Appendix \ref{appB}.
-\end{appendix}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Acknowledgements                         %%
-%% should be provided in the                %%
-%% Acknowledgements section.                %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{acks}[Acknowledgments]
@@
 \end{acks}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Funding information, if any,             %%
-%% should be provided in the                %%
-%% funding section.                         %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{funding}
-The first author was supported by NSF Grant DMS-??-??????.
-
-The second author was supported in part by NIH Grant ???????????.
-\end{funding}
 

### 2024-02-02 inst/very_simplified_likelihood.Rmd
more work

@@
 }
+MLE(0,9)
 MLE(3,9)
+MLE(3,7)
 ```
@@
 f(a,r,n,log=FALSE))
+```
+
+```{r}
+out <- list()
+for(n in 1:10){
+  out[[n]]  <- rep(NA,n+1)
+  for(r in 0:n){
+    if(r==0){
+      jj <- 1
+    } else if(r==n){n
+      jj <- 0
+    } else {
+      jj <- MLE(r,n)$maximum
+    }  
+    out[[n]][r+1] <- jj
+  }
+} 
+out
+```
+
+```{r}
+plotterp <- function(...){
+plot(NA,xlim=c(1,10),ylim=c(0,1),type="n",xlab="n",ylab=expression(hat(a)))
+for(n in 1:10){
+  for(r in 0:n){   points(n,out[[n]][r+1],pch=16) }
+  for(r in 0:n){   text(n+0.2,out[[n]][r+1],r,cex=0.5,col='gray') }
+}
+}
+plotterp()
+pdf(file="dotprobs.pdf")
+plotterp()
+dev.off()
 ```
@@
 
-```{r}
+```{r,cache=TRUE}
+plotter <- function(...){
 a <- seq(from=0,to=1,by=0.01)
-plot(a,f(a,3,8,log=FALSE))
+n <- 7
+jj <- f(a,3,n,log=FALSE)
+plot(a,jj/max(jj,na.rm=TRUE),type='n')
+grid()
+for(r in seq(from=0,to=n)){
+  y <- f_vec_a(a,r,n,log=FALSE)
+  if(r==0){y[length(y)] <- 1}  
+  y <- y/max(y,na.rm=TRUE)
+  if(r>0){y[length(y)] <- 0}
+  points(a,y,type='l',lwd=4,col=rainbow(n+1)[r+1])
+  segments(x0=r/(n),y0=0.9,y1=1)
+  text(x=r/(n),y=0.8,r)
+}
+}
+plotter()
+pdf(file="ninelikes.pdf")
+plotter()
+dev.off()
 ```

### 2024-02-09 inst/very_simplified_likelihood.Rmd
park run using VSL

@@
 
-## Masterchef
+## Parkrun
 
-Look at the dotchart in the _MasterChef_ part of the `hyper2`
-vignette.  We could use a Very Simplified Likelihood to estimate, say,
-Laura's BT strength assuming that the other competitors are clones.
+```{r readresults}
+results <- read.table("parkrun_results.txt",header=TRUE)
+results
+```
+
+Above we see three parkrun results.  In parkrun number 148, I placed
+229 out of a field of 318.
+
+```{r definelikelihood}
+
+L <- function(a,results){
+   out <- a*0
+   for(i in seq_len(nrow(results))){
+   jj <- f_vec_a(a,r=results$place[i]-1,results$runners[i]-1,log=TRUE)
+   out <- out + jj
+   }
+   out[is.infinite(out)] <- NA
+   return(out-max(out,na.rm=TRUE))
+}
+```
+
+```{r calculatelike,cache=TRUE}
+a <- seq(from=0.15,to=0.5,by=0.01)
+La <- L(a,results)
+```
+
+```{r plotlike}
+plot(a,La)
+abline(h=c(0,-2))
+```
+
+```{r splitresults}
+supp2 <- function(vec,place,runners){
+  a_stirling <- vec[1]
+  a_auckland <- vec[2]
+  out <- 0
+  for(i in 1:3){
+    out <- out + f_vec_a(a_stirling,r=place[i]-1,runners[i]-1,log=TRUE)
+  }
+  for(i in 4:10){
+    out <- out + f_vec_a(a_auckland,r=place[i]-1,runners[i]-1,log=TRUE)
+  }
+  return(out)
+}
+```
+
+```{r}
+print(supp2(c(0.1,0.3),results$place,results$runners))
+supp2(c(0.1,0.3),results$place,results$runners)
+x <- seq(from=0.1,to=0.5,by=0.005)
+jj <- as.matrix(expand.grid(x,x))
+L <- apply(jj,1,supp2,results$place,results$runners)
+L <- L - max(L,na.rm=TRUE)
+L <- matrix(L,length(x),length(x))
+par(pty='s')
+contour(x,x,L,asp=1,xlim=range(x),ylim=range(x))
+abline(0,1)
+```

### 2024-02-10 inst/very_simplified_likelihood.Rmd
consistentification

--- 2024-02-02 inst/very_simplified_likelihood.Rmd
@@
 
-f <- function(a,r,n,log=FALSE){ # vectorized in 'a', treats 'r' as a vector of independent observations
+f_vec_arn <- function(a,r,n,log=FALSE){ ## vectorized in 'a'; treats 'r' and 'n' as
+                                ## vectors of independent observations
+
+  M <- cbind(r,n)
 
@@
     out <- 0
-    for(rr in r){out <- out + f_vec_a(a,rr,n,log=TRUE)}
+    for(i in seq_len(nrow(M))){out <- out + f_vec_a(a,r=M[i,1],n=M[i,2],log=TRUE)}
   } else {
     out <- 1
-    for(rr in r){out <- out * f_vec_a(a,rr,n,log=FALSE)}
+    for(i in seq_len(nrow(M))){out <- out * f_vec_a(a,r=M[i,1],n=M[i,2],log=FALSE)}
   }
@@
     d <- 1e-6
-   optimize(f,c(d,1-d),r=r,n=n,log=TRUE,maximum=TRUE)
+    optimize(f_vec_arn,c(d,1-d),r=r,n=n,log=TRUE,maximum=TRUE)
 }
@@
 (B-1)/prod(B+(5:8)),
-f(a,r,n,log=FALSE))
+f_vec_arn(a,r,n,log=FALSE))
 ```
@@
 
-```{r,cache=TRUE}
+```{r,label=likevis,cache=TRUE}
 plotter <- function(...){
@@
 a <- a[,seq_len(ncol(a)-1)]
-d1 <- as.numeric(as.matrix(a)["Ocon",])
-d2 <- as.numeric(as.matrix(a)["Gasly",])
-d1[is.na(d1)] <- 22
-d2[is.na(d2)] <- 22
-d1
-d2
-MLE(d1,22)
-MLE(d2,22)
+d_ocon  <- as.numeric(as.matrix(a)["Ocon",])
+d_gasly <- as.numeric(as.matrix(a)["Gasly",])
+d_ocon [is.na(d_ocon )] <- 22
+d_gasly[is.na(d_gasly)] <- 22
+d_ocon
+d_gasly
+MLE(d_ocon ,22)
+MLE(d_gasly,22)
 ```
@@
 ```{r 2samptest}
-both <- c(d1,d2)
+both <- c(d_ocon,d_gasly)
 bothmax <- MLE(both,22)$maximum
 
-d1max <- MLE(d1,22)$maximum
-d2max <- MLE(d2,22)$maximum
-f(bothmax,d1,22,log=TRUE)
-f(bothmax,d2,22,log=TRUE)
-f(d1max,d1,22,log=TRUE)
-f(d2max,d2,22,log=TRUE)
+d_ocon_max  <- MLE(d_ocon ,22)$maximum
+d_gasly_max <- MLE(d_gasly,22)$maximum
+f_vec_arn(bothmax,d_ocon ,22,log=TRUE)
+f_vec_arn(bothmax,d_gasly,22,log=TRUE)
+f_vec_arn(d_ocon_max ,d_ocon ,22,log=TRUE)
+f_vec_arn(d_gasly_max,d_gasly,22,log=TRUE)
 
-Lambda <- (f(d2max,d1,22,log=TRUE)+f(d2max,d2,22,log=TRUE)) -(f(bothmax,d1,22,log=TRUE)+f(bothmax,d2,22,log=TRUE))
+Lambda <- (
+   (f_vec_arn(d_gasly_max  ,d_ocon,22,log=TRUE)+f_vec_arn(d_gasly_max  ,d_gasly,22,log=TRUE)) -
+   (f_vec_arn(bothmax      ,d_ocon,22,log=TRUE)+f_vec_arn(bothmax      ,d_gasly,22,log=TRUE))
+)
 Lambda
@@
   d2max <- MLE(d2,n)$maximum
-  Lambda <- (f(d2max,d1,n,log=TRUE)+f(d2max,d2,n,log=TRUE)) -(f(bothmax,d1,n,log=TRUE)+f(bothmax,d2,n,log=TRUE))
+  Lambda <- (
+  (f_vec_arn(d2max  ,d1,n,log=TRUE)+f_vec_arn(d2max  ,d2,n,log=TRUE)) -
+  (f_vec_arn(bothmax,d1,n,log=TRUE)+f_vec_arn(bothmax,d2,n,log=TRUE))
+)
   pchisq(-2*Lambda,df=1,lower.tail=FALSE)
@@
 
-## Masterchef
+## Parkrun
 
-Look at the dotchart in the _MasterChef_ part of the `hyper2`
-vignette.  We could use a Very Simplified Likelihood to estimate, say,
-Laura's BT strength assuming that the other competitors are clones.
+```{r readresults}
+results <- read.table("parkrun_results.txt",header=TRUE)
+results
+```
+
+Above we see three parkrun results.  In parkrun number 148, I placed
+229 out of a field of 318.
+
+```{r definelikelihood}
+
+L <- function(a,results){
+   out <- a*0
+   for(i in seq_len(nrow(results))){
+   jj <- f_vec_a(a,r=results$place[i]-1,results$runners[i]-1,log=TRUE)
+   out <- out + jj
+   }
+   out[is.infinite(out)] <- NA
+   return(out)
+}
+```
+
+```{r calculatelikes,cache=TRUE}
+a <- seq(from=0.15,to=0.5,by=0.01)
+La <- L(a,results)
+La <- La-max(La,na.rm=TRUE)
+La2 <- f_vec_arn(a,r=results$place-1,results$runners-1,log=TRUE)
+La2 <- La2 - max(La2,na.rm=TRUE)
+```
+
+```{r plotlike}
+plot(a,La)
+points(a,La2)
+abline(h=c(0,-2))
+```
+
+```{r optimsinglelike}
+optimize(function(a){L(a,results)},c(0.3,0.4),maximum=TRUE)
+```
+
+```{r splitresults}
+supp2 <- function(vec,place,runners){
+  a_stirling <- vec[1]
+  a_auckland <- vec[2]
+  out <- 0
+  for(i in 1:3){
+    out <- out + f_vec_a(a_stirling,r=place[i]-1,runners[i]-1,log=TRUE)
+  }
+  for(i in 4:10){
+    out <- out + f_vec_a(a_auckland,r=place[i]-1,runners[i]-1,log=TRUE)
+  }
+  return(out)
+}
+```
+
+```{r,label=calcbothlikes,cache=FALSE}
+x <- seq(from=0.1,to=0.8,by=0.01)
+jj <- as.matrix(expand.grid(x,x))
+L <- apply(jj,1,supp2,results$place,results$runners)
+L <- L - max(L,na.rm=TRUE)
+L <- matrix(L,length(x),length(x))
+```
+
+```{r,label=plotbothlikes}
+par(pty='s')
+contour(x,x,L,asp=1,xlim=range(x),ylim=range(x),levels=-(0:5))
+abline(0,1)
+```
+
+```{r, label=onedimopt}
+f <- function(a){supp2(c(a,a),results$place,results$runners)}
+a <- seq(from=0.1,to=0.6,by=0.05)
+alike <- sapply(a,f)
+plot(a,alike,type='b')
+optimize(f,c(0.1,0.6),maximum=TRUE)
+```
+
+```{r, label=twodimopt}
+f2 <- function(v){supp2(v,results$place,results$runners)}
+optim(par=c(0.6,0.3),f2,control=list(fnscale=-1))
+```
+
+```{r,label=showchisquareasymptotic}
+pchisq(2*(7381.496-7380.314),df=1,lower.tail=FALSE)
+```
+
+## Education
+
+The focal competitor
+
+```{r}
+rank <- c( 9, 7, 2, 3,2 ,8 ,5 ,4 ,9 )
+class_size <- c(12, 17,23, 9,13,14,13,12,15)
+course <- c("rings and modules","group theory","calculus","linear algebra",
+"differential equations","topology","special relativity","fluid mechanics","Lie algebra")
+category=c("pure","pure","applied","applied","applied","pure","applied","applied","pure")
+data.frame(course,category,rank,class_size)
+```
+
+```{r}
+a <- seq(from=0.1,by=0.01,to=0.7)
+wp <- category=='pure'
+wa <- category=='applied'
+```
+
+Pure first
+
+```{r}
+Lpure <- f_vec_arn(a,rank[wp]-1,class_size[wp]-1,log=TRUE)
+Lpure <- Lpure - max(Lpure,na.rm=TRUE)
+plot(a,Lpure)
+optimize(function(a){f_vec_arn(a,rank[wp]-1,class_size[wp]-1,log=TRUE)},c(0.3,0.6),maximum=TRUE)
+optimize(function(a){f_vec_arn(a,rank[wa]-1,class_size[wa]-1,log=TRUE)},c(0.3,0.6),maximum=TRUE)
+```
+
+```{r}
+supp2_ed <- function(vec,place1,place2,runners1,runners2){
+  out <- 0
+  M1 <- cbind(place1,runners1)
+  M2 <- cbind(place2,runners2)
+  for(i in seq_len(nrow(M1))){
+    out <- out + f_vec_arn(vec[1],r=M1[i,1]-1,n=M1[i,2]-1,log=TRUE)
+  }
+  for(i in seq_len(nrow(M2))){
+    out <- out + f_vec_arn(vec[2],r=M2[i,1]-1,n=M2[i,2]-1,log=TRUE)
+  }
+  return(out)
+}
+```
+
+```{r,cache=TRUE}
+x <- seq(from=0.1,to=0.95,by=0.01)
+jj <- as.matrix(expand.grid(x,x))
+L <- apply(jj,1,supp2_ed,rank[wp]-1,rank[wa]-1,class_size[wp]-1,class_size[wa]-1)
+L <- L - max(L,na.rm=TRUE)
+L <- matrix(L,length(x),length(x))
+```
+
+```{r}
+par(pty='s')
+contour(x,x,L,asp=1,xlim=range(x),ylim=range(x),levels=-(0:5),
+xlab='pure strength',ylab='applied strength')
+abline(0,1)
+```
+
+```{r}
+jj1 <- 
+optim(
+   c(.4,.8),
+   function(vec){
+      supp2_ed(vec,rank[wp]-1,rank[wa]-1,class_size[wp]-1,class_size[wa]-1)},
+   control=list(fnscale = -1)
+)
+jj1
+```
+
+```{r}
+jj2 <-
+optimize(function(v){supp2_ed(c(v,v),rank[wp]-1,rank[wa]-1,class_size[wp]-1,class_size[wa]-1)},
+c(0.1,0.9),maximum=TRUE)
+jj2
+```
+
+```{r}
+pchisq(2*(jj1$value-jj2$objective),lower.tail=FALSE,df=1)
+```

### 2024-02-11 inst/very_simplified_likelihood.Rmd
streamlining

@@
 a <- seq(from=0.15,to=0.5,by=0.01)
-La <- L(a,results)
-La <- La-max(La,na.rm=TRUE)
-La2 <- f_vec_arn(a,r=results$place-1,results$runners-1,log=TRUE)
-La2 <- La2 - max(La2,na.rm=TRUE)
+La <- f_vec_arn(a,r=results$place-1,results$runners-1,log=TRUE)
+La <- La - max(La,na.rm=TRUE)
 ```
@@
 plot(a,La)
-points(a,La2)
 abline(h=c(0,-2))
@@
 ```{r optimsinglelike}
-optimize(function(a){L(a,results)},c(0.3,0.4),maximum=TRUE)
+optimize(function(a){L(a,results)},c(0.2,0.4),maximum=TRUE)
+```
+
+```{r calculatelikes1,cache=TRUE}
+a <- seq(from=0.005,to=0.6,by=0.005)
+La <- f_vec_arn(a,r=results$place[1]-1,results$runners[1]-1,log=TRUE)
+La <- La - max(La,na.rm=TRUE)
+plot(a,La,type='b')
+abline(h=c(0,-2))
 ```
@@
 f <- function(a){supp2(c(a,a),results$place,results$runners)}
-a <- seq(from=0.1,to=0.6,by=0.05)
+a <- seq(from=0.1,to=0.45,by=0.01)
 alike <- sapply(a,f)
-plot(a,alike,type='b')
+plot(a,alike-max(alike,na.rm=TRUE),type='b')
+abline(h=c(0,-2))
 optimize(f,c(0.1,0.6),maximum=TRUE)
@@
 ```{r,label=showchisquareasymptotic}
+7380.314-7381.496
 pchisq(2*(7381.496-7380.314),df=1,lower.tail=FALSE)

### 2024-02-11 inst/ejs-sample.tex
rewording

@@
 function~\cite{luce1959,plackett1975} is a generalization of the
-Bradley-Terry~\cite{bradley1952} in which non-negative strengths are
-assigned to each competitor; without loss of generality, if the
+Bradley-Terry model~\cite{bradley1952} in which non-negative strengths
+are assigned to each competitor; without loss of generality, if the
 finishing order is $1,2,\ldots,n$ and competitor $i$ has strength
@@
 \begin{equation}\label{plackettluce}
-\prod_{i=1}^n\frac{p_i}{\sum_{j=i}^np_i}
+\prod_{i=1}^n\frac{p_i}{\sum_{j=i}^np_j}
 \end{equation}

### 2024-02-12 inst/very_simplified_likelihood.Rmd
typo
typos

@@
 ahead of $a$, so $0\leqslant r\leqslant n$.  The initial field
-strength is $a+nb=n+(n-1)a$.
+strength is $a+nb=1+b(n-1)$.
 
@@
 
+This is
+
+\[
+\frac{(B-1)\Gamma(B+n-r-1)}{\Gamma(B+n)}
+\]
+
 ## Placing last
@@
      } else {   # not last
-       out <- log(B-1) - sum(log(B + B+(n-r-1):(n-1)))
+       out <- log(B-1) - sum(log(B + (n-r-1):(n-1)))
      }
@@
 that $b_1$ and $b_2$ both have strength $b$.  Then "coming second"
-means that the finishing order was $b_1,a,b_2$ or $b_2,a,b_1$.  Noting
-that these events are disjoint, the probability would be
+means that the finishing order was $b_1\succ a\succ b_2$ or $b_2\succ
+a\succ b_1$.  Noting that these events are disjoint, the probability
+would be
 
 \[\mbox{prob of coming second}   =
-\operatorname{P}(b_1,a,b_2)      +
-\operatorname{P}(b_2,a,b_1)      =
+\operatorname{P}(b_1\succ a\succ b_2)      +
+\operatorname{P}(b_2\succ a\succ b_1)      =
 \frac{b}{a+2b}\cdot\frac{a}{a+b} +
@@
 Similarly for the focal competitor coming last we need to sum the
-probabilities of finishing orders $b_1,b_2,a$ and $b_2,b_1,a$:
+probabilities of finishing orders $b_1\succ b_2\succ a$ and $b_2\succ
+b_1\succ a$:
 
 \[\mbox{prob of coming third}=
-\operatorname{P}(b_1,b_2,a) +
-\operatorname{P}(b_2,b_1,a) =
+\operatorname{P}(b_1\succ b_2\succ a) +
+\operatorname{P}(b_2\succ b_1\succ a) =
 \frac{b}{a+2b}\cdot\frac{b}{a+b}+
@@
 \mbox{prob of coming first}      =
-\operatorname{P}(a,b_1,b_2)      +
-\operatorname{P}(a,b_2,b_1)      =
+\operatorname{P}(a\succ b_1\succ b_2)      +
+\operatorname{P}(a\succ b_2\succ b_1)      =
 \frac{b}{a+2b}\cdot\frac{b}{b+b} +
@@
 
-```{r,label=likevis,cache=TRUE}
+```{r,label=likevis,cache=F}
 plotter <- function(...){
@@
 n <- 7
-jj <- f(a,3,n,log=FALSE)
+jj <- f_vec_arn(a,3,n,log=FALSE)
 plot(a,jj/max(jj,na.rm=TRUE),type='n')
@@
 
-```{r calculatelikes,cache=TRUE}
+```{r calculatelikes,cache=F}
 a <- seq(from=0.15,to=0.5,by=0.01)
@@
 
-```{r calculatelikes1,cache=TRUE}
+```{r calculatelikes1,cache=F}
 a <- seq(from=0.005,to=0.6,by=0.005)
@@
 
-```{r,label=calcbothlikes,cache=FALSE}
+```{r,label=calcbothlikes,cache=F}
 x <- seq(from=0.1,to=0.8,by=0.01)
@@
 
-```{r,cache=TRUE}
+```{r,cache=F}
 x <- seq(from=0.1,to=0.95,by=0.01)

### 2024-02-12 inst/very_simplified_likelihood.Rmd
alternative f_single()

@@
 
+f_single2 <- function(a,r,n,log=FALSE){  # not vectorised
+  B <- 1/(1-a)
+  if(log){
+     if(r==n){  # last place
+       out <- lgamma(B)-lgamma(B+n)
+     } else {   # not last
+       out <- log(B-1) - lfactorial(B+n-1) + lfactorial(B+n-r-2)
+     }
+   } else {  # not-log 
+     if(r==n){
+      out <- gamma(B)/gamma(B+n)
+     } else {
+       out <- (B-1)/prod(B+(n-r-1):(n-1))
+     }
+   }	
+   return(out)
+}
+
 f_vec_a <- function(a,r,n, log=FALSE){  # vectorised in 'a' but not in 'r'
@@
 
-f_vec_arn <- function(a,r,n,log=FALSE){ ## vectorized in 'a'; treats 'r' and 'n' as
+f_vec_arn <- function(a,r,n,log=FALSE){
+  ## vectorized in 'a'; treats 'r' and 'n' as
   ## vectors of independent observations
@@
 
-```{r,label=calcbothlikes,cache=F}
+```{r,label=calcbothlikes,cache=TRUE}
 x <- seq(from=0.1,to=0.8,by=0.01)
@@
 
-```{r,cache=F}
+```{r,label=fishbish,cache=TRUE}
 x <- seq(from=0.1,to=0.95,by=0.01)
@@
 
-```{r}
+```{r,label=contourstrength}
 par(pty='s')
@@
 
-```{r}
+```{r,label=defsupp2}
 jj1 <- 
@@
 
-```{r}
+```{r,label=findmaxlike}
 jj2 <-
@@
 
-```{r}
+```{r,label=shochisw}
 pchisq(2*(jj1$value-jj2$objective),lower.tail=FALSE,df=1)

### 2024-02-13 inst/very_simplified_likelihood.Rmd
simplifications and corrections

@@
 \[
-\frac{(B-1)\Gamma(B+n-r-1)}{\Gamma(B+n)}
+\frac{(B-1)(B+n-r-2)!}{(B+n-1)!}
 \]
@@
 \[
-\frac{1}{(B+3)(B+2)(B+1)B}=\frac{(B-1)!}{(B+3)!}=\frac{\Gamma(B)}{\Gamma(B+4)}
+\frac{1}{(B+3)(B+2)(B+1)B}=\frac{(B-1)!}{(B+3)!}
 \]
@@
 \[
-\frac{1}{(B+n-1)(B+n-2)\cdots(B+1)B}=\frac{(B-1)!}{(B+n)!}=\frac{\Gamma(B)}{\Gamma(B+n+1)}
+\frac{1}{(B+n-1)(B+n-2)\cdots(B+1)B}=
+\frac{(B-1)!}{(B+n-1)!}
 \]
@@
 
-```{r}
-f_single <- function(a,r,n,log=FALSE){  # not vectorised
+```{r,label=definefsingle}
+f_single_old <- function(a,r,n,log=FALSE){  # not vectorised
   B <- 1/(1-a)
@@
 
-f_single2 <- function(a,r,n,log=FALSE){  # not vectorised
+f_single_old2 <- function(a,r,n,log=FALSE){  # not vectorised
   B <- 1/(1-a)
@@
      } else {   # not last
-       out <- log(B-1) - lfactorial(B+n-1) + lfactorial(B+n-r-2)
+      out <- log(B-1) - lgamma(B+n) + lgamma(B+n-r-1)
      }
@@
      }
+   }	
+   return(out)
+}
+
+f_single <- function(a,r,n,log=FALSE){  # not vectorised
+  B <- 1/(1-a)
+  if(log){
+      out <- log(B-1) - lgamma(B+n) + lgamma(B+n-r-1)
+   } else {  # not-log 
+      out <- (B-1)/prod(B+(n-r-1):(n-1))
    }	
@@
 }
+```
+
+```{r,label=defmle}
 
@@
 
-```{r}
+```{r,label=verifytwo}
 a <- 0.23423434
@@
 c(
-b^3*a/((a+9*b) * (a+8*b) * (a+7*b) * (a+6*b)),
-(B-1)/prod(B+(5:8)),
-f_vec_arn(a,r,n,log=FALSE))
+way1 = b^3*a/((a+9*b) * (a+8*b) * (a+7*b) * (a+6*b)),
+way2 = (B-1)/prod(B+(5:8)),
+way3 = f_vec_arn(a,r,n,log=FALSE)
+)
 ```
 
-```{r}
+```{r,label=makejj}
 out <- list()
@@
 
-```{r}
+```{r,label=plotjj}
 plotterp <- function(...){
@@
 ```{r calculatelikes,cache=F}
-a <- seq(from=0.15,to=0.5,by=0.01)
+a <- seq(from=0.25,to=0.65,by=0.01)
 La <- f_vec_arn(a,r=results$place-1,results$runners-1,log=TRUE)
@@
 ```{r plotlike}
-plot(a,La)
+plot(a,La,ylab='support')
 abline(h=c(0,-2))
@@
 ```{r calculatelikes1,cache=F}
-a <- seq(from=0.005,to=0.6,by=0.005)
+a <- seq(from=0.005,to=0.75,by=0.01)
 La <- f_vec_arn(a,r=results$place[1]-1,results$runners[1]-1,log=TRUE)
 La <- La - max(La,na.rm=TRUE)
-plot(a,La,type='b')
+plot(a,La,type='b',ylab='fishy')
 abline(h=c(0,-2))
@@
 f <- function(a){supp2(c(a,a),results$place,results$runners)}
-a <- seq(from=0.1,to=0.45,by=0.01)
+a <- seq(from=0.2,to=0.6,by=0.01)
 alike <- sapply(a,f)
-plot(a,alike-max(alike,na.rm=TRUE),type='b')
+plot(a,alike-max(alike,na.rm=TRUE),type='b',ylab='blong')
 abline(h=c(0,-2))
@@
 ```{r}
-a <- seq(from=0.1,by=0.01,to=0.7)
+a <- seq(from=0.2,by=0.01,to=0.8)
 wp <- category=='pure'
@@
 Lpure <- Lpure - max(Lpure,na.rm=TRUE)
-plot(a,Lpure)
+plot(a,Lpure,ylab='pure strength')
 optimize(function(a){f_vec_arn(a,rank[wp]-1,class_size[wp]-1,log=TRUE)},c(0.3,0.6),maximum=TRUE)

### 2024-02-21 inst/very_simplified_likelihood.Rmd
use hyper3()

--- 2024-02-12 inst/very_simplified_likelihood.Rmd
@@
 library("pracma")
+library("magrittr")
 ```
@@
 
+```{r usehyper3}
+a <- seq(from=0,to=1,by=0.01)
+fhyper3 <- function(r,n){
+  out <- hyper3()
+  out['a'] <- 1
+  out['b'] <- r
+  for(i in (n-r):n){
+    out[c(a=1,b=i)] %<>% dec
+  }
+  return(out)
+}
+
+plot(NA,xlim=c(0,1),ylim=c(0,1),type='n')
+M <- cbind(a=a,b=1-a)
+for(r in 0:7){
+  y <- loglik(M,fhyper3(r,n),log=FALSE)
+  y <- y/max(y, na.rm=TRUE)
+points(M[,2],y,type="l",lwd=8)
+}
+```
+
 # Formula 1.

### 2024-02-21 inst/very_simplified_likelihood.Rmd
bugfix

@@
 ```{r usehyper3}
+n <- 7
 a <- seq(from=0,to=1,by=0.01)

### 2024-02-21 inst/very_simplified_likelihood.Rmd
other bugfix

@@
   y <- y/max(y, na.rm=TRUE)
-points(M[,2],y,type="l",lwd=8)
+points(M[,1],y,type="l",lwd=8)
 }

### 2024-02-22 inst/very_simplified_likelihood.Rmd
tidyup, remove old function defs

@@
 ```{r,label=definefsingle}
-f_single_old <- function(a,r,n,log=FALSE){  # not vectorised
-  B <- 1/(1-a)
-  if(log){
-     if(r==n){  # last place
-       out <- lgamma(B)-lgamma(B+n)
-     } else {   # not last
-       out <- log(B-1) - sum(log(B + (n-r-1):(n-1)))
-     }
-   } else {  # not-log 
-     if(r==n){
-      out <- gamma(B)/gamma(B+n)
-     } else {
-       out <- (B-1)/prod(B+(n-r-1):(n-1))
-     }
-   }	
-   return(out)
-}
-
-f_single_old2 <- function(a,r,n,log=FALSE){  # not vectorised
-  B <- 1/(1-a)
-  if(log){
-     if(r==n){  # last place
-       out <- lgamma(B)-lgamma(B+n)
-     } else {   # not last
-      out <- log(B-1) - lgamma(B+n) + lgamma(B+n-r-1)
-     }
-   } else {  # not-log 
-     if(r==n){
-      out <- gamma(B)/gamma(B+n)
-     } else {
-       out <- (B-1)/prod(B+(n-r-1):(n-1))
-     }
-   }	
-   return(out)
-}
-
 f_single <- function(a,r,n,log=FALSE){  # not vectorised
@@
 MLE(3,7)
+
+MLE_analytic <- function(r,n){
+uniroot(
+  f=function(B){psigamma(B-n+r-1)-psigamma(B+n)+1/(1-B)},
+  interval=c(1.0001,1000)
+  )
+}
+
 ```
@@
 
-```{r}
+```{r focaleducation}
 rank <- c( 9, 7, 2, 3,2 ,8 ,5 ,4 ,9 )
@@
 
-```{r}
+```{r seppurefromapplied}
 a <- seq(from=0.2,by=0.01,to=0.8)
@@
 
-```{r}
+```{r dopurefurst}
 Lpure <- f_vec_arn(a,rank[wp]-1,class_size[wp]-1,log=TRUE)
@@
 
-```{r}
+```{r definesupp2ed}
 supp2_ed <- function(vec,place1,place2,runners1,runners2){

### 2024-02-22 inst/very_simplified_likelihood.Rmd
derivatives and MLE

@@
 
+```{r label=defdiff}
+fdash <- function(a,r,n,log=TRUE){
+	B <- 1/(1-a)
+	out <- (1/(B-1) - psigamma(B+n) + psigamma(B+n-r-1))*B^2
+	if(!log){out <- out * f_vec_arn(a,r,n,log=FALSE)}
+	return(out)
+}
+```
+
+Now verify `fdash()`:
+
+```{r label=verifydiff}
+a <- 0.34
+d <- 1e-3
+r <- 4
+n <- 10
+c(
+  numerical  = (f_single(a+d/2,r,n,log=TRUE)-f_single(a-d/2,r,n,log=TRUE))/d,
+  analytical = fdash(a,r,n)
+)
+
+c(
+  numerical  = (f_single(a+d/2,r,n,log=FALSE)-f_single(a-d/2,r,n,log=FALSE))/d,
+  analytical = fdash(a,r,n,log=FALSE)
+)
+```
+
+Now use the derivative to find the maximum likelihood point
+
 ```{r,label=defmle}
-
-MLE <- function(r,n){
+MLE <- function(r,n,give=FALSE){
     d <- 1e-6
-    optimize(f_vec_arn,c(d,1-d),r=r,n=n,log=TRUE,maximum=TRUE)
-}
-MLE(0,9)
-MLE(3,9)
-MLE(3,7)
-
-MLE_analytic <- function(r,n){
-uniroot(
-  f=function(B){psigamma(B-n+r-1)-psigamma(B+n)+1/(1-B)},
-  interval=c(1.0001,1000)
-  )
+     jj <- optimize(f_vec_arn,c(d,1-d),r=r,n=n,log=TRUE,maximum=TRUE)
+	 if(!give){jj <- jj$maximum}
+	 return(jj)
 }
 
+MLEa <- function(r,n,give=FALSE){
+	jj <- uniroot(
+		f=function(a){fdash(a,r,n)},
+		interval=c(0.01,0.99))
+	if(!give){jj <- jj$root}
+	return(jj)
+}
+
+rbind(
+c(MLE(1,9),MLEa(1,9)),
+c(MLE(3,9),MLEa(3,9)),
+c(MLE(3,7),MLEa(3,7))
+)
 ```
@@
     } else {
-      jj <- MLE(r,n)$maximum
+      jj <- MLE(r,n)
     }  
@@
 
-```{r usehyper3}
+```{r usehyper3,cache=TRUE}
 n <- 7
@@
 both <- c(d_ocon,d_gasly)
-bothmax <- MLE(both,22)$maximum
+bothmax <- MLE(both,22)
 
-d_ocon_max  <- MLE(d_ocon ,22)$maximum
-d_gasly_max <- MLE(d_gasly,22)$maximum
+d_ocon_max  <- MLE(d_ocon ,22)
+d_gasly_max <- MLE(d_gasly,22)
 f_vec_arn(bothmax,d_ocon ,22,log=TRUE)
@@
   both <- c(d1,d2)
-  bothmax <- MLE(both,n)$maximum
+  bothmax <- MLE(both,n)
 
-  d1max <- MLE(d1,n)$maximum
-  d2max <- MLE(d2,n)$maximum
+  d1max <- MLE(d1,n)
+  d2max <- MLE(d2,n)
   Lambda <- (

### 2024-02-22 inst/very_simplified_likelihood.Rmd
edge cases (r=0 -> p=1; r=n -> p=0

@@
 	d <- 1e-6
-     jj <- optimize(f_vec_arn,c(d,1-d),r=r,n=n,log=TRUE,maximum=TRUE)
-	 if(!give){jj <- jj$maximum}
-	 return(jj)
+	if(r==n){
+		out <- 0
+	} else if(r==0){
+		out <- 1
+	} else {
+		out <- optimize(f_vec_arn,c(d,1-d),r=r,n=n,log=TRUE,maximum=TRUE)
+		if(!give){out <- out$maximum}
+	}
+	return(out)
 }
@@
 MLEa <- function(r,n,give=FALSE){
-	jj <- uniroot(
-		f=function(a){fdash(a,r,n)},
-		interval=c(0.01,0.99))
-	if(!give){jj <- jj$root}
-	return(jj)
+	if(r==n){
+		out <- 0
+	} else if(r==0){
+		out <- 1
+	} else {
+		out <- uniroot(f=function(a){fdash(a,r,n)},interval=c(0.01,0.99))
+		if(!give){out <- out$root}
+	}
+	return(out)
 }

### 2024-02-23 inst/very_simplified_likelihood.Rmd
include Stirling

--- 2024-02-22 inst/very_simplified_likelihood.Rmd
@@
 \[
-\frac{(B-1)(B+n-r-2)!}{(B+n-1)!}
+\mathcal{L}_{r,n}(a)=\frac{(B-1)(B+n-r-2)!}{(B+n-1)!},\qquad B=1/(1-a)
 \]
+
+Then, using Stirling [viz $n!\sim n^ne^{-n}\sqrt{2\pi n}$] we get
+
+\[
+\log(B-1) + ((B+n-r-2)\log(B+n-r-2)-(B+n-r-2)+\frac{\log(B+n-r-2)}{2})
+          - ((B+n  -1)\log(B+n  -1)-(B+n  -1)+\frac{\log(B+n  -1)}{2})
+\]
+
+If you give Sage this:
+
+```
+var('B n r')
+pretty_print_default(True)
+X=log(B-1) + (
++ ((B+n-r-2)*log(B+n-r-2)-(B+n-r-2)+log(B+n-r-2)/2)    
+- ((B+n  -1)*log(B+n  -1)-(B+n  -1)+log(B+n  -1)/2)
+)
+X.taylor(n,Infinity,2)
+```
+
+It returns this:
+
+\[
+-(r+1)\log n
+-\frac{(2B-3)r-r^2+2B-2}{2n}
+-\frac{3(2B-3)r^2-6B^2-6(B^2-3B+2)r+12B-5}{12n^2}
++\log(B-1)+\mathcal{O}(n^{-3})
+\]
+
+After simplifying and taking an asymptotic expansion we find
+
+\[
+\mathcal{S}=\log\mathcal{L}_{r,n}(a)=\log(B-1)-B\frac{r+1}{n} + \frac{B^2-B(5+r^2)/2}{n^2}
++ K + \mathcal{O}\left(n^{-3}\right)
+\]
+
+(where $K$ is an arbitrary constant).  If we then neglect terms of
+order $n^{-2}$ and above, viz $\log(B-1)-B\frac{r+1}{n}$, observing
+that $\partial^2\mathcal{S}/\partial B^2=-(B-1)^{-2}<0$, the evaluate
+would be $\hat{a}=\frac{n}{n+r+1}$ [although this seems to be a poor
+approximation to the exact evaluate]
 
@@
 }
+
+fapprox <- function(a,r,n,log=FALSE){
+	if(log){
+		return(log(B-1)-B*(r+1)/n)
+	} else {
+		return(B-1 - (r+1)*exp(B)/n)
+	}
+}
 ```
@@
 	d <- 1e-6
-     jj <- optimize(f_vec_arn,c(d,1-d),r=r,n=n,log=TRUE,maximum=TRUE)
-	 if(!give){jj <- jj$maximum}
-	 return(jj)
+	out <- optimize(f_vec_arn,c(d,1-d),r=r,n=n,log=TRUE,maximum=TRUE)
+	if(!give){out <- out$maximum}
+	return(out)
 }
@@
 MLEa <- function(r,n,give=FALSE){
-	jj <- uniroot(
-		f=function(a){fdash(a,r,n)},
-		interval=c(0.01,0.99))
-	if(!give){jj <- jj$root}
-	return(jj)
+	out <- uniroot(f=function(a){fdash(a,r,n)},interval=c(0.01,0.99))
+	if(!give){out <- out$root}
+	return(out)
 }
@@
 c(MLE(3,7),MLEa(3,7))
+)
+```
+
+```{r usestirling}
+rbind(
+c(MLE(10 , 100),100 / 111),
+c(MLE(100,1000),1000/1101),
+c(MLE(500,5000),5000/5501)
+)
+
+rbind(
+c(MLEa(10 , 100),100 / 111),
+c(MLEa(100,1000),1000/1101),
+c(MLEa(500,5000),5000/5501)
 )

### 2024-02-23 inst/very_simplified_likelihood.Rmd
more work

@@
 
-After simplifying and taking an asymptotic expansion we find
+Note that if we use instead $n~\sim n^ne^{-n}\sqrt{2\pi
+n}\left(1+\frac{1}{12n}\right)$, then only terms of
+$\mathcal{O}(n^{-2})$ and above are changed; the first-order terms and
+indeed the zeroth order terms are unaffected.  Anyway, after
+simplifying and taking an asymptotic expansion to order $n^{-1}$, we
+find, to first order,
 
 \[
-\mathcal{S}=\log\mathcal{L}_{r,n}(a)=\log(B-1)-B\frac{r+1}{n} + \frac{B^2-B(5+r^2)/2}{n^2}
-+ K + \mathcal{O}\left(n^{-3}\right)
+\mathcal{S}=\log\mathcal{L}_{r,n}(a)=\log(B-1)-B\frac{r+1}{n} 
++ K + \mathcal{O}\left(n^{-2}\right)
 \]
 
-(where $K$ is an arbitrary constant).  If we then neglect terms of
-order $n^{-2}$ and above, viz $\log(B-1)-B\frac{r+1}{n}$, observing
-that $\partial^2\mathcal{S}/\partial B^2=-(B-1)^{-2}<0$, the evaluate
-would be $\hat{a}=\frac{n}{n+r+1}$ [although this seems to be a poor
-approximation to the exact evaluate]
+(where $K$ is an arbitrary constant).  Observing that
+$\partial^2\mathcal{S}/\partial B^2=-(B-1)^{-2}<0$, the evaluate would
+be unique; we have
+$\hat{B}=\frac{n+r+1}{r+1}+\mathcal{O}\left(n^{-1}\right)$, or
+alternatively $\hat{a}=\frac{n}{n+r+1}$.  I present some numerical
+results below.
 
@@
 
-(denominator has $n+1$ terms).  In R:
+(denominator has $n+1$ terms).  Observe that this agrees with the
+$r<n$ case above (substituting $r=n$).  We may introduce some R idiom:
 
@@
 
-MLEa <- function(r,n,give=FALSE){
-	out <- uniroot(f=function(a){fdash(a,r,n)},interval=c(0.01,0.99))
+MLEa <- function(r,n,give=FALSE,small=1e-4){
+	out <- uniroot(f=function(a){fdash(a,r,n)},interval=c(small,1-small))
 	if(!give){out <- out$root}
@@
 
+showdiff <- function(x,y){c(way1=x,way2=y,diff=x-y,logratio=log(x/y))}
+
 rbind(
-c(MLE(1,9),MLEa(1,9)),
-c(MLE(3,9),MLEa(3,9)),
-c(MLE(3,7),MLEa(3,7))
+showdiff(MLE(1,9),MLEa(1,9)),
+showdiff(MLE(3,9),MLEa(3,9)),
+showdiff(MLE(3,7),MLEa(3,7))
 )
@@
 rbind(
-c(MLE(10 , 100),100 / 111),
-c(MLE(100,1000),1000/1101),
-c(MLE(500,5000),5000/5501)
+showdiff(MLE(10, 100), 100 / 111),
+showdiff(MLE(10, 500), 500 / 511),
+showdiff(MLE(10,1000),1000 /1011),
+showdiff(MLE(10,5000),5000 /5011)
+)
+```
+
+```{r usestirlinga}
+rbind(
+showdiff(MLEa(10, 100), 100 / 111),
+showdiff(MLEa(10, 500), 500 / 511),
+showdiff(MLEa(10,1000),1000 /1011),
+showdiff(MLEa(10,5000),5000 /5011)
 )
 
-rbind(
-c(MLEa(10 , 100),100 / 111),
-c(MLEa(100,1000),1000/1101),
-c(MLEa(500,5000),5000/5501)
-)
 ```
@@
   points(a,y,type='l',lwd=4,col=rainbow(n+1)[r+1])
-  segments(x0=r/(n),y0=0.9,y1=1)
-  text(x=r/(n),y=0.8,r)
 }
@@
 ```{r,label=fishbish,cache=TRUE}
-x <- seq(from=0.1,to=0.95,by=0.01)
+x <- seq(from=0.1,to=0.99,by=0.01)
 jj <- as.matrix(expand.grid(x,x))

### 2024-02-23 inst/ejs-sample.tex
work

@@
 
-\begin{equation}\label{likeforrn}
-  \mathcal{L}_r(a)=\begin{cases}
+\begin{eqnarray}\label{likeforrn1}
+  \mathcal{L}_{r,n}(a) &=& 
 \frac{b}{a+ n   b}\cdot
@@
 \frac{b}{a+(n-2)b}\cdots
-\frac{a}{a+(n-r)b}\qquad &{r<n}\\
-\frac{b}{a+ n   b}\cdot
-\frac{b}{a+(n-1)b}\cdot
-\frac{b}{a+(n-2)b}\cdots
-\frac{b}{a+b}\qquad &{r=n}
-  \end{cases}
+\frac{a}{a+(n-r)b}\\ \label{likeforrn2}
+&=& \frac{B-1}{(B+n-1)(B+n-2)\cdots(B+n-r-1)}\\ \label{likeforrn3}
+&=& \frac{(B-1)(B+n-r-2)!}{(B+n-1)!}
+\end{eqnarray}
+
+\noindent where $B=1/b$.  Note carefully that these expressions are
+likelihoods, not probabilities, and do not sum to 1.  Asymptotic
+analysis of expression~\ref{likeforrn3} shows that
+
+\begin{equation}
+\log\mathcal{L}_{r,n}(a)=\log(B-1)-B\frac{r+1}{n} 
++ K + \mathcal{O}\left(n^{-2}\right)
 \end{equation}
 
-Although these expressions do have closed-form analytical forms, it is
-not helpful in numerical work because they tend to be very unstable,
-especially where $kb\simeq n$ for some integer $k$, as limiting forms
-have to be used.  Note carefully that the expressions in
-equation~\ref{likeforrn} are likelihoods, not probabilities, and do
-not sum to 1.
+Using $\mathcal{S}=\log(B-1)-B\frac{r+1}{n}$ as an approximate support
+thus incurs an error of only $\mathcal{O}(n^{-2})$.  Observing that
+$\partial^2\mathcal{S'}/\partial B^2=-(B-1)^{-2}<0$, the evaluate
+would be unique; we have
+$\hat{B}=\frac{n+r+1}{r+1}+\mathcal{O}\left(n^{-1}\right)$, or
+alternatively $\hat{a}=\frac{n}{n+r+1}$.  I present some numerical
+results below.
+
+Previous related work presented the case $n=2$ [in current notation].
+Considering the case $n=7$ gives us $7+1=8$ distinct likelihood
+functions, Figure~\ref{ninelikes}.  We see an unanticipated asymmetry
+between $r$ and $n-r$.  For $r=3$ one might expect that
+$\hat{a}=\frac{1}{2}$ on the grounds that the focal competitor's
+performance was median.  However, $\hat{a}\simeq 0.64$.
 
@@
 \end{figure}
-
-Previous related work presented the case $n=2$ [in current notation].
-Considering the case $n=7$ gives us $7+1=8$ distinct likelihood
-functions, Figure~\ref{ninelikes}.  We see an unanticipated asymmetry
-between $r$ and $n-r$, and in particular that
-$\hat{a}\geqslant\frac{r}{n}$, with equality only in the edge-cases
-$r=0$ or $n$.  For $r=3$ one might expect that $\hat{a}=\frac{1}{2}$
-on the grounds that the focal competitor's performance was median.
-However, $\hat{a}\simeq 0.514$.
 

### 2024-02-23 inst/very_simplified_likelihood.Rmd
work

@@
   points(a,y,type='l',lwd=4,col=rainbow(n+1)[r+1])
+  abline(v=0.64)
 }

### 2024-02-25 inst/very_simplified_likelihood.Rmd
logistic regression

@@
 ```{r calculatelikes1,cache=F}
-a <- seq(from=0.005,to=0.75,by=0.01)
+a <- seq(from=0.005,to=0.85,by=0.01)
 La <- f_vec_arn(a,r=results$place[1]-1,results$runners[1]-1,log=TRUE)
@@
 pchisq(2*(7381.496-7380.314),df=1,lower.tail=FALSE)
+```
+
+Above shows that there is no evidence to suggest that the Stirling
+strength differs from the Auckland strength.  Now let's try a
+logistic-style regression:
+
+```{r logisticsupportparkrun,cache=FALSE}
+results <- read.table("parkrun_results.txt",header=TRUE)
+results
+supp_logistic <- function(vec){
+  alpha <- vec[1]
+  beta <- vec[2]
+  out <- 0
+  for(i in seq_len(nrow(results))){
+     LO <- alpha + beta*i
+     strength <- exp(LO)/(1+exp(LO))
+     out <- out + f_vec_arn(strength,results$place[i]-1,results$runners[i]-1,log=TRUE)
+  }
+  return(out+13400)
+}
+
+c(
+supp_logistic(c(0,0)),
+supp_logistic(c(0,0.01)),
+supp_logistic(c(0,-0.01))
+)
+
+```
+
+```{r optimlogist, cache=FALSE}
+optim(par=c(0,0),fn=supp_logistic,control=list(fnscale=-1))
+optim(par=c(0,0),fn=function(v){supp_logistic(vec=c(v[1],0))},control=list(fnscale=-1))
 ```

### 2024-02-28 inst/very_simplified_likelihood.Rmd
in silico formula 1

@@
 
-## Placing last
+# Placing last
 
@@
 
-## Note on the differences between probability and likelihood
+# Note on the differences between probability and likelihood
 
@@
 
-# Formula 1.
+# Formula 1: Ocon vs Gasly
 
@@
 
-## Mathematical Olympiad
+# Formula 1: Formula 1 drivers: career arc
+
+```{r getnumgetfoc}
+readstring <- function(year){read.table(paste("formula1_",year,".txt",sep=""))}
+getfoc <- function(year,comp="Perez"){  # get focal competitor
+  M <- as.matrix(readstring(year))
+  out <- suppressWarnings(as.numeric(M[comp,seq_len(ncol(M)-1)]))
+  out[is.na(out)] <- nrow(M)
+  return(out)
+}
+getnum <- function(year){nrow(readstring(year))} # number of competitors
+
+perez <- lapply(2011:2023,getfoc)
+print(perez)
+y <- unlist(lapply(seq_along(perez),function(i){mean(perez[[i]])}))
+x <- 2011:2023
+plot(x,y)
+summary(lm(y~x))
+```
+
+```{r checofirst}
+checo_like <- function(a){
+  out <- a*0
+  for(year in 2011:2023){ out <- out + f_vec_arn(a,getfoc(year)-1,getnum(year)-1,log=TRUE) }
+  return(out)
+}
+
+a <- seq(from=0.45,to=0.62,by=0.01)
+cL <- checo_like(a)
+cL <- cL - max(cL)
+plot(a,cL,type='b')
+abline(h=c(0,-2))
+```
+
+```{r definef1log}
+f1_logistic <- function(vec){
+  alpha <- vec[1]
+  beta  <- vec[2]
+  out <- 0
+  for(year in 2011:2023){
+     LO <- alpha + beta*(year-2011)
+     strength <- exp(LO)/(1+exp(LO))
+     out <- out + f_vec_arn(strength,getfoc(year)-1,getnum(year)-1,log=TRUE)
+  }
+  return(out)
+}
+```
+
+```{r calculatelogisticcontou,cache=TRUE}
+a <- seq(from=-0.8,to=0.0,by=0.01)
+b <- seq(from=0,to=0.2,by=0.01)
+jj <- as.matrix(expand.grid(a,b))
+L <- apply(jj,1,f1_logistic)
+L <- L-max(L)
+L <- matrix(L,length(a),length(b))
+```
+
+```{r showfilledc}
+L <- pmax(L,-40)
+contour(a,b,L)
+```
+
+```{r uselogf1}
+f1_logistic(c(0,0))
+f1_logistic(c(0,0.0001))
+f1_logistic(c(0.001,0))
+```
+
+```{r optimizelof1,cache=TRUE}
+o <- optim(c(-0.2,0.1),fn=f1_logistic, control=list(fnscale = -1),hessian=TRUE)
+```
+
+```{r showjj}
+o
+o$par
+o$hessian
+eigen(o$hessian)
+best <- o$par
+f1_logistic(best)
+
+jj <- best
+jj[1] <- jj[1]*1.01
+f1_logistic(jj) - f1_logistic(best)
+
+jj <- best
+jj[1] <- jj[1]*0.99
+f1_logistic(jj) - f1_logistic(best)
+
+jj <- best
+jj[2] <- jj[2]*1.00001
+f1_logistic(jj) - f1_logistic(best)
+
+jj <- best
+jj[2] <- jj[2]*0.99
+f1_logistic(jj) - f1_logistic(best)
+```
+
+```{r testlogf1,cache=TRUE}
+o_free <- optim(c(-0.2,0.1),fn=f1_logistic, control=list(fnscale = -1),hessian=TRUE)
+o_constrained <- optim(c(-0.2,0.1),fn=function(v){
+v[2] <- 0
+return(f1_logistic(v))}, control=list(fnscale = -1),hessian=TRUE)
+```
+
+```{r showpvalue}
+pchisq(2*(o_free$value - o_constrained$value),df=1,lower.tail=FALSE)
+```
+
+# in silico formula
+
+```{r insn}
+table(replicate(100,which(rrace3(pn=c(a=1,b=10),ps=c(a=0.9,b=0.1))=='a')))
+```
+
+```{r insilicoestmation,cache=TRUE}
+n <- 1000
+M <- matrix(0,n,4)
+alpha_true <- 4
+beta_true <- -6
+for(i in seq_len(n)){
+    x <- i/n
+    LO <- alpha_true + beta_true*x
+    p <- exp(LO)/(1+exp(LO))
+    nn <- round(runif(1,10,100))
+    M[i,1] <- x  # regressor
+    M[i,2] <- which(rrace3(pn=c(a=1,b=nn),ps=c(a=p,b=1-p))=='a')
+    M[i,3] <- nn+1
+    M[i,4] <- p
+}
+head(M)
+tail(M)
+```
+
+```{r testran,cache=FALSE}
+runif(10)
+```
+
+```{r definelikeinsilico}
+likeinsilico <- function(vec,M){
+  alpha <- vec[1]
+  beta  <- vec[2]
+  S <- 0
+  for(i in seq_len(nrow(M))){
+    x <- M[i,1]
+    LO <- alpha + beta * x
+    p <- exp(LO)/(1+exp(LO))
+    S <- S + f_vec_arn(p,M[i,2]-1,M[i,3]-1,log=TRUE)
+  }
+return(S)
+}
+```
+
+```{r testlikeininsilicco,cache=TRUE}
+likeinsilico(c(4,-6),M)
+likeinsilico(c(4,-6.5),M)
+likeinsilico(c(4,-5.5),M)
+```
+
+```{r optimize_insilicco,cache=TRUE}
+optim(par=c(4,-6),likeinsilico,control=list(fnscale= -1),M=M)
+```
+
+```{r makeZcontourdataset,cache=TRUE}
+a <- seq(from=3.8,to=4.2,by=0.1)
+b <- seq(from=-6.1,to=-5.6,by=0.1)
+Z <- as.matrix(expand.grid(a,b))
+Z <- apply(Z,1,function(x){likeinsilico(x,M=M)})
+Z <- matrix(Z,length(a),length(b))
+```
+
+```{r doZcontour}
+Z <- Z-max(Z)
+contour(a,b,Z,levels=-c(0:10,15,20,30))
+points(alpha_true,beta_true,pch=16,cex=4)
+grid()
+persp(a,b,Z)
+```
+
+# Mathematical Olympiad
 
@@
 
-## Parkrun
+# Parkrun
 

### 2024-02-28 inst/ejs-sample.tex
talk about Checo

@@
 
-\begin{equation}
+\begin{equation}\label{asymptotic}
 \log\mathcal{L}_{r,n}(a)=\log(B-1)-B\frac{r+1}{n} 
@@
 competition winner is the competitor who accumulates the most points
-after the final race.  However, as argued by Hankin~\cite{hankin2023},
-the drivers' {\em ranks} are statistically sufficient for analysis of
+after the final race.  However, as argued by
+Hankin~\cite{hankin2023_formula1points}, the drivers' {\em ranks} are
+statistically sufficient for analysis using Plackett-Luce approaches.
+The methods discussed above may be used to assess Bradley-Terry
+strengths of Formula 1 racing driver Sergio P\'{e}rez (``Checo'') over
+his career from 2011 to 2023.  Checo's Formula 1 performance began
+somewhat inauspiciously in 2011 as second driver for Sauber for whom
+he finished no higher than P7, but his performance rose subsequently,
+culminating in his achievement of a startling 20 podium finishes in
+2022-3.  Noting that Checo raced against 73 distinct competitors [this
+  rendering estimation of the other players' Plackett-Luce strengths
+  effectively impossible], we nevertheless are in a position to
+estimate his career-span Bradley-Terry strength and estimate it as
 
-The methods discussed above may be used to assess Bradley-Terry
-strengths of Formula 1 racing drivers.  Let us consider 2018 season as
-an example and first consider Lewis Hamilton.
+we consider the following inference
+problem:
 
-Lewis Hamilton is a
+\begin{equation}
+  b = b(t) = \frac{e^{\alpha + \beta t}}{1+e^{\alpha + \beta t}},\qquad\alpha,\beta\in\mathbb{R}
+\end{equation}
+
+(that is, a logistic regression of generalizeBradley-Terry strength
+$b$ against time).
+
+There seems to be no easy analytical solution to
+this, although Equation~\ref{asymptotic} would give
+\begin{equation}
+  \label{asymptoticcheco}
+  \log\mathcal{L}(B)=m\log(B-1) -B\left(\frac{r_1+1}{n_1}+\ldots+\frac{r_m+1}{n_m}\right)
+\end{equation}
+
+over the whole $m$ seasons, where $r_i$ is Checo's rank in race
 

### 2024-03-01 inst/very_simplified_likelihood.Rmd
sensible park run null

@@
 author: "Robin Hankin"
-output: html_document
+output: bookdown::html_document2
 ---
@@
 
-```{r testran,cache=FALSE}
-runif(10)
-```
-
 ```{r definelikeinsilico}
@@
 ```{r optimize_insilicco,cache=TRUE}
-optim(par=c(4,-6),likeinsilico,control=list(fnscale= -1),M=M)
+optim_ab <- optim(par=c(4,-6),likeinsilico,control=list(fnscale= -1),M=M)
+optim_ab
 ```
@@
 ```{r makeZcontourdataset,cache=TRUE}
-a <- seq(from=3.8,to=4.2,by=0.1)
-b <- seq(from=-6.1,to=-5.6,by=0.1)
+a <- seq(from=3.8,to=4.2,by=0.01)
+b <- seq(from=-6.4,to=-5.6,by=0.01)
 Z <- as.matrix(expand.grid(a,b))
@@
 Z <- Z-max(Z)
+Z <- pmax(Z,-10)
 contour(a,b,Z,levels=-c(0:10,15,20,30))
 points(alpha_true,beta_true,pch=16,cex=4)
+jj <- optim_ab$par
+points(jj[1],jj[2],pch=4,cex=4,col='red')
 grid()
@@
 
-```{r definelikelihood}
-
+```{r definesupportparkrun}
 L <- function(a,results){
@@
 
-```{r plotlike}
-plot(a,La,ylab='support')
-abline(h=c(0,-2))
+```{r optimsinglelike}
+(optparkrun <- optimize(function(a){L(a,results)},c(0.2,0.8),maximum=TRUE))
+f <- function(a){L(a,results)-2}
 ```
 
-```{r optimsinglelike}
-optimize(function(a){L(a,results)},c(0.2,0.4),maximum=TRUE)
+```{r plotparkrunlike, fig.cap="Support curve for author's generalized Bradley-Terry strength.  Evaluate of $\\hat{a}\\simeq 0.448$ has a support of zero"}
+plot(a,La,xlab="generalized Bradley-Terry strength",ylab="support")
+abline(h=c(0,-2))
+abline(v=0.5)
 ```
+
+Figure \@ref(fig:plotparkrunlike) shows a support curve for the
+author's generalized Bradley-Terry strength, normalized so that
+$\mathcal{S}(\hat{a})=0$.  We see a credible interval [that is, values
+of $a$ for which $\mathcal{S}(a)\geqslant -2$] of about
 

### 2024-03-01 inst/ejs-sample.tex
start to write up for ejs

--- 2024-02-23 inst/ejs-sample.tex
@@
 \begin{abstract}
-
 Here a simplification of the Plackett-Luce likelihood functions for
@@
 focal competitor.  We assign Placket-Luce strength of $a$ to the focal
-competitor and $b$ to all others; we require $a+b=1$.  This device
-furnishes a reasonable, intuitive, and computationally tractable
-likelihood function that I apply to a range of applications including
-Formula 1 motor racing and the International Mathematical Olympiad.
-The results are difficult to obtain with other methods.
+competitor and $b$ to all others; we require $a+b=1$.  The associated
+inference problem appears to have a wide range of applications and I
+present some analysis of datasets drawn from Formula 1 motor racing
+and community initiative Parkrun.
 \end{abstract}
@@
 
-\begin{equation}
+\begin{equation}\label{asymptotic}
 \log\mathcal{L}_{r,n}(a)=\log(B-1)-B\frac{r+1}{n}
@@
 $\hat{B}=\frac{n+r+1}{r+1}+\mathcal{O}\left(n^{-1}\right)$, or
-alternatively $\hat{a}=\frac{n}{n+r+1}$.  I present some numerical
-results below.
+alternatively $\hat{a}=\frac{n}{n+r+1}$.  For repeated observations
+$r_i,n_i$, $1\leqslant i\leqslant m$ we would have
+\begin{equation}
+  \hat{a} = 1-
+  \frac{\frac{r_1+1}{n_1}+\cdots+\frac{r_m+1}{n_m}}{m}
+  = 1-\overline{\left(\frac{r+1}{n}\right)}.
+\end{equation}
+
+(that is, one minus the mean ratio of rank to number of [cloned]
+competitors).  I present some numerical results below.
 
@@
 \begin{figure}[t]
-\includegraphics[width=4in]{ninelikes}  % produced by very_simplified_likelihood.Rmd
+\includegraphics[width=4in]{ninelikes}  % ninelikes.pdf produced by very_simplified_likelihood.Rmd
 \caption{Likelihood functions for $r=0(1)8$\label{ninelikes}}
@@
 
+Here I present three examples that showcase likelihood
+function~\ref{likeforrn3}.  Formula 1 education, parkrun
+
+\subsection{Park run}
+
+``Parkrun'' is a distributed community initiative that organises
+weekly timed 5 km runs/walks in parks worldwide~\cite{hindley2020}.  A
+typical event will have 200 participants.  Table~\ref{parkruntable}
+shows the author has completed a total of 15 parkruns to date, and
+from the first pair of numbers we see that 238 runners attended that
+particular parkrun, of whom the author placed 173.
+
+\begin{table*}[t]
+  \caption{Parkrun results}
+\label{parkruntable}
+\begin{tabular}{ccccccccccc}
+\hline
+rank   &173  & 165 & 172 & 199 & 181 & 229 & 177 & 222 & 206 & 142 \\
+runners&238  & 238 & 196 & 242 & 242 & 318 & 259 & 305 & 297 & 241 \\ \\ \\
+rank   & 118 & 224 & 128 & 115 & 183 &     &     &     &     &\\
+runners& 179 & 338 & 203 & 245 & 254 &     &     &     &     &\\
+\hline
+\end{tabular}
+\end{table*}
+
+We may consider the author to have an unknown generalized Bradley
+Terry strength $a$: Figure~\ref{parkrunsupport} shows a support curve
+for $a$ with the evaluate, $\hat{a}\simeq 0.448$ shown.  We also see
+two units of support~\cite{edwards1992} shown as a dotted line
+illustrating a support interval of about 0.315 - 0.567; we may be
+reasonably confident that the author's true strength lies in this
+range.  In particular, note that a reasonable $H_0\colon
+a=\frac{1}{2}$ may not be rejected, the support for $H_0$ being only
+susceptible to very minor improvement [$\simeq 0.351$] by moving to
+the evaluate.
+
+\begin{figure}[t]
+\includegraphics[width=4in]{plotparkrun}  % plotparkrun.pdf made in very_simplified_likelihood.Rmd
+\caption{Author's strength\label{parkrunsupport}}
+\end{figure}
+
+\subsection{Education}
+
+One issue encountered in educational research is that of
+confidentiality of examination scores, and individual students'
+results are generally regarded as sensitive and private information.
+Nonetheless, it is common for students to possess two pertinent and
+informative observations: the total enrollment in a specific class and
+their individual ranking within that class; table~\ref{educationtable}
+shows an apocryphal dataset.  What could a student infer from such a
+dataset?
+
+\begin{table*}[t]
+  \caption{Educational dataset}
+\label{educationtable}
+\begin{tabular}{llll}
+\hline
+                   course&category   &rank&class size\\ \hline
+      rings and modules  &  pure     & 9  &      12\\
+           group theory  &  pure     & 7  &      17\\
+               calculus  & applied   & 2  &      23\\
+         linear algebra  & applied   & 3  &       9\\
+ differential equations  & applied   & 2  &      13\\
+               topology  &   pure    & 8  &      14\\
+     special relativity  & applied   & 5  &      13\\
+        fluid mechanics  & applied   & 4  &      12\\
+            Lie algebra  &   pure    & 9  &      15\\
+     \hline
+\end{tabular}
+\end{table*}
+
+One natural question would be to wonder whether the student was bettar
+at pure or applied mathematics and, if so, to quantify the difference.
+Figure~\ref{plotpureandapplied} shows contours of support [again
+  normalised so the support of the evaluate is zero] for the dataset,
+as a function of pure strength on the horizontal axis and applied
+strength on the vertical axis.  A null of equal strengths appears as a
+diagonal line.  We see that the maximum likelihood estimate is
+$\hat{p}_\mathrm{pure}=0.567$, $\hat{p}_\mathrm{applied}=0.887$.
+However, a constrained optimization shows that the maximum support
+along the null diagonal is about $-2.43$.  Edwards's two units of
+support criterion is met, and we are justified in concluding that the
+student is indeed more competent at applied than pure mathematics.
+Alternatively, we may observe that $2\mathcal{S}=4.86$ is in the tail
+region of its asymptotic $\chi^2_1$ distribution, corresponding to a
+$p$-value of about $0.028$.
+
+\begin{figure}
+\includegraphics[width=4in]{plotpureandapplied}  % plotpureandapplied.pdf produced by very_simplified_likelihood.Rmd
+\caption{Contours of equal support for pure (horizontal) and applied
+  (vertical) strength; null of equal strengths shown as a diagonal
+  line  \label{plotpureandapplied}}
+\end{figure}
+
+\subsection{Formula 1}
+
 Formula 1 motor racing is an important and prestigious motor sport
@@
 competition winner is the competitor who accumulates the most points
-after the final race.  However, as argued by Hankin~\cite{hankin2023},
-the drivers' {\em ranks} are statistically sufficient for analysis of
+after the final race.  However, as argued by
+Hankin~\cite{hankin2023_formula1points}, the drivers' {\em ranks} are
+statistically sufficient for analysis using Plackett-Luce approaches.
+The methods presented here may be used to assess competitive strengths
+of Formula 1 racing driver Sergio P\'{e}rez (``Checo'') over his
+career from 2011 to 2023.  Checo's Formula 1 performance began
+somewhat inauspiciously as second driver for Sauber for whom he
+finished no higher than P7 in 2011, but his performance rose
+subsequently, culminating in his achievement of a startling 20 podium
+finishes in 2022-3.  Noting that Checo raced against 73 distinct
+competitors (this rendering estimation of the other players'
+Plackett-Luce strengths effectively impossible~\cite{hankin2020}), we
+are nevertheless in a position to consider competitive strength using
+the modified likelihood approach presented here.  We consider the
+following inference problem:
 
-The methods discussed above may be used to assess Bradley-Terry
-strengths of Formula 1 racing drivers.  Let us consider 2018 season as
-an example and first consider Lewis Hamilton.
+\begin{equation}
+  a = a(t) = \frac{e^{\alpha + \beta t}}{1+e^{\alpha + \beta t}},\qquad\alpha,\beta\in\mathbb{R}
+\end{equation}
 
-Lewis Hamilton is a
+(that is, a logistic regression of generalized Bradley-Terry strength
+$a$ against time, with $\alpha$ and $\beta$ being linear
+coefficients).  Contours of equal support are shown in
+Figure~\ref{showchecolike}, again normalized so that
+$\mathcal{S}(\hat{\alpha},\hat{\beta})=0$.  We see the expected
+elliptical contours, and further $H_0\colon\beta=0$ may be rejected in
+favour of $\alpha=-0.27$, $\beta=0.081$, on the grounds that
+$\mathcal{S}(\alpha,0)\leqslant 9.2$.  Alternatively we may observe
+that $2\mathcal{S}=18.4$ would correspond to a $p$-value of about
+$1.7\times 10^{-5}$ on its asymptotic $\chi^2_1$ null distribution.
 
-Pierre Gasly and Esteban Ocon have a long-standing rivalry and
-comparisons between them typically compare disparate metrics such as
-number of podiums, and highest grid position in a season.
+\begin{figure}[t]
+\includegraphics[width=4in]{showchecolike}  % showchecolike.pdf produced by very_simplified_likelihood.Rmd
+\caption{Likelihood functions for $\alpha,\beta$ for Checo's racing
+  career 2011-2023\label{showchecolike}}
+\end{figure}
 
-\begin{acks}[Acknowledgments]
-The authors would like to thank the anonymous referees, an Associate
-Editor and the Editor for their constructive comments that improved the
-quality of this paper.
-\end{acks}
+\section{Conclusions and further work}
+
+Here a simplification of the Plackett-Luce likelihood functions for
+order statistics was proposed, some simple properties obtained, and
+inference problems from three disciplines given.  The results seem to
+be difficult to obtain any other way.  The assumptions furnish a
+reasonable, intuitive, and computationally tractable likelihood
+function with wide applicability.  Further work might include the
+analysis of more than one focal competitor.
+
+% \begin{acks}[Acknowledgments]
+% The authors would like to thank the anonymous referees, an Associate
+% Editor and the Editor for their constructive comments that improved the
+% quality of this paper.
+% \end{acks}
 

### 2024-03-01 inst/very_simplified_likelihood.Rmd
start to write up for ejs

@@
 L <- pmax(L,-40)
+showchec <- function(...){
+contour(a,b,L,xlab=expression(alpha),ylab=expression(beta))
+}
+showchec()
+pdf(file="showchecolike.pdf")
 contour(a,b,L)
+dev.off()
 ```
@@
 ```{r showpvalue}
+o_free
+o_constrained
 pchisq(2*(o_free$value - o_constrained$value),df=1,lower.tail=FALSE)
@@
 
-```{r definesupportparkrun}
-L <- function(a,results){
-   out <- a*0
-   for(i in seq_len(nrow(results))){
-   jj <- f_vec_a(a,r=results$place[i]-1,results$runners[i]-1,log=TRUE)
-   out <- out + jj
-   }
-   out[is.infinite(out)] <- NA
-   return(out)
-}
+```{r optimsinglelike,cache=TRUE}
 ```
 
-```{r calculatelikes,cache=F}
-a <- seq(from=0.25,to=0.65,by=0.01)
-La <- f_vec_arn(a,r=results$place-1,results$runners-1,log=TRUE)
-La <- La - max(La,na.rm=TRUE)
+```{r calculate_parksupp,cache=TRUE}
+S_parkrun <- function(a){f_vec_arn(a,r=results$place-1,results$runners-1,log=TRUE)}
+(optparkrun <- optimize(S_parkrun, c(0.2,0.8),maximum=TRUE))
+
 ```
 
-```{r optimsinglelike}
-(optparkrun <- optimize(function(a){L(a,results)},c(0.2,0.8),maximum=TRUE))
-f <- function(a){L(a,results)-2}
+```{r calcparka,cache=TRUE}
+a <- seq(from=0.25,to=0.65,by=0.01)
+S_parkrun_a <- S_parkrun(a) - optparkrun$objective
+uniroot(function(a){S_parkrun(a) - optparkrun$objective+2},c(0.1,0.5))$root
+uniroot(function(a){S_parkrun(a) - optparkrun$objective+2},c(0.5,0.9))$root
+optparkrun$objective - S_parkrun(0.5)
 ```
 
-```{r plotparkrunlike, fig.cap="Support curve for author's generalized Bradley-Terry strength.  Evaluate of $\\hat{a}\\simeq 0.448$ has a support of zero"}
-plot(a,La,xlab="generalized Bradley-Terry strength",ylab="support")
-abline(h=c(0,-2))
-abline(v=0.5)
+```{r plotparkrunlike, fig.cap="Support curve for author's generalized Bradley-Terry strength.
+Curve is normalised so that the evaluate of $\\hat{a}\\simeq 0.448$ has a support of zero.
+Dotted line shows $\\mathcal{S}=-2$; intersection of this with the  curve shows that the support interval runs from about 0.315 to 0.567"}
+plotparkrun <- function(...){
+plot(a,S_parkrun_a,xlab="generalized Bradley-Terry strength",ylab="support",type='b',ylim=c(-5,0))
+abline(h=c(0,-2),lty=2)
+segments(x0=0.5,y0=-1.5,y1=0)
+segments(x0=optparkrun$maximum,y0=-1,y1=0)
+text(optparkrun$maximum,-0.8,expression(hat(a)==0.448),pos=2)
+text(0.5,-1.5,expression(H[o]:a==0.5),pos=2)
+}
+plotparkrun()
+pdf("plotparkrun.pdf")
+plotparkrun()
+dev.off()
 ```
@@
 author's generalized Bradley-Terry strength, normalized so that
-$\mathcal{S}(\hat{a})=0$.  We see a credible interval [that is, values
-of $a$ for which $\mathcal{S}(a)\geqslant -2$] of about
+$\mathcal{S}(\hat{a})=0$.  We see a support interval [that is,
+$\left\lbrace a\colon\mathcal{S}(a)\geqslant -2\right\rbrace$] of
+about 0.315 to 0.567.  We might also observe that the support for
+$H_0\colon a=\frac{1}{2}$ [that is, that the author has a 50\% chance
+of running faster than a randomly chosen Parkrun competitor] is about
+0.35, less than Edwards's two units of support criterion.  We thus see
+some support for Hankin's assertion that he is "as fit as anyone
+else there".
 
@@
 ```{r,label=contourstrength}
+plotpureandapplied <- function(...){
 par(pty='s')
@@
 abline(0,1)
+}
+plotpureandapplied()
+pdf(file="plotpureandapplied.pdf")
+plotpureandapplied()
+dev.off()
 ```

### 2024-03-04 inst/ejs-sample.tex
rewording

@@
 One issue encountered in educational research is that of
-confidentiality of examination scores, and individual students'
-results are generally regarded as sensitive and private information.
-Nonetheless, it is common for students to possess two pertinent and
-informative observations: the total enrollment in a specific class and
-their individual ranking within that class; table~\ref{educationtable}
-shows an apocryphal dataset.  What could a student infer from such a
-dataset?
+confidentiality of examination scores, and individual results are
+generally regarded as sensitive and private information.
+Nevertheless, it is common a student to possess some informative
+observations: the total enrollment in a specific class and their
+individual ranking within that class;  table~\ref{educationtable}
+shows an apocryphal dataset.   What could a student infer from such
+data?
 
@@
 
-One natural question would be to wonder whether the student was bettar
+One natural question would be to wonder whether the student was better
 at pure or applied mathematics and, if so, to quantify the difference.

### 2024-03-04 inst/race3.Rmd

@@
 Looking at the first column we see that the constructors are Mercedes
-("`Merc`"), Red Bull Racing Honda ("`RBRH`"), and so on (see file
-`inst/constructor_names.txt` for a list of abbreviations).  Each
-constructor fields two cars in each race.  Looking at the next column,
-we see the order statistic for Bahrain (`BHR`).  We see that Mercedes
-came first and third, Red Bull came second and fifth, Ferrari sixth
-and eighth, and so on.
+RedBull, and so on (file `inst/constructor_names.txt` summarises the
+constructors' lifespans).  Each constructor fields two cars in each
+race.  Looking at the next column, we see the order statistic for
+Bahrain (`BHR`).  We see that Mercedes came first and third, RedBull
+came second and fifth, Ferrari sixth and eighth, and so on.
 
@@
 Further, we may test the hypothesis that the big four (viz Mercedes,
-Ferrari, Red Bull Racing Honda, and McLaren-Mercedes) are equal:
+Ferrari, RedBull, and McLaren) are equal:
 
 ```{r,label=testsameconstructor,cache=TRUE}
-samep.test(constructor_2021,c("Merc","Ferrari","RBRH","MM")) # ~10 mins
+samep.test(constructor_2021,c("Mercedes","Ferrari","RedBull","McLaren")) # ~10 mins
 ```
@@
 ```{r,label=testsameconstructorlowscores,cache=TRUE}
-samep.test(constructor_2021,c("ATH","ARRF","AR","AMM")) # ~ 10 mins
+samep.test(constructor_2021,c("Alpine","AlphaTauri","AstonMartin")) # ~ 10 mins
 ```
@@
 
-```{r,label=changenames,cache=TRUE}
-constructor_2020_and_2021 <- psubs(constructor_2020,"Merc","Merc2020") + psubs(constructor_2021,"Merc","Merc2021")
+```{r,label=changenames,cache=FALSE}
+constructor_2020_and_2021 <- psubs(constructor_2020,"Mercedes","Merc2020") + psubs(constructor_2021,"Mercedes","Merc2021")
 ```

### 2024-03-05 inst/skating.Rmd
typo

@@
 Slutskaya's one) have cost her more likelihood-based strength than her
-extra "1" gave her; note also that the two judges who placed Highes
+extra "1" gave her; note also that the two judges who placed Hughes
 fourth (viz `J2` and `J4`) placed Slutskaya first.  Also:

### 2024-03-08 inst/vsl_masterchef.Rmd
work on very simplified likelihood

---
title: "Very simplified likelihood: masterchef"
author: "Robin Hankin"
output: bookdown::html_document2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("pracma")
library("magrittr")
```

```{r hexsticker, out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
```

This follows on from `vsl_teams.Rmd` and `vsl_partial_ranks.Rmd`.  But
the application of those ideas to _MasterChef_ is flawed: as the
competition progresses, Brent and Laura face increasingly skilled
competitors, for the weakest ones have been eliminated.  And it is not
clear to what extent that observation interferes with the assumtion of
constant non-focal competitor strength.

```{r,label=define_f_range_setup,include=FALSE}
f_single_old <- function(a,r,n){  # not vectorised
	B <- 1/(1-a)
    (B-1)/prod(B+(n-r-1):(n-1)) * exp( lfactorial(n)-lfactorial(n-r))
}

f_single <- function(a,r,n){
	B <- 1/(1-a)
    exp(log(B-1)-lgamma(B+n)+lgamma(B+n-r-1)+lfactorial(n)-lfactorial(n-r))
}
f_vec_a <- function(a,r,n){  # vectorised in 'a' but not in 'r'
	sapply(a,function(a){f_single(a,r=r,n=n)})
}

```

```{r label=show_f_range}
L_range <- function(a,possible,n){ # r1=4 -> came fifth [four b clones ahead]  
    out <- a*0
    for(i in (possible)){
        out <- out + f_vec_a(a,i-1,n-1) }
    return(out)
}
```

Two teams of size $m$:

```{r}
`L_win` <- function(a,m){
    b <- 1-a
    ((m-2)*b+1)/((2*m-2)*b+1)}

`L_lose` <- function(a,m){1-L_win(a,m)}
```

# Laura in  Australian MasterChef series 6

* week 1
  +  top 3 out of 25 but not the winner: (2:3)/24

* week 2
  +  not top 3, not bottom 3: (4:20)/23
  +  team lose (11 v 11)
  +  not bottom 6: (1:5)/11

* week 3
  +  top 3 out of 21: (1:3)/21
  +  team win (10 v 10)

* week 4
  +  not top 3, not bottom 3:  (4:16)/19
  +  team win (9 v 9)

* week 5
  +  not top 3, not bottom 4:  (4:13)/17
  +  4 teams of 4; Laura came in second team.

* week 6
  + bottom 3, not eliminated: (13:14)/15
  + two teams; Laura came in top team

* week 7
  + top3: (1:3) /14 
  + 2 teams of 6, losing team

* week 8
  + bottom 2, not eliminated, (13/15)
  + 2 teams of 5, losing team

* week 9
  + not bottom 5 out of 11 (1:6)/11
  + two teams of 5 (again); losing team
  + of the 5 in the losing team, came 4th

* week 10
  + top 3 out of 9
  + 4 teams of 2, Laura came in top team

* week 11
  + top 3 out of 7

* Finals
  + bottom 3 out of 6     (4:6)/6
  + not eliminated out of 3
  + top 3 out of 5   (1:3)/5
  + top out of 4  (1)/4
  + top 2 out of 3 (1:2)/3
  + not the winner

# Brent in  Australian MasterChef series 6

* Week 1
  + Bottom 9 out of 24 (16:24)/24
  + elimination out of 9, not in bottom 3 (1:6)/9

* Week 2
  + out of 23, not in top 3, not in bottom 3 (4:20)/23
  + team win; two teams of 11
  + out of 11, in the bottom 6 (6:11)/11
  + out of 6, not in bottom 3 (1:3)/6

* week 3
  + out of 21, not bottom 3, not top 3 (4:18)/21
  + bottom 3 out of 18 (16:18)/18
  + two teams of 10, losing team
  + out of 10, bottom 6 (5:10)/10
  + out of 6, in top 3 (1:3)/6

* week 4
  + out of 19, in top 3, (1:3)/19
  + out of 18, in bottom 3 (16:18)/18
  + out of 3, did not get eliminated (1:2)/3

* Week 5
  + out of 17, top 3 (1:3)/17
  + four teams of 4, in losing team
  + out of 4, did not get eliminaed (1:3)/4

* week 6
  + out of 15, not in bottom 3, not in top 3 (4:12)/15
  + two teams of 7, in losing team
  + out of 7, not in bottom 3 (1:4)/7

* week 7
  + out of 13, not in top 3 not in bottom 3 (4:10)/13
  + two teams of 6, losing team
  + out of 6, bottom 2 (5:6)/6
  + out of 2, not eliminated (1)/2

* week 8 
  + (2:4)/11
  + 2 teams of 5, winning team

* week 9
  + bottom 5, (7:11)/11
  + out of 5, not bottom 3 (1:2)/5
  + 2 teams of 5, losing team
  + out of 5, not bottom 2 (1:3)/5

* week 10
  + out of 9, not in top 3 or bottom 4 (4:5)/9
  + 4 teams of 2, in third or fourth team (?)
  + out of 4, not eliminted (1:3)/4

* week 11
  + out of 7, bottom 4 (4:7)/7
  + out of 4, not bottom 2 (1:2)/4

* Finals
  + out of 6, top 3 (1:3)/6
  + out of 5, bottom 2 (4:5)/5
  + out of 2, not eliminated (1)/2
  + out of 4, (2:3)/4
  + out of 3, not eliminated (1:2)/3
  + out of 3, not eliminatged (1:2)/3
  + overall winner (1)/3

```{r label=defLauraLike}
L_laura <- function(a){
    b <- 1-a
    out <- a*0 + 1
    out <- out * L_range(a,2:3,24)    # week 1
    out <- out * L_range(a,4:20,24)   # week 2
    out <- out * L_lose(a,11)         # week 2
    out <- out * L_range(a,1:5,11)    # week 2
    out <- out * L_range(a,1:3,21)    # week 3
    out <- out * L_win(a,10)          # week 3
    out <- out * L_range(a,4:16,19)   # week 4
    out <- out * L_win(a,9)           # week 4
    out <- out * 4*b*(1+2*b)/
                 (1+14*b)/(1+10*b)    # week 5
    out <- out * L_range(a,14:15,16)  # week 6
    out <- out * L_win(a,7)           # week 6
    out <- out * L_range(a,1:3,13)    # week 7
    out <- out * L_lose(a,6)          # week 7
    out <- out * L_range(a, 10, 11)   # week 8
    out <- out * L_lose(a,5)          # week 8
    out <- out * L_range(a,1:6,11)    # week 9
    out <- out * L_lose(a,5)          # week 9
    out <- out * L_range(a,4,5)       # week 9
    out <- out * L_range(a,1:3,9)     # week 10
    out <- out * 1/(7-6*a)            # week 10
    out <- out * L_range(a,1:3,7)     #  week 11
    out <- out * L_range(a,4:6,6)     # finals
    out <- out * L_range(a,1:2,3)     # finals
    out <- out * L_range(a,1:3,5)     # finals
    out <- out * L_range(a,1,4)       # finals
    out <- out * L_range(a,1:2,3)     # finals
    out <- out * L_range(a,2,2)       # finals
    return(out)
}
```
```{r label=optimizeLauras,cache=TRUE}
evaluate <- optimize(L_laura,c(0.1,0.9),maximum=TRUE)$max
small <- 1e-4
second <- (2*log(L_laura(evaluate))-log(L_laura(evaluate-small))-log(L_laura(evaluate+small)))/small^2
evaluate
second
f <- function(x){x*(1-x)}
(2*log(f(evaluate))-log(f(evaluate-small))-log(f(evaluate+small)))/small^2
```

```{r label=plotlaurassupport}
a <- seq(from=0,to=1,by=0.01)
L <- L_laura(a)
L <- L/max(L,na.rm=TRUE)
plot(a,L)
abline(v=evaluate)
plot(a,log(L),ylim=c(-5,0))
abline(h= -2)
abline(v=evaluate)
```

### 2024-03-08 inst/vsl_partial_ranks.Rmd
work on very simplified likelihood

 ---
title: "Very simplified likelihood: partial ranks"
author: "Robin Hankin"
output: bookdown::html_document2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("pracma")
library("magrittr")
library("knitr")
```

```{r hexsticker, out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
```

This follows on from `very_simplified_likelihood.Rmd`.
Consider a race between a competitor of Bradley-Terry strength $a$ and
$n$ competitors each of strength $b$; we require $a+b=1$.  NB we have
$n+1$ competitors in total, 1 of strength $a$ and $n$ of strength $b$.
An observation is indexed by $r$, the number of $b$ clones finishing
ahead of $a$, so $0\leqslant r\leqslant n$.  The initial field
strength is $a+nb=1+b(n-1)$.

```{r,label=definefsingle}
f_single <- function(a,r,n){
	B <- 1/(1-a)
    exp(log(B-1)-lgamma(B+n)+lgamma(B+n-r-1)+lfactorial(n)-lfactorial(n-r))
}
f_vec_a <- function(a,r,n){  # vectorised in 'a' but not in 'r'
	sapply(a,function(a){f_single(a,r=r,n=n)})
}

f_range <- function(a,possible,n){ # r1=4 -> came fifth [four b clones ahead]  
    out <- a*0
    for(i in (possible)){
        out <- out + f_vec_a(a,i-1,n-1) }
    return(out)
}
```

verify verify verify

```{r verifyverifyverify}
f_range(0.2323423,1:10,10)-1
```

Suppose there are 30 in the class, we know our bro came between 5th
and 29th:

```{r plot4tto9th}
a <- seq(from=0,by=0.1,to=1)
L <- f_range(a,possible=5:29,30)
plot(a,L,type='b')
plot(a,log(L),type='b')
abline(h=-2)
```

What I want to do now is compare students with different grades:

```{r}
M <- t(matrix(c(
    "4-9","10-14", "50",
    "5-10","1-4", "23",
    "23-40","23-40", "100",
    "1-3","5-9", "100"),3,4))
rownames(M) <- c("Algebra","Calculus","Geometry","Topology")
colnames(M) <- c("Alice range","Bob range", "class size")
kable(M,caption='asdf')
```

```{r}
alice <- t(matrix(c(
    4,9,50,
    5,10,23,
    23,40,100,
    1,3,100),nrow=3))
dimnames(alice) <- list(topic=rownames(M),alice_results=c("low","high","class"))

bob <- t(matrix(c(
    30,34,50,
    17,19,23,
    70,90,100,
    66,88,100),nrow=3))
dimnames(bob) <- list(topic=rownames(M),bob_results=c("low","high","class"))
```

```{r showalicebob}
alice
bob
```

```{r definemakelike}
makelike <- function(a,M){
    L <- a*0 + 1
    for(i in seq_len(nrow(M))){
        L <- L * f_range(a,M[i,1]:M[i,2],M[i,3])
    }
    return(L)
}
```

```{r plotalicebobsep}
s <- seq(from=0,to=1,by=0.02)
La <- makelike(s,alice)
Lb <- makelike(s,bob)
plot(s,La/max(La,na.rm=TRUE))
plot(s,Lb/max(Lb,na.rm=TRUE))
```

```{r deflikeboth}
Likeboth <- function(a){makelike(a,alice)*makelike(a,bob)}
plot(s,Likeboth(s),col='red',pch=16)
optimize(Likeboth,c(0.4,0.6),maximum=TRUE)
```

```{r contouralicebob,cache=TRUE}
s_alice <- seq(from=0.1,to=0.95,by=0.01)
s_bob   <- seq(from=0.1,to=0.95,by=0.01)
like_sep <- function(v){log(makelike(v[1],alice)) + log(makelike(v[2],bob))}
L <- apply(as.matrix(expand.grid(s_alice,s_bob)),1,like_sep)
L <- L - max(L)
```

```{r contourbothalicebob}
par(pty='s')
contour(s_alice,s_bob,
        matrix(L,length(s_alice),length(s_bob)),
        asp=1,levels = -(0:7)
        )
abline(0,1)

```

```{r dobothoptims,cache=TRUE}
H_alt <- optim(c(0.8,0.4),like_sep,control=list(fnscale=-1))
H_null <- optim(c(0.8,0.4),function(v){
    v[2] <- v[1]
    like_sep(v)},control=list(fnscale=-1))
pchisq(2*(H_alt$value-H_null$value),df=1,lower.tail=FALSE)
```

### 2024-03-08 inst/vsl_teams.Rmd
work on very simplified likelihood

---
title: "Very simplified likelihood"
author: "Robin Hankin"
output: bookdown::html_document2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("pracma")
library("magrittr")
```

```{r hexsticker, out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
```

This follows on from `very_simplified_likelihood.Rmd`.  Consider some
form of comparison between a competitor of Bradley-Terry strength $a$
and $n$ competitors each of strength $b$; we require $a+b=1$.  NB we
have $n+1$ competitors in total, 1 of strength $a$ and $n$ of strength
$b$.  An observation is indexed by $r$, the number of $b$ clones
finishing ahead of $a$, so $0\leqslant r\leqslant n$.  The initial
field strength is $a+nb=1+b(n-1)$.

Suppose we split the $n+1$ competitors into two teams.  The team with
our bro has $r+1$ members and the other one has size $n-r$ members.
Then, using standard notation

\begin{equation}
\frac{a+rb;(n-r)b}{a+nb}
\end{equation}

Let us suppose that the teams are of equal size:  $1+r=n-r=m$, say:

\begin{equation}=
\frac{a+(m-1)b;mb}{a+(2m-1)b}=\frac{mb (a-b); mb}{2mb+(a-b)}=\frac{(m-2)b+1;mb}{(2m-2)b+1}
\end{equation}

Let's try $m=15$:

```{r}
m <- 15
a <- seq(from=0,to=1,by=0.001)
b <- 1-a
L_win <- ((m-2)*b+1)/((2*m-2)*b+1)
L_lose<- (m*b)      /((2*m-2)*b+1)
plot(a,L_win ,type='b',ylim=0:1,main="Focal competitor's team wins")
plot(a,L_lose,type='b',ylim=0:1,main="Focal competitor's team loses")
S_win <- log(L_win)
S_win <- S_win-max(S_win)
S_lose <- log(L_lose)
S_lose <- S_lose-max(S_lose)
plot(a,S_win,type='b',main="Win")
plot(a,S_lose,type='b',main="Lose")
logodds <- log(a/(1-a))
plot(logodds,S_win,main="win")
plot(logodds,S_lose,main="lose")
```

Now in week 4 Laura came in the second team of four (of four members
each).  I get the teams's strengths are $\beta=4b$ and
$\alpha=3b+a=1+2b$, so

\begin{eqnarray}
L &=& \frac{\alpha}{3\alpha+\beta}\cdot\frac{\beta}{2\alpha+\beta}\\
  &=& \frac{4b}{12b + (1+2b)}\cdot\frac{1+2b}{8b+(1+2b)}\\
  &=& \frac{4b}{1+14b}\cdot\frac{1+2b}{1+10b}
\end{eqnarray}

```{r plotsecondteam}
a <- seq(from=0,to=1,by=0.001)

f <- function(a){
    b <- 1-a    
    4*b*(1+2*b)/(1+14*b)/(1+10*b)
}
L <- f(a)
L <- L/max(L,na.rm=TRUE)
plot(a, L,type='b')
plot(a, log(L),type='b',ylim=c(-5,0))
abline(h= -2)
```

### 2024-03-11 inst/formula1_start_vs_finish.Rmd
more work on formula 1 start vs finish position

---
title: "formula1_start_vs_finish.Rmd"
output: html_document
date: "2024-03-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
```

Formula 1 grand prix racing, 2023 season.

```{r loadfinish}
setwd("~/rstudio/hyper2/inst")
finish <- read.table("formula1_2023.txt",header=TRUE)
finish <- finish[,seq_len(ncol(finish)-1)]
finish[,1:5]
```

```{r loadgrid}
grid <- read.table("formula1_starting_2023.txt",header=TRUE)
grid[,1:5]
```

```{r definemakeH3}
`makeH3` <- function(i,lambda){
  suppressWarnings(a <- setNames(as.numeric(finish[,1]),rownames(finish)))
  suppressWarnings(b <- setNames(as.numeric(grid  [,1]),rownames(finish)))
  wanted <- !is.na(b)
  a <- a[wanted]
  b <- b[wanted]
  a[is.na(a)] <- 0 # did not finish
  H <- ordervec2supp(a)
  for(i in seq_along(b)){H[names(b)[i]] <- as.weight(lambda^(i-1))}
  return(H)
}

likelam <- function(lambda){
  H <- hyper3()
  for(i in seq_len(ncol(finish))){H <- H + makeH3(1,lambda=lambda)}
  return(H)
}
```

```{r evaluatelamrange,cache=TRUE}

f <- function(lambda){
  H <- likelam(lambda)
  loglik(maxp(H,n=4),H)
}
f(0.9)
f(0.901)
```

```{r evalam,cache=TRUE}
lam <- seq(from=0.9,to=1.05,len=5)
like <- sapply(lam,f)
```

```{r plotlikelam}
plot(lam,like,type='b')
```

### 2024-03-12 inst/hankin-ejs.tex
follow instructions on ejs website

--- 2024-03-04 inst/ejs-sample.tex
@@
 indexed by~$r$, the number of strength $b$ clones finishing ahead
-of~$a$, and because we have $n+1$ competitors $0\leqslant r\leqslant
-n$.  The initial field strength is $a+nb=n+(n-1)a$.  The Plackett-Luce
-likelihood function~\cite{plackett1975,luce1959} for this situation
-would be
+of~$a$, and because we have $n+1$ competitors %%%%%%%%%%%%%%%%%%%%%%
+$0\leqslant r\leqslant n$.  The initial field strength is
+$a+nb=n+(n-1)a$.  The Plackett-Luce likelihood
+function~\cite{luce1959,plackett1975} for this situation would be
 
@@
 \log\mathcal{L}_{r,n}(a)=\log(B-1)-B\frac{r+1}{n}
-+ K + \mathcal{O}\left(n^{-2}\right)
++ K + \mathcal{O}\left(n^{-2}\right).
 \end{equation}
@@
 (that is, one minus the mean ratio of rank to number of [cloned]
-competitors).  I present some numerical results below.
-
-Previous related work presented the case $n=2$ [in current notation].
-Considering the case $n=7$ gives us $7+1=8$ distinct likelihood
-functions, Figure~\ref{ninelikes}.  We see an unanticipated asymmetry
-between $r$ and $n-r$.  For $r=3$ one might expect that
-$\hat{a}=\frac{1}{2}$ on the grounds that the focal competitor's
-performance was median.  However, $\hat{a}\simeq 0.64$.
+competitors).  Previous related work presented the case $n=2$ [in
+  current notation].  Considering the case $n=7$ gives us $7+1=8$
+distinct likelihood functions, Figure~\ref{ninelikes}.  We see an
+unanticipated asymmetry between $r$ and $n-r$.  For $r=3$ one might
+expect that $\hat{a}=\frac{1}{2}$ on the grounds that the focal
+competitor's performance was median.  However, $\hat{a}\simeq 0.64$.
 
@@
 generally regarded as sensitive and private information.
-Nevertheless, it is common a student to possess some informative
+Nevertheless, it is common for a student to possess some informative
 observations: the total enrollment in a specific class and their
-individual ranking within that class;  table~\ref{educationtable}
-shows an apocryphal dataset.   What could a student infer from such
-data?
+individual ranking within that class; table~\ref{educationtable} shows
+an apocryphal dataset.  What could a student infer from such data?
 
@@
 
-One natural question would be to wonder whether the student was better
-at pure or applied mathematics and, if so, to quantify the difference.
+One might wonder whether the student was better at pure or applied
+mathematics and, if so, to quantify the difference.
 Figure~\ref{plotpureandapplied} shows contours of support [again
@@
 as a function of pure strength on the horizontal axis and applied
-strength on the vertical axis.  A null of equal strengths appears as a
+strength on the vertical axis.  A null of equal strengths appears as
+the
 diagonal line.  We see that the maximum likelihood estimate is

### 2024-03-13 inst/hankin-ejs.tex
tidyup in prep. for submission

@@
-\documentclass[ejs]{imsart}
+\documentclass[ejs,noshowframe]{imsart}
 
@@
 
-\arxiv{2010.00000}
 \startlocaldefs
@@
 inference problem appears to have a wide range of applications and I
-present some analysis of datasets drawn from Formula 1 motor racing
+present some analysis of datasets drawn from Formula~1 motor racing
 and community initiative Parkrun.
@@
 \kwd[Primary ]{6202}
-\kwd{00X00}
-\kwd[; secondary ]{62F30}
-\end{keyword}
-
-\begin{keyword}
-\kwd{First keyword}
-\kwd{second keyword}
+\kwd{Statistical inference under constraints}
+\kwd[ ; secondary: Plackett-Luce likelihood functions; ]{62F30}
 \end{keyword}
@@
 distinct likelihood functions, Figure~\ref{ninelikes}.  We see an
-unanticipated asymmetry between $r$ and $n-r$.  For $r=3$ one might
+unanticipated asymmetry between $r$ and $n-r$.  For $r=4$ one might
 expect that $\hat{a}=\frac{1}{2}$ on the grounds that the focal
-competitor's performance was median.  However, $\hat{a}\simeq 0.64$.
+competitor's performance was median.  However, $\hat{a}\simeq 0.5886$.
 
@@
 \includegraphics[width=4in]{ninelikes}  % ninelikes.pdf produced by very_simplified_likelihood.Rmd
-\caption{Likelihood functions for $r=0(1)8$\label{ninelikes}}
+\caption{Likelihood functions for $r=0(1)8$\label{ninelikes}.  Value of $r$ indicated in colour by its curve; vertical line shows $\hat{a}\simeq 0.5886$ for the case $r=4$}
 \end{figure}
@@
 \begin{supplement}
-\stitle{Title of Supplement A}
-\sdescription{Short description of Supplement A.}
-\end{supplement}
-\begin{supplement}
-\stitle{Title of Supplement B}
-\sdescription{Short description of Supplement B.}
+\stitle{Very simplified likelihood}
+\sdescription{A terse rmarkdown document that  recreates the figures and salient results of the submission}
 \end{supplement}

### 2024-03-13 inst/very_simplified_likelihood.Rmd
tidyup
tidyup in prep. for submission

@@
 
+\[
+\mbox{prob of coming third} =
+\frac{b}{a+nb}\cdot
+\frac{b}{a+(n-1)b}\cdot
+\frac{a}{a+(n-2)b}\cdot
+\frac{b}{(n-2)b}\cdots
+\frac{b}{2b}\cdot\frac{b}{b}
+=\frac{n!}{(n-2)!}\cdot X
+\]
+
+Now consider the case where there are $r$ clones coming ahead of the
+focal competitor:
+
+\[
+\mbox{prob of having $r$ clones ahead}
+=\frac{n!}{(n-r)!}\cdot X
+\]
+
 ## Visuals
@@
 a <- seq(from=0,to=1,by=0.01)
-n <- 7
+n <- 8
 jj <- f_vec_arn(a,3,n,log=FALSE)
-plot(a,jj/max(jj,na.rm=TRUE),type='n')
+plot(a,jj/max(jj,na.rm=TRUE),type='n',xlab=expression(a),ylab="likelihood")
 grid()
+rain <- rainbow(n+1)
 for(r in seq(from=0,to=n)){
@@
   if(r>0){y[length(y)] <- 0}
-  points(a,y,type='l',lwd=4,col=rainbow(n+1)[r+1])
-  abline(v=0.64)
+  points(a,y,type='l',lwd=4,col=rain[r+1])
 }
+abline(v=MLE(r=4,n=8))
+text(0.07,0.95,"r=8",col=rain[9])
+text(0.20,0.95,"r=7",col=rain[8])
+text(0.26,0.91,"r=6",col=rain[7])
+text(0.29,0.83,"r=5",col=rain[6])
+text(0.33,0.76,"r=4",col=rain[5])
+text(0.35,0.64,"r=3",col=rain[4])
+text(0.39,0.53,"r=2",col=rain[3])
+text(0.45,0.41,"r=1",col=rain[2])
+text(0.63,0.22,"r=0",col=rain[1])
 }
+
 plotter()
@@
 dev.off()
+```
+
+```{r shown4n8mle}
+MLE(r=4,n=8)
 ```
@@
 
-Above we see three parkrun results.  In parkrun number 148, I placed
+Above we see some parkrun results.  In parkrun number 148, I placed
 229 out of a field of 318.
-
-```{r optimsinglelike,cache=TRUE}
-```
 

### 2024-03-13 inst/formula1_start_vs_finish.Rmd
bugfixes

@@
 ```{r definemakeH3}
-`makeH3` <- function(i,lambda){
-  suppressWarnings(a <- setNames(as.numeric(finish[,1]),rownames(finish)))
-  suppressWarnings(b <- setNames(as.numeric(grid  [,1]),rownames(finish)))
+replacenonzerowithorder <- function(x){
+  nonzero <- x!=0
+  x[nonzero] <- order(x[nonzero])
+  return(x)
+}
+
+order_non_NA_NZ <- function(x){ # leaves NA and 0 entries unchanged; replaces other elements with their order
+  wanted <- !is.na(x)
+  x[wanted] <- replacenonzerowithorder(x[wanted])
+  return(x)
+}
+
+topn <- 15
+finish <- finish[seq_len(topn),]
+options(use_alabama=TRUE)
+
+`makeH3` <- function(i,lambda,topn){
+  suppressWarnings(a <- setNames(as.numeric(finish[,i]),rownames(finish)))
+  suppressWarnings(b <- setNames(as.numeric(grid  [,i]),rownames(finish)))
   wanted <- !is.na(b)
@@
   a[is.na(a)] <- 0 # did not finish
+  a <- order_non_NA_NZ(a)
+  a <- a[seq_len(topn)]
   H <- ordervec2supp(a)
-  for(i in seq_along(b)){H[names(b)[i]] <- as.weight(lambda^(i-1))}
+  for(j in seq_along(b)){
+    if(!is.na(names(b)[j])){H[names(b)[j]] <- as.weight(lambda^b[j])}
+  }
   return(H)
@@
   H <- hyper3()
-  for(i in seq_len(ncol(finish))){H <- H + makeH3(1,lambda=lambda)}
+  for(i in seq_len(ncol(finish))){H <- H + makeH3(i,lambda=lambda,topn=topn)}
   return(H)
@@
 
-```{r evaluatelamrange,cache=TRUE}
-
+```{r evaluatelambbda,cache=TRUE}
 f <- function(lambda){
   H <- likelam(lambda)
-  loglik(maxp(H,n=4),H)
+  m <- maxp(H,n=1)
+  return(c(loglik(m,H),lambda,m))
 }
-f(0.9)
-f(0.901)
 ```
@@
 ```{r evalam,cache=TRUE}
-lam <- seq(from=0.9,to=1.05,len=5)
-like <- sapply(lam,f)
+lam <- seq(from=0.85, to=1.05, len=15)
+bigmat <- sapply(lam,f)
+bigmat
 ```
 
-```{r plotlikelam}
-plot(lam,like,type='b')
+```{r plotlikelambda}
+if(FALSE){plot(lam,like-max(like),type='b')
+abline(h=-2)
+abline(v=1)
+}
 ```

### 2024-03-19 inst/very_simplified_likelihood.Rmd
work on lane advantage in olympic athletics

@@
 
+# Athletics
+
+```{r loadolymp}
+O <- read.table("olympic_athletics_mens_100m.txt",header=TRUE)
+head(O)
+```
+
+```{r analuyzelmp,cache=TRUE}
+jjf <- function(x){f_vec_arn(x,O$rank,O$n,log=TRUE)}
+system.time(optolym <- optimize(jjf, c(0.45,0.5), maximum=TRUE))
+```
+
+```{r}
+optolym
+jjf(0.5)
+jjf(optolym$maximum)
+(LR <- 2*(jjf(optolym$maximum)-jjf(0.5)))
+pchisq(LR,df=1,lower.tail=FALSE)
+```
+
+```{r makeolymplike}
+a <- seq(from=0.3,to=0.65,len=45)
+L <- f_vec_arn(a,O$rank,O$n,log=TRUE)
+Lmax <- jjf(optolym$maximum)
+Lmax
+L <- L-Lmax
+```
+
+```{r findcredint}
+(a_lower <- uniroot(function(x){jjf(x)+2-Lmax},interval=c(0.3,0.5))$root)
+(a_upper <- uniroot(function(x){jjf(x)+2-Lmax},interval=c(0.5,0.6))$root)
+```
+
+```{r plotolymp}
+plot(a,L,type='l',ylab="log-likelihood",lwd=2)
+abline(h=c(0,-2))
+ahat <- optolym$maximum
+segments(x0=ahat,y0=-1.5,y1=0)
+text(ahat,-0.91,expression(hat(a)==0.573),pos=2)
+abline(v=0.5,lty=3)
+
+segments(x0=a_lower,y0=-1.5,y1=-3.5)
+segments(x0=a_upper,y0=-1.5,y1=-3.5)
+
+text(x=a_lower,y=-3,"0.380",pos=4)
+text(x=a_upper,y=-3,"0.563",pos=2)
+```
+
 # Parkrun

### 2024-03-19 inst/hankin-ejs.tex
minor rewording and improvement to figures

@@
 One entity is of particular interest to us: the ``focal entity'', or
-focal competitor.  We assign Placket-Luce strength of $a$ to the focal
+focal competitor.  We assign Placket-Luce strength $a$ to the focal
 competitor and $b$ to all others; we require $a+b=1$.  The associated
 inference problem appears to have a wide range of applications and I
-present some analysis of datasets drawn from Formula~1 motor racing
-and community initiative Parkrun.
+present some analysis of datasets drawn from the fields of Olympic
+athletics, education, and Formula~1 motor racing.
 \end{abstract}
@@
 likelihood strengths identify the correct ranking with a probability
-of essentially zero.  The situation is worse in cases such as the
-International Mathematical Olympiad, which requires us to consider
-over 100 competitors.
+of essentially zero.  The situation is worse in cases such as formula
+1 motor racing considered below, which requires us to consider
+${\mathcal O}(100)$ competitors.
 
@@
 $\hat{B}=\frac{n+r+1}{r+1}+\mathcal{O}\left(n^{-1}\right)$, or
-alternatively $\hat{a}=\frac{n}{n+r+1}$.  For repeated observations
-$r_i,n_i$, $1\leqslant i\leqslant m$ we would have
+alternatively $\hat{a}=\frac{n}{n+r+1}$.  Compare this with the result
+of asking ``what is the probability $p$ that the focal competitor
+beats a randomly chosen opponent?'' to which the binomial maximum
+likelihood estimate is simply $\frac{n-r}{n}$, agreeing to within
+$\mathcal{O}(n^{-2})$.  For repeated observations $r_i,n_i$,
+$1\leqslant i\leqslant m$ we would have
 \begin{equation}
@@
 (that is, one minus the mean ratio of rank to number of [cloned]
-competitors).  Previous related work presented the case $n=2$ [in
-  current notation].  Considering the case $n=7$ gives us $7+1=8$
-distinct likelihood functions, Figure~\ref{ninelikes}.  We see an
-unanticipated asymmetry between $r$ and $n-r$.  For $r=4$ one might
-expect that $\hat{a}=\frac{1}{2}$ on the grounds that the focal
-competitor's performance was median.  However, $\hat{a}\simeq 0.5886$.
+competitors); compare the binomial result of $1-\left.\sum
+r_i\right/\sum n_i$.
+
+Previous related work presented the case $n=2$ [in current notation].
+Considering the case $n=7$ gives us $7+1=8$ distinct likelihood
+functions, Figure~\ref{ninelikes}.  We see an unanticipated asymmetry
+between $r$ and $n-r$.  For $r=4$ one might expect that
+$\hat{a}=\frac{1}{2}$ on the grounds that the focal competitor's
+performance was median.  However, $\hat{a}\simeq 0.5886$.
 
@@
 
-\section{Use-cases}
+Below, I present three examples that showcase likelihood
+function~\ref{likeforrn3}, drawn from the fields of athletics,
+education, and Formula~1 motor racing.
 
-Here I present three examples that showcase likelihood
-function~\ref{likeforrn3}.  Formula 1 education, parkrun
+\subsection{Olympic track and field athletics}
 
-\subsection{Park run}
-
-``Parkrun'' is a distributed community initiative that organises
-weekly timed 5 km runs/walks in parks worldwide~\cite{hindley2020}.  A
-typical event will have 200 participants.  Table~\ref{parkruntable}
-shows the author has completed a total of 15 parkruns to date, and
-from the first pair of numbers we see that 238 runners attended that
-particular parkrun, of whom the author placed 173.
-
-\begin{table*}[t]
-  \caption{Parkrun results}
-\label{parkruntable}
-\begin{tabular}{ccccccccccc}
-\hline
-rank   &173  & 165 & 172 & 199 & 181 & 229 & 177 & 222 & 206 & 142 \\
-runners&238  & 238 & 196 & 242 & 242 & 318 & 259 & 305 & 297 & 241 \\ \\ \\
-rank   & 118 & 224 & 128 & 115 & 183 &     &     &     &     &\\
-runners& 179 & 338 & 203 & 245 & 254 &     &     &     &     &\\
-\hline
-\end{tabular}
-\end{table*}
-
-We may consider the author to have an unknown generalized Bradley
-Terry strength $a$: Figure~\ref{parkrunsupport} shows a support curve
-for $a$ with the evaluate, $\hat{a}\simeq 0.448$ shown.  We also see
-two units of support~\cite{edwards1992} shown as a dotted line
-illustrating a support interval of about 0.315 - 0.567; we may be
-reasonably confident that the author's true strength lies in this
-range.  In particular, note that a reasonable $H_0\colon
-a=\frac{1}{2}$ may not be rejected, the support for $H_0$ being only
-susceptible to very minor improvement [$\simeq 0.351$] by moving to
-the evaluate.
+One long-standing contention in the athletic world is that of lane
+bias~\cite{munro2022}.  Running in the inner lane is frequently held
+to be a disadvantage; mechanisms include biomechanical issues arising
+from centrifugal acceleration, to competitors' inability to see their
+rivals.  Here attention is restricted to heats, for which lane placing
+is random.  I consider every heat in Olympic running from 2004-2020 (a
+total of 41 heats, each of 8 or 9 competitors); only competitors'
+ranks are used.  Figure~\ref{plotolymp} shows a log-likelihood curve
+for the Bradley-Terry strength of the focal competitor, here the
+runner in the innermost lane.  We see that the evaluate of $0.574$ is
+very close to the neutral value of $0.5$, for which the support is
+only $0.16$, far short of Edwards's two units of support criterion for
+rejection~\cite{edwards1972}.  This dataset contains no evidence to
+support lane bias.
 
 \begin{figure}[t]
-\includegraphics[width=4in]{plotparkrun}  % plotparkrun.pdf made in very_simplified_likelihood.Rmd
-\caption{Author's strength\label{parkrunsupport}}
+\includegraphics[width=4in]{plotolymp} % produced by very_simplified_likelihood.Rmd
+\caption{Log-likelihood curve for the Bradley-Terry strength of the
+  lane-1 runner, olympic qualifying heats 2012-2020 \label{plotolymp}}
 \end{figure}
@@
 
-One issue encountered in educational research is that of
-confidentiality of examination scores, and individual results are
-generally regarded as sensitive and private information.
-Nevertheless, it is common for a student to possess some informative
-observations: the total enrollment in a specific class and their
-individual ranking within that class; table~\ref{educationtable} shows
-an apocryphal dataset.  What could a student infer from such data?
+In educational research, confidentiality of examination scores is
+often an issue, and individual results are generally regarded as
+sensitive and private information.  Nevertheless, it is common for a
+student to possess some informative observations: the total enrollment
+in a specific class and their individual ranking within that class;
+table~\ref{educationtable} shows an apocryphal dataset.  What could a
+student infer from such data?
 
@@
 strength on the vertical axis.  A null of equal strengths appears as
-the
-diagonal line.  We see that the maximum likelihood estimate is
+the diagonal line.  We see that the maximum likelihood estimate is
 $\hat{p}_\mathrm{pure}=0.567$, $\hat{p}_\mathrm{applied}=0.887$.
@@
 along the null diagonal is about $-2.43$.  Edwards's two units of
-support criterion is met, and we are justified in concluding that the
-student is indeed more competent at applied than pure mathematics.
-Alternatively, we may observe that $2\mathcal{S}=4.86$ is in the tail
-region of its asymptotic $\chi^2_1$ distribution, corresponding to a
-$p$-value of about $0.028$.
+support criterion~\cite{edwards1972} is met, and we are justified in
+concluding that the student is indeed more competent at applied than
+pure mathematics.  Alternatively, we may observe that
+$2\mathcal{S}=4.86$ is in the tail region of its asymptotic $\chi^2_1$
+distribution, corresponding to a $p$-value of about $0.028$.
 
@@
 \includegraphics[width=4in]{showchecolike}  % showchecolike.pdf produced by very_simplified_likelihood.Rmd
-\caption{Likelihood functions for $\alpha,\beta$ for Checo's racing
+\caption{Log-likelihood contours for $\alpha,\beta$ for Checo's racing
   career 2011-2023\label{showchecolike}}

### 2024-03-19 inst/very_simplified_likelihood.Rmd
minor rewording and improvement to figures

@@
 
-```{r showfilledc}
+```{r showfilledcwithdot}
 L <- pmax(L,-40)
 showchec <- function(...){
-contour(a,b,L,xlab=expression(alpha),ylab=expression(beta))
+contour(a,b,L,xlab=expression(alpha),ylab=expression(beta),levels=-c(2,5*(1:5)))
+points(-0.27,0.0813,pch=16,cex=2)
 }
@@
 pdf(file="showchecolike.pdf")
-contour(a,b,L)
+showchec(a,b,L)
 dev.off()
@@
 ```{r plotolymp}
+plotolymp <- function(...){
 plot(a,L,type='l',ylab="log-likelihood",lwd=2)
@@
 text(x=a_upper,y=-3,"0.563",pos=2)
+}
+plotolymp()
+pdf(file="plotolymp.pdf")
+plotolymp()
+dev.off()
 ```

### 2024-03-19 inst/formula1_start_vs_finish.Rmd
improvements

--- 2024-03-11 inst/formula1_start_vs_finish.Rmd
@@
 
-```{r definemakeH3}
-`makeH3` <- function(i,lambda){
-  suppressWarnings(a <- setNames(as.numeric(finish[,1]),rownames(finish)))
-  suppressWarnings(b <- setNames(as.numeric(grid  [,1]),rownames(finish)))
+```{r definereplacenonzerowithorder}
+replacenonzerowithorder <- function(x){
+  nonzero <- x!=0
+  x[nonzero] <- rank(x[nonzero])
+  return(x)
+}
+
+order_non_NA_NZ <- function(x){ # leaves NA and 0 entries unchanged; replaces other elements with their order
+  wanted <- !is.na(x)
+  x[wanted] <- replacenonzerowithorder(x[wanted])
+  return(x)
+}
+
+topn <- 15
+finish <- finish[seq_len(topn),]
+options(use_alabama=TRUE)
+
+`makeH3` <- function(i,lambda,topn,finish,grid){
+  suppressWarnings(a <- setNames(as.numeric(finish[,i]),rownames(finish)))
+  suppressWarnings(b <- setNames(as.numeric(grid  [,i]),rownames(finish)))
   wanted <- !is.na(b)
@@
   a[is.na(a)] <- 0 # did not finish
+  a <- a[seq_len(topn)]
+  a <- order_non_NA_NZ(a)
   H <- ordervec2supp(a)
-  for(i in seq_along(b)){H[names(b)[i]] <- as.weight(lambda^(i-1))}
+  for(j in seq_along(b)){
+    if(!is.na(names(b)[j])){H[names(b)[j]] <- as.weight(lambda^b[j])}
+  }
   return(H)
@@
 
-likelam <- function(lambda){
+likelam <- function(lambda,topn,finish,grid){
   H <- hyper3()
-  for(i in seq_len(ncol(finish))){H <- H + makeH3(1,lambda=lambda)}
+  for(i in seq_len(ncol(finish))){H <- H + makeH3(i,lambda=lambda,topn=topn,finish=finish,grid=grid)}
   return(H)
@@
 
-```{r evaluatelamrange,cache=TRUE}
-
-f <- function(lambda){
-  H <- likelam(lambda)
-  loglik(maxp(H,n=4),H)
+```{r evaluatelambbda,cache=TRUE}
+f <- function(lambda,topn,finish,grid){
+  H <- likelam(lambda,topn,finish,grid)
+  m <- maxp(H,n=1)
+  return(c(loglik(m,H),lambda,m))
 }
-f(0.9)
-f(0.901)
 ```
 
-```{r evalam,cache=TRUE}
-lam <- seq(from=0.9,to=1.05,len=5)
-like <- sapply(lam,f)
+```{r evalamlam,cache=TRUE}
+system.time(lam <- seq(from=0.85, to=1.01, len=15))
+system.time(bigmat <- sapply(lam,f,topn=topn,finish=finish,grid=grid))
 ```
 
-```{r plotlikelam}
+```{r showbigmat}
+head(bigmat)
+```
+
+```{r max_alpha,cache=TRUE}
+system.time(alpha_MLE <- optimize(function(a){f(a,topn=topn,finish=finish,grid=grid)[1]},c(0.95,0.99),maximum=TRUE))
+alpha_MLE
+```
+
+```{r plotlikelambda}
+like <- bigmat[1,]
+like <- like - max(like)
 plot(lam,like,type='b')
+abline(h=c(0,-2))
+abline(v=1)
+segments(x0=0.9654,y0=-5,y1=0)
+grid()
 ```
+
+```{r showmaxalpha}
+if(FALSE){print(alpha_MLE)}
+```
+
+```{r conventionalhyper2max,cache=TRUE}
+H2 <- ordertable2supp(finish)
+mH2 <- maxp(H2)
+```
+
+```{r}
+dotchart(mH2,pch=16)
+dotchart(log(mH2),pch=16)
+```
+
+# Above was 2023, now do the 2022 season
+
+```{r loadfinish22}
+setwd("~/rstudio/hyper2/inst")
+finish <- read.table("formula1_2022.txt",header=TRUE)
+finish22 <- finish[,seq_len(ncol(finish)-1)]
+finish22[,1:5]
+grid22 <- read.table("formula1_starting_2022.txt",header=TRUE)
+grid22[,1:5]
+```
+
+```{r mlecalc22, cache=TRUE}
+system.time(
+    alpha_MLE22 <- optimize(function(a){f(a,topn=topn,finish=finish22,grid=grid22)[1]},c(0.95,0.99),maximum=TRUE)
+    )
+alpha_MLE22
+```
+
+```{r evalam22,cache=TRUE}
+lam <- seq(from=0.90, to=1.05, len=15)
+system.time(bigmat <- sapply(lam,f,topn=topn,finish=finish22,grid=grid22))
+```
+
+```{r showbigmat22}
+head(bigmat)
+```

### 2024-03-22 inst/hankin-ejs.Rmd
more for git to ignore

--- 2024-03-19 inst/very_simplified_likelihood.Rmd
@@
 ---
-title: "Very simplified likelihood"
+title: "Recreation of figures and results in the manuscript"
 author: "Robin Hankin"
@@
 
-```{r hexsticker, out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
-knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
-```
+This terse document reproduces the results presented in the associated
+manuscript.
 
-Consider a race between a competitor of Bradley-Terry strength $a$ and
-$n$ competitors each of strength $b$; we require $a+b=1$.  NB we have
-$n+1$ competitors in total, 1 of strength $a$ and $n$ of strength $b$.
-An observation is indexed by $r$, the number of $b$ clones finishing
-ahead of $a$, so $0\leqslant r\leqslant n$.  The initial field
-strength is $a+nb=1+b(n-1)$.
-
-Suppose our bro comes first (zero $b$'s ahead of him):
-
-\[
-\mathcal{L}(a)=\frac{a}{a+nb}=\frac{1-b}{1+(n-1)b}
-\]
-
-Comes second, one $b$ ahead of him:
-
-\[
-\mathcal{L}(a)=\frac{b}{a+nb}\cdot\frac{a}{a+(n-1)b}
-\]
-
-Third, two ahead of him:
-
-\[
-\mathcal{L}(a)=
-\frac{b}{a+ n   b}\cdot
-\frac{b}{a+(n-1)b}\cdot
-\frac{a}{a+(n-2)b}
-\]
-
-Fourth, three ahead:
-
-\[
-\mathcal{L}(a)=
-\frac{b}{a+ n   b}\cdot
-\frac{b}{a+(n-1)b}\cdot
-\frac{b}{a+(n-2)b}\cdot
-\frac{a}{a+(n-3)b}
-\]
-
-Placing after $r<n$ of the $n$ clones of strength $b$ finish:
-
-\[
-\mathcal{L}(a)=
-\frac{b^ra}{
-(a+nb)(a+(n-1)b)(a+(n-2)b)\cdots(a+(n-r)b)}
-=
-\frac{B-1}{(B+n-1)(B+n-2)\cdots(B+n-r-1)}
-\]
-
-where $B=1/b$ [also using the fact that $a+b=1$].
-
-This is
-
-\[
-\mathcal{L}_{r,n}(a)=\frac{(B-1)(B+n-r-2)!}{(B+n-1)!},\qquad B=1/(1-a)
-\]
-
-Then, using Stirling [viz $n!\sim n^ne^{-n}\sqrt{2\pi n}$] we get
-
-\[
-\log(B-1) + ((B+n-r-2)\log(B+n-r-2)-(B+n-r-2)+\frac{\log(B+n-r-2)}{2})
-          - ((B+n  -1)\log(B+n  -1)-(B+n  -1)+\frac{\log(B+n  -1)}{2})
-\]
-
-If you give Sage this:
+## Asymptotic likelihood expression {-}
 
@@
 
-(where $K$ is an arbitrary constant).  Observing that
-$\partial^2\mathcal{S}/\partial B^2=-(B-1)^{-2}<0$, the evaluate would
-be unique; we have
-$\hat{B}=\frac{n+r+1}{r+1}+\mathcal{O}\left(n^{-1}\right)$, or
-alternatively $\hat{a}=\frac{n}{n+r+1}$.  I present some numerical
-results below.
-
-# Placing last
-
-Placing last is finishing  after $n=r$ clones finish:
-
-\[
-n=1\longrightarrow\mathcal{L}=b\]
-
-\[
-n=2\longrightarrow\mathcal{L}=\frac{b}{a+2b}\cdot\frac{b}{a+b}=\frac{b^2}{1+b}
-=\frac{1}{B(B+1)}
-\]
-
-\[
-n=3\longrightarrow\mathcal{L}=\frac{b}{a+3b}\cdot\frac{b}{a+2b}\cdot\frac{b}{a+b}=\frac{b^3}{(1+2b)(1+b)}=\frac{1}{B(B+1)(B+2)}
-\]
-
-\[
-n=4\longrightarrow\mathcal{L}=\frac{b}{a+4b}\cdot\frac{b}{a+3b}\cdot\frac{b}{a+2b}\cdot\frac{b}{a+b}=
-\frac{b^4}{(1+3b)(1+2b)(1+b)}=
-\frac{1}{B(B+1)(B+2)(B+3)}
-\]
-
-For $n=4$ we have equivalently
-
-\[
-\frac{1}{(B+3)(B+2)(B+1)B}=\frac{(B-1)!}{(B+3)!}
-\]
-
-In general we would have
-
-\[
-\frac{1}{(B+n-1)(B+n-2)\cdots(B+1)B}=
-\frac{(B-1)!}{(B+n-1)!}
-\]
-
-(denominator has $n+1$ terms).  Observe that this agrees with the
-$r<n$ case above (substituting $r=n$).  We may introduce some R idiom:
+## Definition of likelihood and log-likelihood in R {-}
 
@@
 
-Now use the derivative to find the maximum likelihood point
+## Maximum likelihood estimation
+
+Use numerical optimization [`optim()`] to find the maximum likelihood
+point [R function `MLE()`] and, as a check, use calculus `uniroot()`
+to find the point of zero derivative; R function `MLEa()`] to find the
+maximum likelihood point:
 
@@
 
+## Produce Maximum Likelihood diagram, Figure 1 {-}
+
 ```{r,label=makejj}
@@
 
-# Note on the differences between probability and likelihood
+## Visuals {-}
 
-The likelihood calculations above can be confusing becuase likelihood
-is not the same as probability.
-
-Consider, for example, $n=2$ so we have 3 competitors.
-
-probability of focal competitor coming first [i.e. $r=0$]:
-
-\[\frac{a}{a+2b}=\frac{1-b}{1+b}\]
-
-Now what is the probability of coming second, that is, $r=1$?
-
-My first thought was:
-
-\[
-\mbox{prob of coming second}   =
-\frac{b}{a+2b}\cdot\frac{a}{a+b}=\frac{b(1-b)}{1+b}\qquad\mbox{WRONG}
-\]
-
-but this is incorrect: there are two clones of (strength) $b$, and the
-likelihood arguments above do not distinguish between their finishing
-order.  We can name the competitors $a$, $b_1$ and $b_2$ and specify
-that $b_1$ and $b_2$ both have strength $b$.  Then "coming second"
-means that the finishing order was $b_1\succ a\succ b_2$ or $b_2\succ
-a\succ b_1$.  Noting that these events are disjoint, the probability
-would be
-
-\[\mbox{prob of coming second}   =
-\operatorname{P}(b_1\succ a\succ b_2)      +
-\operatorname{P}(b_2\succ a\succ b_1)      =
-\frac{b}{a+2b}\cdot\frac{a}{a+b} +
-\frac{b}{a+2b}\cdot\frac{a}{a+b} = \frac{2(1-b)}{1+b}
-\]
-
-Similarly for the focal competitor coming last we need to sum the
-probabilities of finishing orders $b_1\succ b_2\succ a$ and $b_2\succ
-b_1\succ a$:
-
-\[\mbox{prob of coming third}=
-\operatorname{P}(b_1\succ b_2\succ a) +
-\operatorname{P}(b_2\succ b_1\succ a) =
-\frac{b}{a+2b}\cdot\frac{b}{a+b}+
-\frac{b}{a+2b}\cdot\frac{b}{a+b}=
-\frac{2b^2}{1+b}.
-\]
-
-Above we see a factor of two difference between the probabilities and
-the likelihoods presented previously.  Note that there is no such
-factor for the probability of the focal competitor finishing first:
-
-\[
-\mbox{prob of coming first}      =
-\operatorname{P}(a\succ b_1\succ b_2)      +
-\operatorname{P}(a\succ b_2\succ b_1)      =
-\frac{b}{a+2b}\cdot\frac{b}{b+b} +
-\frac{b}{a+2b}\cdot\frac{b}{b+b} = \frac{1-b}{1+b}
-\]
-
-Now we see that the probabilities sum to
-
-\[
-\frac{1-b}{1+b} + \frac{2b^2}{1+b} + \frac{2(1-b)}{1+b}=1
-\]
-
-Now try $r=3$ [i.e. 4 competitors].  Taking things one step at a time:
-
-\[
-\mbox{prob of coming first} =
-3!\cdot\frac{a}{a+3b}\cdot\frac{b}{3b}\cdot\frac{b}{2b}\cdot\frac{b}{b}=
-\frac{1-b}{1+2b}
-\]
-
-\[
-\mbox{prob of coming second} =
-3!\cdot\frac{b}{a+3b}\cdot\frac{a}{a+2b}\cdot\frac{b}{2b}\cdot\frac{b}{b}=
-3\cdot\frac{b(1-b)}{(1+2b)(1+b)}
-\]
-
-\[
-\mbox{prob of coming third} =
-3!\cdot\frac{b}{a+3b}\cdot\frac{b}{a+2b}\cdot\frac{a}{a+b}\cdot\frac{b}{b}
-=
-6\cdot\frac{b^2(1-b)}{(1+2b)(1+b)}
-\]
-\[
-\mbox{prob of coming fourth (last)} =
-3!\cdot\frac{b}{a+3b}\cdot\frac{b}{a+2b}\cdot\frac{b}{a+b}\cdot\frac{a}{a}
-=
-6\cdot\frac{b^3}{(1+2b)(1+b)}
-\]
-
-The sum would be $[(1+b)(1-b) + 3b(1-b) + 6b^2(1-b) +
-6b^3]/[(1+b)(1+2b)]=1$.
-
-Now consider the general case.  Again, $n$ clones of strength $b$ and
-one of strength $a=1-b$, $n+1$ competitors.
-
-\[
-\mbox{prob of coming first} =
-n!\cdot\frac{b}{a+nb}\cdot\frac{b}{nb}\cdot\frac{b}{(n-1)b}\cdots\frac{b}{b}
-=
-\frac{b}{1+(n-1)b}
-\]
-
-\[
-\mbox{prob of coming second} =
-n!\cdot\frac{b}{a+nb}\cdot\frac{a}{a+(n-1)b}\cdot\frac{b}{(n-1)b}\cdot\frac{b}{(n-2)b}\cdots\frac{b}{2b}\cdot\frac{b}{b}
-=n\cdot X
-\]
-
-\[
-\mbox{prob of coming third} =
-\frac{b}{a+nb}\cdot
-\frac{b}{a+(n-1)b}\cdot
-\frac{a}{a+(n-2)b}\cdot
-\frac{b}{(n-2)b}\cdots
-\frac{b}{2b}\cdot\frac{b}{b}
-=\frac{n!}{(n-2)!}\cdot X
-\]
-
-Now consider the case where there are $r$ clones coming ahead of the
-focal competitor:
-
-\[
-\mbox{prob of having $r$ clones ahead}
-=\frac{n!}{(n-r)!}\cdot X
-\]
-
-## Visuals
-
-Now some likelihood visuals.  Suppose $r=4,n=8$:
+Produce the rainbow-coloured likelihood plot
 
@@
 
-```{r shown4n8mle}
-MLE(r=4,n=8)
-```
+### Alternative production of rainbow figure, as a check {-}
 
@@
 
-# Formula 1: Ocon vs Gasly
-
-```{r maxvscheco}
-a <- read.table("formula1_2023.txt",header=TRUE)
-a <- a[,seq_len(ncol(a)-1)]
-d_ocon  <- as.numeric(as.matrix(a)["Ocon",])
-d_gasly <- as.numeric(as.matrix(a)["Gasly",])
-d_ocon [is.na(d_ocon )] <- 22
-d_gasly[is.na(d_gasly)] <- 22
-d_ocon
-d_gasly
-MLE(d_ocon ,22)
-MLE(d_gasly,22)
-```
-
-two sample test:
-
-```{r 2samptest}
-both <- c(d_ocon,d_gasly)
-bothmax <- MLE(both,22)
-
-d_ocon_max  <- MLE(d_ocon ,22)
-d_gasly_max <- MLE(d_gasly,22)
-f_vec_arn(bothmax,d_ocon ,22,log=TRUE)
-f_vec_arn(bothmax,d_gasly,22,log=TRUE)
-f_vec_arn(d_ocon_max ,d_ocon ,22,log=TRUE)
-f_vec_arn(d_gasly_max,d_gasly,22,log=TRUE)
-
-Lambda <- (
-   (f_vec_arn(d_gasly_max  ,d_ocon,22,log=TRUE)+f_vec_arn(d_gasly_max  ,d_gasly,22,log=TRUE)) -
-   (f_vec_arn(bothmax      ,d_ocon,22,log=TRUE)+f_vec_arn(bothmax      ,d_gasly,22,log=TRUE))
-)
-Lambda
-pchisq(-2*Lambda,df=1,lower.tail=FALSE)
-```
-
-# Formula 1: Formula 1 drivers: career arc
+# Formula 1: Checo's career arc
 
@@
 plot(x,y)
-summary(lm(y~x))
 ```
@@
 
-# in silico formula
-
-```{r insn}
-table(replicate(100,which(rrace3(pn=c(a=1,b=10),ps=c(a=0.9,b=0.1))=='a')))
-```
-
-```{r insilicoestmation,cache=TRUE}
-n <- 1000
-M <- matrix(0,n,4)
-alpha_true <- 4
-beta_true <- -6
-for(i in seq_len(n)){
-    x <- i/n
-    LO <- alpha_true + beta_true*x
-    p <- exp(LO)/(1+exp(LO))
-    nn <- round(runif(1,10,100))
-    M[i,1] <- x  # regressor
-    M[i,2] <- which(rrace3(pn=c(a=1,b=nn),ps=c(a=p,b=1-p))=='a')
-    M[i,3] <- nn+1
-    M[i,4] <- p
-}
-head(M)
-tail(M)
-```
-
-```{r definelikeinsilico}
-likeinsilico <- function(vec,M){
-  alpha <- vec[1]
-  beta  <- vec[2]
-  S <- 0
-  for(i in seq_len(nrow(M))){
-    x <- M[i,1]
-    LO <- alpha + beta * x
-    p <- exp(LO)/(1+exp(LO))
-    S <- S + f_vec_arn(p,M[i,2]-1,M[i,3]-1,log=TRUE)
-  }
-return(S)
-}
-```
-
-```{r testlikeininsilicco,cache=TRUE}
-likeinsilico(c(4,-6),M)
-likeinsilico(c(4,-6.5),M)
-likeinsilico(c(4,-5.5),M)
-```
-
-```{r optimize_insilicco,cache=TRUE}
-optim_ab <- optim(par=c(4,-6),likeinsilico,control=list(fnscale= -1),M=M)
-optim_ab
-```
-
-```{r makeZcontourdataset,cache=TRUE}
-a <- seq(from=3.8,to=4.2,by=0.01)
-b <- seq(from=-6.4,to=-5.6,by=0.01)
-Z <- as.matrix(expand.grid(a,b))
-Z <- apply(Z,1,function(x){likeinsilico(x,M=M)})
-Z <- matrix(Z,length(a),length(b))
-```
-
-```{r doZcontour}
-Z <- Z-max(Z)
-Z <- pmax(Z,-10)
-contour(a,b,Z,levels=-c(0:10,15,20,30))
-points(alpha_true,beta_true,pch=16,cex=4)
-jj <- optim_ab$par
-points(jj[1],jj[2],pch=4,cex=4,col='red')
-grid()
-persp(a,b,Z)
-```
-
-# Mathematical Olympiad
-
-```{r readoly}
-sametest <- function(d1,d2,n){
-  both <- c(d1,d2)
-  bothmax <- MLE(both,n)
-
-  d1max <- MLE(d1,n)
-  d2max <- MLE(d2,n)
-  Lambda <- (
-  (f_vec_arn(d2max  ,d1,n,log=TRUE)+f_vec_arn(d2max  ,d2,n,log=TRUE)) -
-  (f_vec_arn(bothmax,d1,n,log=TRUE)+f_vec_arn(bothmax,d2,n,log=TRUE))
-)
-  pchisq(-2*Lambda,df=1,lower.tail=FALSE)
-}
-
-a <- as.matrix(read.table("olympiad.txt"))
-nrow(a)
-d1 <- a["AUS",]
-d2 <- a["NZL",]
-d1
-d2
-sametest(d1,d2,100)
-a["USA",]
-a["CHN",]
-sametest(a["USA",],a["CHN",],100)
-sametest(a["USA",],a["CHN",],5)
-```
-
 # Athletics
@@
 
-# Parkrun
-
-```{r readresults}
-results <- read.table("parkrun_results.txt",header=TRUE)
-results
-```
-
-Above we see some parkrun results.  In parkrun number 148, I placed
-229 out of a field of 318.
-
-```{r calculate_parksupp,cache=TRUE}
-S_parkrun <- function(a){f_vec_arn(a,r=results$place-1,results$runners-1,log=TRUE)}
-(optparkrun <- optimize(S_parkrun, c(0.2,0.8),maximum=TRUE))
-
-```
-
-```{r calcparka,cache=TRUE}
-a <- seq(from=0.25,to=0.65,by=0.01)
-S_parkrun_a <- S_parkrun(a) - optparkrun$objective
-uniroot(function(a){S_parkrun(a) - optparkrun$objective+2},c(0.1,0.5))$root
-uniroot(function(a){S_parkrun(a) - optparkrun$objective+2},c(0.5,0.9))$root
-optparkrun$objective - S_parkrun(0.5)
-```
-
-```{r plotparkrunlike, fig.cap="Support curve for author's generalized Bradley-Terry strength.
-Curve is normalised so that the evaluate of $\\hat{a}\\simeq 0.448$ has a support of zero.
-Dotted line shows $\\mathcal{S}=-2$; intersection of this with the  curve shows that the support interval runs from about 0.315 to 0.567"}
-plotparkrun <- function(...){
-plot(a,S_parkrun_a,xlab="generalized Bradley-Terry strength",ylab="support",type='b',ylim=c(-5,0))
-abline(h=c(0,-2),lty=2)
-segments(x0=0.5,y0=-1.5,y1=0)
-segments(x0=optparkrun$maximum,y0=-1,y1=0)
-text(optparkrun$maximum,-0.8,expression(hat(a)==0.448),pos=2)
-text(0.5,-1.5,expression(H[o]:a==0.5),pos=2)
-}
-plotparkrun()
-pdf("plotparkrun.pdf")
-plotparkrun()
-dev.off()
-```
-
-Figure \@ref(fig:plotparkrunlike) shows a support curve for the
-author's generalized Bradley-Terry strength, normalized so that
-$\mathcal{S}(\hat{a})=0$.  We see a support interval [that is,
-$\left\lbrace a\colon\mathcal{S}(a)\geqslant -2\right\rbrace$] of
-about 0.315 to 0.567.  We might also observe that the support for
-$H_0\colon a=\frac{1}{2}$ [that is, that the author has a 50\% chance
-of running faster than a randomly chosen Parkrun competitor] is about
-0.35, less than Edwards's two units of support criterion.  We thus see
-some support for Hankin's assertion that he is "as fit as anyone
-else there".
-
-```{r calculatelikes1,cache=F}
-a <- seq(from=0.005,to=0.85,by=0.01)
-La <- f_vec_arn(a,r=results$place[1]-1,results$runners[1]-1,log=TRUE)
-La <- La - max(La,na.rm=TRUE)
-plot(a,La,type='b',ylab='fishy')
-abline(h=c(0,-2))
-```
-
-```{r splitresults}
-supp2 <- function(vec,place,runners){
-  a_stirling <- vec[1]
-  a_auckland <- vec[2]
-  out <- 0
-  for(i in 1:3){
-    out <- out + f_vec_a(a_stirling,r=place[i]-1,runners[i]-1,log=TRUE)
-  }
-  for(i in 4:10){
-    out <- out + f_vec_a(a_auckland,r=place[i]-1,runners[i]-1,log=TRUE)
-  }
-  return(out)
-}
-```
-
-```{r,label=calcbothlikes,cache=TRUE}
-x <- seq(from=0.1,to=0.8,by=0.01)
-jj <- as.matrix(expand.grid(x,x))
-L <- apply(jj,1,supp2,results$place,results$runners)
-L <- L - max(L,na.rm=TRUE)
-L <- matrix(L,length(x),length(x))
-```
-
-```{r,label=plotbothlikes}
-par(pty='s')
-contour(x,x,L,asp=1,xlim=range(x),ylim=range(x),levels=-(0:5))
-abline(0,1)
-```
-
-```{r, label=onedimopt}
-f <- function(a){supp2(c(a,a),results$place,results$runners)}
-a <- seq(from=0.2,to=0.6,by=0.01)
-alike <- sapply(a,f)
-plot(a,alike-max(alike,na.rm=TRUE),type='b',ylab='blong')
-abline(h=c(0,-2))
-optimize(f,c(0.1,0.6),maximum=TRUE)
-```
-
-```{r, label=twodimopt}
-f2 <- function(v){supp2(v,results$place,results$runners)}
-optim(par=c(0.6,0.3),f2,control=list(fnscale=-1))
-```
-
-```{r,label=showchisquareasymptotic}
-7380.314-7381.496
-pchisq(2*(7381.496-7380.314),df=1,lower.tail=FALSE)
-```
-
-Above shows that there is no evidence to suggest that the Stirling
-strength differs from the Auckland strength.  Now let's try a
-logistic-style regression:
-
-```{r logisticsupportparkrun,cache=FALSE}
-results <- read.table("parkrun_results.txt",header=TRUE)
-results
-supp_logistic <- function(vec){
-  alpha <- vec[1]
-  beta <- vec[2]
-  out <- 0
-  for(i in seq_len(nrow(results))){
-     LO <- alpha + beta*i
-     strength <- exp(LO)/(1+exp(LO))
-     out <- out + f_vec_arn(strength,results$place[i]-1,results$runners[i]-1,log=TRUE)
-  }
-  return(out+13400)
-}
-
-c(
-supp_logistic(c(0,0)),
-supp_logistic(c(0,0.01)),
-supp_logistic(c(0,-0.01))
-)
-
-```
-
-```{r optimlogist, cache=FALSE}
-optim(par=c(0,0),fn=supp_logistic,control=list(fnscale=-1))
-optim(par=c(0,0),fn=function(v){supp_logistic(vec=c(v[1],0))},control=list(fnscale=-1))
-```
-
-## Education
-
-The focal competitor
+## Education {-}
 

### 2024-03-22 inst/hankin-ejs.tex
typos

@@
 
-Using $\mathcal{S}=\log(B-1)-B\frac{r+1}{n}$ as an approximate support
+Using $\mathcal{S}'=\log(B-1)-B\frac{r+1}{n}$ as an approximate support
 thus incurs an error of only $\mathcal{O}(n^{-2})$.  Observing that
@@
 \caption{Log-likelihood curve for the Bradley-Terry strength of the
-  lane-1 runner, olympic qualifying heats 2012-2020 \label{plotolymp}}
+  lane-1 runner, Olympic qualifying heats 2012-2020 \label{plotolymp}}
 \end{figure}

### 2024-03-25 inst/hankin-ejs.tex
rewording to include actual event

@@
 rivals.  Here attention is restricted to heats, for which lane placing
-is random.  I consider every heat in Olympic running from 2004-2020 (a
-total of 41 heats, each of 8 or 9 competitors); only competitors'
-ranks are used.  Figure~\ref{plotolymp} shows a log-likelihood curve
-for the Bradley-Terry strength of the focal competitor, here the
-runner in the innermost lane.  We see that the evaluate of $0.574$ is
-very close to the neutral value of $0.5$, for which the support is
-only $0.16$, far short of Edwards's two units of support criterion for
-rejection~\cite{edwards1972}.  This dataset contains no evidence to
-support lane bias.
+is random.  I consider every heat in Men's 100m Olympic running from
+2004-2020 (a total of 41 heats, each of 8 or 9 competitors); only
+competitors' ranks are used.  Figure~\ref{plotolymp} shows a
+log-likelihood curve for the Bradley-Terry strength of the focal
+competitor, here the runner in the innermost lane.  We see that the
+evaluate of $0.574$ is very close to the neutral value of $0.5$, for
+which the support is only $0.16$, far short of Edwards's two units of
+support criterion for rejection~\cite{edwards1972}.  This dataset
+contains no evidence to support lane bias.
 
@@
 (that is, a logistic regression of generalized Bradley-Terry strength
-$a$ against time, with $\alpha$ and $\beta$ being linear
+$a$ against time $t$, with $\alpha$ and $\beta$ being linear
 coefficients).  Contours of equal support are shown in

### 2024-03-25 inst/hankin-ejs.Rmd
minor rewording
typo: change printed value of evaluate

@@
 
+Following is a `sage`-compatible input:
+
 ```
@@
 
-It returns this:
+`sage` returns this:
 
@@
 segments(x0=ahat,y0=-1.5,y1=0)
-text(ahat,-0.91,expression(hat(a)==0.573),pos=2)
+text(ahat,-0.91,expression(hat(a)==0.474),pos=2)
 abline(v=0.5,lty=3)

### 2024-03-28 inst/parkrun_idea.Rmd
whitespace

--- 2024-02-28 inst/ejs-sample.tex
@@
-\documentclass[ejs]{imsart}
+---
+title: "Underdispersion in the context of community athletics"
+output: bookdown::html_document2
+author: "Robin K. S. Hankin"
+bibliography: "hyper2.bib"
+---
 
-%% Packages
-\RequirePackage{amsthm,amsmath,amsfonts,amssymb}
-\RequirePackage[numbers]{natbib}
-%\RequirePackage[authoryear]{natbib}%% uncomment this for author-year citations
-\RequirePackage[colorlinks,citecolor=blue,urlcolor=blue]{hyperref}
-\RequirePackage{graphicx}
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = TRUE)
+```
 
-\arxiv{2010.00000}
-\startlocaldefs
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% Uncomment next line to change            %%
-%% the type of equation numbering           %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\numberwithin{equation}{section}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% For Axiom, Claim, Corollary, Hypothesis, %%
-%% Lemma, Theorem, Proposition              %%
-%% use \theoremstyle{plain}                 %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\theoremstyle{plain}
-\newtheorem{axiom}{Axiom}
-\newtheorem{claim}[axiom]{Claim}
-\newtheorem{theorem}{Theorem}[section]
-\newtheorem{lemma}[theorem]{Lemma}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% For Assumption, Definition, Example,     %%
-%% Notation, Property, Remark, Fact         %%
-%% use \theoremstyle{definition}            %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\theoremstyle{definition}
-\newtheorem{definition}[theorem]{Definition}
-\newtheorem*{example}{Example}
-\newtheorem*{fact}{Fact}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% For Case use \theoremstyle{remark}       %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\theoremstyle{remark}
-\newtheorem{case}{Case}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Please put your definitions here:        %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\endlocaldefs
+```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
+knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+```
 
-\begin{document}
-\begin{frontmatter}
-\title{A simplified Plackett-Luce likelihood function for rank orders}
-\runtitle{Simplified Plackett-Luce likelihood functions}
-%\thankstext{T1}{A sample additional note to the title.}
+To cite the `hyper2` package in publications, please use @hankin2017_rmd.
 
-\begin{aug}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Only one address is permitted per author. %%
-%% Only division, organization and e-mail is %%
-%% included in the address.                  %%
-%% Additional information can be included in %%
-%% the Acknowledgments section if necessary. %%
-%% ORCID can be inserted by command:         %%
-%% \orcid{0000-0000-0000-0000}               %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\author[A]{\fnms{Robin}~\snm{Hankin}\ead[label=e1]{hankin.robin@gmail.com}},
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Addresses                                %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\address[A]{Computer Science and Mathematics,
-University of Stirling\printead[presep={,\ }]{e1}}
+```{r deff}
+library("hyper2")
 
-\end{aug}
+f <- function(n,a){
+  p <- 1:n
+  p <- (p-min(p))/(max(p)-min(p))
+  jj <- exp(a*p)
+  jj/sum(jj)  
+}
 
-\begin{abstract}
+getlike <- function(v){
+  alpha <- v[1]
+  beta <- v[2]
+ c(
+tabulate(replicate(10000,which(rrace(setNames(f(26,alpha),letters))==letters[beta])),nbins=26)
+,alpha,beta)
+}
+```
 
-Here a simplification of the Plackett-Luce likelihood functions for
-order statistics is proposed.  We consider a field of $n+1$ entities
-that are ordered in some way [the preferred example is competitors in
-  a running race, ordered by time of crossing the finishing line].
-One entity is of particular interest to us: the ``focal entity'', or
-focal competitor.  We assign Placket-Luce strength of $a$ to the focal
-competitor and $b$ to all others; we require $a+b=1$.  This device
-furnishes a reasonable, intuitive, and computationally tractable
-likelihood function that I apply to a range of applications including
-Formula 1 motor racing and the International Mathematical Olympiad.
-The results are difficult to obtain with other methods.
-\end{abstract}
+```{r calculateLikemat,cache=TRUE}
 
-\begin{keyword}[class=MSC]
-\kwd[Primary ]{6202}
-\kwd{00X00}
-\kwd[; secondary ]{62F30}
-\end{keyword}
+alpha_try <- 0:6
+beta_try <- 1:26
 
-\begin{keyword}
-\kwd{First keyword}
-\kwd{second keyword}
-\end{keyword}
+M <- as.matrix(expand.grid(alpha=alpha_try, beta=beta_try))
+system.time(L <- t(apply(M,1,getlike)))
+L[,1:26] <- L[,1:26]/mean(rowSums(L[,1:26]))
+colnames(L) <- c(letters,'alpha','beta')
+head(L)
 
-\end{frontmatter}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Please use \tableofcontents for articles %%
-%% with 50 pages and more                   %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\tableofcontents
+```
 
-\section{Introduction}
+## Now 10 in-silico races:
+```{r doinsilico,cache=TRUE}
+alphatrue <- 3
+betatrue <- 4
+set.seed(0)
+D <- replicate(30,which(rrace(setNames(f(26,alphatrue),letters))==letters[betatrue]))
+sort(D)
+```
 
-Ranked data arises whenever multiple entities are sorted by some
-criterion; the preferred interpretation is a race in which the
-sufficient statistic is the order of competitors crossing the
-finishing line.  The Plackett-Luce likelihood
-function~\cite{luce1959,plackett1975} is a generalization of the
-Bradley-Terry model~\cite{bradley1952} in which non-negative strengths
-are assigned to each competitor; without loss of generality, if the
-finishing order is $1,2,\ldots,n$ and competitor $i$ has strength
-$p_i\geqslant 0$, then the Plackett-Luce likelihood function will be
-proportional to
+```{r trylike,cache=TRUE}
+likelihood <- apply(L[,D],1,prod)
+likelihood <- matrix(likelihood,length(alpha_try),length(beta_try))
+```
 
-\begin{equation}\label{plackettluce}
-\prod_{i=1}^n\frac{p_i}{\sum_{j=i}^np_j}
-\end{equation}
+```{r showlike}
+matplot(likelihood)
+filled.contour(alpha_try,beta_try,(likelihood))
+table(likelihood==0)
+plot(f(26,alphatrue))
+abline(v=betatrue)
+```
 
-\noindent where the strengths are normalised so $\sum p_i=1$.
-However, estimation of the strengths is somewhat difficult if $n$ is
-even moderately large: even with as few as 20 competitors following
-Zipf's law~\cite{zipf1949} [a typical assumption for Bradley-Terry
-  strengths], Hankin~\cite{hankin2017_rmd} shows that maximum
-likelihood strengths identify the correct ranking with a probability
-of essentially zero.  The situation is worse in cases such as the
-International Mathematical Olympiad, which requires us to consider
-over 100 competitors.
+\subsection{Park run}
 
-Consider a race between a competitor of Bradley-Terry strength $a$ and
-$n$ cloned competitors, each of strength $b=1-a$.  An observation is
-indexed by~$r$, the number of strength $b$ clones finishing ahead
-of~$a$, and because we have $n+1$ competitors $0\leqslant r\leqslant
-n$.  The initial field strength is $a+nb=n+(n-1)a$.  The Plackett-Luce
-likelihood function~\cite{plackett1975,luce1959} for this situation
-would be
+``Parkrun'' is a distributed community initiative that organises
+weekly timed 5 km runs/walks in parks worldwide~\cite{hindley2020}.  A
+typical event will have 200 participants.  Table~\ref{parkruntable}
+shows the author has completed a total of 15 parkruns to date, and
+from the first pair of numbers we see that 238 runners attended that
+particular parkrun, of whom the author placed 173.
 
-\begin{eqnarray}\label{likeforrn1}
-  \mathcal{L}_{r,n}(a) &=& 
-\frac{b}{a+ n   b}\cdot
-\frac{b}{a+(n-1)b}\cdot
-\frac{b}{a+(n-2)b}\cdots
-\frac{a}{a+(n-r)b}\\ \label{likeforrn2}
-&=& \frac{B-1}{(B+n-1)(B+n-2)\cdots(B+n-r-1)}\\ \label{likeforrn3}
-&=& \frac{(B-1)(B+n-r-2)!}{(B+n-1)!}
-\end{eqnarray}
+\begin{table*}[t]
+  \caption{Parkrun results}
+\label{parkruntable}
+\begin{tabular}{ccccccccccc}
+\hline
+rank   &173  & 165 & 172 & 199 & 181 & 229 & 177 & 222 & 206 & 142 \\
+runners&238  & 238 & 196 & 242 & 242 & 318 & 259 & 305 & 297 & 241 \\ \\ \\
+rank   & 118 & 224 & 128 & 115 & 183 &     &     &     &     &\\
+runners& 179 & 338 & 203 & 245 & 254 &     &     &     &     &\\
+\hline
+\end{tabular}
+\end{table*}
 
-\noindent where $B=1/b$.  Note carefully that these expressions are
-likelihoods, not probabilities, and do not sum to 1.  Asymptotic
-analysis of expression~\ref{likeforrn3} shows that
+We may consider the author to have an unknown generalized Bradley
+Terry strength $a$: Figure~\ref{parkrunsupport} shows a support curve
+for $a$ with the evaluate, $\hat{a}\simeq 0.448$ shown.  We also see
+two units of support~\cite{edwards1992} shown as a dotted line
+illustrating a support interval of about 0.315 - 0.567; we may be
+reasonably confident that the author's true strength lies in this
+range.  In particular, note that a reasonable $H_0\colon
+a=\frac{1}{2}$ may not be rejected, the support for $H_0$ being only
+susceptible to very minor improvement [$\simeq 0.351$] by moving to
+the evaluate.
 
-\begin{equation}\label{asymptotic}
-\log\mathcal{L}_{r,n}(a)=\log(B-1)-B\frac{r+1}{n} 
-+ K + \mathcal{O}\left(n^{-2}\right)
-\end{equation}
-
-Using $\mathcal{S}=\log(B-1)-B\frac{r+1}{n}$ as an approximate support
-thus incurs an error of only $\mathcal{O}(n^{-2})$.  Observing that
-$\partial^2\mathcal{S'}/\partial B^2=-(B-1)^{-2}<0$, the evaluate
-would be unique; we have
-$\hat{B}=\frac{n+r+1}{r+1}+\mathcal{O}\left(n^{-1}\right)$, or
-alternatively $\hat{a}=\frac{n}{n+r+1}$.  I present some numerical
-results below.
-
-Previous related work presented the case $n=2$ [in current notation].
-Considering the case $n=7$ gives us $7+1=8$ distinct likelihood
-functions, Figure~\ref{ninelikes}.  We see an unanticipated asymmetry
-between $r$ and $n-r$.  For $r=3$ one might expect that
-$\hat{a}=\frac{1}{2}$ on the grounds that the focal competitor's
-performance was median.  However, $\hat{a}\simeq 0.64$.
-
+```{asis}
 \begin{figure}[t]
-\includegraphics[width=4in]{ninelikes}  % produced by very_simplified_likelihood.Rmd
-\caption{Likelihood functions for $r=0(1)8$\label{ninelikes}}
+\includegraphics[width=4in]{plotparkrun}  % plotparkrun.pdf made in very_simplified_likelihood.Rmd
+\caption{Author's strength\label{parkrunsupport}}
 \end{figure}
-
-\begin{figure}[t]
-\includegraphics[width=4in]{dotprobs}  % produced by very_simplified_likelihood.Rmd
-\caption{Maximum likelihood estimator $\hat{a}$ for $n=1(1)10$ and $r=0(1)n$
-  \label{dotprobs}}
-\end{figure}
-
-\section{Use-cases}
-
-Formula 1 motor racing is an important and prestigious motor sport
-\citep{codling2017,jenkins2010}.  Season ranking is based on a points
-allocation system wherein competitors are awarded points based on race
-finishing order; points accumulate additively.  The overall
-competition winner is the competitor who accumulates the most points
-after the final race.  However, as argued by
-Hankin~\cite{hankin2023_formula1points}, the drivers' {\em ranks} are
-statistically sufficient for analysis using Plackett-Luce approaches.
-The methods discussed above may be used to assess Bradley-Terry
-strengths of Formula 1 racing driver Sergio P\'{e}rez (``Checo'') over
-his career from 2011 to 2023.  Checo's Formula 1 performance began
-somewhat inauspiciously in 2011 as second driver for Sauber for whom
-he finished no higher than P7, but his performance rose subsequently,
-culminating in his achievement of a startling 20 podium finishes in
-2022-3.  Noting that Checo raced against 73 distinct competitors [this
-  rendering estimation of the other players' Plackett-Luce strengths
-  effectively impossible], we nevertheless are in a position to
-estimate his career-span Bradley-Terry strength and estimate it as
-
-we consider the following inference
-problem:
-
-\begin{equation}
-  b = b(t) = \frac{e^{\alpha + \beta t}}{1+e^{\alpha + \beta t}},\qquad\alpha,\beta\in\mathbb{R}
-\end{equation}
-
-(that is, a logistic regression of generalizeBradley-Terry strength
-$b$ against time).
-
-There seems to be no easy analytical solution to
-this, although Equation~\ref{asymptotic} would give
-\begin{equation}
-  \label{asymptoticcheco}
-  \log\mathcal{L}(B)=m\log(B-1) -B\left(\frac{r_1+1}{n_1}+\ldots+\frac{r_m+1}{n_m}\right)
-\end{equation}
-
-over the whole $m$ seasons, where $r_i$ is Checo's rank in race
-
-Pierre Gasly and Esteban Ocon have a long-standing rivalry and
-comparisons between them typically compare disparate metrics such as
-number of podiums, and highest grid position in a season.
-
-\begin{acks}[Acknowledgments]
-The authors would like to thank the anonymous referees, an Associate
-Editor and the Editor for their constructive comments that improved the
-quality of this paper.
-\end{acks}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Supplementary Material, including data   %%
-%% sets and code, should be provided in     %%
-%% {supplement} environment with title      %%
-%% and short description. It cannot be      %%
-%% available exclusively as external link.  %%
-%% All Supplementary Material must be       %%
-%% available to the reader on Project       %%
-%% Euclid with the published article.       %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{supplement}
-\stitle{Title of Supplement A}
-\sdescription{Short description of Supplement A.}
-\end{supplement}
-\begin{supplement}
-\stitle{Title of Supplement B}
-\sdescription{Short description of Supplement B.}
-\end{supplement}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                  The Bibliography                       %%
-%%                                                         %%
-%%  imsart-???.bst  will be used to                        %%
-%%  create a .BBL file for submission.                     %%
-%%                                                         %%
-%%  Note that the displayed Bibliography will not          %%
-%%  necessarily be rendered by Latex exactly as specified  %%
-%%  in the online Instructions for Authors.                %%
-%%                                                         %%
-%%  MR numbers will be added by VTeX.                      %%
-%%                                                         %%
-%%  Use \cite{...} to cite references in text.             %%
-%%                                                         %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%% if your bibliography is in bibtex format, uncomment commands:
-\bibliographystyle{imsart-number} % Style BST file (imsart-number.bst or imsart-nameyear.bst)
-\bibliography{hyper2}       % Bibliography file (usually '*.bib')
-
-\end{document}
+```

### 2024-03-28 inst/parkrun_idea.Rmd
reparametrization

@@
 
-f <- function(n,a){
-  p <- 1:n
-  p <- (p-min(p))/(max(p)-min(p))
-  jj <- exp(a*p)
-  jj/sum(jj)  
+f <- function(n,alpha){
+  jj <- alpha^(1:n)
+  return(jj/sum(jj))
 }
@@
 
-alpha_try <- 0:6
+alpha_try <- seq(from=1,to=6,len=10)
 beta_try <- 1:26
@@
 system.time(L <- t(apply(M,1,getlike)))
+L_integer <- L  # counts
 L[,1:26] <- L[,1:26]/mean(rowSums(L[,1:26]))
@@
 head(L)
+```
 
+```{r makecontL}
+colSums(L)
+mL <- function(i){matrix(L[,i],length(alpha_try),length(beta_try))}
+filled.contour(alpha_try,beta_try,mL(1))
+filled.contour(alpha_try,beta_try,mL(5))
+filled.contour(alpha_try,beta_try,mL(10))
+filled.contour(alpha_try,beta_try,mL(15))
+filled.contour(alpha_try,beta_try,mL(20))
+filled.contour(alpha_try,beta_try,mL(26))
+```
+
+```{r}
+plot(table(L_integer[,1:26]))
 ```
@@
 ## Now 10 in-silico races:
-```{r doinsilico,cache=TRUE}
-alphatrue <- 3
-betatrue <- 4
-set.seed(0)
-D <- replicate(30,which(rrace(setNames(f(26,alphatrue),letters))==letters[betatrue]))
+```{r doinsilico,cache=FALSE}
+alphatrue <- 1.3
+betatrue <- 20   # 13th letter = "m"
+set.seed(1)
+D <- replicate(10,which(rrace(setNames(f(26,alphatrue),letters))==letters[betatrue]))
 sort(D)
+D <- 10:20
+D
 ```
 
-```{r trylike,cache=TRUE}
-likelihood <- apply(L[,D],1,prod)
+```{r trylikeinsilico,cache=FALSE}
+likelihood <- apply(L[,D],1,prod) + 0.00
 likelihood <- matrix(likelihood,length(alpha_try),length(beta_try))
@@
 ```{r showlike}
-matplot(likelihood)
+round(alpha_try,2)
+matplot(alpha_try,likelihood,type='b')
+matplot(beta_try,t(likelihood),type='b')
+matplot(alpha_try,log(likelihood)-log(max(likelihood)),type='b',ylim=c(-20,0))
+matplot(beta_try,log(t(likelihood))-log(max(likelihood)),type='b',ylim=c(-20,0))
 filled.contour(alpha_try,beta_try,(likelihood))
@@
 abline(v=betatrue)
+contour(alpha_try,beta_try,likelihood)
+abline(v=alphatrue)
+abline(h=betatrue)
+contour(alpha_try,beta_try,log(likelihood) - log(max(likelihood)),levels=-(0:10))
+abline(v=alphatrue)
+abline(h=betatrue)
 ```
@@
 
-``Parkrun'' is a distributed community initiative that organises
+"Parkrun" is a distributed community initiative that organises
 weekly timed 5 km runs/walks in parks worldwide~\cite{hindley2020}.  A
@@
 illustrating a support interval of about 0.315 - 0.567; we may be
-reasonably confident that the author's true strength lies in this
+reasonably confident that the author's true BT strength lies in this
 range.  In particular, note that a reasonable $H_0\colon

### 2024-03-28 inst/parkrun_idea.Rmd
save the matrix, optionally calculate from scratch

@@
  c(
-tabulate(replicate(10000,which(rrace(setNames(f(26,alpha),letters))==letters[beta])),nbins=26)
+tabulate(replicate(1000,which(rrace(setNames(f(26,alpha),letters))==letters[beta])),nbins=26)
 ,alpha,beta)
@@
 
-```{r calculateLikemat,cache=TRUE}
-
+```{r calculateLikemat,cache=FALSE}
 alpha_try <- seq(from=1,to=6,len=10)
 beta_try <- 1:26
-
+calculate_from_scratch <- FALSE
+if(calculate_from_scratch){
     M <- as.matrix(expand.grid(alpha=alpha_try, beta=beta_try))
     system.time(L <- t(apply(M,1,getlike)))
+    colnames(L) <- c(letters,'alpha','beta')
+    write.table(L,file="likelihoodmatrix.txt")
+} else {
+    L <- read.table("likelihoodmatrix.txt")
+}
 L_integer <- L  # counts
 L[,1:26] <- L[,1:26]/mean(rowSums(L[,1:26]))
-colnames(L) <- c(letters,'alpha','beta')
-head(L)
+head(L_integer)
 ```
@@
 ```{r}
-plot(table(L_integer[,1:26]))
+#plot(table(L_integer[,1:26]))
 ```

### 2024-03-28 inst/hankin-ejs.tex
typo

@@
 competitor, here the runner in the innermost lane.  We see that the
-evaluate of $0.574$ is very close to the neutral value of $0.5$, for
+evaluate of $0.474$ is very close to the neutral value of $0.5$, for
 which the support is only $0.16$, far short of Edwards's two units of

### 2024-03-28 inst/parkrun_idea.Rmd
tweaks

@@
 f <- function(n,alpha){
-  jj <- alpha^(1:n)
+  jj <- 1/alpha^(1:n)
   return(jj/sum(jj))
@@
 ```{r calculateLikemat,cache=FALSE}
-alpha_try <- seq(from=1,to=6,len=10)
+alpha_try <- seq(from=1,to=6,len=20)
 beta_try <- 1:26
@@
 ```{r doinsilico,cache=FALSE}
-alphatrue <- 1.3
-betatrue <- 20   # 13th letter = "m"
+alphatrue <- 2.3
+betatrue <- 4  
 set.seed(1)
@@
 sort(D)
-D <- 10:20
-D
 ```

### 2024-04-03 inst/hankin-ejs.tex
rewording

@@
 sensitive and private information.  Nevertheless, it is common for a
-student to possess some informative observations: the total enrollment
-in a specific class and their individual ranking within that class;
-table~\ref{educationtable} shows an apocryphal dataset.  What could a
-student infer from such data?
+student to possess some informative observations, for example the
+total enrollment in a specific class and their individual ranking
+within that class.  Table~\ref{educationtable} shows an apocryphal
+dataset: what could a student infer from such data?
 
@@
 \begin{supplement}
-\stitle{Very simplified likelihood}
-\sdescription{A terse rmarkdown document that  recreates the figures and salient results of the submission}
+  \stitle{R code source files}
+  \sdescription{
+    File {\tt hankin-ejs.Rmd} is a terse rmarkdown document that
+    recreates the figures and salient results of the submission.
+  }
 \end{supplement}

### 2024-04-05 inst/hankin-ejs.tex
typos

@@
 alternatively $\hat{a}=\frac{n}{n+r+1}$.  Compare this with the result
-of asking ``what is the probability $p$ that the focal competitor
+of asking ``what is the probability that the focal competitor
 beats a randomly chosen opponent?'' to which the binomial maximum
@@
 Previous related work presented the case $n=2$ [in current notation].
-Considering the case $n=7$ gives us $7+1=8$ distinct likelihood
+Considering the case $n=8$ gives us $8+1=9$ distinct likelihood
 functions, Figure~\ref{ninelikes}.  We see an unanticipated asymmetry

### 2024-04-05 inst/parkrun_idea.Rmd
cite underdispersion
bugfix: there was something wrong with read.table()

--- 2024-03-28 inst/parkrun_idea.Rmd
@@
 
+@paul1995 state that "underdispersion will be seen very rarely in
+real-life situations".
+
 ```{r deff}
@@
 
-```{r calculateLikemat,cache=TRUE}
-
-alpha_try <- seq(from=1,to=6,len=10)
+```{r calculateLikemat,cache=FALSE}
+alpha_try <- seq(from=1,to=6,len=20)
 beta_try <- 1:26
-
+calculate_from_scratch <- FALSE
+if(calculate_from_scratch){
     M <- as.matrix(expand.grid(alpha=alpha_try, beta=beta_try))
     system.time(L <- t(apply(M,1,getlike)))
+    colnames(L) <- c(letters,'alpha','beta')
+    saveRDS(L,file="likelihoodmatrix.Rdata")
+} else {
+    L <- readRDS("likelihoodmatrix.Rdata")
+}
 L_integer <- L  # counts
 L[,1:26] <- L[,1:26]/mean(rowSums(L[,1:26]))
-colnames(L) <- c(letters,'alpha','beta')
+head(L_integer)
 head(L)
@@
 ```{r doinsilico,cache=FALSE}
-alphatrue <- 1.3
-betatrue <- 20   # 13th letter = "m"
+alphatrue <- 2.3
+betatrue <- 11 
 set.seed(1)
-D <- replicate(10,which(rrace(setNames(f(26,alphatrue),letters))==letters[betatrue]))
+D <- replicate(17,which(rrace(setNames(f(26,alphatrue),letters))==letters[betatrue]))
 sort(D)
-D <- 10:20
-D
+plot(table(D))
 ```

### 2024-04-10 inst/formula1_start_vs_finish.Rmd
another plot for 2022

@@
 
-```{r}
+```{r maketwodotcharts}
 dotchart(mH2,pch=16)
 dotchart(log(mH2),pch=16)
+```
+
+```{r showanexampleoff,cache=TRUE}
+o <- likelam(lambda=0.9, topn=15,finish=finish,grid=grid)
+loglik(equalp(o),o)
+
+o <- likelam(lambda=0.95, topn=15,finish=finish,grid=grid)
+loglik(equalp(o),o)
+
 ```
@@
 ```
+
+```{r plotbigmat22}
+like <- bigmat[1,]
+like <- like - max(like)
+plot(lam,like,type='b')
+abline(h=c(0,-2))
+```

### 2024-04-12 inst/formula1_start_vs_finish.Rmd
typo

@@
 ```{r max_alpha,cache=TRUE}
-system.time(alpha_MLE <- optimize(function(a){f(a,topn=topn,finish=finish,grid=grid)[1]},c(0.95,0.99),maximum=TRUE))
+system.time(alpha_MLE <- optimize(function(a){f(a,topn=topn,finish=finish,grid=grid)[1]},c(0.90,0.95),maximum=TRUE))
 alpha_MLE

### 2024-04-12 inst/formula1_sequential_correlation.Rmd
start work on sequential correlatoin

---
title: "formula1: sequential correlation"
output: html_document
date: "2024-03-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
```

Formula 1 grand prix racing, 2022 season.  We wonder whether placing
well at a race helps the team in some way in the next race.  The
"help" might be an $\alpha$ weighting term for just the winners, or a
$\alpha^n$ term for placing $n$-th in the previous race.  I would have
thought $\alpha>1$ [the help accruing from a win is positive] but it
is possible that $\alpha<1$ which would correspond to it being a
hindrance in some way.  Perhaps winning leads to overconfidence or
complacency in the next one?

```{r loadfinish}
setwd("~/rstudio/hyper2/inst")
finish <- read.table("formula1_2023.txt",header=TRUE)
finish <- finish[,seq_len(ncol(finish)-1)]
finish[1:9,1:5]
```

```{r starthyper2}
f <- function(i){ # this is column 'i' of finish
    ordertable2supp(finish[,i,drop=FALSE])
}
g <- function(i){
    suppressWarnings(out <- as.numeric(finish[,i,drop=TRUE]))
    out[is.na(out)] <- length(out)
    names(out) <- rownames(finish)
    return(out)
}
```

```{r muckaboutwithlikelihood}
	weightedlikelihood <- function(alpha){
		out <- hyper3()
		for(i in seq(from=2,to=ncol(finish))){
			H <- f(i)
			previous <- g(i-1)
			H[names(previous)] <- as.weight(alpha^previous)
			out <- out + H
		}
		return(out)
}
```

```{r doweightedlike,cache=TRUE}
H <- weightedlikelihood(0.9)
mH <- maxp(H)
```

```{r showweightedlike}
mH
dotchart(mH)
loglik(mH,H)
```

```{r doweightedlike91,cache=TRUE}
H91 <- weightedlikelihood(0.91)
mH91 <- maxp(H91)
```

```{r showweightedlike91}
mH91
dotchart(mH91)
loglik(mH91,H91)
```

```{r definelikealpha}
likealpha <- function(alpha){
	H <- weightedlikelihood(alpha)
	mH <- maxp(H)
	return(loglik(mH,H))
}
```

```{r uselikealpha,cache=TRUE}
alpha <- seq(from=0.95,to=1.02,by=0.01)
likes <- sapply(alpha,likealpha)
```

```{r plotalpha}
plot(alpha,likes-max(likes),type='b')
abline(h=c(0,-2))
abline(v=1)
```

### 2024-04-12 inst/hankin-ejs.Rmd
whitespace

@@
 
-Following is a `sage`-compatible input:
+First we use Stirling's approximation $\log(n!)\sim
+(n+1/2)\log(n)-n+\log(2\pi)/2$ to simplify equation 4 of the manuscript
+Following is a `sage`-compatible input [the $\sqrt{2\pi}$ bit
+cancels]:
 
@@
 
+where $K$ is a constant.
+
 ## Definition of likelihood and log-likelihood in R {-}
+
+Below I give R idiom for the ikelihood expression given by equation 4
+in the manuscript.  Some inline comments are included.
 
@@
 }
+```
 
+```{r definefapprox}
 fapprox <- function(a,r,n,log=FALSE){
@@
 ```
+
+Above, `fapprox()` gives the asymptotic expression (5) in the
+manuscript.
 
@@
 
-Now verify `fdash()`:
+Notation is $\mathcal{L}$ for likelihood and
+$\mathcal{S}=\log\mathcal{L}$ for log-likelihood (support).  Above we
+see `fdash()` implements either $\frac{\partial\mathcal{S}}{\partial
+a}$ or $\frac{\partial\mathcal{L}}{\partial a}$ [depending on the
+value of argument `log`.  We see that
+
+$$
+\frac{\partial\mathcal{S}}{\partial a}=
+B^2\left(\frac{1}{B-1}-\psi(B+n)+\psi(B+n-r-1)\right)$$
+
+and $\frac{\partial\mathcal{L}}{\partial a}$ is of course just
+$\frac{\mathcal{L}\partial\mathcal{S}}{\partial a}$.  Now, numerical
+verification of `fdash()`:
 
@@
 
-## Maximum likelihood estimation
+Above we see very close agreemen between numerical and analytical
+results, for both `log=TRUE` and `log=FALSE`.
+
+## Maximum likelihood estimation {-}
 
@@
 
+Above we see very close agreement between the two methods.  We can get
+a feel for the accuracy of the asymptotic result numerically using
+similar methods:
+
 ```{r usestirling}
@@
 
-Now verification
+Now another verification, this time comparing the R implementation
+against two different direct numerical implementations of equation 4
+of the manuscript:
 
@@
 ```
+
+Above we see close agreement.
 
@@
 
-Pure first
+Following lines show optimization of the _marginal_ Bradley Terry
+strength for pure and applied mathematics:
 
@@
 plot(a,Lpure,ylab='pure strength')
-optimize(function(a){f_vec_arn(a,rank[wp]-1,class_size[wp]-1,log=TRUE)},c(0.3,0.6),maximum=TRUE)
-optimize(function(a){f_vec_arn(a,rank[wa]-1,class_size[wa]-1,log=TRUE)},c(0.3,0.6),maximum=TRUE)
+optimize(function(a){f_vec_arn(a,rank[wp]-1,class_size[wp]-1,log=TRUE)},c(0.1,0.9),maximum=TRUE)
+optimize(function(a){f_vec_arn(a,rank[wa]-1,class_size[wa]-1,log=TRUE)},c(0.1,0.9),maximum=TRUE)
 ```
+
+[the evaluates above do not appear in the manuscript] Now we define a
+likelihood function for the two subjects jointly:
 
@@
 
+Following code shows _joint_ maximization of likelihood:
+
 ```{r,label=defsupp2}
@@
 
+Following code shows _constrained_ maximization of likelihood
+[constrained, that is, to follow the diagonal null line]:
+
 ```{r,label=findmaxlike}
@@
 
+Extra support is thus $-86.9449 + 89.3732=2.4283$ (to 4 d.p.)  Now
+calculate the asymptotic $p$-value obtained from Wilks's theorem:
+
 ```{r,label=shochisw}

### 2024-04-15 inst/hankin-sts.tex
repurpose for Statistical Science (STS)

--- 2024-04-05 inst/hankin-ejs.tex
@@
 One entity is of particular interest to us: the ``focal entity'', or
-focal competitor.  We assign Placket-Luce strength $a$ to the focal
+focal competitor.  We assign Plackett-Luce strength $a$ to the focal
 competitor and $b$ to all others; we require $a+b=1$.  The associated
@@
 Zipf's law~\cite{zipf1949} [a typical assumption for Bradley-Terry
-  strengths], Hankin~\cite{hankin2017_rmd} shows that maximum
-likelihood strengths identify the correct ranking with a probability
-of essentially zero.  The situation is worse in cases such as formula
-1 motor racing considered below, which requires us to consider
-${\mathcal O}(100)$ competitors.
+  strengths] and say 20 independent complete rank observations,
+Hankin~\cite{hankin2017_rmd} shows that maximum likelihood strengths
+identify the correct ranking with a probability of essentially zero.
+The situation is worse in cases such as formula 1 motor racing
+considered below, which requires us to consider ${\mathcal O}(100)$
+competitors.
 
@@
 However, a constrained optimization shows that the maximum support
-along the null diagonal is about $-2.43$.  Edwards's two units of
-support criterion~\cite{edwards1972} is met, and we are justified in
-concluding that the student is indeed more competent at applied than
-pure mathematics.  Alternatively, we may observe that
+along the null diagonal is about $-2.43$, at
+${p}_\mathrm{pure}=0.567={p}_\mathrm{applied}=0.713$.  Edwards's two
+units of support criterion~\cite{edwards1972} is met, and we are
+justified in concluding that the student is indeed more competent at
+applied than pure mathematics.  Alternatively, we may observe that
 $2\mathcal{S}=4.86$ is in the tail region of its asymptotic $\chi^2_1$

### 2024-04-15 inst/hankin-sts.tex
repurpose for Statistical Science (STS)

@@
-\documentclass[ejs,noshowframe]{imsart}
+%% Template for the submission to:
+%%   Statistical Science [STS]
+%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% In this template, the places where you   %%
+%% need to fill in your information are     %%
+%% indicated by '???'.                      %%
+%%                                          %%
+%% Please do not use \input{...} to include %%
+%% other tex files. Submit your LaTeX       %%
+%% manuscript as one .tex document.         %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\documentclass[sts]{imsart}
 
@@
 %\RequirePackage[authoryear]{natbib}%% uncomment this for author-year citations
-\RequirePackage[colorlinks,citecolor=blue,urlcolor=blue]{hyperref}
-\RequirePackage{graphicx}
+%\RequirePackage[colorlinks,citecolor=blue,urlcolor=blue]{hyperref}%% uncomment this for coloring bibliography citations and linked URLs
+\RequirePackage{graphicx}%% uncomment this for including figures
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\theoremstyle{plain}
-\newtheorem{axiom}{Axiom}
-\newtheorem{claim}[axiom]{Claim}
-\newtheorem{theorem}{Theorem}[section]
-\newtheorem{lemma}[theorem]{Lemma}
+%\theoremstyle{plain}
+%\newtheorem{???}{???}
+%\newtheorem*{???}{???}
+%\newtheorem{???}{???}[???]
+%\newtheorem{???}[???]{???}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %% Notation, Property, Remark, Fact         %%
-%% use \theoremstyle{definition}            %%
+%% use \theoremstyle{remark}                %%
 %%                                          %%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\theoremstyle{definition}
-\newtheorem{definition}[theorem]{Definition}
-\newtheorem*{example}{Example}
-\newtheorem*{fact}{Fact}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% For Case use \theoremstyle{remark}       %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\theoremstyle{remark}
-\newtheorem{case}{Case}
+%\theoremstyle{remark}
+%\newtheorem{???}{???}
+%\newtheorem*{???}{???}
+%\newtheorem{???}{???}[???]
+%\newtheorem{???}[???]{???}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \endlocaldefs
@@
 \begin{document}
+
 \begin{frontmatter}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                          %%
+%% Enter the title of your article here     %%
+%%                                          %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \title{A simplified Plackett-Luce likelihood function for rank orders}
+%\title{A sample article title with some additional note\thanksref{T1}}
 \runtitle{Simplified Plackett-Luce likelihood functions}
-%\thankstext{T1}{A sample additional note to the title.}
+%\thankstext{T1}{A sample of additional note to the title.}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Only one address is permitted per author. %%
-%% Only division, organization and e-mail is %%
-%% included in the address.                  %%
-%% Additional information can be included in %%
-%% the Acknowledgments section if necessary. %%
 %% ORCID can be inserted by command:         %%
-%% \orcid{0000-0000-0000-0000}               %%
+\orcid{0000-0001-5982-0415}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\author[A]{\fnms{Robin}~\snm{Hankin}\ead[label=e1]{hankin.robin@gmail.com}},
+\author[A]{\fnms{Robin Hankin}~\snm{not sure what this is}\ead[label=e1]{hankin.robin@gmail.com}}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\address[A]{Computer Science and Mathematics,
-University of Stirling\printead[presep={,\ }]{e1}}
-
+\address[A]{University of Stirlling\printead[presep={\ }]{e1}.}
 \end{aug}
@@
 \begin{abstract}
+
 Here a simplification of the Plackett-Luce likelihood functions for
@@
 
-\begin{keyword}[class=MSC]
-\kwd[Primary ]{6202}
+\begin{keyword}
 \kwd{Statistical inference under constraints}
-\kwd[ ; secondary: Plackett-Luce likelihood functions; ]{62F30}
 \end{keyword}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Please use \tableofcontents for articles %%
-%% with 50 pages and more                   %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\tableofcontents
+%%%% Main text entry area:
 
@@
 \frac{b}{a+ n   b}\cdot
-\frac{b}{a+(n-1)b}\cdot
-\frac{b}{a+(n-2)b}\cdots
-\frac{a}{a+(n-r)b}\\ \label{likeforrn2}
-&=& \frac{B-1}{(B+n-1)(B+n-2)\cdots(B+n-r-1)}\\ \label{likeforrn3}
-&=& \frac{(B-1)(B+n-r-2)!}{(B+n-1)!}
+\frac{b}{a+(n-1)b}\cdots\frac{a}{a+(n-r)b}\nonumber\\
+&=& \frac{B-1}{(B+n-1)(B+n-2)\cdots(B+n-r-1)}\nonumber\\ 
+&=& \frac{(B-1)(B+n-r-2)!}{(B+n-1)!}\label{likeforrn3}
 \end{eqnarray}
@@
 \log\mathcal{L}_{r,n}(a)=\log(B-1)-B\frac{r+1}{n}
-+ K + \mathcal{O}\left(n^{-2}\right).
+\mathcal{O}\left(n^{-2}\right)
 \end{equation}
 
-Using $\mathcal{S}'=\log(B-1)-B\frac{r+1}{n}$ as an approximate support
-thus incurs an error of only $\mathcal{O}(n^{-2})$.  Observing that
+(up to an arbitrary additive constant).  Using
+$\mathcal{S}'=\log(B-1)-B\frac{r+1}{n}$ as an approximate support thus
+incurs an error of only $\mathcal{O}(n^{-2})$.  Observing that
 $\partial^2\mathcal{S'}/\partial B^2=-(B-1)^{-2}<0$, the evaluate
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% Single Appendix:                         %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\begin{appendix}
+%\section*{???}%% if no title is needed, leave empty \section*{}.
+%\end{appendix}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% Multiple Appendixes:                     %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\begin{appendix}
+%\section{???}
+%
+%\section{???}
+%
+%\end{appendix}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% Support information, if any,             %%
+%% should be provided in the                %%
+%% Acknowledgements section.                %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\begin{acks}[Acknowledgments]
+% The authors would like to thank ...
+%\end{acks}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% Funding information, if any,             %%
+%% should be provided in the                %%
+%% funding section.                         %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\begin{funding}
+% The first author was supported by ...
+%
+% The second author was supported in part by ...
+%\end{funding}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% Supplementary Material, including data   %%
+%% sets and code, should be provided in     %%
+%% {supplement} environment with title      %%
+%% and short description. It cannot be      %%
+%% available exclusively as external link.  %%
+%% All Supplementary Material must be       %%
+%% available to the reader on Project       %%
+%% Euclid with the published article.       %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\begin{supplement}
+%\stitle{???}
+%\sdescription{???.}
+%\end{supplement}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                  The Bibliography                       %%
+%%                                                         %%
+%%  imsart-???.bst  will be used to                        %%
+%%  create a .BBL file for submission.                     %%
+%%                                                         %%
+%%  Note that the displayed Bibliography will not          %%
+%%  necessarily be rendered by Latex exactly as specified  %%
+%%  in the online Instructions for Authors.                %%
+%%                                                         %%
+%%  MR numbers will be added by VTeX.                      %%
+%%                                                         %%
+%%  Use \cite{...} to cite references in text.             %%
+%%                                                         %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%% if your bibliography is in bibtex format, uncomment commands:
+%\bibliographystyle{imsart-number} % Style BST file (imsart-number.bst or imsart-nameyear.bst)
+%\bibliography{bibliography}       % Bibliography file (usually '*.bib')
+
+%% or include bibliography directly:
+% \begin{thebibliography}{}
+% \bibitem{b1}
+% \end{thebibliography}
+
 \end{document}

### 2024-04-15 inst/hankin-sts.tex
typos

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\author[A]{\fnms{Robin Hankin}~\snm{not sure what this is}\ead[label=e1]{hankin.robin@gmail.com}}
+\author[A]{\fnms{Robin K. S.}~\snm{Hankin}\ead[label=e1]{hankin.robin@gmail.com}}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\address[A]{University of Stirlling\printead[presep={\ }]{e1}.}
+\address[A]{University of Stirling\printead[presep={\ }]{e1}.}
 \end{aug}
@@
   \sdescription{
-    File {\tt hankin-ejs.Rmd} is a terse rmarkdown document that
+    File {\tt hankin-sts.Rmd} is a terse rmarkdown document that
     recreates the figures and salient results of the submission.

### 2024-04-16 inst/hankin-sts.tex
Plackett--Luce not Plackett-Luce

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\title{A simplified Plackett-Luce likelihood function for rank orders}
+\title{A simplified Plackett--Luce likelihood function for rank orders}
 %\title{A sample article title with some additional note\thanksref{T1}}
-\runtitle{Simplified Plackett-Luce likelihood functions}
+\runtitle{Simplified Plackett--Luce likelihood functions}
 %\thankstext{T1}{A sample of additional note to the title.}
@@
 
-Here a simplification of the Plackett-Luce likelihood functions for
+Here a simplification of the Plackett--Luce likelihood functions for
 order statistics is proposed.  We consider a field of $n+1$ entities
@@
 One entity is of particular interest to us: the ``focal entity'', or
-focal competitor.  We assign Plackett-Luce strength $a$ to the focal
+focal competitor.  We assign Plackett--Luce strength $a$ to the focal
 competitor and $b$ to all others; we require $a+b=1$.  The associated
@@
 sufficient statistic is the order of competitors crossing the
-finishing line.  The Plackett-Luce likelihood
+finishing line.  The Plackett--Luce likelihood
 function~\cite{luce1959,plackett1975} is a generalization of the
-Bradley-Terry model~\cite{bradley1952} in which non-negative strengths
+Bradley--Terry model~\cite{bradley1952} in which non-negative strengths
 are assigned to each competitor; without loss of generality, if the
 finishing order is $1,2,\ldots,n$ and competitor $i$ has strength
-$p_i\geqslant 0$, then the Plackett-Luce likelihood function will be
+$p_i\geqslant 0$, then the Plackett--Luce likelihood function will be
 proportional to
@@
 even moderately large: even with as few as 20 competitors following
-Zipf's law~\cite{zipf1949} [a typical assumption for Bradley-Terry
+Zipf's law~\cite{zipf1949} [a typical assumption for Bradley--Terry
   strengths] and say 20 independent complete rank observations,
@@
 
-Consider a race between a competitor of Bradley-Terry strength $a$ and
+Consider a race between a competitor of Bradley--Terry strength $a$ and
 $n$ cloned competitors, each of strength $b=1-a$.  An observation is
@@
 $0\leqslant r\leqslant n$.  The initial field strength is
-$a+nb=n+(n-1)a$.  The Plackett-Luce likelihood
+$a+nb=n+(n-1)a$.  The Plackett--Luce likelihood
 function~\cite{luce1959,plackett1975} for this situation would be
@@
 is random.  I consider every heat in Men's 100m Olympic running from
-2004-2020 (a total of 41 heats, each of 8 or 9 competitors); only
+2004--2020 (a total of 41 heats, each of 8 or 9 competitors); only
 competitors' ranks are used.  Figure~\ref{plotolymp} shows a
-log-likelihood curve for the Bradley-Terry strength of the focal
+log-likelihood curve for the Bradley--Terry strength of the focal
 competitor, here the runner in the innermost lane.  We see that the
@@
 \includegraphics[width=4in]{plotolymp} % produced by very_simplified_likelihood.Rmd
-\caption{Log-likelihood curve for the Bradley-Terry strength of the
+\caption{Log-likelihood curve for the Bradley--Terry strength of the
   lane-1 runner, Olympic qualifying heats 2012-2020 \label{plotolymp}}
@@
 Hankin~\cite{hankin2023_formula1points}, the drivers' {\em ranks} are
-statistically sufficient for analysis using Plackett-Luce approaches.
+statistically sufficient for analysis using Plackett--Luce approaches.
 The methods presented here may be used to assess competitive strengths
@@
 subsequently, culminating in his achievement of a startling 20 podium
-finishes in 2022-3.  Noting that Checo raced against 73 distinct
+finishes in 2022--3.  Noting that Checo raced against 73 distinct
 competitors (this rendering estimation of the other players'
-Plackett-Luce strengths effectively impossible~\cite{hankin2020}), we
+Plackett--Luce strengths effectively impossible~\cite{hankin2020}), we
 are nevertheless in a position to consider competitive strength using
@@
 
-(that is, a logistic regression of generalized Bradley-Terry strength
+(that is, a logistic regression of generalized Bradley--Terry strength
 $a$ against time $t$, with $\alpha$ and $\beta$ being linear
@@
 \caption{Log-likelihood contours for $\alpha,\beta$ for Checo's racing
-  career 2011-2023\label{showchecolike}}
+  career 2011--2023\label{showchecolike}}
 \end{figure}
@@
 
-Here a simplification of the Plackett-Luce likelihood functions for
+Here a simplification of the Plackett--Luce likelihood functions for
 order statistics was proposed, some simple properties obtained, and

### 2024-04-19 inst/kka.Rmd
typo

@@
 $(\lambda-1)(p_1-p_2)=0$, that is, if the white advantage is
-nonexistant, or if the players have the same strength.
+nonexistent, or if the players have the same strength.
 

### 2024-05-02 inst/parkrun_idea.Rmd
more thoughts on exponential strengths

@@
 real-life situations".
+
+Ranked data arises whenever multiple entities are sorted by some
+criterion; the preferred interpretation is a race in which the
+sufficient statistic is the order of competitors crossing the
+finishing line.  The Plackett--Luce likelihood function
+[@luce1959;@plackett1975] is a generalization of the Bradley--Terry
+model [@bradley1952] in which non-negative strengths are assigned to
+each competitor; without loss of generality, if the finishing order is
+$1,2,\ldots,n$ and competitor $i$ has strength $p_i\geqslant 0$, then
+the Plackett--Luce likelihood function will be proportional to
+
+\begin{equation}\label{plackettluce}
+\prod_{i=1}^n\frac{p_i}{\sum_{j=i}^np_j}
+\end{equation}
+
+\noindent where the strengths are normalised so $\sum p_i=1$.  Here we
+consider Plackett--Luce likelihood functions in which
+$p_i\propto\alpha^{-i}$ for some (generally unknown) $\alpha\geqslant
+1$.  The reason for the minus sign is so that, if $\alpha\gg 1$, the
+rank of competitor $i$ will be close to $i$.  Sometimes it will be
+convenient to parametrize in terms of $\beta=\alpha^{-1}$.
+
+Here, observations are restricted to knowledge of the rank of a single
+competitor.  Such observations are common in practice: examples would
+include an educational context in which a known number of people
+attempt an examination and the only information one actually has is
+one's own rank ["you came fifteenth out of a class of 85"]:
+confidentiality considerations mean that knowledge of classmates'
+performance is censored.  One might complete a foot race and learn
+that you placed, say, 194${}^\mathrm{th}$ out of 340 entries.  In this
+case the identities and abilities of the other runners are of no
+interest.  Other examples might include Formula 1 motor racing, in
+which a driver might race in multiple seasons in fields of 22-23:
+@hankin2024 notes that Sergio Perez has raced against a total of 73
+distinct competitors, rendering conventional Plackett-Luce techniques
+intractable.
+
+# Simple cases
+
+### Two competitors {-}
+The simplest non-trivial case would be two competitors with strengths
+proportional to $1,\alpha^{-1}$, the unit-sum constraint meaning that
+the strengths are
+$\left(\frac{\alpha}{1+\alpha},\frac{1}{1+\alpha}\right)$; these are
+the probabilities of competitor 1 coming first and second.
+
+### Three competitors {-}
+
+With three competitors we will do the general case first; let's say
+the strengths are $a,b,c, a+b+c=1$.
+
+Competitor $a$ comes first means the order was either $a\succ b\succ
+c$ or $a\succ c\succ b$, that is with probability
+
+$$
+\frac{a}{a+b+c}\cdot\frac{b}{b+c} + 
+\frac{a}{a+b+c}\cdot\frac{c}{b+c}
+=\frac{a}{a+b+c}
+$$
+
+Competitor $a$ comes second means $b\succ a\succ c$ or $c\succ a\succ b$, that is
+
+$$
+\frac{b}{a+b+c}\cdot\frac{a}{a+c} + 
+\frac{c}{a+b+c}\cdot\frac{a}{a+b}
+$$
+
+Competitor $a$ comes third means $b\succ c\succ a$ or $c\succ b\succ a$, that is
+
+$$
+\frac{b}{a+b+c}\cdot\frac{c}{a+c} + 
+\frac{c}{a+b+c}\cdot\frac{b}{a+b}
+$$
+
+Now the strengths of $a,b,c$ are are proportional to
+$1,\alpha^{-1},\alpha^{-2}$ respectively, so they are
+$\left(\frac{\alpha^2}{1+\alpha+\alpha^2},\frac{\alpha}{1+\alpha+\alpha^2},\frac{1}{1+\alpha+\alpha^2}\right)$.
+
+**Competitor ${\mathbf a}$** comes first with probability
+$\frac{\alpha^2}{1+\alpha+\alpha^2}$.  Second with probability
+
+$$
+\frac{\alpha(1+2\alpha^2+\alpha^3)}{(1+\alpha)(1+\alpha^2)(1+\alpha+\alpha^2)}
+$$
+
+[interesting observation: this has a maximum at $a\simeq 1.35271>1$ of
+about $0.3466 > \frac{1}{3}$].  He comes last with a probability of
+
+$$
+\frac{1+\alpha+2\alpha^2}{(1+\alpha)(1+\alpha^2)(1+\alpha+\alpha^2)}
+$$
+
+**Competitor ${\mathbf b}$** comes first with probability
+$\frac{\alpha}{1+\alpha+\alpha^2}$.  Second with probability (after
+simplification) of
+
+$$
+\frac{1+\alpha^3}{(1+\alpha)(1+\alpha+\alpha^2)}
+$$
+
+and third with probability $\frac{\alpha}{1+\alpha+\alpha^2}$.
+
+**Competitor ${\mathbf c}$** comes first with probability
+$\frac{1}{1+\alpha+\alpha^2}$.  Second with probability (after
+simplification) of
+
+$$
+\frac{\alpha(1+2\alpha+\alpha^3)}{(1+\alpha)(1+\alpha^2)(1+\alpha+\alpha^2)}
+$$
+
+and last with probability
+
+$$
+\frac{\alpha^3(2+\alpha+\alpha^2)}{(1+\alpha)(1+\alpha^2)(1+\alpha+\alpha^2)}
+$$
 
@@
 
-## Now 10 in-silico races:
+## Now some in-silico races
+
 ```{r doinsilico,cache=FALSE}
-alphatrue <- 2.3
+alphatrue <- 1.3
 betatrue <- 11 
 set.seed(1)
-D <- replicate(17,which(rrace(setNames(f(26,alphatrue),letters))==letters[betatrue]))
+D <- replicate(7,which(rrace(setNames(f(26,alphatrue),letters))==letters[betatrue]))
 sort(D)
@@
 ```{r trylikeinsilico,cache=FALSE}
-likelihood <- apply(L[,D],1,prod) + 0.00
+likelihood <- apply(0.00+L[,D],1,prod) + 0.00
 likelihood <- matrix(likelihood,length(alpha_try),length(beta_try))
@@
 round(alpha_try,2)
-matplot(alpha_try,likelihood,type='b')
-matplot(beta_try,t(likelihood),type='b')
+matplot(alpha_try,likelihood*1e14,type='b')
+matplot(beta_try,t(likelihood)*1e14,type='b')
 matplot(alpha_try,log(likelihood)-log(max(likelihood)),type='b',ylim=c(-20,0))
 matplot(beta_try,log(t(likelihood))-log(max(likelihood)),type='b',ylim=c(-20,0))
-filled.contour(alpha_try,beta_try,(likelihood))
+filled.contour(alpha_try,beta_try,(likelihood)*1e14)
 table(likelihood==0)
@@
 abline(h=betatrue)
+```
+
+```{r try2}
+likelihood <- apply(L[,c(11,18)],1,prod)
+likelihood <- matrix(likelihood,length(alpha_try),length(beta_try))
+contour(alpha_try,beta_try,likelihood)
+contour(-1/alpha_try,beta_try,likelihood)
+filled.contour(
+    alpha_try,beta_try,log(likelihood)-max(log(likelihood )),
+    color.palette = function(n) hcl.colors(n, "YlOrRd", rev = TRUE)
+)
+
 ```

### 2024-05-08 inst/paper.tex
Add JSS's paper.tex as per issue #153

--- 2023-10-29 inst/jss4832.tex
@@
 \documentclass[article]{jss}
+
+\usepackage{orcidlink,thumbpdf,lmodern}
+\graphicspath{{Figures/}}
 
@@
 \Abstract{
-
-  The \code{hyper2} package provides functionality to work with
+  The \pkg{hyper2} package provides functionality to work with
   extensions of the Bradley-Terry probability model such as
@@
   entities (monsters).  The package allows one to use relatively
-  natural R idiom to manipulate such likelihood functions.  Here, I
+  natural \proglang{R}~idiom to manipulate such likelihood functions.  Here, I
   present a generalization of \code{hyper2} in which multiple entities
   are constrained to have identical Bradley-Terry strengths.  A new
-  \proglang{S3} class \code{hyper3}, along with associated methods, is
+  \proglang{S3} class `\code{hyper3}', along with associated methods, is
   motivated and introduced.  Three datasets are analysed, each
   analysis furnishing new insight, and each highlighting different
-  capabilities of the package.  } \Keywords{Plackett-Luce,
-  Bradley-Terry, Mann-Whitney} \Plainkeywords{Plackett-Luce,
-  Bradley-Terry, Mann-Whitney}
+  capabilities of the package.  }
+
+\Keywords{Plackett-Luce, Bradley-Terry, Mann-Whitney}
+\Plainkeywords{Plackett-Luce, Bradley-Terry, Mann-Whitney}
 
@@
 
-The \code{hyper2} package~\citep{hankin2009,hankin2017} furnishes
+The \pkg{hyper2} package \citep{hankin2009,hankin2017} furnishes
 computational support for generalized
-Plackett-Luce~\citep{plackett1975} likelihood functions.  The
+Plackett-Luce \citep{plackett1975} likelihood functions.  The
 preferred interpretation is a race (as in track and field athletics):
@@
 $p_1,\ldots, p_6$; the probability that $i$ beats $j$ is
-$p_i/(p_i+p_j)$.  It is conventional to normalise so that the total
+$p_i/(p_i+p_j)$.  It is conventional to normalize so that the total
 strength is unity, and to identify a competitor with his strength.
@@
 Plackett-Luce likelihood function would be
-
+%
 \begin{equation}\label{PL_like}
@@
 \end{equation}
-
+%
 \citet{mollica2014} call this a forward ranking process on the grounds
@@
 available in a range of languages.  \cite{hunter2004}, for example,
-presents results in \proglang{MATLAB} [although he works with a
-nonlinear extension to account for ties]; \cite{allison1994} present
+presents results in \proglang{MATLAB} (although he works with a
+nonlinear extension to account for ties); \cite{allison1994} present
 related work for ranking statistics in \proglang{SAS} and
-\cite{maystre2022} has released a \proglang{python} package for
+\cite{maystre2022} has released a \proglang{Python} package for
 Luce-type choice datasets.
 
-However, the majority of software is written in the \proglang{R}
-computer language~\citep{rcore2023}, which includes extensive
+However, the majority of software is written in the \proglang{R}~computer language \citep{rcore2023}, which includes extensive
 functionality for working with such likelihood functions:
@@
 perspective.  The \pkg{BradleyTerry} package \citep{firth2005}
-considers pairwise comparisons using \proglang{glm} but cannot deal
+considers pairwise comparisons using \code{glm} but cannot deal
 with ties or player-specific predictors; the \pkg{BradleyTerry2}
@@
 interface to create and manipulate likelihood functions over the
-simplex ${S}_n=\left\lbrace\left(p_1,\ldots,p_n\right)\left|p_i\geq
-0,\sum p_i=1\right.\right\rbrace$.  A further extension in the package
+simplex ${S}_n=\left\lbrace\left(p_1,\ldots,p_n\right)\mid p_i\geq
+0,\sum p_i=1\right\rbrace$.  A further extension in the package
 generalizes this likelihood function to functions of ${\mathbf
 p}=(p_1,\ldots,p_n)$ with
-
+%
 \begin{equation}\label{hyper2likelihood}
@@
 \end{equation}
-
-\noindent where~$\mathcal{O}$ is a set of observations and~$s$ a
+%
+where~$\mathcal{O}$ is a set of observations and~$s$ a
 subset of~$\left\{1,2,\ldots,n\right\}$; numbers~$n_s$ are integers
@@
 thereby admitting a wide range of non-standard nulls such as order
-statistics on the $p_i$~\citep{hankin2017}.  It becomes possible to
+statistics on the $p_i$ \citep{hankin2017}.  It becomes possible to
 analyse a wider range of likelihood functions than standard
-Plackett-Luce~\citep{turner2020}.  For example, results involving
+Plackett-Luce \citep{turner2020}.  For example, results involving
 incomplete order statistics or teams are tractable.  Further, the
 introduction of reified entities (monsters) allows one to consider
-draws~\citep{hankin2009}, noncompetitive tactics such as
-collusion~\citep{hankin2020}, and the phenomenon of team cohesion
+draws \citep{hankin2009}, noncompetitive tactics such as
+collusion \citep{hankin2020}, and the phenomenon of team cohesion
 wherein the team becomes stronger than the sum of its
-parts~\citep{hankin2010}.  Recent versions of the package include
+parts \citep{hankin2010}.  Recent versions of the package include
 experimental functionality (\code{cheering()}) to investigate the
@@
 there are six runners 1-6 but we happen to know that three of the
-runners (1,2,3) are clones of strength $p_a$, two of the runners (4,5)
+runners (1, 2, 3) are clones of strength $p_a$, two of the runners (4, 5)
 have strength $p_b$, and the final runner (6) is of strength $p_c$.
-We normalise so $p_a+p_b+p_c=1$.  The runners race and the finishing
+We normalize so $p_a+p_b+p_c=1$.  The runners race and the finishing
 order is:
-
+%
 $$a\succ c\succ b\succ a\succ a \succ b$$
-
+%
 Thus the winner was $a$, second place for $c$, third for $b$, and so
@@
 fifth; $b$ came third and sixth, and $c$ came second.  The
-Plackett-Luce likelihood function for $p_a,p_b,p_c$ would be
-
+Plackett-Luce likelihood function for $p_a, p_b, p_c$ would be
+%
 \begin{equation}\label{plackettluce}
@@
 \end{equation}
-
+%
 Here I consider such generalized Plackett-Luce likelihood functions,
@@
 method that used a complicated mapping system from integers to
-competitor names.  Current methodology [following commit
-  \code{51a8b46}] is to use sets of character strings which represent
+competitor names.  Current methodology (following commit
+  \code{51a8b46}) is to use sets of character strings which represent
 the competitors directly; this allows for easier combination of
 observations including different competitors.}.  The declarations
-
+%
 \begin{verbatim}
@@
 \end{verbatim}
-
-show how the \proglang{map} class of the Standard Template Library is
-used with \proglang{weightedplayervector} objects mapping strings to
+%
+show how the `\code{map}' class of the Standard Template Library is
+used with `\code{weightedplayervector}' objects mapping strings to
 long doubles (specifically, mapping player names to their
-multiplicities), and objects of class \proglang{hyper3} are maps from
-a \proglang{weightedplayervector} object to long doubles.  One
+multiplicities), and objects of class `\code{hyper3}' are maps from
+a `\code{weightedplayervector}' object to long doubles.  One
 advantage of this is efficiency: Search, removal, and insertion
@@
 the first term in Equation~\ref{plackettluce}:
-
+%
 \begin{verbatim}
@@
 \end{verbatim}
-
+%
 Above, we understand \code{n} and \code{d} to represent numerator and
 denominator respectively.  Object \code{L} is an object of class
-\code{hyper3}; it may be evaluated at points in probability space
-[that is, a vector \code{[a,b,c]} of nonnegative values with unit sum]
+`\code{hyper3}'; it may be evaluated at points in probability space
+(that is, a vector \code{[a,b,c]} of nonnegative values with unit sum)
 using standard \proglang{R} idiom wrapping \proglang{C++} back end.
@@
 
-The package includes an \proglang{S3} class \code{hyper3} for this
+The package includes an \proglang{S}3 class `\code{hyper3}' for this
 type of object; extraction and replacement methods use \pkg{disordR}
-discipline~\citep{hankin2022}.  Package idiom for creating such
+discipline \citep{hankin2022}.  Package idiom for creating such
 objects uses named vectors:
-
+%
 \begin{Schunk}
@@
 \end{Schunk}
-
+%
 Above, we see object \code{LL} is a log-likelihood function of the
@@
 \code{loglik()}:
-
+%
 \begin{Schunk}
@@
 \end{Schunk}
-
+%
 Thus we prefer $H_1$ over $H_2$ with about 3.5 units of support,
 satisfying the standard two units of support
-criterion~\citep{edwards1972}, and we conclude that our observation
+criterion \citep{edwards1972}, and we conclude that our observation
 [in this case, that one of the three clones of player $a$ beat the $b$
 twins and the singleton $c$] furnishes strong support against $H_2$ in
-favour of $H_1$.
+favor of $H_1$.
 
@@
 Equation~\ref{PL_like}:
-
+%
 \begin{Schunk}
@@
 \end{Schunk}
-
+%
 (the package gives extensive documentation at
@@
 dispatching to a specialist \code{hyper3} method:
-
+%
 \begin{Schunk}
@@
 \end{Schunk}
-
+%
 (function \code{maxp()} uses standard optimization techniques to
@@
 The package provides a number of statistical tests on likelihood
-functions, modified from~\citet{hankin2017} to work with \code{hyper3}
+functions, modified from~\citet{hankin2017} to work with `\code{hyper3}'
 objects.  For example, we may assess the hypothesis that all three
-players are of equal strength [viz $H_0\colon
-p_a=p_b=p_c=\frac{1}{3}$]:
-
+players are of equal strength (viz $H_0\colon
+p_a=p_b=p_c=\frac{1}{3}$):
+%
 \begin{Schunk}
@@
 \end{Schunk}
-
+%
 showing, perhaps unsurprisingly, that this small dataset is consistent
@@
 
-Arithmetic operations are implemented for \code{hyper3} objects in
-much the same way as for \code{hyper2} objects: independent
+Arithmetic operations are implemented for `\code{hyper3}' objects in
+much the same way as for `\code{hyper2}' objects: independent
 observations may be combined using the overloaded \code{+} operator;
@@
 
-The original motivation for \code{hyper3} was the analysis of Formula
-1 motor racing, and the package accordingly includes wrappers for
+The original motivation for \code{hyper3} was the analysis of
+Formula~1 motor racing, and the package accordingly includes wrappers for
 \code{ordervec2supp()} such as \code{ordertable2supp3()} and
@@
 
-\section{Exact analytical solution for some simple generalized
-Plackett-Luce likelihood functions}
+\section{Exact analytical solution for some simple generalized Plackett-Luce likelihood functions}
 
@@
 $a\succ b\succ a$.  The Plackett-Luce likelihood function would be
-
+%
 \begin{equation}\label{aba}
@@
 \end{equation}
-
+%
 and in this case we know that $a+b=1$ so this is equal to
@@
 given by
-
+%
 \begin{equation}\label{aba_mle}
@@
 \end{equation}
-
+%
 and this will be zero at $\sqrt{2}-1$; we also note that
@@
 $0\leq a\leq 1$: the root is a maximum.
-
+%
 \begin{Schunk}
@@
 \end{Schunk}
-
+%
 Above, we see close agreement with the theoretical value of
@@
 we have
-
+%
 \begin{equation}
@@
 \end{equation}
-
+%
 as a standard support function which has a maximum value of zero when
@@
 Bradley-Terry strengths:
-
+%
 \begin{Schunk}
 \begin{Sinput}
-R> a <- 1/2 # null
+R> a <- 1/2
 R> (S_delta <- log(a * (1 - a)/(1 + a)) - log(3 - 2 * sqrt(2)))
@@
 \end{Schunk}
-
+%
 Thus the additional support gained in moving from $a=\frac{1}{2}$ to
@@
 \cite{edwards1972} and apply Wilks's theorem for a $p$~value:
-
+%
 \begin{Schunk}
@@
 \end{Schunk}
-
+%
 The $p$~value is about 0.81, exceeding 0.05; thus we have no strong
@@
 Figure~\ref{ABA_likelihood}.
-
-\begin{figure}[htbp]
-  \begin{center}
+%
 \begin{Schunk}
@@
 R> S <- function(a){log(a * (1 - a) / ((1 + a) * (3 - 2 * sqrt(2))))}
-R> plot(a, S(a), type = 'b',xlab=expression(p[a]),ylab="support")
+R> plot(a, S(a), type = "b", xlab = expression(p[a]), ylab = "support")
 R> abline(h = c(0, -2))
-R> abline(v = c(0.02438102, 0.9524271), col = 'red')
+R> abline(v = c(0.02438102, 0.9524271), col = "red")
 R> abline(v = sqrt(2) - 1)
@@
 \end{Schunk}
-\includegraphics{jss_hyper3-figaba}
+%
+\begin{figure}[t!]
+\centering
+\includegraphics[trim = 0 5 0 20,clip]{jss_hyper3-figaba}
 \caption{A\label{ABA_likelihood} support function for $p_a$ with observation $a\succ b\succ a$.}
-\end{center}
 \end{figure}
@@
 \subsubsection{Fisher information}
+
 If we have two clones of $a$ and a singleton $b$, then there are three
@@
 Fisher information for such observations is
-
+%
 \begin{equation}
@@
 \end{equation}
-
+%
 which has a minimum of about $6.21$ at at about $a=0.522$.  We can
@@
 
-\begin{figure}[htbp]
-  \begin{center}
-\includegraphics{jss_hyper3-figabbabb}
+\begin{figure}[t!]
+\centering
+\includegraphics[trim = 0 5 0 20,clip]{jss_hyper3-figabbabb}
 \caption{Likelihood \label{ABAetc_likelihood} functions for
@@
 line represents two units of support}
-\end{center}
 \end{figure}
@@
 {\em top $n$ ranking}), there is another nontrivial order statistic,
-viz $a\succ b\succ\left\lbrace a,b\right\rbrace$ [thus one of the two
+viz $a\succ b\succ\left\lbrace a,b\right\rbrace$ (thus one of the two
   $a$'s won, one of the $b$'s came second, and one of each of $a$ and
-  $b$ failed to finish].  Now
-
+  $b$ failed to finish.  Now
+%
 \begin{equation}
@@
 \end{equation}
-
+%
 (see how the likelihood function is actually simpler than for the
@@
 0.586$:
-
+%
 \begin{Schunk}
 \begin{Sinput}
-R> maxp(ordervec2supp3(c("a", "b"), nonfinishers=c("a", "b")))
+R> maxp(ordervec2supp3(c("a", "b"), nonfinishers = c("a", "b")))
 \end{Sinput}
@@
 \end{Schunk}
-
+%
 The Fisher information for such observations has a minimum of
@@
 
-\section{An alternative to the Mann-Whitney test using generalized
-Plackett-Luce likelihood}
+\section{An alternative to the Mann-Whitney test using generalized Plackett-Luce likelihood}
 
@@
 order statistic of the type considered above, the
-Mann-Whitney-Wilcoxon test~\citep{mann1947,wilcoxon1945} assesses a
-null of identity of underlying distributions~\citep{ahmad1996}.
-Consider the chorioamnion dataset~\citep{hollander2013}, used in
+Mann-Whitney-Wilcoxon test \citep{mann1947,wilcoxon1945} assesses a
+null of identity of underlying distributions \citep{ahmad1996}.
+Consider the chorioamnion dataset \citep{hollander2013}, used in
 \code{wilcox.test.Rd}:
-
+%
 \begin{Schunk}
@@
 \end{Schunk}
-
+%
 Here we see a measure of permeability of the human placenta at term
@@
 The order statistic is straightforward to calculate:
-
+%
 \begin{Schunk}
@@
 \end{Schunk}
-
-Then object \code{os} is converted to a \code{hyper3} object, again
-with \code{ordervec2supp3()}, which may be assessed using the Method
-of Support:
-
+%
+Then object \code{os} is converted to a `\code{hyper3}' object, again
+with \code{ordervec2supp3()}, which may be assessed using the method
+of support:
+%
 \begin{Schunk}
@@
 \end{Schunk}
-
+%
 Above, we use generic function \code{equalp.test()} to test the null
@@
 
-\begin{figure}[htbp]
-  \begin{center}
-\includegraphics{jss_hyper3-plotwilcoxlike}
+\begin{figure}[t!]
+\centering
+\includegraphics[trim = 0 5 0 40,clip]{jss_hyper3-plotwilcoxlike}
 \caption{A\label{wilcox_likelihood} support function for the Bradley-Terry
@@
 exclude $H_0\colon p_a=0.5$ (dotted line), is also shown.}
-\end{center}
 \end{figure}
@@
 javelin, 2020 Olympics:
-
+%
 \begin{Schunk}
@@
 \end{Schunk}
-
+%
 Thus Chopra threw 87.03m on his first throw, 87.58m on his second, and
@@
 \code{attemptstable2supp3()}:
-
+%
 \begin{Schunk}
@@
 \end{Schunk}
-
+%
 Above we see that Chopra threw the longest and second-longest throws
@@
 throw of 86.67m, and so on (\code{NA} entries correspond to
-no-throws.)  The attempts table may be converted to a \code{hyper3}
+no-throws.)  The attempts table may be converted to a `\code{hyper3}'
 object, again using function \code{attemptstable2supp3()} but this
-time passing \code{give.supp=TRUE}:
-
+time passing \code{give.supp = TRUE}:
+%
 \begin{Schunk}
@@
 \end{Schunk}
-
+%
 Above, object \code{javelin} is a \code{hyper3} likelihood function,
@@
 of $\frac{1}{8}$ ($p=0.1$).
-
-\begin{figure}[htbp]
-  \begin{center}
+%
 \begin{Schunk}
@@
 \begin{Sinput}
-R> dotchart(mj, pch = 16,xlab="Estimated Bradley-Terry strength")
+R> dotchart(mj, pch = 16, xlab = "Estimated Bradley-Terry strength")
 \end{Sinput}
 \end{Schunk}
-\includegraphics{jss_hyper3-testthejav}
+%
+\begin{figure}[t!]
+\centering
+\includegraphics[trim = 0 5 0 40,clip]{jss_hyper3-testthejav}
 \caption{Maximum \label{javelinbradleyterry} likelihood estimate for javelin throwers' Bradley-Terry strengths.}
-\end{center}
 \end{figure}
@@
 training or selection perspective we might follow \cite{hankin2017}
-and observe that log-contrasts~\citep{ohagan2004} appear to have
+and observe that log-contrasts \citep{ohagan2004} appear to have
 approximately Gaussian likelihood functions for observations of the
@@
 
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics{jss_hyper3-plottheloglikcont}
+\begin{figure}[t!]
+\centering
+\includegraphics[trim = 0 5 0 20,clip]{jss_hyper3-plottheloglikcont}
 \caption{Profile \label{profliklogcont} likelihood for log-contrast
@@
 thus the null is not rejected.}
-\end{center}
 \end{figure}
@@
 use a generalized Plackett-Luce model to assess the constructors'
-performance.  The following table, included in the \code{hyper2}
+performance.  The following table, included in the \pkg{hyper2}
 package as a dataset, shows rankings for the first 9 venues of the
 2021 season:
-
+%
 \begin{Schunk}
@@
 \end{Schunk}
-
+%
 Above, we see that Mercedes (``\code{Merc}'') came first and third at
@@
 retain a skilled driver being part of the object of inference.  The
-associated generalized Plackett-Luce \code{hyper3} object is easily
+associated generalized Plackett-Luce `\code{hyper3}' object is easily
 constructed using package idiom, in this case
@@
 Plackett-Luce strengths of the constructors:
-
+%
 \begin{Schunk}
@@
 \end{Schunk}
-
+%
 Above, we see the strength of Mercedes falling from about 0.38 in 2020
@@
 years, keeping track of the year:
-
+%
 \begin{Schunk}
 \begin{Sinput}
-R> H <- (
-+       psubs(constructor_2020, "Merc", "Merc2020") +
+R> H <- (psubs(constructor_2020, "Merc", "Merc2020") +
 +      psubs(constructor_2021, "Merc", "Merc2021")
@@
 \end{Schunk}
-
+%
 Above, we use generic function \code{psubs()} to change the name of
 Mercedes from \code{Merc} to \code{Merc2020} and \code{Merc2021}
-respectively.  Note the use of ``\code{+}" to represent addition of
+respectively.  Note the use of ``\code{+}'' to represent addition of
 log-likelihoods, corresponding to the assumption of conditional
@@
 idiom would be to use generic function \code{samep.test()}:
-
+%
 \begin{Schunk}
@@
 \end{Schunk}
-
+%
 Above, we see strong evidence for a real decrease in the strength of
@@
 Implementing this in \proglang{R} is accomplished via a \proglang{C++}
-back-end making use of the \proglang{STL} ``map" class which offers
+back-end making use of the \proglang{STL} `\code{map}' class which offers
 efficiency advantages, especially for large objects.
@@
 accommodated, either using likelihoods comprising sums of
-Plackett-Luce probabilities~\citep{hankin2017}; or the introduction of
-a reified draw entity~\citep{hankin2010}.
+Plackett-Luce probabilities \citep{hankin2017}; or the introduction of
+a reified draw entity \citep{hankin2010}.
 
@@
 \code{hyper3} approach and the Mann-Whitney-Wilcoxon test, including
-the characterisation of the power function of both tests.  The package
+the characterization of the power function of both tests.  The package
 could easily be extended to allow non-integer multiplicities, which
 might prove useful in the context of reified Bradley Terry
-techniques~\citep{hankin2020}.
+techniques \citep{hankin2020}.
 
-\bibliography{jss4832}
+\bibliography{ref}
 \end{document}

### 2024-05-08 inst/paper.tex
issue #153, bullet point 1

@@
 
-The package includes an \proglang{S}3 class `\code{hyper3}' for this
-type of object; extraction and replacement methods use \pkg{disordR}
-discipline \citep{hankin2022}.  Package idiom for creating such
-objects uses named vectors:
+Functionality for working with generalized Plackett-Luce likelihoods
+is provided in the package \pkg{hyper2}~\citep{hankin2017,hankin2009},
+available from the Comprehensive \proglang{R} Archive Network (CRAN)
+at \url{https://CRAN.R-project.org/hyper2}.  The package includes an
+\proglang{S}3 class `\code{hyper3}' for this type of object;
+extraction and replacement methods use \pkg{disordR} discipline
+\citep{hankin2022}.  Package idiom for creating such objects uses
+named vectors:
 %

### 2024-05-08 inst/paper.tex
issue #153, bullet point 2
issue #153, bullet point 3 (affiliation)

@@
 
-\author{Robin K. S. Hankin\\Auckland University of Technology}
+\author{Robin K. S. Hankin~\orcidlink{0000-0001-5982-0415}\\
+  The University of Stirling}
 \title{Generalized Plackett-Luce Likelihoods}
@@
   Robin K. S. Hankin\\
-  School of Engineering and Mathematical Sciences\\
-  Auckland University of Technology\\
-  Wellesley Street\\
-  Auckland, New Zealand\\
+  The University of Stirling\\
+  Scotland\\
   E-mail: \email{hankin.robin@gmail.com}\\
-  URL: \url{https://academics.aut.ac.nz/robin.hankin}
+  URL: \url{https://www.stir.ac.uk/people/1966824}
 }

### 2024-05-08 inst/paper.tex
issue #153, bullet point 4 (shorten long section titles)

@@
 
-\subsection{Computational methodology for generalized Plackett-Luce likelihood functions}
+\subsection{Computational methodology}
 
@@
 
-\section{Exact analytical solution for some simple generalized Plackett-Luce likelihood functions}
+\section{Exact analytical solutions}
 
@@
 
-\section{An alternative to the Mann-Whitney test using generalized Plackett-Luce likelihood}
+\section{An alternative to the Mann-Whitney test}
 
@@
 
-\subsection{A generalization of the Mann-Whitney test using
-generalized Plackett-Luce likelihood}
+\subsection{A generalization of the Mann-Whitney test}
 
@@
 
-\section{Formula 1 motor racing: The Constructors' Championship}\label{f1}
+\section{Formula 1 motor racing}\label{f1}
 

### 2024-05-16 inst/f1points.Rmd
change affiliation

--- 2023-04-24 inst/f1points.Rmd
@@
   - name: Robin K. S. Hankin
-    affiliation: Auckland University of Technology
+    affiliation: University of Stirling
     address:
-      2-14 Wakefield Street
-      Auckland, New Zealand
+		Stirling, Scotland FK9 4LH
     email: \email{hankin.robin@gmail.com}

### 2024-05-16 inst/jss4832.tex
add some citations which had been deleted accidentally
change affiliation

@@
 
-\author{Robin K. S. Hankin\\Auckland University of Technology}
+\author{Robin K. S. Hankin\\University of Stirling}
 \title{Generalized Plackett-Luce Likelihoods}
@@
   Robin K. S. Hankin\\
-  School of Engineering and Mathematical Sciences\\
-  Auckland University of Technology\\
-  Wellesley Street\\
-  Auckland, New Zealand\\
+  University of Stirling\\
   E-mail: \email{hankin.robin@gmail.com}\\
@@
 
-\bibliography{jss4832}
+\bibliography{hyper2}
 \end{document}

### 2024-05-16 inst/hankin_american_statistician.tex
work on issue #232

--- 2024-04-16 inst/hankin-sts.tex
@@
-%% Template for the submission to:
-%%   Statistical Science [STS]
-%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% In this template, the places where you   %%
-%% need to fill in your information are     %%
-%% indicated by '???'.                      %%
-%%                                          %%
-%% Please do not use \input{...} to include %%
-%% other tex files. Submit your LaTeX       %%
-%% manuscript as one .tex document.         %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\documentclass[sts]{imsart}
-
-%% Packages
+\documentclass{article}
 \RequirePackage{amsthm,amsmath,amsfonts,amssymb}
 \RequirePackage[numbers]{natbib}
-%\RequirePackage[authoryear]{natbib}%% uncomment this for author-year citations
-%\RequirePackage[colorlinks,citecolor=blue,urlcolor=blue]{hyperref}%% uncomment this for coloring bibliography citations and linked URLs
 \RequirePackage{graphicx}%% uncomment this for including figures
-
-\startlocaldefs
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% Uncomment next line to change            %%
-%% the type of equation numbering           %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\numberwithin{equation}{section}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% For Axiom, Claim, Corollary, Hypothesis, %%
-%% Lemma, Theorem, Proposition              %%
-%% use \theoremstyle{plain}                 %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\theoremstyle{plain}
-%\newtheorem{???}{???}
-%\newtheorem*{???}{???}
-%\newtheorem{???}{???}[???]
-%\newtheorem{???}[???]{???}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% For Assumption, Definition, Example,     %%
-%% Notation, Property, Remark, Fact         %%
-%% use \theoremstyle{remark}                %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\theoremstyle{remark}
-%\newtheorem{???}{???}
-%\newtheorem*{???}{???}
-%\newtheorem{???}{???}[???]
-%\newtheorem{???}[???]{???}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Please put your definitions here:        %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\endlocaldefs
+\usepackage{setspace} \doublespacing
 
@@
 
-\begin{frontmatter}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% Enter the title of your article here     %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \title{A simplified Plackett--Luce likelihood function for rank orders}
-%\title{A sample article title with some additional note\thanksref{T1}}
-\runtitle{Simplified Plackett--Luce likelihood functions}
-%\thankstext{T1}{A sample of additional note to the title.}
-
-\begin{aug}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% ORCID can be inserted by command:         %%
-\orcid{0000-0001-5982-0415}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\author[A]{\fnms{Robin K. S.}~\snm{Hankin}\ead[label=e1]{hankin.robin@gmail.com}}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Addresses                                %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\address[A]{University of Stirling\printead[presep={\ }]{e1}.}
-\end{aug}
+%\author{Robin K. S. Hankin}
+\maketitle
 
@@
 \end{abstract}
-
-\begin{keyword}
-\kwd{Statistical inference under constraints}
-\end{keyword}
-
-\end{frontmatter}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%% Main text entry area:
 
@@
 \begin{figure}[t]
+  \begin{centering}
 \includegraphics[width=4in]{ninelikes}  % ninelikes.pdf produced by very_simplified_likelihood.Rmd
-\caption{Likelihood functions for $r=0(1)8$\label{ninelikes}.  Value of $r$ indicated in colour by its curve; vertical line shows $\hat{a}\simeq 0.5886$ for the case $r=4$}
+\caption{Likelihood functions for $r=0(1)8$\label{ninelikes}.  Value
+  of $r$ indicated in colour by its curve; vertical line shows
+  $\hat{a}\simeq 0.5886$ for the case $r=4$}
+\end{centering}
 \end{figure}
@@
 \begin{figure}[t]
+  \begin{centering}
 \includegraphics[width=4in]{dotprobs}  % produced by very_simplified_likelihood.Rmd
-\caption{Maximum likelihood estimator $\hat{a}$ for $n=1(1)10$ and $r=0(1)n$
-  \label{dotprobs}}
+\caption{Maximum likelihood estimator $\hat{a}$ for $n=1(1)10$
+  and\label{dotprobs}} $r=0(1)n$
+\end{centering}
 \end{figure}
@@
 \begin{figure}[t]
+  \begin{centering}
 \includegraphics[width=4in]{plotolymp} % produced by very_simplified_likelihood.Rmd
@@
   lane-1 runner, Olympic qualifying heats 2012-2020 \label{plotolymp}}
+\end{centering}
 \end{figure}
@@
 \begin{figure}
+\begin{centering}
 \includegraphics[width=4in]{plotpureandapplied}  % plotpureandapplied.pdf produced by very_simplified_likelihood.Rmd
@@
   line \label{plotpureandapplied}}
+\end{centering}
 \end{figure}
@@
 \begin{figure}[t]
+  \begin{centering}
 \includegraphics[width=4in]{showchecolike}  % showchecolike.pdf produced by very_simplified_likelihood.Rmd
@@
   career 2011--2023\label{showchecolike}}
+\end{centering}
 \end{figure}
@@
 
-% \begin{acks}[Acknowledgments]
-% The authors would like to thank the anonymous referees, an Associate
-% Editor and the Editor for their constructive comments that improved the
-% quality of this paper.
-% \end{acks}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Supplementary Material, including data   %%
-%% sets and code, should be provided in     %%
-%% {supplement} environment with title      %%
-%% and short description. It cannot be      %%
-%% available exclusively as external link.  %%
-%% All Supplementary Material must be       %%
-%% available to the reader on Project       %%
-%% Euclid with the published article.       %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{supplement}
-  \stitle{R code source files}
-  \sdescription{
-    File {\tt hankin-sts.Rmd} is a terse rmarkdown document that
-    recreates the figures and salient results of the submission.
-  }
-\end{supplement}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                  The Bibliography                       %%
-%%                                                         %%
-%%  imsart-???.bst  will be used to                        %%
-%%  create a .BBL file for submission.                     %%
-%%                                                         %%
-%%  Note that the displayed Bibliography will not          %%
-%%  necessarily be rendered by Latex exactly as specified  %%
-%%  in the online Instructions for Authors.                %%
-%%                                                         %%
-%%  MR numbers will be added by VTeX.                      %%
-%%                                                         %%
-%%  Use \cite{...} to cite references in text.             %%
-%%                                                         %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%% if your bibliography is in bibtex format, uncomment commands:
-\bibliographystyle{imsart-number} % Style BST file (imsart-number.bst or imsart-nameyear.bst)
-\bibliography{hyper2}       % Bibliography file (usually '*.bib')
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Single Appendix:                         %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\begin{appendix}
-%\section*{???}%% if no title is needed, leave empty \section*{}.
-%\end{appendix}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Multiple Appendixes:                     %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\begin{appendix}
-%\section{???}
-%
-%\section{???}
-%
-%\end{appendix}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Support information, if any,             %%
-%% should be provided in the                %%
-%% Acknowledgements section.                %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\begin{acks}[Acknowledgments]
-% The authors would like to thank ...
-%\end{acks}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Funding information, if any,             %%
-%% should be provided in the                %%
-%% funding section.                         %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\begin{funding}
-% The first author was supported by ...
-%
-% The second author was supported in part by ...
-%\end{funding}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Supplementary Material, including data   %%
-%% sets and code, should be provided in     %%
-%% {supplement} environment with title      %%
-%% and short description. It cannot be      %%
-%% available exclusively as external link.  %%
-%% All Supplementary Material must be       %%
-%% available to the reader on Project       %%
-%% Euclid with the published article.       %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\begin{supplement}
-%\stitle{???}
-%\sdescription{???.}
-%\end{supplement}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                  The Bibliography                       %%
-%%                                                         %%
-%%  imsart-???.bst  will be used to                        %%
-%%  create a .BBL file for submission.                     %%
-%%                                                         %%
-%%  Note that the displayed Bibliography will not          %%
-%%  necessarily be rendered by Latex exactly as specified  %%
-%%  in the online Instructions for Authors.                %%
-%%                                                         %%
-%%  MR numbers will be added by VTeX.                      %%
-%%                                                         %%
-%%  Use \cite{...} to cite references in text.             %%
-%%                                                         %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%% if your bibliography is in bibtex format, uncomment commands:
-%\bibliographystyle{imsart-number} % Style BST file (imsart-number.bst or imsart-nameyear.bst)
-%\bibliography{bibliography}       % Bibliography file (usually '*.bib')
-
-%% or include bibliography directly:
-% \begin{thebibliography}{}
-% \bibitem{b1}
-% \end{thebibliography}
-
+\bibliographystyle{plain}
+\bibliography{hyper2}    
 \end{document}

### 2024-05-22 inst/hankin_american_statistician.tex
change MS to APA style bibliography as per issue #232

@@
 \RequirePackage{amsthm,amsmath,amsfonts,amssymb}
-\RequirePackage[numbers]{natbib}
 \RequirePackage{graphicx}%% uncomment this for including figures
 \usepackage{setspace} \doublespacing
+\usepackage{apacite}
 
@@
   strengths] and say 20 independent complete rank observations,
-Hankin~\cite{hankin2017_rmd} shows that maximum likelihood strengths
+\citeA{hankin2017_rmd} shows that maximum likelihood strengths
 identify the correct ranking with a probability of essentially zero.
@@
 Formula 1 motor racing is an important and prestigious motor sport
-\citep{codling2017,jenkins2010}.  Season ranking is based on a points
+\cite{codling2017,jenkins2010}.  Season ranking is based on a points
 allocation system wherein competitors are awarded points based on race
@@
 after the final race.  However, as argued by
-Hankin~\cite{hankin2023_formula1points}, the drivers' {\em ranks} are
+\citeA{hankin2023_formula1points}, the drivers' {\em ranks} are
 statistically sufficient for analysis using Plackett--Luce approaches.
@@
 
-\bibliographystyle{plain}
+\bibliographystyle{apacite}
 \bibliography{hyper2}    

### 2024-05-27 inst/grad_weight.Rmd
rewrite and bugfix

@@
----
 title: "Differentiation with respect to weights for a `hyper3` object"
@@
 bibliography: hyper2.bib  # copy vignettes/hyper2.bib to inst/
----
-
 ```{r setup, include=FALSE}
@@
 library("disordR")
-```
-
+library("magrittr")
 ```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
 knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
-```
-
 To cite the `hyper2` package in publications, please use @hankin2017_rmd.
@@
 and $p_2$ winning $b$ times.  We use the following likelihood function:
-
-$$
 \frac{(wp_1)^ap_2^b}{(wp_1+p_2)^{a+b}}
-$$
-
 where $w$ represents some form of competitive advantage present in the
@@
 follows:
-
 ```{r,label=definefw}
@@
     H
-}
-```
-
 Thus
-
 ```{r,label=showfinuse}
 f(w=1.1)
-```
-
 Here I consider how we may differentiate this support function with respect to $w$:
-
 ```{r,label=definegws}
@@
   alpha <- drop(powers(jj))
-    powers(jj) <- 1
-    alpha*sum(probs[names(probs) %in% wanted])/loglik(probs,jj,log=FALSE)
-}
-```
-
+  powers(jj) <- -1
+  jj[setNames(1,wanted)] <- +1
+  pnames(jj) <- pnames(H)
+  alpha*loglik(probs,jj,log=FALSE)
 Function `grad_weight_single()` takes a `hyper3` object `H`, a vector
@@
 each bracket the weight appears in, and add the results.  For example:
-
 ```{r,label=testthederiv}
@@
 probs <- c(p1=0.35, p2=0.65)
-
 wpv <- c(p1=w)
@@
 d1 <- grad_weight_single(H, probs, wpv, wanted)
-
 wpv <- c(p2=1,p1=w)
@@
 d2 <- grad_weight_single(H, probs, wpv, wanted)
-
 print(c(d1,d2))
-
 LHS <- (d1+d2)                                          # differential from grad_weight_single()
 RHS <- (loglik(probs,f(w+dw)) - loglik(probs,f(w)))/dw  # differential via numerics
-
 print(c(lhs=LHS,rhs=RHS,diff=RHS-LHS))
-```
-
 above we see very close agreement.
-
 # Slightly more complicated example
-
 ```{r slightlymorecomplicated}
@@
   H <- hyper3()
-  H[c(p1=1 , p2=w , p3=w , p4=1)] <- 5
-  H
-}
+  H[c(p1=1 , p2=w , p3=w )] <- -5
+  H[c(p1=1)] <- 2
+  H[c(p2=2)] <- 1
+  H[c(p2=w,p3=w)] <- 2
 w <- 1.4
@@
 HH <- f(w)
-HH
-probs <- c(p1=0.1,p2=0.2,p3=0.3,p4=0.4)
-wpv <- c(p1=1 , p2=w , p3=w , p4=1)
-LHS <- grad_weight_single(HH, probs, wpv, c("p2","p3"))
+probs <- c(p1=0.1,p2=0.6,p3=0.3)
+d1 <- grad_weight_single(HH, probs, c(p1=1 , p2=w , p3=w), "p2")
+d2 <- grad_weight_single(HH, probs, c(p1=1 , p2=w , p3=w), "p3")
+d3 <- grad_weight_single(HH, probs, c(       p2=w , p3=w), "p2")
+d4 <- grad_weight_single(HH, probs, c(       p2=w , p3=w), "p3")
+LHS <- d1+d2+d3+d4
 RHS <-  (loglik(probs,f(w=w+dw)) - loglik(probs,f(w=w)))/dw
 c(RHS,LHS,LHS-RHS)
-```
-
 ## An example from chess
-
 Two chess players, 1, 2, and 3, play chess.  Our observations (chess
@@
 Consider the following likelihood function:
-
-$$
 \frac{(w p_1)^a\cdot(D(p_1+p_2))^b\cdot p_2^c}{(w p_1+D(p_1+p_2)+p_2)^{a+b+c}}
@@
 \frac{ w p_3}{wp_3+p_1}
-$$
-
 ```{r chessexample}
-
 w <- 1.2
 D <- 0.3
-
 brackfunc <- function(w,D){
@@
   b9 = c(p3=w,p1=1)
-)
-}
-
-wanted <- list("p1",NULL,NULL,"p1","p2",NULL,"p2","p3","p3")
-
 `f2` <- function(w,D, a=3,b=12,c=1,d=5,e=6,f=6){
@@
   return(H)
-}
-
 H <- f2(w=w, D=D)
-H
-```
-
-$$
 \frac{(w p_1)^a\cdot(D(p_1+p_2))^b\cdot p_2^c}{(w p_1+D(p_1+p_2)+p_2)^{a+b+c}}
@@
 \frac{ w p_3}{wp_3+p_1}
-$$
-
 Differentiate with respect to `w`:
-
 ```{r diffwrtw}
-diff <- 0
-dw <- 1e-4
 probs <- c(p1=0.15 , p2=0.45, p3=0.4)
-B <- brackfunc(w,D)
-for(i in seq_along(wanted)){
-   diff <- diff +  grad_weight_single(H, probs, wpv=B[[i]],wanted=wanted[[i]])
-}
-
+diff <- (
+ grad_weight_single(H, probs, c(p1=w), "p1") +
+ grad_weight_single(H, probs, c(p1=w+D,p2=1+D), "p1") +
+ grad_weight_single(H, probs, c(p2=w), "p2") +
+ grad_weight_single(H, probs, c(p1=1+D,p2=w+D), "p2") +
+ grad_weight_single(H, probs, c(p3=w), "p3")+
+ grad_weight_single(H, probs, c(p3=w,p1=1),"p3")
 LHS <- diff  # using grad_weight_single()
@@
 c(LHS,RHS,LHS-RHS)
-```
-
 We may even be a little clever  and use a central difference:
-
 ```{r cleverusecentral}
+if(FALSE){
 RHS2 <- (loglik(probs,f2(w=w+dw/2,D=D)) - loglik(probs,f2(w=w-dw/2,D=D)))/dw  # differential via numerics
 c(LHS,RHS2,LHS-RHS2)
-```
-
 If we wish to differentiate with respect to `D` we need to change the `wanted` list:
-
 ```{r diffwrtD}
+if(FALSE){
 o <- c("p1","p2")
@@
 wantedD <- list(n,o,n,o,n,n,o,n,n)
-
 diff <- 0
@@
    diff <- diff +  grad_weight_single(H, probs, wpv=B[[i]],wanted=wantedD[[i]])
-}
-
 LHS <- diff  # using grad_weight_single()
@@
 c(LHS,RHS,LHS-RHS)
-```
-
 ### References {-}

### 2024-05-28 inst/grad_weight.Rmd
pursuant to commit 1d7c34c

--- 2024-01-26 inst/grad_weight.Rmd
@@
 library("disordR")
+library("magrittr")
 ```
@@
 
-$$
-\frac{(wp_1)^ap_2^b}{(wp_1+p_2)^{a+b}}
-$$
+$$\frac{(wp_1)^ap_2^b}{(wp_1+p_2)^{a+b}}$$
 
@@
 
-Here I consider how we may differentiate this support function with respect to $w$:
+Here I consider how we may differentiate this support function with
+respect to $w$:
 
@@
   alpha <- drop(powers(jj))
-    powers(jj) <- 1
-    alpha*sum(probs[names(probs) %in% wanted])/loglik(probs,jj,log=FALSE)
+  powers(jj) <- -1
+  jj[setNames(1,wanted)] <- +1
+  pnames(jj) <- pnames(H)
+  alpha*loglik(probs,jj,log=FALSE)
   }
@@
 
-print(c(d1,d2))
-
 LHS <- (d1+d2)                                          # differential from grad_weight_single()
-RHS <- (loglik(probs,f(w+dw)) - loglik(probs,f(w)))/dw  # differential via numerics
-
+RHS <- (loglik(probs,f(w+dw/2)) - loglik(probs,f(w-dw/2)))/dw  # differential via numerics
 print(c(lhs=LHS,rhs=RHS,diff=RHS-LHS))
@@
   H <- hyper3()
-  H[c(p1=1 , p2=w , p3=w , p4=1)] <- 5
-  H
+  H[c(p1=1 , p2=w , p3=w )] <- -5
+  H[c(p1=1)] <- 2
+  H[c(p2=2)] <- 1
+  H[c(p2=w,p3=w)] <- 2
+  return(H)
 }
+
 w <- 1.4
 dw <- 1e-6
-HH <- f(w)
-HH
-probs <- c(p1=0.1,p2=0.2,p3=0.3,p4=0.4)
-wpv <- c(p1=1 , p2=w , p3=w , p4=1)
-LHS <- grad_weight_single(HH, probs, wpv, c("p2","p3"))
-RHS <-  (loglik(probs,f(w=w+dw)) - loglik(probs,f(w=w)))/dw
+(HH <- f(w))
+probs <- c(p1=0.1,p2=0.6,p3=0.3)
+d1 <- grad_weight_single(HH, probs, c(p1=1 , p2=w , p3=w), "p2")
+d2 <- grad_weight_single(HH, probs, c(p1=1 , p2=w , p3=w), "p3")
+d3 <- grad_weight_single(HH, probs, c(       p2=w , p3=w), "p2")
+d4 <- grad_weight_single(HH, probs, c(       p2=w , p3=w), "p3")
+LHS <- d1+d2+d3+d4
+RHS <- (loglik(probs,f(w=w+dw/2)) - loglik(probs,f(w=w-dw/2)))/dw
   c(RHS,LHS,LHS-RHS)
@@
 ```{r chessexample}
-
 w <- 1.2
@@
 }
-
-wanted <- list("p1",NULL,NULL,"p1","p2",NULL,"p2","p3","p3")
 
@@
   H <- f2(w=w, D=D)
-H
+
 ```
@@
 ```{r diffwrtw}
-diff <- 0
-dw <- 1e-4
 probs <- c(p1=0.15 , p2=0.45, p3=0.4)
-B <- brackfunc(w,D)
-for(i in seq_along(wanted)){
-   diff <- diff +  grad_weight_single(H, probs, wpv=B[[i]],wanted=wanted[[i]])
-}
+diff <- (
+ grad_weight_single(H, probs, c(p1=w), "p1") +
+ grad_weight_single(H, probs, c(p1=w+D,p2=1+D), "p1") +
+ grad_weight_single(H, probs, c(p2=w), "p2") +
+ grad_weight_single(H, probs, c(p1=1+D,p2=w+D), "p2") +
+ grad_weight_single(H, probs, c(p3=w), "p3")+
+ grad_weight_single(H, probs, c(p3=w,p1=1),"p3")
+ )
 
@@
 
-If we wish to differentiate with respect to `D` we need to change the `wanted` list:
-
-```{r diffwrtD}
-o <- c("p1","p2")
-n <- NULL
-wantedD <- list(n,o,n,o,n,n,o,n,n)
-
-diff <- 0
-dD <- 1e-5
-B <- brackfunc(w,D)
-for(i in seq_along(wanted)){
-   diff <- diff +  grad_weight_single(H, probs, wpv=B[[i]],wanted=wantedD[[i]])
-}
-
-LHS <- diff  # using grad_weight_single()
-RHS <- (loglik(probs,f2(w,D=D+dD)) - loglik(probs,f2(w,D)))/dD  # differential via numerics
-c(LHS,RHS,LHS-RHS)
-```
+If we wish to differentiate with respect to `D` we need a different technique (TBA).
 

### 2024-05-28 inst/grad_weight.Rmd
whitespace

@@
 diff <- (
- grad_weight_single(H, probs, c(p1=w), "p1") +
+ grad_weight_single(H, probs, c(p1=w)         , "p1") +
  grad_weight_single(H, probs, c(p1=w+D,p2=1+D), "p1") +
- grad_weight_single(H, probs, c(p2=w), "p2") +
+ grad_weight_single(H, probs, c(p2=w)         , "p2") +
  grad_weight_single(H, probs, c(p1=1+D,p2=w+D), "p2") +
- grad_weight_single(H, probs, c(p3=w), "p3")+
- grad_weight_single(H, probs, c(p3=w,p1=1),"p3")
+ grad_weight_single(H, probs, c(p3=w)         , "p3") +
+ grad_weight_single(H, probs, c(p3=w,p1=1)    , "p3")
  )

### 2024-05-28 inst/grad_weight.Rmd
multiple probabilities in the argument

@@
   powers(jj) <- -1
-  jj[setNames(1,wanted)] <- +1
+  jj[setNames(rep(1,length(wanted)),wanted)] <- +1
   pnames(jj) <- pnames(H)
@@
 
-If we wish to differentiate with respect to `D` we need a different technique (TBA).
+If we wish to differentiate with respect to `D` we need a slight
+generalization, specifically that the `wanted` argument is length 2:
+
+```{r label=diffD}
+diff <- (
+ grad_weight_single(H, probs, c(p1=  D,p2=  D), c("p1","p2")) +
+ grad_weight_single(H, probs, c(p1=w+D,p2=1+D), c("p1","p2")) +
+ grad_weight_single(H, probs, c(p1=1+D,p2=w+D), c("p1","p2"))
+ )
+
+LHS <- diff  # using grad_weight_single()
+dD <- 1e-4
+RHS <- (loglik(probs,f2(w,D=D+dD/2)) - loglik(probs,f2(w=w,D=D-dD/2)))/dD  # d/dD via numerics
+c(LHS,RHS,LHS-RHS)
+```
 

### 2024-05-28 inst/grad_weight.Rmd
a more exacting example

@@
 
+Come to think of it, all the `wanted` arguments are the same length
+(two).  A more exacting test would be:
+
+$$
+\frac{(p_1)^3\,(w p_2)^4}{(p_1+wp_2)^7}\cdot
+\frac{p_2^5\,(wp_1+wp_3)^2}{(wp_1+p_2+wp_3)^7}
+$$
+
+```{r label=moreexacting}
+m <- function(w){
+  H <- hyper3()
+  H[c(p1=1)] <- 3
+  H[c(p2=w)] <- 4
+  H[c(p1=1,p2=w)] <- -7
+  H[c(p2=1)] <- 5
+  H[c(p1=w,p3=w)] <- 2
+  H[c(p1=w,p2=1,p3=w)] <- -7
+  H
+}
+w <- 1.888
+dw <- 1e-4
+probs <- c(p1=0.4,p2=0.1, p3=0.5)
+m(w)
+
+H <- m(w)
+loglik(probs,H)
+
+diffD <- (
+ grad_weight_single(H, probs, c(p2=w          ), "p2"        ) +
+ grad_weight_single(H, probs, c(p1=1,p2=w     ), "p2"        ) +
+ grad_weight_single(H, probs, c(p1=w,p3=w     ), c("p1","p3")) +
+ grad_weight_single(H, probs, c(p1=w,p2=1,p3=w), c("p1","p3"))
+ )
+
+LHS <- (loglik(probs,m(w+dw/2)) -loglik(probs,m(w-dw/2)))/dw
+RHS <- diffD
+c(LHS,RHS,LHS-RHS)
+```
+
 ### References {-}

### 2024-05-29 inst/rock_paper_scissors_monster.Rmd
bugfix

@@
      H3[jj] %<>% inc(hw[i,i])
-     H3[setNames(1,tt[i])] %<>% inc(hw[i,i])
 

### 2024-07-29 inst/monster_vs_lambda.Rmd
more detail

@@
 title: "Monster or Lambda?  A likelihood analysis"
+subtitle: "Takes about an hour to run without cache"
 author: "R. K. S. Hankin"
@@
 library("hyper2")
+library("magrittr")
 ```
@@
 
-To cite the `hyper2` package in publications, please use @hankin2017_rmd;
-takes about an hour to process without cache.  Here I consider
-pairwise comparisons with a home-ground advantage and wish to compare
-two likelihood functions.  First, the monster parametrization:
+To cite the `hyper2` package in publications, please use
+@hankin2017_rmd.  Here I consider pairwise comparisons with a
+home-ground advantage and wish to compare two likelihood functions.
+Draws are excluded.  First, the monster parametrization:
 
@@
 
-## Start simple: two teams
+# Start simple: two teams {-}
+
+### Monster parametrization {-}
 
 ```{r,echo=FALSE}
-jj <- matrix(c(NA, 6+2i ,6+5i, NA),2,2,byrow=TRUE)
+jj <- matrix(c(NA, 7+2i ,6+5i, NA),2,2,byrow=TRUE)
 teams <- LETTERS[1:2]
@@
 
-Above we see that A played at home 3 times, winning 1 and losing 2
-[row 1 column 2 is $1+2i$]; B played at home 9 times, winning 4 and
-losing 5 [row 2 column 1 is $4+5i$].  In its most general form, the
-$2\times 2$ result matrix would be
+Above we see that A played at home 9 times, winning 7 and losing 2
+[row 1 column 2 is $7+2i$]; B played at home 11 times, winning 6 and
+losing 5 [row 2 column 1 is $6+5i$] (observe that the real components
+correspond to home wins and the imaginary components correspond to
+away wins.  Thus A wins 7 times at home and 5 times away).  In its
+most general form, the $2\times 2$ result matrix would be
 
@@
 
-and the log-likelihood function would be, algebraically,
+and the monster log-likelihood function would be, algebraically,
 
@@
 
-Just to check, we evaluate the Hessian matrix and find
+Observe that the point
+$\left(\hat{A},\hat{B},\hat{M}=1-\hat{A}-\hat{B}\right)$ might be
+inadmissable.  For example, if $\beta,\delta>0$ and $\alpha=\gamma=0$
+then $\hat{A}=\hat{B}=1$ and $\hat{M}=-1$.  This would correspond to
+both teams winning all their games at home and losing none.
+
+Just to check the well-definedness of the ML estimate, we evaluate the
+Hessian matrix and find
 
@@
 
+### Lambda parametrization {-}
+
 Now the $\lambda$ parameterization.  The likelihood function for the
@@
 -(\gamma+\delta)\log(A+\lambda B)$ and even after simplifying [we may
-assume $A+B=1$] the result does not appear to have a simple analytical
-solution; we must proceed numerically.  To do so,
-
-we need to define a
+assume $B=1-A$] the result of maximizing $\mathcal{S}$ with respect to
+$A$ and $\lambda$ does not appear to have a simple analytical
+solution; we must proceed numerically.  To do so, we need to define a
 slight generalization of `home_away3()`:

### 2024-08-13 inst/table_tennis_serve.Rmd
magrittr package now needed explicitly
pair3()

@@
 library("hyper2",quietly=TRUE)
+library("magrittr")
 (H <- hyper2(pnames=c("S","a","b","c")))
@@
 
-so the likelihood function is uniform, for the only term is raised to
-the zeroth power (players' names are not capitalised in R idiom for
-clarity).  Now we have to input the data, and we will go through the
-four scorelines above, one by one.  Note that this approach implicitly
-assumes independence of scores.
+so the likelihood function is uniform.  We have players `a`, `b`, `c`
+(players' names are not capitalised in R idiom for clarity) and a
+reified entity `S` representing the strength of the serve.  Now we
+have to input the data, and we will go through the four scorelines
+above, one by one.  Note that this approach implicitly assumes
+independence of scores.
 
-### A vs B, A serves, score 5-1:
+### A vs B, A serves, score 5-1: {-}
 ```{r,label=abaserve}
@@
 
-### A vs B, B serves, score 1-3:
+* A vs B, B serves, score 1-3:
 ```{r,label=abbserve}
@@
 
-### A vs C, A serves, score 4-1:
+* A vs C, A serves, score 4-1:
 ```{r,label=acaserve}
@@
 
-### A vs C, C serves, score 1-2:
+* A vs C, C serves, score 1-2:
 ```{r,label=accserve}
@@
 ```
-
-## Likelihood function
 
@@
 
-## Assessment of B vs C
-
-The maximum likelihood estimate for strengths allows us to assess what
-might happen when B plays C (which was not observed).  For
-convenience, here is the evaluate:
-
-```{r,label=maxtab,cache=TRUE}
-maxp(table_tennis)
+```{r, label=equalplayerstrength,cache=TRUE}
+samep.test(table_tennis,c("a","b","c"))
 ```
 
-If B serves, we would have $B+S:C\simeq 0.14+0.45:0.15\equiv 59:15$,
-or B wins with probability of about $0.78$.  If C serves we have
-$B:C+S\simeq 0.14:0.16+0.45\equiv 14:61$, so B wins with a probability
-of about $0.19$.
+Above, `samep.test()` complains because it is performing a
+one-dimensional optimization (we may ignore this issue).  The result
+of the test is that there is no evidence to suggest a difference
+between players `a`, `b`, and `c`.
 
@@
 particular excludes zero).
+
+# Assessment of B vs C
+
+The maximum likelihood estimate for strengths allows us to assess what
+might happen when B plays C (which was not observed).  For
+convenience, here is the evaluate:
+
+```{r,label=maxtab,cache=TRUE}
+maxp(table_tennis)
+```
+
+If B serves, we would have $B+S:C\simeq 0.14+0.45:0.15\equiv 59:15$,
+or B wins with probability of about $0.78$.  If C serves we have
+$B:C+S\simeq 0.14:0.16+0.45\equiv 14:61$, so B wins with a probability
+of about $0.19$.
 
@@
 
+# Weighted likelihoods
+
+We now try `hyper3` formalism:
+
+```{r,label=calcproflike,cache=TRUE}
+M_home <- matrix( c(NA, 5 , 4, 3, NA, 0, 2,0,NA),byrow=TRUE,3,3)
+jj <- letters[1:3]
+dimnames(M_home) <- list(jj,jj)
+
+M_away <- matrix( c(NA, 1 , 1, 1, NA, 0, 1,0,NA),byrow=TRUE,3,3)
+jj <- letters[1:3]
+dimnames(M_away) <- list(jj,jj)
+M_home
+M_away
+
+home_away3(M_home,M_away,lambda=1.8)
+
+f <- function(lam){
+    maxp(home_away3(M_home,M_away,lambda=lam),give=TRUE)$value
+}
+```
+
+```{r,makelamlike,cache=TRUE}
+lam <- exp(seq(from=log(0.6),to=log(20),len=9))
+L <- sapply(lam,f)
+```
+
+```{r,optimlamlike,cache=TRUE}
+maxf <- optimize(f,c(2,5),maximum=TRUE)
+maxf
+```
+
+```{r plotlikelam}
+L <- L-max(L)
+plot(lam,L,type='b',pch=16,log="x")
+abline(h=c(0,-2))
+```
+
+How does the situation change if we have some more observations?
+Suppose `c` plays at home to `d`, and loses.  If we use
+$\hat{\lambda}\simeq 3.35$:
+
+```{r, label=newdatacd,cache=TRUE}
+(Hnew <- home_away3(M_home,M_away,lambda=3.35) + pair3(c=0,d=1,lambda=3.35))
+maxp(Hnew)
+```
+
+Above we see that our _estimate_ would be that `d` has a BT strength
+of 1 and the other players have a strength of zero.  Is there strong
+evidence that the teams' strength differ?
+
+```{r,label=newdatatest,cache=TRUE}
+equalp.test(Hnew)
+```
+
+Above we see that there is no evidence that the teams' strengths do in
+fact differ.
+
 ### Package dataset

### 2024-08-15 inst/rock_paper_scissors_monster.Rmd
dirichlet3()

@@
 H3seven <- function(lambda){
-  H3 <- hyper3()
+    out <- hyper3()
     for(i in 1:6){
-     jj <- 1  # diagonal players do not have lambda-advantage
-     names(jj) <- tt[i]
-     H3[jj] %<>% inc(hw[i,i])
-
-     jj <- lambda
-     names(jj) <- tt[i+1]
-     H3[jj] %<>% inc(hw[i,i+1])
-
-     jj <- c(1,lambda)
-     names(jj) <- c(tt[i],tt[i+1])
-     H3[jj] %<>% dec(hw[i,i] + hw[i,i+1])
-  }  # i loop closes
-
-  ## Now do final one:
-  jj <- 1
-  names(jj) <- tt[7]
-  H3[jj] %<>% inc(hw[7,7])
-
-  jj <- lambda
-  names(jj) <- tt[1]
-  H3[jj] %<>% inc(hw[7,1])
-
-  jj <- c(1,lambda)
-  names(jj) <- c(tt[7],tt[1])
-  H3[jj] %<>% dec(hw[7,7] + hw[7,1])
-
-  return(H3)
-}  # function H3seven() defn closes
+        out %<>% inc(dirichlet3(powers=hw[i,i:(i+1)],lambda=c(1,lambda)))
+    }
+    out %<>% inc(dirichlet3(powers=hw[7,c(7,1)],lambda=c(1,lambda)))
+    return(out)
+}
+H3seven(lambda=1.66)
 ```
 
-Now we can execute it:
-
-```{r,label=executeh3seven}
-H3 <- H3seven(lambda=1.66)
-H3
-```
-
-```{r,label=defineproflikeh3}
+```{r,label=defineproflikeh3seven}
 profl <- function(lambda){

### 2024-08-15 R/hyper3.R

@@ -650,17 +650,22 @@ stop("not yet written")
     return(x)
 }
 
-`pair3` <- function(..., lambda){ # pair3(g=5, h=2, lambda=1.88)
-    jj <- list(...)
-    stopifnot(length(jj) == 2)
-    home_player <- names(jj)[1]
-    away_player <- names(jj)[2]
-    home_wins <- jj[[1]]
-    away_wins <- jj[[2]]
+`dirichlet3` <- function(powers, lambda=NULL){
+    if(is.null(names(powers))){
+        stop("dirichlet3d() requires powers to be a _named_ vector")
+    }
+    if(!is.null(names(lambda)) & !identical(names(lambda),names(powers))){
+        stop("names(lambda) must match names(powers), if supplied")
+    }
+    if(length(lambda) < length(powers)){
+        lambda <- c(lambda,rep(1,length(powers)-length(lambda)))
+    }
+
     out <- hyper3()
-    out[setNames(lambda, home_player)] %<>% inc(home_wins)
-    out[setNames(1     , away_player)] %<>% inc(away_wins)
-    out[setNames(c(lambda, 1), c(home_player, away_player))] %<>% dec(home_wins + away_wins)
+    for(i in seq_along(powers)){
+        out[setNames(lambda[i],names(powers)[i])] %<>% inc(powers[i])
+    }
+    out[setNames(lambda,names(powers))] %<>% dec(sum(powers))
     return(out)
 }
 

### 2024-08-15 tests/testthat/test_dirichlet.R

@@ -1,0 +1,32 @@
+test_that("dirichlet() and dirichlet3() tests", {
+
+    f <- function(n){
+        jj <- setNames(sample(n,n),sample(letters[seq_len(n)]))
+        expect_true(dirichlet(alpha=jj+1) == dirichlet(powers=jj))
+    }
+
+    sapply(4:9,f)
+
+    
+    expect_true(
+        dirichlet(powers=c(t=4,r=6,p=2)) ==
+        hyper2(list("r","t",c("t","p","r"),"p") ,c(6,4,-12,2))
+    )
+
+
+    expect_error(dirichlet3(1:5))  # no names attribute
+    
+    jj <- setNames(4:6,c("o","a","w"))
+    expect_error(dirichlet3(jj,setNames(0.8 + 1:3,c("o","w","a")))) # names different order
+    
+    expect_true(
+        dirichlet3(jj,7) ==
+        hyper3(list(c(a=1),c(a=1,o=7,w=1),c(o=7),c(w=1)), powers=c(5,-15,4,6))
+    )
+    
+    expect_true(
+        dirichlet3(jj,lambda=1:3) ==
+        hyper3(list(c(a=2),c(a=2,o=1,w=3),c(o=1),c(w=3)),powers=c(5,-15,4,6))
+    )
+
+} )

### 2024-08-16 inst/hankin-AustrianJournalofStatistics.tex
work on issue #239

--- 2024-05-22 inst/hankin_american_statistician.tex
@@
-\documentclass{article}
+\documentclass[article]{ajs}
 \RequirePackage{amsthm,amsmath,amsfonts,amssymb}
 \RequirePackage{graphicx}%% uncomment this for including figures
-\usepackage{setspace} \doublespacing
-\usepackage{apacite}
 
-\begin{document}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\title{A simplified Plackett--Luce likelihood function for rank orders}
-%\author{Robin K. S. Hankin}
-\maketitle
+%% almost as usual
+\author{Robin K. S. Hankin\,\orcidlink{0000-0001-5982-0415}\\
+  University of Stirling\\ Scotland}
+\title{A Simplified Plackett-Luce Likelihood Function for Rank Orders}
 
-\begin{abstract}
+%% for pretty printing and a nice hypersummary also set:
+\Plainauthor{Robin K. S. Hankin} %% comma-separated
+\Plaintitle{A Simplified Plackett-Luce Likelihood Function for Rank Orders}
+
+%% an abstract and keywords
+\Abstract{
 
@@
 athletics, education, and Formula~1 motor racing.
-\end{abstract}
+
+}
+\Keywords{keywords, comma-separated, non-capitalized, \proglang{R}}
+\Plainkeywords{keywords, comma-separated, non-capitalized,, R} %% without formatting
+%% at least one keyword must be supplied
+
+%% publication information
+%% NOTE: Typically, this can be left commented and will be filled out by the technical editor
+%% \Volume{50}
+%% \Issue{9}
+%% \Month{June}
+%% \Year{2012}
+%% \Submitdate{2012-06-04}
+%% \Acceptdate{2012-06-04}
+%% \setcounter{page}{1}
+\Pages{1--xx}
+
+%% The address of (at least) one author should be given
+%% in the following format:
+\Address{
+  Robin K. S. Hankin\\
+  University of Stirling\\
+  Scotland\\
+  E-mail: \email{hankin.robin@gmail.com}\\
+  URL: \url{https://www.stir.ac.uk/people/1966824}
+}
+%% It is also possible to add a telephone and fax number
+%% before the e-mail in the following format:
+%% Telephone: +43/512/507-7103
+%% Fax: +43/512/507-2851
+
+%% for those who use Sweave please include the following line (with % symbols):
+%% need no \usepackage{Sweave.sty}
+
+%% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{document}
+
+%% include your article here, just as usual
 
@@
   strengths] and say 20 independent complete rank observations,
-\citeA{hankin2017_rmd} shows that maximum likelihood strengths
+\cite{hankin2017_rmd} shows that maximum likelihood strengths
 identify the correct ranking with a probability of essentially zero.
@@
 
-Consider a race between a competitor of Bradley--Terry strength $a$ and
-$n$ cloned competitors, each of strength $b=1-a$.  An observation is
-indexed by~$r$, the number of strength $b$ clones finishing ahead
-of~$a$, and because we have $n+1$ competitors %%%%%%%%%%%%%%%%%%%%%%
-$0\leqslant r\leqslant n$.  The initial field strength is
-$a+nb=n+(n-1)a$.  The Plackett--Luce likelihood
-function~\cite{luce1959,plackett1975} for this situation would be
+\cite{hankin2024_hyper3} considers Plackett-Luce likelihood functions
+for a race between a competitor of Bradley--Terry strength $a$ and
+$n=2$ or $n=3$ cloned competitors, each of strength $b=1-a$.  Here we
+consider the generalized case with arbitrary $n$.  Under these
+circumstances, an observation is indexed by~$r$, the number of
+strength $b$ clones finishing ahead of~$a$, and because we have $n+1$
+competitors %%%%%%%%%%%%%%%%%%%%%% $0\leqslant r\leqslant n$.  The
+initial field strength is $a+nb=n+(n-1)a$.  The Plackett--Luce
+likelihood function~\citep{luce1959,plackett1975} for this situation
+would be
 
@@
 One long-standing contention in the athletic world is that of lane
-bias~\cite{munro2022}.  Running in the inner lane is frequently held
+bias~\citep{munro2022}.  Running in the inner lane is frequently held
 to be a disadvantage; mechanisms include biomechanical issues arising
@@
 which the support is only $0.16$, far short of Edwards's two units of
-support criterion for rejection~\cite{edwards1972}.  This dataset
+support criterion for rejection~\citep{edwards1972}.  This dataset
 contains no evidence to support lane bias.
@@
 Formula 1 motor racing is an important and prestigious motor sport
-\cite{codling2017,jenkins2010}.  Season ranking is based on a points
+\citep{codling2017,jenkins2010}.  Season ranking is based on a points
 allocation system wherein competitors are awarded points based on race
@@
 after the final race.  However, as argued by
-\citeA{hankin2023_formula1points}, the drivers' {\em ranks} are
+\cite{hankin2023_formula1points}, the drivers' {\em ranks} are
 statistically sufficient for analysis using Plackett--Luce approaches.
@@
 finishes in 2022--3.  Noting that Checo raced against 73 distinct
-competitors (this rendering estimation of the other players'
-Plackett--Luce strengths effectively impossible~\cite{hankin2020}), we
-are nevertheless in a position to consider competitive strength using
-the modified likelihood approach presented here.  We consider the
-following inference problem:
+competitors---this rendering estimation of the other players'
+Plackett--Luce strengths effectively
+impossible~\citep{hankin2020}---we are nevertheless in a position to
+consider competitive strength using the modified likelihood approach
+presented here.  We consider the following inference problem:
 
@@
 
-\bibliographystyle{apacite}
+%\bibliographystyle{plainat}
 \bibliography{hyper2}
+
 \end{document}

### 2024-08-16 inst/hankin-AustrianJournalofStatistics.tex
typos and tweaks

@@
 finishing line.  The Plackett--Luce likelihood
-function~\cite{luce1959,plackett1975} is a generalization of the
-Bradley--Terry model~\cite{bradley1952} in which non-negative strengths
-are assigned to each competitor; without loss of generality, if the
-finishing order is $1,2,\ldots,n$ and competitor $i$ has strength
-$p_i\geqslant 0$, then the Plackett--Luce likelihood function will be
-proportional to
+function~\citep{luce1959,plackett1975} is a generalization of the
+Bradley--Terry model~\citep{bradley1952} in which non-negative
+strengths are assigned to each competitor; without loss of generality,
+if the finishing order is $1,2,\ldots,n$ and competitor $i$ has
+strength $p_i\geqslant 0$, then the Plackett--Luce likelihood function
+will be proportional to
 
@@
 even moderately large: even with as few as 20 competitors following
-Zipf's law~\cite{zipf1949} [a typical assumption for Bradley--Terry
+Zipf's law~\citep{zipf1949} [a typical assumption for Bradley--Terry
   strengths] and say 20 independent complete rank observations,
@@
 
-\cite{hankin2024_hyper3} considers Plackett-Luce likelihood functions
-for a race between a competitor of Bradley--Terry strength $a$ and
-$n=2$ or $n=3$ cloned competitors, each of strength $b=1-a$.  Here we
-consider the generalized case with arbitrary $n$.  Under these
-circumstances, an observation is indexed by~$r$, the number of
-strength $b$ clones finishing ahead of~$a$, and because we have $n+1$
+Consider a Plackett-Luce likelihood functions for a race between a
+competitor of Bradley--Terry strength $a$ and $n\geqslant 1$ cloned
+competitors, each of strength $b=1-a$.  Under these circumstances, an
+observation is indexed by~$r$, the number of strength $b$ clones
+finishing ahead of~$a$, and because we have $n+1$
 competitors %%%%%%%%%%%%%%%%%%%%%% $0\leqslant r\leqslant n$.  The
-initial field strength is $a+nb=n+(n-1)a$.  The Plackett--Luce
+initial field strength is $a+nb=n-(n-1)a$.  The Plackett--Luce
 likelihood function~\citep{luce1959,plackett1975} for this situation
@@
 \log\mathcal{L}_{r,n}(a)=\log(B-1)-B\frac{r+1}{n}
-\mathcal{O}\left(n^{-2}\right)
++\mathcal{O}\left(n^{-2}\right)
 \end{equation}
@@
 
-Previous related work presented the case $n=2$ [in current notation].
-Considering the case $n=8$ gives us $8+1=9$ distinct likelihood
-functions, Figure~\ref{ninelikes}.  We see an unanticipated asymmetry
-between $r$ and $n-r$.  For $r=4$ one might expect that
-$\hat{a}=\frac{1}{2}$ on the grounds that the focal competitor's
-performance was median.  However, $\hat{a}\simeq 0.5886$.
+Previous related work \citep{hankin2024_hyper3} presented the case
+$n=2$ [in current notation].  Considering the case $n=8$ gives us
+$8+1=9$ distinct likelihood functions, Figure~\ref{ninelikes}.  We see
+an unanticipated asymmetry between $r$ and $n-r$.  For $r=4$ one might
+expect that $\hat{a}=\frac{1}{2}$ on the grounds that the focal
+competitor's performance was median.  However, $\hat{a}\simeq 0.5886$.
 
@@
 evaluate of $0.474$ is very close to the neutral value of $0.5$, for
-which the support is only $0.16$, far short of Edwards's two units of
-support criterion for rejection~\citep{edwards1972}.  This dataset
+which the support is only $0.16$, far short of the standard two units
+of support criterion for rejection~\citep{edwards1972}.  This dataset
 contains no evidence to support lane bias.
@@
 along the null diagonal is about $-2.43$, at
-${p}_\mathrm{pure}=0.567={p}_\mathrm{applied}=0.713$.  Edwards's two
-units of support criterion~\cite{edwards1972} is met, and we are
-justified in concluding that the student is indeed more competent at
-applied than pure mathematics.  Alternatively, we may observe that
+${p}_\mathrm{pure}=0.567={p}_\mathrm{applied}=0.713$.  The two units
+of support criterion~\citep{edwards1972} is met, and we are justified
+in concluding that the student is indeed more competent at applied
+than pure mathematics.  Alternatively, we may observe that
 $2\mathcal{S}=4.86$ is in the tail region of its asymptotic $\chi^2_1$

### 2024-08-16 inst/hankin-AustrianJournalofStatistics.tex
typo

@@
 along the null diagonal is about $-2.43$, at
-${p}_\mathrm{pure}=0.567={p}_\mathrm{applied}=0.713$.  The two units
-of support criterion~\citep{edwards1972} is met, and we are justified
-in concluding that the student is indeed more competent at applied
-than pure mathematics.  Alternatively, we may observe that
+${p}_\mathrm{pure}={p}_\mathrm{applied}=0.713$.  The two units of
+support criterion~\citep{edwards1972} is met, and we are justified in
+concluding that the student is indeed more competent at applied than
+pure mathematics.  Alternatively, we may observe that
 $2\mathcal{S}=4.86$ is in the tail region of its asymptotic $\chi^2_1$

### 2024-08-20 inst/hankin-AustrianJournalofStatistics.tex
more work on issue #239

@@
 competitor and $b$ to all others; we require $a+b=1$.  The associated
-inference problem appears to have a wide range of applications and I
-present some analysis of datasets drawn from the fields of Olympic
-athletics, education, and Formula~1 motor racing.
+inference problem has a wide range of applications and I present some
+analysis of datasets drawn from the fields of Olympic athletics,
+education, and Formula~1 motor racing.
 
@@
 strength $p_i\geqslant 0$, then the Plackett--Luce likelihood function
-will be proportional to
+for this observation will be proportional to
 
@@
 
-Consider a Plackett-Luce likelihood functions for a race between a
-competitor of Bradley--Terry strength $a$ and $n\geqslant 1$ cloned
-competitors, each of strength $b=1-a$.  Under these circumstances, an
-observation is indexed by~$r$, the number of strength $b$ clones
-finishing ahead of~$a$, and because we have $n+1$
-competitors %%%%%%%%%%%%%%%%%%%%%% $0\leqslant r\leqslant n$.  The
-initial field strength is $a+nb=n-(n-1)a$.  The Plackett--Luce
-likelihood function~\citep{luce1959,plackett1975} for this situation
-would be
+\subsection{Simplified Plackett-Luce likelihoods}
+
+As noted, likelihood function~\ref{plackettluce} becomes unmanageable
+with $n$ even moderately large and some form of simplification is
+needed.  To that end, consider a Plackett-Luce likelihood function for
+a race between a competitor of Bradley-Terry strength $a$ and
+$n\geqslant 1$ cloned competitors, each of strength $b=1-a$.  Under
+these circumstances, an observation is indexed by~$r$, the number of
+strength $b$ clones finishing ahead of~$a$, and because we have $n+1$
+competitors $0\leqslant r\leqslant n$.  The initial field strength is
+$a+nb=n-(n-1)a$.  The Plackett--Luce likelihood
+function~\citep{luce1959,plackett1975} for this situation would be
 
@@
 \noindent where $B=1/b$.  Note carefully that these expressions are
-likelihoods, not probabilities, and do not sum to 1.  Asymptotic
-analysis of expression~\ref{likeforrn3} shows that
+likelihoods, not probabilities, and do not sum to 1.  Previous related
+work \citep{hankin2024_hyper3} presented the case $n=2$ [in current
+  notation].  Considering the case $n=8$ gives us $8+1=9$ distinct
+likelihood functions, Figure~\ref{ninelikes}.  We see an unanticipated
+asymmetry between $r$ and $n-r$.  For $r=4$ one might expect that
+$\hat{a}=\frac{1}{2}$ on the grounds that the focal competitor's
+performance was median.  However, $\hat{a}\simeq 0.5886$.
+Figure~\ref{dotprobs} shows the maximum likelihood estimator $\hat{a}$
+for various values of $n,r$, obtained by direct numerical maximization
+of equation~\ref{likeforrn3}.
+
+\begin{figure}[t]
+  \begin{centering}
+\includegraphics[width=4in]{ninelikes}  % ninelikes.pdf produced by very_simplified_likelihood.Rmd
+\caption{Likelihood functions for $r=0(1)8$\label{ninelikes}.  Value
+  of $r$ indicated in colour by its curve; vertical line shows
+  $\hat{a}\simeq 0.5886$ for the case $r=4$}
+\end{centering}
+\end{figure}
+
+\subsection{Asymptotic analysis}
+
+Analysis of expression~\ref{likeforrn3} for fixed $n$ shows that the
+support $\mathcal{S}=\log\mathcal{L}_{r,n}(a)$ is asymptotically
 
 \begin{equation}\label{asymptotic}
-\log\mathcal{L}_{r,n}(a)=\log(B-1)-B\frac{r+1}{n}
+\mathcal{S}=
+%\log\mathcal{L}_{r,n}(a)=
+\log(B-1)-B\frac{r+1}{n}
 +\mathcal{O}\left(n^{-2}\right)
@@
 (up to an arbitrary additive constant).  Using
-$\mathcal{S}'=\log(B-1)-B\frac{r+1}{n}$ as an approximate support thus
-incurs an error of only $\mathcal{O}(n^{-2})$.  Observing that
-$\partial^2\mathcal{S'}/\partial B^2=-(B-1)^{-2}<0$, the evaluate
+$\mathcal{S}^*=\log(B-1)-B\frac{r+1}{n}$ as an approximate
+support thus incurs an error of only $\mathcal{O}(n^{-2})$.  Observing
+that $\partial^2\mathcal{S}^*/\partial B^2=-(B-1)^{-2}<0$, the evaluate
 would be unique; we have
@@
 alternatively $\hat{a}=\frac{n}{n+r+1}$.  Compare this with the result
-of asking ``what is the probability that the focal competitor
-beats a randomly chosen opponent?'' to which the binomial maximum
-likelihood estimate is simply $\frac{n-r}{n}$, agreeing to within
+of asking ``what is the probability that the focal competitor beats a
+randomly chosen opponent?'' to which the binomial maximum likelihood
+estimate is simply $\frac{n-r}{n}$, agreeing to within
 $\mathcal{O}(n^{-2})$.  For repeated observations $r_i,n_i$,
@@
 r_i\right/\sum n_i$.
-
-Previous related work \citep{hankin2024_hyper3} presented the case
-$n=2$ [in current notation].  Considering the case $n=8$ gives us
-$8+1=9$ distinct likelihood functions, Figure~\ref{ninelikes}.  We see
-an unanticipated asymmetry between $r$ and $n-r$.  For $r=4$ one might
-expect that $\hat{a}=\frac{1}{2}$ on the grounds that the focal
-competitor's performance was median.  However, $\hat{a}\simeq 0.5886$.
-
-\begin{figure}[t]
-  \begin{centering}
-\includegraphics[width=4in]{ninelikes}  % ninelikes.pdf produced by very_simplified_likelihood.Rmd
-\caption{Likelihood functions for $r=0(1)8$\label{ninelikes}.  Value
-  of $r$ indicated in colour by its curve; vertical line shows
-  $\hat{a}\simeq 0.5886$ for the case $r=4$}
-\end{centering}
-\end{figure}
 
@@
   lane-1 runner, Olympic qualifying heats 2012-2020 \label{plotolymp}}
+\end{centering}
+\end{figure}
+
+\subsection{Park run}
+
+``Parkrun'' is a distributed community initiative that organises
+weekly timed 5 km runs/walks in parks worldwide~\citep{hindley2020}.
+A typical event will have 200 participants.  Table~\ref{parkruntable}
+shows the author has completed a total of 15 parkruns to date,
+comprising 10 runs in Auckland and 5 at Stirling.  From the first pair
+of numbers we see that 238 runners attended that particular parkrun,
+of whom the author placed 173.
+
+\begin{table}[t]
+  \centering
+  \caption{Parkrun results}
+\label{parkruntable}
+\begin{tabular}{ccccccccccc}
+  \\
+ & \multicolumn{10}{c}{Auckland}\\
+rank    & 163  & 155 & 162 & 189 & 171 & 219 & 167 & 212 & 206 & 142 \\
+runners & 238  & 238 & 196 & 242 & 242 & 318 & 259 & 305 & 297 & 241 \\ \\ \\
+\end{tabular}
+\begin{tabular}{cccccc}
+  & \multicolumn{5}{c}{Stirling}\\
+rank   & 118  & 224 & 128 & 115 & 183 \\
+runners& 179  & 338 & 203 & 245 & 254 \\
+\end{tabular}
+\end{table}
+
+Is there evidence that the author's performance differs between
+Auckland and Stirling?  Figure~\ref{parkruncontour} shows contours of
+support [again normalised so the support of the evaluate is zero] for
+the parkrun dataset, as a function of Auckland strength on the
+horizontal axis and Stirling strength on the vertical axis.  A null of
+equal strengths appears as the diagonal line.  We see that the maximum
+likelihood estimate is $\hat{p}_\mathrm{Auck}=0.427$,
+$\hat{p}_\mathrm{Stir}=0.408$.  However, a constrained optimization
+shows that the maximum support along the null diagonal is about
+$-2.43$, at ${p}_\mathrm{Auck}={p}_\mathrm{Stir}\simeq 0.409$.  The
+two units of support criterion~\citep{edwards1972} is met, and we are
+justified in concluding that the student is indeed more competent at
+applied than pure mathematics.  Alternatively, we may observe that
+$2\mathcal{S}=4.86$ is in the tail region of its asymptotic $\chi^2_1$
+distribution, corresponding to a $p$-value of about $0.028$.
+
+\begin{figure}[t]
+  \begin{centering}
+\includegraphics[width=4in]{parkruncontour} % produced by very_simplified_likelihood.Rmd
+\caption{Support contours for Bradley-Terry strength
+  at \label{parkruncontour} Auckland and Stirling.  Null of equal
+  strengths shown as a diagonal line}
 \end{centering}
@@
 along the null diagonal is about $-2.43$, at
-${p}_\mathrm{pure}={p}_\mathrm{applied}=0.713$.  The two units of
+${p}_\mathrm{pure}={p}_\mathrm{applied}\simeq 0.713$.  The two units of
 support criterion~\citep{edwards1972} is met, and we are justified in

### 2024-08-20 inst/very_simplified_likelihood.Rmd
more work on issue #239

@@
 ```{r,label=plotbothlikes}
+parkrun_contour <- function(...){
     par(pty='s')
-contour(x,x,L,asp=1,xlim=range(x),ylim=range(x),levels=-(0:5))
+    contour(x,x,L,asp=1,xlim=range(x),ylim=range(x),levels=-(0:5),xlab="Auckland",ylab="Stirling")
     abline(0,1)
+}
+parkrun_contour()
+pdf(file="parkruncontour.pdf")
+parkrun_contour()
+dev.off()
 ```
@@
 ```
+
+# Park run, version 2.
+
+```{r}
+
+rank_auck    <- c(177, 222, 206, 142, 118, 224, 128, 115, 183)
+runners_auck <- c(259, 305, 297, 241, 179, 338, 203, 245, 254)
+
+rank_stir <-    c(145, 157, 156, 191, 172, 174, 173, 165, 172, 199, 181, 229)
+runners_stir <- c(193, 240, 197, 257, 227, 234, 238, 238, 196, 242, 242, 318)
+```
+
+```{r}
+support_vec <- function(a,ranks,runners){
+    out <- 0
+    for(i in seq_along(ranks)){
+        out <- out + f_single(a, ranks[i],runners[i],log=TRUE)
+    }
+    return(out)
+}
+
+support_ij <- function(v){
+    auck <- v[1]
+    stir <- v[2]
+    return(
+        support_vec(auck, rank_auck, runners_auck) +
+        support_vec(stir, rank_stir, runners_stir)
+    )
+}
+
+x <- seq(from=0.1,to=0.8,by=0.01)
+jj <- as.matrix(expand.grid(x,x))
+L <- apply(jj,1,support_ij)
+L <- L - max(L,na.rm=TRUE)
+L <- matrix(L,length(x),length(x))
+```
+
+```{r}
+parkrun_contour <- function(...){
+    par(pty='s')
+    contour(x,x,L,asp=1,xlim=range(x),ylim=range(x),levels=-(0:5),xlab="Auckland",ylab="Stirling")
+    abline(0,1)
+}
+parkrun_contour()
+```
+
+```{r}
+M <- matrix(c(
+    sum(rank_auck-1),sum(runners_auck-rank_auck),
+    sum(rank_stir-1),sum(runners_stir-rank_stir)),2,2,byrow=TRUE)
+dimnames(M) <- list(venue=c("Auckland","Stirling"),better=c(T,F))
+M
+fisher.test(M)
+```

### 2024-08-21 inst/very_simplified_likelihood.Rmd
replace with actual data
start work on #240

@@
 $n+1$ competitors in total, 1 of strength $a$ and $n$ of strength $b$.
-An observation is indexed by $r$, the number of $b$ clones finishing
-ahead of $a$, so $0\leqslant r\leqslant n$.  The initial field
-strength is $a+nb=1+b(n-1)$.
-
-Suppose our bro comes first (zero $b$'s ahead of him):
+An observation is indexed by $R$, the rank of the focal competitor: we
+have $1\leqslant R\leqslant n+1$.  Alternatively we might consider
+$r=R-1$, the number of $b$ clones finishing ahead of $a$, so
+$0\leqslant r\leqslant n$.  The initial field strength is
+$a+nb=1+b(n-1)$.
+
+$$r=R-1$$
+
+Suppose our bro comes first (zero $b$'s ahead of him) [$r=0, R=1$]:
 
@@
 
-Comes second, one $b$ ahead of him:
+Comes second [$r=1,R=2$], one $b$ ahead of him:
 
@@
 
-Third, two ahead of him:
+Third [$r=2, R=3$]; two ahead of him:
 
@@
 
-Fourth, three ahead:
+Fourth [$r=3, R=4$], three ahead:
 
@@
 
+\[
+\mathcal{L}_{R,n}(a)=\frac{(B-1)(B+n-R-1)!}{(B+n-1)!},\qquad B=1/(1-a)
+\]
+
+\[
+=\frac{(B-1)\Gamma(B+n-R)}{\Gamma(B+n)},\qquad B=1/(1-a)
+\]
+
 Then, using Stirling [viz $n!\sim n^ne^{-n}\sqrt{2\pi n}$] we get
@@
 \log(B-1) + ((B+n-r-2)\log(B+n-r-2)-(B+n-r-2)+\frac{\log(B+n-r-2)}{2})
+          - ((B+n  -1)\log(B+n  -1)-(B+n  -1)+\frac{\log(B+n  -1)}{2})
+\]
+
+\[
+\log(B-1) + ((B+n-R-1)\log(B+n-R-1)-(B+n-R-1)+\frac{\log(B+n-R-1)}{2})
           - ((B+n  -1)\log(B+n  -1)-(B+n  -1)+\frac{\log(B+n  -1)}{2})
@@
 \]
+
+Now using $R$ not $r$:
+
+```
+sage: var('B n R')
+(B, n, R)
+pretty_print_default(False)
+X = log(B-1) + (
+  + ((B+n-R-1)*log(B+n-R-1)-(B+n-R-1)+log(B+n-R-1)/2)
+  - ((B+n  -1)*log(B+n  -1)-(B+n  -1)+log(B+n  -1)/2)
+  )
+X
+(B - R + n - 1)*log(B - R + n - 1) - (B + n - 1)*log(B + n - 1) + R + 1/2*log(B - R + n - 1) - 1/2*log(B + n - 1) + log(B - 1)
+X.taylor(n,Infinity,2).full_simplify()
+-1/12*(12*R*n^2*log(n) + 3*(2*B - 1)*R^2 - 2*R^3 - 12*n^2*log(B - 1) - 6*(B^2 - B)*R + 6*((2*B - 1)*R - R^2)*n)/n^2
+```
 
@@
 results below.
+
+Now some Mathematica:
+
+```
+f[B_, n_, R_] := Log[(B - 1)*Gamma[B + n - R]/Gamma[B + n]] 
+Assuming[B > 1 && p > 0 && p < 1, 
+ Series[f[B, n, n*p], {n, Infinity, 1}]]
+```
+
+We get
+
+$$
+\log\left(
+(B-1)(1-p)^{B-1/2} +
+\frac{
+	(B-1)(1-6B+6B^2)(1-p)^{B-1/2}p
+}{
+	12(1-p)n
+}
+\right)
+$$
+
+or
+
+$$
+\log\left(
+(B-1)(1-p)^{B-1/2} \left[1+
+\frac{(1-6B+6B^2)p}{2(1-p)n}\right]\right)
+$$
+
+$$
+\log
+(B-1) + B\log(1-p) + 
+\frac{(1-6B+6B^2)p}{2(1-p)n}\right]\right)
+$$
 
@@
 ```{r}
-
-rank_auck    <- c(177, 222, 206, 142, 118, 224, 128, 115, 183)
-runners_auck <- c(259, 305, 297, 241, 179, 338, 203, 245, 254)
-
-rank_stir <-    c(145, 157, 156, 191, 172, 174, 173, 165, 172, 199, 181, 229)
-runners_stir <- c(193, 240, 197, 257, 227, 234, 238, 238, 196, 242, 242, 318)
+a <- read.table("parkrun_results.txt",header=TRUE)
+
+rank_auck    <- a$place  [a$venue == "auckland"]
+runners_auck <- a$runners[a$venue == "auckland"]
+
+rank_stir    <- a$place  [a$venue == "stirling"]
+runners_stir <- a$runners[a$venue == "stirling"]
+
+rank_auck    
+runners_auck
+
+rank_stir 
+runners_stir
+
 ```

### 2024-08-21 inst/very_simplified_likelihood.Rmd
further work on issue #140

@@
 (B-1)(1-p)^{B-1/2} \left[1+
-\frac{(1-6B+6B^2)p}{2(1-p)n}\right]\right)
+\frac{(1-6B+6B^2)p}{12(1-p)n}\right]\right)
 $$
@@
 (B-1) + B\log(1-p) +
-\frac{(1-6B+6B^2)p}{2(1-p)n}\right]\right)
+n^{-1}
+\frac{(1-6B+6B^2)p}{12(1-p)} +{\mathcal O}\left(n^{-2}\right)
 $$
 
+$$
+\log
+(B-1) + B\log(1-p) +
+\frac{1}{12}n^{-1}\frac{p}{1-p}(1-6B+6B^2) + {\mathcal O}\left(n^{-2}\right)
+$$
+
 # Placing last
 
-Placing last is finishing  after $n=r$ clones finish:
-
-\[
-n=1\longrightarrow\mathcal{L}=b\]
-
-\[
-n=2\longrightarrow\mathcal{L}=\frac{b}{a+2b}\cdot\frac{b}{a+b}=\frac{b^2}{1+b}
+Placing last is finishing after $n=r$ clones finish, that is, $R=n+1$
+[me plus two clones, I come third].
+
+\[
+n=1,R=2\longrightarrow\mathcal{L}=b\]
+
+\[
+n=2,R=3\longrightarrow\mathcal{L}=\frac{b}{a+2b}\cdot\frac{b}{a+b}=\frac{b^2}{1+b}
 =\frac{1}{B(B+1)}
@@
 \[
-n=3\longrightarrow\mathcal{L}=\frac{b}{a+3b}\cdot\frac{b}{a+2b}\cdot\frac{b}{a+b}=\frac{b^3}{(1+2b)(1+b)}=\frac{1}{B(B+1)(B+2)}
-\]
-
-\[
-n=4\longrightarrow\mathcal{L}=\frac{b}{a+4b}\cdot\frac{b}{a+3b}\cdot\frac{b}{a+2b}\cdot\frac{b}{a+b}=
+n=3,R=4\longrightarrow\mathcal{L}=\frac{b}{a+3b}\cdot\frac{b}{a+2b}\cdot\frac{b}{a+b}=\frac{b^3}{(1+2b)(1+b)}=\frac{1}{B(B+1)(B+2)}
+\]
+
+\[
+n=4,R=n+1\longrightarrow\mathcal{L}=\frac{b}{a+4b}\cdot\frac{b}{a+3b}\cdot\frac{b}{a+2b}\cdot\frac{b}{a+b}=
 \frac{b^4}{(1+3b)(1+2b)(1+b)}=
@@
 \[
-\frac{1}{(B+3)(B+2)(B+1)B}=\frac{(B-1)!}{(B+3)!}
+\frac{1}{(B+3)(B+2)(B+1)B}=\frac{(B-1)!}{(B+3)!}=\frac{\Gamma(B)}{\Gamma(B+2)}
 \]
@@
 \frac{1}{(B+n-1)(B+n-2)\cdots(B+1)B}=
-\frac{(B-1)!}{(B+n-1)!}
+\frac{(B-1)!}{(B+n-1)!}=
+\frac{\Gamma(B)}{\Gamma(B+n)}
 \]
@@
 ```{r,label=definefsingle}
-f_single <- function(a,r,n,log=FALSE){  # not vectorised
+f_single <- function(a,R,n,log=FALSE){  # not vectorised
+    stopifnot(R>0)
 	B <- 1/(1-a)
 	if(log){
-		out <- log(B-1) - lgamma(B+n) + lgamma(B+n-r-1)
+		out <- log(B-1) + lgamma(B+n-R) - lgamma(B+n)
 	} else {  # not-log
-		out <- (B-1)/prod(B+(n-r-1):(n-1))
+		out <- (B-1)/prod(B+(n-R):(n-1))
 	}
@@
 
-f_vec_a <- function(a,r,n, log=FALSE){  # vectorised in 'a' but not in 'r'
-	sapply(a,function(a){f_single(a,r=r,n=n,log=log)})
-}
-
-f_vec_arn <- function(a,r,n,log=FALSE){
-	## vectorized in 'a'; treats 'r' and 'n' as
+f_vec_a <- function(a,R,n, log=FALSE){  # vectorised in 'a' but not in 'R'
+	sapply(a,function(a){f_single(a,R=R,n=n,log=log)})
+}
+
+f_vec_arn <- function(a,R,n,log=FALSE){
+	## vectorized in 'a'; treats 'R' and 'n' as
 	## vectors of independent observations
 
-	M <- cbind(r,n)
+	M <- cbind(R,n)
 
@@
 		out <- 0
-		for(i in seq_len(nrow(M))){out <- out + f_vec_a(a,r=M[i,1],n=M[i,2],log=TRUE)}
+		for(i in seq_len(nrow(M))){out <- out + f_vec_a(a,R=M[i,1],n=M[i,2],log=TRUE)}
 	} else {
 		out <- 1
-		for(i in seq_len(nrow(M))){out <- out * f_vec_a(a,r=M[i,1],n=M[i,2],log=FALSE)}
+		for(i in seq_len(nrow(M))){out <- out * f_vec_a(a,R=M[i,1],n=M[i,2],log=FALSE)}
 	}
@@
 
-fapprox <- function(a,r,n,log=FALSE){
+fapprox <- function(a,R,n,log=FALSE){
 	if(log){
-		return(log(B-1)-B*(r+1)/n)
+		return(log(B-1)-B*R/n)
 	} else {
-		return(B-1 - (r+1)*exp(B)/n)
+		return((B-1)/exp(R*B/n))
 	}
@@
 ```{r label=defdiff}
-fdash <- function(a,r,n,log=TRUE){
+fdash <- function(a,R,n,log=TRUE){
 	B <- 1/(1-a)
-	out <- (1/(B-1) - psigamma(B+n) + psigamma(B+n-r-1))*B^2
-	if(!log){out <- out * f_vec_arn(a,r,n,log=FALSE)}
+	out <- (1/(B-1) - psigamma(B+n) + psigamma(B+n-R))*B^2
+	if(!log){out <- out * f_vec_arn(a,R,n,log=FALSE)}
 	return(out)
@@
 d <- 1e-3
-r <- 4
+R <- 4
 n <- 10
 c(
-  numerical  = (f_single(a+d/2,r,n,log=TRUE)-f_single(a-d/2,r,n,log=TRUE))/d,
-  analytical = fdash(a,r,n)
+  numerical  = (f_single(a+d/2,R,n,log=TRUE)-f_single(a-d/2,R,n,log=TRUE))/d,
+  analytical = fdash(a,R,n)
 )
@@
 c(
-  numerical  = (f_single(a+d/2,r,n,log=FALSE)-f_single(a-d/2,r,n,log=FALSE))/d,
-  analytical = fdash(a,r,n,log=FALSE)
+  numerical  = (f_single(a+d/2,R,n,log=FALSE)-f_single(a-d/2,R,n,log=FALSE))/d,
+  analytical = fdash(a,R,n,log=FALSE)
 )
@@
 ```{r,label=defmle}
-MLE <- function(r,n,give=FALSE){
+MLE <- function(R,n,give=FALSE){
 	d <- 1e-6
-	out <- optimize(f_vec_arn,c(d,1-d),r=r,n=n,log=TRUE,maximum=TRUE)
+	out <- optimize(f_vec_arn,c(d,1-d),R=R,n=n,log=TRUE,maximum=TRUE)
 	if(!give){out <- out$maximum}
@@
 
-MLEa <- function(r,n,give=FALSE,small=1e-4){
-	out <- uniroot(f=function(a){fdash(a,r,n)},interval=c(small,1-small))
+MLEa <- function(R,n,give=FALSE,small=1e-4){
+	out <- uniroot(f=function(a){fdash(a,R,n)},interval=c(small,1-small))
 	if(!give){out <- out$root}
@@
 rbind(
-showdiff(MLE(1,9),MLEa(1,9)),
 showdiff(MLE(3,9),MLEa(3,9)),
@@
 n <- 9
-r <- 3
+R <- 4
 c(
@@
 way2 = (B-1)/prod(B+(5:8)),
-way3 = f_vec_arn(a,r,n,log=FALSE)
+way3 = f_vec_arn(a,R,n,log=FALSE)
 )
@@
   out[[n]]  <- rep(NA,n+1)
-  for(r in 0:n){
-    if(r==0){
+  for(R in seq_len(n+1)){
+    if(R==1){
       jj <- 1
-    } else if(r==n){n
+    } else if(R == n+1){
       jj <- 0
     } else {
-      jj <- MLE(r,n)
-    }
-    out[[n]][r+1] <- jj
+      jj <- MLE(R,n)
+    }
+    out[[n]][R] <- jj
   }
@@
 for(n in 1:10){
-  for(r in 0:n){   points(n,out[[n]][r+1],pch=16) }
-  for(r in 0:n){   text(n+0.2,out[[n]][r+1],r,cex=0.5,col='gray') }
+  for(R in seq_len(n+1)){   points(n,out[[n]][R],pch=16) }
+  for(R in seq_len(n+1)){   text(n+0.2,out[[n]][R],R,cex=0.5,col='gray') }
 }
@@
 rain <- rainbow(n+1)
-for(r in seq(from=0,to=n)){
-  y <- f_vec_a(a,r,n,log=FALSE)
-  if(r==0){y[length(y)] <- 1}
+for(R in 1:n){
+  y <- f_vec_a(a,R,n,log=FALSE)
+  if(R==1){y[length(y)] <- 1}
   y <- y/max(y,na.rm=TRUE)
-  if(r>0){y[length(y)] <- 0}
-  points(a,y,type='l',lwd=4,col=rain[r+1])
-}
-abline(v=MLE(r=4,n=8))
-text(0.07,0.95,"r=8",col=rain[9])
-text(0.20,0.95,"r=7",col=rain[8])
-text(0.26,0.91,"r=6",col=rain[7])
-text(0.29,0.83,"r=5",col=rain[6])
-text(0.33,0.76,"r=4",col=rain[5])
-text(0.35,0.64,"r=3",col=rain[4])
-text(0.39,0.53,"r=2",col=rain[3])
-text(0.45,0.41,"r=1",col=rain[2])
-text(0.63,0.22,"r=0",col=rain[1])
+  if(R>1){y[length(y)] <- 0}
+  points(a,y,type='l',lwd=4,col=rain[R])
+}
+abline(v=MLE(R=5,n=8))
+text(0.07,0.95,"R=9",col=rain[9])
+text(0.07,0.95,"R=8",col=rain[8])
+text(0.20,0.95,"R=7",col=rain[7])
+text(0.26,0.91,"R=6",col=rain[6])
+text(0.29,0.83,"R=5",col=rain[5])
+text(0.33,0.76,"R=4",col=rain[4])
+text(0.35,0.64,"R=3",col=rain[3])
+text(0.39,0.53,"R=2",col=rain[2])
+text(0.45,0.41,"R=1",col=rain[1])
 }
@@
 ```{r shown4n8mle}
-MLE(r=4,n=8)
+MLE(R=5,n=8)
 ```
@@
   out <- a*0
-  for(year in 2011:2023){ out <- out + f_vec_arn(a,getfoc(year)-1,getnum(year)-1,log=TRUE) }
+  for(year in 2011:2023){ out <- out + f_vec_arn(a,getfoc(year),getnum(year)-1,log=TRUE) }
   return(out)
@@
      strength <- exp(LO)/(1+exp(LO))
-     out <- out + f_vec_arn(strength,getfoc(year)-1,getnum(year)-1,log=TRUE)
+     out <- out + f_vec_arn(strength,getfoc(year),getnum(year)-1,log=TRUE)
   }
@@
 ```{r findcredint}
-(a_lower <- uniroot(function(x){jjf(x)+2-Lmax},interval=c(0.3,0.5))$root)
-(a_upper <- uniroot(function(x){jjf(x)+2-Lmax},interval=c(0.5,0.6))$root)
+(a_lower <- uniroot(function(x){jjf(x)+2-Lmax},interval=c(0.1,0.5))$root)
+(a_upper <- uniroot(function(x){jjf(x)+2-Lmax},interval=c(0.5,0.9))$root)
 ```
@@
 a <- seq(from=0.005,to=0.85,by=0.01)
-La <- f_vec_arn(a,r=results$place[1]-1,results$runners[1]-1,log=TRUE)
+La <- f_vec_arn(a,R=results$place[1],results$runners[1]-1,log=TRUE)
 La <- La - max(La,na.rm=TRUE)
@@
   for(i in 1:3){
-    out <- out + f_vec_a(a_stirling,r=place[i]-1,runners[i]-1,log=TRUE)
+    out <- out + f_vec_a(a_stirling,R=place[i],runners[i]-1,log=TRUE)
   }
   for(i in 4:10){
-    out <- out + f_vec_a(a_auckland,r=place[i]-1,runners[i]-1,log=TRUE)
+    out <- out + f_vec_a(a_auckland,R=place[i],runners[i]-1,log=TRUE)
   }
@@
 plot(a,Lpure,ylab='pure strength')
-optimize(function(a){f_vec_arn(a,rank[wp]-1,class_size[wp]-1,log=TRUE)},c(0.3,0.6),maximum=TRUE)
-optimize(function(a){f_vec_arn(a,rank[wa]-1,class_size[wa]-1,log=TRUE)},c(0.3,0.6),maximum=TRUE)
+optimize(function(a){f_vec_arn(a,rank[wp],class_size[wp]-1,log=TRUE)},c(0.3,0.6),maximum=TRUE)
+optimize(function(a){f_vec_arn(a,rank[wa],class_size[wa]-1,log=TRUE)},c(0.3,0.6),maximum=TRUE)
 ```
@@
   for(i in seq_len(nrow(M1))){
-    out <- out + f_vec_arn(vec[1],r=M1[i,1]-1,n=M1[i,2]-1,log=TRUE)
+    out <- out + f_vec_arn(vec[1],R=M1[i,1],n=M1[i,2]-1,log=TRUE)
   }
   for(i in seq_len(nrow(M2))){
-    out <- out + f_vec_arn(vec[2],r=M2[i,1]-1,n=M2[i,2]-1,log=TRUE)
+    out <- out + f_vec_arn(vec[2],R=M2[i,1],n=M2[i,2]-1,log=TRUE)
   }
@@
    function(vec){
-      supp2_ed(vec,rank[wp]-1,rank[wa]-1,class_size[wp]-1,class_size[wa]-1)},
+      supp2_ed(vec,rank[wp],rank[wa],class_size[wp]-1,class_size[wa]-1)},
    control=list(fnscale = -1)

### 2024-08-22 inst/very_simplified_likelihood.Rmd
further work on issue #240

@@
 rain <- rainbow(n+1)
-for(R in 1:n){
+for(R in seq_len(n+1)){
   y <- f_vec_a(a,R,n,log=FALSE)
@@
 text(0.07,0.95,"R=9",col=rain[9])
-text(0.07,0.95,"R=8",col=rain[8])
-text(0.20,0.95,"R=7",col=rain[7])
-text(0.26,0.91,"R=6",col=rain[6])
-text(0.29,0.83,"R=5",col=rain[5])
-text(0.33,0.76,"R=4",col=rain[4])
-text(0.35,0.64,"R=3",col=rain[3])
-text(0.39,0.53,"R=2",col=rain[2])
-text(0.45,0.41,"R=1",col=rain[1])
+text(0.20,0.95,"R=8",col=rain[8])
+text(0.26,0.91,"R=7",col=rain[7])
+text(0.29,0.83,"R=6",col=rain[6])
+text(0.33,0.76,"R=5",col=rain[5])
+text(0.35,0.64,"R=4",col=rain[4])
+text(0.39,0.53,"R=3",col=rain[3])
+text(0.45,0.41,"R=2",col=rain[2])
+text(0.54,0.21,"R=1",col=rain[1])
 }
@@
     p <- exp(LO)/(1+exp(LO))
-    S <- S + f_vec_arn(p,M[i,2]-1,M[i,3]-1,log=TRUE)
+    S <- S + f_vec_arn(p,R=M[i,2],n=M[i,3]-1,log=TRUE)
   }
@@
 ```{r calculate_parksupp,cache=TRUE}
-S_parkrun <- function(a){f_vec_arn(a,r=results$place-1,results$runners-1,log=TRUE)}
+S_parkrun <- function(a){f_vec_arn(a,R=results$place,results$runners-1,log=TRUE)}
 (optparkrun <- optimize(S_parkrun, c(0.2,0.8),maximum=TRUE))

### 2024-08-22 inst/very_simplified_likelihood.Rmd
one parkrun analysis, not two

@@
 Above shows that there is no evidence to suggest that the Stirling
-strength differs from the Auckland strength.  Now let's try a
+strength differs from the Auckland strength.
+
+## Fisher parkrun
+
+```{r splitruns}
+a <- read.table("parkrun_results.txt",header=TRUE)
+
+rank_auck    <- a$place  [a$venue == "auckland"]
+runners_auck <- a$runners[a$venue == "auckland"]
+
+rank_stir    <- a$place  [a$venue == "stirling"]
+runners_stir <- a$runners[a$venue == "stirling"]
+```
+
+```{r fisherparkrun}
+M <- matrix(c(
+    sum(rank_auck-1),sum(runners_auck-rank_auck),
+    sum(rank_stir-1),sum(runners_stir-rank_stir)),2,2,byrow=TRUE)
+dimnames(M) <- list(venue=c("Auckland","Stirling"),better=c(T,F))
+M
+fisher.test(M)
+```
+
+## Logistic parkrun
+
+Now let's try a
 logistic-style regression:
@@
 ```
-
-# Park run, version 2.
-
-```{r}
-a <- read.table("parkrun_results.txt",header=TRUE)
-
-rank_auck    <- a$place  [a$venue == "auckland"]
-runners_auck <- a$runners[a$venue == "auckland"]
-
-rank_stir    <- a$place  [a$venue == "stirling"]
-runners_stir <- a$runners[a$venue == "stirling"]
-
-rank_auck    
-runners_auck
-
-rank_stir 
-runners_stir
-
-```
-
-```{r}
-support_vec <- function(a,ranks,runners){
-    out <- 0
-    for(i in seq_along(ranks)){
-        out <- out + f_single(a, ranks[i],runners[i],log=TRUE)
-    }
-    return(out)
-}
-
-support_ij <- function(v){
-    auck <- v[1]
-    stir <- v[2]
-    return(
-        support_vec(auck, rank_auck, runners_auck) +
-        support_vec(stir, rank_stir, runners_stir)
-    )
-}
-
-x <- seq(from=0.1,to=0.8,by=0.01)
-jj <- as.matrix(expand.grid(x,x))
-L <- apply(jj,1,support_ij)
-L <- L - max(L,na.rm=TRUE)
-L <- matrix(L,length(x),length(x))
-```
-
-```{r}
-parkrun_contour <- function(...){
-    par(pty='s')
-    contour(x,x,L,asp=1,xlim=range(x),ylim=range(x),levels=-(0:5),xlab="Auckland",ylab="Stirling")
-    abline(0,1)
-}
-parkrun_contour()
-```
-
-```{r}
-M <- matrix(c(
-    sum(rank_auck-1),sum(runners_auck-rank_auck),
-    sum(rank_stir-1),sum(runners_stir-rank_stir)),2,2,byrow=TRUE)
-dimnames(M) <- list(venue=c("Auckland","Stirling"),better=c(T,F))
-M
-fisher.test(M)
-```

### 2024-09-03 inst/hankin-AustrianJournalofStatistics.tex
replace data with actual observations from parkrun_results.txt

@@
 A typical event will have 200 participants.  Table~\ref{parkruntable}
-shows the author has completed a total of 15 parkruns to date,
-comprising 10 runs in Auckland and 5 at Stirling.  From the first pair
-of numbers we see that 238 runners attended that particular parkrun,
-of whom the author placed 173.
+shows the author has completed a total of 21 parkruns to date,
+comprising 9 runs in Auckland and 12 at Stirling.  From the first pair
+of numbers we see that 259 runners attended that particular parkrun at
+Auckland, of whom the author placed 177.
 
@@
 \label{parkruntable}
-\begin{tabular}{ccccccccccc}
-  \\
- & \multicolumn{10}{c}{Auckland}\\
-rank    & 163  & 155 & 162 & 189 & 171 & 219 & 167 & 212 & 206 & 142 \\
-runners & 238  & 238 & 196 & 242 & 242 & 318 & 259 & 305 & 297 & 241 \\ \\ \\
+\begin{tabular}{cccccccccc}\\
+ & \multicolumn{9}{c}{Auckland}\\
+rank   & 177& 222& 206& 142& 118& 224& 128& 115& 183\\
+runners& 259& 305& 297& 241& 179& 338& 203& 245& 254\\ \\
 \end{tabular}
-\begin{tabular}{cccccc}
-  & \multicolumn{5}{c}{Stirling}\\
-rank   & 118  & 224 & 128 & 115 & 183 \\
-runners& 179  & 338 & 203 & 245 & 254 \\
+\begin{tabular}{ccccccccccccc}
+  & \multicolumn{12}{c}{Stirling}\\
+rank   &  193& 240& 197& 257& 227& 234& 238& 238& 196& 242& 242& 318\\
+runners&  145& 157& 156& 191& 172& 174& 173& 165& 172& 199& 181& 229\\
 \end{tabular}

### 2024-09-03 inst/very_simplified_likelihood.Rmd
typos

@@
 
-```{r analuyzelmp,cache=TRUE}
-jjf <- function(x){f_vec_arn(x,O$rank,O$n,log=TRUE)}
-system.time(optolym <- optimize(jjf, c(0.45,0.5), maximum=TRUE))
+```{r analyze_athletics_olymp,cache=TRUE}
+jjf <- function(x){f_vec_arn(x,O$rank,O$n-1,log=TRUE)}
+system.time(optolym <- optimize(jjf, c(0.45,0.65), maximum=TRUE))
 ```
@@
 
-```{r makeolymplike}
-a <- seq(from=0.3,to=0.65,len=45)
-L <- f_vec_arn(a,O$rank,O$n,log=TRUE)
+```{r make_athletic_like}
+a <- seq(from=0.35,to=0.65,len=45)
+L <- f_vec_arn(a,O$rank,O$n-1,log=TRUE)
 Lmax <- jjf(optolym$maximum)
@@
 segments(x0=a_upper,y0=-1.5,y1=-3.5)
-
+grid()
 text(x=a_lower,y=-3,"0.380",pos=4)

### 2024-09-03 inst/hankin-AustrianJournalofStatistics.tex
fixes #240

@@
 Here a simplification of the Plackett--Luce likelihood functions for
-order statistics is proposed.  We consider a field of $n+1$ entities
+order statistics is proposed.  We consider a field of $n$ entities
 that are ordered in some way [the preferred example is competitors in
@@
 \Keywords{keywords, comma-separated, non-capitalized, \proglang{R}}
-\Plainkeywords{keywords, comma-separated, non-capitalized,, R} %% without formatting
+\Plainkeywords{keywords, comma-separated, non-capitalized, R} %% without formatting
 %% at least one keyword must be supplied
@@
 a race between a competitor of Bradley-Terry strength $a$ and
-$n\geqslant 1$ cloned competitors, each of strength $b=1-a$.  Under
-these circumstances, an observation is indexed by~$r$, the number of
-strength $b$ clones finishing ahead of~$a$, and because we have $n+1$
-competitors $0\leqslant r\leqslant n$.  The initial field strength is
-$a+nb=n-(n-1)a$.  The Plackett--Luce likelihood
-function~\citep{luce1959,plackett1975} for this situation would be
+$n-1\geqslant 1$ cloned competitors, each of strength $b=1-a$.  Under
+these circumstances, an observation is indexed by~$r$, the rank of the
+focal competitor, and because we have $n$ competitors $1\leqslant
+r\leqslant n$.  The initial field strength is $a+(n-1)b=1+(n-2)b$.  The
+Plackett--Luce likelihood function~\citep{luce1959,plackett1975} for
+this situation would be
 
@@
   \mathcal{L}_{r,n}(a) &=&
-\frac{b}{a+ n   b}\cdot
-\frac{b}{a+(n-1)b}\cdots\frac{a}{a+(n-r)b}\nonumber\\
+\frac{b}{1+ (n-2)b}\cdot
+\frac{b}{a+(n-3)b}\cdots\frac{a}{a+(n-r-1)b}\nonumber\\
 &=& \frac{B-1}{(B+n-1)(B+n-2)\cdots(B+n-r-1)}\nonumber\\ 
@@
 likelihoods, not probabilities, and do not sum to 1.  Previous related
-work \citep{hankin2024_hyper3} presented the case $n=2$ [in current
+work \citep{hankin2024_hyper3} presented the case $n=3$ [in current
   notation].  Considering the case $n=8$ gives us $8+1=9$ distinct
 likelihood functions, Figure~\ref{ninelikes}.  We see an unanticipated
-asymmetry between $r$ and $n-r$.  For $r=4$ one might expect that
+asymmetry between $r$ and $n-r$.  For $r=5$ one might expect that
 $\hat{a}=\frac{1}{2}$ on the grounds that the focal competitor's
@@
 
-Analysis of expression~\ref{likeforrn3} for fixed $n$ shows that the
-support $\mathcal{S}=\log\mathcal{L}_{r,n}(a)$ is asymptotically
+Analysis of expression~\ref{likeforrn3} for large $n$ and $r=pn$ for
+fixed $p$ shows that the support
+$\mathcal{S}=\log\mathcal{L}_{r,n}(a)$ is asymptotically
 
@@
 %\log\mathcal{L}_{r,n}(a)=
-\log(B-1)-B\frac{r+1}{n}
-+\mathcal{O}\left(n^{-2}\right)
+\log(B-1) + B\log(1-r/n)
++\mathcal{O}\left(n^{-1}\right)
 \end{equation}
@@
 (up to an arbitrary additive constant).  Using
-$\mathcal{S}^*=\log(B-1)-B\frac{r+1}{n}$ as an approximate
-support thus incurs an error of only $\mathcal{O}(n^{-2})$.  Observing
-that $\partial^2\mathcal{S}^*/\partial B^2=-(B-1)^{-2}<0$, the evaluate
-would be unique; we have
-$\hat{B}=\frac{n+r+1}{r+1}+\mathcal{O}\left(n^{-1}\right)$, or
-alternatively $\hat{a}=\frac{n}{n+r+1}$.  Compare this with the result
-of asking ``what is the probability that the focal competitor beats a
-randomly chosen opponent?'' to which the binomial maximum likelihood
-estimate is simply $\frac{n-r}{n}$, agreeing to within
-$\mathcal{O}(n^{-2})$.  For repeated observations $r_i,n_i$,
-$1\leqslant i\leqslant m$ we would have
+$\mathcal{S}^*=\log(B-1)+B\log(1-r/n)$ as an approximate support thus
+incurs an error of only $\mathcal{O}(n^{-1})$.  Observing that
+$\partial^2\mathcal{S}^*/\partial B^2=-(B-1)^{-2}<0$, the evaluate
+would be unique; we have $\hat{B}=1-\log(1-r/n)^{-1}$, or
+alternatively $\hat{a}=(1-\log(1-r/n))^{-1}$.  Compare this with the
+result of asking ``what is the probability that the focal competitor
+beats a randomly chosen opponent?'' to which the binomial maximum
+likelihood estimate is $\frac{n-r}{n}$.  For repeated independent
+observations $r_i,n_i$, $1\leqslant i\leqslant N$ we would have
+
 \begin{equation}
-  \hat{a} = 1-
-  \frac{\frac{r_1+1}{n_1}+\cdots+\frac{r_m+1}{n_m}}{m}
-  = 1-\overline{\left(\frac{r+1}{n}\right)}.
+  \hat{a} =   \frac{N}{N+\sum\log(1-r_i/n_i)};
 \end{equation}
 
-(that is, one minus the mean ratio of rank to number of [cloned]
-competitors); compare the binomial result of $1-\left.\sum
-r_i\right/\sum n_i$.
+compare the binomial result of $1-\left.\sum r_i\right/\sum n_i$.
+Below, I present three examples that showcase likelihood
+function~\ref{likeforrn3}, drawn from the fields of athletics,
+education, and Formula~1 motor racing.
 
@@
 \includegraphics[width=4in]{dotprobs}  % produced by very_simplified_likelihood.Rmd
-\caption{Maximum likelihood estimator $\hat{a}$ for $n=1(1)10$
-  and\label{dotprobs}} $r=0(1)n$
+\caption{Maximum likelihood estimator $\hat{a}$ for $n=2(1)10$
+  and\label{dotprobs}} $r=1(1)n$
 \end{centering}
 \end{figure}
-
-Below, I present three examples that showcase likelihood
-function~\ref{likeforrn3}, drawn from the fields of athletics,
-education, and Formula~1 motor racing.
 

### 2024-09-03 inst/very_simplified_likelihood.Rmd
fixes #240

@@
 Consider a race between a competitor of Bradley-Terry strength $a$ and
-$n$ competitors each of strength $b$; we require $a+b=1$.  NB we have
-$n+1$ competitors in total, 1 of strength $a$ and $n$ of strength $b$.
-An observation is indexed by $R$, the rank of the focal competitor: we
-have $1\leqslant R\leqslant n+1$.  Alternatively we might consider
-$r=R-1$, the number of $b$ clones finishing ahead of $a$, so
-$0\leqslant r\leqslant n$.  The initial field strength is
-$a+nb=1+b(n-1)$.
-
-$$r=R-1$$
-
-Suppose our bro comes first (zero $b$'s ahead of him) [$r=0, R=1$]:
-
-\[
-\mathcal{L}(a)=\frac{a}{a+nb}=\frac{1-b}{1+(n-1)b}
-\]
-
-Comes second [$r=1,R=2$], one $b$ ahead of him:
-
-\[
-\mathcal{L}(a)=\frac{b}{a+nb}\cdot\frac{a}{a+(n-1)b}
-\]
-
-Third [$r=2, R=3$]; two ahead of him:
-
-\[
-\mathcal{L}(a)=
+$n-1$ competitors each of strength $b$; we require $a+b=1$.  NB we
+have $n$ competitors in total, 1 of strength $a$ and $n-1$ of strength
+$b$.  An observation is indexed by $r$, the rank of the focal
+competitor: we have $1\leqslant r\leqslant n$.  The initial field
+strength is $a+(n-1)b=1+b(n-2)b$.
+
+Suppose our bro comes first (zero $b$'s ahead of him) [$r=1$]:
+
+\[
+\mathcal{L}(a)=\frac{a}{a+(n-1)b}=\frac{1-b}{1+(n-2)b}
+\]
+
+Comes second [$r=2$], one $b$ ahead of him:
+
+\[
+\mathcal{L}(a)=\frac{b}{a+(n-1)b}\cdot\frac{a}{a+(n-2)b}=
+\frac{b(1-b)}{(1+(n-2)b)\cdot(1+(n-3)b)}
+\]
+
+Third [$r=3$]; two ahead of him:
+
+\begin{eqnarray}
+\mathcal{L}(a) &=&
 \frac{b}{a+ n   b}\cdot
 \frac{b}{a+(n-1)b}\cdot
-\frac{a}{a+(n-2)b}
-\]
-
-Fourth [$r=3, R=4$], three ahead:
+\frac{a}{a+(n-2)b}\\
+&=&
+\frac{b^2(1-b)\rule{0mm}{10mm}
+}{
+(1+(n-2)b)\cdot(1+(n-3)b)\cdot(1+(n-4)b)}
+\end{eqnarray}
+
+Fourth [$r=4$], three clones finish before the focal competitor:
 
@@
 \mathcal{L}(a)=
-\frac{b}{a+ n   b}\cdot
-\frac{b}{a+(n-1)b}\cdot
-\frac{b}{a+(n-2)b}\cdot
-\frac{a}{a+(n-3)b}
-\]
-
-Placing after $r<n$ of the $n$ clones of strength $b$ finish:
-
-\[
-\mathcal{L}(a)=
-\frac{b^ra}{
-(a+nb)(a+(n-1)b)(a+(n-2)b)\cdots(a+(n-r)b)}
-=
-\frac{B-1}{(B+n-1)(B+n-2)\cdots(B+n-r-1)}
-\]
-
-where $B=1/b$ [also using the fact that $a+b=1$].
-
-This is
-
-\[
-\mathcal{L}_{r,n}(a)=\frac{(B-1)(B+n-r-2)!}{(B+n-1)!},\qquad B=1/(1-a)
-\]
-
-\[
-\mathcal{L}_{R,n}(a)=\frac{(B-1)(B+n-R-1)!}{(B+n-1)!},\qquad B=1/(1-a)
-\]
-
-\[
-=\frac{(B-1)\Gamma(B+n-R)}{\Gamma(B+n)},\qquad B=1/(1-a)
-\]
-
-Then, using Stirling [viz $n!\sim n^ne^{-n}\sqrt{2\pi n}$] we get
-
-\[
-\log(B-1) + ((B+n-r-2)\log(B+n-r-2)-(B+n-r-2)+\frac{\log(B+n-r-2)}{2})
-          - ((B+n  -1)\log(B+n  -1)-(B+n  -1)+\frac{\log(B+n  -1)}{2})
-\]
-
-\[
-\log(B-1) + ((B+n-R-1)\log(B+n-R-1)-(B+n-R-1)+\frac{\log(B+n-R-1)}{2})
-          - ((B+n  -1)\log(B+n  -1)-(B+n  -1)+\frac{\log(B+n  -1)}{2})
-\]
-
-If you give Sage this:
-
-```
-var('B n r')
-pretty_print_default(True)
-X=log(B-1) + (
-+ ((B+n-r-2)*log(B+n-r-2)-(B+n-r-2)+log(B+n-r-2)/2)
-- ((B+n  -1)*log(B+n  -1)-(B+n  -1)+log(B+n  -1)/2)
-)
-X.taylor(n,Infinity,2)
-```
-
-It returns this:
-
-\[
--(r+1)\log n
--\frac{(2B-3)r-r^2+2B-2}{2n}
--\frac{3(2B-3)r^2-6B^2-6(B^2-3B+2)r+12B-5}{12n^2}
-+\log(B-1)+\mathcal{O}(n^{-3})
-\]
-
-Now using $R$ not $r$:
-
-```
-sage: var('B n R')
-(B, n, R)
-pretty_print_default(False)
-X = log(B-1) + (
-  + ((B+n-R-1)*log(B+n-R-1)-(B+n-R-1)+log(B+n-R-1)/2)
-  - ((B+n  -1)*log(B+n  -1)-(B+n  -1)+log(B+n  -1)/2)
-  )
-X
-(B - R + n - 1)*log(B - R + n - 1) - (B + n - 1)*log(B + n - 1) + R + 1/2*log(B - R + n - 1) - 1/2*log(B + n - 1) + log(B - 1)
-X.taylor(n,Infinity,2).full_simplify()
--1/12*(12*R*n^2*log(n) + 3*(2*B - 1)*R^2 - 2*R^3 - 12*n^2*log(B - 1) - 6*(B^2 - B)*R + 6*((2*B - 1)*R - R^2)*n)/n^2
-```
-
-Note that if we use instead $n~\sim n^ne^{-n}\sqrt{2\pi
-n}\left(1+\frac{1}{12n}\right)$, then only terms of
-$\mathcal{O}(n^{-2})$ and above are changed; the first-order terms and
-indeed the zeroth order terms are unaffected.  Anyway, after
-simplifying and taking an asymptotic expansion to order $n^{-1}$, we
-find, to first order,
-
-\[
-\mathcal{S}=\log\mathcal{L}_{r,n}(a)=\log(B-1)-B\frac{r+1}{n}
-+ K + \mathcal{O}\left(n^{-2}\right)
-\]
-
-(where $K$ is an arbitrary constant).  Observing that
-$\partial^2\mathcal{S}/\partial B^2=-(B-1)^{-2}<0$, the evaluate would
-be unique; we have
-$\hat{B}=\frac{n+r+1}{r+1}+\mathcal{O}\left(n^{-1}\right)$, or
-alternatively $\hat{a}=\frac{n}{n+r+1}$.  I present some numerical
-results below.
+\frac{b^3(1-b)
+}{
+(1+(n-2)b)\cdot(1+(n-3)b)\cdot(1+(n-4)b)\cdot(1+(n-5)b)}
+\]
+
+Placing with general rank $r$, $1\leqslant r\leqslant n$.  Rank $r$
+means $r-1$ clones finish ahead of the focal competitor:
+
+\begin{eqnarray}
+\mathcal{L}(a) &=&
+\frac{b^{r-1}(1-b)
+}{
+\underbrace{(1+(n-2)b)\cdot(1+(n-3)b)\cdots(1+(n-r-1)b)}_{\mbox{$r$ terms}}
+}
+\\&=&
+\frac{B-1\rule{1mm}{10mm}}{(n+B-2)\cdot(n+B-3)\cdots(n+B-r-1)}
+\end{eqnarray}
+
+where $B=1/b$.  Recall that $r$ is the rank and $r-1$ clones finish
+ahead of the FC.  This is
+
+\[
+\mathcal{L}_{r,n}(a)=\frac{(B-1)(n+B-r-2)!}{(n+B-2)!},\qquad B=1/(1-a)
+\]
+
+\[
+=\frac{(B-1)\Gamma(n+B-r-1)}{\Gamma(n+B-1)},\qquad B=1/(1-a)
+\]
 
@@
 ```
-f[B_, n_, R_] := Log[(B - 1)*Gamma[B + n - R]/Gamma[B + n]] 
+f[B_, n_, r_] := 
+ Log[(B - 1)] + Log[Gamma[n + B - r - 1]/Gamma[n + B - 1]]
 Assuming[B > 1 && p > 0 && p < 1, 
@@
 
-We get
-
-$$
-\log\left(
-(B-1)(1-p)^{B-1/2} +
-\frac{
-	(B-1)(1-6B+6B^2)(1-p)^{B-1/2}p
-}{
-	12(1-p)n
-}
-\right)
-$$
-
-or
-
-$$
-\log\left(
-(B-1)(1-p)^{B-1/2} \left[1+
-\frac{(1-6B+6B^2)p}{12(1-p)n}\right]\right)
-$$
-
-$$
-\log
-(B-1) + B\log(1-p) +
-n^{-1}
-\frac{(1-6B+6B^2)p}{12(1-p)} +{\mathcal O}\left(n^{-2}\right)
-$$
-
-$$
-\log
-(B-1) + B\log(1-p) +
-\frac{1}{12}n^{-1}\frac{p}{1-p}(1-6B+6B^2) + {\mathcal O}\left(n^{-2}\right)
-$$
-
-# Placing last
-
-Placing last is finishing after $n=r$ clones finish, that is, $R=n+1$
-[me plus two clones, I come third].
-
-\[
-n=1,R=2\longrightarrow\mathcal{L}=b\]
-
-\[
-n=2,R=3\longrightarrow\mathcal{L}=\frac{b}{a+2b}\cdot\frac{b}{a+b}=\frac{b^2}{1+b}
-=\frac{1}{B(B+1)}
-\]
-
-\[
-n=3,R=4\longrightarrow\mathcal{L}=\frac{b}{a+3b}\cdot\frac{b}{a+2b}\cdot\frac{b}{a+b}=\frac{b^3}{(1+2b)(1+b)}=\frac{1}{B(B+1)(B+2)}
-\]
-
-\[
-n=4,R=n+1\longrightarrow\mathcal{L}=\frac{b}{a+4b}\cdot\frac{b}{a+3b}\cdot\frac{b}{a+2b}\cdot\frac{b}{a+b}=
-\frac{b^4}{(1+3b)(1+2b)(1+b)}=
-\frac{1}{B(B+1)(B+2)(B+3)}
-\]
-
-For $n=4$ we have equivalently
-
-\[
-\frac{1}{(B+3)(B+2)(B+1)B}=\frac{(B-1)!}{(B+3)!}=\frac{\Gamma(B)}{\Gamma(B+2)}
-\]
-
-In general we would have
-
-\[
-\frac{1}{(B+n-1)(B+n-2)\cdots(B+1)B}=
-\frac{(B-1)!}{(B+n-1)!}=
-\frac{\Gamma(B)}{\Gamma(B+n)}
-\]
-
-(denominator has $n+1$ terms).  Observe that this agrees with the
-$r<n$ case above (substituting $r=n$).  We may introduce some R idiom:
-
-```{r,label=definefsingle}
-f_single <- function(a,R,n,log=FALSE){  # not vectorised
-    stopifnot(R>0)
+gives
+
+\[
+\mathcal{S}=\log\mathcal{L}_{r,n}(a)=\log(B-1) + (B-3/2)\log(1-r/n) + 
+\mathcal{O}\left(n^{-1}\right)
+\]
+
+(where $K$ is an arbitrary constant).  Here, $\mathcal{S}$ is the
+support (log-likelihood).  Observing that
+$\partial^2\mathcal{S}/\partial B^2=-(B-1)^{-2}<0$, the evaluate would
+be unique.
+
+# R idiom
+
+```{r, label=definefsingle}
+f_single <- function(a, r, n) {  # not vectorised
+    stopifnot(r>0)
 B <- 1/(1-a)
-	if(log){
-		out <- log(B-1) + lgamma(B+n-R) - lgamma(B+n)
-	} else {  # not-log
-		out <- (B-1)/prod(B+(n-R):(n-1))
-	}
+  log(B-1) + lgamma(n+B-r-1) - lgamma(n+B-1)
+}
+
+f_vec_a <- function(a,r,n){  # vectorised in 'a' but not in 'R'
+	sapply(a,function(a){f_single(a,r=r,n=n)})
+}
+
+f_vec_arn <- function(a,r,n){
+	## vectorized in 'a'; treats 'r' and 'n' as
+	## vectors of independent observations
+
+	M <- cbind(r,n)
+
+    out <- 0
+    for(i in seq_len(nrow(M))){out <- out + f_vec_a(a,r=M[i,1],n=M[i,2])}
 	return(out)
@@
 
-f_vec_a <- function(a,R,n, log=FALSE){  # vectorised in 'a' but not in 'R'
-	sapply(a,function(a){f_single(a,R=R,n=n,log=log)})
-}
-
-f_vec_arn <- function(a,R,n,log=FALSE){
-	## vectorized in 'a'; treats 'R' and 'n' as
-	## vectors of independent observations
-
-	M <- cbind(R,n)
-
-	if(log){
-		out <- 0
-		for(i in seq_len(nrow(M))){out <- out + f_vec_a(a,R=M[i,1],n=M[i,2],log=TRUE)}
-	} else {
-		out <- 1
-		for(i in seq_len(nrow(M))){out <- out * f_vec_a(a,R=M[i,1],n=M[i,2],log=FALSE)}
-	}
-	return(out)
-}
-
-fapprox <- function(a,R,n,log=FALSE){
-	if(log){
-		return(log(B-1)-B*R/n)
-	} else {
-		return((B-1)/exp(R*B/n))
-	}
+fapprox <- function(a,r,n){
+    stop("not written")
 }
@@
 ```{r label=defdiff}
-fdash <- function(a,R,n,log=TRUE){
+fdash <- function(a,r,n){  # dS/da = dS/dB*dB/da
 	B <- 1/(1-a)
-	out <- (1/(B-1) - psigamma(B+n) + psigamma(B+n-R))*B^2
-	if(!log){out <- out * f_vec_arn(a,R,n,log=FALSE)}
-	return(out)
+    (1/(B-1) + psigamma(n+B-r-1) - psigamma(n+B-1))*B^2
 }
@@
 d <- 1e-3
-R <- 4
+r <- 4
 n <- 10
 c(
-  numerical  = (f_single(a+d/2,R,n,log=TRUE)-f_single(a-d/2,R,n,log=TRUE))/d,
-  analytical = fdash(a,R,n)
+  numerical  = (f_single(a+d/2, r, n) - f_single(a-d/2, r, n))/d,
+  analytical = fdash(a,r,n)
 )
 
-c(
-  numerical  = (f_single(a+d/2,R,n,log=FALSE)-f_single(a-d/2,R,n,log=FALSE))/d,
-  analytical = fdash(a,R,n,log=FALSE)
-)
 ```
@@
 ```{r,label=defmle}
-MLE <- function(R,n,give=FALSE){
+MLE <- function(r,n,give=FALSE){
 	d <- 1e-6
-	out <- optimize(f_vec_arn,c(d,1-d),R=R,n=n,log=TRUE,maximum=TRUE)
+	out <- optimize(f_vec_arn, c(d, 1-d), r=r, n=n, maximum=TRUE)
 	if(!give){out <- out$maximum}
@@
 
-MLEa <- function(R,n,give=FALSE,small=1e-4){
-	out <- uniroot(f=function(a){fdash(a,R,n)},interval=c(small,1-small))
+MLEa <- function(r,n,give=FALSE,small=1e-4){
+	out <- uniroot(f=function(a){fdash(a,r,n)},interval=c(small,1-small))
 	if(!give){out <- out$root}
@@
 
-```{r usestirling}
-rbind(
-showdiff(MLE(10, 100), 100 / 111),
-showdiff(MLE(10, 500), 500 / 511),
-showdiff(MLE(10,1000),1000 /1011),
-showdiff(MLE(10,5000),5000 /5011)
-)
-```
-
-```{r usestirlinga}
-rbind(
-showdiff(MLEa(10, 100), 100 / 111),
-showdiff(MLEa(10, 500), 500 / 511),
-showdiff(MLEa(10,1000),1000 /1011),
-showdiff(MLEa(10,5000),5000 /5011)
-)
-
-```
-
 Now verification
@@
 B <- 1/b
-n <- 9
-R <- 4
+n <- 9  # 9 runners
+r <- 4  # FC came 4th
 c(
@@
 way2 = (B-1)/prod(B+(5:8)),
-way3 = f_vec_arn(a,R,n,log=FALSE)
+way3 = f_vec_arn(a,r,n)
 )
@@
   out[[n]]  <- rep(NA,n+1)
-  for(R in seq_len(n+1)){
-    if(R==1){
+  for(r in seq_len(n+1)){
+    if(r==1){
       jj <- 1
-    } else if(R == n+1){
+    } else if(r == n+1){
       jj <- 0
     } else {
-      jj <- MLE(R,n)
-    }
-    out[[n]][R] <- jj
+      jj <- MLE(r,n)
+    }
+    out[[n]][r] <- jj
   }
@@
 plotterp <- function(...){
-plot(NA,xlim=c(1,10),ylim=c(0,1),type="n",xlab="n",ylab=expression(hat(a)))
-for(n in 1:10){
-  for(R in seq_len(n+1)){   points(n,out[[n]][R],pch=16) }
-  for(R in seq_len(n+1)){   text(n+0.2,out[[n]][R],R,cex=0.5,col='gray') }
+plot(NA,xlim=c(2,10),ylim=c(0,1),type="n",xlab="n",ylab=expression(hat(a)))
+for(n in 2:10){
+  for(r in seq_len(n)){ points(n,out[[n]][r],pch=16,col='black') }
+  for(r in seq_len(n)){ text(n+0.2,out[[n]][r],r,cex=0.5,col='gray') }
 }
@@
 The likelihood calculations above can be confusing becuase likelihood
-is not the same as probability.
-
-Consider, for example, $n=2$ so we have 3 competitors.
-
-probability of focal competitor coming first [i.e. $r=0$]:
+is not the same as probability.  Consider, for example, $n=3$ so we
+have 3 competitors: the focal competitor of strength $a$ and two
+clones each of strength $b=1-a$.
+
+probability of focal competitor coming first [i.e. $r=1$]:
 
@@
 
-Now what is the probability of coming second, that is, $r=1$?
+Now what is the probability of coming second, that is, $r=2$?
 
@@
 \mbox{prob of coming second}   =
-\frac{b}{a+2b}\cdot\frac{a}{a+b}=\frac{b(1-b)}{1+b}\qquad\mbox{WRONG}
-\]
-
-but this is incorrect: there are two clones of (strength) $b$, and the
+\frac{b}{a+2b}\cdot\frac{a}{a+b}=\frac{b(1-b)}{1+b}\qquad{\mbox{WARNING}}
+\]
+
+but this is *incorrect*: there are two clones of strength $b$, and the
 likelihood arguments above do not distinguish between their finishing
 order.  We can name the competitors $a$, $b_1$ and $b_2$ and specify
-that $b_1$ and $b_2$ both have strength $b$.  Then "coming second"
-means that the finishing order was $b_1\succ a\succ b_2$ or $b_2\succ
-a\succ b_1$.  Noting that these events are disjoint, the probability
-would be
+that $b_1$ and $b_2$ both have strength $b$.  Then "the focal
+competitor came second" means that the finishing order was $b_1\succ
+a\succ b_2$ or $b_2\succ a\succ b_1$.  Noting that these events are
+disjoint, the probability would be
 
@@
 
-Similarly for the focal competitor coming last we need to sum the
-probabilities of finishing orders $b_1\succ b_2\succ a$ and $b_2\succ
-b_1\succ a$:
+[see the factor of two difference].  Similarly for the focal
+competitor coming last we need to sum the probabilities of finishing
+orders $b_1\succ b_2\succ a$ and $b_2\succ b_1\succ a$:
 
@@
 
-Now try $r=3$ [i.e. 4 competitors].  Taking things one step at a time:
+Now try $n=4$ [i.e. 4 competitors: the FC and three clones].  Taking
+things one step at a time:
 
@@
 
-Now consider the general case.  Again, $n$ clones of strength $b$ and
-one of strength $a=1-b$, $n+1$ competitors.
+Now consider the general case.  This time we consider $n$ clones of
+strength $b$ and one of strength $a=1-b$, $n+1$ competitors.
 
@@
 
-Now consider the case where there are $r$ clones coming ahead of the
+Now consider the case where there are $p$ clones coming ahead of the
 focal competitor:
@@
 \mbox{prob of having $r$ clones ahead}
-=\frac{n!}{(n-r)!}\cdot X
+=\frac{n!}{(n-p)!}\cdot X
 \]
@@
 a <- seq(from=0,to=1,by=0.01)
-n <- 8
-jj <- f_vec_arn(a,3,n,log=FALSE)
+n <- 9 # the FC and 8 clones
+jj <- exp(f_vec_arn(a,3,n))
 plot(a,jj/max(jj,na.rm=TRUE),type='n',xlab=expression(a),ylab="likelihood")
 grid()
-rain <- rainbow(n+1)
-for(R in seq_len(n+1)){
-  y <- f_vec_a(a,R,n,log=FALSE)
-  if(R==1){y[length(y)] <- 1}
+rain <- rainbow(n)
+for(r in seq_len(n)){
+  y <- exp(f_vec_a(a,r,n))
+  if(r==1){y[length(y)] <- 1}
   y <- y/max(y,na.rm=TRUE)
-  if(R>1){y[length(y)] <- 0}
-  points(a,y,type='l',lwd=4,col=rain[R])
-}
-abline(v=MLE(R=5,n=8))
-text(0.07,0.95,"R=9",col=rain[9])
-text(0.20,0.95,"R=8",col=rain[8])
-text(0.26,0.91,"R=7",col=rain[7])
-text(0.29,0.83,"R=6",col=rain[6])
-text(0.33,0.76,"R=5",col=rain[5])
-text(0.35,0.64,"R=4",col=rain[4])
-text(0.39,0.53,"R=3",col=rain[3])
-text(0.45,0.41,"R=2",col=rain[2])
-text(0.54,0.21,"R=1",col=rain[1])
+  if(r>1){y[length(y)] <- 0}
+  points(a,y,type='l',lwd=4,col=rain[r])
+}
+abline(v=MLE(r=5,n=8))
+text(0.09,0.95,"r=9",col=rain[9])
+text(0.18,0.93,"r=8",col=rain[8])
+text(0.25,0.90,"r=7",col=rain[7])
+text(0.29,0.83,"r=6",col=rain[6])
+text(0.33,0.76,"r=5",col=rain[5])
+text(0.35,0.65,"r=4",col=rain[4])
+text(0.39,0.54,"r=3",col=rain[3])
+text(0.45,0.41,"r=2",col=rain[2])
+text(0.59,0.21,"r=1",col=rain[1])
 }
@@
 ```{r shown4n8mle}
-MLE(R=5,n=8)
+MLE(r=5,n=8)
 ```
@@
 d_gasly_max <- MLE(d_gasly,22)
-f_vec_arn(bothmax,d_ocon ,22,log=TRUE)
-f_vec_arn(bothmax,d_gasly,22,log=TRUE)
-f_vec_arn(d_ocon_max ,d_ocon ,22,log=TRUE)
-f_vec_arn(d_gasly_max,d_gasly,22,log=TRUE)
+f_vec_arn(bothmax,d_ocon ,22)
+f_vec_arn(bothmax,d_gasly,22)
+f_vec_arn(d_ocon_max ,d_ocon ,22)
+f_vec_arn(d_gasly_max,d_gasly,22)
 
 Lambda <- (
-   (f_vec_arn(d_gasly_max  ,d_ocon,22,log=TRUE)+f_vec_arn(d_gasly_max  ,d_gasly,22,log=TRUE)) -
-   (f_vec_arn(bothmax      ,d_ocon,22,log=TRUE)+f_vec_arn(bothmax      ,d_gasly,22,log=TRUE))
+   (f_vec_arn(d_gasly_max  ,d_ocon,22)+f_vec_arn(d_gasly_max  ,d_gasly,22)) -
+   (f_vec_arn(bothmax      ,d_ocon,22)+f_vec_arn(bothmax      ,d_gasly,22))
 )
@@
   out <- a*0
-  for(year in 2011:2023){ out <- out + f_vec_arn(a,getfoc(year),getnum(year)-1,log=TRUE) }
+  for(year in 2011:2023){ out <- out + f_vec_arn(a,getfoc(year),getnum(year)) }
   return(out)
@@
      strength <- exp(LO)/(1+exp(LO))
-     out <- out + f_vec_arn(strength,getfoc(year),getnum(year)-1,log=TRUE)
+     out <- out + f_vec_arn(strength,getfoc(year),getnum(year))
   }
@@
   Lambda <- (
-  (f_vec_arn(d2max  ,d1,n,log=TRUE)+f_vec_arn(d2max  ,d2,n,log=TRUE)) -
-  (f_vec_arn(bothmax,d1,n,log=TRUE)+f_vec_arn(bothmax,d2,n,log=TRUE))
+  (f_vec_arn(d2max  ,d1,n) + f_vec_arn(d2max  ,d2,n)) -
+  (f_vec_arn(bothmax,d1,n) + f_vec_arn(bothmax,d2,n))
 )
@@
 ```{r analyze_athletics_olymp,cache=TRUE}
-jjf <- function(x){f_vec_arn(x,O$rank,O$n-1,log=TRUE)}
+jjf <- function(x){f_vec_arn(x,O$rank,O$n)}
 system.time(optolym <- optimize(jjf, c(0.45,0.65), maximum=TRUE))
@@
 a <- seq(from=0.35,to=0.65,len=45)
-L <- f_vec_arn(a,O$rank,O$n-1,log=TRUE)
+L <- f_vec_arn(a,O$rank,O$n)
 Lmax <- jjf(optolym$maximum)
@@
 
-```{r calculatelikes1,cache=F}
+```{r calculatelikes1, cache=F}
 a <- seq(from=0.005,to=0.85,by=0.01)
-La <- f_vec_arn(a,R=results$place[1],results$runners[1]-1,log=TRUE)
+La <- f_vec_arn(a,r=results$place[1],results$runners[1])
 La <- La - max(La,na.rm=TRUE)
@@
   for(i in 1:3){
-    out <- out + f_vec_a(a_stirling,R=place[i],runners[i]-1,log=TRUE)
+    out <- out + f_vec_a(a_stirling,r=place[i],runners[i])
   }
   for(i in 4:10){
-    out <- out + f_vec_a(a_auckland,R=place[i],runners[i]-1,log=TRUE)
+    out <- out + f_vec_a(a_auckland,r=place[i],runners[i])
   }
@@
      strength <- exp(LO)/(1+exp(LO))
-     out <- out + f_vec_arn(strength,results$place[i]-1,results$runners[i]-1,log=TRUE)
+     out <- out + f_vec_arn(strength,results$place[i],results$runners[i])
   }
@@
 ```{r dopurefurst}
-Lpure <- f_vec_arn(a,rank[wp]-1,class_size[wp]-1,log=TRUE)
+Lpure <- f_vec_arn(a,rank[wp],class_size[wp])
 Lpure <- Lpure - max(Lpure,na.rm=TRUE)
 plot(a,Lpure,ylab='pure strength')
-optimize(function(a){f_vec_arn(a,rank[wp],class_size[wp]-1,log=TRUE)},c(0.3,0.6),maximum=TRUE)
-optimize(function(a){f_vec_arn(a,rank[wa],class_size[wa]-1,log=TRUE)},c(0.3,0.6),maximum=TRUE)
+optimize(function(a){f_vec_arn(a,rank[wp],class_size[wp])},c(0.3,0.6),maximum=TRUE)
+optimize(function(a){f_vec_arn(a,rank[wa],class_size[wa])},c(0.3,0.6),maximum=TRUE)
 ```
@@
   for(i in seq_len(nrow(M1))){
-    out <- out + f_vec_arn(vec[1],R=M1[i,1],n=M1[i,2]-1,log=TRUE)
+    out <- out + f_vec_arn(vec[1],r=M1[i,1],n=M1[i,2])
   }
   for(i in seq_len(nrow(M2))){
-    out <- out + f_vec_arn(vec[2],R=M2[i,1],n=M2[i,2]-1,log=TRUE)
+    out <- out + f_vec_arn(vec[2],r=M2[i,1],n=M2[i,2])
   }

### 2024-09-17 inst/very_simplified_likelihood.Rmd
bugfix

@@
     p <- exp(LO)/(1+exp(LO))
-    S <- S + f_vec_arn(p,R=M[i,2],n=M[i,3]-1,log=TRUE)
+    S <- S + f_vec_arn(p,r=M[i,2],n=M[i,3]-1)
   }
@@
 grid()
-persp(a,b,Z)
+#persp(a,b,Z)
 ```
@@
 segments(x0=ahat,y0=-1.5,y1=0)
-text(ahat,-0.91,expression(hat(a)==0.573),pos=2)
+text(ahat,-0.91,expression(hat(a)==0.511),pos=2)
 abline(v=0.5,lty=3)
@@
 grid()
-text(x=a_lower,y=-3,"0.380",pos=4)
-text(x=a_upper,y=-3,"0.563",pos=2)
+text(x=a_lower,y=-3,"0.414",pos=4)
+text(x=a_upper,y=-3,"0.602",pos=2)
 }
@@
 ```{r calculate_parksupp,cache=TRUE}
-S_parkrun <- function(a){f_vec_arn(a,R=results$place,results$runners-1,log=TRUE)}
+S_parkrun <- function(a){f_vec_arn(a, r=results$place, results$runners)}
 (optparkrun <- optimize(S_parkrun, c(0.2,0.8),maximum=TRUE))
-
 ```
@@
   out <- 0
-  for(i in 1:3){
+  for(i in 1:12){
     out <- out + f_vec_a(a_stirling,r=place[i],runners[i])
   }
-  for(i in 4:10){
+  for(i in 13:21){
     out <- out + f_vec_a(a_auckland,r=place[i],runners[i])
@@
     par(pty='s')
-    contour(x,x,L,asp=1,xlim=range(x),ylim=range(x),levels=-(0:5),xlab="Auckland",ylab="Stirling")
+    contour(x,x,L,asp=1,xlim=range(x),ylim=range(x),levels=-(0:5),xlab="2023",ylab="2024")
     abline(0,1)
@@
 "differential equations","topology","special relativity","fluid mechanics","Lie algebra")
-category=c("pure","pure","applied","applied","applied","pure","applied","applied","pure")
+category <- c("pure","pure","applied","applied","applied","pure","applied","applied","pure")
 data.frame(course,category,rank,class_size)
@@
 wa <- category=='applied'
+```
+
+```{r dofishereducation}
+pure_betterthanme    <- sum(rank[wp]-1)
+pure_worsethanme     <- sum(class_size[wp] - rank[wp])
+applied_betterthanme <- sum(rank[wa]-1)
+applied_worsethanme  <-  sum(class_size[wa] - rank[wa])
+M <- matrix(c(pure_betterthanme,pure_worsethanme,applied_betterthanme,applied_worsethanme),2,2,byrow=TRUE)
+dimnames(M) <- list(type=c("pure","applied"),beatsme=c(T,F))
+M
+fisher.test(M)
 ```

### 2024-09-17 inst/hankin-AustrianJournalofStatistics.tex
corrections

@@
 beats a randomly chosen opponent?'' to which the binomial maximum
-likelihood estimate is $\frac{n-r}{n}$.  For repeated independent
+likelihood estimate is $\frac{r-1}{n-r}$.  For repeated independent
 observations $r_i,n_i$, $1\leqslant i\leqslant N$ we would have
@@
 \begin{equation}
-  \hat{a} =   \frac{N}{N+\sum\log(1-r_i/n_i)};
+  \hat{a} =   \frac{1}{1-\sum\log(1-r_i/n_i)/N};
 \end{equation}
 
-compare the binomial result of $1-\left.\sum r_i\right/\sum n_i$.
-Below, I present three examples that showcase likelihood
+compare the binomial result of $\sum(r_i-1)/\sum n_i$.  Below, I
+present three examples that showcase likelihood
 function~\ref{likeforrn3}, drawn from the fields of athletics,
@@
 
-\subsection{Olympic track and field athletics}
-
-One long-standing contention in the athletic world is that of lane
-bias~\citep{munro2022}.  Running in the inner lane is frequently held
-to be a disadvantage; mechanisms include biomechanical issues arising
-from centrifugal acceleration, to competitors' inability to see their
-rivals.  Here attention is restricted to heats, for which lane placing
-is random.  I consider every heat in Men's 100m Olympic running from
-2004--2020 (a total of 41 heats, each of 8 or 9 competitors); only
-competitors' ranks are used.  Figure~\ref{plotolymp} shows a
-log-likelihood curve for the Bradley--Terry strength of the focal
-competitor, here the runner in the innermost lane.  We see that the
-evaluate of $0.474$ is very close to the neutral value of $0.5$, for
-which the support is only $0.16$, far short of the standard two units
-of support criterion for rejection~\citep{edwards1972}.  This dataset
-contains no evidence to support lane bias.
-
-\begin{figure}[t]
-  \begin{centering}
-\includegraphics[width=4in]{plotolymp} % produced by very_simplified_likelihood.Rmd
-\caption{Log-likelihood curve for the Bradley--Terry strength of the
-  lane-1 runner, Olympic qualifying heats 2012-2020 \label{plotolymp}}
-\end{centering}
-\end{figure}
+% \subsection{Olympic track and field athletics}
+% 
+% One long-standing contention in the athletic world is that of lane
+% bias~\citep{munro2022}.  Running in the inner lane is frequently held
+% to be a disadvantage; mechanisms include biomechanical issues arising
+% from centrifugal acceleration, to competitors' inability to see their
+% rivals.  Here attention is restricted to heats, for which lane placing
+% is random.  I consider every heat in Men's 100m Olympic running from
+% 2004--2020 (a total of 41 heats, each of 8 or 9 competitors); only
+% competitors' ranks are used.  Figure~\ref{plotolymp} shows a
+% log-likelihood curve for the Bradley--Terry strength of the focal
+% competitor, here the runner in the innermost lane.  We see that the
+% evaluate of $0.511$ is very close to the neutral value of $0.5$, for
+% which the support is only $0.029$, far short of the standard two units
+% of support criterion for rejection~\citep{edwards1972}.  This dataset
+% contains no evidence to support lane bias.
+% 
+% \begin{figure}[t]
+%   \begin{centering}
+% \includegraphics[width=4in]{plotolymp} % produced by very_simplified_likelihood.Rmd
+% \caption{Log-likelihood curve for the Bradley--Terry strength of the
+%   lane-1 runner, Olympic qualifying heats 2012-2020 \label{plotolymp}}
+% \end{centering}
+% \end{figure}
 
@@
 shows the author has completed a total of 21 parkruns to date,
-comprising 9 runs in Auckland and 12 at Stirling.  From the first pair
-of numbers we see that 259 runners attended that particular parkrun at
-Auckland, of whom the author placed 177.
+comprising 9 runs in 2023 and 12 in 2024.  From the first pair of
+numbers we see that 259 runners attended that particular parkrun in
+2023, of whom the author placed 177.
 
@@
 \begin{tabular}{cccccccccc}\\
- & \multicolumn{9}{c}{Auckland}\\
+ & \multicolumn{9}{c}{2023}\\
 rank   & 177& 222& 206& 142& 118& 224& 128& 115& 183\\
@@
 \begin{tabular}{ccccccccccccc}
-  & \multicolumn{12}{c}{Stirling}\\
+  & \multicolumn{12}{c}{2024}\\
 rank   &  193& 240& 197& 257& 227& 234& 238& 238& 196& 242& 242& 318\\
@@
 Is there evidence that the author's performance differs between
-Auckland and Stirling?  Figure~\ref{parkruncontour} shows contours of
+years?  Figure~\ref{parkruncontour} shows contours of
 support [again normalised so the support of the evaluate is zero] for
-the parkrun dataset, as a function of Auckland strength on the
-horizontal axis and Stirling strength on the vertical axis.  A null of
+the parkrun dataset, as a function of 2023 strength on the
+horizontal axis and 2024 strength on the vertical axis.  A null of
 equal strengths appears as the diagonal line.  We see that the maximum
-likelihood estimate is $\hat{p}_\mathrm{Auck}=0.427$,
-$\hat{p}_\mathrm{Stir}=0.408$.  However, a constrained optimization
+likelihood estimate is $\hat{p}_\mathrm{2023}=0.427$,
+$\hat{p}_\mathrm{2024}=0.408$.  However, a constrained optimization
 shows that the maximum support along the null diagonal is about
-$-2.43$, at ${p}_\mathrm{Auck}={p}_\mathrm{Stir}\simeq 0.409$.  The
+$-2.43$, at ${p}_\mathrm{2023}={p}_\mathrm{2024}\simeq 0.409$.  The
 two units of support criterion~\citep{edwards1972} is met, and we are
@@
 
+Now consider table~\ref{fisherparkrun}, giving a summary of whether
+nonfocal ocmpetitors place ahead of the author, split by venue.  A
+two-sided Fisher test gives a $p$-value of about $5\times 10^{-14}$,
+surely an unreasonable conclusion given the author's perceived
+performance stagnation.
+
 \begin{figure}[t]
@@
 \caption{Support contours for Bradley-Terry strength
-  at \label{parkruncontour} Auckland and Stirling.  Null of equal
-  strengths shown as a diagonal line}
+  for \label{parkruncontour} 2023 and 2024.  Null of equal strengths
+  shown as a diagonal line}
 \end{centering}
 \end{figure}
+
+\begin{table}[h]
+    \begin{tabular}{ l c r }
+       & \multicolumn{2}{c}{Faster?}\\
+       & yes  & no  \\
+  2023 & 1506 & 806 \\
+  2024 & 2102 & 708
+    \end{tabular}
+    \caption{Two-way contingency\label{fisherparkrun} table}
+\end{table}
 
@@
                topology  &   pure    & 8  &      14\\
-     special relativity  & applied   & 5  &      13\\
+     special relativity  & applied   & 3  &      13\\
         fluid mechanics  & applied   & 4  &      12\\
@@
 the diagonal line.  We see that the maximum likelihood estimate is
-$\hat{p}_\mathrm{pure}=0.567$, $\hat{p}_\mathrm{applied}=0.887$.
+$\hat{p}_\mathrm{pure}=0.542$, $\hat{p}_\mathrm{applied}=0.852$.
 However, a constrained optimization shows that the maximum support
-along the null diagonal is about $-2.43$, at
-${p}_\mathrm{pure}={p}_\mathrm{applied}\simeq 0.713$.  The two units of
+along the null diagonal is about $-2.06$, at
+${p}_\mathrm{pure}={p}_\mathrm{applied}\simeq 0.683$.  The two units of
 support criterion~\citep{edwards1972} is met, and we are justified in
@@
 pure mathematics.  Alternatively, we may observe that
-$2\mathcal{S}=4.86$ is in the tail region of its asymptotic $\chi^2_1$
-distribution, corresponding to a $p$-value of about $0.028$.
+$2\mathcal{S}=4.11$ is in the tail region of its asymptotic $\chi^2_1$
+distribution, corresponding to a $p$-value of about $0.042$.
 

### 2024-09-17 inst/very_simplified_likelihood.Rmd
corrections

@@
 ```{r focaleducation}
-rank <- c( 9, 7, 2, 3,2 ,8 ,5 ,4 ,9 )
-class_size <- c(12, 17,23, 9,13,14,13,12,15)
+rank       <- c( 9,  7,  2, 3,  2,  8,  3,  4,  9)
+class_size <- c(12, 17, 23, 9, 13, 14, 13, 12, 15)
 course <- c("rings and modules","group theory","calculus","linear algebra",
@@
 plot(a,Lpure,ylab='pure strength')
-optimize(function(a){f_vec_arn(a,rank[wp],class_size[wp])},c(0.3,0.6),maximum=TRUE)
-optimize(function(a){f_vec_arn(a,rank[wa],class_size[wa])},c(0.3,0.6),maximum=TRUE)
+optimize(function(a){f_vec_arn(a,rank[wp],class_size[wp])},c(0.3,0.9),maximum=TRUE)
+optimize(function(a){f_vec_arn(a,rank[wa],class_size[wa])},c(0.3,0.9),maximum=TRUE)
 ```
@@
 jj <- as.matrix(expand.grid(x,x))
-L <- apply(jj,1,supp2_ed,rank[wp]-1,rank[wa]-1,class_size[wp]-1,class_size[wa]-1)
+L <- apply(jj,1,supp2_ed,rank[wp],rank[wa],class_size[wp],class_size[wa])
 L <- L - max(L,na.rm=TRUE)
@@
    function(vec){
-      supp2_ed(vec,rank[wp],rank[wa],class_size[wp]-1,class_size[wa]-1)},
+      supp2_ed(vec,rank[wp],rank[wa],class_size[wp],class_size[wa])},
    control=list(fnscale = -1)
@@
 jj2 <-
-optimize(function(v){supp2_ed(c(v,v),rank[wp]-1,rank[wa]-1,class_size[wp]-1,class_size[wa]-1)},
+optimize(function(v){supp2_ed(c(v,v),rank[wp],rank[wa],class_size[wp],class_size[wa])},
 c(0.1,0.9),maximum=TRUE)
@@
 ```{r,label=shochisw}
+jj1$value
+jj2$objective
+jj1$value-jj2$objective
 pchisq(2*(jj1$value-jj2$objective),lower.tail=FALSE,df=1)

### 2024-09-18 inst/hankin-AustrianJournalofStatistics.tex
typos

@@
 work \citep{hankin2024_hyper3} presented the case $n=3$ [in current
-  notation].  Considering the case $n=8$ gives us $8+1=9$ distinct
-likelihood functions, Figure~\ref{ninelikes}.  We see an unanticipated
-asymmetry between $r$ and $n-r$.  For $r=5$ one might expect that
-$\hat{a}=\frac{1}{2}$ on the grounds that the focal competitor's
-performance was median.  However, $\hat{a}\simeq 0.5886$.
-Figure~\ref{dotprobs} shows the maximum likelihood estimator $\hat{a}$
-for various values of $n,r$, obtained by direct numerical maximization
-of equation~\ref{likeforrn3}.
+  notation] but here arbitrary $n$ may be considered.  Taking the case
+$n=8$, for example, gives us $8+1=9$ distinct likelihood functions,
+Figure~\ref{ninelikes}.  We see an unanticipated asymmetry between $r$
+and $n-r$.  For $r=5$ one might expect that $\hat{a}=\frac{1}{2}$ on
+the grounds that the focal competitor's performance was median.
+However, $\hat{a}\simeq 0.5886$.  Figure~\ref{dotprobs} shows the
+maximum likelihood estimator $\hat{a}$ for various values of $n,r$,
+obtained by direct numerical maximization of
+equation~\ref{likeforrn3}.
 
@@
 
-Is there evidence that the author's performance differs between
-years?  Figure~\ref{parkruncontour} shows contours of
-support [again normalised so the support of the evaluate is zero] for
-the parkrun dataset, as a function of 2023 strength on the
-horizontal axis and 2024 strength on the vertical axis.  A null of
-equal strengths appears as the diagonal line.  We see that the maximum
-likelihood estimate is $\hat{p}_\mathrm{2023}=0.427$,
-$\hat{p}_\mathrm{2024}=0.408$.  However, a constrained optimization
-shows that the maximum support along the null diagonal is about
-$-2.43$, at ${p}_\mathrm{2023}={p}_\mathrm{2024}\simeq 0.409$.  The
-two units of support criterion~\citep{edwards1972} is met, and we are
-justified in concluding that the student is indeed more competent at
-applied than pure mathematics.  Alternatively, we may observe that
-$2\mathcal{S}=4.86$ is in the tail region of its asymptotic $\chi^2_1$
-distribution, corresponding to a $p$-value of about $0.028$.
-
-Now consider table~\ref{fisherparkrun}, giving a summary of whether
-nonfocal ocmpetitors place ahead of the author, split by venue.  A
-two-sided Fisher test gives a $p$-value of about $5\times 10^{-14}$,
+Is there evidence that the author's performance differs between years?
+Figure~\ref{parkruncontour} shows contours of support [again
+  normalised so the support of the evaluate is zero] for the parkrun
+dataset, as a function of 2023 strength on the horizontal axis and
+2024 strength on the vertical axis.  A null of equal strengths appears
+as the diagonal line.  We see that the maximum likelihood estimate is
+$\hat{p}_\mathrm{2023}=0.413$, $\hat{p}_\mathrm{2024}=0.485$.
+However, a constrained optimization shows that the maximum support
+along the null diagonal is about $-0.22$, at
+${p}_\mathrm{2023}={p}_\mathrm{2024}\simeq 0.441$.  The two units of
+support criterion~\citep{edwards1972} is not met, and we conclude that
+there is no evidence for a change in competitive strength between the
+years.  Now consider table~\ref{fisherparkrun}, giving a summary of
+whether nonfocal competitors place ahead of the author, split by year.
+A two-sided Fisher test gives a $p$-value of about $5\times 10^{-14}$,
 surely an unreasonable conclusion given the author's perceived

### 2024-10-11 inst/very_simplified_likelihood.Rmd
off-by-one error fixing

@@
 
-Now some likelihood visuals.  Suppose $r=4,n=8$:
+Now some likelihood visuals.  Suppose $n=9$:
 
@@
 }
-abline(v=MLE(r=5,n=8))
+abline(v=MLE(r=5,n=9))
 text(0.09,0.95,"r=9",col=rain[9])
@@
 ```{r shown4n8mle}
-MLE(r=5,n=8)
+MLE(r=5,n=9)
 ```

### 2024-10-18 inst/home_advantage.Rmd
formatting of time taken to run without cache
mention hyper3

@@
 ---
-title: "Quantifying the home ground advantage with `hyper2` objects"
+title: "Quantifying the home ground advantage with `hyper2` and `hyper3` objects"
 author: "R. K. S. Hankin"
@@
 
-To cite the `hyper2` package in publications, please use @hankin2017_rmd
-(takes about fifteen minutes to run without cache).  Here I analyse a
-small synthetic dataset for home ground advantage, then go on to
-discuss a dataset presented by Agresti.  But first, the smallest
-nontrivial example is a 2x2 case:
+(takes about fifteen minutes to run without cache).
+
+To cite the `hyper2` package in publications, please use
+@hankin2017_rmd.  Here I analyse a small synthetic dataset for home
+ground advantage, then go on to discuss a dataset presented by
+Agresti.  But first, the smallest nontrivial example is a 2x2 case:
 

### 2024-10-18 inst/ternaryplot_hyper2.Rmd
consistentification

@@
 ---
-title: "Ternary plots for hyper2 samples"
+title: "Ternary plots for `hyper2` samples"
 author: "Robin Hankin"
@@
 
+(takes about fifteen minutes to run without cache).
+
 To cite the `hyper2` package in publications please use @hankin2017_rmd.
@@
 ```{r loadlibraries}
-library("hyper2")
-library("Ternary")
+suppressPackageStartupMessages(library("Ternary"))
+suppressPackageStartupMessages(library("hyper2"))
 ```

### 2024-10-18 inst/ternaryplot_hyper2.Rmd
fixes #137

@@
 ---
-title: "Ternary plots for `hyper2` samples"
+title: "Ternary plots for `hyper2` and `hyper3` samples"
 author: "Robin Hankin"
@@
 
-To cite the `hyper2` package in publications please use @hankin2017_rmd.
-Here I show how to use the excellent `Ternary` package [@smith2017] to
-generate triangular plots for `hyper2` likelihood functions with size
-3.  First load the `Ternary` and `hyper2` packages:
+To cite the `hyper2` package in publications please use
+@hankin2017_rmd.  Here I show how to use the excellent `Ternary`
+package [@smith2017] to generate triangular plots for `hyper2` and
+`hyper3` likelihood functions with size 3.
+
+First load the `Ternary` and `hyper2` packages:
 
@@
 
-## The `chess` dataset
+# The `chess` dataset (`hyper2`)
 
@@
 
-## The `icons` dataset
+# The `icons` dataset (`hyper2`)
 
@@
 ```
+
+# Synthetic race data (`hyper3`)
+
+Here we use `rrace3()` to generate repeated results from a race
+between three runners.
+
+```{r singlerrace3example}
+set.seed(0)
+(randomrace <- noquote(rrace3(pn = c(a=6, b=3, c=8), ps = c(a=0.3,b=0.5,c=0.2))))
+(S <- ordervec2supp3(randomrace))
+maxp(S)
+```
+
+Above we see `randomrace` being an order vector for a race with six
+`a` clones, 3 `b` clones and 2 `c` clones of PL strengths 0.3, 0.5 and
+0.2 respectively.  Thus the first place was a `b`, second place was an
+`a`, third place a `c`, and so on.  Object `S` is support function for
+this realisation and we also see the maimum likelihood estimate for
+the three strengths.  Now we repeatedly run races with these PL
+strengths and calculate the maximum likelihood estimate for each
+realisation:
+
+```{r repeatedrrace3, cache = TRUE}
+x <- t(replicate(50,maxp(ordervec2supp3(rrace3(pn = c(a=16, b=13, c=18), ps = c(a=0.3,b=0.5,c=0.2))))))
+```
+
+```{r showx}
+head(x)
+```
+
+```{r ternaryrrace3}
+TernaryPlot(atip="a", btip="b", ctip="c")
+TernaryPoints(x, pch=16, cex=0.5)
+TernaryPoints(c(0.3,0.5,0.2),pch=16,col="red")
+```
+
+Above, we see the ensemble of estimates of the players' Plackett-Luce
+strengths as black dots, and the true value of the strengths
+(viz. `c(a=0.3,b=0.5,c=0.2)`) in red.  The estimates cluster around the true value.  It might be interesting to assess whether the estimates 
+estimates are Dirichlet and, if so, what parameters the distribution
+possesses.

### 2024-10-18 inst/ternaryplot_hyper2.Rmd
use bookdown so figures can be numbered

@@
 author: "Robin Hankin"
-output: html_document
+output: bookdown::html_document2
 bibliography: hyper2.bib  # copy vignettes/hyper2.bib to inst/
@@
 Using the `TernaryPlot()` and `TernaryPoints()` functions is straightforward:
-```{r plotdata}
+```{r plotdata, fig.cap="Ternary plot of estimated Plackett-Luce strengths of three chess players, using synthetic data"}
 jj <- pnames(chess)
@@
 
-Above we see black circles being samples from the posterior and the
-red point is the maximum likelihood estimate for the three players'
-strengths.  We can adapt this technique to higher-dimensional `hyper2`
-objects.
+In figure \@ref(fig:plotdata) we see black circles being samples from
+the posterior and the red point is the maximum likelihood estimate for
+the three players' strengths.  We can adapt this technique to
+higher-dimensional `hyper2` objects.
 
@@
 
-```{r, iconsmatplot}
+```{r iconsmatplot, fig.cap="Metropolis-Hastings random sampling of the `icons` dataset: the six Bradley-Terry strengths of each icon"}
 matplot(x,type="l",lty=1)
@@
 
-```{r, selectjustbigones}
+```{r selectjustbigones, fig.cap='Ternary plot of the icons dataset, showing Norfolk Broads ("NB", top), Polar Bears ("PB", lower right), and everything else ("other", lower left).
+Small black dots show MLEs of synthetic (Metropolis-Hastings) datasets; large red dot shows MLE for the original dataset'}
 wanted <- c(1,3) # NB,PB
-xx <- cbind(x[,wanted],fillup=rowSums(x[,-wanted]))
-TernaryPlot(atip="NB",btip="PB",ctip="other")
-TernaryPoints(xx,cex=0.5)
+xx <- cbind(x[, wanted], fillup=rowSums(x[, -wanted]))
+TernaryPlot(atip="NB", btip="PB", ctip="other")
+TernaryPoints(xx, cex=0.5, pch=16)
+jj <- icons_maxp[wanted]
+TernaryPoints(c(jj,1-sum(jj)), pch=16, col="red")
 ```
+
+Figures \@ref(fig:iconsmatplot) and \@ref(fig:selectjustbigones) show
+different aspects of the BT strengths of the six icons of the `icons`
+dataset.
 
@@
 
-```{r ternaryrrace3}
+```{r ternaryrrace3, fig.cap="Maximum likelihood estimates of synthetic datasets generated by the `rrace3()` random race function"}
 TernaryPlot(atip="a", btip="b", ctip="c")
@@
 
-Above, we see the ensemble of estimates of the players' Plackett-Luce
+Figure \@ref(fig:ternaryrrace3) shows an ensemble of estimates of the players' Plackett-Luce
 strengths as black dots, and the true value of the strengths

### 2024-10-18 inst/ternaryplot_hyper2.Rmd
observation about nature of MH sampling
consistentification

@@
 
-(takes about fifteen minutes to run without cache).
+<font size="1"> (takes about fifteen minutes to run without cache)
+</font>
 
@@
 different aspects of the BT strengths of the six icons of the `icons`
-dataset.
+dataset.  Note that figure \@ref(fig:selectjustbigones) is potentially
+misleading: due to the sequential correlation of MH sampling, repeated
+observations are overplotted and thus under-represented.
 

### 2024-10-18 inst/carcinoma.Rmd
consistentify "time to process" information

--- 2023-11-01 inst/carcinoma.Rmd
@@
 
-To cite the `hyper2` package in publications, please use @hankin2017
-(takes maybe twenty minutes to run without cache.  The value of
-`carcinoma_maxp` is calculated separately [discussed below], and that
-takes a few days to run).
+<font size="1"> (takes about fifteen minutes to run without cache; the
+value of `carcinoma_maxp` is calculated separately [discussed below],
+and that takes a few days to run.  </font>
 
-This is a short analysis of the carcinoma dataset presented by
-Agresti, table 13.1, p542.  Seven pathologists classified each of 118
-slides on the presence or absence of carcinoma in the uterine cervix.
+To cite the `hyper2` package in publications, please use
+@hankin2017_rmd.  This is a short analysis of the carcinoma dataset
+presented by Agresti, table 13.1, p542.  Seven pathologists classified
+each of 118 slides on the presence or absence of carcinoma in the
+uterine cervix.
 

### 2024-10-18 inst/global_liveability_ranking.Rmd
consistentify "time to process" information

@@
 
-To cite the `hyper2` package in publications, please use @hankin2017_rmd;
-takes about an hour to process without cache.
+<font size="1"> (takes about an hour to run without cache)
+</font>
 
-_This analysis is essentially without value, see the very end for why_
+*This analysis is essentially without value, see the
+very end for why*
 
-[The global liveability
+To cite the `hyper2` package in publications, please use
+@hankin2017_rmd.  [The global liveability
 ranking](https://en.wikipedia.org/wiki/Global_Liveability_Ranking)

### 2024-10-18 inst/kka.Rmd
clarification that white_draw3() is a package function
consistentify "time to process" information

@@
 
-To cite the `hyper2` package in publications, please use @hankin2017_rmd;
-this `Rmd` file takes about 45 minutes to process without cache.  Here
-we define dataset `kka` which is the observed scorelines between
-players Karpov, Kasparov, and Anand.  It is used to calculate
-likelihood functions: `karpov_kasparov_anand`, `kka_3whites` and
-`kka_3draws`.  This Rmd file supercedes files
+<font size="1"> (takes about forty five minutes to run without cache)
+</font>
+
+To cite the `hyper2` package in publications, please use
+@hankin2017_rmd.  Here we define dataset `kka` which is the observed
+scorelines between players Karpov, Kasparov, and Anand.  It is used to
+calculate likelihood functions: `karpov_kasparov_anand`, `kka_3whites`
+and `kka_3draws`.  This Rmd file supercedes files
 `inst/karpov_kasparov_anand.R` and `inst/kka_3draws.R` and
@@
 
-We can use the bespoke function `white_draw3()` to make a likelihood
-function:
+We can use the bespoke function `hyper2::white_draw3()` to make a
+likelihood function:
 

### 2025-01-17 inst/hankin-AustrianJournalofStatistics.tex
off-by-one error fixing
typo

@@
   of $r$ indicated in colour by its curve; vertical line shows
-  $\hat{a}\simeq 0.5886$ for the case $r=4$}
+  $\hat{a}\simeq 0.5886$ for the case $r=5$}
 \end{centering}
@@
 beats a randomly chosen opponent?'' to which the binomial maximum
-likelihood estimate is $\frac{r-1}{n-r}$.  For repeated independent
+likelihood estimate is $\frac{r-1}{n-1}$.  For repeated independent
 observations $r_i,n_i$, $1\leqslant i\leqslant N$ we would have

### 2025-01-20 inst/hankin-AustrianJournalofStatistics.tex
minor tweaks to satisfy AJS style guides

@@
 
-Here a simplification of the Plackett--Luce likelihood functions for
-order statistics is proposed.  We consider a field of $n$ entities
+Here a new simplification of the Plackett--Luce likelihood functions
+for order statistics is proposed.  We consider a field of $n$ entities
 that are ordered in some way [the preferred example is competitors in
@@
   notation] but here arbitrary $n$ may be considered.  Taking the case
-$n=8$, for example, gives us $8+1=9$ distinct likelihood functions,
+$n=9$, for example, gives us 9 distinct likelihood functions,
 Figure~\ref{ninelikes}.  We see an unanticipated asymmetry between $r$
@@
 \includegraphics[width=4in]{ninelikes}  % ninelikes.pdf produced by very_simplified_likelihood.Rmd
-\caption{Likelihood functions for $r=0(1)8$\label{ninelikes}.  Value
+\caption{Likelihood functions for $r=1,2,\ldots 9$\label{ninelikes}.  Value
   of $r$ indicated in colour by its curve; vertical line shows
@@
 
-Analysis of expression~\ref{likeforrn3} for large $n$ and $r=pn$ for
-fixed $p$ shows that the support
-$\mathcal{S}=\log\mathcal{L}_{r,n}(a)$ is asymptotically
+Even though expression ~\ref{likeforrn3} represents a simplification
+of the Plackett-Luce likelihood function, it becomes intractable even
+for moderate values of $n$ and $r$.  Analysis of
+expression~\ref{likeforrn3} for large $n$ and $r=pn$ for fixed $p$
+shows that the support $\mathcal{S}=\log\mathcal{L}_{r,n}(a)$ is
+asymptotically
 
@@
 ${p}_\mathrm{2023}={p}_\mathrm{2024}\simeq 0.441$.  The two units of
-support criterion~\citep{edwards1972} is not met, and we conclude that
+support criterion~\citep{edwards1992} is not met, and we conclude that
 there is no evidence for a change in competitive strength between the
@@
 ${p}_\mathrm{pure}={p}_\mathrm{applied}\simeq 0.683$.  The two units of
-support criterion~\citep{edwards1972} is met, and we are justified in
+support criterion~\citep{edwards1992} is met, and we are justified in
 concluding that the student is indeed more competent at applied than
@@
 %\bibliographystyle{plainat}
-\bibliography{hyper2}
+\bibliography{simplified.bib}
 

### 2025-01-20 inst/hankin-AustrianJournalofStatistics.tex
cite R as specified in the AJS style guide

@@
 maximum likelihood estimator $\hat{a}$ for various values of $n,r$,
-obtained by direct numerical maximization of
-equation~\ref{likeforrn3}.
+obtained by direct numerical maximization, in the R programming
+language \cite{rcore2024}, of equation~\ref{likeforrn3}.
 

### 2025-01-22 inst/hankin-AustrianJournalofStatistics.tex
typo

@@
 inference problem has a wide range of applications and I present some
-analysis of datasets drawn from the fields of Olympic athletics,
+analysis of datasets drawn from the fields of amateur athletics,
 education, and Formula~1 motor racing.

### 2025-01-30 inst/jss4832.tex
instructions as to how to produce the PDF

@@
 % -*- mode: noweb; noweb-default-code-mode: R-mode; -*-
+
+% Notes:
+
+% To bibtex this file, copy hyper2.bib from the vignettes/ directory
+% of the repo (which is under version control).  To create the pdf
+% files for the figures, source file "jss4832_replication.R" which
+% will create them.  There are slight differences between the JSS
+% paper and the manuscript here and there in the citations.
+
 \documentclass[article]{jss}
@@
 
-The \code{hyper2} package~\citep{hankin2009,hankin2017} furnishes
+The \code{hyper2} package~\citep{hankin2017_rmd} furnishes
 computational support for generalized
@@
 thereby admitting a wide range of non-standard nulls such as order
-statistics on the $p_i$~\citep{hankin2017}.  It becomes possible to
+statistics on the $p_i$~\citep{hankin2017_rmd}.  It becomes possible to
 analyse a wider range of likelihood functions than standard
@@
 introduction of reified entities (monsters) allows one to consider
-draws~\citep{hankin2009}, noncompetitive tactics such as
+draws~\citep{hankin2020}, noncompetitive tactics such as
 collusion~\citep{hankin2020}, and the phenomenon of team cohesion
@@
 
-Existing \code{hyper2} formalism as described by \citet{hankin2017}
+Existing \code{hyper2} formalism as described by \citet{hankin2017_rmd}
 cannot represent Equation~\ref{plackettluce}, because
@@
 and in this case we need multisets\footnote{Note that the version of
-\code{hyper2} presented by \citet{hankin2017} and reviewed
+\code{hyper2} presented by \citet{hankin2017_rmd} and reviewed
 by~\cite{turner2020} used integer-valued sets together with a print
@@
 The package provides a number of statistical tests on likelihood
-functions, modified from~\citet{hankin2017} to work with \code{hyper3}
+functions, modified from~\citet{hankin2017_rmd} to work with \code{hyper3}
 objects.  For example, we may assess the hypothesis that all three
@@
 Vad}} + p_{\mbox{\tiny Ves}}\right)\simeq 0.74$.  Indeed, from a
-training or selection perspective we might follow \cite{hankin2017}
+training or selection perspective we might follow \cite{hankin2017_rmd}
 and observe that log-contrasts~\citep{ohagan2004} appear to have
@@
 accommodated, either using likelihoods comprising sums of
-Plackett-Luce probabilities~\citep{hankin2017}; or the introduction of
+Plackett-Luce probabilities~\citep{hankin2017_rmd}; or the introduction of
 a reified draw entity~\citep{hankin2010}.

### 2025-01-30 inst/podium.Rmd
magrittr package needed
more detail

@@
 library("hyper2")
+library("magrittr")
 options("digits" = 5)
@@
 
-To cite the `hyper2` package in publications, please use @hankin2017_rmd.
-This short document defines the `podium()` function which tests the
-hypothesis that a Formula 1 driver's being on the podium at one venue
-affects his performance on the next.  Data from the 2007 season are
-used.  It uses `hyper3` objects from the `hyper2` package.  A detailed
-use-case for the top six drivers (Alonso, Hamilton, Heidfeld, Kubica,
-Massa, and Raikkonen), and only the first five venues (Austria,
+<font size="1"> (almost all the time is spent on chunk `podlikevec7`,
+which takes about one minute per evaluation) </font>
+
+To cite the `hyper2` package in publications, please use
+@hankin2017_rmd.  This short document defines the `podium()` function
+which tests the hypothesis that a Formula 1 driver's being on the
+podium at one venue affects his performance on the next.  Data from
+the 2007 season are used.  It uses `hyper3` objects from the `hyper2`
+package.  A detailed use-case for the top six drivers (Alonso,
+Hamilton, Heidfeld, Kubica, Massa, and Raikkonen^[Umlauts pose special
+problems for Rmd files]), and only the first five venues (Austria,
 Malaysia, Bahrain, Spain, and Monaco) is given.
@@
 
-Define a subset of it:
+The first race has standard Plackett-Luce likelihood:
+
+$$a\succ b\succ c\succ d\succ e\succ f$$
+
+has a PL likelihood of
+
+$$
+\frac{a}{a+b+c+d+e+f}\cdot
+\frac{b}{  b+c+d+e+f}\cdot
+\frac{c}{    c+d+e+f}\cdot
+\frac{d}{      d+e+f}\cdot
+\frac{e}{        e+f}
+$$
+
+However, if the result were
+
+$$\overline{a}\succ b\succ\overline{c}\succ\overline{d}\succ e\succ f$$
+
+(where an overbar signifies that the driver was on the podium after
+the previous race), the likelihood function would be
+
+$$
+\frac{\lambda a}{\lambda a+b+\lambda c+\lambda d+e+f}\cdot
+\frac{        b}{          b+\lambda c+\lambda d+e+f}\cdot
+\frac{\lambda c}{            \lambda c+\lambda d+e+f}\cdot
+\frac{\lambda d}{                      \lambda d+e+f}\cdot
+\frac{        e}{                                e+f}\cdot
+$$
+
+and the null of no podium effect would be $H_0\colon\lambda=0$.  Next,
+we will take dataset `formula1_2007` and define a subset of it which
+we will call `b`:
 
@@
 
+Taking Monaco (`MON`) as an example, the preceding race was in spain
+(`ESP`) at which Massa, Hamilton, and Alonso finished on the podium.
+The order statistic would be
+
+$$
+\text{Raikkonen}\succ
+\overline{\text{Hamilton}}\succ
+\overline{\text{Massa}}\succ
+\overline{\text{Alonso}}\succ
+\text{Kubica}\succ
+\text{Heidfeld}
+$$
+
+where again an overbar signifies that the driver was on the podium in
+the previous race.  The likelihood function for Monaco would thus be
+
+$$
+\begin{split}
+\operatorname{\mathcal{L}}\left(\lambda;
+p_\text{Raikkonen},\ldots,p_\text{Heidfeld}
+\right)
+& =
+\frac{p_\text{Raikkonen}}{p_\text{Raikkonen}+\lambda p_\text{Hamilton}+\lambda p_\text{Massa}+\lambda p_\text{Alonso}+p_\text{Kubica}+p_\text{Heidfeld}}\\
+&{\qquad} \cdot\frac{\lambda p_\text{Hamilton}}{\lambda  p_\text{Hamilton}+\lambda p_\text{Massa}+\lambda p_\text{Alonso}+p_\text{Kubica}+p_\text{Heidfeld}}\\
+&{\qquad}\cdot\frac{\lambda p_\text{Massa}}{\lambda p_\text{Massa}+\lambda p_\text{Alonso}+p_\text{Kubica}+p_\text{Heidfeld}}\\
+&{\qquad}\cdot\frac{\lambda p_\text{Alonso}}{\lambda p_\text{Alonso}+p_\text{Kubika}+p_\text{Heidfeld}}\cdot
+\frac{p_\text{Kubica}}{p_\text{Kubica}+p_\text{Heidfeld}}
+\end{split}
+$$
+
 ```{r,usepodiu,cache=TRUE}
-Ha <- podium(1.4,b)
-Hb <- podium(1.5,b)
-ma <- maxp(Ha,give=TRUE,n=1)
-mb <- maxp(Hb,give=TRUE,n=1)
+Ha <- podium(1.4, b)
+Hb <- podium(2.1, b)
+ma <- maxp(Ha, give=0, n=1)
+mb <- maxp(Hb, give=0, n=1)
 ```
@@
 Ha
-Hb
+summary(Hb)
 ma
@@
 ```
+
+Above we see that passing $\lambda$ values of 1.4 and 2.1 result in
+different likelihoods.  The difference of about 0.4 units of support
+is not significant, failing to reach the two-units-of-support
+criterion of Edwards.
 
@@
 date()
-lam <- seq(from=0.2,to=1.2,len=7)
+lam <- seq(from=0.2,to=1.2,len=15)
 like <- lapply(lam,podlike)

### 2025-01-30 inst/rock_paper_scissors_monster.Rmd
consistentify "time to process" information
more explanation

@@
 
-To cite the `hyper2` package in publications, please use @hankin2017_rmd;
-takes about an hour to run without cache.  Here I use the `hyper2`
-package to quantify the strength of a circular rock-paper-scissors
-effect using the `hyper2` package, improving on previous results
-[@west2008] which merely established its existence.  I use two
-datasets, a three-player scoreline and a synthetic dataset referring
-to wear on a seven-tooth gear wheel.
+<font size="1"> (takes about an hour to run without cache)
+</font>
 
-## Rock Paper Scissors
+### Abstract {-}
+
+To cite the `hyper2` package in publications, please use
+@hankin2017_rmd.  Here I use the `hyper2` package to quantify the
+strength of a circular rock-paper-scissors effect using the `hyper2`
+package, improving on previous results [@west2008] which merely
+established its existence.  I use two datasets, a three-player
+scoreline and a synthetic dataset referring to wear on a seven-tooth
+gear wheel.
+
+# Introduction
 
@@
 
-## RPS in chess
+# RPS in chess
 
@@
 
-Estimating $p_M$ is also straightforward:
+Above we see a p-value of about 1.2\%, sufficient to reject the null
+of zero monster strength^[Note that diagonal entries of `chess_table`
+correspond to procyclic wins and off-diagonal entries correspond to
+anti-cyclic wins.  This is due to the fact that `chess_table` is of
+the same type as the seven-toothed gear wheel dataset considered
+below, in which prograde "wins" appear on the diagonal and retrograde
+wins appear one place to the right (wrapped if necessary)].
+
+Estimating $M$, the strength of the monster, is also straightforward.
+First we consider the case where the probability of a procyclic win is
+known to be $p$ (a "procyclic win" is a win that would be expected
+under a strong monster).
 
 $$
-Prob(\mathrm{procyclic}\ \mathrm{win})=\frac{\alpha+M}{2\alpha+M}
+\operatorname{Prob}(\mathrm{procyclic}\ \mathrm{win})=\frac{\alpha+M}{2\alpha+M}
 $$
 
-(a "procyclic win" is a win that would be expected under a strong
-monster).  Here we also have $3\alpha+M=1$, so to estimate the
-strength of the monster we would solve
+Here we also have $3\alpha+M=1$, so to estimate the strength of the
+monster we would solve
 
@@
 
-### Profile likelihood
+# Profile likelihood
 
@@
 
-### Analysis of seven-tooth gear with `hyper3`
+## Analysis of seven-tooth gear with `hyper3`
 

### 2025-02-05 inst/collusion_chess_draws.Rnw
consistentification of .bib file

@@
 modified and extended many times, for example to account for multi-way
-competitions such as {\em MasterChef Australia}~\citep{hankin2017},
+competitions such as {\em MasterChef Australia}~\citep{hankin2017_rnw},
 and non-competitive situations such as consumer
@@
 The analysis presented here uses the framework of the
-R~\citep{rcore2019} \pkg{hyper2} package~\citep{hankin2017} as a
+R~\citep{rcore2019} \pkg{hyper2} package~\citep{hankin2017_rnw} as a
 convenient and unified interface.  Because gradients are available,
@@
 
-\bibliography{chess}
+\bibliography{hyper2}
 \end{document}

### 2025-02-06 inst/home_advantage.Rmd
consistentify "time to process" information
home_away_table()

@@
 ground advantage, then go on to discuss a dataset presented by
-Agresti.  But first, the smallest nontrivial example is a 2x2 case:
+Agresti.  At the end, I show how a results table can be created from
+scratch using bespoke function `home_away_table()`.
+
+But first, the smallest nontrivial example is a 2x2 case:
 
@@
 
+# Creating a home/away results table from scratch
+
+Function `home_away_table()` takes a data frame of results and
+produces a table amenable to analysis by `home_away()` or
+`home_away3()`.
+
+```{r showdata}
+a <- read.table("celtic_rangers_livingston_falkirk.txt", header=TRUE)
+head(a)
+(aF <- home_away_table(a, give=FALSE))
+(aT <- home_away_table(a, give=TRUE))
+```
+
+```{r doscottish,cache=TRUE}
+(H <- home_away(aF))
+(mH <- maxp(H))
+```
+
+```{r testrangersandceltic, cache=TRUE}
+samep.test(H,c("Rangers","Celtic"))
+
+```{r  testhomezero, cache=TRUE}
+specificp.gt.test(H,"home",0)
+```
+
+```{r dohome3, cache=TRUE}
+home_away3(aF,lambda=1.88)
+maxp(home_away3(aF,lambda=1.88),give=1)
+maxp(home_away3(aF,lambda=1.11),give=1)
+```
+
+```{r calcsuppscotty,cache=TRUE}
+f <- function(l){ maxp(home_away3(aF,l),give=TRUE,n=1)$likes}
+lam <- seq(from=1.4,to=2.3,len=10)
+date()
+loglike <- sapply(lam,f)
+date()
+```
+
+```{r optscotty,cache=TRUE}
+(o <- optimize(f,c(1.1,4.6),maximum=TRUE))
+```
+
+```{r scottychisq,cache=TRUE}
+o
+maxL <- f(o$maximum)
+W <- maxL - f(1)    # f(1) being the likelihood of the null
+W
+pchisq(2*W,df=1,lower.tail=FALSE)
+```
+
+```{r scotmax}
+(H <- home_away3(home_away_table(a,give=FALSE),o$maximum))  # o$maximum ~= 1.353
+maxp(H,give=1)
+```
+
+```{r}
+plot(lam,loglike-max(loglike),type='b')
+abline(h=c(0,-2))
+abline(v=1)
+grid()
+```
+
+```{r rememberhyper2}
+(mH2 <- round(mH*1000))
+```
+
+```{r sfasdf, cache=TRUE}
+o$maximum
+mH3 <- maxp(home_away3(home_away_table(a,give=FALSE),o$maximum),give=FALSE)
+round(mH3*1000)
+```
+
+Now consider Celtic vs Falkirk, Celtic at home and also  Falkirk at home.
+
+### (Celtic at home) {-}
+* hyper2: P(Celtic;Falkirk) = $\frac{390+67; 56}{390 + 56 + 67}\simeq (0.89; 0.11)$
+* hyper3: P(Celtic;Falkirk) = $\frac{1.77*412; 74}{1.77*412+74}\simeq(0.92; 0.08)$
+
+actual scoreline:  61-8 or about $(0.88; 0.12)$
+
+###  (Celtic away)
+* hyper2: P(Celtic;Falkirk) = $\frac{390; 56+67}{390 + 56 + 67}\simeq (0.76;0.24)$
+* hyper3: P(Celtic;Falkirk) = $\frac{412; 1.77*74}{412 + 1.77*74}\simeq(0.76; 24)$
+
+actual scoreline: 43-17 or about $(0.72; 0.28)$
+
 ###  References

### 2025-02-06 R/hyper2.R

@@ -1301,3 +1301,31 @@ rorder_single <- function(p){
     names(out) <- NULL
     return(out)
 }
+
+`home_away_table` <- function(a, give = FALSE, teams){
+    if(missing(teams)){ teams <- unique(c(a$hometeam, a$awayteam)) }
+    homewins <- matrix(0, length(teams), length(teams))
+    dimnames(homewins) <- list(`@home` = teams, `@away` = teams)
+    awaywins <- homewins
+    draws    <- homewins
+    for(i in seq_len(nrow(a))){
+        if(a$homescore[i] > a$awayscore[i]){ # home win
+            homewins[a$hometeam[i], a$awayteam[i]] %<>% add(1)
+        } else if(a$homescore[i] < a$awayscore[i]){  # away win
+            awaywins[a$hometeam[i], a$awayteam[i]] %<>% add(1)
+        } else { # draw
+            draws[a$hometeam[i], a$awayteam[i]] %<>% add(1)
+        }
+    }
+
+    diag(homewins) <- NA
+    diag(awaywins) <- NA
+    diag(draws)    <- NA
+    
+    if(give){
+        return(list(home_wins = homewins, away_wins = awaywins, draws = draws))
+    } else {
+        return(homewins + 1i*awaywins)
+    }
+}
+

### 2025-02-07 inst/home_advantage.Rmd
chi-square analysis

@@
 library("hyper2",quietly=TRUE)
+library("magrittr")
 options(warn = defaultW)
@@
 head(a)
-(aF <- home_away_table(a, give=FALSE))
-(aT <- home_away_table(a, give=TRUE))
+teams <- c("Rangers","Celtic","Falkirk","Livingston")
+(aF <- home_away_table(a, give=FALSE,teams = teams))
+(aT <- home_away_table(a, give=TRUE, teams = teams))
 ```
+
+(from here on we ignore draws and consider only won games)
 
@@
 (H <- home_away(aF))
+pnames(H) <- c(teams,"home")
+H
 (mH <- maxp(H))
@@
 
-Now consider Celtic vs Falkirk, Celtic at home and also  Falkirk at home.
+# Rangers vs Falkirk {-}
 
-### (Celtic at home) {-}
+**(Rangers at home)**
+
+* hyper2: P(Rangers;Falkirk) = $\frac{448+67; 56}{448 + 56 + 67}\simeq (0.902; 0.098)$
+* hyper3: P(Rangers;Falkirk) = $\frac{1.77*448; 56}{1.77*448+56}\simeq(0.934; 0.066)$
+
+actual scoreline: 69(R) / 2(F) or about $(0.972; 0.028)$. Expectation
+would be (64;7) (H2) or (66.3;4.7) (H3)
+
+*(Rangers away)*
+
+* hyper2: P(Rangers;Falkirk) = $\frac{448; 56+67}{448 + 56 + 67}\simeq (0.78;0.22)$
+* hyper3: P(Rangers;Falkirk) = $\frac{448; 1.77*56}{448 + 1.77*56}\simeq(0.82; 0.18)$
+
+actual scoreline: 57(R) /17(F) - or about $(0.77; 0.23)$
+
+# Rangers vs Livingston {-}
+
+**(Rangers at home)**
+
+* hyper2: P(Rangers;Livingston) = $\frac{448+67; 39}{448 + 39 + 67}\simeq (0.930; 0.070)$
+* hyper3: P(Rangers;Livingston) = $\frac{1.77*448; 39}{1.77*448+39}\simeq(0.953; 0.047)$
+
+actual scoreline:  15(R) / 0(L) or  $(1; 0)$
+
+*(Rangers away)*
+
+* hyper2: P(Rangers;Livingston) = $\frac{448; 39+67}{448 + 39 + 67}\simeq (0.81;0.19)$
+* hyper3: P(Rangers;Livingston) = $\frac{448; 1.77*39}{448 + 1.77*39}\simeq(0.86; 0.14)$
+
+actual scoreline: 12(R) /2(L) - or about $(0.86; 0.14)$
+
+# Celtic vs Falkirk {-}
+
+*(Celtic at home)*
+
 * hyper2: P(Celtic;Falkirk) = $\frac{390+67; 56}{390 + 56 + 67}\simeq (0.89; 0.11)$
@@
 
-###  (Celtic away)
+*(Celtic away)*
+
 * hyper2: P(Celtic;Falkirk) = $\frac{390; 56+67}{390 + 56 + 67}\simeq (0.76;0.24)$
-* hyper3: P(Celtic;Falkirk) = $\frac{412; 1.77*74}{412 + 1.77*74}\simeq(0.76; 24)$
+* hyper3: P(Celtic;Falkirk) = $\frac{412; 1.77*74}{412 + 1.77*74}\simeq(0.76; 0.24)$
 
 actual scoreline: 43-17 or about $(0.72; 0.28)$
+
+# Livingston vs Falkirk {-}
+
+*(Livingston at home)*
+
+* hyper2: P(Livingston;Falkirk) = $\frac{39+67; 56}{39 + 67 + 56}\simeq (0.65; 0.35)$
+* hyper3: P(Livingston;Falkirk) = $\frac{1.77*42; 74}{1.77*42+74}\simeq(0.50; 0.50)$
+
+actual scoreline: 3-9 or about $(0.25; 0.75)$
+
+*(Livingston away)*
+
+* hyper2: P(Livingston;Falkirk) = $\frac{39; 56+67}{39 + 56 + 67}\simeq (0.24;0.76)$
+* hyper3: P(Livingston;Falkirk) = $\frac{42; 1.77*74}{42 + 1.77*74}\simeq(0.24; 0.76)$
+
+actual scoreline: 6-8 or about $(0.43; 0.57)$
+
+# Expectations
+
+First, hyper2:
+
+```{r}
+mH
+mH["Rangers"]
+mH["home"]
+mH3
+lambda <- o$maximum
+h2_exp <- matrix(0,4,4)
+rownames(h2_exp) <- teams
+colnames(h2_exp) <- teams
+h3_exp <- h2_exp
+n_matches <- Re(aF) + Im(aF)
+for(hometeam in teams){
+    for(awayteam in teams){
+
+        h2_exp[hometeam,awayteam] <- (
+            n_matches[hometeam,awayteam] * (mH[hometeam] + mH["home"])/
+            (mH[hometeam] + mH[awayteam] + mH["home"])
+        )
+        h2_exp[hometeam,awayteam] %<>%
+            add(1i*
+                n_matches[hometeam,awayteam] * (mH[awayteam])/
+                (mH[hometeam] + mH[awayteam] + mH["home"])
+                )
+
+        h3_exp[hometeam,awayteam] <- (
+            n_matches[hometeam,awayteam] * (lambda* mH3[hometeam])/
+            (lambda* mH3[hometeam] + mH3[awayteam])
+        )
+
+        h3_exp[hometeam,awayteam] %<>%
+            add(1i*
+                n_matches[hometeam,awayteam] * (mH3[awayteam])/
+                (lambda* mH3[hometeam] + mH3[awayteam])
+        )
+    }
+}
+aF
+round(h2_exp,1)
+round(h3_exp,1)
+
+makev <- function(M){
+    out <- c(Re(M),Im(M))
+    out[!is.na(out)]
+}
+
+(obs <- makev(aF))
+(exp2 <- makev(h2_exp))
+(exp3 <- makev(h3_exp))
+
+plot((obs-exp2)^2/exp2)
+plot((obs-exp3)^2/exp3)
+
+chisq2 <- sum((obs-exp2)^2/exp2)
+chisq3 <- sum((obs-exp3)^2/exp3)
+
+pchisq(chisq2,df=12-4,lower.tail=FALSE)
+pchisq(chisq3,df=12-4,lower.tail=FALSE)
+
+par(pty='s')
+plot(obs,exp2,log="xy",asp=1,xlim=c(1,100),ylim=c(1,100))
+abline(0,1)
+plot(obs,exp3,log="xy",asp=1,xlim=c(1,100),ylim=c(1,100))
+abline(0,1)
+```
 

### 2025-02-07 inst/home_advantage.Rmd
refactoring

@@
 o
-maxL <- f(o$maximum)
+lambda_baseball <- o$maximum
+maxL <- f(lambda_baseball)
 W <- maxL - f(1)    # f(1) being the likelihood of the null
@@
 ```{r, label=hyper3_baseball,cache=TRUE}
-jj1  <- seq(from=log(o$maximum),to=log(1.9),len=7)
-jj2  <- seq(from=log(o$maximum),to=log(0.8),len=7)
+jj1  <- seq(from=log(lambda_baseball),to=log(1.9),len=7)
+jj2  <- seq(from=log(lambda_baseball),to=log(0.8),len=7)
 l <- sort(unique(exp(c(jj1,jj2))))
-lam  <- sort(c(l,o$maximum))
+lam  <- sort(c(l,lambda_baseball))
 L <- sapply(lam,f) - maxL
@@
 o
-maxL <- f(o$maximum)
+scot_max_lambda <- o$maximum
+maxL <- f(scot_max_lambda)
+
 W <- maxL - f(1)    # f(1) being the likelihood of the null
@@
 ```{r scotmax}
-(H <- home_away3(home_away_table(a,give=FALSE),o$maximum))  # o$maximum ~= 1.353
+(H <- home_away3(home_away_table(a,give=FALSE),scot_max_lambda))  # o$maximum ~= 1.353
 maxp(H,give=1)
@@
 o$maximum
-mH3 <- maxp(home_away3(home_away_table(a,give=FALSE),o$maximum),give=FALSE)
+mH3 <- maxp(home_away3(home_away_table(a,give=FALSE),scot_max_lambda),give=FALSE)
 round(mH3*1000)
@@
 mH3
-lambda <- o$maximum
+lambda <- scot_max_lambda
 h2_exp <- matrix(0,4,4)
@@
 
+```{r defineexpectation2}
+expectation2 <- function(probs, results){
+    teams <- names(probs[names(probs) != "home"])
+    h2_exp <- matrix(0,4,4)
+    rownames(h2_exp) <- teams
+    colnames(h2_exp) <- teams
+    n_matches <- Re(aF) + Im(aF)
+    for(hometeam in teams){
+        for(awayteam in teams){
+
+            h2_exp[hometeam,awayteam] <- (
+                n_matches[hometeam,awayteam] * (mH[hometeam] + mH["home"])/
+                (mH[hometeam] + mH[awayteam] + mH["home"])
+            )
+
+            h2_exp[hometeam,awayteam] %<>%
+                add(1i*
+                    n_matches[hometeam,awayteam] * (mH[awayteam])/
+                    (mH[hometeam] + mH[awayteam] + mH["home"])
+                    )
+        }
+    }
+    return(h2_exp)
+}
+round(expectation2(mH,aF),1)
+```
+
+```{r defineexp3}
+
+expectation3 <- function(probs, results, lambda){
+    teams <- names(probs)
+    h3_exp <- matrix(0,4,4)
+    rownames(h3_exp) <- teams
+    colnames(h3_exp) <- teams
+    n_matches <- Re(aF) + Im(aF)
+    for(hometeam in teams){
+        for(awayteam in teams){
+
+            h3_exp[hometeam,awayteam] <- (
+                n_matches[hometeam,awayteam] * (lambda* mH3[hometeam])/
+                (lambda* mH3[hometeam] + mH3[awayteam])
+            )
+
+            h3_exp[hometeam,awayteam] %<>%
+                add(1i*
+                    n_matches[hometeam,awayteam] * (mH3[awayteam])/
+                    (lambda* mH3[hometeam] + mH3[awayteam])
+                    )
+        }
+    }
+    return(h3_exp)
+}
+
+round(expectation3(mH3[c("Rangers","Celtic","Falkirk","Livingston")],aF,scot_max_lambda),1)
+```
+
 ###  References

### 2025-02-07 inst/monster_vs_lambda.Rmd
consistentify "time to process" information
include theoretical result

@@
 title: "Monster or Lambda?  A likelihood analysis"
-subtitle: "Takes about an hour to run without cache"
 author: "R. K. S. Hankin"
@@
 ```
+
+<font size="1"> (takes about an hour to run without cache)
+</font>
 
@@
 assume $B=1-A$] the result of maximizing $\mathcal{S}$ with respect to
-$A$ and $\lambda$ does not appear to have a simple analytical
-solution; we must proceed numerically.  To do so, we need to define a
-slight generalization of `home_away3()`:
+$A$ and $\lambda$ gives the following result:
+
+$$
+\hat{A} = \frac{\sqrt{\alpha\gamma}}{\sqrt{\alpha\gamma} + \sqrt{\beta\delta}}
+$$
+
+$$
+\hat{\lambda} = \sqrt{\frac{\alpha\delta}{\beta\gamma}}
+$$
 

### 2025-02-08 inst/hyper3.Rmd
expectations

@@
 $$
+
+If we consider $p_1$ at home and $p_2$ away, we have $a+b$ matches.
+Using the maximum likelihood estimates
+$\hat{p_1},\hat{p_2},\hat{\lambda}$, the expected number of wins by
+$p_1$ is
+
+$$
+(a+b)\frac{\hat{\lambda}\hat{p_1}}{\hat{\lambda}\hat{p_1} + \hat{p_2}}
+=
+(a+b)\cdot
+\frac{
+\sqrt{\frac{ad}{bc}}\frac{\sqrt{ac}}{\sqrt{ac}+\sqrt{bd}}
+  }{
+\sqrt{\frac{ad}{bc}}\frac{\sqrt{ac}}{\sqrt{ac}+\sqrt{bd}} + \frac{\sqrt{bd}}{\sqrt{ac}+\sqrt{bd}}
+}
+=\cdots=a
+$$
+
+Also the expected number of matches won by $p_2$ when $p_1$ is at home
+can be shown to be $b$, and similarly when $p_2$ is at home, the
+expected number of victories is $c,d$ respectively.  Thus, at least in
+the two-team case, the expectations under the ML estimates are exactly
+the observations.
+
+### `hyper2` idiom for likelihood functions {-}
+
 Likelihoods of this form are easily specified using package idiom.

### 2025-02-08 inst/home_advantage.Rmd
weight comparison work of Davidson and Beaver

@@
 abline(h=c(0,-2))
+```
+
+### Analysis of  Work by Davidson and Beaver
+
+```{r}
+DB <- # Davidson and Beaver
+matrix(c(
+    NA, 32+18i, 36+14i, 47+03i, 47+03i,
+14+36i,     NA, 34+16i, 43+07i, 46+04i,
+06+44i, 14+36i,     NA, 34+16i, 40+10i,
+02+48i, 07+43i, 12+38i,     NA, 28+22i,
+01+49i, 02+48i, 05+45i, 10+40i,     NA
+),5,5,byrow=TRUE
+)
+jj <- c("90g","95g","100g","105g","110g")
+dimnames(DB) <- list(`@home` = jj, `@away` = jj)
+DB
+```
+
+```{r first_second_likefun, cache=TRUE}
+f <- function(l){ maxp(home_away3(DB,l),give=TRUE,n=1)$likes}
+lam <- seq(from=0.6,to=1,len=19)
+date()
+loglike <- sapply(lam,f)
+date()
+```
+
+```{r plotfirstsecond}
+plot(lam,loglike-max(loglike),type='b')
+abline(h=c(0,-2))
+abline(v=1)
 ```

### 2025-02-10 inst/home_advantage.Rmd
merge home_advantage.Rmd and home_advantage3.Rmd

--- 2025-02-07 inst/home_advantage.Rmd
@@
 To cite the `hyper2` package in publications, please use
-@hankin2017_rmd.  Here I analyse a small synthetic dataset for home
-ground advantage, then go on to discuss a dataset presented by
-Agresti.  At the end, I show how a results table can be created from
-scratch using bespoke function `home_away_table()`.
+@hankin2017_rmd.
 
-But first, the smallest nontrivial example is a 2x2 case:
+Objects of class \code{hyper3} are a generalization
+of \code{hyper2} objects that allow the brackets to contain weighted
+probabilities.  Likelihood functions are defined on non-negative
+$p_1,\ldots, p_n$ subject to the unit-sum constraint $\sum p_i=1$.
+Given known weights $w^i_j$ with $1\leq i\leq j$ we have
 
-```{r,label=smallestexample}
-M <- matrix(c(
-    NA, 1+5i, 3+11i, NA),
-    nrow=2,ncol=2,byrow=TRUE)
+$$\mathcal{L}\left(p_1,\ldots, p_n\right)=\prod_j\left(\sum_{i=1}^n w^i_jp_i\right)^{n_j}.$$
 
-teams <- LETTERS[seq_len(2)]
+As a motivating example, suppose two teams (players) with
+Bradley-Terry strengths $p_1,p_2$ play football where we quantify the
+home-ground advantage with a term $\lambda$.  If $p_1$ plays at home
+$a+b$ times with $a$ wins and $b$ losses, and plays away [so $p_2$ is
+at home] $c+d$ times with $c$ wins and $d$ losses, then a sensible
+likelihood function might be
+
+$$\mathcal{L}(p_1,p_2,\lambda;A,B,C,D)=
+\underbrace{
+\overbrace{\left(\frac{\lambda p_1}{\lambda p_1 + p_2}\right)^{A}}^{\mbox{$A$ home wins for $p_1$}}\quad
+\overbrace{\left(\frac{p_2        }{\lambda p_1 + p_2}\right)^{B}}^{\mbox{$B$ away wins for $p_2$}}
+}_\mbox{$p_1$ plays at home to $p_2$ $A+B$ times}\quad
+\underbrace{
+\overbrace{\left(\frac{\lambda p_2}{p_1 + \lambda p_2}\right)^{C}}^{\mbox{$C$ home wins for $p_2$}}\quad
+\overbrace{\left(\frac{        p_1}{p_1 + \lambda p_2}\right)^{D}}^{\mbox{$D$ away wins for $p_1$}}
+}_\mbox{$p_2$ plays at home to $p_1$ $C+D$ times}
+$$
+
+where we understand that $p_1+p_2=1$, $p_1,p_2,\lambda\geqslant 0$.
+Elementary techniques allow us to identify a maximum and we find
+
+$$
+\hat{p_1}=\frac{\sqrt{AD}}{\sqrt{AD}+\sqrt{BC}}\qquad
+\hat{p_2}=\frac{\sqrt{BC}}{\sqrt{AD}+\sqrt{BC}}\qquad
+\hat{\lambda}=\sqrt{\frac{AC}{BD}}
+$$
+
+Note that the evaluate is unchanged if $A,B,C,D$ are all multiplied by
+a positive constant.
+
+# Aside: `hyper2` analysis of the same scoreline
+
+Now the likelihood function would be
+
+$$\mathcal{L}(p_1,p_2,\lambda;A,B,C,D)=
+\underbrace{
+\overbrace{\left(\frac{p_1+M}{p_1 + p_2+M}\right)^{A}}^{\mbox{$A$ home wins for $p_1$}}\quad
+\overbrace{\left(\frac{p_2  }{p_1 + p_2+M}\right)^{B}}^{\mbox{$B$ away wins for $p_2$}}
+}_\mbox{$p_1$ plays at home to $p_2$ $A+B$ times}\quad
+\underbrace{
+\overbrace{\left(\frac{p_2+M}{p_1 + p_2+M}\right)^{C}}^{\mbox{$C$ home wins for $p_2$}}\quad
+\overbrace{\left(\frac{p_1  }{p_1 + p_2+M}\right)^{D}}^{\mbox{$D$ away wins for $p_1$}}
+}_\mbox{$p_2$ plays at home to $p_1$ $C+D$ times}
+$$
+
+The maximum would be
+
+$$
+\hat{p_1}=\frac{D}{C+D}\qquad
+\hat{p_2}=\frac{B}{A+B}\qquad
+\hat{M}=\frac{AC-BD}{(A+B)(C+D)}
+$$
+
+although it is not clear to me what the _constrained_ optimal point
+[that is, requiring $M\geq 0$] is.
+
+## Numerical analysis
+
+```{r label=defineABCD}
+A <- 9
+B <- 5
+C <- 7
+D <- 3
+
+(p1hat <- sqrt(A*D)/(sqrt(A*D) + sqrt(B*C)))
+(p2hat <- sqrt(B*C)/(sqrt(A*D) + sqrt(B*C)))
+
+(lam_hat <- sqrt((A*C)/(B*D)))
+```
+
+$$
+\hat{p_1}=\frac{\sqrt{21}}{\sqrt{21} + \sqrt{35}}\simeq 0.47\qquad
+\hat{p_2}=\frac{\sqrt{63}}{\sqrt{21} + \sqrt{35}}\simeq 0.53\qquad
+\hat{\lambda}=\sqrt{\frac{AC}{BD}}=2.05.
+$$
+
+Moving to the monster formulation we have
+
+$$
+\hat{p_1}=\frac{3}{7+3} = 0.3\qquad
+\hat{p_2}=\frac{5}{5+9}\simeq  0.36\qquad
+\hat{M}=\frac{63-21}{14\cdot 10} = 0.3
+$$
+
+Keeping $A=9,B=5, C=7,D=3$, and $\lambda=2.05$ [assumed for the moment
+to be known, estimation techniques are discussed later], we can
+represent this information in a standard format:
+
+```{r,label=usehomeaway3a}
+library("hyper2",quietly=TRUE)
+M <- matrix(c(NA, A+B*1i ,C+D*1i, NA),2,2,byrow=TRUE)
+teams <- c("p1","p2")
 dimnames(M) <- list("@home" = teams,"@away"=teams)
 dimnames(M) <- list("@home" = teams,"@away"=teams)
-M
-home_away(M)
+print(M)
 ```
 
-Above we see (from `M[1,2]`) that A played at home $1+5=6$ times,
-winning 1 and losing 5; and B played at home $3+11=14$ times, winning
-3 and losing 11.   A slightly more complicated example:
+Above, object `M` has real parts being home wins and imaginary parts
+being away wins.  Appropriate package idiom to specify a likelihood
+function might be to use bespoke function `home_away3()`:
+
+```{r,label=ha3calc,cache=TRUE}
+(H <- home_away3(M,lambda=2.05))
+```
+
+Keeping $A=9,B=5, C=7,D=3$, and $\lambda=2.05$ [assumed for the moment
+to be known, estimation techniques are discussed later], we may
+estimate $p_1$ and $p_2$ using maximum likelihood, `maxp()` in package
+idiom:
+
+```{r,label=findtheevaluateh3,cache=TRUE}
+H
+maxp(H)
+```
+
+Further, we can test whether the players are in fact of equal strength:
+
+```{r,label=testequalitynow,cache=TRUE}
+equalp.test(H)
+```
+
+Showing convincingly that we may reject the null that $p_1=p_2$ and
+assert that the players do in fact differ in strength, at least with
+this value of $\lambda$.  Observe that a naive analysis would have
+$p_1$ winning $w=A+C=`r A+C`$ games and losing $l=B+D=`r B+D`$ games,
+from which we would obtain $\hat{p_1}=`r A+C`/`r A+B+C+D`$; a
+likelihood ratio for $H_0\colon p_1=0.5$ would be
+
+$$\frac
+{
+  {{w+l}\choose {w\, l}}\left(\frac{w}{w+l}\right)^w\left(\frac{l}{w+l}\right)^l
+}{
+  {{w+l}\choose {w\, l}}\left(\frac{1}{2}\right)^w\left(\frac{1}{2}\right)^l
+}=\frac{2^{w+l}w^wl^l}{(w+l)^{w+l}}\simeq 1.94,
+$$
+
+not significant.
+
+Now, how to estimate $\lambda$?
+
+```{r,label=estimatelambda3,cache=TRUE}
+f <- function(lambda){maxp(home_away3(M,lambda=lambda),give=TRUE)$value}
+f(2)
+f(2.1)
+lam <- seq(from=1,to=10,len=17)
+Supp <- sapply(lam,f)
+```
+
+```{r,label=plotsup}
+plot(log(lam),Supp-max(Supp),type="b")
+```
+
+Or even
+
+```{r,label=optimlambda,cache=TRUE}
+op <- optimize(f,interval=c(2,3),maximum=TRUE)
+op
+```
+
+We can proceed in two ways.  Firstly we can use $\hat{\lambda}$ directly:
+
+```{r,ept3,cache=TRUE}
+equalp.test(home_away3(M,lambda=op$maximum))
+```
+
+The flaw in this analysis is that it is conditional on the estimated
+value of $\lambda$ [which is not too bad, IMO].  Secondly we can
+perform a fully 2D analysis, which allows for the fact that the
+evaluate might have a different value of $\lambda$ from the
+$\hat{\lambda}$:
+
+```{r,label=loopysuppcalc,cache=TRUE}
+probs <- seq(from=0.1,to=0.8,len=17)
+jj <- as.matrix(expand.grid(lam,probs))
+S <- rep(NA,nrow(jj))
+for(i in seq_len(nrow(jj))){
+    lambda <- jj[i,1]
+    p1 <- jj[i,2]
+    p <- c(p1,1-p1)
+    names(p) <- c("p1","p2")
+   S[i] <- loglik(p=p,H=home_away3(M,lambda=lambda),log=TRUE)
+}
+S <- matrix(S,length(lam),length(probs))
+```
+
+```{r}
+contour(lam,probs,S,levels=seq(from=-20,to=-10,by=0.5),xlab="lambda",ylab="p1")
+abline(h=0.5)
+```
+
+#  Ternary representation
+
+```{r,label=useternary}
+library(Ternary)
+
+par(mar = rep(0.2, 4))
+TernaryPlot(alab = "p1", blab = "p2", clab = "M")
+
+f <- function(p1,p2,M) {
+pmax(-20,log(M+p1)*A + log(p2)*B - log(M+p1+p2)*(A+B+C+D) + log(p2+M)*C + log(p1)*D )
+}
+FunctionToContour <- function(a,b,c){f(p1=a, p2=b, M=c)}
+
+# Compute and plot colour tiles
+values <- TernaryPointValues(FunctionToContour, resolution = 24L)
+ColourTernary(values)
+
+# Add contour lines
+TernaryContour(FunctionToContour, resolution = 36L)
+```
+
+A slightly more complicated example:
 
@@
 pie(baseball_maxp)
+
 ```
@@
 
+# Appendix
+
+Here is a transcript of a mathematica session which establishes the
+estimates above.  First the `hyper2` likelihood:
+
+```
+rhankin@rhrpi4:~ $ wolfram
+Mathematica 12.2.0 Kernel for Linux ARM (32-bit)
+Copyright 1988-2021 Wolfram Research, Inc.
+
+In[1]:= L = Log[M+p1]*A +Log[p2]*B- Log[M+p1+p2]*(A+B+C+D) + Log[p2+M]*C + Log[p1]*D
+
+Out[1]= D Log[p1] + A Log[M + p1] + B Log[p2] + C Log[M + p2] - (A + B + C + D) Log[M + p1 + p2]
+
+In[2]:= LL = L /. {p2 -> 1-p1-M}
+
+Out[2]= C Log[1 - p1] + B Log[1 - M - p1] + D Log[p1] + A Log[M + p1]
+
+In[3]:= dp = D[LL,p1]//FullSimplify
+
+            C      D         B          A
+Out[3]=  ------- + -- + ----------- + ------
+         -1 + p1   p1   -1 + M + p1   M + p1
+
+In[4]:= dM = D[LL,M]//FullSimplify
+
+              B          A
+Out[4]= ----------- + ------
+         -1 + M + p1   M + p1
+
+In[5]:= Solve[{dp==0,dM==0},{p1,M}]
+
+                   D            A C - B D
+Out[5]=  {{p1 -> -----, M -> ---------------}}
+                 C + D       (A + B) (C + D)
+
+In[6]:= p2 = 1-p1-M /. {%} //FullSimplify
+
+             B
+Out[6]=  {{-----}}
+           A + B
+
+```
+
+Now the `hyper3` likelihood:
+
+```
+rhankin@rhrpi4:~ $ wolfram
+Mathematica 12.2.0 Kernel for Linux ARM (32-bit)
+Copyright 1988-2021 Wolfram Research, Inc.
+
+In[1]:= L = A*Log[lambda*p1] + B*Log[p2] -(A+B)*Log[lambda*p1 + p2] + C*Log[lambda*p2] + D*Log[p1] -(C+D)*Log[p1 + lambda*p2]
+
+Out[1]= D Log[p1] + A Log[lambda p1] + B Log[p2] + C Log[lambda p2] - (A + B) Log[lambda p1 + p2] - (C + D) Log[p1 + lambda p2]
+
+In[2]:= LL = L /. {p2 -> 1-p1}
+
+Out[2]= B Log[1 - p1] + C Log[lambda (1 - p1)] + D Log[p1] + A Log[lambda p1] - (C + D) Log[lambda (1 - p1) + p1] -
+
+>    (A + B) Log[1 - p1 + lambda p1]
+
+In[3]:= dp1 = D[LL,p1] //FullSimplify
+
+         B + C    A + D   (A + B) (-1 + lambda)    (C + D) (-1 + lambda)
+Out[3]= ------- + ----- - --------------------- + -----------------------
+        -1 + p1    p1     1 + (-1 + lambda) p1    lambda + p1 - lambda p1
+
+In[4]:= dlam = D[LL,lambda]//FullSimplify
+
+          A        C           (A + B) p1           (C + D) (-1 + p1)
+Out[4]= ------ + ------ - -------------------- + -----------------------
+        lambda   lambda   1 + (-1 + lambda) p1   lambda + p1 - lambda p1
+
+In[5]:= Solve[{dp1==0,dlam==0},{p1,lambda}]//FullSimplify
+
+[snip]
+
+In[6]:= %[[1]]
+
+                        1                     Sqrt[A] Sqrt[C]
+Out[6]= {p1 -> -------------------, lambda -> ---------------}
+                   Sqrt[B] Sqrt[C]            Sqrt[B] Sqrt[D]
+               1 + ---------------
+                   Sqrt[A] Sqrt[D]
+
+```
+
 ###  References

### 2025-02-11 inst/home_advantage.Rmd
slight rewording

@@
 
-Showing convincingly that we may reject the null that $p_1=p_2$ and
-assert that the players do in fact differ in strength, at least with
+Showing that we fail to reject the null that $p_1=p_2$, at least with
 this value of $\lambda$.  Observe that a naive analysis would have
@@
 
-Agresti considers a dataset of seven baseball teams who play one
-another repeatedly.  Each game is played at the home ground of one
-team, the other playing away.  The dataset is as follows:
+@agresti2002 (p438) considers a dataset of seven baseball teams who
+play one another repeatedly.  Each game is played at the home ground
+of one team, the other playing away.  The dataset is as follows:
 
@@
 M <- baseball_table # saves typing
+M # for convenience
 ```

### 2025-02-11 inst/eurovision.Rmd
consistentification
distinguish hankin2017 from hankin2017_rnw
time to process without cache

@@
 author: "R. K. S. Hankin"
-output: bookdown::html_document2
-bibliography: jss_hyper3.bib
+bibliography: hyper2.bib  # copy vignettes/hyper2.bib to inst/
 ---
@@
 
-To cite the `hyper2` package in publications, please use @hankin2017a.
-The file `eurovision.txt`, used below, is copied from "Eurovision Song
-Contest 2009," Wikipedia, accessed May 13, 2018.  It refers to
-semi-final 1.  More documentation is given in `eurovision.Rd` [type
-`help(euro2009)` at the `R` prompt].
+ <font size=1.4px> (takes about 30 minutes to run without cache, most
+of the time taken processing the `maximi` chunk)</font>
+
+To cite the `hyper2` package in publications, please use
+@hankin2017_rmd.  The file `eurovision.txt`, used below, is copied
+from "Eurovision Song Contest 2009," Wikipedia, accessed May 13, 2018.
+It refers to semi-final 1.  More documentation is given in
+`eurovision.Rd` [type `help(euro2009)` at the `R` prompt].
 

### 2025-02-11 inst/sushi.Rmd
sushi

---
title: "Sushi monster"
author: "R. K. S. Hankin"
output: bookdown::html_document2
bibliography: hyper2.bib  # copy vignettes/hyper2.bib to inst/
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("partitions")
library("pdftools")
```

```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
```

To cite the `hyper2` package in publications, please use @hankin2017_rmd.

```{r}
a <- as.matrix(read.table("sushi.txt"))
```

```{r,defcalculate_sushi_H}
make_sushi_H <- function(a, strength){
    h <- c(1, strength) # strength of support
    sushi_types <- c("ebi", "anago", "maguro", "ika", "uni", "sake", "tamago", "toro", "tekka", "kappa")
    ec <- # equivalence classes
        c(`ebi`  = 1, `anago` = 2, `maguro` = 1, `ika` = 1, `uni` = 1,
          `sake` = 1, `tamago` = 1, `toro` = 2, `tekka-maki` = 2,
          `kappa-maki` = 1)

    H <- hyper3()

    for(i in seq_len(nrow(a))){
        H <- H + cheering3(sushi_types[a[i,]], e=ec, h=h)
    }
    return(H)
}
```

```{r tryH, cache=TRUE}
wanted <- sample(seq_len(nrow(a)), 10, replace=TRUE, prob=a[,1])
H1888 <- make_sushi_H(a[wanted,-1], strength=1.888)
H1000 <- make_sushi_H(a[wanted,-1], strength=1.000)
```

```{r maxpsushiH,cache=TRUE}
mHa <- maxp(H1888,give=1)
mHb <- maxp(H1000,give=1)
```

```{r showsushiH}
mHa
mHb
```

## References {-}

### 2025-02-11 inst/home_advantage.Rmd

@@
 
+### Analysis of  Work by Davidson and Beaver
+
+```{r}
+DB <- # Davidson and Beaver
+matrix(c(
+    NA, 32+18i, 36+14i, 47+03i, 47+03i,
+14+36i,     NA, 34+16i, 43+07i, 46+04i,
+06+44i, 14+36i,     NA, 34+16i, 40+10i,
+02+48i, 07+43i, 12+38i,     NA, 28+22i,
+01+49i, 02+48i, 05+45i, 10+40i,     NA
+),5,5,byrow=TRUE
+)
+jj <- c("90g","95g","100g","105g","110g")
+dimnames(DB) <- list(`@home` = jj, `@away` = jj)
+DB
+```
+
+```{r first_second_likefun, cache=TRUE}
+f <- function(l){ maxp(home_away3(DB,l),give=TRUE,n=1)$likes}
+lam <- seq(from=0.6,to=1,len=19)
+date()
+loglike <- sapply(lam,f)
+date()
+```
+
+```{r plotfirstsecond}
+plot(lam,loglike-max(loglike),type='b')
+abline(h=c(0,-2))
+abline(v=1)
+```
+
 # Creating a home/away results table from scratch

### 2025-02-11 inst/red_bus_blue_bus.Rmd

@@
 
+To cite the `hyper2` package in publications, please use
+@hankin2017_rmd.
+
 In this short discussion document I discuss three related problems:
@@
 The red bus-blue bus problem was originally stated by @chipman1960 and
-discussed by @davidson1970 and @alvo2010 (page 153).
-Debussy-Beethoven was @debreu1960.  I have lost the original reference
-to the Holiday problem.
+discussed by @davidson1970 and @alvo2014 (page 153).
+Debussy-Beethoven was @debreu1960.  The Holiday problem is discussed
+in @tverskey1972 (in the form of trips to Paris or Rome), attributed
+to "L. J. Savage"; in this document I use Florida and California,
+which I did not make up, but cannot find the original reference.
 

### 2025-02-11 inst/sushi.Rmd
remove unnecessary packages
consistentification

@@
 library("hyper2")
-library("partitions")
-library("pdftools")
 ```
@@
 
+```{r calclike, cache=TRUE}
+f <- function(strength){
+   H <- make_sushi_H(a[wanted,-1], strength=strength)
+   maxp(H,give=1)$`Log-likelihood`
+}
+s <- seq(from=0.4,to=2,len=8)
+L <- sapply(s,f)
+```
+
+```{r}
+plot(s,L-max(L),type='b')
+```
+
 ## References {-}

### 2025-02-13 inst/universities.Rmd
consistentification
distinguish hankin2017 from hankin2017_rnw
use better magrittr idiom

@@
 author: "R. K. S. Hankin"
-bibliography: jss_hyper3.bib
+bibliography: hyper2.bib # copy from vignettes/
 ---
@@
 
-To cite the `hyper2` package in publications, please use @hankin2017a.
+To cite the `hyper2` package in publications, please use @hankin2017_rmd.
 Here I analyse a dataset on NZ university rankings, taken from the
@@
 universities <- hyper2(pnames=universities_table[1:8,1])
-for(i in 2013:2020){universities %<>% `+`(ordertable2supp(f(universities_table,i)))}
+for(i in 2013:2020){universities %<>% add(ordertable2supp(f(universities_table,i)))}
 universities

### 2025-02-13 inst/home_advantage.Rmd
Rangers and Celtic

@@
 o
-scot_max_lambda <- o$maximum
-maxL <- f(scot_max_lambda)
+RCLF3_lambda_max <- o$maximum
+maxL <- f(RCLF3_lambda_max)
 
@@
 ```{r scotmax}
-(H <- home_away3(home_away_table(a,give=FALSE),scot_max_lambda))  # o$maximum ~= 1.353
+(H <- home_away3(home_away_table(a,give=FALSE),RCLF3_lambda_max))  # o$maximum ~= 1.353
 maxp(H,give=1)
@@
 ```{r sfasdf, cache=TRUE}
-o$maximum
-mH3 <- maxp(home_away3(home_away_table(a,give=FALSE),scot_max_lambda),give=FALSE)
-round(mH3*1000)
+RCLF3_lambda_max <- o$maximum
+RCLF3_max <- maxp(home_away3(home_away_table(a,give=FALSE),RCLF3_lambda_max),give=FALSE)
+round(RCLF3_max * 1000)
 ```
@@
 mH["home"]
-mH3
-lambda <- scot_max_lambda
+RCLF3_max
+lambda <- RCLF3_lambda_max
 h2_exp <- matrix(0,4,4)
@@
         h3_exp[hometeam,awayteam] <- (
-            n_matches[hometeam,awayteam] * (lambda* mH3[hometeam])/
-            (lambda* mH3[hometeam] + mH3[awayteam])
+            n_matches[hometeam,awayteam] * (lambda* RCLF3_max[hometeam])/
+            (lambda* RCLF3_max[hometeam] + RCLF3_max[awayteam])
         )
@@
             add(1i*
-                n_matches[hometeam,awayteam] * (mH3[awayteam])/
-                (lambda* mH3[hometeam] + mH3[awayteam])
+                n_matches[hometeam,awayteam] * (RCLF3_max[awayteam])/
+                (lambda* RCLF3_max[hometeam] + RCLF3_max[awayteam])
         )
@@
             h3_exp[hometeam,awayteam] <- (
-                n_matches[hometeam,awayteam] * (lambda* mH3[hometeam])/
-                (lambda* mH3[hometeam] + mH3[awayteam])
+                n_matches[hometeam,awayteam] * (lambda* RCLF3_max[hometeam])/
+                (lambda* RCLF3_max[hometeam] + RCLF3_max[awayteam])
             )
@@
                 add(1i*
-                    n_matches[hometeam,awayteam] * (mH3[awayteam])/
-                    (lambda* mH3[hometeam] + mH3[awayteam])
+                    n_matches[hometeam,awayteam] * (RCLF3_max[awayteam])/
+                    (lambda* RCLF3_max[hometeam] + RCLF3_max[awayteam])
                     )
@@
 
-round(expectation3(mH3[c("Rangers","Celtic","Falkirk","Livingston")],aF,scot_max_lambda),1)
+round(expectation3(RCLF3_max[c("Rangers","Celtic","Falkirk","Livingston")],aF,RCLF3_lambda_max),1)
 ```
@@
 
-Following lines create `baseball.rda`, residing in the `data/`
+Following lines create `RCLF.rda`, residing in the `data/`
 directory of the package.
@@
 ```{r,label=savesolingdataset}
-save(baseball_table, baseball, baseball_maxp, 
-  file="baseball.rda")
+RCLF3_table <- home_away_table(a, give=TRUE, teams = teams)
+
+save(RCLF3_table,  RCLF3_max, RCLF3_lambda_max, file="RCLF.rda")
 ```

### 2025-02-13 inst/home_advantage.Rmd
consistentification

@@
 maxp(H,give=1)
+RCLF3 <- H
 ```
@@
 
-save(RCLF3_table,  RCLF3_max, RCLF3_lambda_max, file="RCLF.rda")
+save(RCLF3_table,  RCLF3, RCLF3_max, RCLF3_lambda_max, file="RCLF.rda")
 ```

### 2025-02-13 inst/sushi.Rmd
better range for lambda

@@
 ```{r tryH, cache=TRUE}
-wanted <- sample(seq_len(nrow(a)), 10, replace=TRUE, prob=a[,1])
+wanted <- sample(seq_len(nrow(a)), 500, replace=TRUE, prob=a[,1])
 H1888 <- make_sushi_H(a[wanted,-1], strength=1.888)
@@
 
-```{r calclike, cache=TRUE}
+```{r calclikesushi, cache=TRUE}
 f <- function(strength){
@@
 }
-s <- seq(from=0.4,to=2,len=8)
+s <- seq(from=1.4,to=7.5,len=5)
 L <- sapply(s,f)

### 2025-02-14 inst/home_advantage.Rmd
consistentification

@@
 RCLF3_lambda_max <- o$maximum
-RCLF3_max <- maxp(home_away3(home_away_table(a,give=FALSE),RCLF3_lambda_max),give=FALSE)
-round(RCLF3_max * 1000)
+RCLF3_maxp <- maxp(home_away3(home_away_table(a,give=FALSE),RCLF3_lambda_max),give=FALSE)
+round(RCLF3_maxp * 1000)
 ```
@@
 mH["home"]
-RCLF3_max
+RCLF3_maxp
 lambda <- RCLF3_lambda_max
@@
         h3_exp[hometeam,awayteam] <- (
-            n_matches[hometeam,awayteam] * (lambda* RCLF3_max[hometeam])/
-            (lambda* RCLF3_max[hometeam] + RCLF3_max[awayteam])
+            n_matches[hometeam,awayteam] * (lambda* RCLF3_maxp[hometeam])/
+            (lambda* RCLF3_maxp[hometeam] + RCLF3_maxp[awayteam])
         )
@@
             add(1i*
-                n_matches[hometeam,awayteam] * (RCLF3_max[awayteam])/
-                (lambda* RCLF3_max[hometeam] + RCLF3_max[awayteam])
+                n_matches[hometeam,awayteam] * (RCLF3_maxp[awayteam])/
+                (lambda* RCLF3_maxp[hometeam] + RCLF3_maxp[awayteam])
         )
@@
             h3_exp[hometeam,awayteam] <- (
-                n_matches[hometeam,awayteam] * (lambda* RCLF3_max[hometeam])/
-                (lambda* RCLF3_max[hometeam] + RCLF3_max[awayteam])
+                n_matches[hometeam,awayteam] * (lambda* RCLF3_maxp[hometeam])/
+                (lambda* RCLF3_maxp[hometeam] + RCLF3_maxp[awayteam])
             )
@@
                 add(1i*
-                    n_matches[hometeam,awayteam] * (RCLF3_max[awayteam])/
-                    (lambda* RCLF3_max[hometeam] + RCLF3_max[awayteam])
+                    n_matches[hometeam,awayteam] * (RCLF3_maxp[awayteam])/
+                    (lambda* RCLF3_maxp[hometeam] + RCLF3_maxp[awayteam])
                     )
@@
 
-round(expectation3(RCLF3_max[c("Rangers","Celtic","Falkirk","Livingston")],aF,RCLF3_lambda_max),1)
+round(expectation3(RCLF3_maxp[c("Rangers","Celtic","Falkirk","Livingston")],aF,RCLF3_lambda_max),1)
 ```
@@
 
-save(RCLF3_table,  RCLF3, RCLF3_max, RCLF3_lambda_max, file="RCLF.rda")
+save(RCLF3_table,  RCLF3, RCLF3_maxp, RCLF3_lambda_max, file="RCLF.rda")
 ```

### 2025-02-14 inst/home_advantage.Rmd
home_draw_away3()

@@
 
-<font size="1"> (takes about fifteen minutes to run without cache)
-</font>
+<font size="1"> most of the time is taken calculating the contour
+graph right at the end; takes about an hour to run without cache for
+n=10 </font>
 
@@
 
-(from here on we ignore draws and consider only won games)
+(We will ignore draws for the moment, and consider only won games)
 
@@
 
+# Bring back draws
+
+```{r, bbdlambdaD,cache=TRUE}
+aT
+home_draw_away3(aT,lambda=1.888,D=0.3)
+f <- function(v){
+    H <- home_draw_away3(aT, lambda = v[1], D=v[2])
+    maxp(H,give=1)$`Log-likelihood`
+}
+n <- 20
+lambda_try <- seq(from=1.6,to=4,len=n)
+D_try <- seq(from=0.3,to=1,len=n)
+V <- as.matrix(expand.grid(lambda_try,D_try))
+LL <- apply(V,1,f)
+```
+
+```{r plotLL}
+LL <- matrix(LL,n,n) - max(LL)
+LL <- pmax(LL,-10)
+contour(lambda_try, D_try, LL,xlab="lambda",ylab="D")
+```
+
 # Appendix

### 2025-02-14 inst/home_advantage.Rmd
work

@@
 ```{r,label=loopysuppcalc,cache=TRUE}
-probs <- seq(from=0.1,to=0.8,len=17)
+probs <- seq(from=0.1,to=0.8,len=2)
 jj <- as.matrix(expand.grid(lam,probs))
@@
 n <- 20
-lambda_try <- seq(from=1.6,to=4,len=n)
-D_try <- seq(from=0.3,to=1,len=n)
+lambda_try <- seq(from=1,to=3,len=n)
+D_try <- seq(from=0.1,to=0.6,len=n)
 V <- as.matrix(expand.grid(lambda_try,D_try))
@@
 contour(lambda_try, D_try, LL,xlab="lambda",ylab="D")
+```
+
+Now find the evaluate:
+
+```{r findtheevaluate, cache=TRUE}
+maxv <- optim(par=c(3,0.7), fn=f, control=list(fnscale = -1))
+```
+
+```{r showtheevaluate}
+maxv
+```
+
+```{r usetheevaluatefindmax, cache=TRUE}
+MLparams <- maxv$par
+mH <- maxp(home_draw_away3(aT, lambda = MLparams[1], D=MLparams[2]),give=1)
+```
+
+```{r}
+mH
 ```

### 2025-02-16 inst/home_advantage.Rmd
whitespace and label all chunks

@@
 
-```{r,label=usehomeaway3a}
+```{r, label=usehomeaway3a}
 library("hyper2",quietly=TRUE)
@@
 
-```{r,label=ha3calc,cache=TRUE}
+```{r, label=ha3calc, cache=TRUE}
 (H <- home_away3(M,lambda=2.05))
@@
 
-```{r,label=findtheevaluateh3,cache=TRUE}
+```{r, label=findtheevaluateh3, cache=TRUE}
 H
@@
 
-```{r,label=testequalitynow,cache=TRUE}
+```{r, label=testequalitynow, cache=TRUE}
 equalp.test(H)
@@
 
-```{r,label=estimatelambda3,cache=TRUE}
+```{r, label=estimatelambda3, cache=TRUE}
 f <- function(lambda){maxp(home_away3(M,lambda=lambda),give=TRUE)$value}
@@
 
-```{r,label=plotsup}
+```{r, label=plotsup}
 plot(log(lam),Supp-max(Supp),type="b")
@@
 
-```{r,label=optimlambda,cache=TRUE}
+```{r, label=optimlambda, cache=TRUE}
 op <- optimize(f,interval=c(2,3),maximum=TRUE)
@@
 
-```{r,ept3,cache=TRUE}
+```{r, ept3, cache=TRUE}
 equalp.test(home_away3(M,lambda=op$maximum))
@@
 
-```{r,label=loopysuppcalc,cache=TRUE}
+```{r, label=loopysuppcalc, cache=TRUE}
 probs <- seq(from=0.1,to=0.8,len=2)
@@
 
-```{r}
+```{r contlam}
 contour(lam,probs,S,levels=seq(from=-20,to=-10,by=0.5),xlab="lambda",ylab="p1")
@@
 
-```{r,label=useternary}
+```{r, label=useternary}
 library(Ternary)
@@
 
-```{r,label=genthedata}
+```{r, label=genthedata}
 home_games_won <- matrix(c(
@@
 
-```{r,label=showhomeawayusingcomplex}
+```{r, label=showhomeawayusingcomplex}
 home_games_won + 1i*away_games_won
@@
 
-```{r,label=makehyper2,cache=TRUE}
+```{r, label=makehyper2, cache=TRUE}
 H <- home_away(home_games_won,away_games_won)
@@
 
-```{r,label=homegroundAgresti,cache=TRUE}
+```{r, label=homegroundAgresti, cache=TRUE}
 samep.test(H,teams)
@@
 
-```{r,label=showvisH,cache=TRUE}
+```{r, label=showvisH, cache=TRUE}
 (mt <- maxp(H))
@@
 
-```{r,label=showprofilelikelihood, cache=TRUE}
+```{r, label=showprofilelikelihood, cache=TRUE}
 home_strength <- seq(from=0.15,to=0.45,length=13)
@@
 
-```{r,label=makeAgrestimatrix}
+```{r, label=makeAgrestimatrix}
 baseball_table <- matrix(c(
@@
 
-```{r,label=makesuppfrommatrix}
+```{r, label=makesuppfrommatrix, cache=TRUE}
 baseball <- home_away(M)
@@
 
-```{r,label=calculatemaxpAgresti,cache=TRUE}
+```{r, label=calculatemaxpAgresti, cache=TRUE}
 baseball_maxp <- maxp(baseball)
@@
 
-```{r,label=showmaxpAgresti}
+```{r, label=showmaxpAgresti}
 dotchart(baseball_maxp,pch=16)
@@
 
-```{r,analyseAgresti,cache=TRUE}
+```{r, analyseAgresti, cache=TRUE}
 specificp.gt.test(baseball,"home",0)
@@
 
-```{r,Agrestiprofsupp,cache=TRUE}
+```{r, Agrestiprofsupp, cache=TRUE}
 home_strength <- seq(from=0.005,to=0.1,length=20)
@@
 
-```{r,label=testhomegroundadvantage,cache=TRUE}
+```{r, label=testhomegroundadvantage, cache=TRUE}
 samep.test(baseball,teams)
@@
 
-```{r}
+```{r, label=defdavidsonbeaver}
 DB <- # Davidson and Beaver
@@
 
-```{r scotmax}
+```{r scotmax, cache=TRUE}
 (H <- home_away3(home_away_table(a,give=FALSE),RCLF3_lambda_max))  # o$maximum ~= 1.353
@@
 
-```{r}
+```{r plotlamloglike}
 plot(lam,loglike-max(loglike),type='b')
@@
 
-```{r}
+```{r, label=oldfirmexpectation}
 mH
@@
 }
-n <- 20
+n <- 2
 lambda_try <- seq(from=1,to=3,len=n)
@@
 
-```{r}
+```{r, label=showmH}
 mH

### 2025-02-16 inst/ternaryplot_hyper2.Rmd
include ternary contour plot of chess

@@
 higher-dimensional `hyper2` objects.
+
+```{r plotternarychess}
+FunctionToContour <- function (a, b, c) {
+    M <- cbind(a, b, c)
+    colnames(M) <- c("Topalov", "Anand", "Karpov")
+    M[] <- pmax(M,0)
+    M <- sweep(M,1,rowSums(M),"/")
+    out <- loglik(M, chess)
+    out[is.infinite(out)] <- NA
+    out <- out - max(out,na.rm = TRUE)
+    out <- pmax(out, -20, na.rm = TRUE)
+    return(out)
+}
+
+originalPar <- par(mar = rep(0, 4))
+TernaryPlot(alab = "Topalov", blab = "Anand", clab = "Karpov")
+values <- TernaryPointValues(FunctionToContour, resolution = 24L)
+ColourTernary(
+  values,
+  legend = signif(seq(max(values), min(values), length.out = 4), 2),
+  bty = "n"
+)
+TernaryContour(FunctionToContour, resolution = 72L, nlevels=10)
+```
 

### 2025-02-16 inst/sushi.Rmd
how long to run without cache
better range

@@
 ```
+
+<font size="1"> most of the time is taken calculating the
+`calclikesushi` chunk.  With a sample size of 500 in chunk `tryH`, and
+`n=5` in `calclikesushi`, it takes about half an hour to run without
+cache. </font>
 
@@
 }
-s <- seq(from=1.4,to=7.5,len=5)
+s <- seq(from=1.4,to=4,len=8)
 L <- sapply(s,f)
@@
 
+```{r label=justtryf1, cache=TRUE}
+f1 <- f(1)
+```
+
+```{r label=showf1}
+f1 - max(L)
+```
+
 ## References {-}

### 2025-02-16 inst/cheering.Rmd
cheering() in use

--- 2024-01-26 inst/plackett_luce_monster.Rmd
@@
 ---
-title: "A generalization of Plackett-Luce likelihoods to incorporate non-independence of competitors"
+title: "Function `cheering()` in the `hyper2` package"
 author: "R. K. S. Hankin"
@@
 To cite the `hyper2` package in publications, please use @hankin2017_rmd.
-Here I modify standard Plackett-Luce likelihood functions to include
-an element of non-independence among the competitors.  There seem to
-be a few distinct statistical models for nonindependence:
 
-* Runner $i$ might wish to support $j$.  If runner $i$ finishes, once
-  his rank is established he might be able to help $j$ in some way.
-  He might cheer his still-running friend, spurring him to run faster
-  and augmenting his competitive strength.
+This document is a distillation and clarification of
+`inst/plackett_luce_monster.Rmd`.
 
-* In general, we can consider $n$ ordered pairs $(i_r,j_r)$, $1\leq
-  r\leq n$, with runner $i_1$ helping runner $j_1$, runner $i_2$
-  helping $j_2$, and so on.  One simple case would be equivalence
-  classes of $[n]=\{1,2,\ldots,n\}$ with each equivalence class
-  comprising mutually supporting runners.  In this model, with a pair
-  of mutually supporting runners only the higher-placed competitor can
-  affect the other.  Each equivalence class might have a different
-  supportive proclivity.
+In the following, I use the ''race'' metaphor: five runners take part
+in a race and arrive in order: $a\succ b\succ c\succ d\succ 5e.
+Consider the following Plackett-Luce likelihood function for this
+observation:
 
-* Runner $i$ might hate runner $j$.  If $i$ finishes, he might expend
-  his spare energy hindering $j$ (by throwing stones or booing, for
-  example).  Note that equivalence classes of mutual hatred might
-  exist.
+\begin{equation}
+\frac{p_a}{p_a + p_b + p_c + p_d + p_e}\cdot
+\frac{p_b}{      p_b + p_c + p_d + p_e}\cdot
+\frac{p_c}{            p_c + p_d + p_e}\cdot
+\frac{p_d}{                  p_d + p_e}\cdot
+\frac{p_e}{                        p_e}
+\end{equation}
 
-* Also, the help or hindrance might have a finite lifetime.  Under one
-  model, if $i$ finishes then he can stay helping (or hindering) $j$
-  wherever $j$ places.  Alternatively, suppose $i$ places $n$th.  Then
-  perhaps he can affect $j$ only place $n+1$, and after a competitor
-  finishes in place $n+1$, then $i$ can exert no further influence.
+Note the use of multiplication between the terms, indicating
+conditional independence.  We may use the multiplicative
+generalization of Bradley-Terry strengths to introduce an element of
+non-independence between the terms.  There are many ways to do this,
+but one simple case would be to define equivalence classes of the
+competitors with each equivalence class comprising mutually supporting
+runners.  Equivalence classes of size one correspond to unsupported
+runners.  The metaphor would be that a runner who has finished the
+race is able to support other members of his equivalence class by
+cheering his teammates, boosting their better performance.  The
+`hyper2` package includes function `cheering()` which implements this
+functionality.
 
-* The help or hindrance discussed above might be contingent on
-  placing: perhaps only the winnner (that is, first-placing
-  competitor) would be able to influence other competitors.  Or maybe
-  competitors with any placing could influence other competitors.
+As an example, consider figure 0.1 This shows a partial probability
+tree diagram for some of the $5!=120$ possible order statistics.  The
+standard Plackett-Luce likelihoods have been modified to account for
+two groups of mutually supporting runners: $\left\lbrace
+a,b,c\right\rbrace$ with support term $\lambda$, and $\left\lbrace
+d,e\right\rbrace$ with support term $\mu$.  From START, the first
+runner to cross the finishing line is associated with standard
+Plackett-Luce.  Taking the top path as an example, we see that the
+likelihood function for $a\succ b\succ c\succ d\succ e$ would be
 
-* Runners $i$ and $j$ might elect to cross the finishing line holding
-  hands, with the apparent tie broken randomly.  This would be a
-  symmetrical relationship.  In general, we could consider
-  hand-holding equivalence classes.
-
-* We might have a resource allocation problem, with runners having a
-  certain [possibly unknown] amount of "strength" but also the ability
-  to expend this strength on objectives other than ranking in the
-  race.  A runner might be able to expend just enough energy to remain
-  slightly ahead of a rival, reserving energy for other purposes.  If
-  runner $i$ wishes to beat runner $j$, and has sufficient resources
-  to do so, then runners $i$ and $j$ will finish consecutively with
-  $i$ ahead of $j$.
-
-# Start simple: three runners
-
-The simplest nontrivial case is three runners, $\left\lbrace
-1,2,3\right\rbrace$ where runner 1 helps runner 2 with multiplier
-$\lambda$ [and that is the only non PL interaction].  Note that this
-is _not_ a symmetric interaction: 1 helps 2, but 2 does not help 1.
-The model is that 1 helps 2 if and only if 1 has finished and 2 has
-not.
-
-123: $\frac{p_1}{p_1+p_2+p_3}\cdot\frac{\lambda p_2}{\lambda p_2+p_3}$
-
-132: $\frac{p_1}{p_1+p_2+p_3}\cdot\frac{p_3}{\lambda p_2+p_3}$
-
-213: $\frac{p_2}{p_1+p_2+p_3}\cdot\frac{p_1}{p_1+p_3}$
-
-231: $\frac{p_2}{p_1+p_2+p_3}\cdot\frac{p_3}{p_1+p_3}$
-
-312: $\frac{p_3}{p_1+p_2+p_3}\cdot\frac{p_1}{p_1+p_2}$
-
-321: $\frac{p_3}{p_1+p_2+p_3}\cdot\frac{p_2}{p_1+p_2}$
-
-(Above, the probabilities clearly sum to 1).  Next simplest: 1 and 2
-mutually support each other with $\lambda$:
-
-123: $\frac{p_1}{p_1+p_2+p_3}\cdot\frac{\lambda p_2}{\lambda p_2+p_3}$
-
-132: $\frac{p_1}{p_1+p_2+p_3}\cdot\frac{p_3}{\lambda p_2+p_3}$
-
-213: $\frac{p_2}{p_1+p_2+p_3}\cdot\frac{\lambda p_1}{p_1+p_3}$
-
-231: $\frac{p_2}{p_1+p_2+p_3}\cdot\frac{p_3}{\lambda p_1+p_3}$
-
-312: $\frac{p_3}{p_1+p_2+p_3}\cdot\frac{p_1}{p_1+p_2}$
-
-321: $\frac{p_3}{p_1+p_2+p_3}\cdot\frac{p_2}{p_1+p_2}$
-
-Now we retain the mutual support between 1 and 2 but allow an
-asymmetry: 1 helps 2 with $\lambda$ as before, but now in addition, 2
-helps 1 with $\mu$:
-
-123: $\frac{p_1}{p_1+p_2+p_3}\cdot\frac{\lambda p_2}{\lambda p_2+p_3}$
-
-132: $\frac{p_1}{p_1+p_2+p_3}\cdot\frac{p_3}{\lambda p_2+p_3}$
-
-213: $\frac{p_2}{p_1+p_2+p_3}\cdot\frac{\mu p_1}{\mu p_1+p_3}$
-
-231: $\frac{p_2}{p_1+p_2+p_3}\cdot\frac{p_3}{\mu p_1+p_3}$
-
-312: $\frac{p_3}{p_1+p_2+p_3}\cdot\frac{p_1}{p_1+p_2}$
-
-321: $\frac{p_3}{p_1+p_2+p_3}\cdot\frac{p_2}{p_1+p_2}$
-
-Now how about 1 must beat 2.  He does not care whether he comes first
-or second, only that he must beat 2.  Player 1 has sufficient energy
-to win if he wants, but is subject to some sort of resource constraint
-(money, perhaps) and wishes to expend only enough money to beat 2.  So
-conventional P-L likelihoods are modified for 1, but only if 2 is
-_still running_.
-
-123: $\frac{\lambda p_1}{\lambda p_1+p_2+p_3}\cdot\frac{p_2}{p_2+p_3}$
-
-132: $\frac{\lambda p_1}{\lambda p_1+p_2+p_3}\cdot\frac{p_3}{p_2+p_3}$
-
-213: $\frac{p_2}{\lambda p_1+p_2+p_3}\cdot\frac{p_1}{p_1+p_3}$
-
-231: $\frac{p_2}{\lambda p_1+p_2+p_3}\cdot\frac{p_3}{p_1+p_3}$
-
-312: $\frac{p_3}{\lambda p_1+p_2+p_3}\cdot\frac{\lambda p_1}{\lambda p_1+p_2}$
-
-321: $\frac{p_3}{\lambda p_1+p_2+p_3}\cdot\frac{p_2}{\lambda p_1+p_2}$
-
-## Four runners
-
-Now we have four runners $\left\lbrace 1,2,3,4\right\rbrace$, again
-with the only interaction being runner 1 helping runner 2.  Thus, in a
-situation where $p_1$ has finished and $p_2$ is still running, we have
-$p_2\longrightarrow\lambda p_2$ [although the $\lambda$ cancels if 2
-is the only competitor still running, that is, if *all* the remaining
-competitors are helped]
-
-1234: $\frac{p_1}{p_1+p_2+p_3+p_4}\cdot\frac{\lambda p_2}{\lambda p_2+p_3+p_4}\cdot\frac{p_3}{p_3+p_4}$
-
-1243: $\frac{p_1}{p_1+p_2+p_3+p_4}\cdot\frac{\lambda p_2}{\lambda p_2+p_3+p_4}\cdot\frac{p_4}{p_3+p_4}$
-
-1324: $\frac{p_1}{p_1+p_2+p_3+p_4}\cdot\frac{p_3}{\lambda p_2+p_3+p_4}\cdot\frac{\lambda p_2}{\lambda p_2+p_4}$
-
-1342: $\frac{p_1}{p_1+p_2+p_3+p_4}\cdot\frac{p_3}{\lambda p_2+p_3+p_4}\cdot\frac{p_4}{\lambda p_2+p_4}$
-
-1324: $\frac{p_1}{p_1+p_2+p_3+p_4}\cdot\frac{p_3}{\lambda p_2+p_3+p_4}\cdot\frac{\lambda p_2}{\lambda p_2+p_4}$
-
-1423: $\frac{p_1}{p_1+p_2+p_3+p_4}\cdot\frac{p_4}{p_1+\lambda p_2+p_3+p_4}\cdot\frac{\lambda p_2}{\lambda p_2+p_3}$
-
-1432: $\frac{p_1}{p_1+p_2+p_3+p_4}\cdot\frac{p_4}{p_1+\lambda p_2+p_3+p_4}\cdot\frac{p_3}{\lambda p_2+p_3}$
-
-2xxx: P-L [no help given by 2 to 1]
-
-3124: $\frac{p_3}{p_1+p_2+p_3+p_4}\cdot\frac{p_1}{p_1+p_2+p_4}\cdot\frac{\lambda p_2}{\lambda p_2+p_4}$
-
-3142: $\frac{p_3}{p_1+p_2+p_3+p_4}\cdot\frac{p_1}{p_1+p_2+p_4}\cdot\frac{p_4}{\lambda p_2+p_4}$
-
-32xx: P-L
-
-3412: P-L [1 helps 2 but when 1 crosses the finishing line, only 2 is
-still running; there are no non-helped competitors]
-
-4123: $\frac{p_4}{p_1+p_2+p_3+p_4}\cdot\frac{p_1}{p_1+p_2+p_3}\cdot\frac{\lambda p_2}{\lambda p_2+p_3}$
-
-4132: $\frac{p_4}{p_1+p_2+p_3+p_4}\cdot\frac{p_1}{p_1+p_2+p_3}\cdot\frac{p_3}{\lambda p_2+p_3}$
-
-42xx: P-L
-
-43xx: P-L
-
-Now 1 and 2 mutually support one another by $\lambda$.  Here I will
-put the last term in, to show explicitly that the $\lambda$ term
-remains in the expressions for numerator and denominator.
-
-1234: $\frac{p_1}{p_1+p_2+p_3+p_4}\cdot\frac{\lambda p_2}{\lambda
-p_2+p_3+p_4}\cdot\frac{p_3}{p_3+p_4}\cdot\frac{p_4}{p_4}$
-
-1243: $\frac{p_1}{p_1+p_2+p_3+p_4}\cdot\frac{\lambda p_2}{\lambda
-p_2+p_3+p_4}\cdot\frac{p_4}{p_3+p_4}\cdot\frac{p_3}{p_3}$
-
-1324: $\frac{p_1}{p_1+p_2+p_3+p_4}\cdot\frac{p_3}{\lambda
-p_2+p_3+p_4}\cdot\frac{\lambda p_2}{\lambda
-p_2+p_4}\cdot\frac{p_4}{p_4}$
-
-1342: $\frac{p_1}{p_1+p_2+p_3+p_4}\cdot\frac{p_3}{\lambda
-p_2+p_3+p_4}\cdot\frac{p_4}{\lambda p_2+p_4}\cdot\frac{\lambda
-p_2}{\lambda p_2}$
-
-1423: $\frac{p_1}{p_1+p_2+p_3+p_4}\cdot\frac{p_4}{\lambda
-p_2+p_3+p_4}\cdot\frac{\lambda p_2}{\lambda
-p_2+p_3}\cdot\frac{p_3}{p_3}$
-
-1432: $\frac{p_1}{p_1+p_2+p_3+p_4}\cdot\frac{p_4}{\lambda
-p_2+p_3+p_4}\cdot\frac{p_3}{\lambda p_2+p_3}\cdot\frac{\lambda
-p_2}{\lambda p_2}$
-
-2134: $\frac{p_2}{p_1+p_2+p_3+p_4}\cdot\frac{\lambda p_1}{\lambda
-p_1+p_3+p_4}\cdot\frac{p_3}{p_3+p_4}\cdot\frac{p_4}{p_4}$
-
-2143: $\frac{p_2}{p_1+p_2+p_3+p_4}\cdot\frac{\lambda p_1}{\lambda
-p_1+p_3+p_4}\cdot\frac{p_4}{p_3+p_4}\cdot\frac{p_3}{p_3}$
-
-2314: $\frac{p_2}{p_1+p_2+p_3+p_4}\cdot\frac{\lambda p_3}{\lambda
-p_1+p_3+p_4}\cdot\frac{\lambda p_1}{\lambda
-p_1+p_4}\cdot\frac{p_4}{p_4}$
-
-2341: $\frac{p_2}{p_1+p_2+p_3+p_4}\cdot\frac{\lambda p_3}{\lambda
-p_1+p_3+p_4}\cdot\frac{p_4}{\lambda p_1+p_4}\cdot\frac{\lambda
-p_1}{\lambda p_1}$
-
-We will calculate a Plackett-Luce likelihood function for four runners
-$\left\lbrace 1,2,3,4\right\rbrace$ but with a reified Bradley-Terry
-monster that corresponds to having subsets of mutual friends who
-support one another.  For example, we might have runners 1 and 2 being
-friends, and runners 3 and 4 being friends: we would represent this as
-$\left\lbrace\left\lbrace 1,2\right\rbrace,\left\lbrace
-3,4\right\rbrace\right\rbrace$.
-
-The runner metaphor is not the only one.  We can apply the resulting
-likelihood functions to other cases including the red bus-blue bus
-problem (see `inst/red_bus_blue_bus.Rmd` for more discussion).
-
-First do three runners, and suppose the equivalence classes are
-$\left\lbrace\left\lbrace 1,2\right\rbrace,\left\lbrace
-3\right\rbrace\right\rbrace$.
-
- 123: $\frac{p_1}{p_1+p_2+p_3}\cdot\frac{p_2+M}{p_2+p_3+M}\qquad\frac{p_1}{p_1+p_2+p_3}\cdot\frac{\lambda p_2}{\lambda p_2+p_3}$
-
- 132: $\frac{p_1}{p_1+p_3+p_2}\cdot\frac{p_3}{p_3+p_2+M}\qquad\frac{p_1}{p_1+p_3+p_2}\cdot\frac{p_3}{p_3+\lambda p_2}$
-
- 213: $\frac{p_2}{p_2+p_1+p_3}\cdot\frac{p_1+M}{p_1+p_3+M}\qquad\frac{p_2}{p_2+p_1+p_3}\cdot\frac{\lambda p_1}{\lambda p_1+p_3}$
-
- 231: $\frac{p_2}{p_2+p_3+p_1}\cdot\frac{p_3}{p_3+p_1+M}\qquad\frac{p_2}{p_2+p_1+p_3}\cdot\frac{\lambda p_1}{p_3+\lambda p_1}$
-
- 312: $\frac{p_3}{p_3+p_1+p_2}\cdot\frac{p_1}{p_2+p_1}\qquad\frac{p_3}{p_3+p_1+p_2}\cdot\frac{\lambda p_1}{p_3+p_1}$
-
- 321: $\frac{p_3}{p_3+p_2+p_1}\cdot\frac{p_2}{p_2+p_1}\qquad\frac{p_3}{p_3+p_2+p_1}\cdot\frac{p_2}{p_2+p_1}$
-
-Now four runners with classes $\left\lbrace\left\lbrace
-1,2\right\rbrace,\left\lbrace 3,4\right\rbrace\right\rbrace$.
-
- 1234: $\frac{1}{1+2+3+4}\cdot\frac{2+M}{2+3+4+M}\cdot\frac{3}{3+4}$
-
- 1243: $\frac{1}{1+2+4+3}\cdot\frac{2+M}{2+4+3+M}\cdot\frac{4}{4+3}$
-
- 1324: $\frac{1}{1+3+2+4}\cdot\frac{3}{3+2+4+M}\cdot\frac{2}{2+4+M}$
-
- 1342: $\frac{1}{1+3+4+2}\cdot\frac{3}{3+4+2+M}\cdot\frac{4+M}{2+4+M}$
-
- 1423: $\frac{1}{1+4+2+3}\cdot\frac{4}{4+2+3+M}\cdot\frac{2}{2+3}$
-
- 1432: $\frac{1}{1+4+3+2}\cdot\frac{4}{4+3+2+M}\cdot\frac{3}{2+3}$
-
- 2134: $\frac{2}{2+1+3+4}\cdot\frac{1+M}{1+3+4+M}\cdot\frac{3}{3+4}$
-
- 2143: $\frac{2}{2+1+4+3}\cdot\frac{1+M}{1+4+3+M}\cdot\frac{4}{4+3}$
-
- 2314: $\frac{2}{2+3+1+4}\cdot\frac{3}{3+1+4+M}\cdot\frac{1}{1+4+M}$
-
- 2341: $\frac{2}{2+3+4+1}\cdot\frac{3}{3+4+1+M}\cdot\frac{4+M}{1+4+M}$
-
- 2413: $\frac{2}{2+4+1+3}\cdot\frac{4}{4+1+3+M}\cdot\frac{1}{1+3}$
-
- 2431: $\frac{2}{2+4+3+1}\cdot\frac{4}{4+3+1+M}\cdot\frac{3}{1+3}$
-
-etc.
-
-Now 123 tethered, 45 tethered:
-
-12345: $\frac{1}{1+2+3+4+5}\cdot\frac{2+M}{2+3+4+5+M}\cdot\frac{3+M}{3+4+5+M}\cdot\frac{4}{4+5}$
-
-12354: $\frac{1}{1+2+3+5+4}\cdot\frac{2+M}{2+3+5+4+M}\cdot\frac{3+M}{3+5+4+M}\cdot\frac{5}{5+4}$
-
-12435: $\frac{1}{1+2+3+5+4}\cdot\frac{2+M}{2+3+5+4+M}\cdot\frac{4}{3+5+4+M}\cdot\frac{3}{3+5+M}$
-
-12453: $\frac{1}{1+2+3+5+4}\cdot\frac{2+M}{2+3+5+4+M}\cdot\frac{4}{3+5+4+M}\cdot\frac{5+M}{5+3+M}$
-
-12534: $\frac{1}{1+2+3+5+4}\cdot\frac{2+M}{2+5+3+4+M}\cdot\frac{5}{3+5+4+M}\cdot\frac{3}{3+4+M}$
-
-12543: $\frac{1}{1+2+3+5+4}\cdot\frac{2+M}{2+5+4+3+M}\cdot\frac{5}{3+5+4+M}\cdot\frac{4+M}{4+3+M}$
-
-Might be better to consider ${5\choose 2\,3}$ arrangements of
-$aaabb\ldots bbaaa$.  FOLLOWING IS NOT RIGHT: THE PROBABILITY MODEL IS
-NOT PRECISELY DEFINED.
-
-aaabb: $\frac{a}{a+a+a+b+b}\cdot\frac{a+M}{a+a+b+b+M}\cdot\frac{a+M}{a+b+b+M}\cdot\frac{b}{b+b+M}$
-
-aabab: $\frac{a}{a+a+a+b+b}\cdot\frac{a+M}{a+a+b+b+M}\cdot\frac{b}{a+b+b+M}\cdot\frac{a}{a+b+M}$
-
-aabba: $\frac{a}{a+a+a+b+b}\cdot\frac{a+M}{a+a+b+b+M}\cdot\frac{b}{a+b+b+M}\cdot\frac{b+M}{b+a+M}$
-
-abaab: $\frac{a}{a+a+a+b+b}\cdot\frac{b}{a+a+b+b+M}\cdot\frac{a}{a+a+b+M}\cdot\frac{a+M}{a+b+M}$
-
-ababa: $\frac{a}{a+a+a+b+b}\cdot\frac{b}{a+a+b+b+M}\cdot\frac{a}{a+a+b+M}\cdot\frac{b}{b+a+M}$
-
-abbaa: $\frac{a}{a+a+a+b+b}\cdot\frac{b}{a+a+b+b+M}\cdot\frac{b+M}{b+a+a+M}\cdot\frac{a}{a+a+M}$
-
-baaab: $\frac{b}{a+a+a+b+b}\cdot\frac{a}{a+a+a+b+M}\cdot\frac{a+M}{a+a+b+M}\cdot\frac{a+M}{a+b+M}$
-
-baaba: $\frac{b}{a+a+a+b+b}\cdot\frac{a}{a+a+a+b+M}\cdot\frac{a+M}{a+a+b+M}\cdot\frac{b}{a+b+M}$
-
-babaa: $\frac{b}{a+a+a+b+b}\cdot\frac{a}{a+a+a+b+M}\cdot\frac{b}{b+a+a+M}\cdot\frac{a}{a+a}$
-
-bbaaa: $\frac{b}{a+a+a+b+b}\cdot\frac{b+M}{a+a+a+b+M}\cdot\frac{a}{a+a+a}\cdot\frac{a}{a+a}$
-
-```{r,label=definefunctionf}
-f <- function (a,b,M){
-return(
-  factorial(3)*factorial(2)*(
-   a/(a+a+a+b+b)*(a+M)/(a+a+b+b+M)*(a+M)/(a+b+b+M)*(b  )/(b+b+M) +
-   a/(a+a+a+b+b)*(a+M)/(a+a+b+b+M)*(b  )/(a+b+b+M)*(a  )/(a+b+M) +
-   a/(a+a+a+b+b)*(a+M)/(a+a+b+b+M)*(b  )/(a+b+b+M)*(b+M)/(b+a+M) +
-   a/(a+a+a+b+b)*(b  )/(a+a+b+b+M)*(a  )/(a+a+b+M)*(a+M)/(a+b+M) +
-   a/(a+a+a+b+b)*(b  )/(a+a+b+b+M)*(a  )/(a+a+b+M)*(b  )/(b+a+M) +
-   a/(a+a+a+b+b)*(b  )/(a+a+b+b+M)*(b+M)/(b+a+a+M)*(a  )/(a+a+M) +
-   b/(a+a+a+b+b)*(a  )/(a+a+a+b+M)*(a+M)/(a+a+b+M)*(a+M)/(a+b+M) +
-   b/(a+a+a+b+b)*(a  )/(a+a+a+b+M)*(a+M)/(a+a+b+M)*(b  )/(a+b+M) +
-   b/(a+a+a+b+b)*(a  )/(a+a+a+b+M)*(b  )/(b+a+a  )*(a  )/(a+a  ) +
-   b/(a+a+a+b+b)*(b+M)/(a+a+a+b+M)*(a  )/(a+a+a  )*(a  )/(a+a  )
-   ) 
-   )
-}
-```
-
-```{r evalat44}
-f(4.4,3.3, 0.2)*12
-f(4.4,3.3, 0.0)*12
-```
-
-```{tikz, label=tikzpartprobtree1,fig.cap = "Partial probability tree for five competitors $a-e$ with $a$ supporting $b$, hyper3 approach", fig.ext = 'png', echo=FALSE}
-\usetikzlibrary{arrows}
-\usetikzlibrary{patterns}
-\begin{tikzpicture}[line cap=round,line join=round,>=triangle 45,x=1cm,y=1cm]
-\fill (0,0) circle[radius=2pt]; % root; paths abcde
-\draw (0,0) -- (2,3); 
-\draw (0,0) -- (2,2);
-\draw (0,0) -- (2,1);
-\draw (0,0) -- (2,0);
-\draw (0,0) -- (2,-1);
-\node at (0,2.5) (eq1) {$a\longrightarrow b$};
-
-\node at (1.5,2.5) {$a$};
-\node at (1.5,1.7) {$b$};
-\node at (1.5,1.0) {$c$};
-\node at (1.5,0.2) {$d$};
-\node at (1.5,-0.5) {$e$};
-
-\node at (3, 3)  {$\frac{a}{a+b+c+d+e}$};
-\node at (3, 2)  {$\frac{b}{a+b+c+d+e}$};
-\node at (3, 1)  {$\frac{c}{a+b+c+d+e}$};
-\node at (3, 0)  {$\frac{d}{a+b+c+d+e}$};
-\node at (3,-1)  {$\frac{e}{a+b+c+d+e}$};
-
-\fill  (4, 3) circle[radius=2pt];  % a finishes; paths bcde
-\fill  (4, 2) circle[radius=2pt];  % terminal node
-\fill  (4, 1) circle[radius=2pt];  % terminal node
-\fill  (4, 0) circle[radius=2pt];  % d finishes; paths abce
-\fill  (4,-1) circle[radius=2pt];  % terminal node
-
-\draw (4,3) -- (5,5); 
-\draw (4,3) -- (5,4); 
-\draw (4,3) -- (5,3); 
-\draw (4,3) -- (5,2);
-
-\node at (4.5,4.6) {$b$};
-\node at (4.5,3.7) {$c$};
-\node at (4.5,3.2) {$d$};
-\node at (4.5,2.7) {$e$};
-
-\newcommand{\lb}{\lambda b }
-\newcommand{\lc}{\lambda c }
-\newcommand{\lee}{\lambda e }
-\node at (6, 5)  {$\frac{\lb}{\lb+c+d+e}$};
-\node at (6, 4)  {$\frac{c}{\lb+c+d+e}$};
-\node at (6, 3)  {$\frac{d  }{\lb+c+d+e}$};
-\node at (6, 2)  {$\frac{e  }{\lb+c+d+e}$};
-
-\fill (7.4, 5) circle[radius=2pt];  % ab finishes; paths cde
-\draw (7.4,5) -- (8.4,6);
-\draw (7.4,5) -- (8.4,5);
-\draw (7.4,5) -- (8.4,4);
-
-\node at (7.9,5.7) {$c$};
-\node at (7.9,5.2) {$d$};
-\node at (7.9,4.7) {$e$};
-\node at (9.3, 6) {$\frac{c}{c+d+e}$};
-\node at (9.3, 5) {$\frac{d  }{c+d+e}$};
-\node at (9.3, 4) {$\frac{e  }{c+d+e}$};
-
-\fill (10.2, 6) circle[radius=2pt];  % abc finishes; paths de
-\draw (10.2, 6) -- (11.2,7);
-\node at (10.7,6.8) {$d$};
-
-\node at (11.8, 7) {$\frac{d}{d+e}$};
-
-\draw (10.2, 6) -- (11.2,6.2);
-\node at (11.8, 6.2) {$\frac{e}{d+e}$};
-\node at (10.7,6.3) {$e$};
-
-\fill (10.2, 5) circle[radius=2pt];
-\draw (10.2, 5) -- (11.2,5.3);
-\node at (11.8, 5.3) {$\frac{c}{c+e}$};
-\node at (10.7,5.3) {$c$};
-
-\draw (10.2, 5) -- (11.2,4.7);
-\node at (11.8, 4.7) {$\frac{e}{c+e}$};
-\node at (10.7,4.7) {$e$};
-
-\fill (10.2, 4) circle[radius=2pt];
-\draw (10.2, 4) -- (11.2,3.9);
-\node at (11.8, 3.9) {$\frac{c}{c+e}$};
-\node at (10.7,4.1) {$c$};
-
-\draw (10.2, 4) -- (11.2,3.1);
-\node at (11.8, 3.1) {$\frac{e}{c+e}$};
-\node at (10.7,3.7) {$e$};
-
-\draw (4, 0) -- (5,1);
-\node at (6, 1)  {$\frac{a}{a+b+c+e}$};
-
-\draw (4, 0) -- (5,0);
-\node at (6, 0)  {$\frac{b}{a+b+c+e}$};
-
-\draw (4, 0) -- (5,-1);
-\node at (6, -1)  {$\frac{c}{a+b+c+e}$};
-
-\draw (4, 0) -- (5,-2);
-\node at (6, -2)  {$\frac{e}{a+b+c+e}$};
-
-\node at (4.5,0.7) {$a$};
-\node at (4.5,0.2) {$b$};
-\node at (4.5,-0.3) {$c$};
-\node at (4.5,-0.8) {$e$};
-
-\fill (7.4, -2) circle[radius=2pt];  % de finishes; paths abc
-\draw (7.4, -2) -- (8.4,-1);
-\draw (7.4, -2) -- (8.4,-2);
-\draw (7.4, -2) -- (8.4,-3);
-
-\node at (7.9,-1.2) {$a$};
-\node at (7.9,-1.8) {$b$};
-\node at (7.9,-2.3) {$c$};
-
-\node at (9.3, -1) {$\frac{a}{a+b+c}$};
-\node at (9.3, -2) {$\frac{b}{a+b+c}$};
-\node at (9.3, -3) {$\frac{c}{a+b+c}$};
-
-\fill (10.2, -1) circle[radius=2pt];  % dea finishes; paths bc
-\fill (10.2, -2) circle[radius=2pt];  % deb finishes; paths ac
-\fill (10.2, -3) circle[radius=2pt];  % dec finishes; paths ab
-
-\draw (10.2, -1) -- (11.2,-0);
-\node at (10.6,-0.4) {$b$};
-\node at (10.6,-0.8) {$c$};
-
-\node at (11.8, 0) {$\frac{\lb}{\lb+c}$};
-\draw (10.2, -1) -- (11.2,-1);
-
-\node at (11.8, -1) {$\frac{c}{\lb+c}$};
-
-\draw (10.2, -2) -- (11.2,-1.6);
-\node at (11.8, -1.6) {$\frac{a}{a+c}$};
-
-\node at (10.6,-1.7) {$a$};
-\node at (10.6,-2) {$c$};
-
-\draw (10.2, -2) -- (11.2,-2.2);
-\node at (11.8, -2.2) {$\frac{c}{a+c}$};
-
-\draw (10.2, -3) -- (11.2,-2.9);
-\node at (11.8, -2.9) {$\frac{a}{a+b}$};
-
-\draw (10.2, -3) -- (11.2,-3.4);
-\node at (11.8, -3.4) {$\frac{b}{a+b}$};
-
-\node at (10.6,-2.8) {$a$};
-\node at (10.6,-3.3) {$b$};
-
-\fill (7.4, 1) circle[radius=2pt];  % da finishes; paths bce
-
-\draw (7.4, 1) -- (8.4,2);
-\draw (7.4, 1) -- (8.4,1);
-\draw (7.4, 1) -- (8.4,0);
-
-\node at (7.9,1.7) {$b$};
-\node at (7.9,1.2) {$c$};
-\node at (7.9,0.7) {$e$};
-
-\node at (9.3, 2) {$\frac{\lb}{\lb+c+e}$};
-\node at (9.3, 1) {$\frac{  c}{\lb+c+e}$};
-\node at (9.3, 0) {$\frac{e  }{\lb+c+e}$};
-
-\fill (10.2, 1) circle[radius=2pt];  % da finishes; paths bce
-\draw (10.2, 1) -- (11.2,2);
-\node at (11.8, 2) {$\frac{\lb}{\lb+e}$};
-\draw (10.2, 1) -- (11.2,1);
-\node at (11.8, 1) {$\frac{e}{\lb+e}$};
-
-\end{tikzpicture}
-```
-
-```{tikz, label=tikzpic2, fig.cap = "Partial probability tree for five competitors $a-e$ with mutually supporting  group $abc$, hyper3 approach.
-If any competitor in the set {$a,b,c$} finishes, they lend their strength to the other competitors in the set who are still running", fig.ext = 'png', echo=FALSE}
-\usetikzlibrary{arrows}
-\usetikzlibrary{patterns}
-\begin{tikzpicture}[line cap=round,line join=round,>=triangle 45,x=1cm,y=1cm]
-\fill (0,0) circle[radius=2pt]; % root; paths abcde
-\draw (0,0) -- (2,3); 
-\draw (0,0) -- (2,2);
-\draw (0,0) -- (2,1);
-\draw (0,0) -- (2,0);
-\draw (0,0) -- (2,-1);
-\node at (0,2.5) (eq1) {$\underbrace{\left\lbrace a,b,c\right\rbrace}_{\lambda}$};
-
-\node at (1.5,2.5) {$a$};
-\node at (1.5,1.7) {$b$};
-\node at (1.5,1.0) {$c$};
-\node at (1.5,0.2) {$d$};
-\node at (1.5,-0.5) {$e$};
-
-\node at (3, 3)  {$\frac{a}{a+b+c+d+e}$};
-\node at (3, 2)  {$\frac{b}{a+b+c+d+e}$};
-\node at (3, 1)  {$\frac{c}{a+b+c+d+e}$};
-\node at (3, 0)  {$\frac{d}{a+b+c+d+e}$};
-\node at (3,-1)  {$\frac{e}{a+b+c+d+e}$};
-
-\fill  (4, 3) circle[radius=2pt];  % a finishes; paths bcde
-\fill  (4, 2) circle[radius=2pt];  % terminal node
-\fill  (4, 1) circle[radius=2pt];  % terminal node
-\fill  (4, 0) circle[radius=2pt];  % d finishes; paths abce
-\fill  (4,-1) circle[radius=2pt];  % terminal node
-
-\draw (4,3) -- (5,5); 
-\draw (4,3) -- (5,4); 
-\draw (4,3) -- (5,3); 
-\draw (4,3) -- (5,2);
-
-\node at (4.5,4.6) {$b$};
-\node at (4.5,3.7) {$c$};
-\node at (4.5,3.2) {$d$};
-\node at (4.5,2.7) {$e$};
-
-\newcommand{\la}{\lambda a }
-\newcommand{\lb}{\lambda b }
-\newcommand{\lc}{\lambda c }
-\newcommand{\lee}{\lambda e }
- \node at (6, 5)  {$\frac{\lb}{\lb+\lc+d+e}$};
- \node at (6, 4)  {$\frac{\lc}{\lb+\lc+d+e}$};
- \node at (6, 3)  {$\frac{d  }{\lb+\lc+d+e}$};
- \node at (6, 2)  {$\frac{e  }{\lb+\lc+d+e}$};
-
- \fill (7.4, 5) circle[radius=2pt];  % ab finishes; paths cde
- \draw (7.4,5) -- (8.4,6);
- \draw (7.4,5) -- (8.4,5);
- \draw (7.4,5) -- (8.4,4);
-
- \node at (7.9,5.7) {$c$};
- \node at (7.9,5.2) {$d$};
- \node at (7.9,4.7) {$e$};
- \node at (9.3, 6) {$\frac{\lc}{\lc+d+e}$};
- \node at (9.3, 5) {$\frac{d  }{\lc+d+e}$};
- \node at (9.3, 4) {$\frac{e  }{\lc+d+e}$};
-
- \fill (10.2, 6) circle[radius=2pt];  % abc finishes; paths de
- \draw (10.2, 6) -- (11.2,7);
- \node at (10.7,6.8) {$d$};
-
- \node at (11.8, 7) {$\frac{d}{d+e}$};
-
- \draw (10.2, 6) -- (11.2,6.2);
- \node at (11.8, 6.2) {$\frac{e}{d+e}$};
- \node at (10.7,6.3) {$e$};
-
- \fill (10.2, 5) circle[radius=2pt];
- \draw (10.2, 5) -- (11.2,5.3);
- \node at (11.8, 5.3) {$\frac{\lc}{\lc+e}$};
- \node at (10.7,5.3) {$c$};
-
- \draw (10.2, 5) -- (11.2,4.7);
- \node at (11.8, 4.7) {$\frac{e}{\lc+e}$};
- \node at (10.7,4.7) {$e$};
-
- \fill (10.2, 4) circle[radius=2pt];
- \draw (10.2, 4) -- (11.2,3.9);
- \node at (11.8, 3.9) {$\frac{\lc}{\lc+e}$};
- \node at (10.7,4.1) {$c$};
-
- \draw (10.2, 4) -- (11.2,3.1);
- \node at (11.8, 3.1) {$\frac{e}{\lc+e}$};
- \node at (10.7,3.7) {$e$};
-
- \draw (4, 0) -- (5,1);
- \node at (6, 1)  {$\frac{a}{a+b+c+e}$};
-
- \draw (4, 0) -- (5,0);
- \node at (6, 0)  {$\frac{b}{a+b+c+e}$};
-
- \draw (4, 0) -- (5,-1);
- \node at (6, -1)  {$\frac{c}{a+b+c+e}$};
-
- \draw (4, 0) -- (5,-2);
- \node at (6, -2)  {$\frac{e}{a+b+c+e}$};
-
- \node at (4.5,0.7) {$a$};
- \node at (4.5,0.2) {$b$};
- \node at (4.5,-0.3) {$c$};
- \node at (4.5,-0.8) {$e$};
-
- \fill (7.4, -2) circle[radius=2pt];  % de finishes; paths abc
- \draw (7.4, -2) -- (8.4,-1);
- \draw (7.4, -2) -- (8.4,-2);
- \draw (7.4, -2) -- (8.4,-3);
-
- \node at (7.9,-1.2) {$a$};
- \node at (7.9,-1.8) {$b$};
- \node at (7.9,-2.3) {$c$};
-
- \node at (9.3, -1) {$\frac{a}{a+b+c}$};
- \node at (9.3, -2) {$\frac{b}{a+b+c}$};
- \node at (9.3, -3) {$\frac{c}{a+b+c}$};
-
- \fill (10.2, -1) circle[radius=2pt];  % dea finishes; paths bc
- \fill (10.2, -2) circle[radius=2pt];  % deb finishes; paths ac
- \fill (10.2, -3) circle[radius=2pt];  % dec finishes; paths ab
-
- \draw (10.2, -1) -- (11.2,-0);
- \node at (10.6,-0.4) {$b$};
- \node at (10.6,-0.8) {$c$};
-
- \node at (11.8, 0) {$\frac{\lb}{\lb+\lc}$};
- \draw (10.2, -1) -- (11.2,-1);
-
- \node at (11.8, -1) {$\frac{c}{\lb+\lc}$};
-
- \draw (10.2, -2) -- (11.2,-1.6);
- \node at (11.8, -1.6) {$\frac{\la}{\la+\lc}$};
-
- \node at (10.6,-1.7) {$a$};
- \node at (10.6,-2) {$c$};
-
- \draw (10.2, -2) -- (11.2,-2.2);
- \node at (11.8, -2.2) {$\frac{\lc}{\la+\lc}$};
-
- \draw (10.2, -3) -- (11.2,-2.9);
- \node at (11.8, -2.9) {$\frac{\la}{\la+\lb}$};
-
- \draw (10.2, -3) -- (11.2,-3.4);
- \node at (11.8, -3.4) {$\frac{\lb}{\la+\lb}$};
-
- \node at (10.6,-2.8) {$a$};
- \node at (10.6,-3.3) {$b$};
-
- \fill (7.4, 1) circle[radius=2pt];  % da finishes; paths bce
-
- \draw (7.4, 1) -- (8.4,2);
- \draw (7.4, 1) -- (8.4,1);
- \draw (7.4, 1) -- (8.4,0);
-
- \node at (7.9,1.7) {$b$};
- \node at (7.9,1.2) {$c$};
- \node at (7.9,0.7) {$e$};
-
- \node at (9.3, 2) {$\frac{\lb}{\lb+\lc+e}$};
- \node at (9.3, 1) {$\frac{\lc}{\lb+\lc+e}$};
- \node at (9.3, 0) {$\frac{e  }{\lb+\lc+e}$};
-
- \fill (10.2, 1) circle[radius=2pt];  % da finishes; paths bce
- \draw (10.2, 1) -- (11.2,2);
- \node at (11.8, 2) {$\frac{\lb}{\lb+e}$};
- \draw (10.2, 1) -- (11.2,1);
- \node at (11.8, 1) {$\frac{e}{\lb+e}$};
-
-\end{tikzpicture}
-```
-
-```{tikz, label=tikz3,fig.cap = "Partial probability tree for five competitors $a$-$e$ with mutually supporting subsets $\\left\\lbrace a,b,c\\right\\rbrace$ [with support term $\\lambda$] and $(de)$ [with support term $\\mu$], hyper3 approach", fig.ext = 'png', echo=FALSE}
+```{tikz, label=tikz3, fig.cap = "Partial probability\\label{tikzabcde} tree for five competitors $a$-$e$ with mutually supporting subsets $\\left\\lbrace a,b,c\\right\\rbrace$ [with support term $\\lambda$] and $(de)$ [with support term $\\mu$], hyper3 approach", fig.ext = 'png', echo=FALSE}
 \usetikzlibrary{arrows}
@@
 
+\begin{equation}\label{asuccb}
+\frac{p_a}{p_a + p_b + p_c + p_d + p_e}\cdot
+\frac{\lambda p_b}{\lambda p_b + \lambda p_c + p_d + p_e}\cdot
+\frac{\lambda p_c}{\lambda p_c + p_d + p_e}\cdot
+\frac{p_d}{p_d + p_e}
+\end{equation}
+
+In equation \ref{asuccb}, the first term is standard Plackett-Luce: at
+this point, nooone has finished and cheering effects are absent.  The
+second term reflects the fact that competitors $b$ and $c$ are
+supported by competitor $a$, who has by this point finished the race
+and is supporting his teammates.
+
+By contrast, the likelihood function for observation $d\succ a\succ
+c\succ b\succ e$ would be
+
+\begin{equation}
+\frac{p_d}{p_a + p_b + p_c + p_d + p_e}\cdot
+\frac{p_a}{p_a + p_b + p_c + \mu p_e}\cdot
+\frac{\lambda p_c}{\lambda p_b + \lambda p_c + p_e}\cdot
+\frac{\lambda p_b}{\lambda p_b + \mu p_e}
+\end{equation}
+
+where this likelihood function reflects the mutual support for
+equivalence class $\left\lbrace d,e\right\rbrace$.  Note that the
+final term reflects the fact that competitors $b$ and $e$ have their
+support active when vying for fourth place, as members of both their
+teams have finished at this point.
+
+This probability model could easily be modified to account for
+specific circumstances.  The cheering effect be asymmetrical (with $a$
+helping $b$ but not vice-versa, for example).  The effect could
+operate only on specific ordered pairs.  Or perhaps the effect might
+have a finite lifetime: if $a$ places $n^\mathrm{th}$, then the
+cheering effect is active only for competitors placing
+$(n+r)^\mathrm{th}$ or above.
+
 # Package idiom
@@
 pchisq(suppdiff*2,df=1,lower.tail=FALSE)
+```
+
+Now use the evaluate for the likelihood function:
+
+```{r calcmaxmax, cache=TRUE}
+maxHmax <- maxp(RB_BB_LF(s = osup$maximum))
+```
+
+```{r showmaxmax}
+maxHmax
 ```

### 2025-02-17 inst/jss_hyper3_real.Rnw
initial rough draft, need to add red bus/blue bus stuff

% -*- mode: noweb; noweb-default-code-mode: R-mode; -*-
\documentclass[article]{jss}
\usepackage{booktabs}
\usepackage{units}
\usepackage{amsmath}
\usepackage{amssymb}

\author{Robin K. S. Hankin\\University of Stirling}
\title{Generalized Bradley-Terry models}
\Plainauthor{Robin K. S. Hankin} %% comma-separated
\Plaintitle{Generalized Bradley-Terry models}

%% an abstract and keywords
\Abstract{

  The \code{hyper2} package provides functionality to work with
  extensions of the Bradley-Terry probability model such as reified
  entities (monsters) that account for known enhancements in
  competitive ability.  Here, I present a generalization of
  \code{hyper2} in which Bradley-Terry strengths may be enhanced using
  a multiplicative factor $\lambda$.  Generally, ordinary
  Bradley-Terry is recovered if $\lambda=1$.  This approach offers
  multiple advantages over reified Bradley-Terry, and several examples
  are given.
}
\Keywords{Plackett-Luce,
  Bradley-Terry, Mann-Whitney} \Plainkeywords{Plackett-Luce,
  Bradley-Terry, Mann-Whitney}

%% publication information
%% NOTE: Typically, this can be left commented and will be filled out by the technical editor
%% \Volume{50}
%% \Issue{9}
%% \Month{June}
%% \Year{2012}
%% \Submitdate{2012-06-04}
%% \Acceptdate{2012-06-04}

%% The address of (at least) one author should be given
%% in the following format:
\Address{
  Robin K. S. Hankin\\
  University of Stirling\\
  E-mail: \email{hankin.robin@gmail.com}\\
  URL: \url{https://academics.aut.ac.nz/robin.hankin}
}

%% for those who use Sweave please include the following line (with % symbols):
%% need no \usepackage{Sweave.sty}

%% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newcommand\headercell[1]{%
   \smash[b]{\begin{tabular}[t]{@{}c@{}} #1 \end{tabular}}}
\newcommand{\draw}{\ensuremath{\mathfrak{D}}}
\newcommand{\cdraw}{\ensuremath{\mathfrak{C}}}
\newcommand{\white}{\ensuremath{\mathfrak{W}}}
\newcommand{\home}{\ensuremath{\mathfrak{H}}}
\newcommand{\oo}{\vphantom{W^{W^{W^W}}}}

\begin{document}

<<setup, cache=FALSE,echo=FALSE>>=
source("loadlibrary.R",echo=FALSE)  # to avoid stupid messages
opts_chunk$set(cache=TRUE, autodep=TRUE)
options(prompt = "R> ")
@

\section{Introduction}

The Bradley-Terry model for datasets involving paired comparisons has
wide uptake in the {R} community.  The canonical problem is to
consider~$n$ players who compete against one another; the basic
inference problem is to estimate
numbers~$\mathbf{p}=\left({p_1,\ldots,p_n}\right)$, $p_i\geqslant 0$,
$\sum p_i=1$ so that
$\Prob(\mbox{$i$~beats~$j$})=\frac{p_i}{p_i+p_j}$.  Information about
the~$p_i$ may be obtained from the results of paired comparisons
between the players.

One generalization of the Bradley-Terry model is due
to~\citet{luce1959}, in which the probability of competitor~$i$
winning in a field of~$j=1,\ldots, n$ is $\frac{p_i}{p_1+\cdots
  +p_n}$.  It is then natural to quantify some ephemeral enhancement
to competitor~$i$ using a multiplicative factor $\lambda$.  Then
$\Prob(\mbox{$i$ wins})=\frac{\lambda p_i}{p_1+\cdots+\lambda
  p_i+\cdots+p_n}$.  Examples of such enhancements include the
home-ground advantage in football and rugby, playing white in chess,
or XXXX.  Note that $\lambda < 1$ would correspond to a decrease in
competitive ability.

Further, one might follow HANKIN and consider teams of players.  In
doubles tennis, for example, $\left\lbrace p_1,p_2\right\rbrace$ might
play $\left\lbrace p_3,p_4\right\rbrace$ and win with probability
$\frac{p_1+p_2}{p_1+p_2+p_3+p_4}$.  It is natural to combine this idea
with that of a multiplicative factor.  For example, suppose $p_1$ has
a special move against $p_3$, and $p_4$ has just been given a
performance-enhancing drug, we might have a likelihood functions
proportional to $\frac{\lambda p_1+ p_2}{p_1+p_2+p_3+\mu p_4}$.  Then
the effectiveness of $p_3$'s special move and the enhancement afforded
by the drug are measured by $\lambda$ and $\mu$ respectively with
values $>1$ corresponding to a positive effect.

In this paper I present a generalization of the {\tt hyper2} suite for
Bradley-Terry likelihoods to accommodate general multiplicative
factors.  The most general likelihood function would be

\begin{equation}
  \prod_{i=1}^N
  \sum_{j=1}^n
 \left(\lambda_{ij}p_i\right)^{w_i}
  \end{equation}

where the $p_i$ are traditional Bradley-Terry strengths and the
$\lambda_{ij}\geqslant 0$ are multiplicative factors that account for
ephemeral or temporary changes in competitive ability.  In practice,
almost all the $\lambda_{ij}$ are zero.  To fix ideas, consider the
data presented in Table~\ref{oldfirm}.  Ignoring draws for the moment,
we have a likelihood proportional to

\begin{equation}
  \left(\frac{\lambda p_R}{\lambda p_R + p_C}\right)^{81}\cdot
  \left(\frac{\lambda p_R}{\lambda p_R + p_F}\right)^{69}\cdots
  \left(\frac{\lambda p_L}{\lambda p_L + p_F}\right)^{81}\cdots
  \left(\frac{p_C}{p_C + \lambda p_R}\right)^{42}\cdots
    \left(\frac{p_F}{p_F + \lambda p_L}\right)^{9}
\end{equation}

In {\tt hyper3} idiom, a term such as $\lambda p_R + p_C$ is
represented as a named numeric vector.  For example, if the home
strength were $\lambda = 1.8$ then the first bracket would be {\tt
  c(Rangers=1.8, Celtic=1)} with an associated power of $-81$.  The
computational methodology is a generalization of that used
in~\cite{hankin2022} except that the multiplicative factors are real
(as opposed to non-negative integer), and also unknown.

\subsection{The package in use}
Package idiom for creating a likelihood function for this dataset
would look something like this: Considering only Rangers and Celtic:
Rangers at home:

<<defineoldfirm, cache=TRUE>>=
oldfirm <- function(lambda){
    H <- hyper3(pnames = c("Rangers", "Celtic"))
    H[c(Rangers = lambda)            ] %<>% add(81) 
    H[c(Celtic = 1)                  ] %<>% add(42) 
    H[c(Rangers = lambda, Celtic = 1)] %<>% subtract(81 + 42)

    H[c(Celtic = lambda)             ] %<>% add(76)
    H[c(Rangers = 1)                 ] %<>% add(46)
    H[c(Celtic = lambda, Rangers = 1)] %<>% subtract(46 + 76)
    return(H)
}
@

Above we see package idiom for creating a log-likelihood function.  We
first create an empty {\tt hyper3} object with an appropriate names
attribute.  The next three lines, using \pkg{magrittr} assignment
pipes, give the results for Rangers at home, and the next three lines
give the result for Celtic at home.  It is then easy to create a
bespoke log-likelihood function for Rangers-Celtic matches:

<<useoldfirm, cache = TRUE>>=
oldfirm(1.88)
@

We see a log-likelihood function for all $82+42+76+46=245$
Rangers-Celtic matches that were not score draws, conditional on
assumed value $\lambda=1.88$.  We may make inferences about the true
value of $\lambda$ using profile support:

<<profilelikeoldfirm, cache=TRUE>>=
f <- function(lambda){maxp(oldfirm(lambda),give=1)$`Log-likelihood`}
lam <- seq(from=0.9, to=3, len=23)
logLike <- sapply(lam,f)
@

Thus {\tt f()} returns the support at the evaluate, conditional on a
supplied value of $\lambda$.  We may use this to create a profile
likelihood:

\begin{figure}[htbp]
  \begin{center}
<<plotproflikeoldfirm, echo=FALSE, fig=TRUE>>=
plot(lam,logLike - max(logLike),type='b')
abline(v=1,col='gray')
segments(x0=1.785,y0=0.1,y1=-1,col='red')
abline(h = c(0,-2),col='gray')
@
\caption{A\label{profilesupportlambda} profile support function for the
  home-team advantage $\lambda$.  Vertical gray line shows
  $H_0\colon\lambda =1$, horizontal lines show the 
  two-units-of-support limits; small red bar shows
  $\hat{\lambda}\simeq 1.785$}
\end{center}
\end{figure}

<<optimoldfirm, cache=TRUE>>=
(lambda_max <-  optimize(f,c(1.5,2.0),maximum=TRUE)$maximum)
@

<<mleoldfirm,cache=TRUE>>=
maxp(oldfirm(lambda_max))
@

Compare with the theoretical value of $\widehat{p_{\mbox{\tiny
      C}}}=\frac{\sqrt{42\cdot 76}}{\sqrt{81\cdot 46}+\sqrt{42\cdot
    76}}\simeq 0.4807$, $\widehat{p_{\mbox{\tiny R}}}
=\frac{\sqrt{81\cdot 46}}{\sqrt{81\cdot 46}+\sqrt{42\cdot 76}}\simeq
0.5193$, and $\widehat{\lambda}=\frac{\sqrt{76\cdot 81}}{\sqrt{46\cdot
    42}}\simeq 1.785$.

\section{Score draws}

About 22\% of matches in Table~\ref{oldfirm} end in a draw and {\tt
  hyper2} formalism may be used to assess this.  Consider the
likelihood function described in Table~\ref{windrawloss}; this has a
number of desirable features.  Firstly, if we consider two players of
identical strengths and nullify the home ground effect (that is,
$\lambda=1$) then the probability of a draw is $\frac{2D}{2+2D}$; the
odds of a draw is simply $D$, regardless of team strength.  Secondly,
considering only won (that is, not score-drawn) games, the probability
of a home win is $\frac{\lambda}{1+\lambda}$, that is, the odds of a
home win is simply $\lambda$, independently of the player
strength. Such arguments give direct operational significance to
$\lambda$ and $D$.

\begin{table}
\centering
\begin{tabular}{l|c|c}\\&
{$i$ plays at home} &
{$i$ plays away} \\
  \hline
  $\Prob{(\mbox{$i$ wins})}$                             &
  $\frac{\lambda p_i}{\lambda p_i + D(p_i + p_j) + p_j}$  &
  $\frac{p_i}{p_i + D(p_i + \lambda p_j) + p_j}$          \\
  $\Prob{(\mbox{draw})}$                                 &  
  $\frac{D(p_i + p_j)}{\lambda p_i+D(p_i+p_j) + p_j}$     &
  $\frac{D(p_i + p_j)}{p_i + D(p_i+p_j) + \lambda p_j}$   \\
  $\Prob{(\mbox{$i$ loses})}$                            &  
  $\frac{p_j}{\lambda p_i + D(p_i+p_j) + p_j}$            &
  $\frac{\lambda p_j}{p_i + D(p_i + p_j) + \lambda p_j}$  \\
\end{tabular}
\caption{Probabilities of a win, draw, loss for the two teams under a
  multiplicative Bradley-Terry model\label{windrawloss} and a lambda thing}
\end{table}

We are now in a position to consider the entirety of
Table~\ref{oldfirm}.  Bespoke function \code{home\_draw\_away3()}
converts data in the form of Table~\ref{oldfirm} to a \code{hyper3}
likelihood function following Table~\ref{windrawloss}; it takes
numerical values for the home advantage $\lambda$ and proclivity to
draw $D$:

<<eg_RCLF3_lf, cache=TRUE>>=
home_draw_away3(RCLF3_table, lambda=1.7, D=0.3)
@

Observe that the resulting likelihood function has two distinct types
of parameters: the plain Bradley-Terry strengths $\mathbf{p}=(p_R,
p_C,p_L,p_F)$, with unit sum; but also the auxiliary parameters
$\lambda,D$.  To make inferences about these parameters we calculate a
profile likelihood function: conditional on $\lambda,D$ we maximize
the likelihood over possible values of $\mathbf{p}$.

<<defineF>>=
f <- function(v){
    H <- home_draw_away3(RCLF3_table, lambda = v[1], D=v[2])
    maxp(H,give=1)$`Log-likelihood`
}
@

\begin{figure}[htbp]
  \begin{center}
<<makecontour, fig=TRUE, cache=TRUE>>=
n <- 2
lambda <- seq(from=1, to=3, len=n)
D <- seq(from=0.1, to=0.6, len=n)
V <- as.matrix(expand.grid(lambda, D))
LL <- apply(V, 1, f)
LL <- LL - max(LL)
contour(lambda, D, matrix(LL, n, n))
@
\caption{Contours of support\label{contourlambdaD} for a grid}
\end{center}
\end{figure}

Thus \code{f()} gives us a likelihood function for
$\mathbf{v}=(\lambda,D)$.  We may apply \code{f()} over a grid and
thereby determine a likelihood surface, shown in
figure~\ref{contourlambdaD}.  Further, we may find the evaluate using
\code{optim()}:

<<findmaxlambdaD, cache=TRUE>>=
maxv <- optim(par=c(2, 0.3), fn=f, control=list(fnscale = -1))$par
maxv
@

<<testlambdazero, cache=TRUE>>=
jj <- optimize(f=function(D){f(c(lambda=1,D))}, interval=c(0.1, 0.4), maximum=TRUE)
jj
@

\section{Other stuff}

functions~\citep{hankin2010,hankin2017_rnw} account for enhancements
of competitive ability through the inclusion of reified entities
(monsters) whose Bradley-Terry strength represents the extra strength.
For example, if a player is at an advantage for some reason then a
reified entity might help him and then
$\Prob(\mbox{$i$~beats~$j$})=\frac{p_i+\mathfrak{M}}{p_i+p_j+\mathfrak{M}}$,
where $\mathfrak{M}$ is a reified entity whose strength helps
competitor $i$.
Such considerations motivate the
likelihood function in Equations~\ref{likefour}, originally set out by
XXXXX and YYYY.

\begin{align}\label{likefour}
\Prob{(\mbox{$i$ wins at home})} &= \frac{\lambda p_i}{\lambda p_i+p_j}\\
\Prob{(\mbox{$i$ wins away})} &= \frac{\lambda p_i}{\lambda p_i+p_j}\\
\Prob{(\mbox{$j$ wins at home})} &= \frac{\lambda p_j}{p_i+\lambda p_j}\\
\Prob{(\mbox{$j$ wins away})} &= \frac{p_j}{\lambda p_i+\lambda p_j}
\end{align}

Here, $\lambda\geqslant 0$ is a multiplicative constant accounting for
the home-ground advantage. If $\lambda>1$ the advantage is positive
and if $\lambda<1$ it is negative and helps the away side.

Here I introduce {\tt hyper3} formalism

However, reified Bradley-Terry techniques suffer from the following
defect, here illustrated using the context of Scottish football, and
the home advantage helping the home team and hindering the away team.
Consider the case of four teams 1-4 and represent the strength of the
home ground advantage as a home monster of strength $\mathfrak{H}$
(draws being discarded for the moment).  Now suppose that
$p_1=p_2=p_A$, and $p_3=p_4=p_B$.  The unit-sum constraint gives
$2p_A+2p_aB+\mathfrak{H}=1$.  The probability that team1 (at home)
beats team 2 (away) is then
$\frac{p_A+\mathfrak{H}}{2p_A+\mathfrak{H}}$; and the probability that
team 3 (home) beats team 4 (away) is
$\frac{p_B+\mathfrak{H}}{2p_B+\mathfrak{H}}$.  The odds ratios would
be $1+\mathfrak{H}/p_A$ and be $1+\mathfrak{H}/p_B$.  Such analysis is
incompatible with the perfectly reasonable requirement that the odds
ratio of home victory between two teams of equal strength be constant;
but here we see that weak teams are predicted to have an overwhelming
home-ground advantage (finding actual datasets that illustrate
precisely this phenomenon is not straightforward: reified
Bradley-Terry methods have a suprising amount of flexibility and can
accommodate wide ranges of strengths by dint of manipulating reified
strengths and real strengths).

However, consider the data presented in table~\ref{oldfirm} in which
we see football results from

\begin{table}
\centering
\begin{tabular}{@{} *{5}{c} @{}}
\headercell{\small{Plays at home,}\\ \small{wins}} &
\multicolumn{4}{c@{}}{\small{Plays away, loses}}\\
\cmidrule(l){2-5}
&   Rangers     &    Celtic & Falkirk & Livingston\\
\midrule
   Rangers      &   NA &    81  &    69   &      15\\
   Celtic       &   76 &    NA  &    61   &      15\\
   Falkirk      &   17 &    17  &    NA   &       6\\
   Livingston   &    2 &     2  &     3   &      NA\\
\end{tabular}
\\
\rule{1mm}{6mm}
\\
\begin{tabular}{@{} *{5}{c} @{}}
\headercell{\small{Plays at home,}\\ \small{draws}} &
\multicolumn{4}{c@{}}{\small{Plays away, draws}}\\
\cmidrule(l){2-5}
&  Rangers     &    Celtic & Falkirk & Livingston\\
\midrule
   Rangers     &       NA  &   44 &     14   &       2\\
   Celtic      &       47  &   NA &     10   &       3\\
   Falkirk     &       12  &   19 &     NA   &       4\\
   Livingston  &       4   &   4  &     4    &      NA\\
   \end{tabular}
\\
\rule{1mm}{6mm}
\\
\begin{tabular}{@{} *{5}{c} @{}}
\headercell{\small{Plays at home,}\\ \small{loses}} &
\multicolumn{4}{c@{}}{\small{Plays away, wins}}\\
\cmidrule(l){2-5}
&  Rangers     &    Celtic & Falkirk & Livingston\\
\midrule
   Rangers     &   NA  &  42  &    2  &       0\\
   Celtic      &   46  &  NA  &    8  &       0\\
   Falkirk     &   57  &  43  &   NA  &       8\\
   Livingston  &   12  &  11  &    9  &      NA\\
  \end{tabular}
\caption{The \label{oldfirm} dominance of the Old Firm: 669 matches
  between Rangers, Celtic, Falkirk and Livingston arranged by home
  wins, draws, and home losses.  Data from {\tt soccerworld}}
\end{table}

\section{Conclusions and further work}

consider Table~\ref{kka}, originally considered
by~\citet{hankin2020}.  Considering only non-drawn games for the
moment, we wish to represent the first-mover advantage using a ``White
monster'', of Bradley-Terry strength \white

\begin{table}
\centering
\begin{tabular}{@{} *{4}{c} @{}}
\headercell{\small{Plays White,}\\ \small{wins}} & \multicolumn{3}{c@{}}{\small{Plays black, loses}}\\
\cmidrule(l){2-4}
& An &  Kr & Ks        \\
\midrule
  An & -  &  18 &   6  \\
  Kr &  7 &  -  &  18  \\
  Ks & 15 &  30 &  -   \\
\end{tabular}\quad
\begin{tabular}{@{} *{4}{c} @{}}
\headercell{\small{Plays W,}\\ \small{draws}} & \multicolumn{3}{c@{}}{\small{Plays black, draws}}\\
 \cmidrule(l){2-4}
& An &  Kr & Ks        \\
\midrule
  An & -  &  20 &  20  \\
  Kr & 29 &  -  &  72  \\
  Ks & 26 &  57 &  -   \\
\end{tabular}\quad
\begin{tabular}{@{} *{4}{c} @{}}
\headercell{\small{Plays W,}\\ \small{loses}} & \multicolumn{3}{c@{}}{\small{Plays black, wins}}\\
\cmidrule(l){2-4}
& An &  Kr & Ks        \\
\midrule
  An & -  &   5 &   11  \\
  Kr & 13 &  -  &    9  \\
  Ks &  2 &   7 &  -   \\
\end{tabular}
\caption{Results of 365 chess games \label{kka} between Grandmaster
  Anand, Grandmaster Karpov, and Grandmaster Kasparov.  For example,
  Anand played white against Karpov a total of $18+20+13=51$ times and
  won 18, drew 20, lost 13}
\end{table}

\begin{table}
\centering
\begin{tabular}{l|cc|cc}\\&
  \multicolumn{2}{c}{$i$ plays at home} &
  \multicolumn{2}{c}{$i$ plays away} \\
 & Reified & $\lambda$ &
   Reified & $\lambda$ \\
  \hline
  $\Prob{(\mbox{$i$ wins})}$  &
  $\frac{p_i+\home\oo}{p_i+p_j+\home}$ &
  $\frac{\lambda p_i}{\lambda p_i+p_j}$ &
  $\frac{p_i}{p_i+p_j+\home}$ &
  $\frac{p_i}{\lambda p_i+p_j}$ \\
  $\Prob{(\mbox{$i$ loses})}$ &
  $\frac{p_j\oo}{p_i+p_j+\home}$   &
  $\frac{p_j}{p_i+\lambda p_j}$  &
  $\frac{p_j+\home}{p_i+p_j+\home}$ &
  $\frac{\lambda p_j}{p_i + \lambda p_j}$
  \\
\end{tabular}
\caption{Probabilities of a win, draw, loss for the two teams under a reified
  Bradley-Terry model\label{probabilities} and a lambda thing}
\end{table}

\section{Conclusions and further work}

Plackett-Luce likelihood functions for rank datasets have been

\bibliography{hyper2}
\end{document}

### 2025-02-17 inst/jss_hyper3_real.Rnw
more work

@@
   \begin{center}
-<<plotproflikeoldfirm, echo=FALSE, fig=TRUE>>=
+<<plotproflikeoldfirm, echo=FALSE>>=
 plot(lam,logLike - max(logLike),type='b')
@@
 
-\begin{figure}[htbp]
-  \begin{center}
-<<makecontour, fig=TRUE, cache=TRUE>>=
-n <- 2
-lambda <- seq(from=1, to=3, len=n)
-D <- seq(from=0.1, to=0.6, len=n)
-V <- as.matrix(expand.grid(lambda, D))
-LL <- apply(V, 1, f)
-LL <- LL - max(LL)
-contour(lambda, D, matrix(LL, n, n))
-@
-\caption{Contours of support\label{contourlambdaD} for a grid}
-\end{center}
-\end{figure}
-
 Thus \code{f()} gives us a likelihood function for
-$\mathbf{v}=(\lambda,D)$.  We may apply \code{f()} over a grid and
-thereby determine a likelihood surface, shown in
-figure~\ref{contourlambdaD}.  Further, we may find the evaluate using
+$\mathbf{v}=(\lambda,D)$.  We may find the evaluate using
 \code{optim()}:
@@
 
-<<testlambdazero, cache=TRUE>>=
+Then
+
+<<showevaluate, cache=TRUE>>=
+maxp(home_draw_away3(RCLF3_table, lambda = maxv[1], D=maxv[2]))
+@
+
+Such estimates have rich predictive power.  For example, if Rangers
+were to play Celtic with both teams away [thus nullifying the home
+  ground advantage: $\lambda=1$ in Table~\ref{windrawloss}], the
+probability of win/draw/loss for Rangers would be
+$\left(\frac{p_R}{(1+D)(p_R+p_C)};\frac{D}{1+D};\frac{p_R}{(1+D)(p_R+p_C)}\right)$,
+estimated at $(0.39;0.27;0.34)$.
+
+\begin{figure}[htbp]
+  \begin{center}
+<<makecontour, cache=TRUE>>=
+n <- 2
+lambda <- seq(from=1, to=3, len=n)
+D <- seq(from=0.1, to=0.6, len=n)
+V <- as.matrix(expand.grid(lambda, D))
+LL <- apply(V, 1, f)
+LL <- pmax(-20, LL - max(LL))
+contour(lambda, D, matrix(LL, n, n),levels = -2*(0:9), xlab=expression(lambda),ylab=expression(D))
+points(maxv[1], maxv[2], pch=16, col='red')
+@
+\caption{Contours of support\label{contourlambdaD} evaluated over
+  a grid.  Maximum likelihood estimate for $(\lambda, D)$ shown as
+  a red dot at the evaluate}
+  \end{center}
+\end{figure}
+
+<<testlambdazero, cache=TRUE, include=FALSE>>=
 jj <- optimize(f=function(D){f(c(lambda=1,D))}, interval=c(0.1, 0.4), maximum=TRUE)
-jj
 @
+
+Further, we may calculate a support function over the
+$(\lambda,D)$-plane, shown in figure~\ref{contourlambdaD}.  For
+example, we may test the hypothesis that $\lambda=1$ [corresponding to
+  the right hand side of figure~\ref{contourlambdaD}] for which the
+support is about \Sexpr{round(f(maxv) - jj$objective,2)}, far
+exceeding Edwards's two units of support per degree of freedom
+criteron.  Alternatively, we can observe that this lies in the tail
+region of its asymptotic null $\chi^2_4$ distribution with a $p$-value
+of about \Sexpr{signif(pchisq(f(maxv) -
+  jj$objective,lower.tail=FALSE,df=2),digits=2)}.
 

### 2025-02-17 inst/salad.Rmd
distinguish hankin2017 from hankin2017_rnw
incorporate Heather Turner's work

@@
 author: "R. K. S. Hankin"
-bibliography: hyper2.bib
+bibliography: hyper2.bib  # copy vignettes/hyper2.bib to inst/
 ---
@@
 
-To cite the `hyper2` package in publications, please use @hankin2017.
+To cite the `hyper2` package in publications, please use @hankin2017_rmd.
 This short document shows how to apply `hyper2` idiom to the `salad`
@@
 
+# Acid analysis
+
+Turner goes on to assess whether a model of tartness using acetic and
+gluconic acids is appropriate.  This is possible using the `hyper2`
+formalism as follows.
+
+```{r}
+features <- data.frame(salad = LETTERS[1:4],
+                       acetic = c(0.5, 0.5, 1, 0),
+                       gluconic = c(0, 10, 0, 10))
+features
+(M <- as.matrix(features[,-1]))
+makestrengths <- function(v){
+    out <- exp(M %*% v)
+    out <- c(out/sum(out))
+    out <- pmax(out,0)
+    names(out) <- LETTERS[1:4]
+    return(out)
+}
+makestrengths(c(2,1))
+f <- function(v){loglik(makestrengths(v),H2)}
+(maxo <- optim(par=c(3, 0.3), fn=f,control=list(fnscale = -1)))
+makestrengths(maxo$par)
+loglik(makestrengths(maxo$pa),H1)
+maxp(H1,give=1)
+```
+
 # References

### 2025-02-18 inst/jss_hyper3_real.Rnw
cheering (tikz diagram is bad atm)

@@
 \usepackage{amssymb}
+\usepackage{tikz}
 
@@
 <<makecontour, cache=TRUE>>=
-n <- 2
+n <- 20
 lambda <- seq(from=1, to=3, len=n)
@@
   jj$objective,lower.tail=FALSE,df=2),digits=2)}.
+
+\section{Non-independence in successive arrivals}
+
+In the following, I use the ``race'' metaphor: five runners take part
+in a race and arrive in order: $a\succ b\succ c\succ d\succ e$.
+Consider the following Plackett-Luce likelihood function for this
+observation:
+
+\begin{equation}
+\frac{p_a}{p_a + p_b + p_c + p_d + p_e}\cdot
+\frac{p_b}{      p_b + p_c + p_d + p_e}\cdot
+\frac{p_c}{            p_c + p_d + p_e}\cdot
+\frac{p_d}{                  p_d + p_e}\cdot
+\frac{p_e}{                        p_e}
+\end{equation}
+
+Note the use of multiplication between the terms, indicating
+conditional independence.  We may use the multiplicative
+generalization of Bradley-Terry strengths to introduce an element of
+non-independence between the terms.  There are many ways to do this,
+but one simple case would be to define equivalence classes of the
+competitors with each equivalence class comprising mutually supporting
+runners.  Equivalence classes of size one correspond to unsupported
+runners.  The metaphor would be that a runner who has finished the
+race is able to support other members of his equivalence class by
+cheering his teammates, boosting their performance.  The \pkg{hyper2}
+package includes function \code{cheering()} which implements this
+functionality.
+
+As an example, consider figure~\ref{tikzabcde}.  This shows a partial
+probability tree diagram for some of the $5!=120$ possible order
+statistics.  The standard Plackett-Luce likelihoods have been modified
+to account for two groups of mutually supporting runners:
+$\left\lbrace a,b,c\right\rbrace$ with support term $\lambda$, and
+$\left\lbrace d,e\right\rbrace$ with support term $\mu$.  From START,
+the first runner to cross the finishing line has standard
+Plackett-Luce probability.  Taking the top path as an example, we see
+that the likelihood function for $a\succ b\succ c\succ d\succ e$ would
+be
+
+\begin{figure}[htbp]
+  \begin{center}
+\usetikzlibrary{arrows}
+\usetikzlibrary{patterns}
+\begin{tikzpicture}[line cap=round,line join=round,>=triangle 45,x=1cm,y=1cm]
+\fill (0,0) circle[radius=2pt]; % root; paths abcde
+\draw (0,0) -- (2, 3) -- (4.8, 3);
+\draw (0,0) -- (2, 2) -- (4.8, 2);
+\draw (0,0) -- (2, 1) -- (4.8, 1);
+\draw (0,0) -- (2, 0) -- (4.8, 0);
+\draw (0,0) -- (2,-1) -- (4.8,-1);
+\node at (0,2.5) (eq1) {$\underbrace{\left\lbrace a,b,c\right\rbrace}_{\lambda}\underbrace{\left\lbrace d,e\right\rbrace}_{\mu}$};
+
+\node at (4.8, 3) {$a$};
+\node at (4.8, 2) {$b$};
+\node at (4.8, 1) {$c$};
+\node at (4.8, 0) {$d$};
+\node at (4.8,-1) {$e$};
+
+\node at (3.4, 3.4)  {$\frac{p_a}{p_a+p_b+p_c+p_d+p_e}$};
+\node at (3.4, 2.4)  {$\frac{p_b}{p_a+p_b+p_c+p_d+p_e}$};
+\node at (3.4, 1.4)  {$\frac{p_c}{p_a+p_b+p_c+p_d+p_e}$};
+\node at (3.4, 0.4)  {$\frac{p_d}{p_a+p_b+p_c+p_d+p_e}$};
+\node at (3.4,-0.6)  {$\frac{p_e}{p_a+p_b+p_c+p_d+p_e}$};
+
+\fill  (4.8, 3) circle[radius=2pt];  % a finishes; paths bcde
+\fill  (4.8, 2) circle[radius=2pt];  % terminal node
+\fill  (4.8, 1) circle[radius=2pt];  % terminal node
+\fill  (4.8, 0) circle[radius=2pt];  % d finishes; paths abce
+\fill  (4.8,-1) circle[radius=2pt];  % terminal node
+
+\draw (4.8,3) -- (5.8,5);
+\draw (4.8,3) -- (5.8,4);
+\draw (4.8,3) -- (5.8,3);
+\draw (4.8,3) -- (5.8,2);
+
+\node at (4.5,4.6) {$b$};
+\node at (4.5,3.7) {$c$};
+\node at (4.5,3.2) {$d$};
+\node at (4.5,2.7) {$e$};
+
+\newcommand{\la}[1]{\lambda{#1}}
+\newcommand{\ld}[1]{\mu    {#1}}
+\node at (6, 5)  {$\frac{\la{p_b}}{\la{p_b}+\la{p_c}+p_d+p_e}$};
+\node at (6, 4)  {$\frac{\la{p_c}}{\la{p_b}+\la{p_c}+p_d+p_e}$};
+\node at (6, 3)  {$\frac{p_d     }{\la{p_b}+\la{p_c}+p_d+p_e}$};
+\node at (6, 2)  {$\frac{p_e     }{\la{p_b}+\la{p_c}+p_d+p_e}$};
+
+\fill (7.4, 5) circle[radius=2pt];  % ab finishes; pahts cde
+\draw (7.4,5) -- (8.4,6);
+\draw (7.4,5) -- (8.4,5);
+\draw (7.4,5) -- (8.4,4);
+
+\node at (7.9,5.7) {$p_c$};
+\node at (7.9,5.2) {$p_d$};
+\node at (7.9,4.7) {$p_e$};
+\node at (9.3, 6) {$\frac{\la{p_c}}{\la{p_c}+p_d+p_e}$};
+\node at (9.3, 5) {$\frac{p_d  }{\la{p_c}+p_d+p_e}$};
+\node at (9.3, 4) {$\frac{p_e  }{\la{p_c}+p_d+p_e}$};
+
+\fill (10.2, 6) circle[radius=2pt];  % abc finishes; paths de
+\draw (10.2, 6) -- (11.2,7);
+\node at (10.7,6.8) {$p_d$};
+
+\node at (11.8, 7) {$\frac{p_d}{p_d+p_e}$};
+
+\draw (10.2, 6) -- (11.2,6.2);
+\node at (11.8, 6.2) {$\frac{p_e}{p_d+p_e}$};
+\node at (10.7,6.3) {$p_e$};
+
+\fill (10.2, 5) circle[radius=2pt];
+\draw (10.2, 5) -- (11.2,5.3);
+\node at (11.8, 5.3) {$\frac{\la{p_c}}{\la{p_c}+\ld{p_e}}$};
+\node at (10.7,5.3) {$p_c$};
+
+\draw (10.2, 5) -- (11.2,4.7);
+\node at (11.8, 4.7) {$\frac{\ld{p_e}}{\la{p_c}+\ld{p_e}}$};
+\node at (10.7,4.7) {$p_e$};
+
+\fill (10.2, 4) circle[radius=2pt];
+\draw (10.2, 4) -- (11.2,3.9);
+\node at (11.8, 3.9) {$\frac{\la{p_c}}{\la{p_c}+\ld{p_e}}$};
+\node at (10.7,4.1) {$p_c$};
+
+\draw (10.2, 4) -- (11.2,3.1);
+\node at (11.8, 3.1) {$\frac{\ld{p_e}}{\la{p_c}+\ld{p_e}}$};
+\node at (10.7,3.7) {$p_e$};
+
+\draw (4, 0) -- (5,1);
+\node at (6, 1)  {$\frac{p_a}{p_a+p_b+p_c+\ld{p_e}}$};
+
+\draw (4, 0) -- (5,0);
+\node at (6, 0)  {$\frac{p_b}{p_a+p_b+p_c+\ld{p_e}}$};
+
+\draw (4, 0) -- (5,-1);
+\node at (6, -1)  {$\frac{p_c}{p_a+p_b+p_c+\ld{p_e}}$};
+
+\draw (4, 0) -- (5,-2);
+\node at (6, -2)  {$\frac{\ld{p_e}}{p_a+p_b+p_c+\ld{p_e}}$};
+
+\node at (4.5,0.7) {$a$};
+\node at (4.5,0.2) {$b$};
+\node at (4.5,-0.3) {$c$};
+\node at (4.5,-0.8) {$e$};
+
+\fill (7.4, -2) circle[radius=2pt];  % de finishes; paths abc
+\draw (7.4, -2) -- (8.4,-1);
+\draw (7.4, -2) -- (8.4,-2);
+\draw (7.4, -2) -- (8.4,-3);
+
+\node at (7.9,-1.2) {$p_a$};
+\node at (7.9,-1.8) {$p_b$};
+\node at (7.9,-2.3) {$p_c$};
+
+\node at (9.3, -1) {$\frac{p_a}{p_a+p_b+c}$};
+\node at (9.3, -2) {$\frac{p_b}{p_a+p_b+c}$};
+\node at (9.3, -3) {$\frac{p_c}{p_a+p_b+c}$};
+
+\fill (10.2, -1) circle[radius=2pt];  % dea finishes; paths bc
+\fill (10.2, -2) circle[radius=2pt];  % deb finishes; paths ac
+\fill (10.2, -3) circle[radius=2pt];  % dec finishes; paths ab
+
+\draw (10.2, -1) -- (11.2,-0);
+\node at (10.6,-0.4) {$b$};
+\node at (10.6,-0.8) {$c$};
+
+\node at (11.8, 0) {$\frac{\la{p_b}}{\la{p_b}+\la{p_c}}$};
+\draw (10.2, -1) -- (11.2,-1);
+
+\node at (11.8, -1) {$\frac{\la{p_c}}{\la{p_b}+\la{p_c}}$};
+
+\draw (10.2, -2) -- (11.2,-1.6);
+\node at (11.8, -1.6) {$\frac{\la{p_a}}{\la{p_a}+\la{p_c}}$};
+
+\node at (10.6,-1.7) {$a$};
+\node at (10.6,-2) {$c$};
+
+\draw (10.2, -2) -- (11.2,-2.2);
+\node at (11.8, -2.2) {$\frac{\la{p_c}}{\la{p_a}+\la{p_c}}$};
+
+\draw (10.2, -3) -- (11.2,-2.9);
+\node at (11.8, -2.9) {$\frac{\la{p_a}}{\la{p_a}+\la{p_b}}$};
+
+\draw (10.2, -3) -- (11.2,-3.4);
+\node at (11.8, -3.4) {$\frac{\la{p_b}}{\la{p_a}+\la{p_b}}$};
+
+\node at (10.6,-2.8) {$a$};
+\node at (10.6,-3.3) {$b$};
+
+\fill (7.4, 1) circle[radius=2pt];  % da finishes; paths bce
+
+\draw (7.4, 1) -- (8.4,2);
+\draw (7.4, 1) -- (8.4,1);
+\draw (7.4, 1) -- (8.4,0);
+
+\node at (7.9,1.7) {$b$};
+\node at (7.9,1.2) {$c$};
+\node at (7.9,0.7) {$e$};
+
+\node at (9.3, 2) {$\frac{\la{p_b}}{\la{p_b}+\la{p_c}+p_e}$};
+\node at (9.3, 1) {$\frac{\la{p_c}}{\la{p_b}+\la{p_c}+p_e}$};
+\node at (9.3, 0) {$\frac{e       }{\la{p_b}+\la{p_c}+p_e}$};
+
+\fill (10.2, 1) circle[radius=2pt];  % da finishes; paths bce
+\draw (10.2, 1) -- (11.2,2);
+\node at (11.8, 2) {$\frac{\la{p_b}}{\la{p_b}+\ld{p_e}}$};
+\draw (10.2, 1) -- (11.2,1);
+\node at (11.8, 1) {$\frac{\ld{p_e}}{\la{p_b}+p_e}$};
+
+\end{tikzpicture}
+\caption{Partial probability\label{tikzabcde} tree for five competitors
+  $a$-$e$ with mutually supporting subsets $\left\lbrace a,b,c\right\rbrace$
+  [with support term $\lambda$] and $(de)$ [with support term $\mu$]}
+\end{center}
+\end{figure}
+
+\begin{equation}\label{asuccb}
+\frac{p_a}{p_a + p_b + p_c + p_d + p_e}\cdot
+\frac{\lambda p_b}{\lambda p_b + \lambda p_c + p_d + p_e}\cdot
+\frac{\lambda p_c}{\lambda p_c + p_d + p_e}\cdot
+\frac{p_d}{p_d + p_e}
+\end{equation}
+
+In equation \ref{asuccb}, the first term is standard Plackett-Luce: at
+this point, no competitor has finished and cheering effects are
+absent.  The second term reflects the fact that competitors $b$ and
+$c$ are supported by competitor $a$, who has by this point finished
+the race and is supporting his teammates.
+
+By contrast, the likelihood function for observation $d\succ a\succ
+c\succ b\succ e$ would be
+
+\begin{equation}
+\frac{p_d}{p_a + p_b + p_c + p_d + p_e}\cdot
+\frac{p_a}{p_a + p_b + p_c + \mu p_e}\cdot
+\frac{\lambda p_c}{\lambda p_b + \lambda p_c + p_e}\cdot
+\frac{\lambda p_b}{\lambda p_b + \mu p_e}
+\end{equation}
+
+where this likelihood function reflects the mutual support for
+equivalence class $\left\lbrace d,e\right\rbrace$.  Note that the
+final term reflects the fact that competitors $b$ and $e$ have their
+support active when vying for fourth place, as members of both their
+teams have finished at this point.
+
+This probability model could easily be modified to account for
+specific circumstances.  The cheering effect be asymmetrical (with $a$
+helping $b$ but not vice-versa, for example).  The effect could
+operate only on specific ordered pairs.  Or perhaps the effect might
+have a finite lifetime: if $a$ places $n^\mathrm{th}$, then the
+cheering effect is active only for competitors placing
+$(n+r)^\mathrm{th}$ or above.
 

### 2025-02-18 inst/loadlibrary.R

@@ -1,3 +1,4 @@
 suppressMessages(library("knitr",quietly=TRUE,verbose=FALSE))
 suppressMessages(library("hyper2",quietly=TRUE,verbose=FALSE))
 suppressMessages(library("magrittr",quietly=TRUE,verbose=FALSE))
+suppressMessages(library("tikzDevice",quietly=TRUE,verbose=FALSE))

### 2025-02-18 inst/jss_hyper3_real.Rnw
work on the diagram

@@
 
-\node at (4.8, 3) {$a$};
-\node at (4.8, 2) {$b$};
-\node at (4.8, 1) {$c$};
-\node at (4.8, 0) {$d$};
-\node at (4.8,-1) {$e$};
+\node at (4.8, 2.8) {$a$};
+\node at (4.8, 1.8) {$b$};
+\node at (4.8, 0.8) {$c$};
+\node at (4.8,-0.2) {$d$};
+\node at (4.8,-1.2) {$e$};
 
@@
 
-\fill  (4.8, 3) circle[radius=2pt];  % a finishes; paths bcde
-\fill  (4.8, 2) circle[radius=2pt];  % terminal node
-\fill  (4.8, 1) circle[radius=2pt];  % terminal node
-\fill  (4.8, 0) circle[radius=2pt];  % d finishes; paths abce
-\fill  (4.8,-1) circle[radius=2pt];  % terminal node
+\fill[red] (4.8, 3) circle[radius=2pt];  % a finishes; paths bcde
+\fill[red] (4.8, 2) circle[radius=2pt];  % terminal node
+\fill[red] (4.8, 1) circle[radius=2pt];  % terminal node
+\fill[red] (4.8, 0) circle[radius=2pt];  % d finishes; paths abce
+\fill[red] (4.8,-1) circle[radius=2pt];  % terminal node
 
-\draw (4.8,3) -- (5.8,5);
-\draw (4.8,3) -- (5.8,4);
-\draw (4.8,3) -- (5.8,3);
-\draw (4.8,3) -- (5.8,2);
-
-\node at (4.5,4.6) {$b$};
-\node at (4.5,3.7) {$c$};
-\node at (4.5,3.2) {$d$};
-\node at (4.5,2.7) {$e$};
+\draw (4.8,3) -- (5.8,5) -- (8.2,5);
+\draw (4.8,3) -- (5.8,4) -- (8.2,4);
+\draw (4.8,3) -- (5.8,3) -- (8.2,3);
+\draw (4.8,3) -- (5.8,2) -- (8.2,2);
 
@@
 \newcommand{\ld}[1]{\mu    {#1}}
-\node at (6, 5)  {$\frac{\la{p_b}}{\la{p_b}+\la{p_c}+p_d+p_e}$};
-\node at (6, 4)  {$\frac{\la{p_c}}{\la{p_b}+\la{p_c}+p_d+p_e}$};
-\node at (6, 3)  {$\frac{p_d     }{\la{p_b}+\la{p_c}+p_d+p_e}$};
-\node at (6, 2)  {$\frac{p_e     }{\la{p_b}+\la{p_c}+p_d+p_e}$};
+\node at (7, 5.4)  {$\frac{\la{p_b}}{\la{p_b}+\la{p_c}+p_d+p_e}$};
+\node at (7, 4.4)  {$\frac{\la{p_c}}{\la{p_b}+\la{p_c}+p_d+p_e}$};
+\node at (7, 3.4)  {$\frac{p_d     }{\la{p_b}+\la{p_c}+p_d+p_e}$};
+\node at (7, 2.4)  {$\frac{p_e     }{\la{p_b}+\la{p_c}+p_d+p_e}$};
 
-\fill (7.4, 5) circle[radius=2pt];  % ab finishes; pahts cde
-\draw (7.4,5) -- (8.4,6);
-\draw (7.4,5) -- (8.4,5);
-\draw (7.4,5) -- (8.4,4);
+\fill[blue] (8.2,5) circle[radius=2pt];  % ab finishes; paths cde
+\fill[blue] (8.2,4) circle[radius=2pt];  % ab finishes; paths cde
+\fill[blue] (8.2,3) circle[radius=2pt];  % ab finishes; paths cde
+\fill[blue] (8.2,2) circle[radius=2pt];  % ab finishes; paths cde
+\draw (8.2,5) -- (8.9,6) -- (10.6,6);
+\draw (8.2,5) -- (8.9,5) -- (10.6,5);
+\draw (8.2,5) -- (8.9,4) -- (10.6,4);
 
-\node at (7.9,5.7) {$p_c$};
-\node at (7.9,5.2) {$p_d$};
-\node at (7.9,4.7) {$p_e$};
-\node at (9.3, 6) {$\frac{\la{p_c}}{\la{p_c}+p_d+p_e}$};
-\node at (9.3, 5) {$\frac{p_d  }{\la{p_c}+p_d+p_e}$};
-\node at (9.3, 4) {$\frac{p_e  }{\la{p_c}+p_d+p_e}$};
+\node at (9.5, 6.4) {$\frac{\la{p_c}}{\la{p_c}+p_d+p_e}$};
+\node at (9.5, 5.4) {$\frac{p_d  }{\la{p_c}+p_d+p_e}$};
+\node at (9.5, 4.4) {$\frac{p_e  }{\la{p_c}+p_d+p_e}$};
 
-\fill (10.2, 6) circle[radius=2pt];  % abc finishes; paths de
-\draw (10.2, 6) -- (11.2,7);
-\node at (10.7,6.8) {$p_d$};
+\fill[green] (10.6, 6) circle[radius=2pt];  % abc finishes; paths de
+\draw (10.6, 6) -- (11.2,7) -- (12.2,7);
+\fill[red] (12.2,7) circle[radius=2pt];
+\node at (11.8, 7.4) {$\frac{p_d}{p_d+p_e}$};
 
-\node at (11.8, 7) {$\frac{p_d}{p_d+p_e}$};
+\draw (10.6, 6) -- (11.2,6.2) -- (12.2,6.2);
+\fill[red] (12.2,6.2) circle[radius=2pt];
+\node at (11.8, 6.6) {$\frac{p_e}{p_d+p_e}$};
 
-\draw (10.2, 6) -- (11.2,6.2);
-\node at (11.8, 6.2) {$\frac{p_e}{p_d+p_e}$};
-\node at (10.7,6.3) {$p_e$};
+\fill[green] (10.6, 5) circle[radius=2pt];
+\draw (10.6, 5) -- (11.2,5.3) -- (12.2, 5.3);
+\fill[red] (12.2,5.3) circle[radius=2pt];
+\node at (11.8, 5.7) {$\frac{\la{p_c}}{\la{p_c}+\ld{p_e}}$};
 
-\fill (10.2, 5) circle[radius=2pt];
-\draw (10.2, 5) -- (11.2,5.3);
-\node at (11.8, 5.3) {$\frac{\la{p_c}}{\la{p_c}+\ld{p_e}}$};
-\node at (10.7,5.3) {$p_c$};
+\draw (10.6, 5) -- (11.2,4.7) -- (12.2, 4.7);
+\fill[red] (12.2,4.7) circle[radius=2pt];
+\node at (11.8, 5.0) {$\frac{\ld{p_e}}{\la{p_c}+\ld{p_e}}$};
 
-\draw (10.2, 5) -- (11.2,4.7);
-\node at (11.8, 4.7) {$\frac{\ld{p_e}}{\la{p_c}+\ld{p_e}}$};
-\node at (10.7,4.7) {$p_e$};
+\fill[green] (10.6, 4) circle[radius=2pt];
+\draw (10.6, 4) -- (11.2,3.9) -- (12.2, 3.9);
+\fill[red] (12.2,3.9) circle[radius=2pt];
+\node at (11.8, 4.3) {$\frac{\la{p_c}}{\la{p_c}+\ld{p_e}}$};
 
-\fill (10.2, 4) circle[radius=2pt];
-\draw (10.2, 4) -- (11.2,3.9);
-\node at (11.8, 3.9) {$\frac{\la{p_c}}{\la{p_c}+\ld{p_e}}$};
-\node at (10.7,4.1) {$p_c$};
+\draw (10.6, 4) -- (11.2,3.1) -- (12.2,3.1);
+\fill[red] (12.2,3.1) circle[radius=2pt];
+\node at (11.8, 3.5) {$\frac{\ld{p_e}}{\la{p_c}+\ld{p_e}}$};
 
-\draw (10.2, 4) -- (11.2,3.1);
-\node at (11.8, 3.1) {$\frac{\ld{p_e}}{\la{p_c}+\ld{p_e}}$};
-\node at (10.7,3.7) {$p_e$};
+\draw (4.8, 0) -- (5.8,1) -- (8.2,1);
+\fill[red] (12.2,1) circle[radius=2pt];
 
-\draw (4, 0) -- (5,1);
-\node at (6, 1)  {$\frac{p_a}{p_a+p_b+p_c+\ld{p_e}}$};
+\node at (7, 1.4)  {$\frac{p_a}{p_a+p_b+p_c+\ld{p_e}}$};
 
-\draw (4, 0) -- (5,0);
-\node at (6, 0)  {$\frac{p_b}{p_a+p_b+p_c+\ld{p_e}}$};
+\draw (4.8, 0) -- (5.8,0) -- (8.2,0);
+\fill[red] (12.2,0) circle[radius=2pt];
+\node at (7, 0.4)  {$\frac{p_b}{p_a+p_b+p_c+\ld{p_e}}$};
 
-\draw (4, 0) -- (5,-1);
-\node at (6, -1)  {$\frac{p_c}{p_a+p_b+p_c+\ld{p_e}}$};
+\draw (4.8, 0) -- (5.8,-1) -- (8.2,-1);
+\node at (7, -0.6)  {$\frac{p_c}{p_a+p_b+p_c+\ld{p_e}}$};
 
-\draw (4, 0) -- (5,-2);
-\node at (6, -2)  {$\frac{\ld{p_e}}{p_a+p_b+p_c+\ld{p_e}}$};
+\draw (4.8, 0) -- (5.8,-2) -- (8.2,-2);
+\node at (7, -1.6)  {$\frac{\ld{p_e}}{p_a+p_b+p_c+\ld{p_e}}$};
 
-\node at (4.5,0.7) {$a$};
-\node at (4.5,0.2) {$b$};
-\node at (4.5,-0.3) {$c$};
-\node at (4.5,-0.8) {$e$};
+\fill[blue] (8.2, -2) circle[radius=2pt];  % de finishes; paths abc
+\fill[blue] (8.2, -1) circle[radius=2pt];  % de finishes; paths abc
+\fill[blue] (8.2,  0) circle[radius=2pt];  % de finishes; paths abc
+\draw (8.2, -2) -- (8.9,-1) -- (10.6,-1);
+\draw (8.2, -2) -- (8.9,-2) -- (10.6,-2);
+\draw (8.2, -2) -- (8.9,-3) -- (10.6,-3);
 
-\fill (7.4, -2) circle[radius=2pt];  % de finishes; paths abc
-\draw (7.4, -2) -- (8.4,-1);
-\draw (7.4, -2) -- (8.4,-2);
-\draw (7.4, -2) -- (8.4,-3);
+\fill[red] (12.2,-1) circle[radius=2pt];
 
-\node at (7.9,-1.2) {$p_a$};
-\node at (7.9,-1.8) {$p_b$};
-\node at (7.9,-2.3) {$p_c$};
+\node at (9.5, -0.6) {$\frac{p_a}{p_a+p_b+p_c}$};
+\node at (9.5, -1.6) {$\frac{p_b}{p_a+p_b+p_c}$};
+\node at (9.5, -2.6) {$\frac{p_c}{p_a+p_b+p_c}$};
 
-\node at (9.3, -1) {$\frac{p_a}{p_a+p_b+c}$};
-\node at (9.3, -2) {$\frac{p_b}{p_a+p_b+c}$};
-\node at (9.3, -3) {$\frac{p_c}{p_a+p_b+c}$};
+\fill[green] (10.6, -1) circle[radius=2pt];  % dea finishes; paths bc
+\fill[green] (10.6, -2) circle[radius=2pt];  % deb finishes; paths ac
+\fill[green] (10.6, -3) circle[radius=2pt];  % dec finishes; paths ab
 
-\fill (10.2, -1) circle[radius=2pt];  % dea finishes; paths bc
-\fill (10.2, -2) circle[radius=2pt];  % deb finishes; paths ac
-\fill (10.2, -3) circle[radius=2pt];  % dec finishes; paths ab
+\draw (10.6, -1) -- (11.2,-0) -- (12.2,0);
 
-\draw (10.2, -1) -- (11.2,-0);
-\node at (10.6,-0.4) {$b$};
-\node at (10.6,-0.8) {$c$};
+\node at (11.8, 0.4) {$\frac{\la{p_b}}{\la{p_b}+\la{p_c}}$};
+\draw (10.6, -1) -- (11.2,-1) -- (12.2,-1);
 
-\node at (11.8, 0) {$\frac{\la{p_b}}{\la{p_b}+\la{p_c}}$};
-\draw (10.2, -1) -- (11.2,-1);
+\node at (11.8, -0.6) {$\frac{\la{p_c}}{\la{p_b}+\la{p_c}}$};
 
-\node at (11.8, -1) {$\frac{\la{p_c}}{\la{p_b}+\la{p_c}}$};
+\draw(10.6, -2) -- (11.2,-1.8) -- (12.2,-1.8);
+\fill[red] (12.2,-1.8) circle[radius=2pt];
 
-\draw (10.2, -2) -- (11.2,-1.6);
-\node at (11.8, -1.6) {$\frac{\la{p_a}}{\la{p_a}+\la{p_c}}$};
+\node at (11.8, -1.4) {$\frac{\la{p_a}}{\la{p_a}+\la{p_c}}$};
 
-\node at (10.6,-1.7) {$a$};
-\node at (10.6,-2) {$c$};
-
-\draw (10.2, -2) -- (11.2,-2.2);
+\draw (10.6, -2) -- (11.2,-2.6) -- (12.2,-2.6);
+\fill[red] (12.2,-2.6) circle[radius=2pt];
 \node at (11.8, -2.2) {$\frac{\la{p_c}}{\la{p_a}+\la{p_c}}$};
 
-\draw (10.2, -3) -- (11.2,-2.9);
-\node at (11.8, -2.9) {$\frac{\la{p_a}}{\la{p_a}+\la{p_b}}$};
+%\draw (10.6, -3) -- (11.2,-2.9);
+%\fill[red] (12.2,-2.9) circle[radius=2pt];
+%\node[red] at (11.8, -2.9) {$\frac{\la{p_a}}{\la{p_a}+\la{p_b}}$};
 
-\draw (10.2, -3) -- (11.2,-3.4);
-\node at (11.8, -3.4) {$\frac{\la{p_b}}{\la{p_a}+\la{p_b}}$};
+\draw (10.6, -3) -- (11.2,-3.4) -- (12.2, -3.4);
+\fill[red] (12.2,-3.4) circle[radius=2pt];
+\node at (11.8, -3) {$\frac{\la{p_b}}{\la{p_a}+\la{p_b}}$};
 
-\node at (10.6,-2.8) {$a$};
-\node at (10.6,-3.3) {$b$};
+\fill[blue] (8.2, 1) circle[radius=2pt];  % da finishes; paths bce
 
-\fill (7.4, 1) circle[radius=2pt];  % da finishes; paths bce
+\draw (8.2,1) -- (8.9,2) -- (10.6,2);
+\draw (8.2,1) -- (8.9,1) -- (10.6,1);
+\draw (8.2,1) -- (8.9,0) -- (10.6,0);
 
-\draw (7.4, 1) -- (8.4,2);
-\draw (7.4, 1) -- (8.4,1);
-\draw (7.4, 1) -- (8.4,0);
-
-\node at (7.9,1.7) {$b$};
-\node at (7.9,1.2) {$c$};
-\node at (7.9,0.7) {$e$};
-
-\node at (9.3, 2) {$\frac{\la{p_b}}{\la{p_b}+\la{p_c}+p_e}$};
-\node at (9.3, 1) {$\frac{\la{p_c}}{\la{p_b}+\la{p_c}+p_e}$};
-\node at (9.3, 0) {$\frac{e       }{\la{p_b}+\la{p_c}+p_e}$};
-
-\fill (10.2, 1) circle[radius=2pt];  % da finishes; paths bce
-\draw (10.2, 1) -- (11.2,2);
-\node at (11.8, 2) {$\frac{\la{p_b}}{\la{p_b}+\ld{p_e}}$};
-\draw (10.2, 1) -- (11.2,1);
-\node at (11.8, 1) {$\frac{\ld{p_e}}{\la{p_b}+p_e}$};
+\node at (9.5, 2.4) {$\frac{\la{p_b}}{\la{p_b}+\la{p_c}+p_e}$};
+\node at (9.5, 1.4) {$\frac{\la{p_c}}{\la{p_b}+\la{p_c}+p_e}$};
+\node at (9.5, 0.4) {$\frac{p_e     }{\la{p_b}+\la{p_c}+p_e}$};
+\fill[green] (10.6, 2) circle[radius=2pt];  % da finishes; paths bce
+\fill[green] (10.6, 1) circle[radius=2pt];  % da finishes; paths bce
+\fill[green] (10.6, 0) circle[radius=2pt];  % da finishes; paths bce
+\draw (10.6, 1) -- (11.2,2) -- (12.2,2);
+\node at (11.8, 2.4) {$\frac{\la{p_b}}{\la{p_b}+\ld{p_e}}$};
+\draw (10.6, 1) -- (11.2,1) -- (12.2,1);
+\node at (11.8, 1.4) {$\frac{\ld{p_e}}{\la{p_b}+p_e}$};
 

### 2025-02-18 inst/jss_hyper3_real.Rnw
further work on probability tree diagram

@@
 \fill (0,0) circle[radius=2pt]; % root; paths abcde
-\draw (0,0) -- (2, 3) -- (4.8, 3);
-\draw (0,0) -- (2, 2) -- (4.8, 2);
-\draw (0,0) -- (2, 1) -- (4.8, 1);
-\draw (0,0) -- (2, 0) -- (4.8, 0);
-\draw (0,0) -- (2,-1) -- (4.8,-1);
+\draw[thick] (0,0) -- (2, 3) -- (4.8, 3);
+\draw[thick] (0,0) -- (2, 2) -- (4.8, 2);
+\draw[thick] (0,0) -- (2, 1) -- (4.8, 1);
+\draw[thick] (0,0) -- (2, 0) -- (4.8, 0);
+\draw[thick] (0,0) -- (2,-1) -- (4.8,-1);
 \node at (0,2.5) (eq1) {$\underbrace{\left\lbrace a,b,c\right\rbrace}_{\lambda}\underbrace{\left\lbrace d,e\right\rbrace}_{\mu}$};
@@
 
-\fill[red] (4.8, 3) circle[radius=2pt];  % a finishes; paths bcde
-\fill[red] (4.8, 2) circle[radius=2pt];  % terminal node
-\fill[red] (4.8, 1) circle[radius=2pt];  % terminal node
-\fill[red] (4.8, 0) circle[radius=2pt];  % d finishes; paths abce
-\fill[red] (4.8,-1) circle[radius=2pt];  % terminal node
+\fill (4.8, 3) circle[radius=2pt];  % a finishes; paths bcde
+\fill (4.8, 2) circle[radius=2pt];  % terminal node
+\fill (4.8, 1) circle[radius=2pt];  % terminal node
+\fill (4.8, 0) circle[radius=2pt];  % d finishes; paths abce
+\fill (4.8,-1) circle[radius=2pt];  % terminal node
 
-\draw (4.8,3) -- (5.8,5) -- (8.2,5);
-\draw (4.8,3) -- (5.8,4) -- (8.2,4);
-\draw (4.8,3) -- (5.8,3) -- (8.2,3);
-\draw (4.8,3) -- (5.8,2) -- (8.2,2);
+\draw[thick] (4.8,3) -- (5.8,5) -- (8.2,5);
+\draw[thick] (4.8,3) -- (5.8,4) -- (8.2,4);
+\draw[thick] (4.8,3) -- (5.8,3) -- (8.2,3);
+\draw[thick] (4.8,3) -- (5.8,2) -- (8.2,2);
 
@@
 
-\fill[blue] (8.2,5) circle[radius=2pt];  % ab finishes; paths cde
-\fill[blue] (8.2,4) circle[radius=2pt];  % ab finishes; paths cde
-\fill[blue] (8.2,3) circle[radius=2pt];  % ab finishes; paths cde
-\fill[blue] (8.2,2) circle[radius=2pt];  % ab finishes; paths cde
-\draw (8.2,5) -- (8.9,6) -- (10.6,6);
-\draw (8.2,5) -- (8.9,5) -- (10.6,5);
-\draw (8.2,5) -- (8.9,4) -- (10.6,4);
+\fill (8.2,5) circle[radius=2pt];  % ab finishes; paths cde
+\fill (8.2,4) circle[radius=2pt];  % ab finishes; paths cde
+\fill (8.2,3) circle[radius=2pt];  % ab finishes; paths cde
+\fill (8.2,2) circle[radius=2pt];  % ab finishes; paths cde
+
+\node at (8.2,4.8) {$b$};
+\node at (8.2,3.8) {$c$};
+\node at (8.2,2.8) {$d$};
+\node at (8.2,1.8) {$e$};
+
+\draw[thick] (8.2,5) -- (8.9,6) -- (10.6,6);
+\draw[thick] (8.2,5) -- (8.9,5) -- (10.6,5);
+\draw[thick] (8.2,5) -- (8.9,4) -- (10.6,4);
 
@@
 
-\fill[green] (10.6, 6) circle[radius=2pt];  % abc finishes; paths de
-\draw (10.6, 6) -- (11.2,7) -- (12.2,7);
-\fill[red] (12.2,7) circle[radius=2pt];
+\fill (10.6, 6) circle[radius=2pt];  % abc finishes; paths de
+\node at (10.6,5.8) {$c$};
+\draw[thick] (10.6, 6) -- (11.2,7) -- (12.2,7);
+\fill (12.2,7) circle[radius=2pt];
+\node[anchor = west] at (12.2,7) {$a\succ b\succ c\succ d\succ e$};
 \node at (11.8, 7.4) {$\frac{p_d}{p_d+p_e}$};
 
-\draw (10.6, 6) -- (11.2,6.2) -- (12.2,6.2);
-\fill[red] (12.2,6.2) circle[radius=2pt];
+\draw[thick] (10.6, 6) -- (11.2,6.2) -- (12.2,6.2);
+\fill (12.2,6.2) circle[radius=2pt];
+\node[anchor = west] at (12.2,6.2) {$a\succ b\succ c\succ e\succ d$};
 \node at (11.8, 6.6) {$\frac{p_e}{p_d+p_e}$};
 
-\fill[green] (10.6, 5) circle[radius=2pt];
-\draw (10.6, 5) -- (11.2,5.3) -- (12.2, 5.3);
-\fill[red] (12.2,5.3) circle[radius=2pt];
+\fill (10.6, 5) circle[radius=2pt];
+\node at (10.6,4.8) {$d$};
+
+\draw[thick] (10.6, 5) -- (11.2,5.3) -- (12.2, 5.3);
+\fill (12.2,5.3) circle[radius=2pt];
+\node[anchor = west] at (12.2,5.3) {$a\succ b\succ d\succ c\succ e$};
 \node at (11.8, 5.7) {$\frac{\la{p_c}}{\la{p_c}+\ld{p_e}}$};
 
-\draw (10.6, 5) -- (11.2,4.7) -- (12.2, 4.7);
-\fill[red] (12.2,4.7) circle[radius=2pt];
+\draw[thick] (10.6, 5) -- (11.2,4.7) -- (12.2, 4.7);
+\fill (12.2,4.7) circle[radius=2pt];
+\node[anchor = west] at (12.2,4.7) {$a\succ b\succ d\succ e\succ c$};
 \node at (11.8, 5.0) {$\frac{\ld{p_e}}{\la{p_c}+\ld{p_e}}$};
 
-\fill[green] (10.6, 4) circle[radius=2pt];
-\draw (10.6, 4) -- (11.2,3.9) -- (12.2, 3.9);
-\fill[red] (12.2,3.9) circle[radius=2pt];
-\node at (11.8, 4.3) {$\frac{\la{p_c}}{\la{p_c}+\ld{p_e}}$};
+\fill (10.6, 4) circle[radius=2pt];
+\node at (10.6,3.8) {$e$};
+\draw[thick] (10.6, 4) -- (11.2,3.9) -- (12.2, 3.9);
+\fill (12.2,3.9) circle[radius=2pt];
+\node[anchor = west] at (12.2,3.9) {$a\succ b\succ e\succ c\succ d$};
+\node at (11.8, 4.3) {$\frac{\la{p_c}}{\la{p_c}+\ld{p_d}}$};
 
-\draw (10.6, 4) -- (11.2,3.1) -- (12.2,3.1);
-\fill[red] (12.2,3.1) circle[radius=2pt];
-\node at (11.8, 3.5) {$\frac{\ld{p_e}}{\la{p_c}+\ld{p_e}}$};
+\draw[thick] (10.6, 4) -- (11.2,3.1) -- (12.2,3.1);
+\fill (12.2,3.1) circle[radius=2pt];
+\node at (11.8, 3.5) {$\frac{\ld{p_d}}{\la{p_c}+\ld{p_d}}$};
+\node[anchor = west] at (12.2,3.1) {$a\succ b\succ e\succ d\succ c$};
 
-\draw (4.8, 0) -- (5.8,1) -- (8.2,1);
-\fill[red] (12.2,1) circle[radius=2pt];
-
+\draw[thick] (4.8, 0) -- (5.8,1) -- (8.2,1);
+\fill (12.2,1) circle[radius=2pt];
+\node[anchor = west] at (12.2,1) {$d\succ a\succ c\succ e\succ b$};
 \node at (7, 1.4)  {$\frac{p_a}{p_a+p_b+p_c+\ld{p_e}}$};
 
-\draw (4.8, 0) -- (5.8,0) -- (8.2,0);
-\fill[red] (12.2,0) circle[radius=2pt];
+\draw[thick] (4.8, 0) -- (5.8,0) -- (8.2,0);
+\fill (12.2,0) circle[radius=2pt];
+\node[anchor = west] at (12.2,0) {$d\succ e\succ a\succ b\succ c$};
 \node at (7, 0.4)  {$\frac{p_b}{p_a+p_b+p_c+\ld{p_e}}$};
 
-\draw (4.8, 0) -- (5.8,-1) -- (8.2,-1);
+\draw[thick] (4.8, 0) -- (5.8,-1) -- (8.2,-1);
 \node at (7, -0.6)  {$\frac{p_c}{p_a+p_b+p_c+\ld{p_e}}$};
 
-\draw (4.8, 0) -- (5.8,-2) -- (8.2,-2);
+\draw[thick] (4.8, 0) -- (5.8,-2) -- (8.2,-2);
 \node at (7, -1.6)  {$\frac{\ld{p_e}}{p_a+p_b+p_c+\ld{p_e}}$};
 
-\fill[blue] (8.2, -2) circle[radius=2pt];  % de finishes; paths abc
-\fill[blue] (8.2, -1) circle[radius=2pt];  % de finishes; paths abc
-\fill[blue] (8.2,  0) circle[radius=2pt];  % de finishes; paths abc
-\draw (8.2, -2) -- (8.9,-1) -- (10.6,-1);
-\draw (8.2, -2) -- (8.9,-2) -- (10.6,-2);
-\draw (8.2, -2) -- (8.9,-3) -- (10.6,-3);
+\fill (8.2, -2) circle[radius=2pt];  % de finishes; paths abc
+\fill (8.2, -1) circle[radius=2pt];  % de finishes; paths abc
+\fill (8.2,  0) circle[radius=2pt];  % de finishes; paths abc
 
-\fill[red] (12.2,-1) circle[radius=2pt];
+\node at (8.2, -2.2) {$e$};
+\node at (8.2, -1.2) {$c$};
+\node at (8.2, -0.2) {$b$};
+\node at (8.2,  0.8) {$a$};
+
+\draw[thick] (8.2, -2) -- (8.9,-1) -- (10.6,-1);
+\draw[thick] (8.2, -2) -- (8.9,-2) -- (10.6,-2);
+\draw[thick] (8.2, -2) -- (8.9,-3) -- (10.6,-3);
+
+\fill (12.2,-1) circle[radius=2pt];
+\node[anchor = west] at (12.2,-1) {$d\succ e\succ a\succ c\succ b$};
 
@@
 
-\fill[green] (10.6, -1) circle[radius=2pt];  % dea finishes; paths bc
-\fill[green] (10.6, -2) circle[radius=2pt];  % deb finishes; paths ac
-\fill[green] (10.6, -3) circle[radius=2pt];  % dec finishes; paths ab
+\fill (10.6, -1) circle[radius=2pt];  % dea finishes; paths bc
+\fill (10.6, -2) circle[radius=2pt];  % deb finishes; paths ac
+\fill (10.6, -3) circle[radius=2pt];  % dec finishes; paths ab
+\node at (10.6,-1.2) {$a$};
+\node at (10.6,-2.2) {$b$};
+\node at (10.6,-3.2) {$c$};
 
-\draw (10.6, -1) -- (11.2,-0) -- (12.2,0);
+\draw[thick] (10.6, -1) -- (11.2,-0) -- (12.2,0);
 
 \node at (11.8, 0.4) {$\frac{\la{p_b}}{\la{p_b}+\la{p_c}}$};
-\draw (10.6, -1) -- (11.2,-1) -- (12.2,-1);
+\draw[thick] (10.6, -1) -- (11.2,-1) -- (12.2,-1);
 
@@
 
-\draw(10.6, -2) -- (11.2,-1.8) -- (12.2,-1.8);
-\fill[red] (12.2,-1.8) circle[radius=2pt];
-
+\draw[thick](10.6, -2) -- (11.2,-1.8) -- (12.2,-1.8);
+\fill (12.2,-1.8) circle[radius=2pt];
+\node[anchor = west] at (12.2,-1.8) {$d\succ e\succ b\succ a\succ c$};
 \node at (11.8, -1.4) {$\frac{\la{p_a}}{\la{p_a}+\la{p_c}}$};
 
-\draw (10.6, -2) -- (11.2,-2.6) -- (12.2,-2.6);
-\fill[red] (12.2,-2.6) circle[radius=2pt];
+\draw[thick] (10.6, -2) -- (11.2,-2.6) -- (12.2,-2.6);
+\fill (12.2,-2.6) circle[radius=2pt];
+\node[anchor = west] at (12.2,-2.6) {$d\succ e\succ b\succ c\succ a$};
 \node at (11.8, -2.2) {$\frac{\la{p_c}}{\la{p_a}+\la{p_c}}$};
 
-%\draw (10.6, -3) -- (11.2,-2.9);
-%\fill[red] (12.2,-2.9) circle[radius=2pt];
-%\node[red] at (11.8, -2.9) {$\frac{\la{p_a}}{\la{p_a}+\la{p_b}}$};
+%\draw[thick] (10.6, -3) -- (11.2,-2.9);
+%\fill (12.2,-2.9) circle[radius=2pt];
+%\node at (11.8, -2.9) {$\frac{\la{p_a}}{\la{p_a}+\la{p_b}}$};
 
-\draw (10.6, -3) -- (11.2,-3.4) -- (12.2, -3.4);
-\fill[red] (12.2,-3.4) circle[radius=2pt];
+\draw[thick] (10.6, -3) -- (11.2,-3.4) -- (12.2, -3.4);
+\fill (12.2,-3.4) circle[radius=2pt];
+\node[anchor = west] at (12.2,-3.4) {$d\succ e\succ c\succ b\succ a$};
+
 \node at (11.8, -3) {$\frac{\la{p_b}}{\la{p_a}+\la{p_b}}$};
 
-\fill[blue] (8.2, 1) circle[radius=2pt];  % da finishes; paths bce
+\fill (8.2, 1) circle[radius=2pt];  % da finishes; paths bce
 
-\draw (8.2,1) -- (8.9,2) -- (10.6,2);
-\draw (8.2,1) -- (8.9,1) -- (10.6,1);
-\draw (8.2,1) -- (8.9,0) -- (10.6,0);
+\draw[thick] (8.2,1) -- (8.9,2) -- (10.6,2);
+\draw[thick] (8.2,1) -- (8.9,1) -- (10.6,1);
+\draw[thick] (8.2,1) -- (8.9,0) -- (10.6,0);
 
@@
 \node at (9.5, 0.4) {$\frac{p_e     }{\la{p_b}+\la{p_c}+p_e}$};
-\fill[green] (10.6, 2) circle[radius=2pt];  % da finishes; paths bce
-\fill[green] (10.6, 1) circle[radius=2pt];  % da finishes; paths bce
-\fill[green] (10.6, 0) circle[radius=2pt];  % da finishes; paths bce
-\draw (10.6, 1) -- (11.2,2) -- (12.2,2);
+\fill (10.6, 2) circle[radius=2pt];  % da finishes; paths bce
+\fill (10.6, 1) circle[radius=2pt];  % da finishes; paths bce
+\fill (10.6, 0) circle[radius=2pt];  % da finishes; paths bce
+
+\node at (10.6, 1.8) {$b$};
+\node at (10.6, 0.8) {$c$};
+\node at (10.6,-0.2) {$e$};
+
+\draw[thick] (10.6, 1) -- (11.2,2) -- (12.2,2);
 \node at (11.8, 2.4) {$\frac{\la{p_b}}{\la{p_b}+\ld{p_e}}$};
-\draw (10.6, 1) -- (11.2,1) -- (12.2,1);
+\fill (12.2,2) circle[radius=2pt];
+\node[anchor = west] at (12.2,2) {$d\succ a\succ c\succ b\succ e$};
+\draw[thick] (10.6, 1) -- (11.2,1) -- (12.2,1);
 \node at (11.8, 1.4) {$\frac{\ld{p_e}}{\la{p_b}+p_e}$};

### 2025-02-19 inst/jss_hyper3_real.Rnw
sushi example

@@
 \end{tikzpicture}
-\caption{Partial probability\label{tikzabcde} tree for five competitors
-  $a$-$e$ with mutually supporting subsets $\left\lbrace a,b,c\right\rbrace$
-  [with support term $\lambda$] and $(de)$ [with support term $\mu$]}
+\caption{Partial probability\label{tikzabcde} tree for five
+  competitors $a$-$e$ with two mutually supporting subsets
+  $\left\lbrace a,b,c\right\rbrace$ and $\left\lbrace d,
+  e\right\rbrace$ having support $\lambda$, $\mu$ respectively}
 \end{center}
@@
 This probability model could easily be modified to account for
-specific circumstances.  The cheering effect be asymmetrical (with $a$
-helping $b$ but not vice-versa, for example).  The effect could
-operate only on specific ordered pairs.  Or perhaps the effect might
-have a finite lifetime: if $a$ places $n^\mathrm{th}$, then the
-cheering effect is active only for competitors placing
-$(n+r)^\mathrm{th}$ or above.
+specific circumstances.  The cheering effect could be asymmetrical
+(with $a$ helping $b$ but not vice-versa, for example).  The effect
+might operate only on certain ordered pairs, following a directed
+graph (cyclic or acyclic) instead of equivalence classes.  Or perhaps
+the effect might have a finite lifetime: if $a$ places
+$n^\mathrm{th}$, then the cheering effect is active only for
+competitors placing $(n+r)^\mathrm{th}$ or above, for some fixed $r$.
+There is no reason that $\lambda$ could not be less than one, which
+would indicate that some competitors inhibit, rather than help,
+others.
+
+The probability model of Figure~\ref{tikzabcde} is applied, in a
+simple one-parameter form, to two examples: Formula 1 motor racing,
+and also a preference rating.
+
+\section{Red bus, blue bus}
+
+The red bus-blue bus problem was originally stated by
+\cite{chipman1960} and discussed by \cite{davidson1970} and
+\cite[p153]{alvo2014}.  was \cite{debreu1960}.  The Holiday problem is
+discussed in \cite{tverskey1972}.  A traveller has a choice of going
+to work by different transport methods.  Here, we can assume that the
+traveller has strong but unknown views on different forms of transport
+which might incorporate a random component.  If considering a car or
+bus, the colour of the bus is of no importance to him.  However, when
+offered the choice between a red bus or a blue bus, the colour becomes
+important (having nothing else to guide his decision).  Now, he does
+not really care what colour the bus is.  If we ask him to rank his
+options, it is highly probable that he will put RB and BB
+consecutively (because they are essentially indistinguishable).  Can
+we quantify the strength of this effect?  To do this, we define a
+bespoke function \code{RB\_BB\_LF()} which returns a \code{hyper3}
+log-likelihood function corresponding to repeated observations of our
+commuter's reported ranks for the five options:
+
+<<defineredbuslf>>=
+`RB_BB_LF` <- function(lambda){
+    ec <- c(C=1, T=2, RB=3, BB=3, W=4) 
+    h <- c(1, 1, lambda, 1)           
+    (
+        cheering3(v=c("RB", "BB", "C" , "T", "W"), e=ec, h=h)*3 + 
+        cheering3(v=c("BB", "RB", "T" , "C", "W"), e=ec, h=h)*2 + 
+        cheering3(v=c("T" , "BB", "RB", "C", "W"), e=ec, h=h)*5 + 
+        cheering3(v=c("W" , "BB", "RB", "T", "C"), e=ec, h=h)*4 + 
+        cheering3(v=c("C" , "RB", "BB", "W", "T"), e=ec, h=h)*3 + 
+        cheering3(v=c("BB", "C" , "RB", "T", "W"), e=ec, h=h)*3
+    )
+}
+@
+
+Above, we see from the function body that he reported $RB\succ BB\succ
+C\succ T\succ W$ three times [first row], and so on; perhaps his
+ranking depends on the weather or how tired he is on any given day.
+Observe that in almost every case he ranks RB and BB consecutively.
+Function \code{RB_BB_LF()} takes argument \code{lambda} that
+quantifies the perceived similarity between RB and BB in the same way
+as function \code{oldfirm()} above.  It is straightforward to find a
+profile support function for $\lambda$,
+Figure~\ref{profilesupportlambdaRBBB}, and formally maximize the
+likelihood:
+
+\begin{figure}[htbp]
+\begin{center}
+<<plotproflike>>=
+o <- function(lambda){maxp(RB_BB_LF(lambda), give=1)$`Log-likelihood`} 
+lambda <- exp(seq(from=log(0.9), to=log(47), len=17)) 
+L <- sapply(lambda, o)
+plot(log(lambda), L-max(L), type="b")
+abline(h=c(0, -2))
+abline(v=0)
+@ 
+\caption{A\label{profilesupportlambdaRBBB} profile support function
+  for the red bus, blue bus parameter $\lambda$.  Vertical line
+  indicates null of $\lambda=1$, horizontal lines show maximum support
+  and $-2$}
+\end{center}
+\end{figure}
+
+<<label=maxlikesim, cache=TRUE>>=
+(osup <- optimize(o, c(10, 40), maximum=TRUE))
+@
+
+So a likelihood ratio test of the null that $\lambda=1$ would be:
+
+<<label=LRT_RB, cache=TRUE>>=
+(suppdiff <- o(osup$maximum) - o(1))
+@
+
+Easily satisfying Edwards's two-units-of-support criterion;
+alternatively, we could observe that this observation lies in the tail
+region of its asymptotic $\chi^2_1$ distribution (Wilks's theorem),
+with a $p$-value of \Sexpr{signif(pchisq(suppdiff*2, df=1,
+  lower.tail=FALSE),3)}.
+
+\subsection{Sushi}
+
+We now apply the ideas above to a dataset obtained from the {\tt
+  preflib} library.
+
+<<readsushi>>=
+a <- as.matrix(read.table("sushi.txt"))
+head(a)
+@
+
+<<defcalculate_sushi_H}>>=
+make_sushi_H <- function(a, strength){
+    h <- c(1, strength) # strength of support
+    sushi_types <- c("ebi", "anago", "maguro", "ika", "uni", "sake", "tamago", "toro", "tekka", "kappa")
+    ec <- # equivalence classes
+        c(`ebi`  = 1, `anago` = 2, `maguro` = 1, `ika` = 1, `uni` = 1,
+          `sake` = 1, `tamago` = 1, `toro` = 2, `tekka-maki` = 2,
+          `kappa-maki` = 1)
+
+    H <- hyper3()
+
+    for(i in seq_len(nrow(a))){
+        H <- H + cheering3(sushi_types[a[i,]], e=ec, h=h)
+    }
+    return(H)
+}
+@
+
+<<tryH,cache=TRUE>>=
+wanted <- sample(seq_len(nrow(a)), 50, replace=TRUE, prob=a[,1])
+H1888 <- make_sushi_H(a[wanted,-1], strength=1.888)
+H1000 <- make_sushi_H(a[wanted,-1], strength=1.000)
+@
+
+<<calclikesushi, cache=TRUE>>=
+f <- function(strength){
+   H <- make_sushi_H(a[wanted,-1], strength=strength)
+   maxp(H,give=1)$`Log-likelihood`
+}
+s <- seq(from=0.4,to=4,len=5)
+L <- sapply(s,f)
+@
+
+<<plotsushi>>=
+plot(s,L-max(L),type='b')
+@
+
+<<justf1>>=
+f1 <- f(1)
+@
+
+<<diffmaxlike>>=
+f1 - max(L)
+@
 

### 2025-02-20 inst/jss_hyper3_real.Rnw
include metadata

@@
 <<readsushi>>=
-a <- as.matrix(read.table("sushi.txt"))
+a <- as.matrix(read.table("sushi.txt",skip=22))
 head(a)
@@
 <<tryH,cache=TRUE>>=
-wanted <- sample(seq_len(nrow(a)), 50, replace=TRUE, prob=a[,1])
+wanted <- sample(seq_len(nrow(a)), 5, replace=TRUE, prob=a[,1])
 H1888 <- make_sushi_H(a[wanted,-1], strength=1.888)

### 2025-02-20 inst/jss_hyper3_real.Rnw
instructions how to process
consistentification

--- 2025-02-19 inst/jss_hyper3_real.Rnw
@@
 \usepackage{tikz}
+
+%  Process with:
+%  
+%  Rscript -e 'library("knitr") ; knit("jss_hyper3_real.Rnw")' && pdflatex jss_hyper3_real && open jss_hyper3_real.pdf
+%
 
@@
 
-<<readsushi>>=
-a <- as.matrix(read.table("sushi.txt"))
-head(a)
-@
-
 <<defcalculate_sushi_H}>>=
@@
 
+<<readsushi>>=
+a <- as.matrix(read.table("sushi.txt",skip=22))
+wanted <- sample(seq_len(nrow(a)), 5, replace=TRUE, prob=a[,1])
+a <- a[wanted, -1]
+@
+
 <<tryH,cache=TRUE>>=
-wanted <- sample(seq_len(nrow(a)), 50, replace=TRUE, prob=a[,1])
-H1888 <- make_sushi_H(a[wanted,-1], strength=1.888)
-H1000 <- make_sushi_H(a[wanted,-1], strength=1.000)
+
+H1888 <- make_sushi_H(a, strength=1.888)
+H1000 <- make_sushi_H(a, strength=1.000)
 @
@@
 f <- function(strength){
-   H <- make_sushi_H(a[wanted,-1], strength=strength)
+   H <- make_sushi_H(a, strength=strength)
    maxp(H,give=1)$`Log-likelihood`
@@
 plot(s,L-max(L),type='b')
-@
-
-<<justf1>>=
-f1 <- f(1)
-@
-
-<<diffmaxlike>>=
-f1 - max(L)
 @

### 2025-02-20 inst/jss_hyper3_real.Rnw
Makefile
consistentification: rename strength to lambda

@@
 %
+%  (also the Makefile has an entry for this)
 
@@
 <<defcalculate_sushi_H}>>=
-make_sushi_H <- function(a, strength){
-    h <- c(1, strength) # strength of support
+make_sushi_H <- function(a, lambda){
+
     sushi_types <- c("ebi", "anago", "maguro", "ika", "uni", "sake", "tamago", "toro", "tekka", "kappa")
@@
     for(i in seq_len(nrow(a))){
-        H <- H + cheering3(sushi_types[a[i,]], e=ec, h=h)
+        H <- H + cheering3(sushi_types[a[i,]], e=ec, h=c(1, lambda))
     }
@@
 <<calclikesushi, cache=TRUE>>=
-f <- function(strength){
-   H <- make_sushi_H(a, strength=strength)
+f <- function(lambda){
+   H <- make_sushi_H(a, lambda=lambda)
    maxp(H,give=1)$`Log-likelihood`

### 2025-02-20 inst/sushi.Rmd
sushi

--- 2025-02-11 inst/sushi.Rmd
@@
 
+<font size="1"> most of the time is taken calculating the
+`calclikesushi` and `maxlike`chunks.  With a sample size of 50 in
+chunk `tryH`, and 11 points on the graph, it takes about half an hour
+to run without cache. </font>
+
 To cite the `hyper2` package in publications, please use @hankin2017_rmd.
@@
 ```{r}
-a <- as.matrix(read.table("sushi.txt"))
+a <- as.matrix(read.table("sushi.txt", skip=22))
+set.seed(0)
+wanted <- sample(seq_len(nrow(a)),50,replace=TRUE,prob=a[,1])
+a <- a[wanted,-1]
+```
+
+```{r}
+sushi_types <- c("ebi", "anago", "maguro", "ika", "uni", "sake", "tamago", "toro", "tekka", "kappa")
+a[] <- sushi_types[a]
+noquote(a)
+    ec <- c("ebi"  = 1, "anago" = 2, "maguro" = 1, "ika" = 1, "uni" = 1,
+          "sake" = 1, "tamago" = 1, "toro" = 2, "tekka" = 2,
+          "kappa" = 1)
+
 ```
@@
 ```{r,defcalculate_sushi_H}
-make_sushi_H <- function(a, strength){
-    h <- c(1, strength) # strength of support
-    sushi_types <- c("ebi", "anago", "maguro", "ika", "uni", "sake", "tamago", "toro", "tekka", "kappa")
-    ec <- # equivalence classes
-        c(`ebi`  = 1, `anago` = 2, `maguro` = 1, `ika` = 1, `uni` = 1,
-          `sake` = 1, `tamago` = 1, `toro` = 2, `tekka-maki` = 2,
-          `kappa-maki` = 1)
-
+make_sushi_H <- function(a, lambda){
     H <- hyper3()
@@
     for(i in seq_len(nrow(a))){
-        H <- H + cheering3(sushi_types[a[i,]], e=ec, h=h)
+        H <- H + cheering3(a[i,], e=ec, h=c(1,lambda))
     }
@@
 ```{r tryH, cache=TRUE}
-wanted <- sample(seq_len(nrow(a)), 10, replace=TRUE, prob=a[,1])
-H1888 <- make_sushi_H(a[wanted,-1], strength=1.888)
-H1000 <- make_sushi_H(a[wanted,-1], strength=1.000)
+H1 <- make_sushi_H(a, lambda=1)
+H2 <- make_sushi_H(a, lambda=2)
 ```
@@
 ```{r maxpsushiH,cache=TRUE}
-mHa <- maxp(H1888,give=1)
-mHb <- maxp(H1000,give=1)
+mH1 <- maxp(H1,give=1)
+mH2 <- maxp(H2,give=1)
 ```
@@
 ```{r showsushiH}
-mHa
-mHb
+mH1
+mH2
 ```
 
-```{r calclike, cache=TRUE}
-f <- function(strength){
-   H <- make_sushi_H(a[wanted,-1], strength=strength)
+```{r calclikesushi, cache=TRUE}
+f <- function(lambda){
+   H <- make_sushi_H(a, lambda = lambda)
    maxp(H,give=1)$`Log-likelihood`
 }
-s <- seq(from=0.4,to=2,len=8)
+
+s <- exp(seq(from = log(0.9), to = log(6), len = 11))
 L <- sapply(s,f)
@@
 
-```{r}
-plot(s,L-max(L),type='b')
+```{r maxlike, cache=TRUE}
+(jj <- optimize(f, 2:3, maximum=TRUE))
+```
+
+```{r plotsushilike}
+plot(log(s), L-jj$objective, type='b')
+```
+
+```{r calc_sushi}
+sushi <- make_sushi_H(a, lambda = jj$maximum)
+```
+
+```{r calc_sushimaxp}
+(sushi_maxp <- maxp(sushi))
+```
+
+### Package dataset {-}
+
+Following lines create `sushi.rda`, residing in the `data/` directory of the package.
+
+```{r,label=savesushidataset}
+sushi_table <- a
+save(sushi_table,  sushi, sushi_maxp, file="sushi.rda")
 ```

### 2025-02-20 inst/sushi.Rmd
sushi consistentification

@@
 To cite the `hyper2` package in publications, please use @hankin2017_rmd.
+
+This short document applies `hyper3` formalism to a dataset
+obtained from the `preflib` library [@MaWa13a] in which ten
+different types of sushi were placed in order of preference by 5000
+judges.  Three of the sushis (maguro, toro, tekka-maki) were made with
+tuna, and it is reasonable to view these three types as being
+exchangeable in much the same way as the red and blue buses.  To
+proceed, we need to define equivalence classes of sushi---one for
+tuna-based sushis and one for everything else.  Of course, one can
+also imagine the opposite phenomenon whereby the judge decides he has
+had enough tuna for the moment and this disincentivises further tuna
+choices.  This mechanism would be modelled by having $\lambda<1$.
 

### 2025-02-20 inst/jss_hyper3_real.Rnw
more work including using data(sushi)

@@
 We now apply the ideas above to a dataset obtained from the {\tt
-  preflib} library.
+  preflib} library~\citep{MaWa13a} in which ten different types of
+sushi were placed in order of preference by 5000 judges.  Three of the
+sushis (maguro, toro, tekka-maki) were made with tuna, and it is
+reasonable to view these three types as being exchangeable in much the
+same way as the red and blue buses.  To proceed, we need to define
+equivalence classes of sushi---one for tuna-based sushis and one for
+everything else.  Of course, one can also imagine the opposite
+phenomenon whereby the judge decides he has had enough tuna for the
+moment and this disincentivises further tuna choices.  This mechanism
+would be modelled by having $\lambda<1$.
 
-<<defcalculate_sushi_H}>>=
-make_sushi_H <- function(a, lambda){
+<<defcalculate_sushi_H>>=
 
-    sushi_types <- c("ebi", "anago", "maguro", "ika", "uni", "sake", "tamago", "toro", "tekka", "kappa")
-    ec <- # equivalence classes
-        c(`ebi`  = 1, `anago` = 2, `maguro` = 1, `ika` = 1, `uni` = 1,
-          `sake` = 1, `tamago` = 1, `toro` = 2, `tekka-maki` = 2,
-          `kappa-maki` = 1)
+ec <- c("ebi"  = 1, "anago" = 2, "maguro" = 1, "ika" = 1, "uni" = 1,
+        "sake" = 1, "tamago" = 1, "toro" = 2, "tekka-maki" = 2,
+        "kappa-maki" = 1)
 
-    H <- hyper3()
-
+make_sushi_H <- function(lambda){    
     for(i in seq_len(nrow(a))){
@@
 }
-@
-
-<<readsushi>>=
-a <- as.matrix(read.table("sushi.txt",skip=22))
-wanted <- sample(seq_len(nrow(a)), 5, replace=TRUE, prob=a[,1])
-a <- a[wanted, -1]
-@
-
-<<tryH,cache=TRUE>>=
-
-H1888 <- make_sushi_H(a, strength=1.888)
-H1000 <- make_sushi_H(a, strength=1.000)
 @

### 2025-02-21 inst/jss_hyper3_real.Rnw
make sushi a package dataset

@@
 \usepackage{tikz}
+\usepackage{wrapfig}
 
@@
 \section{Introduction}
+
+\begin{wrapfigure}{r}{0.1\textwidth}
+    \centering
+    \includegraphics[width=0.1\textwidth]{hyper2.png}\\
+\end{wrapfigure}
 
@@
 
+Implementation in \code{hyper3} formalism is straightforward.  Object
+\code{sushi\_eq\_class} specifies the two equivalence classes
+
+<<show_sushi_eq_classes>>=
+sushi_eq_classes
+head(noquote(sushi_table))
+@
+
+We can now use this to define a bespoke likelihood function, using
+$\lambda$ as  the argument:
+
 <<defcalculate_sushi_H>>=
-
-ec <- c("ebi"  = 1, "anago" = 2, "maguro" = 1, "ika" = 1, "uni" = 1,
-        "sake" = 1, "tamago" = 1, "toro" = 2, "tekka-maki" = 2,
-        "kappa-maki" = 1)
-
 make_sushi_H <- function(lambda){
-    for(i in seq_len(nrow(a))){
-        H <- H + cheering3(sushi_types[a[i,]], e=ec, h=c(1, lambda))
+    H <- hyper3()
+    for(i in seq_len(nrow(sushi_table))){
+        H <- H + cheering3(sushi_table[i,], e=sushi_eq_classes, h=c(1, lambda))
     }
@@
 
+Then, as before, we may define a profile likelihood function of $\lambda$:
+
 <<calclikesushi, cache=TRUE>>=
 f <- function(lambda){
-   H <- make_sushi_H(a, lambda=lambda)
+   H <- make_sushi_H(lambda)
    maxp(H,give=1)$`Log-likelihood`
 }
-s <- seq(from=0.4,to=4,len=5)
+s <- seq(from=0.4,to=3,len=2)
 L <- sapply(s,f)

### 2025-02-21 inst/sushi.Rmd
make sushi a package dataset

@@
 noquote(a)
-    ec <- c("ebi"  = 1, "anago" = 2, "maguro" = 1, "ika" = 1, "uni" = 1,
+sushi_eq_classes <-
+    c("ebi"  = 1, "anago" = 2, "maguro" = 1, "ika" = 1, "uni" = 1,
       "sake" = 1, "tamago" = 1, "toro" = 2, "tekka" = 2,
       "kappa" = 1)
-
 ```
@@
     for(i in seq_len(nrow(a))){
-        H <- H + cheering3(a[i,], e=ec, h=c(1,lambda))
+        H <- H + cheering3(a[i,], e=sushi_eq_classes, h=c(1,lambda))
     }
@@
 sushi_table <- a
-save(sushi_table,  sushi, sushi_maxp, file="sushi.rda")
+save(sushi_table,  sushi, sushi_maxp, sushi_eq_classes, file="sushi.rda")
 ```

### 2025-02-21 inst/jss_hyper3_real.Rnw
more tweaking

@@
 
-Implementation in \code{hyper3} formalism is straightforward.  Object
-\code{sushi\_eq\_class} specifies the two equivalence classes
+Implementation in \code{hyper3} formalism is straightforward.  The
+dataset used is \code{sushi\_table}, part of the \code{hyper2} package:
+
+<<showsushitable>>=
+head(noquote(sushi_table))
+@
+
+Thus, from the top line, we see that the first judge placed tamago
+first, followed by anago and ika.  The equivalence classes are given
+by
 
@@
 sushi_eq_classes
-head(noquote(sushi_table))
 @
 
-We can now use this to define a bespoke likelihood function, using
+so class 2 is the tuna group and class 1 everything else.  We can now
+use these datapoint to define a bespoke likelihood function, using
 $\lambda$ as the argument:
@@
 }
-s <- seq(from=0.4,to=3,len=2)
+s <- exp(seq(from = log(0.9), to = log(10), len = 12))
 L <- sapply(s,f)
@@
 <<plotsushi>>=
-plot(s,L-max(L),type='b')
+plot(log(s),L-max(L),type='b',xlab=expression(log(lambda)), ylab='Log-likelihood')
 @

### 2025-02-22 inst/sushi.Rmd
name column headings of sushi_table

@@
 a <- a[wanted,-1]
+colnames(a) <- c("first","second","third","fourth","fifth","sixth","seventh","eighth","ninth","tenth")
 ```

### 2025-02-22 R/preftable.R
preftable

@@ -1,0 +1,83 @@
+`ordertable_to_preftable` <- function(M){
+  choices <- rownames(M)
+  judges <- colnames(M)
+  n <- ncol(M) # number of judges
+  m <- nrow(M) # number of choices
+  out <- matrix(NA, nrow = n, ncol = m)
+  rownames(out) <- NULL
+  colnames(out) <- paste0("c",seq_len(m))
+  
+  for(i in seq_len(n)){
+    out[i,] <- choices[order(M[,i])] # the meat
+  }
+  return(as.preftable(out))
+}
+
+`as.preftable` <- function(M){
+  if(is.preftable(M)){return(M)}
+  if(!is.matrix(M)){stop("argument M must be a matrix")}
+  n <- nrow(M) # number of judges
+  choices <- sort(M[1,])
+  for(i in seq_len(n)[-1]){
+    stopifnot(all(sort(M[i,]) == choices))
+  }
+  
+  if(is.null(rownames(M))){
+    rownames(M) <- paste0("judge", seq_len(n))
+  }
+  
+  class(M) <- "preftable"
+  return(M)
+}
+
+is.preftable <- function(M){inherits(M, "preftable")}
+  
+`print.preftable` <- function(x, ...){
+  cat("A preftable object:\n")
+  print(noquote(unclass(x)))
+  return(invisible(x))
+}
+
+`preftable_to_ranktable` <- function(x) {
+  stopifnot(is.preftable(x))
+  n <- nrow(x) # number of judges
+  m <- ncol(x) # number of choices
+ 
+  choices <- sort(x[1,])
+  out <- matrix(NA, nrow = n, ncol = m)
+  rownames(out) <- rownames(x)
+  colnames(out) <- choices
+  
+  for(i in seq_len(n)){
+    out[i,] <- rank(x[i,]) # the meat
+  }
+  class(out) <- "ranktable"
+  return(out)
+}
+
+`preftable_to_supp` <- function(x){
+  H <- hyper2()
+  for(i in seq_len(nrow(x))){
+    H <- H + rankvec_likelihood(x[i,])
+  }
+  return(H)
+}
+
+if(FALSE){
+a1 <- preftable_to_supp(as.preftable(sushi_table[1:5,]))
+a2 <- ordertable2supp(ranktable_to_ordertable(preftable_to_ranktable(as.preftable(sushi_table[1:5,]))))
[…]

### 2025-02-23 inst/ordertable_to_ranktable.Rmd
whitespace typo
S3 classes ordertable and ranktable

@@
 
-# Toy dataset
+### Toy dataset {-}
 
@@
 
-Note that it is possible to view object `RT` directly by removing
-class `ranktable`:
+The default print method used for `RT` as used above obscures the
+internal structure of the object.  Note that it is possible to view
+object `RT` directly by removing class `ranktable` with `unclass()`:
 
 ```{r viewdirect}
-class(RT)
-class(RT) <- "matrix"  # removes "ranktable" from class list
-RT
+unclass(RT)
 ```
 
-Note carefully that the second row of `RT` (which corresponds to the
-second judge) does *not* match the second column of `OT`.  The way to
-read the second row of `RT` is as follows.  We see the entries `1 3 4
-5 2` indicating that the first choice was competitor 1 [viz `a`], the
-second choice was competitor 3 [viz `c`], the third choice was
-competitor 4 [viz `d`], fourth choice was competitor 5 [`e`] and fifth
-choice was competitor 2 [`b`].  So the rows are *ranks* (see the
-discussion in `rrank.Rd`).  But it is very tempting to read the second
-row of `RT` as "judge 2 placed competitor `b` in third place", which
-is wrong because this is interpreting the row as an *order*.  This is
-why the print method is there.
+(it is easier to start by considering the second row; the first row is
+a bit confusing).  Note carefully that the second row of `RT` (which
+corresponds to the second judge) does *not* match the second column of
+`OT`.  The way to read the second row of `RT` is as follows.  We see
+the entries `1 3 4 5 2` indicating that the first choice was
+competitor 1 [viz `a`], the second choice was competitor 3 [viz `c`],
+the third choice was competitor 4 [viz `d`], fourth choice was
+competitor 5 [`e`] and fifth choice was competitor 2 [`b`].  So the
+rows are *ranks* (see the discussion in `rrank.Rd`).  But it is very
+tempting to read the second row of `RT` as "judge 2 placed competitor
+`b` in third place", which is wrong because this is interpreting the
+row as an *order*.  This is why the print method is there.
 
-# Volvo ocean racing dataset
+### Volvo ocean racing dataset {-}
 

### 2025-02-23 inst/skating.Rmd
rewording
consistentify "time to process" information
S3 classes ordertable and ranktable

@@
 
-To cite the `hyper2` package in publications, please use @hankin2017_rmd.
-This document analyses Bradley-Terry strengths of the competitors in
-the Ladies' Figure Skating at the 2002 Winter Olympics.  In the
-package, dataframe `skating_table` is copied from Lock and Lock (see
-skating.Rd for more details).  It also creates file `data/skating.rda`
-which contains R objects `skating`, `skating_table`, and
-`skating_maxp`.
+<font size="1"> (takes about twenty minutes to run without cache)
+</font>
+
+To cite the `hyper2` package in publications, please use
+@hankin2017_rmd).  This document analyses Bradley-Terry strengths of
+the competitors in the Ladies' Figure Skating at the 2002 Winter
+Olympics.  In the package, dataframe `skating_table` is copied from
+Lock and Lock (see skating.Rd for more details).  It also creates file
+`data/skating.rda` which contains R objects `skating`,
+`skating_table`, and `skating_maxp`.
 
@@
 set.seed(0)
-skating_table <- read.table("skating.txt",header=TRUE)
-skating_table
+skating_table <- ordertable(read.table("skating.txt", header=TRUE))
 ```
@@
 ```{r,convertordertosupport}
-head(ordertable2supp(skating_table))
+summary(suppfun(ordertable(skating_table)))
 ```
@@
 ```{r,unconstrainedlikes, cache=TRUE}
-mgv <- maxp(skating, give=TRUE, n=10)$value
+options(use_alabama = TRUE)
+mgv <- maxp(skating, give=TRUE, n=1)$value
 mgv
@@
 ```{r,label=saveskatingdataset}
-save(skating_table,skating,skating_maxp,file="skating.rda")
+save(skating_table, skating, skating_maxp, file="skating.rda")
 ```

### 2025-02-23 inst/soling.Rmd
S3 classes ordertable and ranktable

--- 2023-10-30 inst/soling.Rmd
@@
 
-To cite the `hyper2` package in publications, please use @hankin2017a.
+To cite the `hyper2` package in publications, please use @hankin2017_rmd.
 Soling is a type of boat raced in the 2000 Summer Olympics.  Here I
@@
 ```{r loadlib}
-soling_table <- read.table("soling.txt")
+soling_table <- ordertable(read.table("soling.txt"))
 soling_table
@@
 ```{r interpretsoling}
-ordertable_to_ranktable(soling_table)
+as.ranktable(soling_table)
 ```
@@
 ```{r solingtolikelihood}
-soling <- ordertable2supp(soling_table)
+soling <- suppfun(soling_table)
 head(soling)

### 2025-02-23 inst/sushi.Rmd
preftable
S3 classes ordertable and ranktable

@@
 
-s <- exp(seq(from = log(0.9), to = log(6), len = 11))
+s <- exp(seq(from = log(0.9), to = log(6), len = 2))
 L <- sapply(s,f)
@@
 ```{r,label=savesushidataset}
-sushi_table <- a
+sushi_table <- as.preftable(a)
 save(sushi_table,  sushi, sushi_maxp, sushi_eq_classes, file="sushi.rda")

### 2025-02-23 R/hyper2.R

@@ -757,6 +757,10 @@ rorder_single <- function(p){
   return(out)
 }
 
+setGeneric("as.ranktable", function(x){standardGeneric("as.ranktable")})
+`as.ranktable` <- function(x){UseMethod("as.ranktable")}
+`as.ranktable.ordertable` <- function(x){ordertable_to_ranktable(x)}
+
 `.allorders` <- function(x){
   out <- perms(length(x))
   matrix(x[out],nrow=length(x))
@@ -1141,6 +1145,18 @@ rorder_single <- function(p){
     return(ranktable_to_ordertable(o))
 }
 
+`ordertable` <- function(x){
+    ## This is the only place class ordertable is set:
+    class(x) <- c("ordertable", setdiff(class(x), "ordertable"))
+    return(x)
+}
+
+`print.ordertable` <- function(x, ...){
+    cat("An ordertable:\n")
+    class(x) <- setdiff(class(x), "ordertable")
+    print(x)
+}
+
 `as.ordertable` <- function(w){
     out <-
         apply(w,2,    
@@ -1154,7 +1170,7 @@ rorder_single <- function(p){
                   return(out)
               } )
     rownames(out) <- rownames(w)
-    return(out)
+    return(ordertable(out))
 }
 
 `ordertable2points` <- function(o,points,totals=TRUE){

### 2025-02-26 inst/jss_hyper3_real.Rnw
include a figure

@@
 
-Then, as before, we may define a profile likelihood function of $\lambda$:
+Then, as before, we may define a profile likelihood function of
+$\lambda$, Figure~\ref{profilesupportsushi}.
 
+\begin{figure}[htbp]
+  \begin{center}
 <<calclikesushi, cache=TRUE>>=
@@
 L <- sapply(s,f)
+plot(log(s),L-max(L),type='b',
+     xlab=expression(log(lambda)), ylab='Log-likelihood')
 @ 
-
-<<plotsushi>>=
-plot(log(s),L-max(L),type='b',xlab=expression(log(lambda)), ylab='Log-likelihood')
-@
+\caption{A\label{profilesupportsushi} profile support function for the
+  exchangeability term $\lambda$.  Vertical gray line shows
+  $H_0\colon\lambda =1$, horizontal lines show the
+  two-units-of-support limits}
+\end{center}
+\end{figure}
 

### 2025-02-26 inst/counterstrike.Rmd
consistentification
distinguish hankin2017 from hankin2017_rnw
All games played by `Cloud9`.

@@
 
-To cite the `hyper2` package in publications, please use @hankin2017a.
+To cite the `hyper2` package in publications, please use @hankin2017_rmd.
 In this short vignette I use the `hyper2` package to analyse an
@@
 
+## A red bus blue bus analysis.
+
+I now analyse every game played by `Cloud9` in the tournament.
+
+https://www.hltv.org/stats/teams/matches/5752/cloud9?event=3247
+
+As far as I can tell this is every match played at ELEAGUE Major 2018  in Boston by `Cloud9`.
+
+```{r}
+a <- read.table("cloud9_boston_2018.txt",header=TRUE)
+a
+```
+
+Above we see from the first line that on 28 Jan 2018, playing `FaZe` on the Inferno map we have a win, with deathorder Skadoodle, RUSH, etc.
+
+We will look at the effect of `Skadoodle` on `Stewie2K`.
+
+```{r}
+ec <-c(autimatic=1, RUSH=1, Skadoodle=2, Stewie2K=2, tarik=1)
+mk <- function(lam){
+  H3 <- hyper3()
+  for(i in seq_len(nrow(a))){
+
+    d <- strsplit(a$deathorder[i],", ")[[1]]
+    H3 <- H3 + cheering3(d, e=ec, help=c(1,lam))
+  }
+  return(H3)
+}
+
+```{r mkmaxp, cache=TRUE}
+maxp(mk(0.5),give=1)
+maxp(mk(1.01),give=1)
+```
+
 ### References {-}

### 2025-02-27 inst/curling.Rmd
S3 classes ordertable and ranktable
start work on issue #244

@@
 ```{r readcurlingtable}
-curling_table  <- read.table("curling.txt",header=TRUE)
+curling_table  <- ordertable(read.table("curling.txt",header=TRUE))
 curling_table
@@
 ```{r,definsupportb,cache=TRUE}
-curling1 <- ordertable2supp(ct)
-head(curling1)
+curling1 <- suppfun(ordertable(ct))
+summary(curling1)
 curling1_maxp <- maxp(curling1)
@@
 }
-head(curling2)
+summary(curling2)
 (curling2_maxp <- maxp(curling2))

### 2025-02-27 inst/skating.Rmd
start work on issue #244

@@
 ```{r,maxshow, cache = TRUE}
-skating <- ordertable2supp(skating_table)
+skating <- suppfun(skating_table)
 options("use_alabama" = TRUE)
@@
 ```{r makelikef}
-o <- ordertable2supp(o)
+o <- suppfun(o)
 o

### 2025-02-27 inst/skeleton.Rmd
start work on issue #244

@@
 ```{r readskeletontable}
-(skeleton_table <- read.table("skeleton2022.txt",header=TRUE))
+(skeleton_table <- ordertable(read.table("skeleton2022.txt",header=TRUE)))
 ```
@@
 ```{r,label=processskeleton,cache=TRUE}
-skeleton <- ordertable2supp(skeleton_table)
+skeleton <- suppfun(skeleton_table)
 skeleton_maxp <- maxp(skeleton)

### 2025-02-27 R/suppfun.R

@@ -1,6 +1,6 @@
 setGeneric("suppfun",function(x){standardGeneric("suppfun")})
 
-`suppfun` <- function(M){UseMethod("suppfun")}
+`suppfun` <- function(x){UseMethod("suppfun")}
 `suppfun.ordertable` <- function(x){ordertable2supp(x)}
 `suppfun.ranktable` <- function(x){"not yet implemented"}
 `suppfun.preftable` <- function(x){"not yet implemented"}

### 2025-02-27 inst/cheering.Rmd

@@
 In the following, I use the ''race'' metaphor: five runners take part
-in a race and arrive in order: $a\succ b\succ c\succ d\succ 5e.
+in a race and arrive in order: $a\succ b\succ c\succ d\succ e$.
 Consider the following Plackett-Luce likelihood function for this
@@
     (
-        cheering3(v=c("RB","BB","C" ,"T","W"),e=ec,h=h)*3 +
-        cheering3(v=c("BB","RB","T" ,"C","W"),e=ec,h=h)*2 +
-        cheering3(v=c("T" ,"BB","RB","C","W"),e=ec,h=h)*2 +
-        cheering3(v=c("W" ,"BB","RB","T","C"),e=ec,h=h)*4 +
-        cheering3(v=c("C" ,"RB","BB","W","T"),e=ec,h=h)*4 +
-        cheering3(v=c("BB","C" ,"RB","T","W"),e=ec,h=h)*3
+        cheering3(v=c("RB","BB","C" ,"T","W"), e=ec, h=h)*3 +
+        cheering3(v=c("W" ,"BB","RB","C","T"), e=ec, h=h)*7 +
+        cheering3(v=c("W" ,"RB","BB","C","T"), e=ec, h=h)*8 +
+        cheering3(v=c("W" ,"BB","RB","T","C"), e=ec, h=h)*7 +
+        cheering3(v=c("W" ,"RB","BB","C","T"), e=ec, h=h)*7 +
+        cheering3(v=c("RB","BB","C" ,"T","W"), e=ec, h=h)*3 +
+        cheering3(v=c("RB","BB","C" ,"T","W"), e=ec, h=h)*3 +
+        cheering3(v=c("BB","RB","T" ,"C","W"), e=ec, h=h)*2 +
+        cheering3(v=c("T" ,"BB","RB","C","W"), e=ec, h=h)*2 +
+        cheering3(v=c("W" ,"BB","RB","T","C"), e=ec, h=h)*4 +
+        cheering3(v=c("C" ,"RB","BB","W","T"), e=ec, h=h)*4 +
+        cheering3(v=c("BB","C" ,"RB","T","W"), e=ec, h=h)*3
     )
@@
 
-```{r,label=proflikeredbus,cache=TRUE}
+```{r,label=proflikeredbusbluebus,cache=TRUE}
 o <- function(s){maxp(RB_BB_LF(s),give=TRUE,n=1)$likes} # optimand
-s <- exp(seq(from=log(1.3),to=log(47),len=17)) # putative similarity measures
+s <- exp(seq(from=log(1.3),to=log(147),len=17)) # putative similarity measures
 L <- sapply(s,o)
@@
 a
-H <- ordertable2supp(a)
+H <- suppfun(ordertable(a))
 ```

### 2025-02-27 inst/counterstrike.Rmd

@@
 }
+```
+
+  Now some profile likelihood
 
@@
 
+```{r calcklikef, cache=TRUE}
+f <- function(l){maxp(mk(l), give=1)$`Log-likelihood`}
+lam <- seq(from = 0.01,to = 1.1, len = 43)
+L <- sapply(lam, f)
+```
+
+```{r plotlikelam}
+plot(lam, L - max(L), type='b')
+plot(log(lam), L - max(L), type='b')
+```
+
+```{r findaopt, cache=TRUE}
+optlam <- optimize(f,c(0.09, 0.14), maximum = TRUE)
+```
+
+```{r showoptlam}
+optlam
+```
+
+```{r}
+maxp(mk(optlam$maximum))
+```
+
 ### References {-}

### 2025-02-27 inst/pentathlon.Rmd
S3 classes ordertable and ranktable

@@
 ```{r readpentathlontable}
-pentathlon_table <- as.matrix(read.table("pentathlon.txt"))
+pentathlon_table <- ordertable(as.matrix(read.table("pentathlon.txt")))
 pentathlon_table
@@
 ```{r showpiechart,cache=TRUE}
-pentathlon <- ordertable2supp(pentathlon_table)
+pentathlon <- suppfun(pentathlon_table)
 pentathlon_maxp <- maxp(pentathlon)

### 2025-02-27 inst/salad.Rmd

@@
    paste("j",formatC(seq_len(nrow(salad)),width=2,format="d",flag="0"),sep="")
-ts <- t(salad)
-ts
+ts <- ordertable(t(salad))
 ```
@@
 ```{r useslickmethod,cache=TRUE}
-(H2 <- ordertable2supp(ts))
+(H2 <- suppfun(ts))
 ```

### 2025-02-27 inst/volvo.Rmd

@@
 ```{r readvolvotable}
-volvo_table <- read.table("volvo.txt")
+volvo_table <- ordertable(read.table("volvo.txt"))
 volvo_table
@@
 ```{r specifynoscore,cache=TRUE}
-volvo <- ordertable2supp(volvo_table,noscore=c("DNF","DNS"))
+volvo <- suppfun(volvo_table, noscore=c("DNF", "DNS"))
 (volvo_maxp <- maxp(volvo))
@@
 ```{r,label=savecurlingdataset}
-save(volvo_table,volvo_maxp,volvo,file="volvo.rda")
+save(volvo_table, volvo_maxp, volvo, file = "volvo.rda")
 ```

### 2025-02-28 inst/eurovision.Rmd
further work on issue #244

@@
     d <- preference[,i,drop=TRUE]
-	eurovision <- eurovision + ordervec2supp(d[!is.na(d)])
+	eurovision <- eurovision + suppfun(d[!is.na(d)])
 } # i loop closes

### 2025-02-28 inst/podium.Rmd
further work on issue #244

@@
      o[o > 0] <- rank(o[o > 0])
-     out <- as.hyper3(ordervec2supp(o))
+     out <- as.hyper3(suppfun(o))
      for (i in seq(from=2,to=nrow(fmat))){

### 2025-02-28 R/hyper2.R

@@ -637,22 +637,7 @@ setGeneric("pnames<-",function(x,value){standardGeneric("pnames<-")})
 
 `head.hyper2` <- function(x,...){ x[head(elements(brackets(x)),...)] }
 
-`rank_likelihood` <- function(M,times=1){
-  M <- rbind(M)  # deals with vectors
-  times <- cbind(seq_len(nrow(M)),times)[,2]
-  if(is.null(colnames(M))){
-    cn <- NA
-  } else {
-    cn <- colnames(M)
-  }
-  out <- hyper2(pnames=cn)
-  for(i in seq_len(nrow(M))){
-      out <- out + ordervec2supp(rev(M[i,,drop=TRUE]))
-  } 
-  return(out)
-}
-
-`rankvec_likelihood` <- function(v,nonfinishers=NULL){
+`rankvec_likelihood` <- function(v, nonfinishers = NULL){
   stopifnot(all(table(v)==1))
   out <- hyper2()
   v <- rev(v)   # first-placed competitor is the first element of v
@@ -702,11 +687,10 @@ rorder_single <- function(p){
 `rrank` <- function(n=1, p, pnames=NULL, fill=FALSE, rnames=NULL){
     if(fill){ p <- fillup(p) }
     if(is.null(pnames)){pnames <- names(p)}
+    if(is.null(pnames)){pnames <- letters[seq_along(p)]}
     out <- t(replicate(n, rrank_single(p)))
-    colnames(out) <- pnames
-    rownames(out) <- rnames
-    class(out) <- "ranktable"
-    return(drop(out))
+    out[] <- pnames[out]
+    return(as.ranktable(out))
 }
 
 `.allorders` <- function(x){
@@ -1106,6 +1090,7 @@ rorder_single <- function(p){
 }
 
 `as.ordertable` <- function(w){
+    if(is.ranktable(w)){return(ranktable_to_ordertable(w))}
     out <-
         apply(w,2,    
               function(x){ # 'x' is a column of w

### 2025-02-28 R/ranktable.R

@@ -1,25 +1,5 @@
-`ranktable_to_printable_object` <- function(x){
-  x <- rbind(x)
-  if(is.null(colnames(x))){
-    cn <- seq_len(ncol(x))
-  } else {
-    cn <- colnames(x)
-  }
-  thing <- matrix(cn[x],ncol=ncol(x))
-  colnames(thing) <- paste("c",seq_len(ncol(x)),sep="")
-  rownames(thing) <- rownames(x)
-  class(thing) <- "noquote"
-  return(thing)
-}
-
-`print.ranktable` <- function(x,...){
-    cat("A ranktable:\n")
-    print(ranktable_to_printable_object(x))
-}
-
 `print.ranktablesummary` <- function(x,...){
-    x <- ranktable_to_printable_object(x)
-    class(x) <- "matrix"
+    x <- unclass(x)
     x <- cbind(x[,1:3],"...",x[,ncol(x)+c(-1,0)])
     x <- noquote(x)
     print(x)
@@ -33,20 +13,52 @@
 }
 
 `ranktable_to_ordertable` <- function(xrank){
-  out <- apply(xrank,1,order)
-  rownames(out) <- colnames(xrank)
-  return(out)
+    stop("not yet implemented.\nIt is harder than it looks.\nI'll get round to it one day.\nProbably.")
 }
 
 `ordertable_to_ranktable` <- function(xorder){
-  stopifnot(all(apply(xorder,2,function(x){all(sort(x) == seq_along(x))})))
-  out <- t(apply(xorder,2,function(x){seq_len(nrow(xorder))[order(x)]}))
-  colnames(out) <- rownames(xorder)
-  class(out) <- "ranktable"
-  return(out)
+  stopifnot(all(apply(xorder, 2, function(x){all(sort(x) == seq_along(x))})))
+  o <- t(apply(xorder, 2, function(x){seq_len(nrow(xorder))[order(x)]}))
+  colnames(o) <- rownames(xorder)
+  o <- rbind(o)
+  if(is.null(colnames(o))){
+    cn <- seq_len(ncol(o))
+  } else {
+    cn <- colnames(o)
+  }
+  out <- matrix(cn[o], ncol = ncol(o))
+  colnames(out) <- paste("c", seq_len(ncol(o)), sep = "")
+  rownames(out) <- rownames(o)
+  return(as.ranktable(out))
 }
 
 setGeneric("as.ranktable", function(x){standardGeneric("as.ranktable")})
+
[…]

### 2025-02-28 R/suppfun.R

@@ -2,5 +2,5 @@ setGeneric("suppfun",function(x){standardGeneric("suppfun")})
 
 `suppfun` <- function(x, ...){UseMethod("suppfun")}
 `suppfun.ordertable` <- function(x, ...){ordertable2supp(x, ...)}
-`suppfun.ranktable` <- function(x, ...){"not yet implemented"}
-`suppfun.preftable` <- function(x, ...){"not yet implemented"}
+`suppfun.character` <- function(x, nonfinishers=NULL, ...){rankvec_likelihood(x, nonfinishers=nonfinishers, ...)}
+`suppfun.numeric` <- function(x, ...){ordervec2supp(x)}

### 2025-02-28 inst/curling.Rmd

@@
     names(x) <- rownames(curling_table)
-    curling2 <- curling2 + ordervec2supp(x[!is.na(x)])
+    curling2 <- curling2 + suppfun(x[!is.na(x)])
 }

### 2025-02-28 inst/eurodance.Rmd
consistentify "time to process" information

@@
 
-To cite the `hyper2` package in publications, please use @hankin2017_rmd;
-takes about 20 minutes to process without cache.  The file
-`eurodance2008.txt`, used below, is copied from "Eurovision Dance
-Contest 2008", Wikipedia, accessed March 31, 2021.  First we specify
-the matrix as appearing in the Wikipedia page:
+<font size="1"> (takes about twenty minutes to run without cache)
+</font>
+
+To cite the `hyper2` package in publications, please use
+@hankin2017_rmd.  The file `eurodance2008.txt`, used below, is copied
+from "Eurovision Dance Contest 2008", Wikipedia, accessed March 31,
+2021.  First we specify the matrix as appearing in the Wikipedia page:
 
@@
 	x <- pref[,i]
-	eurodance %<>% `+`(ordervec2supp(x[!is.na(x)]))
+	eurodance %<>% `+`(suppfun(x[!is.na(x)]))
 }  # i loop closes
@@
 	names(jjstar) <- names(jury)
-	LL[i] <- loglik(indep(eurodance_maxp),ordervec2supp(jjstar),log=FALSE)
+	LL[i] <- loglik(indep(eurodance_maxp),suppfun(jjstar),log=FALSE)
 }
@@
 plot(sort(LL),log="y")
-p <- loglik(indep(eurodance_maxp),ordervec2supp(jury),log=FALSE)
+p <- loglik(indep(eurodance_maxp),suppfun(jury),log=FALSE)
 abline(h=p)

### 2025-02-28 inst/formula1.Rmd
consistentify "time to process" information

@@
 title: "formula1"
-output: html_document
+author: "R. K. S. Hankin"
+output: bookdown::html_document2
+bibliography: hyper2.bib  # copy from vignettes/
 ---
@@
 
-**This Rmd file takes about two hours to run without cache.**
+<font size="1"> (takes about two hours to run without cache)
+</font>
 
-Formula 1 racing is an important and prestigious motor sport.  Here I
-analyse a series of racing results using the `hyper2` package which
-implements the Plackett-Luce likelihood function.
+To cite the `hyper2` package in publications, please use
+@hankin2017_rmd.  Formula 1 racing is an important and prestigious
+motor sport.  Here I analyse a series of racing results using the
+`hyper2` package which implements the Plackett-Luce likelihood
+function.
 
@@
 	}
-	L2017 <- L2017 + ordervec2supp(o)
+	L2017 <- L2017 + suppfun(o)
 }  # for(i) loop closes
@@
 	cond <- details$conditions[i]
-        jj <- ordervec2supp(y)
+        jj <- suppfun(y)
 	if(cond == "dry"){
@@
 	names(y) <- rownames(f2017)
-	HP <- HP + pwa(ordervec2supp(y),details$pole[i],"pole")
+	HP <- HP + pwa(suppfun(y),details$pole[i],"pole")
 } # i loop closes

### 2025-02-28 inst/formula1_points.Rmd

@@
  for(i in seq_len(ncol(o))){
-	H <- H + ordervec2supp(o[,i])
+	H <- H + suppfun(o[,i])
 }

### 2025-02-28 inst/ordertable_to_ranktable.Rmd

--- 2024-01-26 inst/ordertable_to_ranktable.Rmd
@@
 ---
-title: "The print method for class ranktable"
+title: "Objects of class `ranktable`"
 author: "R. K. S. Hankin"
@@
 
-To cite the `hyper2` package in publications, please use @hankin2017_rmd.
-Rank tables can be confusing.  The package includes a range of
-functionality for dealing with them, including a bespoke print method.
-Here, I discuss two cases: firstly, a tiny toy example, and secondly
-the volvo ocean racing dataset.
+To cite the `hyper2` package in publications, please use
+@hankin2017_rmd.  Rank tables can be confusing.  The package includes
+a range of functionality for dealing with them.  Here, I discuss two
+cases: firstly, a tiny toy example, and secondly the volvo ocean
+racing dataset.
 
-# Toy dataset
+### Toy dataset {-}
 
@@
 colnames(OT) <- c("J1","J2")
+OT <- ordertable(OT)
 OT
@@
 
-Note that it is possible to view object `RT` directly by removing
-class `ranktable`:
-
-```{r viewdirect}
-class(RT)
-class(RT) <- "matrix"  # removes "ranktable" from class list
-RT
-```
-
-Note carefully that the second row of `RT` (which corresponds to the
-second judge) does *not* match the second column of `OT`.  The way to
-read the second row of `RT` is as follows.  We see the entries `1 3 4
-5 2` indicating that the first choice was competitor 1 [viz `a`], the
-second choice was competitor 3 [viz `c`], the third choice was
-competitor 4 [viz `d`], fourth choice was competitor 5 [`e`] and fifth
-choice was competitor 2 [`b`].  So the rows are *ranks* (see the
-discussion in `rrank.Rd`).  But it is very tempting to read the second
-row of `RT` as "judge 2 placed competitor `b` in third place", which
-is wrong because this is interpreting the row as an *order*.  This is
-why the print method is there.
-
-# Volvo ocean racing dataset
+### Volvo ocean racing dataset {-}
 
@@
 `[.ranktable()` (right now I am not 100% sure that this is a good
-idea).  Note that simply restoring the `ranktable` class can be
-misleading:
-
-```{r nowmakerank3}
-xx <- x[,1:3]
-xx
-class(xx) <- "ranktable"
-xx
-```
-
-In the above, `NA`s are introduced because a `ranktable` object uses
-the column names to store names of the competitors, and of course
-extracting the first three columns means keeping only the first three
-column names.  Observe that extraction after passing to
-`ranktable_to_printable_object()` works nicely:
-
-```{r showprintableobject}
-ranktable_to_printable_object(x)[,1:3]
-```
+idea).
 

### 2025-03-02 inst/powerboat.Rmd
update to new-style ordertable class

@@
 powerboat_table[20:21,8] <- 0  # above does not apply to points
+powerboat_table <- ordertable(powerboat_table)
 powerboat_table

### 2025-03-02 R/hyper2.R

@@ -1083,15 +1083,16 @@ rorder_single <- function(p){
 }
 
 `ordertable` <- function(x){
+    x <- as.matrix(x)
+    if(is.null(rownames(x))){stop("rownames must be supplied")}
     ## This is the only place class ordertable is set:
-    class(x) <- c("ordertable", setdiff(class(x), "ordertable"))
+    class(x) <- c("ordertable", "matrix")
     return(x)
 }
 
 `print.ordertable` <- function(x, ...){
     cat("An ordertable:\n")
-    class(x) <- setdiff(class(x), "ordertable")
-    print(x)
+    print(noquote(unclass(x)))
 }
 
 `as.ordertable` <- function(w){

### 2025-03-02 inst/skating.Rmd
extraction

@@
 Object `skating_table` is an order table.  It is structured so that
-each competitor is a row, and each judge is a column.  Function
-`rank_likelihood()` considers each row to be a race [or a judge], so
-we need to take a transpose.  The following R idiom is plausible but
-**incorrect**:
-
-```{r,skatingwrong}
-skating_incorrect <- rank_likelihood(t(skating_table)) # incorrect
-```
-
-The above is incorrect because we are applying a rank technique to an
-order table.  Correct analysis follows.  One good thing to do is to
-present the dataset as an order table;
+each competitor is a row, and each judge is a column.  One good thing
+to do is to present the dataset as an order table;
 
 ```{r,coercetoranktable}
-ordertable_to_ranktable(skating_table)
+as.ranktable(skating_table)
 ```
@@
 
+Now we consider only the top seven competitors (that is, the top seven
+according to Judge 1).
+
 ```{r justtop7}
-skating_table
-o <- ordertable_to_ranktable(skating_table)
-o
+(o <- skating_table[1:7])
 ```
 
-```{r setclasstomatrix}
-class(o) <- "matrix"
-o
-```
-
-Select just the top seven:
-```{r selectthetop7}
-o <- o[,1:7]
-o
-```
-
-and now replace ranks with internal ranks (that is, rank among the top
-seven):
-
-```{r replaceranks}
-o <- t(apply(o,1,rank))
-colnames(o) <- rownames(skating_table)[1:7]
-o
-```
-
-Now coece to a `ranktable`:
-
-```{r coercetoranktble}
-class(o) <- "ranktable"
-o
-```
-
-Thence to an order table:
-
 ```{r thenceto}
-o <- ranktable_to_ordertable(o)
-o
+(o <- as.ranktable(o))
 ```

### 2025-03-02 R/hyper2.R

@@ -1285,3 +1285,4 @@ rorder_single <- function(p){
     }
 }
 
+`[.ordertable` <- function(x, ...){ ordertable(apply(unclass(x)[...,], 2, rank)) }

### 2025-03-02 inst/skating.Rmd
tweaks

@@
 skating_table <- ordertable(read.table("skating.txt", header=TRUE))
+skating_table
 ```
@@
 ```{r,convertordertosupport}
-summary(suppfun(ordertable(skating_table)))
+summary(suppfun(skating_table))
 ```
@@
 ```{r,threemedallists}
-head(skating_table,2)
+skating_table[1:3]
 ```
@@
 ```{r, tableofresults}
-hughes    <- unlist(skating_table[1,])
-slutskaya <- unlist(skating_table[2,])
+hughes    <- unclass(skating_table)[1,]
+slutskaya <- unclass(skating_table)[2,]
 table(hughes)

### 2025-03-03 inst/cook_repeated.Rmd
consistentify "time to process" information
bugfix

@@
 
-To cite the `hyper2` package in publications, please use @hankin2017_rmd;
-takes about five minutes without cache.  Suppose there are five people
-in a family.  Each day for five days, one person cooks a meal.  After
-the fifth day, each person considers the four meals cooked by other
-family members, and puts them in order of preference.  This process is
-repeated and here I consider how to interpret such data.
+<font size="1"> (takes about five minutes to run without cache)
+</font>
+
+To cite the `hyper2` package in publications, please use
+@hankin2017_rmd.  Suppose there are five people in a family.  Each day
+for five days, one person cooks a meal.  After the fifth day, each
+person considers the four meals cooked by other family members, and
+puts them in order of preference.  This process is repeated and here I
+consider how to interpret such data.
 
@@
 (day1 <- cooking_table[1,])
-H_row1 <- rank_likelihood(day1[day1>0])
+(H_row1 <- suppfun(names(sort(day1[day1 > 0]))))
 ```
@@
 jj <- cooking_table[i,] 
-    H <- H + rank_likelihood(jj[jj>0])
+    H <- H + suppfun(sort(jj[jj > 0]))
 }
@@
 and boys systematically vote boys higher, than warranted.  To test
-this hypothesis, we posit two reified entities, `girl` and `boy`.
+this hypothesis, we either posit two reified entities, `girl` and `boy`.
 Entity `girl` helps a girl when another girl is voting for her, and
@@
 
-```{r girlsvotinggirls,cache=TRUE}
-cooking_table <-
-	as.matrix(read.table("cooking_weekly.txt", header=TRUE))
-cooking_table <- cooking_table[rowSums(cooking_table==0)==1,]
-boys <-c("robin","zac")
-girls <- c("lesley","alice","annabel")
-cooks <- colnames(cooking_table)
-P <- hyper2()
-P0 <- hyper2()
-for(i in seq_len(nrow(cooking_table))){  # iterate over rows
-    x <- cooking_table[i,] 
-	if(sum(x==0)>1){break}
-    names(x) <- cooks
-	if(names(x[x==0]) %in% boys){  # if the voter is a boy
-		x <- x[x>0]  # remove the voter
-		Pjj <- pwa(rank_likelihood(x),names(x)[names(x) %in% boys], chameleon="boy")
-	} else if (names(x[x==0]) %in% girls){ # voter a girl
-		x <- x[x>0]
-		Pjj <- pwa(rank_likelihood(x),names(x)[names(x) %in% girls], chameleon="girl")
-	} else {
-		stop()
-	}
-    P <- P + Pjj
-}
-mP <- maxp(P)
-```
-
-```{r plotapie}
-mP
-pie(mP)
-```
-
-Now test whether the girls preferentially vote:
-
-```{r girlspref, cache=TRUE}
-specificp.gt.test(P,"girl",0)
-```
+Or we use multipliers and `hyper3` idiom (not yet done)
 

### 2025-03-03 inst/curling.Rmd
typo

@@
 for(i in seq_len(ncol(curling_table))){
-    x <- curling_table[,i]
+    x <- unclass(curling_table)[,i]
     names(x) <- rownames(curling_table)

### 2025-03-03 inst/formula1.Rmd
remove old-style flawed code

@@
 
-Now we can test the hypothesis that Hamilton's wet strength `S` is zero:
-
-```{r,testwetstrengthzero,cache=TRUE}
-pnames(H)
-H9 <- keep_flawed(H,c("S",pnames(H)[1:8]))
-specificp.gt.test(H9,"S",0)
-```
-
-now profile likelihood:
-
-```{r,doproflikn,cache=TRUE}
-H9
-maxp(H9)
-p <- seq(from=0.11,to=0.13,len=50)
-u01 <- profsupp(H9,"Hamilton",p,n=01)
-plot(p,u01)
-u20 <- profsupp(H9,"Hamilton",p,n=20)
-plot(p,u20)
-u50 <- profsupp(H9,"Hamilton",p,n=50)
-plot(p,u50)
-uu <- pmax(u01,u20,u50)
-plot(p,uu)
-```
+Now in principle we can test the hypothesis that Hamilton's wet
+strength `S` is zero but I have not done this yet.
 

### 2025-03-03 inst/jss_hyper3_real.Rnw
start discussing counterstrike

@@
 
+\subsection{Counterstrike}
+
+{\em Counter Strike: Global Offensive (CS:GO)} is a multiplayer
+first-person shooter game in which two teams of five compete in an
+immersive virtual reality combat environment.  Gameplay is typically
+complex, including features such as cooperative teamwork, within-team
+communication, and real-time strategic fluidity.
+
+Here I analyse a dataset taken from recent esports tournament {\em
+  ELEAGUE Major: Boston 2018}.  This Valve-sponsored Major featured 24
+elite teams competing for a million dollar prize pool; peak concurrent
+viewership exceeded $1.3\times 10^6$, the highest in CS:GO history at
+the time.  I study the performance of tournament champions {\em
+  Cloud9} who, as underdogs, narrowly defeated aggressive European
+rivals {\em FaZe Clan} in the final, which is widely considered to be
+the
+
+The dataset used is the deathorder
+
 \section{Other stuff}

### 2025-03-03 inst/jss_hyper3_real.Rnw
tiny change

@@
   Cloud9} who, as underdogs, narrowly defeated aggressive European
-rivals {\em FaZe Clan} in the final, which is widely considered to be
-the
+rivals {\em FaZe Clan} in the final.
 
-The dataset used is the deathorder
+The dataset used is the order statistic for number of deaths by each
+of the five {\em Cloud9} players.  Deaths are
 

### 2025-03-04 inst/formula1_points.Rmd
code does not work with new-style ranktables

@@
 ```{r howclosetoreality,cache=TRUE}
-maxp(rank_likelihood(random_table)) # should be close to mL2017
+maxp(suppfun(random_table)) # should be close to mL2017
 ```
 
-```{r checkagain}
+```
 a <- random_table
@@
 
-```{r thuswesee}
+```
 jj <- sort(p,decreasing=TRUE)
@@
 
-```{r,dothework,cache=TRUE}
+```
 n <- 200

### 2025-03-04 inst/jss_hyper3_real.Rnw
more work on counterstrike

@@
 
+\newcommand{\skad}{{\sc Skadoodle}}
+\newcommand{\rush}{{\sc RUSH}}
+\newcommand{\tarik}{{\sc tarik}}
+\newcommand{\stew}{{\sc Stewie2K}}
+
 {\em Counter Strike: Global Offensive (CS:GO)} is a multiplayer
@@
 immersive virtual reality combat environment.  Gameplay is typically
-complex, including features such as cooperative teamwork, within-team
-communication, and real-time strategic fluidity.
-
-Here I analyse a dataset taken from recent esports tournament {\em
+complex, including features such as cooperative teamwork and real-time
+strategic fluidity.
+
+Here I analyse a dataset taken from recent e-sports tournament {\em
   ELEAGUE Major: Boston 2018}.  This Valve-sponsored Major featured 24
@@
 
-The dataset used is the order statistic for number of deaths by each
-of the five {\em Cloud9} players.  Deaths are
+Typically when a player is killed, the team suffers a number of
+strategic disadvantages; player deaths are thus a quantifiable proxy
+for both individual ability and also both teams' tactical execution.
+Here I analyse the order statistic for number of deaths of each of the
+five {\em Cloud9} players.
+
+<<echo=FALSE>>=
+a <- read.table("cloud9_boston_2018.txt", header=TRUE)
+a[,-1]
+@
+
+From the first line we see that {\em Cloud9} played {\em FaZe Clan} on
+the {\em inferno} map, won the match, and \skad\ had the least number
+of deaths, followed by \rush, then \tarik, and so on.  Consider two
+{\em Cloud9} players: \skad\ and \stew.  These teammates, although
+comparable in skill [we fail to reject $H_0\colon p_{\mbox{\tiny\sc
+      Skadoodle}}\neq p_{\mbox{\tiny\sc Stewie2k}}$, Plackett-Luce,
+  $p=0.28$] have playing styles that stand in dramatic contrast.
+Commentators characterise \skad\ as emphasising reactive map control
+while \stew\ is more of a disruptive presence engaging in
+high-variance play.
+
+Given that these two players offer equally attractive and accessible
+targets to the opposing team, it is possible that some tactical
+situations might result in a systematic preference for targetting
+\skad, and some a preference for killing \stew.  If this is so, the
+deaths will be overdispersed and one would have a red bus-blue bus
+situation, but with $\lambda<1$.
+
+To assess this, we follow exactly the same protocol as for the sushi
+preference dataset but with modified Plackett-Luce likelihoods using
+the deathorder statistic from the dataset above, and equivalence
+classes comprising \skad\ and \stew, and everybody else:
+
+<<>>=
+ec <-c(autimatic=1, RUSH=1, Skadoodle=2, Stewie2K=2, tarik=1)
+@
+
+<<defredbusfunc, echo=FALSE>>=
+mk <- function(lam){
+  H3 <- hyper3()
+  for(i in seq_len(nrow(a))){
+
+    d <- strsplit(a$deathorder[i],", ")[[1]]
+    H3 <- H3 + cheering3(d, e=ec, help=c(1,lam))
+  }
+  return(H3)
+}
+f <- function(l){maxp(mk(l), give=1)$`Log-likelihood`}
+@
+
+Find the MLE for $\lambda$:
+
+<<findopt, cache=TRUE, echo=FALSE>>=
+optlam <- optimize(f,c(0.09, 0.14), maximum = TRUE)
+evaluate <- optlam$maximum
+support <- optlam$objective - f(1)
+pvalue <- pchisq(support,df=1,lower.tail=FALSE)
+@
+
+<<calcklikelam, cache=TRUE, echo=FALSE>>=
+lam <- exp(seq(from = -4,to = log(1.1), len = 23))
+L <- sapply(lam, f)
+@
+
+<<findinterval, cache=TRUE, echo=FALSE>>=
+g <- function(l){f(l) - optlam$objective  + 2}
+jj1 <- uniroot(g, interval = exp(c(-4,-3)))
+jj2 <- uniroot(g, interval = exp(c(-1,-0)))
+lowerlimit <- jj1$root
+upperlimit <- jj2$root
+@
+
+\begin{figure}[htbp]
+  \begin{center}
+<<plotlikelam, echo=FALSE>>=
+plot(log(lam), L - max(L), type='b')
+segments(x0=log(optlam$maximum),y0 = -0.5, y1 = 0.1,col='red')
+abline(h=c(0,-2), col = 'gray')
+segments(x0 = log(jj1$root), y0 = -2.2, y1 = -1.8, col = 'blue')
+segments(x0 = log(jj2$root), y0 = -2.2, y1 = -1.8, col = 'blue')
+abline(v = 0, lty = 2)
+@ 
+\caption{A\label{profsupcounterstrike} profile support function for
+  the red bus-blue bus parameter $\log\lambda$ applied to mutually
+  supporting equivalence class
+  $\left\lbrace\mbox{\skad},\mbox{\stew}\right\rbrace$.  Vertical gray
+  line shows $H_0\colon\lambda =1$, for which the support is about
+  $\Sexpr{round(support,2)}$ relative to the evaluate
+  $\hat{\lambda}\simeq\Sexpr{round(evaluate,3)}$ (small red bar),
+  corresponding to a $p$-value of about $\Sexpr{round(pvalue,3)}$.
+  Horizontal lines show the two-units-of-support credible interval of
+  $(\Sexpr{round(lowerlimit,3)},\Sexpr{round(upperlimit,3)})$, marked
+  in blue}
+\end{center}
+\end{figure}
 

### 2025-03-05 inst/jester.Rmd
consistentify "time to process" information
use new-style suppfun()

@@
 ```
+
+<font size="1"> (takes a "long time" to run without cache)
+</font>
 
@@
 jester <- hyper2()
-system.time(for(i in seq_len(nrow(a))){jester <- jester + rank_likelihood(f(a[i,]))})
+system.time(for(i in seq_len(nrow(a))){jester <- jester + suppfun(f(a[i,]))})
 ```

### 2025-03-05 inst/jss_hyper3_real.Rnw
remove unnecessary work

@@
 \end{figure}
-
-\section{Other stuff}
-
-functions~\citep{hankin2010,hankin2017_rnw} account for enhancements
-of competitive ability through the inclusion of reified entities
-(monsters) whose Bradley-Terry strength represents the extra strength.
-For example, if a player is at an advantage for some reason then a
-reified entity might help him and then
-$\Prob(\mbox{$i$~beats~$j$})=\frac{p_i+\mathfrak{M}}{p_i+p_j+\mathfrak{M}}$,
-where $\mathfrak{M}$ is a reified entity whose strength helps
-competitor $i$.
-Such considerations motivate the
-likelihood function in Equations~\ref{likefour}, originally set out by
-XXXXX and YYYY.
-
-\begin{align}\label{likefour}
-\Prob{(\mbox{$i$ wins at home})} &= \frac{\lambda p_i}{\lambda p_i+p_j}\\
-\Prob{(\mbox{$i$ wins away})} &= \frac{\lambda p_i}{\lambda p_i+p_j}\\
-\Prob{(\mbox{$j$ wins at home})} &= \frac{\lambda p_j}{p_i+\lambda p_j}\\
-\Prob{(\mbox{$j$ wins away})} &= \frac{p_j}{\lambda p_i+\lambda p_j}
-\end{align}
-
-Here, $\lambda\geqslant 0$ is a multiplicative constant accounting for
-the home-ground advantage. If $\lambda>1$ the advantage is positive
-and if $\lambda<1$ it is negative and helps the away side.
-
-Here I introduce {\tt hyper3} formalism
-
-However, reified Bradley-Terry techniques suffer from the following
-defect, here illustrated using the context of Scottish football, and
-the home advantage helping the home team and hindering the away team.
-Consider the case of four teams 1-4 and represent the strength of the
-home ground advantage as a home monster of strength $\mathfrak{H}$
-(draws being discarded for the moment).  Now suppose that
-$p_1=p_2=p_A$, and $p_3=p_4=p_B$.  The unit-sum constraint gives
-$2p_A+2p_aB+\mathfrak{H}=1$.  The probability that team1 (at home)
-beats team 2 (away) is then
-$\frac{p_A+\mathfrak{H}}{2p_A+\mathfrak{H}}$; and the probability that
-team 3 (home) beats team 4 (away) is
-$\frac{p_B+\mathfrak{H}}{2p_B+\mathfrak{H}}$.  The odds ratios would
-be $1+\mathfrak{H}/p_A$ and be $1+\mathfrak{H}/p_B$.  Such analysis is
-incompatible with the perfectly reasonable requirement that the odds
-ratio of home victory between two teams of equal strength be constant;
-but here we see that weak teams are predicted to have an overwhelming
-home-ground advantage (finding actual datasets that illustrate
-precisely this phenomenon is not straightforward: reified
-Bradley-Terry methods have a suprising amount of flexibility and can
-accommodate wide ranges of strengths by dint of manipulating reified
-strengths and real strengths).
-
-However, consider the data presented in table~\ref{oldfirm} in which
-we see football results from
 

### 2025-03-05 inst/notthewinner.Rmd
use new-style suppfun()

@@
 (H <- 
-	rank_likelihood(c(a=3,c=2,b=1,d=4)) + 
-	rank_likelihood(c(b=2,c=3,d=1    )) + 
-	rank_likelihood(c(d=2,a=3,c=1    )) + 
-	rank_likelihood(c(b=1,d=2        )) )
+	suppfun(c(a=3,c=2,b=1,d=4)) + 
+	suppfun(c(b=2,c=3,d=1    )) + 
+	suppfun(c(d=2,a=3,c=1    )) + 
+	suppfun(c(b=1,d=2        )) )
 ```
@@
 	names(jj) <- letters[seq_along(jj)]
-	SL[[i]] <- H + rank_likelihood(jj)
+	SL[[i]] <- H + suppfun(jj)
 }

### 2025-03-05 inst/plackett_luce_monster.Rmd
better range
pdftools library not needed

@@
 library("partitions")
-library("pdftools")
 ```
@@
 
+Now use the evaluate for the likelihood function:
+
+```{r calcmaxmax}
+maxHmax <- maxp(RB_BB_LF(s = osup$maximum))
+```
+
+```{r showmaxmax}
+maxHmax
+```
+
 # University ranking analysis

### 2025-03-05 inst/powerboat.Rmd
bugfixes

--- 2024-01-26 inst/powerboat.Rmd
@@
 ```{r readpowerboattable}
-powerboat_table <- read.table("powerboat.txt",header=TRUE)
+jj <- read.table("powerboat.txt",header=TRUE)
 ```
@@
 
-```{r}
+```{r dotable}
+powerboat_table <- ordertable(jj[,1:7])
 powerboat_table[powerboat_table==0] <- "DNS"
-powerboat_table[20:21,8] <- 0  # above does not apply to points
 powerboat_table
+```
+
+```{r dopoints}
+points <- jj[,8]
+names(points) <- rownames(powerboat_table)
+points <- points[order(names(points))]
+points
 ```
@@
 ```{r,definsupportb,cache=TRUE}
-powerboat <- ordertable2supp(powerboat_table[,1:7])
-head(powerboat)
+powerboat <- suppfun(powerboat_table)
 (powerboat_maxp <- maxp(powerboat))
+```
+
+```{r makeplots}
 pie(powerboat_maxp)
 par(pty="s")
-points <- as.numeric(powerboat_table[,8])
-names(points) <- rownames(powerboat_table)
-points
-points <- points[order(names(points))]
-points
-plot(powerboat_maxp,points*max(powerboat_maxp)/max(points),xlab="likelihood",ylab="points",pch=16)
+plot(powerboat_maxp,points/sum(points),xlab="likelihood",ylab="points",pch=16)
 abline(0,1)

### 2025-03-06 inst/pudding.Rmd
tweaks

@@
 p[2 ,4:6] <- c(10,10,27)  # 1v3
-p[4 ,4:6] <- c(10,10,24)  # 1v4
-p[7 ,4:6] <- c(10,10,20)  # 1v5
+p[4 ,4:6] <- c(10,10,34)  # 1v4
+p[7 ,4:6] <- c(20,20,10)  # 1v5
 p[11,4:6] <- c(10,10,31)  # 1v6
@@
 Hp <- convert(p)[[2]]
+options(use_alabama = TRUE)
 mHp <- maxp(Hp)
@@
 
-```{r,label=savecurlingdataset}
-save(pudding_table,pudding,pudding_maxp,curling2_maxp,file="pudding.rda")
+```{r,label=savepuddingdataset}
+save(pudding_table, pudding, pudding_maxp, file = "pudding.rda")
 ```

### 2025-03-06 inst/counterstrike.Rmd
more work

@@
 
-As far as I can tell this is every match played at ELEAGUE Major 2018  in Boston by `Cloud9`.
+As far as I can tell this is every match played at ELEAGUE Major 2018
+in Boston by `Cloud9`.
 
@@
 
-Above we see from the first line that on 28 Jan 2018, playing `FaZe` on the Inferno map we have a win, with deathorder Skadoodle, RUSH, etc.
+Above we see from the first line that on 28 Jan 2018, playing `FaZe`
+on the Inferno map we have a win, with deathorder Skadoodle, RUSH,
+etc.
 
 We will look at the effect of `Skadoodle` on `Stewie2K`.
+
+```{r, counterh2, echo=TRUE}
+H <- hyper2()
+  for(i in seq_len(nrow(a))){
+      d <- strsplit(a$deathorder[i],", ")[[1]]
+      H <- H + suppfun(d)
+  }
+H
+samep.test(H,c("Skadoodle","Stewie2K"))
+```
 
@@
 
-  Now some profile likelihood
-
-```{r mkmaxp, cache=TRUE}
-maxp(mk(0.5),give=1)
-maxp(mk(1.01),give=1)
-```
-
-```{r calcklikef, cache=TRUE}
-f <- function(l){maxp(mk(l), give=1)$`Log-likelihood`}
-lam <- seq(from = 0.01,to = 1.1, len = 43)
-L <- sapply(lam, f)
-```
-
-```{r plotlikelam}
-plot(lam, L - max(L), type='b')
-plot(log(lam), L - max(L), type='b')
-```
+Find the MLE for $\lambda$:
 
@@
 
-```{r}
+And the BT strengths at the evaluate:
+
+```{r BTatevaluate, cache=TRUE}
 maxp(mk(optlam$maximum))
@@
 
+Now some profile likelihood
+
+```{r mkmaxp, cache=TRUE}
+maxp(mk(0.5),give=1)
+maxp(mk(1.01),give=1)
+```
+
+```{r calcklikelam, cache=TRUE}
+f <- function(l){maxp(mk(l), give=1)$`Log-likelihood`}
+lam <- exp(seq(from = -4,to = log(1.1), len = 33))
+L <- sapply(lam, f)
+```
+
+```{r findinterval, cache=TRUE}
+g <- function(l){f(l) - optlam$objective  + 2}
+jj1 <- uniroot(g, interval = exp(c(-4,-3)))
+jj2 <- uniroot(g, interval = exp(c(-1,-0)))
+jj1
+jj2
+```
+
+```{r plotlikelam}
+plot(log(lam), L - max(L), type='l')
+segments(x0=log(optlam$maximum),y0 = -0.5, y1 = 0.1,col='red')
+abline(h=c(0,-2), col = 'gray')
+segments(x0 = log(jj1$root), y0 = -2.2, y1 = -1.8, col = 'blue')
+segments(x0 = log(jj2$root), y0 = -2.2, y1 = -1.8, col = 'blue')
+abline(v = 0, lty = 2)
+```
+
 ### References {-}

### 2025-03-06 inst/jss_hyper3_real.Rnw
more reworking

@@
   p_i+\cdots+p_n}$.  Examples of such enhancements include the
-home-ground advantage in football and rugby, playing white in chess,
-or XXXX.  Note that $\lambda < 1$ would correspond to a decrease in
-competitive ability.
+home-ground advantage in football and rugby, or playing white in
+chess~\citep{hankin2020}.  Note that $\lambda < 1$ would correspond to
+a decrease in competitive ability.
 
-Further, one might follow HANKIN and consider teams of players.  In
-doubles tennis, for example, $\left\lbrace p_1,p_2\right\rbrace$ might
-play $\left\lbrace p_3,p_4\right\rbrace$ and win with probability
-$\frac{p_1+p_2}{p_1+p_2+p_3+p_4}$.  It is natural to combine this idea
-with that of a multiplicative factor.  For example, suppose $p_1$ has
-a special move against $p_3$, and $p_4$ has just been given a
-performance-enhancing drug, we might have a likelihood functions
-proportional to $\frac{\lambda p_1+ p_2}{p_1+p_2+p_3+\mu p_4}$.  Then
-the effectiveness of $p_3$'s special move and the enhancement afforded
-by the drug are measured by $\lambda$ and $\mu$ respectively with
-values $>1$ corresponding to a positive effect.
+Further, one might follow~\cite{hankin2017_rnw} and consider teams of
+players.  In doubles tennis, for example, $\left\lbrace
+p_1,p_2\right\rbrace$ might play $\left\lbrace p_3,p_4\right\rbrace$
+and win with probability
+$\frac{p_1+p_2}{p_1+p_2+p_3+p_4}$~\citep{hankin2010}.  It is natural to
+combine this idea with that of a multiplicative factor.  For example,
+suppose $p_1$ has a special move against $p_3$, and $p_4$ has just
+been given a performance-enhancing drug, we might have a likelihood
+functions proportional to $\frac{\lambda p_1+ p_2}{\lambda
+  p_1+p_2+p_3+\mu p_4}$.  Then the effectiveness of $p_3$'s special
+move and the enhancement afforded by the drug are measured by
+$\lambda$ and $\mu$ respectively with values $>1$ corresponding to a
+positive effect.
 
@@
 
+\begin{table}
+\centering
+\begin{tabular}{@{} *{5}{c} @{}}
+\headercell{\small{Plays at home,}\\ \small{wins}} &
+\multicolumn{4}{c@{}}{\small{Plays away, loses}}\\
+\cmidrule(l){2-5}
+&   Rangers     &    Celtic & Falkirk & Livingston\\
+\midrule
+   Rangers      &   NA &    81  &    69   &      15\\
+   Celtic       &   76 &    NA  &    61   &      15\\
+   Falkirk      &   17 &    17  &    NA   &       6\\
+   Livingston   &    2 &     2  &     3   &      NA\\
+\end{tabular}
+\\
+\rule{1mm}{6mm}
+\\
+\begin{tabular}{@{} *{5}{c} @{}}
+\headercell{\small{Plays at home,}\\ \small{draws}} &
+\multicolumn{4}{c@{}}{\small{Plays away, draws}}\\
+\cmidrule(l){2-5}
+&  Rangers     &    Celtic & Falkirk & Livingston\\
+\midrule
+   Rangers     &       NA  &   44 &     14   &       2\\
+   Celtic      &       47  &   NA &     10   &       3\\
+   Falkirk     &       12  &   19 &     NA   &       4\\
+   Livingston  &       4   &   4  &     4    &      NA\\
+   \end{tabular}
+\\
+\rule{1mm}{6mm}
+\\
+\begin{tabular}{@{} *{5}{c} @{}}
+\headercell{\small{Plays at home,}\\ \small{loses}} &
+\multicolumn{4}{c@{}}{\small{Plays away, wins}}\\
+\cmidrule(l){2-5}
+&  Rangers     &    Celtic & Falkirk & Livingston\\
+\midrule
+   Rangers     &   NA  &  42  &    2  &       0\\
+   Celtic      &   46  &  NA  &    8  &       0\\
+   Falkirk     &   57  &  43  &   NA  &       8\\
+   Livingston  &   12  &  11  &    9  &      NA\\
+  \end{tabular}
+\caption{The \label{oldfirm} dominance of the Old Firm: 669 matches
+  between Rangers, Celtic, Falkirk and Livingston arranged by home
+  wins, draws, and home losses.  Data from {\tt soccerworld}}
+\end{table}
+
 \begin{equation}
@@
 computational methodology is a generalization of that used
-in~\cite{hankin2022} except that the multiplicative factors are real
-(as opposed to non-negative integer), and also unknown.
+in~\cite{hankin2024_hyper3} in which the multiplicative factors are
+known non-negative integers.
 
@@
 support is about \Sexpr{round(f(maxv) - jj$objective,2)}, far
-exceeding Edwards's two units of support per degree of freedom
-criteron.  Alternatively, we can observe that this lies in the tail
-region of its asymptotic null $\chi^2_4$ distribution with a $p$-value
-of about \Sexpr{signif(pchisq(f(maxv) -
+exceeding the two units of support per degree of freedom
+criteron~\citep{edwards1992}.  Alternatively, we can observe that this
+lies in the tail region of its asymptotic null $\chi^2_4$ distribution
+with a $p$-value of about \Sexpr{signif(pchisq(f(maxv) -
   jj$objective,lower.tail=FALSE,df=2),digits=2)}.
@@
 The probability model of Figure~\ref{tikzabcde} is applied, in a
-simple one-parameter form, to two examples: Formula 1 motor racing,
-and also a preference rating.
+simple one-parameter form, to two datasets: one a set of preferences
+of different types of sushi, and one in the field of e-sports.
 
@@
 \cite{chipman1960} and discussed by \cite{davidson1970} and
-\cite[p153]{alvo2014}.  was \cite{debreu1960}.  The Holiday problem is
-discussed in \cite{tverskey1972}.  A traveller has a choice of going
+\cite[p153]{alvo2014}.  
+by \cite{tverskey1972}.
+The canonical problem is as follows. A traveller has a choice of going
 to work by different transport methods.  Here, we can assume that the
 traveller has strong but unknown views on different forms of transport
-which might incorporate a random component.  If considering a car or
-bus, the colour of the bus is of no importance to him.  However, when
-offered the choice between a red bus or a blue bus, the colour becomes
-important (having nothing else to guide his decision).  Now, he does
-not really care what colour the bus is.  If we ask him to rank his
-options, it is highly probable that he will put RB and BB
+which might incorporate a random component.  If considering whether to
+use a car or a bus, the colour of the bus is of no importance to him.
+Such preferences are problematic for Bradley-Terry: If our subject is
+indifferent between cars and buses, say, $p_{C} = p_{RB}$ and also
+indifferent to colour of bus, so $p_{RB}=p_{BB}$, the following
+paradox occurs.  Bradley-Terry strengths must be equal so $p_{C} =
+p_{RB} = p_{BB}=\frac{1}{3}$.  Thus $\Prob(C| \left\lbrace
+C,RB\right\rbrace)=\frac{1}{2}$.  Now we introduce a blue bus as an
+option.  Any realistic choice model will have $\Prob(C | \left\lbrace
+C,RB,BB\right\rbrace)=\frac{1}{2}$ but Bradley-Terry gives us
+$\Prob(C| \left\lbrace C,RB\right\rbrace)=\frac{1}{3}$.  Various
+solutions to this problem have been proposed.  \cite{tverskey1972},
+for example, presents an approach involving overlapping aspects;
+
+If we ask him to
+rank his options, it is highly probable that he will put RB and BB
 consecutively (because they are essentially indistinguishable).  Can
@@
 
-Easily satisfying Edwards's two-units-of-support criterion;
-alternatively, we could observe that this observation lies in the tail
-region of its asymptotic $\chi^2_1$ distribution (Wilks's theorem),
-with a $p$-value of \Sexpr{signif(pchisq(suppdiff*2, df=1,
+Easily satisfying \citeauthor{edwards1992}'s two-units-of-support
+criterion; alternatively, we could observe that this observation lies
+in the tail region of its asymptotic $\chi^2_1$ distribution (Wilks's
+theorem), with a $p$-value of \Sexpr{signif(pchisq(suppdiff*2, df=1,
   lower.tail=FALSE),3)}.
@@
 
-To assess this, we follow exactly the same protocol as for the sushi
-preference dataset but with modified Plackett-Luce likelihoods using
-the deathorder statistic from the dataset above, and equivalence
-classes comprising \skad\ and \stew, and everybody else:
-
-<<>>=
-ec <-c(autimatic=1, RUSH=1, Skadoodle=2, Stewie2K=2, tarik=1)
-@
-
 <<defredbusfunc, echo=FALSE>>=
@@
 @
-
-Find the MLE for $\lambda$:
 
@@
 
-\begin{table}
-\centering
-\begin{tabular}{@{} *{5}{c} @{}}
-\headercell{\small{Plays at home,}\\ \small{wins}} &
-\multicolumn{4}{c@{}}{\small{Plays away, loses}}\\
-\cmidrule(l){2-5}
-&   Rangers     &    Celtic & Falkirk & Livingston\\
-\midrule
-   Rangers      &   NA &    81  &    69   &      15\\
-   Celtic       &   76 &    NA  &    61   &      15\\
-   Falkirk      &   17 &    17  &    NA   &       6\\
-   Livingston   &    2 &     2  &     3   &      NA\\
-\end{tabular}
-\\
-\rule{1mm}{6mm}
-\\
-\begin{tabular}{@{} *{5}{c} @{}}
-\headercell{\small{Plays at home,}\\ \small{draws}} &
-\multicolumn{4}{c@{}}{\small{Plays away, draws}}\\
-\cmidrule(l){2-5}
-&  Rangers     &    Celtic & Falkirk & Livingston\\
-\midrule
-   Rangers     &       NA  &   44 &     14   &       2\\
-   Celtic      &       47  &   NA &     10   &       3\\
-   Falkirk     &       12  &   19 &     NA   &       4\\
-   Livingston  &       4   &   4  &     4    &      NA\\
-   \end{tabular}
-\\
-\rule{1mm}{6mm}
-\\
-\begin{tabular}{@{} *{5}{c} @{}}
-\headercell{\small{Plays at home,}\\ \small{loses}} &
-\multicolumn{4}{c@{}}{\small{Plays away, wins}}\\
-\cmidrule(l){2-5}
-&  Rangers     &    Celtic & Falkirk & Livingston\\
-\midrule
-   Rangers     &   NA  &  42  &    2  &       0\\
-   Celtic      &   46  &  NA  &    8  &       0\\
-   Falkirk     &   57  &  43  &   NA  &       8\\
-   Livingston  &   12  &  11  &    9  &      NA\\
-  \end{tabular}
-\caption{The \label{oldfirm} dominance of the Old Firm: 669 matches
-  between Rangers, Celtic, Falkirk and Livingston arranged by home
-  wins, draws, and home losses.  Data from {\tt soccerworld}}
-\end{table}
-
-\section{Conclusions and further work}
-
-consider Table~\ref{kka}, originally considered
-by~\citet{hankin2020}.  Considering only non-drawn games for the
-moment, we wish to represent the first-mover advantage using a ``White
-monster'', of Bradley-Terry strength \white
-
-\begin{table}
-\centering
-\begin{tabular}{@{} *{4}{c} @{}}
-\headercell{\small{Plays White,}\\ \small{wins}} & \multicolumn{3}{c@{}}{\small{Plays black, loses}}\\
-\cmidrule(l){2-4}
-& An &  Kr & Ks        \\
-\midrule
-  An & -  &  18 &   6  \\
-  Kr &  7 &  -  &  18  \\
-  Ks & 15 &  30 &  -   \\
-\end{tabular}\quad
-\begin{tabular}{@{} *{4}{c} @{}}
-\headercell{\small{Plays W,}\\ \small{draws}} & \multicolumn{3}{c@{}}{\small{Plays black, draws}}\\
- \cmidrule(l){2-4}
-& An &  Kr & Ks        \\
-\midrule
-  An & -  &  20 &  20  \\
-  Kr & 29 &  -  &  72  \\
-  Ks & 26 &  57 &  -   \\
-\end{tabular}\quad
-\begin{tabular}{@{} *{4}{c} @{}}
-\headercell{\small{Plays W,}\\ \small{loses}} & \multicolumn{3}{c@{}}{\small{Plays black, wins}}\\
-\cmidrule(l){2-4}
-& An &  Kr & Ks        \\
-\midrule
-  An & -  &   5 &   11  \\
-  Kr & 13 &  -  &    9  \\
-  Ks &  2 &   7 &  -   \\
-\end{tabular}
-\caption{Results of 365 chess games \label{kka} between Grandmaster
-  Anand, Grandmaster Karpov, and Grandmaster Kasparov.  For example,
-  Anand played white against Karpov a total of $18+20+13=51$ times and
-  won 18, drew 20, lost 13}
-\end{table}
-
-\begin{table}
-\centering
-\begin{tabular}{l|cc|cc}\\&
-  \multicolumn{2}{c}{$i$ plays at home} &
-  \multicolumn{2}{c}{$i$ plays away} \\
- & Reified & $\lambda$ &
-   Reified & $\lambda$ \\
-  \hline
-  $\Prob{(\mbox{$i$ wins})}$  &
-  $\frac{p_i+\home\oo}{p_i+p_j+\home}$ &
-  $\frac{\lambda p_i}{\lambda p_i+p_j}$ &
-  $\frac{p_i}{p_i+p_j+\home}$ &
-  $\frac{p_i}{\lambda p_i+p_j}$ \\
-  $\Prob{(\mbox{$i$ loses})}$ &
-  $\frac{p_j\oo}{p_i+p_j+\home}$   &
-  $\frac{p_j}{p_i+\lambda p_j}$  &
-  $\frac{p_j+\home}{p_i+p_j+\home}$ &
-  $\frac{\lambda p_j}{p_i + \lambda p_j}$
-  \\
-\end{tabular}
-\caption{Probabilities of a win, draw, loss for the two teams under a reified
-  Bradley-Terry model\label{probabilities} and a lambda thing}
-\end{table}
+To assess this, we follow exactly the same protocol as for the sushi
+preference dataset but using the deathorder statistic from the dataset
+above.  We specify two equivalence classes: one comprising \skad\ and
+\stew, and one everyone else.  Figure~\ref{profsupcounterstrike} gives
+a support function for $\lambda$ showing, among other things, that we
+may reject the hypothesis that $\lambda=1$ in favour of $\lambda<1$ by
+a likelihood ratio of about $\Sexpr{round(support,2)}$.  We observe
+that this exceeds the two units of support criterion of
+\cite{edwards1992} [asymptotic $p$-value
+  $\simeq\Sexpr{round(pvalue,3)}$] so we may infer that there is
+indeed some systematic mechanism---perhaps a positive feedback
+loop---which tends to magnify differences in death statistic.
 

### 2025-03-06 inst/jss_hyper3_real.Rnw
typo

@@
 elite teams competing for a million dollar prize pool; peak concurrent
-viewership exceeded $1.3\times 10^6$, the highest in CS:GO history at
-the time.  I study the performance of tournament champions {\em
-  Cloud9} who, as underdogs, narrowly defeated aggressive European
+viewership exceeded $1.3\times 10^6$, the highest in {\em CS:GO}
+history at the time.  I study the performance of tournament champions
+{\em Cloud9} who, as underdogs, narrowly defeated aggressive European
 rivals {\em FaZe Clan} in the final.

### 2025-03-07 inst/jss_hyper3_real.Rnw
more work

@@
 likelihood function described in Table~\ref{windrawloss}; this has a
-number of desirable features.  Firstly, if we consider two players of
-identical strengths and nullify the home ground effect (that is,
+number of desirable features not present in the reified entity
+approach of~\citet{hankin2020}.  Firstly, if we consider two players
+of identical strengths and nullify the home ground effect (that is,
 $\lambda=1$) then the probability of a draw is $\frac{2D}{2+2D}$; the
@@
 \caption{Probabilities of a win, draw, loss for the two teams under a
-  multiplicative Bradley-Terry model\label{windrawloss} and a lambda thing}
+  two-parameter multiplicative Bradley-Terry model\label{windrawloss}}
 \end{table}
@@
 
-\begin{figure}[htbp]
-  \begin{center}
-<<makecontour, cache=TRUE>>=
+<<makecontourdata, cache=TRUE>>=
 n <- 20
@@
 LL <- pmax(-20, LL - max(LL))
-contour(lambda, D, matrix(LL, n, n),levels = -2*(0:9), xlab=expression(lambda),ylab=expression(D))
+@
+
+<<>>=
+lambda
+@
+
+\begin{figure}[htbp]
+  \begin{center}
+<<docontourplot, cache=FALSE>>=
+contour(lambda, D, matrix(LL, n, n),levels = -2*(1:9), xlab=expression(lambda),ylab=expression(D))
 points(maxv[1], maxv[2], pch=16, col='red')
@@
 
-The red bus-blue bus problem was originally stated by
-\cite{chipman1960} and discussed by \cite{davidson1970} and
-\cite[p153]{alvo2014}.  
-by \cite{tverskey1972}.
-The canonical problem is as follows. A traveller has a choice of going
-to work by different transport methods.  Here, we can assume that the
-traveller has strong but unknown views on different forms of transport
-which might incorporate a random component.  If considering whether to
-use a car or a bus, the colour of the bus is of no importance to him.
-Such preferences are problematic for Bradley-Terry: If our subject is
+The classic red bus-blue bus problem was originally stated by
+\cite{mcfadden1980} although different metaphors have been employed
+for the same phenomenon by \cite{tverskey1972}.  The canonical problem
+is as follows.  A traveller has a choice of going to work by different
+transport methods.  Here, we can assume that the traveller has strong
+but unknown views on different forms of transport which might
+incorporate a random component.  If considering whether to use a car
+or a bus, the colour of the bus is of no importance to him.  Such
+preferences are problematic for Bradley-Terry: If our subject is
 indifferent between cars and buses, say, $p_{C} = p_{RB}$ and also
@@
 solutions to this problem have been proposed.  \cite{tverskey1972},
-for example, presents an approach involving overlapping aspects;
+for example, presents an approach involving overlapping ``aspects'' of
+the elements of choice set; \cite{mcfadden1980} suggests using a
+generalized extreme-value model.
 
-If we ask him to
-rank his options, it is highly probable that he will put RB and BB
-consecutively (because they are essentially indistinguishable).  Can
-we quantify the strength of this effect?  To do this, we define a
-bespoke function \code{RB\_BB\_LF()} which returns a \code{hyper3}
-log-likelihood function corresponding to repeated observations of our
-commuter's reported ranks for the five options:
+Here I sugest an alternative in the form of a generalization of
+Plackett-Luce likelihood functions for ranking options, using the
+multiplicative generalization to Bradley-Terry strengths presented
+above.  If we ask our respondent to rank his options, it is highly
+probable that he will put RB and BB consecutively (because they are
+essentially indistinguishable).  Can we quantify the strength of this
+effect?  To do this, we define a bespoke function \code{RB\_BB\_LF()}
+which returns a \code{hyper3} log-likelihood function corresponding to
+repeated observations of our commuter's reported ranks for the five
+options:
 
@@
 quantifies the perceived similarity between RB and BB in the same way
-as function \code{oldfirm()} above.  It is straightforward to find a
-profile support function for $\lambda$,
-Figure~\ref{profilesupportlambdaRBBB}, and formally maximize the
-likelihood:
+as function \code{oldfirm()} above.
 
-\begin{figure}[htbp]
-\begin{center}
-<<plotproflike>>=
+<<definefunctiono>>=
 o <- function(lambda){maxp(RB_BB_LF(lambda), give=1)$`Log-likelihood`} 
-lambda <- exp(seq(from=log(0.9), to=log(47), len=17)) 
-L <- sapply(lambda, o)
-plot(log(lambda), L-max(L), type="b")
-abline(h=c(0, -2))
-abline(v=0)
 @
-\caption{A\label{profilesupportlambdaRBBB} profile support function
-  for the red bus, blue bus parameter $\lambda$.  Vertical line
-  indicates null of $\lambda=1$, horizontal lines show maximum support
-  and $-2$}
-\end{center}
-\end{figure}
+
+Above, we define function \code{o()} which returns the support for a
+given value of $\lambda$ at the evaluate [over permissible Bradley
+  Terry strengths for the options].  We can treat \code{o()} as a
+support function for $\lambda$, and then maximize it:
 
@@
 <<label=LRT_RB, cache=TRUE>>=
-(suppdiff <- o(osup$maximum) - o(1))
+(suppdiff <- osup$objective - o(lambda = 1))
 @
 
-Easily satisfying \citeauthor{edwards1992}'s two-units-of-support
-criterion; alternatively, we could observe that this observation lies
-in the tail region of its asymptotic $\chi^2_1$ distribution (Wilks's
-theorem), with a $p$-value of \Sexpr{signif(pchisq(suppdiff*2, df=1,
-  lower.tail=FALSE),3)}.
+We can thus improve the hypothesis that $\lambda=1$, corresponding to
+an unaltered Plackett-Luce system, by over 5.6 units of support by
+dint of choosing instead $\lambda = 10.49$.  This easily satisfies
+\citeauthor{edwards1992}'s two-units-of-support criterion;
+alternatively, we could observe that this observation lies in the tail
+region of its asymptotic $\chi^2_1$ distribution (Wilks's theorem),
+with a $p$-value of \Sexpr{signif(pchisq(suppdiff*2, df=1,
+  lower.tail=FALSE),3)}.  It is straightforward to find a profile
+support function for $\lambda$, Figure~\ref{profilesupportlambdaRBBB}.
+We see that support of $\log\lambda$ is closely quadratic over a broad
+range.
+
+\begin{figure}[htbp]
+\begin{center}
+<<plotproflike>>=
+lambda <- exp(seq(from=log(0.9), to=log(67), len=19)) 
+L <- sapply(lambda, o)
+plot(log(lambda), L-max(L), type="b")
+abline(h=c(0, -2))
+abline(v=0)
+@ 
+\caption{A\label{profilesupportlambdaRBBB} profile support function
+  for the red bus, blue bus parameter $\lambda$, synthetic dataset.
+  Vertical line indicates null of $\lambda=1$, horizontal lines show
+  maximum support and $-2$}
+\end{center}
+\end{figure}
 
@@
 
-\begin{figure}[htbp]
-  \begin{center}
-<<calclikesushi, cache=TRUE>>=
+<<calclikesushi,echo=FALSE, cache=TRUE>>=
 f <- function(lambda){
@@
 }
-s <- exp(seq(from = log(0.9), to = log(10), len = 12))
+s <- exp(seq(from = log(0.9), to = log(10), len = 23))
 L <- sapply(s,f)
+@
+
+\begin{figure}[htbp]
+  \begin{center}
+<<plotlikesushi>>=
 plot(log(s),L-max(L),type='b',
@@
 \caption{A\label{profilesupportsushi} profile support function for the
-  exchangeability term $\lambda$.  Vertical gray line shows
-  $H_0\colon\lambda =1$, horizontal lines show the
-  two-units-of-support limits}
+  exchangeability term $\lambda$ in the case of sushi preferences.
+  Vertical gray line shows $H_0\colon\lambda =1$, horizontal lines
+  show the two-units-of-support limits}
 \end{center}
@@
 mk <- function(lam){
+    e <- c(Skadoodle=2, RUSH=1, tarik=1, autimatic=1, Stewie2K=2)
     H3 <- hyper3()
     for(i in seq_len(nrow(a))){
-
         d <- strsplit(a$deathorder[i],", ")[[1]]
-    H3 <- H3 + cheering3(d, e=ec, help=c(1,lam))
+        H3 <- H3 + cheering3(d, e=e, help=c(1, lam))
     }
@@
 
-<<findopt, cache=TRUE, echo=FALSE>>=
+<<findopt, cache=TRUE, echo=TRUE>>=
 optlam <- optimize(f,c(0.09, 0.14), maximum = TRUE)
@@
 support <- optlam$objective - f(1)
-pvalue <- pchisq(support,df=1,lower.tail=FALSE)
+pvalue <- pchisq(support, df = 1, lower.tail = FALSE)
 @
 
-<<calcklikelam, cache=TRUE, echo=FALSE>>=
+<<calcklikelambda, cache=TRUE, echo=FALSE>>=
 lam <- exp(seq(from = -4,to = log(1.1), len = 23))
@@
 abline(h=c(0,-2), col = 'gray')
-segments(x0 = log(jj1$root), y0 = -2.2, y1 = -1.8, col = 'blue')
-segments(x0 = log(jj2$root), y0 = -2.2, y1 = -1.8, col = 'blue')
+segments(x0 = log(lowerlimit), y0 = -2.2, y1 = -1.8, col = 'blue')
+segments(x0 = log(upperlimit), y0 = -2.2, y1 = -1.8, col = 'blue')
 abline(v = 0, lty = 2)

### 2025-03-10 inst/jss_hyper3_real.Rnw
more work

@@
 
-  The \code{hyper2} package provides functionality to work with
-  extensions of the Bradley-Terry probability model such as reified
-  entities (monsters) that account for known enhancements in
-  competitive ability.  Here, I present a generalization of
-  \code{hyper2} in which Bradley-Terry strengths may be enhanced using
-  a multiplicative factor $\lambda$.  Generally, ordinary
-  Bradley-Terry is recovered if $\lambda=1$.  This approach offers
-  multiple advantages over reified Bradley-Terry, and several examples
-  are given.
-}
-\Keywords{Plackett-Luce,
-  Bradley-Terry, Mann-Whitney} \Plainkeywords{Plackett-Luce,
-  Bradley-Terry, Mann-Whitney}
+  A generalization of \code{hyper2} is presented in which
+  Bradley-Terry strengths may be modified using a multiplicative
+  factor $\lambda$; Generally, ordinary Bradley-Terry is recovered if
+  $\lambda=1$.  This approach offers multiple advantages over other
+  generalizations such as reified Bradley-Terry.  It becomes possible
+  to represent non-independence of successive observations in
+  Plackett-Luce likelihood functions which allows a parametrization of
+  the ``red bus, blue bus'' problem in preference analysis.  I present
+  examples drawn from the fields of food preference and the e-sport
+  {\em Counterstrike}.  }
+
+\Keywords{Plackett-Luce, Bradley-Terry, Mann-Whitney}
+\Plainkeywords{Plackett-Luce, Bradley-Terry, Mann-Whitney}
 
@@
 and win with probability
-$\frac{p_1+p_2}{p_1+p_2+p_3+p_4}$~\citep{hankin2010}.  It is natural to
-combine this idea with that of a multiplicative factor.  For example,
-suppose $p_1$ has a special move against $p_3$, and $p_4$ has just
-been given a performance-enhancing drug, we might have a likelihood
-functions proportional to $\frac{\lambda p_1+ p_2}{\lambda
-  p_1+p_2+p_3+\mu p_4}$.  Then the effectiveness of $p_3$'s special
-move and the enhancement afforded by the drug are measured by
-$\lambda$ and $\mu$ respectively with values $>1$ corresponding to a
-positive effect.
-
-In this paper I present a generalization of the {\tt hyper2} suite for
-Bradley-Terry likelihoods to accommodate general multiplicative
-factors.  The most general likelihood function would be
+$\frac{p_1+p_2}{p_1+p_2+p_3+p_4}$~\citep{hankin2010}.  It is natural
+to combine this idea with that of a multiplicative factor.  For
+example, suppose $p_1$ has a special move against $p_3$, and $p_4$ has
+just been given a performance-enhancing drug.  If so, we might have
+likelihood function
+
+\begin{equation}\label{like1234}
+  \mathcal{L}{\left(p_1,p_2,p_3,p_4;\lambda,\mu\right)}\propto
+  \frac{
+    \left(\lambda p_1 + p_2\right)^{w_{12}}\cdot
+    \left(p_3 + \mu p_4\right)^{w_{34}}
+    }{
+    \left(\lambda p_1+p_2+p_3+\mu p_4\right)^{w_{12} + w_{34}}
+    }
+\end{equation}
+
+where $w_{12}$ and $w_{34}$ are the number of wins by each team
+respectively.  Then $\lambda$ and $\mu$ represent the effectiveness of
+$p_1$'s special move and the enhancement afforded to $p_4$ by the
+drug, with values $>1$ corresponding to a positive effect.  In this
+paper I generalize the {\tt hyper2} suite for Bradley-Terry
+likelihoods~\citep{hankin2017_rnw} to accommodate this class of
+multiplicative factors.  The most general likelihood function would be
+proportional to
 
@@
   \sum_{j=1}^n
- \left(\lambda_{ij}p_i\right)^{w_i}
+ \left(\lambda_{ij}p_j\right)^{w_i}
   \end{equation}
 
-where the $p_i$ are traditional Bradley-Terry strengths and the
-$\lambda_{ij}\geqslant 0$ are multiplicative factors that account for
-ephemeral or temporary changes in competitive ability.  In practice,
-almost all the $\lambda_{ij}$ are zero.  To fix ideas, consider the
-data presented in Table~\ref{oldfirm}.  Ignoring draws for the moment,
-we have a likelihood proportional to
+where the $p_j$ are traditional Bradley-Terry strengths, $p_j\geqslant
+0$, $\sum p_j=1$, and the $\lambda_{ij}\geqslant 0$ are multiplicative
+factors that account for ephemeral or temporary changes in competitive
+ability.  The data $w_1,\ldots w_N$ correspond to the number of
+``wins'' and number of matches, as per $w_{12},w_{34}$ in
+equation~\ref{like1234}.  In practice, almost all the $\lambda_{ij}$
+are zero.  To fix ideas, consider the data presented in
+Table~\ref{oldfirm}.  Ignoring draws for the moment, we have a
+likelihood proportional to
 
@@
 
-\begin{equation}
+\begin{equation}\label{firstfewRCFL}
   \left(\frac{\lambda p_R}{\lambda p_R + p_C}\right)^{81}\cdot
@@
 
-In {\tt hyper3} idiom, a term such as $\lambda p_R + p_C$ is
-represented as a named numeric vector.  For example, if the home
-strength were $\lambda = 1.8$ then the first bracket would be {\tt
-  c(Rangers=1.8, Celtic=1)} with an associated power of $-81$.  The
-computational methodology is a generalization of that used
-in~\cite{hankin2024_hyper3} in which the multiplicative factors are
-known non-negative integers.
+where the home-ground advantage is represented by~$\lambda$.  In {\tt
+  hyper3} idiom, a term such as $\lambda p_R + p_C$ is represented as
+a named numeric vector.  For example, if the home strength were
+$\lambda = 1.8$ then the first bracket would be {\tt c(Rangers=1.8,
+  Celtic=1)} with an associated power of $-81$.  The computational
+methodology is a generalization of that used
+in~\cite{hankin2024_hyper3} in which the multiplicative factors
+correspond to number of clones.  However, in~\cite{hankin2024_hyper3}
+the factors are known non-negative integers, and here they are, in
+general, unknown non-negative real numbers.
 
 \subsection{The package in use}
-Package idiom for creating a likelihood function for this dataset
-would look something like this: Considering only Rangers and Celtic:
-Rangers at home:
+Package idiom for creating a likelihood function for this dataset is
+straightforward.  Considering first only Rangers-Celtic matches:
 
@@
     H <- hyper3(pnames = c("Rangers", "Celtic"))
-    H[c(Rangers = lambda)            ] %<>% add(81)
-    H[c(Celtic = 1)                  ] %<>% add(42)
+    H[c(Rangers = lambda            )] %<>%      add(81     )
+    H[c(                  Celtic = 1)] %<>%      add(     42)
     H[c(Rangers = lambda, Celtic = 1)] %<>% subtract(81 + 42)
 
-    H[c(Celtic = lambda)             ] %<>% add(76)
-    H[c(Rangers = 1)                 ] %<>% add(46)
+    H[c(Celtic = lambda             )] %<>%      add(     76)
+    H[c(                 Rangers = 1)] %<>%      add(46     )
     H[c(Celtic = lambda, Rangers = 1)] %<>% subtract(46 + 76)
@@
 Rangers-Celtic matches that were not score draws, conditional on
-assumed value $\lambda=1.88$.  We may make inferences about the true
+assumed value $\lambda=1.88$.  This corresponds to the relevant terms
+of equation~\ref{firstfewRCFL}.  We may make inferences about the true
 value of $\lambda$ using profile support:
@@
 supplied value of $\lambda$.  We may use this to create a profile
-likelihood:
+likelihood (figure~\ref{profilesupportlambda}), or to find the maximum
+likelihood estimate for $\lambda$:
 
@@
 @
-\caption{A\label{profilesupportlambda} profile support function for the
-  home-team advantage $\lambda$.  Vertical gray line shows
+\caption{A\label{profilesupportlambda} profile support function for
+  the home-team advantage $\lambda$.  Vertical gray line shows
   $H_0\colon\lambda =1$, horizontal lines show the
@@
 @
+
+Then using this value we may estimate the Bradley-Terry strengths of
+Rangers and Celtic:
 
@@
 We are now in a position to consider the entirety of
-Table~\ref{oldfirm}.  Bespoke function \code{home\_draw\_away3()}
-converts data in the form of Table~\ref{oldfirm} to a \code{hyper3}
-likelihood function following Table~\ref{windrawloss}; it takes
-numerical values for the home advantage $\lambda$ and proclivity to
-draw $D$:
+Table~\ref{oldfirm}.  Bespoke package function
+\code{home\_draw\_away3()} converts data in the form of
+Table~\ref{oldfirm} to a \code{hyper3} likelihood function following
+Table~\ref{windrawloss}; it takes numerical values for the home
+advantage $\lambda$ and proclivity to draw $D$:
 
@@
 probability of win/draw/loss for Rangers would be
-$\left(\frac{p_R}{(1+D)(p_R+p_C)};\frac{D}{1+D};\frac{p_R}{(1+D)(p_R+p_C)}\right)$,
+$\left(\frac{p_R}{(1+D)(p_R+p_C)};\frac{D}{1+D};\frac{p_C}{(1+D)(p_R+p_C)}\right)$,
 estimated at $(0.39;0.27;0.34)$.
 
-<<makecontourdata, cache=TRUE>>=
+<<makecontourdata, echo=FALSE, cache=TRUE>>=
 n <- 20
@@
 LL <- pmax(-20, LL - max(LL))
-@
-
-<<>>=
-lambda
 @
@@
 In the following, I use the ``race'' metaphor: five runners take part
-in a race and arrive in order: $a\succ b\succ c\succ d\succ e$.
-Consider the following Plackett-Luce likelihood function for this
-observation:
+in a race and arrive in order $a\succ b\succ c\succ d\succ e$.  Thus
+competitor $a$ wins, $b$ comes second, and so on.  Consider the
+following Plackett-Luce likelihood function for this observation:
 
@@
 Note the use of multiplication between the terms, indicating
-conditional independence.  We may use the multiplicative
+conditional independence.  Now we use the multiplicative
 generalization of Bradley-Terry strengths to introduce an element of
 non-independence between the terms.  There are many ways to do this,
-but one simple case would be to define equivalence classes of the
+but one simple way would be to define equivalence classes of the
 competitors with each equivalence class comprising mutually supporting
-runners.  Equivalence classes of size one correspond to unsupported
 runners.  The metaphor would be that a runner who has finished the
 race is able to support other members of his equivalence class by
-cheering his teammates, boosting their performance.  The \pkg{hyper2}
-package includes function \code{cheering()} which implements this
-functionality.
+cheering his still-running teammates, boosting their performance.  The
+\pkg{hyper2} package includes function \code{cheering()} which
+implements this functionality.
 
@@
 for the same phenomenon by \cite{tverskey1972}.  The canonical problem
-is as follows.  A traveller has a choice of going to work by different
-transport methods.  Here, we can assume that the traveller has strong
-but unknown views on different forms of transport which might
-incorporate a random component.  If considering whether to use a car
-or a bus, the colour of the bus is of no importance to him.  Such
+is as follows.  A commuter has a choice of going to work by different
+transport methods.  Here, we can assume that he has possibly
+strong---but unknown---views on different forms of transport which
+might incorporate a random component.  If considering whether to use a
+car or a bus, the colour of the bus is of no importance to him.  Such
 preferences are problematic for Bradley-Terry: If our subject is
-indifferent between cars and buses, say, $p_{C} = p_{RB}$ and also
-indifferent to colour of bus, so $p_{RB}=p_{BB}$, the following
+indifferent between cars and buses, then $p_{C} = p_{RB}$.  If he is
+also indifferent to colour of bus, so $p_{RB}=p_{BB}$, the following
 paradox occurs.  Bradley-Terry strengths must be equal so $p_{C} =
@@
 C,RB,BB\right\rbrace)=\frac{1}{2}$ but Bradley-Terry gives us
-$\Prob(C| \left\lbrace C,RB\right\rbrace)=\frac{1}{3}$.  Various
-solutions to this problem have been proposed.  \cite{tverskey1972},
-for example, presents an approach involving overlapping ``aspects'' of
-the elements of choice set; \cite{mcfadden1980} suggests using a
-generalized extreme-value model.
+$\frac{1}{3}$.  Various solutions to this problem have been proposed.
+\cite{tverskey1972}, for example, presents an approach involving
+overlapping ``aspects'' of the elements of choice set;
+\cite{mcfadden1980} suggests using a generalized extreme-value model.
 
@@
 ranking depends on the weather or how tired he is on any given day.
-Observe that in almost every case he ranks RB and BB consecutively.
-Function \code{RB_BB_LF()} takes argument \code{lambda} that
-quantifies the perceived similarity between RB and BB in the same way
-as function \code{oldfirm()} above.
+Vector \code{ec} places the options into equivalence classes;
+equivalence class 3 comprises the two buses and the other classes are
+unsupported as they are of size one.  Vector \code{h} encodes the
+strength of the nonindependence.  Observe that in almost every case he
+ranks RB and BB consecutively.  Function \code{RB_BB_LF()} takes
+argument \code{lambda} that quantifies the perceived similarity
+between RB and BB in the same way as function \code{oldfirm()} above.
 
@@
 with a $p$-value of \Sexpr{signif(pchisq(suppdiff*2, df=1,
-  lower.tail=FALSE),3)}.  It is straightforward to find a profile
-support function for $\lambda$, Figure~\ref{profilesupportlambdaRBBB}.
-We see that support of $\log\lambda$ is closely quadratic over a broad
-range.
-
-\begin{figure}[htbp]
-\begin{center}
-<<plotproflike>>=
+  lower.tail=FALSE),3)}.
+
+%It is straightforward to find a profile
+%support function for $\lambda$, Figure~\ref{profilesupportlambdaRBBB}.
+%We see that support of $\log\lambda$ is closely quadratic over a broad
+%range.
+
+%\begin{figure}[htbp]
+%\begin{center}
+<<plotproflike, echo=FALSE>>=
+if(FALSE){
     lambda <- exp(seq(from=log(0.9), to=log(67), len=19)) 
@@
     abline(v=0)
-@ 
-\caption{A\label{profilesupportlambdaRBBB} profile support function
-  for the red bus, blue bus parameter $\lambda$, synthetic dataset.
-  Vertical line indicates null of $\lambda=1$, horizontal lines show
-  maximum support and $-2$}
-\end{center}
-\end{figure}
+}
+@ 
+%\caption{A\label{profilesupportlambdaRBBB} profile support function
+%  for the red bus, blue bus parameter $\lambda$, synthetic dataset.
+%  Vertical line indicates null of $\lambda=1$, horizontal lines show
+%  maximum support and $-2$}
+%\end{center}
+%\end{figure}
 
@@
 <<showsushitable>>=
-head(noquote(sushi_table))
+noquote(head(sushi_table))
 @
@@
 so class 2 is the tuna group and class 1 everything else.  We can now
-use these datapoint to define a bespoke likelihood function, using
+define a bespoke likelihood function for these observations, using
 $\lambda$ as the argument:
@@
 
+<<findstuffsushi,echo=FALSE,cache=TRUE>>=
+osushi <- optimize(f, c(2,4), maximum=TRUE)
+@
+
+<<tinysushi,echo=FALSE>>=
+sushi_maxlambda <- osushi$maximum
+@
+
+<<findothersushistuff, echo=FALSE, cache=TRUE>>=
+g <- function(lambda){f(lambda) - osushi$objective + 2}
+sushi_lower <- uniroot(g, interval = exp(c(0,1)), tol = 0.01)$root
+sushi_upper <- uniroot(g, interval = exp(c(1,2)), tol = 0.01)$root
+@
+
 \begin{figure}[htbp]
   \begin{center}
-<<plotlikesushi>>=
+<<plotlikesushi, echo=FALSE>>=
 plot(log(s),L-max(L),type='b',
      xlab=expression(log(lambda)), ylab='Log-likelihood')
+abline(v=0,col='gray')
+abline(h=c(0,-2),col='gray')
+segments(x0 = log(sushi_maxlambda), y0 =  0.1, y1 = -1.0, col = 'red' )
+segments(x0 = log(sushi_lower    ), y0 = -2.2, y1 = -1.8, col = 'blue')
+segments(x0 = log(sushi_upper    ), y0 = -2.2, y1 = -1.8, col = 'blue')
 @ 
@@
   Vertical gray line shows $H_0\colon\lambda =1$, horizontal lines
-  show the two-units-of-support limits}
+  show the two-units-of-support limits with the credible interval of
+  $(\Sexpr{round(sushi_lower,2)},\Sexpr{round(sushi_upper,2)})$,
+  marked in blue.  Red bar shows evaluate at
+  $\lambda\simeq\Sexpr{round(sushi_maxlambda,2)}$}
 \end{center}
@@
 comparable in skill [we fail to reject $H_0\colon p_{\mbox{\tiny\sc
-      Skadoodle}}\neq p_{\mbox{\tiny\sc Stewie2k}}$, Plackett-Luce,
+      Skadoodle}} = p_{\mbox{\tiny\sc Stewie2k}}$, Plackett-Luce,
   $p=0.28$] have playing styles that stand in dramatic contrast.
@@
 
-<<findopt, cache=TRUE, echo=TRUE>>=
-optlam <- optimize(f,c(0.09, 0.14), maximum = TRUE)
-evaluate <- optlam$maximum
-support <- optlam$objective - f(1)
-pvalue <- pchisq(support, df = 1, lower.tail = FALSE)
-@
-
-<<calcklikelambda, cache=TRUE, echo=FALSE>>=
-lam <- exp(seq(from = -4,to = log(1.1), len = 23))
+<<csgo_findopt, cache=TRUE, echo=FALSE>>=
+csgo_optimize <- optimize(f,c(0.09, 0.14), maximum = TRUE, tol = 0.01)
+csgo_optlam   <- csgo_optimize$maximum
+csgo_support  <- csgo_optimize$objective - f(1)
+csgo_pvalue   <- pchisq(csgo_support, df = 1, lower.tail = FALSE)
+@
+
+<<csgo_calcklikelambda, cache=TRUE, echo=FALSE>>=
+lam <- exp(seq(from = -4, to = log(1.1), len = 23))
 L <- sapply(lam, f)
@@
 
-<<findinterval, cache=TRUE, echo=FALSE>>=
-g <- function(l){f(l) - optlam$objective  + 2}
-jj1 <- uniroot(g, interval = exp(c(-4,-3)))
-jj2 <- uniroot(g, interval = exp(c(-1,-0)))
-lowerlimit <- jj1$root
-upperlimit <- jj2$root
+<<cgso_findinterval, cache=TRUE, echo=FALSE>>=
+g <- function(l){f(l) - csgo_optimize$objective  + 2}
+csgo_lower <- uniroot(g, interval = exp(c(-4,-3)))$root
+csgo_upper <- uniroot(g, interval = exp(c(-1,-0)))$root
 @
@@
   \begin{center}
-<<plotlikelam, echo=FALSE>>=
+<<csgo_plotlikelam, echo=FALSE>>=
 plot(log(lam), L - max(L), type='b')
-segments(x0=log(optlam$maximum),y0 = -0.5, y1 = 0.1,col='red')
-abline(h=c(0,-2), col = 'gray')
-segments(x0 = log(lowerlimit), y0 = -2.2, y1 = -1.8, col = 'blue')
-segments(x0 = log(upperlimit), y0 = -2.2, y1 = -1.8, col = 'blue')
-abline(v = 0, lty = 2)
+segments(x0 = log(csgo_optlam),y0 = -0.5, y1 =  0.1, col='red')
+segments(x0 = log(csgo_lower), y0 = -2.2, y1 = -1.8, col = 'blue')
+segments(x0 = log(csgo_upper), y0 = -2.2, y1 = -1.8, col = 'blue')
+abline(v = 0, col = 'gray')
+abline(h = c(0, -2), col = 'gray')
 @ 
@@
   line shows $H_0\colon\lambda =1$, for which the support is about
-  $\Sexpr{round(support,2)}$ relative to the evaluate
-  $\hat{\lambda}\simeq\Sexpr{round(evaluate,3)}$ (small red bar),
-  corresponding to a $p$-value of about $\Sexpr{round(pvalue,3)}$.
-  Horizontal lines show the two-units-of-support credible interval of
-  $(\Sexpr{round(lowerlimit,3)},\Sexpr{round(upperlimit,3)})$, marked
+  $\Sexpr{round(csgo_support,2)}$ relative to the evaluate
+  $\hat{\lambda}\simeq\Sexpr{round(csgo_optlam,3)}$ (small red bar),
+  corresponding to a $p$-value of about
+  $\Sexpr{round(csgo_pvalue,3)}$.  Horizontal lines show the
+  two-units-of-support credible interval of
+  $(\Sexpr{round(csgo_lower,3)},\Sexpr{round(csgo_upper,3)})$, marked
   in blue}
@@
 may reject the hypothesis that $\lambda=1$ in favour of $\lambda<1$ by
-a likelihood ratio of about $\Sexpr{round(support,2)}$.  We observe
-that this exceeds the two units of support criterion of
+a likelihood ratio of about $\Sexpr{round(csgo_support,2)}$.  We
+observe that this exceeds the two units of support criterion of
 \cite{edwards1992} [asymptotic $p$-value
-  $\simeq\Sexpr{round(pvalue,3)}$] so we may infer that there is
+  $\simeq\Sexpr{round(csgo_pvalue,3)}$] so we may infer that there is
 indeed some systematic mechanism---perhaps a positive feedback
@@
 
-Plackett-Luce likelihood functions for rank datasets have been
+Bradley-Terry strengths $p_1,\ldots,p_n$, $p_i\geqslant 0$, $\sum
+p_i=1$ are generalized with non-negative multiplicative $\lambda$-
+factors that represent temporary or ephemeral changes in competitive
+strength.  The {\tt hyper2} R package implements the resulting
+likelihood functions, and furnishes a wide range of functionality
+geared towards assessing meaningful nulls using likelihood methods.
+
+The standard Plackett-Luce likelihood function
+$\prod_{i=1}^{n}p_{(i)}\left[\sum_{j=i}^n p_{(j)}\right]^{-1}$ over
+Bradley-Terry strengths assumes conditional independence of successive
+rank observations $p_{(1)},\ldots p_{(n)}$, an assumption which is
+broken in the current work.  As indicated above, multiplicative
+$\lambda$- factors may be used to implement many different
+non-independence schemes and comparing these might be interesting.
+
+Further work might include an individualised multiplicative draw
+factor, giving rise to likelihood functions such as
+$\frac{p_1^{w_1};(\lambda_1p_1+\lambda_2p_2)^{d_{12}};
+  p_2^{w_2}}{\left(p_1(1+\lambda_1)+p_2(1+\lambda_2)\right)^{w_1+d_{12}+w_2}}$.
+One might also consider multiplicative factors for the strengths of
+{\em reified} entities.  These modifications would be directly
+representable with existing {\tt hyper3} idiom.
 

### 2025-03-13 inst/jss_hyper3_real.Rnw
minor tweaks

@@
   University of Stirling\\
-  E-mail: \email{hankin.robin@gmail.com}\\
-  URL: \url{https://academics.aut.ac.nz/robin.hankin}
+  E-mail: \email{hankin.robin@gmail.com}
 }
@@
 The Bradley-Terry model for datasets involving paired comparisons has
-wide uptake in the {R} community.  The canonical problem is to
-consider~$n$ players who compete against one another; the basic
-inference problem is to estimate
+wide uptake in the {R} community.  The canonical case is to
+consider~$n$ players who compete against one another in pairs; the
+basic inference problem is to estimate
 numbers~$\mathbf{p}=\left({p_1,\ldots,p_n}\right)$, $p_i\geqslant 0$,
@@
 where $w_{12}$ and $w_{34}$ are the number of wins by each team
-respectively.  Then $\lambda$ and $\mu$ represent the effectiveness of
-$p_1$'s special move and the enhancement afforded to $p_4$ by the
-drug, with values $>1$ corresponding to a positive effect.  In this
+respectively.  Then $\lambda$ would represent the effectiveness of
+$p_1$'s special move, and $\mu$ the enhancement afforded to $p_4$ by
+the drug; values $>1$ would correspond to a positive effect.  In this
 paper I generalize the {\tt hyper2} suite for Bradley-Terry
@@
 criteron~\citep{edwards1992}.  Alternatively, we can observe that this
-lies in the tail region of its asymptotic null $\chi^2_4$ distribution
+lies in the tail region of its asymptotic null $\chi^2_2$ distribution
 with a $p$-value of about \Sexpr{signif(pchisq(f(maxv) -
@@
 probability tree diagram for some of the $5!=120$ possible order
-statistics.  The standard Plackett-Luce likelihoods have been modified
-to account for two groups of mutually supporting runners:
-$\left\lbrace a,b,c\right\rbrace$ with support term $\lambda$, and
-$\left\lbrace d,e\right\rbrace$ with support term $\mu$.  From START,
-the first runner to cross the finishing line has standard
-Plackett-Luce probability.  Taking the top path as an example, we see
-that the likelihood function for $a\succ b\succ c\succ d\succ e$ would
-be
+statistics for five competitors.  The standard Plackett-Luce
+likelihoods have been modified to account for two groups of mutually
+supporting runners: $\left\lbrace a,b,c\right\rbrace$ with support
+term $\lambda$, and $\left\lbrace d,e\right\rbrace$ with support term
+$\mu$.  From START, the first runner to cross the finishing line has
+standard Plackett-Luce probability.  Taking the top path as an
+example, we see that the likelihood function for $a\succ b\succ c\succ
+d\succ e$ would be
 

### 2025-03-13 inst/jss_hyper3_real.Rnw
label final node
slight rewriting

@@
 
-  A generalization of \code{hyper2} is presented in which
-  Bradley-Terry strengths may be modified using a multiplicative
-  factor $\lambda$; Generally, ordinary Bradley-Terry is recovered if
-  $\lambda=1$.  This approach offers multiple advantages over other
-  generalizations such as reified Bradley-Terry.  It becomes possible
-  to represent non-independence of successive observations in
-  Plackett-Luce likelihood functions which allows a parametrization of
-  the ``red bus, blue bus'' problem in preference analysis.  I present
-  examples drawn from the fields of food preference and the e-sport
-  {\em Counterstrike}.  }
+  A generalization of the Bradley-Terry model is presented in which
+  strengths may be modified using a multiplicative factor $\lambda$;
+  Generally, ordinary Bradley-Terry is recovered if $\lambda=1$.  This
+  approach offers multiple advantages over other generalizations such
+  as reified Bradley-Terry.  It becomes possible to represent
+  non-independence of successive observations in Plackett-Luce
+  likelihood functions which allows a parametrization of the ``red
+  bus, blue bus'' problem in preference analysis.  I present examples
+  drawn from the fields of food preference and the e-sport {\em
+    Counterstrike}.  }
 
@@
 basic inference problem is to estimate
-numbers~$\mathbf{p}=\left({p_1,\ldots,p_n}\right)$, $p_i\geqslant 0$,
-$\sum p_i=1$ so that
-$\Prob(\mbox{$i$~beats~$j$})=\frac{p_i}{p_i+p_j}$.  Information about
-the~$p_i$ may be obtained from the results of paired comparisons
-between the players.
+numbers~$\left({p_1,\ldots,p_n}\right)$, $p_i\geqslant 0$, $\sum
+p_i=1$ so that $\Prob(\mbox{$i$~beats~$j$})=\frac{p_i}{p_i+p_j}$.
+Information about the~$p_i$ may be obtained from the results of paired
+comparisons between the players.
 
@@
 the drug; values $>1$ would correspond to a positive effect.  In this
-paper I generalize the {\tt hyper2} suite for Bradley-Terry
-likelihoods~\citep{hankin2017_rnw} to accommodate this class of
-multiplicative factors.  The most general likelihood function would be
-proportional to
+paper I generalize the {\tt hyper2} R package~\citep{hankin2017_rnw},
+available on CRAN~\citep{rcore2025}, for Bradley-Terry likelihoods to
+accommodate this class of multiplicative factors.  The most general
+likelihood function would be proportional to
 
@@
 \\
-\rule{1mm}{6mm}
+\rule{0mm}{16mm}
 \\
@@
 
-where the home-ground advantage is represented by~$\lambda$.  In {\tt
-  hyper3} idiom, a term such as $\lambda p_R + p_C$ is represented as
-a named numeric vector.  For example, if the home strength were
-$\lambda = 1.8$ then the first bracket would be {\tt c(Rangers=1.8,
-  Celtic=1)} with an associated power of $-81$.  The computational
-methodology is a generalization of that used
-in~\cite{hankin2024_hyper3} in which the multiplicative factors
-correspond to number of clones.  However, in~\cite{hankin2024_hyper3}
-the factors are known non-negative integers, and here they are, in
-general, unknown non-negative real numbers.
+where the home-ground advantage is represented by~$\lambda$.  In
+package idiom, a term such as $\lambda p_R + p_C$ is represented as a
+named numeric vector.  For example, if the home strength were $\lambda
+= 1.8$ then the first bracket would be {\tt c(Rangers=1.8, Celtic=1)}
+with an associated power of $-81$.  The computational methodology is a
+generalization of that used in~\cite{hankin2024_hyper3} in which the
+multiplicative factors correspond to number of clones.  However,
+in~\cite{hankin2024_hyper3} the factors are known non-negative
+integers, and here they are, in general, unknown non-negative real
+numbers.
 
@@
 \fill (12.2,7) circle[radius=2pt];
-\node[anchor = west] at (12.2,7) {$a\succ b\succ c\succ d\succ e$};
+\node at (12.3,6.8) {$d$};
+\node[anchor = west] at (12.5,7) {$a\succ b\succ c\succ d\succ e$};
 \node at (11.8, 7.4) {$\frac{p_d}{p_d+p_e}$};
@@
 \fill (12.2,6.2) circle[radius=2pt];
-\node[anchor = west] at (12.2,6.2) {$a\succ b\succ c\succ e\succ d$};
+\node at (12.3,6.0) {$e$};
+\node[anchor = west] at (12.5,6.2) {$a\succ b\succ c\succ e\succ d$};
 \node at (11.8, 6.6) {$\frac{p_e}{p_d+p_e}$};
@@
 \fill (12.2,5.3) circle[radius=2pt];
-\node[anchor = west] at (12.2,5.3) {$a\succ b\succ d\succ c\succ e$};
+\node at (12.3,5.1) {$c$};
+\node[anchor = west] at (12.5,5.3) {$a\succ b\succ d\succ c\succ e$};
 \node at (11.8, 5.7) {$\frac{\la{p_c}}{\la{p_c}+\ld{p_e}}$};
@@
 \fill (12.2,4.7) circle[radius=2pt];
-\node[anchor = west] at (12.2,4.7) {$a\succ b\succ d\succ e\succ c$};
+\node at (12.3,4.5) {$e$};
+\node[anchor = west] at (12.5,4.7) {$a\succ b\succ d\succ e\succ c$};
 \node at (11.8, 5.0) {$\frac{\ld{p_e}}{\la{p_c}+\ld{p_e}}$};
@@
 \fill (12.2,3.9) circle[radius=2pt];
-\node[anchor = west] at (12.2,3.9) {$a\succ b\succ e\succ c\succ d$};
+\node at (12.3,3.7) {$c$};
+\node[anchor = west] at (12.5,3.9) {$a\succ b\succ e\succ c\succ d$};
 \node at (11.8, 4.3) {$\frac{\la{p_c}}{\la{p_c}+\ld{p_d}}$};
@@
 \node at (11.8, 3.5) {$\frac{\ld{p_d}}{\la{p_c}+\ld{p_d}}$};
-\node[anchor = west] at (12.2,3.1) {$a\succ b\succ e\succ d\succ c$};
+\node at (12.3,2.9) {$d$};
+\node[anchor = west] at (12.5,3.1) {$a\succ b\succ e\succ d\succ c$};
 
@@
 \fill (12.2,1) circle[radius=2pt];
-\node[anchor = west] at (12.2,1) {$d\succ a\succ c\succ e\succ b$};
+\node at (12.3,0.8) {$e$};
+\node[anchor = west] at (12.5,1.0) {$d\succ a\succ c\succ e\succ b$};
 \node at (7, 1.4)  {$\frac{p_a}{p_a+p_b+p_c+\ld{p_e}}$};
@@
 \fill (12.2,0) circle[radius=2pt];
-\node[anchor = west] at (12.2,0) {$d\succ e\succ a\succ b\succ c$};
+\node at (12.3, -0.2) {$b$};
+\node[anchor = west] at (12.5, 0) {$d\succ e\succ a\succ b\succ c$};
 \node at (7, 0.4)  {$\frac{p_b}{p_a+p_b+p_c+\ld{p_e}}$};
@@
 \fill (12.2,-1) circle[radius=2pt];
-\node[anchor = west] at (12.2,-1) {$d\succ e\succ a\succ c\succ b$};
+\node at (12.3,-1.2) {$c$};
+\node[anchor = west] at (12.5,-1) {$d\succ e\succ a\succ c\succ b$};
 
@@
 \fill (12.2,-1.8) circle[radius=2pt];
-\node[anchor = west] at (12.2,-1.8) {$d\succ e\succ b\succ a\succ c$};
+\node at (12.3, -2.0) {$a$};
+\node[anchor = west] at (12.5, -1.8) {$d\succ e\succ b\succ a\succ c$};
 \node at (11.8, -1.4) {$\frac{\la{p_a}}{\la{p_a}+\la{p_c}}$};
@@
 \fill (12.2,-2.6) circle[radius=2pt];
-\node[anchor = west] at (12.2,-2.6) {$d\succ e\succ b\succ c\succ a$};
+\node at (12.3, -2.8) {$c$};
+\node[anchor = west] at (12.5,-2.6) {$d\succ e\succ b\succ c\succ a$};
 \node at (11.8, -2.2) {$\frac{\la{p_c}}{\la{p_a}+\la{p_c}}$};
@@
 \fill (12.2,-3.4) circle[radius=2pt];
-\node[anchor = west] at (12.2,-3.4) {$d\succ e\succ c\succ b\succ a$};
+\node at (12.3, -3.6) {$b$};
+\node[anchor = west] at (12.5,-3.4) {$d\succ e\succ c\succ b\succ a$};
 
@@
 \fill (12.2,2) circle[radius=2pt];
-\node[anchor = west] at (12.2,2) {$d\succ a\succ c\succ b\succ e$};
+\node at (12.3, 1.8) {$b$};
+\node[anchor = west] at (12.5, 2) {$d\succ a\succ c\succ b\succ e$};
 \draw[thick] (10.6, 1) -- (11.2,1) -- (12.2,1);
@@
 
-Then, as before, we may define a profile likelihood function of
-$\lambda$, Figure~\ref{profilesupportsushi}.
-
 <<calclikesushi,echo=FALSE, cache=TRUE>>=
@@
 
-<<findstuffsushi,echo=FALSE,cache=TRUE>>=
+% chunk 'findstuffsushi' takes about 10-20 minutes to run
+
+<<findstuffsushi,echo=FALSE, cache=TRUE>>=
 osushi <- optimize(f, c(2,4), maximum=TRUE)
@@
 
-<<tinysushi,echo=FALSE>>=
-sushi_maxlambda <- osushi$maximum
-@
+% chunk tinysushi' takes about 5 minutes to run
+
+<<tinysushi, echo=FALSE, cache=TRUE>>=
+sushimaxlambda <- osushi$maximum
+sushisupport <- osushi$objective - f(1)
+@
+
+% chunk 'findotherstuffsushi' takes about 10 minutes to run
 
@@
 @
+
+(above, we note that the return value {\tt H} is the sum of successive
+log-likelihood functions of each row of the table, assumed to be
+independent).  Then, as before, we may define a profile likelihood
+function of $\lambda$, Figure~\ref{profilesupportsushi}.  We
+reject~$H_0\colon\lambda=1$ in favour of
+$H_A\colon\lambda=\Sexpr{round(sushimaxlambda,2)}$ with about
+$\Sexpr{round(sushisupport,2)}$ units of support.  Alternatively, we
+may observe that this lies in the tail region of its asymptotic
+$\chi^2_1$ distribution with a $p$-value of about
+$\Sexpr{round(pchisq(sushisupport,df=1,lower.tail=FALSE),3)}$.  We can
+therefore be reasonably sure that tuna sushi are somehow analagous to
+red and blue buses, in the sense that all three tuna items tend to
+occupy closely separated places in judges' preference ranks.
 
@@
 abline(h=c(0,-2),col='gray')
-segments(x0 = log(sushi_maxlambda), y0 =  0.1, y1 = -1.0, col = 'red' )
+segments(x0 = log(sushimaxlambda), y0 =  0.1, y1 = -1.0, col = 'red' )
 segments(x0 = log(sushi_lower    ), y0 = -2.2, y1 = -1.8, col = 'blue')
@@
   marked in blue.  Red bar shows evaluate at
-  $\lambda\simeq\Sexpr{round(sushi_maxlambda,2)}$}
+  $\lambda\simeq\Sexpr{round(sushimaxlambda,2)}$}
 \end{center}

### 2025-03-14 inst/jss_hyper3_real.Rnw
various typos spotted by Gavin

@@
 to combine this idea with that of a multiplicative factor.  For
-example, suppose $p_1$ has a special move against $p_3$, and $p_4$ has
-just been given a performance-enhancing drug.  If so, we might have
-likelihood function
+example, suppose $p_1$ has just learned a new special technique, and
+$p_4$ has just been given a performance-enhancing drug.  If so, we
+might have likelihood function
 
@@
 respectively.  Then $\lambda$ would represent the effectiveness of
-$p_1$'s special move, and $\mu$ the enhancement afforded to $p_4$ by
-the drug; values $>1$ would correspond to a positive effect.  In this
-paper I generalize the {\tt hyper2} R package~\citep{hankin2017_rnw},
-available on CRAN~\citep{rcore2025}, for Bradley-Terry likelihoods to
-accommodate this class of multiplicative factors.  The most general
-likelihood function would be proportional to
+$p_1$'s special technique, and $\mu$ the enhancement afforded to $p_4$
+by the drug; values $>1$ would correspond to a positive effect.  In
+this paper I generalize the {\tt hyper2} R
+package~\citep{hankin2017_rnw}, available on CRAN~\citep{rcore2025},
+for Bradley-Terry likelihoods to accommodate this class of
+multiplicative factors.  The most general likelihood function would be
+proportional to
 
@@
   \end{tabular}
-\caption{The \label{oldfirm} dominance of the Old Firm: 669 matches
+\caption{The \label{oldfirm} dominance of the Old Firm: 769 matches
   between Rangers, Celtic, Falkirk and Livingston arranged by home
@@
   \left(\frac{\lambda p_R}{\lambda p_R + p_F}\right)^{69}\cdots
-  \left(\frac{\lambda p_L}{\lambda p_L + p_F}\right)^{81}\cdots
+  \left(\frac{\lambda p_L}{\lambda p_L + p_F}\right)^{3}\cdots
   \left(\frac{p_C}{p_C + \lambda p_R}\right)^{42}\cdots
@@
   $\frac{\lambda p_i}{\lambda p_i + D(p_i + p_j) + p_j}$  &
-  $\frac{p_i}{p_i + D(p_i + \lambda p_j) + p_j}$          \\
+  $\frac{p_i}{p_i + D(p_i + p_j) + p_j}$          \\
   $\Prob{(\mbox{draw})}$                                 &
@@
 <<docontourplot, cache=FALSE>>=
-contour(lambda, D, matrix(LL, n, n),levels = -2*(1:9), xlab=expression(lambda),ylab=expression(D))
-points(maxv[1], maxv[2], pch=16, col='red')
+contour(lambda, D, matrix(LL, n, n), levels = -2*(1:9),
+        xlab = expression(lambda), ylab = expression(D))
+points(maxv[1], maxv[2], pch = 16, col = 'red')
+abline(v=1, col = 'gray')
 @
@@
   a grid.  Maximum likelihood estimate for $(\lambda, D)$ shown as
-  a red dot at the evaluate}
+  a red dot at the evaluate.  Vertical gray line shows $H_0\colon\lambda=1$}
   \end{center}
@@
 example, we may test the hypothesis that $\lambda=1$ [corresponding to
-  the right hand side of figure~\ref{contourlambdaD}] for which the
+  the left hand side of figure~\ref{contourlambdaD}] for which the
 support is about \Sexpr{round(f(maxv) - jj$objective,2)}, far
@@
 \frac{p_a}{p_a + p_b + p_c + \mu p_e}\cdot
-\frac{\lambda p_c}{\lambda p_b + \lambda p_c + p_e}\cdot
+\frac{\lambda p_c}{\lambda p_b + \lambda p_c + \mu p_e}\cdot
 \frac{\lambda p_b}{\lambda p_b + \mu p_e}
@@
 simple one-parameter form, to two datasets: one a set of preferences
-of different types of sushi, and one in the field of e-sports.
+of different types of sushi, and one in the field of esports.
 
@@
 
-Here I sugest an alternative in the form of a generalization of
+Here I suggest an alternative in the form of a generalization of
 Plackett-Luce likelihood functions for ranking options, using the
@@
 
-Here I analyse a dataset taken from recent e-sports tournament {\em
+Here I analyse a dataset taken from recent esports tournament {\em
   ELEAGUE Major: Boston 2018}.  This Valve-sponsored Major featured 24

### 2025-03-18 inst/jss_hyper3_real.Rnw
more suggestions from Gavin

@@
 for Bradley-Terry likelihoods to accommodate this class of
-multiplicative factors.  The most general likelihood function would be
+multiplicative factors with new \proglang{S3} class {\tt hyper3} and
+associated methods.
+
+The most general likelihood function would be
 proportional to
@@
 simple one-parameter form, to two datasets: one a set of preferences
-of different types of sushi, and one in the field of esports.
-
-\section{Red bus, blue bus}
+of different types of sushi, and one in the field of esports.  First,
+however, we discuss a synthetic example showing how generalizing
+Placket-Luce to admit non-independence of successive arrivals can be
+applied to an apparently different situation.
+
+\subsection{Red bus, blue bus}
 
@@
 
-Here I suggest an alternative in the form of a generalization of
-Plackett-Luce likelihood functions for ranking options, using the
-multiplicative generalization to Bradley-Terry strengths presented
-above.  If we ask our respondent to rank his options, it is highly
-probable that he will put RB and BB consecutively (because they are
-essentially indistinguishable).  Can we quantify the strength of this
-effect?  To do this, we define a bespoke function \code{RB\_BB\_LF()}
-which returns a \code{hyper3} log-likelihood function corresponding to
+Here I suggest an alternative in the form of a one-parameter
+generalization of Plackett-Luce likelihood functions for ranking
+options, using the multiplicative generalization to Bradley-Terry
+strengths presented above.  As a synthetic example, we specify options
+RB and BB for red bus and blue bus together options C, T and W for
+car, train, and walking.
+
+Specifically, it seems reasonable to assume that if we ask our
+respondent to rank his options, it is highly probable that he will put
+RB and BB consecutively (because they are essentially
+indistinguishable).  Can we quantify the strength of this effect?  To
+do this, we define a bespoke function \code{RB\_BB\_LF()} which
+returns a \code{hyper3} log-likelihood function corresponding to
 repeated observations of our commuter's reported ranks for the five
-options:
+options.
 
@@
     (
-        cheering3(v=c("RB", "BB", "C" , "T", "W"), e=ec, h=h)*3 + 
-        cheering3(v=c("BB", "RB", "T" , "C", "W"), e=ec, h=h)*2 + 
-        cheering3(v=c("T" , "BB", "RB", "C", "W"), e=ec, h=h)*5 + 
-        cheering3(v=c("W" , "BB", "RB", "T", "C"), e=ec, h=h)*4 + 
-        cheering3(v=c("C" , "RB", "BB", "W", "T"), e=ec, h=h)*3 + 
-        cheering3(v=c("BB", "C" , "RB", "T", "W"), e=ec, h=h)*3
+        cheering3(v=c("RB", "BB", "C" , "T", "W"), e = ec, h = h) * 3 + 
+        cheering3(v=c("BB", "RB", "T" , "C", "W"), e = ec, h = h) * 2 + 
+        cheering3(v=c("T" , "BB", "RB", "C", "W"), e = ec, h = h) * 5 + 
+        cheering3(v=c("W" , "BB", "RB", "T", "C"), e = ec, h = h) * 4 + 
+        cheering3(v=c("C" , "RB", "BB", "W", "T"), e = ec, h = h) * 3 + 
+        cheering3(v=c("BB", "C" , "RB", "T", "W"), e = ec, h = h) * 3
     )
@@
 moment and this disincentivises further tuna choices.  This mechanism
-would be modelled by having $\lambda<1$.
-
-Implementation in \code{hyper3} formalism is straightforward.  The
-dataset used is \code{sushi\_table}, part of the \code{hyper2} package:
+would be modelled by having $\lambda<1$.  Implementation in
+\code{hyper3} formalism is straightforward.  The dataset used is
+\code{sushi\_table}, part of the \code{hyper2} package:
 
@@
 strength.  The {\tt hyper2} R package implements the resulting
-likelihood functions, and furnishes a wide range of functionality
-geared towards assessing meaningful nulls using likelihood methods.
+likelihood functions using \proglang{S3} class {\tt hyper3}, and
+furnishes a wide range of functionality geared towards assessing
+meaningful nulls using likelihood methods.
 
@@
 $\lambda$- factors may be used to implement many different
-non-independence schemes and comparing these might be interesting.
+non-independence schemes.
 

### 2025-03-19 inst/jss_hyper3_real.Rnw
cite Plackett

@@
 
-\begin{equation}
+\begin{equation}\label{mostgeneral}
   \prod_{i=1}^N
@@
   between Rangers, Celtic, Falkirk and Livingston arranged by home
-  wins, draws, and home losses.  Data from {\tt soccerworld}}
+  wins, draws, and home losses.  Data from {\tt stats.football.co.uk}}
 \end{table}
@@
 
+Equation~\ref{mostgeneral} is very general and may be applied to a
+range of situations in which standard Bradley-Terry strengths are
+modified in some way.  In the following section, I apply it to
+Plackett-Luce likelihoods~\citep{plackett1975} likelihood function for
+rank data.
+
 \section{Non-independence in successive arrivals}
@@
 competitor $a$ wins, $b$ comes second, and so on.  Consider the
-following Plackett-Luce likelihood function for this observation:
+following Plackett-Luce~\citep{plackett1975} likelihood function for
+this observation:
 

### 2025-03-19 inst/fatigue.Rmd
need magrittr package
soft link for hyper2.bib

@@
 output: bookdown::html_document2
-bibliography: hyper2.bib  # copy vignettes/hyper2.bib to inst/
+bibliography: hyper2.bib
 ---

### 2025-03-19 inst/retain.Rmd
bugfixes
soft link for hyper2.bib

@@
 
-```{r}
-jj <- skating_table[1:3,1:6]
+```{r subsetofskating}
+jj <- unclass(skating_table)[1:3,1:6]
 rownames(jj) <- letters[1:3]
@@
 
-```{r}
+```{r convexp}
 a <- apply(jj,2,rank)
@@
 
-```{r}
+```{r coerceto}
 OT <- ordertable2supp(a)
@@
 
-```{r}
+```{r keepflaw}
 keep_flawed(OT, c("a","b"))
@@
 
-```{r}
+```{r keeponlya}
 keep_flawed(OT,"a")

### 2025-03-19 inst/salad.Rmd
soft link for hyper2.bib

@@
 author: "R. K. S. Hankin"
-bibliography: hyper2.bib  # copy vignettes/hyper2.bib to inst/
+bibliography: hyper2.bib
 ---
@@
 for(i in seq_len(nrow(salad))){
-   H1 <-  H1 + ordervec2supp(as.matrix(salad)[i,])
+   H1 <-  H1 + suppfun(as.matrix(salad)[i,])
 }

### 2025-03-19 inst/skeleton.Rmd
bowdlerise
soft link for hyper2.bib

@@
 ```
+
+(takes about four hours to run without cache)
 
@@
 
-```{r,label=datamatrixfucksthingsup}
+```{r,label=datamatrixbuggersthingsup, error=TRUE}
 head(data.matrix(skeleton_table))

### 2025-03-19 inst/table_tennis_serve.Rmd
dirichlet3()
soft link for hyper2.bib

@@
 ```{r, label=newdatacd,cache=TRUE}
-(Hnew <- home_away3(M_home,M_away,lambda=3.35) + pair3(c=0,d=1,lambda=3.35))
+(Hnew <- home_away3(M_home,M_away,lambda=3.35) + dirichlet3(c(c=0,d=1),lambda=3.35))
 maxp(Hnew)

### 2025-03-19 inst/surfing.Rmd
surfing citations

@@
 
-@farley2013 analyse surfing results using one-way ANOVA techniques and
+@farley2015 analyse surfing results using one-way ANOVA techniques and
 report comparisons between the top 10 and bottom 10 surfers in the

### 2025-03-20 inst/jss_hyper3_real.Rnw
typo
JSS house style

@@
-% -*- mode: noweb; noweb-default-code-mode: R-mode; -*-
 \documentclass[article]{jss}
+
+%% -- LaTeX packages and custom commands ---------------------------------------
+
+%% recommended packages
+\usepackage{orcidlink,thumbpdf,lmodern}
+
+%% another package (only for this demo article)
+\usepackage{framed}
 \usepackage{booktabs}
@@
 
-%  Process with:
-%  
-%  Rscript -e 'library("knitr") ; knit("jss_hyper3_real.Rnw")' && pdflatex jss_hyper3_real && open jss_hyper3_real.pdf
-%
-%  (also the Makefile has an entry for this)
-
-\author{Robin K. S. Hankin\\University of Stirling}
+%% new custom commands
+\newcommand{\class}[1]{`\code{#1}'}
+\newcommand{\fct}[1]{\code{#1()}}
+\newcommand\headercell[1]{%
+   \smash[b]{\begin{tabular}[t]{@{}c@{}} #1 \end{tabular}}}
+
+%% For Sweave-based articles about R packages:
+%% need no \usepackage{Sweave}
+\SweaveOpts{engine=R, eps=FALSE, keep.source = TRUE}
+<<preliminaries, echo=FALSE, results=hide>>=
+options(prompt = "R> ", continue = "+  ", width = 70, useFancyQuotes = FALSE)
+library("hyper2")
+library("magrittr")
+howmany <- 23  # set to a smaller value to save time
+@
+
+%% -- Article metainformation (author, title, ...) -----------------------------
+
+%% - \author{} with primary affiliation (and optionally ORCID link)
+%% - \Plainauthor{} without affiliations
+%% - Separate authors by \And or \AND (in \author) or by comma (in \Plainauthor).
+%% - \AND starts a new line, \And does not.
+\author{Robin K. S. Hankin~\orcidlink{0000-0001-5982-0415}\\University of Stirling}
+\Plainauthor{Robin K. S Hankin}
+
+%% - \title{} in title case
+%% - \Plaintitle{} without LaTeX markup (if any)
+%% - \Shorttitle{} with LaTeX markup (if any), used as running title
+
 \title{Generalized Bradley-Terry models}
-\Plainauthor{Robin K. S. Hankin} %% comma-separated
-\Plaintitle{Generalized Bradley-Terry models}
-
-%% an abstract and keywords
+
+%% - \Abstract{} almost as usual
 \Abstract{
-
   A generalization of the Bradley-Terry model is presented in which
@@
   bus, blue bus'' problem in preference analysis.  I present examples
-  drawn from the fields of food preference and the e-sport {\em
-    Counterstrike}.  }
-
-\Keywords{Plackett-Luce, Bradley-Terry, Mann-Whitney}
-\Plainkeywords{Plackett-Luce, Bradley-Terry, Mann-Whitney}
-
-%% publication information
-%% NOTE: Typically, this can be left commented and will be filled out by the technical editor
-%% \Volume{50}
-%% \Issue{9}
-%% \Month{June}
-%% \Year{2012}
-%% \Submitdate{2012-06-04}
-%% \Acceptdate{2012-06-04}
-
-%% The address of (at least) one author should be given
-%% in the following format:
+  drawn from the fields of food preference and the esport {\em
+    Counterstrike}.
+}
+
+%% - \Keywords{} with LaTeX markup, at least one required
+%% - \Plainkeywords{} without LaTeX markup (if necessary)
+%% - Should be comma-separated and in sentence case.
+\Keywords{JSS, style guide, comma-separated, not capitalized, \proglang{R}}
+\Plainkeywords{JSS, style guide, comma-separated, not capitalized, R}
+\Keywords{Plackett-Luce, Bradley-Terry, \proglang{R}}
+\Plainkeywords{Plackett-Luce, Bradley-Terry, R}
+
+%% - \Address{} of at least one author
+%% - May contain multiple affiliations for each author
+%%   (in extra lines, separated by \emph{and}\\).
+%% - May contain multiple authors for the same affiliation
+%%   (in the same first line, separated by comma).
 \Address{
@@
 
-%% for those who use Sweave please include the following line (with % symbols):
-%% need no \usepackage{Sweave.sty}
-
-%% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\newcommand\headercell[1]{%
-   \smash[b]{\begin{tabular}[t]{@{}c@{}} #1 \end{tabular}}}
-\newcommand{\draw}{\ensuremath{\mathfrak{D}}}
-\newcommand{\cdraw}{\ensuremath{\mathfrak{C}}}
-\newcommand{\white}{\ensuremath{\mathfrak{W}}}
-\newcommand{\home}{\ensuremath{\mathfrak{H}}}
-\newcommand{\oo}{\vphantom{W^{W^{W^W}}}}
-
 \begin{document}
 
-<<setup, cache=FALSE,echo=FALSE>>=
-source("loadlibrary.R",echo=FALSE)  # to avoid stupid messages
-opts_chunk$set(cache=TRUE, autodep=TRUE)
-options(prompt = "R> ")
-@
+%% -- Introduction -------------------------------------------------------------
+
+%% - In principle "as usual".
+%% - But should typically have some discussion of both _software_ and _methods_.
+%% - Use \proglang{}, \pkg{}, and \code{} markup throughout the manuscript.
+%% - If such markup is in (sub)section titles, a plain text version has to be
+%%   added as well.
+%% - All software mentioned should be properly \cite-d.
+%% - All abbreviations should be introduced.
+%% - Unless the expansions of abbreviations are proper names (like "Journal
+%%   of Statistical Software" above) they should be in sentence case (like
+%%   "generalized linear models" below).
 
 \section{Introduction}
-
 \begin{wrapfigure}{r}{0.1\textwidth}
@@
 might have likelihood function
-
+%
 \begin{equation}\label{like1234}
@@
 \end{equation}
-
+%
 where $w_{12}$ and $w_{34}$ are the number of wins by each team
@@
 by the drug; values $>1$ would correspond to a positive effect.  In
-this paper I generalize the {\tt hyper2} R
+this paper I generalize the \pkg{hyper2} R
 package~\citep{hankin2017_rnw}, available on CRAN~\citep{rcore2025},
 for Bradley-Terry likelihoods to accommodate this class of
-multiplicative factors with new \proglang{S3} class {\tt hyper3} and
+multiplicative factors with new \proglang{S3} \class{hyper3} and
 associated methods.
@@
 proportional to
-
+%
 \begin{equation}\label{mostgeneral}
@@
 \end{equation}
-
+%
 where the $p_j$ are traditional Bradley-Terry strengths, $p_j\geqslant
@@
 likelihood proportional to
-
-\begin{table}
+%
+\begin{table}[t!]
 \centering
@@
 \end{table}
-
+%
 \begin{equation}\label{firstfewRCFL}
@@
 \end{equation}
-
+%
 where the home-ground advantage is represented by~$\lambda$.  In
@@
 integers, and here they are, in general, unknown non-negative real
-numbers.
+numbers.  The \class{hyper3} class enforces \pkg{disordR}
+discipline~\citep{hankin2022_disordR}.
 
@@
 straightforward.  Considering first only Rangers-Celtic matches:
-
+%
 <<defineoldfirm, cache=TRUE>>=
@@
 @
-
+%
 Above we see package idiom for creating a log-likelihood function.  We
@@
 bespoke log-likelihood function for Rangers-Celtic matches:
-
+%
 <<useoldfirm, cache = TRUE>>=
@@
 @
-
+%
 We see a log-likelihood function for all $82+42+76+46=245$
@@
 value of $\lambda$ using profile support:
-
+%
 <<profilelikeoldfirm, cache=TRUE>>=
 f <- function(lambda){maxp(oldfirm(lambda),give=1)$`Log-likelihood`}
-lam <- seq(from=0.9, to=3, len=23)
+lam <- seq(from=0.9, to=3, len=howmany)
 logLike <- sapply(lam,f)
 @
-
-Thus {\tt f()} returns the support at the evaluate, conditional on a
+%
+Thus \fct{f} returns the support at the evaluate, conditional on a
 supplied value of $\lambda$.  We may use this to create a profile
@@
 Rangers and Celtic:
-
+%
 <<mleoldfirm,cache=TRUE>>=
@@
 @
-
+%
 Compare with the theoretical value of $\widehat{p_{\mbox{\tiny
@@
 
-\begin{table}
+\begin{table}[t!]
 \centering
@@
 \end{table}
-
+%
 We are now in a position to consider the entirety of
@@
 advantage $\lambda$ and proclivity to draw $D$:
-
+%
 <<eg_RCLF3_lf, cache=TRUE>>=
@@
 @
-
+%
 Observe that the resulting likelihood function has two distinct types
@@
 the likelihood over possible values of $\mathbf{p}$.
-
+%
 <<defineF>>=
@@
 @
-
+%
 Thus \code{f()} gives us a likelihood function for
@@
 \code{optim()}:
-
+%
 <<findmaxlambdaD, cache=TRUE>>=
@@
 @
-
+%
 Then
-
+%
 <<showevaluate, cache=TRUE>>=
@@
 @
-
+%
 Such estimates have rich predictive power.  For example, if Rangers
@@
 estimated at $(0.39;0.27;0.34)$.
-
+%
 <<makecontourdata, echo=FALSE, cache=TRUE>>=
-n <- 20
+n <- howmany
 lambda <- seq(from=1, to=3, len=n)
@@
 @
-
+%
 Further, we may calculate a support function over the
@@
   the left hand side of figure~\ref{contourlambdaD}] for which the
-support is about \Sexpr{round(f(maxv) - jj$objective,2)}, far
+support is about \Sexpr{round(f(maxv)-jj$objective,2)}, far
 exceeding the two units of support per degree of freedom
@@
 lies in the tail region of its asymptotic null $\chi^2_2$ distribution
-with a $p$-value of about \Sexpr{signif(pchisq(f(maxv) -
-  jj$objective,lower.tail=FALSE,df=2),digits=2)}.
+with a $p$-value of about \Sexpr{signif(pchisq(f(maxv)-jj$objective,lower.tail=FALSE,df=2),digits=2)}.
 
@@
 this observation:
-
+%
 \begin{equation}
@@
 \end{equation}
-
+%
 Note the use of multiplication between the terms, indicating
@@
 
-\begin{figure}[htbp]
+\begin{figure}[t!]
   \begin{center}
@@
 \end{equation}
-
+%
 In equation \ref{asuccb}, the first term is standard Plackett-Luce: at
@@
 c\succ b\succ e$ would be
-
+%
 \begin{equation}
@@
 \end{equation}
-
+%
 where this likelihood function reflects the mutual support for
@@
 options.
-
+%
 <<defineredbuslf>>=
@@
 @
-
+%
 Above, we see from the function body that he reported $RB\succ BB\succ
@@
 between RB and BB in the same way as function \code{oldfirm()} above.
-
+%
 <<definefunctiono>>=
@@
 @ 
-
+%
 Above, we define function \code{o()} which returns the support for a
@@
 support function for $\lambda$, and then maximize it:
-
+%
 <<label=maxlikesim, cache=TRUE>>=
@@
 @ 
-
+%
 So a likelihood ratio test of the null that $\lambda=1$ would be:
-
+%
 <<label=LRT_RB, cache=TRUE>>=
@@
 @
-
+%
 We can thus improve the hypothesis that $\lambda=1$, corresponding to
@@
 region of its asymptotic $\chi^2_1$ distribution (Wilks's theorem),
-with a $p$-value of \Sexpr{signif(pchisq(suppdiff*2, df=1,
-  lower.tail=FALSE),3)}.
-
+with a $p$-value of \Sexpr{signif(pchisq(suppdiff*2,df=1,lower.tail=FALSE),3)}.
+%
 %It is straightforward to find a profile
@@
 %range.
-
+%
 %\begin{figure}[htbp]
@@
 \code{hyper3} formalism is straightforward.  The dataset used is
-\code{sushi\_table}, part of the \code{hyper2} package:
-
+\code{sushi\_table}, part of the \pkg{hyper2} package:
+%
 <<showsushitable>>=
@@
 @
-
+%
 Thus, from the top line, we see that the first judge placed tamago
@@
 by
-
+%
 <<show_sushi_eq_classes>>=
@@
 @
-
+%
 so class 2 is the tuna group and class 1 everything else.  We can now
@@
 $\lambda$ as the argument:
-
+%
 <<defcalculate_sushi_H>>=
@@
 @
-
+%
 <<calclikesushi,echo=FALSE, cache=TRUE>>=
@@
 }
-s <- exp(seq(from = log(0.9), to = log(10), len = 23))
+s <- exp(seq(from = log(0.9), to = log(10), len = howmany))
 L <- sapply(s,f)
 @ 
-
+%
 % chunk 'findstuffsushi' takes about 10-20 minutes to run
-
+%
 <<findstuffsushi,echo=FALSE, cache=TRUE>>=
@@
 @ 
-
+%
 % chunk tinysushi' takes about 5 minutes to run
-
+%
 <<tinysushi, echo=FALSE, cache=TRUE>>=
@@
 @
-
+%
 % chunk 'findotherstuffsushi' takes about 10 minutes to run
-
+%
 <<findothersushistuff, echo=FALSE, cache=TRUE>>=
@@
 @ 
-
+%
 (above, we note that the return value {\tt H} is the sum of successive
@@
 five {\em Cloud9} players.
-
+%
 <<echo=FALSE>>=
@@
 @
-
+%
 From the first line we see that {\em Cloud9} played {\em FaZe Clan} on
@@
 situation, but with $\lambda<1$.
-
+%
 <<defredbusfunc, echo=FALSE>>=
@@
 @
-
+%
 <<csgo_findopt, cache=TRUE, echo=FALSE>>=
@@
 @
-
+%
 <<csgo_calcklikelambda, cache=TRUE, echo=FALSE>>=
-lam <- exp(seq(from = -4, to = log(1.1), len = 23))
+lam <- exp(seq(from = -4, to = log(1.1), len = howmany))
 L <- sapply(lam, f)
 @
-
+%
 <<cgso_findinterval, cache=TRUE, echo=FALSE>>=
 g <- function(l){f(l) - csgo_optimize$objective  + 2}
-csgo_lower <- uniroot(g, interval = exp(c(-4,-3)))$root
-csgo_upper <- uniroot(g, interval = exp(c(-1,-0)))$root
+csgo_lower <- uniroot(g, interval = exp(c(-4,-3)), tol = 0.01)$root
+csgo_upper <- uniroot(g, interval = exp(c(-1,-0)), tol = 0.01)$root
 @
@@
 factors that represent temporary or ephemeral changes in competitive
-strength.  The {\tt hyper2} R package implements the resulting
+strength.  The \pkg{hyper2} R package implements the resulting
 likelihood functions using \proglang{S3} class {\tt hyper3}, and

### 2025-03-20 inst/jss_hyper3_real.Rnw
more JSS house style

@@
 Table~\ref{oldfirm}.  Bespoke package function
-\code{home\_draw\_away3()} converts data in the form of
+\fct{home\_draw\_away3} converts data in the form of
 Table~\ref{oldfirm} to a \code{hyper3} likelihood function following
@@
 %
-Thus \code{f()} gives us a likelihood function for
+Thus \fct{f} gives us a likelihood function for
 $\mathbf{v}=(\lambda,D)$.  We may find the evaluate using
-\code{optim()}:
+\fct{optim}:
 %
@@
 cheering his still-running teammates, boosting their performance.  The
-\pkg{hyper2} package includes function \code{cheering()} which
+\pkg{hyper2} package includes function \fct{cheering} which
 implements this functionality.
@@
 indistinguishable).  Can we quantify the strength of this effect?  To
-do this, we define a bespoke function \code{RB\_BB\_LF()} which
+do this, we define a bespoke function \fct{RB\_BB\_LF} which
 returns a \code{hyper3} log-likelihood function corresponding to
@@
 strength of the nonindependence.  Observe that in almost every case he
-ranks RB and BB consecutively.  Function \code{RB_BB_LF()} takes
+ranks RB and BB consecutively.  Function \fct{RB_BB_LF} takes
 argument \code{lambda} that quantifies the perceived similarity
-between RB and BB in the same way as function \code{oldfirm()} above.
+between RB and BB in the same way as function \fct{oldfirm} above.
 %
@@
 %
-Above, we define function \code{o()} which returns the support for a
+Above, we define function \fct{o} which returns the support for a
 given value of $\lambda$ at the evaluate [over permissible Bradley
-  Terry strengths for the options].  We can treat \code{o()} as a
+  Terry strengths for the options].  We can treat \fct{o} as a
 support function for $\lambda$, and then maximize it:

### 2025-03-20 inst/jss_hyper3_real.Rnw
JSS bibliography style [capialization]

@@
 home-ground advantage in football and rugby, or playing white in
-chess~\citep{hankin2020}.  Note that $\lambda < 1$ would correspond to
+chess~\citep{hankin2020_jss}.  Note that $\lambda < 1$ would correspond to
 a decrease in competitive ability.
 
-Further, one might follow~\cite{hankin2017_rnw} and consider teams of
+Further, one might follow~\cite{hankin2017_rnw_jss} and consider teams of
 players.  In doubles tennis, for example, $\left\lbrace
@@
 this paper I generalize the \pkg{hyper2} R
-package~\citep{hankin2017_rnw}, available on CRAN~\citep{rcore2025},
+package~\citep{hankin2017_rnw_jss}, available on CRAN~\citep{rcore2025},
 for Bradley-Terry likelihoods to accommodate this class of
@@
 with an associated power of $-81$.  The computational methodology is a
-generalization of that used in~\cite{hankin2024_hyper3} in which the
+generalization of that used in~\cite{hankin2024_hyper3_jss} in which the
 multiplicative factors correspond to number of clones.  However,
-in~\cite{hankin2024_hyper3} the factors are known non-negative
+in~\cite{hankin2024_hyper3_jss} the factors are known non-negative
 integers, and here they are, in general, unknown non-negative real
 numbers.  The \class{hyper3} class enforces \pkg{disordR}
-discipline~\citep{hankin2022_disordR}.
+discipline~\citep{hankin2022_disordR_jss}.
 
@@
 number of desirable features not present in the reified entity
-approach of~\citet{hankin2020}.  Firstly, if we consider two players
+approach of~\citet{hankin2020_jss}.  Firstly, if we consider two players
 of identical strengths and nullify the home ground effect (that is,

### 2025-03-20 inst/jss_hyper3_real.Rnw
more JSS house style bib changes (capitalization)

@@
 One generalization of the Bradley-Terry model is due
-to~\citet{luce1959}, in which the probability of competitor~$i$
+to~\citet{luce1959_jss}, in which the probability of competitor~$i$
 winning in a field of~$j=1,\ldots, n$ is $\frac{p_i}{p_1+\cdots
@@
 modified in some way.  In the following section, I apply it to
-Plackett-Luce likelihoods~\citep{plackett1975} likelihood function for
+Plackett-Luce likelihoods~\citep{plackett1975_jss} likelihood function for
 rank data.
@@
 competitor $a$ wins, $b$ comes second, and so on.  Consider the
-following Plackett-Luce~\citep{plackett1975} likelihood function for
+following Plackett-Luce~\citep{plackett1975_jss} likelihood function for
 this observation:
@@
 \cite{mcfadden1980} although different metaphors have been employed
-for the same phenomenon by \cite{tverskey1972}.  The canonical problem
+for the same phenomenon by \cite{tverskey1972_jss}.  The canonical problem
 is as follows.  A commuter has a choice of going to work by different

### 2025-03-20 inst/jss_hyper3_real.Rnw
broader citation of previous work

@@
 The Bradley-Terry model for datasets involving paired comparisons has
-wide uptake in the {R} community.  The canonical case is to
-consider~$n$ players who compete against one another in pairs; the
-basic inference problem is to estimate
+wide uptake in the {R} community [see, for example,
+\citet{turner2012,turner2020} and references therein].  The canonical
+case is to consider~$n$ players who compete against one another in
+pairs; the basic inference problem is to estimate
 numbers~$\left({p_1,\ldots,p_n}\right)$, $p_i\geqslant 0$, $\sum
@@
 
-About 22\% of matches in Table~\ref{oldfirm} end in a draw and {\tt
-  hyper2} formalism may be used to assess this.  Consider the
-likelihood function described in Table~\ref{windrawloss}; this has a
-number of desirable features not present in the reified entity
-approach of~\citet{hankin2020_jss}.  Firstly, if we consider two players
-of identical strengths and nullify the home ground effect (that is,
-$\lambda=1$) then the probability of a draw is $\frac{2D}{2+2D}$; the
-odds of a draw is simply $D$, regardless of team strength.  Secondly,
-considering only won (that is, not score-drawn) games, the probability
-of a home win is $\frac{\lambda}{1+\lambda}$, that is, the odds of a
-home win is simply $\lambda$, independently of the player
-strength. Such arguments give direct operational significance to
-$\lambda$ and $D$.
+About 22\% of matches in Table~\ref{oldfirm} end in a draw and
+\pkg{hyper2} formalism may be used to assess this.  Consider the
+likelihood function described in Table~\ref{windrawloss}; this is not
+easily represented as a generalized linear model using the techniques
+of~\citet{turner2020}.  Relatedlikelihood functions for drawn matches
+were considered using the reified entity technique
+of~\citet{hankin2020_jss} in which a drawn match was considered to be
+won by a fictitious entity (a ``draw monster'') who possessed his own
+BT strength.  However, the likelihood function of
+table~\ref{windrawloss} has a number of desirable featureas not
+present in the reified entity approach of~\citet{hankin2020_jss}.
+Firstly, if we consider two players of identical strengths and nullify
+the home ground effect (that is, $\lambda=1$) then the probability of
+a draw is $\frac{2D}{2+2D}$; the odds of a draw is simply $D$,
+regardless of team strength.  Secondly, considering only won (that is,
+not score-drawn) games, the probability of a home win is
+$\frac{\lambda}{1+\lambda}$, that is, the odds of a home win is simply
+$\lambda$, independently of the player strength. Such arguments give
+direct operational significance to $\lambda$ and $D$.
 
@@
 modified in some way.  In the following section, I apply it to
-Plackett-Luce likelihoods~\citep{plackett1975_jss} likelihood function for
-rank data.
+Plackett-Luce \citep{plackett1975_jss} likelihood functions for
+rank data
 
@@
 %
-Note the use of multiplication between the terms, indicating
-conditional independence.  Now we use the multiplicative
-generalization of Bradley-Terry strengths to introduce an element of
-non-independence between the terms.  There are many ways to do this,
-but one simple way would be to define equivalence classes of the
-competitors with each equivalence class comprising mutually supporting
-runners.  The metaphor would be that a runner who has finished the
-race is able to support other members of his equivalence class by
-cheering his still-running teammates, boosting their performance.  The
-\pkg{hyper2} package includes function \fct{cheering} which
-implements this functionality.
+Note the use of multiplication between the terms, indicating the
+standard assumption~\citep{turner2020} of conditional independence.
+Now we use the multiplicative generalization of Bradley-Terry
+strengths to introduce an element of non-independence between the
+terms.  There are many ways to do this, but one simple way would be to
+define equivalence classes of the competitors with each equivalence
+class comprising mutually supporting runners.  The metaphor would be
+that a runner who has finished the race is able to support other
+members of his equivalence class by cheering his still-running
+teammates, boosting their performance.  The \pkg{hyper2} package
+includes function \fct{cheering} which implements this functionality.
 

### 2025-03-20 inst/jss_hyper3_real.Rnw
typo

@@
 strength of the nonindependence.  Observe that in almost every case he
-ranks RB and BB consecutively.  Function \fct{RB_BB_LF} takes
+ranks RB and BB consecutively.  Function \fct{RB\_BB\_LF} takes
 argument \code{lambda} that quantifies the perceived similarity

### 2025-03-21 inst/jss_hyper3_real.Rnw
typo
$p$~value not $p$-value

@@
 lies in the tail region of its asymptotic null $\chi^2_2$ distribution
-with a $p$-value of about \Sexpr{signif(pchisq(f(maxv)-jj$objective,lower.tail=FALSE,df=2),digits=2)}.
+with a $p$~value of about \Sexpr{signif(pchisq(f(maxv)-jj$objective,lower.tail=FALSE,df=2),digits=2)}.
 
@@
 $\frac{1}{3}$.  Various solutions to this problem have been proposed.
-\cite{tverskey1972}, for example, presents an approach involving
+\cite{tverskey1972_jss}, for example, presents an approach involving
 overlapping ``aspects'' of the elements of choice set;
@@
 region of its asymptotic $\chi^2_1$ distribution (Wilks's theorem),
-with a $p$-value of \Sexpr{signif(pchisq(suppdiff*2,df=1,lower.tail=FALSE),3)}.
+with a $p$~value of \Sexpr{signif(pchisq(suppdiff*2,df=1,lower.tail=FALSE),3)}.
 %
@@
 may observe that this lies in the tail region of its asymptotic
-$\chi^2_1$ distribution with a $p$-value of about
+$\chi^2_1$ distribution with a $p$~value of about
 $\Sexpr{round(pchisq(sushisupport,df=1,lower.tail=FALSE),3)}$.  We can
@@
   $\hat{\lambda}\simeq\Sexpr{round(csgo_optlam,3)}$ (small red bar),
-  corresponding to a $p$-value of about
+  corresponding to a $p$~value of about
   $\Sexpr{round(csgo_pvalue,3)}$.  Horizontal lines show the
@@
 observe that this exceeds the two units of support criterion of
-\cite{edwards1992} [asymptotic $p$-value
+\cite{edwards1992} [asymptotic $p$~value
   $\simeq\Sexpr{round(csgo_pvalue,3)}$] so we may infer that there is

### 2025-03-21 inst/jss_hyper3_real.Rnw
Equation not equation

@@
 ``wins'' and number of matches, as per $w_{12},w_{34}$ in
-equation~\ref{like1234}.  In practice, almost all the $\lambda_{ij}$
+Equation~\ref{like1234}.  In practice, almost all the $\lambda_{ij}$
 are zero.  To fix ideas, consider the data presented in
@@
 assumed value $\lambda=1.88$.  This corresponds to the relevant terms
-of equation~\ref{firstfewRCFL}.  We may make inferences about the true
+of Equation~\ref{firstfewRCFL}.  We may make inferences about the true
 value of $\lambda$ using profile support:
@@
 %
-In equation \ref{asuccb}, the first term is standard Plackett-Luce: at
+In Equation \ref{asuccb}, the first term is standard Plackett-Luce: at
 this point, no competitor has finished and cheering effects are

### 2025-03-21 inst/jss_hyper3_real.Rnw
jss house style whitespace

@@
 %
-<<defineoldfirm, cache=TRUE>>=
+<<defineoldfirm, cache = TRUE>>=
 oldfirm <- function(lambda){
@@
 <<profilelikeoldfirm, cache=TRUE>>=
-f <- function(lambda){maxp(oldfirm(lambda),give=1)$`Log-likelihood`}
-lam <- seq(from=0.9, to=3, len=howmany)
-logLike <- sapply(lam,f)
+f <- function(lambda){maxp(oldfirm(lambda), give = 1)$`Log-likelihood`}
+lam <- seq(from = 0.9, to = 3, len = howmany)
+logLike <- sapply(lam, f)
 @
@@
 <<plotproflikeoldfirm, echo=FALSE>>=
-plot(lam,logLike - max(logLike),type='b')
-abline(v=1,col='gray')
-segments(x0=1.785,y0=0.1,y1=-1,col='red')
-abline(h = c(0,-2),col='gray')
+plot(lam, logLike - max(logLike), type = 'b')
+abline(v = 1, col = 'gray')
+segments(x0 = 1.785, y0 = 0.1, y1 = -1, col = 'red')
+abline(h = c(0, -2), col = 'gray')
 @
@@
 <<optimoldfirm, cache=TRUE>>=
-(lambda_max <-  optimize(f,c(1.5,2.0),maximum=TRUE)$maximum)
+(lambda_max <-  optimize(f, c(1.5, 2.0), maximum = TRUE)$maximum)
 @
@@
 %
-<<mleoldfirm,cache=TRUE>>=
+<<mleoldfirm, cache=TRUE>>=
 maxp(oldfirm(lambda_max))
@@
 %
-<<eg_RCLF3_lf, cache=TRUE>>=
-home_draw_away3(RCLF3_table, lambda=1.7, D=0.3)
+<<eg_RCLF3_lf, cache = TRUE>>=
+home_draw_away3(RCLF3_table, lambda = 1.7, D = 0.3)
 @
@@
 f <- function(v){
-    H <- home_draw_away3(RCLF3_table, lambda = v[1], D=v[2])
-    maxp(H,give=1)$`Log-likelihood`
+    H <- home_draw_away3(RCLF3_table, lambda = v[1], D = v[2])
+    maxp(H, give = 1)$`Log-likelihood`
 }
@@
 %
-<<findmaxlambdaD, cache=TRUE>>=
-maxv <- optim(par=c(2, 0.3), fn=f, control=list(fnscale = -1))$par
+<<findmaxlambdaD, cache = TRUE>>=
+maxv <- optim(par=c(2, 0.3), fn = f, control = list(fnscale = -1))$par
 maxv
@@
 %
-<<showevaluate, cache=TRUE>>=
-maxp(home_draw_away3(RCLF3_table, lambda = maxv[1], D=maxv[2]))
+<<showevaluate, cache = TRUE>>=
+maxp(home_draw_away3(RCLF3_table, lambda = maxv[1], D = maxv[2]))
 @
@@
 %
-<<makecontourdata, echo=FALSE, cache=TRUE>>=
+<<makecontourdata, echo = FALSE, cache = TRUE>>=
 n <- howmany
-lambda <- seq(from=1, to=3, len=n)
-D <- seq(from=0.1, to=0.6, len=n)
+lambda <- seq(from = 1, to = 3, len = n)
+D <- seq(from = 0.1, to = 0.6, len = n)
 V <- as.matrix(expand.grid(lambda, D))
@@
   \begin{center}
-<<docontourplot, cache=FALSE>>=
+<<docontourplot, cache = FALSE>>=
 contour(lambda, D, matrix(LL, n, n), levels = -2*(1:9),
@@
 
-<<testlambdazero, cache=TRUE, include=FALSE>>=
-jj <- optimize(f=function(D){f(c(lambda=1,D))}, interval=c(0.1, 0.4), maximum=TRUE)
+<<testlambdazero, cache = TRUE, include = FALSE>>=
+jj <- optimize(f = function(D){f(c(lambda=1, D))}, interval = c(0.1, 0.4), maximum = TRUE)
 @
@@
 `RB_BB_LF` <- function(lambda){
-    ec <- c(C=1, T=2, RB=3, BB=3, W=4) 
+    ec <- c(C = 1, T = 2, RB = 3, BB = 3, W = 4) 
     h <- c(1, 1, lambda, 1)           
@@
 <<definefunctiono>>=
-o <- function(lambda){maxp(RB_BB_LF(lambda), give=1)$`Log-likelihood`} 
+o <- function(lambda){maxp(RB_BB_LF(lambda), give = 1)$`Log-likelihood`} 
 @ 
@@
 %
-<<label=maxlikesim, cache=TRUE>>=
-(osup <- optimize(o, c(10, 40), maximum=TRUE))
+<<label = maxlikesim, cache = TRUE>>=
+(osup <- optimize(o, c(10, 40), maximum = TRUE))
 @ 
@@
 %
-<<label=LRT_RB, cache=TRUE>>=
+<<label = LRT_RB, cache = TRUE>>=
 (suppdiff <- osup$objective - o(lambda = 1))
@@
 %\begin{center}
-<<plotproflike, echo=FALSE>>=
+<<plotproflike, echo = FALSE>>=
 if(FALSE){
-    lambda <- exp(seq(from=log(0.9), to=log(67), len=19)) 
+    lambda <- exp(seq(from = log(0.9), to = log(67), len = 19)) 
     L <- sapply(lambda, o)
-    plot(log(lambda), L-max(L), type="b")
-    abline(h=c(0, -2))
-    abline(v=0)
+    plot(log(lambda), L - max(L), type = "b")
+    abline(h = c(0, -2))
+    abline(v = 0)
 }
@@
     for(i in seq_len(nrow(sushi_table))){
-        H <- H + cheering3(sushi_table[i,], e=sushi_eq_classes, h=c(1, lambda))
+        H <- H + cheering3(sushi_table[i,], e = sushi_eq_classes, h = c(1, lambda))
     }
@@
 %
-<<calclikesushi,echo=FALSE, cache=TRUE>>=
+<<calclikesushi, echo = FALSE, cache = TRUE>>=
 f <- function(lambda){
    H <- make_sushi_H(lambda)
-   maxp(H,give=1)$`Log-likelihood`
+   maxp(H, give = 1)$`Log-likelihood`
 }
 s <- exp(seq(from = log(0.9), to = log(10), len = howmany))
-L <- sapply(s,f)
+L <- sapply(s, f)
 @ 
@@
 %
-<<findstuffsushi,echo=FALSE, cache=TRUE>>=
-osushi <- optimize(f, c(2,4), maximum=TRUE)
+<<findstuffsushi, echo = FALSE, cache = TRUE>>=
+osushi <- optimize(f, c(2, 4), maximum = TRUE)
 @ 
@@
 %
-<<tinysushi, echo=FALSE, cache=TRUE>>=
+<<tinysushi, echo = FALSE, cache = TRUE>>=
 sushimaxlambda <- osushi$maximum
@@
 %
-<<findothersushistuff, echo=FALSE, cache=TRUE>>=
+<<findothersushistuff, echo = FALSE, cache = TRUE>>=
 g <- function(lambda){f(lambda) - osushi$objective + 2}
-sushi_lower <- uniroot(g, interval = exp(c(0,1)), tol = 0.01)$root
-sushi_upper <- uniroot(g, interval = exp(c(1,2)), tol = 0.01)$root
+sushi_lower <- uniroot(g, interval = exp(c(0, 1)), tol = 0.01)$root
+sushi_upper <- uniroot(g, interval = exp(c(1, 2)), tol = 0.01)$root
 @ 
@@
   \begin{center}
-<<plotlikesushi, echo=FALSE>>=
-plot(log(s),L-max(L),type='b',
-     xlab=expression(log(lambda)), ylab='Log-likelihood')
-abline(v=0,col='gray')
-abline(h=c(0,-2),col='gray')
+<<plotlikesushi, echo = FALSE>>=
+plot(log(s), L - max(L), type = 'b',
+     xlab = expression(log(lambda)), ylab = 'Log-likelihood')
+abline(v = 0, col = 'gray')
+abline(h = c(0, -2), col = 'gray')
 segments(x0 = log(sushimaxlambda), y0 =  0.1, y1 = -1.0, col = 'red' )
@@
 %
-<<echo=FALSE>>=
-a <- read.table("cloud9_boston_2018.txt", header=TRUE)
-a[,-1]
+<<echo = FALSE>>=
+a <- read.table("cloud9_boston_2018.txt", header = TRUE)
+a[, -1]
 @
@@
 %
-<<defredbusfunc, echo=FALSE>>=
+<<defredbusfunc, echo = FALSE>>=
 mk <- function(lam){
-    e <- c(Skadoodle=2, RUSH=1, tarik=1, autimatic=1, Stewie2K=2)
+    e <- c(Skadoodle = 2, RUSH = 1, tarik = 1, autimatic = 1, Stewie2K = 2)
     H3 <- hyper3()
@@
         d <- strsplit(a$deathorder[i],", ")[[1]]
-        H3 <- H3 + cheering3(d, e=e, help=c(1, lam))
+        H3 <- H3 + cheering3(d, e = e, help = c(1, lam))
     }
@@
 %
-<<csgo_findopt, cache=TRUE, echo=FALSE>>=
-csgo_optimize <- optimize(f,c(0.09, 0.14), maximum = TRUE, tol = 0.01)
+<<csgo_findopt, cache = TRUE, echo = FALSE>>=
+csgo_optimize <- optimize(f, c(0.09, 0.14), maximum = TRUE, tol = 0.01)
 csgo_optlam   <- csgo_optimize$maximum
@@
 g <- function(l){f(l) - csgo_optimize$objective  + 2}
-csgo_lower <- uniroot(g, interval = exp(c(-4,-3)), tol = 0.01)$root
-csgo_upper <- uniroot(g, interval = exp(c(-1,-0)), tol = 0.01)$root
+csgo_lower <- uniroot(g, interval = exp(c(-4, -3)), tol = 0.01)$root
+csgo_upper <- uniroot(g, interval = exp(c(-1, -0)), tol = 0.01)$root
 @
@@
 plot(log(lam), L - max(L), type='b')
-segments(x0 = log(csgo_optlam),y0 = -0.5, y1 =  0.1, col='red')
+segments(x0 = log(csgo_optlam),y0 = -0.5, y1 =  0.1, col = 'red')
 segments(x0 = log(csgo_lower), y0 = -2.2, y1 = -1.8, col = 'blue')

### 2025-03-31 inst/jss_hyper3_real.Rnw
typos

@@
 BT strength.  However, the likelihood function of
-table~\ref{windrawloss} has a number of desirable featureas not
+table~\ref{windrawloss} has a number of desirable features not
 present in the reified entity approach of~\citet{hankin2020_jss}.
@@
 Plackett-Luce \citep{plackett1975_jss} likelihood functions for
-rank data
+rank data.
 

### 2025-03-31 inst/jss_hyper3_real.Rnw
comment out unused chunk

@@
 %
-<<profilelikeoldfirm, cache=TRUE>>=
+<<profilelikeoldfirm, cache = TRUE>>=
 f <- function(lambda){maxp(oldfirm(lambda), give = 1)$`Log-likelihood`}
@@
 %\begin{center}
-<<plotproflike, echo = FALSE>>=
-if(FALSE){
-    lambda <- exp(seq(from = log(0.9), to = log(67), len = 19)) 
-    L <- sapply(lambda, o)
-    plot(log(lambda), L - max(L), type = "b")
-    abline(h = c(0, -2))
-    abline(v = 0)
-}
-@ 
+%<<plotproflike, echo = FALSE>>=
+%if(FALSE){
+%    lambda <- exp(seq(from = log(0.9), to = log(67), len = 19)) 
+%    L <- sapply(lambda, o)
+%    plot(log(lambda), L - max(L), type = "b")
+%    abline(h = c(0, -2))
+%    abline(v = 0)
+%}
+%@ 
 %\caption{A\label{profilesupportlambdaRBBB} profile support function

### 2025-04-01 inst/jss_hyper3_real.Rnw
fig = TRUE needed

@@
   \begin{center}
-<<plotproflikeoldfirm, echo=FALSE>>=
+<<plotproflikeoldfirm, echo=FALSE, fig = TRUE>>=
 plot(lam, logLike - max(logLike), type = 'b')
@@
   \begin{center}
-<<docontourplot, cache = FALSE>>=
+<<docontourplot, cache = FALSE, fig = TRUE>>=
 contour(lambda, D, matrix(LL, n, n), levels = -2*(1:9),

### 2025-04-01 inst/hankin-AustrianJournalofStatistics.tex
issue #239: easily-fixed referee comments

@@
 \frac{b}{a+(n-3)b}\cdots\frac{a}{a+(n-r-1)b}\nonumber\\
-&=& \frac{B-1}{(B+n-1)(B+n-2)\cdots(B+n-r-1)}\nonumber\\ 
+&=& \frac{B-1}{(B+n-2)\cdots(B+n-r-1)}\nonumber\\ 
 &=& \frac{(B-1)(B+n-r-2)!}{(B+n-1)!}\label{likeforrn3}
@@
   & \multicolumn{12}{c}{2024}\\
-rank   &  193& 240& 197& 257& 227& 234& 238& 238& 196& 242& 242& 318\\
-runners&  145& 157& 156& 191& 172& 174& 173& 165& 172& 199& 181& 229\\
+rank      &  145& 157& 156& 191& 172& 174& 173& 165& 172& 199& 181& 229\\
+runners   &  193& 240& 197& 257& 227& 234& 238& 238& 196& 242& 242& 318\\
 \end{tabular}
@@
 along the null diagonal is about $-2.06$, at
-${p}_\mathrm{pure}={p}_\mathrm{applied}\simeq 0.683$.  The two units of
-support criterion~\citep{edwards1992} is met, and we are justified in
-concluding that the student is indeed more competent at applied than
-pure mathematics.  Alternatively, we may observe that
-$2\mathcal{S}=4.11$ is in the tail region of its asymptotic $\chi^2_1$
-distribution, corresponding to a $p$-value of about $0.042$.
+${p}_\mathrm{pure}={p}_\mathrm{applied}\simeq 0.683$.  The two units
+of support criterion~\citep{edwards1992} is met, and we are justified
+in concluding that the student indeed has a higher ranking in the
+applied modules than the pure modules.  Alternatively, we may observe
+that $2\mathcal{S}=4.11$ is in the tail region of its asymptotic
+$\chi^2_1$ distribution, corresponding to a $p$-value of about
+$0.042$.
 

### 2025-04-01 inst/jss_hyper3_real.Rnw
tuna types are maguro,toro,tekka

@@
 named numeric vector.  For example, if the home strength were $\lambda
-= 1.8$ then the first bracket would be {\tt c(Rangers=1.8, Celtic=1)}
+= 1.8$ then one bracket would be {\tt c(Rangers=1.8, Celtic=1)}
 with an associated power of $-81$.  The computational methodology is a
@@
 easily represented as a generalized linear model using the techniques
-of~\citet{turner2020}.  Relatedlikelihood functions for drawn matches
+of~\citet{turner2020}.  Related likelihood functions for drawn matches
 were considered using the reified entity technique
@@
 
-<<testlambdazero, cache = TRUE, include = FALSE>>=
+<<testlambdazero, cache = TRUE, eval = TRUE, echo = FALSE>>=
 jj <- optimize(f = function(D){f(c(lambda=1, D))}, interval = c(0.1, 0.4), maximum = TRUE)
@@
   \begin{center}
-<<plotlikesushi, echo = FALSE>>=
+<<plotlikesushi, echo = FALSE, fig = TRUE>>=
 plot(log(s), L - max(L), type = 'b',

### 2025-04-01 inst/sushi.Rmd
soft link for hyper2.bib
tuna types are maguro,toro,tekka

--- 2025-02-22 inst/sushi.Rmd
@@
 sushi_eq_classes <-
-    c("ebi"  = 1, "anago" = 2, "maguro" = 1, "ika" = 1, "uni" = 1,
+    c("ebi"  = 1, "anago" = 1, "maguro" = 2, "ika" = 1, "uni" = 1,
       "sake" = 1, "tamago" = 1, "toro" = 2, "tekka" = 2,
@@
 
-```{r calc_sushi}
+```{r calc_sushi, cache=TRUE}
 sushi <- make_sushi_H(a, lambda = jj$maximum)
@@
 
-```{r calc_sushimaxp}
+```{r calc_sushimaxp, cache=TRUE}
 (sushi_maxp <- maxp(sushi))

### 2025-04-01 inst/jss_hyper3_real.Rnw
capitalizations

@@
 supplied value of $\lambda$.  We may use this to create a profile
-likelihood (figure~\ref{profilesupportlambda}), or to find the maximum
+likelihood (Figure~\ref{profilesupportlambda}), or to find the maximum
 likelihood estimate for $\lambda$:
@@
 
-<<testlambdazero, cache = TRUE, eval = TRUE, echo = FALSE>>=
+<<testlambdazero, cache = TRUE, eval = TRUE, echo = FALSE, fig = TRUE>>=
 jj <- optimize(f = function(D){f(c(lambda=1, D))}, interval = c(0.1, 0.4), maximum = TRUE)
@@
 Further, we may calculate a support function over the
-$(\lambda,D)$-plane, shown in figure~\ref{contourlambdaD}.  For
+$(\lambda,D)$-plane, shown in Figure~\ref{contourlambdaD}.  For
 example, we may test the hypothesis that $\lambda=1$ [corresponding to
-  the left hand side of figure~\ref{contourlambdaD}] for which the
+  the left hand side of Figure~\ref{contourlambdaD}] for which the
 support is about \Sexpr{round(f(maxv)-jj$objective,2)}, far
@@
 
-As an example, consider figure~\ref{tikzabcde}.  This shows a partial
+As an example, consider Figure~\ref{tikzabcde}.  This shows a partial
 probability tree diagram for some of the $5!=120$ possible order

### 2025-04-02 inst/jss_hyper3_real.Rnw
typos

@@
 
-<<testlambdazero, cache = TRUE, eval = TRUE, echo = FALSE, fig = TRUE>>=
+<<testlambdazero, cache = TRUE, eval = TRUE, echo = FALSE>>=
 jj <- optimize(f = function(D){f(c(lambda=1, D))}, interval = c(0.1, 0.4), maximum = TRUE)
@@
   \begin{center}
-<<csgo_plotlikelam, echo=FALSE>>=
+<<csgo_plotlikelam, echo = FALSE, fig = TRUE>>=
 plot(log(lam), L - max(L), type='b')

### 2025-04-02 inst/exponential_BT.Rmd
kickoff for exponential BT

---
title: "exponential_BT"
author: "robin hankin"
date: "2025-04-02"
output: html_document

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(hyper2)
library(stringr)
set.seed(0)
```

Basic idea is to have $n$ competitors with BT strengths $\beta, \beta
x,\ldots \beta x^{n-1}$ with $x\in[0,1]$.  The unit sum constraint
gives $\beta=\frac{1-x}{1-x^n}$ but this usually cancels out.  One of
the competitors is the _focal_ competitor with BT strength $\beta
x^{r-1}$ for some $r$ with $1\leqslant r < n$.  Thus $r=1$ means that
the focal competitor is the strongest among the $n$, $r=2$ means he is
the second, and so on.

```{r}
BT <- function(n,x){
    out <- x^(0:(n-1))
    if(x != 1){
        out <- out*(1-x)/(1-x^n)
    } else {
        out <- out/n
    }
    names(out) <- paste0("p",str_pad(1:n,ceiling(log10(n))))
    out
}
BT(4,0.5)
sum(BT(4,0.5))
```

We will consider $n=9$, $x=0.5$, $r=4$ and make repeated observations:

```{r}
(jj <- BT(9,0.5))
o <- tabulate(replicate(1e4,which(rrace(jj) == "p4")))
o
plot(o,type='h')
```

```{r, cache=TRUE}
getprob <- function(n, r, x, N=1e4){
    jj <- BT(n, x)
    wanted <- paste0("p", str_pad(r ,ceiling(log10(n))))
    c(r=r, x=x, setNames(tabulate(replicate(N, which(rrace(jj) == wanted)),nbins=n), paste0("p", 1:n)))
}
rbind(
getprob(9,1,0.5,N=100),
getprob(9,2,0.5,N=100),
getprob(9,3,0.5,N=100),
getprob(9,4,0.5,N=100),
getprob(9,5,0.5,N=100),
getprob(9,6,0.5,N=100),
getprob(9,7,0.5,N=100),
getprob(9,8,0.5,N=100),
getprob(9,9,0.5,N=100)
)

rbind(
getprob(9,1,0.2,N=100),
getprob(9,2,0.2,N=100),
getprob(9,3,0.2,N=100),
getprob(9,4,0.2,N=100),
getprob(9,5,0.2,N=100),
getprob(9,6,0.2,N=100),
getprob(9,7,0.2,N=100),
getprob(9,8,0.2,N=100),
getprob(9,9,0.2,N=100)
)
rbind(
getprob(9,1,0.9,N=100),
getprob(9,2,0.9,N=100),
getprob(9,3,0.9,N=100),
getprob(9,4,0.9,N=100),
getprob(9,5,0.9,N=100),
getprob(9,6,0.9,N=100),
getprob(9,7,0.9,N=100),
getprob(9,8,0.9,N=100),
getprob(9,9,0.9,N=100)
)
```

```{r makeII,cache=TRUE}
r_try <- 1:9
x_try <- seq(from = 0.1, to = 1, by = 0.1)
jj <- as.matrix(expand.grid(r_try, x_try))
II <- t(apply(jj,1,function(v){getprob(9,v[1],v[2],N=1e4)}))
head(II)
```

```{r makesupp, cache=TRUE}
r_true <- 3
x_true <- 0.85
o <- getprob(9, r_true, x_true, 17)  # observations
o[1:2]
o[-(1:2)]

support_func <- function(o,v){
    v <- v[-(1:2)]
    v <- v/sum(v)
    dmultinom(o[-(1:2)], prob=v, log=TRUE)
}
support <- rep(0,nrow(II))
for(i in seq_len(nrow(II))){
    support[i] <- support_func(o,II[i,])
}
```

```{r}
support <- support-max(support)
support <- pmax(support,-10)
support  <- matrix(support,length(r_try),length(x_try))
contour(r_try, x_try, support, xlab="r", ylab="x", levels = -(1:5)*2)
abline(v = r_true)
abline(h = x_true)
abline(v = 1:9, col = 'gray', lty = 3, lwd = 0.3)
filled.contour(r_try, x_try, support, xlab = "r", ylab = "x")
```

```{R showjj}
jj <- cbind(II[,1:2],c(support))
jj[jj[,2] == 1,]
```

#  Analysis from a single observation

We translate $\alpha\in [0,1]$ to a number from 1 to $n$ inclusive
with $\lceil n\alpha\rceil$

Write a likelihood function for 9,12 [that is, came 9th in a class of
12] There are 8 better and 3 worse than the focal student.

We need a likelihood function for the observation (9) of $\alpha,x$.
Let's try again.

```{r ninetwelve, cache=TRUE}
n <- 12
a_try <- seq(from=0.01,to=0.99,len=10)
p_try <- ceiling(n * a_try)
x_try <- seq(from = 0.1, to = 1, by = 0.1)
jj <- as.matrix(expand.grid(p_try, x_try))
II <- t(apply(jj,1,function(v){getprob(n, v[1], v[2], N=1e4)}))
head(II)
tail(II)
S912 <- matrix(II[,11],length(a_try),length(x_try))
```

```{r show912}
S912 <- S912/max(S912)
contour(p_try,a_try,log(S912))
```

# Variable class sizes

Let's try this on the apocryphal education dataset

```{r}
M <- matrix(c(
    9, 7, 2, 3, 2, 8, 3, 4, 9,
    12,17,23,09,13,14,13,12,15
),nrow=2,byrow=TRUE)
M
```

### 2025-04-02 inst/exponential_BT.Rmd
more work on exponential model

@@
 
-```{r ninetwelve, cache=TRUE}
-n <- 12
+```{r ninetwelve}
 a_try <- seq(from=0.01,to=0.99,len=10)
+x_try <- seq(from = 0.1, to = 1, by = 0.1)
+single_like_fun <- function(n,a){
     p_try <- ceiling(n * a_try)
-x_try <- seq(from = 0.1, to = 1, by = 0.1)
     jj <- as.matrix(expand.grid(p_try, x_try))
     II <- t(apply(jj,1,function(v){getprob(n, v[1], v[2], N=1e4)}))
-head(II)
-tail(II)
-S912 <- matrix(II[,11],length(a_try),length(x_try))
+    matrix(II[,ceiling(a*n)+2],length(a_try),length(x_try))
+}
+```
+
+```{r makesupp912, cache=TRUE}
+allsupp <- list(
+    log(single_like_fun(12, a = 6/12)),
+    log(single_like_fun(17, a = 7/17)),
+    log(single_like_fun(23, a =12/23)),
+    log(single_like_fun(09, a = 5/09)),
+    log(single_like_fun(06, a = 7/13)),
+    log(single_like_fun(08, a = 4/18)),
+    log(single_like_fun(13, a = 8/13))
+)
 ```
@@
 ```{r show912}
-S912 <- S912/max(S912)
-contour(p_try,a_try,log(S912))
+S <- Reduce("+",allsupp)
+contour(a_try, x_try, S-max(S))
 ```

### 2025-04-03 inst/exponential_BT.Rmd
more work

--- 2025-04-02 inst/exponential_BT.Rmd
@@
 jj <- as.matrix(expand.grid(r_try, x_try))
-II <- t(apply(jj,1,function(v){getprob(9,v[1],v[2],N=1e4)}))
+II <- t(apply(jj, 1, function(v){getprob(9, v[1], v[2], N=1e4)}))
 head(II)
@@
 
-support_func <- function(o,v){
+support_func <- function(o, v){
     v <- v[-(1:2)]
@@
 support <- support-max(support)
-support <- pmax(support,-10)
+support <- pmax(support, -10)
 support  <- matrix(support,length(r_try),length(x_try))
@@
 
-We need a likelihood function for the observation (9) of $\alpha,x$.
-Let's try again.
+```{r ninetwelve}
+a_try <- seq(from=0.01,to=0.99,len=10)
+x_try <- seq(from = 0.1, to = 1, by = 0.1)
 
-```{r ninetwelve, cache=TRUE}
-n <- 12
-a_try <- seq(from=0.01,to=0.99,len=10)
+single_like_fun <- function(n, a, N=1e4){
     p_try <- ceiling(n * a_try)
-x_try <- seq(from = 0.1, to = 1, by = 0.1)
     jj <- as.matrix(expand.grid(p_try, x_try))
-II <- t(apply(jj,1,function(v){getprob(n, v[1], v[2], N=1e4)}))
-head(II)
-tail(II)
-S912 <- matrix(II[,11],length(a_try),length(x_try))
+    II <- t(apply(jj,1,function(v){getprob(n, v[1], v[2], N)}))
+    matrix(II[, ceiling(a*n) + 2], length(a_try), length(x_try))
+}
 ```
 
-```{r show912}
-S912 <- S912/max(S912)
-contour(p_try,a_try,log(S912))
+```{r makesupp912, cache=TRUE}
+allsupp <- list(
+    log(single_like_fun(12, a =  8/12)),  # pure
+    log(single_like_fun(17, a = 12/17)),  # pure
+    log(single_like_fun(23, a = 18/23)),  # pure
+    log(single_like_fun(14, a =  5/14)),  # applied
+    log(single_like_fun(13, a =  3/13)),  # applied
+    log(single_like_fun(11, a =  2/11)),  # applied
+    log(single_like_fun(13, a =  3/13))   # applied
+)
 ```
 
-# Variable class sizes
+```{r showallsupp}
+S <- Reduce("+",allsupp)
+S <- pmax(S-max(S),-10)
+contour(a_try, x_try, S, levels = -1*(1:7))
+jj <- which(S == max(S), arr.ind = TRUE)
+points(a_try[jj[1]], x_try[jj[2]], pch=16, col = 'red')
+```
 
-Let's try this on the apocryphal education dataset
+```{r showbothsupp}
+S_pure    <- Reduce("+", allsupp[1:3])
+S_applied <- Reduce("+", allsupp[4:7])
+S_all     <- Reduce("+", allsupp[1:7])
+max(S_all)
+max(S_pure)
+max(S_applied)
+max(S_pure) + max(S_applied)
+max(S_pure) + max(S_applied) - max(S_all)
+pchisq(max(S_pure) + max(S_applied) - max(S_all),df=2,lower.tail=FALSE)
+```
+
+# Parkrun
+
+```{r parkrunlikefun1, cache=TRUE}
+
+# In this chunk, each execution of single_like_fun(...,N=100) takes ~1
+# minute
+
+ print(system.time(parkrun01 <- single_like_fun(259, 177/259, N=1e3)))
+ print(system.time(parkrun02 <- single_like_fun(305, 222/305, N=1e3)))
+ print(system.time(parkrun03 <- single_like_fun(297, 206/297, N=1e3)))
+#print(system.time(parkrun04 <- single_like_fun(241, 142/241, N=1e1)))
+#print(system.time(parkrun05 <- single_like_fun(179, 118/179, N=1e1)))
+#print(system.time(parkrun06 <- single_like_fun(338, 224/338, N=1e1)))
+#print(system.time(parkrun07 <- single_like_fun(203, 128/203, N=1e1)))
+```
 
 ```{r}
-M <- matrix(c(
-    9, 7, 2, 3, 2, 8, 3, 4, 9,
-    12,17,23,09,13,14,13,12,15
-),nrow=2,byrow=TRUE)
-M
+parkrun01
+parkrun02
+parkrun03
+#parkrun04
+#parkrun05
+#parkrun06
+#parkrun07
+sum(parkrun01)
+sum(parkrun02)
+sum(parkrun03)
+parkrun01 * parkrun02 *parkrun03
+sum(parkrun01 * parkrun02 *parkrun03)
 ```

### 2025-04-03 inst/exponential_BT.Rmd
increase sample size

@@
 
- print(system.time(parkrun01 <- single_like_fun(259, 177/259, N=1e3)))
- print(system.time(parkrun02 <- single_like_fun(305, 222/305, N=1e3)))
- print(system.time(parkrun03 <- single_like_fun(297, 206/297, N=1e3)))
-#print(system.time(parkrun04 <- single_like_fun(241, 142/241, N=1e1)))
-#print(system.time(parkrun05 <- single_like_fun(179, 118/179, N=1e1)))
-#print(system.time(parkrun06 <- single_like_fun(338, 224/338, N=1e1)))
-#print(system.time(parkrun07 <- single_like_fun(203, 128/203, N=1e1)))
+ print(system.time(parkrun01 <- single_like_fun(259, 177/259, N=4e3)))
+ print(system.time(parkrun02 <- single_like_fun(305, 222/305, N=4e3)))
+ print(system.time(parkrun03 <- single_like_fun(297, 206/297, N=4e3)))
+ print(system.time(parkrun04 <- single_like_fun(241, 142/241, N=4e3)))
+ print(system.time(parkrun05 <- single_like_fun(179, 118/179, N=4e3)))
+ print(system.time(parkrun06 <- single_like_fun(338, 224/338, N=4e3)))
+ print(system.time(parkrun07 <- single_like_fun(203, 128/203, N=4e3)))
 ```
@@
 parkrun03
-#parkrun04
-#parkrun05
-#parkrun06
-#parkrun07
+parkrun04
+parkrun05
+parkrun06
+parkrun07
 sum(parkrun01)
@@
 ```
+
+1

### 2025-04-04 inst/exponential_BT.Rmd
refactored: minimal working example

@@
 ---
-title: "exponential_BT"
+title: "exponential Bradley-Terry"
 author: "robin hankin"
@@
 
-```{r}
+```{r defineBT}
 BT <- function(n,x){
@@
 
-```{r}
+```{r showtabulate}
 (jj <- BT(9,0.5))
@@
 
-```{r, cache=TRUE}
+```{r definegetprob, cache=TRUE}
 getprob <- function(n, r, x, N=1e4){
@@
 
-```{r makeII,cache=TRUE}
+```{r makeII, cache=TRUE}
 r_try <- 1:9
@@
 ```{r ninetwelve}
-a_try <- seq(from=0.01,to=0.99,len=10)
-x_try <- seq(from = 0.1, to = 1, by = 0.1)
-
-single_like_fun <- function(n, a, N=1e4){
+single_like_fun <- function(n, a,          # n competitors, places a [e.g a=3 -> third]
+                            a_try, x_try,  # values to try
+                            N=1e2){        # N is number of trials
     p_try <- ceiling(n * a_try)
@@
 ```{r makesupp912, cache=TRUE}
+a_try <- seq(from = 0.01, to = 0.99, len = 10)
+x_try <- seq(from = 0.2, to = 1, by = 0.1)
 allsupp <- list(
-    log(single_like_fun(12, a =  8/12)),  # pure
-    log(single_like_fun(17, a = 12/17)),  # pure
-    log(single_like_fun(23, a = 18/23)),  # pure
-    log(single_like_fun(14, a =  5/14)),  # applied
-    log(single_like_fun(13, a =  3/13)),  # applied
-    log(single_like_fun(11, a =  2/11)),  # applied
-    log(single_like_fun(13, a =  3/13))   # applied
+    log(single_like_fun(12, a =  8/12, a_try, x_try)),  # pure
+    log(single_like_fun(17, a = 12/17, a_try, x_try)),  # pure
+    log(single_like_fun(23, a = 18/23, a_try, x_try)),  # pure
+    log(single_like_fun(14, a =  5/14, a_try, x_try)),  # applied
+    log(single_like_fun(13, a =  3/13, a_try, x_try)),  # applied
+    log(single_like_fun(11, a =  2/11, a_try, x_try)),  # applied
+    log(single_like_fun(13, a =  3/13, a_try, x_try))   # applied
 )
@@
 
-```{r parkrunlikefun1, cache=TRUE}
+In the following chunks, each execution of
+`single_like_fun(...,N=1000)` takes  about 5 minutes
 
-# In this chunk, each execution of single_like_fun(...,N=100) takes ~1
-# minute
+```{r sethowmany}
+howmany <- 1000
+```
 
- print(system.time(parkrun01 <- single_like_fun(259, 177/259, N=4e3)))
- print(system.time(parkrun02 <- single_like_fun(305, 222/305, N=4e3)))
- print(system.time(parkrun03 <- single_like_fun(297, 206/297, N=4e3)))
- print(system.time(parkrun04 <- single_like_fun(241, 142/241, N=4e3)))
- print(system.time(parkrun05 <- single_like_fun(179, 118/179, N=4e3)))
- print(system.time(parkrun06 <- single_like_fun(338, 224/338, N=4e3)))
- print(system.time(parkrun07 <- single_like_fun(203, 128/203, N=4e3)))
+```{r parkrunlikefunobs1, cache = TRUE}
+print(system.time(parkrun01 <- single_like_fun(259, 177/259, N = howmany, a_try, x_try)))
 ```
@@
 parkrun01
-parkrun02
-parkrun03
-parkrun04
-parkrun05
-parkrun06
-parkrun07
-sum(parkrun01)
-sum(parkrun02)
-sum(parkrun03)
-parkrun01 * parkrun02 *parkrun03
-sum(parkrun01 * parkrun02 *parkrun03)
 ```
 
-1
+```{r parkrunlikefunobs2, cache = TRUE}
+ print(system.time(parkrun02 <- single_like_fun(305, 222/305, N = howmany, a_try, x_try)))
+```
+
+```{r}
+parkrun02
+```
+
+```{r parkrunlikefunobs3, cache = TRUE}
+print(system.time(parkrun03 <- single_like_fun(297, 206/297, N = howmany, a_try, x_try)))
+```
+
+```{r}
+parkrun03
+```
+
+```{r parkrunlikefunobs4, cache = TRUE}
+print(system.time(parkrun04 <- single_like_fun(241, 142/241, N = howmany, a_try, x_try)))
+```
+
+```{r}
+parkrun04
+```
+
+```{r parkrunlikefunobs5, cache = TRUE}
+print(system.time(parkrun05 <- single_like_fun(179, 118/179, N = howmany, a_try, x_try)))
+```
+
+```{r}
+parkrun05
+```
+
+```{r parkrunlikefunobs6, cache = TRUE}
+print(system.time(parkrun06 <- single_like_fun(338, 224/338, N = howmany, a_try, x_try)))
+```
+
+```{r}
+parkrun06
+```
+
+```{r parkrunlikefunobs7, cache = TRUE}
+print(system.time(parkrun07 <- single_like_fun(203, 128/203, N = howmany, a_try, x_try)))
+```
+
+```{r}
+parkrun07
+```
+
+```{r}
+c(
+sum(parkrun01),
+sum(parkrun02),
+sum(parkrun03),
+sum(parkrun04),
+sum(parkrun05),
+sum(parkrun06),
+sum(parkrun07)
+)
+```
+
+```{r}
+contour(a_try,x_try,parkrun05)
+contour(a_try,x_try,parkrun06)
+parkrun01*parkrun02*parkrun03*parkrun04*parkrun05*parkrun06*parkrun07
+```

### 2025-04-07 inst/exponential_BT.Rmd
lando

@@
 ```
+
+## Formula 1
+
+```{r checo}
+n <- 20 # 20 drivers
+multiple_like_fun <- function(n,           # n competitors
+                            a_try, x_try,  # values to try
+                            N=1e2){        # N is number of trials
+    p_try <- ceiling(n * a_try)
+    jj <- as.matrix(expand.grid(p_try, x_try))
+    t(apply(jj,1,function(v){getprob(n, v[1], v[2], N)}))
+}
+```
+
+```{r checo2023}
+o <- c(6,8,3,5,2,1,2,4,2,2,20,3,2,5,1,3,4,1,4,2,6,6,10) # lando norris 2024
+```
+
+```{r trymultx, cache=TRUE}
+II <- multiple_like_fun(20, a_try,x_try,1e4)
+head(II)
+nrow(II)
+length(a_try)
+length(x_try)
+```
+
+```{r calcmultx, cache=TRUE}
+like <- list()
+dim(II)
+o
+for(i in seq_along(o)){
+    like[[i]] <-  matrix(II[, o[i] + 2], length(a_try), length(x_try))
+}
+```
+
+```{r showlike}
+like
+alllike <- Reduce("*",like)
+alllike 
+alllike/max(alllike)
+S <- log(alllike)
+S <- S-max(S)
+S <- pmax(S,-15)
+contour(S,levels = -(0:6)*2, lwd=c(1,4,1,1,1,1,1))
+filled.contour(S)
+```

### 2025-04-08 inst/imo.Rmd
start work on IMO analysis

---
title: "imo"
author: "robin hankin"
date: "2025-04-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(hyper2)
library(stringr)
```

## R Markdown

```{r}
oecd <- read.csv("oecd.txt",header=FALSE)$V1
oecd
length(oecd)
oecd <- oecd[!(oecd %in% c("Chile", "Democratic People's Republic of Korea", "Turkiye"))]

years <- 2001:2024

jj <- read.table("imo.txt", header=TRUE)
ranks <- list()
for(i in seq_along(years)){
    ranks[[i]] <- subset(jj, jj$year == years[[i]])$country
}
names(ranks) <- years
```

```{r}
for(i in seq_along(ranks)){
    jj <- ranks[[i]]
    ranks[[i]] <- jj[jj %in% oecd]
}
ranks
```

```{r}
uk <- seq_along(ranks)*0
for(i in seq_along(ranks)){
    uk[i] <- which(ranks[[i]] == "Australia")
}
uk
```

```{r checo}
BT <- function(n,x){
    out <- x^(0:(n-1))
    if(x != 1){
        out <- out*(1-x)/(1-x^n)
    } else {
        out <- out/n
    }
    names(out) <- paste0("p",str_pad(1:n,ceiling(log10(n))))
    out
}

getprob <- function(n, r, x, N=1e2){
    jj <- BT(n, x)
    wanted <- paste0("p", str_pad(r ,ceiling(log10(n))))
    c(r=r, x=x, setNames(tabulate(replicate(N, which(rrace(jj) == wanted)),nbins=n), paste0("p", 1:n)))
}
n <- length(oecd)
a_try <- seq(from = 0.01, to = 0.99, len = 20)
x_try <- seq(from = 0.2, to = 1, by = 0.05)
multiple_like_fun <- function(n,           # n competitors
                            a_try, x_try,  # values to try
                            N=1e2){        # N is number of trials
    p_try <- ceiling(n * a_try)
    jj <- as.matrix(expand.grid(p_try, x_try))
    t(apply(jj,1,function(v){getprob(n, v[1], v[2], N)}))
}
```

```{r trymultxy, cache=TRUE}
print(system.time(II <- multiple_like_fun(length(oecd), a_try, x_try, 1e4)))
head(II)
nrow(II)
length(a_try)
length(x_try)
```

```{r calcmultxy, cache=TRUE}
like <- list()
dim(II)
o <- uk # o for observation
o
for(i in seq_along(o)){
    like[[i]] <-  matrix(II[, o[i] + 2], length(a_try), length(x_try))
}
```

```{r showlike}
alllike <- Reduce("*",like)
alllike 
alllike/max(alllike)
S <- log(alllike)
S <- S-max(S)
S <- pmax(S,-15)
round(S,2)
contour(S,levels = -(0:6)*2, lwd=c(1,4,1,1,1,1,1))
filled.contour(S)
```

### 2025-04-08 inst/imo.Rmd
consistentification

@@
 ```{r}
-oecd <- read.csv("oecd.txt",header=FALSE)$V1
+oecd <- as.character(read.csv("oecd.txt",header=FALSE)$V1)
 oecd
-length(oecd)
+
 oecd <- oecd[!(oecd %in% c("Chile", "Democratic People's Republic of Korea", "Turkiye"))]
@@
 }
-ranks
 ```
@@
 alllike <- Reduce("*",like)
-alllike 
-alllike/max(alllike)
 S <- log(alllike)
 S <- S-max(S)
-S <- pmax(S,-15)
-round(S,2)
-contour(S,levels = -(0:6)*2, lwd=c(1,4,1,1,1,1,1))
+S <- pmax(S, -9.999)
+jj <- S
+jj[jj < -9.99] <- NA
+round(jj,2)
+contour(a_try, x_try, S,
+        xlab = "BT strength (scaled)", ylab = "x",
+        levels = -(0:6)*2, lwd=c(1,4,1,1,1,1,1))
+dim(S)
+contour(a_try[1:8], x_try[6:17], S[1:8, 6:17], levels = -(0:6)*2, lwd=c(1,4,1,1,1,1,1))
 filled.contour(S)
 ```
+
+```{r}
+like[[1]]
+```

### 2025-04-09 inst/parkrun_idea.Rmd
page reference
another reference to underdispersion
more work on issue #239
consistentification

--- 2024-03-28 inst/parkrun_idea.Rmd
@@
 ---
-title: "Underdispersion in the context of community athletics"
-output: bookdown::html_document2
-author: "Robin K. S. Hankin"
-bibliography: "hyper2.bib"
+title: "Park run with exponential Bradley-Terry"
+author: "robin hankin"
+date: "2025-04-02"
+output: html_document
+
 ---
@@
 knitr::opts_chunk$set(echo = TRUE)
+library(hyper2)
+library(stringr)
+set.seed(0)
 ```
 
-```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
-knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+Basic idea is to have $n$ competitors with BT strengths $\beta, \beta
+x,\ldots \beta x^{n-1}$ with $x\in[0,1]$.  The unit sum constraint
+gives $\beta=\frac{1-x}{1-x^n}$ but this usually cancels out.  One of
+the competitors is the _focal_ competitor with BT strength $\beta
+x^{r-1}$ for some $r$ with $1\leqslant r < n$.  Thus $r=1$ means that
+the focal competitor is the strongest among the $n$, $r=2$ means he is
+the second, and so on.
+
+```{r defineBT}
+BT <- function(n,x){
+    out <- x^(0:(n-1))
+    if(x != 1){
+        out <- out*(1-x)/(1-x^n)
+    } else {
+        out <- out/n
+    }
+    names(out) <- paste0("p",str_pad(1:n,ceiling(log10(n))))
+    out
+}
+BT(4,0.5)
+sum(BT(4,0.5))
 ```
 
-To cite the `hyper2` package in publications, please use @hankin2017_rmd.
+We will consider $n=9$, $x=0.5$, $r=4$ and make repeated observations:
 
-```{r deff}
-library("hyper2")
+```{r showtabulate}
+(jj <- BT(9,0.5))
+o <- tabulate(replicate(1e4,which(rrace(jj) == "p4")))
+o
+plot(o,type='h')
+```
 
-f <- function(n,alpha){
-  jj <- 1/alpha^(1:n)
-  return(jj/sum(jj))
-}
-
-getlike <- function(v){
-  alpha <- v[1]
-  beta <- v[2]
- c(
-tabulate(replicate(1000,which(rrace(setNames(f(26,alpha),letters))==letters[beta])),nbins=26)
-,alpha,beta)
+```{r definegetprob, cache=TRUE}
+getprob <- function(n, r, x, N=1e4){
+    jj <- BT(n, x)
+    wanted <- paste0("p", str_pad(r ,ceiling(log10(n))))
+    c(r=r, x=x, setNames(tabulate(replicate(N, which(rrace(jj) == wanted)),nbins=n), paste0("p", 1:n)))
 }
@@
 
-```{r calculateLikemat,cache=FALSE}
-alpha_try <- seq(from=1,to=6,len=20)
-beta_try <- 1:26
-calculate_from_scratch <- FALSE
-if(calculate_from_scratch){
-    M <- as.matrix(expand.grid(alpha=alpha_try, beta=beta_try))
-    system.time(L <- t(apply(M,1,getlike)))
-    colnames(L) <- c(letters,'alpha','beta')
-    write.table(L,file="likelihoodmatrix.txt")
-} else {
-    L <- read.table("likelihoodmatrix.txt")
+We translate $\alpha\in [0,1]$ to a number from 1 to $n$ inclusive
+with $\lceil n\alpha\rceil$
+
+Write a likelihood function for 9,12 [that is, came 9th in a class of
+12] There are 8 better and 3 worse than the focal student.
+
+```{r ninetwelve}
+single_like_fun <- function(n, a,          # n competitors, places a [e.g a=3 -> third]
+                            a_try, x_try,  # values to try
+                            N=1e2){        # N is number of trials
+    p_try <- ceiling(n * a_try)
+    jj <- as.matrix(expand.grid(p_try, x_try))
+    II <- t(apply(jj,1,function(v){getprob(n, v[1], v[2], N)}))
+    matrix(II[, ceiling(a*n) + 2], length(a_try), length(x_try))
 }
-L_integer <- L  # counts
-L[,1:26] <- L[,1:26]/mean(rowSums(L[,1:26]))
-head(L_integer)
 ```
 
-```{r makecontL}
-colSums(L)
-mL <- function(i){matrix(L[,i],length(alpha_try),length(beta_try))}
-filled.contour(alpha_try,beta_try,mL(1))
-filled.contour(alpha_try,beta_try,mL(5))
-filled.contour(alpha_try,beta_try,mL(10))
-filled.contour(alpha_try,beta_try,mL(15))
-filled.contour(alpha_try,beta_try,mL(20))
-filled.contour(alpha_try,beta_try,mL(26))
+```{r makesupp912, cache=TRUE}
+a_try <- seq(from = 0.01, to = 0.99, len = 10)
+x_try <- seq(from = 0.2, to = 1, by = 0.1)
+```
+
+In the following chunks, each execution of
+`single_like_fun(...,N=1000)` takes  about 5 minutes
+
+```{r sethowmany}
+howmany <- 100
+```
+
+```{r parkrunlikefunobs1, cache = TRUE}
+print(system.time(parkrun01 <- single_like_fun(259, 177/259, N = howmany, a_try, x_try)))
 ```
@@
 ```{r}
-#plot(table(L_integer[,1:26]))
+parkrun01
 ```
 
-## Now 10 in-silico races:
-```{r doinsilico,cache=FALSE}
-alphatrue <- 2.3
-betatrue <- 4  
-set.seed(1)
-D <- replicate(10,which(rrace(setNames(f(26,alphatrue),letters))==letters[betatrue]))
-sort(D)
+```{r parkrunlikefunobs2, cache = TRUE}
+ print(system.time(parkrun02 <- single_like_fun(305, 222/305, N = howmany, a_try, x_try)))
 ```
 
-```{r trylikeinsilico,cache=FALSE}
-likelihood <- apply(L[,D],1,prod) + 0.00
-likelihood <- matrix(likelihood,length(alpha_try),length(beta_try))
+```{r}
+parkrun02
 ```
 
-```{r showlike}
-round(alpha_try,2)
-matplot(alpha_try,likelihood,type='b')
-matplot(beta_try,t(likelihood),type='b')
-matplot(alpha_try,log(likelihood)-log(max(likelihood)),type='b',ylim=c(-20,0))
-matplot(beta_try,log(t(likelihood))-log(max(likelihood)),type='b',ylim=c(-20,0))
-filled.contour(alpha_try,beta_try,(likelihood))
-table(likelihood==0)
-plot(f(26,alphatrue))
-abline(v=betatrue)
-contour(alpha_try,beta_try,likelihood)
-abline(v=alphatrue)
-abline(h=betatrue)
-contour(alpha_try,beta_try,log(likelihood) - log(max(likelihood)),levels=-(0:10))
-abline(v=alphatrue)
-abline(h=betatrue)
+```{r parkrunlikefunobs3, cache = TRUE}
+print(system.time(parkrun03 <- single_like_fun(297, 206/297, N = howmany, a_try, x_try)))
 ```
 
-\subsection{Park run}
+```{r}
+parkrun03
+```
 
-"Parkrun" is a distributed community initiative that organises
-weekly timed 5 km runs/walks in parks worldwide~\cite{hindley2020}.  A
-typical event will have 200 participants.  Table~\ref{parkruntable}
-shows the author has completed a total of 15 parkruns to date, and
-from the first pair of numbers we see that 238 runners attended that
-particular parkrun, of whom the author placed 173.
+```{r parkrunlikefunobs4, cache = TRUE}
+print(system.time(parkrun04 <- single_like_fun(241, 142/241, N = howmany*10, a_try, x_try)))
+```
 
-\begin{table*}[t]
-  \caption{Parkrun results}
-\label{parkruntable}
-\begin{tabular}{ccccccccccc}
-\hline
-rank   &173  & 165 & 172 & 199 & 181 & 229 & 177 & 222 & 206 & 142 \\
-runners&238  & 238 & 196 & 242 & 242 & 318 & 259 & 305 & 297 & 241 \\ \\ \\
-rank   & 118 & 224 & 128 & 115 & 183 &     &     &     &     &\\
-runners& 179 & 338 & 203 & 245 & 254 &     &     &     &     &\\
-\hline
-\end{tabular}
-\end{table*}
+```{r}
+parkrun04
+```
 
-We may consider the author to have an unknown generalized Bradley
-Terry strength $a$: Figure~\ref{parkrunsupport} shows a support curve
-for $a$ with the evaluate, $\hat{a}\simeq 0.448$ shown.  We also see
-two units of support~\cite{edwards1992} shown as a dotted line
-illustrating a support interval of about 0.315 - 0.567; we may be
-reasonably confident that the author's true BT strength lies in this
-range.  In particular, note that a reasonable $H_0\colon
-a=\frac{1}{2}$ may not be rejected, the support for $H_0$ being only
-susceptible to very minor improvement [$\simeq 0.351$] by moving to
-the evaluate.
+```{r parkrunlikefunobs5, cache = TRUE}
+print(system.time(parkrun05 <- single_like_fun(179, 118/179, N = howmany, a_try, x_try)))
+```
 
-```{asis}
-\begin{figure}[t]
-\includegraphics[width=4in]{plotparkrun}  % plotparkrun.pdf made in very_simplified_likelihood.Rmd
-\caption{Author's strength\label{parkrunsupport}}
-\end{figure}
+```{r}
+parkrun05
 ```
+
+```{r parkrunlikefunobs6, cache = TRUE}
+print(system.time(parkrun06 <- single_like_fun(338, 224/338, N = howmany, a_try, x_try)))
+```
+
+```{r}
+parkrun06
+```
+
+```{r parkrunlikefunobs7, cache = TRUE}
+print(system.time(parkrun07 <- single_like_fun(203, 128/203, N = howmany, a_try, x_try)))
+```
+
+```{r}
+parkrun07
+```
+
+```{r}
+c(
+sum(parkrun01),
+sum(parkrun02),
+sum(parkrun03),
+sum(parkrun04),
+sum(parkrun05),
+sum(parkrun06),
+sum(parkrun07)
+)
+```
+
+```{r}
+contour(a_try,x_try,parkrun05)
+contour(a_try,x_try,parkrun06)
+parkrun01*parkrun02*parkrun03*parkrun04*parkrun05*parkrun06*parkrun07
+```

### 2025-04-22 inst/jss_hyper3_real.Rnw
typos

@@
 this paper I generalize the \pkg{hyper2} R
-package~\citep{hankin2017_rnw_jss}, available on CRAN~\citep{rcore2025},
-for Bradley-Terry likelihoods to accommodate this class of
-multiplicative factors with new \proglang{S3} \class{hyper3} and
-associated methods.
+package~\citep{hankin2017_rnw_jss}, available on
+CRAN~\citep{rcore2025}, for Bradley-Terry likelihoods to accommodate
+this class of multiplicative factors with new \proglang{S3} class
+\class{hyper3} and associated methods.
 
@@
 \subsection{The package in use}
-Package idiom for creating a likelihood function for this dataset is
-straightforward.  Considering first only Rangers-Celtic matches:
+Creating a likelihood function for this dataset is straightforward.
+As an illustration of package idiom, one way of representing a
+likelihood function for Rangers-Celtic matches (here, ignoring draws)
+would be:
 %
@@
 supplied value of $\lambda$.  We may use this to create a profile
-likelihood (Figure~\ref{profilesupportlambda}), or to find the maximum
-likelihood estimate for $\lambda$:
+likelihood curve (Figure~\ref{profilesupportlambda}), or to find the
+maximum likelihood estimate for $\lambda$:
 
@@
 \cite{mcfadden1980} although different metaphors have been employed
-for the same phenomenon by \cite{tverskey1972_jss}.  The canonical problem
-is as follows.  A commuter has a choice of going to work by different
-transport methods.  Here, we can assume that he has possibly
+for the same phenomenon by \cite{tverskey1972_jss}.  The canonical
+problem is as follows.  A commuter has a choice of going to work by
+different transport methods.  Here, we can assume that he has possibly
 strong---but unknown---views on different forms of transport which
@@
 \cite{tverskey1972_jss}, for example, presents an approach involving
-overlapping ``aspects'' of the elements of choice set;
-\cite{mcfadden1980} suggests using a generalized extreme-value model.
+overlapping ``aspects'' of the choice set; \cite{mcfadden1980}
+suggests using a generalized extreme-value model.
 
@@
 strengths presented above.  As a synthetic example, we specify options
-RB and BB for red bus and blue bus together options C, T and W for
-car, train, and walking.
+RB and BB for red bus and blue bus together with options C, T and W
+for car, train, and walking.
 
@@
 <<findstuffsushi, echo = FALSE, cache = TRUE>>=
-osushi <- optimize(f, c(2, 4), maximum = TRUE)
+osushi <- optimize(f, c(1, 4), maximum = TRUE)
 @ 
@@
 $\chi^2_1$ distribution with a $p$~value of about
-$\Sexpr{round(pchisq(sushisupport,df=1,lower.tail=FALSE),3)}$.  We can
+$\Sexpr{round(pchisq(2*sushisupport,df=1,lower.tail=FALSE),3)}$.  We can
 therefore be reasonably sure that tuna sushi are somehow analagous to

### 2025-04-22 inst/imo.Rmd
better title
variable number of competitors

@@
 ---
-title: "imo"
+title: "International Mathematical Olympiad using Plackett-Luce"
 author: "robin hankin"
@@
 ```
+
+# Some thoughts about including variable numbers of competitors
+
+Above we used OECD countries for Australia to compete against.  Here I
+use all the competitors:
+
+```{r}
+a <- read.table("imo.txt", header = TRUE)
+min(a$year)
+max(a$year)
+which(subset(a,a$year == 2015)$country == "Australia")
+competitors <- function(year,a){ table(a$year)[as.character(year)]   }
+competitors(2013, a)
+rank <- function(place, y, a){
+    which(subset(a,a$year == y)$country == place)
+}
+
+rank("Australia",2001,a)
+rank("Australia",2002,a)
+rank("Australia",2024,a)
+cbind(place = sapply(2001:2024,function(y){rank("Australia",y,a)}),
+      competitors = table(a$year)
+      )
+```
+
+Now try New Zealand:
+
+```{r imo_nz}
+cbind(place = sapply(2001:2024,function(y){rank("New Zealand",y,a)}),
+      competitors = table(a$year)
+      )
+```

### 2025-04-22 inst/jss_hyper3_real.Rnw
better make it 4 decimal places

@@
 $\chi^2_1$ distribution with a $p$~value of about
-$\Sexpr{round(pchisq(2*sushisupport,df=1,lower.tail=FALSE),3)}$.  We can
+$\Sexpr{round(pchisq(2*sushisupport,df=1,lower.tail=FALSE),4)}$.  We can
 therefore be reasonably sure that tuna sushi are somehow analagous to

### 2025-04-27 inst/carcinoma.Rmd
typo

@@
 value of `carcinoma_maxp` is calculated separately [discussed below],
-and that takes a few days to run.  </font>
+and that takes a few days to run).  </font>
 

### 2025-04-27 inst/volvo.Rmd
soft link for hyper2.bib
ordertable() now deprecated

@@
 ```{r readvolvotable}
-volvo_table <- ordertable(read.table("volvo.txt"))
+volvo_table <- read.table("volvo.txt")
 volvo_table
@@
 finish" and `DNS` "did not start".  We can convert the table to a
-support function with `ordertable2supp()`, remembering to specify the
+support function with `suppfun()`, remembering to specify the
 noscore options:

### 2025-09-30 inst/exponential_BT.Rmd
Hamilton not Norris

@@
 ```{r checo2023}
-o <- c(6,8,3,5,2,1,2,4,2,2,20,3,2,5,1,3,4,1,4,2,6,6,10) # lando norris 2024
+
+o <- c(5,5,2,6,6,4,2,3,8,3,4,4,6,6,3,5,2,8,7,9)  # Lewis Hamilton 2023
+# o <- c(7,9,9,9,6,6,7,4,3,4,1,3,1,8,5,9,6,4,10,2,12,4) # Lewis Hamilton 2024
+# o <- c(6,8,3,5,2,1,2,4,2,2,20,3,2,5,1,3,4,1,4,2,6,6,10) # lando norris 2024
 ```
@@
 
-```{r showlike}
+```{r showlike, cache=FALSE}
+options(digits=5)
 like

### 2025-10-02 inst/exponential_BT.Rmd
more work

@@
 r_try <- 1:9
-x_try <- seq(from = 0.1, to = 1, by = 0.1)
+x_try <- seq(from = 0.1, to = 1, by = 0.05)
 jj <- as.matrix(expand.grid(r_try, x_try))
@@
 x_true <- 0.85
-o <- getprob(9, r_true, x_true, 17)  # observations
+set.seed(0)
+o <- getprob(9, r_true, x_true, 100)  # observations
 o[1:2]
@@
 for(i in seq_len(nrow(II))){
-    support[i] <- support_func(o,II[i,])
+    support[i] <- support_func(o, II[i,])
 }
@@
 ```{r}
+x_try
+length(support)
 support <- support-max(support)
@@
 support  <- matrix(support,length(r_try),length(x_try))
-contour(r_try, x_try, support, xlab="r", ylab="x", levels = -(1:5)*2)
+contour(r_try, x_try, support, xlab="r", ylab="x", levels = -c(1,(1:5)*2), lwd=c(1,3,1,1,1,1))
+abline(v=1:9, lwd=0.3, col='gray')
 abline(v = r_true)
-abline(h = x_true)
 abline(v = 1:9, col = 'gray', lty = 3, lwd = 0.3)
 filled.contour(r_try, x_try, support, xlab = "r", ylab = "x")
+```
+
+```{r contourLO}
+x_try
+wanted <- 8:18
+LO <- function(p){log(p/(1-p))}
+contour(r_try, LO(x_try[wanted]), support[,wanted], xlab="r", ylab="Log-odds of x", levels = -c(1,(1:5)*2), lwd=c(1,3,1,1,1,1))
 ```

### 2025-10-02 inst/hankin-exponential_BT.tex
more work

--- 2025-04-01 inst/hankin-AustrianJournalofStatistics.tex
@@
 \Plainauthor{Robin K. S. Hankin} %% comma-separated
-\Plaintitle{A Simplified Plackett-Luce Likelihood Function for Rank Orders}
+\Plaintitle{A Parametrized Plackett-Luce Likelihood Function for Rank Orders}
 
@@
 that are ordered in some way [the preferred example is competitors in
-  a running race, ordered by time of crossing the finishing line].
-One entity is of particular interest to us: the ``focal entity'', or
-focal competitor.  We assign Plackett--Luce strength $a$ to the focal
-competitor and $b$ to all others; we require $a+b=1$.  The associated
-inference problem has a wide range of applications and I present some
-analysis of datasets drawn from the fields of amateur athletics,
-education, and Formula~1 motor racing.
+  a running race, ordered by time of crossing the finishing line].  We
+assign Plackett--Luce strengths $\alpha x,\alpha x^2\ldots,\alpha x^n$
+where $0\leq x\leqslant 1$ is a common ratio and $\alpha$ a
+normalising constant.  One entity is of particular interest to us: the
+``focal entity'', or focal competitor whose strength is $\alpha x^r$
+for some unknown value of $r$ with $1\leqslant r\leqslant n$.  We
+consider the following inference problem: given the order statistic
+for the focal competitor [first, second, etc], and the total number of
+competitors, what might we infer about $x$ and $r$?  Such problems
+arise when non-focal competitors' performances are censored, unknown,
+or irrelevant.  }
 
-}
 \Keywords{keywords, comma-separated, non-capitalized, \proglang{R}}
@@
 competitors.
+
+\subsection{Some simple cases}
+
+We consider the smallest non-trivial cases $n=3$ and $n=4$.  For $n=3$
+we have Plackett-Luce strengths $\alpha x,\alpha x^2,\alpha x^3$ [the
+  extra factor of $x$ allows us to identify the $r$ in ``$\alpha x^r$''
+  with the rank of the competitor, the best being rank 1].  Suppose
+the focal competitor came first, the likelihood function would be
+
+\begin{equation}
+  \mathcal{L}(x,r)=\frac{x^{r-1}}{1+x+x^2}\qquad r=1,2,3
+\end{equation}
+
+If the focal competitor came second we would have two Plackett-Luce
+probabilities to add, corresponding to $\alpha x^i$ or $\alpha x^j$
+coming first where $i,j\neq r$.  For example, if $r=1$ we would have
+two finishing orders that are consistent with the observation: $2\succ
+1\succ 3$ and $3\succ 1\succ 2$.  These are disjoint so the likelihood
+will be proportional to
+
+$$
+\frac{x^2}{x+x^2+x^3}\cdot\frac{x}{x+x^3} + 
+\frac{x^3}{x+x^2+x^3}\cdot\frac{x}{x+x^2}
+$$
+
+After simplification:
+
+\begin{equation}
+  \mathcal{L}(x,r)=
+  \begin{cases}
+    \frac{x  }{x+x^2+x^3}\left(\frac{x^2}{x+x^3}   + \frac{x^3}{x+x^2}\right)\qquad r=1\\
+    \frac{x^2}{x+x^2+x^3}\left(\frac{x  }{x^2+x^3} + \frac{x^3}{x+x^2}\right)\qquad r=2\\
+    \frac{x^3}{x+x^2+x^3}\left(\frac{x  }{x^2+x^3} + \frac{x^2}{x+x^3}\right)\qquad r=3
+\end{cases}
+\end{equation}
+
+The $n=4$ case is harder as we potentially have six consistent
+orderings for each observation.  As an example, if the focal
+competitor comes third we have $1
 

### 2025-10-07 inst/exponential_BT.Rmd
simple cases

@@
 the second, and so on.
+
+# Some simple cases
+
+We consider the smallest non-trivial cases $n=3$ and $n=4$.  For $n=3$
+we have Plackett-Luce strengths $\alpha x,\alpha x^2,\alpha x^3$ [the
+  extra factor of $x$ allows us to identify the $r$ in ``$\alpha x^r$''
+  with the rank of the competitor, the best being rank 1].  Suppose
+the focal competitor came first, the likelihood function would be
+
+\begin{equation}
+  \mathcal{L}(x,r)=\frac{x^{r-1}}{1+x+x^2}\qquad r=1,2,3
+\end{equation}
+
+```{r plot123first}
+f <- function(x,r){ x^r/(x + x^2 + x^3) }
+
+x <- seq(from=0.01 , to=1, by=0.01)
+plot(c(-3,0),c(-3,1),type='n', xlab="log x", ylab="support",
+     main = "support for x,r", sub="Three competitors, focal competitor came FIRST"
+     )
+points(log(x),log(f(x,1)),type="l",col="black")
+points(log(x),log(f(x,2)),type="l", col="red")
+points(log(x),log(f(x,3)),type="l", col="blue")
+
+abline(h=c(0,-2))
+
+legend("topright",lty=1, col=c("black","red","blue"),
+       legend=c("r=1","r=2","r=3"))
+
+```
+
+If the focal competitor came second we would have two Plackett-Luce
+probabilities to add, corresponding to different order statistics.
+$\alpha x^i$ or $\alpha x^j$ coming first where $i,j\neq r$.  For
+example, if $r=1$ we would have two finishing orders that are
+consistent with the observation: $2\succ 1\succ 3$ and $3\succ 1\succ
+2$.  These are disjoint so the likelihood will be proportional to
+
+$$
+\frac{x^2}{x+x^2+x^3}\cdot\frac{x}{x+x^3} + 
+\frac{x^3}{x+x^2+x^3}\cdot\frac{x}{x+x^2}
+$$
+
+After simplification:
+
+\begin{equation}
+  \mathcal{L}(x,r)=
+  \begin{cases}
+    \frac{x  }{x+x^2+x^3}\left(\frac{x^2}{x+x^3}   + \frac{x^3}{x+x^2}\right)\qquad r=1\\
+    \frac{x^2}{x+x^2+x^3}\left(\frac{x  }{x^2+x^3} + \frac{x^3}{x+x^2}\right)\qquad r=2\\
+    \frac{x^3}{x+x^2+x^3}\left(\frac{x  }{x^2+x^3} + \frac{x^2}{x+x^3}\right)\qquad r=3
+\end{cases}
+\end{equation}
+
+```{r plot123second}
+like <- function(x,r){
+  S <- x + x^2 + x^3
+  switch(r,
+         '1' = x^2/S*x^1/(x^1+x^3) + x^3/S*x^1/(x^1+x^2), # 213+312
+         '2' = x^1/S*x^2/(x^2+x^3) + x^3/S*x^2/(x^1+x^2), # 123+321
+         '3' = x^1/S*x^3/(x^2+x^3) + x^2/S*x^3/(x^1+x^3)  # 132+231
+  )
+}
+
+x <- seq(from=0.01 , to=1, by=0.01)
+plot(c(-3,0),c(-3,1),type='n', xlab="log x", ylab="support",
+     main = "support for x,r", sub="Three competitors, focal competitor came SECOND"
+     )
+points(log(x),log(like(x,1)),type="l",col="black")
+points(log(x),log(like(x,2)),type="l", col="red")
+points(log(x),log(like(x,3)),type="l", col="blue")
+
+abline(h=c(0,-2))
+
+legend("topright",lty=1, col=c("black","red","blue"),
+       legend=c("r=1","r=2","r=3"))
+
+```
+
+Now the focal competitor comes last:
+
+```{r plot123third}
+like <- function(x,r){
+  S <- x + x^2 + x^3
+  switch(r,
+         '1' = x^3/S*x^2/(x^1+x^2) + x^2/S*x^3/(x^1+x^3), # 213+312
+         '2' = x^1/S*x^3/(x^2+x^3) + x^3/S*x^1/(x^1+x^2), # 123+321
+         '3' = x^1/S*x^2/(x^2+x^3) + x^2/S*x^1/(x^1+x^3)  # 132+231
+  )
+}
+
+x <- seq(from=0.01 , to=1, by=0.01)
+plot(c(-3,0),c(-3,1),type='n', xlab="log x", ylab="support",
+     main = "support for x,r", sub="Three competitors, focal competitor came THIRD"
+     )
+points(log(x),log(like(x,1)),type="l",col="black")
+points(log(x),log(like(x,2)),type="l", col="red")
+points(log(x),log(like(x,3)),type="l", col="blue")
+
+abline(h=c(0,-2))
+
+legend("topright",lty=1, col=c("black","red","blue"),
+       legend=c("r=1","r=2","r=3"))
+
+```
+
+The $n=4$ case is harder as we potentially have six consistent
+orderings for each observation.  As an example, if the focal
+competitor comes third we have $1
 

### 2025-10-13 inst/exponential_BT.Rmd
simple cases

--- 2025-10-02 inst/exponential_BT.Rmd
@@
 the second, and so on.
+
+# Some simple cases
+
+We consider the smallest non-trivial cases $n=3$ and $n=4$.  For $n=3$
+we have Plackett-Luce strengths $\alpha x,\alpha x^2,\alpha x^3$ [the
+  extra factor of $x$ allows us to identify the $r$ in ``$\alpha x^r$''
+  with the rank of the competitor, the best being rank 1].  Suppose
+the focal competitor came first, the likelihood function would be
+
+\begin{equation}
+  \mathcal{L}(x,r)=\frac{x^{r-1}}{1+x+x^2}\qquad r=1,2,3
+\end{equation}
+
+```{r plot123first}
+worker_123first <- function(...){
+f <- function(x,r){ x^r/(x + x^2 + x^3) }
+
+x <- seq(from=0.01 , to=1, by=0.01)
+plot(c(-3,0), c(-3,0), type='n',
+     xlab="log x", ylab="log-likelihood",main="(a)")
+points(log(x),log(f(x,1)), type="l", col="black")
+points(log(x),log(f(x,2)), type="l", col="red"  )
+points(log(x),log(f(x,3)), type="l", col="blue" )
+
+abline(h=0, lty=2)
+abline(h = -2)
+e <- exp(1)
+r2reject <- ((e^2-1)-sqrt(e^4-2*e^2-3))/2
+segments(x0=log(r2reject), y0=-2.2,y1=-1.8)
+
+r3reject <- (1+sqrt(1+4*(e^2-1)))/(2*(e^2-1))
+segments(x0=log(r3reject), y0=-2.2,y1=-1.8)
+legend(x= -3, y= - 0.5, lty=1, col=c("black","red","blue"),
+       legend=c("r=1","r=2","r=3"))
+}
+e <- exp(1)
+r2reject <- ((e^2-1)-sqrt(e^4-2*e^2-3))/2
+print(r2reject)
+r3reject <- (1+sqrt(1+4*(e^2-1)))/(2*(e^2-1))
+print(r3reject)
+
+worker_123first()
+pdf(file="123first.pdf")
+worker_123first()
+dev.off()
+
+```
+
+If the focal competitor came second we would have two Plackett-Luce
+probabilities to add, corresponding to different order statistics.
+$\alpha x^i$ or $\alpha x^j$ coming first where $i,j\neq r$.  For
+example, if $r=1$ we would have two finishing orders that are
+consistent with the observation: $2\succ 1\succ 3$ and $3\succ 1\succ
+2$.  These are disjoint so the likelihood will be proportional to
+
+$$
+\frac{x^2}{x+x^2+x^3}\cdot\frac{x}{x+x^3} + 
+\frac{x^3}{x+x^2+x^3}\cdot\frac{x}{x+x^2}
+$$
+
+After simplification:
+
+\begin{equation}
+  \mathcal{L}(x,r)=
+  \begin{cases}
+    \frac{x  }{x+x^2+x^3}\left(\frac{x^2}{x+x^3}   + \frac{x^3}{x+x^2}\right)\qquad r=1\\
+    \frac{x^2}{x+x^2+x^3}\left(\frac{x  }{x^2+x^3} + \frac{x^3}{x+x^2}\right)\qquad r=2\\
+    \frac{x^3}{x+x^2+x^3}\left(\frac{x  }{x^2+x^3} + \frac{x^2}{x+x^3}\right)\qquad r=3
+\end{cases}
+\end{equation}
+
+```{r plot123second}
+worker_123second <- function(...){
+like <- function(x,r){
+  S <- x + x^2 + x^3
+  switch(r,
+         '1' = x^2/S*x^1/(x^1+x^3) + x^3/S*x^1/(x^1+x^2), # 213+312
+         '2' = x^1/S*x^2/(x^2+x^3) + x^3/S*x^2/(x^1+x^2), # 123+321
+         '3' = x^1/S*x^3/(x^2+x^3) + x^2/S*x^3/(x^1+x^3)  # 132+231
+  )
+}
+
+x <- seq(from=0.01, to=1, by=0.01)
+plot(c(-3,0),c(-3,0), type='n',
+     xlab="log x", ylab="log-likelihood", main="(b)")
+points(log(x),log(like(x,1)), type="l", col="black")
+points(log(x),log(like(x,2)), type="l", col="red"  )
+points(log(x),log(like(x,3)), type="l", col="blue" )
+
+abline(h=0,lty=2)
+abline(h=-2)
+
+legend(x = -3, y = -0.5, lty=1, col=c("black","red","blue"),
+       legend=c("r=1", "r=2", "r=3"))
+}
+worker_123second()
+pdf(file="123second.pdf")
+worker_123second()
+dev.off()
+```
+
+Now the focal competitor comes last:
+
+```{r plot123third}
+worker_123third <- function(...){
+like <- function(x,r){
+  S <- x + x^2 + x^3
+  switch(r,
+         '1' = x^3/S*x^2/(x^1+x^2) + x^2/S*x^3/(x^1+x^3), # 213+312
+         '2' = x^1/S*x^3/(x^2+x^3) + x^3/S*x^1/(x^1+x^2), # 123+321
+         '3' = x^1/S*x^2/(x^2+x^3) + x^2/S*x^1/(x^1+x^3)  # 132+231
+  )
+}
+
+x <- seq(from=0.01, to=1, by=0.01)
+plot(c(-3,0),c(-3,0),type='n',
+     xlab="log x", ylab="log-likelihood", main="(c)")
+points(log(x),log(like(x,1)), type="l", col="black")
+points(log(x),log(like(x,2)), type="l", col="red"  )
+points(log(x),log(like(x,3)), type="l", col="blue" )
+
+abline(h=0,lty=2)
+abline(h=-2)
+
+legend(x = -3, y = -0.5, lty=1, col=c("black", "red", "blue"),
+       legend=c("r=1", "r=2", "r=3"))
+}
+
+worker_123third()
+pdf(file="123third.pdf")
+worker_123third()
+dev.off()
+```
+
+```{r allthree}
+worker_allthree <- function(...){
+op <- par(mfrow=c(1,3))
+worker_123first()
+worker_123second()
+worker_123third()
+dev.off()
+par(op)
+}
+worker_allthree()
+pdf(file="123all.pdf",height=4)
+worker_allthree()
+dev.off()
+```
+The $n=4$ case is harder as we potentially have six consistent
+orderings for each observation.  As an example, if the focal
+competitor comes third we have $1$
 

### 2025-10-13 inst/hankin-exponential_BT.tex
simple cases

@@
 If the focal competitor came second we would have two Plackett-Luce
-probabilities to add, corresponding to $\alpha x^i$ or $\alpha x^j$
-coming first where $i,j\neq r$.  For example, if $r=1$ we would have
-two finishing orders that are consistent with the observation: $2\succ
-1\succ 3$ and $3\succ 1\succ 2$.  These are disjoint so the likelihood
-will be proportional to
+probabilities to add, corresponding to different order statistics.
+$\alpha x^i$ or $\alpha x^j$ coming first where $i,j\neq r$.  For
+example, if $r=1$ we would have two finishing orders that are
+consistent with the observation: $2\succ 1\succ 3$ and $3\succ 1\succ
+2$.  These are disjoint so the likelihood will be proportional to
 
@@
   \begin{cases}
-    \frac{x  }{x+x^2+x^3}\left(\frac{x^2}{x+x^3}   + \frac{x^3}{x+x^2}\right)\qquad r=1\\
-    \frac{x^2}{x+x^2+x^3}\left(\frac{x  }{x^2+x^3} + \frac{x^3}{x+x^2}\right)\qquad r=2\\
-    \frac{x^3}{x+x^2+x^3}\left(\frac{x  }{x^2+x^3} + \frac{x^2}{x+x^3}\right)\qquad r=3
+    \frac{x^1}{x^1+x^2+x^3}\left(\frac{x^2}{x^1+x^3} + \frac{x^3}{x^1+x^2}\right)\qquad r=1\\
+    \frac{x^2}{x^1+x^2+x^3}\left(\frac{x^1}{x^2+x^3} + \frac{x^3}{x^1+x^2}\right)\qquad r=2\\
+    \frac{x^3}{x^1+x^2+x^3}\left(\frac{x^1}{x^2+x^3} + \frac{x^2}{x^1+x^3}\right)\qquad r=3
 \end{cases}
@@
 
+\begin{figure}[t]
+  \begin{centering}
+\includegraphics[width=4in]{123all}  % 123all.pdf produced by exponential_BT.Rmd
+\caption{Log-likelihood functions \label{123all} for support
+  $\in\left\lbrace 1, 2,3\right\rbrace\times\mathbb{R}$}
+\end{centering}
+\end{figure}
+
 The $n=4$ case is harder as we potentially have six consistent
 orderings for each observation.  As an example, if the focal
-competitor comes third we have $1
+competitor comes third we have $1$
 

### 2025-10-17 inst/exponential_BT.Rmd
more work

@@
 orderings for each observation.  As an example, if the focal
-competitor comes third we have $1$
+competitor comes third we have
+
+\begin{equation}
+\newcommand{pl}[4]{
+\frac{x^{#1}}{x^{#1} + x^{#2} + x^{#3} + x^{#4}}\cdot
+\frac{x^{#2}}{         x^{#2} + x^{#3} + x^{#4}}\cdot
+\frac{x^{#3}}{                  x^{#3} + x^{#4}}}
+
+\end{equation}
+
+SDAdsfDSFFDSAfdsafdsff asdf dfs dfs dfs dsff dsfds fds fds sfd s fs sd sdf sdf sdf sdf sdf sdf sdf sdfs dfsdf sdsdfsa asdf saf
+
+\begin{eqnarray}
+\mathcal{L}(x;r=1)
+\quad &=& \quad\hphantom{+}
+            \pl{2}{3}{1}{4}\\
+&{}&\quad + \pl{2}{4}{1}{3}\\
+&{}&\quad + \pl{3}{2}{1}{4}\\
+&{}&\quad + \pl{3}{4}{1}{2}\\
+&{}&\quad + \pl{4}{2}{1}{3}\\
+&{}&\quad + \pl{4}{3}{1}{2}
+\end{eqnarray}
+
+\begin{eqnarray}
+\mathcal{L}(x;r=2) 
+\quad &=& \quad\hphantom{+}
+            \pl{1}{3}{2}{4}\\
+&{}&\quad + \pl{1}{4}{2}{3}\\
+&{}&\quad + \pl{3}{1}{2}{4}\\
+&{}&\quad + \pl{3}{4}{2}{1}\\
+&{}&\quad + \pl{4}{1}{2}{3}\\
+&{}&\quad + \pl{4}{3}{2}{1}
+\end{eqnarray}
+
+and so on.
+
+```{r fourcompetitors}
+worker <- function(...){
+PL <- function(v){
+  stopifnot(length(v) == 4)
+  v[1]/(v[1]+v[2]+v[3]+v[4]) *
+  v[2]/(     v[2]+v[3]+v[4]) *
+  v[3]/(          v[3]+v[4])
+}
+
+f1 <- function(v){  # focal competitor came third, r=1
+  return(
+  PL(v^c(2,3,1,4)) +
+  PL(v^c(2,4,1,3)) +
+  PL(v^c(3,2,1,4)) +
+  PL(v^c(3,4,1,2)) +
+  PL(v^c(4,2,1,3)) +
+  PL(v^c(4,3,1,2))
+  )
+}
+f2 <- function(v){  # focal competitor came third, r=2
+  return(
+  PL(v^c(1,3,2,4)) +
+  PL(v^c(1,4,2,3)) +
+  PL(v^c(3,1,2,4)) +
+  PL(v^c(3,4,2,1)) +
+  PL(v^c(4,1,2,3)) +
+  PL(v^c(4,3,2,1))
+  )
+}
+f3 <- function(v){  # focal competitor came third, r=2
+  return(
+  PL(v^c(1,2,3,4)) +
+  PL(v^c(1,4,3,2)) +
+  PL(v^c(2,1,3,4)) +
+  PL(v^c(2,4,3,1)) +
+  PL(v^c(4,1,3,2)) +
+  PL(v^c(4,2,3,1))
+  )
+}
+f4 <- function(v){  # focal competitor came third, r=4
+  return(
+  PL(v^c(1,2,4,3)) +
+  PL(v^c(1,3,4,2)) +
+  PL(v^c(2,1,4,3)) +
+  PL(v^c(2,3,4,1)) +
+  PL(v^c(3,1,4,2)) +
+  PL(v^c(3,2,4,1))
+  )
+}
+
+o <- 1:4
+
+plot(c(-3,0),c(-3,0), type='n',
+     xlab="log x", ylab="log-likelihood", main="Focal competitor came third")
+x <- seq(from=0.01,to=1,by=0.001)
+
+points(log(x),log(sapply(x,f1)), type="l", col="black")
+points(log(x),log(sapply(x,f2)), type="l", col="red")
+points(log(x),log(sapply(x,f3)), type="l", col="blue")
+points(log(x),log(sapply(x,f4)), type="l", col="green")
+
+abline(h=0,lty=2)
+abline(h = -2, lty=1)
+
+legend("topright",lty=1,col=c("black","red","blue","green"),
+       legend=c("r=1","r=2","r=3","r=4"), bg="white")
+}
+
+worker()
+pdf(file="1234third.pdf")
+worker()
+dev.off()
+```
+
+# More stuff
 

### 2025-10-17 inst/hankin-exponential_BT.tex
more work

@@
 orderings for each observation.  As an example, if the focal
-competitor comes third we have $1$
+competitor comes third we have
+
+The $n=4$ case is harder as we potentially have six consistent
+orderings for each observation.  As an example, if the focal
+competitor comes third we have
+
+SDAdsfDSFFDSAfdsafdsff asdf dfs dfs dfs dsff dsfds fds fds sfd s fs sd
+sdf sdf sdf sdf sdf sdf sdf sdfs dfsdf sdsdfsa asdf saf
+
+\newcommand{\plackett}[4]{
+\frac{x^{#1}}{x^{#1} + x^{#2} + x^{#3} + x^{#4}}\cdot
+\frac{x^{#2}}{         x^{#2} + x^{#3} + x^{#4}}\cdot
+\frac{x^{#3}}{                  x^{#3} + x^{#4}}}
+
+\begin{eqnarray}
+\mathcal{L}(x;r=1)
+\quad &=& \quad\hphantom{+}
+            \plackett{2}{3}{1}{4}\\
+&{}&\quad + \plackett{2}{4}{1}{3}\\
+&{}&\quad + \plackett{3}{2}{1}{4}\\
+&{}&\quad + \plackett{3}{4}{1}{2}\\
+&{}&\quad + \plackett{4}{2}{1}{3}\\
+&{}&\quad + \plackett{4}{3}{1}{2}
+\end{eqnarray}
+
+%\begin{eqnarray}
+%\mathcal{L}(x;r=2) 
+%\quad &=& \quad\hphantom{+}
+%            \pl{1}{3}{2}{4}\\
+%&{}&\quad + \pl{1}{4}{2}{3}\\
+%&{}&\quad + \pl{3}{1}{2}{4}\\
+%&{}&\quad + \pl{3}{4}{2}{1}\\
+%&{}&\quad + \pl{4}{1}{2}{3}\\
+%&{}&\quad + \pl{4}{3}{2}{1}
+%\end{eqnarray}
+
+and so on.
+
+\begin{figure}[t]
+  \begin{centering}
+\includegraphics[width=4in]{1234third}  % 1234third.pdf produced by
+                                       % very_simplified_likelihood.Rmd
+\caption{Likelihood function for the observation that the focal
+  competitor came third out of four, showing
+  $r=1,2,3,4$\label{1234third}.  Value of $r$ indicated in colour by
+  its curve; dotted line shows (asymptotic) maximum likelihood
+  estimate (viz, $r=3, x\longrightarrow 0$)}
+\end{centering}
+\end{figure}
 

### 2025-10-22 inst/exponential_BT.Rmd
slight reworking

@@
 ```{r fourcompetitors}
-worker <- function(...){
 PL <- function(v){
@@
 
+worker <- function(...){
+
 o <- 1:4
@@
 abline(h=0,lty=2)
-abline(h = -2, lty=1)
+abline(h = -2, lty=1, col="gray")
 
-legend("topright",lty=1,col=c("black","red","blue","green"),
-       legend=c("r=1","r=2","r=3","r=4"), bg="white")
-}
+legend("topright",lty=1,col=c("black", "red", "blue", "green"),
+       legend=c("r=1", "r=2", "r=3", "r=4"), bg="white")
+
+f1_solve <- function(x){log(sapply(exp(x), f1)) + 2}
+f2_solve <- function(x){log(sapply(exp(x), f2)) + 2}
+f4_solve <- function(x){log(sapply(exp(x), f4)) + 2}
+
+root1 <- uniroot(f1_solve, c(-3,-0.1))$root
+root2 <- uniroot(f2_solve, c(-3,-0.1))$root
+root4 <- uniroot(f4_solve, c(-3,-0.1))$root
+
+segments(x0=root1, y0=-2.1, y1=-1.9, col="gray")
+segments(x0=root2, y0=-2.1, y1=-1.9, col="gray")
+segments(x0=root4, y0=-2.1, y1=-1.9, col="gray")
+
+print(c(root1, root2, root4))
+} # worker() def closes
 
@@
 dev.off()
+
 ```
+
+# Try 7 competitors, focal competitor comes third
+
+```{r}
+
+library(partitions)
+perms(7) -> x
+## first r=1
+
+jj <- x[,x[3,] == 1]
 

### 2025-10-22 inst/hankin-exponential_BT.tex
slight reworking

@@
 \quad &=& \quad\hphantom{+}
-            \plackett{2}{3}{1}{4}\\
-&{}&\quad + \plackett{2}{4}{1}{3}\\
-&{}&\quad + \plackett{3}{2}{1}{4}\\
-&{}&\quad + \plackett{3}{4}{1}{2}\\
-&{}&\quad + \plackett{4}{2}{1}{3}\\
+            \plackett{2}{3}{1}{4}\nonumber\\
+&{}&\quad + \plackett{2}{4}{1}{3}\nonumber\\
+&{}&\quad + \plackett{3}{2}{1}{4}\nonumber\\
+&{}&\quad + \plackett{3}{4}{1}{2}\nonumber\\
+&{}&\quad + \plackett{4}{2}{1}{3}\nonumber\\
 &{}&\quad + \plackett{4}{3}{1}{2}
@@
 \includegraphics[width=4in]{1234third}  % 1234third.pdf produced by
-                                       % very_simplified_likelihood.Rmd
-\caption{Likelihood function for the observation that the focal
-  competitor came third out of four, showing
+                                        % exponential_BT.Rmd
+\caption{Support (log-likelihood) function for the observation that
+  the focal competitor came third out of four, showing
   $r=1,2,3,4$\label{1234third}.  Value of $r$ indicated in colour by
   its curve; dotted line shows (asymptotic) maximum likelihood
-  estimate (viz, $r=3, x\longrightarrow 0$)}
+  estimate, viz $r=3, x\longrightarrow 0$.  Gray line at $y=-2$ shows
+  the two-units-of-support limit; solutions of $\mathcal{S}=-2$
+  indicated by short vertical gray lines.  For example, looking at the
+  red line, we may reject the hypothesis $r=2,
+  x\leqslant\exp(-1.93)\simeq 0.147$ (in favour of $r=3, x=0^+$).
+  Observe that all four lines meet at $x=1$, at which point all four
+  competitors have identical Plackett-Luce strengths and this
+  observation is uninformative as to the value of $r$}
 \end{centering}

### 2025-11-11 inst/exponential_BT.Rmd
more work

@@
 library(stringr)
+library("partitions")
+library("magrittr")
 set.seed(0)
@@
 
-# Try 7 competitors, focal competitor comes third
+# Try 8 competitors, focal competitor comes third
 
-```{r}
+```{r calconetoeight, cache=TRUE}
+pl <- function(v){
+  out <- 1
+  n <- length(v)
+  for(i in seq_len(n-1)){
+    out <- out*v[i]
+    out <- out/sum(v[seq(from=i, to=n)])
+  }
+  return(out)
+}
 
-library(partitions)
-perms(7) -> x
-## first r=1
+f <- function(x, place, no_of_comp){
+    M <- perms(no_of_comp)
+    out <- rep(0, no_of_comp)
+    for(i in seq_len(ncol(M))){
+        out[M[place,i]] %<>% `+`(pl(x^M[,i]))
+    }
+    return(out)
+}
 
-jj <- x[,x[3,] == 1]
+n <- 8
+place <- 3
+x <- seq(from = 0.1, to=1, by=0.01)
+L <- matrix(0, length(x), n)
+for(i in seq_along(x)){
+    L[i,] <- f(x[i], place, n)
+}
+```
+
+```{r onetoeight}
+worker <- function(...){
+    matplot(log(x),log(L), type='l', lty=1,ylim=c(-4,0))
+}
+pdf(file = "onetoeight.pdf")
+worker()
+dev.off()
+worker()
+```
 
@@
 
-```{r definegetprob, cache=TRUE}
+```{r definegetprob, caasdf=TRUE}
 getprob <- function(n, r, x, N=1e4){

### 2025-11-11 inst/hankin-exponential_BT.tex
more work

@@
 
-The $n=4$ case is harder as we potentially have six consistent
-orderings for each observation.  As an example, if the focal
-competitor comes third we have
-
-The $n=4$ case is harder as we potentially have six consistent
-orderings for each observation.  As an example, if the focal
-competitor comes third we have
-
-SDAdsfDSFFDSAfdsafdsff asdf dfs dfs dfs dsff dsfds fds fds sfd s fs sd
-sdf sdf sdf sdf sdf sdf sdf sdfs dfsdf sdsdfsa asdf saf
+The general case with $n$ competitors is harder as we potentially have
+$(n-1)!$ orderings for each observation.  Taking $n=4$ as an example,
+if the focal competitor comes third we have
 
@@
 
-and so on.
+and so on; see figure~\ref{1234third}.
 
@@
 
-\subsection{Simplified Plackett-Luce likelihoods}
+\subsection{The case $x=1$}
 
 As noted, likelihood function~\ref{plackettluce} becomes unmanageable
-with $n$ even moderately large and some form of simplification is
-needed.  To that end, consider a Plackett-Luce likelihood function for
+with~$n$ even moderately large and we may gain additional insight by
+considering a limiting case.  To that end, if $x=1$, all competitors
+have the same Plackett-Luce strength.
+
+To that end, consider a Plackett-Luce likelihood function for
 a race between a competitor of Bradley-Terry strength $a$ and
@@
 focal competitor, and because we have $n$ competitors $1\leqslant
-r\leqslant n$.  The initial field strength is $a+(n-1)b=1+(n-2)b$.  The
-Plackett--Luce likelihood function~\citep{luce1959,plackett1975} for
-this situation would be
+r\leqslant n$.  The initial field strength is $a+(n-1)b=1+(n-2)b$.
+The Plackett--Luce likelihood function~\citep{luce1959,plackett1975}
+for this situation would be
 

### 2025-11-24 inst/hankin-exponential_BT.tex
remove work on cloned identical nonfocal competitors

@@
 
-\subsection{The case $x=1$}
-
-As noted, likelihood function~\ref{plackettluce} becomes unmanageable
-with~$n$ even moderately large and we may gain additional insight by
-considering a limiting case.  To that end, if $x=1$, all competitors
-have the same Plackett-Luce strength.
-
-To that end, consider a Plackett-Luce likelihood function for
-a race between a competitor of Bradley-Terry strength $a$ and
-$n-1\geqslant 1$ cloned competitors, each of strength $b=1-a$.  Under
-these circumstances, an observation is indexed by~$r$, the rank of the
-focal competitor, and because we have $n$ competitors $1\leqslant
-r\leqslant n$.  The initial field strength is $a+(n-1)b=1+(n-2)b$.
-The Plackett--Luce likelihood function~\citep{luce1959,plackett1975}
-for this situation would be
-
-\begin{eqnarray}\label{likeforrn1}
-  \mathcal{L}_{r,n}(a) &=&
-\frac{b}{1+ (n-2)b}\cdot
-\frac{b}{a+(n-3)b}\cdots\frac{a}{a+(n-r-1)b}\nonumber\\
-&=& \frac{B-1}{(B+n-2)\cdots(B+n-r-1)}\nonumber\\ 
-&=& \frac{(B-1)(B+n-r-2)!}{(B+n-1)!}\label{likeforrn3}
-\end{eqnarray}
-
-\noindent where $B=1/b$.  Note carefully that these expressions are
-likelihoods, not probabilities, and do not sum to 1.  Previous related
-work \citep{hankin2024_hyper3} presented the case $n=3$ [in current
-  notation] but here arbitrary $n$ may be considered.  Taking the case
-$n=9$, for example, gives us 9 distinct likelihood functions,
-Figure~\ref{ninelikes}.  We see an unanticipated asymmetry between $r$
-and $n-r$.  For $r=5$ one might expect that $\hat{a}=\frac{1}{2}$ on
-the grounds that the focal competitor's performance was median.
-However, $\hat{a}\simeq 0.5886$.  Figure~\ref{dotprobs} shows the
-maximum likelihood estimator $\hat{a}$ for various values of $n,r$,
-obtained by direct numerical maximization, in the R programming
-language \cite{rcore2024}, of equation~\ref{likeforrn3}.
-
-\begin{figure}[t]
-  \begin{centering}
-\includegraphics[width=4in]{ninelikes}  % ninelikes.pdf produced by very_simplified_likelihood.Rmd
-\caption{Likelihood functions for $r=1,2,\ldots 9$\label{ninelikes}.  Value
-  of $r$ indicated in colour by its curve; vertical line shows
-  $\hat{a}\simeq 0.5886$ for the case $r=5$}
-\end{centering}
-\end{figure}
-
-\subsection{Asymptotic analysis}
-
-Even though expression ~\ref{likeforrn3} represents a simplification
-of the Plackett-Luce likelihood function, it becomes intractable even
-for moderate values of $n$ and $r$.  Analysis of
-expression~\ref{likeforrn3} for large $n$ and $r=pn$ for fixed $p$
-shows that the support $\mathcal{S}=\log\mathcal{L}_{r,n}(a)$ is
-asymptotically
-
-\begin{equation}\label{asymptotic}
-\mathcal{S}=
-%\log\mathcal{L}_{r,n}(a)=
-\log(B-1) + B\log(1-r/n)
-+\mathcal{O}\left(n^{-1}\right)
-\end{equation}
-
-(up to an arbitrary additive constant).  Using
-$\mathcal{S}^*=\log(B-1)+B\log(1-r/n)$ as an approximate support thus
-incurs an error of only $\mathcal{O}(n^{-1})$.  Observing that
-$\partial^2\mathcal{S}^*/\partial B^2=-(B-1)^{-2}<0$, the evaluate
-would be unique; we have $\hat{B}=1-\log(1-r/n)^{-1}$, or
-alternatively $\hat{a}=(1-\log(1-r/n))^{-1}$.  Compare this with the
-result of asking ``what is the probability that the focal competitor
-beats a randomly chosen opponent?'' to which the binomial maximum
-likelihood estimate is $\frac{r-1}{n-1}$.  For repeated independent
-observations $r_i,n_i$, $1\leqslant i\leqslant N$ we would have
-
-\begin{equation}
-  \hat{a} =   \frac{1}{1-\sum\log(1-r_i/n_i)/N};
-\end{equation}
-
-compare the binomial result of $\sum(r_i-1)/\sum n_i$.  Below, I
-present three examples that showcase likelihood
-function~\ref{likeforrn3}, drawn from the fields of athletics,
-education, and Formula~1 motor racing.
-
-\begin{figure}[t]
-  \begin{centering}
-\includegraphics[width=4in]{dotprobs}  % produced by very_simplified_likelihood.Rmd
-\caption{Maximum likelihood estimator $\hat{a}$ for $n=2(1)10$
-  and\label{dotprobs}} $r=1(1)n$
-\end{centering}
-\end{figure}
-
-% \subsection{Olympic track and field athletics}
-% 
-% One long-standing contention in the athletic world is that of lane
-% bias~\citep{munro2022}.  Running in the inner lane is frequently held
-% to be a disadvantage; mechanisms include biomechanical issues arising
-% from centrifugal acceleration, to competitors' inability to see their
-% rivals.  Here attention is restricted to heats, for which lane placing
-% is random.  I consider every heat in Men's 100m Olympic running from
-% 2004--2020 (a total of 41 heats, each of 8 or 9 competitors); only
-% competitors' ranks are used.  Figure~\ref{plotolymp} shows a
-% log-likelihood curve for the Bradley--Terry strength of the focal
-% competitor, here the runner in the innermost lane.  We see that the
-% evaluate of $0.511$ is very close to the neutral value of $0.5$, for
-% which the support is only $0.029$, far short of the standard two units
-% of support criterion for rejection~\citep{edwards1972}.  This dataset
-% contains no evidence to support lane bias.
-% 
-% \begin{figure}[t]
-%   \begin{centering}
-% \includegraphics[width=4in]{plotolymp} % produced by very_simplified_likelihood.Rmd
-% \caption{Log-likelihood curve for the Bradley--Terry strength of the
-%   lane-1 runner, Olympic qualifying heats 2012-2020 \label{plotolymp}}
-% \end{centering}
-% \end{figure}
-
-\subsection{Park run}
-
-``Parkrun'' is a distributed community initiative that organises
-weekly timed 5 km runs/walks in parks worldwide~\citep{hindley2020}.
-A typical event will have 200 participants.  Table~\ref{parkruntable}
-shows the author has completed a total of 21 parkruns to date,
-comprising 9 runs in 2023 and 12 in 2024.  From the first pair of
-numbers we see that 259 runners attended that particular parkrun in
-2023, of whom the author placed 177.
-
-\begin{table}[t]
-  \centering
-  \caption{Parkrun results}
-\label{parkruntable}
-\begin{tabular}{cccccccccc}\\
- & \multicolumn{9}{c}{2023}\\
-rank   & 177& 222& 206& 142& 118& 224& 128& 115& 183\\
-runners& 259& 305& 297& 241& 179& 338& 203& 245& 254\\ \\
-\end{tabular}
-\begin{tabular}{ccccccccccccc}
-  & \multicolumn{12}{c}{2024}\\
-rank      &  145& 157& 156& 191& 172& 174& 173& 165& 172& 199& 181& 229\\
-runners   &  193& 240& 197& 257& 227& 234& 238& 238& 196& 242& 242& 318\\
-\end{tabular}
-\end{table}
-
-Is there evidence that the author's performance differs between years?
-Figure~\ref{parkruncontour} shows contours of support [again
-  normalised so the support of the evaluate is zero] for the parkrun
-dataset, as a function of 2023 strength on the horizontal axis and
-2024 strength on the vertical axis.  A null of equal strengths appears
-as the diagonal line.  We see that the maximum likelihood estimate is
-$\hat{p}_\mathrm{2023}=0.413$, $\hat{p}_\mathrm{2024}=0.485$.
-However, a constrained optimization shows that the maximum support
-along the null diagonal is about $-0.22$, at
-${p}_\mathrm{2023}={p}_\mathrm{2024}\simeq 0.441$.  The two units of
-support criterion~\citep{edwards1992} is not met, and we conclude that
-there is no evidence for a change in competitive strength between the
-years.  Now consider table~\ref{fisherparkrun}, giving a summary of
-whether nonfocal competitors place ahead of the author, split by year.
-A two-sided Fisher test gives a $p$-value of about $5\times 10^{-14}$,
-surely an unreasonable conclusion given the author's perceived
-performance stagnation.
-
-\begin{figure}[t]
-  \begin{centering}
-\includegraphics[width=4in]{parkruncontour} % produced by very_simplified_likelihood.Rmd
-\caption{Support contours for Bradley-Terry strength
-  for \label{parkruncontour} 2023 and 2024.  Null of equal strengths
-  shown as a diagonal line}
-\end{centering}
-\end{figure}
-
-\begin{table}[h]
-    \begin{tabular}{ l c r }
-       & \multicolumn{2}{c}{Faster?}\\
-       & yes  & no  \\
-  2023 & 1506 & 806 \\
-  2024 & 2102 & 708
-    \end{tabular}
-    \caption{Two-way contingency\label{fisherparkrun} table}
-\end{table}
-
-\subsection{Education}
-
-In educational research, confidentiality of examination scores is
-often an issue, and individual results are generally regarded as
-sensitive and private information.  Nevertheless, it is common for a
-student to possess some informative observations, for example the
-total enrollment in a specific class and their individual ranking
-within that class.  Table~\ref{educationtable} shows an apocryphal
-dataset: what could a student infer from such data?
-
-\begin{table*}[t]
-  \caption{Educational dataset}
-\label{educationtable}
-\begin{tabular}{llll}
-\hline
-                   course&category   &rank&class size\\ \hline
-      rings and modules  &  pure     & 9  &      12\\
-           group theory  &  pure     & 7  &      17\\
-               calculus  & applied   & 2  &      23\\
-         linear algebra  & applied   & 3  &       9\\
- differential equations  & applied   & 2  &      13\\
-               topology  &   pure    & 8  &      14\\
-     special relativity  & applied   & 3  &      13\\
-        fluid mechanics  & applied   & 4  &      12\\
-            Lie algebra  &   pure    & 9  &      15\\
-     \hline
-\end{tabular}
-\end{table*}
-
-One might wonder whether the student was better at pure or applied
-mathematics and, if so, to quantify the difference.
-Figure~\ref{plotpureandapplied} shows contours of support [again
-  normalised so the support of the evaluate is zero] for the dataset,
-as a function of pure strength on the horizontal axis and applied
-strength on the vertical axis.  A null of equal strengths appears as
-the diagonal line.  We see that the maximum likelihood estimate is
-$\hat{p}_\mathrm{pure}=0.542$, $\hat{p}_\mathrm{applied}=0.852$.
-However, a constrained optimization shows that the maximum support
-along the null diagonal is about $-2.06$, at
-${p}_\mathrm{pure}={p}_\mathrm{applied}\simeq 0.683$.  The two units
-of support criterion~\citep{edwards1992} is met, and we are justified
-in concluding that the student indeed has a higher ranking in the
-applied modules than the pure modules.  Alternatively, we may observe
-that $2\mathcal{S}=4.11$ is in the tail region of its asymptotic
-$\chi^2_1$ distribution, corresponding to a $p$-value of about
-$0.042$.
-
-\begin{figure}
-\begin{centering}
-\includegraphics[width=4in]{plotpureandapplied}  % plotpureandapplied.pdf produced by very_simplified_likelihood.Rmd
-\caption{Contours of equal support for pure (horizontal) and applied
-  (vertical) strength; null of equal strengths shown as a diagonal
-  line \label{plotpureandapplied}}
-\end{centering}
-\end{figure}
-
-\subsection{Formula 1}
-
-Formula 1 motor racing is an important and prestigious motor sport
-\citep{codling2017,jenkins2010}.  Season ranking is based on a points
-allocation system wherein competitors are awarded points based on race
-finishing order; points accumulate additively.  The overall
-competition winner is the competitor who accumulates the most points
-after the final race.  However, as argued by
-\cite{hankin2023_formula1points}, the drivers' {\em ranks} are
-statistically sufficient for analysis using Plackett--Luce approaches.
-The methods presented here may be used to assess competitive strengths
-of Formula 1 racing driver Sergio P\'{e}rez (``Checo'') over his
-career from 2011 to 2023.  Checo's Formula 1 performance began
-somewhat inauspiciously as second driver for Sauber for whom he
-finished no higher than P7 in 2011, but his performance rose
-subsequently, culminating in his achievement of a startling 20 podium
-finishes in 2022--3.  Noting that Checo raced against 73 distinct
-competitors---this rendering estimation of the other players'
-Plackett--Luce strengths effectively
-impossible~\citep{hankin2020}---we are nevertheless in a position to
-consider competitive strength using the modified likelihood approach
-presented here.  We consider the following inference problem:
-
-\begin{equation}
-  a = a(t) = \frac{e^{\alpha + \beta t}}{1+e^{\alpha + \beta t}},\qquad\alpha,\beta\in\mathbb{R}
-\end{equation}
-
-(that is, a logistic regression of generalized Bradley--Terry strength
-$a$ against time $t$, with $\alpha$ and $\beta$ being linear
-coefficients).  Contours of equal support are shown in
-Figure~\ref{showchecolike}, again normalized so that
-$\mathcal{S}(\hat{\alpha},\hat{\beta})=0$.  We see the expected
-elliptical contours, and further $H_0\colon\beta=0$ may be rejected in
-favour of $\alpha=-0.27$, $\beta=0.081$, on the grounds that
-$\mathcal{S}(\alpha,0)\leqslant 9.2$.  Alternatively we may observe
-that $2\mathcal{S}=18.4$ would correspond to a $p$-value of about
-$1.7\times 10^{-5}$ on its asymptotic $\chi^2_1$ null distribution.
-
-\begin{figure}[t]
-  \begin{centering}
-\includegraphics[width=4in]{showchecolike}  % showchecolike.pdf produced by very_simplified_likelihood.Rmd
-\caption{Log-likelihood contours for $\alpha,\beta$ for Checo's racing
-  career 2011--2023\label{showchecolike}}
-\end{centering}
-\end{figure}
-
 \section{Conclusions and further work}

### 2025-11-24 inst/exponential_BT.Rmd
more work

@@
 
-```{r}
+```{r plottoycontour}
 x_try
@@
 support  <- matrix(support,length(r_try),length(x_try))
-contour(r_try, x_try, support, xlab="r", ylab="x", levels = -c(1,(1:5)*2), lwd=c(1,3,1,1,1,1))
-abline(v=1:9, lwd=0.3, col='gray')
-abline(v = r_true)
-abline(v = 1:9, col = 'gray', lty = 3, lwd = 0.3)
+wanted <- x_try >= 0.5
+worker <- function(...){
+contour(r_try, x_try[wanted], support[,wanted], xlab="r", ylab="x", levels = -c(1,(1:5)*2), lwd=c(1,3,1,1,1,1))
+points(r_true,x_true,pch=16,col='red',cex=2)
+jj <- support[,wanted]
+maxplace <- which(jj==max(jj), arr.ind=TRUE)
+points(r_try[maxplace[1]], x_try[wanted][maxplace[2]], pch=16, col='blue', cex=2)
+}
+worker()
+pdf(file="toyexample.pdf")
+worker()
+dev.off()
+
 filled.contour(r_try, x_try, support, xlab = "r", ylab = "x")
@@
 S <- pmax(S,-15)
-contour(S,levels = -(0:6)*2, lwd=c(1,4,1,1,1,1,1))
+contour(a_try, x_try, S, levels = -(0:6)*1, lwd=c(1,4,1,1,1,1,1))
+
+worker <- function(...){
+    wanted <- a_try < 0.5
+    contour(a_try[wanted], x_try, S[wanted,], levels = -(0:6)*2, lwd=c(1,4,1,1,1,1,1))
+    jj <- S[wanted,]
+    m <- which(jj == max(jj), arr.ind=TRUE)
+    points(a_try[wanted][m[1]], x_try[m[2]], pch=16, cex=2, col='red')
+}
+worker()
+pdf(file="formula1.pdf")
+worker()
+dev.off()
 filled.contour(S)

### 2025-11-24 inst/hankin-exponential_BT.tex
more work

@@
 
-\begin{equation}
+\begin{equation}\label{aftersim}
   \mathcal{L}(x,r)=
@@
 &{}&\quad + \plackett{4}{2}{1}{3}\nonumber\\
-&{}&\quad + \plackett{4}{3}{1}{2}
+&{}&\quad + \plackett{4}{3}{1}{2}\label{requalfour}
 \end{eqnarray}
@@
 
+\section{The case of unknown $r$}
+
+Although figure~\ref{1234third} shows a perfectly respectable
+log-likelihood function, in practice we often do not know the value of
+$r$, and need a joint likelihood function for $(r,x)$.  It is
+convenient to normalize $r$ and use $s=r/n$ for consistency between
+different values of $n$, and to treat $s$ as a continuous variable.
+The only reasonable way to calulate such a likelihood function appears
+to be numerical; fortunately this is straightforward.
+
+We leverage the {\tt hyper2} R package~\cite{hankin2017_rnw} which
+includes extensive numerical functionality to simulate Plackett-Luce
+situations, and in this case function {\tt rrace()} simulates a race
+between competitors of known Plackett-Luce strengths.  To obtain a
+likelihood function for a specific value of $(x,s)$ at a particular
+observed rank $r$, we run repeated {\em in-silico} races between
+competitors of Plackett-Luce strengths $x^1,x^2,\ldots x^n$, and
+observe the frequency with which competitor $\lceil ns\rceil$ places
+$r$.  For example, consider $x=0.1, r=3$.  We might simulate this as
+follows:
+
+\label{rsession}
+\begin{verbatim}
+n <- 9  # number of competitors
+s <- 4  # observation: focal competitor places fourth
+
+v <- seq_len(n)
+rrank(15, p=0.1^v, as.character(v))
+#> A ranktable:
+#>       c1 c2 c3 c4 c5 c6 c7 c8 c9
+#>  [1,] 2  1  4  3  5  6  7  8  9 
+#>  [2,] 1  2  3  5  4  6  7  9  8 
+#>  [3,] 1  2  3  4  5  6  7  8  9 
+#>  [4,] 1  2  3  4  5  6  7  8  9 
+#>  [5,] 1  2  4  3  5  6  7  9  8 
+#>  [6,] 1  2  3  4  5  6  7  8  9 
+#>  [7,] 1  3  2  4  5  6  7  8  9 
+#>  [8,] 1  2  3  4  5  7  6  8  9 
+#>  [9,] 3  1  2  5  4  9  6  7  8 
+#> [10,] 1  2  3  4  5  6  7  8  9 
+#> [11,] 1  2  3  4  5  6  7  8  9 
+#> [12,] 1  2  3  5  4  6  7  9  8 
+#> [13,] 1  3  2  4  5  6  7  8  9
+#> [14,] 1  2  3  4  5  7  6  8  9 
+#> [15,] 1  2  5  4  4  6  8  7  9 
+\end{verbatim}
+
+Above we see 15 independent Plackett-Luce races; for example the first
+one indicates that competitor 2 came first, 1 came second, 4 came
+third, and so on.  Looking at the fourth column [headed \verb+c4+,
+  ``came fourth''], we see that
+
+\begin{equation}
+  \mathcal{L}(0.1,3)\propto 1,\qquad
+  \mathcal{L}(0.1,4)\propto 11,\qquad
+  \mathcal{L}(0.1,5)\propto 3
+\end{equation}
+
+The maximum likelihood estimate for $s$ would therefore be $4/9\simeq
+0.44$, conditional on $x=0.1$.  To estimate $x$ we would need repeated
+observations, and here we note that small values of $x$ lead to
+underdispersed results.
+
+To validate the suggested likelihood method, we set $x=0.85$ and
+$r=3$, that is, competitor $i$ has strength $\propto 0.85^i,
+1\leqslant i\leqslant 9$; the focal competitor has the third strongest
+Plackett-Luce strength of $0.85^3\simeq 0.61$.  The results of running
+{\tt rrace()} 100 times with $n=9$ are given in
+table~\ref{results100insilico}.
+
+\begin{table}
+  \centering
+  \begin{tabular}{llllllllll}
+    place & 1  & 2  &  3 &  4 &  5 &  6 &  7 & 8 &  9  \\
+    count & 14 & 13 & 18 & 12 & 13 &  9 & 11 & 8 &  2
+  \end{tabular}
+  \caption{Results of repeated \label{results100insilico}
+    Plackett-Luce racing with $x=0.85, r=3$; thus the focal competitor
+    is the third strongest.  We see that the focal competitor placed
+    first 14 times, second 13 times, and so on}
+\end{table}
+
+Now, we can run repeated simulations like those on page~\pageref{rsession}, but
+with $n=10^4$ trials; as support we use $x$ running from 0.1 to 1 and
+$r=1,2,\ldots 9$: figure~\ref{toyexample}.  We see reasonable
+agreement between the true values of $x$ and $r$ and the estimates.
+Note the odd shape of the contours and the non-independence of the
+parameters.
+
+\begin{figure}[t]
+  \begin{centering}
+\includegraphics[width=4in]{toyexample}  % toyexample.pdf produced by exponential_BT.Rmd
+\caption{Contours of log-likelihood for $x$ and $r$.  Red point shows
+  true values of parameters, blue dot shows maximum likelihood\label{toyexample}
+  estimate}
+\end{centering}
+\end{figure}
+
+\section{Formula 1}
+
+Formula 1 racing is an important and prestigious motor sport
+\citep{codling2017,jenkins2010}.  In a typical season, $\sim 20$
+competitors race at $\sim 20$ venues and here we consider the 2023
+season, using Lewis Hamilton as the focal competitor.  His results are
+shown in table~\ref{lewishamilton}; Qatar and the US are disregarded
+as he did not finish.
+
+\begin{table}
+  \centering
+  \begin{tabular}{lllllllll}
+venue & BHR & SAU & AUS & AZE & MIA & MON & ESP & CAN\\
+place & 5   & 5   & 2   & 6   & 6   & 4   & 2   & 3  \\ \\
+venue & AUT & GBR & HUN & BEL & NED & ITA & SIN & JPN\\ 
+place & 8   & 3   & 4   & 4   & 6   & 6   & 3   & 5  \\ \\
+venue & MXC & SAP & LVG & ABU \\
+place & 2   & 8   & 7   & 9
+\end{tabular}
+  \caption{Lewis Hamilton's results for the 2023 season. \label{lewishamilton}
+    We see that he finished fifth in Bahrain, fifth in Saudi Arabia, second in Austria, and so on}
+\end{table}
+
+Again using numerical simulation, amounting to repeated evaluation of
+generalised version of equations \ref{aftersim} and \ref{requalfour}
+(this time with $b=10^4$ trials per point), we obtain
+figure~\ref{formula1}.  From this, we see that $r$ is less than about
+0.25 and $x$ is between about 0.4 and 0.8.  Recall that, in this
+model, $x=0.5$ implies that the Plackett-Luce strength of each
+competitor is about twice that of the next.
+
+\begin{figure}[t]
+  \begin{centering}
+\includegraphics[width=4in]{formula1}  % formula1.pdf produced by exponential_BT.Rmd
+\caption{Contours of log-likelihood functions \label{formula1}.  Red circle shows
+  maximum likelihood estimate}
+\end{centering}
+\end{figure}
+
 \section{Conclusions and further work}
 
-Here a simplification of the Plackett--Luce likelihood functions for
+Here a simplification of the Plackett-Luce likelihood functions for
 order statistics was proposed, some simple properties obtained, and

### 2025-11-25 inst/skating.Rmd
soft link for hyper2.bib
typo

@@
 To cite the `hyper2` package in publications, please use
-@hankin2017_rmd).  This document analyses Bradley-Terry strengths of
+@hankin2017_rmd.  This document analyses Bradley-Terry strengths of
 the competitors in the Ladies' Figure Skating at the 2002 Winter

### 2025-11-25 inst/hankin-exponential_BT.tex
instructions for auxiliary files not under version control
more writing

@@
-\documentclass[article]{ajs}
+\documentclass[article]{ajs}  % need to download ajs.cls, oesg.png, ajs.bst from the Austrian Journal of Statistics github repo
 \RequirePackage{amsthm,amsmath,amsfonts,amssymb}
@@
 even moderately large: even with as few as 20 competitors following
-Zipf's law~\citep{zipf1949} [a typical assumption for Bradley--Terry
+Zipf's law~\citep{zipf1949} [a typical assumption for Bradley-Terry
   strengths] and say 20 independent complete rank observations,
@@
 
+In many situations, we are not interested in the $n-1$-dimensional
+simplex of BT strengths, viz $\left\lbrace(p_1,\ldots,p_n)\colon
+p_i\geqslant 0,\sum p_i=1\right\rbrace$, but instead wish to make
+inferences about one particular entity.  For example, we might be
+interested in whether Mercedes's apparent decline in performance in
+the 2023-4 season could be due to chance.  Further examples are given
+in previous analysis by the same author \citep{hankin2010,
+  hankin2017_rmd, hankin2020, hankin2023_formula1points}.  In these
+cases, we are interested in one of the competitors and assess a number
+of hypotheses about that competitor; the strengths of the others are
+nuisance parameters.  We designate this entity as the {\em focal
+ competitor}.
+
+If we have $n$ competitors, optimizing the Bradley--Terry model over
+$n-1$ degrees of freedom is, as noted above, a formidable task even
+for moderate $n$; this suggests that some form of dimension-reducing
+parametrization might be useful.  One option might be to assign the
+focal competitor a BT strength proportional to $a\in [0,1]$ and
+$b=1-a$ to each of $n-1$ competitors; the normalization constant would
+be $n+(n-1)a$.  However, this option does not seem to be
+realistic~\citep{kuenzer2025}.
+
+The next natural suggestion would be to have $n$ competitors with
+strengths $\beta, \beta x,\ldots \beta x^{n-1}$ with $0 < x\leqslant
+1$.  The unit sum constraint gives $\beta=\frac{1-x}{1-x^n}$ but this
+constant of proportionality cancels out in Bradley-Terry likelihoods.
+The focal competitor has strength $\beta x^{r}$ for some (unknown) $r$
+with $1\leqslant r\leqslant n$.  Thus $r=1$ means that the focal
+competitor is the strongest among the $n$, $r=2$ means he is the
+second, and so on.
+
+\subsection{Theoretical results}
+
+A Plackett-Luce observation is of the form $p_{(1)}\succ
+p_{(2)}\succ\cdots\succ p_{(n)}$, where $p_{(i)}$ denotes the
+competitor in place $i $ for $1\leqslant i\leqslant n$.  If competitor
+$j$ has strength $\beta x^{j}$, then the probability of observing
+$p_1\succ p_2\succ\cdots\succ p_n$ will be, according to
+equation~\ref{plackettluce},
+
+\begin{equation}
+f(x)=\frac{1}{1+\cdots+x^{n-1}}\cdot
+\frac{1}{1+\cdots+x^{n-2}}\cdots
+\frac{1}{1+x^{2}}\cdot\frac{1}{1+x}=
+\frac{1}{[n]_x!}
+\end{equation}
+
+(here $[n]_x!$ is the $q$-factorial).  We see that
+$f(1)=\frac{1}{n!}$, $f'(x)>0$ and $f(0)=1$; so small $x$ corresponds
+to a high probability of observing the ``natural order'' in which
+competitor $j$ places $j$-th, $1\leqslant j\leqslant n$.  Further, we
+observe that $f(x)=1-(n-1)x + {\mathcal O}(x^2)$.  It is interesting
+to note that the probability of observing the reverse order, that is
+$p_n\succ p_{n-1}\succ\cdots\succ p_1$, is $f(1/x)$ for $x>0$.
+
 \subsection{Some simple cases}
@@
 we have Plackett-Luce strengths $\alpha x,\alpha x^2,\alpha x^3$ [the
-  extra factor of $x$ allows us to identify the $r$ in ``$\alpha x^r$''
-  with the rank of the competitor, the best being rank 1].  Suppose
-the focal competitor came first, the likelihood function would be
+  extra factor of $x$ allows us to identify the $r$ in ``$\alpha
+  x^r$'' with the nominal rank of the competitor, the best being rank
+  1].  Suppose the focal competitor came first, the likelihood
+function would be
 
 \begin{equation}
-  \mathcal{L}(x,r)=\frac{x^{r-1}}{1+x+x^2}\qquad r=1,2,3
+  \mathcal{L}(x,r)=\frac{x^{r-1}}{1+x+x^2}\qquad r=1,2,3.
 \end{equation}
@@
 probabilities to add, corresponding to different order statistics.
-$\alpha x^i$ or $\alpha x^j$ coming first where $i,j\neq r$.  For
-example, if $r=1$ we would have two finishing orders that are
+For example, if $r=1$ we would have two finishing orders that are
 consistent with the observation: $2\succ 1\succ 3$ and $3\succ 1\succ
@@
 
-We leverage the {\tt hyper2} R package~\cite{hankin2017_rnw} which
+We leverage the {\tt hyper2} R package~\citep{hankin2017_rmd} which
 includes extensive numerical functionality to simulate Plackett-Luce
@@
 
-Above we see 15 independent Plackett-Luce races; for example the first
-one indicates that competitor 2 came first, 1 came second, 4 came
-third, and so on.  Looking at the fourth column [headed \verb+c4+,
-  ``came fourth''], we see that
+Above we see a bespoke {\tt hyper2} object, of class {\tt ranktable},
+designed to display rank data for use with Bradley-Terry likelihoods.
+We see 15 independent Plackett-Luce races; for example the first one
+indicates that competitor 2 came first, 1 came second, 4 came third,
+and so on.  Looking at the fourth column [headed \verb+c4+, ``came
+  fourth''], we see that
 
@@
 observations, and here we note that small values of $x$ lead to
-underdispersed results.
+underdispersed results, for the focal competitor would come $r$-th
+with probability $\geqslant x^1+x^2+\ldots x^r$
 

### 2025-11-27 inst/exponential_BT.Rmd
note on time to process
analysis of skating

@@
 ```
+
+<font size="1">
+(takes maybe two or three hours to run without cache; each
+<tt>parkrunlikefunobs</tt> chunk takes maybe thirty minutes).
+</font>
 
@@
 ```
+
+# Skating
+
+```{r, startskating, cache=TRUE}
+sk <- skating
+pnames(sk) <- names(sort(skating_maxp, decreasing=TRUE))
+maxp(sk)
+```
+
+```{r moreskating, cache=TRUE}
+skater_strengths <- function(x){
+    p <- BT(23, x)
+    names(p) <- pnames(sk)
+    return(p)
+}
+skater_strengths(0.5)
+```
+
+```{r doskating, cache=TRUE}
+gvec <- seq(from=0.5, by=0.01, to=0.65)
+f <- function(g){loglik(skater_strengths(g), sk)}
+L <- sapply(gvec, f)
+L <- L-max(L)
+```
+
+```{r}
+plot(gvec,L,type="b")
+abline(h = c(0,-2))
+grid()
+```
+
+```{r testskating, cache=TRUE}
+skater_strengths(0.57)
+skating_test <- knownp.test(sk, skater_strengths(0.57))
+skating_test
+```
+
+```{r}
+xn <- skating_test$null_estimate
+xa <- skating_test$alternative_estimate
+xn <- log10(xn)
+xa <- log10(xa)
+par(pty="s")
+plot(xn,xa,xlim=c(-6,0),ylim=c(-6,0))
+abline(0,1)
+```

### 2025-11-27 inst/hankin-exponential_BT.tex
typo
more work

@@
 p_{(2)}\succ\cdots\succ p_{(n)}$, where $p_{(i)}$ denotes the
-competitor in place $i $ for $1\leqslant i\leqslant n$.  If competitor
+competitor in place $i$ for $1\leqslant i\leqslant n$.  If competitor
 $j$ has strength $\beta x^{j}$, then the probability of observing
-$p_1\succ p_2\succ\cdots\succ p_n$ will be, according to
+$p_1\succ p_2\succ\cdots\succ p_n$ for fixed $n$ will be, according to
 equation~\ref{plackettluce},
@@
 
-(here $[n]_x!$ is the $q$-factorial).  We see that
-$f(1)=\frac{1}{n!}$, $f'(x)>0$ and $f(0)=1$; so small $x$ corresponds
-to a high probability of observing the ``natural order'' in which
-competitor $j$ places $j$-th, $1\leqslant j\leqslant n$.  Further, we
-observe that $f(x)=1-(n-1)x + {\mathcal O}(x^2)$.  It is interesting
-to note that the probability of observing the reverse order, that is
-$p_n\succ p_{n-1}\succ\cdots\succ p_1$, is $f(1/x)$ for $x>0$.
+(here $[n]_x!$ is the $q$-factorial).  We see that $f(0)=1$, $f'(x)<0$
+and $f(1)=\frac{1}{n!}$; so small $x$ corresponds to a high
+probability of observing the ``natural order'' in which competitor $j$
+places $j$-th, $1\leqslant j\leqslant n$.  Further, we observe that
+$f(x)=1-(n-1)x + {\mathcal O}(x^2)$.  If $x=1/2$, it can be shown that
+the probability of observing the natural order is asymptotically $\sim
+2^{-n}$.  It is interesting to note that the probability of observing
+the reverse order, that is $p_n\succ p_{n-1}\succ\cdots\succ p_1$, is
+$f(1/x)$ for $x>0$.
+
+The rest of this paper considers the following inference problem: We
+consider $n$ competitors with Bradley-Terry strengths proportional to
+$x^i, 1\leqslant i\leqslant n$.  Our object of inference is the focal
+competitor with strength $\propto x^r$, with $r$ unknown.  The
+competitors are ranked randomly according to a Plackett-Luce model; we
+observe only the rank of the focal competitor.  What may we infer
+about $(x,r)$?
 
@@
 \includegraphics[width=4in]{123all}  % 123all.pdf produced by exponential_BT.Rmd
-\caption{Log-likelihood functions \label{123all} for support
-  $\in\left\lbrace 1, 2,3\right\rbrace\times\mathbb{R}$}
+\caption{The case $n=3$.  Three log-likelihood
+  functions \label{123all}, one for each of the three possible
+  observations of the rank $r$ of the focal competitor, presented as a
+  function of $\left\lbrace 1, 2,3\right\rbrace\times\mathbb{R}^-$.
+  (a), $r=1$; (b), $r=2$; (c), $r=3$}
 \end{centering}
@@
 different values of $n$, and to treat $s$ as a continuous variable.
-The only reasonable way to calulate such a likelihood function appears
-to be numerical; fortunately this is straightforward.
+The only reasonable way to calculate such a likelihood function
+appears to be numerical; fortunately this is straightforward.
 
@@
 observe the frequency with which competitor $\lceil ns\rceil$ places
-$r$.  For example, consider $x=0.1, r=3$.  We might simulate this as
-follows:
+$j$.  For example, consider $x=0.1, j=4$: each competitor has 10 times
+the strength of the next, and the focal competitor placed fourth.  We
+might simulate this as follows:
 
@@
 n <- 9 # number of competitors
-s <- 4  # observation: focal competitor places fourth
+       # observation: focal competitor places fourth
 
@@
 observations, and here we note that small values of $x$ lead to
-underdispersed results, for the focal competitor would come $r$-th
-with probability $\geqslant x^1+x^2+\ldots x^r$
+underdispersed results.
 
@@
 $r=3$, that is, competitor $i$ has strength $\propto 0.85^i,
-1\leqslant i\leqslant 9$; the focal competitor has the third strongest
-Plackett-Luce strength of $0.85^3\simeq 0.61$.  The results of running
-{\tt rrace()} 100 times with $n=9$ are given in
+1\leqslant i\leqslant 9$; the focal competitor has $r=3$, that is, has
+the third strongest Plackett-Luce strength of $0.85^3\simeq 0.61$.
+The results of running {\tt rrace()} 100 times with $n=9$ are given in
 table~\ref{results100insilico}.
@@
 
-Now, we can run repeated simulations like those on page~\pageref{rsession}, but
-with $n=10^4$ trials; as support we use $x$ running from 0.1 to 1 and
-$r=1,2,\ldots 9$: figure~\ref{toyexample}.  We see reasonable
-agreement between the true values of $x$ and $r$ and the estimates.
-Note the odd shape of the contours and the non-independence of the
-parameters.
+Now, we can run repeated simulations like those on
+page~\pageref{rsession}, but with $n=10^4$ trials; as support we use
+$x$ running from 0.1 to 1 and $r=1,2,\ldots 9$.
+Figure~\ref{toyexample} gives a visual representation of the
+log-likelihood.  We see reasonable agreement between the true values
+of $x$ and $r$, indicated with a red dot, and the estimates.  Note the
+odd shape of the contours and the non-independence of the parameters.
 
@@
 Formula 1 racing is an important and prestigious motor sport
-\citep{codling2017,jenkins2010}.  In a typical season, $\sim 20$
-competitors race at $\sim 20$ venues and here we consider the 2023
+\citep{codling2017,jenkins2010}.  In a typical season, ${\sim} 20$
+competitors race at ${\sim} 20$ venues and here we consider the 2023
 season, using Lewis Hamilton as the focal competitor.  His results are
@@
 (this time with $b=10^4$ trials per point), we obtain
-figure~\ref{formula1}.  From this, we see that $r$ is less than about
+figure~\ref{formula1}.  From this, we see that $s$ is less than about
 0.25 and $x$ is between about 0.4 and 0.8.  Recall that, in this
@@
 
-Here a simplification of the Plackett-Luce likelihood functions for
-order statistics was proposed, some simple properties obtained, and
-inference problems from three disciplines given.  The results seem to
+Here a parametrized Plackett-Luce likelihood function for order
+statistics was proposed, some simple properties obtained, and
+inference problems from motor sport given.  The results seem to
 be difficult to obtain any other way.  The assumptions furnish a

### 2025-11-28 inst/javelin.Rmd
soft link for hyper2.bib
thinko

@@
 dotchart(m2,pch=16)
-abline(0,1)
 ordertransplot(m1,m2)

### 2025-12-03 inst/tennis.Rmd
soft link for hyper2.bib
consistentification: use specificp.gt.test() if possible

@@
 ```{r specificptesthere,cache=TRUE}
-specificp.test(tennis_ghost,"G",0.0001)  # nonzero value needed to avoid triggering known R bug
+specificp.gt.test(tennis_ghost,"G",0)  
 ```

### 2025-12-03 inst/tennis.Rmd
more work on hyper3

@@
 
-```{r printH}
-H3 <- as.hyper3(H)
-H3
+```{r defineadd12vs34}
 
-## 1&2 vs 3&4;  scoreline 9-2 (again), with lambda=1.2
-
-f <- function(lambda){
+## 1&2 vs 3&4;  scoreline 9-2 (again)
+add12vs34 <- function(lambda){
+    H3 <- hyper3()
     H3[c(p1=lambda,p2=lambda)] %<>% inc(9)
-  H3[c(p3=1,p2=1)] %<>% dec(2)
+    H3[c(p3=1,p2=1)] %<>% inc(2)
     H3[c(p1=lambda,p2=lambda,p3=1,p2=1)] %<>% dec(11)
-  return(H3)
+    return(H3 + as.hyper3(H))
 }
-f(1.1443345)
-f(1.888888)
 ```
 
-maximize:
+```{r useadd12vs34, cache=TRUE}
+add12vs34(1.88)
+maxp(add12vs34(2.0), give=TRUE)$value
+maxp(add12vs34(2.1), give=TRUE)$value
+```
 
-```{r,maxH3,cache=TRUE}
-tennis_H3_maxp <- maxp(H3)
-tennis_H3_maxp
+```{r defineprofilelikelihood12vs34}
+L <- function(lambda){maxp(add12vs34(lambda), give=TRUE)$value}
+```
+
+```{r useprofilelikelihood12vs34, cache=TRUE}
+lam <- exp(seq(from=log(0.9), to=log(60), len=20))
+support <- sapply(lam, L)
+```
+
+```{r optimize12v34, cache=TRUE}
+o <- optimize(L, interval=c(exp(1.5),exp(2.5)), maximum=TRUE)
+o
+lambda_max <- o$maximum
+```
+
+```{r plotproflike}
+support <- support - max(support)
+plot(log(lam), support,
+     pch=16, type="b",
+     xlab=expression(log(lambda)), ylab="log-Likelihood"
+     )
+abline(h=c(0,-2))
+abline(v=c(0,log(lambda_max)), lty=2)
+```
+
+```{r maxtennis3, cache=TRUE}
+tennis_hyper3_maxp <- maxp(add12vs34(lambda_max))
 ```
@@
 
-```{r}
+```{r, showallmax}
 tennis_maxp
 tennis_ghost_maxp
+tennis_hyper3_maxp
+```
+
+Visuals:
+
+```{r plotallmax}
+x1 <- tennis_maxp
+x2 <- tennis_ghost_maxp[-1]
+x2 <- x2/sum(x2)
+x3 <- tennis_hyper3_maxp
+x1
+x2
+x3
+```
+
+Above, observe how x2 and x3 are identical: evidently, in this case,
+`hyper2` formalism is formally identical to `hyper3` formalism.
+Visually:
+
+```{r showall}
+par(pty="s")
+l <- c(0.15,0.3)
+plot(x1, x2, asp=1, xlim=l, ylim=l, pch=16)
+abline(0,1)
 ```
@@
 ```{r,label=saveghosttennisdataset}
-save(tennis,tennis_maxp,tennis_ghost, tennis_ghost_maxp,file="tennis.rda")
+save(tennis, tennis_maxp, tennis_ghost, tennis_ghost_maxp, file="tennis.rda")
 ```

### 2025-12-05 inst/hankin-exponential_BT.tex
work

@@
 
-\subsection{Theoretical results}
+\section{Theoretical results}
 
@@
 
-\section{Formula 1}
+\section{Case study: parkrun}
+
+``Parkrun'' is a distributed community initiative that organises
+weekly timed 5 km runs/walks in parks worldwide~\citep{hindley2020}.
+A typical event will have 200 participants but for the purposes of
+this study, we will consider only male competitors aged 55 and over, a
+demographic to which the author considers himself to belong.
+Table~\ref{parkruntable} shows 7 completed parkruns.  From the first
+pair of numbers we see that 43 runners from our demographic attended
+that particular parkrun, of whom the author placed 33.
+
+\begin{table}[t]
+  \centering
+  \caption{Parkrun results}
+\label{parkruntable}
+\begin{tabular}{cccccccc}\\
+rank   & 33 & 35 & 35 & 37 & 32 & 30 & 37\\
+runners& 43 & 49 & 59 & 51 & 56 & 55 & 58
+\end{tabular}
+\end{table}
+
+Using numerical simulation, in this case repeated evaluation of
+equation \ref{aftersim} and a generalized version of equation
+\ref{requalfour}, with $b=5\times 10^4$ trials per point, we obtain
+figure~\ref{doesnotexist}.  From this, we see that $s$ is less than
+about XXXXXX and $x$ is between about SXXXXASDFASDFADF and WWWWW.
+SDDFDSF
+
+Recall that, in this model, $x=0.5$ implies that the Plackett-Luce
+strength of each competitor is twice that of the next; and if
+$x=0.25$, we can expect
+
+\section{Case study: Formula 1}
 
@@
 model, $x=0.5$ implies that the Plackett-Luce strength of each
-competitor is about twice that of the next.
+competitor is twice that of the next; and if $x=0.25$, we can expect
+Hamilton to be better than a randomly chosen competitor with
+probability $1-0.25=0.75$.
 
@@
 \includegraphics[width=4in]{formula1}  % formula1.pdf produced by exponential_BT.Rmd
-\caption{Contours of log-likelihood functions \label{formula1}.  Red circle shows
-  maximum likelihood estimate}
+\caption{Contours of log-likelihood function for Lewis Hamilton's
+  Formula One performance over the 2023 season. \label{formula1} Red
+  circle shows maximum likelihood estimate}
 \end{centering}

### 2025-12-05 inst/parkrun_idea.Rmd
another approach
work

@@
 ```{r sethowmany}
-howmany <- 100
+howmany <- 1000
 ```
@@
 ```{r parkrunlikefunobs1, cache = TRUE}
-print(system.time(parkrun01 <- single_like_fun(259, 177/259, N = howmany, a_try, x_try)))
+print(system.time(parkrun01 <- single_like_fun(43, 33/43, N = howmany, a_try, x_try)))
 ```
@@
 ```{r parkrunlikefunobs2, cache = TRUE}
- print(system.time(parkrun02 <- single_like_fun(305, 222/305, N = howmany, a_try, x_try)))
+ print(system.time(parkrun02 <- single_like_fun(49, 35/49, N = howmany, a_try, x_try)))
 ```
@@
 ```{r parkrunlikefunobs3, cache = TRUE}
-print(system.time(parkrun03 <- single_like_fun(297, 206/297, N = howmany, a_try, x_try)))
+print(system.time(parkrun03 <- single_like_fun(59, 35/59, N = howmany, a_try, x_try)))
 ```
@@
 ```{r parkrunlikefunobs4, cache = TRUE}
-print(system.time(parkrun04 <- single_like_fun(241, 142/241, N = howmany*10, a_try, x_try)))
+print(system.time(parkrun04 <- single_like_fun(51, 37/51, N = howmany, a_try, x_try)))
 ```
@@
 
-```{r parkrunlikefunobs5, cache = TRUE}
-print(system.time(parkrun05 <- single_like_fun(179, 118/179, N = howmany, a_try, x_try)))
+```{r parkrunlikefunobs4, cache = TRUE}
+print(system.time(parkrun05 <- single_like_fun(56, 32/56, N = howmany, a_try, x_try)))
 ```
@@
 
-```{r parkrunlikefunobs6, cache = TRUE}
-print(system.time(parkrun06 <- single_like_fun(338, 224/338, N = howmany, a_try, x_try)))
+```{r parkrunlikefunobs4, cache = TRUE}
+print(system.time(parkrun04 <- single_like_fun(55, 30/55, N = howmany, a_try, x_try)))
 ```
@@
 
-```{r parkrunlikefunobs7, cache = TRUE}
-print(system.time(parkrun07 <- single_like_fun(203, 128/203, N = howmany, a_try, x_try)))
+```{r parkrunlikefunobs4, cache = TRUE}
+print(system.time(parkrun04 <- single_like_fun(58, 37/58, N = howmany, a_try, x_try)))
 ```
@@
 sum(parkrun03),
-sum(parkrun04),
-sum(parkrun05),
-sum(parkrun06),
-sum(parkrun07)
+sum(parkrun04)
 )
@@
 ```{r}
-contour(a_try,x_try,parkrun05)
-contour(a_try,x_try,parkrun06)
-parkrun01*parkrun02*parkrun03*parkrun04*parkrun05*parkrun06*parkrun07
+contour(a_try,x_try,parkrun01)
+contour(a_try,x_try,parkrun02)
+Likelihood <-
+    parkrun01*
+    parkrun02*
+    parkrun03*
+    parkrun04*
+    parkrun05*
+    parkrun06*
+    parkrun07
+contour(a_try,x_try, Likelihood)
 ```
+
+```{r}
+M <- matrix(c(
+259, 177,
+305, 222,
+297, 206,
+241, 142,
+179, 118,
+338, 224,
+203, 128),ncol=2,byrow=TRUE)
+M
+plot(M[,2]/M[,1])
+```

### 2025-12-06 inst/pwa3.Rmd
pwa3() example

---
title: "Person with advantage functionality for `hyper3` objects: `pwa3()`"
output: bookdown::html_document2
author: "Robin K. S. Hankin"
link-citations: true
bibliography: hyper2.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
```

```{r}
pwa3
```

To cite the `hyper2` package in publications, please use @hankin2017_rmd.

```{r definelamf}
lam <- function(lambda){
  race_nohelp <-
    race(letters[c(1,2,4,3)]) + 
    race(letters[c(3,1,2,4)]) +
    race(letters[c(2,4,1,3)])

  race_help <- 
    race(letters[c(2,3,1,4)]) |> pwa3("b", lambda) + 
    race(letters[c(1,3,4,2)]) |> pwa3("b", lambda) +
    race(letters[c(3,2,1,4)]) |> pwa3("b", lambda)

  maxp(as.hyper3(race_nohelp) + race_help, give=TRUE)$value
}
```

```{r calclamf, cache=TRUE}
l <- exp(seq(from=log(0.02), to=log(8), len=17))
like <- sapply(l,lam)
```

```{r optimizelamf, cache=TRUE}
optimal_sol <- optimize(lam, c(0.01, 3), maximum=TRUE)
```

```{r showoptimlam}
optimal_sol
```

```{r lamfminus2, cache=TRUE}
f <- function(l){lam(l) - optimal_sol$objective + 2}
lower <- uniroot(f, interval = c(0.01, 1))
upper <- uniroot(f, interval = c(1, 10))
```

```{r showroots}
lower
upper
```

```{r plotlamf}
plot(log(l), like - optimal_sol$objective, type='b')
abline(h=c(0, -2))
abline(v = log(optimal_sol$maximum))
segments(x0 = log(c(lower$root, upper$root)), y0 = -2.25, y1 = -1.75)
```

### 2025-12-06 inst/pwa3.Rmd
better example

@@
 
-To cite the `hyper2` package in publications, please use @hankin2017_rmd.
+To cite the `hyper2` package in publications, please use
+@hankin2017_rmd.  Here we analyse some simple Plackett-Luce
+observations, using `hyper3` likelihood functions.
+
+Our observations are a total of six order statistics for four
+competitors `a,b,c,d`.  These competitors race three times under
+Plackett-Luce probabilities.  They then race three times, but with
+competitor `b` having some advantage (steroids?), parametrized by his
+Plackett-Luce strength being $\lambda b$.
+
+Our object of inference is $\lambda$; the Plackett-Luce strengths
+`a,b,c,d` are nuisance parameters.
 
@@
 lam <- function(lambda){
+
   race_nohelp <-
-    race(letters[c(1,2,4,3)]) + 
-    race(letters[c(3,1,2,4)]) +
-    race(letters[c(2,4,1,3)])
+    race(c("b","a","d","c")) + 
+    race(c("c","b","a","d")) + 
+    race(c("b","d","a","c"))
 
   race_help <- 
-    race(letters[c(2,3,1,4)]) |> pwa3("b", lambda) + 
-    race(letters[c(1,3,4,2)]) |> pwa3("b", lambda) +
-    race(letters[c(3,2,1,4)]) |> pwa3("b", lambda)
+    race(c("a","c","d","b")) |> pwa3("b", lambda) + 
+    race(c("c","d","b","a")) |> pwa3("b", lambda) + 
+    race(c("d","c","a","b")) |> pwa3("b", lambda)
 
@@
 
+Above we see that in the unaided case (the first three races), `b`
+does quite well, coming first twice and second once.  In the aided
+case, `b` does much worse, coming either third or fourth.  Evidently,
+the "help" actually hinders `b` and we suspect that $\lambda<1$.
+
 ```{r calclamf, cache=TRUE}
-l <- exp(seq(from=log(0.02), to=log(8), len=17))
+l <- exp(seq(from=log(0.0002), to=log(2), len=7))
 like <- sapply(l,lam)
@@
 f <- function(l){lam(l) - optimal_sol$objective + 2}
-lower <- uniroot(f, interval = c(0.01, 1))
-upper <- uniroot(f, interval = c(1, 10))
+lower <- uniroot(f, interval = c(exp(-8), exp(-7)))
+upper <- uniroot(f, interval = c(exp(-2), exp( 0)))
 ```
@@
 
-```{r plotlamf}
+```{r plotlamf, echo=TRUE,fig.cap="Log-likelihood function for lambda"}
 plot(log(l), like - optimal_sol$objective, type='b')
-abline(h=c(0, -2))
+abline(h = c(0, -2))
 abline(v = log(optimal_sol$maximum))
-segments(x0 = log(c(lower$root, upper$root)), y0 = -2.25, y1 = -1.75)
+abline(v = 0)
+segments(x0 = log(c(lower$root, upper$root)), y0 = -2.25, y1 = -1.75, col='gray')
 ```
+
+Figure \@ref(fig:plotlamf) shows

### 2025-12-07 inst/pwa3.Rmd
pwa3() formalism

@@
 pwa3
+pwa3_single
 ```
@@
   race_help <- 
-    race(c("a","c","d","b")) |> pwa3("b", lambda) + 
-    race(c("c","d","b","a")) |> pwa3("b", lambda) + 
-    race(c("d","c","a","b")) |> pwa3("b", lambda)
+    race(c("a","c","d","b")) |> pwa3(b = lambda) + 
+    race(c("c","d","b","a")) |> pwa3(b = lambda) + 
+    race(c("d","c","a","b")) |> pwa3(b = lambda)
 
@@
 ```{r calclamf, cache=TRUE}
-l <- exp(seq(from=log(0.0002), to=log(2), len=7))
+l <- exp(seq(from=log(0.0002), to=log(2), len=17))
 like <- sapply(l,lam)
@@
 
-```{r showroots}
-lower
-upper
-```
-
 ```{r plotlamf, echo=TRUE,fig.cap="Log-likelihood function for lambda"}
-plot(log(l), like - optimal_sol$objective, type='b')
+plot(log(l), like - optimal_sol$objective,
+	     type='b', pch=16,
+	     xlab = expression(log(lambda)),
+	     ylab = "log-likelihood")
 abline(h = c(0, -2))
@@
 
-Figure \@ref(fig:plotlamf) shows
+Figure \@ref(fig:plotlamf) shows a log-likelihood function for
+$\log\lambda$.  We see that the evaluate is about
+$\lambda=\exp(`r round(log(optimal_sol$maximum),2)`)\simeq
+`r round(optimal_sol$maximum,3)`$.  Further, the support interval
+runs from about
+$\exp(`r round(log(lower$root),2)`)\simeq `r round(lower$root,6)`$
+to about
+$\exp(`r round(log(upper$root),2)`)\simeq `r round(upper$root,2)`$
+thereby excluding the null of $\lambda=1$
+which has a support of about `r round(lam(1)-optimal_sol$objective,2)`.

### 2025-12-07 R/hyper3.R

@@ -724,7 +724,7 @@ stop("not yet written")
     return(out)
 }
 
-"pwa3" <- function(H3, pwa, lambda){
+"pwa3_single" <- function(H3, pwa, lambda){
     stopifnot(pwa %in% pnames(H3))
     stopifnot(length(pwa) == 1)
     H3 <- as.hyper3(H3)
@@ -734,4 +734,12 @@ stop("not yet written")
         W[[i]][B[[i]] %in% pwa] <- lambda
     }
     return(hyper3_bw(B, W, powers(H3), pnames(H3)))
+}
+
+"pwa3" <- function(H3, ...){
+    l <- list(...)
+    for(i in seq_along(l)){
+        H3 <- pwa3_single(H3, names(l[i]), l[[i]])
+    }
+    return(H3)
 }

### 2025-12-07 inst/pwa3.Rmd
another case study

@@
 ```
+
+```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
+knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
+```
+
+<font size="1"> (takes about five minutes to run without cache)
+</font>
 
@@
 	     ylab = "log-likelihood")
-abline(h = c(0, -2))
+abline(h = c(0, -2), col='gray')
 abline(v = log(optimal_sol$maximum))
-abline(v = 0)
+abline(v = 0, lty=2)
 segments(x0 = log(c(lower$root, upper$root)), y0 = -2.25, y1 = -1.75, col='gray')
@@
 which has a support of about `r round(lam(1)-optimal_sol$objective,2)`.
+
+# Another example {-}
+
+In this case, competitor `a` receives help $\lambda$ or $\lambda^n$ if
+he gets $n$ sets of help.  Below, in function `lam2()` we see that, in
+general, the more help `a` gets, the better he does.  See how a
+negative power corresponds to some sort of hindrance or penalty, and a
+zero power (of course) corresponds to no help at all.
+
+```{r lam2def}
+lam2 <- function(lam){
+
+  race_help <- 
+    race(c("d","c","a","b")) |> pwa3(a = lam^0 ) +
+    race(c("c","b","d","a")) |> pwa3(a = lam^-3) + 
+    race(c("c","b","d","a")) |> pwa3(a = lam^-2) + 
+    race(c("c","b","a","d")) |> pwa3(a = lam^-1) + 
+    race(c("a","c","b","d")) |> pwa3(a = lam   ) + 
+    race(c("a","c","b","d")) |> pwa3(a = lam^3 ) +
+    race(c("a","b","d","c")) |> pwa3(a = lam   ) +
+    race(c("c","a","d","b")) |> pwa3(a = lam^2 )
+
+    maxp(race_help, give=TRUE)$value
+}
+```
+
+```{r calclam2f, cache=TRUE}
+l <- exp(seq(from=log(0.8), to=log(19.1), len=11))
+like2 <- sapply(l,lam2)
+```
+
+```{r}
+plot(log(l),like2-max(like2),type="b",pch=16)
+abline(h=c(0,-2), col='gray')
+abline(v=0, lty=2)
+```
+
+# Another other example {-}
+
+In this example we see help being given to two competitors at the same
+time.
+
+```{r lam3def}
+
+myr <- function(a1,a2,a3,a4,a5){race(letters[c(a1,a2,a3,a4,a5)])}
+
+lam3 <- function(l){
+  race_help <- 
+    myr(3,1,2,4,5) |> pwa3(a=l, b=l) + 
+    myr(2,3,1,4,5) |> pwa3(a=l, b=l) + 
+    myr(3,1,5,4,2) |> pwa3(a=1, b=1) +   # no help!
+    myr(4,2,5,3,1) |> pwa3(a=1, b=1) +   # no help!
+    myr(3,1,2,4,5) |> pwa3(a=l, b=l)
+
+    maxp(race_help, give=TRUE)$value
+}
+```
+
+```{r calclam3, cache=TRUE}
+l <- exp(seq(from=log(0.5), to=log(84.1), len=13))
+like3 <- sapply(l,lam3)
+```
+
+```{r plotlam3}
+plot(log(l),like3-max(like3), type="b", pch=16)
+abline(h=c(0,-2), col='gray')
+abline(v=0, lty=2)
+```
+
+### References {-}

### 2025-12-08 inst/pwa3.Rmd
another example

@@
 
+# Yet another example {-}
+
+Here we consider help in turn to each of five competitors.  From the
+function below, if someone gets help they generally do quite well.
+
+```{r lam4def}
+
+myr <- function(a1,a2,a3,a4,a5){race(letters[c(a1,a2,a3,a4,a5)])}
+
+lam4 <- function(x, givelike=FALSE){
+    race_help <- 
+        myr(3,1,2,4,5) |> pwa3(a=x) +
+        myr(2,3,1,4,5) |> pwa3(b=x) +
+        myr(3,1,5,4,2) |> pwa3(c=x) +
+        myr(4,2,5,3,1) |> pwa3(d=x) +
+        myr(3,5,2,4,1) |> pwa3(e=x)
+    if(givelike){
+        return(race_help)
+    } else {
+        return(maxp(race_help, give=TRUE)$value)
+    }
+}
+```
+
+```{r calclam4, cache=TRUE}
+l <- exp(seq(from= -0.5, to=log(250), len=15))
+like4 <- sapply(l,lam4)
+```
+
+```{r optimizelamf4, cache=TRUE}
+optimal_sol4 <- optimize(lam4, exp(2:3), maximum=TRUE)
+```
+
+It makes sense to consider our likelihood function at the estimate for
+$\lambda$:
+
+```{r maxp4}
+H <- lam4(optimal_sol4$maximum, givelike=TRUE)
+H
+summary(H)
+maxp(H)
+```
+
+Above we see that the evaluate is very high for `c` having a BT
+strength of about 0.6.  This is consistent with our observations for
+twice we see `c` winning [that is, coming first] without any help.
+
+```{r plotlam4}
+plot(log(l),like4-max(like4), type="b", pch=16)
+abline(h=c(0,-2), col='gray')
+abline(v=0, lty=2)
+abline(v = log(optimal_sol4$maximum))
+```
+
 ### References {-}

### 2025-12-09 inst/pwa3.Rmd
pole position analysis

@@
 
-```{r maxp4}
+```{r plotlam4}
+plot(log(l),like4-max(like4), type="b", pch=16)
+abline(h=c(0,-2), col='gray')
+abline(v=0, lty=2)
+abline(v = log(optimal_sol4$maximum))
+```
+
+```{r maxp4,cache=TRUE}
 H <- lam4(optimal_sol4$maximum, givelike=TRUE)
@@
 twice we see `c` winning [that is, coming first] without any help.
+Is there statistical evidence to support this assertion?
 
-```{r plotlam4}
-plot(log(l),like4-max(like4), type="b", pch=16)
-abline(h=c(0,-2), col='gray')
-abline(v=0, lty=2)
-abline(v = log(optimal_sol4$maximum))
+The simplest approach would be to consider the likelihood function at
+$\hat{\lambda}\simeq 16.705$:
+
+```{r stat4, cache=TRUE}
+f <- function(o){samep.test(H, c(o, "c"))$statistic}
+stat4 <- c(a=f("a"), b=f("b"), d=f("d"), e=f("e"))
+```
+
+```{r}
+stat4
+```
+
+Above we see that there is no strong evidence for `c` being the
+strongest, on the grounds that we are unable to reject $H_0\colon
+p_b=p_c$ [the support difference is about 0.74, short of the two-units
+of support criterion required for rejection], at least with
+$\lambda=16.705$.
+
+# Another further idea {-}
+
+Now we have a model for the starting grid in formula 1.   In the
+following function we specify the finishing order and also the
+starting grid order.
+
+```{r lam5}
+
+tolam <- function(lam,o){setNames(lam^(seq_along(o)-1),o)}
+
+lam5 <- function(lam, givelike=FALSE){
+    race_help <- 
+        race(c("a","b","d","c","e")) |> pwa3(tolam(lam, c("a","b","c","d","e"))) +
+        race(c("a","c","b","d","e")) |> pwa3(tolam(lam, c("c","b","a","d","e"))) +
+        race(c("a","d","c","b","e")) |> pwa3(tolam(lam, c("a","c","b","d","e"))) +
+        race(c("e","c","b","a","d")) |> pwa3(tolam(lam, c("a","c","b","e","d"))) +
+        race(c("b","c","e","a","d")) |> pwa3(tolam(lam, c("c","b","e","d","a"))) +
+        race(c("a","d","c","b","e")) |> pwa3(tolam(lam, c("a","b","c","d","e")))
+
+    if(givelike){
+        return(race_help)
+    } else {
+        return(maxp(race_help, give=TRUE)$value)
+    }
+}
+```
+
+```{r lam5calculator, cache=TRUE}
+l <- exp(seq(from= -2, to=log(1.5), len=15))
+like5 <- sapply(l,lam5)
+optimal_sol5 <- optimize(lam5, exp(c(-1,0)), maximum=TRUE)
+```
+
+```{r lamfminus5, cache=TRUE}
+f <- function(l){lam5(l) - optimal_sol5$objective + 2}
+lower <- uniroot(f, interval = c(exp(-1.5), exp(-1.0)))
+upper <- uniroot(f, interval = c(exp(-0.5), exp(+0.5)))
+```
+
+```{r plotlamf5, echo=TRUE, fig.cap="Log-likelihood function for lambda, the pole position measure"}
+plot(log(l), like5 - max(like5), type='b', pch=16)
+abline(h = c(0, -2), col='gray')
+abline(v = log(optimal_sol5$maximum))
+abline(v = 0, lty=2)
+segments(x0 = log(c(lower$root, upper$root)), y0 = -2.5, y1 = -1.5, col='red')
 ```

### 2025-12-09 inst/grid.Rmd
grid placing work

---
title: "The starting grid in Formula 1: 2025"
author: "R. K. S. Hankin"
output: bookdown::html_document2
bibliography: hyper2.bib  # copy from vignettes/

---

```{r setup, include=FALSE}
set.seed(0)
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
options("digits" = 5)
```

```{r out.width='20%', out.extra='style="float:right; padding:10px"',echo=FALSE}
knitr::include_graphics(system.file("help/figures/hyper2.png", package = "hyper2"))
```

To cite the `hyper2` package in publications, please use @hankin2017_rmd.

```{r}
x <- readLines("grid_2025.txt")
```

```{r}
drivers <- function(i){strsplit(x[i], " ")[[1]][-1]}
venue   <- function(i){strsplit(x[i], " ")[[1]][ 1]}
drivers(11)
allvenues <- sapply(1:24,venue)
alldrivers <- NULL
for(i in 1:24){alldrivers <- c(alldrivers,drivers(i))}
alldrivers <- sort(unique(alldrivers))
M <- matrix(0,nrow=24, ncol=length(alldrivers))
colnames(M) <- alldrivers
rownames(M) <- allvenues
for(i in 1:24){M[i,] <- alldrivers %in%  drivers(i)    }
M
rowSums(M)
colSums(M)   
```

Above we see that Colapinto and Doohan were the only drivers that did
not compete at every venue.  We will have to remove these two drivers
from the data:

```{r}
grid_order <- list()
unwanted <- c("Colapinto", "Doohan")
for(i in 1:24){
    jj <- drivers(i)
    grid_order[[i]] <- jj[!(jj %in% unwanted)]
}
grid_order
```

We are going to use the 2025 finishing dataset:

```{r, read2025, cache=TRUE}
a <- read.table("~/rstudio/hyper2/inst/formula1_2025.txt")
a <- a[1:19,-ncol(a)]   # remove Colapinto and Doohan; and the points
```

We need to iterate through each venue, and use `pwa3()` to modify the
likelihood function for each one.  Then add the log-likelihoods for
each venue together.

```{r f1griddef}

geometric <- function(lambda, x){setNames(lambda^(seq_along(x)-1),x)}

f1grid <- function(lambda){

    H <- list()
    for(i in 1:24){
        vec <- as.numeric(a[,1])
        vec[is.na(vec)] <- 0
        jj <- setNames(vec, rownames(a))
        H[[i]] <- ordervec2supp(jj) |> pwa3(geometric(lambda,grid_order[[i]]))
    }
    out <- hyper3()
    for(i in 1:24){
        out <- out + H[[i]]
    }
    return(out)
}
```

```{r calcpoll, cache=TRUE}
jj <- f1grid(0.5)
```

```{r maxcalc, cache=TRUE}
mH <- maxp(jj)
```

```{r}
mH
```

### 2025-12-11 inst/grid.Rmd
likelihood function plot for F1

@@
 ```{r calcpoll, cache=TRUE}
-jj <- f1grid(0.5)
+supptime  <- system.time(jj <- f1grid(0.5))
 ```
@@
 ```{r maxcalc, cache=TRUE}
-mH <- maxp(jj)
+maxtime <- system.time(mH <- maxp(jj))
 ```
@@
 ```{r}
+supptime
+maxtime
 mH
+pie(mH)
 ```
+
+```{r domore, cache=TRUE}
+o <- function(lambda){maxp(f1grid(lambda), give=TRUE)$value}
+timeo1 <- system.time(l1 <- o(1.01))
+timeo2 <- system.time(l2 <- o(1.02))
+```
+
+```{r}
+timeo1
+timeo2
+l1
+l2
+```
+
+```{r domanyL, cache=TRUE}
+lam <- seq(from=0.9, to=1, len=17)
+L <- sapply(lam, o)
+```
+
+```{r plotmanyL}
+plot(lam, L - max(L), pch=16, type="b")
+```

### 2025-12-12 inst/surfing_jsa.tex
various changes to maintain the surfing work

@@
 \begin{document}
+
+(this document is based on \citep{driver2023}).
 
@@
 because the likelihood function is exponential family and as such is
-amenable to analysis via general linear models~\citep{turner2012}.
+amenable to analysis via general linear models~\citep{turner2012_nomarkup}.
 However, this is not applicable when more than two entities are
@@
 \bibliographystyle{agsm}
-\bibliography{chess}
+\bibliography{hyper2}
 

### 2025-12-12 inst/grid.Rmd
more output

--- 2025-12-09 inst/grid.Rmd
@@
     for(i in 1:24){
-        vec <- as.numeric(a[,1])
+        suppressWarnings(vec <- as.numeric(a[,1]))
         vec[is.na(vec)] <- 0
@@
 
-```{r calcpoll, cache=TRUE}
-jj <- f1grid(0.5)
+```{r calcpolltime, cache=TRUE}
+supptime  <- system.time(jj <- f1grid(0.5))
 ```
 
-```{r maxcalc, cache=TRUE}
-mH <- maxp(jj)
+```{r maxcalctime, cache=TRUE}
+maxtime <- system.time(mH <- maxp(jj))
+```
+
+```{r showresultstime}
+supptime
+maxtime
+mH
+pie(mH)
+```
+
+```{r domore, cache=TRUE}
+o <- function(lambda){maxp(f1grid(lambda), give=TRUE)$value}
+timeo1 <- system.time(l1 <- o(1.01))
+timeo2 <- system.time(l2 <- o(1.02))
 ```
@@
 ```{r}
-mH
+timeo1
+timeo2
+l1
+l2
 ```
+
+```{r domanyLikes, cache=TRUE}
+lam <- seq(from=0.9, to=1, len=17)
+L <- sapply(lam, o)
+```
+
+```{r plotF1lamf, echo=TRUE, fig.cap="Log-likelihood function for lambda, the pole position measure"}
+plot(lam, L - max(L), pch=16, type="b")
+```
+
+Figure \@ref(fig:plotF1lamf) shows a log-likelihood function for
+$\lambda$.  It is very irregular, indicating that the optimization
+routine is having difficulty finding a global maximum for all values
+of $\lambda$.

### 2025-12-12 inst/skating.Rmd
typo
cite Lock and Lock

@@
 output: bookdown::html_document2
+link-citations: true
 bibliography: hyper2.bib
@@
 Olympics.  In the package, dataframe `skating_table` is copied from
-Lock and Lock (see skating.Rd for more details).  It also creates file
-`data/skating.rda` which contains R objects `skating`,
-`skating_table`, and `skating_maxp`.
+Lock and Lock -@lock2003.  It also creates file `data/skating.rda`
+which contains R objects `skating`, `skating_table`, and
+`skating_maxp`.
 
@@
 
-Object `skating_table` is an order table.  It is structured so that
+Object `skating_table` is an _order table_.  It is structured so that
 each competitor is a row, and each judge is a column.  One good thing
-to do is to present the dataset as an order table;
+to do is to present the dataset as a _rank table_:
 

### 2025-12-14 R/ranktable.R
fixes #41

@@ -13,7 +13,10 @@
 }
 
 `ranktable_to_ordertable` <- function(xrank){
-    stop("not yet implemented.\nIt is harder than it looks.\nI'll get round to it one day.\nProbably.")
+    out <- apply(xrank, 1, order)
+    rownames(out) <- sort(unique(as.vector(xrank)))
+    colnames(out) <- rownames(xrank)
+    return(as.ordertable(out))
 }
 
 `ordertable_to_ranktable` <- function(xorder){

### 2025-12-14 tests/testthat/test_ranktable.R

@@ -4,7 +4,7 @@ test_that("ranktable tests", {
     expect_output(print(x))
     expect_output(print(summary(x)))
 
-    expect_error(ranktable_to_ordertable(x))
+    expect_true(all(as.ranktable(as.ordertable(x)) == x))
 
     expect_silent(ordertable_to_ranktable(pentathlon_table))
 

### 2025-12-14 inst/skating.Rmd
all competitors have strictly positive strength

@@
 ```
+
+(we note in passing that the Plackett-Luce strengths of every
+competitors is strictly positive.  We can see this by observing that:
+
+\begin{alignat*}{4}
+\textbf{luka}
+&\succ_3\mbox{kopak}      &&\succ_1\mbox{babiakova} &&\succ_3\mbox{giunchi}   &&\succ_1\mbox{hegel}      \\
+&\succ_3\mbox{soldatova}  &&\succ_6\mbox{gusmeroli} && \succ_2\mbox{meier}    &&\succ_3\mbox{hubert}     \\
+&\succ_4\mbox{onda}       &&\succ_1\mbox{kiashenko} &&\succ_3\mbox{fontana}   &&\succ_2\mbox{maniachenko}\\
+&\succ_3\mbox{volchkova}  &&\succ_2\mbox{kettunen}  &&\succ_2\mbox{sebestyen} &&\succ_4\mbox{robinson}   \\
+&\succ_3\mbox{butyrskaya} &&\succ_2\mbox{suguri}    &&\succ_1\mbox{cohen}     &&\succ_2\mbox{kwan}       \\
+&\succ_1\mbox{slutskaya}  &&\succ_2\mbox{hughes}  \\
+&\succ_1\textbf{luka}
+\end{alignat*}
+
+and if any competitor has zero BT strength, at least one of the above
+successors is false with probability one).
 

### 2025-12-16 inst/skating.Rmd
extraction methods for ordertable objects

@@
 kwan.  We can gain some understanding of this result by looking at the
-raw judging results for the three medallists:
+raw judging results for the three medallists.
+
+# Extraction methods for ordertables
+
+The entire ordertable is given here again for convenience:
+
+```{r,tableagain}
+skating_table
+```
+
+Suppose we are interested only in `hughes`, `slutskaya`, and `kwan`;
+we thus focus on the first three lines of `skating_table`.  If we were
+to consider the entries of the first three lines, see how `hughes`
+came fourth according to `J2`.  This cannot be converted to a
+ranktable because the judges' observations are not sequential.  Each
+column should be a permutation of $\left\lbrace 1,2,3\right\rbrace$.
+To do this, we use the extraction method:
 
 ```{r,threemedallists}
-skating_table[1:3]
+skating_table[1:3,]
+```
+
+Above we see just the order for these three competitors.  This may be
+converted to a ranktable, or indeed a likelihood function:
+
+```{r,threemedallistsstuff}
+as.ranktable(skating_table[1:3,])
+suppfun(skating_table[1:3,])
 ```
@@
 ```{r justtop7}
-(o <- skating_table[1:7])
+(o <- skating_table[1:7,])
 ```

### 2025-12-16 R/hyper2.R

@@ -1281,4 +1281,6 @@ rorder_single <- function(p){
     }
 }
 
-`[.ordertable` <- function(x, ...){ ordertable(apply(unclass(x)[...,], 2, rank)) }
+`[.ordertable` <- function(x, i, j, ..., drop = FALSE) {
+   as.ordertable(NextMethod("[", drop = drop))
+}

### 2025-12-23 inst/javelin.Rmd
attemptstable objects

@@
 
-```{r,label=loadjavelinandconvert}
-javelin_table <- read.table("javelin.txt",header=TRUE)
+```{r, label=loadjavelinandconvert}
+javelin_table <- as.attemptstable(read.table("javelin.txt", header=TRUE))
 a <- javelin_table # saves typing
@@
 We may convert this to a named vector with elements being the throw
-distances, and names being the competitors, using
-`attemptstable2supp3()`:
+distances, and names being the competitors:
 
-```{r,label=converttosupp3,cache=TRUE}
-javelin_vector <- attemptstable2supp3(a,decreasing=TRUE,give.supp=FALSE)
+```{r, label=converttosupp3, cache=TRUE}
+javelin_vector <- as.vector(javelin_table)
 javelin_vector
@@
 n <- length(levels(as.factor(names(o))))
-plot(o,pch=16,col=rainbow(n)[as.factor(names(o))])
-legend("topright",pch=16,col=rainbow(n),legend=levels(as.factor(names(o))))
+plot(o, pch=16, col=rainbow(n)[as.factor(names(o))])
+legend("topright", pch=16, col=rainbow(n), legend=levels(as.factor(names(o))))
 ```
 
-Now convert the attempts table to a `hyper3` object, again using
-function `attemptstable2supp3()` but this time pass `give.supp=TRUE`.
-However, note that there are two reasonable interpretations: firstly,
-we simply ignore no-throws [in the package docs, these are sometimes
-referred to as "DNF", or "Did not finish", following Formula 1
-terminology]:
+Now convert the attempts table to a `hyper3` object, using function
+\code{suppfun()}.  However, note that there are two reasonable
+interpretations: firstly, we simply ignore no-throws [in the package
+docs, these are sometimes referred to as "DNF", or "Did not finish",
+following Formula 1 terminology] and to do this we pass
+`dnf.last=FALSE`:
 
-```{r,label=dothething2,cache=TRUE}    	   
-javelin1 <- ordervec2supp3(v=names(o)[!is.na(o)])
+```{r, label=dothething2, cache=TRUE}
+javelin1 <- suppfun(javelin_table, dnf.last=FALSE)
 H1 <- javelin1 # saves typing
@@
 Secondly, we may treat no-throws as being inferior to any actual
-measured throw.  To do this it is easiest to pass `give.supp=TRUE` to
-`attemptstable2supp3()`:
+measured throw.  We specify `dnf.last=TRUE`:
 
 ```{r,label=dothething1,cache=TRUE}    	   
-javelin2 <- attemptstable2supp3(a,decreasing=TRUE,give.supp=TRUE)
+javelin2 <- suppfun(javelin_table, dnf.last=TRUE)
 H2 <- javelin2 # saves typing
@@
 
-```{r,plotm}
-rbind(one=m1,two=m2)
+```{r, label=plotm}
+rbind(one=m1, two=m2)
 pie(m1)
 pie(m2)
-dotchart(m1,pch=16)
-dotchart(m2,pch=16)
-ordertransplot(m1,m2)
+dotchart(m1, pch=16)
+dotchart(m2, pch=16)
+ordertransplot(m1, m2)
 ```
 
-```{r,dothetests,cache=TRUE}
+```{r, dothetests, cache=TRUE}
 equalp.test(javelin1)
@@
 ```{r,label=savejavelindataset}
-save(javelin1,javelin2,javelin_table,javelin_vector,javelin1_maxp,javelin2_maxp,  file="javelin.rda")
+save(javelin1, javelin2, javelin_table, javelin_vector, javelin1_maxp, javelin2_maxp, file="javelin.rda")
 ```

### 2025-12-23 R/hyper3.R

@@ -486,22 +486,6 @@ stop("not yet written")
     return(out)
 }
 
-`attemptstable2supp3` <- function(a, decreasing=TRUE, give.supp=TRUE, dnf.last=TRUE){
-    o <- c(matrix(suppressWarnings(as.numeric(c(a,recursive=TRUE))),nrow(a),ncol(a)))
-    names(o) <- c(matrix(rownames(a),nrow(a),ncol(a)))
-    o <- sort(o,na.last=TRUE,decreasing=decreasing)
-    if(give.supp){
-        if(dnf.last){
-            nf <- names(o)[is.na(o)]
-        } else {
-            nf <- NULL
-        }
-        return(ordervec2supp3(v=names(o)[!is.na(o)], nonfinishers=nf))
-    } else {
-        return(o)
-    }
-}
-
 `args2ordervec` <- function(...){
     l <- list(...)
     if(any(names(l) == "") && !all(names(l) == "")){

### 2025-12-28 inst/collusion3.Rmd
need the magrittr package explicitly now
whitespace

@@
 
-```{r,label=loadjavelinandconvert}
-cur_matches <- read.table("stockholm1962_matches.txt",header=FALSE)
-colnames(cur_matches) <- c("white","black","result")
+```{r, label=loadandconvert}
+cur_matches <- read.table("stockholm1962_matches.txt", header=FALSE)
+colnames(cur_matches) <- c("white", "black", "result")
 head(cur_matches)
@@
 
-```{r,try11and1888,cache=TRUE}
-H3 <- f(1.1,1.888)
+```{r, try11and1888, cache=TRUE}
+H3 <- f(1.1, 1.888)
 ```
@@
 
-```{r,testH3,cache=TRUE}
+```{r, testH3, cache=TRUE}
 equalp.test(H3,n=1)
@@
 
-```{r,comparelikelihoods,cache=TRUE}
-m1121 <- maxp(f(1.1,2.1),n=1,give=TRUE)
-m1123 <- maxp(f(1.1,2.3),n=1,give=TRUE)
+```{r, comparelikelihoods, cache=TRUE}
+m1121 <- maxp(f(1.1,2.1), n=1, give=TRUE)
+m1123 <- maxp(f(1.1,2.3), n=1, give=TRUE)
 ```
 
-```{r,showmaxpwithlam}
+```{r, showmaxpwithlam}
 pie(m1121$par)
@@
 
-```{r,label=defineprofilelike,cache=FALSE}
+```{r, label=defineprofilelike, cache=FALSE}
 profl <- function(v){
@@
 
-```{r,calcproflike,cache=TRUE}
+```{r, calcproflike, cache=TRUE}
 profl(c(1.1,2.1))
-#profl(c(1.2,2.1))
 ```
 
-```{r,calcmanyproflike,cache=TRUE} 
+```{r, calcmanyproflike, cache=TRUE} 
 np <- 4
-lambda <- seq(from=0.1,to=1.5,len=np)
-D <- seq(from=0.1,to=0.4,len=np)
-jj <- as.matrix(expand.grid(lambda,D))
-LL <- apply(jj,1,profl)
+lambda <- seq(from=0.1, to=1.5, len=np)
+D <- seq(from=0.1, to=0.4, len=np)
+jj <- as.matrix(expand.grid(lambda, D))
+LL <- apply(jj, 1, profl)
 LL

### 2025-12-28 inst/javelin.Rmd
typos

@@
 Above we see that Chopra threw the longest and second-longest throws
-of 87.58m and 87.03 respectively; Vadlejch threw the third-longest
-throw of 86.67m, and so on.  `NA` entries correspond to no-throws.  It
-is easy to plot the data in visual form:
+of 87.58 and 87.03 respectively (all measurements are in meters);
+Vadlejch threw the third-longest throw of 86.67, and so on.  `NA`
+entries correspond to no-throws.  It is easy to plot the data in
+visual form:
 

### 2025-12-29 inst/javelin.Rmd
start talking about max in BT

--- 2025-12-23 inst/javelin.Rmd
@@
 
-Above, we see that Chopra threw 87.03m on his first throw, 87.58m on
-his second, and so on.  Vadlejch threw 83.98m on his first throw, etc.
-We may convert this to a named vector with elements being the throw
-distances, and names being the competitors:
+Above, we see that Chopra threw 87.03 on his first throw, 87.58 on his
+second, and so on (all measurements are in meters).  Vadlejch threw
+83.98 on his first throw, etc.  We may convert this to a named vector
+with elements being the throw distances, and names being the
+competitors:
 
@@
 Above we see that Chopra threw the longest and second-longest throws
-of 87.58m and 87.03 respectively; Vadlejch threw the third-longest
-throw of 86.67m, and so on.  `NA` entries correspond to no-throws.  It
-is easy to plot the data in visual form:
+of 87.58 and 87.03 respectively;
+Vadlejch threw the third-longest throw of 86.67, and so on.  `NA`
+entries correspond to no-throws.  It is easy to plot the data in
+visual form:
 
@@
 
+# Discussion: maximum throw vs Bradley-Terry philosophy
+
+The competition is won by the player who has the longest throw: the
+sufficient statistic would be the maximum of the six attempts.
+Bradley-Terry cannot handle this if different players have different
+variance: short throws are disregarded.
+
+As an example, suppose two throwers have normally distributed
+distances; thrower A has a smaller mean distance than thrower B but a
+larger variance.  Numerically:
+
+```{r}
+set.seed(0)
+thrower_A <- round(rnorm(6, 80.2, 2.6), 2)
+thrower_B <- round(rnorm(6, 80.3, 1.3), 2)
+D <- as.attemptstable(as.data.frame(rbind(thrower_A, thrower_B)))
+D
+apply(D, 1, max)
+```
+
+Above we see that thrower A wins, despite having a smaller mean
+distance.
+
 ### Package dataset {-}

### 2025-12-31 inst/javelin.Rmd
print method for attemptstables

@@
 a <- javelin_table # saves typing
+options("bold_personal_best" = FALSE)
 javelin_table
@@
 second, and so on (all measurements are in meters).  Vadlejch threw
-83.98 on his first throw, etc.  We may convert this to a named vector
-with elements being the throw distances, and names being the
-competitors:
+83.98 on his first throw, etc.  Print option `bold_personal_best`
+controls whether the personal (row-wise) best attempt is printed in
+bold.  This does not work in rmarkdown documents but does work on the
+terminal and Rstudio console.  We may convert `javelin_table` to a
+named vector with elements being the throw distances, and names being
+the competitors:
 

### 2025-12-31 R/attemptstable.R

@@ -1,4 +1,4 @@
- `as.attemptstable` <- function(x){
+`as.attemptstable` <- function(x){
     x <- as.data.frame(x)
     class(x) <- c("attemptstable", "data.frame")  # the only place
     return(x)
@@ -6,9 +6,38 @@
 
 `print.attemptstable` <- function(x, ...){
     cat("An attemptstable:\n")
-    class(x) <- "data.frame"
-    x[is.na(x)] <- "X"
-    print(x)
+    co <- capture.output(as.data.frame(x))
+    
+    cat(co[1]) # colnames
+    cat("\n")
+    for(i in 2:length(co)){
+        string <- gsub("NA", " X", co[i])
+        jj <- strsplit(string, " ")[[1]][-1]
+        suppressWarnings(p <- as.numeric(jj[nchar(jj) > 0]))
+        if(!isFALSE(getOption("bold_personal_best"))){
+            string <- rowfiddle(string, which.max(p))
+        }
+        cat(string)
+        cat("\n")
+    }
+}
+
+rowfiddle <- function(v, n){  # takes a string, eg v="Vesely     79.73   80.3  85.44      X  84.98      X"
+    bold_start <- "\033[1m"
+    bold_end   <- "\033[22m"
+
+    jj <- rle(strsplit(v, "")[[1]] == " ")
+    start_position <- sum(jj$lengths[seq_len(n*2    )])
+      end_position <- sum(jj$lengths[seq_len(n*2 + 1)])
+
+    paste(
+        substr(v,1,start_position),
+        bold_start,
+        substr(v,start_position+1, end_position),
+        bold_end,
+        substr(v,end_position+1, nchar(v)),
+        sep = ""
+    )
 }
 
 `suppfun.attemptstable` <- function(x, ...){
